### 2021-04-22 types2021/EasyChair3.5/easychair.tex
types abstract

% easychair.tex,v 3.5 2017/03/15

\documentclass{easychair}
%\documentclass[EPiC]{easychair}
%\documentclass[EPiCempty]{easychair}
%\documentclass[debug]{easychair}
%\documentclass[verbose]{easychair}
%\documentclass[notimes]{easychair}
%\documentclass[withtimes]{easychair}
%\documentclass[a4paper]{easychair}
%\documentclass[letterpaper]{easychair}

\usepackage{doc}

% use this if you have a long article and want to create an index
% \usepackage{makeidx}

% In order to save space or manage large tables or figures in a
% landcape-like text, you can use the rotating and pdflscape
% packages. Uncomment the desired from the below.
%
% \usepackage{rotating}
% \usepackage{pdflscape}

% Some of our commands for this guide.
%
\newcommand{\easychair}{\textsf{easychair}}
\newcommand{\miktex}{MiK{\TeX}}
\newcommand{\texniccenter}{{\TeX}nicCenter}
\newcommand{\makefile}{\texttt{Makefile}}
\newcommand{\latexeditor}{LEd}

%\makeindex

%% Front Matter
%%
% Regular title as in the article class.
%
\title{The {\easychair} Class File\\
       Documentation and Guide for Authors%
\thanks{Other people who contributed to this document include Maria Voronkov
  (Imperial College and EasyChair) and Graham Gough (The University of
  Manchester).}}

% Authors are joined by \and. Their affiliations are given by \inst, which indexes
% into the list defined using \institute
%
\author{
Serguei A. Mokhov\inst{1}\thanks{Designed and implemented the class style}
\and
    Geoff Sutcliffe\inst{2}\thanks{Did numerous tests and provided a lot of suggestions}
\and
   Andrei Voronkov\inst{3}\inst{4}\inst{5}\thanks{Masterminded EasyChair and created versions
     3.0--3.5 of the class style}
}

% Institutes for affiliations are also joined by \and,
\institute{
  Concordia University,
  Montreal, Quebec, Canada\\
  \email{mokhov@cse.concordia.ca}
\and
   University of Miami,
   Miami, Florida, U.S.A.\\
   \email{geoff@cs.miami.edu}\\
\and
   University of Manchester,
   Manchester, U.K.\\
   \email{andrei@voronkov.com}\\
\and
   Chalmers University of Technology,
   Gothenburg, Sweden
\and
   EasyChair
 }

%  \authorrunning{} has to be set for the shorter version of the authors' names;
% otherwise a warning will be rendered in the running heads. When processed by
% EasyChair, this command is mandatory: a document without \authorrunning
% will be rejected by EasyChair

\authorrunning{Mokhov, Sutcliffe and Voronkov}

% \titlerunning{} has to be set to either the main title or its shorter
% version for the running heads. When processed by
% EasyChair, this command is mandatory: a document without \titlerunning
% will be rejected by EasyChair
\titlerunning{The {\easychair} Class File}

\begin{document}

\maketitle

\begin{abstract}
  In order to ease the lives of authors, editors, and trees, we present an
  easy-to-read guide to the easy-to-use {\easychair} {\LaTeX2e} document style
  class for EasyChair-based electronic and on-paper publishing of workshop and conference
  proceedings.
\end{abstract}

% The table of contents below is added for your convenience. Please do not use
% the table of contents if you are preparing your paper for publication in the
% EPiC Series or Kalpa Publications series

\setcounter{tocdepth}{2}
{\small
\tableofcontents}

%\section{To mention}
%
%Processing in EasyChair - number of pages.
%
%Examples of how EasyChair processes papers. Caveats (replacement of EC
%class, errors).

%------------------------------------------------------------------------------
\section{Introduction}
\label{sect:introduction}

The {\easychair} class was designed to be easy to use, and specifically favors
electronic and on-paper publishing by the EasyChair conference system
\cite{easychair}, including the
\href{http://www.easychair.org/publications/EPiC}{EasyChair EPiC
  Series} and \href{http://www.easychair.org/publications/Kalpa}{EasyChair Kalpa
  Publications}. 
EasyChair is a free conference management system that is flexible, easy to use,
and has many features to make it suitable for various conference models. It is
currently probably the most commonly used conference management system
\cite{easychair}.

The {\easychair} class was designed according to some requirements, which
are described in Appendix~\ref{sect:easychair-requirements}. 
An article that occupies approximately 15 LNCS-formatted pages
takes up approximately 14 {\easychair} pages.

\begin{figure}[tb]
	\begin{centering}
	\includegraphics[width=0.5\textwidth]{logoEC}
	\caption{EasyChair logo}
	\label{fig:easychair-logo}
	\end{centering}
\end{figure}

%------------------------------------------------------------------------------
\section{Typesetting}
\label{sect:typesetting}

Typesetting with {\easychair} is, well, easy.  Just by using the
document class entry in the document's preamble as follows:
\verb+\documentclass{easychair}+ the typesetting work is nearly done.
The {\easychair} class is a relatively conservative extension of the
standard \texttt{article} class, so most of the environments, section
headers, etc. defined by \texttt{article} are available.

%------------------------------------------------------------------------------
\subsection{Generalities}
\label{sect:generalities}

The following are the general default parameters {\easychair}
introduces into the typesetting aspect of articles. If you use
\easychair\ for proceedings or other kinds of publishing through
EasyChair, do not alter these -- papers deviating from the formatting
standards will be rejected by EasyChair.

\begin{enumerate}
\item
The default paper size is US letter. It can be explicitly set to A4 
(\texttt{a4paper}) or letter (\texttt{letterpaper}) paper in the
document class entry, e.g.:\\\verb+\documentclass[a4paper]{easychair}+

\item
The print area for both letter and A4 paper sizes is 145x224 mm. This size
has been selected to allow for inexpensive printing using our current
print-on-demand publisher.

\item
The base font is Computer Modern. The base font size is 10pt. If you
use any other font size, there is no guarantee that the produced
document will look nice or fit into our standard page size.

\item
The references list is condensed. The default bibliography styles, such as
\texttt{plain}, \texttt{abbrv}, and \texttt{alpha}, are suggested.

\item
PNG, JPG, and PDF images are supported, i.e., those that are supported
by the standard \texttt{graphicx} package \cite{graphicx-package}, and
render nicely in online versions of PDF documents.  This document
shows some examples of JPG and PDF images, for example in
Figure~\ref{fig:easychair-logo}. If the papers are designed for
publishing in print, the images should be at least 300dpi in
resolution.

\end{enumerate}

%------------------------------------------------------------------------------
\subsection{Front Matter}
\label{sect:front-matter}

The front matter of an {\easychair} article follows the \texttt{article}
style, augmented with the \verb+\titlerunning+ and \verb+\authorrunning+
commands for use by authors. 
For the \verb+\author+ command with multiple authors, use \verb+\and+ to
separate authors from different institutions, as done in this document.
Institutions are defined using \verb+\institute+ in a similar way and affiliations of authors assigned using \verb+\inst+.
Authors must set the \verb+\titlerunning+ and \verb+\authorrunning+.
For example, the front matter of this document defined the authors
and title as follows:

%---------------------------------------------------------------

\noindent\small
\begin{verbatim}
\title{The {\easychair} Class File\\
       Documentation and Guide for Authors%
\thanks{Other people who contributed to this document include Maria Voronkov
  (Imperial College and EasyChair) and Graham Gough (The University of
  Manchester).}}

% Authors are joined by \and. Their affiliations are given by \inst, which indexes
% into the list defined using \institute
%
\author{
Serguei A. Mokhov\inst{1}\thanks{Designed and implemented the class style}
\and
   Geoff Sutcliffe\inst{2}\thanks{Did numerous tests and provided a lot of suggestions}
\and
   Andrei Voronkov\inst{3}\inst{4}\inst{5}\thanks{Masterminded EasyChair and created 
     versions 3.0--3.5 of the class style}
}

% Institutes for affiliations are also joined by \and,
\institute{
  Concordia University,
  Montreal, Quebec, Canada\\
  \email{mokhov@cse.concordia.ca}
\and
   University of Miami,
   Miami, Florida, U.S.A.\\
   \email{geoff@cs.miami.edu}\\
\and
   University of Manchester,
   Manchester, U.K.\\
   \email{andrei@voronkov.com}\\
\and
   Chalmers University of Technology,
   Gothenburg, Sweden
\and
   EasyChair
 }

\authorrunning{Mokhov, Sutcliffe and Voronkov}
\titlerunning{The {\easychair} Class File}
\end{verbatim}
\normalsize

%------------------------------------------------------------------------------
\subsection{Page Numbering}
\label{sect:page-numbering}

Page numbers are at the bottom of every page. Authors must leave the
page numbers in as-is. When EasyChair proceedings and Procedia volumes are processed by
EasyChair, the correct volume page numbers will be inserted
automatically.

%------------------------------------------------------------------------------
\subsection{Section Headings and Capitalization}
\label{sect:section-headings}

Section and paragraph headings in {\easychair} are invoked via the standard 
commands, such as
\verb+\section+,
\verb+\subsection+,
\verb+\subsubsection+, and
\verb+\paragraph+.
Generally, every non-trivial word in a heading must be capitalized according to
general capitalization guidelines. A reasonable rule to use is that
all prepositions, coordinating conjunctions and articles having four
or fewer letters should not be capitalized. If you do not know what it
means, simply do not capitalize the following words:
\textit{amid, anti, as, at, atop, but, by, down, for, from, in, into, like,
near, next, of, off, on, onto, out, over, pace, past, per, plus, qua,
save, than, till, to, up, upon, via, with, for, and, nor, but, or,
yet, so, a, an, the.} For example, if you want to call your paper
``oldest but goldest'', then the proper title for it is ``Oldest but
Goldest''. ``Oldest but goldest'' is wrong (since ``goldest'' is not
capitalized) and ``Oldest But Goldest'' is wrong (``but'' should stay
lower-cased since it belongs to the list of words above.) Needless to
say, ``OLDEST BUT GOLDEST'' is very wrong.

Paragraph headings should not be capitalized and should have a
trailing period. That is, you should write

\begin{quote}
\verb|\paragraph{EasyChair is cool.}|
\end{quote}
rather than

\begin{quote}
\verb|\paragraph{EasyChair is cool}|
\end{quote}
unless your aim was to write something like

\begin{quote}
\verb|\paragraph{EasyChair is cool} when you use it for publishing.| 
\end{quote}
Of course, you are welcome to replace the trailing period by any other
punctuation mark, for example

\begin{quote}
\verb|\paragraph{EasyChair is cool!}|
\end{quote}

See the examples in this document, e.g.,
Section~\ref{sect:typesetting} is a section, this 
(Section~\ref{sect:section-headings}) is a subsection, and
Section~\ref{sect:subsubsection-headings} is a subsubsection.

%------------------------------------------------------------------------------
\subsubsection{Subsubsection Header}
\label{sect:subsubsection-headings}

This is a subsubsection.

\paragraph{Paragraph header.}

This is a paragraph. 
One way of saving space when hyper-references are not essential is to 
use paragraphs instead of subsubsections.

%------------------------------------------------------------------------------
\subsection{Mathematics}
\label{sect:mathematics}

Mathematics can be done inline for simple things, e.g., an equation
$x = 0$, possibly with super and subscripts, e.g., $x^2_k \approx 27$,
Greek letters, e.g., $\alpha \cup \Theta \ne \gamma$, etc.
Larger formulae must be done using {\tt \verb|\|[}~~{\tt \verb|\|]}
bracketing, e.g.,
\[
\int_{0}^{1} x dx = \left[ \frac{1}{2}x^2 \right]_{0}^{1} = \frac{1}{2}
\]
or using {\tt \verb|\|begin\{equation\}} and {\tt \verb|\|end\{equation\}} for
numbered equations, e.g.,
\begin{equation}
e^x = \sum_{n=0}^\infty \frac{x^n}{n!} = \lim_{n\rightarrow\infty} (1+x/n)^n
\end{equation}

Use {\tt \verb|\|begin\{align*\}} and {\tt \verb|\|end\{align*\}} (or without
the {\tt *} include number) to align equations, e.g.,
\begin{align*}
x^2 + y^2 &= 1 \\
y &= \sqrt{1 - x^2}
\end{align*}

Fonts, using \verb|\matcal| and others can also be used in the math mode: $\mathcal{ALC}$.

%------------------------------------------------------------------------------
\subsection{Tables}
\label{sect:tables}

Table~\ref{tab:ltbexample} shows an example of a table of data that was
conveniently available (i.e., the data has nothing to do with
{\easychair}) apart from being related to the research of two of its
authors.

\begin{table}[htp]
	\begin{centering}
		\begin{tabular}{lrrrrrrrr}
		\hline
		ATP System            & LTB & Avg  &Prfs & SOTA & \multicolumn{1}{c}{$\mu$} & CYC & MZR & SMO \\
		                      & /100& time & out & Con. & Eff. & /35 & /40 & /25 \\
		\hline
		Vampire-LTB 11.0      &  69 & 24.5 &  69 & 0.37 & 28.1 &  23 &  22 &  24 \\
		iProver-SInE 0.7      &  67 & 76.5 &   0 & 0.36 &  8.8 &  28 &  14 &  25 \\
		SInE 0.4              &  64 & 75.3 &  64 & 0.32 &  8.5 &  26 &  13 &  25 \\
		leanCoP-SInE 2.1      &  35 &110.8 &  35 & 0.23 &  3.2 &  23 &   1 &  11 \\
		E-LTB 1.1pre          &  18 & 63.4 &   0 & 0.21 &  2.8 &   7 &   9 &   2 \\
		EP-LTB 1.1pre         &  18 & 77.8 &  18 & 0.21 &  2.3 &   7 &   9 &   2 \\
		E-KRH'-LTB 1.1.3      &   0 &   -- &  -- &   -- &   -- &   0 &   0 &   0 \\
		\hline
		\end{tabular}
		\caption{LTB division results}
		\label{tab:ltbexample}
	\end{centering}
\end{table}

%------------------------------------------------------------------------------
\subsection{References}
\label{sect:references}

References must be provided in a {\tt .bib} file, so that \BibTeX\ can
be used to generate the references in a consistent style in a volume.
The preferred styles are {\tt plain} and {\tt alpha}.
For example, the references for this paper are generated from the
lines
\begin{verbatim}
    \bibliographystyle{plain}
    \bibliography{easychair}
\end{verbatim}
and a way to compose the entires, e.g. citing this class style~\cite{easychair-latex-class}
is below:
\small
\begin{verbatim}
    @misc
    {
      easychair-latex-class,
      author       = {Serguei A. Mokhov and Geoff Sutcliffe and Andrei Voronkov},
      title        = {The {\sf easychair} Class File Documentation and Guide
                      for Authors},
      year         = {2008--2011},
      howpublished = {[online]},
      note         = {Available at \url{http://easychair.org/easychair.zip}}
    }
\end{verbatim}
\normalsize

%------------------------------------------------------------------------------
\section{Installation and Usage Instructions}
\label{sect:installation-usage}

%------------------------------------------------------------------------------
\subsection{Installation}

The ``installation'' of the {\easychair} document class is easy.
Download the latest version of the \texttt{easychair.zip} package 
from \url{http://www.easychair.org/easychair.zip}
and unzip it in the directory where you will prepare your paper.
You will get the following files, out of which you may need to keep only 
the \texttt{easychair.cls} style class if you are familiar with the rest 
of the files and do not require them to get started.
We are also working to make {\easychair} available from CTAN \cite{ctan},
such that it can be installed with the popular \TeX Live \cite{texlive} and
{\miktex} \cite{miktex} \LaTeX\ package management systems.

\begin{itemize}
\item
\texttt{easychair.cls} -- the class file that this is all about.

\item
\texttt{easychair-a4.pdf} -- the PDF version of this guide rendered using 
\texttt{a4paper} option.

\item
\texttt{easychair.tex} -- the {\LaTeX} source of this guide, 
and
\texttt{easychair.bib} -- the supporting bibliography entries found starting 
on page~\pageref{sect:bib}.

\item
\texttt{logoEC.pdf} -- the PDF version of the EasyChair logo rendered in
Figure~\ref{fig:easychair-logo} and
\texttt{throneEC.jpg} -- the JPG version of the easy throne rendered in 
Figure~\ref{fig:easythrone}.

\end{itemize}

%------------------------------------------------------------------------------
\subsection{Required Packages}

The {\easychair} class relies only on packages deemed standard and shipped by
most {\LaTeX} distributions in the worlds of Linux (current \texttt{texlive} \cite{texlive}
or older \texttt{tetex}), MacOS X,
and Windows (via Cygwin or {\miktex}).
If for some reason your distribution is old or doesn't have the packages
listed below, you can always obtain a copy from CTAN \cite{ctan}. Note
that EasyChair loads this packages automatically so you don't have to
use \verb|\usepackage| to use any of these packages.

\begin{itemize}
\item
\texttt{url} \cite{url-package} (included also by \texttt{hyperref} automatically) -- to provide URL rendering support for the
monospaced font, which takes care of special characters as well as line 
wrapping.

\item
\texttt{hyperref} \cite{hyperref-package} -- to allow hyperlinking of URLs and
cross references within an article.
Its options are set to either \verb+letterpaper+ or \verb+a4paper+, depending
on the \verb+\documentclass+ options.

\item
\texttt{graphicx} \cite{graphicx-package} -- the standard package for rendering
PNG, JPG, and PDF graphic images, primarily in \texttt{figure} environments.

\item
optional \texttt{mathptmx} \cite{mathptmx-package} -- Times base font for compactness
(use with the \texttt{withtimes} {\easychair} option).

\item
\texttt{listings} \cite{listings-package} -- to allow highlighted source code 
listing styles.

\item
\texttt{latexsym} \cite{latexsym-package} -- to provide common math and other 
symbols.

\item
\texttt{amsthm} \cite{amsthm-package} -- to provide {\AmS} theorem-like 
environments.

\item
\texttt{empheq} \cite{empheq-package} -- to provide equation environments, etc.

\item
\texttt{geometry} \cite{geometry-package} -- to set {\easychair} margins, 
outlined in Section~\ref{sect:generalities}.

\item
\texttt{lastpage} \cite{lastpage-package} -- to allow computationally
referencing the last page.

\item
\texttt{fancyhdr} \cite{fancyhdr-package} -- for running heads.

\item
\texttt{footmisc} \cite{footmisc-package} -- to ensure that footnotes are
always at the bottom.

\item
optional \texttt{makeidx} \cite{makeidx-package} -- for index generation
(use with the \texttt{thesis} {\easychair} option).

\item
\texttt{eso-pic} \cite{eso-pic-package} -- for draft versions and checking page
overlflows vs. a border drawn around the headers, footers, and the main body of
the article.

\end{itemize}

%------------------------------------------------------------------------------
\subsection{Recommended Packages}
\label{sect:recommended-packages}

Here is a list of some packages that this guide's authors have experimented 
with, and which are suitable for inclusion if needed by article authors.
These packages must be loaded using \verb+\usepackage+.
In general, authors may use any standard packages provided they do not change
the basic layout and font settings established by the {\easychair} class.
Such packages must be provided with the submission of articles.

\begin{itemize}
\item
\texttt{rotating} \cite{rotating-package} -- to rotate floats (figures and
tables) on the page, when wide tables or figures do not fit in portrait layout.

\item
\texttt{pdflscape} \cite{pdflscape-package} -- similar to \texttt{rotating}, 
but also allows rotating text to make it conveniently viewable in a PDF 
viewer that supports individual rotated pages.
A possible disadvantage is that a page break is forced, which may create
gaps before or after the landscape page.

\item
\texttt{algorithm2e} \cite{algorithm2e-package} -- provides a figure-like
algorithm environment for formal algorithm presentation with highlighting.

\end{itemize}

%------------------------------------------------------------------------------
\subsection{Compiling}
\label{sect:compiling}

\texttt{pdflatex} \cite{pdflatex-instructions} is the preferred tool for
producing PDF files with {\easychair} class documents.
The author kit (\texttt{easychair.zip}) includes some minimal automation 
that authors can use at their discretion.

\begin{itemize}
\item
Linux and UNIX-like platforms (also works under Cygwin and MacOS X):
A {\makefile} is provided for the GNU \texttt{make} \cite{gmake} utility,
so this document can be compiled by typing \texttt{make} at the terminal 
prompt (on the systems where both GNU and non-GNU versions of \texttt{make} 
are installed, one may need to use \texttt{gmake}).

\item
Microsoft Windows:
{\texniccenter} \cite{texniccenter} or {\latexeditor} \cite{led} and
{\miktex} \cite{miktex} as their backend are common tools
for {\LaTeX} processing under Microsoft Windows. 
The former provide a GUI front-end to {\LaTeX}, and the latter is the 
Windows native-compiled binaries and standard packages with 
a comprehensive package update tool. 
The \texttt{easychair.tcp} project file is provided for {\texniccenter} users,
as well as \texttt{easychair.lpr} for {\latexeditor} users.

\item
MacOS X:
TeXShop \cite{texshop} is a tool for {\LaTeX} processing under Mac OS X.
It provides a GUI front-end to {\LaTeX}. The backend can be installed
through the \texttt{fink} \cite{fink} repository or the Darwin Ports.
\end{itemize}

Some authors use packages that require the use of \texttt{latex}
instead of \texttt{pdflatex}, most notably \texttt{pstricks} and its
derivatives. If switching to modern packages, such as
\texttt{tikz} is not an option, these authors can use the {\easychair}
class with \texttt{latex}.

%------------------------------------------------------------------------------
\subsection{Bug Reports}
\label{sec:bug-reports}

Please report bugs, errors, and omissions you find with the {\easychair} 
class to its current maintainer, Andrei Voronkov,
at \url{andrei@voronkov.com}. Any \emph{constructive} feedback is
always welcome. If anybody has a very hard question, we might be
unable to answer it without Sergei. If anybody knows his whereabouts,
Geoff and Andrei have the ransom money ready.

%------------------------------------------------------------------------------

\section{Avoiding Text Overflows}

When used in EasyChair, the {\easychair} class is intended for
publishing, including printing. This implies that the text of your
document should not overflow the page dimensions. This section gives
some practical advice on how you can avoid text overflows. When you
run \texttt{pdflatex} or \texttt{latex} on your files, you sometimes
get warnings like

\small
\begin{verbatim}
  LaTeX Warning: Overfull \hbox (7.43138pt too wide) in paragraph at lines 994--994
\end{verbatim}
\normalsize
or

\small
\begin{verbatim}
  LaTeX Warning: Float too large for page by 55.21666pt on input line 638.
\end{verbatim}
\normalsize
They mean what they say: your text overflows the allowed space. When
you prepare a PDF document that will only be used online or as a
draft, in many cases you can ignore these warnings, since they only
make your document uglier than it could be but it remains still
readable. However, for documents that are supposed to be printed,
overflows make a lot of difference. There is no universal recipe on
how to make your text fit into the page. However, there are some
common cases that can be easily fixed using the tricks described
here.

\subsection{Using the \texttt{debug} Option to Check for Overflows}

A general guideline is that you should first check how the alleged
overflows looks in reality. For example, some LaTeX overflows are
invisible or hardly noticeable for people and overlows by
less than 4pt look OK anyhow. The {\easychair} class has an option
\texttt{debug} that adds to each page of your documents a red frame
corresponding to the allowed page
dimensions. Figure~\ref{fig:redframe} on page~\pageref{fig:redframe}
shows an example of a page created using this style. With such a red
frame, you can clearly see any overflows apart from very minor ones.

\begin{figure}
  \begin{centering}
    \includegraphics[width=0.89\textwidth]{debug.pdf}
    \caption{A page of a document created using the \texttt{debug}
      option} 
    \label{fig:redframe}
  \end{centering}
\end{figure}

\subsection{Tables}

Many page overflows happen because of large tables. In many case these
overflows can be easily removed by slightly reducing padding added by
\LaTeX\ to every column. It is controlled by the \LaTeX\ command
\verb|\tabcolsep| whose value by default is 6pt. Even small changes in
the value of this command may give drastic reductions in the width of
tables. This is illustrated in Figure~\ref{fig:tabcolsep} on
page~\pageref{fig:tabcolsep}. Note though that there is no free lunch:
smaller values for this command may result in lower redability.

%------------------------------------------------------------
\begin{figure}[tb]\small
\begin{center}
  \begin{tabular}{lrrrrrrrr}
    \hline
    ATP System            & LTB & Avg  &Prfs & SOTA & \multicolumn{1}{c}{$\mu$} & CYC & MZR & SMO \\
   \hline
    Vampire-LTB 11.0      &  69 & 24.5 &  69 & 0.37 & 28.1 &  23 &  22 &  24 \\
    iProver-SInE 0.7      &  67 & 76.5 &   0 & 0.36 &  8.8 &  28 &  14 &  25 \\
   \hline
  \end{tabular}
\end{center}

\begin{center}\renewcommand{\tabcolsep}{5pt}
  \begin{tabular}{lrrrrrrrr}
    \hline
    ATP System            & LTB & Avg  &Prfs & SOTA & \multicolumn{1}{c}{$\mu$} & CYC & MZR & SMO \\
   \hline
    Vampire-LTB 11.0      &  69 & 24.5 &  69 & 0.37 & 28.1 &  23 &  22 &  24 \\
    iProver-SInE 0.7      &  67 & 76.5 &   0 & 0.36 &  8.8 &  28 &  14 &  25 \\
   \hline
  \end{tabular}
\end{center}

\begin{center}\renewcommand{\tabcolsep}{3pt}
  \begin{tabular}{lrrrrrrrr}
    \hline
    ATP System            & LTB & Avg  &Prfs & SOTA & \multicolumn{1}{c}{$\mu$} & CYC & MZR & SMO \\
   \hline
    Vampire-LTB 11.0      &  69 & 24.5 &  69 & 0.37 & 28.1 &  23 &  22 &  24 \\
    iProver-SInE 0.7      &  67 & 76.5 &   0 & 0.36 &  8.8 &  28 &  14 &  25 \\
   \hline
  \end{tabular}
\end{center}

\begin{center}\renewcommand{\tabcolsep}{1pt}
  \begin{tabular}{lrrrrrrrr}
    \hline
    ATP System            & LTB & Avg  &Prfs & SOTA & \multicolumn{1}{c}{$\mu$} & CYC & MZR & SMO \\
   \hline
    Vampire-LTB 11.0      &  69 & 24.5 &  69 & 0.37 & 28.1 &  23 &  22 &  24 \\
    iProver-SInE 0.7      &  67 & 76.5 &   0 & 0.36 &  8.8 &  28 &  14 &  25 \\
   \hline
  \end{tabular}
\end{center}
\normalsize

\caption{Original table and tables with \texttt{tabcolsep} set to 5pt,
  3pt, and 1pt
  \label{fig:tabcolsep}}

\end{figure}
%------------------------------------------------------------

\subsection{Images}

Images included using \verb|\includegraphics| are easy to resize since
one can specify the size of the result explicitly. For example,
Figure~\ref{fig:easythrone} shows three copies of the same image
having different sizes obtained using the following commands:

\begin{verbatim}
  \includegraphics[width=0.5\textwidth]{throneEC.jpg}
  \includegraphics[width=0.3\textwidth]{throneEC.jpg}
  \includegraphics[width=0.15\textwidth]{throneEC.jpg}
\end{verbatim}

\subsection{A Universal Recipe}

\LaTeX\ has a very powerful weapon for reducing the size of almost
anythings. More precisely, it can reduce anything producing what
\LaTeX\ considers a box. This weapon is called
\verb|\scalebox|. Consider an example (check the source of this file
to see how it was produced).

\begin{center}
\begin{tabular}{|c|c|}
\hline \Huge
\begin{tabular}[b]{cr}
year & users \\ \hline
2007 &    47,753 \\
 2008 &    114,494 \\
 2009 &    207,506 \\
 2010 &   371,054 \\
\end{tabular}
&
\includegraphics[width=0.28\textwidth]{chairEC}\\
\hline
\multicolumn{2}{|c|}{The number of users of EasyChair and one of its
  logos,}\\
\multicolumn{2}{|c|}{scaled to the number of users in 2010} \\
\hline
\end{tabular}
\end{center}
This is what happens when we put (almost) the same \LaTeX\ code in 
\verb|\scalebox{0.55923}{...}| to scale it down to the number of users
in 2009:

\begin{center}
\scalebox{0.55923}{%
\begin{tabular}{|c|c|}
\hline \Huge
\begin{tabular}[b]{cr}
year & users \\ \hline
2007 &    47,753 \\
 2008 &    114,494 \\
 2009 &    207,506 \\
 2010 &   371,054 \\
\end{tabular}
&
\includegraphics[width=0.28\textwidth]{chairEC}\\
\hline
\multicolumn{2}{|c|}{The number of users of EasyChair and one of its
  logos,}\\
\multicolumn{2}{|c|}{scaled to the number of users in 2009} \\
\hline
\end{tabular}}
\end{center}
We can scale it down even further to the 2008 figure using
\verb|\scalebox{0.30856}{...}|:

\begin{center}
\scalebox{0.30856}{%
\begin{tabular}{|c|c|}
\hline \Huge
\begin{tabular}[b]{cr}
year & users \\ \hline
2007 &    47,753 \\
 2008 &    114,494 \\
 2009 &    207,506 \\
 2010 &   371,054 \\
\end{tabular}
&
\includegraphics[width=0.28\textwidth]{chairEC}\\
\hline
\multicolumn{2}{|c|}{The number of users of EasyChair and one of its
  logos,}\\
\multicolumn{2}{|c|}{scaled to the number of users in 2008} \\
\hline
\end{tabular}}
\end{center}
or further down to 2007:

\begin{center}
\scalebox{0.12870}{%
\begin{tabular}{|c|c|}
\hline \Huge
\begin{tabular}[b]{cr}
year & users \\ \hline
2007 &    47,753 \\
 2008 &    114,494 \\
 2009 &    207,506 \\
 2010 &   371,054 \\
\end{tabular}
&
\includegraphics[width=0.28\textwidth]{chairEC}\\
\hline
\multicolumn{2}{|c|}{The number of users of EasyChair and one of its
  logos,}\\
\multicolumn{2}{|c|}{scaled to the number of users in 2008} \\
\hline
\end{tabular}}
\end{center}

This size reduction technique is very efficient: using the right scale
you may post your whole article on Twitter in a single tweet. However,
it may also may parts of your text virtually unreadable with an
unfortunate side effect of annoying reviewers.

\section{Submitting Your Article Through EasyChair}

This section is intended only for the authors and editors of EasyChair
proceedings and Procedia volumes.

When you prepare an article for either of these, it should be
submitted through EasyChair. EasyChair automates the submission
process as much as possible and goes to a great length to ensure that
your article can be published and printed. Publication for EasyChair
means much more than just putting a PDF of your article online. It
collects a some meta-information about the article to classify it,
find similar articles, make it easily searchable, and index it in
various Web services, such as DBLP. This section explains how
EasyChair processes your article.

\subsection{Submitting the Article}

Submitting the article is easy. All you should do is to put the source
of your article in a single zip file. The source must contain all
auxiliary files required to create a PDF file of your article: this
includes images, bibliography, and all non-standard \LaTeX\ packages
you used\footnote{ 
  A non-standard \LaTeX\ package is a package that is not included in
  CTAN. 
} For example, suppose that your main \LaTeX\ file is
\texttt{main.tex}, it inputs another file \texttt{macros.tex} and uses
the file \texttt{biblio.bib} to produce the bibliography. Suppose it
also uses two images \texttt{images/easy.jpg} and
\texttt{images/easy.jpg}. Then you should create a zip archive
containing all these files. Suppose all these files are put in a
directory \texttt{mypaper} on your computer, where \texttt{images} is
a subdirectory of \texttt{mypaper}

On almost any operating system (Linux,
Windows, or Mac) you can achieve this by using the following sequence
of commands:

\begin{verbatim}
cd mypaper
zip -r mypaper.zip *
\end{verbatim}
This will create a zip archive \texttt{mypaper.zip} including all
files in the directory \texttt{mypaper} and its subdirectories.

\subsection{EPiC Series and Kalpa Publications}
\label{subsec:EPiC}

If you publish your article in the EasyChair EPiC series, our aim is to produce
a nicely looking EPiC Series header, see Figure~\ref{fig:epicheader} on page
\pageref{fig:epicheader} for an idea. The images and information in the header
will be added by EasyChair automatically and depend on the subseries and the
volume information.

\begin{figure}
  \begin{centering}
    \includegraphics[width=0.89\textwidth]{epic.pdf}
    \caption{This document with an EPiC Series header} 
    \label{fig:epicheader}
  \end{centering}
\end{figure}

To ensure that the layout of your document will not change when the header is
added, you should use the option \texttt{EPiC} as follows:

\small
\begin{verbatim}
\documentclass[EPiC]{easychair}
\end{verbatim}
\normalsize
This will reserve space for the header on the top of the first page of your
article.

%------------------------------------------------------------------------------
\section{Future Work}
\label{sect:future-work}

We plan to further strengthen the {\easychair} class and promote it for 
electronic publishing for EasyChair-powered conferences and workshops,
and take over the world, as shown in Figure~\ref{fig:easythrone}. We
aim at creating a new model of \emph{affordable publishing}, where
anybody can become a low-cost publisher. The
\href{http://www.easychair.org/publications/EPiC}{EasyChair EPiC
  Series} and
the
\href{http://www.easychair.org/publications/Kalpa}{EasyChair Kalpa Publications} 
are our first contributions to affordable open access publishing.

\begin{figure}[tb]
  \begin{centering}
    \includegraphics[width=0.5\textwidth]{throneEC.jpg}
    \includegraphics[width=0.3\textwidth]{throneEC.jpg}
    \includegraphics[width=0.15\textwidth]{throneEC.jpg}
  \end{centering}
  \caption{Why one should use EasyChair}
  \label{fig:easythrone}
\end{figure}

%------------------------------------------------------------------------------
\section{Acknowledgments}
\label{sect:acks}

\begin{itemize}
\item
Aleksander Kosenkov for the graphics that are used here and on the EasyChair 
website~\cite{easychair}.

\item
Graham Gough (The University of Manchester) for implementing title-related
commands.

\item
Thomas F. Sturm (Universit\"at der Bundeswehr M\"unchen) for implementing the
\LaTeX\ part of the headers used for the EasyChair EPiC series.

\item
The CTAN \cite{ctan} and {\LaTeX} communities \cite{texniccenter,miktex}.

\item
Guilin Qi, Jasmin Christian Blanchette, Leslie Lamport, Uwe Pfeiffer,
and others for constructive feedback on the style, most of which had been
incorporated in version 2 of the class style.
\end{itemize}

\label{sect:bib}
\bibliographystyle{plain}
%\bibliographystyle{alpha}
%\bibliographystyle{unsrt}
%\bibliographystyle{abbrv}
\bibliography{easychair}

%------------------------------------------------------------------------------
\appendix
\section{{\easychair} Requirements Specification}
\label{sect:easychair-requirements}

The following high-level requirements were set for the development of 
the {\easychair} class, and were refined as development went along.

\begin{enumerate}
\item
The style should be easy to use. 
The average {\LaTeX} user should not need to read a long manual.

\item
It should be economical in space but the text should be nice-to-read.

\item
It should use fonts producing a reasonable-quality PDF.

\item
The bibliography should produce hyperlinks.

\item
Sections should produce menu sections in PDF.

\item
The text should look good on both A4 and letter paper.

\item
The style should be single-column for convenience of scrolling.

\item
The print area should be convenient for printing using print-on-demand publishers.

\item
Running heads.
\end{enumerate}

%------------------------------------------------------------------------------
% Index
%\printindex

%------------------------------------------------------------------------------
\end{document}

### 2021-04-22 types2021/abstract.tex
types abstract

\documentclass{easychair}

\usepackage{doc}
\usepackage{scalerel}
\usepackage{amsfonts}

\newcommand{\easychair}{\textsf{easychair}}
\newcommand{\miktex}{MiK{\TeX}}
\newcommand{\texniccenter}{{\TeX}nicCenter}
\newcommand{\makefile}{\texttt{Makefile}}
\newcommand{\latexeditor}{LEd}
\newcommand{\emptycon}{\scaleobj{.75}\bullet}
\newcommand{\ext}{\triangleright}
\newcommand{\arri}{\Rightarrow}

\newcommand{\ToS}{\mathsf{ToS}}
\newcommand{\U}{\mathsf{U}}
\newcommand{\Code}{\mathsf{Code}}
\newcommand{\Ty}{\mathsf{Ty}}
\newcommand{\V}{\mathsf{V}}
\newcommand{\C}{\mathsf{C}}

\title{Using Two-Level Type Theory for Staged Compilation %
  \thanks{The author was supported by the European Union, co-financed
    by the European Social Fund (EFOP-3.6.3-VEKOP-16-2017-00002).}}

\author{
Andr\'as Kov\'acs
}

% Institutes for affiliations are also joined by \and,
\institute{
  E\"otv\"os Lor\'and University,
  Budapest, Hungary \\
  \email{kovacsandras@inf.elte.hu}
}

\authorrunning{Kov\'acs}
\titlerunning{Using Two-Level Type Theory for Staged Compilation}
\pagenumbering{gobble}
\begin{document}

\maketitle

Two-level type theory \cite{twolevel} (2LTT) is a system for performing
metatheoretic constructions and reasoning involving certain object-level type
theories. Such reasoning is always possible by simply embedding object theories
in metatheories, but in that case we have to explicitly handle a deluge of
technical details about the object theory, most notably substitutions. If
everything that we aim to do is natural with respect to object-level
substitution, we can instead use 2LTT, which can be viewed as a notation for
working with presheaves over the object-level category of substitutions.

Likewise in metaprogramming, there is a spectrum: we can simply write programs
which output raw source code, or use staging instead, which is safer and more
convenient, but also restricted in some ways. In the current work we observe
that 2LTT is a powerful model for generative staged compilation.

\emph{Basic rules of 2LTT}. We have universes $\U^s_{i}$, where $s \in \{0,1\}$,
denoting a \emph{stage} or level in the 2LTT sense, and $i \in \mathbb{N}$
denotes a usual level index of \emph{sizing} hierarchies. The two dimensions of
indexing are orthogonal, and we will elide the $i$ indices in the following. We
assume Russell-style universes. For each $\Gamma \vdash A : \U^0$, we have
$\Gamma \vdash \Code\,A : \U^1$. Quoting: for each $\Gamma \vdash t : A$, we
have $\Gamma \vdash\,<\!t\!> : \Code\,A$. Unquoting: for each $\Gamma \vdash t :
\Code\,A$, we have $\Gamma \vdash\,\sim\!t : A$. Moreover, quoting and unquoting
form an isomorphism up to definitional equality. $\U^0$ and $\U^1$ can be closed
under arbitrary additional type formers.

The idea of staging is the following: given a closed $A : \U^0$ with a closed $t
: A$ in 2LTT, there are unique $A'$ and $t'$ in the object theory, which become
definitionally equal to $A$ and $t$ respectively after being embedded in
2LTT. In short, every meta-level construction can be computed away. Annekov et
al. \cite{twolevel} showed this without uniqueness of $A'$ and $t'$. We can get
uniqueness as well, using the normalization of 2LTT: by induction on the
(unique) normal forms of $A$ and $t$, we can show that they cannot contain
meta-level subterms. This shows that normalization is a sound staging algorithm,
but in practice we do not want to compute full normal forms; we want to compute
\emph{meta-level redexes only}. This can be done with a variation of standard
normalization-by-evaluation \cite{abel2013normalization,
  Wieczorek:2018:CFN:3176245.3167091} which also keeps track of stages.

\subsubsection*{Applications}

\textbf{Control over inlining and compile-time computation.} We can define two
variations of the polymorphic identity function for object-level types:
\begin{alignat*}{3}
  & \mathit{id} : (A : \U^0) \to A \to A \hspace{2em}&& \mathit{id'} : (A : \Code\,\U^0) \to \Code\,\sim\!A \to \Code\,\sim\!A\\
  & \mathit{id} = \lambda\,A\,x.\,x      && \mathit{id'} = \lambda A\,x.x
\end{alignat*}
The second version is evaluated at compile time. For example,
$\sim\!(\mathit{id'}<\!\mathsf{Bool^0}\!>\,<\!\mathsf{true^0}\!>)$ can be used
in object-level code, which is computed to $\mathsf{true^0}$ by staging. We can
also freely use induction on meta-level values to generate object-level code,
including types. Hence, 2LTT supports full dependent types (with universes and
large elimination) in staging.

\textbf{Monomorphization.} We assume now that the object language is a
\emph{simple type theory}. In this case, there is no universe $\U^0$ in the
object-level, so there is no $\Code\,\U^0$, but we can still freely include a
meta-level type $\Ty^0$ whose terms are identified with object-level types. Now,
meta-level functions can be used for quantification over object-level types, as in
$\mathit{id} : (A : \Ty^0) \to \Code\,\sim\!A \to \Code\,\sim\!A$. However,
since the object theory is simply typed and monomorphic, all polymorphism is
guaranteed to compute away during staging.

\textbf{Control over lambda lifting and closure creation}. We assume now a
dependent type theory on both levels, but with a \emph{first-order function
type} on the object level. This is defined by splitting $\U^0$ to a universe
$\V^0$ which is closed under inductive types but not functions, and a universe
$\C^0$ which has $V^0$ as a sub-universe, and is closed under functions with
domains in $\V^0$ and codomains in $\C^0$. The object theory supports
compilation which requires only lambda lifting, but no closures. On its own, the
object theory is fairly restricted, but together with staging we have a
remarkably expressive system. Then, we can close $\V^0$ under a separate type
former of closure-based functions, thereby formally distinguishing
lambda-liftable functions from closure-based functions. This enables typed
analysis of various optimization and fusion techniques. E.g.\ we get guaranteed
closure-freedom in code output if a certain fusion technique can be formalized
with only first-order function types. In particular, this may obviate the need
for arity analysis \cite{DBLP:journals/cl/Breitner18} in fold-based fusion.

\textbf{Memory layout control}. We assume again a dependent theory on both
levels, but now index $\U^0$ over \emph{memory layouts}. For example,
$\U^0\,\mathsf{erased}$ may contain runtime-erased types, and
$\U^0\,(\mathsf{word64} \times \mathsf{word64})$ may contain types represented
as unboxed pairs of machine words. We assume a meta-level type of
layouts. Hence, we can abstract over layouts, but after staging every layout
will be concrete and canonical in the output. This can be viewed as a more
powerful version of \emph{levity polymorphism} in GHC \cite{levity}, and a way
to retain both dependent types and non-uniform memory layouts in the object
theory.

\subsubsection*{Potential extensions}

\textbf{More stages, stage polymorphism}. The standard presheaf semantics of
2LTT can be extended to more levels in a straightforward way. It seems feasible
to also allow quantifying over all smaller levels, at a given level.

\textbf{Stage inference}. $\Code$ preserves all negative type formers up to
definitional isomorphism \cite{twolevel}, e.g. $\Code\,(A \to B) \simeq
(\Code\,A \to \Code\,B)$. This can be used to support inference for staging
annotations, by automatically inserting transports along preservation
isomorphisms during elaboration. The previous
$\sim\!(\mathit{id'}<\!\mathsf{Bool^0}\!>\,<\!\mathsf{true^0}\!>)$ example could
be simply written as $\mathit{id'}\,\mathsf{Bool^0}\,\mathsf{true^0}$ in the
surface language, and elaboration would transport $\mathit{id'}$ appropriately.
We demonstrated the practical feasibility of such stage inference in a prototype
implementation.

\textbf{Induction on $\Code$}. Basic 2LTT supports \emph{any model} of the
object theory in the presheaf semantics, it does not assume that we have
presheaves over the initial model (syntax). Hence, $\Code\,A$ is a black box
without elimination principles. However, if we are interested in staged
compilation, we can assume the object-level to be syntactic and consistently add
operations on $\Code\,A$ which rely on that assumption, e.g. conversion
checking, pattern matching, or induction on normal forms of object-level
expressions.

\bibliographystyle{plain}
\bibliography{references}

\end{document}

### 2021-04-22 src/Elaboration.hs

@@ -1,4 +1,4 @@
--- {-# options_ghc -Wno-unused-imports #-}
+{-# options_ghc -Wno-unused-imports -Wno-type-defaults #-}
 
 module Elaboration where
 
@@ -21,7 +21,7 @@ import ElabState
 import InCxt
 import Exceptions
 
--- import Debug.Trace
+
 
 --------------------------------------------------------------------------------
 
@@ -106,7 +106,9 @@ subtype0 cxt a cv a' cv' = do
 
 -- Coercion in U1 (explicit)
 coe1 :: Dbg => Cxt -> S.Tm1 -> V.Ty -> V.Ty -> IO S.Tm1
-coe1 cxt t a a' = maybe t id <$> go cxt t a a' where
+coe1 cxt t a a' = do
+  traceShowM ("coe", t, quote1 cxt a, quote1 cxt a')
+  maybe t id <$> go cxt t a a' where
 
   -- ugly helper for record structural rule
   goRec1 :: Dbg => Cxt -> S.Tm1 -> Int -> V.Close (Fields S.Tm1)
@@ -310,7 +312,9 @@ check0 cxt topT topA cv = case (topT, forceFU1 topA, forceCV cv) of
 
   (t, topA, cv) -> do
     (t, a, cv') <- infer0 cxt t
+    traceShowM ("infer0-d", t, quote1 cxt a, cv', quote1 cxt topA, cv)
     decorateEx cxt topT $ subtype0 cxt a cv' topA cv
+    traceM "coerced"
     pure t
 
 checkRec0 :: Dbg => Cxt -> [(Span, P.Tm)] -> IO (Fields S.Ty)
@@ -455,7 +459,10 @@ infer0 cxt topT = case topT of
       V.Lift cv a -> do
         let t' = S.down t
         pure (t', a, cv)
+
+      -- PRUNING
       a -> do
+        traceM "suuss"
         cv <- freshCV
         a' <- eval1 cxt <$!> freshMeta cxt (S.U (U0 cv))
         t  <- S.down <$!> coe1 cxt t a (V.Lift cv a')
@@ -467,16 +474,20 @@ infer1 cxt topT = case topT of
   P.Lam _ (bindToName cxt -> x) i a t -> case i of
     P.Named{}  -> elabError cxt topT NoNamedLambdaInference
     P.NoName i -> do
+      traceM "inferlam"
       a <- tyAnnot cxt a U1
       let va = eval1 cxt a
[…]

### 2021-04-22 src/Syntax.hs

@@ -16,7 +16,10 @@ data Tm0
   | Down Tm1
   | Field0 Tm0 Name Int
   | RecCon0 (Fields Tm0)
-  | Wk0 Tm0
+  | Add Tm0 Tm0
+  | Mul Tm0 Tm0
+  | Sub Tm0 Tm0
+  | IntLit Int
   deriving Show
 
 data Tm1
@@ -39,6 +42,7 @@ data Tm1
   | Inserted MetaVar Locals
   | Wk1 Tm1
   | Meta MetaVar
+  | Int
   deriving Show
 
 data Locals

### 2021-04-22 src/Unification.hs

@@ -1,3 +1,4 @@
+{-# options_ghc -Wno-unused-imports -Wno-type-defaults #-}
 
 module Unification where
 
@@ -11,6 +12,8 @@ import Evaluation
 import Exceptions
 import Values
 import qualified Syntax as S
+
+import Debug.Trace
 
 {-
 TODO:
@@ -75,19 +78,19 @@ invert gamma sp = do
       go SId =
         (0, mempty)
       go (SField1 sp x n) = do
-        throw CantUnify
+        impossible -- throw CantUnify
       go (SApp1 sp t i) =
         let (!dom, !ren) = go sp in
         case forceFU1 t of
           Var1 (Lvl x) | IM.notMember x ren ->
             ((,) $$! (dom + 1) $$! (IM.insert x dom ren))
           _ ->
-            throw CantUnify
+            impossible -- throw CantUnify
 
   let (!dom, !ren) = go sp
   pure $ PRen dom gamma ren
 
-renameSp :: MetaVar -> ConvState -> PartialRenaming -> S.Tm1 -> Spine -> IO S.Tm1
+renameSp :: Dbg => MetaVar -> ConvState -> PartialRenaming -> S.Tm1 -> Spine -> IO S.Tm1
 renameSp m st pren t sp = let
   go1   t = rename1 m st pren t;     {-# inline go1 #-}
   goSp sp = renameSp m st pren t sp; {-# inline goSp #-}
@@ -96,7 +99,7 @@ renameSp m st pren t sp = let
     SApp1 t u i   -> S.App1 <$!> goSp t <*!> go1 u <*!> pure i
     SField1 t x n -> S.Field1  <$!> goSp t <*!> pure x <*!> pure n
 
-rename0 :: MetaVar -> ConvState -> PartialRenaming -> Val0 -> IO S.Tm0
+rename0 :: Dbg => MetaVar -> ConvState -> PartialRenaming -> Val0 -> IO S.Tm0
 rename0 m st pren t = let
   go0 = rename0 m st pren; {-# inline go0 #-}
   go1 = rename1 m st pren; {-# inline go1 #-}
@@ -132,8 +135,12 @@ rename0 m st pren t = let
     Case t cs     -> S.Case <$!> go0 t <*!> goCases cs
[…]

### 2021-04-22 types2021/abstract.tex
update abstract

@@
 definitionally equal to $A$ and $t$ respectively after being embedded in
-2LTT. In short, every meta-level construction can be computed away. Annekov et
-al. \cite{twolevel} showed this without uniqueness of $A'$ and $t'$. We can get
-uniqueness as well, using the normalization of 2LTT: by induction on the
-(unique) normal forms of $A$ and $t$, we can show that they cannot contain
-meta-level subterms. This shows that normalization is a sound staging algorithm,
-but in practice we do not want to compute full normal forms; we want to compute
-\emph{meta-level redexes only}. This can be done with a variation of standard
-normalization-by-evaluation \cite{abel2013normalization,
+2LTT. In short, every meta-level construction can be computed away, and only
+object-level constructions remain in the result. Annekov et al. \cite{twolevel}
+showed this without uniqueness of $A'$ and $t'$ (as a conservativity theorem for
+2LTT). We can get uniqueness as well, using the normalization of 2LTT: by
+induction on the (unique) normal forms of $A$ and $t$, we can show that they
+cannot contain meta-level subterms. This shows that normalization is a sound
+staging algorithm, but in practice we do not want to compute full normal forms;
+we want to compute \emph{meta-level redexes only}. This can be done with a
+variation of standard normalization-by-evaluation \cite{abel2013normalization,
   Wieczorek:2018:CFN:3176245.3167091} which also keeps track of stages.
@@
 $\C^0$ which has $V^0$ as a sub-universe, and is closed under functions with
-domains in $\V^0$ and codomains in $\C^0$. The object theory supports
+domains in $\V^0$ and codomains in $\C^0$. This object theory supports
 compilation which requires only lambda lifting, but no closures. On its own, the

### 2021-04-22 app/Main.hs
abstract fix

@@ -88,16 +88,18 @@ test str = do
   displayState
 
 p1 = unlines [
-  "Alg  = [B: MTy, true: B, false: B]",
-  "Bool = (A : Alg) → A.B",
-  "id : Bool → Bool = λ b A. b [A.B, A.true, A.false]",
+  -- "Alg  = [B: MTy, true: B, false: B]",
+  -- "Bool = (A : Alg) → A.B",
+  -- "id : Bool → Bool = λ b A. b [A.B, A.true, A.false]",
 
-  "NatAlg = [N : MTy, zero : N, suc: N → N]",
-  "Nat    = (A : NatAlg) → A.N",
-  "zero   = λ (A : NatAlg). A.zero",
-  "suc : Nat → Nat = λ n A. A.suc (n A)",
-  "id  : Nat → Nat = λ n A. n [A.N, A.zero, A.suc]",
-  "add : Nat → Nat → Nat = λ a b A. a [A.N, b A, A.suc]"
+  -- "NatAlg = [N : MTy, zero : N, suc: N → N]",
+  -- "Nat    = (A : NatAlg) → A.N",
+  -- "zero   = λ (A : NatAlg). A.zero",
+  -- "suc : Nat → Nat = λ n A. A.suc (n A)",
+  -- "id  : Nat → Nat = λ n A. n [A.N, A.zero, A.suc]",
+  -- "add : Nat → Nat → Nat = λ a b A. a [A.N, b A, A.suc]",
+
+  "foo : (A : CTy) → A → A = λ A x. x"
 
   -- "foo = λ x. 10 + x"
 

### 2021-04-22 src/Presyntax.hs

@@ -50,6 +50,7 @@ data Tm
   | Case Pos Tm Pos [(Span, [Bind], Tm)]
   | Hole Pos
 
+
   | Int Span
   | IntLit Span Int
   | Add Tm Tm
@@ -84,6 +85,7 @@ span t = Span (left t) (right t) where
     Add l r             -> left l
     Mul l r             -> left l
     Sub l r             -> left l
+    Int (Span l r)      -> l
 
   right :: Tm -> Pos
   right = \case
@@ -110,3 +112,4 @@ span t = Span (left t) (right t) where
     Add _ r             -> right r
     Mul _ r             -> right r
     Sub _ r             -> right r
+    Int (Span l r)      -> r

### 2021-04-22 src/Pretty.hs

@@ -3,3 +3,9 @@ module Pretty where
 
 import Common
 import Syntax
+
+
+-- -- | Wrap in parens if expression precedence is lower than
+-- --   enclosing expression precedence.
+-- par :: Int -> Int -> ShowS -> ShowS
+-- par p p' = showParen (p' < p)

### 2021-04-22 src/Unification.hs

@@ -13,8 +13,6 @@ import Exceptions
 import Values
 import qualified Syntax as S
 
-import Debug.Trace
-
 {-
 TODO:
   - eta-short unification solutions:

### 2021-09-21 efop_abstract/abstract.tex
add types and efop materials

\documentclass{easychair}

\usepackage{doc}
\usepackage{scalerel}
\usepackage{amsfonts}

\newcommand{\easychair}{\textsf{easychair}}
\newcommand{\miktex}{MiK{\TeX}}
\newcommand{\texniccenter}{{\TeX}nicCenter}
\newcommand{\makefile}{\texttt{Makefile}}
\newcommand{\latexeditor}{LEd}
\newcommand{\emptycon}{\scaleobj{.75}\bullet}
\newcommand{\ext}{\triangleright}
\newcommand{\arri}{\Rightarrow}

\newcommand{\ms}[1]{\mathsf{#1}}
\newcommand{\ToS}{\mathsf{ToS}}
\newcommand{\U}{\mathsf{U}}
\newcommand{\Code}{\mathsf{Code}}
\newcommand{\Ty}{\mathsf{Ty}}
\newcommand{\V}{\mathsf{V}}
\newcommand{\C}{\mathsf{C}}
\newcommand{\Bool}{\ms{Bool}}

\title{Staged Compilation and Generativity%
  \thanks{The author was supported by the European Union, co-financed
    by the European Social Fund (EFOP-3.6.3-VEKOP-16-2017-00002).}}

\author{
Andr\'as Kov\'acs
}

% Institutes for affiliations are also joined by \and,
\institute{
  E\"otv\"os Lor\'and University,
  Budapest, Hungary \\
  \email{kovacsandras@inf.elte.hu}
}

\authorrunning{Kov\'acs}
\titlerunning{Staged Compilation and Generativity}
\pagenumbering{gobble}
\begin{document}

\maketitle

The purpose of staged compilation is to write code-generating programs in a safe
and ergonomic way. Although it is always possible to write metaprograms by
simply manipulating strings or deeply embedded syntax trees, this is often
error-prone and tedious. Staging is a way to have more guarantees about the
safety and well-typing of metaprograms, and also a way to integrate object-level
and meta-level syntaxes more organically.

Two-level type theory (2LTT) \cite{twolevel} was originally developed for the
purpose of doing synthetic homotopy theory, by adding a metaprogramming layer on
top of homotopy type theory \cite{hottbook}. However, it turns out that 2LTT is
also a great framework for metaprogramming and staging in general; it is
applicable to a wide range of object theories.

There is a natural semantics to 2LTT which justifies the metaprogramming view:
this is the \emph{presheaf model} of 2LTT. Here, meta-level types are presheaves
over the underlying category of the object theory. Hence, every meta-level
construction must be stable under the object-level morphisms.

\paragraph{Generativity}Generativity means that we can only generate code, but not look inside
object-level syntax and make decisions based on that. Generativity simplifies
staging, and it is often enforced in practical implementations
\cite{kiselyov14metaocaml}. However, non-generative staging provides additional
power and flexibility. A simple example for a non-generative feature is
conversion checking. Assume that given an object-level type $A$, $\Code\,A$ is
the type of meta-level programs which compute $A$-expressions. Conversion
checking may be postulated as $\ms{conversion}_A : (t\,u : \Code\,A) \to (t = u)
+ (t \neq u)$. We aim to investigate generativity in the presheaf model. We
observe that it depend on the choice of morphisms in the object theory.

If base morphisms are \emph{substitutions}, then $\ms{conversion}$ does not hold
in the model, because definitional inequality is not stable under substitution. For
example, inequal variables may become equal after substitution.

If base morphisms are \emph{weakenings}, then $\ms{conversion}$
holds. However, in this case the object theory can only support features which
can be specified with weakenings. For instance, this allows simple types at at
the object level, without any $\beta$-rules, but it does not allow polymorphism
or dependent types. This is still sufficient for many staging applications.

\paragraph{Yoneda lemma} It is worth to note that we get generativity statements
in the presheaf model from the Yoneda lemma. For example, the natural
transformation $\ms{y}\,\Bool_0 \Rightarrow \Bool_1$ must be a constant map in the
model. How is it possible then to have non-generativity? The answer is that
the Yoneda lemma can only meaningfully apply to \emph{types} (as opposed to
contexts) in the presheaf model, if context extension is preserved by Yoneda
embedding. If base morphisms are weakenings, the base category does not have all
binary products, and context extension is not preserved by $\ms{y}$. Hence, the
singleton context $\emptycon,\ms{y}\,\Bool_0$ is not representable, and the
Yoneda lemma does not say anything about dependency on $\ms{y}\,\Bool_0$.

\bibliographystyle{plain}
\bibliography{references}

\end{document}

### 2021-09-21 efop_abstract/ext_abstract.tex
add types and efop materials

% Dear Author,

% fill the fields marked with dots,

% please comments or remarks: dcs0@inf.elte.hu

%

%

\documentclass[11pt,a4paper,twoside]{article}
\usepackage{latexsym,amssymb}

\usepackage{doc}
\usepackage{scalerel}
\usepackage{amsfonts}
\usepackage{hyperref}
\usepackage{amsmath}

%% sorry, but it's very inconvenient to not define commands.
%% I undefine them at the bottom of the file
\newcommand{\msf}[1]{\mathsf{#1}}
\newcommand{\mi}[1]{\mathit{#1}}
\newcommand{\U}{\mathsf{U}}
\newcommand{\Code}{\mathsf{Code}}
\newcommand{\Ty}{\mathsf{Ty}}
\newcommand{\Bool}{\msf{Bool}}
\newcommand{\true}{\msf{true}}
\newcommand{\id}{\msf{id}}
\newcommand{\qtm}[1]{\langle #1\rangle}

%

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%  PLEASE, ONLY IN ENGLISH                                         %
%  PLEASE, USE LATEX2E                                             %

%                                                                  %

%  PLEASE, DO NOT DEFINE NEW COMMANDS                              %

%  USE NEWTHEOREMS BELOW, PLEASE, IF YOU WRITE THEOREMS AND SUCH   %

%                                                                  %

%  PLEASE, AT MOST TWO PAGES                                       %

%  PLEASE, CHECK THE NUMBER OF PAGES BEFORE SUBMISSION             %

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%

\newtheorem{theo}{Theorem}
\newtheorem{defi}[theo]{Definition}
\newtheorem{axio}[theo]{Axiom}
\newtheorem{clai}[theo]{Claim}
\newtheorem{coll}[theo]{Collorary}
\newtheorem{lemm}[theo]{Lemma}
\newtheorem{conj}[theo]{Conjecture}
\newtheorem{hypo}[theo]{Hypothesis}
\newtheorem{rema}[theo]{Remark}
\newcommand{\proof}{\noindent{\bf Proof:}\hspace{0.2cm}}

\setcounter{theo}{0}

%

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\textwidth 15.0cm
\textheight 22.0cm
\oddsidemargin 0.4cm
\evensidemargin 0.4cm
\topmargin  0.0cm
\frenchspacing

\pagestyle{myheadings}

\markboth{Developments in Computer Science,%
June 17 -- 19, 2021, ELTE,  Hungary}
{Developments in Computer Science,
June 17 -- 19, 2021, ELTE, Hungary}
%
%
\begin{document}
%
%
\begin{center}{\Large\bf
Staged Compilation and Generativity
}\end{center}
%
\begin{center}{\large\bf\noindent
András Kovács
}\\[2mm]
Eötvös Loránd University, Department of Programming Languages and Compilers
\\[1mm]\texttt{
kovacsandras@inf.elte.hu
}\end{center}
%
%
\vspace*{7mm}
%
%

\subsection*{Overview}

The purpose of staged compilation is to write code-generating programs in a safe
and ergonomic way. Although it is always possible to write metaprograms by
simply manipulating strings or deeply embedded syntax trees, this is often
error-prone and tedious. Staging is a way to have more guarantees about the
safety and well-typing of metaprograms, and also a way to integrate object-level
and meta-level syntaxes more organically.

Two-level type theory (2LTT) \cite{twolevel} was originally developed for the
purpose of doing synthetic homotopy theory, by adding a metaprogramming layer on
top of homotopy type theory \cite{hottbook}. However, it turns out that 2LTT is
also a great framework for metaprogramming and staging in general, and it is
applicable to a wide range of theories, both on the object and meta level.

There is a simple semantics to 2LTT which justifies the metaprogramming view:
this is the \emph{presheaf model} of 2LTT. Here, meta-level types are presheaves
over the underlying category of the object theory. Hence, every meta-level
construction must be stable under the object-level morphisms. The advantage of
working in 2LTT stems from stability under morphisms: if every construction is
automatically stable, it becomes possible to omit explicit handling of base
morphisms. More concretely, from the staging perspective, this means that we
never have to deal with scoping, renaming, substitution or de Bruijn indices in
the object syntax, when working in 2LTT.
\\\\
\textbf{Generativity.} Generativity means that we can only generate code, but
not make decisions based on the internal structure of object-level
syntax. Generativity simplifies staging, and it is often enforced in practical
implementations \cite{kiselyov14metaocaml}. However, non-generative staging
provides additional power and flexibility. A simple example for a non-generative
feature is conversion checking. This can be viewed as an axiom in 2LTT, which
says that meta-level (``strict'') equality of object-level values is decidable.
We aim to investigate semantics of non-generativity in the following.

\subsection*{Basic Rules and Usage of 2LTT}

To illustrate using 2LTT for staging, we specify a simple variant of 2LTT where
we have exactly the same dependent type theory for the object-level and
meta-level theories.

We have universes $\U^s_{i}$, where $s \in \{0,1\}$, denoting a \emph{stage} or
level in the 2LTT sense, and $i \in \mathbb{N}$ denotes a usual level index of
\emph{sizing} hierarchies. The two dimensions of indexing are orthogonal, and we
will elide the $i$ indices in the following. We assume Russell-style universes.
Both $\U_0$ and $\U_1$ may be closed under arbitrary type formers, but
eliminators in each \emph{only target the same universe}, i.e.\ elimination
cannot cross universes. We have the following operations:
\begin{itemize}
\item
  For $A : \U_0$, we have $\Code\,A : \U_1$. This is the type of meta-level
  programs which return object-level code with type $A$.
\item
  Quoting: for $A : \U_0$ and $t : A$ we have $\qtm{t} : \Code\,A$. In other
  words, for every object-level term we have a metaprogram which immediately
  returns that term.
\item Splicing: for $t : \Code\,A$, we have $\sim\!t : A$. This means running
  a metaprogram and inserting its result into object-level code.
\item We also know that quoting/splicing is a definitional isomorphism, so
  $\qtm{\sim\!t} = t$ and  $\sim\!\qtm{t} = t$.
\end{itemize}

A \emph{staging algorithm} takes as input a closed term $t : A$ where $A :
\U_0$, and splices the results of all metaprograms, so that we get an output
term which is free of splices. This can be implemented using variations of
normalization-by-evaluation \cite{abel2013normalization} which track current
stages. We do not detail staging algorithms here.

Let's look at some examples. We have the object-level identity function as usual:
\begin{alignat*}{3}
  &\msf{id_0} : (A : \U_0) \to A \to A\\
  &\msf{id_0} := \lambda\,A\,x.\,x
\end{alignat*}
Staging does not do anything with $\msf{id_0}$, since it has no
splices. Likewise if we apply $\msf{id_0}$ to object-level values, as in $\msf{id_0}\,\Bool\,\msf{true}$. We also have the meta-level version:
\begin{alignat*}{3}
  &\msf{id_1} : (A : \U_1) \to A \to A\\
  &\msf{id_1} := \lambda\,A\,x.\,x
\end{alignat*}
Note that this also works on object-level values, because of quoting:
\[
   \sim\!(\msf{id_0}\,(\Code\,\Bool_0)\,\qtm{\true_0}) : \Bool_0
\]
Staging the above term computes to $\sim\!\qtm{\true_0}$, which in turn
computes to $\true_0$. Thus, $\id_1$ is compile-time evaluated. There's a third
version, which is a specialized version of $\id_1$: it's also evaluated at compile time,
but it only works on object-level types:
\begin{alignat*}{3}
  &\msf{id_{Code}} : (A : \Code\,\U_0) \to \Code\,(\sim\!A) \to \Code\,(\sim\!A)\\
  &\msf{id_{Code}} := \lambda\,A\,x.\,x
\end{alignat*}
Now, $\sim\!(\id_{\Code}\,\qtm{\Bool_0}\,\qtm{\true_0})$ also stages to
$\true_0$. Meta-functions which are restricted to $\Code$ are also useful when
we want to define functions which are partially evaluated at compile time. For
example, if we want to inline a function argument for object-level list mapping:
\begin{alignat*}{3}
  &\rlap{$\msf{map} : (A\,B : \Code\,\U_0) \to (\Code(\sim\!A) \to \Code(\sim\!B))$}\\
  &\hspace{2em}\rlap{$\to \Code(\msf{List_0}\,(\sim\!A)) \to \Code(\msf{List_0}\,(\sim\!B)))$}\\
  &\msf{map} := \lambda\,A\,B\,f\,\mi{as}.\,\qtm{\msf{foldr_0}\,
    (\lambda\,a\,bs.\,\msf{cons_0}\,(\sim\!(f\,\qtm{a}))\,bs)\,\msf{nil_0}\,(\sim\!as)}
\end{alignat*}

\subsection*{Presheaf Model}

Why consider the presheaf model?  The reason is that it's the simplest semantics
which justifies the metaprogramming view of 2LTT. It is \emph{not} the same
thing as the staging algorithm, which is based on normalization-by-evaluation,
which is much more complicated to formalize \cite{bocquet2021induction}. The
presheaf model is fairly simple as far as models go, so it's interesting to see
which staging features can be justified with it. We skip presenting the whole
presheaf model. For details, we refer the reader to
\cite[Section~1.2]{huber-thesis}.

We give some examples for interpreting constructions in the model. We present the
results up to isomorphism, with some simplifications. We assume now that
object-level morphisms are substitutions. We have $\Bool_0 : \U_0$ and $\Bool_1
: \U_1$.
\begin{itemize}
  \item A closed function $t : \Bool_1 \to \Bool_1$ becomes a metatheoretical
    function in $\mathbb{B} \to \mathbb{B}$.
  \item A closed function $t : \Bool_0 \to \Bool_0$ becomes a closed object-theoretic
    function in $\Bool \to \Bool$.
  \item A closed function $t : \Code\,\Bool_0 \to \Bool_1$ becomes a function
    which maps a $\Bool$ term in any context to $\mathbb{B}$, such that the
    function commutes with object-theoretic substitution. For example if we have
    a variable $x$, we can substitute it with any term before feeding it to the
    semantic $t$, and the result is the same. In fact, this means that $t$
    cannot depend on its term argument, hence $t$ is specified simply by a
    $\mathbb{B}$.
  \item
    A closed function $t : \Bool_1 \to \Code\,\Bool_0$ becomes simply a pair of
    closed $\Bool$ terms.
\end{itemize}
\noindent\textbf{Yoneda lemma.} The Yoneda lemma is a general statement which
restricts the way meta-level values can depend on object-level ones. First, note that any
object-level typing context $\Gamma$ can be mapped to a presheaf, by taking the
sets of parallel substitutions into $\Gamma$. This is the \emph{Yoneda
embedding} of $\Gamma$, denoted by $\msf{y}\Gamma$. The Yoneda lemma says
that we have the following isomorphism of sets:
\[
   (\msf{y}\,\Gamma \Rightarrow \Delta) \simeq |\Delta|\,\Gamma
\]
where $\Rightarrow$ means a natural transformation, and $|\Delta|\,\Gamma$
denotes the set that we get by evaluating the $\Delta$ presheaf at the
object-level $\Gamma$ context. From this, what we essentially get is that any
2LTT term $\Gamma \vdash t : A$, such that $\Gamma$ is essentially interpreted
as $\msf{y}\,\Gamma'$ for some $\Gamma'$, is interpreted as an element of
$|A|\Gamma'$. We call $\Gamma$ \emph{representable} is there is such $\Gamma'$.

In particular, if $\Gamma \vdash t : \Bool_1$ and $\Gamma$ representable, then
since $|\Bool_1|\,\Gamma' = \mathbb{B}$, $t$ is simply an element of
$\mathbb{B}$ in the semantics, and cannot depend on the typing context.

In short, whether the Yoneda lemma applies to a given term, depends on whether
the typing context is representable. In turn, the representability of the context
depends on what morphisms are in the object theory. We consider two options.
\\\\
\textbf{1. Morphisms are substitutions.} In this case, $\msf{y}$ preserves
context extension, i.e.\ $\msf{y}\,(\Gamma,\,x : A) \simeq (\msf{y}\,\Gamma,\,x
: \msf{y}A)$ in the presheaf model. That's because a substitution which targets
$(\Gamma,\,x : A)$ is equivalent to a pair of substitutions, targeting $\Gamma$
and $A$ respectively. Therefore, if we have $x_1 : A_1,\,x_2 : A_2,\,...,\, x_i :
A_i \vdash t : B$, such that all $A_i$ are representable, the entire context is
also representable, and the Yoneda lemma applies.

Which types are representable? For starters, every type of the form $\Code\,A$,
since $\Code$ in the model is essentially interpreted as $\msf{y}$ (eliding the
formal complications arising from possible dependencies of $A$ on the context).
Also, if we have $x : A$ in a context, where $A : \U_0$, that context extension
is also interpreted as extension with $\msf{y}A$. In short: if the context
only has $\Code$ types or types in $\U_0$, it is representable.

This greatly limits non-generative features in the model. Consider adding the
axiom which says that meta-level equality of object-level values is decidable:
$x : \Code\,A,\,y : \Code\,A \vdash \msf{conversion}_A : (x = y) + (x = y \to
\bot)$.  This is a simple non-generative axiom, since any constructive
interpretation must look inside $\Code$-s. This axiom is false if object
morphisms are substitutions. That's because the context is representable, so we
can simplify using the Yoneda lemma. The statement that we get in the model is
that ``two terms are either equal, or they are inequal and remain inequal after
arbitrarily substitutions''. Now, if we pick two \emph{variables} $x$ and $y$
such that $x \neq y$, then they are not equal, but they can be also made
equal by substituting both variables with the same term.

Can we repair this? One possibility is to have decidable equality only for
\emph{closed} terms. However, the syntax of 2LTT provides no way to talk about
closed terms. Instead we'd have to use a closed modality
\cite{bocquet2021induction}. This would be interesting to investigate in future
work.
\\\\
\textbf{2. Morphisms are weakenings.} In this case, object morphisms are are
so-called \emph{order-preserving embeddings}, meaning that a morphism can drop
zero or more entries from a context, so morphisms are essentially bitmasks which
mark a sub-context. The action of weakening embeds terms in larger
contexts. Moreover, $\msf{y}$ does not preserve context extension. Hence,
typing contexts are not necessarily representable, even if they only contain object-level
bindings. So the Yoneda-reduction of dependencies generally does not apply.

Now, $\msf{conversion}$ is fine, because inequality of terms is stable under
weakening. On the other hand, by only having weakening in the object theory, the
range of supported object theories is greatly restricted. For example, we can't
have $\beta$-reduction for functions in the equational theory, since that's
specified using substitution. Likewise, dependent types are out, since the
typing of dependent elimination involves substitution.

Simple type theories still work, if we only have weakening in their equational
theory. For the perspective of staging, this is fine, because in \emph{code
generation} we care about the intensional definition of programs, and we do not
want to equate $\beta$-reducts, since a primary use-case of staging is to
improve runtime performance, hence distinguish between possibly
$\beta$-convertible programs.

It appears that if morphisms are weakenings, then the presheaf model is
compatible with a wide range of non-generative axioms. For example, we can also
postulate \emph{countability} of $\Code\,A$, i.e.\ that there are injections
$\msf{index}_A : \Code\,A \to \msf{Nat}_1$. In the presheaf model, the indexing function
works by enumerating maximally strengthened terms, which are stable under
weakening.
\\\\
\textbf{Other ways of justifying non-generativity.} An alternative solution
would be to use something other than the presheaf model to justify
non-generative axioms. For example: could we use the staging algorithm itself,
i.e.\ does normalization-by-evaluation support non-generativity? It seems
likely, as semantic values need only be stable under weakening. This remains
future work.

\begin{thebibliography}{1}

\bibitem{abel2013normalization}
Andreas Abel.
\newblock {\em Normalization by Evaluation: Dependent Types and
  Impredicativity}.
\newblock PhD thesis, Ludwig-Maximilians-Universit{\"a}t M{\"u}nchen, 2013.
\newblock Habilitation thesis.

\bibitem{twolevel}
Danil Annenkov, Paolo Capriotti, Nicolai Kraus, and Christian Sattler.
\newblock Two-level type theory and applications.
\newblock {\em ArXiv e-prints}, may 2019.

\bibitem{bocquet2021induction}
Rafa{\"e}l Bocquet, Ambrus Kaposi, and Christian Sattler.
\newblock Induction principles for type theories, internally to presheaf
  categories.
\newblock {\em arXiv preprint arXiv:2102.11649}, 2021.

\bibitem{huber-thesis}
Simon Huber.
\newblock {\em Cubical Interpretations of Type Theory}.
\newblock PhD thesis, University of Gothenburg, 2016.

\bibitem{kiselyov14metaocaml}
Oleg Kiselyov.
\newblock The design and implementation of {BER} metaocaml - system
  description.
\newblock In Michael Codish and Eijiro Sumii, editors, {\em Functional and
  Logic Programming - 12th International Symposium, {FLOPS} 2014, Kanazawa,
  Japan, June 4-6, 2014. Proceedings}, volume 8475 of {\em Lecture Notes in
  Computer Science}, pages 86--102. Springer, 2014.

\bibitem{hottbook}
The {Univalent Foundations Program}.
\newblock {\em Homotopy Type Theory: Univalent Foundations of Mathematics}.
\newblock \url{https://homotopytypetheory.org/book}, Institute for Advanced
  Study, 2013.
\end{thebibliography}

%% --------------------------------------------------------------------------------

\let\msf\undefined
\let\mi\undefined
\let\U\undefined
\let\Code\undefined
\let\Ty\undefined
\let\Bool\undefined
\let\true\undefined
\let\id\undefined
\let\qtm\undefined

%% --------------------------------------------------------------------------------

\end{document}

### 2021-09-21 app/Main.hs

@@ -61,10 +61,10 @@ displayState = do
       putStrLn (show x ++ " : " ++ showTm1Top a)
       putStrLn ("  = " ++ showTm1Top t)
       nl
-      putStrLn "-- NORMAL FORM"
-      putStrLn (show x ++ " : " ++ showTm1Top' a)
-      putStrLn ("  = " ++ showTm1Top' t)
-      nl
+      -- putStrLn "-- NORMAL FORM"
+      -- putStrLn (show x ++ " : " ++ showTm1Top' a)
+      -- putStrLn ("  = " ++ showTm1Top' t)
+      -- nl
     TETyCon{} -> do
       putStrLn "<tycon not supported>"
     TEDataCon{} -> do
@@ -94,31 +94,46 @@ p1 = unlines [
 
   "VTy : U1 = U0 Val",
   "CTy : U1 = U0 Comp",
-  "Pair : VTy → VTy → VTy = λ A B . [fst: A, snd: B]",
+  "the : (A : U1) → A → A = λ A x. x",
+  -- "Pair : VTy → VTy → VTy = λ A B . [fst: A, snd: B]",
 
   -- "let-insertion"
   "CPair : VTy → VTy → U1 = λ A B. (P : VTy) → (A → B → P) → P",
-  "dup : {A : U0 Val} → A → CPair A A = λ a P p. let v := a; p a a",
+  -- "dup : {A : U0 Val} → A → CPair A A = λ a P p. let v := a; p a a",
 
-  -- -- here      Template Haskell
-  -- -- <_>        [| _ |]
-  -- -- ~_         $(_)
+  -- -- -- here      Template Haskell
+  -- -- -- <_>        [| _ |]
+  -- -- -- ~_         $(_)
 
   "Eq : {A : U1} → A → A → U1",
   "  = λ {A} x y. (P : A → U1) → P x → P y",
   "refl : {A : U1}{x :A} → Eq {A} x x = λ P px. px",
 
-  "id    : {A : U1} → A → A = λ x. x",
-  "const : {A B : U1} → A → B → A = λ a b. a",
-  "comp  : {A B C : U1} → (B → C) → (A → B) → A → C = λ f g x. f (g x)",
+  -- "id    : {A : U1} → A → A = λ x. x",
+  -- "const : {A B : U1} → A → B → A = λ a b. a",
[…]

### 2021-09-21 src/Exceptions.hs

@@ -2,7 +2,6 @@
 module Exceptions (throwIO, catch, throw, module Exceptions) where
 
 import Control.Exception
-import qualified Data.ByteString as B
 
 import Common
 import qualified Syntax    as S
@@ -23,8 +22,8 @@ data SolutionEx
 data UnifyEx
   = Unify0 V.Val0 V.Val0
   | Unify1 V.Val1 V.Val1
-  | forall a. (Show a) => UnifyEq a a
   | SolutionError V.Val1 V.Val1 SolutionEx
+  | UnifyFieldName Name Name
 deriving instance Show UnifyEx
 
 -- | Unification exception in local unification context.
@@ -34,10 +33,12 @@ data UnifyInner = UnifyInner Unif.Cxt UnifyEx
 -- | Exception thrown during elaboration
 data ElabEx
   = UnifyOuter V.Val1 V.Val1 UnifyInner
-  | NoSuchField S.Tm1 RawName
+  | NoSuchFieldName S.Tm1 RawName
+  | NoSuchFieldIx   S.Tm1 Int
   | NoSuchArgument RawName
   | NameNotInScope RawName
   | IcitMismatch Icit Icit
+  | NegativeFieldIndex
   | NoImplicitLam0
   | ExpectedVal S.Tm1
   | FieldNameMismatch Name Name
@@ -48,7 +49,7 @@ data ElabEx
   | ExpectedNonEmptyRecCon
   | ExpectedType S.Ty
   | CantInferTuple
-  | ExpectedRecord S.Ty
+  | ExpectedRecord V.Ty
   | ExpectedRuntimeType S.Ty
   | CantInferRec1
   | CantInfer

### 2021-09-21 src/Presyntax.hs

@@ -23,6 +23,11 @@ data Bind
   = Bind Span
   | DontBind
 
+data Proj
+  = ProjIx Span Int
+  | ProjName Span
+  deriving Show
+
 instance Show Bind where
   show (Bind x) = show x
   show DontBind = "_"
@@ -46,7 +51,7 @@ data Tm
   | RecCon Span [(Span, Tm)]
   | EmptyRec Span                 -- overloads both tt and Top
   | Tuple Span [Tm]
-  | Field Tm Span
+  | Field Tm Proj
   | Case Pos Tm Pos [(Span, [Bind], Tm)]
   | Hole Pos
   | Int Span
@@ -59,6 +64,7 @@ deriving instance Show Tm
 
 span :: Tm -> Span
 span t = Span (left t) (right t) where
+
   left :: Tm -> Pos
   left = \case
     Var (Span l _)      -> l
@@ -74,7 +80,7 @@ span t = Span (left t) (right t) where
     RecCon (Span l _) _ -> l
     EmptyRec (Span l _) -> l
     Tuple (Span l _) _  -> l
-    Field t _           -> left t
+    Field t pr          -> left t
     Case l _ _ _        -> l
     Hole l              -> l
     IntLit (Span l _) _ -> l
@@ -90,30 +96,31 @@ span t = Span (left t) (right t) where
 
   right :: Tm -> Pos
   right = \case
-    Var (Span _ r)      -> r
-    Let0 _ _ _ _ t      -> right t
-    Let1 _ _ _ _ t      -> right t
-    Pi _ _ _ _ t        -> right t
-    Lam _ _ _ _ t       -> right t
-    App _ t _           -> right t
-    Lift _ t            -> right t
-    Up (Span _ r) _     -> r
-    Down _ t            -> right t
-    Rec (Span _ r) _    -> r
-    RecCon (Span _ r) _ -> r
-    EmptyRec (Span _ r) -> r
-    Tuple (Span _ r) _  -> r
-    Field _ (Span _ r)  -> r
-    Case _ _ r []       -> r
-    Case _ _ r ts       -> case last ts of (_, _, t) -> right t
-    Hole r              -> r
-    IntLit (Span _ r) _ -> r
-    Add _ r             -> right r
-    Mul _ r             -> right r
-    Sub _ r             -> right r
-    Int (Span l r)      -> r
-    CV (Span l r)       -> r
-    U0 l t              -> right t
-    U1 (Span l r)       -> r
[…]

### 2021-09-21 src/Pretty.hs

@@ -112,28 +112,42 @@ lamBind x i = icit i brace id (name x)
 recCon0 :: [Name] -> Fields Tm0 -> ShowS
 recCon0 ns = \case
   FNil           -> id
-  FCons x t FNil -> name x . (" = "++). tm0 tmp ns t
-  FCons x t ts   -> name x . (" = "++). tm0 tmp ns t . (", "++) . recCon0 ns ts
+  FCons x t FNil
+    | NEmpty <- x -> tm0 tmp ns t
+    | True        -> name x . (" = "++). tm0 tmp ns t
+  FCons x t ts
+    | NEmpty <- x -> tm0 tmp ns t . (", "++) . recCon0 ns ts
+    | True        -> name x . (" = "++). tm0 tmp ns t . (", "++) . recCon0 ns ts
 
 recCon1 :: [Name] -> Fields Tm1 -> ShowS
 recCon1 ns = \case
   FNil           -> id
-  FCons x t FNil -> name x . (" = "++). tm1 tmp ns t
-  FCons x t ts   -> name x . (" = "++). tm1 tmp ns t . (", "++) . recCon1 ns ts
+  FCons x t FNil
+    | NEmpty <- x -> tm1 tmp ns t
+    | True        -> name x . (" = "++). tm1 tmp ns t
+  FCons x t ts
+    | NEmpty <- x -> tm1 tmp ns t . (", "++) . recCon1 ns ts
+    | True        -> name x . (" = "++). tm1 tmp ns t . (", "++) . recCon1 ns ts
 
 rec0 :: [Name] -> Fields Tm1 -> ShowS
 rec0 ns = \case
   FNil           -> id
-  FCons x t FNil -> name x . (" : "++). tm1 tmp ns t
-  FCons x t ts   -> name x . (" : "++). tm1 tmp ns t . (", "++) . rec0 ns ts
+  FCons x t FNil
+    | NEmpty <- x -> tm1 tmp ns t
+    | True        -> name x . (" : "++). tm1 tmp ns t
+  FCons x t ts
+    | NEmpty <- x -> tm1 tmp ns t . (", "++) . rec0 ns ts
+    | True        -> name x . (" : "++). tm1 tmp ns t . (", "++) . rec0 ns ts
 
 rec1 :: [Name] -> Fields Tm1 -> ShowS
 rec1 ns = \case
   FNil -> id
-  FCons (fresh ns -> x) t FNil ->
[…]

### 2021-09-22 efop_abstract/ext_abstract.tex
typo
typo again

@@
 as $\msf{y}\,\Gamma'$ for some $\Gamma'$, is interpreted as an element of
-$|A|\Gamma'$. We call $\Gamma$ \emph{representable} is there is such $\Gamma'$.
+$|A|\Gamma'$. We call $\Gamma$ \emph{representable} if there is such $\Gamma'$.
 
@@
 Simple type theories still work, if we only have weakening in their equational
-theory. For the perspective of staging, this is fine, because in \emph{code
+theory. From the perspective of staging, this is fine, because in \emph{code
 generation} we care about the intensional definition of programs, and we do not

### 2022-02-03 paper/paper.tex
stuff, start paper notes

%%
%% This is file `sample-acmsmall.tex',
%% generated with the docstrip utility.
%%
%% The original source files were:
%%
%% samples.dtx  (with options: `acmsmall')
%%
%% IMPORTANT NOTICE:
%%
%% For the copyright see the source file.
%%
%% Any modified versions of this file must be renamed
%% with new filenames distinct from sample-acmsmall.tex.
%%
%% For distribution of the original source see the terms
%% for copying and modification in the file samples.dtx.
%%
%% This generated file may be distributed as long as the
%% original source files, as listed above, are part of the
%% same distribution. (The sources need not necessarily be
%% in the same archive or directory.)
%%
%%
%% Commands for TeXCount
%TC:macro \cite [option:text,text]
%TC:macro \citep [option:text,text]
%TC:macro \citet [option:text,text]
%TC:envir table 0 1
%TC:envir table* 0 1
%TC:envir tabular [ignore] word
%TC:envir displaymath 0 word
%TC:envir math 0 word
%TC:envir comment 0 0
%%
%%
%% The first command in your LaTeX source must be the \documentclass command.
\documentclass[acmsmall]{acmart}

%%
%% \BibTeX command to typeset BibTeX logo in the docs
\AtBeginDocument{%
  \providecommand\BibTeX{{%
    \normalfont B\kern-0.5em{\scshape i\kern-0.25em b}\kern-0.8em\TeX}}}

%% Rights management information.  This information is sent to you
%% when you complete the rights form.  These commands have SAMPLE
%% values in them; it is your responsibility as an author to replace
%% the commands and values with those provided to you when you
%% complete the rights form.
\setcopyright{acmcopyright}
\copyrightyear{2018}
\acmYear{2018}
\acmDOI{XXXXXXX.XXXXXXX}

%%
%% These commands are for a JOURNAL article.
\acmJournal{JACM}
\acmVolume{37}
\acmNumber{4}
\acmArticle{111}
\acmMonth{8}

%%
%% Submission ID.
%% Use this when submitting an article to a sponsored event. You'll
%% receive a unique submission ID from the organizers
%% of the event, and this ID should be used as the parameter to this command.
%%\acmSubmissionID{123-A56-BU3}

%%
%% The majority of ACM publications use numbered citations and
%% references.  The command \citestyle{authoryear} switches to the
%% "author year" style.
%%
%% If you are preparing content for an event
%% sponsored by ACM SIGGRAPH, you must use the "author year" style of
%% citations and references.
%% Uncommenting
%% the next command will enable that style.
%%\citestyle{acmauthoryear}

%% --------------------------------------------------------------------------------

\usepackage{xcolor}
\usepackage{mathpartir}
\usepackage{todonotes}
\presetkeys{todonotes}{inline}{}
\usepackage{scalerel}
%% \usepackage{amssymb}

\newcommand{\mit}[1]{\mathit{#1}}
\newcommand{\msf}[1]{\mathsf{#1}}
\newcommand{\mbb}[1]{\mathbb{#1}}
\newcommand{\mbf}[1]{\mathbf{#1}}
\newcommand{\bs}[1]{\boldsymbol{#1}}

%% --------------------------------------------------------------------------------

%%
%% end of the preamble, start of the body of the document source.
\begin{document}

%%
%% The "title" command has an optional parameter,
%% allowing the author to define a "short title" to be used in page headers.
\title{Using Two-Level Type Theory for Staged Compilation}

%%
%% The "author" command and its associated commands are used to define
%% the authors and their affiliations.
%% Of note is the shared affiliation of the first two authors, and the
%% "authornote" and "authornotemark" commands
%% used to denote shared contribution to the research.
\author{András Kovács}
\email{kovacsandras@inf.elte.hu}
\orcid{0000-0002-6375-9781}
\affiliation{%
  \institution{Eötvös Loránd University}
  \country{Hungary}
  \city{Budapest}
}

%%
%% By default, the full list of authors will be used in the page
%% headers. Often, this list is too long, and will overlap
%% other information printed in the page headers. This command allows
%% the author to define a more concise list
%% of authors' names for this purpose.
%% \renewcommand{\shortauthors}{Trovato and Tobin, et al.}

%%
%% The abstract is a short summary of the work to be presented in the
%% article.
\begin{abstract}
  A clear and well-documented \LaTeX\ document is presented as an
  article formatted for publication by ACM in a conference proceedings
  or journal publication. Based on the ``acmart'' document class, this
  article presents and explains many of the common variations, as well
  as many of the formatting elements an author may use in the
  preparation of the documentation of their work.
\end{abstract}

%%
%% The code below is generated by the tool at http://dl.acm.org/ccs.cfm.
%% Please copy and paste the code instead of the example below.
%%
\begin{CCSXML}
\end{CCSXML}

%%
%% Keywords. The author(s) should pick words that accurately describe
%% the work being presented. Separate the keywords with commas.
\keywords{type theory, two-level type theory, staged compilation}

%%
%% This command processes the author and affiliation and title
%% information and builds the first part of the formatted document.
\maketitle

\section{Introduction}
ACM's consolidated article template, introduced in 2017, provides a
consistent \LaTeX\ style for use across ACM publications, and
incorporates accessibility and metadata-extraction functionality
necessary for future Digital Library endeavors. Numerous ACM and
SIG-specific \LaTeX\ templates have been examined, and their unique
features incorporated into this single new template.

If you are new to publishing with ACM, this document is a valuable
guide to the process of preparing your work for publication. If you
have published with ACM before, this document provides insight and
instruction into more recent changes to the article template.

The ``\verb|acmart|'' document class can be used to prepare articles
for any ACM publication --- conference or journal, and for any stage
of publication, from review to final ``camera-ready'' copy, to the
author's own version, with {\itshape very} few changes to the source.

\section{Template Overview}
As noted in the introduction, the ``\verb|acmart|'' document class can
be used to prepare many different kinds of documentation --- a
double-blind initial submission of a full-length technical paper, a
two-page SIGGRAPH Emerging Technologies abstract, a ``camera-ready''
journal article, a SIGCHI Extended Abstract, and more --- all by
selecting the appropriate {\itshape template style} and {\itshape
  template parameters}.

This document will explain the major features of the document
class. For further information, the {\itshape \LaTeX\ User's Guide} is
available from
\url{https://www.acm.org/publications/proceedings-template}.

\subsection{Template Styles}

The primary parameter given to the ``\verb|acmart|'' document class is
the {\itshape template style} which corresponds to the kind of publication
or SIG publishing the work. This parameter is enclosed in square
brackets and is a part of the {\verb|documentclass|} command:
\begin{verbatim}
  \documentclass[STYLE]{acmart}
\end{verbatim}

Journals use one of three template styles. All but three ACM journals
use the {\verb|acmsmall|} template style:
\begin{itemize}
\item {\texttt{acmsmall}}: The default journal template style.
\item {\texttt{acmlarge}}: Used by JOCCH and TAP.
\item {\texttt{acmtog}}: Used by TOG.
\end{itemize}

The majority of conference proceedings documentation will use the {\verb|acmconf|} template style.
\begin{itemize}
\item {\texttt{acmconf}}: The default proceedings template style.
\item{\texttt{sigchi}}: Used for SIGCHI conference articles.
\item{\texttt{sigchi-a}}: Used for SIGCHI ``Extended Abstract'' articles.
\item{\texttt{sigplan}}: Used for SIGPLAN conference articles.
\end{itemize}

\subsection{Template Parameters}

In addition to specifying the {\itshape template style} to be used in
formatting your work, there are a number of {\itshape template parameters}
which modify some part of the applied template style. A complete list
of these parameters can be found in the {\itshape \LaTeX\ User's Guide.}

Frequently-used parameters, or combinations of parameters, include:
\begin{itemize}
\item {\texttt{anonymous,review}}: Suitable for a ``double-blind''
  conference submission. Anonymizes the work and includes line
  numbers. Use with the \texttt{\acmSubmissionID} command to print the
  submission's unique ID on each page of the work.
\item{\texttt{authorversion}}: Produces a version of the work suitable
  for posting by the author.
\item{\texttt{screen}}: Produces colored hyperlinks.
\end{itemize}

This document uses the following string as the first command in the
source file:
\begin{verbatim}
\documentclass[acmsmall]{acmart}
\end{verbatim}

\section{Modifications}

Modifying the template --- including but not limited to: adjusting
margins, typeface sizes, line spacing, paragraph and list definitions,
and the use of the \verb|\vspace| command to manually adjust the
vertical spacing between elements of your work --- is not allowed.

{\bfseries Your document will be returned to you for revision if
  modifications are discovered.}

\section{Typefaces}

The ``\verb|acmart|'' document class requires the use of the
``Libertine'' typeface family. Your \TeX\ installation should include
this set of packages. Please do not substitute other typefaces. The
``\verb|lmodern|'' and ``\verb|ltimes|'' packages should not be used,
as they will override the built-in typeface families.

\section{Title Information}

The title of your work should use capital letters appropriately -
\url{https://capitalizemytitle.com/} has useful rules for
capitalization. Use the {\verb|title|} command to define the title of
your work. If your work has a subtitle, define it with the
{\verb|subtitle|} command.  Do not insert line breaks in your title.

If your title is lengthy, you must define a short version to be used
in the page headers, to prevent overlapping text. The \verb|title|
command has a ``short title'' parameter:
\begin{verbatim}
  \title[short title]{full title}
\end{verbatim}

\section{Authors and Affiliations}

Each author must be defined separately for accurate metadata
identification.  As an exception, multiple authors may share one
affiliation. Authors' names should not be abbreviated; use full first
names wherever possible. Include authors' e-mail addresses whenever
possible.

Grouping authors' names or e-mail addresses, or providing an ``e-mail
alias,'' as shown below, is not acceptable:
\begin{verbatim}
  \author{Brooke Aster, David Mehldau}
  \email{dave,judy,steve@university.edu}
  \email{firstname.lastname@phillips.org}
\end{verbatim}

The \verb|authornote| and \verb|authornotemark| commands allow a note
to apply to multiple authors --- for example, if the first two authors
of an article contributed equally to the work.

If your author list is lengthy, you must define a shortened version of
the list of authors to be used in the page headers, to prevent
overlapping text. The following command should be placed just after
the last \verb|\author{}| definition:
\begin{verbatim}
  \renewcommand{\shortauthors}{McCartney, et al.}
\end{verbatim}
Omitting this command will force the use of a concatenated list of all
of the authors' names, which may result in overlapping text in the
page headers.

The article template's documentation, available at
\url{https://www.acm.org/publications/proceedings-template}, has a
complete explanation of these commands and tips for their effective
use.

Note that authors' addresses are mandatory for journal articles.

\section{Rights Information}

Authors of any work published by ACM will need to complete a rights
form. Depending on the kind of work, and the rights management choice
made by the author, this may be copyright transfer, permission,
license, or an OA (open access) agreement.

Regardless of the rights management choice, the author will receive a
copy of the completed rights form once it has been submitted. This
form contains \LaTeX\ commands that must be copied into the source
document. When the document source is compiled, these commands and
their parameters add formatted text to several areas of the final
document:
\begin{itemize}
\item the ``ACM Reference Format'' text on the first page.
\item the ``rights management'' text on the first page.
\item the conference information in the page header(s).
\end{itemize}

Rights information is unique to the work; if you are preparing several
works for an event, make sure to use the correct set of commands with
each of the works.

The ACM Reference Format text is required for all articles over one
page in length, and is optional for one-page articles (abstracts).

\section{CCS Concepts and User-Defined Keywords}

Two elements of the ``acmart'' document class provide powerful
taxonomic tools for you to help readers find your work in an online
search.

The ACM Computing Classification System ---
\url{https://www.acm.org/publications/class-2012} --- is a set of
classifiers and concepts that describe the computing
discipline. Authors can select entries from this classification
system, via \url{https://dl.acm.org/ccs/ccs.cfm}, and generate the
commands to be included in the \LaTeX\ source.

User-defined keywords are a comma-separated list of words and phrases
of the authors' choosing, providing a more flexible way of describing
the research being presented.

CCS concepts and user-defined keywords are required for for all
articles over two pages in length, and are optional for one- and
two-page articles (or abstracts).

\section{Sectioning Commands}

Your work should use standard \LaTeX\ sectioning commands:
\verb|section|, \verb|subsection|, \verb|subsubsection|, and
\verb|paragraph|. They should be numbered; do not remove the numbering
from the commands.

Simulating a sectioning command by setting the first word or words of
a paragraph in boldface or italicized text is {\bfseries not allowed.}

\section{Tables}

The ``\verb|acmart|'' document class includes the ``\verb|booktabs|''
package --- \url{https://ctan.org/pkg/booktabs} --- for preparing
high-quality tables.

Table captions are placed {\itshape above} the table.

Because tables cannot be split across pages, the best placement for
them is typically the top of the page nearest their initial cite.  To
ensure this proper ``floating'' placement of tables, use the
environment \textbf{table} to enclose the table's contents and the
table caption.  The contents of the table itself must go in the
\textbf{tabular} environment, to be aligned properly in rows and
columns, with the desired horizontal and vertical rules.  Again,
detailed instructions on \textbf{tabular} material are found in the
\textit{\LaTeX\ User's Guide}.

Immediately following this sentence is the point at which
Table~\ref{tab:freq} is included in the input file; compare the
placement of the table here with the table in the printed output of
this document.

\begin{table}
  \caption{Frequency of Special Characters}
  \label{tab:freq}
  \begin{tabular}{ccl}
    \toprule
    Non-English or Math&Frequency&Comments\\
    \midrule
    \O & 1 in 1,000& For Swedish names\\
    $\pi$ & 1 in 5& Common in math\\
    \$ & 4 in 5 & Used in business\\
    $\Psi^2_1$ & 1 in 40,000& Unexplained usage\\
  \bottomrule
\end{tabular}
\end{table}

To set a wider table, which takes up the whole width of the page's
live area, use the environment \textbf{table*} to enclose the table's
contents and the table caption.  As with a single-column table, this
wide table will ``float'' to a location deemed more
desirable. Immediately following this sentence is the point at which
Table~\ref{tab:commands} is included in the input file; again, it is
instructive to compare the placement of the table here with the table
in the printed output of this document.

\begin{table*}
  \caption{Some Typical Commands}
  \label{tab:commands}
  \begin{tabular}{ccl}
    \toprule
    Command &A Number & Comments\\
    \midrule
    \texttt{{\char'134}author} & 100& Author \\
    \texttt{{\char'134}table}& 300 & For tables\\
    \texttt{{\char'134}table*}& 400& For wider tables\\
    \bottomrule
  \end{tabular}
\end{table*}

Always use midrule to separate table header rows from data rows, and
use it only for this purpose. This enables assistive technologies to
recognise table headers and support their users in navigating tables
more easily.

\section{Math Equations}
You may want to display math equations in three distinct styles:
inline, numbered or non-numbered display.  Each of the three are
discussed in the next sections.

\subsection{Inline (In-text) Equations}

\begin{math}
  \lim_{n\rightarrow \infty}x=0
\end{math},
set here in in-line math style, looks slightly different when
set in display style.  (See next section).

\subsection{Display Equations}
A numbered display equation---one set off by vertical space from the
text and centered horizontally---is produced by the \textbf{equation}
environment. An unnumbered display equation is produced by the
\textbf{displaymath} environment.

Again, in either environment, you can use any of the symbols and
structures available in \LaTeX\@; this section will just give a couple
of examples of display equations in context.  First, consider the
equation, shown as an inline equation above:
\begin{equation}
  \lim_{n\rightarrow \infty}x=0
\end{equation}
Notice how it is formatted somewhat differently in
the \textbf{displaymath}
environment.  Now, we'll enter an unnumbered equation:
\begin{displaymath}
  \sum_{i=0}^{\infty} x + 1
\end{displaymath}
and follow it with another numbered equation:
\begin{equation}
  \sum_{i=0}^{\infty}x_i=\int_{0}^{\pi+2} f
\end{equation}
just to demonstrate \LaTeX's able handling of numbering.

\section{Figures}

Your figures should contain a caption which describes the figure to
the reader.

Figure captions are placed {\itshape below} the figure.

Every figure should also have a figure description unless it is purely
decorative. These descriptions convey what’s in the image to someone
who cannot see it. They are also used by search engine crawlers for
indexing images, and when images cannot be loaded.

A figure description must be unformatted plain text less than 2000
characters long (including spaces).  {\bfseries Figure descriptions
  should not repeat the figure caption – their purpose is to capture
  important information that is not already provided in the caption or
  the main text of the paper.} For figures that convey important and
complex new information, a short text description may not be
adequate. More complex alternative descriptions can be placed in an
appendix and referenced in a short figure description. For example,
provide a data table capturing the information in a bar chart, or a
structured list representing a graph.  For additional information
regarding how best to write figure descriptions and why doing this is
so important, please see
\url{https://www.acm.org/publications/taps/describing-figures/}.

\subsection{The ``Teaser Figure''}

A ``teaser figure'' is an image, or set of images in one figure, that
are placed after all author and affiliation information, and before
the body of the article, spanning the page. If you wish to have such a
figure in your article, place the command immediately before the
\verb|\maketitle| command:
\begin{verbatim}
  \begin{teaserfigure}
    \includegraphics[width=\textwidth]{sampleteaser}
    \caption{figure caption}
    \Description{figure description}
  \end{teaserfigure}
\end{verbatim}

\section{Citations and Bibliographies}

The use of \BibTeX\ for the preparation and formatting of one's
references is strongly recommended. Authors' names should be complete
--- use full first names (``Donald E. Knuth'') not initials
(``D. E. Knuth'') --- and the salient identifying features of a
reference should be included: title, year, volume, number, pages,
article DOI, etc.

The bibliography is included in your source document with these two
commands, placed just before the \verb|\end{document}| command:
\begin{verbatim}
  \bibliographystyle{ACM-Reference-Format}
  \bibliography{bibfile}
\end{verbatim}
where ``\verb|bibfile|'' is the name, without the ``\verb|.bib|''
suffix, of the \BibTeX\ file.

\section{Acknowledgments}

Identification of funding sources and other support, and thanks to
individuals and groups that assisted in the research and the
preparation of the work should be included in an acknowledgment
section, which is placed just before the reference section in your
document.

This section has a special environment:
\begin{verbatim}
  \begin{acks}
  ...
  \end{acks}
\end{verbatim}
so that the information contained therein can be more easily collected
during the article metadata extraction phase, and to ensure
consistency in the spelling of the section heading.

Authors should not prepare this section as a numbered or unnumbered {\verb|\section|}; please use the ``{\verb|acks|}'' environment.

%% The acknowledgments section is defined using the "acks" environment
%% (and NOT an unnumbered section). This ensures the proper
%% identification of the section in the article metadata, and the
%% consistent spelling of the heading.
\begin{acks}
To Robert, for the bagels and explaining CMYK and color spaces.
\end{acks}

%%
%% The next two lines define the bibliography style to be used, and
%% the bibliography file.
\bibliographystyle{ACM-Reference-Format}
\bibliography{references}

\end{document}
\endinput
%%
%% End of file `sample-acmsmall.tex'.

### 2022-02-03 src/Common.hs

@@ -1,45 +1,64 @@
 
 module Common (
     module Common
-  , FlatParse.Stateful.Span(..)
+  , B.ByteString
   , FlatParse.Stateful.Pos(..)
-  , module Data.Coerce
+  , FlatParse.Stateful.Span(..)
+  , unpackUTF8
+  , packUTF8
+  , coerce
   , module Control.Monad
   , module Data.Kind
   , module Lens.Micro
   ) where
 
-import Data.Kind
-import Control.Monad
-import Data.Foldable
-import GHC.Exts
 import qualified Data.ByteString as B
+import Control.Monad
+import Data.Bits
+import Data.Coerce
+import Data.Foldable
+import Data.Kind
+import Data.List
+import FlatParse.Stateful
+import GHC.Exts
 import Lens.Micro
 
-import Data.Bits
-import Data.Hashable
-import FNV164
-import FlatParse.Stateful
-import Data.Coerce
--- import GHC.Stack
+#ifdef DEBUG
+import GHC.Stack
+#endif
 
-import qualified Debug.Trace as Trace
-
+-- Debug printing, toggled by "debug" cabal flag
 --------------------------------------------------------------------------------
 
-trace :: String -> a -> a
-trace = Trace.trace
+-- define DEBUG
 
-traceShow :: Show a => a -> b -> b
-traceShow = Trace.traceShow
+#ifdef DEBUG
+type Dbg = HasCallStack
 
-traceM :: Applicative m => String -> m ()
-traceM = Trace.traceM
--- traceM _ = pure ()
+debug :: [String] -> IO ()
+debug strs = U.io $ putStrLn (intercalate " | " strs ++ " END")
 
-traceShowM :: (Show a, Applicative f) => a -> f ()
-traceShowM a = Trace.traceShowM a
--- traceShowM a = pure ()
+debugging :: IO () -> IO ()
+debugging act = act
+{-# inline debugging #-}
+#else
+type Dbg = () :: Constraint
+
+debug :: [String] -> IO ()
+debug strs = pure ()
+{-# inline debug #-}
+
+debugging :: IO () -> IO ()
+debugging _ = pure ()
+{-# inline debugging #-}
+#endif
+
+debug' :: [String] -> IO ()
+debug' strs = putStrLn (intercalate " | " strs ++ " END")
+
+debugging' :: IO () -> IO ()
+debugging' act = act
+{-# inline debugging' #-}
 
 --------------------------------------------------------------------------------
 
[…]

### 2022-02-03 src/Lexer.hs

@@ -1,5 +1,4 @@
 
-
 module Lexer where
 
 import FlatParse.Stateful hiding (Parser, runParser, string, char, cut, err)
@@ -26,7 +25,6 @@ data Error'
   | ExactIndent Int
   | IndentMore  Int
   | Imprecise [Expected]
-
   deriving Show
 
 data Error = Error !Pos !Error'

### 2022-02-03 staged/Main.hs

@@ -1,0 +1,182 @@
+
+module Main where
+
+-- import qualified Data.ByteString.Char8 as B
+import qualified Data.Array.Dynamic.L as D
+-- import qualified Data.HashMap.Strict  as M
+-- import qualified FlatParse.Stateful as FP
+-- import qualified Data.Set as S
+
+-- import qualified Syntax              as S
+-- import qualified Values              as V
+-- import qualified Presyntax           as P
+-- import qualified Evaluation          as Eval
+
+import Common
+-- import Cxt
+import ElabState
+-- import InCxt
+import Exceptions
+import Elaboration
+import Parser
+import Lexer
+import Pretty
+
+import System.Exit
+
+
+--------------------------------------------------------------------------------
+
+displayState :: IO ()
+displayState = do
+  let nl = putStrLn ""
+
+  nl
+  putStrLn "Metacontext"
+  putStrLn (replicate 60 '-')
+  nl
+  D.forIx metaCxt \i -> \case
+    Unsolved a -> do
+      putStrLn ("?" ++ show i ++ " : " ++ showVal1Top a)
+      nl
+    Solved t a -> do
+      putStrLn ("?" ++ show i ++ " : " ++ showVal1Top a)
+      putStrLn ("  = " ++ showVal1Top t)
+      nl
+
+  nl
+  putStrLn "Top environment"
+  putStrLn (replicate 60 '-')
+  nl
+  D.for top \case
+    TEDef0 a va t vt cv x _ -> do
+      putStrLn (show x ++ " : " ++ showVal1Top va)
+      putStrLn ("  := " ++ showTm0Top t)
+      nl
+      putStrLn "-- STAGED"
+      putStrLn (show x ++ " : " ++ showVal1Top va)
+      putStrLn ("  := " ++ showTm0Top' t)
+      nl
+    TEDef1 a va t vt x _ -> do
+      putStrLn (show x ++ " : " ++ showTm1Top a)
+      putStrLn ("  = " ++ showTm1Top t)
+      nl
+      putStrLn "-- NORMAL FORM"
+      putStrLn (show x ++ " : " ++ showTm1Top' a)
+      putStrLn ("  = " ++ showTm1Top' t)
+      nl
+    TETyCon{} -> do
[…]

### 2022-02-09 paper/paper.tex
stuff

@@
-%%
-%% This is file `sample-acmsmall.tex',
-%% generated with the docstrip utility.
-%%
-%% The original source files were:
-%%
-%% samples.dtx  (with options: `acmsmall')
-%%
-%% IMPORTANT NOTICE:
-%%
-%% For the copyright see the source file.
-%%
-%% Any modified versions of this file must be renamed
-%% with new filenames distinct from sample-acmsmall.tex.
-%%
-%% For distribution of the original source see the terms
-%% for copying and modification in the file samples.dtx.
-%%
-%% This generated file may be distributed as long as the
-%% original source files, as listed above, are part of the
-%% same distribution. (The sources need not necessarily be
-%% in the same archive or directory.)
-%%
-%%
-%% Commands for TeXCount
-%TC:macro \cite [option:text,text]
-%TC:macro \citep [option:text,text]
-%TC:macro \citet [option:text,text]
-%TC:envir table 0 1
-%TC:envir table* 0 1
-%TC:envir tabular [ignore] word
-%TC:envir displaymath 0 word
-%TC:envir math 0 word
-%TC:envir comment 0 0
-%%
-%%
-%% The first command in your LaTeX source must be the \documentclass command.
+%% build: latexmk -pdf -pvc paper.tex
+
 \documentclass[acmsmall]{acmart}
+\raggedbottom
 
-%%
 %% \BibTeX command to typeset BibTeX logo in the docs
@@
 
-%%
 %% These commands are for a JOURNAL article.
@@
 
+\newcommand{\Lift}{{\Uparrow}}
+\newcommand{\spl}{{\sim}}
+\newcommand{\qut}[1]{\langle #1\rangle}
+
+\renewcommand{\U}{\msf{U}}
+\newcommand{\Con}{\msf{Con}}
+\newcommand{\Sub}{\msf{Sub}}
+\newcommand{\Ty}{\msf{Ty}}
+\newcommand{\Tm}{\msf{Tm}}
+
+\newcommand{\Bool}{\msf{Bool}}
+\newcommand{\true}{\msf{true}}
+\newcommand{\false}{\msf{false}}
+\newcommand{\List}{\msf{List}}
+\newcommand{\nil}{\msf{nil}}
+\newcommand{\cons}{\msf{cons}}
+\newcommand{\Nat}{\msf{Nat}}
+\newcommand{\zero}{\msf{zero}}
+\newcommand{\suc}{\msf{suc}}
+\renewcommand{\tt}{\msf{tt}}
+
+\theoremstyle{remark}
+\newtheorem{notation}{Notation}
+
 %% --------------------------------------------------------------------------------
@@
 
-%%
-%% The "title" command has an optional parameter,
-%% allowing the author to define a "short title" to be used in page headers.
-\title{Using Two-Level Type Theory for Staged Compilation}
+\title{Staged Compilation With Two-Level Type Theory}
 
-%%
 %% The "author" command and its associated commands are used to define
@@
 
-%%
-%% By default, the full list of authors will be used in the page
-%% headers. Often, this list is too long, and will overlap
-%% other information printed in the page headers. This command allows
-%% the author to define a more concise list
-%% of authors' names for this purpose.
-%% \renewcommand{\shortauthors}{Trovato and Tobin, et al.}
-
-%%
-%% The abstract is a short summary of the work to be presented in the
-%% article.
 \begin{abstract}
-  A clear and well-documented \LaTeX\ document is presented as an
-  article formatted for publication by ACM in a conference proceedings
-  or journal publication. Based on the ``acmart'' document class, this
-  article presents and explains many of the common variations, as well
-  as many of the formatting elements an author may use in the
-  preparation of the documentation of their work.
+  The aim of staged compilation is to enable metaprogramming in a way such that
+  we have guarantees about the well-formedness of code output, and we can also
+  mix together object-level and meta-level code in a concise and convenient
+  manner. In this work, we observe that two-level type theory (2LTT), a system
+  originally devised for the purpose of synthetic homotopy theory, also serves
+  as a system for staged compilation dependent types. 2LTT has numerous good
+  properties for this use case: it has a concise specification, well-developed
+  algebraic and categorical model theory, and it supports a wide range of
+  language features both at the object and the meta level. First, we give an
+  overview of 2LTT's features and applications in staging. Then, we present a
+  staging algorithm and provide a proof of correctness. Our algorithm is
+  ``staging-by-evaluation'', analogously to the technique of
+  normalization-by-evaluation, in that staging is given by the evaluation of
+  2LTT syntax in a semantic domain. Staging together with its correctness
+  constitutes a proof of strong conservativity of 2LLT over the object
+  theory. To our knowledge, this is the first system for staged compilation which
+  supports full dependent types and unrestricted staging for types.
 \end{abstract}
 
-%%
-%% The code below is generated by the tool at http://dl.acm.org/ccs.cfm.
-%% Please copy and paste the code instead of the example below.
-%%
 \begin{CCSXML}
@@
 
-%%
-%% Keywords. The author(s) should pick words that accurately describe
-%% the work being presented. Separate the keywords with commas.
 \keywords{type theory, two-level type theory, staged compilation}
-
-%%
-%% This command processes the author and affiliation and title
-%% information and builds the first part of the formatted document.
 \maketitle
@@
 \section{Introduction}
-ACM's consolidated article template, introduced in 2017, provides a
-consistent \LaTeX\ style for use across ACM publications, and
-incorporates accessibility and metadata-extraction functionality
-necessary for future Digital Library endeavors. Numerous ACM and
-SIG-specific \LaTeX\ templates have been examined, and their unique
-features incorporated into this single new template.
 
-If you are new to publishing with ACM, this document is a valuable
-guide to the process of preparing your work for publication. If you
-have published with ACM before, this document provides insight and
-instruction into more recent changes to the article template.
+The purpose of staged compilation is to write code-generating programs in a
+safe, ergonomic and expressive way. It is always possible to do ad-hoc code
+generation, by simply manipulating strings or syntax trees in a sufficiently
+expressive programming language. However, these approaches tend to suffer from
+verbosity, non-reusability and lack of safety. In staged compilation, there are
+certain \emph{restrictions} on which metaprograms are expressible. Usually,
+staged systems enforce typing disciple, prohibit arbitrary manipulation of
+object-level scopes, and often they also prohibit accessing the internal
+structure of object expressions. On the other hand, we get \emph{guarantees}
+about the well-scoping or well-typing of the code output, and we are also able
+to use concise syntax for embedding object-level code.
 
-The ``\verb|acmart|'' document class can be used to prepare articles
-for any ACM publication --- conference or journal, and for any stage
-of publication, from review to final ``camera-ready'' copy, to the
-author's own version, with {\itshape very} few changes to the source.
+\emph{Two-level type theory}, or 2LTT in short, was described by Annekov,
+Capriotti, Kraus and Sattler \cite{twolevel}, building on ideas from Vladimir
+Voevodsky \cite{hts}. The motivation was to allow convenient metatheoretical
+reasoning about a certain mathematical language (homotopy type theory), and to
+enable concise and modular ways to extend the language with axioms.
 
-\section{Template Overview}
-As noted in the introduction, the ``\verb|acmart|'' document class can
-be used to prepare many different kinds of documentation --- a
-double-blind initial submission of a full-length technical paper, a
-two-page SIGGRAPH Emerging Technologies abstract, a ``camera-ready''
-journal article, a SIGCHI Extended Abstract, and more --- all by
-selecting the appropriate {\itshape template style} and {\itshape
-  template parameters}.
+It turns out that metamathematical convenience closely corresponds to to
+metaprogramming convenience: 2LTT can be directly and effectively employed
+in staged compilation. Moreover, semantic ideas underlying 2LTT are also
+directly applicable to the theory of staging.
 
-This document will explain the major features of the document
-class. For further information, the {\itshape \LaTeX\ User's Guide} is
-available from
-\url{https://www.acm.org/publications/proceedings-template}.
+\subsection{Contributions}
 
-\subsection{Template Styles}
-
-The primary parameter given to the ``\verb|acmart|'' document class is
-the {\itshape template style} which corresponds to the kind of publication
-or SIG publishing the work. This parameter is enclosed in square
-brackets and is a part of the {\verb|documentclass|} command:
-\begin{verbatim}
-  \documentclass[STYLE]{acmart}
-\end{verbatim}
-
-Journals use one of three template styles. All but three ACM journals
-use the {\verb|acmsmall|} template style:
 \begin{itemize}
-\item {\texttt{acmsmall}}: The default journal template style.
-\item {\texttt{acmlarge}}: Used by JOCCH and TAP.
-\item {\texttt{acmtog}}: Used by TOG.
+  \item In \ref{TODO} we present an informal syntax of two-level type theory, a
+    dependent type theory with staging features. We look at basic use-cases
+    involving inlining control, partial evaluation and fusion optimizations. We
+    also describe several feature variations, enabling applications in
+    monomorphization and memory layout control.
+  \item In \ref{TODO}, following \cite{twolevel}, we present a formal syntax of
+    2LTT and the object theory (the target theory of code generation). We recall
+    the standard presheaf model of 2LTT, which lies over the syntactic category
+    of the object theory. We show that the evaluation of 2LTT syntax in the presheaf
+    model yields a staging algorithm.
+  \item In \ref{TODO} we show correctness of staging, consisting of
+    \begin{itemize}
+    \item \emph{Stability:} staging the output of staging has no action.
+    \item \emph{Soundness:} the output of staging is convertible to the input.
+    \item \emph{Completeness:} convertible programs produce convertible staging outputs.
+    \end{itemize}
+    Staging together with its correctness can be viewed as a \emph{strong
+    conservativity} theorem of 2LTT over the object theory. Intuitively, this
+    means that the possible object-level constructions in 2LTT are in bijection
+    with the constructions in the object theory, and staging witnesses that
+    meta-level constructions can be always computed away. This improves on the
+    weak notion of conservativity shown in \cite{capriotti2017models} and
+    \cite{twolevel}.
+  \item To our knowledge, this is the first description of a language which
+    supports staging in the presence of full-blown dependent types, with
+    universes and large elimination. Moreover, we allow unrestricted staging
+    for types, so that types can be computed by metaprograms at compile time.
 \end{itemize}
 
-The majority of conference proceedings documentation will use the {\verb|acmconf|} template style.
+\section{A Tour of Two-Level Type Theory}
+
+In this section, we provide a short overview of 2LTT. We work in the informal
+syntax of a dependently typed language which resembles Agda \cite{TODO}. We focus
+on examples and informal explanations here; the formal details will be presented
+in Section \cite{TODO}.
+
+\begin{notation}
+We use the following notations throughout the paper. $(x : A) \to B$ denotes a
+dependent function type, where $x$ may occur in $B$. We use $\lambda\,x.\,t$ for
+abstraction. A $\Sigma$-type is written as $(x : A) \times B$, with pairing as
+$(t,\,u)$, and we may use pattern matching notation on pairs, e.g.\ as in
+$\lambda\,(x,\,y).\,t$. The unit type is $\top$ with element $\tt$. We will also
+use Agda-style notation for implicit arguments, where $t : \{x : A\} \to B$
+implies that the first argument to $t$ is inferred by default, and we can
+override this by writing a $t \{u\}$ application. We may also implicitly
+quantify over arguments (in the style of Idris and Haskell), for example
+when declaring $\mit{id} : A \to A$ with the assumption that $A$ is universally
+quantified.
+\end{notation}
+
+\subsection{Rules of 2LTT}
+
+\paragraph{Universes} We have universes $\U_{i,j}$, where $i \in \{0,1\}$, and $j \in \mbb{N}$.
+The $i$ index denotes stages, where $0$ is the runtime (object-level) stage, and
+$1$ is the compile time (meta-level) stage. The $j$ index denotes universe sizes
+in the usual sense of type theory. We assume Russell-style universes, with
+$\U_{i,j} : \U_{i, j+1}$. However, for the sake of brevity we will usually omit
+the $j$ indices in this section, and simply write $\U_0$ or $\U_1$.
 \begin{itemize}
-\item {\texttt{acmconf}}: The default proceedings template style.
-\item{\texttt{sigchi}}: Used for SIGCHI conference articles.
-\item{\texttt{sigchi-a}}: Used for SIGCHI ``Extended Abstract'' articles.
-\item{\texttt{sigplan}}: Used for SIGPLAN conference articles.
+\item $\U_0$ can be viewed as the \emph{universe of object-level or runtime types}.
+    Each closed type $A : \U_0$ can be staged to an actual type in the object language
+    (the language of the staging output).
+  \item $\U_1$ can be viewed as the \emph{universe of meta-level or static types}. If we
+    have $A : \U_1$, then $A$ is guaranteed to be only present at compile time,
+    and will be staged away. Elements $a : A$ are likewise computed away.
 \end{itemize}
 
-\subsection{Template Parameters}
+\paragraph{Type formers} $\U_0$ and $\U_1$ may be closed under arbitrary type formers,
+such as functions, $\Sigma$-types, identity types or inductive types in general.
+However, all constructors and eliminators in type formers must stay at the same
+stage. For example:
+\begin{itemize}
+  \item Function domain and codomain types must be at the same stage.
+  \item If we have $\Nat_0 : \U_0$ for the runtime type of natural numbers,
+        we can only map from it to a type in $\U_0$ by recursion or induction.
+\end{itemize}
+It is not required that we have the \emph{same} type formers at both stages. As we will
+see in Section \cite{TODO}, simpler object languages have the advantage that they are
+easier to process during downstream compilation.
 
-In addition to specifying the {\itshape template style} to be used in
-formatting your work, there are a number of {\itshape template parameters}
-which modify some part of the applied template style. A complete list
-of these parameters can be found in the {\itshape \LaTeX\ User's Guide.}
+\paragraph{Moving between stages} At this point, our system is rather limited, since
+there is no interaction between the stages. We add such interaction via the follow
+three operations.
+\begin{itemize}
+\item \emph{Lifting:} for $A : \U_0$, we have $\Lift A : \U_1$.  From the
+  staging point of view, $\Lift A$ is the type of metaprograms which compute
+  to runtime expressions of type $A$.
+\item \emph{Quoting:} for $A : \U_0$ and $t : A$, we have $\qut{t} :\,\Lift A$.
+  A quoted term $\qut{t}$ represents the metaprogram which immediately computes
+  to $t$.
+\item \emph{Splicing:} for $A : \U_0$ and $t :\,\Lift A$, we have $\spl t : A$.
+  During staging, the metaprogram in the splice is executed, and the resulting
+  expression is inserted into the output.
+\item Quoting and splicing are definitional inverses, i.e.\ we have $\spl\qut{t} = t$ and
+  $\qut{\spl t}=t$ as definitional equalities.
+\end{itemize}
+Note that none of these three operations can be expressed as functions, since
+function types cannot cross between stages.
 
-Frequently-used parameters, or combinations of parameters, include:
-\begin{itemize}
-\item {\texttt{anonymous,review}}: Suitable for a ``double-blind''
-  conference submission. Anonymizes the work and includes line
-  numbers. Use with the \texttt{\acmSubmissionID} command to print the
-  submission's unique ID on each page of the work.
-\item{\texttt{authorversion}}: Produces a version of the work suitable
-  for posting by the author.
-\item{\texttt{screen}}: Produces colored hyperlinks.
-\end{itemize}
+Informally, if we have a closed program $t : A$ with $A : \U_0$, \emph{staging}
+means computing all metaprograms and recursively replacing all splices in $t$ and $A$
+with the resulting runtime expressions. The rules of 2LTT ensure that this is possible,
+and we always get a splice-free runtime program after staging.
 
-This document uses the following string as the first command in the
-source file:
-\begin{verbatim}
-\documentclass[acmsmall]{acmart}
-\end{verbatim}
+\begin{notation} We may disambiguate type formers at different stages by using $0$ or $1$
+subscripts. For example, $\Nat_1 : \U_1$ is distinguished from $\Nat_0 : \U_0$,
+and likewise we may write $\zero_0 : \Nat_0$ and so on. For function and
+$\Sigma$ types, the stage is usually easy to infer, so we do not annotate
+them. For example, the type $\Nat_0 \to \Nat_0$ must be at the runtime stage,
+since the domain and codomain types are at that stage, and we know that the
+function type former stays within a single stage. We may also omit stage annotations
+from $\lambda$ and pairing.
+\end{notation}
 
-\section{Modifications}
+\subsection{Basic Examples}
 
-Modifying the template --- including but not limited to: adjusting
-margins, typeface sizes, line spacing, paragraph and list definitions,
-and the use of the \verb|\vspace| command to manually adjust the
-vertical spacing between elements of your work --- is not allowed.
+In 2LTT, we may have several different polymorphic identity functions. First, we
+consider the usual identity functions at each stage:
+\begin{alignat*}{3}
+  & \mit{id}_0 : (A : \U_0) \to A \to A\hspace{2em} && \mit{id}_1 : (A : \U_1) \to A \to A\\
+  & \mit{id}_0 := \lambda\,A\,x.x       && \mit{id}_1 := \lambda\,A\,x.x
+\end{alignat*}
+An $\mit{id}_0$ application will simply appear in staging output as it is. In
+contrast, $\mit{id}_1$ can be used as a compile-time evaluated function, because
+the staging operations allow us to freely apply $\mit{id}_1$ to runtime
+arguments. For example, $\mit{id}_1\,(\Lift\,\Bool_0)\,\qut{\true_0}$ has type
+$\Lift \Bool_0$, therefore $\spl(\mit{id}_1\,(\Lift\,\Bool_0)\,\qut{\true_0})$
+has type $\Bool_0$. We can stage this expression as follows:
+\[
+\spl(\mit{id}_1\,(\Lift\,\Bool_0)\,\qut{\true_0}) = \spl\qut{\true_0} = \true_0
+\]
+There is another identity function, which computes at compile time, but which
+can be only used on runtime arguments:
+\begin{alignat*}{3}
+  & \mit{id_\Lift} : (A : \Lift\U_0) \to \Lift (\spl A) \to \Lift (\spl A)\\
+  & \mit{id_\Lift} := \lambda\,A\,x.x
+\end{alignat*}
+Note that since $A : \Lift\U_0$, we have $\spl A : \U_0$, hence $\Lift (\spl A)
+: \U_1$.  Also, $\Lift\U_0 : \U_1$, so all function domain and codomain types in
+the type of $\mit{id_\Lift}$ are at the same stage. Now, we may write
+$\spl(\mit{id_\Lift}\,\qut{\Bool_0}\,\qut{\true_0})$ for a term which is staged
+to $\true_0$. In this specific case $\mit{id_\Lift}$ has no practical advantage
+over $\mit{id}_1$, but in some cases we really have to quantify over
+$\Lift\U_0$. This brings us the next example.
 
-{\bfseries Your document will be returned to you for revision if
-  modifications are discovered.}
+Assume $\List_0 : \U_0 \to \U_0$ with $\nil_0 : (A : \U_0) \to \List_0\,A$,
+$\cons_0 : (A : \U_0) \to A \to \List_0\,A$ and $\msf{foldr}_0 : (A\,B : \U_0)
+\to (A \to B \to B) \to B \to \List_0\,A \to B$. We define a map function which
+``inlines'' its function argument:
+\begin{alignat*}{3}
+  & \mit{map} : (A\,B : \Lift\U_0) \to ((\Lift(\spl A) \to \Lift(\spl B))
+      \to \Lift(\List_0\,(\spl A)) \to \Lift(\List_0\,(\spl B))\\
+  & \mit{map} := \lambda\,A\,B\,f\,\mit{as}.\,
+      \qut{\msf{foldr}_0\,
+        (\spl A)\,(\spl B)\,
+        (\lambda\,a\,\mit{bs}.\, \cons_0\,(\spl B)\,(\spl(f\,\qut{a}))\,\mit{bs})\,
+        (\nil_0\,(\spl B))\,
+        (\spl as)
+        }
+\end{alignat*}
 
-\section{Typefaces}
+%% we shall work in the syntax of a dependently typed language
+%% which resembles Agda.
 
-The ``\verb|acmart|'' document class requires the use of the
-``Libertine'' typeface family. Your \TeX\ installation should include
-this set of packages. Please do not substitute other typefaces. The
-``\verb|lmodern|'' and ``\verb|ltimes|'' packages should not be used,
-as they will override the built-in typeface families.
+\subsection{Informal Syntax}
 
-\section{Title Information}
+\section{Variations \& Applications}
 
-The title of your work should use capital letters appropriately -
-\url{https://capitalizemytitle.com/} has useful rules for
-capitalization. Use the {\verb|title|} command to define the title of
-your work. If your work has a subtitle, define it with the
-{\verb|subtitle|} command.  Do not insert line breaks in your title.
+\subsection{Fusion}
 
-If your title is lengthy, you must define a short version to be used
-in the page headers, to prevent overlapping text. The \verb|title|
-command has a ``short title'' parameter:
-\begin{verbatim}
-  \title[short title]{full title}
-\end{verbatim}
+\subsection{Monomorphization}
 
-\section{Authors and Affiliations}
+\subsection{Levity Polymorphism}
 
-Each author must be defined separately for accurate metadata
-identification.  As an exception, multiple authors may share one
-affiliation. Authors' names should not be abbreviated; use full first
-names wherever possible. Include authors' e-mail addresses whenever
-possible.
+\subsection{Typed Closures}
 
-Grouping authors' names or e-mail addresses, or providing an ``e-mail
-alias,'' as shown below, is not acceptable:
-\begin{verbatim}
-  \author{Brooke Aster, David Mehldau}
-  \email{dave,judy,steve@university.edu}
-  \email{firstname.lastname@phillips.org}
-\end{verbatim}
+\section{Formal Syntax \& Models}
 
-The \verb|authornote| and \verb|authornotemark| commands allow a note
-to apply to multiple authors --- for example, if the first two authors
-of an article contributed equally to the work.
+\subsection{Metatheory}
+\subsection{Models}
 
-If your author list is lengthy, you must define a shortened version of
-the list of authors to be used in the page headers, to prevent
-overlapping text. The following command should be placed just after
-the last \verb|\author{}| definition:
-\begin{verbatim}
-  \renewcommand{\shortauthors}{McCartney, et al.}
-\end{verbatim}
-Omitting this command will force the use of a concatenated list of all
-of the authors' names, which may result in overlapping text in the
-page headers.
+\section{Staging by Evaluation}
 
-The article template's documentation, available at
-\url{https://www.acm.org/publications/proceedings-template}, has a
-complete explanation of these commands and tips for their effective
-use.
+\subsection{Presheaf Model of 2LTT}
 
-Note that authors' addresses are mandatory for journal articles.
+\subsection{Closed staging}
+\subsection{Yoneda, representability, intensional analysis}
 
-\section{Rights Information}
+\section{Stability}
 
-Authors of any work published by ACM will need to complete a rights
-form. Depending on the kind of work, and the rights management choice
-made by the author, this may be copyright transfer, permission,
-license, or an OA (open access) agreement.
+\subsection{Open staging}
 
-Regardless of the rights management choice, the author will receive a
-copy of the completed rights form once it has been submitted. This
-form contains \LaTeX\ commands that must be copied into the source
-document. When the document source is compiled, these commands and
-their parameters add formatted text to several areas of the final
-document:
-\begin{itemize}
-\item the ``ACM Reference Format'' text on the first page.
-\item the ``rights management'' text on the first page.
-\item the conference information in the page header(s).
-\end{itemize}
+\section{Soundness}
 
-Rights information is unique to the work; if you are preparing several
-works for an event, make sure to use the correct set of commands with
-each of the works.
+\subsection{Internal Language \& Features}
+\subsection{The Logical Relation}
+\subsection{Externalization \& Soundness}
 
-The ACM Reference Format text is required for all articles over one
-page in length, and is optional for one-page articles (abstracts).
+\section{Related Work}
 
-\section{CCS Concepts and User-Defined Keywords}
+Igarashi, Kiselyov et al, MetaML, MetaOCaml, Carette, TH, Scala ppl (?)
+PE lit ?
+2LTT, Voevodsky,
 
-Two elements of the ``acmart'' document class provide powerful
-taxonomic tools for you to help readers find your work in an online
-search.
+\section{Future Work \& Conclusions}
 
-The ACM Computing Classification System ---
-\url{https://www.acm.org/publications/class-2012} --- is a set of
-classifiers and concepts that describe the computing
-discipline. Authors can select entries from this classification
-system, via \url{https://dl.acm.org/ccs/ccs.cfm}, and generate the
-commands to be included in the \LaTeX\ source.
-
-User-defined keywords are a comma-separated list of words and phrases
-of the authors' choosing, providing a more flexible way of describing
-the research being presented.
-
-CCS concepts and user-defined keywords are required for for all
-articles over two pages in length, and are optional for one- and
-two-page articles (or abstracts).
-
-\section{Sectioning Commands}
-
-Your work should use standard \LaTeX\ sectioning commands:
-\verb|section|, \verb|subsection|, \verb|subsubsection|, and
-\verb|paragraph|. They should be numbered; do not remove the numbering
-from the commands.
-
-Simulating a sectioning command by setting the first word or words of
-a paragraph in boldface or italicized text is {\bfseries not allowed.}
-
-\section{Tables}
-
-The ``\verb|acmart|'' document class includes the ``\verb|booktabs|''
-package --- \url{https://ctan.org/pkg/booktabs} --- for preparing
-high-quality tables.
-
-Table captions are placed {\itshape above} the table.
-
-Because tables cannot be split across pages, the best placement for
-them is typically the top of the page nearest their initial cite.  To
-ensure this proper ``floating'' placement of tables, use the
-environment \textbf{table} to enclose the table's contents and the
-table caption.  The contents of the table itself must go in the
-\textbf{tabular} environment, to be aligned properly in rows and
-columns, with the desired horizontal and vertical rules.  Again,
-detailed instructions on \textbf{tabular} material are found in the
-\textit{\LaTeX\ User's Guide}.
-
-Immediately following this sentence is the point at which
-Table~\ref{tab:freq} is included in the input file; compare the
-placement of the table here with the table in the printed output of
-this document.
-
-\begin{table}
-  \caption{Frequency of Special Characters}
-  \label{tab:freq}
-  \begin{tabular}{ccl}
-    \toprule
-    Non-English or Math&Frequency&Comments\\
-    \midrule
-    \O & 1 in 1,000& For Swedish names\\
-    $\pi$ & 1 in 5& Common in math\\
-    \$ & 4 in 5 & Used in business\\
-    $\Psi^2_1$ & 1 in 40,000& Unexplained usage\\
-  \bottomrule
-\end{tabular}
-\end{table}
-
-To set a wider table, which takes up the whole width of the page's
-live area, use the environment \textbf{table*} to enclose the table's
-contents and the table caption.  As with a single-column table, this
-wide table will ``float'' to a location deemed more
-desirable. Immediately following this sentence is the point at which
-Table~\ref{tab:commands} is included in the input file; again, it is
-instructive to compare the placement of the table here with the table
-in the printed output of this document.
-
-\begin{table*}
-  \caption{Some Typical Commands}
-  \label{tab:commands}
-  \begin{tabular}{ccl}
-    \toprule
-    Command &A Number & Comments\\
-    \midrule
-    \texttt{{\char'134}author} & 100& Author \\
-    \texttt{{\char'134}table}& 300 & For tables\\
-    \texttt{{\char'134}table*}& 400& For wider tables\\
-    \bottomrule
-  \end{tabular}
-\end{table*}
-
-Always use midrule to separate table header rows from data rows, and
-use it only for this purpose. This enables assistive technologies to
-recognise table headers and support their users in navigating tables
-more easily.
-
-\section{Math Equations}
-You may want to display math equations in three distinct styles:
-inline, numbered or non-numbered display.  Each of the three are
-discussed in the next sections.
-
-\subsection{Inline (In-text) Equations}
-
-\begin{math}
-  \lim_{n\rightarrow \infty}x=0
-\end{math},
-set here in in-line math style, looks slightly different when
-set in display style.  (See next section).
-
-\subsection{Display Equations}
-A numbered display equation---one set off by vertical space from the
-text and centered horizontally---is produced by the \textbf{equation}
-environment. An unnumbered display equation is produced by the
-\textbf{displaymath} environment.
-
-Again, in either environment, you can use any of the symbols and
-structures available in \LaTeX\@; this section will just give a couple
-of examples of display equations in context.  First, consider the
-equation, shown as an inline equation above:
-\begin{equation}
-  \lim_{n\rightarrow \infty}x=0
-\end{equation}
-Notice how it is formatted somewhat differently in
-the \textbf{displaymath}
-environment.  Now, we'll enter an unnumbered equation:
-\begin{displaymath}
-  \sum_{i=0}^{\infty} x + 1
-\end{displaymath}
-and follow it with another numbered equation:
-\begin{equation}
-  \sum_{i=0}^{\infty}x_i=\int_{0}^{\pi+2} f
-\end{equation}
-just to demonstrate \LaTeX's able handling of numbering.
-
-\section{Figures}
-
-Your figures should contain a caption which describes the figure to
-the reader.
-
-Figure captions are placed {\itshape below} the figure.
-
-Every figure should also have a figure description unless it is purely
-decorative. These descriptions convey what’s in the image to someone
-who cannot see it. They are also used by search engine crawlers for
-indexing images, and when images cannot be loaded.
-
-A figure description must be unformatted plain text less than 2000
-characters long (including spaces).  {\bfseries Figure descriptions
-  should not repeat the figure caption – their purpose is to capture
-  important information that is not already provided in the caption or
-  the main text of the paper.} For figures that convey important and
-complex new information, a short text description may not be
-adequate. More complex alternative descriptions can be placed in an
-appendix and referenced in a short figure description. For example,
-provide a data table capturing the information in a bar chart, or a
-structured list representing a graph.  For additional information
-regarding how best to write figure descriptions and why doing this is
-so important, please see
-\url{https://www.acm.org/publications/taps/describing-figures/}.
-
-\subsection{The ``Teaser Figure''}
-
-A ``teaser figure'' is an image, or set of images in one figure, that
-are placed after all author and affiliation information, and before
-the body of the article, spanning the page. If you wish to have such a
-figure in your article, place the command immediately before the
-\verb|\maketitle| command:
-\begin{verbatim}
-  \begin{teaserfigure}
-    \includegraphics[width=\textwidth]{sampleteaser}
-    \caption{figure caption}
-    \Description{figure description}
-  \end{teaserfigure}
-\end{verbatim}
-
-\section{Citations and Bibliographies}
-
-The use of \BibTeX\ for the preparation and formatting of one's
-references is strongly recommended. Authors' names should be complete
---- use full first names (``Donald E. Knuth'') not initials
-(``D. E. Knuth'') --- and the salient identifying features of a
-reference should be included: title, year, volume, number, pages,
-article DOI, etc.
-
-The bibliography is included in your source document with these two
-commands, placed just before the \verb|\end{document}| command:
-\begin{verbatim}
-  \bibliographystyle{ACM-Reference-Format}
-  \bibliography{bibfile}
-\end{verbatim}
-where ``\verb|bibfile|'' is the name, without the ``\verb|.bib|''
-suffix, of the \BibTeX\ file.
-
-\section{Acknowledgments}
-
-Identification of funding sources and other support, and thanks to
-individuals and groups that assisted in the research and the
-preparation of the work should be included in an acknowledgment
-section, which is placed just before the reference section in your
-document.
-
-This section has a special environment:
-\begin{verbatim}
-  \begin{acks}
-  ...
-  \end{acks}
-\end{verbatim}
-so that the information contained therein can be more easily collected
-during the article metadata extraction phase, and to ensure
-consistency in the spelling of the section heading.
-
-Authors should not prepare this section as a numbered or unnumbered {\verb|\section|}; please use the ``{\verb|acks|}'' environment.
-
-%% The acknowledgments section is defined using the "acks" environment
-%% (and NOT an unnumbered section). This ensures the proper
-%% identification of the section in the article metadata, and the
-%% consistent spelling of the heading.
-\begin{acks}
-To Robert, for the bagels and explaining CMYK and color spaces.
-\end{acks}
-
-%%
-%% The next two lines define the bibliography style to be used, and
-%% the bibliography file.
 \bibliographystyle{ACM-Reference-Format}
@@
 \endinput
-%%
-%% End of file `sample-acmsmall.tex'.

### 2022-02-09 paper/paper.tex
stuff

@@
 \renewcommand{\tt}{\msf{tt}}
+\newcommand{\fst}{\msf{fst}}
+\newcommand{\snd}{\msf{snd}}
+\newcommand{\mylet}{\msf{let}}
+\newcommand{\emptycon}{\scaleobj{.75}\bullet}
+
+\newcommand{\blank}{{\mathord{\hspace{1pt}\text{--}\hspace{1pt}}}}
 
@@
   expression is inserted into the output.
+  \begin{notation} Splicing binds stronger than any operation, including function
+    application. For instance, $\spl f\,x$ is parsed as $(\spl f)\,x$.
+  \end{notation}
+
 \item Quoting and splicing are definitional inverses, i.e.\ we have $\spl\qut{t} = t$ and
@@
 
+\paragraph{Remark}
+Why do we use the index $0$ for the runtime stage? The reason is that it is not
+difficult to generalize 2LTT to multi-level type theory, by allowing to lift
+types from $\U_i$ to $\U_{i+1}$. In the semantics, this can be modeled by
+having a 2LTT whose object theory is once again a 2LTT, and doing this in an
+iterated fashion. But there must be necessarily a bottom-most object theory;
+hence our stage indexing scheme. For now though, we leave the multi-level
+generalization to future work.
+
 \begin{notation} We may disambiguate type formers at different stages by using $0$ or $1$
@@
 
-\subsection{Basic Examples}
+\subsection{Staged Programming in 2LTT}
 
@@
 \begin{alignat*}{3}
-  & \mit{id_\Lift} : (A : \Lift\U_0) \to \Lift (\spl A) \to \Lift (\spl A)\\
+  & \mit{id_\Lift} : (A : \Lift\U_0) \to \Lift\,\spl A \to \Lift\,\spl A\\
   & \mit{id_\Lift} := \lambda\,A\,x.x
 \end{alignat*}
-Note that since $A : \Lift\U_0$, we have $\spl A : \U_0$, hence $\Lift (\spl A)
+Note that since $A : \Lift\U_0$, we have $\spl A : \U_0$, hence $\Lift\,\spl A
 : \U_1$.  Also, $\Lift\U_0 : \U_1$, so all function domain and codomain types in
@@
 \begin{alignat*}{3}
-  & \mit{map} : (A\,B : \Lift\U_0) \to ((\Lift(\spl A) \to \Lift(\spl B))
-      \to \Lift(\List_0\,(\spl A)) \to \Lift(\List_0\,(\spl B))\\
+  & \mit{map} : (A\,B : \Lift\U_0) \to (\Lift\,\spl A \to \Lift\,\spl B)
+      \to \Lift(\List_0\,\spl A) \to \Lift(\List_0\,\spl B)\\
   & \mit{map} := \lambda\,A\,B\,f\,\mit{as}.\,
       \qut{\msf{foldr}_0\,
-        (\spl A)\,(\spl B)\,
-        (\lambda\,a\,\mit{bs}.\, \cons_0\,(\spl B)\,(\spl(f\,\qut{a}))\,\mit{bs})\,
-        (\nil_0\,(\spl B))\,
-        (\spl as)
+        \spl A\,\spl B\,
+        (\lambda\,a\,\mit{bs}.\, \cons_0\,\spl B\,\spl(f\,\qut{a})\,\mit{bs})\,
+        (\nil_0\,\spl B)\,
+        \spl as
         }
 \end{alignat*}
+This $\mit{map}$ function works with quantification over $\Lift \U_0$ but not
+over $\U_1$, because $\List_0$ expects type parameters in $\U_0$, and there is
+no generic way to convert from $\U_1$ to $\U_0$. Now, assuming
+$\blank\!+_0\!\blank : \Nat_0 \to \Nat_0 \to \Nat_0$ and $\mit{ns} :
+\List_0\,\Nat_0$, we have the following staging behavior:
+\begin{alignat*}{3}
+  & &&\spl(\mit{map}\,\qut{\Nat_0}\,\qut{\Nat_0}\,(\lambda\,n.\,\qut{\spl n +_0 10})\,\qut{\mit{ns}})\\
+  & =\hspace{1em}&&\spl\qut{\msf{foldr}_0\,
+        \spl \qut{\Nat_0}\,\spl \qut{\Nat_0}\,
+        (\lambda\,a\,\mit{bs}.\, \cons_0\,\spl B\,\spl\qut{\spl\qut{a} +_0 10}\,\mit{bs})\,
+        (\nil_0\,\spl \qut{\Nat_0})\,
+        \spl\qut{\mit{ns}}}\\
+  & =\hspace{1em} &&\msf{foldr_0}\,\Nat_0\,\Nat_0\,\,(\lambda\,a\,bs.\,a +_0 10)\,(\nil_0\,\Nat_0)\,\mit{ns}
+\end{alignat*}
+
+By using meta-level functions and lifted types, we already have control over
+inlining. However, if we want to do more complicated meta-level computation, it
+is convenient to use recursion or induction on meta-level type formers. A
+classic example in staged compilation is the power function for natural numbers,
+which evaluates the exponent at compile time. We assume the iterator function
+$\msf{iter_1} : \{A : \U_1\} \to \Nat_1 \to (A \to A) \to A \to A$, and runtime
+multiplication as $\blank\!*_0\!\blank$.
+\begin{alignat*}{3}
+  &\mit{exp} : \Nat_1 \to \Lift \Nat_0 \to \Lift \Nat_0 \\
+  &\mit{exp} := \lambda\,x\,y.\,
+  \msf{iter}_1\,x\,(\lambda\,n.\,\qut{\spl y *_0 \spl n})\,\qut{1}
+\end{alignat*}
+Now, $\spl(\mit{exp}\,3\,\qut{n})$ stages to $n *_0 n *_0 n *_0 1$ by the computation rules
+of $\msf{iter_1}$ and the staging operations.
+
+We can also stage \emph{types}. Below, we use iteration to compute the type of
+vectors with static length, as a nested pair type.
+\begin{alignat*}{3}
+  &\mit{Vec} : \Nat_1 \to \Lift \U_0 \to \Lift \U_0\\
+  &\mit{Vec} := \lambda\,n\,A.\,\msf{iter}_1\,n\,(\lambda\,B.\,\qut{\spl A \times \spl B})\,\qut{\top_0}
+\end{alignat*}
+With this definition, $\spl(\mit{Vec}\,3\,\qut{\Nat_0})$ stages to $\Nat_0
+\times (\Nat_0 \times (\Nat_0 \times \top_0))$. Now, we can use \emph{induction}
+on $\Nat_1$ to implement a map function. For readability, we use an Agda-style
+pattern matching definition below (instead of the elimination principle).
+\begin{alignat*}{6}
+  &\hspace{-3em}\rlap{$\mit{map} : (n : \Nat_1) \to (\Lift\,\spl A \to \Lift\,\spl B) \to \Lift(\mit{Vec}\,n\,A) \to \Lift(\mit{Vec}\,n\,B)$}\\
+  &\hspace{-3em}\mit{map}\,&&\zero_1      && f\,&&\mit{as} := \qut{\tt_0}\\
+  &\hspace{-3em}\mit{map}\,&&(\suc_1\,n)\,&& f\,&&\mit{as} :=
+     \qut{(\spl(f\,\qut{\fst_0\,\spl\mit{as}}),\,\mit{map}\,n\,f\,\qut{\snd_0\,\spl \mit{as}})}
+\end{alignat*}
+This definition inlines the mapping function for each projected element of the
+vector. For instance, staging $\spl(\mit{map}\,2\,(\lambda\,n.\,\qut{\spl n +_0
+  10})\,\qut{\mit{ns}})$ yields $(\fst_0\,\mit{ns} +_0
+10,\,(\fst_0(\snd_0\,\mit{ns}) +_0 10,\,\tt_0))$. Sometimes, we do not want to
+duplicate the code of the mapping function. In such cases, we can use
+\emph{let-insertion} \cite{TODO}, a standard technique in staged compilation. If
+we bind a runtime expression to a runtime variable, and only use that variable
+in subsequent staging, only the variable itself can be duplicated, as an
+expression. One solution is to do an ad-hoc let-insertion:
+\begin{alignat*}{6}
+  &   && \mylet_0\,f := \lambda\,n.\, n +_0 10\,\,\msf{in}\,\,
+         \spl(\mit{map}\,2\,(\lambda\,n.\,\qut{f\,\spl n})\,\qut{\mit{ns}}) \\
+  & =\,\,&&  \mylet_0\,f := \lambda\,n.\, n +_0 10\,\,\msf{in}\,\,
+          (f\,(\fst_0\,\mit{ns}),\,(f\,(\fst_0(\snd_0\,\mit{ns})),\,\tt_0))
+\end{alignat*}
+ Alternatively, we can define $\mit{map}$ so that it performs let-insertion, and
+ we can switch between the two versions as needed.
+
+More generally, we are free to use dependent types at the meta-level, so we can
+reproduce more complicated staging examples. Any well-typed interpreter can be
+rephrased as a \emph{partial evaluator}, as long as we have sufficient type
+formers. For instance, we may write a partial evaluator for a simply typed
+lambda calculus. We sketch the implementation in the following. First, we
+inductively define contexts, types and terms:
+\begin{alignat*}{6}
+  & \Ty : \U_1  \hspace{2em} \Con : \U_1 \hspace{2em} \Tm : \Con \to \Ty \to \U_1
+\end{alignat*}
+Then we define the interpretation functions:
+\begin{alignat*}{6}
+  & \msf{EvalTy}  &&: \Ty \to \Lift \U_0 \\
+  & \msf{EvalCon} &&: \Con \to \U_1 \\
+  & \msf{EvalTm}  &&: \Tm\,\Gamma\,A \to \msf{EvalCon}\,\Gamma \to \Lift\,\spl(\msf{EvalTy}\,A)
+\end{alignat*}
+Types are necessarily computed to runtime types, since we can only stage to such types.
+Contexts are computed as follows:
+\begin{alignat*}{4}
+  &\msf{EvalCon}\,\,\msf{empty}                &&:= \top_1 \\
+  &\msf{EvalCon}\,\,(\msf{extend}\,\Gamma\,A) &&:= \msf{EvalCon}\,\Gamma \times (\Lift \spl(\msf{EvalTy}\,A))
+\end{alignat*}
+This is an nice example for the usage of \emph{partially static
+data}\cite{TODO}: semantic contexts are \emph{static} lists storing
+\emph{runtime} expressions. This allows us to completely eliminate environment
+lookups in the staging output: an embedded lambda expression is staged to the
+corresponding lambda expression in the runtime language. This is similar to the
+partial evaluator presented in Idris 1 \cite{TODO}. However, in contrast to
+2LTT, Idris 1 does not provide a formal guarantee that partial evaluation does
+not get stuck.
+
+\subsection{Properties of Lifting}
+
+preservation properties
+\\\\
+stage inference
+
+\subsection{Restricting the Object Language}
+
+monomorphization
+\\\\
+memory layout control
+\\\\
+typed closures
+
+\subsection{Fusion}
+2LTT is also a natural setting for a wide range of \emph{fusion
+optimizations}. We only describe here a simple example, \emph{foldr-build
+fusion} for lists, which is notably used in GHC Haskell \cite{TODO}. The idea is
+the following: we want to eliminate intermediate lists, e.g.\ fuse repeated
+mapping to a single traversal. For this, we exploit that there is an isomorphic
+\emph{meta-level} representation for runtime lists which supports more
+definitional equations. Starting from $\Lift (\List_0\,A)$, we can use
+Böhm-Berarducci encoding under the lifting to get
+\[ \Lift((L : \U_0) \to (A \to L \to L) \to L \to L) \]
+However, $\Lift$ preserves function types, so the above is definitionally
+isomorphic to
+\[ (L : \Lift\,U_0) \to (\Lift\,A \to \Lift\,\spl L \to \Lift\,\spl L)
+\to \Lift\,\spl L \to \Lift\,\spl L. \] Now, we may abbreviate the above type as
+$\msf{BBList}\,A$. We write fusible functions on $\msf{BBList}\,A$, and when we
+want to fuse a definition, we convert from $\Lift\,(\List_0\,A)$ to
+$\msf{BBList}\,A$, apply operations on that representation, then convert back in
+the end.
+
+Now, in 2LTT it is not provable that Böhm-Berarducci encoding is an isomorphism \cite{TODO},
+so it certainly does not hold that back-and forth conversion between the two
+representations is definitionally the identity function. In GHC Haskell, this
+equality is added as a \emph{rewrite rule}. This makes it possible to hide the
+back-and-forth conversions from library users, since inverse conversions will
+hopefully cancel out and cause no additional overhead. A similar mechanism could be
+used in 2LTT as well. However, such implicit conversions are not used in most
+languages which support fusion. For example, iterators in the Rust language
+\cite{TODO}, which use a form of stream fusion, require explicit conversion to
+iterators.
+
+We may compare GHC and 2LTT with respect to the implementation of fusion. In
+2LTT, fusion happens through meta-level $\beta$-reduction, which is formally
+guaranteed to go through, and which can be also computed very efficiently. In
+GHC, all of the necessary computation is realized by rewrite rules and
+general-purpose optimization passes. This is in practice very fragile and
+non-composable. It requires carefully tuning every individual collection of
+rewrite rules, and also ordering rule applications in specific ways, while
+keeping details of the compiler optimization passes in mind. And when fusion
+fails to go through, it is usually a performance \emph{pessimization}.
+
+\subsection{Properties of Lifting}
 

### 2022-02-13 paper/paper.tex
stuff

@@
 \documentclass[acmsmall]{acmart}
-\raggedbottom
+%% \raggedbottom
 
@@
 \usepackage{scalerel}
+\usepackage{bm}
 %% \usepackage{amssymb}
@@
 \newcommand{\bs}[1]{\boldsymbol{#1}}
+\newcommand{\wh}[1]{\widehat{#1}}
+\newcommand{\ext}{\triangleright}
+\newcommand{\Code}{\msf{Code}}
+\newcommand{\El}{\msf{El}}
+\newcommand{\lam}{\msf{lam}}
+\newcommand{\app}{\msf{app}}
+\newcommand{\NatElim}{\msf{NatElim}}
+\newcommand{\y}{\msf{y}}
 
@@
 \newcommand{\qut}[1]{\langle #1\rangle}
+
+\newcommand{\mbbc}{\mbb{C}}
+\newcommand{\mbbo}{\mbb{O}}
+\newcommand{\ob}{_\mbbo}
 
@@
 \newcommand{\Tm}{\msf{Tm}}
+\newcommand{\Cono}{\msf{Con}_{\mbbo}}
+\newcommand{\Subo}{\msf{Sub}_{\mbbo}}
+\newcommand{\Tyo}{\msf{Ty}_{\mbbo}}
+\newcommand{\Tmo}{\msf{Tm}_{\mbbo}}
+\newcommand{\hCon}{\wh{\msf{Con}}}
+\newcommand{\hSub}{\wh{\msf{Sub}}}
+\newcommand{\hTy}{\wh{\msf{Ty}}}
+\newcommand{\hTm}{\wh{\msf{Tm}}}
+
+\newcommand{\p}{\mathsf{p}}
+\newcommand{\q}{\mathsf{q}}
 
@@
 \newcommand{\emptycon}{\scaleobj{.75}\bullet}
+\newcommand{\id}{\msf{id}}
 
+\newcommand{\Set}{\mathsf{Set}}
+\newcommand{\Rep}{\msf{Rep}}
 \newcommand{\blank}{{\mathord{\hspace{1pt}\text{--}\hspace{1pt}}}}
+\newcommand{\emb}[1]{\ulcorner#1\urcorner}
+
+\newcommand{\Stage}{\msf{Stage}}
+\newcommand{\hato}{\bm\hat{\mbbo}}
+\newcommand{\ev}{\mbb{E}}
 
@@
   originally devised for the purpose of synthetic homotopy theory, also serves
-  as a system for staged compilation dependent types. 2LTT has numerous good
+  as a system for staged compilation with dependent types. 2LTT has numerous good
   properties for this use case: it has a concise specification, well-developed
@@
   normalization-by-evaluation, in that staging is given by the evaluation of
-  2LTT syntax in a semantic domain. Staging together with its correctness
-  constitutes a proof of strong conservativity of 2LLT over the object
-  theory. To our knowledge, this is the first system for staged compilation which
-  supports full dependent types and unrestricted staging for types.
+  2LTT syntax in a semantic domain. The staging algorithm together with its
+  correctness proof constitutes a proof of strong conservativity of 2LLT over
+  the object theory. To our knowledge, this is the first system for staged
+  compilation which supports full dependent types and unrestricted staging for
+  types.
 \end{abstract}
@@
     involving inlining control, partial evaluation and fusion optimizations. We
-    also describe several feature variations, enabling applications in
-    monomorphization and memory layout control.
+    also describe feature variations, enabling applications in monomorphization
+    and memory layout control.
   \item In \ref{TODO}, following \cite{twolevel}, we present a formal syntax of
@@
     Staging together with its correctness can be viewed as a \emph{strong
-    conservativity} theorem of 2LTT over the object theory. Intuitively, this
-    means that the possible object-level constructions in 2LTT are in bijection
-    with the constructions in the object theory, and staging witnesses that
-    meta-level constructions can be always computed away. This improves on the
-    weak notion of conservativity shown in \cite{capriotti2017models} and
+    conservativity} theorem of 2LTT over the object theory. This means that the
+    possible object-level constructions in 2LTT are in bijection with the
+    constructions in the object theory, and staging witnesses that meta-level
+    constructions can be always computed away. This improves on the weak notion
+    of conservativity shown in \cite{capriotti2017models} and
     \cite{twolevel}.
@@
 
-In this section, we provide a short overview of 2LTT. We work in the informal
-syntax of a dependently typed language which resembles Agda \cite{TODO}. We focus
-on examples and informal explanations here; the formal details will be presented
-in Section \cite{TODO}.
+In this section, we provide a short overview of 2LTT and its potential
+applications in staging. We work in the informal syntax of a dependently typed
+language which resembles Agda \cite{TODO}. We focus on examples and informal
+explanations here; the formal details will be presented in Section \cite{TODO}.
 
@@
 abstraction. A $\Sigma$-type is written as $(x : A) \times B$, with pairing as
-$(t,\,u)$, and we may use pattern matching notation on pairs, e.g.\ as in
-$\lambda\,(x,\,y).\,t$. The unit type is $\top$ with element $\tt$. We will also
-use Agda-style notation for implicit arguments, where $t : \{x : A\} \to B$
-implies that the first argument to $t$ is inferred by default, and we can
-override this by writing a $t \{u\}$ application. We may also implicitly
-quantify over arguments (in the style of Idris and Haskell), for example
-when declaring $\mit{id} : A \to A$ with the assumption that $A$ is universally
-quantified.
+$(t,\,u)$, projections as $\fst$ and $\snd$, and we may use pattern matching
+notation on pairs, e.g.\ as in $\lambda\,(x,\,y).\,t$. The unit type is $\top$
+with element $\tt$. We will also use Agda-style notation for implicit arguments,
+where $t : \{x : A\} \to B$ implies that the first argument to $t$ is inferred
+by default, and we can override this by writing a $t \{u\}$ application. We may
+also implicitly quantify over arguments (in the style of Idris and Haskell), for
+example when declaring $\mit{id} : A \to A$ with the assumption that $A$ is
+universally quantified.
 \end{notation}
@@
 
-\paragraph{Universes} We have universes $\U_{i,j}$, where $i \in \{0,1\}$, and $j \in \mbb{N}$.
-The $i$ index denotes stages, where $0$ is the runtime (object-level) stage, and
-$1$ is the compile time (meta-level) stage. The $j$ index denotes universe sizes
-in the usual sense of type theory. We assume Russell-style universes, with
+\paragraph{Universes}
+We have universes $\U_{i,j}$, where $i \in \{0,1\}$, and $j \in \mbb{N}$.  The
+$i$ index denotes stages, where $0$ is the runtime (object-level) stage, and $1$
+is the compile time (meta-level) stage. The $j$ index denotes universe sizes in
+the usual sense of type theory. We assume Russell-style universes, with
 $\U_{i,j} : \U_{i, j+1}$. However, for the sake of brevity we will usually omit
-the $j$ indices in this section, and simply write $\U_0$ or $\U_1$.
+the $j$ indices in this section, since sizing is orthogonal to our use-cases and
+examples.
 \begin{itemize}
@@
 \end{itemize}
-It is not required that we have the \emph{same} type formers at both stages. As we will
-see in Section \cite{TODO}, simpler object languages have the advantage that they are
-easier to process during downstream compilation.
+It is not required that we have the \emph{same} type formers at both stages. We
+will discuss setups with different languages at different stages in Seection
+\ref{todo}.
 
-\paragraph{Moving between stages} At this point, our system is rather limited, since
-there is no interaction between the stages. We add such interaction via the follow
-three operations.
+\paragraph{Moving between stages}
+At this point, our system is rather limited, since there is no interaction
+between the stages. We enable such interaction through the following three
+operations.
 \begin{itemize}
@@
 over $\mit{id}_1$, but in some cases we really have to quantify over
-$\Lift\U_0$. This brings us the next example.
+$\Lift\U_0$. This brings us to the next example.
 
@@
 \end{alignat*}
-This $\mit{map}$ function works with quantification over $\Lift \U_0$ but not
+This $\mit{map}$ function can be defined with quantification over $\Lift \U_0$ but not
 over $\U_1$, because $\List_0$ expects type parameters in $\U_0$, and there is
@@
 we bind a runtime expression to a runtime variable, and only use that variable
-in subsequent staging, only the variable itself can be duplicated, as an
-expression. One solution is to do an ad-hoc let-insertion:
+in subsequent staging, only the variable itself can be duplicated. One solution
+is to do an ad-hoc let-insertion:
 \begin{alignat*}{6}
@@
 \end{alignat*}
-Types are necessarily computed to runtime types, since we can only stage to such types.
+Types are necessarily computed to runtime types, e.g.\ an embedded
+representation of the natural number type is evaluated to $\qut{\Nat_0}$.
 Contexts are computed as follows:
@@
 \end{alignat*}
-This is an nice example for the usage of \emph{partially static
+This is a nice example for the usage of \emph{partially static
 data}\cite{TODO}: semantic contexts are \emph{static} lists storing
@@
 
-\subsection{Properties of Lifting}
+\subsection{Properties of Lifting \& Binding Time Improvements}
 
-preservation properties
-\\\\
-stage inference
+We describe more generally the action of $\Lift$ on type formers. First, $\Lift$
+preserves negative type formers up to definitional isomorphism \cite{twolevel}:
+\begin{alignat*}{5}
+  \Lift((x : A) \to B\,x) &\simeq ((x : \Lift A) \to \Lift\,(B\,\spl x))\\
+  \Lift ((x : A) \times B) &\simeq ((x : \Lift A) \times \Lift (B\,\spl x))\\
+  \Lift \top_0 &\simeq \top_1
+\end{alignat*}
+For function types, the preservation maps are the following:
+\begin{alignat*}{5}
+  & \hspace{-6em}\rlap{$\mit{pres}_\to : \Lift((x : A) \to B\,x) \to ((x : \Lift A) \to \Lift\,(B\,\spl x))$}\\
+  & \hspace{-6em}\mit{pres}_\to\,f     &&:= \lambda\,x.\,\,\qut{f\,\spl x}\\
+  & \hspace{-6em}\mit{pres}_\to^{-1}\,f &&:= \qut{\lambda\,x.\,\,\spl(f\,\qut{x})}
+\end{alignat*}
+With this, we have that $\mit{pres}_\to\,(\mit{pres}_\to^{-1}\,f)$ is
+definitionally equal to $f$, and also the other way around. Preservation maps
+for $\Sigma$ and $\top$ work analogously.
 
-\subsection{Restricting the Object Language}
+By rewriting a 2LTT program left-to-right along preservation maps, we perform
+what is termed a \emph{binding time improvement} in the partial evaluation
+literature \cite{TODO}. Note that the output of $\mit{pres}_{\to}$ uses a
+meta-level $\lambda$, while going the other way we introduce a runtime
+binder. Meta-level function and $\Sigma$ types support more computation during
+staging, so in many cases it is beneficial to use the improved forms. In some
+cases though we may want to use unimproved forms, to limit the size of generated
+code. This is similar to what we have seen with let-insertion. For a minimal
+example, consider the following unimproved version of $\mit{id}_\Lift$:
+\begin{alignat*}{5}
+  & \mit{id}_\Lift : (A : \Lift \U_0) \to \Lift(\spl A \to \spl A) \\
+  & \mit{id}_\Lift := \lambda\,A.\,\,\qut{\lambda\,x.\,x}
+\end{alignat*}
+This can be used at the runtime stage as
+$\spl(\mit{id}_\Lift\,\qut{\Bool_0})\,\true_0$, which is staged to
+$(\lambda\,x.\,x)\,\true_0$.  This introduces a useless $\beta$-redex, so in
+this case the improved version is clearly preferable.
 
-monomorphization
-\\\\
-memory layout control
-\\\\
-typed closures
+For inductive types in general we only get oplax preservation. For example, we
+have $\Bool_1 \to \Lift \Bool_0$, defined as
+$\lambda\,b.\,\msf{if}\,b\,\msf{then}\,\qut{\true_0}\,\msf{then}\,\qut{\false_0}$.
+In the staging literature, this is called ``serialization'' \cite{TODO}, or
+``lifting'' in the context of Template Haskell \cite{TODO}. In the other
+direction, we can only define constant functions from $\Lift \Bool_0$ to
+$\Bool_1$.
 
-\subsection{Fusion}
-2LTT is also a natural setting for a wide range of \emph{fusion
-optimizations}. We only describe here a simple example, \emph{foldr-build
-fusion} for lists, which is notably used in GHC Haskell \cite{TODO}. The idea is
-the following: we want to eliminate intermediate lists, e.g.\ fuse repeated
-mapping to a single traversal. For this, we exploit that there is an isomorphic
-\emph{meta-level} representation for runtime lists which supports more
-definitional equations. Starting from $\Lift (\List_0\,A)$, we can use
-Böhm-Berarducci encoding under the lifting to get
+The lack of elimination principles for $\Lift A$ means that we cannot inspect
+the internal structure of expressions. This is called ``generativity'' in
+staging \cite{TODO}. We will briefly discuss non-generative staging in Section
+\ref{TODO}.
+
+In particular, we have no serialization map from $\Nat_1 \to \Nat_1$ to
+$\Lift(\Nat_0 \to \Nat_0)$. However, when $A : \U_1$ is \emph{finite}, and $B :
+\U_1$ can be serialized, then $A \to B$ can be serialized, because it is
+equivalent to a finite product. For instance, $\Bool_1 \to \Nat_1 \simeq \Nat_1
+\times \Nat_1$.  In 2LTT, $A$ is called \emph{cofibrant} \cite{TODO}: this means
+that for each $B$, $A \to \Lift\,B$ is equivalent to $\Lift C$ for some
+$C$. This formalizes the so-called ``trick'' in partial evaluation, which
+improves binding times by $\eta$-expanding functions out of finite sums
+\cite{TODO}.
+
+\subsubsection{Fusion}
+Fusion optimizations can be viewed as binding time improvement techniques for
+general inductive types. The basic idea is that by lambda-encoding an inductive
+type, it is brought to a form which can be binding-time improved. For instance,
+consider foldr-build fusion for lists, which is employed in GHC Haskell
+\cite{TODO}. Starting from $\Lift (\List_0\,A)$, we use Böhm-Berarducci
+encoding \cite{TODO} under the lifting to get
 \[ \Lift((L : \U_0) \to (A \to L \to L) \to L \to L) \]
-However, $\Lift$ preserves function types, so the above is definitionally
-isomorphic to
+which is isomorphic to
 \[ (L : \Lift\,U_0) \to (\Lift\,A \to \Lift\,\spl L \to \Lift\,\spl L)
-\to \Lift\,\spl L \to \Lift\,\spl L. \] Now, we may abbreviate the above type as
-$\msf{BBList}\,A$. We write fusible functions on $\msf{BBList}\,A$, and when we
-want to fuse a definition, we convert from $\Lift\,(\List_0\,A)$ to
-$\msf{BBList}\,A$, apply operations on that representation, then convert back in
-the end.
+    \to \Lift\,\spl L \to \Lift\,\spl L. \]
+Alternatively, for \emph{stream fusion}, we embed $\List\,A$ into the
+coinductive colists (i.e.\ the possibly infinite lists), and use a terminal
+lambda-encoding. The embedding into the ``larger'' structure enables some staged
+optimizations which are otherwise not possible, such as fusion for the
+$\msf{zip}$ function \cite{TODO}. However, the price we pay is that converting
+back to lists from colists is not necessarily total.
 
-Now, in 2LTT it is not provable that Böhm-Berarducci encoding is an isomorphism \cite{TODO},
-so it certainly does not hold that back-and forth conversion between the two
-representations is definitionally the identity function. In GHC Haskell, this
-equality is added as a \emph{rewrite rule}. This makes it possible to hide the
-back-and-forth conversions from library users, since inverse conversions will
-hopefully cancel out and cause no additional overhead. A similar mechanism could be
-used in 2LTT as well. However, such implicit conversions are not used in most
-languages which support fusion. For example, iterators in the Rust language
-\cite{TODO}, which use a form of stream fusion, require explicit conversion to
-iterators.
+We do not detail the implementation of fusion in 2LTT here. In short, 2LTT is a
+natural setting for a wide range of fusion setups. A major advantage of fusion
+in 2LTT is the formal guarantee of staging, in contrast to implementations where
+compile-time computation relies on ad-hoc user annotations and general-purpose
+optimization passes. For instance, fusion in GHC relies on rewrite rules and
+inlining annotations which have to be carefully tuned and ordered, and it is
+quite possible to get pessimized code via failed fusion.
 
-We may compare GHC and 2LTT with respect to the implementation of fusion. In
-2LTT, fusion happens through meta-level $\beta$-reduction, which is formally
-guaranteed to go through, and which can be also computed very efficiently. In
-GHC, all of the necessary computation is realized by rewrite rules and
-general-purpose optimization passes. This is in practice very fragile and
-non-composable. It requires carefully tuning every individual collection of
-rewrite rules, and also ordering rule applications in specific ways, while
-keeping details of the compiler optimization passes in mind. And when fusion
-fails to go through, it is usually a performance \emph{pessimization}.
+\subsubsection{Inferring staging operations}
+During bidirectional elaboration \cite{TODO}, we can use the preservation
+isomorphisms and the quote-splice isomorphism as a coercive subtyping
+system. When elaboration needs to compare an inferred and an expected type, it
+may insert transports along isomorphisms. We implemented this feature in our
+prototype. It additionally supports Agda-style implicit arguments and pattern
+unification, so it can elaborate the following definition:
+\begin{alignat*}{3}
+  & \mit{map} : \{A\,B : \Lift\U_0\} \to (\Lift A \to \Lift B)
+      \to \Lift(\List_0\,A) \to \Lift(\List_0\,B)\\
+  & \mit{map} := \lambda\,f\,\mit{as}.\,
+      \msf{foldr}_0
+        (\lambda\,a\,\mit{bs}.\, \cons_0\,(f\,a)\,\mit{bs})\,
+        \nil_0\,
+        as
+\end{alignat*}
+We may go a bit further, and also add the coercive subtyping rule $\U_0 \leq
+\U_1$, witnessed by $\Lift$. Then, the type of $\mit{map}$ can be written as
+$\{A\,B : \Lift\U_0\} \to (A \to B) \to \List_0\,A \to \List_0\,B$. However,
+here the elaborator has to make a choice, whether to elaborate to improved or
+unimproved types. In this case, the fully unimproved type would be
+\[ \{A\,B : \Lift\U_0\} \to \Lift((\spl A \to \spl B) \to \List_0\,\spl A \to \List_0\,\spl B). \]
+It seems to the author of this paper that improved types are a sensible default,
+and we can insert explicit lifting when we want to opt for unimproved
+types. This is also available in our prototype.
 
-\subsection{Properties of Lifting}
+\subsection{Variations of Object-Level Languages}
 
-%% we shall work in the syntax of a dependently typed language
-%% which resembles Agda.
+%% In principle, 2LTT supports any sensible type-theoretic feature on the meta
+%% level. On the object level though, we have yet more freedom in choosing type
+%% formers and even the equational theory of terms. We may choose to omit
+%% $\beta\eta$-rules from the object language.
 
-\subsection{Informal Syntax}
+%% \todo{merge closure typing into rep poly?}
+%% \todo{range of object languages?}
+%% \todo{effectful, etc object languages?}
 
-\section{Variations \& Applications}
+In the following, we consider variations on object-level languages, with a focus
+on applications in downstream compilation after staging. Adding restrictions or
+more distinctions to the object language can make it easier to optimize and
+compile it.
 
-\subsection{Fusion}
+%% Simpler object languages are easier to optimize and compile, but they are often
+%% more tedious to directly program in. 2LTT can be used as a principled way to
+%% extended languages with abstraction facilities, through metaprogramming. We
+%% look at several setups with restricted object languages.
 
-\subsection{Monomorphization}
+\subsubsection{Monomorphization}
 
-\subsection{Levity Polymorphism}
+In this case, the object language is simply typed, so every type is known
+statically. This makes it easy to assign different memory layouts to different
+types, and generate code accordingly for each type. Moving to 2LTT, we still
+want to abstract over runtime types at compile time, so we use the following
+setup.
+\begin{itemize}
+\item We have a \emph{jugdment}, written as $A\,\msf{type_0}$, for well-formed
+  runtime types. Runtime types may be closed under simple type formers.
+\item We have a type $\Ty_0 : \U_1$ in lieu of the previous $\Lift \U_0$.
+\item For each $A\,\msf{type}_0$, we have $\Lift A : \Ty_0$.
+\item We have quoting and splicing for types and terms. For types, we send $A\,
+  \msf{type}_0$ to $\qut{A} : \Ty_0$. For terms, we send $t : A$ to $\qut{t} :
+  \Lift A$.
+\end{itemize}
+Despite the restriction to simple types at runtime, we can still write arbitrary
+higher-rank polymorphic functions in this setup, such as a function with type
+$((A : \Ty_0) \to \Lift\,\spl A \to \Lift \spl A) \to \Lift\,\Bool_0$. This
+function can be only applied to statically known arguments in runtime
+code, so the polymorphism can be staged away. The main restriction that
+programmers have to keep in mind is that polymorphic functions cannot be stored
+inside runtime data types.
 
-\subsection{Typed Closures}
+This setup is fairly similar to the well-known monomorphization models in the
+C++ and Rust programming languages. The evident difference is that in 2LTT we
+can use a full type theory at compile time, as opposed to languages which are
+less expressive and less principled. Additionally, we will see in Section
+\ref{TODO} that this setup is compatible with an \emph{induction principle} for
+$\Ty_0$, so that we can analyze the structure of runtime types.
 
-\section{Formal Syntax \& Models}
+\subsubsection{Memory representation polymorphism}
+
+This refines monomorphization, in that types are not directly identified with
+memory representations, but instead representations are internalized in 2LTT as
+a meta-level type, and runtime types are indexed over representations.
+\begin{itemize}
+\item We have $\Rep : \U_1$ as the type of memory representations. We have
+  considerable freedom in the specification of $\Rep$. A simple setup may
+  distinguish references from unboxed products, i.e.\ we have $\msf{Ref} : \Rep$
+  and $\msf{Prod} : \Rep \to \Rep \to \Rep$, and additionally we may assume
+  any desired primitive machine representation as a $\Rep$.
+\item We have Russell-style $\U_{0,j} : \Rep \to \U_{0, j+1}\,r$, where $r$ is
+  some chosen runtime representation for types; often this would mark types as
+  erased at runtime. We leave the meta-level $\U_{1,j}$ hierarchy unchanged.
+\item We may introduce unboxed $\Sigma$ types and primitive machine types in the
+  runtime language. For $A : \U_{0}\,r$ and $B : A \to \U_{0}\,r'$, we may have
+  $(x : A) \times B\,x : \U_{0}\,(\msf{Prod\,r\,r'})$. Thus, we have
+  type dependency, but we do not have dependency in memory representations.
+\end{itemize}
+Since $\Rep$ is meta-level, there is no way to abstract over it at runtime, and
+during staging all $\Rep$ indices are computed to concrete canonical
+representations. This is a way to reconcile dependent types with control over
+memory layouts.  The unboxed flavor of $\Sigma$ ends up with a statically known
+flat memory representation, computed from the representations of the fields.
+
+In principle, it should be possible to have \emph{dependent memory layouts};
+this would be a more advanced variation on type-passing polymorphism
+\cite{TODO}, where layouts of objects may depend on runtime data. For
+example, length-prefixed flat arrays would not be a primitive notion, but simply
+defined as $(n : \Nat) \times \msf{Vec}\,n\,A$. However, it seems to be highly
+challenging to implement runtime systems and precise garbage collection
+generically over dependently typed memory layouts. See \texttt{sixten}
+\cite{TODO} for an experimental implementation with conservative garbage
+collection.
+
+\section{Formal 2LTT}
+
+In this section we switch to a formal description of 2LTT, which we will use in
+the subsequent sections to define the staging algorithm and prove its
+correctness. First we describe the metatheory that we work in.
 
 \subsection{Metatheory}
-\subsection{Models}
+TODO
 
-\section{Staging by Evaluation}
+\subsection{Models and Syntax of 2LTT}
+We use an algebraic specification for models, and specify the syntax as the
+initial model. We only handle well-formed syntactic objects and only define
+operations which respect syntactic definitional equality; in fact, we identify
+definitional equality with metatheoretic equality. First, we define the
+structural scaffolding of 2LTT without type formers.
 
-\subsection{Presheaf Model of 2LTT}
+\begin{definition} A model of \textbf{basic 2LTT} consists of the following.
+\begin{itemize}
+\item
+  A category $\mbbc$ with a terminal object. We denote the set of objects as
+  $\Con_{\mbbc} : \Set$ and use capital Greek letters starting from $\Gamma$ to
+  refer to objects. The set of morphisms is $\Sub_{\mbbc} : \Con_{\mbbc} \to
+  \Con_{\mbbc} \to \Set$, and we use $\sigma$, $\delta$ and so on to refer to
+  morphisms. The terminal object is written as $\emptycon$ with unique morphism
+  $\epsilon : \Sub_\mbbc\,\Gamma\,\emptycon$. We omit the $\mbbc$ subscript if
+  it is clear from context.
+\item
+  For each $i \in \{0,1\}$ and $j \in \mbb{N}$, we have $\Ty_{i,j} : \Con
+  \to \Set$ and $\Tm_{i,j} : (\Gamma : \Con) \to \Ty_{i,j}\,\Gamma \to \Set$,
+  where $\Ty$ is a presheaf over $\mbbc$ and $\Tm_{i,j}$ is a presheaf over the
+  category of elements of $\Ty_{i,j}$. This means that both types ($\Ty$) and
+  terms ($\Tm$) can be substituted, and substitution has functorial action. We
+  use $A$, $B$, $C$ to refer to types and $t$, $u$, $v$ to refer to terms, and
+  use $A[\sigma]$ and $t[\sigma]$ for substituting types and
+  terms. Additionally, for each $\Gamma : \Con$ and $A : \Ty_{i,j}\,\Gamma$, we
+  have the extended object $\Gamma \ext A : \Con$ such that there is a natural
+  isomorphism $\Sub\,\Gamma\,(\Delta\ext A) \simeq ((\sigma :
+  \Sub\,\Gamma\,\Delta) \times \Tm_{i,j}\,\Gamma\,(A[\sigma]))$.
+\item
+  For each $j$ we have a \emph{lifting structure}, consisting of a natural
+  transformation $\Lift : \Ty_{0,j}\,\Gamma \to \Ty_{1,j}\,\Gamma$, and an
+  invertible natural transformation $\qut{\blank} : \Tm_{0,j}\,\Gamma\,A \to
+  \Tm_{1,j}\,\Gamma\,(\Lift A)$, with inverse $\spl\blank$.
+\end{itemize}
+\end{definition}
 
-\subsection{Closed staging}
-\subsection{Yoneda, representability, intensional analysis}
+In short, for each $i$ and $j$ we have a family structure in the sense of
+categories-with-families \cite{TODO}, such that there is a family morphism from
+each $(\Ty_{0,j},\,\Tm_{0,j})$ to $(\Ty_{1,j},\,\Tm_{1,j})$ in the sense of
+\cite{TODO}.
 
-\section{Stability}
+The following notions are derivable in any model:
+\begin{itemize}
+\item
+  By moving left-to-right along $\Sub\,\Gamma\,(\Delta\ext A) \simeq ((\sigma :
+  \Sub\,\Gamma\,\Delta) \times \Tm_{i,j}\,\Gamma\,(A[\sigma]))$,
+  and starting from the identity morphism $\id : \Sub\,(\Gamma\ext A)\,(\Gamma\ext A)$, we recover
+  the \emph{weakening substitution} $\p : \Sub\,(\Gamma\ext A)\,\Gamma$ and the \emph{zero variable}
+  $\q : \Tm_{i,j}\,(\Gamma\ext A)\,(A[\p])$.
+\item
+  By weakening $\q$, we recover a notion of variables as De Bruijn
+  indices. In general, the $n$-th De Bruijn index is defined as $\q[\p^{n}]$,
+  where $\p^{n}$ denotes $n$-fold composition.
+\item
+  By moving right-to-left along $\Sub\,\Gamma\,(\Delta\ext A) \simeq
+  ((\sigma : \Sub\,\Gamma\,\Delta) \times \Tm_{i,j}\,\Gamma\,(A[\sigma]))$, we
+  recover the operation which extends a morphism with a term. In the syntax
+  (initial model), this justifies the view of $\Sub$ as a list of terms, i.e.\ a
+  parallel substitution. We denote the extension operation as $(\sigma,\,t)$ for
+  $\sigma : \Sub\,\Gamma\,\Delta$.
+\end{itemize}
 
-\subsection{Open staging}
+\begin{notation}
+De Bruijn indices are rather hard to read, so we will often use nameful notation
+for binders and substitutions. For example, we may write $\Gamma \ext (x : A)
+\ext (y : B)$ for a context, and subsequently write $B[y \mapsto t]$ for
+substituting the $x$ variable for some term $t : \Tm_{i,j}\,\Gamma\,A$. Using
+nameless notation, we would instead have $B : \Ty_{i,j}\,(\Gamma \ext A)$ and
+$B[\id,\,t]$; here we recover single substitution by extending the identity
+substitution $\id : \Sub\,\Gamma\,\Gamma$ with $t$.
 
-\section{Soundness}
+We may also use implicit weakening: if a type or term is in a $\Gamma$ context,
+we may use it in an extended $\Gamma \ext A$ context without marking the
+weakening substitution.
+\end{notation}
 
-\subsection{Internal Language \& Features}
-\subsection{The Logical Relation}
-\subsection{Externalization \& Soundness}
+\begin{definition} A \textbf{model of 2LTT} is a model of basic 2LTT
+which supports certain type formers. For the sake of brevity, we only present
+our results for a small collection of type formers. However, we will argue that
+our results easily extend to any general notion of inductive type formers. We
+specify type formers in the following.  We assume that all type formers are
+natural, i.e.\ stable under substitution.
+\begin{itemize}
+\item \emph{Universes}. For each $i$ and $j$, we have a Coquand-style universe
+  \cite{TODO} in $\Ty_{i,j}$. This consists of $\U_{i,j} : \Ty_{i,j+1}\,\Gamma$,
+  together with $\El : \Tm_{i,j+1}\,\Gamma\,\U_{i,j} \to \Ty_{i,j}\,\Gamma$ and $\Code$, where $\Code$
+  and $\El$ are inverses.
+\item \emph{$\Sigma$-types}. We have $\Sigma\,(x : A)\,B : \Ty_{i,j}\,\Gamma$ for $A : \Ty_{i,j}\,\Gamma$
+  and $B : \Ty_{i,j}\,(\Gamma \ext (x : A))$, together with a natural isomorphism of pairing and projections:
+  \[ \Tm_{i,j}\,\Gamma\,(\Sigma\,A\,B) \simeq ((t : \Tm_{i,j}\,\Gamma\,A) \times \Tm_{i,j}\,\Gamma\,(B[x\mapsto t]) \]
+  We write $(t,\,u)$ for pairing and $\fst$ and $\snd$ for projections.
+\item \emph{Function types}. We have $\Pi\,(x : A)\,B : \Ty_{i,j}\,\Gamma$ for $A :
+  \Ty_{i,j}\,\Gamma$ and $B : \Ty_{i,j}\,(\Gamma \ext (x : A))$, together with
+  $\app : \Tm_{i,j}\,\Gamma\,(\Pi\,A\,B) \to \Tm_{i,j}\,(\Gamma \ext (x :
+  A))\,B$ and its inverse $\lam$.
+\item \emph{Natural numbers}. We have $\Nat_{i,j} : \Ty_{i,j}\,\Gamma$, $\zero_{i,j} : \Tm_{i,j}\,\Gamma\,\Nat_{i,j}$,
+  and $\suc_{i,j} : \Tm_{i,j}\,\Gamma\,\Nat_{i,j} \to \Tm_{i,j}\,\Gamma\,\Nat_{i,j}$. The eliminator is the following.
+  \begin{alignat*}{5}
+    & \NatElim :\hspace{0.5em}
+                &&  (P &&: \Ty_{i,k}\,(\Gamma \ext (n : \Nat_{i,j}) \\
+    &           &&  (z &&: \Tm_{i,k}\,\Gamma\,(P[n\mapsto \zero_{i,j}])) \\
+    &           &&  (s &&: \Tm_{i,k}\,(\Gamma \ext (n : \Nat_{i,j}) \ext (\mit{pn} : P[n\mapsto n]))\,(P[n\mapsto \suc_{i,j}\,n]))\\
+    &           &&  (t &&: \Tm_{i,j}\,\Gamma\,\Nat_{i,j}) \\
+    &\hspace{3em}\to        &&&& \Tm_{i,j}\,\Gamma\,(P[n\mapsto t]))
+  \end{alignat*}
+  We also have the $\beta$-rules:
+  \begin{alignat*}{5}
+    & \NatElim\,P\,z\,s\,\zero_{i,j}     &&= z \\
+    & \NatElim\,P\,z\,s\,(\suc_{i,j}\,t) &&= s[n \mapsto t,\,\mit{pn} \mapsto \NatElim\,P\,z\,s\,t]
+  \end{alignat*}
+  Note that we can eliminate from a $j$ level to any $k$ level.
+\end{itemize}
+\end{definition}
 
-\section{Related Work}
+\begin{definition}
+The \textbf{syntax of 2LTT} is defined to be the initial model of 2LTT. The syntax
+can be assumed to exist, since the notion of model is algebraic; the
+specification can be expressed using a suitable notion of algebraic signatures,
+such as one for essentially algebraic theories \cite{TODO} or as a finitary
+quotient inductive-inductive signature \cite{TODO}. The initiality of syntax
+directly yields a notion of \emph{recursion on syntax}, and it can be shown that
+initiality also implies an \emph{induction principle} \cite{TODO}.
+\end{definition}
 
-Igarashi, Kiselyov et al, MetaML, MetaOCaml, Carette, TH, Scala ppl (?)
-PE lit ?
-2LTT, Voevodsky,
+The examples in Section \ref{TODO} can be faithfully represented in the formal
+syntax. The only notable difference is that the formal syntax uses Coquand-style
+universes instead of Russell-style ones. We use the former because it is much
+easier to model in the semantics in Section \ref{TODO}. We note though that it
+is straightforward to elaborate from a Russell-style surface syntax to the
+formal version, by inserting $\El$ and $\Code$ as required.
 
-\section{Future Work \& Conclusions}
+\subsubsection{Comparison to Annekov et al.}
+Comparing our models to the primary reference on 2LTT \cite{twolevel}, the main
+difference is the handling of ``sizing'' levels. In ibid.\ there is a cumulative
+lifting from $\Ty_{i,j}$ to $\Ty_{i,j+1}$, which we do not assume. Instead,
+we allow elimination from $\Nat_{i,j}$ into any $k$ level. This means that we
+can manually define lifting maps from $\Nat_{i,j}$ to $\Nat_{i,j+1}$ by
+elimination. This is more similar to e.g.\ Agda, where we do not have cumulativity,
+but we can define explicit lifting from $\Nat_j : \Set_j$ to $\Nat_k : \Set_k$.
+
+In \cite{twolevel}, ``two-level type theory'' specifically refers to the setup
+where the object-level is a homotopy type theory and the meta level is an
+extensional type theory. In contrast, we allow a wider range of setups under the
+2LTT umbrella. Annekov et al.\ also considers a range of additional
+strengthenings and extension of 2LTT \cite[Section~2.4]{twolevel}, most of which
+are of use in synthetic homotopy theory. We do not assume any of these, and
+stick to the most basic formulation of 2LTT.
+
+\subsection{Models and Syntax of the Object Theory}
+
+We also need to specify the object theory, which serves as the output language
+of staging. In general, the object theory corresponding to a particular flavor
+of 2LTT is simply the type theory that supports only the object-level $\Ty_{0,j}$
+hierarchy with its type formers.
+
+\begin{definition}
+A \textbf{model of the object theory} is a category-with-families, with types and
+terms indexed over $j \in \mbb{N}$, supporting Coquand-style universes $\U_j$,
+type formers $\Pi$, $\Sigma$ and $\Nat_j$, with elimination from $\Nat_j$ to
+any $k$ level.
+\end{definition}
+
+\begin{definition}
+Like before, the \textbf{syntax of the object theory} is the initial model.
+\end{definition}
+
+\begin{notation}
+From now on, by default we use $\Con$, $\Sub$, $\Ty$ and $\Tm$ to refer to sets
+in the syntax of 2LTT. We use $\mbbo$ to refer to the object syntax, and
+$\Con\ob$, $\Sub\ob$, $\Ty\ob$ and $\Tm\ob$ to refer to its underlying sets.
+\end{notation}
+
+\begin{definition}
+We have an \textbf{embedding} of object syntax syntax into 2LTT syntax, consisting
+of the following functions:
+\begin{alignat*}{4}
+  & \emb{\blank} : \Con\ob \to \Con\\
+  & \emb{\blank} : \Sub\ob\,\Gamma\,\Delta \to \Sub\,\emb{\Gamma}\,\emb{\Delta}\\
+  & \emb{\blank} : \Ty_{\mbbo j}\,\Gamma \to \Ty_{0,j}\,\emb{\Gamma}\\
+  & \emb{\blank} : \Tm_{\mbbo j}\,\Gamma\,A \to \Tm_{0,j}\,\emb{\Gamma}\,\emb{A}
+\end{alignat*}
+Embedding maps all type and term formers to the corresponding ones in 2LTT, and
+strictly preserves all structure.
+\end{definition}
+
+\section{The Staging Algorithm}
+
+In this section we specify what we mean by staging, then define a staging
+algorithm.
+
+\begin{definition}
+A \textbf{staging algorithm} consists of two functions:
+  \begin{alignat*}{3}
+    & \Stage : \Ty_{0,j}\,\emb{\Gamma} \to \Ty_{\mbbo j}\,\Gamma \\
+    & \Stage : \Tm_{0,j}\,\emb{\Gamma}\,A \to \Tm_{\mbbo j}\,\Gamma\,(\Stage\,A)
+  \end{alignat*}
+Note that we can stage open types and terms as long as their contexts are purely
+object-level. By \emph{closed staging} we mean staging only for closed types and
+terms.
+\end{definition}
+\begin{definition}
+  A staging algorithm $\Stage$ is \textbf{correct} if the following properties hold:
+  \begin{itemize}
+  \item \emph{Soundness:} $\emb{\Stage\,A} = A$ and $\emb{\Stage\,t} = t$.
+  \item \emph{Completeness:} staging respects definitional equality.
+  \item \emph{Stability:} $\Stage\,\emb{A} = A$ and $\Stage\,\emb{t} = t$.
+  \end{itemize}
+\end{definition}
+We make some remarks on correctness. First, we get completeness for free in our
+algebraic setup, since \emph{all} functions definable in the metatheory must
+respect definitional equality.
+
+Second, note that soundness and stability together is the statement that
+embedding is invertible on types and terms. This is a statement of
+\emph{conservativity} of 2LTT over the object theory. In \cite{TODO} a
+significantly weaker conservativity theorem is shown, which expresses that there
+exists a function from $\Tm_{0,j}\,\emb{\Gamma}\,\emb{A}$ to $\Tm_{\mbbo
+  j}\,\Gamma\,A$.
+
+Lastly, we note that the correctness terminology is borrowed from previous works
+on normalization-by-evaluation, where the soundness-completeness-stability trio
+has been used several times \cite{TODO}. In \cite{cubicalnbe}, stability is
+instead called ``idempotence''.
+
+\subsection{The Presheaf Model}
+
+We observe that the presheaf model described in \cite[Section~2.5.3]{twolevel}
+immediately yields a closed staging algorithm, by evaluation of 2LTT types and
+terms in the model. In this model, contexts are presheaves over the syntactic
+category of the object theory. We call the model $\hato$ and denote its
+components by putting hats on 2LTT components. We summarize the key components
+in the following.
+
+\begin{notation}
+In this section, we switch to naming elements of $\Cono$ as $a$, $b$ and $c$,
+and elements of $\Subo$ as $f$, $g$, and $h$, to avoid name clashing with
+contexts and substitutions in the presheaf model.
+\end{notation}
+
+\subsubsection{The syntactic category and the meta-level fragment}
+
+\begin{definition} $\bs{\wh{\Con} : \Set_{\omega+1}}$ is defined as the set of presheaves
+over $\mbbo$. A $\Gamma : \widehat{\Con}$ has an action on objects $|\Gamma| :
+\Cono \to \Set_\omega$ and an action on morphisms $\blank[\blank] : |\Gamma|\,b
+\to \Subo\,a\,b \to |\Gamma|\,a$, such that $\gamma[\id] = \gamma$ and
+$\gamma[f\circ g] = \gamma[f][g]$.
+\begin{notation}
+We reuse the substitution notation $\blank[\blank]$ for the action on morphisms.
+Also, we use lowercase $\gamma$ and $\delta$ to denote elements of $|\Gamma|\,a$
+and $|\Delta|\,a$ respectively.
+\end{notation}
+\end{definition}
+\begin{definition} $\bs{\hSub\,\Gamma\,\Delta : \Set_\omega}$ is the set of natural transformations
+from $\Gamma$ to $\Delta$. A $\sigma : \hSub\,\Gamma\,\Delta$, has an action on
+objects $|\sigma| : \{a : \Cono\} \to |\Gamma|\,a \to |\Delta|\,a$ such that
+$|\sigma|\,(\gamma[f]) = (|\sigma|\,\gamma)[f]$.
+\end{definition}
+
+\begin{definition}
+$\bs{\hTy_{1,j}\,\Gamma : \Set_\omega}$ is the set of \emph{displayed presheaves}
+over $\Gamma$; see e.g.\ \cite{TODO}. This is equivalent to the set of
+presheaves over the category of elements of $\Gamma$, but it is usually more
+convenient in calculations. An $A : \hTy\,\Gamma$ has an action on objects
+$|A| : \{a : \Cono\} \to |\Gamma|\,a \to \Set_j$ and an action on
+morphisms $\blank[\blank] : |A|\,\gamma \to (f : \Subo\,a\,b) \to
+|A|\,(\gamma[f])$, such that $\alpha[\id] = \alpha$ and $\alpha[f \circ g] = \alpha[f][g]$.
+
+\begin{notation} We write $\alpha$ and $\beta$ respectively for elements of $|A|\,\gamma$ and $|B|\,\gamma$.
+\end{notation}
+\end{definition}
+
+\begin{definition}
+  $\bs{\hTm_{1,j}\,\Gamma\,A : \Set_\omega}$ is the set of sections of the
+  displayed presheaf $A$. This can be viewed as a dependently typed analogue of a natural transformation.
+  A $t : \hTm_{1,j}\,\Gamma\,A$ has an action on objects $|t| : \{a\} \to (\gamma : |\Gamma|\,a) \to |A|\,\gamma$,
+  such that $|t|\,(\gamma[f]) = (|t|\,\gamma)[f]$.
+\end{definition}
+
+%% \begin{definition}
+%% $\bs{\wh{\U}_{1,j} : \hTy_{1,j+1}\,\Gamma}$ is defined by
+%% $|\wh{\U}_{1,j}|\,\{a\}\,\_ = \hTy_{1,j}\,(\y\,a)$, and defining action on
+%% morphisms by $\hTy$ substitution. Note that the universe levels check out in
+%% this definition, since by unfolding $\hTy_{1,j}\,(\y\,a)$ we can show that it
+%% is in $\Set_j$.
+%% \end{definition}
+
+Using the above definitions, we can model the syntactic category of 2LTT, and
+also the meta-level family structure and all meta-level type formers. For an
+exposition in previous literature, see \cite{TODO} and \cite{TODO}.
+
+\subsubsection{The object-level fragment}
+
+We move on to modeling the object-level syntactic fragment of 2LTT. We make some
+preliminary definitions. First, note the types in the object theory yield a
+presheaf, and terms yield a displayed presheaf over them; this immediately
+follows from the specification of a family structure in a cwf. Hence, we do a
+bit of a name overloading, and have $\Ty_{\mbbo j} : \hCon$ and $\Tm_{\mbbo j} : \hTy\,\Ty_{\mbbo j}$.
+
+\begin{definition}
+$\bs{\hTy_{0,j}\,\Gamma : \Set_\omega}$ is defined as $\hSub\,\Gamma\,\Ty_{\mbbo j}$,
+and $\bs{\hTm_{0,j}\,\Gamma\,A : \Set_\omega}$ is defined as $\hTm\,\Gamma\,(\Tm_{\mbbo j}[A])$.
+\end{definition}
+
+For illustration, if $A : \hTy_{0,j}\,\Gamma$, then $A :
+\hSub\,\Gamma\,\Ty_{\mbbo j}$, so $|A| : \{a : \Cono\} \to |\Gamma|\,a \to
+\Ty_{\mbbo j}\,a$. In other words, the action of $A$ on objects maps a semantic
+context to a syntactic object-level type. Likewise, for $t :
+\hTm_{0,j}\,\Gamma\,A$, we have $|t| : (\gamma : |\Gamma|\,a) \to \Tm_{\mbbo
+  j}\,a\,(|A|\,\gamma)$, so we get a syntactic object-level term as output.
+
+Using the above definitions and following \cite{TODO}, we can model all type
+formers in $\hTy_{0,j}$. Intuitively, that is because $\hTy_{0,j}$ and $\hTm_{0,j}$
+return types and terms, so we can reuse the types and terms in the object theory.
+
+\subsubsection{Lifting}
+
+\begin{definition}
+$\bs{\wh{\Lift} A : \hTy_{1,j}\,\Gamma}$ is defined as $\Tm_{\mbbo j}[A]$. With
+  this, we get that $\hTm_{0,j}\,\Gamma\,A$ is equal to
+  $\hTm_{1,j}\,\Gamma\,(\wh{\Lift}\,A)$. Hence, we can define both quoting and
+  splicing as identity functions in the model.
+\end{definition}
+
+\subsection{Staging by Evaluation}
+
+\begin{definition} The \textbf{evaluation morphism, denoted $\bs{\ev}$} is
+a model morphism from the syntax of 2LTT to $\hato$, arising from the initiality
+of the syntax. In other words, $\ev$ is defined by recursion on the syntax and
+strictly preserves all structure.
+\end{definition}
+
+Note that $\wh{\emptycon}$ is defined as the terminal presheaf, which is
+constantly $\top$. This leads us to the following definition.
+
+\begin{definition} \textbf{Closed staging} is defined as follows.
+\begin{alignat*}{4}
+  & \Stage : \Ty_{0,j}\,\emptycon \to \Ty_{\mbbo j}\,\emptycon \hspace{2em} && \Stage : \Tm_{0,j}\,\emptycon\,A \to \Tm_{\mbbo j}\,\emptycon\,(\Stage\,A) \\
+  & \Stage\,A := |\ev\,A|\,\{\emptycon\}\,\tt && \Stage\,t := |\ev\,t|\,\{\emptycon\}\,\tt
+\end{alignat*}
+This is well-typed, since $|\ev\,A|\{\emptycon\} : |\ev\,\emptycon|\,\emptycon \to \Ty_{\mbbo
+  j}\,\emptycon$, so $|\ev\,A|\{\emptycon\} : \top \to \Ty_{\mbbo
+  j}\,\emptycon$, and $|\ev\,t|\,\{\emptycon\} : \top \to \Tm_{\mbbo
+  j}\,\emptycon\,(|\ev\,A|\,\tt)$.
+\end{definition}
+
+What about general (open) staging though? Given $A : \Ty_{0,j}\,\emb{\Gamma}$,
+we get $|\ev\,A|\,\{\Gamma\} : |\emb{\Gamma}|\,\Gamma \to \Ty_{\mbbo j}$. We
+need a semantic context with type $|\emb{\Gamma}|\,\Gamma$, in order to extract
+an object-level type. It turns out that such ``generic'' semantic contexts fall
+out from a proof of stability for $\ev$.
+
+\subsection{Stability and Open Staging}
+
+\subsection{Algorithm Extraction and Efficiency}
+
+\section{Yoneda, Representability and Generativity}
+
+\section{Soundness of Staging}
+
+%% \subsection{Internal Language \& Features}
+%% \subsection{The Logical Relation}
+%% \subsection{Externalization \& Soundness}
+
+%% \section{Related Work}
+
+%% Igarashi, Kiselyov et al, MetaML, MetaOCaml, Carette, TH, Scala ppl (?)
+%% PE lit ?
+%% 2LTT, Voevodsky,
+
+%% \section{Future Work \& Conclusions}
 

### 2022-02-14 paper/paper.tex
stuff

@@
 
+Stability means that staging an object-level construction does nothing. However,
+it turns out that the induction motive for contexts is necessarily the existence
+of generic semantic contexts. We define a $\blank^P$ family of functions by
+induction on object syntax. The induction motives are as follows.
+\begin{alignat*}{4}
+  & (\Gamma : \Cono)^P                    &&: |\ev\,\emb{\Gamma}|\,\Gamma\\
+  & (\sigma : \Subo\,\Gamma\,\Delta)^P    &&: \Delta^P[\sigma] = |\ev\,\emb{\sigma}|\,\Gamma^P\\
+  & (A      : \Ty_{\mbbo j}\,\Gamma)^P      &&: A = |\ev\,\emb{A}|\,\Gamma^P\\
+  & (t      : \Tm_{\mbbo j}\,\Gamma\,A)^P   &&: t = |\ev\,\emb{t}|\,\Gamma^P
+\end{alignat*}
+The $\blank^P$ interpretation of all structure is straightforward. In
+particular, we do not have to show preservation of any definitional equality,
+since types, terms and substitutions are all interpreted as proof-irrelevant
+equations.
+
+TODO example
+
+\begin{definition} We define \textbf{open staging} as follows.
+\begin{alignat*}{4}
+  & \Stage : \Ty_{0,j}\,\emb{\Gamma} \to \Ty_{\mbbo j}\,\Gamma \hspace{2em} && \Stage : \Tm_{0,j}\,\emb{\Gamma}\,A \to \Tm_{\mbbo j}\,\Gamma\,(\Stage\,A) \\
+  & \Stage\,A := |\ev\,A|\,\Gamma^P && \Stage\,t := |\ev\,t|\,\Gamma^P
+\end{alignat*}
+\end{definition}
+
+\begin{theorem} Open staging is stable.
+\end{theorem}
+\begin{proof} For $A : \Ty_{\mbbo j}$, $\Stage\,\emb{A}$ is by definition $|\ev\,\emb{A}|\,\Gamma^P$,
+  hence by $A^P$ it is equal to $A$. Likewise, $\Stage\,\emb{t}$ is equal to $t$ by $t^P$.
+\end{proof}
+
+%% Δᴾ[σ] = |F⌜σ⌝| Γᴾ
+%% (A : TyB Γ)ᴾ    : A = |F⌜A⌝| Γᴾ
+%% (t : TmB Γ A)ᴾ  : t = |F⌜t⌝| Γᴾ
+
 \subsection{Algorithm Extraction and Efficiency}

### 2022-02-16 paper/paper.tex
paper progress

@@
 \newcommand{\ev}{\mbb{E}}
+\newcommand{\re}{\mbb{R}}
 
@@
 \newtheorem{notation}{Notation}
+
+\newcommand{\whset}{\wh{\Set}}
+\newcommand{\rexti}{\re_{\ext_1}^{-1}}
 
@@
   manner. In this work, we observe that two-level type theory (2LTT), a system
-  originally devised for the purpose of synthetic homotopy theory, also serves
-  as a system for staged compilation with dependent types. 2LTT has numerous good
-  properties for this use case: it has a concise specification, well-developed
-  algebraic and categorical model theory, and it supports a wide range of
-  language features both at the object and the meta level. First, we give an
-  overview of 2LTT's features and applications in staging. Then, we present a
-  staging algorithm and provide a proof of correctness. Our algorithm is
-  ``staging-by-evaluation'', analogously to the technique of
+  originally devised for the purpose of developing synthetic homotopy theory,
+  also serves as a system for staged compilation with dependent types. 2LTT has
+  numerous good properties for this use case: it has a concise specification,
+  well-developed algebraic and categorical model theory, and it supports a wide
+  range of language features both at the object and the meta level. First, we
+  give an overview of 2LTT's features and applications in staging. Then, we
+  present a staging algorithm and provide a proof of correctness. Our algorithm
+  is ``staging-by-evaluation'', analogously to the technique of
   normalization-by-evaluation, in that staging is given by the evaluation of
@@
 At this point, our system is rather limited, since there is no interaction
-between the stages. We enable such interaction through the following three
-operations.
+between the stages. We enable such interaction through the following operations.
 \begin{itemize}
@@
   staging point of view, $\Lift A$ is the type of metaprograms which compute
-  to runtime expressions of type $A$.
+  runtime expressions of type $A$.
 \item \emph{Quoting:} for $A : \U_0$ and $t : A$, we have $\qut{t} :\,\Lift A$.
-  A quoted term $\qut{t}$ represents the metaprogram which immediately computes
-  to $t$.
+  A quoted term $\qut{t}$ represents the metaprogram which immediately yields
+  $t$.
 \item \emph{Splicing:} for $A : \U_0$ and $t :\,\Lift A$, we have $\spl t : A$.
@@
 with the resulting runtime expressions. The rules of 2LTT ensure that this is possible,
-and we always get a splice-free runtime program after staging.
+and we always get a splice-free object program after staging.
 
@@
 \end{alignat*}
-This is a nice example for the usage of \emph{partially static
+This is an example for the usage of \emph{partially static
 data}\cite{TODO}: semantic contexts are \emph{static} lists storing
@@
 By rewriting a 2LTT program left-to-right along preservation maps, we perform
-what is termed a \emph{binding time improvement} in the partial evaluation
+what is termed \emph{binding time improvement} in the partial evaluation
 literature \cite{TODO}. Note that the output of $\mit{pres}_{\to}$ uses a
@@
 $((A : \Ty_0) \to \Lift\,\spl A \to \Lift \spl A) \to \Lift\,\Bool_0$. This
-function can be only applied to statically known arguments in runtime
-code, so the polymorphism can be staged away. The main restriction that
-programmers have to keep in mind is that polymorphic functions cannot be stored
-inside runtime data types.
+function can be only applied to statically known arguments, so the polymorphism
+can be staged away. The main restriction that programmers have to keep in mind
+is that polymorphic functions cannot be stored inside runtime data types.
 
@@
 this would be a more advanced variation on type-passing polymorphism
-\cite{TODO}, where layouts of objects may depend on runtime data. For
-example, length-prefixed flat arrays would not be a primitive notion, but simply
-defined as $(n : \Nat) \times \msf{Vec}\,n\,A$. However, it seems to be highly
-challenging to implement runtime systems and precise garbage collection
-generically over dependently typed memory layouts. See \texttt{sixten}
-\cite{TODO} for an experimental implementation with conservative garbage
-collection.
+\cite{TODO}, where layouts of objects may depend on runtime data. For example,
+length-prefixed flat arrays would not be a primitive notion, but simply defined
+as $(n : \Nat) \times \msf{Vec}\,n\,A$. However, it seems to be highly
+challenging to implement runtime systems and precise garbage collection in this
+setting. See \texttt{sixten} \cite{TODO} for an experimental implementation of
+dependent memory layouts.
 
@@
 \item
-  By moving right-to-left along $\Sub\,\Gamma\,(\Delta\ext A) \simeq
-  ((\sigma : \Sub\,\Gamma\,\Delta) \times \Tm_{i,j}\,\Gamma\,(A[\sigma]))$, we
-  recover the operation which extends a morphism with a term. In the syntax
-  (initial model), this justifies the view of $\Sub$ as a list of terms, i.e.\ a
-  parallel substitution. We denote the extension operation as $(\sigma,\,t)$ for
-  $\sigma : \Sub\,\Gamma\,\Delta$.
+  By moving right-to-left along $\Sub\,\Gamma\,(\Delta\ext A) \simeq ((\sigma :
+  \Sub\,\Gamma\,\Delta) \times \Tm_{i,j}\,\Gamma\,(A[\sigma]))$, we recover the
+  operation which extends a morphism with a term. In the syntax (initial model),
+  this justifies the view of $\Sub$ as a type of lists of terms, i.e.\ a
+  parallel substitutions. We denote the extension operation as $(\sigma,\,t)$
+  for $\sigma : \Sub\,\Gamma\,\Delta$.
 \end{itemize}
@@
   \end{alignat*}
-  Note that we can eliminate from a $j$ level to any $k$ level.
+  Note that we can eliminate from a level $j$ to any level $k$.
 \end{itemize}
@@
 type formers $\Pi$, $\Sigma$ and $\Nat_j$, with elimination from $\Nat_j$ to
-any $k$ level.
+any level $k$.
 \end{definition}
@@
 Second, note that soundness and stability together is the statement that
-embedding is invertible on types and terms. This is a statement of
+embedding is a bijection on types and terms. This is a statement of
 \emph{conservativity} of 2LTT over the object theory. In \cite{TODO} a
@@
 \begin{definition} $\bs{\wh{\Con} : \Set_{\omega+1}}$ is defined as the set of presheaves
-over $\mbbo$. A $\Gamma : \widehat{\Con}$ has an action on objects $|\Gamma| :
+over $\mbbo$. $\Gamma : \wh{\Con}$ has an action on objects $|\Gamma| :
 \Cono \to \Set_\omega$ and an action on morphisms $\blank[\blank] : |\Gamma|\,b
@@
 \begin{definition} $\bs{\hSub\,\Gamma\,\Delta : \Set_\omega}$ is the set of natural transformations
-from $\Gamma$ to $\Delta$. A $\sigma : \hSub\,\Gamma\,\Delta$, has an action on
-objects $|\sigma| : \{a : \Cono\} \to |\Gamma|\,a \to |\Delta|\,a$ such that
+from $\Gamma$ to $\Delta$. $\sigma : \hSub\,\Gamma\,\Delta$ has action
+$|\sigma| : \{a : \Cono\} \to |\Gamma|\,a \to |\Delta|\,a$ such that
 $|\sigma|\,(\gamma[f]) = (|\sigma|\,\gamma)[f]$.
@@
 
-\begin{notation} We write $\alpha$ and $\beta$ respectively for elements of $|A|\,\gamma$ and $|B|\,\gamma$.
+\begin{notation}
+  We write $\alpha$ and $\beta$ respectively for elements of $|A|\,\gamma$ and
+  $|B|\,\gamma$.
 \end{notation}
@@
   $\bs{\hTm_{1,j}\,\Gamma\,A : \Set_\omega}$ is the set of sections of the
-  displayed presheaf $A$. This can be viewed as a dependently typed analogue of a natural transformation.
-  A $t : \hTm_{1,j}\,\Gamma\,A$ has an action on objects $|t| : \{a\} \to (\gamma : |\Gamma|\,a) \to |A|\,\gamma$,
-  such that $|t|\,(\gamma[f]) = (|t|\,\gamma)[f]$.
+  displayed presheaf $A$. This can be viewed as a dependently typed analogue of
+  a natural transformation.  A $t : \hTm_{1,j}\,\Gamma\,A$ has action $|t| :
+  \{a\} \to (\gamma : |\Gamma|\,a) \to |A|\,\gamma$ such that $|t|\,(\gamma[f])
+  = (|t|\,\gamma)[f]$.
 \end{definition}
@@
 
+We also look at he empty context and context extension with meta-level types,
+as these will appear in subsequent definitions.
+
+\begin{definition}
+$\bs{{\widehat{\emptycon}} : \hCon}$ is defined as the presheaf
+which is constantly $\top$, i.e.\ $|\wh{\emptycon}|\,\_ = \top$.
+\end{definition}
+
+\begin{definition} For $A : \hTy_{1,j}\,\Gamma$, we define $\bs{\Gamma\,\,\wh{\ext}\,A}$
+pointwise by $|\Gamma\,\,\wh{\ext}\,A|\,a := (\gamma : |\Gamma|\,a) \times |A|\,\gamma$
+and $(\gamma,\,\alpha)[f] := (\gamma[f],\,\alpha[f])$.
+\end{definition}
+
 Using the above definitions, we can model the syntactic category of 2LTT, and
@@
 
+\begin{definition} For $A : \Ty_{0,j}\,\Gamma$, we define $\bs{\Gamma\,\,\wh{\ext}\,A}$
+as $|\Gamma\,\,\wh{\ext}\,A|\,a := (\gamma : |\Gamma|\,a) \times \Tm_{\mbbo
+  j}\,a\,(|A|\,\gamma)$ and $(\gamma,\,t)[f] := (\gamma[f],\,t[f])$. Thus,
+extending a semantic context with an object-level type appends an object-level
+term.
+\end{definition}
+
 Using the above definitions and following \cite{TODO}, we can model all type
-formers in $\hTy_{0,j}$. Intuitively, that is because $\hTy_{0,j}$ and $\hTm_{0,j}$
-return types and terms, so we can reuse the types and terms in the object theory.
+formers in $\hTy_{0,j}$. Intuitively, that is because $\hTy_{0,j}$ and
+$\hTm_{0,j}$ return types and terms, so we can reuse the type and term formers
+in the object theory.
 
@@
 
-\subsection{Staging by Evaluation}
+\subsection{Closed Staging}
 
 \begin{definition} The \textbf{evaluation morphism, denoted $\bs{\ev}$} is
-a model morphism from the syntax of 2LTT to $\hato$, arising from the initiality
+a 2LTT model morphism from the syntax of 2LTT to $\hato$, arising from the initiality
 of the syntax. In other words, $\ev$ is defined by recursion on the syntax and
@@
 we get $|\ev\,A|\,\{\Gamma\} : |\emb{\Gamma}|\,\Gamma \to \Ty_{\mbbo j}$. We
-need a semantic context with type $|\emb{\Gamma}|\,\Gamma$, in order to extract
-an object-level type. It turns out that such ``generic'' semantic contexts fall
-out from a proof of stability for $\ev$.
+need an element of $|\emb{\Gamma}|\,\Gamma$, in order to extract an object-level
+type. Such ``generic'' semantic contexts should be possible to construct, because
+elements of $|\emb{\Gamma}|\,\Gamma$ are essentially lists of object-level terms
+in $\Gamma$, by Definition \ref{TODO}, so $|\emb{\Gamma}|\,\Gamma$ should be
+isomorphic to $\Subo\,\Gamma\,\Gamma$.  It turns out that this falls out from
+the stability proof of $\ev$.
 
@@
 
-Stability means that staging an object-level construction does nothing. However,
-it turns out that the induction motive for contexts is necessarily the existence
-of generic semantic contexts. We define a $\blank^P$ family of functions by
-induction on object syntax. The induction motives are as follows.
+Stability means that staging an object-level construction does nothing. We
+define a family of functions $\blank^P$ by induction on object syntax. The
+induction motives are as follows.
 \begin{alignat*}{4}
@@
 \end{alignat*}
-The $\blank^P$ interpretation of all structure is straightforward. In
-particular, we do not have to show preservation of any definitional equality,
-since types, terms and substitutions are all interpreted as proof-irrelevant
-equations.
-
-TODO example
+We look at the interpretation of contexts.
+\begin{itemize}
+\item  For $\emptycon^P$, we need an element of $|\ev\,\emb{\emptycon}|\,\emptycon$, hence
+  an element of $\top$, so we define $\emptycon^P$ as $\tt$.
+\item
+  For $(\Gamma \ext A)^P$, we need an element of
+    \[(\gamma : |\ev{\emb{\Gamma}}|\,(\Gamma\ext A)) \times \Tm_{\mbbo j}\,(\Gamma \ext A)\,(|\ev\emb{A}|\,\gamma).\]
+  We have $\Gamma^P : |\ev{\emb{\Gamma}}|\,\Gamma$, which we can weaken as
+  $\Gamma^P[\p] : |\ev{\emb{\Gamma}}|\,(\Gamma \ext A)$, so we set the first
+  projection of the result as $\Gamma^P[\p]$. For the second projection, the
+  goal type can be simplified as follows:
+  \begin{alignat*}{4}
+    &       &&\Tm_{\mbbo j}\,(\Gamma \ext A)\,(|\ev\emb{A}|\,(\Gamma^P[\p])) && \\
+    & =\,\, &&\Tm_{\mbbo j}\,(\Gamma \ext A)\,((|\ev\emb{A}|\,\Gamma^P)[\p]) &&\hspace{2em} \text{by naturality of $\ev\emb{A}$}\\
+    & =\,\, &&\Tm_{\mbbo j}\,(\Gamma \ext A)\,(A[\p])                        &&\hspace{2em} \text{by $A^P$}
+  \end{alignat*}
+  We have the zero de Bruijn variable $\q : \Tm_{\mbbo j}\,(\Gamma \ext
+  A)\,(A[\p])$.  Hence, we define $(\Gamma \ext A)^P$ as $(\Gamma^P[\p],\,\q)$.
+\end{itemize}
+
+Thus, a generic semantic context $\Gamma^P : |\ev\emb{\Gamma}|\,\Gamma$ is just
+a list of variables, corresponding to the identity substitution $\id :
+\Subo\,\Gamma\,\Gamma$ which maps each variable to itself.
+
+The rest of the $\blank^P$ interpretation is straightforward and we omit it
+here. In particular, we do not have to show preservation of any definitional
+equality, since types, terms and substitutions are all interpreted as
+proof-irrelevant equations.
 
@@
 
-%% Δᴾ[σ] = |F⌜σ⌝| Γᴾ
-%% (A : TyB Γ)ᴾ    : A = |F⌜A⌝| Γᴾ
-%% (t : TmB Γ A)ᴾ  : t = |F⌜t⌝| Γᴾ
-
-\subsection{Algorithm Extraction and Efficiency}
+\subsection{Implementation and Efficiency}
+
+We have defined staging in a constructive metatheory, so it is possible to
+extract an implementation from it. We discuss its operational behavior and
+potential optimizations in the following. First, consider the computational
+part of evaluation:
+\begin{alignat*}{3}
+  & |\blank| \circ \ev : \Ty_{1 j}\,\Gamma \to \{\Delta : \Cono\} \to |\ev\,\Gamma|\,\Delta \to \Set_j\\
+  & |\blank| \circ \ev : \Ty_{0 j}\,\Gamma \to \{\Delta : \Cono\} \to |\ev\,\Gamma|\,\Delta \to \Ty_{\mbbo j}\,\Delta\\
+  & |\blank| \circ \ev : \Tm_{1 j}\,\Gamma\,A \to \{\Delta : \Cono\} \to (\gamma : |\ev\,\Gamma|\,\Delta) \to
+    |\ev\,A|\,\gamma \\
+  & |\blank| \circ \ev : \Tm_{0 j}\,\Gamma\,A \to \{\Delta : \Cono\} \to (\gamma : |\ev\,\Gamma|\,\Delta) \to
+    \Tm_{\mbbo\,j}\,\Delta\,(|\ev\,A|\,\gamma)
+\end{alignat*}
+Operationally, we interpret syntactic types or terms in a semantic environment
+$\gamma : |\ev\,\Gamma|\,\Delta$. These environments are lists containing
+object-level terms and semantic values. The meta-level semantic values are
+represented using meta-level inductive types and the function space of the
+metatheory.
+\begin{itemize}
+  \item $|\Nat_1|\,\_$ is simply the set of meta-level natural numbers.
+  \item $|\Sigma_1\,A\,B|\,\gamma$ is simply a set of pairs of values.
+  \item In the non-dependent case, a function type $A\,\wh{\to}\,B$ is defined as the presheaf
+        exponential \cite{TODO}. The computational part of $|A\,\wh{\to}\,B|\,\{\Delta\}\,\gamma$
+        is given by a function with type
+        \[ (\Theta : \Cono) \to (\sigma : \Subo\,\Theta\,\Delta) \to |A|\,(\gamma[\sigma]) \to |B|\,(\gamma[\sigma])        \]
+        This may be also familiar as the semantic implication from the Kripke
+        semantics of intuitionistic logics. Whenever we evaluate a function
+        application, we supply an extra $\id : \Subo\,\Delta\,\Delta$. This may
+        incur cost via the $\gamma[\id]$ restrictions in a naive
+        implementation, but this is easy to optimize, by introducing a formal
+        representation of $\id$, such that $\gamma[\id]$ immediately computes to
+        $\gamma$. The dependent case is analogous operationally.
+\end{itemize}
+In summary, in the meta-level fragment of 2LTT, $\ev$ yields a reasonably efficient
+computation of closed values, which reuses functions from the ambient
+metatheory. Alternatively, instead of using ambient functions, we could use our
+own implementation of function closures during staging.
+
+In contrast, we have a bit of an efficiency problem in the handling of object-level
+terms and binders: whenever we go under an object-level binder we have to weaken
+the current semantic context. In other words, when moving from $\Delta$ to
+$\Delta \ext A$, we have to shift $\gamma : |\ev\,\Gamma|\,\Delta$ to
+$\gamma[\p] : |\ev\,\Gamma|\,(\Delta \ext A)$. This cannot be avoided by an easy
+optimization trick.
+
+The exact same efficiency issue occurs in formal presheaf-based
+normalization-by-evaluation \cite{TODO}. However, in practical implementations
+\cite{TODO} this issue can be fully solved by using De Bruijn \emph{levels} in
+the semantic domain, thus arranging that weakening has no operational cost on
+semantic values. This corresponds to Abel's ``liftable terms'' presentation
+\cite{TODO}. We can use the same solution in staging, by using de Bruijn levels
+and delayed weakenings in object-level terms during staging. It is a somewhat
+unfortunate mismatch that indices are far more convenient in formalization, but
+levels are more efficient in practice.
+
+In our prototype implementation, we use the above optimization with De Bruijn
+levels, and we also switch to \emph{syntax-directed} staging. In the formal
+staging algorithm, since types are tracked, quotes and splices have no
+operational significance; recall that they are interpreted as identity functions
+in Definition \ref{TODO}. However, it seems more efficient to not compute types
+at all and instead rely on quotes and splices to keep track of the current
+stage.
+
+\subsubsection{Caching} In a production-strength implementation we would need some
+caching mechanism, to avoid excessive duplication of function code. For example,
+if we use the staged $\mit{map}$ function multiple times with the same type and
+function arguments, we do not want to generate separate code for each
+usage. De-duplicating object-level code with \emph{function types} is
+usually safe, since function bodies become closed top-level code after closure
+conversion. We leave this to future work.
 
@@
 
-%% \subsection{Internal Language \& Features}
-%% \subsection{The Logical Relation}
-%% \subsection{Externalization \& Soundness}
-
-%% \section{Related Work}
-
-%% Igarashi, Kiselyov et al, MetaML, MetaOCaml, Carette, TH, Scala ppl (?)
-%% PE lit ?
-%% 2LTT, Voevodsky,
-
-%% \section{Future Work \& Conclusions}
+In this section we prove soundness of staging. We build a proof-relevant logical
+relation between the evaluation morphism $\ev$ and a \emph{restriction}
+morphism, which restricts 2LTT syntax to object-level contexts. The relational
+interpretations of $\Ty_{0,j}$ and $\Tm_{0,j}$ will yield the soundness
+property.
+
+\subsection{Working in $\hat{\mbbo}$}
+
+We have seen that 2LTT can be modeled in the presheaf topos
+$\hato$. Additionally, $\hato$ supports all type formers of extensional type
+theory \cite{TODO}, and certain other structures which are stable under
+object-level substitution. Since almost all constructions in this section must
+be stable under object-level substitution, it makes sense to work internally to
+$\hato$. This technique has been previously used in normalization proofs
+\cite{TODO} and also in the metatheory of cubical type theories \cite{TODO}.
+
+When we work internally in a model of a type theory, we do not explicitly refer
+to contexts, types, and substitutions. For example, when working in Agda, we do
+not refer to Agda's typing contexts. Instead, we only work with terms, and use
+functions and universes to abstract over types and semantic contexts. Hence, we
+have to convert along certain isomorphisms when we switch between the internal
+and external views. In the following, we summarize features in $\hato$ and also
+the internal-external conversions.
+
+We write $\whset_j$ for ordinal-indexed Russell-style universes. Formally, we
+have Coquand-style universes, but for the sake of brevity we omit $\El$ and
+$\Code$ from internal syntax. Universes are cumulative, and closed under $\Pi$,
+$\Sigma$, extensional identity $\blank\!=\!\blank$ and inductive types. We use
+the same conventions as in Notation \ref{TODO}.
+
+\noindent The basic scheme for internalization is as follows:
+\begin{alignat*}{3}
+  & \Gamma : \hCon                  && \hspace{1em}\text{is internalized as}\hspace{1em} \Gamma : \whset_\omega\\
+  & \sigma : \hSub\,\Gamma\,\Delta  && \hspace{1em}\text{is internalized as}\hspace{1em} \sigma : \Gamma \to \Delta\\
+  & A : \hTy_{1,j}\,\Gamma           && \hspace{1em}\text{is internalized as}\hspace{1em} A      : \Gamma \to \whset_j \\
+  & t : \hTm_{1,j}\,\Gamma\,A        && \hspace{1em}\text{is internalized as}\hspace{1em} t      : (\gamma : \Gamma) \to A\,\gamma
+\end{alignat*}
+We internalize object-level types and terms as $\Ty_{\mbbo j} : \whset_0$ and
+$\Tm_{\mbbo j} : \Ty_{\mbbo j} \to \whset_0$.  We also use an internal view of
+$\ev$ which maps 2LTT syntax to internal values; i.e.\ we compose $\ev$ with
+internalization. Below, note that the input to $\ev$ is \emph{external}, so we
+mark $\ev$ as being parametrized by external input.
+\begin{alignat*}{3}
+  &\ev\,(\Gamma &&: \Con)                      &&: \whset_\omega \\
+  &\ev\,(\sigma &&: \Sub\,\Gamma\,\Delta)      &&: \ev\,\Gamma \to \ev\,\Delta \\
+  &\ev\,(A      &&: \Ty_{1,j}\,\Gamma)          &&: \ev\,\Gamma \to \whset_j\\
+  &\ev\,(t      &&: \Tm_{1,j}\,\Gamma)          &&: (\gamma : \ev\,\Gamma) \to \ev\,A\,\gamma\\
+  &\ev\,(A      &&: \Ty_{0,j}\,\Gamma)          &&: \ev\,\Gamma \to \Ty_{\mbbo_j}\\
+  &\ev\,(t      &&: \Tm_{0,j}\,\Gamma\,A)       &&: (\gamma : \ev\,\Gamma) \to \Tm_{\mbbo j}\,(\ev\,A\,\gamma)
+\end{alignat*}
+The purely object-level syntactic fragment of 2LTT can be internalized as
+follows. We define externally the presheaf of object-level types as
+$|\Ty_{0,j}|\,a := \Ty_{0,j}\,\emb{a}$, and the displayed presheaf of
+object-level terms over $\Ty_{0,j}$ as $|\Tm_{0,j}|\,\{a\}\,A :=
+  \Tm_{0,j}\,\emb{a}\,A$. Hence, internally we have $\Ty_{0,j} : \whset_0$ and
+  $\Tm_{0,j} : \Ty_{0,j} \to \whset_0$.
+
+Now, the embedding operation $\emb{\blank}$ can be also internalized on types
+and terms, as $\emb{\blank} : \Ty_{\mbbo j} \to \Ty_{0,j}$, and $\emb{\blank}
+:\Tm_{\mbbo\,j}\,A \to \Tm_{0,j}\,\emb{A}$.
+
+\subsection{The Restriction Morphism}
+
+We define a family of functions $\re$ from the 2LTT syntax to objects in
+$\hato$. This will be related to the evaluation morphism $\ev$ subsequently, in
+the relational interpretation. In short, $\re$ restricts 2LTT syntax so that it
+can only depend on object-level contexts, i.e.\ contexts given as
+$\emb{\Gamma}$. Below, we specify the types of the $\re$ operations internally,
+and the $|\blank|$ components of the operations externally. The naturality of
+$|\blank|$ is straightforward in each case.
+\begin{alignat*}{5}
+  & \re\,(\Gamma : \Con) : \whset_0 && \re\,(\sigma : \Sub\,\Gamma\,\Delta) : \re\,\Gamma \to \re\,\Delta \\
+  & |\re\,\Gamma|\,a := \Sub\,\emb{a}\,\Gamma && |\re\,\sigma|\,\gamma := \sigma \circ \gamma \\
+  & \\
+  & \re\,(A : \Ty_{1,j}\,\Gamma) : \re\,\gamma \to \whset_0 && \re\,(t : \Tm_{1,j}\,\Gamma\,A) : (\gamma : \re\,\Gamma) \to \re\,A\,\gamma \\
+  & |\re\,A|\,\{a\}\,\gamma := \Tm_{1,j}\,\emb{a}\,(A[\gamma])\hspace{2em} && |\re\,t|\,\gamma := t[\gamma]\\
+  & \\
+  & \ev\,(A : \Ty_{0,j}\,\Gamma) : \re\,\Gamma \to \Ty_{0,j}\hspace{2em} && \ev\,(t : \Tm_{0,j}\,\Gamma\,A) : (\gamma : \re\,\Gamma) \to \Tm_{0,j}\,(\re\,A\,\gamma)\\
+  & |\ev\,A|\,\gamma := A[\gamma] && |\ev\,t|\,\gamma := t[\gamma]
+\end{alignat*}
+
+\subsubsection{Preservation properties of $\re$}
+First, we note that $\re$ strictly preserves $\id$, $\blank\circ\blank$ and
+type/term substitution, and it preserves $\emptycon$ and $\blank\ext_1\blank$ up
+to isomorphism. We have the following isomorphisms internally to $\hato$:
+\begin{alignat*}{3}
+  & \re_{\emptycon} &&: \re\,\emptycon \simeq \top \\
+  & \re_{\ext_1}   &&: \re\,(\Gamma \ext_1 A) \simeq ((\gamma : \re\,\Gamma) \times \re\,A\,\gamma)
+\end{alignat*}
+\begin{notation}
+  When we have an isomorphism $f : A \simeq B$, we may write $f$ for the
+  function in $A \to B$, and $f^{-1} : B \to A$ for its inverse.
+\end{notation}
+The above preservation properties mean that $\re$ is a \emph{pseudomorphism} in
+the sense of \cite{gluing}, between the syntactic cwf given by
+$(\Ty_{1,j},\,\Tm_{1,j})$ and the corresponding cwf structure in $\hato$. In
+ibid.\ there is an analysis of such cwf morphisms, from which we obtain the
+following additional preservation properties:
+\begin{itemize}
+\item Meta-level negative type formers are preserved up to isomorphism. In particular, we have
+  \[ \re_{\Sigma} : \re\,(\Sigma\,A\,B)\,\gamma
+  \simeq ((\alpha : \re\,A) \times \re\,B\,(\rexti\,(\gamma,\,\alpha))). \]
+  To provide a higher-level explanation, $\re\,(\Sigma\,A\,B)\,\gamma$ is
+  essentially the presheaf of 2LTT terms with type $\Sigma_1\,A\,B$, but the
+  terms are restricted to object-level contexts. However, we can still perform
+  pairing and projection on such restricted terms; hence the preservation property.
+
+\item Meta-level $\Pi$ types and universes are preserved in a lax way. For $\Pi$, we have
+  \[ \re_{\app} : \re\,(\Pi\,A\,B)\,\gamma \to (\alpha : \re\,A\,\gamma) \to \re\,B\,(\rexti\,(\gamma,\,\alpha))
+  \]
+  such that $\re_{\app}\,(\re\,(\lam\,t)\,\gamma)\,\alpha =
+  \re\,t\,(\rexti\,(\gamma,\,\alpha))$. In this case, we can apply a restricted
+  term with a $\Pi$-type to a restricted term, but we cannot do
+  lambda-abstraction, because that would require extending the context with a
+  meta-level binder.  For $\U_{1,j}$, we have
+  \[
+    \re_{\Code} : \re\,\U_{1,j}\,\gamma \to \whset_j
+  \]
+  such that $\re_{\Code}\,(\re\,(\El\,t)\,\gamma) = \re\,t\,\gamma$. We skip
+  $\Code$ on the right hand side here, since we assume Russell-style notation
+  internally to $\hato$, i.e.\ we omit $\Code$ and $\El$. Here, we only
+  have lax preservation simply because $\whset_j$ is much larger than
+  the the set of syntactic 2LTT types, so not every semantic $\whset_j$
+  have a syntactic representation.
+
+\item Meta-level positive (inductive) types are preserved in an oplax way. In the case of
+  natural numbers, we have
+  \[ \re_{\mbb{N}} : \mbb{N} \to \re\,\Nat_1\,\gamma. \]
+  This is a ``serialization'' map: from a metatheoretic natural number we
+  compute a numeral as a closed 2LTT term. This works analogously for any
+  inductive type. The restriction to object-level contexts is irrelevant when
+  computing closed terms.
+\end{itemize}
+\paragraph{Action on lifting} We have an isomorphism $\re\,(\Lift\,A)\,\gamma \simeq
+\Tm_{0,j}\,(\re\,A\,\gamma)$. This is given by quoting and splicing: we convert
+between restricted meta-level terms with type $\Lift\,A$ and restricted
+object-level terms with type $A$. Hence, we simply write the components of this
+isomorphism as $\spl\blank$ and $\qut{\blank}$, as internal analogues of the
+external operations. With this, we also have $\re\,\qut{t}\,\gamma =
+\qut{\re\,t\,\gamma}$ and $\re\,\spl\,t\,\gamma = \spl(\re\,t\,\gamma)$.
+
+\paragraph{Action on $\blank\ext_0\blank$} Lastly, $\blank\ext_0\blank$ is preserved
+up to isomorphism:
+\[
+  \re_{\ext_0} : \re\,(\Gamma \ext_0 A) \simeq ((\gamma : \re\,\Gamma)\times \Tm_{0,j}\,(\re\,A\,\gamma))
+\]
+This is because substitutions targeting $\Gamma \ext_0 A$ are the same as
+pairs of substitutions and terms, by the specification of $\blank\ext_0\blank$.
+
+\subsection{The Logical Relation}
+
+Internally to $\hato$, we define by induction on the syntax of 2LTT a
+proof-relevant logical relation interpretation, called $\blank^R$. The induction
+motives are specified as follows.
+\begin{alignat*}{5}
+  &(\Gamma &&: \Con)^R && : \ev\,\Gamma \to \re\,\Gamma \to \whset_\omega\\
+  &(\sigma &&: \Sub\,\Gamma\,\Delta)^R &&: \Gamma^R\,\gamma\,\gamma' \to \Delta^R\,(\ev\,\sigma\,\gamma)\,(\re\,\sigma\,\gamma')\\
+  &(A      &&: \Ty_{1,j}\,\Gamma)^R &&: \Gamma^R\,\gamma\,\gamma' \to \ev\,A\,\gamma \to \re\,A\,\gamma' \to \whset_j\\
+  &(t      &&: \Tm_{1,j}\,\Gamma\,A)^R &&: (\gamma^R : \Gamma^R\,\gamma\,\gamma') \to A^R\,\gamma^R\,(\ev\,t\,\gamma)\,(\re\,t\,\gamma')\\
+  &(A      &&: \Ty_{0,j}\,\Gamma)^R &&: \Gamma^R\,\gamma\,\gamma' \to \emb{\ev\,A\,\gamma} = \re\,A\,\gamma' \\
+  &(t      &&: \Tm_{0,j}\,\Gamma\,A)^R &&: \Gamma^R\,\gamma\,\gamma' \to \emb{\ev\,t\,\gamma} = \re\,t\,\gamma'
+\end{alignat*}
+For $\Con$, $\Sub$, and meta-level types and terms, this is a fairly standard
+logical relation interpretation: contexts are mapped to relations, types to
+dependent relations, and substitutions and terms respect relations. For
+object-level types and terms, we get soundness statements: evaluation via $\ev$
+followed by embedding back to 2LTT is the same as restricting 2LTT syntax to
+object-level contexts.
 

### 2022-02-20 paper/paper.tex
finish soundness

@@
 \documentclass[acmsmall]{acmart}
-%% \raggedbottom
+\raggedbottom
 
@@
 \newcommand{\rexti}{\re_{\ext_1}^{-1}}
+
+\newcommand{\rel}{^{\approx}}
 
@@
 
-In this section we specify what we mean by staging, then define a staging
-algorithm.
+In this section we specify what we mean by a staging algorithm, then proceed to
+define one.
 
@@
 terms in the model. In this model, contexts are presheaves over the syntactic
-category of the object theory. We call the model $\hato$ and denote its
-components by putting hats on 2LTT components. We summarize the key components
-in the following.
+category of the object theory. We denote the components of the model by putting
+hats on 2LTT components, e.g.\ as in $\hCon$. We summarize the key components in
+the following.
 
@@
 $\gamma[f\circ g] = \gamma[f][g]$.
+
 \begin{notation}
-We reuse the substitution notation $\blank[\blank]$ for the action on morphisms.
-Also, we use lowercase $\gamma$ and $\delta$ to denote elements of $|\Gamma|\,a$
-and $|\Delta|\,a$ respectively.
+Above, we reused the substitution notation $\blank[\blank]$ for the action on
+morphisms.  Also, we use lowercase $\gamma$ and $\delta$ to denote elements of
+$|\Gamma|\,a$ and $|\Delta|\,a$ respectively.
 \end{notation}
+
 \end{definition}
@@
 The rest of the $\blank^P$ interpretation is straightforward and we omit it
-here. In particular, we do not have to show preservation of any definitional
-equality, since types, terms and substitutions are all interpreted as
-proof-irrelevant equations.
+here. In particular, preservation of definitional equalities is automatic, since
+types, terms and substitutions are all interpreted as proof-irrelevant
+equations.
 
@@
 \end{alignat*}
-We internalize object-level types and terms as $\Ty_{\mbbo j} : \whset_0$ and
-$\Tm_{\mbbo j} : \Ty_{\mbbo j} \to \whset_0$.  We also use an internal view of
+\subsubsection{Object-theoretic syntax} The syntax of the object theory is clearly
+fully stable under object-theoretic substitution, so we can internalize all of
+its type and term formers. We internalize object-level types and terms as
+$\Ty_{\mbbo j} : \whset_0$ and $\Tm_{\mbbo j} : \Ty_{\mbbo j} \to
+\whset_0$. $\Ty_{\mbbo j}$ is closed under type formers. For instance, we have
+\[ \Nat_j : \Ty_{\mbbo j}\hspace{1.5em}\zero_j : \Tm_{\mbbo j}\,\Nat_j\hspace{1.5em}
+    \suc_j : \Tm_{\mbbo j}\,\Nat_j \to \Tm_{\mbbo j}\,\Nat_j \]
+together with $\NatElim$, and likewise we have all other type formers.
+\subsubsection{Internal $\ev$} We also use an internal view of
 $\ev$ which maps 2LTT syntax to internal values; i.e.\ we compose $\ev$ with
@@
 \end{alignat*}
+\subsubsection{Object-level fragment of 2LTT}
 The purely object-level syntactic fragment of 2LTT can be internalized as
@@
 \Tm_{0,j}\,\emb{a}\,A$. Hence, internally we have $\Ty_{0,j} : \whset_0$ and
-  $\Tm_{0,j} : \Ty_{0,j} \to \whset_0$.
-
+$\Tm_{0,j} : \Ty_{0,j} \to \whset_0$. $\Ty_{0,j}$ is closed under all type formers,
+analogously as we have seen for $\Ty_{\mbbo j}$.
+
+\subsubsection{Embedding}
 Now, the embedding operation $\emb{\blank}$ can be also internalized on types
 and terms, as $\emb{\blank} : \Ty_{\mbbo j} \to \Ty_{0,j}$, and $\emb{\blank}
-:\Tm_{\mbbo\,j}\,A \to \Tm_{0,j}\,\emb{A}$.
+:\Tm_{\mbbo\,j}\,A \to \Tm_{0,j}\,\emb{A}$. Embedding strictly preserves all structure.
 
@@
 We define a family of functions $\re$ from the 2LTT syntax to objects in
-$\hato$. This will be related to the evaluation morphism $\ev$ subsequently, in
-the relational interpretation. In short, $\re$ restricts 2LTT syntax so that it
-can only depend on object-level contexts, i.e.\ contexts given as
-$\emb{\Gamma}$. Below, we specify the types of the $\re$ operations internally,
-and the $|\blank|$ components of the operations externally. The naturality of
-$|\blank|$ is straightforward in each case.
+$\hato$. This will be related to the evaluation morphism $\ev$ in the relational
+interpretation. In short, $\re$ restricts 2LTT syntax so that it can only depend
+on object-level contexts, i.e.\ contexts given as $\emb{\Gamma}$.
+
+\begin{definition} We specify the types of the \textbf{restriction operations} internally, and the $|\blank|$
+components of the operations externally. The naturality of $|\blank|$ is
+straightforward in each case.
 \begin{alignat*}{5}
   & \re\,(\Gamma : \Con) : \whset_0 && \re\,(\sigma : \Sub\,\Gamma\,\Delta) : \re\,\Gamma \to \re\,\Delta \\
-  & |\re\,\Gamma|\,a := \Sub\,\emb{a}\,\Gamma && |\re\,\sigma|\,\gamma := \sigma \circ \gamma \\
+  & |\re\,\Gamma|\,\Delta := \Sub\,\emb{\Delta}\,\Gamma && |\re\,\sigma|\,\gamma := \sigma \circ \gamma \\
   & \\
   & \re\,(A : \Ty_{1,j}\,\Gamma) : \re\,\gamma \to \whset_0 && \re\,(t : \Tm_{1,j}\,\Gamma\,A) : (\gamma : \re\,\Gamma) \to \re\,A\,\gamma \\
-  & |\re\,A|\,\{a\}\,\gamma := \Tm_{1,j}\,\emb{a}\,(A[\gamma])\hspace{2em} && |\re\,t|\,\gamma := t[\gamma]\\
+  & |\re\,A|\,\{\Delta\}\,\gamma := \Tm_{1,j}\,\emb{\Delta}\,(A[\gamma])\hspace{2em} && |\re\,t|\,\gamma := t[\gamma]\\
   & \\
-  & \ev\,(A : \Ty_{0,j}\,\Gamma) : \re\,\Gamma \to \Ty_{0,j}\hspace{2em} && \ev\,(t : \Tm_{0,j}\,\Gamma\,A) : (\gamma : \re\,\Gamma) \to \Tm_{0,j}\,(\re\,A\,\gamma)\\
-  & |\ev\,A|\,\gamma := A[\gamma] && |\ev\,t|\,\gamma := t[\gamma]
-\end{alignat*}
+  & \re\,(A : \Ty_{0,j}\,\Gamma) : \re\,\Gamma \to \Ty_{0,j}\hspace{2em} && \re\,(t : \Tm_{0,j}\,\Gamma\,A) : (\gamma : \re\,\Gamma) \to \Tm_{0,j}\,(\re\,A\,\gamma)\\
+  & |\re\,A|\,\gamma := A[\gamma] && |\re\,t|\,\gamma := t[\gamma]
+\end{alignat*}
+\end{definition}
 
@@
 \end{alignat*}
-\begin{notation}
-  When we have an isomorphism $f : A \simeq B$, we may write $f$ for the
+
+\begin{notation} When we have an isomorphism $f : A \simeq B$, we may write $f$ for the
   function in $A \to B$, and $f^{-1} : B \to A$ for its inverse.
 \end{notation}
+
+\begin{notation}
+  We can use a pattern matching notation on isomorphisms. For example, if $f : A
+  \simeq B$, then we may write $(\lambda\,(f\,a).\,t) : B \to C$, and likewise
+  $(\lambda\,(f^{-1}\,b).\,t) : A \to C$, where the function bodies can refer to
+  the bound $a : A$ and $b : B$ variables.
+\end{notation}
+
 The above preservation properties mean that $\re$ is a \emph{pseudomorphism} in
@@
 \begin{itemize}
-\item Meta-level negative type formers are preserved up to isomorphism. In particular, we have
+\item Meta-level $\Sigma$ types are preserved up to isomorphism, so we have
   \[ \re_{\Sigma} : \re\,(\Sigma\,A\,B)\,\gamma
   \simeq ((\alpha : \re\,A) \times \re\,B\,(\rexti\,(\gamma,\,\alpha))). \]
-  To provide a higher-level explanation, $\re\,(\Sigma\,A\,B)\,\gamma$ is
-  essentially the presheaf of 2LTT terms with type $\Sigma_1\,A\,B$, but the
-  terms are restricted to object-level contexts. However, we can still perform
-  pairing and projection on such restricted terms; hence the preservation property.
-
+  The semantic values of $\re\,(\Sigma\,A\,B)\,\gamma$
+  are 2LTT terms with type $(\Sigma\,A\,B)[\gamma]$, restricted to object-level contexts.
+  We can still perform pairing and projection with such restricted terms; hence
+  the preservation property.
 \item Meta-level $\Pi$ types and universes are preserved in a lax way. For $\Pi$, we have
@@
   \]
-  such that $\re_{\app}\,(\re\,(\lam\,t)\,\gamma)\,\alpha =
-  \re\,t\,(\rexti\,(\gamma,\,\alpha))$. In this case, we can apply a restricted
-  term with a $\Pi$-type to a restricted term, but we cannot do
+  such that $\re_{\app}\,(\re\,t\,\gamma)\,\alpha =
+  \re\,(\app\,t)\,(\rexti\,(\gamma,\,\alpha))$. In this case, we can apply a
+  restricted term with a $\Pi$-type to a restricted term, but we cannot do
   lambda-abstraction, because that would require extending the context with a
@@
   \[
-    \re_{\Code} : \re\,\U_{1,j}\,\gamma \to \whset_j
+    \re_{\El} : \re\,\U_{1,j}\,\gamma \to \whset_j
   \]
-  such that $\re_{\Code}\,(\re\,(\El\,t)\,\gamma) = \re\,t\,\gamma$. We skip
-  $\Code$ on the right hand side here, since we assume Russell-style notation
-  internally to $\hato$, i.e.\ we omit $\Code$ and $\El$. Here, we only
-  have lax preservation simply because $\whset_j$ is much larger than
-  the the set of syntactic 2LTT types, so not every semantic $\whset_j$
-  have a syntactic representation.
+  such that $\re_{\El}\,(\re\,t\,\gamma) = \re\,(\El\,t)\,\gamma$. Here, we only
+  have lax preservation simply because $\whset_j$ is much larger than the the
+  set of syntactic 2LTT types, so not every semantic $\whset_j$ has a syntactic
+  representation.
 
@@
   natural numbers, we have
-  \[ \re_{\mbb{N}} : \mbb{N} \to \re\,\Nat_1\,\gamma. \]
+  \[ \re_{\mbb{N}} : \mbb{N} \to \re\,\Nat_{1,j}\,\gamma. \]
   This is a ``serialization'' map: from a metatheoretic natural number we
-  compute a numeral as a closed 2LTT term. This works analogously for any
-  inductive type. The restriction to object-level contexts is irrelevant when
-  computing closed terms.
+  compute a numeral as a closed 2LTT term. This works analogously for other
+  inductive types, including parameterized ones. For an example, from a semantic
+  list of restricted terms we would get a syntactic term with list type,
+  containing the same restricted terms.
 \end{itemize}
-\paragraph{Action on lifting} We have an isomorphism $\re\,(\Lift\,A)\,\gamma \simeq
+\subsubsection{Action on lifting} We have an isomorphism $\re\,(\Lift\,A)\,\gamma \simeq
 \Tm_{0,j}\,(\re\,A\,\gamma)$. This is given by quoting and splicing: we convert
 between restricted meta-level terms with type $\Lift\,A$ and restricted
-object-level terms with type $A$. Hence, we simply write the components of this
-isomorphism as $\spl\blank$ and $\qut{\blank}$, as internal analogues of the
-external operations. With this, we also have $\re\,\qut{t}\,\gamma =
+object-level terms with type $A$. Hence, we write components of this isomorphism
+as $\spl\blank$ and $\qut{\blank}$, as internal analogues of the external
+operations. With this, we also have $\re\,\qut{t}\,\gamma =
 \qut{\re\,t\,\gamma}$ and $\re\,\spl\,t\,\gamma = \spl(\re\,t\,\gamma)$.
 
-\paragraph{Action on $\blank\ext_0\blank$} Lastly, $\blank\ext_0\blank$ is preserved
-up to isomorphism:
+\subsubsection{Action on $\blank\ext_0\blank$} We have preservation up to isomorphism:
 \[
@@
 
+\subsubsection{Action on object-level types and terms} $\re$ preserves all structure
+in the object-level fragment of 2LTT. This follows from the $\re$ specification:
+an external type $A : \Ty_{0,j}\,\Gamma$ is directly internalized as an element
+of $\Ty_{0,j}$, and the same happens for $t : \Tm_{0,j}\,\Gamma\,A$.
+
 \subsection{The Logical Relation}
@@
 Internally to $\hato$, we define by induction on the syntax of 2LTT a
-proof-relevant logical relation interpretation, called $\blank^R$. The induction
+proof-relevant logical relation interpretation, written as $\blank^{\approx}$. The induction
 motives are specified as follows.
 \begin{alignat*}{5}
-  &(\Gamma &&: \Con)^R && : \ev\,\Gamma \to \re\,\Gamma \to \whset_\omega\\
-  &(\sigma &&: \Sub\,\Gamma\,\Delta)^R &&: \Gamma^R\,\gamma\,\gamma' \to \Delta^R\,(\ev\,\sigma\,\gamma)\,(\re\,\sigma\,\gamma')\\
-  &(A      &&: \Ty_{1,j}\,\Gamma)^R &&: \Gamma^R\,\gamma\,\gamma' \to \ev\,A\,\gamma \to \re\,A\,\gamma' \to \whset_j\\
-  &(t      &&: \Tm_{1,j}\,\Gamma\,A)^R &&: (\gamma^R : \Gamma^R\,\gamma\,\gamma') \to A^R\,\gamma^R\,(\ev\,t\,\gamma)\,(\re\,t\,\gamma')\\
-  &(A      &&: \Ty_{0,j}\,\Gamma)^R &&: \Gamma^R\,\gamma\,\gamma' \to \emb{\ev\,A\,\gamma} = \re\,A\,\gamma' \\
-  &(t      &&: \Tm_{0,j}\,\Gamma\,A)^R &&: \Gamma^R\,\gamma\,\gamma' \to \emb{\ev\,t\,\gamma} = \re\,t\,\gamma'
+  &(\Gamma &&: \Con)\rel && : \ev\,\Gamma \to \re\,\Gamma \to \whset_\omega\\
+  &(\sigma &&: \Sub\,\Gamma\,\Delta)\rel &&: \Gamma\rel\,\gamma\,\gamma' \to \Delta\rel\,(\ev\,\sigma\,\gamma)\,(\re\,\sigma\,\gamma')\\
+  &(A      &&: \Ty_{1,j}\,\Gamma)\rel &&: \Gamma\rel\,\gamma\,\gamma' \to \ev\,A\,\gamma \to \re\,A\,\gamma' \to \whset_j\\
+  &(t      &&: \Tm_{1,j}\,\Gamma\,A)\rel &&: (\gamma\rel : \Gamma\rel\,\gamma\,\gamma') \to A\rel\,\gamma\rel\,(\ev\,t\,\gamma)\,(\re\,t\,\gamma')\\
+  &(A      &&: \Ty_{0,j}\,\Gamma)\rel &&: \Gamma\rel\,\gamma\,\gamma' \to \emb{\ev\,A\,\gamma} = \re\,A\,\gamma' \\
+  &(t      &&: \Tm_{0,j}\,\Gamma\,A)\rel &&: \Gamma\rel\,\gamma\,\gamma' \to \emb{\ev\,t\,\gamma} = \re\,t\,\gamma'
 \end{alignat*}
@@
 logical relation interpretation: contexts are mapped to relations, types to
-dependent relations, and substitutions and terms respect relations. For
-object-level types and terms, we get soundness statements: evaluation via $\ev$
-followed by embedding back to 2LTT is the same as restricting 2LTT syntax to
-object-level contexts.
+dependent relations, and substitutions and terms respect relations. We will only have
+modest complications in meta-level type formers, because we will need to sometimes use
+the lax/oplax preservations properties of $\re$. For object-level types and
+terms, we get soundness statements: evaluation via $\ev$ followed by embedding
+back to 2LTT is the same as restriction to object-level contexts. We summarize
+the $\blank\rel$ interpretation in the following.
+
+\subsubsection{Syntactic category and terminal object}
+Here, we simply have $\id\rel\,\gamma\rel := \gamma\rel$ and $(\sigma \circ
+\delta)\rel\,\gamma\rel := \sigma\rel\,(\delta\rel\,\gamma\rel)$. The terminal
+object is interpreted as $\emptycon\rel\,\gamma\,\gamma' := \top$.
+
+\subsubsection{Meta-level family structure}
+We interpret context extension and type/term substitution as follows. Note the
+usage of the pattern matching notation on the $\rexti$ isomorphism.
+\begin{alignat*}{4}
+  & (\Gamma \ext_1 A)\rel\,(\gamma,\,\alpha)\,(\rexti(\gamma',\,\alpha')) && :=
+    (\gamma\rel : \Gamma\rel\,\gamma\,\gamma') \times A\rel\,\gamma\rel\,\alpha\,\alpha'\\
+  & (A[\sigma])\rel\,\gamma\rel\,\alpha\,\alpha' && := A\rel\,(\sigma\rel\,\gamma\rel)\,\alpha\,\alpha'\\
+  & (t[\sigma])\rel\,\gamma\rel\                 && := t\rel\,(\sigma\rel\,\gamma\rel)
+\end{alignat*}
+It is enough to specify the $\blank\rel$ action on extended substitutions
+$(\sigma,\,t) : \Sub\,\Gamma\,(\Delta\ext A)$, $\p : \Sub\,(\Gamma\ext
+A)\,\Gamma$, $\p : \Sub\,(\Gamma\ext A)\,\Gamma$ and $\q : \Tm_{1,j}\,(\Gamma
+\ext A)\,(A[\p])$. The category-with-families equations hold evidently.
+\begin{alignat*}{4}
+  & (\sigma,\,t)\rel\,\gamma\rel                 && := (\sigma\rel\,\gamma\rel,\,t\rel\,\gamma\rel)\\
+  & \p\rel\,(\gamma\rel,\,t\rel)                 && := \gamma\rel\\
+  & \q\rel\,(\gamma\rel,\,t\rel)                 && := t\rel
+\end{alignat*}
+
+\subsubsection{Meta-level natural numbers} First, we have to define a relation:
+\begin{alignat*}{4}
+  & (\Nat_{1,j})\rel : \Gamma\rel\,\gamma\,\gamma' \to \mbb{N} \to \re\,\Nat_{1,j}\,\gamma' \to \whset_j\\
+  & (\Nat_{1,j})\rel\,\gamma\rel\,n\,n' := (\re_{\mbb{N}}\,n = n')
+\end{alignat*}
+Note that evaluation sends $\Nat_{1,j}$ to the semantic type of natural numbers,
+i.e.\ $\ev\,\Nat_{1,j} = \mbb{N}$. We refer to the serialization map
+$\re_{\mbb{N}} : \mbb{N} \to \re\,\Nat_{1,j}\,\gamma'$ above. In short,
+$(\Nat_{1,j})\rel$ expresses \emph{canonicity}: $n'$ is canonical precisely if
+it is of the form $\re_{\mbb{N}}\,n$ for some $n$, hence a finite successor of
+zero.
+
+For $\zero\rel$ and $\suc\rel$, we need to show that serialization respects
+$\zero$ and $\suc$ respectively, which is evident.
+
+Let us look at elimination. We need to define the following:
+\begin{alignat*}{4}
+  &(\msf{NatElim}\,P\,z\,s\,n)\rel\,\gamma\rel :
+  P\rel\,(\gamma\rel,\,n\rel\,\gamma\rel)\,(\ev\,(\msf{NatElim}\,P\,z\,s\,n)\,\gamma)\,
+                                           (\re\,(\msf{NatElim}\,P\,z\,s\,n)\,\gamma')
+\end{alignat*}
+Unfolding $\ev$, we can further compute this to the following:
+\begin{alignat*}{4}
+  &(\msf{NatElim}\,P\,z\,s\,n)\rel\,\gamma\rel : \\
+  &\hspace{2em}P\rel\,(\gamma\rel,\,n\rel\,\gamma\rel)\,
+  (\msf{NatElim}\,(\lambda\,n.\,\ev\,P\,(\gamma,\,n))\,
+                  (\ev\,z\,\gamma)\,
+                  (\lambda\,n\,\mit{pn}.\,\ev\,s\,((\gamma,\,n),\,\mit{pn}))
+                  (\ev\,n\,\gamma))\\
+  &\hspace{8em}(\re\,(\msf{NatElim}\,P\,z\,s\,n)\,\gamma')
+\end{alignat*}
+In short, we need to show that $\msf{NatElim}$ preserves relations. Here we
+switch to the external view temporarily. By Definition \ref{TODO}, we know that
+\[
+   |\re\,(\msf{NatElim}\,P\,z\,s\,n)|\,\gamma' = (\msf{NatElim}\,P\,z\,s\,n))[\gamma'].
+\]
+At the same time, we have $n\rel\,\gamma\rel : \re_{\mbb{N}}\,(\ev\,n\,\gamma) =
+\re\,n\,\gamma'$, hence we know externally that $|\ev\,n|\,\gamma = n[\gamma']$.
+In other words, $n[\gamma']$ is canonical and is obtained as the serialization
+of $\ev\,n\,\gamma$. Therefore, $(\msf{NatElim}\,P\,z\,s\,n))[\gamma']$ is
+definitionally equal to $|\ev\,n|\,\gamma$-many applications of $s$ to
+$z$, and we can use $|\ev\,n|\,\gamma$-many applications of $s\rel$ to $z\rel$
+to witness the preservation. The $\beta$-rules for $\msf{NatElim}$ are also respected
+by this definition.
+
+\subsubsection{Meta-level $\Sigma$-types} We define relatedness pointwise. Pairing and projection
+are interpreted as meta-level pairing and projection.
+\begin{alignat*}{3}
+  &(\Sigma\,A\,B)\rel\,\gamma\rel : ((\alpha : \ev\,A\,\gamma) \times \ev\,B\,(\gamma,\,\alpha))
+             \to \re\,(\Sigma\,A\,B)\,\gamma' \to \whset_j\\
+  &(\Sigma\,A\,B)\rel\,\gamma\rel\,(\alpha,\,\beta)\,(\re_{\Sigma}^{-1}(\alpha',\,\beta')) :=
+             (\alpha\rel : A\rel\,\gamma\rel\,\alpha\,\alpha') \times B\rel\,(\gamma\rel,\,\alpha\rel)\,\beta\,\beta'
+\end{alignat*}
+
+\subsubsection{Meta-level $\Pi$-types} We again use a pointwise definition. Note that we need to use
+$\re_{\app}$ to apply $t' : \re\,(\Pi\,A\,B)\,\gamma'$ to $\alpha'$.
+\begin{alignat*}{3}
+  &(\Pi\,A\,B)\rel\,\gamma\rel : ((\alpha : \ev\,A\,\gamma) \to \ev\,B\,(\gamma,\,\alpha))
+             \to \re\,(\Pi\,A\,B)\,\gamma' \to \whset_j\\
+  &(\Pi\,A\,B)\rel\,\gamma\rel\,t\,t' :=
+    (\alpha : \ev\,A\,\gamma)(\alpha' : \re\,A\,\gamma')(\alpha\rel : A\rel\,\gamma\rel\,\alpha\,\alpha') \to
+    B\rel\,(\gamma\rel,\,\alpha\rel)\,(t\,\alpha)\,(\re_{\app}\,t'\,\alpha')
+\end{alignat*}
+For abstraction and application, we use a curry-uncurry definition:
+\begin{alignat*}{4}
+  &(\lam\,t)\rel\,\gamma\rel                &&:= \lambda\,\alpha\,\alpha'\,\alpha\rel.\,t\rel\,(\gamma\rel,\,\alpha\rel)\\
+  &(\app\,t)\rel\,(\gamma\rel,\,\alpha\rel) &&:= t\rel\,\gamma\rel\,\alpha\,\alpha'\,\alpha\rel
+\end{alignat*}
+
+\subsubsection{Meta-level universes} We interpret $\U_{1,j}$ as a semantic relation space:
+\begin{alignat*}{3}
+  &(\U_{1,j})\rel\,\gamma\rel : \whset_j \to \re\,\U_{1,j}\,\gamma' \to \whset_{j+1} \\
+  &(\U_{1,j})\rel\,\gamma\rel\,t\,t' := t \to \re_{\El}\,t' \to \whset_{j}
+\end{alignat*}
+Note that we have
+\begin{alignat*}{3}
+  & (\El\,t)\rel   &&: (\gamma\rel : \Gamma\rel\,\gamma\,\gamma') \to \ev\,t\,\gamma \to \re\,(\El\,t)\,\gamma' \to \whset_j\\
+  & (\Code\,t)\rel &&: (\gamma\rel : \Gamma\rel\,\gamma\,\gamma') \to \ev\,t\,\gamma \to \re\,(\El\,t)\,\gamma' \to \whset_j.
+\end{alignat*}
+The types coincide because of the equation $\re\,(\El\,t)\,\gamma' =
+\re_{\El}\,(\re\,t\,\gamma')$. Therefore we can interpret $\El$ and $\Code$ as
+identity maps, as $(\El\,t)\rel := t\rel$ and $(\Code\,t)\rel := t\rel$.
+
+\subsubsection{Object-level family structure} We interpret extended contexts as follows.
+\begin{alignat*}{3}
+  & (\Gamma \ext_0 A)\rel\,(\gamma,\,\alpha)\,(\rexti\,(\gamma',\,\alpha')) := (\gamma\rel : \Gamma\rel\,\gamma\,\gamma') \times (\emb{\alpha} = \alpha')
+\end{alignat*}
+Note that $\alpha : \Tm_{\mbbo j}\,(\ev\,A\,\gamma)$, so $\emb{\alpha} :
+\Tm_{0,j}\,\emb{\ev\,A\,\gamma}$, but since $A\rel\,\gamma\rel :
+\emb{\ev\,A\,\gamma} = \re\,A\,\gamma'$, we also have $\emb{\alpha} :
+\Tm_{0,j}\,(\re\,A\,\gamma')$. Thus, the equation $\emb{\alpha} = \alpha'$ is
+well-typed. For type substitution, we need
+\[
+  (A[\sigma])\rel\,\gamma\rel : \emb{\ev\,(A[\sigma])\,\gamma} = \re\,(A[\sigma])\,\gamma'.
+\]
+The goal type computes to $\emb{\ev\,A\,(\ev\,\sigma\,\gamma)} =
+\re\,A\,(\re\,\sigma\,\gamma')$. This is obtained directly from
+$A\rel\,(\sigma\rel\,\gamma\rel)$.  Similarly, we have
+\begin{alignat*}{4}
+  & (t[\sigma])\rel\,\gamma\rel &&:= t\rel\,(\sigma\rel\,\gamma\rel) \\
+  & (\sigma,\,t)\rel\,\gamma\rel &&:= (\sigma\rel\,\gamma\rel,\,t\rel\,\gamma\rel) \\
+  & \p\rel\,(\gamma\rel,\,\alpha\rel) &&:= \gamma\rel \\
+  & \q\rel\,(\gamma\rel,\,\alpha\rel) &&:= \alpha\rel. \\
+\end{alignat*}
+
+\subsubsection{Lifting structure}
+\begin{alignat*}{3}
+  &(\Lift\,A)\rel : \Gamma\rel\,\gamma\,\gamma' \to \Tm_{\mbbo j}\,(\ev\,A\,\gamma) \to \re\,(\Lift\,A)\,\gamma' \to \whset_j\\
+  &(\Lift\,A)\rel\,\gamma\rel\,t\,t' := (\emb{t} = \spl t')
+\end{alignat*}
+This is well-typed by $A\rel\,\gamma\rel : \emb{\ev\,A\,\gamma} = \re\,A\,\gamma'$, which implies
+that $\emb{t} : \Tm_{\mbbo j}\,(\re\,A\,\gamma')$. For $\qut{t}$, we need
+\[  \qut{t}\rel\,\gamma\rel : \emb{\ev\,\qut{t}\,\gamma} = \spl(\re\,\qut{t}\,\gamma'). \]
+The goal type can be further computed to $\emb{\ev\,t\,\gamma} = \re\,t\,\gamma'$, which we prove
+by $t\rel\,\gamma\rel$. For splicing, we need
+\[
+  (\spl t)\rel\,\gamma\rel : \emb{\ev\,\spl t\,\gamma} = \re\,\spl t\,\gamma'
+\]
+where the goal type computes to $\emb{F\,t\,\gamma} = \spl(\re\,t\,\gamma')$,
+but this again follows directly from $t\rel\,\gamma\rel$.
+
+\subsubsection{Object-level type formers}
+Lastly, object-level type formers are straightforward. For types, we need
+$\emb{\ev\,A\,\gamma} = \re\,A\,\gamma'$, and for terms, we need
+$\emb{\ev\,A\,\gamma} = \re\,A\,\gamma'$. Note that $\ev$ and $\emb{\blank}$
+preserve all structure, and $\re$ preserves all structure on object-level types
+and terms. Hence, each object-level $A$ and $t$ case in $\blank\rel$ trivially
+follows from induction hypotheses.
+
+This concludes the definition of the $\blank\rel$ interpretation.
+
+\subsection{Soundness}
+
+\begin{definition} First, we introduce shorthands for
+external operations that can be extracted from $\blank\rel$.
+\begin{itemize}
+  \item For $\Gamma : \Con$, we get $|\Gamma\rel| : \{\Delta : \Cono\} \to |\ev\,\Gamma|\,\Delta \to
+    \Sub\,\emb{\Delta}\,\Gamma \to \Set_j$.
+  \item For $A : \Ty_{0,j}\,\Gamma$, we get $|A\rel| : |\Gamma\rel|\,\gamma\,\gamma' \to \emb{\ev\,A\,\gamma} = A[\gamma']$.
+  \item For $t : \Tm_{0,j}\,\Gamma\,A$, we get $|t\rel| : |\Gamma\rel|\,\gamma\,\gamma' \to \emb{\ev\,t\,\gamma} = t[\gamma']$.
+\end{itemize}
+Since $\blank\rel$ was defined in $\hato$, we also know that the above are all
+stable under object-theoretic substitution.
+\end{definition}
+
+\begin{theorem}[Soundness for generic contexts]
+For each $\Gamma : \Cono$, we have ${\Gamma^P}\rel : |\Gamma\rel|\,\Gamma^P\,\id$.
+\end{theorem}
+\begin{proof}
+We define ${\blank^P}\rel$ by induction on object-theoretic contexts.
+${\emptycon^P}\rel : \top$ is defined trivially as $\tt$. For ${(\Gamma \ext
+  A)^P}\rel$, we need $(\gamma\rel : |\Gamma|\rel\,(\Gamma^P[\p])\,\p) \times
+(\emb{\q} = \q)$. We get ${\Gamma^P}\rel : |\Gamma\rel|\,\Gamma^P\,\id$. Because
+of the naturality of $|\Gamma\rel|$, this can be weakened to ${\Gamma^P}\rel[\p]
+: |\Gamma\rel|\,(\Gamma^P[\p])\,\p$. Also, $\emb{\q} = \q$ holds immediately.
+\end{proof}
+
+\begin{theorem}[Soundness of staging] The open staging algorithm from Definition \ref{TODO} is sound.
+\end{theorem}
+\begin{proof}
+\mbox{}
+\begin{itemize}
+\item For $A : \Ty_{0,j}\,\emb{\Gamma}$, we have that $\emb{\Stage\,A} =
+\emb{|\ev\,A|\,\Gamma^P}$ by the definition of $\Stage$, and moreover we have
+$|A\rel|\,{\Gamma^P}\rel : \emb{|\ev\,A|\,\Gamma^P} = A[\id]$, hence $\emb{\Stage\,A} = A$.
+
+\item For $t : \Tm_{0,j}\,\emb{\Gamma}\,A$, using $|t\rel|\,{\Gamma^P}\rel : \emb{|\ev\,t|\,\Gamma^P} = t[\id]$, we likewise
+  have $\emb{\Stage\,t} = \emb{|\ev\,t|\,\Gamma^P} = t[\id] = t$.
+\end{itemize}
+\end{proof}
+
+\begin{corollary}[Conservativity of 2LTT]
+From the soundness and stability of staging, we get that $\emb{\blank}$ is
+bijective on types and terms, i.e.\ $\Ty_{\mbbo j}\,\Gamma \simeq \Ty_{0,j}\,\emb{\Gamma}$ and
+$\Tm_{\mbbo j}\,\Gamma\,A \simeq \Tm_{0,j}\,\emb{\Gamma}\,\emb{A}$.
+\end{corollary}
+
+%% \begin{alignat*}{3}
+%%   & \hspace{1em} \emb{\Stage\,A} \hspace{2em}&& \\
+%%   & = \emb{|\ev\,A|\,\Gamma^P}   \hspace{2em}&& \text{by the definition of $\Stage$} \\
+%%   & = A                          \hspace{2em}&& \text{by $|A\rel|\,{\Gamma^P}\rel : \emb{|\ev\,A|\,\Gamma^P} = A[\id]$}
+%% \end{alignat*}
+
+%% to show:
+
+%%   ⌜Stage A⌝  = A
+%%   ⌜|F A| Γᴾ⌝ = A
+
+%%     |Aᴿ| Γᴾᴿ : ⌜|F A| Γᴾ⌝ = A[id] = A OK
+
+%%   ⌜Stage t⌝ = t
+%%   ⌜|F t| Γᴾ⌝ = t
+
+%%     |tᴿ| Γᴾᴿ :  ⌜|F t| Γᴾ⌝ = t[id] = t OK
 

### 2022-02-20 paper/paper.tex
stuff

@@
 \newcommand{\rel}{^{\approx}}
+\newcommand{\yon}{\msf{y}}
 
@@
   numerous good properties for this use case: it has a concise specification,
-  well-developed algebraic and categorical model theory, and it supports a wide
-  range of language features both at the object and the meta level. First, we
-  give an overview of 2LTT's features and applications in staging. Then, we
-  present a staging algorithm and provide a proof of correctness. Our algorithm
-  is ``staging-by-evaluation'', analogously to the technique of
-  normalization-by-evaluation, in that staging is given by the evaluation of
-  2LTT syntax in a semantic domain. The staging algorithm together with its
-  correctness proof constitutes a proof of strong conservativity of 2LLT over
-  the object theory. To our knowledge, this is the first system for staged
-  compilation which supports full dependent types and unrestricted staging for
-  types.
+  well-developed model theory, and it supports a wide range of language features
+  both at the object and the meta level. First, we give an overview of 2LTT's
+  features and applications in staging. Then, we present a staging algorithm and
+  provide a proof of correctness. Our algorithm is ``staging-by-evaluation'',
+  analogously to the technique of normalization-by-evaluation, in that staging
+  is given by the evaluation of 2LTT syntax in a semantic domain. The staging
+  algorithm together with its correctness proof constitutes a proof of strong
+  conservativity of 2LLT over the object theory. To our knowledge, this is the
+  first system for staged compilation which supports full dependent types and
+  unrestricted staging for types.
 \end{abstract}
@@
 
-\section{Introduction}
+\section{Introduction}\label{sec:introduction}
 
@@
 
-\subsection{Contributions}
+\subsection{Overview \& Contributions}
 
 \begin{itemize}
-  \item In \ref{TODO} we present an informal syntax of two-level type theory, a
-    dependent type theory with staging features. We look at basic use-cases
-    involving inlining control, partial evaluation and fusion optimizations. We
-    also describe feature variations, enabling applications in monomorphization
-    and memory layout control.
-  \item In \ref{TODO}, following \cite{twolevel}, we present a formal syntax of
-    2LTT and the object theory (the target theory of code generation). We recall
-    the standard presheaf model of 2LTT, which lies over the syntactic category
-    of the object theory. We show that the evaluation of 2LTT syntax in the presheaf
-    model yields a staging algorithm.
-  \item In \ref{TODO} we show correctness of staging, consisting of
-    \begin{itemize}
-    \item \emph{Stability:} staging the output of staging has no action.
-    \item \emph{Soundness:} the output of staging is convertible to the input.
-    \item \emph{Completeness:} convertible programs produce convertible staging outputs.
+  \item In Section \ref{sec:tour-of-2ltt} we present an informal syntax of two-level
+    type theory, a dependent type theory with staging features. We look at basic
+    use cases involving inlining control, partial evaluation and fusion
+    optimizations. We also describe feature variations, enabling applications in
+    monomorphization and memory layout control.
+  \item In Section \ref{sec:formal-2ltt}, we specify the formal notion of models and
+    syntax for the variant of 2LTT used in this paper, largely following \cite{twolevel}.
+  \item In Section \ref{sec:staging-algorithm}, we review the standard presheaf
+    model of 2LTT \cite[Section~2.5.3]{twolevel}, which lies over the syntactic
+    category of the object theory.  We show that evaluation in this model yields
+    a staging algorithm for closed types and terms. We then prove stability for
+    staging, which means that staging the output of staging has no action. Then,
+    we extend staging to open types and terms which may depend on object-level
+    contexts.
+  \item
+    In Section \ref{sec:soundness} we show soundness of staging, which means that
+    the output of staging is convertible to the input.  Staging together with
+    its stability and soundness can be viewed as a \emph{strong conservativity}
+    theorem of 2LTT over the object theory. This means that the possible
+    object-level constructions in 2LTT are in bijection with the constructions
+    in the object theory, and staging witnesses that meta-level constructions
+    can be always computed away. This strengthens the weak notion of
+    conservativity shown in \cite{capriotti2017models} and \cite{twolevel}.
+  \item
+    In Section \ref{sec:intensional-analysis}, we discuss possible interpretations
+    of intensional code analysis in the presheaf model.
+  \item To our knowledge, this is the first account of staged compilation in the
+    presence of full-blown dependent types, with universes and large
+    elimination. In particular, we allow unrestricted staging for types, so that
+    types can be computed by metaprograms at compile time.
 \end{itemize}
-    Staging together with its correctness can be viewed as a \emph{strong
-    conservativity} theorem of 2LTT over the object theory. This means that the
-    possible object-level constructions in 2LTT are in bijection with the
-    constructions in the object theory, and staging witnesses that meta-level
-    constructions can be always computed away. This improves on the weak notion
-    of conservativity shown in \cite{capriotti2017models} and
-    \cite{twolevel}.
-  \item To our knowledge, this is the first description of a language which
-    supports staging in the presence of full-blown dependent types, with
-    universes and large elimination. Moreover, we allow unrestricted staging
-    for types, so that types can be computed by metaprograms at compile time.
-\end{itemize}
-
-\section{A Tour of Two-Level Type Theory}
+
+\section{A Tour of Two-Level Type Theory}\label{sec:tour-of-2ltt}
 
@@
 applications in staging. We work in the informal syntax of a dependently typed
-language which resembles Agda \cite{TODO}. We focus on examples and informal
-explanations here; the formal details will be presented in Section \cite{TODO}.
-
-\begin{notation}
+language which resembles Agda \cite{agdadocs}. We focus on examples and informal
+explanations here; the formal details will be presented in Section \ref{sec:formal-2ltt}.
+
+\begin{notation}\label{basic-notation}
 We use the following notations throughout the paper. $(x : A) \to B$ denotes a
@@
 
-\paragraph{Universes}
+\subsubsection{Universes}
 We have universes $\U_{i,j}$, where $i \in \{0,1\}$, and $j \in \mbb{N}$.  The
@@
 
-\paragraph{Type formers} $\U_0$ and $\U_1$ may be closed under arbitrary type formers,
+\subsubsection{Type formers} $\U_0$ and $\U_1$ may be closed under arbitrary type formers,
 such as functions, $\Sigma$-types, identity types or inductive types in general.
@@
 It is not required that we have the \emph{same} type formers at both stages. We
-will discuss setups with different languages at different stages in Seection
-\ref{todo}.
-
-\paragraph{Moving between stages}
+will discuss variations of the object-level languages in Section \ref{sec:variations}.
+
+\subsubsection{Moving between stages}
 At this point, our system is rather limited, since there is no interaction
@@
   expression is inserted into the output.
+
   \begin{notation} Splicing binds stronger than any operation, including function
-    application. For instance, $\spl f\,x$ is parsed as $(\spl f)\,x$.
+    application. For instance, $\spl f\,x$ is parsed as $(\spl f)\,x$. We borrow
+    this notation from MetaML \cite{metaml}.
   \end{notation}
@@
 duplicate the code of the mapping function. In such cases, we can use
-\emph{let-insertion} \cite{TODO}, a standard technique in staged compilation. If
-we bind a runtime expression to a runtime variable, and only use that variable
-in subsequent staging, only the variable itself can be duplicated. One solution
-is to do an ad-hoc let-insertion:
+\emph{let-insertion}, a technique in staged compilation (see
+\cite{staging-with-delimited-control} for an overview). If we bind a runtime
+expression to a runtime variable, and only use that variable in subsequent
+staging, only the variable itself can be duplicated. One solution is to do an
+ad-hoc let-insertion:
 \begin{alignat*}{6}
@@
 \end{alignat*}
-This is an example for the usage of \emph{partially static
-data}\cite{TODO}: semantic contexts are \emph{static} lists storing
-\emph{runtime} expressions. This allows us to completely eliminate environment
-lookups in the staging output: an embedded lambda expression is staged to the
-corresponding lambda expression in the runtime language. This is similar to the
-partial evaluator presented in Idris 1 \cite{TODO}. However, in contrast to
-2LTT, Idris 1 does not provide a formal guarantee that partial evaluation does
-not get stuck.
+This is an example for the usage of \emph{partially static data} \cite{partial-evaluation}:
+semantic contexts are \emph{static} lists storing \emph{runtime}
+expressions. This allows us to completely eliminate environment lookups in the
+staging output: an embedded lambda expression is staged to the corresponding
+lambda expression in the runtime language. This is similar to the partial
+evaluator presented in Idris 1 \cite{scrap-your-inefficient-engine}. However, in
+contrast to 2LTT, Idris 1 does not provide a formal guarantee that partial
+evaluation does not get stuck.
 
@@
 We describe more generally the action of $\Lift$ on type formers. First, $\Lift$
-preserves negative type formers up to definitional isomorphism \cite{twolevel}:
+preserves negative type formers up to definitional isomorphism \cite[Section~2.3]{twolevel}:
 \begin{alignat*}{5}
@@
 what is termed \emph{binding time improvement} in the partial evaluation
-literature \cite{TODO}. Note that the output of $\mit{pres}_{\to}$ uses a
+literature \cite[Chapter~12]{partial-evaluation}. Note that the output of $\mit{pres}_{\to}$ uses a
 meta-level $\lambda$, while going the other way we introduce a runtime
@@
 $\lambda\,b.\,\msf{if}\,b\,\msf{then}\,\qut{\true_0}\,\msf{then}\,\qut{\false_0}$.
-In the staging literature, this is called ``serialization'' \cite{TODO}, or
-``lifting'' in the context of Template Haskell \cite{TODO}. In the other
-direction, we can only define constant functions from $\Lift \Bool_0$ to
-$\Bool_1$.
+In the staging literature, this is called ``serialization'' or ``lifting''
+\cite{template-haskell,metaml}.  In the other direction, we can only define
+constant functions from $\Lift \Bool_0$ to $\Bool_1$.
 
 The lack of elimination principles for $\Lift A$ means that we cannot inspect
-the internal structure of expressions. This is called ``generativity'' in
-staging \cite{TODO}. We will briefly discuss non-generative staging in Section
-\ref{TODO}.
+the internal structure of expressions. We will briefly discuss ways to lift this
+restriction in Section \ref{sec:intensional-analysis}.
 
@@
 equivalent to a finite product. For instance, $\Bool_1 \to \Nat_1 \simeq \Nat_1
-\times \Nat_1$.  In 2LTT, $A$ is called \emph{cofibrant} \cite{TODO}: this means
-that for each $B$, $A \to \Lift\,B$ is equivalent to $\Lift C$ for some
-$C$. This formalizes the so-called ``trick'' in partial evaluation, which
-improves binding times by $\eta$-expanding functions out of finite sums
-\cite{TODO}.
+\times \Nat_1$.  In 2LTT, $A$ is called \emph{cofibrant}
+\cite[Section~3.4]{twolevel}: this means that for each $B$, $A \to \Lift\,B$ is
+equivalent to $\Lift C$ for some $C$. This is the 2LTT formalization of the
+so-called ``trick'' in partial evaluation, which improves binding times by
+$\eta$-expanding functions out of finite sums \cite{eta-expansion-trick}.
 
@@
 type, it is brought to a form which can be binding-time improved. For instance,
-consider foldr-build fusion for lists, which is employed in GHC Haskell
-\cite{TODO}. Starting from $\Lift (\List_0\,A)$, we use Böhm-Berarducci
-encoding \cite{TODO} under the lifting to get
+consider foldr-build fusion for lists \cite{short-cut}, which is employed in GHC
+Haskell.  Starting from $\Lift (\List_0\,A)$, we use Böhm-Berarducci encoding
+\cite{boehm-berarducci} under the lifting to get
 \[ \Lift((L : \U_0) \to (A \to L \to L) \to L \to L) \]
@@
     \to \Lift\,\spl L \to \Lift\,\spl L. \]
-Alternatively, for \emph{stream fusion}, we embed $\List\,A$ into the
+Alternatively, for \emph{stream fusion} \cite{stream-fusion}, we embed $\List\,A$ into the
 coinductive colists (i.e.\ the possibly infinite lists), and use a terminal
@@
 optimizations which are otherwise not possible, such as fusion for the
-$\msf{zip}$ function \cite{TODO}. However, the price we pay is that converting
+$\msf{zip}$ function. However, the price we pay is that converting
 back to lists from colists is not necessarily total.
@@
 \subsubsection{Inferring staging operations}
-During bidirectional elaboration \cite{TODO}, we can use the preservation
-isomorphisms and the quote-splice isomorphism as a coercive subtyping
-system. When elaboration needs to compare an inferred and an expected type, it
-may insert transports along isomorphisms. We implemented this feature in our
-prototype. It additionally supports Agda-style implicit arguments and pattern
-unification, so it can elaborate the following definition:
+During bidirectional elaboration, we can use the preservation isomorphisms and
+the quote-splice isomorphism as a coercive subtyping system. When elaboration
+needs to compare an inferred and an expected type, it may insert transports
+along isomorphisms. We implemented this feature in our prototype. It
+additionally supports Agda-style implicit arguments and pattern unification, so
+it can elaborate the following definition:
 \begin{alignat*}{3}
@@
 \subsection{Variations of Object-Level Languages}
+\label{sec:variations}
 
@@
 less expressive and less principled. Additionally, we will see in Section
-\ref{TODO} that this setup is compatible with an \emph{induction principle} for
+\ref{sec:intensional-analysis} that this setup is compatible with an induction principle for
 $\Ty_0$, so that we can analyze the structure of runtime types.
@@
 this would be a more advanced variation on type-passing polymorphism
-\cite{TODO}, where layouts of objects may depend on runtime data. For example,
+\cite{intensional-poly}, where layouts of objects may depend on runtime data. For example,
 length-prefixed flat arrays would not be a primitive notion, but simply defined
@@
 challenging to implement runtime systems and precise garbage collection in this
-setting. See \texttt{sixten} \cite{TODO} for an experimental implementation of
+setting. See \texttt{sixten} \cite{sixten} for an experimental implementation of
 dependent memory layouts.
 
-\section{Formal 2LTT}
+\section{Formal 2LTT}\label{sec:formal-2ltt}
 
@@
 
-\subsection{Metatheory}
-TODO
+\subsection{Metatheory} We work in an extensional type theory with
+a cumulative ordinal-indexed universe hierarchy $\Set_j$. For type formers, we
+use Notation \ref{basic-notation}. We also use extensional equality, written as
+$\blank\!=\!\blank$. We will sometimes use $\Set_\omega$ and $\Set_{\omega+1}$.
+This is mainly a convenience feature when modeling contexts. Types are indexed
+by $\mbb{N}$ sizes, so we usually model a type at finite level $j$ using
+$\Set_j$, but contexts contain types at possibly different levels. The easy
+solution is to model contexts using $\Set_\omega$, which is large enough for all
+constituent types. Alternatively, we could formally track the sizing levels for
+contexts as well, but that adds inessential noise to the specification.
 
@@
 definitional equality with metatheoretic equality. First, we define the
-structural scaffolding of 2LTT without type formers.
+structural scaffolding of 2LTT without type formers. The following specification
+is agnostic about the sizes of $\Set$-s involved, i.e.\ we can model underlying
+sorts using any metatheoretic $\Set_j$.
 
@@
   isomorphism $\Sub\,\Gamma\,(\Delta\ext A) \simeq ((\sigma :
-  \Sub\,\Gamma\,\Delta) \times \Tm_{i,j}\,\Gamma\,(A[\sigma]))$.
+  \Sub\,\Gamma\,\Delta) \times \Tm_{i,j}\,\Gamma\,(A[\sigma]))$. We will sometimes
+  use $\blank\!\ext_0\!\blank$ and $\blank\!\ext_1\!\blank$ to disambiguate object-level
+  and meta-level context extensions.
 \item
@@
 In short, for each $i$ and $j$ we have a family structure in the sense of
-categories-with-families \cite{TODO}, such that there is a family morphism from
-each $(\Ty_{0,j},\,\Tm_{0,j})$ to $(\Ty_{1,j},\,\Tm_{1,j})$ in the sense of
-\cite{TODO}.
+categories-with-families \cite{Dybjer96internaltype}, such that there is a
+family morphism from each $(\Ty_{0,j},\,\Tm_{0,j})$ to $(\Ty_{1,j},\,\Tm_{1,j})$.
 
@@
 \begin{notation}
-De Bruijn indices are rather hard to read, so we will often use nameful notation
+De Bruijn indices are rather hard to read, so we will sometimes use nameful notation
 for binders and substitutions. For example, we may write $\Gamma \ext (x : A)
@@
 \begin{definition} A \textbf{model of 2LTT} is a model of basic 2LTT
-which supports certain type formers. For the sake of brevity, we only present
-our results for a small collection of type formers. However, we will argue that
-our results easily extend to any general notion of inductive type formers. We
-specify type formers in the following.  We assume that all type formers are
-natural, i.e.\ stable under substitution.
+which supports certain type formers. For the sake of brevity, we concretely
+specify here a small collection of type formers. However, we will argue that our
+results extend to more general notions of inductive types. We specify type
+formers in the following. We assume that all type formers are natural,
+i.e.\ stable under substitution.
 \begin{itemize}
@@
 \begin{definition}
-The \textbf{syntax of 2LTT} is defined to be the initial model of 2LTT. The syntax
-can be assumed to exist, since the notion of model is algebraic; the
-specification can be expressed using a suitable notion of algebraic signatures,
-such as one for essentially algebraic theories \cite{TODO} or as a finitary
-quotient inductive-inductive signature \cite{TODO}. The initiality of syntax
-directly yields a notion of \emph{recursion on syntax}, and it can be shown that
-initiality also implies an \emph{induction principle} \cite{TODO}.
+The \textbf{syntax of 2LTT} is defined to be the initial model of 2LTT, where
+all underlying sets are in $\Set_0$. The syntax can be assumed to exist, since
+the notion of model is algebraic; the specification can be expressed using a
+suitable notion of algebraic signatures, such as one for essentially algebraic
+theories \cite{TODO} or as a finitary quotient inductive-inductive signature
+\cite{TODO}. The initiality of syntax directly yields a notion of
+\emph{recursion on syntax}, and it can be shown that initiality also implies an
+\emph{induction principle} \cite{TODO}.
 \end{definition}
@@
 extensional type theory. In contrast, we allow a wider range of setups under the
-2LTT umbrella. Annekov et al.\ also considers a range of additional
+2LTT umbrella. Ibid.\ also considers a range of additional
 strengthenings and extension of 2LTT \cite[Section~2.4]{twolevel}, most of which
@@
 
-\section{The Staging Algorithm}
+\section{The Staging Algorithm}\label{sec:staging-algorithm}
 
@@
 
-We observe that the presheaf model described in \cite[Section~2.5.3]{twolevel}
-immediately yields a closed staging algorithm, by evaluation of 2LTT types and
-terms in the model. In this model, contexts are presheaves over the syntactic
-category of the object theory. We denote the components of the model by putting
-hats on 2LTT components, e.g.\ as in $\hCon$. We summarize the key components in
-the following.
+We review the presheaf model described in \cite[Section~2.5.3]{twolevel}.  This
+will directly yield a closed staging algorithm, by evaluation of 2LTT types and
+terms in the model. Here, contexts are presheaves over the syntactic category of
+the object theory. We denote the components of the model by putting hats on 2LTT
+components, e.g.\ as in $\hCon$. We summarize the key components in the
+following.
 
@@
 $\gamma : |\ev\,\Gamma|\,\Delta$. These environments are lists containing
-object-level terms and semantic values. The meta-level semantic values are
-represented using meta-level inductive types and the function space of the
-metatheory.
+object-level terms and semantic values. The semantic values are represented
+using metatheoretic inductive types and function types.
 \begin{itemize}
-  \item $|\Nat_1|\,\_$ is simply the set of meta-level natural numbers.
+  \item $|\Nat_1|\,\_$ is simply the set of natural numbers $\mbb{N}$.
   \item $|\Sigma_1\,A\,B|\,\gamma$ is simply a set of pairs of values.
@@
 
-\section{Yoneda, Representability and Generativity}
-
-\section{Soundness of Staging}
+\section{Soundness of Staging}\label{sec:soundness}
 
@@
 theory \cite{TODO}, and certain other structures which are stable under
-object-level substitution. Since almost all constructions in this section must
-be stable under object-level substitution, it makes sense to work internally to
-$\hato$. This technique has been previously used in normalization proofs
-\cite{TODO} and also in the metatheory of cubical type theories \cite{TODO}.
+object-level substitution. As all constructions in this section must be stable,
+it makes sense to work internally to $\hato$. This style has been previously
+used in normalization proofs \cite{TODO} and also in the metatheory of cubical
+type theories \cite{TODO}.
 
@@
 straightforward in each case.
+\begingroup
+%% \allowdisplaybreaks
 \begin{alignat*}{5}
@@
 \end{alignat*}
+\endgroup
 \end{definition}
@@
 terms, we get soundness statements: evaluation via $\ev$ followed by embedding
-back to 2LTT is the same as restriction to object-level contexts. We summarize
+back to 2LTT is the same as restriction to object-level contexts. We describe
 the $\blank\rel$ interpretation in the following.
@@
 $(\Nat_{1,j})\rel$ expresses \emph{canonicity}: $n'$ is canonical precisely if
-it is of the form $\re_{\mbb{N}}\,n$ for some $n$, hence a finite successor of
-zero.
-
-For $\zero\rel$ and $\suc\rel$, we need to show that serialization respects
+it is of the form $\re_{\mbb{N}}\,n$ for some $n$.
+
+For $\zero\rel$ and $\suc\rel$, we need to show that serialization preserves
 $\zero$ and $\suc$ respectively, which is evident.
@@
 $z$, and we can use $|\ev\,n|\,\gamma$-many applications of $s\rel$ to $z\rel$
-to witness the preservation. The $\beta$-rules for $\msf{NatElim}$ are also respected
+to witness the goal type. The $\beta$-rules for $\msf{NatElim}$ are also respected
 by this definition.
@@
 \re\,A\,(\re\,\sigma\,\gamma')$. This is obtained directly from
-$A\rel\,(\sigma\rel\,\gamma\rel)$.  Similarly, we have
-\begin{alignat*}{4}
-  & (t[\sigma])\rel\,\gamma\rel &&:= t\rel\,(\sigma\rel\,\gamma\rel) \\
-  & (\sigma,\,t)\rel\,\gamma\rel &&:= (\sigma\rel\,\gamma\rel,\,t\rel\,\gamma\rel) \\
-  & \p\rel\,(\gamma\rel,\,\alpha\rel) &&:= \gamma\rel \\
-  & \q\rel\,(\gamma\rel,\,\alpha\rel) &&:= \alpha\rel. \\
+$A\rel\,(\sigma\rel\,\gamma\rel)$.  Similarly,
+\begin{alignat*}{6}
+  & (t[\sigma])\rel\,\gamma\rel &&:= t\rel\,(\sigma\rel\,\gamma\rel) && \p\rel\,(\gamma\rel,\,\alpha\rel) &&:= \gamma\rel \\
+  & (\sigma,\,t)\rel\,\gamma\rel &&:= (\sigma\rel\,\gamma\rel,\,t\rel\,\gamma\rel)\hspace{2em}&& \q\rel\,(\gamma\rel,\,\alpha\rel) &&:= \alpha\rel
 \end{alignat*}
@@
 external operations that can be extracted from $\blank\rel$.
-\begin{itemize}
-  \item For $\Gamma : \Con$, we get $|\Gamma\rel| : \{\Delta : \Cono\} \to |\ev\,\Gamma|\,\Delta \to
-    \Sub\,\emb{\Delta}\,\Gamma \to \Set_j$.
-  \item For $A : \Ty_{0,j}\,\Gamma$, we get $|A\rel| : |\Gamma\rel|\,\gamma\,\gamma' \to \emb{\ev\,A\,\gamma} = A[\gamma']$.
-  \item For $t : \Tm_{0,j}\,\Gamma\,A$, we get $|t\rel| : |\Gamma\rel|\,\gamma\,\gamma' \to \emb{\ev\,t\,\gamma} = t[\gamma']$.
-\end{itemize}
+\begin{alignat*}{5}
+  & \text{For}\hspace{1em} \Gamma &&: \Con \hspace{1em}&&\text{we get}\hspace{1em}
+         |\Gamma\rel| &&: \{\Delta : \Cono\} \to |\ev\,\Gamma|\,\Delta \to \Sub\,\emb{\Delta}\,\Gamma \to \Set_j\\
+  & \text{For}\hspace{1em} A &&: \Ty_{0,j}\,\Gamma \hspace{1em}&&\text{we get}\hspace{1em} |A\rel| &&: |\Gamma\rel|\,\gamma\,\gamma' \to \emb{\ev\,A\,\gamma} = A[\gamma']\\
+  & \text{For}\hspace{1em} t &&: \Tm_{0,j}\,\Gamma\,A \hspace{1em}&&\text{we get}\hspace{1em}|t\rel| &&: |\Gamma\rel|\,\gamma\,\gamma' \to \emb{\ev\,t\,\gamma} = t[\gamma']
+\end{alignat*}
 Since $\blank\rel$ was defined in $\hato$, we also know that the above are all
@@
 
-%% \begin{alignat*}{3}
-%%   & \hspace{1em} \emb{\Stage\,A} \hspace{2em}&& \\
-%%   & = \emb{|\ev\,A|\,\Gamma^P}   \hspace{2em}&& \text{by the definition of $\Stage$} \\
-%%   & = A                          \hspace{2em}&& \text{by $|A\rel|\,{\Gamma^P}\rel : \emb{|\ev\,A|\,\Gamma^P} = A[\id]$}
-%% \end{alignat*}
-
-%% to show:
-
-%%   ⌜Stage A⌝  = A
-%%   ⌜|F A| Γᴾ⌝ = A
-
-%%     |Aᴿ| Γᴾᴿ : ⌜|F A| Γᴾ⌝ = A[id] = A OK
-
-%%   ⌜Stage t⌝ = t
-%%   ⌜|F t| Γᴾ⌝ = t
-
-%%     |tᴿ| Γᴾᴿ :  ⌜|F t| Γᴾ⌝ = t[id] = t OK
+\subsubsection{Alternative presentations}
+The above conservativity proof could be rephrased as an instance of more modern
+techniques which let us implicitly handle stability under 2LTT substitutions as
+well, not just $\mbbo$ substitutions. This amounts to working in a \emph{modal}
+internal language, where modalities control the interaction of the different base
+categories.
+\begin{itemize}
+\item Using \emph{synthetic Tait computability} \cite{TODO}, we work in the internal language
+      of presheaves over the glued category $\msf{Id}_{\hato}\downarrow \re$.
+\item Using \emph{relative induction} \cite{TODO}, we work in the modal type theory
+      obtained from the dependent right adjoint functor $\re^* : \wh{\msf{2LTT}} \to \hato$,
+      where $\wh{\msf{2LTT}}$ denotes presheaves over the syntactic category of 2LTT.
+\end{itemize}
+We do not use either of these in this paper, for the following reasons. First,
+the author of the paper is not sufficiently familiar with the above
+techniques. Second, the task at hand is not too technically difficult, so using
+advanced techniques is not essential. Compare e.g.\ normalization for cubical
+type theories, which is infeasible to handle without the more synthetic
+presentations \cite{TODO}.
+
+\section{Intensional Analysis}\label{sec:intensional-analysis}
+
+We briefly discuss intensional analysis in this section. This means analyzing
+the internal structure of object-level terms, i.e.\ values of $\Lift\,A$. Disallowing
+intensional analysis is a major simplification, which is sometimes enforced
+in implementations, for example in MetaML \cite{metaml}.
+
+If we have sufficiently expressive inductive types in the meta-level language,
+it is possible to simply use inductive deeply embedded syntaxes, which can be
+analyzed; our example in Section \ref{TODO} for partially evaluating simply
+typed lambda calculus is like this. However, native metaprogramming features
+could be more concise and convenient than deep embeddings. Hence, intensional
+analysis could provide additional convenience.
+
+In the following, we look at the semantics of intensional analysis in the
+standard presheaf models. We only discuss basic semantics here, and leave
+practical considerations to future work.
+
+\begin{definition}
+The \textbf{Yoneda embedding} is a functor from $\mbbo$ to $\hato$, sending
+$\Gamma : \Cono$ to $\msf{y}\Gamma : \hCon$, such that $|\msf{y}\Gamma|\,\Delta
+= \Subo\,\Delta\,\Gamma$. We say that $\Gamma : \hCon$ is \emph{representable}
+if it is isomorphic to a Yoneda-embedded context.
+\end{definition}
+
+\begin{lemma}[Yoneda lemma]
+We have $\hSub\,(\msf{y}\Gamma)\,\Delta \simeq |\Delta|\,\Gamma$, and also
+$\hTm_{1,j}\,(\msf{y}\Gamma)\,A \simeq |A|\,\id$, where $\id :
+\Subo\,\Gamma\,\Gamma$.
+\end{lemma}
+
+The Yoneda lemma restricts possible dependencies on representable contexts.  For
+example, consider the staging behavior of $t : \Tm_{1}\,(\emptycon \ext (x :
+\Bool_0))\,\Bool_1$. Staging $t$ yields essentially a natural transformation
+$\hSub\,(\ev(\emptycon \ext \Bool_0))\,\mbb{B}$, where $\mbb{B}$ is the
+metatheoretic type with two elements, considered as a constant presheaf.  Now,
+if $\ev(\emptycon\,\ext\, \Bool_0)$ is representable, then
+$\hSub\,(\ev(\emptycon \ext \Bool_0))\,\mbb{B} \simeq \mbb{B}$, so $t$ can be
+staged in at most two different ways.
+
+Which contexts are representable? In the specific 2LTT in this paper,
+$\emb{\Gamma}$ is always representable in the presheaf
+model. $|\ev{\emb{\Gamma}}|\,\Delta$ contains lists of object-theoretic terms,
+hence $\ev{\emb{\Gamma}} \simeq \yon \Gamma$.  This implies that intensional
+analysis is \emph{not compatible} with the standard presheaf model, for our
+2LTT.
+
+Consider a very basic feature of intensional analysis, \emph{decidability of
+definitional equality} for object-level Boolean expressions.
+\[
+  \msf{decEq} : (x\,y : \Lift\,\Bool_0) \to (x =_1 y) +_1 (x \neq_1 y)
+\]
+However, Yoneda lemma says that $\msf{decEq}$ cannot actually decide
+definitional equality. We expect that $\msf{decEq}$ allows us to define
+non-constant maps from $\Lift\,\Bool_0$ to $\Bool_1$, but the Yoneda lemma says
+that we only have two terms with distinct semantics in $\Tm_1\,(\emptycon \ext_0
+\Bool_0)\,\Bool_1$.
+
+A more direct way to show infeasibility of $\msf{decEq}$ in $\hato$ is to note
+that definitional \emph{inequality} for object-level terms is not stable under
+substitution, since inequal variables can be mapped to equal terms.
+
+\subsubsection{Stability under weakening only}
+We can move to different 2LTT setups, where stability under substitution is not
+required. We may have \emph{weakenings} only as morphisms in $\mbbo$.  A
+weakening from $\Gamma$ to $\Delta$ describes how $\Gamma$ can be obtained by
+inserting zero or more entries to $\Delta$. A wide range of intensional analysis
+features are stable under weakenings. For example, $\msf{decEq}$ now holds in
+$\hato$ whenever object-theoretic definitional equality is in fact decidable.
+We also dodge the Yoneda lemma, since $\ev\,\emb{\Gamma}$ is not necessarily
+representable anymore: $|\ev\,\emb{\Gamma}|\,\Delta$ is still a set of lists of
+terms, but weakenings from $\Delta$ to $\Gamma$ are not lists of terms.
+
+As a trade-off, if there is no notion of substitution in the specification of the
+object theory, it is not possible to specify dependent or polymorphic types
+there. We need a substitution operation to describe dependent elimination or
+polymorphic instantiation.
+
+However, this is still sufficient for many practical use cases, for example the
+monomorphization setup in Section \ref{TODO}. Here, it makes sense to only
+having weakening in the object theory, but no $\beta\eta$-rules. We care about
+the intension of the staging output; presumably we are doing monomorphization
+because we care about code performance, so in 2LTT we do not want to
+definitionally equate object-level programs with different performance
+characteristics.
+
+\subsubsection{Closed modalities}
+Another option is to add a \emph{closed} or \emph{crisp} modality
+\cite{licata2018internal} for tracking object-level values which will become
+closed terms after staging. Since closed terms are not affected by substitution,
+we are free to analyze their intension. Such analysis can be viewed as an
+internalization of canonicity for closed object-level terms. For example, a
+closed $\Bool_0$ term will be either $\true_0$ or $\false_0$ after staging, so
+it makes sense to be able to eliminate from it to $\Bool_1$. This can be also
+viewed as a safe implementation of a \emph{run} function in the style of MetaML
+\cite{TODO}, where \emph{scope extrusion} is prevented by the closed modality.
+
+C-style \emph{function pointers} could be also an interesting use case, since
+these have to become closed after staging (as they cannot capture closure
+environments).
+
+\section{Related Work}\label{sec:related-work}
+
+2LTT
+
+PE and staging
+
+We don't do binding time analysis
+
+Two-level LC is a fragment of 2LTT
+
+Igarashi
+
+\section{Conclusions and Future Work}\label{sec:conclusions}
 

### 2022-02-21 paper/paper.tex
stuff

@@
 \newcommand{\false}{\msf{false}}
+\newcommand{\True}{\msf{True}}
+\newcommand{\False}{\msf{False}}
 \newcommand{\List}{\msf{List}}
@@
 %% end of the preamble, start of the body of the document source.
+\hypersetup{draft}
 \begin{document}
@@
   numerous good properties for this use case: it has a concise specification,
-  well-developed model theory, and it supports a wide range of language features
+  well-behaved model theory, and it supports a wide range of language features
   both at the object and the meta level. First, we give an overview of 2LTT's
@@
   conservativity of 2LLT over the object theory. To our knowledge, this is the
-  first system for staged compilation which supports full dependent types and
+  first description of staged compilation which supports full dependent types and
   unrestricted staging for types.
@@
 
-\subsection{Overview \& Contributions}
+\subsection{Overview \& Contributions}\label{sec:overview}
 
@@
 
-\subsection{Staged Programming in 2LTT}
+\subsection{Staged Programming in 2LTT}\label{sec:staged-programming}
 
@@
 
-\subsubsection{Monomorphization}
+\subsubsection{Monomorphization}\label{sec:monomorphization}
 
@@
 definitional equality with metatheoretic equality. First, we define the
-structural scaffolding of 2LTT without type formers. The following specification
-is agnostic about the sizes of $\Set$-s involved, i.e.\ we can model underlying
-sorts using any metatheoretic $\Set_j$.
+structural scaffolding of 2LTT without type formers, mostly following
+\cite{twolevel}. The specification is agnostic about the sizes of sets involved,
+i.e.\ we can model underlying sorts using any metatheoretic $\Set_j$.
 
@@
 \item \emph{Universes}. For each $i$ and $j$, we have a Coquand-style universe
-  \cite{TODO} in $\Ty_{i,j}$. This consists of $\U_{i,j} : \Ty_{i,j+1}\,\Gamma$,
+  \cite{coquand2018canonicity} in $\Ty_{i,j}$. This consists of $\U_{i,j} : \Ty_{i,j+1}\,\Gamma$,
   together with $\El : \Tm_{i,j+1}\,\Gamma\,\U_{i,j} \to \Ty_{i,j}\,\Gamma$ and $\Code$, where $\Code$
@@
 suitable notion of algebraic signatures, such as one for essentially algebraic
-theories \cite{TODO} or as a finitary quotient inductive-inductive signature
-\cite{TODO}. The initiality of syntax directly yields a notion of
-\emph{recursion on syntax}, and it can be shown that initiality also implies an
-\emph{induction principle} \cite{TODO}.
-\end{definition}
-
-The examples in Section \ref{TODO} can be faithfully represented in the formal
-syntax. The only notable difference is that the formal syntax uses Coquand-style
-universes instead of Russell-style ones. We use the former because it is much
-easier to model in the semantics in Section \ref{TODO}. We note though that it
-is straightforward to elaborate from a Russell-style surface syntax to the
-formal version, by inserting $\El$ and $\Code$ as required.
+theories \cite{partialhorn} or as a finitary quotient inductive-inductive
+signature \cite{kaposi2019constructing}. The initiality of syntax directly
+yields a notion of \emph{recursion on syntax}, and it can be shown that
+initiality also implies an \emph{induction principle}.
+\end{definition}
+
+The examples in Section \ref{sec:tour-of-2ltt} can be faithfully represented in
+the formal syntax. The only notable difference is that the formal syntax uses
+Coquand-style universes instead of Russell-style ones. We use the former because
+it is much easier to model in the semantics in Section \ref{sec:staging-algorithm}. We note
+though that it is straightforward to elaborate from a Russell-style surface
+syntax to the formal version, by inserting $\El$ and $\Code$ as required.
 
@@
 \end{alignat*}
-Embedding maps all type and term formers to the corresponding ones in 2LTT, and
-strictly preserves all structure.
+Embedding strictly preserves all structure.
 \end{definition}
@@
 
-\begin{definition}
+\begin{definition}\label{def:staging}
 A \textbf{staging algorithm} consists of two functions:
@@
   A staging algorithm $\Stage$ is \textbf{correct} if the following properties hold:
-  \begin{itemize}
+  \begin{enumerate}
   \item \emph{Soundness:} $\emb{\Stage\,A} = A$ and $\emb{\Stage\,t} = t$.
@@
   \item \emph{Stability:} $\Stage\,\emb{A} = A$ and $\Stage\,\emb{t} = t$.
-  \end{itemize}
+  \end{enumerate}
 \end{definition}
@@
 embedding is a bijection on types and terms. This is a statement of
-\emph{conservativity} of 2LTT over the object theory. In \cite{TODO} a
+\emph{conservativity} of 2LTT over the object theory. In \cite{twolevel} a
 significantly weaker conservativity theorem is shown, which expresses that there
@@
 on normalization-by-evaluation, where the soundness-completeness-stability trio
-has been used several times \cite{TODO}. In \cite{cubicalnbe}, stability is
-instead called ``idempotence''.
+has been used \cite{altenkirch1995categorical,altenkirch2016normalisation}. In
+\cite{cubicalnbe}, stability is instead called ``idempotence''.
 
@@
 will directly yield a closed staging algorithm, by evaluation of 2LTT types and
-terms in the model. Here, contexts are presheaves over the syntactic category of
-the object theory. We denote the components of the model by putting hats on 2LTT
-components, e.g.\ as in $\hCon$. We summarize the key components in the
-following.
+terms in the model. First, we give a higher-level overview of this semantics.
+
+TODO explanation here.
+%% Here, contexts are presheaves over the syntactic category of
+%% the object theory.
+
+We proceed to summarize the key components of the model in a bit more detail. We
+denote the components of the model by putting hats on 2LTT components, e.g.\ as
+in $\hCon$. We summarize the key components in the following.
 
@@
 \begin{definition}
-$\bs{\hTy_{1,j}\,\Gamma : \Set_\omega}$ is the set of \emph{displayed presheaves}
-over $\Gamma$; see e.g.\ \cite{TODO}. This is equivalent to the set of
-presheaves over the category of elements of $\Gamma$, but it is usually more
-convenient in calculations. An $A : \hTy\,\Gamma$ has an action on objects
-$|A| : \{a : \Cono\} \to |\Gamma|\,a \to \Set_j$ and an action on
-morphisms $\blank[\blank] : |A|\,\gamma \to (f : \Subo\,a\,b) \to
-|A|\,(\gamma[f])$, such that $\alpha[\id] = \alpha$ and $\alpha[f \circ g] = \alpha[f][g]$.
+$\bs{\hTy_{1,j}\,\Gamma : \Set_\omega}$ is the set of \emph{displayed
+presheaves} over $\Gamma$; see e.g.\ \cite[Chapter~1.2]{huber-thesis}. This is
+equivalent to the set of presheaves over the category of elements of $\Gamma$,
+but it is usually more convenient in calculations. An $A : \hTy\,\Gamma$ has
+an action on objects $|A| : \{a : \Cono\} \to |\Gamma|\,a \to \Set_j$ and an
+action on morphisms $\blank[\blank] : |A|\,\gamma \to (f : \Subo\,a\,b) \to
+|A|\,(\gamma[f])$, such that $\alpha[\id] = \alpha$ and $\alpha[f \circ g] =
+\alpha[f][g]$.
 
@@
 
-%% \begin{definition}
-%% $\bs{\wh{\U}_{1,j} : \hTy_{1,j+1}\,\Gamma}$ is defined by
-%% $|\wh{\U}_{1,j}|\,\{a\}\,\_ = \hTy_{1,j}\,(\y\,a)$, and defining action on
-%% morphisms by $\hTy$ substitution. Note that the universe levels check out in
-%% this definition, since by unfolding $\hTy_{1,j}\,(\y\,a)$ we can show that it
-%% is in $\Set_j$.
-%% \end{definition}
-
 We also look at he empty context and context extension with meta-level types,
@@
 
-\begin{definition}
+\begin{definition}\label{def:psh-emptycon}
 $\bs{{\widehat{\emptycon}} : \hCon}$ is defined as the presheaf
@@
 
-\begin{definition} For $A : \hTy_{1,j}\,\Gamma$, we define $\bs{\Gamma\,\,\wh{\ext}\,A}$
+\begin{definition}\label{def:psh-ext} For $A : \hTy_{1,j}\,\Gamma$, we define $\bs{\Gamma\,\,\wh{\ext}\,A}$
 pointwise by $|\Gamma\,\,\wh{\ext}\,A|\,a := (\gamma : |\Gamma|\,a) \times |A|\,\gamma$
@@
 Using the above definitions, we can model the syntactic category of 2LTT, and
-also the meta-level family structure and all meta-level type formers. For an
-exposition in previous literature, see \cite{TODO} and \cite{TODO}.
+also the meta-level family structure and all meta-level type formers. For
+expositions in previous literature, see \cite[Chapter~1.2]{huber-thesis} and
+\cite[Section~4.1]{Hofmann97syntaxand}.
 
@@
 
-Using the above definitions and following \cite{TODO}, we can model all type
+Using the above definitions and following \cite{twolevel}, we can model all type
 formers in $\hTy_{0,j}$. Intuitively, that is because $\hTy_{0,j}$ and
@@
 
-\begin{definition}
+\begin{definition}\label{def:psh-lifting}
 $\bs{\wh{\Lift} A : \hTy_{1,j}\,\Gamma}$ is defined as $\Tm_{\mbbo j}[A]$. With
@@
 
-\begin{definition} The \textbf{evaluation morphism, denoted $\bs{\ev}$} is
+\begin{definition} The \textbf{evaluation morphism, denoted $\bs{\ev}$}, is
 a 2LTT model morphism from the syntax of 2LTT to $\hato$, arising from the initiality
@@
 elements of $|\emb{\Gamma}|\,\Gamma$ are essentially lists of object-level terms
-in $\Gamma$, by Definition \ref{TODO}, so $|\emb{\Gamma}|\,\Gamma$ should be
+in $\Gamma$, by Definitions \ref{def:psh-ext} and \ref{def:psh-emptycon}, so $|\emb{\Gamma}|\,\Gamma$ should be
 isomorphic to $\Subo\,\Gamma\,\Gamma$.  It turns out that this falls out from
@@
 
-\begin{definition} We define \textbf{open staging} as follows.
+\begin{definition}\label{def:open-staging} We define \textbf{open staging} as follows.
 \begin{alignat*}{4}
@@
   \item In the non-dependent case, a function type $A\,\wh{\to}\,B$ is defined as the presheaf
-        exponential \cite{TODO}. The computational part of $|A\,\wh{\to}\,B|\,\{\Delta\}\,\gamma$
+        exponential. The computational part of $|A\,\wh{\to}\,B|\,\{\Delta\}\,\gamma$
         is given by a function with type
@@
 The exact same efficiency issue occurs in formal presheaf-based
-normalization-by-evaluation \cite{TODO}. However, in practical implementations
-\cite{TODO} this issue can be fully solved by using De Bruijn \emph{levels} in
-the semantic domain, thus arranging that weakening has no operational cost on
-semantic values. This corresponds to Abel's ``liftable terms'' presentation
-\cite{TODO}. We can use the same solution in staging, by using de Bruijn levels
-and delayed weakenings in object-level terms during staging. It is a somewhat
-unfortunate mismatch that indices are far more convenient in formalization, but
-levels are more efficient in practice.
+normalization-by-evaluation, such as in \cite{kaposinbe} and
+\cite{coquand2018canonicity}. However, in practical implementations this issue
+can be fully solved by using De Bruijn \emph{levels} in the semantic domain,
+thus arranging that weakening has no operational cost on semantic values; see
+\cite{coquand1996algorithm}, \cite{modulartc} and \cite{untypedtc} for
+implementations in this style. We can use the same solution in staging, by using
+de Bruijn levels and delayed weakenings in object-level terms during staging. It
+is a somewhat unfortunate mismatch that indices are far more convenient in
+formalization, but levels are more efficient in practice.
 
@@
 operational significance; recall that they are interpreted as identity functions
-in Definition \ref{TODO}. However, it seems more efficient to not compute types
+in Definition \ref{def:psh-lifting}. However, it seems more efficient to not compute types
 at all and instead rely on quotes and splices to keep track of the current
@@
 $\hato$. Additionally, $\hato$ supports all type formers of extensional type
-theory \cite{TODO}, and certain other structures which are stable under
-object-level substitution. As all constructions in this section must be stable,
-it makes sense to work internally to $\hato$. This style has been previously
-used in normalization proofs \cite{TODO} and also in the metatheory of cubical
-type theories \cite{TODO}.
+theory, and certain other structures which are stable under object-level
+substitution. As all constructions in this section must be stable, it makes
+sense to work internally to $\hato$. This style has been previously used in
+normalization proofs \cite{coquand2018canonicity} and also in the metatheory of
+cubical type theories
+\cite{licata2018internal,orton_et_al:LIPIcs:2016:6564,DBLP:conf/csl/CavalloMS20}.
 
@@
 $\Sigma$, extensional identity $\blank\!=\!\blank$ and inductive types. We use
-the same conventions as in Notation \ref{TODO}.
+the same conventions as in Notation \ref{basic-notation}.
 
@@
 
-\begin{definition} We specify the types of the \textbf{restriction operations} internally, and the $|\blank|$
+\begin{definition}\label{def:restriction} We specify the types of the \textbf{restriction operations} internally, and the $|\blank|$
 components of the operations externally. The naturality of $|\blank|$ is
@@
 In short, we need to show that $\msf{NatElim}$ preserves relations. Here we
-switch to the external view temporarily. By Definition \ref{TODO}, we know that
+switch to the external view temporarily. By Definition \ref{def:restriction}, we know that
 \[
@@
 
-\begin{theorem}[Soundness of staging] The open staging algorithm from Definition \ref{TODO} is sound.
+\begin{theorem}[Soundness of staging]
+The open staging algorithm from Definition \ref{def:open-staging} is sound.
 \end{theorem}
@@
 
-\begin{corollary}[Conservativity of 2LTT]
+\begin{corollary}[Conservativity of 2LTT]\label{conservativity}
 From the soundness and stability of staging, we get that $\emb{\blank}$ is
@@
 \begin{itemize}
-\item Using \emph{synthetic Tait computability} \cite{TODO}, we work in the internal language
+\item Using \emph{synthetic Tait computability} \cite{sterlingthesis}, we work in the internal language
       of presheaves over the glued category $\msf{Id}_{\hato}\downarrow \re$.
-\item Using \emph{relative induction} \cite{TODO}, we work in the modal type theory
+\item Using \emph{relative induction} \cite{bocquet2021relative}, we work in the modal type theory
       obtained from the dependent right adjoint functor $\re^* : \wh{\msf{2LTT}} \to \hato$,
@@
 type theories, which is infeasible to handle without the more synthetic
-presentations \cite{TODO}.
+presentations \cite{cubicalnbe}.
 
@@
 it is possible to simply use inductive deeply embedded syntaxes, which can be
-analyzed; our example in Section \ref{TODO} for partially evaluating simply
+analyzed; our example in Section \ref{sec:staged-programming} for partially evaluating simply
 typed lambda calculus is like this. However, native metaprogramming features
@@
 \begin{lemma}[Yoneda lemma]
-We have $\hSub\,(\msf{y}\Gamma)\,\Delta \simeq |\Delta|\,\Gamma$, and also
+We have $\hSub\,(\msf{y}\Gamma)\,\Delta \simeq |\Delta|\,\Gamma$
+\cite[Section~III.2]{maclane98categories}. Also, we have
 $\hTm_{1,j}\,(\msf{y}\Gamma)\,A \simeq |A|\,\id$, where $\id :
@@
 However, this is still sufficient for many practical use cases, for example the
-monomorphization setup in Section \ref{TODO}. Here, it makes sense to only
+monomorphization setup in Section \ref{sec:monomorphization}. Here, it makes sense to only
 having weakening in the object theory, but no $\beta\eta$-rules. We care about
@@
 
-\subsubsection{Closed modalities}
+\subsubsection{Closed modalities}\label{sec:closed-modalities}
 Another option is to add a \emph{closed} or \emph{crisp} modality
@@
 viewed as a safe implementation of a \emph{run} function in the style of MetaML
-\cite{TODO}, where \emph{scope extrusion} is prevented by the closed modality.
+\cite{metaml}, where \emph{scope extrusion} is prevented by the closed modality.
 
@@
 
-2LTT
-
-PE and staging
-
-We don't do binding time analysis
-
-Two-level LC is a fragment of 2LTT
-
-Igarashi
+The work Annekov et al.\ on two-level type theory \cite{twolevel}, building on
+the thesis of Capriotti \cite{capriotti2017models} and ideas from Voevodsky
+\cite{hts}, is the foremost inspiration in the current work. Interestingly, the
+above works do mention metaprogramming as a source of intuition for 2LTT, but
+they only briefly touch on this aspect, and in \cite{twolevel} the main focus is
+on extensions of basic 2LTT which do not have staging semantics
+anymore. Ibid.\ conjectures the strong conservativity of 2LTT in Proposition
+2.18, and suggests categorical glueing as proof strategy. In this work, we do
+essentially that: our presheaf model and the logical relation model could be
+merged into a single instance of gluing along the $\re$ morphism; this merges
+staging and its soundness proof. We split this construction to two parts in order
+to get a self-contained presentation of the staging algorithm.
+
+Our notation for quoting and splicing is borrowed from MetaML \cite{metaml}.  In
+the following, we compare 2LTT to MetaML, MetaOCaml \cite{kiselyov14metaocaml}
+and typed Template Haskell \cite{typed-th}.
+
+The main difference is the explicit stage tracking via universes. Existing
+systems all support type lifting (usually called $\msf{Code}$), quoting and
+splicing, but lifting does not change the universe of types. For example, we can
+write a function with type $\Bool \to \msf{Code}\,\Bool$, and the two $\Bool$
+occurrences refer to the exact same type.
+
+This yields some semantic and operational confusion down the line, and some
+additional disambiguation is needed for stages of binders. For example, in typed
+Template Haskell, top-level binders behave as runtime binders when used under a
+quote, but behave as static binders otherwise. Given a top-level definition $b =
+\True$, if we write a top-level definition $\msf{expr} = [||\,b\,||]$, then $b$
+is a \emph{variable} in the staging output, but if we write $\msf{expr} =
+\msf{if}\,b\,\msf{then}\,[||\,\True\,||]\,\msf{else}\,[||\,\False\,||]$, then
+$b$ is considered to be a static value. In contrast, 2LTT does not distinguish
+top-level and local binders, and in fact it has no syntactic or scope-based
+restrictions; everything is enforced by typing.
+
+In MetaML and MetaOCaml, there is a \emph{run} function for evaluating closed
+object-level terms: as we mentioned in Section \ref{sec:closed-modalities}, we
+believe that the proper way to implement it in 2LTT would be to use a closed
+modality. This use case is closely related to \cite{licata2018internal}. In
+ibid.\ a semantics is sketched for this modality in a presheaf topos, and we
+could quite directly adapt it to our presheaf model of 2LTT. That said, working
+out the practical details of this modality remains future work.
+
+Many existing staged systems also support \emph{side effects} at compile time,
+while our version of 2LTT does not. Here, general considerations about the
+mixing of dependent types and effects should apply; see
+e.g.\ \cite{fire-triangle}. However, it should be possible to use standard
+effect systems such as monads at the meta level.
+
+Additionally, 2LTT only supports two stages, while the other systems all
+allow countable stages. It remains future work to develop NLTT (N-level type
+theory), but this extension does appear to be straightforward. In the
+specification of NLTT, we would simply need to move from $i \in \{0,1\}$ to
+$i \in \mbb{N}$, although in the semantics there could be more subtleties.
+
+The multi-stage calculus of Kawata and Igarashi \cite{multi-stage-calculus}
+supports staging operations and some type dependencies. However, there are major
+differences to the current work. First, ibid.\ does not have universes or large
+elimination, nor does it allow staged manipulation of type expressions. Second,
+it does not support staging in the sense of Definition \ref{def:staging};
+rather, it is a system for runtime code manipulation and evaluation. In other
+words, programs may freely manipulate code fragments during execution, in more
+liberal ways than in 2LTT, but there is no judgment which ensures that a purely
+``object-level'' program can be extracted from a multi-stage program.
+
+We emphasize that a distinction exists between staged compilation and staged
+computation. For the former, we expect staging in the sense of Definition
+\ref{def:staging}; this is essential if we want to use staging as a lowering
+step in a compilation pipeline. Staged computation on the other hand may support
+runtime code construction and evaluation, but may not guarantee that these
+features can be computed away.
 

### 2022-02-22 paper/paper.tex
stuff

@@
 %% end of the preamble, start of the body of the document source.
-\hypersetup{draft}
+%% \hypersetup{draft}
 \begin{document}
@@
     staging, which means that staging the output of staging has no action. Then,
-    we extend staging to open types and terms which may depend on object-level
+    we extend staging to open types and terms, which may depend on object-level
     contexts.
@@
     of intensional code analysis in the presheaf model.
+  \item We provide a small prototype implementation which we briefly discuss in
+    Section \ref{sec:implementation}.
   \item To our knowledge, this is the first account of staged compilation in the
@@
 C++ and Rust programming languages. The evident difference is that in 2LTT we
-can use a full type theory at compile time, as opposed to languages which are
-less expressive and less principled. Additionally, we will see in Section
+can use a full type theory at compile time. Additionally, we will see in Section
 \ref{sec:intensional-analysis} that this setup is compatible with an induction principle for
@@
 will directly yield a closed staging algorithm, by evaluation of 2LTT types and
-terms in the model. First, we give a higher-level overview of this semantics.
-
-TODO explanation here.
-%% Here, contexts are presheaves over the syntactic category of
-%% the object theory.
+terms in the model.
+
+We give a higher-level overview first. Presheaves can be viewed as a
+generalization of sets, as ``variable sets'' which may vary over a base
+category. A set-based model of 2LTT yields a standard interpreter, where every
+syntactic object is mapped to the obvious semantic counterpart. The presheaf
+model over the category of contexts and substitutions in $\mbbo$ generalizes
+this to an interpreter where semantic values may depend on object-theoretic
+contexts. This allows us to handle object-theoretic \emph{types and terms}
+during interpretation, since these necessarily depend on their contexts.
+
+In the presheaf model, every semantic construction must be \emph{stable} under
+object-level substitution, or using alternative terminology, must be
+\emph{natural} with respect to substitution. Semantic values can be viewed as
+runtime objects during interpretation, and naturality means that
+object-theoretic substitutions can be \emph{applied} to such runtime objects,
+acting on embedded object-theoretic types and terms. Stability captures the core
+idea of staging:
+\begin{itemize}
+\item On one hand, every syntactic rule and construction in 2LTT must
+      be stable, which restricts the range of metaprograms
+      that can be written in 2LTT. For example, we cannot make decisions
+      based on sizes of the object-theoretic scopes, since these sizes may be changed
+      by substitution.
+\item On the other hand, if only stable constructions are possible, we never
+      have to \emph{prove} stability, or even explicitly mention object-theoretic
+      contexts, substitutions and variables. The tedious details of deep embeddings
+      can be replaced with a handful of staging operations and typing rules.
+\end{itemize}
+In principle, the base category in a 2LTT can be any category with finite
+products; it is not required that it is a category of contexts and substitutions.
+If the base category has more structure, we get more interesting object
+theories, but at the same time stability becomes more onerous, because we must
+be stable under \emph{more} morphisms. If the base category is simpler, with
+fewer morphisms, then the requirement of stability is less restrictive. In
+Section \ref{sec:intensional-analysis}, we will look at a concrete example for this.
 
@@
 follows. We define externally the presheaf of object-level types as
-$|\Ty_{0,j}|\,a := \Ty_{0,j}\,\emb{a}$, and the displayed presheaf of
-object-level terms over $\Ty_{0,j}$ as $|\Tm_{0,j}|\,\{a\}\,A :=
-\Tm_{0,j}\,\emb{a}\,A$. Hence, internally we have $\Ty_{0,j} : \whset_0$ and
+$|\Ty_{0,j}|\,\Gamma := \Ty_{0,j}\,\emb{\Gamma}$, and the displayed presheaf of
+object-level terms over $\Ty_{0,j}$ as $|\Tm_{0,j}|\,\{\Gamma\}\,A :=
+\Tm_{0,j}\,\emb{\Gamma}\,A$. Hence, internally we have $\Ty_{0,j} : \whset_0$ and
 $\Tm_{0,j} : \Ty_{0,j} \to \whset_0$. $\Ty_{0,j}$ is closed under all type formers,
@@
 interpretation. In short, $\re$ restricts 2LTT syntax so that it can only depend
-on object-level contexts, i.e.\ contexts given as $\emb{\Gamma}$.
+on object-level contexts, contexts given as $\emb{\Gamma}$.
 
@@
 Lastly, object-level type formers are straightforward. For types, we need
-$\emb{\ev\,A\,\gamma} = \re\,A\,\gamma'$, and for terms, we need
-$\emb{\ev\,A\,\gamma} = \re\,A\,\gamma'$. Note that $\ev$ and $\emb{\blank}$
-preserve all structure, and $\re$ preserves all structure on object-level types
-and terms. Hence, each object-level $A$ and $t$ case in $\blank\rel$ trivially
-follows from induction hypotheses.
+$\emb{\ev\,A\,\gamma} = \re\,A\,\gamma'$, and likewise for terms. Note that
+$\ev$ and $\emb{\blank}$ preserve all structure, and $\re$ preserves all
+structure on object-level types and terms. Hence, each object-level $A$ and $t$
+case in $\blank\rel$ trivially follows from induction hypotheses.
 
@@
 it is possible to simply use inductive deeply embedded syntaxes, which can be
-analyzed; our example in Section \ref{sec:staged-programming} for partially evaluating simply
-typed lambda calculus is like this. However, native metaprogramming features
-could be more concise and convenient than deep embeddings. Hence, intensional
-analysis could provide additional convenience.
+analyzed; our example in Section \ref{sec:staged-programming} for partially
+evaluating simply typed lambda calculus is like this. However, native
+metaprogramming features could be more concise and convenient than deep
+embeddings, similarly to how staged programming is more convenient than code
+generation with deeply embedded syntaxes. Hence, intensional analysis could
+provide additional convenience.
 
@@
 \]
-However, Yoneda lemma says that $\msf{decEq}$ cannot actually decide
-definitional equality. We expect that $\msf{decEq}$ allows us to define
-non-constant maps from $\Lift\,\Bool_0$ to $\Bool_1$, but the Yoneda lemma says
-that we only have two terms with distinct semantics in $\Tm_1\,(\emptycon \ext_0
+However, the Yoneda lemma says that $\msf{decEq}$ cannot actually decide
+definitional equality. We expect that $\msf{decEq}$ lets us define non-constant
+maps from $\Lift\,\Bool_0$ to $\Bool_1$, but the Yoneda lemma implies that we
+only have two terms with distinct semantics in $\Tm_1\,(\emptycon \ext_0
 \Bool_0)\,\Bool_1$.
@@
 monomorphization setup in Section \ref{sec:monomorphization}. Here, it makes sense to only
-having weakening in the object theory, but no $\beta\eta$-rules. We care about
+have weakening in the object theory, but no $\beta\eta$-rules. We care about
 the intension of the staging output; presumably we are doing monomorphization
@@
 closed terms after staging. Since closed terms are not affected by substitution,
-we are free to analyze their intension. Such analysis can be viewed as an
-internalization of canonicity for closed object-level terms. For example, a
-closed $\Bool_0$ term will be either $\true_0$ or $\false_0$ after staging, so
-it makes sense to be able to eliminate from it to $\Bool_1$. This can be also
-viewed as a safe implementation of a \emph{run} function in the style of MetaML
-\cite{metaml}, where \emph{scope extrusion} is prevented by the closed modality.
+we are free to analyze them. Such analysis can be viewed as an internalization
+of canonicity for closed object-level terms. For example, a closed $\Bool_0$
+term will be either $\true_0$ or $\false_0$ after staging, so it makes sense to
+be able to eliminate from it to $\Bool_1$. This can be also viewed as a safe
+implementation of a \emph{run} function in the style of MetaML \cite{metaml}.
+In our case \emph{scope extrusion} would be prevented by the closed modality.
 
 C-style \emph{function pointers} could be also an interesting use case, since
-these have to become closed after staging (as they cannot capture closure
+these must become closed after staging (as they cannot capture closure
 environments).
@@
 anymore. Ibid.\ conjectures the strong conservativity of 2LTT in Proposition
-2.18, and suggests categorical glueing as proof strategy. In this work, we do
+2.18, and suggests categorical gluing as proof strategy. In this work, we do
 essentially that: our presheaf model and the logical relation model could be
@@
 
+The multi-stage calculus of Kawata and Igarashi \cite{multi-stage-calculus}
+supports staging operations and type dependencies. However, there are major
+differences to the current work. First, ibid.\ does not have universes or large
+elimination, nor does it allow staged manipulation of type expressions. Second,
+it does not support staging in the sense of Definition \ref{def:staging};
+rather, it is a system for runtime code manipulation and evaluation. In other
+words, programs may manipulate code fragments during execution, in more liberal
+ways than in 2LTT, but there is no judgment which ensures that a purely
+``object-level'' program can be extracted from a multi-stage program. We make a
+disctinction between staged compilation and staged computation. For the former,
+we expect staging in the sense of Definition \ref{def:staging}; this is
+essential if we want to use staging as a lowering step in a compilation
+pipeline. Staged computation on the other hand may support runtime code
+construction and evaluation, but may not guarantee that these features can be
+entirely computed away.
+
+Idris 1 supports compile-time partial evaluation, through \texttt{static}
+annotations on function arguments \cite{scrap-your-inefficient-engine}. This can
+be used for dependently typed code generation, but Idris 1 does not guarantee
+that partial evaluation adequately progresses. For example, we can pass a neutral
+value as static argument in Idris 1, while in 2LTT static function arguments
+are always canonical during staging.
+
 Our notation for quoting and splicing is borrowed from MetaML \cite{metaml}.  In
@@
 
-The main difference is the explicit stage tracking via universes. Existing
-systems all support type lifting (usually called $\msf{Code}$), quoting and
-splicing, but lifting does not change the universe of types. For example, we can
-write a function with type $\Bool \to \msf{Code}\,\Bool$, and the two $\Bool$
+2LTT explicitly track stages of types, in contrast to the mentioned
+systems. There, we have type lifting (usually called $\msf{Code}$), quoting and
+splicing, but lifting does not change the universe of types. We can write a
+function with type $\Bool \to \msf{Code}\,\Bool$, and the two $\Bool$
 occurrences refer to the exact same type.
 
-This yields some semantic and operational confusion down the line, and some
-additional disambiguation is needed for stages of binders. For example, in typed
-Template Haskell, top-level binders behave as runtime binders when used under a
-quote, but behave as static binders otherwise. Given a top-level definition $b =
+This yields some operational confusion down the line, and additional
+disambiguation is needed for stages of binders. For example, in typed Template
+Haskell, top-level binders behave as runtime binders when used under a quote,
+but behave as static binders otherwise. Given a top-level definition $b =
 \True$, if we write a top-level definition $\msf{expr} = [||\,b\,||]$, then $b$
@@
 
+2LTT supports staging for types and dependent types, while to our knowledge no
+previous system for staged compilation does the same. It appears to the author
+that the main missing component in previous systems is \emph{dependent types at
+the meta level}, rather than staging operations; staging operations in 2LTT and
+MetaML are already quite similar. For example, any interesting operation on
+the static-length vectors in Section \ref{sec:staged-programming} requires
+meta-level dependent types, even if the object language is simply typed.
+
 In MetaML and MetaOCaml, there is a \emph{run} function for evaluating closed
@@
 ibid.\ a semantics is sketched for this modality in a presheaf topos, and we
-could quite directly adapt it to our presheaf model of 2LTT. That said, working
+could directly adapt it to our presheaf model of 2LTT. That said, working
 out the practical details of this modality remains future work.
+
+Additionally, 2LTT only supports two stages, while the other noted systems allow
+countable stages. It remains future work to develop NLTT (N-level type theory),
+but this extension does appear to be straightforward. In the specification of
+NLTT, we would simply need to move from $i \in \{0,1\}$ to $i \in \mbb{N}$,
+although in the semantics there could be more subtleties.
 
@@
 
-Additionally, 2LTT only supports two stages, while the other systems all
-allow countable stages. It remains future work to develop NLTT (N-level type
-theory), but this extension does appear to be straightforward. In the
-specification of NLTT, we would simply need to move from $i \in \{0,1\}$ to
-$i \in \mbb{N}$, although in the semantics there could be more subtleties.
-
-The multi-stage calculus of Kawata and Igarashi \cite{multi-stage-calculus}
-supports staging operations and some type dependencies. However, there are major
-differences to the current work. First, ibid.\ does not have universes or large
-elimination, nor does it allow staged manipulation of type expressions. Second,
-it does not support staging in the sense of Definition \ref{def:staging};
-rather, it is a system for runtime code manipulation and evaluation. In other
-words, programs may freely manipulate code fragments during execution, in more
-liberal ways than in 2LTT, but there is no judgment which ensures that a purely
-``object-level'' program can be extracted from a multi-stage program.
-
-We emphasize that a distinction exists between staged compilation and staged
-computation. For the former, we expect staging in the sense of Definition
-\ref{def:staging}; this is essential if we want to use staging as a lowering
-step in a compilation pipeline. Staged computation on the other hand may support
-runtime code construction and evaluation, but may not guarantee that these
-features can be computed away.
+\section{Implementation}\label{sec:implementation}
 
@@
 
+TODO: psh explain, intro overview upd, conclusion, implementation description, related work type staging
+
 \bibliographystyle{ACM-Reference-Format}
+\interlinepenalty=10000
 \bibliography{references}

### 2022-02-22 paper/paper.tex
start demo impl

@@
 \documentclass[acmsmall]{acmart}
+%% \documentclass[nonacm,acmsmall]{acmart}
 %% \raggedbottom
@@
 %% complete the rights form.
+
 \setcopyright{acmcopyright}
@@
 %% These commands are for a JOURNAL article.
-\acmJournal{JACM}
-\acmVolume{37}
-\acmNumber{4}
-\acmArticle{111}
-\acmMonth{8}
+
+%% \acmJournal{JACM}
+%% \acmVolume{37}
+%% \acmNumber{4}
+%% \acmArticle{111}
+%% \acmMonth{8}
 
@@
 
-2LTT explicitly track stages of types, in contrast to the mentioned
+2LTT explicitly tracks stages of types, in contrast to the mentioned
 systems. There, we have type lifting (usually called $\msf{Code}$), quoting and
@@
 
-This yields some operational confusion down the line, and additional
+This causes some operational confusion down the line, and additional
 disambiguation is needed for stages of binders. For example, in typed Template

### 2022-02-22 paper_proto/Elaboration.hs

@@ -1,0 +1,162 @@
+
+module Elaboration (check, infer) where
+
+import Control.Exception
+import Control.Monad
+
+import qualified Data.Map.Strict as M
+
+import Common
+import Cxt
+import Errors
+import Evaluation
+import Metacontext
+import Syntax
+import Unification
+import Value
+
+import qualified Presyntax as P
+
+--------------------------------------------------------------------------------
+
+freshMeta :: Cxt -> VTy -> IO Tm
+freshMeta cxt a = do
+  let ~closed = eval [] $ closeTy (path cxt) (quote (lvl cxt) a)
+  m <- newMeta closed
+  pure $ AppPruning (Meta m) (pruning cxt)
+
+unifyCatch :: Cxt -> Val -> Val -> IO ()
+unifyCatch cxt t t' =
+  unify (lvl cxt) t t'
+  `catch` \UnifyError ->
+    throwIO $ Error cxt $ CantUnify (quote (lvl cxt) t) (quote (lvl cxt) t')
+
+-- | Insert fresh implicit applications.
+insert' :: Cxt -> IO (Tm, VTy) -> IO (Tm, VTy)
+insert' cxt act = go =<< act where
+  go (t, va) = case force va of
+    VPi x Impl a b -> do
+      m <- freshMeta cxt a
+      let mv = eval (env cxt) m
+      go (App t m Impl, b $$ mv)
+    va -> pure (t, va)
+
+-- | Insert fresh implicit applications to a term which is not
+--   an implicit lambda (i.e. neutral).
+insert :: Cxt -> IO (Tm, VTy) -> IO (Tm, VTy)
+insert cxt act = act >>= \case
+  (t@(Lam _ Impl _), va) -> pure (t, va)
+  (t               , va) -> insert' cxt (pure (t, va))
+
+-- | Insert fresh implicit applications until we hit a Pi with
+--   a particular binder name.
+insertUntilName :: Cxt -> Name -> IO (Tm, VTy) -> IO (Tm, VTy)
+insertUntilName cxt name act = go =<< act where
+  go (t, va) = case force va of
+    va@(VPi x Impl a b) -> do
+      if x == name then
+        pure (t, va)
+      else do
+        m <- freshMeta cxt a
+        let mv = eval (env cxt) m
[…]

### 2022-02-22 paper_proto/Main.hs

@@ -1,0 +1,94 @@
+
+module Main where
+
+import Control.Exception
+import System.Environment
+import System.Exit
+
+import Common
+import Cxt
+import Errors
+import Evaluation
+import Metacontext
+import Parser
+import Pretty
+import Elaboration
+
+import qualified Presyntax as P
+
+--------------------------------------------------------------------------------
+
+helpMsg = unlines [
+  "usage: elabzoo-pruning [--help|nf|type]",
+  "  --help : display this message",
+  "  elab   : read & elaborate expression from stdin",
+  "  nf     : read & typecheck expression from stdin, print its normal form and type",
+  "  type   : read & typecheck expression from stdin, print its type"]
+
+mainWith :: IO [String] -> IO (P.Tm, String) -> IO ()
+mainWith getOpt getRaw = do
+
+  let elab = do
+        (t, file) <- getRaw
+        infer (emptyCxt (initialPos file)) t
+          `catch` \e -> displayError file e >> exitSuccess
+
+  reset
+  getOpt >>= \case
+    ["--help"] -> putStrLn helpMsg
+    ["nf"]   -> do
+      (t, a) <- elab
+      putStrLn $ showTm0 $ nf [] t
+      putStrLn "  :"
+      putStrLn $ showTm0 $ quote 0 a
+    ["type"] -> do
+      (t, a) <- elab
+      putStrLn $ showTm0 $ quote 0 a
+    ["elab"] -> do
+      (t, a) <- elab
+      displayMetas
+      putStrLn $ showTm0 t
+    _ -> putStrLn helpMsg
+
+main :: IO ()
+main = mainWith getArgs parseStdin
+
+-- | Run main with inputs as function arguments.
+main' :: String -> String -> IO ()
+main' mode src = mainWith (pure [mode]) ((,src) <$> parseString src)
+
+
+ex1 :: IO ()
+ex1 = main' "elab" $ unlines [
+
+  "-- Examples for unification with ordered metacontext & pruning",
+  "--------------------------------------------------------------------------------",
+
+  "-- we use Eq to test unification",
+  "let Eq : {A : U} → A → A → U",
+  "    = λ {A} x y. (P : A → U) → P x → P y;",
+  "let refl : {A : U}{x : A} → Eq {A} x x",
+  "    = λ _ px. px;",
+
+  "-- inline type annotation",
[…]

### 2022-02-22 paper_proto/Parser.hs

@@ -1,0 +1,131 @@
+
+module Parser (parseString, parseStdin) where
+
+import Control.Applicative hiding (many, some)
+import Control.Monad
+import Data.Char
+import Data.Void
+import System.Exit
+import Text.Megaparsec
+
+import qualified Text.Megaparsec.Char       as C
+import qualified Text.Megaparsec.Char.Lexer as L
+
+import Common
+import Presyntax
+
+--------------------------------------------------------------------------------
+
+type Parser = Parsec Void String
+
+ws :: Parser ()
+ws = L.space C.space1 (L.skipLineComment "--") (L.skipBlockComment "{-" "-}")
+
+withPos :: Parser Tm -> Parser Tm
+withPos p = SrcPos <$> getSourcePos <*> p
+
+lexeme     = L.lexeme ws
+symbol s   = lexeme (C.string s)
+char c     = lexeme (C.char c)
+parens p   = char '(' *> p <* char ')'
+braces p   = char '{' *> p <* char '}'
+pArrow     = symbol "→" <|> symbol "->"
+pBind      = pIdent <|> symbol "_"
+
+keyword :: String -> Bool
+keyword x = x == "let" || x == "λ" || x == "U"
+
+pIdent :: Parser Name
+pIdent = try $ do
+  x <- takeWhile1P Nothing isAlphaNum
+  guard (not (keyword x))
+  x <$ ws
+
+pKeyword :: String -> Parser ()
+pKeyword kw = do
+  C.string kw
+  (takeWhile1P Nothing isAlphaNum *> empty) <|> ws
+
+pAtom :: Parser Tm
+pAtom  =
+      withPos (    (Var <$> pIdent)
+               <|> (U <$ char 'U')
+               <|> (Hole <$ char '_'))
+  <|> parens pTm
+
+pArg :: Parser (Either Name Icit, Tm)
+pArg =  (try $ braces $ do {x <- pIdent; char '='; t <- pTm; pure (Left x, t)})
+    <|> ((Right Impl,) <$> (char '{' *> pTm <* char '}'))
+    <|> ((Right Expl,) <$> pAtom)
+
+pSpine :: Parser Tm
+pSpine = do
+  h <- pAtom
+  args <- many pArg
+  pure $ foldl (\t (i, u) -> App t u i) h args
+
[…]

### 2022-02-22 paper_proto/Unification.hs

@@ -1,0 +1,225 @@
+
+module Unification (unify) where
+
+import Control.Monad
+import Control.Exception
+import Data.IORef
+import qualified Data.IntMap as IM
+import qualified Data.IntSet as IS
+
+import Common
+import Errors
+import Evaluation
+import Metacontext
+import Syntax
+import Value
+
+--------------------------------------------------------------------------------
+
+-- partial renaming from Γ to Δ
+data PartialRenaming = PRen {
+    occ :: Maybe MetaVar   -- optional occurs check
+  , dom :: Lvl             -- size of Γ
+  , cod :: Lvl             -- size of Δ
+  , ren :: IM.IntMap Lvl}  -- mapping from Δ vars to Γ vars
+
+-- lift : (σ : PRen Γ Δ) → PRen (Γ, x : A[σ]) (Δ, x : A)
+lift :: PartialRenaming -> PartialRenaming
+lift (PRen occ dom cod ren) = PRen occ (dom + 1) (cod + 1) (IM.insert (unLvl cod) dom ren)
+
+-- skip : PRen Γ Δ → PRen Γ (Δ, x : A)
+skip :: PartialRenaming -> PartialRenaming
+skip (PRen occ dom cod ren) = PRen occ dom (cod + 1) ren
+
+-- | invert : (Γ : Cxt) → (spine : Sub Δ Γ) → PRen Γ Δ
+--   Optionally returns a pruning of nonlinear spine entries, if there's any.
+invert :: Lvl -> Spine -> IO (PartialRenaming, Maybe Pruning)
+invert gamma sp = do
+
+  let go :: Spine -> IO (Lvl, IS.IntSet, IM.IntMap Lvl, Pruning, Bool)
+      go []             = pure (0, mempty, mempty, [], True)
+      go (sp :> (t, i)) = do
+        (dom, domvars, ren, pr, isLinear) <- go sp
+        case force t of
+          VVar (Lvl x) -> case IS.member x domvars of
+            True  -> pure (dom + 1, domvars,             IM.delete x ren,     Nothing : pr, False   )
+            False -> pure (dom + 1, IS.insert x domvars, IM.insert x dom ren, Just i  : pr, isLinear)
+          _ -> throwIO UnifyError
+
+  (dom, domvars, ren, pr, isLinear) <- go sp
+  pure (PRen Nothing dom gamma ren, pr <$ guard isLinear)
+
+
+-- | Remove some arguments from a closed iterated Pi type.
[…]

### 2022-02-24 paper/paper.tex
impl progress

@@
 
-Note that $\wh{\emptycon}$ is defined as the terminal presheaf, which is
-constantly $\top$. This leads us to the following definition.
-
 \begin{definition} \textbf{Closed staging} is defined as follows.
@@
 \end{alignat*}
-This is well-typed, since $|\ev\,A|\{\emptycon\} : |\ev\,\emptycon|\,\emptycon \to \Ty_{\mbbo
-  j}\,\emptycon$, so $|\ev\,A|\{\emptycon\} : \top \to \Ty_{\mbbo
-  j}\,\emptycon$, and $|\ev\,t|\,\{\emptycon\} : \top \to \Tm_{\mbbo
+Note that $\wh{\emptycon}$ is defined as the terminal presheaf, which is
+constantly $\top$. Also, $|\ev\,A|\{\emptycon\} : |\ev\,\emptycon|\,\emptycon \to \Ty_{\mbbo
+  j}\,\emptycon$, therefore $|\ev\,A|\{\emptycon\} : \top \to \Ty_{\mbbo
+  j}\,\emptycon$ and $|\ev\,t|\,\{\emptycon\} : \top \to \Tm_{\mbbo
   j}\,\emptycon\,(|\ev\,A|\,\tt)$.
@@
 
-Stability means that staging an object-level construction does nothing. We
+Stability means that staging an object-level type or term does nothing. We
 define a family of functions $\blank^P$ by induction on object syntax. The
@@
   \item $|\Sigma_1\,A\,B|\,\gamma$ is simply a set of pairs of values.
-  \item In the non-dependent case, a function type $A\,\wh{\to}\,B$ is defined as the presheaf
+  \item A non-dependent function type $A\,\wh{\to}\,B$ is defined as the presheaf
         exponential. The computational part of $|A\,\wh{\to}\,B|\,\{\Delta\}\,\gamma$
@@
         representation of $\id$, such that $\gamma[\id]$ immediately computes to
-        $\gamma$. The dependent case is analogous operationally.
+        $\gamma$. The case of dependent functions is analogous operationally.
 \end{itemize}

### 2022-02-24 paper_proto/Evaluation.hs

@@ -12,20 +12,34 @@ infixl 8 $$
 
 vApp :: Val -> Val -> Icit -> Val
 vApp t ~u i = case t of
-  VLam _ _ t  -> t $$ u
-  VFlex  m sp -> VFlex m  (sp :> (u, i))
-  VRigid x sp -> VRigid x (sp :> (u, i))
-  _           -> impossible
+  VLam _ _ _ t -> t $$ u
+  VFlex  m sp  -> VFlex m  (SApp sp u i)
+  VRigid x sp  -> VRigid x (SApp sp u i)
+  _            -> impossible
 
 vAppSp :: Val -> Spine -> Val
 vAppSp t = \case
-  []           -> t
-  sp :> (u, i) -> vApp (vAppSp t sp) u i
+  SId         -> t
+  SApp sp u i -> vApp (vAppSp t sp) u i
+  SSplice sp  -> vSplice (vAppSp t sp)
+
+vQuote :: Val -> Val
+vQuote = \case
+  VFlex  m (SSplice sp) -> VFlex m sp
+  VRigid x (SSplice sp) -> VRigid x sp
+  t                     -> VQuote t
+
+vSplice :: Val -> Val
+vSplice = \case
+  VQuote t    -> t
+  VFlex m sp  -> VFlex m (SSplice sp)
+  VRigid x sp -> VRigid x (SSplice sp)
+  _           -> impossible
 
 vMeta :: MetaVar -> Val
 vMeta m = case lookupMeta m of
-  Solved v _ -> v
-  Unsolved{} -> VMeta m
+  Solved v _ _ -> v
+  Unsolved{}   -> VMeta m
 
 vAppPruning :: Env -> Val -> Pruning -> Val
 vAppPruning env ~v pr = case (env, pr) of
@@ -41,18 +55,23 @@ vVar env x = error $ "index out of env: "
 
 eval :: Env -> Tm -> Val
 eval env = \case
-  Var x           -> vVar env x
-  App t u i       -> vApp (eval env t) (eval env u) i
-  Lam x i t       -> VLam x i (Closure env t)
-  Pi x i a b      -> VPi x i (eval env a) (Closure env b)
-  Let _ _ t u     -> eval (env :> eval env t) u
-  U               -> VU
-  Meta m          -> vMeta m
-  AppPruning t pr -> vAppPruning env (eval env t) pr
+  Var x             -> vVar env x
[…]

### 2022-02-24 paper_proto/Main.hs

@@ -19,35 +19,50 @@ import qualified Presyntax as P
 --------------------------------------------------------------------------------
 
 helpMsg = unlines [
-  "usage: elabzoo-pruning [--help|nf|type]",
+  "usage: 2ltt [--help] COMMAND",
+  "",
+  "Options:",
   "  --help : display this message",
-  "  elab   : read & elaborate expression from stdin",
-  "  nf     : read & typecheck expression from stdin, print its normal form and type",
-  "  type   : read & typecheck expression from stdin, print its type"]
+  "  --show-insertions : displayed inserted implicit lambdas and applications",
+  "",
+  "All commands read input from stdin",
+  "Commands:",
+  "elab         : print elaboration output",
+  "elab-verbose : print elaboration output, show metavariables and inserted implicits",
+  "stage        : print staging output",
+  "nf           : print normal form",
+  "type         : print normal type of input"
+  ]
 
 mainWith :: IO [String] -> IO (P.Tm, String) -> IO ()
 mainWith getOpt getRaw = do
 
   let elab = do
         (t, file) <- getRaw
-        infer (emptyCxt (initialPos file)) t
+        inferS (emptyCxt (initialPos file)) t S0
           `catch` \e -> displayError file e >> exitSuccess
 
   reset
   getOpt >>= \case
     ["--help"] -> putStrLn helpMsg
-    ["nf"]   -> do
-      (t, a) <- elab
-      putStrLn $ showTm0 $ nf [] t
-      putStrLn "  :"
-      putStrLn $ showTm0 $ quote 0 a
-    ["type"] -> do
-      (t, a) <- elab
-      putStrLn $ showTm0 $ quote 0 a
     ["elab"] -> do
       (t, a) <- elab
+      putStrLn "-- metavariables "
+      putStrLn "------------------------------------------------------------"
[…]

### 2022-02-24 paper_proto/Parser.hs

@@ -22,7 +22,11 @@ ws :: Parser ()
 ws = L.space C.space1 (L.skipLineComment "--") (L.skipBlockComment "{-" "-}")
 
 withPos :: Parser Tm -> Parser Tm
-withPos p = SrcPos <$> getSourcePos <*> p
+withPos p = do
+  pos <- getSourcePos
+  p >>= \case
+    t@Pos{} -> pure t
+    t       -> pure $ Pos (coerce pos) t
 
 lexeme     = L.lexeme ws
 symbol s   = lexeme (C.string s)
@@ -33,7 +37,7 @@ pArrow     = symbol "→" <|> symbol "->"
 pBind      = pIdent <|> symbol "_"
 
 keyword :: String -> Bool
-keyword x = x == "let" || x == "λ" || x == "U"
+keyword x = x == "let" || x == "λ" || x == "U0" || x == "U1"
 
 pIdent :: Parser Name
 pIdent = try $ do
@@ -48,9 +52,12 @@ pKeyword kw = do
 
 pAtom :: Parser Tm
 pAtom  =
-      withPos (    (Var <$> pIdent)
-               <|> (U <$ char 'U')
-               <|> (Hole <$ char '_'))
+      withPos (    (Var    <$> pIdent)
+               <|> (U S0   <$  symbol "U0")
+               <|> (U S1   <$  symbol "U1")
+               <|> (Hole   <$  char '_')
+               <|> (Quote  <$> (char '<' *> pTm <* char '>'))
+               <|> (Splice <$> (char '[' *> pTm <* char ']')))
   <|> parens pTm
 
 pArg :: Parser (Either Name Icit, Tm)
@@ -58,17 +65,24 @@ pArg =  (try $ braces $ do {x <- pIdent; char '='; t <- pTm; pure (Left x, t)})
     <|> ((Right Impl,) <$> (char '{' *> pTm <* char '}'))
     <|> ((Right Expl,) <$> pAtom)
 
-pSpine :: Parser Tm
-pSpine = do
+pApps :: Parser Tm
+pApps = do
   h <- pAtom
   args <- many pArg
   pure $ foldl (\t (i, u) -> App t u i) h args
 
-pLamBinder :: Parser (Name, Either Name Icit)
+pLift :: Parser Tm
+pLift = Lift <$> (char '^' *> pAtom)
+
[…]

### 2022-02-24 paper_proto/Presyntax.hs

@@ -4,12 +4,15 @@ module Presyntax where
 import Common
 
 data Tm
-  = Var Name                       -- x
-  | Lam Name (Either Name Icit) Tm -- \x. t | \{x}. t | \{x = y}. t
-  | App Tm Tm (Either Name Icit)   -- t u  | t {u} | t {x = u}
-  | U                              -- U
-  | Pi Name Icit Tm Tm             -- (x : A) -> B | {x : A} -> B
-  | Let Name Tm Tm Tm              -- let x : A = t; u
-  | SrcPos SourcePos Tm            -- source position for error reporting
-  | Hole                           -- _
+  = Var Name                                   -- x
+  | Lam Name (Maybe Tm) (Either Name Icit)Tm   -- \x. t | \{x}. t | \{x = y}. t
+  | App Tm Tm (Either Name Icit)               -- t u  | t {u} | t {x = u}
+  | U Stage                                    -- U i
+  | Pi Name Icit Tm Tm                         -- (x : A) -> B | {x : A} -> B
+  | Let Stage Name (Maybe Tm) Tm Tm
+  | Pos (DontShow SourcePos) Tm                -- source position for error reporting
+  | Hole                                       -- _
+  | Lift Tm
+  | Quote Tm
+  | Splice Tm
   deriving Show

### 2022-02-24 paper_proto/Syntax.hs

@@ -20,8 +20,8 @@ revPruning = RevPruning . reverse
 -- | A "context zipper", used for efficiently creating types for fresh metas.
 data Path
   = Here
-  | Define Path Name ~Ty ~Tm
-  | Bind Path Name ~Ty
+  | Define Path Name ~Ty Stage ~Tm
+  | Bind Path Name ~Ty Stage
   deriving Show
 
 -- | Convert type in context to a closed iterated Pi type.  Note: we need `Tm`
@@ -30,17 +30,24 @@ data Path
 --   anything.
 closeTy :: Path -> Ty -> Ty
 closeTy mcl b = case mcl of
-  Here             -> b
-  Bind mcl x a     -> closeTy mcl (Pi x Expl a b)
-  Define mcl x a t -> closeTy mcl (Let x a t b)
+  Here               -> b
+  Bind mcl x a s     -> closeTy mcl (Pi x Expl a b)
+  Define mcl x a s t -> closeTy mcl (Let s x a t b)
 
 data Tm
   = Var Ix
-  | Lam Name Icit Tm
+  | Lam Name Icit Tm Tm
   | App Tm Tm Icit
-  | AppPruning Tm Pruning  -- ^ Used for applying an inserted or pruned meta to a mask of the scope.
-  | U
+  | AppPruning Tm Pruning  -- ^ Used for applying a pruned meta to a mask of the scope.
+  | InsertedMeta MetaVar Pruning
   | Pi Name Icit Ty Ty
-  | Let Name Ty Tm Tm
+  | Let Stage Name Ty Tm Tm
   | Meta MetaVar
+
+  | U Stage
+  | Quote Tm
+  | Splice Tm
+  | Lift Ty
+
+  | Wk Tm -- ^ Explicit weakening, used in subtyping coercions
   deriving Show

### 2022-02-25 paper/paper.tex
stuff

@@
 
-In principle, it should be possible to have \emph{dependent memory layouts};
-this would be a more advanced variation on type-passing polymorphism
-\cite{intensional-poly}, where layouts of objects may depend on runtime data. For example,
-length-prefixed flat arrays would not be a primitive notion, but simply defined
-as $(n : \Nat) \times \msf{Vec}\,n\,A$. However, it seems to be highly
-challenging to implement runtime systems and precise garbage collection in this
-setting. See \texttt{sixten} \cite{sixten} for an experimental implementation of
-dependent memory layouts.
+%% In principle, it should be possible to have \emph{dependent memory layouts};
+%% this would be a more advanced variation on type-passing polymorphism
+%% \cite{intensional-poly}, where layouts of objects may depend on runtime data. For example,
+%% length-prefixed flat arrays would not be a primitive notion, but simply defined
+%% as $(n : \Nat) \times \msf{Vec}\,n\,A$. However, it seems to be highly
+%% challenging to implement runtime systems and precise garbage collection in this
+%% setting. See \texttt{sixten} \cite{sixten} for an experimental implementation of
+%% dependent memory layouts.
 
@@
 denote the components of the model by putting hats on 2LTT components, e.g.\ as
-in $\hCon$. We summarize the key components in the following.
+in $\hCon$.
 
@@
 elimination, nor does it allow staged manipulation of type expressions. Second,
-it does not support staging in the sense of Definition \ref{def:staging};
-rather, it is a system for runtime code manipulation and evaluation. In other
-words, programs may manipulate code fragments during execution, in more liberal
-ways than in 2LTT, but there is no judgment which ensures that a purely
-``object-level'' program can be extracted from a multi-stage program. We make a
-disctinction between staged compilation and staged computation. For the former,
-we expect staging in the sense of Definition \ref{def:staging}; this is
-essential if we want to use staging as a lowering step in a compilation
-pipeline. Staged computation on the other hand may support runtime code
-construction and evaluation, but may not guarantee that these features can be
-entirely computed away.
+it does not support staged compilation in the sense of Definition
+\ref{def:staging}; rather, it is a system for runtime code manipulation and
+evaluation (which we may call staged \emph{computation}). In other words,
+programs may manipulate code fragments during execution, in more liberal ways
+than in 2LTT, but there is no judgment which ensures that a purely
+``object-level'' program can be extracted from a multi-stage program.
+
+%% We make a
+%% disctinction between staged compilation and staged computation. For the former,
+%% we expect staging in the sense of Definition \ref{def:staging}; this is
+%% essential if we want to use staging as a lowering step in a compilation
+%% pipeline. Staged computation on the other hand may support runtime code
+%% construction and evaluation, but may not guarantee that these features can be
+%% entirely computed away.
 

### 2022-02-25 paper_proto/Errors.hs

@@ -1,56 +1,0 @@
-
-module Errors where
-
-import Control.Exception
-import Text.Printf
-
-import Common
-import Cxt
-import Syntax
-
---------------------------------------------------------------------------------
-
-data UnifyError = UnifyError
-  deriving (Show, Exception)
-
-data ElabError
-  = NameNotInScope Name
-  | CantUnify Tm Tm
-  | InferNamedLam
-  | NoNamedImplicitArg Name
-  | IcitMismatch Icit Icit
-  | InferStage
-  deriving (Show, Exception)
-
-data Error = Error Cxt ElabError
-  deriving (Show, Exception)
-
-displayError :: String -> Error -> IO ()
-displayError file (Error cxt e) = do
-
-  let SourcePos path (unPos -> linum) (unPos -> colnum) = pos cxt
-      lnum = show linum
-      lpad = map (const ' ') lnum
-      msg = case e of
-        NameNotInScope x ->
-          "Name not in scope: " ++ x
-        CantUnify t t'   ->
-          ("Cannot unify expected type\n\n" ++
-           "  " ++ showTm cxt t' ++ "\n\n" ++
-           "with inferred type\n\n" ++
-           "  " ++ showTm cxt t)
-        InferNamedLam ->
-          "Cannot infer type for lambda with named argument"
-        NoNamedImplicitArg name ->
-          "No named implicit argument with name " ++ name
-        IcitMismatch i i' -> printf (
-          "Function icitness mismatch: expected %s, got %s.")
-          (show i) (show i')
-        InferStage ->
-          "Cannot infer stage"
-
-  printf "%s:%d:%d:\n" path linum colnum
-  printf "%s |\n"    lpad
-  printf "%s | %s\n" lnum (lines file !! (linum - 1))
-  printf "%s | %s\n" lpad (replicate (colnum - 1) ' ' ++ "^")
-  printf "%s\n" msg

### 2022-02-25 paper_proto/Metacontext.hs

@@ -1,48 +1,0 @@
-
-module Metacontext where
-
-import Data.IORef
-import System.IO.Unsafe
-import qualified Data.IntMap as IM
-
-import Common
-import Value
-
---------------------------------------------------------------------------------
-
-data MetaEntry
-  = Solved Val ~VTy Stage  -- ^ Solution, type.
-  | Unsolved ~VTy Stage    -- ^ Type.
-
-nextMetaVar :: IORef MetaVar
-nextMetaVar = unsafeDupablePerformIO $ newIORef 0
-{-# noinline nextMetaVar #-}
-
-newMeta :: VTy -> Stage -> IO MetaVar
-newMeta ~a st = do
-  m <- readIORef nextMetaVar
-  writeIORef nextMetaVar $! m + 1
-  modifyIORef mcxt $ IM.insert (coerce m) (Unsolved a st)
-  pure m
-
-type MCxt = IM.IntMap MetaEntry
-
-mcxt :: IORef MCxt
-mcxt = unsafeDupablePerformIO $ newIORef mempty
-{-# noinline mcxt #-}
-
-readMeta :: MetaVar -> IO MetaEntry
-readMeta m = do
-  ms <- readIORef mcxt
-  case IM.lookup (coerce m) ms of
-    Just e  -> pure e
-    Nothing -> impossible
-
-lookupMeta :: MetaVar -> MetaEntry
-lookupMeta = unsafeDupablePerformIO . readMeta
-
--- | Reset all mutable refs to initial state.
-reset :: IO ()
-reset = do
-  writeIORef nextMetaVar 0
-  writeIORef mcxt mempty

### 2022-02-25 paper_proto/Pretty.hs

@@ -1,142 +1,0 @@
-
-module Pretty (showTm0, showTm, showTopTm, displayMetas) where
-
-import Control.Monad
-import Data.IORef
-import Text.Printf
-
-import qualified Data.IntMap.Strict as IM
-
-import Common
-import Evaluation
-import Metacontext
-import Syntax
-
---------------------------------------------------------------------------------
-
-fresh :: [Name] -> Name -> Name
-fresh ns "_" = "_"
-fresh ns x   = if elem x ns then x else go (1::Int)
-  where go n | elem (x ++ show n) ns = go (n + 1)
-             | True                  = x ++ show n
-
-
--- printing precedences
-atomp = 3  :: Int -- U, var
-appp  = 2  :: Int -- application
-pip   = 1  :: Int -- pi
-letp  = 0  :: Int -- let, lambda
-
--- Wrap in parens if expression precedence is lower than
--- enclosing expression precedence.
-par :: Int -> Int -> ShowS -> ShowS
-par p p' = showParen (p' < p)
-
-braces :: ShowS -> ShowS
-braces s = ('{':).s.('}':)
-
-assign :: Stage -> ShowS
-assign = \case S0 -> (":="++); _ -> ("="++)
-
-piBind ns x Expl a = ('(':) . (x++) . (" : "++) . go letp ns a .(')':)
-piBind ns x Impl a = braces  ((x++) . (" : "++) . go letp ns a)
-
-lamBind x Impl = braces (x++)
-lamBind x Expl = (x++)
-
-goPr :: Int -> [Name] -> [Name] -> Tm -> Pruning -> ShowS
-goPr p topNs ns t pr = goPr' p ns pr (0 :: Int) where
-  goPr' p ns pr x = case (ns, pr) of
-    ([]      , []           ) -> go p topNs t
-    (ns :> n , pr :> Just i ) -> par p appp $ goPr' appp ns pr (x + 1) . (' ':)
-                                 . icit i braces id (case n of "_" -> (("@"++show x)++); n -> (n++))
-    (ns :> n , pr :> Nothing) -> goPr' appp ns pr (x + 1)
-    _                         -> impossible
-
-goIx :: [Name] -> Ix -> ShowS
-goIx ns topIx = go ns topIx where
-  go [] _ = impossible
-  go ("_":ns) 0 = (("@"++show topIx)++)
-  go (n:ns)   0 = (n++)
-  go (n:ns)   x = go ns (x - 1)
-
[…]

### 2022-02-25 paper_proto/Syntax.hs

@@ -1,53 +1,0 @@
-
-module Syntax where
-
-import Common
-
-type Ty = Tm
-
--- | A `Pruning` represents a spine of variables, which contains a subsequence
---   of all variables in scope. A `Just` represents application to a var, a `Nothing`
---   skips over a var.
-type Pruning = [Maybe Icit]
-
--- | A reversed pruning. Used for pruning Pi domains, where we have to iterate
---   inside-out.
-newtype RevPruning = RevPruning Pruning
-
-revPruning :: Pruning -> RevPruning
-revPruning = RevPruning . reverse
-
--- | A "context zipper", used for efficiently creating types for fresh metas.
-data Path
-  = Here
-  | Define Path Name ~Ty Stage ~Tm
-  | Bind Path Name ~Ty Stage
-  deriving Show
-
--- | Convert type in context to a closed iterated Pi type.  Note: we need `Tm`
---   and `Ty` in path in order to make this operation efficient. With this, we
---   can simply move things over from `Path` without having to rename or quote
---   anything.
-closeTy :: Path -> Ty -> Ty
-closeTy mcl b = case mcl of
-  Here               -> b
-  Bind mcl x a s     -> closeTy mcl (Pi x Expl a b)
-  Define mcl x a s t -> closeTy mcl (Let s x a t b)
-
-data Tm
-  = Var Ix
-  | Lam Name Icit Tm Tm
-  | App Tm Tm Icit
-  | AppPruning Tm Pruning  -- ^ Used for applying a pruned meta to a mask of the scope.
-  | InsertedMeta MetaVar Pruning
-  | Pi Name Icit Ty Ty
-  | Let Stage Name Ty Tm Tm
-  | Meta MetaVar
-
-  | U Stage
-  | Quote Tm
-  | Splice Tm
-  | Lift Ty
-
-  | Wk Tm -- ^ Explicit weakening, used in subtyping coercions
-  deriving Show

### 2022-02-25 paper_proto/Value.hs

@@ -1,29 +1,0 @@
-
-module Value where
-
-import Common
-import Syntax
-
-type Env     = [Val]
-data Closure = Closure Env Tm
-type VTy     = Val
-
-data Spine
-  = SId
-  | SApp Spine ~Val Icit
-  | SSplice Spine
-
-data Val
-  = VFlex MetaVar Spine
-  | VRigid Lvl Spine
-  | VLam Name Icit ~VTy {-# unpack #-} Closure
-  | VPi Name Icit ~VTy {-# unpack #-} Closure
-  | VU Stage
-  | VLift Val
-  | VQuote Val
-
-pattern VVar :: Lvl -> Val
-pattern VVar x = VRigid x SId
-
-pattern VMeta :: MetaVar -> Val
-pattern VMeta m = VFlex m SId

### 2022-02-28 old/types2021/abstract.tex
reorganize folders

--- 2021-04-22 types2021/abstract.tex
@@
 
-Two-level type theory \cite{twolevel} (2LTT) is a system for performing
-metatheoretic constructions and reasoning involving certain object-level type
+Two-level type theory \cite{twolevel} (2LTT) is a system for performing certain
+metatheoretic constructions and reasoning involving object-level type
 theories. Such reasoning is always possible by simply embedding object theories
@@
 object-level constructions remain in the result. Annekov et al. \cite{twolevel}
-showed this without uniqueness of $A'$ and $t'$ (as a conservativity theorem for
-2LTT). We can get uniqueness as well, using the normalization of 2LTT: by
-induction on the (unique) normal forms of $A$ and $t$, we can show that they
+only showed mere existence of $A'$ and $t'$ (as a conservativity theorem for
+2LTT). We can get unique existence as well, using the normalization of 2LTT:
+by induction on the (unique) normal forms of $A$ and $t$, we can show that they
 cannot contain meta-level subterms. This shows that normalization is a sound
@@
 meta-level type $\Ty^0$ whose terms are identified with object-level types. Now,
-meta-level functions can be used for quantification over object-level types, as in
-$\mathit{id} : (A : \Ty^0) \to \Code\,\sim\!A \to \Code\,\sim\!A$. However,
-since the object theory is simply typed and monomorphic, all polymorphism is
-guaranteed to compute away during staging.
+meta-level functions can be used for quantification over object-level types, as
+in $\mathit{id} : (A : \Ty^0) \to \Code\,A \to \Code\,A$. However, since the
+object theory is simply typed and monomorphic, all polymorphism is guaranteed to
+compute away during staging.
 
@@
 $\V^0$ which is closed under inductive types but not functions, and a universe
-$\C^0$ which has $V^0$ as a sub-universe, and is closed under functions with
+$\C^0$ which has $\V^0$ as a sub-universe, and is closed under functions with
 domains in $\V^0$ and codomains in $\C^0$. This object theory supports
@@
 without elimination principles. However, if we are interested in staged
-compilation, we can assume the object-level to be syntactic and consistently add
+compilation, we can assume the object level to be syntactic and consistently add
 operations on $\Code\,A$ which rely on that assumption, e.g. conversion

### 2022-02-28 notes/ClosedMonadicEval.hs

@@ -1,73 +1,0 @@
-{-# language OverloadedStrings #-}
-
-import GHC.Stack
-import Data.Maybe
-import Data.String
-
-type Name = String
-
-data Tm
-  = Var Name
-  | App Tm Tm
-  | Lam Name Tm
-  | Let Name Tm Tm
-  | Bind Name Tm Tm
-  | Return Tm
-  | Hello                  -- Hello : M ()
-  | Tt                     -- Tt : Unit
-  deriving (Eq, Show)
-
-data Val
-  = VVar Name
-  | VApp Val Val
-  | VLam Env Name Tm
-  | VBind Env Name Tm Tm
-  | VReturn Val
-  | VHello
-  | VTt
-  deriving (Eq, Show)
-
-type Env = [(Name, Val)]
-
---------------------------------------------------------------------------------
-
-instance IsString Tm where fromString = Var
-($$)  = App; infixl 7 $$
-(>>>) = Bind "_"; infixl 1 >>>
-
---------------------------------------------------------------------------------
-
-impossible :: HasCallStack => a
-impossible = error "impossible"
-
--- Pure closed evaluation
-eval :: Env -> Tm -> Val
-eval e = \case
-  Var x      -> fromJust $ lookup x e
-  App t u    -> case eval e t of
-                  VLam e' x t -> eval ((x, eval e u):e') t
-                  _           -> impossible
-  Lam x t    -> VLam e x t
-  Let x t u  -> eval ((x, eval e t):e) u
-  Bind x t u -> VBind e x t u
-  Return t   -> VReturn (eval e t)
-  Hello      -> VHello
-  Tt         -> VTt
-
--- Perform 1 layer of IO
-runIO :: Val -> IO Val
-runIO = \case
-  VReturn t     -> pure t
-  VBind e x t u -> do {t <- runIO (eval e t); runIO (eval ((x, t):e) u)}
-  VHello        -> VTt <$ putStrLn "hello"
-  _             -> impossible
-
-run :: Tm -> IO Val
-run = runIO . eval []
-
---------------------------------------------------------------------------------
-
-p1 :: Tm
-p1 =
-  Let "f" (Lam "m" $ Hello >>> "m") $
-  "f" $$ Hello

### 2022-02-28 notes/FixFusion.hs

@@ -1,56 +1,0 @@
-
-{-# language
-  StandaloneDeriving, DeriveFunctor, GADTs, TypeFamilies, AllowAmbiguousTypes, TypeApplications,
-  Strict, LambdaCase, DataKinds, PolyKinds, RankNTypes, ScopedTypeVariables, BlockArguments,
-  PartialTypeSignatures, UnicodeSyntax
-  #-}
-
-{-# options_ghc -Wno-unused-imports #-}
-
-newtype Fix f = Fix {fold :: ∀ x. (f x → x) → x}
-
-newtype Comp f g a = Comp {unComp :: f (g a)}
-
-data ListF a b = NilF | ConsF a b deriving (Show, Functor)
-
-fuse1 :: (∀ x. (g x → x) → (f x → x)) → Fix f → Fix g
-fuse1 f g = Fix \con → fold g (f con)
-
-fuse2 :: (∀ x. (g x → x) → (f (a → x) → a → x)) → Fix f → a → Fix g
-fuse2 f g a = Fix \con → fold g (f con) a
-
-type List a = Fix (ListF a)
-
-fuse3 :: (∀ x. (g x → x) → (f (h x) → h x)) → Fix f → h (Fix g)
-fuse3 f g = _
-
-rev :: List a -> List a
-rev as = Fix \con →
-  fold as
-    (\case NilF      → \acc → acc
-           ConsF a k → \acc → k (con (ConsF a acc)))
-    (con NilF)
-
-
-
-toList :: List a -> [a]
-toList as = fold as \case
-  NilF       → []
-  ConsF a as → a : as
-
-fromList :: [a] -> List a
-fromList as = Fix \con -> foldr (\a x -> con (ConsF a x)) (con NilF) as
-
-instance Show a => Show (List a) where
-  show = show . toList
-
-
-
-
--- data ListF a k = NilF | ConsF a k deriving (Show, Functor)
-
--- fold :: Functor f ⇒ (f a → a) -> Fix f → a
--- fold f (Fix ff) = let go = fold f in f (go <$> ff)
-
--- fuse1 :: (Functor f, Functor g) ⇒ (∀ x. (g x → x) → (f x → x)) → Fix f → Fix g
--- fuse1 f (Fix ff) = _

### 2022-02-28 notes/Fusion.hs

@@ -1,319 +1,0 @@
-
-{-# language
-  StandaloneDeriving, DeriveFunctor, GADTs, TypeFamilies, AllowAmbiguousTypes, TypeApplications,
-  Strict, LambdaCase, DataKinds, PolyKinds, RankNTypes, ScopedTypeVariables, BlockArguments,
-  PartialTypeSignatures
-  #-}
-
-{-# options_ghc -Wno-unused-imports #-}
-
-module Fusion (foo)  where
-
-import Control.Monad
-import Prelude hiding (zipWith, foldr, foldl, sum, zip, reverse)
-import qualified Prelude as P
-import GHC.Exts hiding (toList)
-
-{-
-Summary:
-
-  - fold fusion
-    + easy, optimal speed without compiler magic, full bind, works on all inductive types
-    - no zip
-
-  - stream fusion
-    + has zip, perhaps easier to freely extend with ad-hoc capabilities
-      (e.g. random access, end access, ranges)
-    - requires either compiler magic (call pattern spec) or mutation (see strymonas)
-      for optimal performance
-    - full bind is incompatible with zip (in the absence of closures)
-      -> weaker bind: Stream s a -> (a -> Stream s' b) -> Stream (Bind a b s s') b
-
-  - hyperfunctions: just a different flavor of stream fusion, not particularly worth it
-
-Questions:
-
-  - how to get rid of call pattern spec with staging
-  - how to mix push/pull in the same system (and do we even want that)
-
--}
-
-
-
-newtype Push a = Push {fold :: forall l. (a -> l -> l) -> l -> l}
-
-instance Functor Push where
-  fmap f (Push as) = Push \c n -> as (\a bs -> (c $! f a) bs) n
-  {-# inline fmap #-}
-
-data Pull a = forall s. Pull (s -> Maybe (a, s)) s
-deriving instance Functor Pull
-
-instance Applicative Push where
-  pure = return
-  (<*>) = ap
-
-instance Monad Push where
-  return a = Push \cons nil -> cons a nil
-  Push as >>= f = Push \c n -> as (\a -> fold (f a) c) n
-
-push :: Foldable t => t a -> Push a
-push ta = Push \cons nil -> P.foldr cons nil ta
[…]

### 2022-02-28 notes/GHCComp.hs

@@ -1,21 +1,0 @@
-{-# language MagicHash #-}
-
-module GHCComp where
-
-import GHC.Exts
-
--- data List = Nil | Cons Int# !List
-
--- foo :: List -> List
--- foo Nil = Nil
--- foo (Cons x xs) = Cons (x *# x) (foo xs)
-
--- foo :: Int# -> Int# -> Int# -> Int# -> Int#
--- foo a b c d = a *# b *# c *# d
-
-bar :: Int# -> Int# -> Int#
-bar a b = a *# b
-{-# noinline bar #-}
-
-foo :: Int# -> Int# -> Int#
-foo a b = a *# (bar a b)

### 2022-02-28 notes/Hyper.hs

@@ -1,106 +1,0 @@
-
-{-# language
-  DeriveFunctor, GADTs, TypeFamilies, AllowAmbiguousTypes, TypeApplications, Strict, LambdaCase,
-  DataKinds, PolyKinds, RankNTypes, ScopedTypeVariables, BlockArguments #-}
-
-module Fusion where
-
--- import GHC.Exts
-
-import Prelude hiding (sum, foldl, foldr, zipWith, reverse, map, filter)
-
-data HF a b =
-  forall s. Pack (s -> forall r. (b -> r) -> ((a -> b) -> s -> r) -> r) s
-
-lift :: (a -> b) -> HF a b
-lift f = Pack (\s left right -> right f s) ()
-
-comp :: HF b c -> HF a b -> HF a c
-comp (Pack f s) (Pack f' s') = Pack
-  (\(s, s') left right -> f s
-    (\c -> left c)
-    (\k s -> f' s'
-      (\b -> left (k b))
-      (\k' s' -> right (k . k') (s, s'))
-    )
-  )
-  (s, s')
-
-fix :: (a -> a) -> a
-fix f = let x = f x in x
-
-run :: HF a a -> a
-run (Pack f s) =
-  (fix $ \go s -> f s (\a -> a) (\k s -> k (go s))) s
-
--- call pattern spec still needed!
-appendHF :: HF a b -> HF a b -> HF a b
-appendHF (Pack f s) (Pack f' s') = Pack
-  (\s end step  -> case s of
-      Left s   -> f s   (\b     -> step (\_ -> b) (Right s'))
-                        (\ab s  -> step ab (Left s))
-      Right s' -> f' s' (\b     -> end b)
-                        (\ab s' -> step ab (Right s'))
-  )
-  (Left s)
-
-cons :: (a -> b) -> HF a b -> HF a b
-cons f (Pack step s) = Pack
-  (\x left right -> case x of
-      Nothing -> right f (Just s)
-      Just s  -> step s
-        (\b -> left b)
-        (\g s -> right g (Just s))
-  )
-  Nothing
-
-newtype Stream a = Stream {unStream :: forall b c. (a -> b -> c) -> c -> HF b c}
-
-map :: (a -> b) -> Stream a -> Stream b
-map f (Stream g) = Stream $ \cons nil -> g (\a b -> cons (f a) b) nil
-
-append :: Stream a -> Stream a -> Stream a
[…]

### 2022-02-28 notes/gcnotes.c

@@ -1,71 +1,0 @@
-
-#include <stdio.h>
-#include <stdint.h>
-#include <stdlib.h>
-
-
-typedef struct {
-  uint64_t payload;
-} obj;
-
-typedef struct {
-  uint64_t* hp;
-  uint64_t* hplim;
-} hp_info;
-
-extern hp_info perform_gc(hp_info hpi, uint64_t* frame) __attribute__ ((cold));
-
-/* int64_t foo(uint64_t* hp, uint64_t* hplim, uint64_t* frame, int64_t a, int64_t b){ */
-/*   return a * b; */
-/* } */
-
-static inline uint64_t get_tag(obj obj){
-  return obj.payload >> 58;
-}
-
-static inline uint64_t* to_ptr(obj obj){
-  return (uint64_t*)(((int64_t)(obj.payload) << 19) >> 16);
-}
-
-static inline obj pack_obj(uint64_t tag, uint64_t* ptr){
-  obj res = {((uint64_t)ptr >> 3) | (tag << 58)};
-  return res;
-}
-
-/* data List = Nil | Cons I64 List */
-
-static inline obj nil(){
-  return pack_obj(0, NULL);
-}
-
-typedef struct {
-  obj res;
-  hp_info hpi;
-} obj_res;
-
-static inline obj_res cons(hp_info hpi, uint64_t* frame, uint64_t x, obj* xs){
-  hp_info hpi2;
-  if (hpi.hp + 2 >= hpi.hplim){
-    hpi2 = perform_gc(hpi, frame);
-  }
-  *(hpi2.hp)     = x;
-  *(hpi2.hp + 1) = xs->payload;
-  obj_res res = {pack_obj(1, hpi2.hp), hpi2};
-  return res;
-}
-
-obj_res map(hp_info hpi, uint64_t* frame, obj* xsptr){
-  uint64_t  tag = get_tag(*xsptr);
-  if (tag == 0){
-    obj_res res = {nil(), hpi};
-    return res;
-  } else {
-    uint64_t* ptr = to_ptr(*xsptr);
-    uint64_t head = *ptr;
-    obj tail      = {*(ptr + 1)};
-    obj_res res1  = map(hpi, frame, &tail);
-    obj_res res2  = cons(res1.hpi, frame, head, &(res1.res));
-    obj_res res3  = cons(res2.hpi, frame, 100, &(res2.res));
-    return res3;
-  }
-}

### 2022-02-28 notes/inline_asm.c

@@ -1,33 +1,0 @@
-
-#include <inttypes.h>
-#include <stdio.h>
-
-register uint64_t* Hp asm("r12");
-
-
-
-int kek(uint64_t* ptr, uint64_t x){
-  *ptr = x;
-  return 0;
-}
-
-uint64_t foo() {
-
-  register uint64_t x asm("rbx") = 0;
-  register uint64_t y asm("rcx") = 20;
-
-  asm volatile
-     ("addq %1, %0; addq %1, %0"
-      : "=r" (y)
-      : "r"  (x));
-
-  return x;
-}
-
-uint64_t bar(uint64_t x){
-  return x + 100;
-}
-
-int main(){
-  printf("%ld\n", foo());
-}

### 2022-02-28 notes/notes.c

@@ -1,46 +1,0 @@
-
-#include <stdio.h>
-#include <stdint.h>
-#include <stdlib.h>
-
-typedef struct {
-  uint8_t  tag     ;  // 6
-  uint8_t  objtype ;  // 3
-  uint8_t  nobjs   ;  // 5
-  uint8_t  nwords  ;  // 5
-  uint64_t ptr     ;  // 45
-} unpacked_obj;
-
-typedef struct {
-  uint64_t payload;
-} packed_obj;
-
-static inline packed_obj pack_obj(unpacked_obj obj){
-  packed_obj res = {
-       (((uint64_t)obj.tag)     << 58)
-     | (((uint64_t)obj.objtype) << 55)
-     | (((uint64_t)obj.nobjs)   << 50)
-     | (((uint64_t)obj.nwords)  << 45)
-     | obj.ptr};
-  return res;
-}
-
-static inline void* to_ptr(packed_obj obj){
-  return (void*)(((int64_t)(obj.payload) << 19) >> 16);
-}
-
-static inline uint8_t get_tag(packed_obj obj){
-  return (uint8_t)((obj.payload) >> 58);
-}
-
-static inline uint64_t compr_ptr(void* ptr){
-  return (((uint64_t) ptr) << 16) >> 19;
-}
-
-int main() {
-  int* ptr = malloc(sizeof(int));
-  *ptr = 100;
-  unpacked_obj test = {10, 0, 0, 0, compr_ptr(ptr)};
-  printf("%d\n", *((int*)(to_ptr(pack_obj(test)))));
-  return 0;
-}

### 2022-02-28 notes/sections.c

@@ -1,45 +1,0 @@
-#include <stdio.h>
-#include <inttypes.h>
-
-/*
-Solution for putting vars in .text:
-
-https://stackoverflow.com/questions/58455300/assembler-warning-with-gcc-warning-when-placing-data-in-text
-
-*/
-
-/* char c1 = 'a'; */
-/* /\* int f1(int n) {return n + 200;} *\/ */
-/* char c2 = 'b'; */
-/* char c3 = 'c'; */
-
-/* long long mynum1 __attribute__ ((section (".text#"))) = 1200000; */
-/* long long mynum2 __attribute__ ((section (".text#"))) = 999999; */
-
-typedef struct {
-  int arity;
-  int foobar;
-} myinfo;
-
-uint64_t fun1_info __attribute__ ((section (".text#"))) = 41;
-uint64_t fun1(uint64_t x){
-  return x + 100 + x;
-}
-
-uint64_t fun2_info __attribute__ ((section (".text#"))) = 42;
-uint64_t fun2(uint64_t x){
-  return x + 300 + x;
-}
-
-myinfo fun3_info __attribute__ ((section (".text#"))) = {.arity = 100, .foobar = 0};
-uint64_t fun3(uint64_t x){
-  return x + 300 + x;
-}
-
-int64_t get_info_table(void* fun){
-  return *((int64_t*)(fun - 8));
-  }
-
-int main(){
-  printf("%ld\n", get_info_table((void*)(&fun2)));
-}

### 2022-02-28 old/mono_staged/src/Common.hs

@@ -1,0 +1,218 @@
+
+module Common (
+    module Common
+  , B.ByteString
+  , FlatParse.Stateful.Pos(..)
+  , FlatParse.Stateful.Span(..)
+  , unpackUTF8
+  , packUTF8
+  , coerce
+  , module Control.Monad
+  , module Data.Kind
+  , module Lens.Micro
+  ) where
+
+import qualified Data.ByteString as B
+import Control.Monad
+import Data.Bits
+import Data.Coerce
+import Data.Foldable
+import Data.Kind
+import Data.List
+import FlatParse.Stateful
+import GHC.Exts
+import Lens.Micro
+
+#ifdef DEBUG
+import GHC.Stack
+#endif
+
+-- Debug printing, toggled by "debug" cabal flag
+--------------------------------------------------------------------------------
+
+-- define DEBUG
+
+#ifdef DEBUG
+type Dbg = HasCallStack
+
+debug :: [String] -> IO ()
+debug strs = U.io $ putStrLn (intercalate " | " strs ++ " END")
+
+debugging :: IO () -> IO ()
+debugging act = act
+{-# inline debugging #-}
+#else
+type Dbg = () :: Constraint
+
+debug :: [String] -> IO ()
+debug strs = pure ()
+{-# inline debug #-}
+
+debugging :: IO () -> IO ()
+debugging _ = pure ()
+{-# inline debugging #-}
+#endif
+
+debug' :: [String] -> IO ()
+debug' strs = putStrLn (intercalate " | " strs ++ " END")
+
+debugging' :: IO () -> IO ()
+debugging' act = act
+{-# inline debugging' #-}
+
+--------------------------------------------------------------------------------
+
+data Fields a
+  = FNil
+  | FCons Name a (Fields a)
+
+instance Functor Fields where
+  fmap f = go where
+    go FNil = FNil
+    go (FCons x a as) = FCons x (f a) (go as)
+  {-# inline fmap #-}
+
+instance Foldable Fields where
+  foldr f z = go where
+    go FNil           = z
+    go (FCons x a fs) = f a $! go fs
+  {-# inline foldr #-}
+  foldl' f z = go z where
+    go acc FNil = acc
+    go acc (FCons x a fs) = go (f acc a) fs
+  {-# inline foldl' #-}
+
+instance Traversable Fields where
+  traverse f = go where
+    go FNil           = pure FNil
[…]

### 2022-02-28 old/mono_staged/src/Cxt/Fields.hs

@@ -1,0 +1,6 @@
+
+module Cxt.Fields (
+  HasLvl(..), HasEnv(..), HasLocals(..), HasPruning(..), HasNameTable(..), HasNames(..),
+  HasSrc(..)) where
+
+import Cxt.Types

### 2022-02-28 old/mono_staged/src/Elaboration.hs

@@ -1,0 +1,965 @@
+{-# options_ghc -Wno-unused-imports -Wno-type-defaults #-}
+
+module Elaboration where
+
+import Control.Monad
+
+import qualified Data.HashMap.Strict as HM
+import qualified FlatParse.Stateful  as FP
+
+import qualified Syntax              as S
+import qualified Values              as V
+import qualified Presyntax           as P
+import qualified Evaluation          as Eval
+import qualified UnifyCxt            as Unif
+import qualified Unification         as Unif
+
+import Common
+import Cxt
+import ElabState
+import InCxt
+import Exceptions
+import Pretty
+
+--------------------------------------------------------------------------------
+
+{-
+High-level design:
+
+- lower : coercing from level 1 to level 0
+
+- check0 :
+  - purely presyntax-directed, matches canonical cases + let
+  - refines checking type if necessary
+  - no need for matching on checking types, because level 0 does not have type-directed
+    elaboration!
+  - why do we even have check0? a) because we can resolve stage ambiguity b) efficiency
+
+- check1 :
+  - Mixed presyntax + type directed, canonical cases + let
+  - performs insertions (<_>, implicit lam)
+
+- infer0 :
+  - purely presyntax-directed, matches *level-ambiguous* canonical cases only
+
+- infer1 :
+  - purely presyntax-directed, matches *level-ambiguous* canonical cases only
+
+- infer :
+  - purely presyntax-directed
+
+
+-}
+
+
+--------------------------------------------------------------------------------
+
+elabError :: Cxt -> P.Tm -> ElabEx -> IO a
+elabError cxt t err = throwIO $ ElabError (unifyCxt cxt) t err
+{-# inline elabError #-}
+
+unify1 :: Cxt -> P.Tm -> V.Val1 -> V.Val1 -> IO ()
+unify1 cxt topT t t' =
[…]

### 2022-02-28 old/mono_staged/src/Exceptions.hs

@@ -1,0 +1,66 @@
+
+module Exceptions (throwIO, catch, throw, module Exceptions) where
+
+import Control.Exception
+
+import Common
+import qualified Syntax    as S
+import qualified Values    as V
+import qualified UnifyCxt  as Unif
+import qualified Presyntax as P
+
+-- | Exception thrown while trying to solve a metavar.
+data SolutionEx
+  = Occurs MetaVar
+  | OutOfScope Lvl
+  | SpineError
+  | NeedExpansion
+  | CSFlexSolution
+  deriving Show
+
+-- | Exception thrown during unification.
+data UnifyEx
+  = Unify0 V.Val0 V.Val0
+  | Unify1 V.Val1 V.Val1
+  | SolutionError V.Val1 V.Val1 SolutionEx
+  | UnifyFieldName Name Name
+deriving instance Show UnifyEx
+
+-- | Unification exception in local unification context.
+data UnifyInner = UnifyInner Unif.Cxt UnifyEx
+  deriving Show
+
+-- | Exception thrown during elaboration
+data ElabEx
+  = UnifyOuter V.Val1 V.Val1 UnifyInner
+  | NoSuchFieldName V.Ty RawName
+  | NoSuchFieldIx   V.Ty Int
+  | NoSuchArgument RawName
+  | NameNotInScope RawName
+  | IcitMismatch Icit Icit
+  | NegativeFieldIndex
+  | NoImplicitLam0
+  | ExpectedVal S.Tm1
+  | FieldNameMismatch Name Name
+  | NoNamedLambdaInference
+  | ExpectedNonEmptyRec
+  | ExpectedEmptyRec
+  | ExpectedEmptyRecCon
+  | ExpectedNonEmptyRecCon
+  | ExpectedType S.Ty
+  | CantInferTuple
+  | ExpectedRecord V.Ty
+  | ExpectedRuntimeType S.Ty
+  | CantInferRec1
+  | CantInfer
+  deriving Show
+
+-- | Elaboration exception in elaboration context.
+data ElabError = ElabError Unif.Cxt P.Tm ElabEx
+  deriving Show
+
+instance Exception SolutionEx
+instance Exception UnifyEx
+instance Exception UnifyInner
+instance Exception ElabEx
+instance Exception ElabError

### 2022-02-28 old/mono_staged/src/InCxt.hs

@@ -1,0 +1,47 @@
+
+module InCxt (
+    eval0, eval1
+  , Eval.forceF0, Eval.forceFU0, Eval.forceF1, Eval.forceFU1
+  , Eval.up, Eval.down
+  , unifyCxt
+  , Eval.app1, (Eval.$$), (Eval.$$$)
+  , quote0, quote1
+  -- , unify0, unify1
+  , Unif.freshMeta
+  , Unif.freshCV
+  ) where
+
+import Common
+import Cxt
+import Syntax
+import Values
+import qualified Evaluation as Eval
+import qualified Unification as Unif
+
+unifyCxt :: Cxt -> Unif.Cxt
+unifyCxt cxt = Unif.Cxt (cxt^.lvl) (cxt^.names)
+{-# inline unifyCxt #-}
+
+eval0 :: Dbg => Cxt -> Tm0 -> Val0
+eval0 cxt t = Eval.eval0 (cxt^.env) t
+{-# inline eval0 #-}
+
+eval1 :: Dbg => Cxt -> Tm1 -> Val1
+eval1 cxt t = Eval.eval1 (cxt^.env) t
+{-# inline eval1 #-}
+
+quote0 :: Dbg => Cxt -> Val0 -> Tm0
+quote0 cxt t = Eval.quote0 (cxt^.lvl) UnfoldNone t
+{-# inline quote0 #-}
+
+quote1 :: Dbg => Cxt -> Val1 -> Tm1
+quote1 cxt t = Eval.quote1 (cxt^.lvl) UnfoldNone t
+{-# inline quote1 #-}
+
+-- unify0 :: Dbg => Cxt -> Val0 -> Val0 -> IO ()
+-- unify0 cxt t t' = Unif.unify0 (unifyCxt cxt) CSRigid t t'
+-- {-# inline unify0 #-}
+
+-- unify1 :: Dbg => Cxt -> Val1 -> Val1 -> IO ()
+-- unify1 cxt t t' = Unif.unify1 (unifyCxt cxt) CSRigid t t'
+-- {-# inline unify1 #-}

### 2022-02-28 old/mono_staged/src/Lexer.hs

@@ -1,0 +1,238 @@
+
+module Lexer where
+
+import FlatParse.Stateful hiding (Parser, runParser, string, char, cut, err)
+
+import qualified FlatParse.Stateful as FP
+import qualified Data.ByteString as B
+import Language.Haskell.TH
+
+import qualified Data.Set as S
+import Data.Char
+import Data.String
+
+--------------------------------------------------------------------------------
+
+data Expected
+  = Lit String
+  | Msg String
+  deriving (Eq, Show, Ord)
+
+instance IsString Expected where fromString = Lit
+
+data Error'
+  = Precise Expected
+  | ExactIndent Int
+  | IndentMore  Int
+  | Imprecise [Expected]
+  deriving Show
+
+data Error = Error !Pos !Error'
+  deriving Show
+
+merge :: Error -> Error -> Error
+merge err@(Error p e) err'@(Error p' e')
+  | p < p'    = err'
+  | p' < p    = err
+  | otherwise = case (e, e') of
+     (ExactIndent _ , _             ) -> err
+     (_             , ExactIndent _ ) -> err'
+     (IndentMore _  , _             ) -> err
+     (_             , IndentMore _  ) -> err'
+     (Precise _     , _             ) -> err
+     (_             , Precise _     ) -> err'
+     (Imprecise ss  , Imprecise ss' ) -> Error p (Imprecise (ss ++ ss'))
+{-# noinline merge #-}
+
+type Parser = FP.Parser Error
+
+prettyError :: B.ByteString -> Error -> String
+prettyError b (Error pos e) =
+
+  let ls       = FP.lines b
+      [(l, c)] = posLineCols b [pos]
+      line     = if 0 <= l && l < length ls then ls !! l else ""
+      linum    = show l
+      lpad     = map (const ' ') linum
+
+      expected (Lit s) = show s
+      expected (Msg s) = s
+
+      err (Precise exp)     = "expected " ++ expected exp
+      err (Imprecise exps)  = "expected " ++ (imprec $ S.toList $ S.fromList exps)
+      err (IndentMore col)  = "expected a token, indented to column " ++ show col ++ " or more"
+      err (ExactIndent col) = "expected a token, indented to column " ++ show col
+
+      imprec :: [Expected] -> String
+      imprec []     = error "impossible"
[…]

### 2022-02-28 old/mono_staged/src/LvlSet.hs

@@ -1,0 +1,56 @@
+{-|
+Bitset for de Bruijn levels. It's currently just a 64-bit 'Int'. We throw an
+error when there are more than 64 local bound variables. If and when the need
+for more than 64 variables arises, we can switch to 128 bit or add a new
+constructor for large sets.
+-}
+
+module LvlSet where
+
+import Data.Coerce
+import Data.Bits
+import Data.List (foldl')
+import Common
+
+newtype LvlSet = LvlSet Int deriving (Eq, Bits) via Int
+
+instance Semigroup LvlSet where
+  (<>) = (.|.)
+  {-# inline (<>) #-}
+
+instance Monoid LvlSet where
+  mempty = LvlSet 0
+  {-# inline mempty #-}
+
+insert :: Lvl -> LvlSet -> LvlSet
+insert (Lvl x) (LvlSet s)
+  | x > 63    = error "LvlSet.insert: element out of range"
+  | otherwise = LvlSet (unsafeShiftL 1 x .|. s)
+{-# inline insert #-}
+
+single :: Lvl -> LvlSet
+single x = insert x mempty
+{-# inline single #-}
+
+delete :: Lvl -> LvlSet -> LvlSet
+delete (Lvl x) (LvlSet s)
+  | x > 63    = error "LvlSet.delete: element out of range"
+  | otherwise = LvlSet (complement (unsafeShiftL 1 x) .&. s)
+{-# inline delete #-}
+
+member :: Lvl -> LvlSet -> Bool
+member (Lvl x) (LvlSet s)
+  | x > 63    = error "LvlSet.member: element out of range"
+  | otherwise = (unsafeShiftL 1 x .&. s) /= 0
+{-# inline member #-}
+
+toList :: LvlSet -> [Lvl]
+toList s = filter (`member` s) (coerce [0..63::Int])
+{-# inline toList #-}
+
+fromList :: [Lvl] -> LvlSet
+fromList = foldl' (flip insert) mempty
+{-# inline fromList #-}
+
+instance Show LvlSet where
+  show = show . toList

### 2022-02-28 old/mono_staged/src/Pretty.hs

@@ -1,0 +1,357 @@
+
+module Pretty (showTm0, showTm1, showTm0Top, showTm1Top,
+               showVal1', showVal1Top',
+               showTm1Top', showTm0Top', showElabError,
+               showVal0, showVal1, showVal0Top, showVal1Top) where
+
+import qualified Data.ByteString.Char8 as B
+import qualified FlatParse.Stateful as FP
+import IO
+import Text.Printf
+
+import Common
+import Syntax
+import Cxt.Fields
+import ElabState
+import Exceptions
+
+import qualified Values     as V
+import qualified Evaluation as Eval
+import qualified Presyntax  as P
+import qualified UnifyCxt   as Unif
+
+--------------------------------------------------------------------------------
+
+showTm0 :: (HasNames cxt [Name]) => cxt -> Tm0 -> String
+showTm0 cxt t = tm0 tmp (cxt^.names) t []
+
+showTm1 :: (HasNames cxt [Name]) => cxt -> Tm1 -> String
+showTm1 cxt t = tm1 tmp (cxt^.names) t []
+
+showVal0 :: (HasNames s [Name], HasLvl s Lvl) => s -> V.Val0 -> String
+showVal0 cxt t = showTm0 cxt $ Eval.quote0 (cxt^.lvl) UnfoldMetas t
+
+showVal1 :: (HasNames s [Name], HasLvl s Lvl) => s -> V.Val1 -> String
+showVal1 cxt t = showTm1 cxt $ Eval.quote1 (cxt^.lvl) UnfoldMetas t
+
+showVal1' :: (HasNames s [Name], HasLvl s Lvl) => s -> V.Val1 -> String
+showVal1' cxt t = showTm1 cxt $ Eval.quote1 (cxt^.lvl) UnfoldAll t
+
+showVal1Top' :: V.Val1 -> String
+showVal1Top' t = showTm1Top $ Eval.quote1 0 UnfoldAll t
+
+showTm1Top' :: Tm1 -> String
+showTm1Top' t = tm1 tmp [] (Eval.quote1 0 UnfoldAll (Eval.eval1 V.Nil t)) []
+
+showTm0Top' :: Tm0 -> String
+showTm0Top' t = tm0 tmp [] (Eval.quote0 0 UnfoldAll (Eval.eval0 V.Nil t)) []
+
+showTm0Top :: Tm0 -> String
+showTm0Top t = tm0 tmp [] t []
+
+showTm1Top :: Tm1 -> String
[…]

### 2022-02-28 old/notes/ClosedMonadicEval.hs

@@ -1,0 +1,73 @@
+{-# language OverloadedStrings #-}
+
+import GHC.Stack
+import Data.Maybe
+import Data.String
+
+type Name = String
+
+data Tm
+  = Var Name
+  | App Tm Tm
+  | Lam Name Tm
+  | Let Name Tm Tm
+  | Bind Name Tm Tm
+  | Return Tm
+  | Hello                  -- Hello : M ()
+  | Tt                     -- Tt : Unit
+  deriving (Eq, Show)
+
+data Val
+  = VVar Name
+  | VApp Val Val
+  | VLam Env Name Tm
+  | VBind Env Name Tm Tm
+  | VReturn Val
+  | VHello
+  | VTt
+  deriving (Eq, Show)
+
+type Env = [(Name, Val)]
+
+--------------------------------------------------------------------------------
+
+instance IsString Tm where fromString = Var
+($$)  = App; infixl 7 $$
+(>>>) = Bind "_"; infixl 1 >>>
+
+--------------------------------------------------------------------------------
+
+impossible :: HasCallStack => a
+impossible = error "impossible"
+
+-- Pure closed evaluation
+eval :: Env -> Tm -> Val
+eval e = \case
+  Var x      -> fromJust $ lookup x e
+  App t u    -> case eval e t of
+                  VLam e' x t -> eval ((x, eval e u):e') t
+                  _           -> impossible
+  Lam x t    -> VLam e x t
+  Let x t u  -> eval ((x, eval e t):e) u
+  Bind x t u -> VBind e x t u
+  Return t   -> VReturn (eval e t)
+  Hello      -> VHello
+  Tt         -> VTt
+
+-- Perform 1 layer of IO
+runIO :: Val -> IO Val
+runIO = \case
+  VReturn t     -> pure t
+  VBind e x t u -> do {t <- runIO (eval e t); runIO (eval ((x, t):e) u)}
+  VHello        -> VTt <$ putStrLn "hello"
+  _             -> impossible
+
+run :: Tm -> IO Val
+run = runIO . eval []
+
+--------------------------------------------------------------------------------
+
+p1 :: Tm
+p1 =
+  Let "f" (Lam "m" $ Hello >>> "m") $
+  "f" $$ Hello

### 2022-02-28 old/notes/MonadicNBE.hs

@@ -1,0 +1,307 @@
+{-
+
+-- Compiler: staging, memory layout control, etc  (small + self-hosting + useful soon)
+
+-- Surface language:
+   1. CBV + side effects (ML)
+   2. CBV + typed effect system:
+        - Eff monad
+        - Eff monad, row of effects
+           f : Int -> Eff [IO, Exc e, MutRef a, MutRef b] Int
+        - (extensible / algebraic / resumable effect handlers)
+
+
+-- Intermediate language:
+
+Assume CBV + printing
+
+   pure language:       admissible substitution (beta-reduce, inline, delta-contraction (CSE))
+   effectful language:  doesn't work
+
+   issue: partial evaluation + effects: don't work together
+          (effectful things cannot be substituted, (reordered, dropped))
+
+   - effects are in a monad
+      - GHC doesn't have a monad in Core, uses token passing to compile IO
+         - unsound
+         - only linear token-passing models (some) IO
+              State hack, runST performance, unsafeInlineIO, ...
+              (mandatory late inlining runST, runRW# etc.)
+
+
+1. Monad in core for everything which is effectful  (M)
+   - guarantees: partial eval + NBE still works, conversion checking modulo laws (efficiently, and in a sound way)
+   - I don't know anything about the intension of my effects
+
+   Optimization: could be State transformer (locally)
+
+      finite number of mutable locations used in lexical
+
+   put x; get     ~ pure x
+   put x; put x;  ~ put x
+
+   throw e; e2 ~ throw e
+
+   limitation of usual State monad: (ST has stronger interface than State)
+          - I formalized ST monad in Agda: (requires parametricity to prove heap well-typed)
+
+       refs : List (MutRef Int)
+       n    : Int
+
+       do
+         write (lookup n refs) 100
+         f refs foo bar             -- out-of-line call
+                                    -- erase all info in symbolic state
+
+       -- ("symbolic" state,    lookup n refs ↦ 100)
+
+   Exceptions:
+      -- something similar
+
+   Call-by-need:
+     implementation: CBV + Eff + mutable refs
+
+     interface for call-by-need:
+
[…]

### 2022-02-28 old/notes/clobber.c

@@ -1,0 +1,20 @@
+
+#include <stdio.h>
+#include <string.h>
+#include <stdlib.h>
+#include <inttypes.h>
+
+extern void unknown (uint64_t* inp);
+/* __attribute__((noinline)) uint64_t foo (){ */
+/*   return 0; */
+/* } */
+
+extern void unknown2 ();
+
+uint64_t foo (uint64_t* buf){
+  /* buf[1] = buf[0]; */
+  /* asm volatile("nop" : : : ); */
+  uint64_t foobar = *buf;
+  unknown2();
+  return *buf + foobar;
+}

### 2022-02-28 old/notes/gccnotes.c

@@ -1,0 +1,91 @@
+
+/* -fno-unwind-tables -fno-exceptions -fno-stack-protector -fomit-frame-pointer */
+
+#include <stdio.h>
+#include <stdint.h>
+#include <stdlib.h>
+#include <string.h>
+#include <signal.h>
+
+#define INLINE static inline __attribute__((always_inline))
+
+// heap objects
+/* -------------------------------------------------------------------------------- */
+
+// layout:
+// pointer: 45 | num_words: 6 | num_objs: 6 | tag: 7
+
+typedef struct {
+  uint64_t out;
+} obj;
+
+INLINE obj* get_ptr(obj x)  {
+  return (obj*)((((int64_t) x.out) >> 16) & (-8));
+}
+
+INLINE uint64_t get_tag(obj x) {
+  return x.out & 127;
+}
+
+INLINE uint64_t get_num_words(obj x){
+  return (x.out >> 13) & 63;
+}
+
+INLINE uint64_t get_num_objs(obj x){
+  return (x.out >> 7) & 63;
+}
+
+INLINE obj pack_obj(obj* ptr, uint64_t nwords, uint64_t nobjs, uint64_t tag){
+  obj res = {(((uint64_t) ptr) << 16) | (nwords << 13) | (nobjs << 7) | tag};
+  return res;
+}
+
+/* -------------------------------------------------------------------------------- */
+
+// A generic stack frame.
+
+struct frame;
+typedef struct __attribute__ ((__packed__)) frame {
+  struct frame* prev; // can be null
+  uint64_t num_objs;
+  obj objs[0];
+} frame;
+
+// A word on the heap. Either an object or unboxed data.
+typedef union {
+  uint64_t word;
+  obj obj;
+} hp_word;
+
+register frame*   fr     asm ("r13");   // frame pointer
+register hp_word* hp     asm ("r14");   // heap pointer       (start of free area)
+register hp_word* hp_lim asm ("r15");   // heap limit pointer (end of free area)
+
+void perform_gc() __attribute__ ((cold));
+
+INLINE hp_word* hp_alloc(uint64_t size){
+  hp_word* res = hp;
+  hp += size;
+  if (hp >= hp_lim) perform_gc();
+  return res;
+}
+
+#define FR_PUSH(size) ({\
+  uint64_t* new_fr = alloca((size+2)*sizeof(uint64_t));	\
+  new_fr[0] = (uint64_t) fr;\
[…]

### 2022-02-28 old/notes/llvmnotes.c

@@ -1,0 +1,56 @@
+
+#include <stdio.h>
+#include <stdlib.h>
+#include <stdint.h>
+
+int64_t foo(int64_t a, int64_t b, int64_t c, int64_t d, int64_t e, int64_t f, int64_t g, int64_t h,
+	    int64_t i, int64_t j, int64_t k, int64_t l, int64_t m, int64_t n, int64_t o, int64_t p){
+  return a * b * c * d * e * f * g * h * i * j * k * l * m * n * o * p;
+}
+
+/* int main(){ */
+/*   printf("%ld\n", foo(1, 2, 3, 4, 5, 6, 7, 8, 9, 10)); */
+/*   return 0; */
+/* } */
+
+
+/* int64_t foo(int64_t a, int64_t b, int64_t c, int64_t d){ */
+/*   return a + b + c + d; */
+/* } */
+
+/* int main(){ */
+/*   printf("%ld\n", foo(1, 2, 3, 4)); */
+/*   return 0; */
+/* } */
+
+/* int64_t fact(int64_t n){ */
+/*   if (n == 0){ */
+/*     return 1; */
+/*   } else { */
+/*     return n * fact(n - 1); */
+/*   } */
+/* } */
+
+int64_t fact_go(int64_t n, int64_t acc){
+  if (n == 0) {
+    return acc;
+  } else {
+    int64_t x = n - 1;
+    int64_t y = n * acc;
+    int64_t res = fact_go(x, y);
+    return res;
+  }
+}
+
+
+/* define dso_local cc 10 i64 @fact_go(i64, i64) #0 { */
+/*   %cmp = icmp eq i64 %0, 0 */
+/*   br i1 %cmp, label %Finish, label %Rec */
+/*   Finish: */
+/*     ret i64 %1 */
+/*   Rec: */
+/*     %x = sub nsw i64 %0, 1 */
+/*     %y = mul nsw i64 %0, %1 */
+/*     %res = tail call i64 @fact_go(i64 %x, i64 %y) */
+/*     ret i64 %res */
+/* } */

### 2022-02-28 old/notes/notes.c

@@ -1,0 +1,46 @@
+
+#include <stdio.h>
+#include <stdint.h>
+#include <stdlib.h>
+
+typedef struct {
+  uint8_t  tag     ;  // 6
+  uint8_t  objtype ;  // 3
+  uint8_t  nobjs   ;  // 5
+  uint8_t  nwords  ;  // 5
+  uint64_t ptr     ;  // 45
+} unpacked_obj;
+
+typedef struct {
+  uint64_t payload;
+} packed_obj;
+
+static inline packed_obj pack_obj(unpacked_obj obj){
+  packed_obj res = {
+       (((uint64_t)obj.tag)     << 58)
+     | (((uint64_t)obj.objtype) << 55)
+     | (((uint64_t)obj.nobjs)   << 50)
+     | (((uint64_t)obj.nwords)  << 45)
+     | obj.ptr};
+  return res;
+}
+
+static inline void* to_ptr(packed_obj obj){
+  return (void*)(((int64_t)(obj.payload) << 19) >> 16);
+}
+
+static inline uint8_t get_tag(packed_obj obj){
+  return (uint8_t)((obj.payload) >> 58);
+}
+
+static inline uint64_t compr_ptr(void* ptr){
+  return (((uint64_t) ptr) << 16) >> 19;
+}
+
+int main() {
+  int* ptr = malloc(sizeof(int));
+  *ptr = 100;
+  unpacked_obj test = {10, 0, 0, 0, compr_ptr(ptr)};
+  printf("%d\n", *((int*)(to_ptr(pack_obj(test)))));
+  return 0;
+}

### 2022-02-28 old/notes/rts.c

@@ -1,0 +1,338 @@
+
+/* -O0 -Wall -fverbose-asm -fno-asynchronous-unwind-tables -fno-stack-protector -foptimize-sibling-calls -fno-plt -fomit-frame-pointer -finline-small-functions -fstrict-aliasing */
+
+#define _GNU_SOURCE
+#include <errno.h>
+#include <signal.h>
+#include <stdint.h>
+#include <stdio.h>
+#include <stdlib.h>
+#include <string.h>
+#include <sys/mman.h>
+#include <sys/resource.h>
+#include <sys/sysinfo.h>
+/* #include <time.h> */
+#include <unistd.h>
+
+// shorthands
+/* -------------------------------------------------------------------------------- */
+
+// stringify a macro
+#define STR1(x) #x
+#define STR(x) STR1(x)
+
+#define INLINE static inline __attribute__((always_inline))
+
+typedef uint64_t u64;
+typedef int64_t  i64;
+
+#define PACKED_STRUCT struct __attribute__ ((__packed__))
+
+
+// heap objects
+/* -------------------------------------------------------------------------------- */
+
+// layout:
+// pointer: 45 | num_words: 6 | num_objs: 6 | tag: 7
+
+typedef struct {
+  u64 out;
+} obj;
+
+INLINE void* get_ptr(obj x)  {
+  return (void*)((((i64) x.out) >> 16) & (-8));
+}
+
+INLINE u64 get_tag(obj x) {
+  return x.out & 127;
+}
+
+INLINE u64 get_num_words(obj x){
+  return (x.out >> 13) & 63;
+}
+
+INLINE u64 get_num_objs(obj x){
+  return (x.out >> 7) & 63;
+}
+
+INLINE obj pack_obj(void* ptr, u64 nwords, u64 nobjs, u64 tag){
+  obj res = {(((u64) ptr) << 16) | (nwords << 13) | (nobjs << 7) | tag};
+  return res;
+}
+
+
+/* -------------------------------------------------------------------------------- */
+
+
+// A generic stack frame.
+struct frame;
+typedef
+struct __attribute__ ((__packed__)) frame {
+  struct frame* prev;     // can be null
+  uint64_t num_objs;
+  obj objs[0];
+} frame;
+
+
+// RTS constants and state
+/* -------------------------------------------------------------------------------- */
+
[…]

### 2022-02-28 old/notes/rtsclang.c

@@ -1,0 +1,338 @@
+
+/* -O0 -Wall -fverbose-asm -fno-asynchronous-unwind-tables -fno-stack-protector -foptimize-sibling-calls -fno-plt -fomit-frame-pointer -finline-small-functions -fstrict-aliasing */
+
+#define _GNU_SOURCE
+#include <errno.h>
+#include <signal.h>
+#include <stdint.h>
+#include <stdio.h>
+#include <stdlib.h>
+#include <string.h>
+#include <sys/mman.h>
+#include <sys/resource.h>
+#include <sys/sysinfo.h>
+/* #include <time.h> */
+#include <unistd.h>
+
+// shorthands
+/* -------------------------------------------------------------------------------- */
+
+// stringify a macro
+#define STR1(x) #x
+#define STR(x) STR1(x)
+
+#define INLINE static inline __attribute__((always_inline))
+
+typedef uint64_t u64;
+typedef int64_t  i64;
+
+#define PACKED_STRUCT struct __attribute__ ((__packed__))
+
+
+// heap objects
+/* -------------------------------------------------------------------------------- */
+
+// layout:
+// pointer: 45 | num_words: 6 | num_objs: 6 | tag: 7
+
+typedef struct {
+  u64 out;
+} obj;
+
+INLINE void* get_ptr(obj x)  {
+  return (void*)((((i64) x.out) >> 16) & (-8));
+}
+
+INLINE u64 get_tag(obj x) {
+  return x.out & 127;
+}
+
+INLINE u64 get_num_words(obj x){
+  return (x.out >> 13) & 63;
+}
+
+INLINE u64 get_num_objs(obj x){
+  return (x.out >> 7) & 63;
+}
+
+INLINE obj pack_obj(void* ptr, u64 nwords, u64 nobjs, u64 tag){
+  obj res = {(((u64) ptr) << 16) | (nwords << 13) | (nobjs << 7) | tag};
+  return res;
+}
+
+
+/* -------------------------------------------------------------------------------- */
+
+
+// A generic stack frame.
+struct frame;
+typedef
+struct __attribute__ ((__packed__)) frame {
+  struct frame* prev;     // can be null
+  uint64_t num_objs;
+  obj objs[0];
+} frame;
+
+
+// RTS constants and state
+/* -------------------------------------------------------------------------------- */
+
[…]

### 2022-02-28 old/notes/sections.c

@@ -1,0 +1,45 @@
+#include <stdio.h>
+#include <inttypes.h>
+
+/*
+Solution for putting vars in .text:
+
+https://stackoverflow.com/questions/58455300/assembler-warning-with-gcc-warning-when-placing-data-in-text
+
+*/
+
+/* char c1 = 'a'; */
+/* /\* int f1(int n) {return n + 200;} *\/ */
+/* char c2 = 'b'; */
+/* char c3 = 'c'; */
+
+/* long long mynum1 __attribute__ ((section (".text#"))) = 1200000; */
+/* long long mynum2 __attribute__ ((section (".text#"))) = 999999; */
+
+typedef struct {
+  int arity;
+  int foobar;
+} myinfo;
+
+uint64_t fun1_info __attribute__ ((section (".text#"))) = 41;
+uint64_t fun1(uint64_t x){
+  return x + 100 + x;
+}
+
+uint64_t fun2_info __attribute__ ((section (".text#"))) = 42;
+uint64_t fun2(uint64_t x){
+  return x + 300 + x;
+}
+
+myinfo fun3_info __attribute__ ((section (".text#"))) = {.arity = 100, .foobar = 0};
+uint64_t fun3(uint64_t x){
+  return x + 300 + x;
+}
+
+int64_t get_info_table(void* fun){
+  return *((int64_t*)(fun - 8));
+  }
+
+int main(){
+  printf("%ld\n", get_info_table((void*)(&fun2)));
+}

### 2022-02-28 old/notes/structret.c

@@ -1,0 +1,15 @@
+
+#include <stdint.h>
+
+typedef struct {
+  int64_t a;
+  int64_t b;
+  /* int64_t c; */
+  /* int64_t d; */
+  /* int64_t e; */
+} mystruct;
+
+mystruct fun(int64_t a, int64_t b, int64_t c){
+  mystruct res = {a * b, b * c};
+  return res;
+}

### 2022-02-28 src/Cxt/Extension.hs

@@ -1,75 +1,0 @@
-
-module Cxt.Extension where
-
-import Common
-import Cxt.Types
-import qualified Values as V
-import qualified Syntax as S
-import qualified Data.HashMap.Strict as M
-import Evaluation
-
-emptyCxt :: RawName -> Cxt
-emptyCxt = Cxt V.Nil 0 S.Empty [] mempty []
-{-# inline emptyCxt #-}
-
-addName :: Name -> NameInfo -> NameTable -> NameTable
-addName NEmpty    _   tbl = tbl
-addName (NName x) inf tbl = M.insert x inf tbl
-addName NX        _   tbl = tbl
-{-# inline addName #-}
-
-bind0 :: Name -> S.Ty -> S.CV -> Cxt -> Cxt
-bind0 x a acv cxt = bind0' x a (eval1 (cxt^.env) a) acv (eval1 (cxt^.env) acv) cxt
-{-# inline bind0 #-}
-
-bind0' :: Name -> S.Ty -> V.Ty -> S.CV -> V.CV -> Cxt -> Cxt
-bind0' x a va acv vacv (Cxt env l loc pr ntbl ns src) =
-  Cxt (V.Snoc0 env l)
-      (l + 1)
-      (S.Bind0 loc a acv)
-      (pr :> S.PEBind0)
-      (addName x (NameInfo0 l va vacv) ntbl)
-      (x:ns)
-      src
-{-# inline bind0' #-}
-
-bind1 :: Name -> S.Ty -> Cxt -> Cxt
-bind1 x a cxt = bind1' x a (eval1 (cxt^.env) a) cxt
-{-# inline bind1 #-}
-
-bind1' :: Name -> S.Ty -> V.Ty -> Cxt -> Cxt
-bind1' x a va (Cxt env l loc pr ntbl ns src) =
-  Cxt (V.Snoc1 env (V.Var1 l))
-      (l + 1)
-      (S.Bind1 loc a)
-      (pr :> S.PEBind1 Expl)
-      (addName x (NameInfo1 l va) ntbl)
-      (x:ns)
-      src
-{-# inline bind1' #-}
-
-newBinder :: Name -> S.Ty -> Cxt -> Cxt
-newBinder x a (Cxt env l loc pr ntbl ns src) =
-  Cxt (V.Snoc1 env (V.Var1 l))
-      (l + 1)
-      (S.Bind1 loc a)
-      (pr :> S.PEBind1 Expl)
[…]

### 2022-02-28 src/Cxt/Types.hs

@@ -1,27 +1,0 @@
-
-module Cxt.Types where
-
-import Lens.Micro.TH
-
-import Common
-import qualified Syntax as S
-import qualified Values as V
-import Data.HashMap.Strict
-
-data NameInfo
- = NameInfo0 Lvl V.Ty V.CV
- | NameInfo1 Lvl V.Ty
-
-type NameTable = HashMap RawName NameInfo
-
-data Cxt = Cxt {
-  cxtEnv       :: V.Env,
-  cxtLvl       :: Lvl,
-  cxtLocals    :: S.Locals,
-  cxtPruning   :: S.Pruning,
-  cxtNameTable :: NameTable,
-  cxtNames     :: [Name],
-  cxtSrc       :: RawName
-  }
-
-makeFields ''Cxt

### 2022-02-28 src/Elaboration.hs

@@ -1,965 +1,0 @@
-{-# options_ghc -Wno-unused-imports -Wno-type-defaults #-}
-
-module Elaboration where
-
-import Control.Monad
-
-import qualified Data.HashMap.Strict as HM
-import qualified FlatParse.Stateful  as FP
-
-import qualified Syntax              as S
-import qualified Values              as V
-import qualified Presyntax           as P
-import qualified Evaluation          as Eval
-import qualified UnifyCxt            as Unif
-import qualified Unification         as Unif
-
-import Common
-import Cxt
-import ElabState
-import InCxt
-import Exceptions
-import Pretty
-
---------------------------------------------------------------------------------
-
-{-
-High-level design:
-
-- lower : coercing from level 1 to level 0
-
-- check0 :
-  - purely presyntax-directed, matches canonical cases + let
-  - refines checking type if necessary
-  - no need for matching on checking types, because level 0 does not have type-directed
-    elaboration!
-  - why do we even have check0? a) because we can resolve stage ambiguity b) efficiency
-
-- check1 :
-  - Mixed presyntax + type directed, canonical cases + let
-  - performs insertions (<_>, implicit lam)
-
-- infer0 :
-  - purely presyntax-directed, matches *level-ambiguous* canonical cases only
-
-- infer1 :
-  - purely presyntax-directed, matches *level-ambiguous* canonical cases only
-
-- infer :
-  - purely presyntax-directed
-
-
--}
-
-
---------------------------------------------------------------------------------
-
-elabError :: Cxt -> P.Tm -> ElabEx -> IO a
-elabError cxt t err = throwIO $ ElabError (unifyCxt cxt) t err
-{-# inline elabError #-}
-
-unify1 :: Cxt -> P.Tm -> V.Val1 -> V.Val1 -> IO ()
-unify1 cxt topT t t' =
[…]

### 2022-02-28 src/Pretty.hs

@@ -1,357 +1,0 @@
-
-module Pretty (showTm0, showTm1, showTm0Top, showTm1Top,
-               showVal1', showVal1Top',
-               showTm1Top', showTm0Top', showElabError,
-               showVal0, showVal1, showVal0Top, showVal1Top) where
-
-import qualified Data.ByteString.Char8 as B
-import qualified FlatParse.Stateful as FP
-import IO
-import Text.Printf
-
-import Common
-import Syntax
-import Cxt.Fields
-import ElabState
-import Exceptions
-
-import qualified Values     as V
-import qualified Evaluation as Eval
-import qualified Presyntax  as P
-import qualified UnifyCxt   as Unif
-
---------------------------------------------------------------------------------
-
-showTm0 :: (HasNames cxt [Name]) => cxt -> Tm0 -> String
-showTm0 cxt t = tm0 tmp (cxt^.names) t []
-
-showTm1 :: (HasNames cxt [Name]) => cxt -> Tm1 -> String
-showTm1 cxt t = tm1 tmp (cxt^.names) t []
-
-showVal0 :: (HasNames s [Name], HasLvl s Lvl) => s -> V.Val0 -> String
-showVal0 cxt t = showTm0 cxt $ Eval.quote0 (cxt^.lvl) UnfoldMetas t
-
-showVal1 :: (HasNames s [Name], HasLvl s Lvl) => s -> V.Val1 -> String
-showVal1 cxt t = showTm1 cxt $ Eval.quote1 (cxt^.lvl) UnfoldMetas t
-
-showVal1' :: (HasNames s [Name], HasLvl s Lvl) => s -> V.Val1 -> String
-showVal1' cxt t = showTm1 cxt $ Eval.quote1 (cxt^.lvl) UnfoldAll t
-
-showVal1Top' :: V.Val1 -> String
-showVal1Top' t = showTm1Top $ Eval.quote1 0 UnfoldAll t
-
-showTm1Top' :: Tm1 -> String
-showTm1Top' t = tm1 tmp [] (Eval.quote1 0 UnfoldAll (Eval.eval1 V.Nil t)) []
-
-showTm0Top' :: Tm0 -> String
-showTm0Top' t = tm0 tmp [] (Eval.quote0 0 UnfoldAll (Eval.eval0 V.Nil t)) []
-
-showTm0Top :: Tm0 -> String
-showTm0Top t = tm0 tmp [] t []
-
-showTm1Top :: Tm1 -> String
[…]

### 2022-02-28 src/Syntax.hs

@@ -1,78 +1,0 @@
-
-module Syntax where
-
-import Common
-
-type Ty = Tm1
-
-data Tm0
-  = Var0 Ix
-  | Top0 Lvl
-  | Let0 Name Ty Tm0 Tm0
-  | Lam0 Name Ty Tm0
-  | App0 Tm0 Tm0
-  | Case Tm0 (Cases Tm0)
-  | Down Tm1
-  | Field0 Tm0 Name Int
-  | RecCon0 (Fields Tm0)
-  | Add Tm0 Tm0
-  | Mul Tm0 Tm0
-  | Sub Tm0 Tm0
-  | IntLit Int
-  | Wk10 Tm0
-  deriving Show
-
-type CV = Tm1
-
-data Tm1
-  = Var1 Ix
-  | Top1 Lvl
-  | Let1 Name Ty Tm1 Tm1
-  | Pi Name Icit Ty Ty
-  | Lam1 Name Icit Ty Tm1
-  | App1 Tm1 Tm1 Icit
-  | Fun Ty Ty CV
-  | U0 CV
-  | U1
-  | Lift CV Ty
-  | CV
-  | Comp
-  | Val
-  | Up Tm0
-  | Rec0 (Fields Ty)
-  | Rec1 (Fields Ty)
-  | RecCon1 (Fields Tm1)
-  | Field1 Tm1 Name Int
-  | TyCon Lvl
-  | DataCon Lvl Int
-  | AppPruning Tm1 Pruning
-  | Wk11 Tm1
-  | Wk01 Tm1
-  | Meta MetaVar
-  | Int
-  | Irrelevant
-  deriving Show
-
-data Locals
-  = Empty
-  | Define Locals Ty Tm1
-  | Bind0 Locals Ty CV
-  | Bind1 Locals Ty
-  deriving Show
-
-data PruneEntry = PESkip | PEBind0 | PEBind1 Icit
-  deriving Show
-
-type Pruning = [PruneEntry]
-newtype RevPruning = RevPruning Pruning
-  deriving stock Show
-
-up :: Tm0 -> Tm1
-up (Down t) = t
-up t        = Up t
-{-# inline up #-}
-
-down :: Tm1 -> Tm0
-down (Up t) = t
-down t      = Down t
-{-# inline down #-}

### 2022-02-28 src/UnifyCxt.hs

@@ -1,20 +1,0 @@
-
-module UnifyCxt where
-
-import Lens.Micro.TH
-
-import Common
-import Values
-import Cxt.Fields
-
-data Cxt = Cxt {cxtLvl :: Lvl, cxtNames :: [Name]}
-  deriving Show
-makeFields ''Cxt
-
-bind :: Name -> Cxt -> Cxt
-bind x (Cxt l xs) = Cxt (l + 1) (x:xs)
-{-# inline bind #-}
-
-topVar :: Cxt -> Val1
-topVar cxt = Var1 (cxt^.lvl)
-{-# inline topVar #-}

### 2022-02-28 src/Values.hs

@@ -1,95 +1,0 @@
-
-module Values where
-
-import Common
-import qualified Syntax as S
-
-data Close a = Close Env a deriving Show
-data Env = Nil | Snoc1 Env ~Val1 | Snoc0 Env Lvl deriving Show
-
-wk1Env :: Env -> Env
-wk1Env = \case Snoc1 env _ -> env; _ -> impossible
-{-# inline wk1Env #-}
-
-wk0Env :: Env -> Env
-wk0Env = \case Snoc0 env _ -> env; _ -> impossible
-{-# inline wk0Env #-}
-
-envLen :: Env -> Int
-envLen = go 0 where
-  go acc Nil = acc
-  go acc (Snoc1 env _) = go (acc + 1) env
-  go acc (Snoc0 env _) = go (acc + 1) env
-
-type Ty = Val1
-type CV = Val1
-
-data Spine
-  = Id
-  | App1 Spine Val1 Icit
-  | Field1 Spine Name Int
-  deriving Show
-
-reverseSpine :: Spine -> Spine
-reverseSpine = go Id where
-  go acc Id               = acc
-  go acc (App1 sp t i)    = go (App1 acc t i) sp
-  go acc (Field1 sp x ix) = go (Field1 acc x ix) sp
-
-spineLen :: Spine -> Lvl
-spineLen = go 0 where
-  go acc Id              = acc
-  go acc (App1 sp _ _)   = go (acc + 1) sp
-  go acc (Field1 sp _ _) = go (acc + 1) sp
-
-
-data UnfoldHead
-  = Top1 Lvl
-  | Solved MetaVar
-  deriving Show
-
-data RigidHead
-  = RHVar1 Lvl
-  | RHTyCon Lvl
-  | RHDataCon Lvl Int
-  deriving (Eq, Show)
-
-data Val0
-  = Var0 Lvl
-  | Top0 Lvl
-  | Let0 Name Ty Val0 {-# unpack #-} (Close S.Tm0)
-  | App0 Val0 Val0
-  | Case Val0 {-# unpack #-} (Close (Cases S.Tm0))
-  | Down Val1
-  | Field0 Val0 Name Int
-  | RecCon0 (Fields Val0)
-  | Lam0 Name Ty {-# unpack #-} (Close S.Tm0)
-  | Add Val0 Val0
-  | Sub Val0 Val0
-  | Mul Val0 Val0
[…]

### 2022-03-02 paper/paper.tex
stuff

@@
 
-\documentclass[acmsmall]{acmart}
+\documentclass[acmsmall,anonymous,review]{acmart}
 %% \documentclass[nonacm,acmsmall]{acmart}
@@
 
-\setcopyright{acmcopyright}
-\copyrightyear{2018}
-\acmYear{2018}
-\acmDOI{XXXXXXX.XXXXXXX}
+%% \setcopyright{acmcopyright}
+%% \copyrightyear{2018}
+%% \acmYear{2018}
+%% \acmDOI{XXXXXXX.XXXXXXX}
 
@@
 certain \emph{restrictions} on which metaprograms are expressible. Usually,
-staged systems enforce typing disciple, prohibit arbitrary manipulation of
+staged systems enforce typing discipline, prohibit arbitrary manipulation of
 object-level scopes, and often they also prohibit accessing the internal
@@
   \item In Section \ref{sec:formal-2ltt}, we specify the formal notion of models and
-    syntax for the variant of 2LTT used in this paper, largely following \cite{twolevel}.
+    syntax for the variant of 2LTT used in this paper, mostly following \cite{twolevel}.
   \item In Section \ref{sec:staging-algorithm}, we review the standard presheaf
@@
     of intensional code analysis in the presheaf model.
-  \item We provide a small prototype implementation which we briefly discuss in
-    Section \ref{sec:implementation}.
+  \item We provide a small standalone implementation of elaboration and staging
+    for a two-level type theory, with code samples that expand on the paper.
   \item To our knowledge, this is the first account of staged compilation in the
@@
 language which resembles Agda \cite{agdadocs}. We focus on examples and informal
-explanations here; the formal details will be presented in Section \ref{sec:formal-2ltt}.
+explanation here; the formal details will be presented in Section \ref{sec:formal-2ltt}.
 
@@
     have $A : \U_1$, then $A$ is guaranteed to be only present at compile time,
-    and will be staged away. Elements $a : A$ are likewise computed away.
+    and will be staged away. Elements of $A$ are likewise computed away.
 \end{itemize}
@@
 $\cons_0 : (A : \U_0) \to A \to \List_0\,A$ and $\msf{foldr}_0 : (A\,B : \U_0)
-\to (A \to B \to B) \to B \to \List_0\,A \to B$. We define a map function which
-``inlines'' its function argument:
+\to (A \to B \to B) \to B \to \List_0\,A \to B$. We define a mapping function which
+inlines its function argument:
 \begin{alignat*}{3}
@@
       \qut{\msf{foldr}_0\,
-        \spl A\,\spl B\,
+        \spl A\,(\List_0 \spl B)\,
         (\lambda\,a\,\mit{bs}.\, \cons_0\,\spl B\,\spl(f\,\qut{a})\,\mit{bs})\,
@@
   & &&\spl(\mit{map}\,\qut{\Nat_0}\,\qut{\Nat_0}\,(\lambda\,n.\,\qut{\spl n +_0 10})\,\qut{\mit{ns}})\\
-  & =\hspace{1em}&&\spl\qut{\msf{foldr}_0\,
-        \spl \qut{\Nat_0}\,\spl \qut{\Nat_0}\,
+  & =\hspace{0.5em}&&\spl\qut{\msf{foldr}_0\,
+        \spl \qut{\Nat_0}\,\spl \qut{\List_0\,\spl\qut{Nat_0}}\,
         (\lambda\,a\,\mit{bs}.\, \cons_0\,\spl B\,\spl\qut{\spl\qut{a} +_0 10}\,\mit{bs})\,
@@
         \spl\qut{\mit{ns}}}\\
-  & =\hspace{1em} &&\msf{foldr_0}\,\Nat_0\,\Nat_0\,\,(\lambda\,a\,bs.\,a +_0 10)\,(\nil_0\,\Nat_0)\,\mit{ns}
+  & =\hspace{0.5em} &&\msf{foldr_0}\,\Nat_0\,(\List_0\,\Nat_0)\,\,(\lambda\,a\,bs.\,\cons_0\,(a +_0 10)\,bs)\,(\nil_0\,\Nat_0)\,\mit{ns}
 \end{alignat*}
@@
 duplicate the code of the mapping function. In such cases, we can use
-\emph{let-insertion}, a technique in staged compilation (see
-\cite{staging-with-delimited-control} for an overview). If we bind a runtime
+\emph{let-insertion}, a technique in staged compilation. If we bind a runtime
 expression to a runtime variable, and only use that variable in subsequent
@@
 \end{alignat*}
- Alternatively, we can define $\mit{map}$ so that it performs let-insertion, and
- we can switch between the two versions as needed.
+We include examples of slightly more sophisticated let-insertion in the supplementary
+code.
 
@@
 formers. For instance, we may write a partial evaluator for a simply typed
-lambda calculus. We sketch the implementation in the following. First, we
-inductively define contexts, types and terms:
+lambda calculus. We sketch the implementation in the following; the full version
+can be found in the supplementary code. First, we inductively define contexts,
+types and terms:
 \begin{alignat*}{6}
@@
 
-We do not detail the implementation of fusion in 2LTT here. In short, 2LTT is a
-natural setting for a wide range of fusion setups. A major advantage of fusion
-in 2LTT is the formal guarantee of staging, in contrast to implementations where
-compile-time computation relies on ad-hoc user annotations and general-purpose
-optimization passes. For instance, fusion in GHC relies on rewrite rules and
-inlining annotations which have to be carefully tuned and ordered, and it is
-quite possible to get pessimized code via failed fusion.
+We do not detail the implementation of fusion in 2LTT here; a small example can
+be found in the supplementary code. In short, 2LTT is a natural setting for a
+wide range of fusion setups. A major advantage of fusion in 2LTT is the formal
+guarantee of staging, in contrast to implementations where compile-time
+computation relies on ad-hoc user annotations and general-purpose optimization
+passes. For instance, fusion in GHC relies on rewrite rules and inlining
+annotations which have to be carefully tuned and ordered, and it is possible to
+get pessimized code via failed fusion.
 
 \subsubsection{Inferring staging operations}
-During bidirectional elaboration, we can use the preservation isomorphisms and
-the quote-splice isomorphism as a coercive subtyping system. When elaboration
-needs to compare an inferred and an expected type, it may insert transports
-along isomorphisms. We implemented this feature in our prototype. It
-additionally supports Agda-style implicit arguments and pattern unification, so
-it can elaborate the following definition:
+We can extract a coercive subtyping system from the staging
+operations, which can be used to infer staging operations, and in particular
+transport definitions along preservation isomorphisms. One choice is to have $A
+\leq \Lift A$, $\Lift A \leq A$, a contravariant-covariant rule for functions
+and a covariant rule for $\Sigma$. During bidirectional elaboration, when we
+need to compare an inferred and an expected type, we can insert coercions.  We
+implemented this feature in our prototype. It additionally supports Agda-style
+implicit arguments and pattern unification, so it can elaborate the following
+definition:
 \begin{alignat*}{3}
@@
 \[ \{A\,B : \Lift\U_0\} \to \Lift((\spl A \to \spl B) \to \List_0\,\spl A \to \List_0\,\spl B). \]
-It seems to the author of this paper that improved types are a sensible default,
-and we can insert explicit lifting when we want to opt for unimproved
-types. This is also available in our prototype.
+It seems that improved types are a sensible default, and we can insert explicit
+lifting when we want to opt for unimproved types. This is also available in our
+prototype.
 
@@
 \label{sec:variations}
-
-%% In principle, 2LTT supports any sensible type-theoretic feature on the meta
-%% level. On the object level though, we have yet more freedom in choosing type
-%% formers and even the equational theory of terms. We may choose to omit
-%% $\beta\eta$-rules from the object language.
-
-%% \todo{merge closure typing into rep poly?}
-%% \todo{range of object languages?}
-%% \todo{effectful, etc object languages?}
 
@@
 more distinctions to the object language can make it easier to optimize and
-compile it.
-
-%% Simpler object languages are easier to optimize and compile, but they are often
-%% more tedious to directly program in. 2LTT can be used as a principled way to
-%% extended languages with abstraction facilities, through metaprogramming. We
-%% look at several setups with restricted object languages.
+compile.
 
@@
 Despite the restriction to simple types at runtime, we can still write arbitrary
-higher-rank polymorphic functions in this setup, such as a function with type
-$((A : \Ty_0) \to \Lift\,\spl A \to \Lift \spl A) \to \Lift\,\Bool_0$. This
-function can be only applied to statically known arguments, so the polymorphism
-can be staged away. The main restriction that programmers have to keep in mind
-is that polymorphic functions cannot be stored inside runtime data types.
-
-This setup is fairly similar to the well-known monomorphization models in the
+higher-rank polymorphic functions, such as a function with type $((A : \Ty_0)
+\to \Lift\,\spl A \to \Lift \spl A) \to \Lift\,\Bool_0$. This function can be
+only applied to statically known arguments, so the polymorphism can be staged
+away. The main restriction that programmers have to keep in mind is that
+polymorphic functions cannot be stored inside runtime data types.
+
+This is fairly similar to the well-known monomorphization models in the
 C++ and Rust programming languages. The evident difference is that in 2LTT we
@@
 
-This refines monomorphization, in that types are not directly identified with
+This refines monomorphization, so that types are not directly identified with
 memory representations, but instead representations are internalized in 2LTT as
@@
 \item We may introduce unboxed $\Sigma$ types and primitive machine types in the
-  runtime language. For $A : \U_{0}\,r$ and $B : A \to \U_{0}\,r'$, we may have
-  $(x : A) \times B\,x : \U_{0}\,(\msf{Prod\,r\,r'})$. Thus, we have
-  type dependency, but we do not have dependency in memory representations.
+  runtime language. For $r : \Rep$, $r' : \Rep$, $A : \U_{0}\,r$ and $B : A \to
+  \U_{0}\,r'$, we may have $(x : A) \times B\,x :
+  \U_{0}\,(\msf{Prod\,r\,r'})$. Thus, we have type dependency, but we do not
+  have dependency in memory representations.
 \end{itemize}
@@
 during staging all $\Rep$ indices are computed to concrete canonical
-representations. This is a way to reconcile dependent types with control over
-memory layouts.  The unboxed flavor of $\Sigma$ ends up with a statically known
-flat memory representation, computed from the representations of the fields.
+representations. This is a way to reconcile dependent types with some amount of
+control over memory layouts. The unboxed flavor of $\Sigma$ ends up with a
+statically known flat memory representation, computed from the representations
+of the fields.
 
@@
 categories-with-families \cite{Dybjer96internaltype}, such that there is a
-family morphism from each $(\Ty_{0,j},\,\Tm_{0,j})$ to $(\Ty_{1,j},\,\Tm_{1,j})$.
-
-The following notions are derivable in any model:
+natural transformation from each $(\Ty_{0,j},\,\Tm_{0,j})$ to $(\Ty_{1,j},\,\Tm_{1,j})$
+which is invertible on terms.
+
+The following notions are derivable:
 \begin{itemize}
@@
 which supports certain type formers. For the sake of brevity, we concretely
-specify here a small collection of type formers. However, we will argue that our
-results extend to more general notions of inductive types. We specify type
-formers in the following. We assume that all type formers are natural,
+specify here only a small collection of type formers. However, we will argue later
+that our results extend to more general notions of inductive types. We specify
+type formers in the following. We assume that all type formers are natural,
 i.e.\ stable under substitution.
@@
 yields a notion of \emph{recursion on syntax}, and it can be shown that
-initiality also implies an \emph{induction principle}.
+initiality also implies an \emph{induction principle} \cite{kaposi2019constructing}.
 \end{definition}
@@
 2LTT umbrella. Ibid.\ also considers a range of additional
-strengthenings and extension of 2LTT \cite[Section~2.4]{twolevel}, most of which
-are of use in synthetic homotopy theory. We do not assume any of these, and
+strengthenings and extensions of 2LTT \cite[Section~2.4]{twolevel}, most of which
+are useful in synthetic homotopy theory. We do not assume any of these, and
 stick to the most basic formulation of 2LTT.
@@
 terms and binders: whenever we go under an object-level binder we have to weaken
-the current semantic context. In other words, when moving from $\Delta$ to
+the current semantic environment. In other words, when moving from $\Delta$ to
 $\Delta \ext A$, we have to shift $\gamma : |\ev\,\Gamma|\,\Delta$ to
@@
 implementations in this style. We can use the same solution in staging, by using
-de Bruijn levels and delayed weakenings in object-level terms during staging. It
-is a somewhat unfortunate mismatch that indices are far more convenient in
-formalization, but levels are more efficient in practice.
+de Bruijn levels and delayed weakenings in object-level terms. It is a somewhat
+unfortunate mismatch that indices are far more convenient in formalization, but
+levels are more efficient in practice.
 
@@
 function arguments, we do not want to generate separate code for each
-usage. De-duplicating object-level code with \emph{function types} is
+usage. De-duplicating object-level code with function types is
 usually safe, since function bodies become closed top-level code after closure
@@
 We define a family of functions $\re$ from the 2LTT syntax to objects in
-$\hato$. This will be related to the evaluation morphism $\ev$ in the relational
+$\hato$. We will relate this to the evaluation morphism $\ev$ in the relational
 interpretation. In short, $\re$ restricts 2LTT syntax so that it can only depend
-on object-level contexts, contexts given as $\emb{\Gamma}$.
+on object-level contexts, which are given as $\emb{\Gamma}$.
 
@@
 
-The above preservation properties mean that $\re$ is a \emph{pseudomorphism} in
-the sense of \cite{gluing}, between the syntactic cwf given by
-$(\Ty_{1,j},\,\Tm_{1,j})$ and the corresponding cwf structure in $\hato$. In
-ibid.\ there is an analysis of such cwf morphisms, from which we obtain the
-following additional preservation properties:
+The preservation properties $\re_{\emptycon}$ and $\re_{\ext_1}$ mean that $\re$
+is a \emph{pseudomorphism} in the sense of \cite{gluing}, between the syntactic
+cwf given by $(\Ty_{1,j},\,\Tm_{1,j})$ and the corresponding cwf structure in
+$\hato$. In ibid.\ there is an analysis of such cwf morphisms, from which we
+obtain the following additional preservation properties:
 \begin{itemize}
@@
 \end{alignat*}
-For $\Con$, $\Sub$, and meta-level types and terms, this is a fairly standard
+For $\Con$, $\Sub$ and meta-level types and terms, this is a fairly standard
 logical relation interpretation: contexts are mapped to relations, types to
 dependent relations, and substitutions and terms respect relations. We will only have
-modest complications in meta-level type formers, because we will need to sometimes use
+modest complications in meta-level type formers because we will sometimes need to use
 the lax/oplax preservations properties of $\re$. For object-level types and
@@
 techniques. Second, the task at hand is not too technically difficult, so using
-advanced techniques is not essential. Compare e.g.\ normalization for cubical
-type theories, which is infeasible to handle without the more synthetic
+advanced techniques is not essential. Contrast e.g.\ normalization for cubical
+type theories, which is not feasible to handle without the more synthetic
 presentations \cite{cubicalnbe}.
@@
 it is possible to simply use inductive deeply embedded syntaxes, which can be
-analyzed; our example in Section \ref{sec:staged-programming} for partially
-evaluating simply typed lambda calculus is like this. However, native
-metaprogramming features could be more concise and convenient than deep
-embeddings, similarly to how staged programming is more convenient than code
-generation with deeply embedded syntaxes. Hence, intensional analysis could
-provide additional convenience.
+analyzed; our example in Section \ref{sec:staged-programming} for embedding a
+simply typed lambda calculus is like this. However, native metaprogramming
+features could be more concise and convenient than deep embeddings, similarly to
+how staged programming is more convenient than code generation with deeply
+embedded syntaxes.
 
@@
 \]
-However, the Yoneda lemma says that $\msf{decEq}$ cannot actually decide
+The Yoneda lemma says that $\msf{decEq}$ cannot actually decide
 definitional equality. We expect that $\msf{decEq}$ lets us define non-constant
@@
 supports staging operations and type dependencies. However, there are major
-differences to the current work. First, ibid.\ does not have universes or large
-elimination, nor does it allow staged manipulation of type expressions. Second,
-it does not support staged compilation in the sense of Definition
-\ref{def:staging}; rather, it is a system for runtime code manipulation and
-evaluation (which we may call staged \emph{computation}). In other words,
-programs may manipulate code fragments during execution, in more liberal ways
-than in 2LTT, but there is no judgment which ensures that a purely
-``object-level'' program can be extracted from a multi-stage program.
+differences to the current work. First, ibid.\ does not have large elimination,
+nor does it allow staged manipulation of type expressions. Second, it does not
+support staged compilation in the sense of Definition \ref{def:staging}; rather,
+it is a system for runtime code manipulation and evaluation (which we may call
+staged \emph{computation}). In other words, programs may manipulate code
+fragments during execution, in more liberal ways than in 2LTT, but there is no
+judgment which ensures that a purely ``object-level'' program can be extracted
+from a multi-stage program.
 
@@
 
-\section{Implementation}\label{sec:implementation}
-
 \section{Conclusions and Future Work}\label{sec:conclusions}
 
-TODO: psh explain, intro overview upd, conclusion, implementation description, related work type staging
+In this work, we have developed the foundational metatheory of 2LTT as a system
+of two-stage compilation, and gave glimpses into potential applications
+and extensions.
+
+We emphasize that variants of 2LTT can serve both as languages where staging
+features are implemented, and also as \emph{metatheories}, not only for the
+formalization for staging techniques, but also more generally for reasoning
+about constructions in object languages. This purpose is prominent in the
+original 2LTT use case in synthetic homotopy theory. The meta-level language in
+a 2LTT can be expressive enough to express general mathematics, and the
+meta-level propositional equality $\blank\!=_1\!\blank$ can be used to prove
+$\beta\eta$-equality of object-level types and terms.
+
+There is much future work in adapting existing staging techniques to 2LTT
+(which, as we mentioned, can double as formalization of said techniques), or
+adding staging where it has not been available previously.
+\begin{itemize}
+\item \emph{Let-insertion techniques}. These allow more precise control over
+  positions of insertion, or smarter selection of such positions. In 2LTT,
+  meta-level continuation monads could be employed to adapt some of the features
+  in \cite{DBLP:journals/jfp/KameyamaKS11}. The automatic let-floating feature
+  of MetaOCaml \cite{DBLP:journals/corr/abs-2201-00495} requires strengthening
+  of terms, which is not stable under substitution, but it is stable under
+  weakening, so it could be viewed as an ``intensional analysis'' feature.
+\item \emph{Generic programming with dependent types}. There is substantial
+  literature in this area
+  (e.g.\ \cite{loh11generic,chapman2010gentle,diehl-thesis,ornaments}), but only
+  in non-staged settings.  2LTT should immediately yield staging for these
+  techniques, assuming that the meta level has sufficient type formers
+  (e.g.\ induction-recursion). We could also try to adapt previous work on
+  generic treatment of partially static data types
+  \cite{DBLP:journals/pacmpl/YallopGK18}; fully internalizing this in 2LTT would
+  also require dependent types.
+\end{itemize}
+
+There are also ways that 2LTT itself could be enhanced.
+\begin{itemize}
+\item \emph{More stages, stage polymorphism}. Currently, there is substantial
+  code duplication if the object-level and meta-level languages are similar.
+  Stage polymorphism could help reduce this. A simple setup could include three
+  stages, the runtime one, the static one, and another one which supports
+  polymorphism over the first two stages.
+\item \emph{Modalities}. We mentioned the closed modality in Section
+  \ref{sec:closed-modalities}.  More generally, there should be many
+  \emph{multimodal} \cite{gratzer20multimodal} type theories supporting
+  staging. 2LTT itself can be viewed as a very simple multimodal theory with
+  $\Lift$ as the only modality, which is also degenerate (because it entails no
+  structural restriction on contexts). We could support multiple object
+  theories, with modalities for lifting them to meta-level or for representing
+  translations between them. We could also have modalities for switching between
+  stability under substitution and stability under weakening.
+\end{itemize}
+
+We could also further develop use-cases of restricted object theories. Here,
+2LTT could provide a convenient way to formalize certain optimizations and
+program translations. In essence, the object-level fragment of 2LTT can serve as
+a \emph{shallowly embedded} target of program translations. For example, we
+may choose a low-level object theory without higher-order functions, and formalize
+closure conversion as an interpretation of an embedded meta-level syntax into
+the object theory.
 

### 2022-03-02 paper/paper.tex
preprint formatting

@@
 
-\documentclass[acmsmall,anonymous,review]{acmart}
-%% \documentclass[nonacm,acmsmall]{acmart}
+%% \documentclass[acmsmall,anonymous,review]{acmart}
+\documentclass[nonacm,acmsmall,review]{acmart}
 %% \raggedbottom
@@
 \subsubsection{Inferring staging operations}
-We can extract a coercive subtyping system from the staging
-operations, which can be used to infer staging operations, and in particular
-transport definitions along preservation isomorphisms. One choice is to have $A
-\leq \Lift A$, $\Lift A \leq A$, a contravariant-covariant rule for functions
-and a covariant rule for $\Sigma$. During bidirectional elaboration, when we
-need to compare an inferred and an expected type, we can insert coercions.  We
+We can extract a coercive subtyping system from the staging operations, which
+can be used for inference, and in particular for automatically transporting
+definitions along preservation isomorphisms. One choice is to have $A \leq \Lift
+A$, $\Lift A \leq A$, a contravariant-covariant rule for functions and a
+covariant rule for $\Sigma$. During bidirectional elaboration, when we need to
+compare an inferred and an expected type, we can insert coercions.  We
 implemented this feature in our prototype. It additionally supports Agda-style
@@
 \item We have Russell-style $\U_{0,j} : \Rep \to \U_{0, j+1}\,r$, where $r$ is
-  some chosen runtime representation for types; often this would mark types as
-  erased at runtime. We leave the meta-level $\U_{1,j}$ hierarchy unchanged.
+  some chosen runtime representation for types; usually we would mark types are
+  erased. We leave the meta-level $\U_{1,j}$ hierarchy unchanged.
 \item We may introduce unboxed $\Sigma$ types and primitive machine types in the
@@
 \item On the other hand, if only stable constructions are possible, we never
-      have to \emph{prove} stability, or even explicitly mention object-theoretic
+      have to prove stability, or even explicitly mention object-theoretic
       contexts, substitutions and variables. The tedious details of deep embeddings
@@
 
-In this work, we have developed the foundational metatheory of 2LTT as a system
-of two-stage compilation, and gave glimpses into potential applications
-and extensions.
-
-We emphasize that variants of 2LTT can serve both as languages where staging
-features are implemented, and also as \emph{metatheories}, not only for the
-formalization for staging techniques, but also more generally for reasoning
-about constructions in object languages. This purpose is prominent in the
-original 2LTT use case in synthetic homotopy theory. The meta-level language in
-a 2LTT can be expressive enough to express general mathematics, and the
+We developed the foundational metatheory of 2LTT as a system of
+two-stage compilation. We view the current results as more like a starting point
+to more thorough investigation into applications and extensions; in this paper
+we only sketched these.
+
+We emphasize that variants of 2LTT can serve as languages where staging
+features can be practically implemented, and also as \emph{metatheories}, not
+only for the formalization for staging techniques, but also more generally for
+reasoning about constructions in object languages. This purpose is prominent in
+the original 2LTT use case in synthetic homotopy theory. The meta-level language
+in a 2LTT can be expressive enough to express general mathematics, and the
 meta-level propositional equality $\blank\!=_1\!\blank$ can be used to prove
 $\beta\eta$-equality of object-level types and terms.
+
+For example, we could use object theories as \emph{shallowly embedded} target
+languages of optimization techniques and program translations. In particular, we
+may choose a low-level object theory without higher-order functions, and
+formalize closure conversion as an interpretation of an embedded meta-level
+syntax into the object theory.
 
@@
   of terms, which is not stable under substitution, but it is stable under
-  weakening, so it could be viewed as an ``intensional analysis'' feature.
+  weakening, so perhaps it could be implemented as an ``intensional analysis''
+  feature.
 \item \emph{Generic programming with dependent types}. There is substantial
@@
 \item \emph{Modalities}. We mentioned the closed modality in Section
-  \ref{sec:closed-modalities}.  More generally, there should be many
-  \emph{multimodal} \cite{gratzer20multimodal} type theories supporting
-  staging. 2LTT itself can be viewed as a very simple multimodal theory with
-  $\Lift$ as the only modality, which is also degenerate (because it entails no
-  structural restriction on contexts). We could support multiple object
-  theories, with modalities for lifting them to meta-level or for representing
-  translations between them. We could also have modalities for switching between
-  stability under substitution and stability under weakening.
+  \ref{sec:closed-modalities}.  More generally, many \emph{multimodal}
+  \cite{gratzer20multimodal} type theories could plausibly support staging. 2LTT
+  itself can be viewed as a very simple multimodal theory with $\Lift$ as the
+  only modality, which is also degenerate (because it entails no structural
+  restriction on contexts). We could support multiple object theories, with
+  modalities for lifting them to the meta level or for representing translations
+  between them. We could also have modalities for switching between stability
+  under substitution and stability under weakening.
+\item \emph{Intensional analysis}. We only touched on the most basic semantics
+  of intensional analysis in Section \ref{sec:intensional-analysis}. It remains
+  a substantial challenge to work out the practicalities. For good ergonomics,
+  we would need something like a ``pattern matching'' operation on object-level
+  terms, or some induction principle which is more flexible than the plain
+  assumption of decidable object-level equality.
 \end{itemize}
-
-We could also further develop use-cases of restricted object theories. Here,
-2LTT could provide a convenient way to formalize certain optimizations and
-program translations. In essence, the object-level fragment of 2LTT can serve as
-a \emph{shallowly embedded} target of program translations. For example, we
-may choose a low-level object theory without higher-order functions, and formalize
-closure conversion as an interpretation of an embedded meta-level syntax into
-the object theory.
 

### 2022-03-03 paper/paper.tex
typos

@@
 \newcommand{\rexti}{\re_{\ext_1}^{-1}}
+\newcommand{\rextizero}{\re_{\ext_0}^{-1}}
 
@@
 This is because substitutions targeting $\Gamma \ext_0 A$ are the same as
-pairs of substitutions and terms, by the specification of $\blank\ext_0\blank$.
+pairs of substitutions and terms, by the specification of $\blank\!\ext_0\!\blank$.
 
@@
 \begin{alignat*}{3}
-  & (\Gamma \ext_0 A)\rel\,(\gamma,\,\alpha)\,(\rexti\,(\gamma',\,\alpha')) := (\gamma\rel : \Gamma\rel\,\gamma\,\gamma') \times (\emb{\alpha} = \alpha')
+  & (\Gamma \ext_0 A)\rel\,(\gamma,\,\alpha)\,(\rextizero\,(\gamma',\,\alpha')) := (\gamma\rel : \Gamma\rel\,\gamma\,\gamma') \times (\emb{\alpha} = \alpha')
 \end{alignat*}
@@
 From the soundness and stability of staging, we get that $\emb{\blank}$ is
-bijective on types and terms, i.e.\ $\Ty_{\mbbo j}\,\Gamma \simeq \Ty_{0,j}\,\emb{\Gamma}$ and
+bijective on types and terms, hence $\Ty_{\mbbo j}\,\Gamma \simeq \Ty_{0,j}\,\emb{\Gamma}$ and
 $\Tm_{\mbbo j}\,\Gamma\,A \simeq \Tm_{0,j}\,\emb{\Gamma}\,\emb{A}$.
@@
 differences to the current work. First, ibid.\ does not have large elimination,
-nor does it allow staged manipulation of type expressions. Second, it does not
+hence it does not support computing types by staging. Second, it does not
 support staged compilation in the sense of Definition \ref{def:staging}; rather,

### 2022-03-23 paper/paper.tex
typos, types2022 abstract

@@
   \Lift((x : A) \to B\,x) &\simeq ((x : \Lift A) \to \Lift\,(B\,\spl x))\\
-  \Lift ((x : A) \times B) &\simeq ((x : \Lift A) \times \Lift (B\,\spl x))\\
+  \Lift ((x : A) \times B\,x) &\simeq ((x : \Lift A) \times \Lift (B\,\spl x))\\
   \Lift \top_0 &\simeq \top_1
@@
 In other words, $n[\gamma']$ is canonical and is obtained as the serialization
-of $\ev\,n\,\gamma$. Therefore, $(\msf{NatElim}\,P\,z\,s\,n))[\gamma']$ is
+of $\ev\,n\,\gamma$. Therefore, $(\msf{NatElim}\,P\,z\,s\,n)[\gamma']$ is
 definitionally equal to $|\ev\,n|\,\gamma$-many applications of $s$ to

### 2022-03-23 types2022/abstract.tex
typos, types2022 abstract

\documentclass{easychair}
\raggedbottom

\usepackage{doc}
\usepackage{xcolor}
\usepackage{mathpartir}
\usepackage{todonotes}
\presetkeys{todonotes}{inline}{}
\usepackage{scalerel}
\usepackage{bm}
\usepackage{amssymb}

\renewcommand{\mit}[1]{\mathit{#1}}
\newcommand{\msf}[1]{\mathsf{#1}}
\newcommand{\mbb}[1]{\mathbb{#1}}
\newcommand{\mbf}[1]{\mathbf{#1}}
\newcommand{\bs}[1]{\boldsymbol{#1}}
\newcommand{\wh}[1]{\widehat{#1}}
\newcommand{\ext}{\triangleright}
\newcommand{\Code}{\msf{Code}}
\newcommand{\El}{\msf{El}}
\newcommand{\lam}{\msf{lam}}
\newcommand{\app}{\msf{app}}
\newcommand{\NatElim}{\msf{NatElim}}
\newcommand{\y}{\msf{y}}
\newcommand\whobj{\wh{\msf{Obj}}}

\newcommand{\Lift}{{\Uparrow}}
\newcommand{\spl}{{\sim}}
\newcommand{\qut}[1]{\langle #1\rangle}

\newcommand{\mbbc}{\mbb{C}}
\newcommand{\mbbo}{\mbb{O}}
\newcommand{\ob}{_\mbbo}

\newcommand{\U}{\msf{U}}
\newcommand{\Con}{\msf{Con}}
\newcommand{\Sub}{\msf{Sub}}
\newcommand{\Ty}{\msf{Ty}}
\newcommand{\Tm}{\msf{Tm}}
\newcommand{\Cono}{\msf{Con}_{\mbbo}}
\newcommand{\Subo}{\msf{Sub}_{\mbbo}}
\newcommand{\Tyo}{\msf{Ty}_{\mbbo}}
\newcommand{\Tmo}{\msf{Tm}_{\mbbo}}
\newcommand{\hCon}{\wh{\msf{Con}}}
\newcommand{\hSub}{\wh{\msf{Sub}}}
\newcommand{\hTy}{\wh{\msf{Ty}}}
\newcommand{\hTm}{\wh{\msf{Tm}}}

\newcommand{\p}{\msf{p}}
\newcommand{\q}{\msf{q}}

\newcommand{\Bool}{\msf{Bool}}
\newcommand{\true}{\msf{true}}
\newcommand{\false}{\msf{false}}
\newcommand{\True}{\msf{True}}
\newcommand{\False}{\msf{False}}
\newcommand{\List}{\msf{List}}
\newcommand{\nil}{\msf{nil}}
\newcommand{\cons}{\msf{cons}}
\newcommand{\Nat}{\msf{Nat}}
\newcommand{\zero}{\msf{zero}}
\newcommand{\suc}{\msf{suc}}
\newcommand{\ttt}{\msf{tt}}
\newcommand{\fst}{\msf{fst}}
\newcommand{\snd}{\msf{snd}}
\newcommand{\mylet}{\msf{let}}
\newcommand{\emptycon}{\scaleobj{.75}\bullet}
\newcommand{\id}{\msf{id}}

\newcommand{\Set}{\msf{Set}}
\newcommand{\Rep}{\msf{Rep}}
\newcommand{\blank}{{\mathord{\hspace{1pt}\text{--}\hspace{1pt}}}}
\newcommand{\emb}[1]{\ulcorner#1\urcorner}

\newcommand{\Stage}{\msf{Stage}}
\newcommand{\hato}{\bm\hat{\mbbo}}
\newcommand{\ev}{\mbb{E}}
\newcommand{\re}{\mbb{R}}

\theoremstyle{remark}
\newtheorem{notation}{Notation}

\newcommand{\whset}{\wh{\Set}}
\newcommand{\rexti}{\re_{\ext_1}^{-1}}
\newcommand{\rextizero}{\re_{\ext_0}^{-1}}
\newcommand{\rel}{^{\approx}}
\newcommand{\yon}{\msf{y}}

\title{Conservativity of Two-Level Type Theory Corresponds to Staged Compilation}

\author{Andr\'as Kov\'acs}

% Institutes for affiliations are also joined by \and,
\institute{
  E\"otv\"os Lor\'and University,
  Budapest, Hungary \\
  \email{kovacsandras@inf.elte.hu}
}

\authorrunning{Kov\'acs}
\titlerunning{Using Two-Level Type Theory for Staged Compilation}
\pagenumbering{gobble}
\begin{document}

\maketitle

%% %% abstract

%% Two-level type theory (2LTT) was originally intended as a tool to make
%% metatheoretical reasoning about constructions in homotopy type theory more
%% convenient, by internalizing an extensional metatheory in a syntactic layer. We
%% observe that variants of 2LTT formalize setups with two-stage compilation.
%% Additionally, the conservativity property of a 2LTT over the object-level theory
%% is equivalent to having a sound and stable staging algorithm, which compiles the
%% object-level fragment of 2LTT to the syntax of the object theory. In previous
%% work on 2LTT by Annekov et al. only a weak form of conservativity is shown,
%% which corresponds to staging without the soundness property. We define staging
%% as evaluation of 2LTT syntax in the presheaf model over the object-level
%% syntactic category. Operationally, this is in similar to
%% normalization-by-evaluation, except that evaluation in the meta-level fragment
%% is closed (i.e. no free variables occur) and thus more efficient. We show
%% stability and soundness of staging; the former is straightforward, and the
%% latter is given by a proof-relevant logical relation internally to the presheaf
%% model. Equivalently, staging and its soundness can be established in a single
%% step, by gluing along the functor which restricts 2LTT syntax to object-level
%% typing contexts.

Two-level type theory (2LTT) \cite{twolevel} was originally intended as a tool
to make metatheoretical reasoning about constructions in homotopy type theory
more convenient, by internalizing an extensional metatheory in a syntactic
layer.
\\\\
\textbf{2LTT as a two-stage language.} First, we observe that basic variants of
2LTT formalize two-stage compilation, where the meta-level syntactic fragment contains
static (compile-time) computations, and \emph{staging} is the algorithm which
performs all static computation, producing object-theoretic syntax as output. The
basic rules are the following. There are two universes, $\U_0$ and $\U_1$, where
$\U_0$ classifies object-level (runtime) types and $\U_1$ classifies meta-level
(compile time) types. There are three \emph{staging operations}.

\emph{Lifting:} for $A : \U_0$, we have $\Lift A : \U_1$. From the staging point
of view, $\Lift A$ is the type of metaprograms which compute runtime expressions
of type $A$.

\emph{Quoting:} for $A : \U_0$ and $t : A$, we have $\qut{t} :\,\Lift A$.  A
quoted term $\qut{t}$ represents the metaprogram which immediately yields $t$.

\emph{Splicing:} for $A : \U_0$ and $t :\,\Lift A$, we have $\spl t : A$.
During staging, the metaprogram in the splice is executed, and the resulting
expression is inserted into the output.

Quoting and splicing are definitional inverses. Also, the above
operations are the \emph{only} way of crossing between stages; all type formers
stay within a single stage, and in particular we cannot eliminate from one stage
to a different one. The staging interpretation of 2LTT remains valid with
arbitrary assumed type formers. Note that all three operations correspond to
features in existing staged systems such as MetaOCaml \cite{kiselyov14metaocaml}
or typed Template Haskell \cite{typed-th}, although none of the existing systems
support staging with dependent types.
\\\\
\textbf{Conservativity as staging.} By conservativity we mean the following.
There is an embedding morphism $\emb{\blank}$ which maps from the object theory
to the object-level syntactic fragment of 2LTT. 2LTT is conservative if
$\emb{\blank}$ is bijective on types and terms,
i.e.\ $\Ty_{\msf{Obj}}\,\Gamma\ \simeq \Tm_{\msf{2LTT}}\,\emb{\Gamma}\,\U_0$ and
$\Tm_{\msf{Obj}}\,\Gamma\,A \simeq \Tm_{\msf{2LTT}}\,\emb{\Gamma}\,\emb{A}$.

A staging algorithm consists of functions $\Stage :
\Tm_{\msf{2LTT}}\,\emb{\Gamma}\,\U_0 \to \Ty_{\msf{Obj}}\,\Gamma$ and $\Stage :
\Tm_{\msf{2LTT}}\,\emb{\Gamma}\,\emb{A} \to \Tm_{\msf{Obj}}\,\Gamma\,A$. We call
$\Stage$ \emph{stable} if $\Stage\,\circ\,\emb{\blank} = \id$, and \emph{sound}
if $\emb{\blank}\,\circ\,\Stage = \id$. Hence, conservativity is the same as
having a sound and stable staging algorithm. In \cite{twolevel}, only a weak
form of conservativity is shown, which corresponds to staging without soundness.
\\\\
\textbf{Staging by evaluation.} We define $\Stage$ as the evaluation of 2LTT
types and terms in the presheaf model over the syntactic category of the object
theory. We call this model $\whobj$. The object-level 2LTT fragment is
interpreted using sets of types and terms in the object theory. Operationally,
this yields closed evaluation for the meta-level 2LTT fragment, and we get naive
weakening for object-theoretic terms. Naive weakening can be inefficient, but in
practice it can be optimized using De Bruijn levels and delayed variable
renamings. The same efficiency issue arises in presheaf-based
normalization-by-evaluation, and the same solution applies there. However,
staging can be overall more efficient, because meta-level evaluation is closed
(no free variables occur in values).
\\\\
\textbf{Soundness.} Stability of staging follows by straightforward induction on the object theory;
soundness requires more effort. We define a \emph{restriction} morphism, which
maps from 2LTT to $\whobj$, restricting the meta-level syntactic fragment so
that it can only depend on object-level typing contexts, that is, contexts given
as $\emb{\Gamma}$ for some $\Gamma$. Then, we show soundness of staging by a
proof-relevant logical relation between the evaluation morphism from 2LTT to
$\whobj$ and the restriction morphism. We define this logical relation in the
internal language of $\whobj$, to avoid the deluge of boilerplate for showing
stability under object-theoretic substitution.

Alternatively, staging together with its soundness can be established in a
single step, by gluing along the restriction morphism. This interpretation can
be more compactly defined using synthetic Tait computability
\cite{sterlingthesis}.
\\\\
\textbf{Intensional analysis.} This means analyzing the internal structure of
object-level terms, i.e.\ values with type $\Lift\,A$. While intensional
analysis can be often simulated with deeply embedded inductive syntaxes at the
meta level, it may be more concise and convenient to use native intensional
analysis features instead. We consider the interpretation of such features in
the presheaf models.

If the object theory has parallel substitutions as syntactic morphisms,
then $\whobj$ does not support intensional analysis. For illustration, consider
decidable equality of $\Lift A$ as an intensional meta-level axiom; this is
essentially decidability of definitional equality of object terms. This
axiom does not hold in $\whobj$, because inequality of object-level terms
is not stable under substitution: inequal variables can be mapped to equal
terms.

However, we may choose to only have \emph{weakenings} as morphisms in the object
syntax. In this case, decidable equaity for $\Lift A$ holds in $\whobj$, since
term inequality is stable under weakening. As a trade-off, if there is no notion
of substitution in the specification of the object theory, it is not possible to
specify dependent or polymorphic types there. This is still sufficient for many
practical use cases, for example when the object theory is simply-typed, in
which case staging also performs \emph{monomorphization}. In this setup, it makes
sense to only have weakening in the equational theory of the object theory, but
no $\beta\eta$-rules and no substitution. The reason is that we do not want
to equate programs with different performance characteristics, when we do
staging in order to improve runtime code performance.
\\\\
\textbf{Future work.} A major line of future work is to connect 2LTT to existing
literature on staged compilation. This would involve formalizing existing tricks
and techniques, such as let-insertion techniques
\cite{DBLP:journals/corr/abs-2201-00495}, fusion, various binding-time
improvements and CPS conversions \cite{partial-evaluation}. Another line is
fleshing out practical details for staging and intensional analysis. In staging,
a production-strength solution should include some form of caching, to reduce
code duplication. In intensional analysis, some form of induction or pattern
matching would be more ergonomic than plain decidability of conversion.

Also, 2LTT could be extended to more general \emph{multimodal}
\cite{gratzer20multimodal} type theories, where modalities represent morphisms
between different object-theoretic syntactic categories, in possibly different
object theories. A simple example is the closed or ``crisp'' modality
\cite{licata2018internal} which can be used to represent closed object terms.

\bibliographystyle{plain}
\bibliography{references}

\end{document}

### 2022-06-09 paper/paper.tex
revisions

@@
 
-%% \documentclass[acmsmall,anonymous,review]{acmart}
-\documentclass[nonacm,acmsmall,review]{acmart}
+\documentclass[acmsmall,anonymous,review]{acmart}
+%% \documentclass[nonacm,acmsmall,review]{acmart}
 %% \raggedbottom
@@
 
+\newcommand{\refl}{\msf{refl}}
 \newcommand{\Bool}{\msf{Bool}}
@@
 \newcommand{\Set}{\mathsf{Set}}
+\newcommand{\Prop}{\mathsf{Prop}}
 \newcommand{\Rep}{\msf{Rep}}
@@
   features and applications in staging. Then, we present a staging algorithm and
-  provide a proof of correctness. Our algorithm is ``staging-by-evaluation'',
+  prove its soundness and stability. Our algorithm is ``staging-by-evaluation'',
   analogously to the technique of normalization-by-evaluation, in that staging
-  is given by the evaluation of 2LTT syntax in a semantic domain. The staging
-  algorithm together with its correctness proof constitutes a proof of strong
-  conservativity of 2LLT over the object theory. To our knowledge, this is the
-  first description of staged compilation which supports full dependent types and
-  unrestricted staging for types.
+  is given by the evaluation of 2LTT syntax in a semantic domain. A sound and
+  stable staging algorithm constitutes a proof of strong conservativity of 2LLT
+  over the object theory. To our knowledge, this is the first description of
+  staged compilation which supports full dependent types and unrestricted
+  staging for types.
 \end{abstract}
@@
 
-It turns out that metamathematical convenience closely corresponds to to
-metaprogramming convenience: 2LTT can be directly and effectively employed
-in staged compilation. Moreover, semantic ideas underlying 2LTT are also
-directly applicable to the theory of staging.
+It turns out that metamathematical convenience closely corresponds to
+metaprogramming convenience: 2LTT can be directly and effectively employed in
+staged compilation. Moreover, semantic ideas underlying 2LTT are also directly
+applicable to the theory of staging.
 
@@
 \begin{itemize}
-  \item In Section \ref{sec:tour-of-2ltt} we present an informal syntax of two-level
-    type theory, a dependent type theory with staging features. We look at basic
-    use cases involving inlining control, partial evaluation and fusion
-    optimizations. We also describe feature variations, enabling applications in
-    monomorphization and memory layout control.
-  \item In Section \ref{sec:formal-2ltt}, we specify the formal notion of models and
-    syntax for the variant of 2LTT used in this paper, mostly following \cite{twolevel}.
-  \item In Section \ref{sec:staging-algorithm}, we review the standard presheaf
+  \item In Section \ref{sec:tour-of-2ltt} we present an informal syntax of
+    two-level type theory, a dependent type theory with support for two-stage
+    compilation. We look at basic use cases involving inlining control, partial
+    evaluation and fusion optimizations. We also describe feature variations,
+    enabling applications in monomorphization and memory layout control.
+  \item In Section \ref{sec:formal-2ltt}, we specify the formal notion of models
+    and syntax for the variant of 2LTT used in this paper. We mostly work with a
+    high-level algebraic presentation, where the syntax is quotiented by
+    conversion. However, we also explain how to extract functions
+    operating on non-quotiented syntax, by interpreting type-theoretic
+    constructions in a setoid model.
+  \item
+    In Section \ref{sec:staging-algorithm}, we review the standard presheaf
     model of 2LTT \cite[Section~2.5.3]{twolevel}, which lies over the syntactic
     category of the object theory. We show that evaluation in this model yields
-    a staging algorithm for closed types and terms. We then prove stability for
-    staging, which means that staging the output of staging has no action. Then,
-    we extend staging to open types and terms, which may depend on object-level
-    contexts.
+    a staging algorithm for closed types and terms. We then extend staging to
+    open types and terms, which may depend on object-level contexts. We show
+    stability of staging, which means that staging is surjective up to
+    conversion, and we also show that staging strictly preserves all type and
+    term formers. Finally, we discuss efficiency and potential practical implementations
+    of staging.
   \item
-    In Section \ref{sec:soundness} we show soundness of staging, which means that
-    the output of staging is convertible to the input.  Staging together with
+    In Section \ref{sec:soundness} we show soundness of staging, which roughly means that
+    the output of staging is convertible to its input. Staging together with
     its stability and soundness can be viewed as a \emph{strong conservativity}
@@
   \item
-    In Section \ref{sec:intensional-analysis}, we discuss possible interpretations
-    of intensional code analysis in the presheaf model.
+    In Section \ref{sec:intensional-analysis}, we discuss possible semantic
+    interpretations of intensional code analysis, i.e.\ the ability to look into
+    the structure of object-level code.
   \item We provide a small standalone implementation of elaboration and staging
     for a two-level type theory, with code samples that expand on the paper.
-  \item To our knowledge, this is the first account of staged compilation in the
-    presence of full-blown dependent types, with universes and large
-    elimination. In particular, we allow unrestricted staging for types, so that
-    types can be computed by metaprograms at compile time.
+  \item To our knowledge, this work is the first formalization and implementation of
+    staged compilation in the presence of full dependent types, with
+    universes and large elimination. In particular, we allow unrestricted
+    staging for types, so that types can be computed by metaprograms at compile
+    time.
 \end{itemize}
@@
 $\U_{i,j} : \U_{i, j+1}$. However, for the sake of brevity we will usually omit
-the $j$ indices in this section, since sizing is orthogonal to our use-cases and
+the $j$ indices in this section, as sizing is orthogonal to our use-cases and
 examples.
@@
 
-\begin{notation} We may disambiguate type formers at different stages by using $0$ or $1$
+\begin{notation}\label{notation:basics} We may disambiguate type formers at different stages by using $0$ or $1$
 subscripts. For example, $\Nat_1 : \U_1$ is distinguished from $\Nat_0 : \U_0$,
@@
 
-In 2LTT, we may have several different polymorphic identity functions. First, we
-consider the usual identity functions at each stage:
+In 2LTT, we may have several different polymorphic identity functions. First,
+consider the usual identity function at each stage:
 \begin{alignat*}{3}
@@
         \spl \qut{\Nat_0}\,\spl \qut{\List_0\,\spl\qut{Nat_0}}\,
-        (\lambda\,a\,\mit{bs}.\, \cons_0\,\spl B\,\spl\qut{\spl\qut{a} +_0 10}\,\mit{bs})\,
+        (\lambda\,a\,\mit{bs}.\, \cons_0\,\spl\qut{\Nat_0}\,\spl\qut{\spl\qut{a} +_0 10}\,\mit{bs})\,
         (\nil_0\,\spl \qut{\Nat_0})\,
@@
 duplicate the code of the mapping function. In such cases, we can use
-\emph{let-insertion}, a technique in staged compilation. If we bind a runtime
+\emph{let-insertion}, a common technique in staged compilation. If we bind a runtime
 expression to a runtime variable, and only use that variable in subsequent
@@
 \end{alignat*}
-We include examples of slightly more sophisticated let-insertion in the supplementary
+We include examples of more sophisticated let-insertion in the supplementary
 code.
@@
 lambda calculus. We sketch the implementation in the following; the full version
-can be found in the supplementary code. First, we inductively define contexts,
-types and terms:
+can be found in the supplementary code. First, we inductively define types,
+contexts and terms:
 \begin{alignat*}{6}
@@
 staging output: an embedded lambda expression is staged to the corresponding
-lambda expression in the runtime language. This is similar to the partial
+lambda expression in the object language. This is similar to the partial
 evaluator presented in Idris 1 \cite{scrap-your-inefficient-engine}. However, in
@@
 
-We describe more generally the action of $\Lift$ on type formers. First, $\Lift$
-preserves negative type formers up to definitional isomorphism \cite[Section~2.3]{twolevel}:
+We describe more generally the action of lifting on type formers. While lifting
+does not have definitional computation rules, it does preserve negative type formers up
+to definitional isomorphism \cite[Section~2.3]{twolevel}:
 \begin{alignat*}{5}
@@
   & \hspace{-6em}\rlap{$\mit{pres}_\to : \Lift((x : A) \to B\,x) \to ((x : \Lift A) \to \Lift\,(B\,\spl x))$}\\
-  & \hspace{-6em}\mit{pres}_\to\,f     &&:= \lambda\,x.\,\,\qut{f\,\spl x}\\
+  & \hspace{-6em}\mit{pres}_\to\,f     &&:= \lambda\,x.\,\,\qut{\spl f\,\spl x}\\
   & \hspace{-6em}\mit{pres}_\to^{-1}\,f &&:= \qut{\lambda\,x.\,\,\spl(f\,\qut{x})}
@@
 literature \cite[Chapter~12]{partial-evaluation}. Note that the output of $\mit{pres}_{\to}$ uses a
-meta-level $\lambda$, while going the other way we introduce a runtime
-binder. Meta-level function and $\Sigma$ types support more computation during
+meta-level $\lambda$, while going the other way introduces a runtime
+binder. Meta-level function types and $\Sigma$ types support more computation during
 staging, so in many cases it is beneficial to use the improved forms. In some
@@
 
-For inductive types in general we only get oplax preservation. For example, we
-have $\Bool_1 \to \Lift \Bool_0$, defined as
+For inductive types in general we do not get full preservation, only maps in one
+direction. For example, we have $\Bool_1 \to \Lift \Bool_0$, defined as
 $\lambda\,b.\,\msf{if}\,b\,\msf{then}\,\qut{\true_0}\,\msf{then}\,\qut{\false_0}$.
@@
   and $\msf{Prod} : \Rep \to \Rep \to \Rep$, and additionally we may assume
-  any desired primitive machine representation as a $\Rep$.
+  any desired primitive machine representation as a value of $\Rep$.
 \item We have Russell-style $\U_{0,j} : \Rep \to \U_{0, j+1}\,r$, where $r$ is
@@
 
-\section{Formal 2LTT}\label{sec:formal-2ltt}
-
-In this section we switch to a formal description of 2LTT, which we will use in
-the subsequent sections to define the staging algorithm and prove its
-correctness. First we describe the metatheory that we work in.
-
-\subsection{Metatheory} We work in an extensional type theory with
-a cumulative ordinal-indexed universe hierarchy $\Set_j$. For type formers, we
-use Notation \ref{basic-notation}. We also use extensional equality, written as
-$\blank\!=\!\blank$. We will sometimes use $\Set_\omega$ and $\Set_{\omega+1}$.
-This is mainly a convenience feature when modeling contexts. Types are indexed
-by $\mbb{N}$ sizes, so we usually model a type at finite level $j$ using
-$\Set_j$, but contexts contain types at possibly different levels. The easy
-solution is to model contexts using $\Set_\omega$, which is large enough for all
-constituent types. Alternatively, we could formally track the sizing levels for
-contexts as well, but that adds inessential noise to the specification.
-
-\subsection{Models and Syntax of 2LTT}
-We use an algebraic specification for models, and specify the syntax as the
-initial model. We only handle well-formed syntactic objects and only define
-operations which respect syntactic definitional equality; in fact, we identify
-definitional equality with metatheoretic equality. First, we define the
+\section{Formal Setup}\label{sec:formal-2ltt}
+
+In this section we describe our approach to formalizing staging and also the
+\emph{extraction} of algorithms from the formalization. There is a tension
+here:
+\begin{itemize}
+  \item On one hand, we would like to do formal work at a higher level of
+    abstraction, in order to avoid a deluge of technical details (for which
+    dependent type theories are infamous).
+  \item On the other hand, at some point we need to talk about lower-level
+    operational aspects and extracted algorithms.
+\end{itemize}
+In the following, we describe our approach to interfacing between the different levels
+of abstraction.
+
+\subsection{The Algebraic Setting}
+
+When defining staging and proving its soundness and stability, we work
+internally in a constructive type theory, using types and terms that are
+quotiented by conversion. In this setting, every construction respects syntactic
+conversion, which is very convenient technically. Here, we use standard
+definitions and techniques from the categorical and algebraic metatheory of type
+theories, and reuse prior results which rely on the same style of
+formalization. Concretely, we use \emph{categories with families} \cite{cwfs}
+which is a commonly used specification of an \emph{explicit substitution
+calculus} of types and terms.
+
+In the rest of the paper we will mostly work in this setting, and we will
+explicitly note when we switch to a different metatheory. We summarize
+the features of this metatheory in the following.
+
+\subsubsection{Metatheory}
+The metatheory is an intensional type theory with a cumulative universe
+hierarchy $\Set_i$, where $i$ is ordinal number such that $i \leq \omega +
+1$. The transfinite $i$ levels are a convenience feature for modeling typing
+contexts, where sometimes we will need universes large enough to fit
+types at arbitrary finite levels. We also assume uniqueness of identity proofs
+and function extensionality. We will omit transports along identity proofs in
+the paper, for the sake of readability. The missing transports can be in
+principle recovered due to Hofmann's conservativity result
+\cite{hofmann95extensional}; see also Winterhalter et
+al. \cite{DBLP:conf/cpp/WinterhalterST19}. We assume $\Pi$, $\Sigma$, $\top$ and
+$\mbb{N}$ types in all universes, and reuse Notation \ref{notation:basics}. We
+also assume certain quotient inductive-inductive types, to represent the syntax of 2LTT
+and the object theory. We will detail these shortly in Sections \ref{sec:2lttmod}
+and \ref{sec:objmod}.
+
+\subsection{Algorithm Extraction}
+
+In the algebraic setting we can \emph{only} talk about properties which
+respect syntactic conversion. However, the main motivation of staged compilation is to
+improve performance and code abstraction, but these notions do \emph{not}
+respect conversion. For instance, rewriting programs along $\beta$-conversion
+can improve performance. Hence, at some point we may want to ``escape'' from
+quotients.
+
+Fortunately, our specified meta type theory can be interpreted in
+\emph{setoids}, as most recently explained by Pujet and Tabareau
+\cite{DBLP:journals/pacmpl/PujetT22}. A setoid is a set together with an
+equivalence relation on it. We interpret closed types as setoids, and
+dependent types as setoid fibrations. For inductive base types such $\Bool$ or
+$\Nat$, the equivalence relation is interpreted as identity. Functions are
+interpreted as equivalence-preserving functions, and two functions are equivalent if
+they are pointwise equivalent. A quotient type $A/R$ is interpreted by extending
+the semantic notion of equivalence for $A$ with the given $R$ relation.
+
+\begin{definition} Assume a closed function $f : A \to B$ defined
+in the mentioned type theory. The \textbf{extraction} of $f$ is
+the underlying function of $f$'s interpretation in the setoid model.
+\end{definition}
+
+In particular, from $f : A/R \to A/R$ we can extract a function which acts on
+the underlying set of $A$'s interpretation. Note that an extracted function can
+be only viewed as an algorithm if the setoid interpretation is itself defined in
+a constructive theory, but that is the case in the cited work
+\cite{DBLP:journals/pacmpl/PujetT22}.
+
+In other words, we can view the usage of constructive quotients as a convenient
+shorthand for working with explicit equivalence relations (an \emph{internal
+language} of setoids). When we reason about the syntax of a type theory, this
+provides a bridge between the categorical/algebraic style and the traditional
+style that uses separate equivalence relations. It makes sense to work with
+quotients internally to a type theory whenever we only want to talk about
+constructions and properties which respect conversion. However, at any point we
+have the option to switch to an external perspective and look at underlying sets
+and the underlying maps.
+
+\subsection{Models and Syntax of 2LTT}\label{sec:2lttmod}
+
+We specify the notion of model for 2LTT in this section. The syntax of 2LTT will
+be defined as the initial model, as a quotient inductive-inductive type,
+following \cite{ttintt}. This yields an explicit substitution calculus,
+quotiented by definitional equality. A major advantage of this representation is
+that it \emph{only} includes well-typed syntax, so we never have to talk about
+``raw'' types and terms.
+
+First, we define the
 structural scaffolding of 2LTT without type formers, mostly following
@@
   it is clear from context.
+
+  In the initial model, an element of $\Con$ is a \emph{typing context}, an
+  element of $\Sub$ is a \emph{parallel substitution}, $\emptycon$ is the
+  \emph{empty typing context}, and $\epsilon$ is the empty list viewed as a
+  parallel substitution. The identity substitution $\id : \Sub\,\Gamma\,\Gamma$
+  maps each variable to itself, and $\sigma \circ \delta$ can be computed
+  by substituting each term in $\sigma$ (which is a list of terms) with $\delta$.
 \item
-  For each $i \in \{0,1\}$ and $j \in \mbb{N}$, we have $\Ty_{i,j} : \Con
-  \to \Set$ and $\Tm_{i,j} : (\Gamma : \Con) \to \Ty_{i,j}\,\Gamma \to \Set$,
-  where $\Ty$ is a presheaf over $\mbbc$ and $\Tm_{i,j}$ is a presheaf over the
-  category of elements of $\Ty_{i,j}$. This means that both types ($\Ty$) and
-  terms ($\Tm$) can be substituted, and substitution has functorial action. We
-  use $A$, $B$, $C$ to refer to types and $t$, $u$, $v$ to refer to terms, and
-  use $A[\sigma]$ and $t[\sigma]$ for substituting types and
-  terms. Additionally, for each $\Gamma : \Con$ and $A : \Ty_{i,j}\,\Gamma$, we
-  have the extended object $\Gamma \ext A : \Con$ such that there is a natural
-  isomorphism $\Sub\,\Gamma\,(\Delta\ext A) \simeq ((\sigma :
-  \Sub\,\Gamma\,\Delta) \times \Tm_{i,j}\,\Gamma\,(A[\sigma]))$. We will sometimes
-  use $\blank\!\ext_0\!\blank$ and $\blank\!\ext_1\!\blank$ to disambiguate object-level
-  and meta-level context extensions.
+  For each $i \in \{0,1\}$ and $j \in \mbb{N}$, we have $\Ty_{i,j} : \Con \to
+  \Set$ and $\Tm_{i,j} : (\Gamma : \Con) \to \Ty_{i,j}\,\Gamma \to \Set$, as
+  sets of types and terms. Both types and terms can be substituted:
+  \begin{alignat*}{3}
+    & \blank\![\blank\!] &&: \Ty_{i,j}\,\Delta \to \Sub\,\Gamma\,\Delta \to \Ty_{i,j}\,\Gamma \\
+    & \blank\![\blank\!] &&: \Tm_{i,j}\,\Delta\,A \to (\sigma : \Sub\,\Gamma\,\Delta) \to \Tm_{i,j}\,\Gamma\,(A[\sigma])
+  \end{alignat*}
+  Additionally, we have $A[\id] = A$ and $A[\sigma \circ \delta] =
+  A[\sigma][\delta]$, and we have the same equations for term substitution as
+  well.
+
+  We also have a \emph{comprehension structure}: for each $\Gamma : \Con$
+  and $A : \Ty_{i,j}\,\Gamma$, we have the extended context $\Gamma \ext A :
+  \Con$ such that there is a natural isomorphism $\Sub\,\Gamma\,(\Delta\ext A)
+  \simeq (\sigma : \Sub\,\Gamma\,\Delta) \times
+  \Tm_{i,j}\,\Gamma\,(A[\sigma])$. We will sometimes use
+  $\blank\!\ext_0\!\blank$ and $\blank\!\ext_1\!\blank$ to disambiguate
+  object-level and meta-level context extensions.
 \item
@@
 
-In short, for each $i$ and $j$ we have a family structure in the sense of
-categories-with-families \cite{Dybjer96internaltype}, such that there is a
-natural transformation from each $(\Ty_{0,j},\,\Tm_{0,j})$ to $(\Ty_{1,j},\,\Tm_{1,j})$
-which is invertible on terms.
+%% In short, for each $i$ and $j$ we have a family structure in the sense of
+%% categories-with-families \cite{Dybjer96internaltype}, such that there is a
+%% natural transformation from each $(\Ty_{0,j},\,\Tm_{0,j})$ to $(\Ty_{1,j},\,\Tm_{1,j})$
+%% which is invertible on terms.
 
@@
   \Sub\,\Gamma\,\Delta) \times \Tm_{i,j}\,\Gamma\,(A[\sigma]))$, we recover the
-  operation which extends a morphism with a term. In the syntax (initial model),
-  this justifies the view of $\Sub$ as a type of lists of terms, i.e.\ a
-  parallel substitutions. We denote the extension operation as $(\sigma,\,t)$
-  for $\sigma : \Sub\,\Gamma\,\Delta$.
+  operation which extends a morphism with a term. In the initial model, this
+  extends a parallel substitution with an extra term, thus justifying the view
+  of substitutions as lists of terms. We denote the extension operation as
+  $(\sigma,\,t)$ for $\sigma : \Sub\,\Gamma\,\Delta$.
 \end{itemize}
@@
 
-\begin{definition} A \textbf{model of 2LTT} is a model of basic 2LTT
-which supports certain type formers. For the sake of brevity, we concretely
-specify here only a small collection of type formers. However, we will argue later
-that our results extend to more general notions of inductive types. We specify
-type formers in the following. We assume that all type formers are natural,
-i.e.\ stable under substitution.
+\begin{definition}
+A \textbf{model of 2LTT} is a model of basic 2LTT which supports certain type
+formers. For the sake of brevity, we specify here only a small collection of
+type formers. However, we will argue later that our results extend to more
+general notions of inductive types. We specify type formers in the following. We
+omit substitution rules for type and term formers; all of these can be specified in
+an evident structural way, e.g.\ as in $(\Lift A)[\sigma] = \Lift(A[\sigma])$.
 \begin{itemize}
@@
   and $B : \Ty_{i,j}\,(\Gamma \ext (x : A))$, together with a natural isomorphism of pairing and projections:
-  \[ \Tm_{i,j}\,\Gamma\,(\Sigma\,A\,B) \simeq ((t : \Tm_{i,j}\,\Gamma\,A) \times \Tm_{i,j}\,\Gamma\,(B[x\mapsto t]) \]
+  \[ \Tm_{i,j}\,\Gamma\,(\Sigma\,(x : A)\,B) \simeq ((t : \Tm_{i,j}\,\Gamma\,A) \times \Tm_{i,j}\,\Gamma\,(B[x\mapsto t]) \]
   We write $(t,\,u)$ for pairing and $\fst$ and $\snd$ for projections.
-\item \emph{Function types}. We have $\Pi\,(x : A)\,B : \Ty_{i,j}\,\Gamma$ for $A :
-  \Ty_{i,j}\,\Gamma$ and $B : \Ty_{i,j}\,(\Gamma \ext (x : A))$, together with
-  $\app : \Tm_{i,j}\,\Gamma\,(\Pi\,A\,B) \to \Tm_{i,j}\,(\Gamma \ext (x :
-  A))\,B$ and its inverse $\lam$.
+\item \emph{Function types}.
+  We have $\Pi\,(x : A)\,B : \Ty_{i,j}\,\Gamma$ for $A : \Ty_{i,j}\,\Gamma$ and
+  $B : \Ty_{i,j}\,(\Gamma \ext (x : A))$, together with abstraction and application
+  \begin{alignat*}{3}
+    & \lam &&: \Tm_{i,j}\,(\Gamma \ext (x : A))\,B \to \Tm_{i,j}\,\Gamma\,(\Pi\,(x : A)\,B) \\
+    & \app &&: \Tm_{i,j}\,\Gamma\,(\Pi\,(x : A)\,B) \to \Tm_{i,j}\,(\Gamma \ext (x : A))\,B
+  \end{alignat*}
+  such that $\lam$ and $\app$ are inverses. This specification of $\app$ is
+  equivalent to the ``traditional'' specification:
+  \[ \app' : (t : \Tm_{i,j}\,\Gamma\,(\Pi\,(x :
+  A)\,B))\to (u : \Tm_{i,j}\,\Gamma\,A) \to \Tm_{i,j}\,\Gamma\,(B[x \mapsto
+    u]).\]
+  The traditional version is definable as $(\app\,t)[\id,\,u]$. We use
+  the $\app$-$\lam$ isomorphism because it is formally more convenient.
+
 \item \emph{Natural numbers}. We have $\Nat_{i,j} : \Ty_{i,j}\,\Gamma$, $\zero_{i,j} : \Tm_{i,j}\,\Gamma\,\Nat_{i,j}$,
@@
 \begin{definition}
-The \textbf{syntax of 2LTT} is defined to be the initial model of 2LTT, where
-all underlying sets are in $\Set_0$. The syntax can be assumed to exist, since
-the notion of model is algebraic; the specification can be expressed using a
-suitable notion of algebraic signatures, such as one for essentially algebraic
-theories \cite{partialhorn} or as a finitary quotient inductive-inductive
-signature \cite{kaposi2019constructing}. The initiality of syntax directly
-yields a notion of \emph{recursion on syntax}, and it can be shown that
-initiality also implies an \emph{induction principle} \cite{kaposi2019constructing}.
-\end{definition}
-
-The examples in Section \ref{sec:tour-of-2ltt} can be faithfully represented in
-the formal syntax. The only notable difference is that the formal syntax uses
-Coquand-style universes instead of Russell-style ones. We use the former because
-it is much easier to model in the semantics in Section \ref{sec:staging-algorithm}. We note
-though that it is straightforward to elaborate from a Russell-style surface
-syntax to the formal version, by inserting $\El$ and $\Code$ as required.
+The \textbf{syntax of 2LTT} is the initial model of 2LTT, defined as a quotient
+inductive-inductive type \cite{kaposi2019constructing}. The signature for this
+type includes four sorts ($\Con$, $\Sub$, $\Ty$, $\Tm$), constructors for
+contexts, explicit substitutions and type/term formers, and equality
+constructors for all specified equations. The syntax comes with an
+\emph{induction principle}, from which we can also derive the \emph{recursion principle} as a
+non-dependent special case, witnessing the initiality of the syntax.
+\end{definition}
 
@@
 
-In \cite{twolevel}, ``two-level type theory'' specifically refers to the setup
-where the object-level is a homotopy type theory and the meta level is an
+In ibid., ``two-level type theory'' specifically refers to the setup
+where the object level is a homotopy type theory and the meta level is an
 extensional type theory. In contrast, we allow a wider range of setups under the
@@
 
-\subsection{Models and Syntax of the Object Theory}
+\subsubsection{Elaborating informal syntax}
+The justification for the usage of the informal syntax in Section
+\ref{sec:tour-of-2ltt} is \emph{elaboration}. Elaboration translates surface
+notation to the formal syntax (often termed ``core syntax'' in the
+implementation of dependently typed programming languages). Elaboration is
+partial: it may throw error on invalid input. Elaboration also requires
+\emph{decidability of conversion} for the formal syntax, since checking
+well-typing requires comparing types for conversion, which in turn requires
+comparing terms.  We do not detail elaboration in this paper, we only make
+some observations.
+\begin{itemize}
+\item Decidability of conversion can be shown through normalization. Since 2LTT
+      is a minor variation on ordinary MLTT, we expect that existing proofs for MLTT
+      can be adapted without issue \cite{coquand2018canonicity,decidableconv}.
+\item We need to elaborate the surface-level Russell-style universe notation to
+      Coquand-style universes with explicit $\El$ and $\Code$ annotations. It is
+      likely that this can be achieved with the use of bidirectional elaboration
+      \cite{DBLP:journals/csur/DunfieldK21}, using two directions (checking and
+      inference) both for types and terms, and inserting $\El$ and $\Code$ when
+      switching between types and terms.
+\item The formal complexity of elaboration varies wildly, depending on how
+      explicit the surface syntax is. For example, Agda-style implicit arguments
+      require metavariables and unification, which dramatically increases the
+      difficulty of formalization.
+\end{itemize}
+
+\subsubsection{The setoid interpretation of the syntax} We need to show
+that the 2LTT syntax, as a particular inductive type, can be modeled in setoids.
+Unfortunately, there is no prior work which presents the setoid interpretation
+of quotient types in the generality of quotient induction-induction. We sketch
+it for our specific use case, but we argue that the interpretation is fairly
+mechanical and could be adapted to any quotient inductive-inductive type.
+We focus on contexts and types here.
+
+We mutually inductively define $\Con : \Set$, $\Con^{\sim} : \Con \to \Con \to \Prop$, $\Ty_{i,j} : \Con \to \Set$
+      and $\Ty^{\sim} : \Ty_{i,j}\,\Gamma \to \Ty_{i,j}\,\Delta \to \Con^{\sim}\,\Gamma\,\Delta \to \Prop$, where
+      $\Con^{\sim}$ and $\Ty^{\sim}$ are proof-irrelevant relations. Note that $\Ty^{\sim}$ is \emph{heterogeneous},
+      since we can relate types in different (but provably convertible) contexts.
+
+      We specify $\emptycon$ and $\blank\ext\blank$ in $\Con$.
+To $\Con^{\sim}$ we add that $\emptycon$ and $\blank\ext\blank$ are congruences, and also that $\Con^{\sim}$ is reflexive, symmetric and transitive.
+
+We add all type formation rules to $\Ty_{i,j}$, and add all quotient equations to
+$\Ty^{\sim}$ as homogeneously indexed rules. For example, the substitution rule
+for $\Nat_0$ is specified as $ \Nat_0[] :
+\Ty^{\sim}\,(\Nat_0[\sigma])\,\Nat_0\,\refl^{\sim} $, where $\sigma :
+\Sub\,\Gamma\,\Delta$ and $\refl^{\sim} : \Con^{\sim}\,\Gamma\,\Gamma$. We also
+add all congruences and reflexivity, symmetry and transitivity to $\Ty^\sim$.
+
+We add a \emph{coercion} rule to types, as $\msf{coerce} :
+\Con^{\sim}\,\Gamma\,\Delta \to \Ty_{i,j}\,\Gamma \to \Ty_{i,j}\,\Delta$. This is probably
+familiar from the traditional specifications of type theories that use
+conversion relations. However, we additionally need a \emph{coherence rule}
+which witnesses $\Ty^{\sim}\,A\,(\msf{coerce}\,p\,A)\,p$ for each $A :
+\Ty_{i,j}\,\Gamma$ and $p : \Con^{\sim}\,\Gamma\,\Delta$. In short, coherence
+says that coercion preserves conversion.
+
+The concise description of this setup is that $(\Con,\,\Con^{\sim})$ is a
+setoid, and $(\Ty_{i,j},\,\Ty^{\sim})$ is a \emph{setoid fibration} over
+$(\Con,\,\Con^{\sim})$, for each $i$ and $j$. This notion of setoid fibration is
+equivalent to the ordinary notion of fibration, when setoids are viewed as
+categories with proof-irrelevant invertible morphisms. This is appropriate,
+since types in general are interpreted as setoid fibrations in the setoid model
+of a type theory.
+
+This scheme generalizes to substitutions and terms in a mechanical way. $\Sub$
+is indexed over two contexts, so it is interpreted as a fibration over
+the setoid product of $\Con$ and $\Con$. $\Tm$ is indexed over $(\Gamma :
+\Con)\times \Ty_{i,j}\,\Gamma$, so it is interpreted as a fibration over a setoid
+$\Sigma$-type. In every conversion relation we take the congruence closure of the specified
+quotient equations. Finally, we can use the induction principle for the thus
+specified sets and relations to interpret the internal elimination principle for
+the 2LTT syntax.
+
+%% We need to show that
+%% the 2LTT syntax, as a particular inductive type, can be modeled in setoids, if
+
+%% The examples in Section \ref{sec:tour-of-2ltt} can be faithfully represented in
+%% the formal syntax. The only notable difference is that the formal syntax uses
+%% Coquand-style universes instead of Russell-style ones. We use the former because
+%% it is much easier to model in the semantics in Section \ref{sec:staging-algorithm}. We note
+%% though that it is straightforward to elaborate from a Russell-style surface
+%% syntax to the formal version, by inserting $\El$ and $\Code$ as required.
+
+\subsection{Models and Syntax of the Object Theory}\label{sec:objmod}
 
@@
 \begin{definition}
-Like before, the \textbf{syntax of the object theory} is the initial model.
+Like before, the \textbf{syntax of the object theory} is the initial model,
+given as a quotient inductive-inductive type, and it can be also interpreted
+in setoids.
 \end{definition}
@@
 \begin{definition}
-We have an \textbf{embedding} of object syntax syntax into 2LTT syntax, consisting
-of the following functions:
+We recursively define the \textbf{embedding} of object syntax into 2LTT syntax,
+which preserves all structure strictly and consists of the following functions:
 \begin{alignat*}{4}
-  & \emb{\blank} : \Con\ob \to \Con\\
-  & \emb{\blank} : \Sub\ob\,\Gamma\,\Delta \to \Sub\,\emb{\Gamma}\,\emb{\Delta}\\
-  & \emb{\blank} : \Ty_{\mbbo j}\,\Gamma \to \Ty_{0,j}\,\emb{\Gamma}\\
-  & \emb{\blank} : \Tm_{\mbbo j}\,\Gamma\,A \to \Tm_{0,j}\,\emb{\Gamma}\,\emb{A}
-\end{alignat*}
-Embedding strictly preserves all structure.
+  & \emb{\blank} : \Con\ob \to \Con  && \emb{\blank} : \Sub\ob\,\Gamma\,\Delta \to \Sub\,\emb{\Gamma}\,\emb{\Delta}\\
+  & \emb{\blank} : \Ty_{\mbbo j}\,\Gamma \to \Ty_{0,j}\,\emb{\Gamma}\hspace{2em} && \emb{\blank} : \Tm_{\mbbo j}\,\Gamma\,A \to \Tm_{0,j}\,\emb{\Gamma}\,\emb{A}
+\end{alignat*}
 \end{definition}
@@
   \begin{alignat*}{3}
-    & \Stage : \Ty_{0,j}\,\emb{\Gamma} \to \Ty_{\mbbo j}\,\Gamma \\
-    & \Stage : \Tm_{0,j}\,\emb{\Gamma}\,A \to \Tm_{\mbbo j}\,\Gamma\,(\Stage\,A)
+    & \Stage : \Ty_{0,j}\,\emb{\Gamma} \to \Ty_{\mbbo j}\,\Gamma\hspace{2em} \Stage : \Tm_{0,j}\,\emb{\Gamma}\,A \to \Tm_{\mbbo j}\,\Gamma\,(\Stage\,A)
   \end{alignat*}
@@
 \begin{definition}
-  A staging algorithm $\Stage$ is \textbf{correct} if the following properties hold:
+  The following properties are of interest when considering a staging algorithm:
   \begin{enumerate}
   \item \emph{Soundness:} $\emb{\Stage\,A} = A$ and $\emb{\Stage\,t} = t$.
-  \item \emph{Completeness:} staging respects definitional equality.
   \item \emph{Stability:} $\Stage\,\emb{A} = A$ and $\Stage\,\emb{t} = t$.
+  \item \emph{Strictness:} the function extracted from $\Stage$ preserves all
+    type and term formers strictly (not just up to conversion).
   \end{enumerate}
 \end{definition}
-We make some remarks on correctness. First, we get completeness for free in our
-algebraic setup, since \emph{all} functions definable in the metatheory must
-respect definitional equality.
-
-Second, note that soundness and stability together is the statement that
-embedding is a bijection on types and terms. This is a statement of
+
+Soundness and stability together state that embedding is a bijection
+on types and terms (up to conversion). This is a statement of
 \emph{conservativity} of 2LTT over the object theory. In \cite{twolevel} a
-significantly weaker conservativity theorem is shown, which expresses that there
-exists a function from $\Tm_{0,j}\,\emb{\Gamma}\,\emb{A}$ to $\Tm_{\mbbo
+significantly weaker conservativity theorem is shown, which only expresses that
+there exists a function from $\Tm_{0,j}\,\emb{\Gamma}\,\emb{A}$ to $\Tm_{\mbbo
   j}\,\Gamma\,A$.
 
-Lastly, we note that the correctness terminology is borrowed from previous works
-on normalization-by-evaluation, where the soundness-completeness-stability trio
-has been used \cite{altenkirch1995categorical,altenkirch2016normalisation}. In
-\cite{cubicalnbe}, stability is instead called ``idempotence''.
+Strictness tells us that staging does not perform any computation in the
+object theory; e.g.\ it does not perform $\beta$-reduction or inline any
+$\msf{let}$-definition. This is practically quite important; for instance,
+\emph{full normalization of 2LTT} is a sound and stable staging algorithm, but
+it is not strict, and it is practically not useful, since it does not
+provide any control over inlining and $\beta$-reduction in code generation.
+
+Note that staging necessarily preserves type and term formers up to conversion,
+since it is defined on quotiented types and terms. More generally, staging
+respects conversion because every function must respect conversion in the
+meta type theory.
 
@@
 We review the presheaf model described in \cite[Section~2.5.3]{twolevel}.  This
-will directly yield a closed staging algorithm, by evaluation of 2LTT types and
+will directly yield a closed staging algorithm, by recursive evaluation of 2LTT types and
 terms in the model.
@@
 object-theoretic substitutions can be \emph{applied} to such runtime objects,
-acting on embedded object-theoretic types and terms. Stability captures the core
-idea of staging:
+acting on embedded object-theoretic types and terms. Stability underlies
+the core trade-off in staging:
 \begin{itemize}
@@
 Using the above definitions and following \cite{twolevel}, we can model all type
-formers in $\hTy_{0,j}$. Intuitively, that is because $\hTy_{0,j}$ and
+formers in $\hTy_{0,j}$. Intuitively, this is because $\hTy_{0,j}$ and
 $\hTm_{0,j}$ return types and terms, so we can reuse the type and term formers
@@
 
-\begin{definition} The \textbf{evaluation morphism, denoted $\bs{\ev}$}, is
-a 2LTT model morphism from the syntax of 2LTT to $\hato$, arising from the initiality
-of the syntax. In other words, $\ev$ is defined by recursion on the syntax and
-strictly preserves all structure.
+\begin{definition}
+The \textbf{evaluation morphism, denoted $\bs{\ev}$} is a morphism of 2LTT
+models, between the syntax of 2LTT and $\hato$. It is defined by recursion on
+the syntax (that is, appealing to the initiality of the syntax), and it strictly
+preserves all structure.
 \end{definition}
@@
 we get $|\ev\,A|\,\{\Gamma\} : |\emb{\Gamma}|\,\Gamma \to \Ty_{\mbbo j}$. We
-need an element of $|\emb{\Gamma}|\,\Gamma$, in order to extract an object-level
-type. Such ``generic'' semantic contexts should be possible to construct, because
+need an element of $|\emb{\Gamma}|\,\Gamma$, in order to obtain object-level
+type. Such ``generic'' semantic environments should be possible to construct, because
 elements of $|\emb{\Gamma}|\,\Gamma$ are essentially lists of object-level terms
@@
 
-\subsection{Stability and Open Staging}
-
-Stability means that staging an object-level type or term does nothing. We
-define a family of functions $\blank^P$ by induction on object syntax. The
+\subsection{Open Staging, Stability and Strictness}
+
+We define a family of functions $\blank^P$ by induction on object syntax, such
+that the interpretation of a context yields a generic semantic environment. The
 induction motives are as follows.
 \begin{alignat*}{4}
-  & (\Gamma : \Cono)^P                    &&: |\ev\,\emb{\Gamma}|\,\Gamma\\
-  & (\sigma : \Subo\,\Gamma\,\Delta)^P    &&: \Delta^P[\sigma] = |\ev\,\emb{\sigma}|\,\Gamma^P\\
-  & (A      : \Ty_{\mbbo j}\,\Gamma)^P      &&: A = |\ev\,\emb{A}|\,\Gamma^P\\
-  & (t      : \Tm_{\mbbo j}\,\Gamma\,A)^P   &&: t = |\ev\,\emb{t}|\,\Gamma^P
+  & (\Gamma : \Cono)^P                    &&: |\ev\,\emb{\Gamma}|\,\Gamma && (\sigma : \Subo\,\Gamma\,\Delta)^P    &&: \Delta^P[\sigma] = |\ev\,\emb{\sigma}|\,\Gamma^P \\
+  & (A      : \Ty_{\mbbo j}\,\Gamma)^P      &&: A = |\ev\,\emb{A}|\,\Gamma^P\hspace{2em} && (t      : \Tm_{\mbbo j}\,\Gamma\,A)^P   &&: t = |\ev\,\emb{t}|\,\Gamma^P
 \end{alignat*}
@@
 
+\begin{theorem} Open staging is strict.
+\end{theorem}
+\begin{proof}
+  This is evident from the definition of the presheaf model, where the action of
+  each type and term former in $\wh{\Ty}_{0,j}$ and $\wh{\Tm}_{0,j}\,\Gamma\,A$ is
+  defined using the corresponding type or term former in the object syntax.
+\end{proof}
+
 \subsection{Implementation and Efficiency}
 
-We have defined staging in a constructive metatheory, so it is possible to
-extract an implementation from it. We discuss its operational behavior and
-potential optimizations in the following. First, consider the computational
-part of evaluation:
+We discuss now the operational behavior of the extracted staging algorithm
+and look at potential adjustments and optimizations that make it more efficient
+or more convenient to implement. Recall the types of the component functions in $\ev$:
 \begin{alignat*}{3}
@@
 \end{alignat*}
-Operationally, we interpret syntactic types or terms in a semantic environment
+We interpret syntactic types or terms in a semantic environment
 $\gamma : |\ev\,\Gamma|\,\Delta$. These environments are lists containing
-object-level terms and semantic values. The semantic values are represented
+a mix of  object-level terms and semantic values. The semantic values are represented
 using metatheoretic inductive types and function types.
@@
         semantics of intuitionistic logics. Whenever we evaluate a function
-        application, we supply an extra $\id : \Subo\,\Delta\,\Delta$. This may
-        incur cost via the $\gamma[\id]$ restrictions in a naive
+        application, we supply an extra $\id : \Subo\,\Delta\,\Delta$.
+        This may incur cost via the $\gamma[\id]$ restrictions in a naive
         implementation, but this is easy to optimize, by introducing a formal
@@
 
-In contrast, we have a bit of an efficiency problem in the handling of object-level
-terms and binders: whenever we go under an object-level binder we have to weaken
-the current semantic environment. In other words, when moving from $\Delta$ to
+In contrast, we have a bit of an efficiency problem in the handling of
+object-level binders: whenever we go under such binder we have to weaken the
+current semantic environment. Concretely, when moving from $\Delta$ to
 $\Delta \ext A$, we have to shift $\gamma : |\ev\,\Gamma|\,\Delta$ to
 $\gamma[\p] : |\ev\,\Gamma|\,(\Delta \ext A)$. This cannot be avoided by an easy
-optimization trick.
-
-The exact same efficiency issue occurs in formal presheaf-based
+optimization trick. For object-level entries in the environment, this is cheap,
+because we just add an extra explicit weakening, but for semantic values
+weakening may need to perform deep traversals.
+
+The same efficiency issue arises in formal presheaf-based
 normalization-by-evaluation, such as in \cite{kaposinbe} and
@@
 \cite{coquand1996algorithm}, \cite{modulartc} and \cite{untypedtc} for
-implementations in this style. We can use the same solution in staging, by using
-de Bruijn levels and delayed weakenings in object-level terms. It is a somewhat
-unfortunate mismatch that indices are far more convenient in formalization, but
-levels are more efficient in practice.
+implementations in this style. We can use the same solution in staging. It is a
+somewhat unfortunate mismatch that indices are far more convenient in
+formalization, but levels are more efficient in practice.
 
 In our prototype implementation, we use the above optimization with De Bruijn
-levels, and we also switch to \emph{syntax-directed} staging. In the formal
-staging algorithm, since types are tracked, quotes and splices have no
-operational significance; recall that they are interpreted as identity functions
-in Definition \ref{def:psh-lifting}. However, it seems more efficient to not compute types
-at all and instead rely on quotes and splices to keep track of the current
-stage.
+levels, and we drop explicit substitutions from the 2LTT core syntax. In the
+object-level syntax, we use De Bruijn levels for variables and closures in
+binders. This lets us completely skip the weakening of environments, which is
+slightly more efficient than the more faithfully extracted algorithm.
+
+Additionally, we use an untyped tagged representation of semantic
+values, since in Haskell (the implementation language) we do not have
+strong enough types to represent typed semantic values. For example, there
+are distinct data constructors storing semantic function values, natural number
+literals and quoted expressions.
 
 \subsubsection{Caching} In a production-strength implementation we would need some
-caching mechanism, to avoid excessive duplication of function code. For example,
-if we use the staged $\mit{map}$ function multiple times with the same type and
+caching mechanism, to avoid excessive duplication of code. For example, if we
+use the staged $\mit{map}$ function multiple times with the same type and
 function arguments, we do not want to generate separate code for each
-usage. De-duplicating object-level code with function types is
-usually safe, since function bodies become closed top-level code after closure
-conversion. We leave this to future work.
+usage. De-duplicating object-level code with function types is usually safe,
+since function bodies become closed top-level code after closure conversion. We
+leave this to future work.
 
@@
 $\hato$. Additionally, $\hato$ supports all type formers of extensional type
-theory, and certain other structures which are stable under object-level
+theory and certain other structures which are stable under object-level
 substitution. As all constructions in this section must be stable, it makes
@@
 \begin{definition} First, we introduce shorthands for
-external operations that can be extracted from $\blank\rel$.
+external operations that can be obtained from $\blank\rel$.
 \begin{alignat*}{5}
@@
 \begin{itemize}
-\item Using \emph{synthetic Tait computability} \cite{sterlingthesis}, we work in the internal language
-      of presheaves over the glued category $\msf{Id}_{\hato}\downarrow \re$.
+\item Using \emph{synthetic Tait computability} \cite{sterlingthesis}, we work
+      in the internal language of the glued category $\msf{Id}_{\hato}\downarrow
+      \re$.
 \item Using \emph{relative induction} \cite{bocquet2021relative}, we work in the modal type theory
@@
 
-C-style \emph{function pointers} could be also an interesting use case, since
-these must become closed after staging (as they cannot capture closure
-environments).
+\emph{Function pointers} in the style of the C programming language could be also
+an interesting use case, since these must become closed after staging (as they
+cannot capture closure environments).
 
@@
 judgment which ensures that a purely ``object-level'' program can be extracted
-from a multi-stage program.
+from a multi-stage program. Similar considerations apply to systems with the S4
+modality (see e.g.\ Davies and Pfenning \cite{DBLP:journals/jacm/DaviesP01}),
+where there is no judgment that guarantees elimination of modal operations.
 

### 2022-06-17 paper/paper.tex
revision, add appendix

@@
 
-\documentclass[acmsmall,anonymous,review]{acmart}
+\documentclass[acmsmall,review]{acmart}
 %% \documentclass[nonacm,acmsmall,review]{acmart}
@@
 \begin{abstract}
-  The aim of staged compilation is to enable metaprogramming in a way such that
-  we have guarantees about the well-formedness of code output, and we can also
-  mix together object-level and meta-level code in a concise and convenient
-  manner. In this work, we observe that two-level type theory (2LTT), a system
-  originally devised for the purpose of developing synthetic homotopy theory,
-  also serves as a system for staged compilation with dependent types. 2LTT has
-  numerous good properties for this use case: it has a concise specification,
-  well-behaved model theory, and it supports a wide range of language features
-  both at the object and the meta level. First, we give an overview of 2LTT's
-  features and applications in staging. Then, we present a staging algorithm and
-  prove its soundness and stability. Our algorithm is ``staging-by-evaluation'',
-  analogously to the technique of normalization-by-evaluation, in that staging
-  is given by the evaluation of 2LTT syntax in a semantic domain. A sound and
-  stable staging algorithm constitutes a proof of strong conservativity of 2LLT
-  over the object theory. To our knowledge, this is the first description of
-  staged compilation which supports full dependent types and unrestricted
-  staging for types.
+The aim of staged compilation is to enable metaprogramming in a way such that we
+have guarantees about the well-formedness of code output, and we can also mix
+together object-level and meta-level code in a concise and convenient manner. In
+this work, we observe that two-level type theory (2LTT), a system originally
+devised for the purpose of developing synthetic homotopy theory, also serves as
+a system for staged compilation with dependent types. 2LTT has numerous good
+properties for this use case: it has a concise specification, well-behaved model
+theory, and it supports a wide range of language features both at the object and
+the meta level. First, we give an overview of 2LTT's features and applications
+in staging. Then, we present a staging algorithm and prove its correctness. Our
+algorithm is ``staging-by-evaluation'', analogously to the technique of
+normalization-by-evaluation, in that staging is given by the evaluation of 2LTT
+syntax in a semantic domain. The staging algorithm together with its correctness
+constitutes a proof of strong conservativity of 2LLT over the object theory. To our
+knowledge, this is the first description of staged compilation which supports
+full dependent types and unrestricted staging for types.
 \end{abstract}
@@
 
-\subsection{Properties of Lifting \& Binding Time Improvements}
+\subsection{Properties of Lifting, Binding Time Improvements}
 
@@
 they are pointwise equivalent. A quotient type $A/R$ is interpreted by extending
-the semantic notion of equivalence for $A$ with the given $R$ relation.
+the semantic notion of equivalence for $A$ with the given $R$ relation. We will
+sketch the interpretation of the 2LTT syntax (given as a quotient inductive-inductive type)
+in Section \ref{sec:2lttsetoid}.
 
@@
 In other words, we can view the usage of constructive quotients as a convenient
-shorthand for working with explicit equivalence relations (an \emph{internal
-language} of setoids). When we reason about the syntax of a type theory, this
-provides a bridge between the categorical/algebraic style and the traditional
-style that uses separate equivalence relations. It makes sense to work with
-quotients internally to a type theory whenever we only want to talk about
-constructions and properties which respect conversion. However, at any point we
-have the option to switch to an external perspective and look at underlying sets
-and the underlying maps.
+shorthand for working with explicit equivalence relations. When we reason about
+the syntax of a type theory, this provides a bridge between the
+categorical/algebraic style and the traditional style that uses separate
+equivalence relations. It makes sense to work with quotients internally to a
+type theory whenever we only want to talk about constructions and properties
+which respect conversion. However, at any point we have the option to switch to
+an external perspective and look at underlying sets and the underlying maps.
 
@@
 
-First, we define the
-structural scaffolding of 2LTT without type formers, mostly following
-\cite{twolevel}. The specification is agnostic about the sizes of sets involved,
-i.e.\ we can model underlying sorts using any metatheoretic $\Set_j$.
+We provide an appendix to this paper, where we present a listing of all rules
+and equations of 2LTT in two styles: first, in a more traditional style with
+derivation rules, and also in a very compact manner using a second-order
+algebraic signature.
+
+First, we define the structural scaffolding of 2LTT without type formers, mostly
+following \cite{twolevel}. The specification is agnostic about the sizes of sets
+involved, i.e.\ we can model underlying sorts using any metatheoretic $\Set_j$.
 
@@
 \item
-  By moving left-to-right along $\Sub\,\Gamma\,(\Delta\ext A) \simeq ((\sigma :
-  \Sub\,\Gamma\,\Delta) \times \Tm_{i,j}\,\Gamma\,(A[\sigma]))$,
+  By moving left-to-right along $\Sub\,\Gamma\,(\Delta\ext A) \simeq (\sigma :
+  \Sub\,\Gamma\,\Delta) \times \Tm_{i,j}\,\Gamma\,(A[\sigma])$,
   and starting from the identity morphism $\id : \Sub\,(\Gamma\ext A)\,(\Gamma\ext A)$, we recover
@@
   indices. In general, the $n$-th De Bruijn index is defined as $\q[\p^{n}]$,
-  where $\p^{n}$ denotes $n$-fold composition.
+  where $\p^{n}$ denotes $n$-fold composition of $p$.
 \item
-  By moving right-to-left along $\Sub\,\Gamma\,(\Delta\ext A) \simeq ((\sigma :
-  \Sub\,\Gamma\,\Delta) \times \Tm_{i,j}\,\Gamma\,(A[\sigma]))$, we recover the
+  By moving right-to-left along $\Sub\,\Gamma\,(\Delta\ext A) \simeq (\sigma :
+  \Sub\,\Gamma\,\Delta) \times \Tm_{i,j}\,\Gamma\,(A[\sigma])$, we recover the
   operation which extends a morphism with a term. In the initial model, this
@@
   of substitutions as lists of terms. We denote the extension operation as
-  $(\sigma,\,t)$ for $\sigma : \Sub\,\Gamma\,\Delta$.
+  $(\sigma,\,t)$ for $\sigma : \Sub\,\Gamma\,\Delta$ and $t : \Tm_{i,j}\,\Gamma\,(A[\sigma])$.
 \end{itemize}
@@
 for binders and substitutions. For example, we may write $\Gamma \ext (x : A)
-\ext (y : B)$ for a context, and subsequently write $B[y \mapsto t]$ for
+\ext (y : B)$ for a context, and subsequently write $B[x \mapsto t]$ for
 substituting the $x$ variable for some term $t : \Tm_{i,j}\,\Gamma\,A$. Using
 nameless notation, we would instead have $B : \Ty_{i,j}\,(\Gamma \ext A)$ and
-$B[\id,\,t]$; here we recover single substitution by extending the identity
+$B[\id,\,t] : \Ty_{i,j}\,\Gamma$; here we recover single substitution by extending the identity
 substitution $\id : \Sub\,\Gamma\,\Gamma$ with $t$.
 
-We may also use implicit weakening: if a type or term is in a $\Gamma$ context,
-we may use it in an extended $\Gamma \ext A$ context without marking the
-weakening substitution.
+We may also leave weakening implicit: if a type or term is in a context
+$\Gamma$, we may use it in an extended context $\Gamma \ext A$ without marking
+the weakening substitution.
 \end{notation}
@@
 \item \emph{$\Sigma$-types}. We have $\Sigma\,(x : A)\,B : \Ty_{i,j}\,\Gamma$ for $A : \Ty_{i,j}\,\Gamma$
-  and $B : \Ty_{i,j}\,(\Gamma \ext (x : A))$, together with a natural isomorphism of pairing and projections:
-  \[ \Tm_{i,j}\,\Gamma\,(\Sigma\,(x : A)\,B) \simeq ((t : \Tm_{i,j}\,\Gamma\,A) \times \Tm_{i,j}\,\Gamma\,(B[x\mapsto t]) \]
+  and $B : \Ty_{i,j}\,(\Gamma \ext (x : A))$, together with a natural isomorphism consisting of pairing and projections:
+  \[ \Tm_{i,j}\,\Gamma\,(\Sigma\,(x : A)\,B) \simeq (t : \Tm_{i,j}\,\Gamma\,A) \times \Tm_{i,j}\,\Gamma\,(B[x\mapsto t]) \]
   We write $(t,\,u)$ for pairing and $\fst$ and $\snd$ for projections.
@@
   \end{alignat*}
-  such that $\lam$ and $\app$ are inverses. This specification of $\app$ is
-  equivalent to the ``traditional'' specification:
+  such that $\lam$ and $\app$ are inverses. Since we have explicit
+  substitutions, this specification of $\app$ is equivalent to the
+  ``traditional'' one:
   \[ \app' : (t : \Tm_{i,j}\,\Gamma\,(\Pi\,(x :
@@
     & \NatElim :\hspace{0.5em}
-                &&  (P &&: \Ty_{i,k}\,(\Gamma \ext (n : \Nat_{i,j}) \\
+                &&  (P &&: \Ty_{i,k}\,(\Gamma \ext (n : \Nat_{i,j}))) \\
     &           &&  (z &&: \Tm_{i,k}\,\Gamma\,(P[n\mapsto \zero_{i,j}])) \\
@@
     &           &&  (t &&: \Tm_{i,j}\,\Gamma\,\Nat_{i,j}) \\
-    &\hspace{3em}\to        &&&& \Tm_{i,j}\,\Gamma\,(P[n\mapsto t]))
+    &\hspace{3em}\to        &&&& \Tm_{i,k}\,\Gamma\,(P[n\mapsto t]))
   \end{alignat*}
@@
 
-\subsubsection{The setoid interpretation of the syntax} We need to show
+\subsubsection{The setoid interpretation of the syntax}\label{sec:2lttsetoid} We need to show
 that the 2LTT syntax, as a particular inductive type, can be modeled in setoids.

### 2022-07-20 paper/paper.tex
before compressions needed bc of citation style

@@
 
-\documentclass[acmsmall,review]{acmart}
+\documentclass[acmsmall,screen]{acmart}
 %% \documentclass[nonacm,acmsmall,review]{acmart}
@@
 %% the next command will enable that style.
-%%\citestyle{acmauthoryear}
+\citestyle{acmauthoryear}
 
@@
 
-\renewcommand{\U}{\msf{U}}
+\newcommand{\U}{\msf{U}}
 \newcommand{\Con}{\msf{Con}}
@@
     a staging algorithm for closed types and terms. We then extend staging to
-    open types and terms, which may depend on object-level contexts. We show
+    open types and terms which may depend on object-level variables. We show
     stability of staging, which means that staging is surjective up to
@@
   \item
-    In Section \ref{sec:soundness} we show soundness of staging, which roughly means that
-    the output of staging is convertible to its input. Staging together with
-    its stability and soundness can be viewed as a \emph{strong conservativity}
-    theorem of 2LTT over the object theory. This means that the possible
-    object-level constructions in 2LTT are in bijection with the constructions
-    in the object theory, and staging witnesses that meta-level constructions
-    can be always computed away. This strengthens the weak notion of
-    conservativity shown in \cite{capriotti2017models} and \cite{twolevel}.
+    In Section \ref{sec:soundness} we show soundness of staging, which roughly
+    means that the output of staging is convertible to its input. Staging
+    together with its stability and soundness can be viewed as a \emph{strong
+    conservativity} theorem of 2LTT over the object theory. This means that the
+    same constructions can be expressed in the object theory and the
+    object-level fragment of 2LTT, up to conversion, and staging witnesses that
+    meta-level constructions can be always computed away. This strengthens the
+    weak notion of conservativity shown in \cite{capriotti2017models} and
+    \cite{twolevel}.
   \item
@@
   \item We provide a small standalone implementation of elaboration and staging
-    for a two-level type theory, with code samples that expand on the paper.
+    for a two-level type theory, with code samples that expand on the paper
+    \cite{staged-demo}.
   \item To our knowledge, this work is the first formalization and implementation of
@@
   & &&\spl(\mit{map}\,\qut{\Nat_0}\,\qut{\Nat_0}\,(\lambda\,n.\,\qut{\spl n +_0 10})\,\qut{\mit{ns}})\\
-  & =\hspace{0.5em}&&\spl\qut{\msf{foldr}_0\,
+  & =\hspace{0.3em}&&\spl\qut{\msf{foldr}_0\,
         \spl \qut{\Nat_0}\,\spl \qut{\List_0\,\spl\qut{Nat_0}}\,
@@
         \spl\qut{\mit{ns}}}\\
-  & =\hspace{0.5em} &&\msf{foldr_0}\,\Nat_0\,(\List_0\,\Nat_0)\,\,(\lambda\,a\,bs.\,\cons_0\,(a +_0 10)\,bs)\,(\nil_0\,\Nat_0)\,\mit{ns}
+  & =\hspace{0.3em} &&\msf{foldr_0}\,\Nat_0\,(\List_0\,\Nat_0)\,\,(\lambda\,a\,bs.\,\cons_0\,(a +_0 10)\,bs)\,(\nil_0\,\Nat_0)\,\mit{ns}
 \end{alignat*}
@@
   &\hspace{-3em}\rlap{$\mit{map} : (n : \Nat_1) \to (\Lift\,\spl A \to \Lift\,\spl B) \to \Lift(\mit{Vec}\,n\,A) \to \Lift(\mit{Vec}\,n\,B)$}\\
-  &\hspace{-3em}\mit{map}\,&&\zero_1      && f\,&&\mit{as} := \qut{\tt_0}\\
-  &\hspace{-3em}\mit{map}\,&&(\suc_1\,n)\,&& f\,&&\mit{as} :=
-     \qut{(\spl(f\,\qut{\fst_0\,\spl\mit{as}}),\,\mit{map}\,n\,f\,\qut{\snd_0\,\spl \mit{as}})}
+  &\hspace{-3em}\mit{map}\,\,&&\zero_1      && f\,&&\mit{as} := \qut{\tt_0}\\
+  &\hspace{-3em}\mit{map}\,\,&&(\suc_1\,n)\,&& f\,&&\mit{as} :=
+     \qut{(\spl(f\,\qut{\fst_0\,\spl\mit{as}}),\,\spl(\mit{map}\,n\,f\,\qut{\snd_0\,\spl \mit{as}}))}
 \end{alignat*}
@@
 We include examples of more sophisticated let-insertion in the supplementary
-code.
+code \cite{staged-demo}.
 
@@
 meta-level $\lambda$, while going the other way introduces a runtime
-binder. Meta-level function types and $\Sigma$ types support more computation during
+binder. Meta-level function types and $\Sigma$-types support more computation during
 staging, so in many cases it is beneficial to use the improved forms. In some
@@
 statically. This makes it easy to assign different memory layouts to different
-types, and generate code accordingly for each type. Moving to 2LTT, we still
+types and generate code accordingly for each type. Moving to 2LTT, we still
 want to abstract over runtime types at compile time, so we use the following
@@
 Despite the restriction to simple types at runtime, we can still write arbitrary
-higher-rank polymorphic functions, such as a function with type $((A : \Ty_0)
+higher-rank polymorphic functions in 2LTT, such as a function with type $((A : \Ty_0)
 \to \Lift\,\spl A \to \Lift \spl A) \to \Lift\,\Bool_0$. This function can be
@@
 
-This is fairly similar to the well-known monomorphization models in the
-C++ and Rust programming languages. The evident difference is that in 2LTT we
-can use a full type theory at compile time. Additionally, we will see in Section
-\ref{sec:intensional-analysis} that this setup is compatible with an induction principle for
-$\Ty_0$, so that we can analyze the structure of runtime types.
+This is fairly similar to the well-known monomorphization models in the C++ and
+Rust programming languages. The evident difference is that in 2LTT we can use a
+full type theory at compile time. Additionally, we will see in Section
+\ref{sec:intensional-analysis} that this setup is compatible with intensional
+analysis, the ability to analyze the structure of object types and terms.
 
@@
   element of $\Sub$ is a \emph{parallel substitution}, $\emptycon$ is the
-  \emph{empty typing context}, and $\epsilon$ is the empty list viewed as a
+  \emph{empty typing context} and $\epsilon$ is the empty list viewed as a
   parallel substitution. The identity substitution $\id : \Sub\,\Gamma\,\Gamma$
@@
 strengthenings and extensions of 2LTT \cite[Section~2.4]{twolevel}, most of which
-are useful in synthetic homotopy theory. We do not assume any of these, and
+are useful in synthetic homotopy theory. We do not assume any of these and
 stick to the most basic formulation of 2LTT.
@@
 partial: it may throw error on invalid input. Elaboration also requires
-\emph{decidability of conversion} for the formal syntax, since checking
-well-typing requires comparing types for conversion, which in turn requires
-comparing terms.  We do not detail elaboration in this paper, we only make
-some observations.
+\emph{decidability of conversion} for the formal syntax, since type checking
+requires comparing types for conversion, which in turn requires comparing terms.
+We do not detail elaboration in this paper, we only make some observations.
 \begin{itemize}
@@
 
-Strictness tells us that staging does not perform any computation in the
-object theory; e.g.\ it does not perform $\beta$-reduction or inline any
-$\msf{let}$-definition. This is practically quite important; for instance,
-\emph{full normalization of 2LTT} is a sound and stable staging algorithm, but
-it is not strict, and it is practically not useful, since it does not
-provide any control over inlining and $\beta$-reduction in code generation.
+Strictness tells us that staging does not perform any computation in the object
+theory; e.g.\ it does not perform $\beta$-reduction. This is practically quite
+important; for instance, \emph{full normalization of 2LTT} is a sound and stable
+staging algorithm, but it is not strict, and it is practically not useful, since
+it does not provide any control over inlining and $\beta$-reduction in code
+generation.
 
@@
 We move on to modeling the object-level syntactic fragment of 2LTT. We make some
-preliminary definitions. First, note the types in the object theory yield a
+preliminary definitions. First, note that types in the object theory yield a
 presheaf, and terms yield a displayed presheaf over them; this immediately
@@
 \end{itemize}
-
 Thus, a generic semantic context $\Gamma^P : |\ev\emb{\Gamma}|\,\Gamma$ is just
@@
 
-However, this is still sufficient for many practical use cases, for example the
-monomorphization setup in Section \ref{sec:monomorphization}. Here, it makes sense to only
-have weakening in the object theory, but no $\beta\eta$-rules. We care about
-the intension of the staging output; presumably we are doing monomorphization
-because we care about code performance, so in 2LTT we do not want to
-definitionally equate object-level programs with different performance
-characteristics.
+However, stability under weakening is still sufficient for many practical use
+cases, for example the monomorphization setup in Section
+\ref{sec:monomorphization}. Here, we only have weakening as definitional
+equality in the object theory, and we do not have $\beta\eta$-rules.
 

### 2022-07-20 paper/paper.tex
camera ready edit

@@
 
-%% Rights management information.  This information is sent to you
-%% when you complete the rights form.  These commands have SAMPLE
-%% values in them; it is your responsibility as an author to replace
-%% the commands and values with those provided to you when you
-%% complete the rights form.
-
-%% \setcopyright{acmcopyright}
-%% \copyrightyear{2018}
-%% \acmYear{2018}
-%% \acmDOI{XXXXXXX.XXXXXXX}
+%%% The following is specific to ICFP '22 and the paper
+%%% 'Staged Compilation with Two-Level Type Theory'
+%%% by András Kovács.
+%%%
+\setcopyright{rightsretained}
+\acmPrice{}
+\acmDOI{10.1145/3547641}
+\acmYear{2022}
+\copyrightyear{2022}
+\acmSubmissionID{icfp22main-p52-p}
+\acmJournal{PACMPL}
+\acmVolume{6}
+\acmNumber{ICFP}
+\acmArticle{110}
+\acmMonth{8}
 
@@
 
+\bibliographystyle{ACM-Reference-Format}
 %%
@@
 \title{Staged Compilation With Two-Level Type Theory}
+\titlenote{The author was supported by the “Application Domain Specific Highly
+  Reliable IT Solutions” project which has been implemented with support from
+  the National Research, Development and Innovation Fund of Hungary, financed
+  under the Thematic Excellence Programme TKP2020-NKA-06 (National Challenges
+  Subprogramme) funding scheme.}
 
@@
 \begin{CCSXML}
+<ccs2012>
+   <concept>
+       <concept_id>10003752.10003790.10011740</concept_id>
+       <concept_desc>Theory of computation~Type theory</concept_desc>
+       <concept_significance>500</concept_significance>
+       </concept>
+   <concept>
+       <concept_id>10011007.10011006.10011041.10011047</concept_id>
+       <concept_desc>Software and its engineering~Source code generation</concept_desc>
+       <concept_significance>500</concept_significance>
+       </concept>
+ </ccs2012>
 \end{CCSXML}
 
+\ccsdesc[500]{Theory of computation~Type theory}
+\ccsdesc[500]{Software and its engineering~Source code generation}
+
 \keywords{type theory, two-level type theory, staged compilation}
+
 \maketitle
@@
 
-\emph{Two-level type theory}, or 2LTT in short, was described by Annekov,
-Capriotti, Kraus and Sattler \cite{twolevel}, building on ideas from Vladimir
-Voevodsky \cite{hts}. The motivation was to allow convenient metatheoretical
-reasoning about a certain mathematical language (homotopy type theory), and to
-enable concise and modular ways to extend the language with axioms.
+\emph{Two-level type theory}, or 2LTT in short, was described in
+\cite{twolevel}, building on ideas from \cite{hts}. The motivation was to allow
+convenient metatheoretical reasoning about a certain mathematical language
+(homotopy type theory), and to enable concise and modular ways to extend the
+language with axioms.
 
@@
 between the stages. We enable such interaction through the following operations.
+Note that none of these operations can be expressed as functions, since function
+types cannot cross between stages.
 \begin{itemize}
@@
 \end{itemize}
-Note that none of these three operations can be expressed as functions, since
-function types cannot cross between stages.
 
@@
 
-This is fairly similar to the well-known monomorphization models in the C++ and
-Rust programming languages. The evident difference is that in 2LTT we can use a
-full type theory at compile time. Additionally, we will see in Section
-\ref{sec:intensional-analysis} that this setup is compatible with intensional
-analysis, the ability to analyze the structure of object types and terms.
+%% This is fairly similar to the well-known monomorphization models in the C++ and
+%% Rust programming languages. The evident difference is that in 2LTT we can use a
+%% full type theory at compile time. Additionally, we will see in Section
+%% \ref{sec:intensional-analysis} that this setup is compatible with intensional
+%% analysis, the ability to analyze the structure of object types and terms.
 
@@
 1$. The transfinite $i$ levels are a convenience feature for modeling typing
-contexts, where sometimes we will need universes large enough to fit
-types at arbitrary finite levels. We also assume uniqueness of identity proofs
-and function extensionality. We will omit transports along identity proofs in
-the paper, for the sake of readability. The missing transports can be in
-principle recovered due to Hofmann's conservativity result
-\cite{hofmann95extensional}; see also Winterhalter et
-al. \cite{DBLP:conf/cpp/WinterhalterST19}. We assume $\Pi$, $\Sigma$, $\top$ and
-$\mbb{N}$ types in all universes, and reuse Notation \ref{notation:basics}. We
-also assume certain quotient inductive-inductive types, to represent the syntax of 2LTT
-and the object theory. We will detail these shortly in Sections \ref{sec:2lttmod}
-and \ref{sec:objmod}.
+contexts, where sometimes we will need universes large enough to fit types at
+arbitrary finite levels. We also assume uniqueness of identity proofs and
+function extensionality. We will omit transports along identity proofs in the
+paper, for the sake of readability. The missing transports can be in principle
+recovered due to Hofmann's conservativity result \cite{hofmann95extensional};
+see also \cite{DBLP:conf/cpp/WinterhalterST19}. We assume $\Pi$, $\Sigma$,
+$\top$ and $\mbb{N}$ types in all universes, and reuse Notation
+\ref{notation:basics}. We also assume certain quotient inductive-inductive
+types, to represent the syntax of 2LTT and the object theory. We will detail
+these shortly in Sections \ref{sec:2lttmod} and \ref{sec:objmod}.
 
@@
 Fortunately, our specified meta type theory can be interpreted in
-\emph{setoids}, as most recently explained by Pujet and Tabareau
+\emph{setoids}, as most recently explained in
 \cite{DBLP:journals/pacmpl/PujetT22}. A setoid is a set together with an
-equivalence relation on it. We interpret closed types as setoids, and
-dependent types as setoid fibrations. For inductive base types such $\Bool$ or
-$\Nat$, the equivalence relation is interpreted as identity. Functions are
-interpreted as equivalence-preserving functions, and two functions are equivalent if
-they are pointwise equivalent. A quotient type $A/R$ is interpreted by extending
-the semantic notion of equivalence for $A$ with the given $R$ relation. We will
-sketch the interpretation of the 2LTT syntax (given as a quotient inductive-inductive type)
-in Section \ref{sec:2lttsetoid}.
-
-\begin{definition} Assume a closed function $f : A \to B$ defined
-in the mentioned type theory. The \textbf{extraction} of $f$ is
-the underlying function of $f$'s interpretation in the setoid model.
-\end{definition}
+equivalence relation on it. We interpret closed types as setoids, and dependent
+types as setoid fibrations. For inductive base types such $\Bool$ or $\Nat$, the
+equivalence relation is interpreted as identity. Functions are interpreted as
+equivalence-preserving functions, and two functions are equivalent if they are
+pointwise equivalent. A quotient type $A/R$ is interpreted by extending the
+semantic notion of equivalence for $A$ with the given $R$ relation. We will
+sketch the interpretation of the 2LTT syntax (given as a quotient
+inductive-inductive type) in Section \ref{sec:2lttsetoid}.
 
@@
 
-In other words, we can view the usage of constructive quotients as a convenient
-shorthand for working with explicit equivalence relations. When we reason about
-the syntax of a type theory, this provides a bridge between the
-categorical/algebraic style and the traditional style that uses separate
-equivalence relations. It makes sense to work with quotients internally to a
-type theory whenever we only want to talk about constructions and properties
-which respect conversion. However, at any point we have the option to switch to
-an external perspective and look at underlying sets and the underlying maps.
+\begin{definition} Assume a closed function $f : A \to B$ defined
+in the mentioned type theory. The \textbf{extraction} of $f$ is
+the underlying function of $f$'s interpretation in the setoid model.
+\end{definition}
+
+In other words, we can view quotients as a convenient shorthand for working with
+explicit equivalence relations. We can work with quotients when we want to enforce
+preservation of equivalence relations, but at any point we have the option to
+switch to an external view and look at underlying sets and functions.
+
+%% In other words, we can view the usage of constructive quotients as a convenient
+%% shorthand for working with explicit equivalence relations. When we reason about
+%% the syntax of a type theory, this provides a bridge between the
+%% categorical/algebraic style and the traditional style that uses separate
+%% equivalence relations.
+%% It makes sense to work with quotients internally to a type theory whenever we
+%% only want to talk about constructions and properties which respect
+%% conversion. However, at any point we have the option to switch to an external
+%% perspective and look at underlying sets and the underlying maps.
 
@@
 internalization. Below, note that the input to $\ev$ is \emph{external}, so we
-mark $\ev$ as being parametrized by external input.
+mark $\ev$ as being parameterized by external input.
 \begin{alignat*}{3}
@@
 analogously as we have seen for $\Ty_{\mbbo j}$.
+
+%% \begin{alignat*}{6}
+%%   &\ev\,(\Gamma &&: \Con)                      &&: \whset_\omega && \ev\,(t      &&: \Tm_{1,j}\,\Gamma)          &&: (\gamma : \ev\,\Gamma) \to \ev\,A\,\gamma\\
+%%   &\ev\,(\sigma &&: \Sub\,\Gamma\,\Delta)      &&: \ev\,\Gamma \to \ev\,\Delta \hspace{3.5em}&& \ev\,(A      &&: \Ty_{0,j}\,\Gamma)          &&: \ev\,\Gamma \to \Ty_{\mbbo_j}\\
+%%   &\ev\,(A      &&: \Ty_{1,j}\,\Gamma)          &&: \ev\,\Gamma \to \whset_j&& \ev\,(t      &&: \Tm_{0,j}\,\Gamma\,A)       &&: (\gamma : \ev\,\Gamma) \to \Tm_{\mbbo j}\,(\ev\,A\,\gamma)
+%% \end{alignat*}
+
+%% \begin{alignat*}{3}
+%%   &\ev\,(\Gamma &&: \Con)                      &&: \whset_\omega \\
+%%   &\ev\,(\sigma &&: \Sub\,\Gamma\,\Delta)      &&: \ev\,\Gamma \to \ev\,\Delta \\
+%%   &\ev\,(A      &&: \Ty_{1,j}\,\Gamma)          &&: \ev\,\Gamma \to \whset_j\\
+%%   &\ev\,(t      &&: \Tm_{1,j}\,\Gamma)          &&: (\gamma : \ev\,\Gamma) \to \ev\,A\,\gamma\\
+%%   &\ev\,(A      &&: \Ty_{0,j}\,\Gamma)          &&: \ev\,\Gamma \to \Ty_{\mbbo_j}\\
+%%   &\ev\,(t      &&: \Tm_{0,j}\,\Gamma\,A)       &&: (\gamma : \ev\,\Gamma) \to \Tm_{\mbbo j}\,(\ev\,A\,\gamma)
+%% \end{alignat*}
 
@@
 \begin{alignat*}{4}
-  & (\sigma,\,t)\rel\,\gamma\rel                 && := (\sigma\rel\,\gamma\rel,\,t\rel\,\gamma\rel)\\
-  & \p\rel\,(\gamma\rel,\,t\rel)                 && := \gamma\rel\\
-  & \q\rel\,(\gamma\rel,\,t\rel)                 && := t\rel
+  & (\sigma,\,t)\rel\,\gamma\rel := (\sigma\rel\,\gamma\rel,\,t\rel\,\gamma\rel) \hspace{2em} \p\rel\,(\gamma\rel,\,t\rel) := \gamma\rel \hspace{2em} \q\rel\,(\gamma\rel,\,t\rel) := t\rel
 \end{alignat*}
@@
 The work Annekov et al.\ on two-level type theory \cite{twolevel}, building on
-the thesis of Capriotti \cite{capriotti2017models} and ideas from Voevodsky
-\cite{hts}, is the foremost inspiration in the current work. Interestingly, the
-above works do mention metaprogramming as a source of intuition for 2LTT, but
-they only briefly touch on this aspect, and in \cite{twolevel} the main focus is
-on extensions of basic 2LTT which do not have staging semantics
-anymore. Ibid.\ conjectures the strong conservativity of 2LTT in Proposition
-2.18, and suggests categorical gluing as proof strategy. In this work, we do
-essentially that: our presheaf model and the logical relation model could be
-merged into a single instance of gluing along the $\re$ morphism; this merges
-staging and its soundness proof. We split this construction to two parts in order
-to get a self-contained presentation of the staging algorithm.
-
-The multi-stage calculus of Kawata and Igarashi \cite{multi-stage-calculus}
+\cite{capriotti2017models} and \cite{hts}, is the foremost inspiration in
+the current work. Interestingly, the above works do mention metaprogramming as a
+source of intuition for 2LTT, but they only briefly touch on this aspect, and in
+\cite{twolevel} the main focus is on extensions of basic 2LTT which do not have
+staging semantics anymore. Ibid.\ conjectures the strong conservativity of 2LTT
+in Proposition 2.18, and suggests categorical gluing as proof strategy. In this
+work, we do essentially that: our presheaf model and the logical relation model
+could be merged into a single instance of gluing along the $\re$ morphism; this
+merges staging and its soundness proof. We split this construction to two parts
+in order to get a self-contained presentation of the staging algorithm.
+
+The multi-stage calculus in \cite{multi-stage-calculus}
 supports staging operations and type dependencies. However, there are major
@@
 from a multi-stage program. Similar considerations apply to systems with the S4
-modality (see e.g.\ Davies and Pfenning \cite{DBLP:journals/jacm/DaviesP01}),
+modality (see e.g.\ \cite{DBLP:journals/jacm/DaviesP01}),
 where there is no judgment that guarantees elimination of modal operations.
@@
 
-\bibliographystyle{ACM-Reference-Format}
 \interlinepenalty=10000

### 2022-07-20 paper/paper.tex
title fix, acmart update

@@
 
-\title{Staged Compilation With Two-Level Type Theory}
+\title{Staged Compilation with Two-Level Type Theory}
 \titlenote{The author was supported by the “Application Domain Specific Highly
@@
 arguments. For example, $\mit{id}_1\,(\Lift\,\Bool_0)\,\qut{\true_0}$ has type
-$\Lift \Bool_0$, therefore $\spl(\mit{id}_1\,(\Lift\,\Bool_0)\,\qut{\true_0})$
+$\Lift \Bool_0$, which implies that $\spl(\mit{id}_1\,(\Lift\,\Bool_0)\,\qut{\true_0})$
 has type $\Bool_0$. We can stage this expression as follows:
@@
   & =\hspace{0.3em}&&\spl\qut{\msf{foldr}_0\,
-        \spl \qut{\Nat_0}\,\spl \qut{\List_0\,\spl\qut{Nat_0}}\,
-        (\lambda\,a\,\mit{bs}.\, \cons_0\,\spl\qut{\Nat_0}\,\spl\qut{\spl\qut{a} +_0 10}\,\mit{bs})\,
+        \spl \qut{\Nat_0}\,\spl \qut{\List_0\,\spl\qut{Nat_0}}\\
+  & && \hspace{3.5em}(\lambda\,a\,\mit{bs}.\, \cons_0\,\spl\qut{\Nat_0}\,\spl\qut{\spl\qut{a} +_0 10}\,\mit{bs})\,
         (\nil_0\,\spl \qut{\Nat_0})\,
@@
 \cite{boehm-berarducci} under the lifting to get
-\[ \Lift((L : \U_0) \to (A \to L \to L) \to L \to L) \]
-which is isomorphic to
-\[ (L : \Lift\,U_0) \to (\Lift\,A \to \Lift\,\spl L \to \Lift\,\spl L)
-    \to \Lift\,\spl L \to \Lift\,\spl L. \]
+\[ \Lift((L : \U_0) \to (A \to L \to L) \to L \to L) \simeq ((L : \Lift\,U_0) \to (\Lift\,A \to \Lift\,\spl L \to \Lift\,\spl L)). \]
 Alternatively, for \emph{stream fusion} \cite{stream-fusion}, we embed $\List\,A$ into the
@@
 
-%% This is fairly similar to the well-known monomorphization models in the C++ and
+%% This is similar to the well-known monomorphization models in the C++ and
 %% Rust programming languages. The evident difference is that in 2LTT we can use a

### 2022-07-20 paper/paper.tex
pi/sigma consistent notation

@@
   erased. We leave the meta-level $\U_{1,j}$ hierarchy unchanged.
-\item We may introduce unboxed $\Sigma$ types and primitive machine types in the
+\item We may introduce unboxed $\Sigma$-types and primitive machine types in the
   runtime language. For $r : \Rep$, $r' : \Rep$, $A : \U_{0}\,r$ and $B : A \to
@@
 \begin{itemize}
-\item Meta-level $\Sigma$ types are preserved up to isomorphism, so we have
+\item Meta-level $\Sigma$-types are preserved up to isomorphism, so we have
   \[ \re_{\Sigma} : \re\,(\Sigma\,A\,B)\,\gamma
@@
   the preservation property.
-\item Meta-level $\Pi$ types and universes are preserved in a lax way. For $\Pi$, we have
+\item Meta-level $\Pi$-types and universes are preserved in a lax way. For $\Pi$, we have
   \[ \re_{\app} : \re\,(\Pi\,A\,B)\,\gamma \to (\alpha : \re\,A\,\gamma) \to \re\,B\,(\rexti\,(\gamma,\,\alpha))

### 2022-07-28 paper/paper.tex
edit

@@
 
-\subsubsection{Type formers} $\U_0$ and $\U_1$ may be closed under arbitrary type formers,
+\subsubsection{Type Formers} $\U_0$ and $\U_1$ may be closed under arbitrary type formers,
 such as functions, $\Sigma$-types, identity types or inductive types in general.
@@
 
-\subsubsection{Moving between stages}
+\subsubsection{Moving Between Stages}
 At this point, our system is rather limited, since there is no interaction
@@
 
-\subsubsection{Inferring staging operations}
+\subsubsection{Inferring Staging Operations}
 We can extract a coercive subtyping system from the staging operations, which
@@
 
-\subsubsection{Memory representation polymorphism}
+\subsubsection{Memory Representation Polymorphism}
 
@@
 
-\subsubsection{Metatheory}
+\paragraph{Metatheory}
 The metatheory is an intensional type theory with a cumulative universe
@@
 
-\subsubsection{Elaborating informal syntax}
+\subsubsection{Elaborating Informal Syntax}
 The justification for the usage of the informal syntax in Section
@@
 
-\subsubsection{The setoid interpretation of the syntax}\label{sec:2lttsetoid} We need to show
+\subsubsection{The Setoid Interpretation of the Syntax}\label{sec:2lttsetoid} We need to show
 that the 2LTT syntax, as a particular inductive type, can be modeled in setoids.
@@
 
-\subsubsection{The syntactic category and the meta-level fragment}
+\subsubsection{The Syntactic Category and the Meta-Level Fragment}
 
@@
 
-\subsubsection{The object-level fragment}
+\subsubsection{The Object-Level Fragment}
 
@@
 
-\subsubsection{Caching} In a production-strength implementation we would need some
+\paragraph{Caching} In a production-strength implementation we would need some
 caching mechanism, to avoid excessive duplication of code. For example, if we
@@
 \end{alignat*}
-\subsubsection{Object-theoretic syntax} The syntax of the object theory is clearly
+\subsubsection{Object-Theoretic Syntax} The syntax of the object theory is clearly
 fully stable under object-theoretic substitution, so we can internalize all of
@@
 \end{alignat*}
-\subsubsection{Object-level fragment of 2LTT}
+\subsubsection{Object-Level Fragment of 2LTT}
 The purely object-level syntactic fragment of 2LTT can be internalized as
@@
 
-\subsubsection{Preservation properties of $\re$}
+\subsubsection{Preservation Properties of $\re$}
 First, we note that $\re$ strictly preserves $\id$, $\blank\circ\blank$ and
@@
 \end{itemize}
-\subsubsection{Action on lifting} We have an isomorphism $\re\,(\Lift\,A)\,\gamma \simeq
+\subsubsection{Action on Lifting} We have an isomorphism $\re\,(\Lift\,A)\,\gamma \simeq
 \Tm_{0,j}\,(\re\,A\,\gamma)$. This is given by quoting and splicing: we convert
@@
 
-\subsubsection{Action on object-level types and terms} $\re$ preserves all structure
+\subsubsection{Action on Object-Level Types and Terms} $\re$ preserves all structure
 in the object-level fragment of 2LTT. This follows from the $\re$ specification:
@@
 
-\subsubsection{Syntactic category and terminal object}
+\subsubsection{Syntactic Category and Terminal Object}
 Here, we simply have $\id\rel\,\gamma\rel := \gamma\rel$ and $(\sigma \circ
@@
 
-\subsubsection{Meta-level family structure}
+\subsubsection{Meta-Level Family Structure}
 We interpret context extension and type/term substitution as follows. Note the
@@
 
-\subsubsection{Meta-level natural numbers} First, we have to define a relation:
+\subsubsection{Meta-Level Natural Numbers} First, we have to define a relation:
 \begin{alignat*}{4}
@@
 
-\subsubsection{Meta-level $\Sigma$-types} We define relatedness pointwise. Pairing and projection
+\subsubsection{Meta-Level $\Sigma$-Types} We define relatedness pointwise. Pairing and projection
 are interpreted as meta-level pairing and projection.
@@
 
-\subsubsection{Meta-level $\Pi$-types} We again use a pointwise definition. Note that we need to use
+\subsubsection{Meta-Level $\Pi$-Types} We again use a pointwise definition. Note that we need to use
 $\re_{\app}$ to apply $t' : \re\,(\Pi\,A\,B)\,\gamma'$ to $\alpha'$.
@@
 
-\subsubsection{Meta-level universes} We interpret $\U_{1,j}$ as a semantic relation space:
+\subsubsection{Meta-Level Universes} We interpret $\U_{1,j}$ as a semantic relation space:
 \begin{alignat*}{3}
@@
 
-\subsubsection{Object-level family structure} We interpret extended contexts as follows.
+\subsubsection{Object-Level Family Structure} We interpret extended contexts as follows.
 \begin{alignat*}{3}
@@
 
-\subsubsection{Lifting structure}
+\subsubsection{Lifting Structure}
 \begin{alignat*}{3}
@@
 
-\subsubsection{Object-level type formers}
+\subsubsection{Object-Level Type Formers}
 Lastly, object-level type formers are straightforward. For types, we need
@@
 
-\subsubsection{Alternative presentations}
+\paragraph{Alternative presentations}
 The above conservativity proof could be rephrased as an instance of more modern
@@
 
-\subsubsection{Stability under weakening only}
+\paragraph{Stability under weakening only}
 We can move to different 2LTT setups, where stability under substitution is not
@@
 
-\subsubsection{Closed modalities}\label{sec:closed-modalities}
+\paragraph{Closed modalities}\label{sec:closed-modalities}
 Another option is to add a \emph{closed} or \emph{crisp} modality
@@
 
-The multi-stage calculus in \cite{multi-stage-calculus}
-supports staging operations and type dependencies. However, there are major
-differences to the current work. First, ibid.\ does not have large elimination,
-hence it does not support computing types by staging. Second, it does not
-support staged compilation in the sense of Definition \ref{def:staging}; rather,
-it is a system for runtime code manipulation and evaluation (which we may call
-staged \emph{computation}). In other words, programs may manipulate code
-fragments during execution, in more liberal ways than in 2LTT, but there is no
-judgment which ensures that a purely ``object-level'' program can be extracted
-from a multi-stage program. Similar considerations apply to systems with the S4
-modality (see e.g.\ \cite{DBLP:journals/jacm/DaviesP01}),
-where there is no judgment that guarantees elimination of modal operations.
+The multi-stage calculus in \cite{multi-stage-calculus} supports staging
+operations and type dependencies. However, there are major differences to the
+current work. First, ibid.\ does not have large elimination, hence it does not
+support computing types by staging. Second, it does not support staged
+compilation in the sense of Definition \ref{def:staging}; rather, it is a system
+for runtime code manipulation and evaluation (which we may call staged
+\emph{computation}). In other words, programs may manipulate code fragments
+during execution, in more liberal ways than in 2LTT, but there is no judgment
+which ensures that a purely ``object-level'' program can be extracted from a
+multi-stage program. Similar considerations apply to systems with the S4
+modality (see e.g.\ \cite{DBLP:journals/jacm/DaviesP01}) where there is no
+judgment that guarantees elimination of modal operations.
 
@@
 the meta level}, rather than staging operations; staging operations in 2LTT and
-MetaML are already quite similar. For example, any interesting operation on
+MetaOCaml are already quite similar. For example, any interesting operation on
 the static-length vectors in Section \ref{sec:staged-programming} requires

### 2022-07-29 paper/paper.tex
edits

@@
 \paragraph{Closed modalities}\label{sec:closed-modalities}
-Another option is to add a \emph{closed} or \emph{crisp} modality
-\cite{licata2018internal} for tracking object-level values which will become
-closed terms after staging. Since closed terms are not affected by substitution,
-we are free to analyze them. Such analysis can be viewed as an internalization
-of canonicity for closed object-level terms. For example, a closed $\Bool_0$
-term will be either $\true_0$ or $\false_0$ after staging, so it makes sense to
-be able to eliminate from it to $\Bool_1$. This can be also viewed as a safe
-implementation of a \emph{run} function in the style of MetaML \cite{metaml}.
-In our case \emph{scope extrusion} would be prevented by the closed modality.
-
-\emph{Function pointers} in the style of the C programming language could be also
-an interesting use case, since these must become closed after staging (as they
+Another option is to add a \emph{closed} or \emph{crisp} modality for modeling
+closed object-level terms \cite{licata2018internal}. Since closed terms are not
+affected by substitution, we are able to analyze them, and we can also recover
+open terms by explicitly abstracting over contexts
+\cite{DBLP:conf/lics/Hofmann99,DBLP:journals/corr/abs-2206-02831}. \emph{Function
+pointers} in the style of the C programming language could be also an
+interesting use case, since these must become closed after staging (as they
 cannot capture closure environments).
+
+%% for tracking object-level values which will become
+%% closed terms after staging. Since closed terms are not affected by substitution,
+%% we are free to analyze them. Such analysis can be viewed as an internalization
+%% of canonicity for closed object-level terms. For example, a closed $\Bool_0$
+%% term will be either $\true_0$ or $\false_0$ after staging, so it makes sense to
+%% be able to eliminate from it to $\Bool_1$. This can be also viewed as a safe
+%% implementation of a \emph{run} function in the style of MetaML \cite{metaml}.
+%% In our case \emph{scope extrusion} would be prevented by the closed modality.
 
@@
 
+In \cite{DBLP:conf/lics/Hofmann99}, it is described how to model higher-order
+abstract syntax of simply typed object languages in presheaves over various
+syntactic categories. In the case where syntactic morphisms are substitutions,
+the internal language described by Hofmann can be viewed as an alternative
+syntax for a 2LTT, where we only have a single family structure and the object
+level is modeled with a Tarski-style universe.
+
 The multi-stage calculus in \cite{multi-stage-calculus} supports staging
@@
 support computing types by staging. Second, it does not support staged
-compilation in the sense of Definition \ref{def:staging}; rather, it is a system
-for runtime code manipulation and evaluation (which we may call staged
-\emph{computation}). In other words, programs may manipulate code fragments
-during execution, in more liberal ways than in 2LTT, but there is no judgment
-which ensures that a purely ``object-level'' program can be extracted from a
-multi-stage program. Similar considerations apply to systems with the S4
-modality (see e.g.\ \cite{DBLP:journals/jacm/DaviesP01}) where there is no
-judgment that guarantees elimination of modal operations.
+compilation in the sense of Definition \ref{def:staging}; there is no judgment
+which ensures that staging operations can be eliminated and a purely
+``object-level'' program can be extracted from a multi-stage program. Rather, the system
+supports \emph{runtime} code generation. There are several modal multi-stage
+systems which support runtime code generation but not staged compilation, such
+as those based on S4 modalities (e.g.\ \cite{DBLP:journals/jacm/DaviesP01}) or
+contextual modalities (\cite{DBLP:journals/pacmpl/JangGMP22}).
 
@@
 
-In MetaML and MetaOCaml, there is a \emph{run} function for evaluating closed
-object-level terms: as we mentioned in Section \ref{sec:closed-modalities}, we
-believe that the proper way to implement it in 2LTT would be to use a closed
-modality. This use case is closely related to \cite{licata2018internal}. In
-ibid.\ a semantics is sketched for this modality in a presheaf topos, and we
-could directly adapt it to our presheaf model of 2LTT. That said, working
-out the practical details of this modality remains future work.
+In MetaOCaml, there is a \emph{run} function for evaluating closed object-level
+terms, however, this can fail during staging since closedness is not statically
+ensured. In our setting, this could be reproduced in a safe way using a closed
+modality; this is part of future work.
+
+%% as we mentioned in Section \ref{sec:closed-modalities}, we
+%% believe that the proper way to implement it in 2LTT would be to use a closed
+%% modality. This use case is closely related to \cite{licata2018internal}. In
+%% ibid.\ a semantics is sketched for this modality in a presheaf topos, and we
+%% could directly adapt it to our presheaf model of 2LTT. That said, working
+%% out the practical details of this modality remains future work.
 

### 2022-07-30 paper/paper.tex
edit

@@
 
-In \cite{DBLP:conf/lics/Hofmann99}, it is described how to model higher-order
-abstract syntax of simply typed object languages in presheaves over various
-syntactic categories. In the case where syntactic morphisms are substitutions,
-the internal language described by Hofmann can be viewed as an alternative
-syntax for a 2LTT, where we only have a single family structure and the object
-level is modeled with a Tarski-style universe.
+\cite{DBLP:conf/lics/Hofmann99} models higher-order abstract syntax of simply
+typed object languages in presheaves over various syntactic categories. The main
+difference between HOAS and 2LTT is that the former represents object-level binders
+using meta-level functions while the latter has a primitive notion of object-level
+binding.
 

### 2022-09-20 paper/paper.tex
typos

@@
   \item We provide a small standalone implementation of elaboration and staging
-    for a two-level type theory, with code samples that expand on the paper
-    \cite{staged-demo}.
+    for a two-level type theory \cite{staged-demo}.
   \item To our knowledge, this work is the first formalization and implementation of
@@
   & =\hspace{0.3em}&&\spl\qut{\msf{foldr}_0\,
-        \spl \qut{\Nat_0}\,\spl \qut{\List_0\,\spl\qut{Nat_0}}\\
+        \spl \qut{\Nat_0}\,(\List_0\,\spl\qut{\Nat_0})\\
   & && \hspace{3.5em}(\lambda\,a\,\mit{bs}.\, \cons_0\,\spl\qut{\Nat_0}\,\spl\qut{\spl\qut{a} +_0 10}\,\mit{bs})\,
@@
         \spl\qut{\mit{ns}}}\\
-  & =\hspace{0.3em} &&\msf{foldr_0}\,\Nat_0\,(\List_0\,\Nat_0)\,\,(\lambda\,a\,bs.\,\cons_0\,(a +_0 10)\,bs)\,(\nil_0\,\Nat_0)\,\mit{ns}
+  & =\hspace{0.3em} &&\msf{foldr_0}\,\Nat_0\,(\List_0\,\Nat_0)\,\,(\lambda\,a\,bs.\,\cons_0\,\Nat_0\,(a +_0 10)\,bs)\,(\nil_0\,\Nat_0)\,\mit{ns}
 \end{alignat*}
@@
 \cite{boehm-berarducci} under the lifting to get
-\[ \Lift((L : \U_0) \to (A \to L \to L) \to L \to L) \simeq ((L : \Lift\,U_0) \to (\Lift\,A \to \Lift\,\spl L \to \Lift\,\spl L)). \]
+\begin{alignat*}{3}
+         &\Lift((L : \U_0) \to (A \to L \to L) \to L \to L)\\
+  \simeq\,\,&((L : \Lift\,U_0) \to (\Lift\,A \to \Lift\,\spl L \to \Lift\,\spl L) \to \Lift\,\spl L \to \Lift\,\spl L).
+\end{alignat*}
 Alternatively, for \emph{stream fusion} \cite{stream-fusion}, we embed $\List\,A$ into the
@@
 
+\section{Data Availability Statement}
+
+A standalone prototype implementation is available as \cite{staged-demo}. It
+implements elaboration and staging for a 2LTT. It includes a tutorial and code
+examples that expand on the contents of Section \ref{sec:tour-of-2ltt}.
+
 \interlinepenalty=10000

### 2023-08-01 partially_static/paper/paper.tex
stackage bump, loads of new notes

%% build: latexmk -pdf -pvc paper.tex

\documentclass[acmsmall,review,screen]{acmart}

%% \setcopyright{acmcopyright}
%% \copyrightyear{2018}
%% \acmYear{2018}
%% \acmDOI{XXXXXXX.XXXXXXX}

\bibliographystyle{ACM-Reference-Format}
\citestyle{acmauthoryear}

%% --------------------------------------------------------------------------------

\usepackage{xcolor}
\usepackage{mathpartir}
\usepackage{todonotes}
\presetkeys{todonotes}{inline}{}
\usepackage{scalerel}
\usepackage{bm}

\newcommand{\mit}[1]{\mathit{#1}}
\newcommand{\msf}[1]{\mathsf{#1}}
\newcommand{\mbb}[1]{\mathbb{#1}}
\newcommand{\mbf}[1]{\mathbf{#1}}
\newcommand{\bs}[1]{\boldsymbol{#1}}
\newcommand{\wh}[1]{\widehat{#1}}
\newcommand{\ext}{\triangleright}
\newcommand{\Code}{\msf{Code}}
\newcommand{\El}{\msf{El}}
\newcommand{\lam}{\msf{lam}}
\newcommand{\app}{\msf{app}}
\newcommand{\NatElim}{\msf{NatElim}}
\newcommand{\y}{\msf{y}}

\newcommand{\Lift}{{\Uparrow}}
\newcommand{\spl}{{\sim}}
\newcommand{\qut}[1]{\langle #1\rangle}

\newcommand{\mbbc}{\mbb{C}}
\newcommand{\mbbo}{\mbb{O}}
\newcommand{\ob}{_\mbbo}

\newcommand{\U}{\msf{U}}
\newcommand{\Con}{\msf{Con}}
\newcommand{\Sub}{\msf{Sub}}
\newcommand{\Ty}{\msf{Ty}}
\newcommand{\Tm}{\msf{Tm}}
\newcommand{\Cono}{\msf{Con}_{\mbbo}}
\newcommand{\Subo}{\msf{Sub}_{\mbbo}}
\newcommand{\Tyo}{\msf{Ty}_{\mbbo}}
\newcommand{\Tmo}{\msf{Tm}_{\mbbo}}
\newcommand{\hCon}{\wh{\msf{Con}}}
\newcommand{\hSub}{\wh{\msf{Sub}}}
\newcommand{\hTy}{\wh{\msf{Ty}}}
\newcommand{\hTm}{\wh{\msf{Tm}}}

\newcommand{\p}{\mathsf{p}}
\newcommand{\q}{\mathsf{q}}

\newcommand{\refl}{\msf{refl}}
\newcommand{\Bool}{\msf{Bool}}
\newcommand{\true}{\msf{true}}
\newcommand{\false}{\msf{false}}
\newcommand{\True}{\msf{True}}
\newcommand{\False}{\msf{False}}
\newcommand{\List}{\msf{List}}
\newcommand{\nil}{\msf{nil}}
\newcommand{\cons}{\msf{cons}}
\newcommand{\Nat}{\msf{Nat}}
\newcommand{\zero}{\msf{zero}}
\newcommand{\suc}{\msf{suc}}
\renewcommand{\tt}{\msf{tt}}
\newcommand{\fst}{\msf{fst}}
\newcommand{\snd}{\msf{snd}}
\newcommand{\mylet}{\msf{let}}
\newcommand{\emptycon}{\scaleobj{.75}\bullet}
\newcommand{\id}{\msf{id}}

\newcommand{\Set}{\mathsf{Set}}
\newcommand{\Prop}{\mathsf{Prop}}
\newcommand{\Rep}{\msf{Rep}}
\newcommand{\blank}{{\mathord{\hspace{1pt}\text{--}\hspace{1pt}}}}
\newcommand{\emb}[1]{\ulcorner#1\urcorner}

\newcommand{\Stage}{\msf{Stage}}
\newcommand{\hato}{\bm\hat{\mbbo}}
\newcommand{\ev}{\mbb{E}}
\newcommand{\re}{\mbb{R}}

\theoremstyle{remark}
\newtheorem{notation}{Notation}

\newcommand{\whset}{\wh{\Set}}
\newcommand{\rexti}{\re_{\ext_1}^{-1}}
\newcommand{\rextizero}{\re_{\ext_0}^{-1}}

\newcommand{\rel}{^{\approx}}
\newcommand{\yon}{\msf{y}}

%% --------------------------------------------------------------------------------

\begin{document}

%%
%% The "title" command has an optional parameter,
%% allowing the author to define a "short title" to be used in page headers.
\title{Partially Static Types and Staged Partial Evaluation in Two-Level Type Theory}

%%
%% The "author" command and its associated commands are used to define
%% the authors and their affiliations.
%% Of note is the shared affiliation of the first two authors, and the
%% "authornote" and "authornotemark" commands
%% used to denote shared contribution to the research.
\author{András Kovács}
\email{kovacsandras@inf.elte.hu}
\orcid{0000-0002-6375-9781}
\affiliation{%
  \institution{Eötvös Loránd University}
  \country{Hungary}
  \city{Budapest}
}

%% The abstract is a short summary of the work to be presented in the
%% article.
\begin{abstract}
  Two-level type theory (2LTT) is a system that can be used for two-stage
  compilation, which allows dependent types both in metaprograms and in
  generated code output. Partially static data types are used in staged
  compilation in situations where only some parts of structures are known at
  compile time. We investigate partially static types and partial evaluation in
  the context of 2LTT. A key question is whether we can mechanically obtain
  partially static types from non-staged data types, and likewise lift
  non-staged operations to operations on partially static data. We show that all
  types and constructions in a Martin-Löf type theory (MLTT) can be given a
  partially static interpretation. Moreover, this interpretation yields a staged
  partial evaluator for MLTT, defined within 2LTT, which can be used to compile
  embedded MLTT syntax to normalized object code. We show that the
  interpretation can be used to partially mechanize the construction of staged
  normalizers for certain algebras. Finally, we compare and investigate free
  extensions of algebras (``frex'') in 2LTT as an alternative generic approach to
  partially static types.
\end{abstract}

\begin{CCSXML}
<ccs2012>
   <concept>
       <concept_id>10003752.10003790.10011740</concept_id>
       <concept_desc>Theory of computation~Type theory</concept_desc>
       <concept_significance>500</concept_significance>
       </concept>
   <concept>
       <concept_id>10011007.10011006.10011041.10011047</concept_id>
       <concept_desc>Software and its engineering~Source code generation</concept_desc>
       <concept_significance>500</concept_significance>
       </concept>
 </ccs2012>
\end{CCSXML}

\ccsdesc[500]{Theory of computation~Type theory}
\ccsdesc[500]{Software and its engineering~Source code generation}
\keywords{type theory, two-level type theory, staged compilation, partial evaluation}
\maketitle

\section{Introduction}\label{sec:introduction}

In most practical programming languages it is possible to write code-generating
code by simply manipulating strings or syntax trees, but this tends to be
tedious, unsafe and non-composable. The purpose of \emph{staged compilation} is
to support metaprogramming with better ergonomics and more safety guarantees. In
two-stage systems, user-written metaprograms are executed at compile time to
produce code that is processed in further compilation.

\emph{Two-level type theory} (2LTT) is a framework which supports two-stage
compilation with strong safety and correctness properties. It is also compatible
with a wide range of object-level and meta-level language features; in
particular it allows dependent types at both levels. 2LTT was originally used in
a purely mathematical context in synthetic homotopy theory \cite{twolevel}, but
it can be also applied in staged compilation \cite{2ltt-staged}.

A common application of staging is to use embedded languages without
interpretative overhead, using \emph{staged interpretation}, or going further,
to optimize embedded programs or partially evaluate them at compile
time. \emph{Partially static data} is commonly used in staged interpreters and
partial evaluators \cite{TODO}. Such data contains a mixture of expressions of
the object language and actual structured data that is computed at compile time.
A simple example is a compile-time list whose elements are object-level expressions.

In this paper, we investigate partially static types and partial evaluation in
the context of 2LTT. In this context, we have a full-powered mathematical
language at our disposal on the meta level. Our object language is a bit simpler
but still highly expressive. Hence, we can internalize much of the formal
reasoning about staging in 2LTT itself, and internalize constructions which were
not feasible in systems with weaker type systems.

\subsection{Overview \& Contributions}\label{sec:overview}
\begin{itemize}
\item In \textbf{Section \ref{sec:2ltt}} we describe the specific variant of 2LTT used
      in this paper and give a short overview of its staging features.
\item In \textbf{Section \ref{sec:ps-interp}} we consider deep embeddings of a
      Martin-Löf type theory (MLTT) into 2LTT, and describe its staged
      interpretation and partial evaluation. The former maps embedded MLTT syntax to
      object-level code. The latter is similar, but it can also perform
      $\beta\eta$-normalization and could serve as a basis for more sophisticated
      compile-time optimization. We define a model of MLTT where each type is
      interpreted as an object-level type (``dynamic'' type) together with a
      partially static type, and there is an embedding of the former into the
      latter.  The interpretation of MLTT syntax in this model corresponds to staged
      partial evaluation.
      \begin{itemize}
        \item  From another point of view, this model explains how to extend all MLTT types
               with neutral values representing object-level expressions, obtaining partially
               static types.
        \item Since MLTT terms are also modeled, any function in MLTT can be
              interpreted as a function operating on partially static types in 2LTT.
      \end{itemize}
\item In \textbf{Section \ref{sec:ps-alg}} we describe an application of the
  partially static interpretation in normalization for algebraic structures.  In
  some cases it is possible to replace an algebra with an equivalent algebra,
  but in which some or all equations hold definitionally. Then, we can take the
  partially static interpretation of such strict algebras. This yields a staged
  normalizer with respect to the definitional equations.

\item In \textbf{Section \ref{sec:frex}} we look at free extensions of algebras,
  called \emph{frex} in prior literature
  \cite{DBLP:journals/pacmpl/YallopGK18}. A key advantage here is that algebras
  can be extended with decidable or even finite sets of variables, which makes
  it possible to compute more normal forms. For example, normalization for free
  groups require decidable equality of variables. Some subtleties arise here
  from the 2LTT staging setup and the fact that we internally verify algebraic
  laws. Free extensions always exist, but only those can be actually interpreted in
  object-level algebras which have a ``fully normalized'' presentation that does not
  use quotients.
\end{itemize}

\section{Overview of Two-Level Type Theory}\label{sec:2ltt}

\section{Staged Interpretation and Normalization for Martin-Löf Type Theory}\label{sec:ps-interp}

\section{Partially Static Types in Normalization of Algebras}\label{sec:ps-alg}

\section{Free Extensions of Algebras in Two-Level Type Theory}\label{sec:frex}

\section{Related Work \& Conclusions}\label{sec:conclusion}

\interlinepenalty=10000
\bibliography{references}

\end{document}
\endinput

### 2023-08-01 hipe-doc/Notes.hs

@@ -1,0 +1,151 @@
+{-# language UnliftedDatatypes, GADTs, MagicHash #-}
+
+module Notes (foo) where
+
+import GHC.Exts hiding (List)
+
+-- data List :: UnliftedType where
+--   Nil :: List
+--   Cons :: Int# -> List -> List
+
+-- foo :: List -> List
+-- foo Nil = Nil
+-- foo (Cons x xs) = Cons (x +# 10#) (foo xs)
+
+data Tree :: UnliftedType where
+  Leaf :: Int# -> Tree
+  Node :: Tree -> Tree -> Tree
+
+foo :: Tree -> Tree
+foo (Leaf x) = Leaf (x +# 10#)
+foo (Node l r) = Node (foo l) (foo r)
+
+
+{-
+
+     {offset
+       cD3: // global
+           if (R2 & 7 != 1) goto cCZ; else goto cCY;
+       cCZ: // global
+           I64[Sp - 16] = cDd;
+           _sCJ::I64 = I64[R2 + 14];
+           R2 = P64[R2 + 6];
+           I64[Sp - 8] = _sCJ::I64;
+           Sp = Sp - 16;
+           call Notes.foo_info(R2) returns to cDd, args: 8, res: 8, upd: 8;
+       cDd: // global
+           Hp = Hp + 24;
+           I64[Hp - 16] = Notes.Cons_con_info;
+           P64[Hp - 8] = R1;
+           I64[Hp] = I64[Sp + 8] + 10;
+           R1 = Hp - 14;
+           Sp = Sp + 16;
+           call (P64[Sp])(R1) args: 8, res: 0, upd: 8;
+       cCY: // global
+           R1 = Notes.Nil_closure+1;
+           call (P64[Sp])(R1) args: 8, res: 0, upd: 8;
+     }
+ },
+
+rbp      = Sp
+r15      = SpLim
+r12      = Hp
+r13      = BaseReg
+r13[856] = HpLim
+r13[904] = HpAlloc
+
+.LcD0:
+	leaq -16(%rbp),%rax
+	cmpq %r15,%rax
+	jb .LcD1
+.LcD2:
+	movq %r14,%rax
[…]

### 2023-08-01 hipe-doc/Notes2.hs

@@ -1,0 +1,126 @@
+{-# language MagicHash, UnliftedDatatypes, UnboxedTuples #-}
+
+module Notes2 (foo) where
+
+import GHC.Exts
+
+-- data Box = Box (Int# -> Int#)
+
+-- foo :: Box -> Int#
+-- foo (Box f) = f 100#
+
+-- foo :: Bool -> Bool
+-- foo True = False
+-- foo x    = x
+
+-- foo :: (Bool -> a) -> Bool -> a
+-- foo f b = f (case b of True -> False; _ -> True)
+
+-- data UBool :: UnliftedType where
+--   UTrue :: UBool
+--   UFalse :: UBool
+
+-- foo :: (UBool -> a) -> UBool -> a
+-- foo f b = f (case b of UTrue -> UFalse; _ -> UTrue)
+
+-- -- pushes call continuation on stack
+-- foo f x = f x x x x x x x x x x
+
+-- foo (x :: Int#)(y :: Int#)(z :: Int#) = (# x, y, z #)
+
+-- bar :: [Int] -> [Int]
+-- bar x = 0 : x
+-- {-# noinline bar #-}
+
+-- foo (x :: [Int])(y :: Int#) =
+--   let !kek = bar x
+--   in I# y : kek
+
+foo (x :: Int#) = x
+
+
+{-
+Passing cont in register in LLVM?
+
+
+
+
+f x =
+  let y = g x;
+  let y' = g y;
+  let y'' = h y y'
+  ret y''
+
+f0[](k, x) =
+  push k;
+  tcall g(&f1, x)
+
+f1[k](y) =
+  push y;
+  tcall g(&f2, y)
+
+f2[k, y](y') =
+  y = pop
+  k = pop
+  tcall h(k, y, y')
+
+
+-- f0[sp](k, x) =
+--   tcall g[sp,k](f1, x)
+
+
+f x = g x x
+
+f[k](x) =
+  tailcall g[k](x, x)    -- hmmm
+
+- if we have tailcall, we just keep the stack or reset it, no pushin
+- if we have non-tail call, we have to push ret addresses anyway
+
+- pushing ret address on non-tail call is just fine
+- we can try to share stax
+
+
+
+
+    ...
+    let y = g x        pass k to g, push
+  k:
+    ...
+
+f [sp] k x =
+  g x \[sp, k] y ->
[…]

### 2023-08-01 hipe-doc/sigbench.cpp

@@ -1,0 +1,614 @@
+#include <assert.h>
+#include <chrono>
+#include <errno.h>
+#include <fcntl.h>
+#include <inttypes.h>
+#include <poll.h>
+#include <pthread.h>
+#include <signal.h>
+#include <stdio.h>
+#include <stdlib.h>
+#include <sys/ioctl.h>
+#include <sys/mman.h>
+#include <sys/syscall.h>
+#include <sys/types.h>
+#include <linux/userfaultfd.h>
+#include <unistd.h>
+#include <string.h>
+
+////////////////////////////////////////////////////////////////////
+// userfaultfd Write Protection Test
+//
+// Aditya Mandaleeka
+// July 28 2020, Hackathon
+//
+// This is a sample program I wrote to learn more about the
+// userfaultfd write protection capabilities introduced
+// recently in the mainline Linux kernel.
+//
+// In addition to learning, I also wanted to compare how UFFD
+// WP compares to using the traditional PROT_NONE+SEGV handler
+// trick in order to catch dirtied memory pages. This is useful
+// for specific VM/GC usecases; one example is for cross-
+// generational write tracking, and another is safepoint polls.
+//
+// Using signal handlers for such control flow "works", but it
+// has problems:
+//     - Handling signals is very expensive and involves a lot
+//       of work on the kernel side every time a signal is
+//       raised. Handling it all in userspace can thus be a
+//       performance win.
+//     - Modifying access protections on memory regions involves
+//       touching the VMA entries in the mm subsystem, which not
+//       only involves taking some low-level locks, but also
+//       leads to a proliferation of VMAs when dealing with a
+//       large address region with permissions that can be dis-
+//       contiguous over the span of the process's lifetime.
+//       The UFFD code avoids the need to deal with VMAs entirely.
+//     - Signal handlers are touchy about what's actually safe
+//       to do inside them, so we have to be extra careful when
+//       doing interesting things from the context of the SEGV
+//       handler, often jumping through extra hoops to make it
+//       safe.
+//
[…]

### 2024-02-03 newpaper/paper.tex
stuff

%% build: latexmk -pdf -pvc paper.tex

\documentclass[acmsmall,screen,review,anonymous]{acmart}
%% \documentclass[nonacm,acmsmall,review]{acmart}
%% \raggedbottom

%% \BibTeX command to typeset BibTeX logo in the docs
\AtBeginDocument{%
  \providecommand\BibTeX{{%
    \normalfont B\kern-0.5em{\scshape i\kern-0.25em b}\kern-0.8em\TeX}}}

%% \setcopyright{rightsretained}
%% \acmPrice{}
%% \acmDOI{10.1145/3547641}
%% \acmYear{2022}
%% \copyrightyear{2022}
%% \acmSubmissionID{icfp22main-p52-p}
%% \acmJournal{PACMPL}
%% \acmVolume{6}
%% \acmNumber{ICFP}
%% \acmArticle{110}
%% \acmMonth{8}

%% These commands are for a JOURNAL article.

%% \acmJournal{JACM}
%% \acmVolume{37}
%% \acmNumber{4}
%% \acmArticle{111}
%% \acmMonth{8}

%%
%% Submission ID.
%% Use this when submitting an article to a sponsored event. You'll
%% receive a unique submission ID from the organizers
%% of the event, and this ID should be used as the parameter to this command.
%%\acmSubmissionID{123-A56-BU3}

\bibliographystyle{ACM-Reference-Format}
%%
%% The majority of ACM publications use numbered citations and
%% references.  The command \citestyle{authoryear} switches to the
%% "author year" style.
%%
%% If you are preparing content for an event
%% sponsored by ACM SIGGRAPH, you must use the "author year" style of
%% citations and references.
%% Uncommenting
%% the next command will enable that style.
\citestyle{acmauthoryear}

%% --------------------------------------------------------------------------------

\usepackage{xcolor}
\usepackage{mathpartir}
\usepackage{todonotes}
\presetkeys{todonotes}{inline}{}
\usepackage{scalerel}
\usepackage{bm}
%% \usepackage{amssymb}

\newcommand{\mit}[1]{\mathit{#1}}
\newcommand{\msf}[1]{\mathsf{#1}}
\newcommand{\mbb}[1]{\mathbb{#1}}
\newcommand{\mbf}[1]{\mathbf{#1}}
\newcommand{\bs}[1]{\boldsymbol{#1}}
\newcommand{\wh}[1]{\widehat{#1}}
\newcommand{\ext}{\triangleright}
\newcommand{\Code}{\msf{Code}}
\newcommand{\El}{\msf{El}}
\newcommand{\lam}{\msf{lam}}
\newcommand{\app}{\msf{app}}
\newcommand{\NatElim}{\msf{NatElim}}
\newcommand{\y}{\msf{y}}

\newcommand{\Lift}{{\Uparrow}}
\newcommand{\spl}{{\sim}}
\newcommand{\qut}[1]{\langle #1\rangle}

\newcommand{\mbbc}{\mbb{C}}
\newcommand{\mbbo}{\mbb{O}}
\newcommand{\ob}{_\mbbo}

\newcommand{\U}{\msf{U}}
\newcommand{\Con}{\msf{Con}}
\newcommand{\Sub}{\msf{Sub}}
\newcommand{\Ty}{\msf{Ty}}
\newcommand{\Tm}{\msf{Tm}}
\newcommand{\Cono}{\msf{Con}_{\mbbo}}
\newcommand{\Subo}{\msf{Sub}_{\mbbo}}
\newcommand{\Tyo}{\msf{Ty}_{\mbbo}}
\newcommand{\Tmo}{\msf{Tm}_{\mbbo}}
\newcommand{\hCon}{\wh{\msf{Con}}}
\newcommand{\hSub}{\wh{\msf{Sub}}}
\newcommand{\hTy}{\wh{\msf{Ty}}}
\newcommand{\hTm}{\wh{\msf{Tm}}}

\newcommand{\p}{\mathsf{p}}
\newcommand{\q}{\mathsf{q}}

\newcommand{\refl}{\msf{refl}}
\newcommand{\Bool}{\msf{Bool}}
\newcommand{\true}{\msf{true}}
\newcommand{\false}{\msf{false}}
\newcommand{\True}{\msf{True}}
\newcommand{\False}{\msf{False}}
\newcommand{\List}{\msf{List}}
\newcommand{\nil}{\msf{nil}}
\newcommand{\cons}{\msf{cons}}
\newcommand{\Nat}{\msf{Nat}}
\newcommand{\zero}{\msf{zero}}
\newcommand{\suc}{\msf{suc}}
\renewcommand{\tt}{\msf{tt}}
\newcommand{\fst}{\msf{fst}}
\newcommand{\snd}{\msf{snd}}
\newcommand{\mylet}{\msf{let}}
\newcommand{\emptycon}{\scaleobj{.75}\bullet}
\newcommand{\id}{\msf{id}}

\newcommand{\Set}{\mathsf{Set}}
\newcommand{\Prop}{\mathsf{Prop}}
\newcommand{\Rep}{\msf{Rep}}
\newcommand{\blank}{{\mathord{\hspace{1pt}\text{--}\hspace{1pt}}}}
\newcommand{\emb}[1]{\ulcorner#1\urcorner}

\newcommand{\Stage}{\msf{Stage}}
\newcommand{\hato}{\bm\hat{\mbbo}}
\newcommand{\ev}{\mbb{E}}
\newcommand{\re}{\mbb{R}}

\theoremstyle{remark}
\newtheorem{notation}{Notation}

\newcommand{\whset}{\wh{\Set}}
\newcommand{\rexti}{\re_{\ext_1}^{-1}}
\newcommand{\rextizero}{\re_{\ext_0}^{-1}}

\newcommand{\rel}{^{\approx}}
\newcommand{\yon}{\msf{y}}

%% --------------------------------------------------------------------------------

%%
%% end of the preamble, start of the body of the document source.
%% \hypersetup{draft}
\begin{document}

\title{Closure-Free Functional Programming in Two-Level Type Theory}
%% \titlenote{}

%% The "author" command and its associated commands are used to define
%% the authors and their affiliations.
%% Of note is the shared affiliation of the first two authors, and the
%% "authornote" and "authornotemark" commands
%% used to denote shared contribution to the research.
\author{András Kovács}
\email{andrask@chalmers.se}
\orcid{0000-0002-6375-9781}
\affiliation{%
  \institution{University of Gothenburg}
  \country{Sweden}
  \city{Gothenburg}
}

\begin{abstract}
There are many abstraction tools in modern functional programming which heavily
rely on general-purpose compiler optimization to achieve adequate
performance. For example, monadic binding is a higher-order function which
yields runtime closures in the absence of sufficient compile-time inlining and
beta-reductions, thereby significantly degrading performance. In current systems
such as the Glasgow Haskell Compiler, there is no strong guarantee that
general-purpose optimization can eliminate abstraction overheads, and users only
have indirect and fragile control over code generation through inlining
directives and compiler options. In this paper we propose using a two-stage
language to simultaneously get strong code generation guarantees and strong
abstraction features. The object language is a simply typed first-order language
where all function calls are statically known, and which can be compiled without
runtime closures. The compile-time language is a dependent type theory. The two
are integrated as a \emph{two-level type theory}. We develop abstraction tools
in this setting and demonstrate two use cases. First, we develop monads and
monad transformers. Second, we develop fusion for push and pull streams. Most of
our results are also adapted to a proof-of-concept library in typed Template
Haskell.
\end{abstract}

\begin{CCSXML}
<ccs2012>
   <concept>
       <concept_id>10003752.10003790.10011740</concept_id>
       <concept_desc>Theory of computation~Type theory</concept_desc>
       <concept_significance>500</concept_significance>
       </concept>
   <concept>
       <concept_id>10011007.10011006.10011041.10011047</concept_id>
       <concept_desc>Software and its engineering~Source code generation</concept_desc>
       <concept_significance>500</concept_significance>
       </concept>
 </ccs2012>
\end{CCSXML}

\ccsdesc[500]{Theory of computation~Type theory}
\ccsdesc[500]{Software and its engineering~Source code generation}

\keywords{two-level type theory, staged compilation}

\maketitle

\section{Introduction}\label{sec:introduction}

\todo{FIGURE OUT: monadic tailrec}
\todo{FIGURE OUT: Traversable}
\todo{DEVELOP: mutual letrec based on computational product types}

Motivation.

Lots of abstraction in FP. Heavy reliance on general-purpose compiler
optimization, without guarantee of code quality.

Closures introduce heap allocation and dynamic control flow. Compilers
try hard to get rid of closures.

Here: use staging instead of closures. Essential use of closures is
remarkably rare.

General-purpose opt is slow, fragile, but automatic.

Metaprog is fast, deterministic, robust but requires user control.

We make metaprog easier and more streamlined.

Defunctionalization. Full defunc requires whole-program compilation.
Our approach does not use defunc.

Apparent closure use-cases can be locally defunctionalized. DList prime example.

Closure use cases:
\begin{itemize}
\item threaded interp, closure interp
\item Higher-order DSL-s (but they could be also staged to FO!)
\item ``Monadic reflection'' in effects
\item ListT working on any monad
\item CPS
\end{itemize}

Summary of contributions, sections, incl Typed TH implementation.

\section{The Object Language}\label{sec:the-object-language}

Describe object language. Simple types + first-order fun + fixpoints + ADT-s.

Show Agda formalization basically, types and syntax.

Show Agda definitional interpreter. Discuss closure-free evaluation

Examples, properties. Maybe big-step reduction? Maybe program equivalence?  Look
at Harper PFPL for program equivalence reference.

Sketch downstream compilation.
\begin{itemize}
\item Eta-expand function definitions to arity.
\item Turn functions in immediate beta-redexes to let-bindings.
\item Lambda-lift non-tail-called functions to top.
\item Keep only-tail-called local functions as join points.
\end{itemize}

Downstream compilation should erase the unit type and implement products as
``flat'' unboxed structures.

Discuss object language design choices.
\begin{itemize}
\item Can be more obviously first-order: e.g.\ no currying, only Val->Val
  functions, or ANF functions. This requires more bureaucracy in
  metaprogramming.
\item Can be more flexible but less obviously first-order: allow function let
  bodies, allow function case bodies. Requires more translation. More convenient
  to program in. In future work.
\end{itemize}

\section{The Staged Language}\label{sec:the-staged-language}

Don't give horizontal-bar rules.

Meta-features:
\begin{itemize}
\item Pi, Sigma, Unit, Bottom, Universes (indices elided), ADT-s. Ind families probably not needed.
\item Agda notation.
\item Type classes informally in Haskell notation?
\end{itemize}

Explain staging ops, staging algorithm. Explain object embedding. Refer to 2LTT
paper for correctness.

Mention stage annotation inference, but we'll not use it often! To make things
explicit.

We should include a generous amount of staging examples, and some traced staging.

Basic staging examples: identity, map, power function. Boolean short-circuiting.

Basic staged classes, Eq, Ord, Show, Monoid, etc. Vectors with computed length.

\section{Programming With Monads}\label{sec:programming-with-monads}

\subsection{Basic Binding Time Improvements}\label{sec:basic-binding-time-improvements}

Notion.
For functions.
For products.
Note that for products there is a computation duplication problem.
Mention let-insertion.

\subsection{The Code Generation Monad}\label{sec:the-code-generation-monad}

Is fundamental to this work.

Definition. Explanation. Define Functor, Applicative, Monad instances.
Define runGen, runGen1, runGenN etc. Define let-insertion.

Basic examples.

Explanation in terms of Church-coding.

Gen Vector mapping or folding example from 2LTT repo. Without Gen, it's a
big pain to get ideal linear-size map code in N. With Gen it's easy-peasy.

\subsection{Monad Transformers}\label{sec:monad-transformers}

Identity, ReaderT, StateT, MaybeT, ExceptT

Strict versions of put, modify, local, runner functions.

MTL-style overloading. MonadReader, MonadState, MonadError, MonadGen, liftGen

Binding time improvement for monads. Up/down overloading.

\subsection{Case Splitting in MonadGen}\label{sec:case-splitting-in-monadgen}

It works for any object ADT.

Introduce sums-of-values (SOV).

Introduce syntax sugar for object case splits in MonadGen blocks.

Basic examples.

\subsection{Joining Control Flow}\label{sec:rejoining-control-flow}

The problem of join points. Motivation in terms of MaybeT Gen.

We can let-bind computations by tripping through up/down of monads,
but that introduces unbox/rebox cost for Maybe \& Either.

Works OK for Reader/State though.

\begin{verbatim}
f : Bool -> MaybeT0 Gen ^Int
f = \b. ~(down do
  x <- case <b> of
    True  -> gen <10>
    False -> gen <20>
  y <- gen <~x + 10>
  z <- gen <~x * 10>
  pure <~x + ~y + ~z>)
\end{verbatim}

Join points using SOV-s. Description of SOV machinery. MonadJoin.
instance MonadJoin Gen. MonadJoin MaybeT.

\subsection{Monadic Tail Recursion}\label{sec:monadic-tail-recursion}

\texttt{replicateM\_}, \texttt{find} functions working in ``any'' monad.
Non-tail recursion is not possible to fuse, return addresses are dynamic.
Tail recursion can return to statically known points.

\subsection{Discussion}

Limitations. We can't put non-value actions into structures, can't accumulate them.

Guarantees/properties. Closure-freedom. No administrative redexes. No boxing in join points.
Only boxing in monadic code comes from ``down''. No code blowup.

\section{Fusion}\label{sec:push-pull-fusion}

Motivation.

\subsection{Push streams}\label{sec:push-streams}

Definition.

\subsection{Pull streams}\label{sec:pull-streams}

Pull streams are definitely not monadic. \texttt{Nat -> Stream a} contains
``infinite'' amount of code internally.

Pull streams are also not selective functors.

We do have monadic binding for lifted types, plus ``case'' matching
on arbitrary object values! This is practically almost as good as
having unlimited monadic bind! We need to assume a ``non-dependency axiom'',
that any $f : \Lift A \to \msf{ValTy}$ must be constant.

\todo{Validate non-dependency in the psh model!}
\todo{Add meta-identity type to the psh model}

%% The purpose of staged compilation is to write code-generating programs in a
%% safe, ergonomic and expressive way. It is always possible to do ad-hoc code
%% generation, by simply manipulating strings or syntax trees in a sufficiently
%% expressive programming language. However, these approaches tend to suffer from
%% verbosity, non-reusability and lack of safety. In staged compilation, there are
%% certain \emph{restrictions} on which metaprograms are expressible. Usually,
%% staged systems enforce typing discipline, prohibit arbitrary manipulation of
%% object-level scopes, and often they also prohibit accessing the internal
%% structure of object expressions. On the other hand, we get \emph{guarantees}
%% about the well-scoping or well-typing of the code output, and we are also able
%% to use concise syntax for embedding object-level code.

%% \emph{Two-level type theory}, or 2LTT in short, was described in
%% \cite{twolevel}, building on ideas from \cite{hts}. The motivation was to allow
%% convenient metatheoretical reasoning about a certain mathematical language
%% (homotopy type theory), and to enable concise and modular ways to extend the
%% language with axioms.

%% It turns out that metamathematical convenience closely corresponds to
%% metaprogramming convenience: 2LTT can be directly and effectively employed in
%% staged compilation. Moreover, semantic ideas underlying 2LTT are also directly
%% applicable to the theory of staging.

%% \subsection{Overview \& Contributions}\label{sec:overview}

%% \begin{itemize}
%%   \item In Section \ref{sec:tour-of-2ltt} we present an informal syntax of
%%     two-level type theory, a dependent type theory with support for two-stage
%%     compilation. We look at basic use cases involving inlining control, partial
%%     evaluation and fusion optimizations. We also describe feature variations,
%%     enabling applications in monomorphization and memory layout control.
%%   \item In Section \ref{sec:formal-2ltt}, we specify the formal notion of models
%%     and syntax for the variant of 2LTT used in this paper. We mostly work with a
%%     high-level algebraic presentation, where the syntax is quotiented by
%%     conversion. However, we also explain how to extract functions
%%     operating on non-quotiented syntax, by interpreting type-theoretic
%%     constructions in a setoid model.
%%   \item
%%     In Section \ref{sec:staging-algorithm}, we review the standard presheaf
%%     model of 2LTT \cite[Section~2.5.3]{twolevel}, which lies over the syntactic
%%     category of the object theory. We show that evaluation in this model yields
%%     a staging algorithm for closed types and terms. We then extend staging to
%%     open types and terms which may depend on object-level variables. We show
%%     stability of staging, which means that staging is surjective up to
%%     conversion, and we also show that staging strictly preserves all type and
%%     term formers. Finally, we discuss efficiency and potential practical implementations
%%     of staging.
%%   \item
%%     In Section \ref{sec:soundness} we show soundness of staging, which roughly
%%     means that the output of staging is convertible to its input. Staging
%%     together with its stability and soundness can be viewed as a \emph{strong
%%     conservativity} theorem of 2LTT over the object theory. This means that the
%%     same constructions can be expressed in the object theory and the
%%     object-level fragment of 2LTT, up to conversion, and staging witnesses that
%%     meta-level constructions can be always computed away. This strengthens the
%%     weak notion of conservativity shown in \cite{capriotti2017models} and
%%     \cite{twolevel}.
%%   \item
%%     In Section \ref{sec:intensional-analysis}, we discuss possible semantic
%%     interpretations of intensional code analysis, i.e.\ the ability to look into
%%     the structure of object-level code.
%%   \item We provide a small standalone implementation of elaboration and staging
%%     for a two-level type theory \cite{staged-demo}.
%%   \item To our knowledge, this work is the first formalization and implementation of
%%     staged compilation in the presence of full dependent types, with
%%     universes and large elimination. In particular, we allow unrestricted
%%     staging for types, so that types can be computed by metaprograms at compile
%%     time.
%% \end{itemize}

%% \section{A Tour of Two-Level Type Theory}\label{sec:tour-of-2ltt}

%% In this section, we provide a short overview of 2LTT and its potential
%% applications in staging. We work in the informal syntax of a dependently typed
%% language which resembles Agda \cite{agdadocs}. We focus on examples and informal
%% explanation here; the formal details will be presented in Section \ref{sec:formal-2ltt}.

%% \begin{notation}\label{basic-notation}
%% We use the following notations throughout the paper. $(x : A) \to B$ denotes a
%% dependent function type, where $x$ may occur in $B$. We use $\lambda\,x.\,t$ for
%% abstraction. A $\Sigma$-type is written as $(x : A) \times B$, with pairing as
%% $(t,\,u)$, projections as $\fst$ and $\snd$, and we may use pattern matching
%% notation on pairs, e.g.\ as in $\lambda\,(x,\,y).\,t$. The unit type is $\top$
%% with element $\tt$. We will also use Agda-style notation for implicit arguments,
%% where $t : \{x : A\} \to B$ implies that the first argument to $t$ is inferred
%% by default, and we can override this by writing a $t \{u\}$ application. We may
%% also implicitly quantify over arguments (in the style of Idris and Haskell), for
%% example when declaring $\mit{id} : A \to A$ with the assumption that $A$ is
%% universally quantified.
%% \end{notation}

%% \subsection{Rules of 2LTT}

%% \subsubsection{Universes}
%% We have universes $\U_{i,j}$, where $i \in \{0,1\}$, and $j \in \mbb{N}$.  The
%% $i$ index denotes stages, where $0$ is the runtime (object-level) stage, and $1$
%% is the compile time (meta-level) stage. The $j$ index denotes universe sizes in
%% the usual sense of type theory. We assume Russell-style universes, with
%% $\U_{i,j} : \U_{i, j+1}$. However, for the sake of brevity we will usually omit
%% the $j$ indices in this section, as sizing is orthogonal to our use-cases and
%% examples.
%% \begin{itemize}
%% \item $\U_0$ can be viewed as the \emph{universe of object-level or runtime types}.
%%     Each closed type $A : \U_0$ can be staged to an actual type in the object language
%%     (the language of the staging output).
%%   \item $\U_1$ can be viewed as the \emph{universe of meta-level or static types}. If we
%%     have $A : \U_1$, then $A$ is guaranteed to be only present at compile time,
%%     and will be staged away. Elements of $A$ are likewise computed away.
%% \end{itemize}

%% \subsubsection{Type Formers} $\U_0$ and $\U_1$ may be closed under arbitrary type formers,
%% such as functions, $\Sigma$-types, identity types or inductive types in general.
%% However, all constructors and eliminators in type formers must stay at the same
%% stage. For example:
%% \begin{itemize}
%%   \item Function domain and codomain types must be at the same stage.
%%   \item If we have $\Nat_0 : \U_0$ for the runtime type of natural numbers,
%%         we can only map from it to a type in $\U_0$ by recursion or induction.
%% \end{itemize}
%% It is not required that we have the \emph{same} type formers at both stages. We
%% will discuss variations of the object-level languages in Section \ref{sec:variations}.

%% \subsubsection{Moving Between Stages}
%% At this point, our system is rather limited, since there is no interaction
%% between the stages. We enable such interaction through the following operations.
%% Note that none of these operations can be expressed as functions, since function
%% types cannot cross between stages.
%% \begin{itemize}
%% \item \emph{Lifting:} for $A : \U_0$, we have $\Lift A : \U_1$.  From the
%%   staging point of view, $\Lift A$ is the type of metaprograms which compute
%%   runtime expressions of type $A$.
%% \item \emph{Quoting:} for $A : \U_0$ and $t : A$, we have $\qut{t} :\,\Lift A$.
%%   A quoted term $\qut{t}$ represents the metaprogram which immediately yields
%%   $t$.
%% \item \emph{Splicing:} for $A : \U_0$ and $t :\,\Lift A$, we have $\spl t : A$.
%%   During staging, the metaprogram in the splice is executed, and the resulting
%%   expression is inserted into the output.

%%   \begin{notation} Splicing binds stronger than any operation, including function
%%     application. For instance, $\spl f\,x$ is parsed as $(\spl f)\,x$. We borrow
%%     this notation from MetaML \cite{metaml}.
%%   \end{notation}

%% \item Quoting and splicing are definitional inverses, i.e.\ we have $\spl\qut{t} = t$ and
%%   $\qut{\spl t}=t$ as definitional equalities.
%% \end{itemize}

%% Informally, if we have a closed program $t : A$ with $A : \U_0$, \emph{staging}
%% means computing all metaprograms and recursively replacing all splices in $t$ and $A$
%% with the resulting runtime expressions. The rules of 2LTT ensure that this is possible,
%% and we always get a splice-free object program after staging.

%% \paragraph{Remark}
%% Why do we use the index $0$ for the runtime stage? The reason is that it is not
%% difficult to generalize 2LTT to multi-level type theory, by allowing to lift
%% types from $\U_i$ to $\U_{i+1}$. In the semantics, this can be modeled by
%% having a 2LTT whose object theory is once again a 2LTT, and doing this in an
%% iterated fashion. But there must be necessarily a bottom-most object theory;
%% hence our stage indexing scheme. For now though, we leave the multi-level
%% generalization to future work.

%% \begin{notation}\label{notation:basics} We may disambiguate type formers at different stages by using $0$ or $1$
%% subscripts. For example, $\Nat_1 : \U_1$ is distinguished from $\Nat_0 : \U_0$,
%% and likewise we may write $\zero_0 : \Nat_0$ and so on. For function and
%% $\Sigma$ types, the stage is usually easy to infer, so we do not annotate
%% them. For example, the type $\Nat_0 \to \Nat_0$ must be at the runtime stage,
%% since the domain and codomain types are at that stage, and we know that the
%% function type former stays within a single stage. We may also omit stage annotations
%% from $\lambda$ and pairing.
%% \end{notation}

%% \subsection{Staged Programming in 2LTT}\label{sec:staged-programming}

%% In 2LTT, we may have several different polymorphic identity functions. First,
%% consider the usual identity function at each stage:
%% \begin{alignat*}{3}
%%   & \mit{id}_0 : (A : \U_0) \to A \to A\hspace{2em} && \mit{id}_1 : (A : \U_1) \to A \to A\\
%%   & \mit{id}_0 := \lambda\,A\,x.x       && \mit{id}_1 := \lambda\,A\,x.x
%% \end{alignat*}
%% An $\mit{id}_0$ application will simply appear in staging output as it is. In
%% contrast, $\mit{id}_1$ can be used as a compile-time evaluated function, because
%% the staging operations allow us to freely apply $\mit{id}_1$ to runtime
%% arguments. For example, $\mit{id}_1\,(\Lift\,\Bool_0)\,\qut{\true_0}$ has type
%% $\Lift \Bool_0$, which implies that $\spl(\mit{id}_1\,(\Lift\,\Bool_0)\,\qut{\true_0})$
%% has type $\Bool_0$. We can stage this expression as follows:
%% \[
%% \spl(\mit{id}_1\,(\Lift\,\Bool_0)\,\qut{\true_0}) = \spl\qut{\true_0} = \true_0
%% \]
%% There is another identity function, which computes at compile time, but which
%% can be only used on runtime arguments:
%% \begin{alignat*}{3}
%%   & \mit{id_\Lift} : (A : \Lift\U_0) \to \Lift\,\spl A \to \Lift\,\spl A\\
%%   & \mit{id_\Lift} := \lambda\,A\,x.x
%% \end{alignat*}
%% Note that since $A : \Lift\U_0$, we have $\spl A : \U_0$, hence $\Lift\,\spl A
%% : \U_1$.  Also, $\Lift\U_0 : \U_1$, so all function domain and codomain types in
%% the type of $\mit{id_\Lift}$ are at the same stage. Now, we may write
%% $\spl(\mit{id_\Lift}\,\qut{\Bool_0}\,\qut{\true_0})$ for a term which is staged
%% to $\true_0$. In this specific case $\mit{id_\Lift}$ has no practical advantage
%% over $\mit{id}_1$, but in some cases we really have to quantify over
%% $\Lift\U_0$. This brings us to the next example.

%% Assume $\List_0 : \U_0 \to \U_0$ with $\nil_0 : (A : \U_0) \to \List_0\,A$,
%% $\cons_0 : (A : \U_0) \to A \to \List_0\,A$ and $\msf{foldr}_0 : (A\,B : \U_0)
%% \to (A \to B \to B) \to B \to \List_0\,A \to B$. We define a mapping function which
%% inlines its function argument:
%% \begin{alignat*}{3}
%%   & \mit{map} : (A\,B : \Lift\U_0) \to (\Lift\,\spl A \to \Lift\,\spl B)
%%       \to \Lift(\List_0\,\spl A) \to \Lift(\List_0\,\spl B)\\
%%   & \mit{map} := \lambda\,A\,B\,f\,\mit{as}.\,
%%       \qut{\msf{foldr}_0\,
%%         \spl A\,(\List_0 \spl B)\,
%%         (\lambda\,a\,\mit{bs}.\, \cons_0\,\spl B\,\spl(f\,\qut{a})\,\mit{bs})\,
%%         (\nil_0\,\spl B)\,
%%         \spl as
%%         }
%% \end{alignat*}
%% This $\mit{map}$ function can be defined with quantification over $\Lift \U_0$ but not
%% over $\U_1$, because $\List_0$ expects type parameters in $\U_0$, and there is
%% no generic way to convert from $\U_1$ to $\U_0$. Now, assuming
%% $\blank\!+_0\!\blank : \Nat_0 \to \Nat_0 \to \Nat_0$ and $\mit{ns} :
%% \List_0\,\Nat_0$, we have the following staging behavior:
%% \begin{alignat*}{3}
%%   & &&\spl(\mit{map}\,\qut{\Nat_0}\,\qut{\Nat_0}\,(\lambda\,n.\,\qut{\spl n +_0 10})\,\qut{\mit{ns}})\\
%%   & =\hspace{0.3em}&&\spl\qut{\msf{foldr}_0\,
%%         \spl \qut{\Nat_0}\,(\List_0\,\spl\qut{\Nat_0})\\
%%   & && \hspace{3.5em}(\lambda\,a\,\mit{bs}.\, \cons_0\,\spl\qut{\Nat_0}\,\spl\qut{\spl\qut{a} +_0 10}\,\mit{bs})\,
%%         (\nil_0\,\spl \qut{\Nat_0})\,
%%         \spl\qut{\mit{ns}}}\\
%%   & =\hspace{0.3em} &&\msf{foldr_0}\,\Nat_0\,(\List_0\,\Nat_0)\,\,(\lambda\,a\,bs.\,\cons_0\,\Nat_0\,(a +_0 10)\,bs)\,(\nil_0\,\Nat_0)\,\mit{ns}
%% \end{alignat*}

%% By using meta-level functions and lifted types, we already have control over
%% inlining. However, if we want to do more complicated meta-level computation, it
%% is convenient to use recursion or induction on meta-level type formers. A
%% classic example in staged compilation is the power function for natural numbers,
%% which evaluates the exponent at compile time. We assume the iterator function
%% $\msf{iter_1} : \{A : \U_1\} \to \Nat_1 \to (A \to A) \to A \to A$, and runtime
%% multiplication as $\blank\!*_0\!\blank$.
%% \begin{alignat*}{3}
%%   &\mit{exp} : \Nat_1 \to \Lift \Nat_0 \to \Lift \Nat_0 \\
%%   &\mit{exp} := \lambda\,x\,y.\,
%%   \msf{iter}_1\,x\,(\lambda\,n.\,\qut{\spl y *_0 \spl n})\,\qut{1}
%% \end{alignat*}
%% Now, $\spl(\mit{exp}\,3\,\qut{n})$ stages to $n *_0 n *_0 n *_0 1$ by the computation rules
%% of $\msf{iter_1}$ and the staging operations.

%% We can also stage \emph{types}. Below, we use iteration to compute the type of
%% vectors with static length, as a nested pair type.
%% \begin{alignat*}{3}
%%   &\mit{Vec} : \Nat_1 \to \Lift \U_0 \to \Lift \U_0\\
%%   &\mit{Vec} := \lambda\,n\,A.\,\msf{iter}_1\,n\,(\lambda\,B.\,\qut{\spl A \times \spl B})\,\qut{\top_0}
%% \end{alignat*}
%% With this definition, $\spl(\mit{Vec}\,3\,\qut{\Nat_0})$ stages to $\Nat_0
%% \times (\Nat_0 \times (\Nat_0 \times \top_0))$. Now, we can use \emph{induction}
%% on $\Nat_1$ to implement a map function. For readability, we use an Agda-style
%% pattern matching definition below (instead of the elimination principle).
%% \begin{alignat*}{6}
%%   &\hspace{-3em}\rlap{$\mit{map} : (n : \Nat_1) \to (\Lift\,\spl A \to \Lift\,\spl B) \to \Lift(\mit{Vec}\,n\,A) \to \Lift(\mit{Vec}\,n\,B)$}\\
%%   &\hspace{-3em}\mit{map}\,\,&&\zero_1      && f\,&&\mit{as} := \qut{\tt_0}\\
%%   &\hspace{-3em}\mit{map}\,\,&&(\suc_1\,n)\,&& f\,&&\mit{as} :=
%%      \qut{(\spl(f\,\qut{\fst_0\,\spl\mit{as}}),\,\spl(\mit{map}\,n\,f\,\qut{\snd_0\,\spl \mit{as}}))}
%% \end{alignat*}
%% This definition inlines the mapping function for each projected element of the
%% vector. For instance, staging $\spl(\mit{map}\,2\,(\lambda\,n.\,\qut{\spl n +_0
%%   10})\,\qut{\mit{ns}})$ yields $(\fst_0\,\mit{ns} +_0
%% 10,\,(\fst_0(\snd_0\,\mit{ns}) +_0 10,\,\tt_0))$. Sometimes, we do not want to
%% duplicate the code of the mapping function. In such cases, we can use
%% \emph{let-insertion}, a common technique in staged compilation. If we bind a runtime
%% expression to a runtime variable, and only use that variable in subsequent
%% staging, only the variable itself can be duplicated. One solution is to do an
%% ad-hoc let-insertion:
%% \begin{alignat*}{6}
%%   &   && \mylet_0\,f := \lambda\,n.\, n +_0 10\,\,\msf{in}\,\,
%%          \spl(\mit{map}\,2\,(\lambda\,n.\,\qut{f\,\spl n})\,\qut{\mit{ns}}) \\
%%   & =\,\,&&  \mylet_0\,f := \lambda\,n.\, n +_0 10\,\,\msf{in}\,\,
%%           (f\,(\fst_0\,\mit{ns}),\,(f\,(\fst_0(\snd_0\,\mit{ns})),\,\tt_0))
%% \end{alignat*}
%% We include examples of more sophisticated let-insertion in the supplementary
%% code \cite{staged-demo}.

%% More generally, we are free to use dependent types at the meta-level, so we can
%% reproduce more complicated staging examples. Any well-typed interpreter can be
%% rephrased as a \emph{partial evaluator}, as long as we have sufficient type
%% formers. For instance, we may write a partial evaluator for a simply typed
%% lambda calculus. We sketch the implementation in the following; the full version
%% can be found in the supplementary code. First, we inductively define types,
%% contexts and terms:
%% \begin{alignat*}{6}
%%   & \Ty : \U_1  \hspace{2em} \Con : \U_1 \hspace{2em} \Tm : \Con \to \Ty \to \U_1
%% \end{alignat*}
%% Then we define the interpretation functions:
%% \begin{alignat*}{6}
%%   & \msf{EvalTy}  &&: \Ty \to \Lift \U_0 \\
%%   & \msf{EvalCon} &&: \Con \to \U_1 \\
%%   & \msf{EvalTm}  &&: \Tm\,\Gamma\,A \to \msf{EvalCon}\,\Gamma \to \Lift\,\spl(\msf{EvalTy}\,A)
%% \end{alignat*}
%% Types are necessarily computed to runtime types, e.g.\ an embedded
%% representation of the natural number type is evaluated to $\qut{\Nat_0}$.
%% Contexts are computed as follows:
%% \begin{alignat*}{4}
%%   &\msf{EvalCon}\,\,\msf{empty}                &&:= \top_1 \\
%%   &\msf{EvalCon}\,\,(\msf{extend}\,\Gamma\,A) &&:= \msf{EvalCon}\,\Gamma \times (\Lift \spl(\msf{EvalTy}\,A))
%% \end{alignat*}
%% This is an example for the usage of \emph{partially static data} \cite{partial-evaluation}:
%% semantic contexts are \emph{static} lists storing \emph{runtime}
%% expressions. This allows us to completely eliminate environment lookups in the
%% staging output: an embedded lambda expression is staged to the corresponding
%% lambda expression in the object language. This is similar to the partial
%% evaluator presented in Idris 1 \cite{scrap-your-inefficient-engine}. However, in
%% contrast to 2LTT, Idris 1 does not provide a formal guarantee that partial
%% evaluation does not get stuck.

%% \subsection{Properties of Lifting, Binding Time Improvements}

%% We describe more generally the action of lifting on type formers. While lifting
%% does not have definitional computation rules, it does preserve negative type formers up
%% to definitional isomorphism \cite[Section~2.3]{twolevel}:
%% \begin{alignat*}{5}
%%   \Lift((x : A) \to B\,x) &\simeq ((x : \Lift A) \to \Lift\,(B\,\spl x))\\
%%   \Lift ((x : A) \times B\,x) &\simeq ((x : \Lift A) \times \Lift (B\,\spl x))\\
%%   \Lift \top_0 &\simeq \top_1
%% \end{alignat*}
%% For function types, the preservation maps are the following:
%% \begin{alignat*}{5}
%%   & \hspace{-6em}\rlap{$\mit{pres}_\to : \Lift((x : A) \to B\,x) \to ((x : \Lift A) \to \Lift\,(B\,\spl x))$}\\
%%   & \hspace{-6em}\mit{pres}_\to\,f     &&:= \lambda\,x.\,\,\qut{\spl f\,\spl x}\\
%%   & \hspace{-6em}\mit{pres}_\to^{-1}\,f &&:= \qut{\lambda\,x.\,\,\spl(f\,\qut{x})}
%% \end{alignat*}
%% With this, we have that $\mit{pres}_\to\,(\mit{pres}_\to^{-1}\,f)$ is
%% definitionally equal to $f$, and also the other way around. Preservation maps
%% for $\Sigma$ and $\top$ work analogously.

%% By rewriting a 2LTT program left-to-right along preservation maps, we perform
%% what is termed \emph{binding time improvement} in the partial evaluation
%% literature \cite[Chapter~12]{partial-evaluation}. Note that the output of $\mit{pres}_{\to}$ uses a
%% meta-level $\lambda$, while going the other way introduces a runtime
%% binder. Meta-level function types and $\Sigma$-types support more computation during
%% staging, so in many cases it is beneficial to use the improved forms. In some
%% cases though we may want to use unimproved forms, to limit the size of generated
%% code. This is similar to what we have seen with let-insertion. For a minimal
%% example, consider the following unimproved version of $\mit{id}_\Lift$:
%% \begin{alignat*}{5}
%%   & \mit{id}_\Lift : (A : \Lift \U_0) \to \Lift(\spl A \to \spl A) \\
%%   & \mit{id}_\Lift := \lambda\,A.\,\,\qut{\lambda\,x.\,x}
%% \end{alignat*}
%% This can be used at the runtime stage as
%% $\spl(\mit{id}_\Lift\,\qut{\Bool_0})\,\true_0$, which is staged to
%% $(\lambda\,x.\,x)\,\true_0$.  This introduces a useless $\beta$-redex, so in
%% this case the improved version is clearly preferable.

%% For inductive types in general we do not get full preservation, only maps in one
%% direction. For example, we have $\Bool_1 \to \Lift \Bool_0$, defined as
%% $\lambda\,b.\,\msf{if}\,b\,\msf{then}\,\qut{\true_0}\,\msf{then}\,\qut{\false_0}$.
%% In the staging literature, this is called ``serialization'' or ``lifting''
%% \cite{template-haskell,metaml}.  In the other direction, we can only define
%% constant functions from $\Lift \Bool_0$ to $\Bool_1$.

%% The lack of elimination principles for $\Lift A$ means that we cannot inspect
%% the internal structure of expressions. We will briefly discuss ways to lift this
%% restriction in Section \ref{sec:intensional-analysis}.

%% In particular, we have no serialization map from $\Nat_1 \to \Nat_1$ to
%% $\Lift(\Nat_0 \to \Nat_0)$. However, when $A : \U_1$ is \emph{finite}, and $B :
%% \U_1$ can be serialized, then $A \to B$ can be serialized, because it is
%% equivalent to a finite product. For instance, $\Bool_1 \to \Nat_1 \simeq \Nat_1
%% \times \Nat_1$.  In 2LTT, $A$ is called \emph{cofibrant}
%% \cite[Section~3.4]{twolevel}: this means that for each $B$, $A \to \Lift\,B$ is
%% equivalent to $\Lift C$ for some $C$. This is the 2LTT formalization of the
%% so-called ``trick'' in partial evaluation, which improves binding times by
%% $\eta$-expanding functions out of finite sums \cite{eta-expansion-trick}.

%% \subsubsection{Fusion}
%% Fusion optimizations can be viewed as binding time improvement techniques for
%% general inductive types. The basic idea is that by lambda-encoding an inductive
%% type, it is brought to a form which can be binding-time improved. For instance,
%% consider foldr-build fusion for lists \cite{short-cut}, which is employed in GHC
%% Haskell.  Starting from $\Lift (\List_0\,A)$, we use Böhm-Berarducci encoding
%% \cite{boehm-berarducci} under the lifting to get
%% \begin{alignat*}{3}
%%          &\Lift((L : \U_0) \to (A \to L \to L) \to L \to L)\\
%%   \simeq\,\,&((L : \Lift\,U_0) \to (\Lift\,A \to \Lift\,\spl L \to \Lift\,\spl L) \to \Lift\,\spl L \to \Lift\,\spl L).
%% \end{alignat*}
%% Alternatively, for \emph{stream fusion} \cite{stream-fusion}, we embed $\List\,A$ into the
%% coinductive colists (i.e.\ the possibly infinite lists), and use a terminal
%% lambda-encoding. The embedding into the ``larger'' structure enables some staged
%% optimizations which are otherwise not possible, such as fusion for the
%% $\msf{zip}$ function. However, the price we pay is that converting
%% back to lists from colists is not necessarily total.

%% We do not detail the implementation of fusion in 2LTT here; a small example can
%% be found in the supplementary code. In short, 2LTT is a natural setting for a
%% wide range of fusion setups. A major advantage of fusion in 2LTT is the formal
%% guarantee of staging, in contrast to implementations where compile-time
%% computation relies on ad-hoc user annotations and general-purpose optimization
%% passes. For instance, fusion in GHC relies on rewrite rules and inlining
%% annotations which have to be carefully tuned and ordered, and it is possible to
%% get pessimized code via failed fusion.

%% \subsubsection{Inferring Staging Operations}
%% We can extract a coercive subtyping system from the staging operations, which
%% can be used for inference, and in particular for automatically transporting
%% definitions along preservation isomorphisms. One choice is to have $A \leq \Lift
%% A$, $\Lift A \leq A$, a contravariant-covariant rule for functions and a
%% covariant rule for $\Sigma$. During bidirectional elaboration, when we need to
%% compare an inferred and an expected type, we can insert coercions.  We
%% implemented this feature in our prototype. It additionally supports Agda-style
%% implicit arguments and pattern unification, so it can elaborate the following
%% definition:
%% \begin{alignat*}{3}
%%   & \mit{map} : \{A\,B : \Lift\U_0\} \to (\Lift A \to \Lift B)
%%       \to \Lift(\List_0\,A) \to \Lift(\List_0\,B)\\
%%   & \mit{map} := \lambda\,f\,\mit{as}.\,
%%       \msf{foldr}_0
%%         (\lambda\,a\,\mit{bs}.\, \cons_0\,(f\,a)\,\mit{bs})\,
%%         \nil_0\,
%%         as
%% \end{alignat*}
%% We may go a bit further, and also add the coercive subtyping rule $\U_0 \leq
%% \U_1$, witnessed by $\Lift$. Then, the type of $\mit{map}$ can be written as
%% $\{A\,B : \Lift\U_0\} \to (A \to B) \to \List_0\,A \to \List_0\,B$. However,
%% here the elaborator has to make a choice, whether to elaborate to improved or
%% unimproved types. In this case, the fully unimproved type would be
%% \[ \{A\,B : \Lift\U_0\} \to \Lift((\spl A \to \spl B) \to \List_0\,\spl A \to \List_0\,\spl B). \]
%% It seems that improved types are a sensible default, and we can insert explicit
%% lifting when we want to opt for unimproved types. This is also available in our
%% prototype.

%% \subsection{Variations of Object-Level Languages}
%% \label{sec:variations}

%% In the following, we consider variations on object-level languages, with a focus
%% on applications in downstream compilation after staging. Adding restrictions or
%% more distinctions to the object language can make it easier to optimize and
%% compile.

%% \subsubsection{Monomorphization}\label{sec:monomorphization}

%% In this case, the object language is simply typed, so every type is known
%% statically. This makes it easy to assign different memory layouts to different
%% types and generate code accordingly for each type. Moving to 2LTT, we still
%% want to abstract over runtime types at compile time, so we use the following
%% setup.
%% \begin{itemize}
%% \item We have a \emph{jugdment}, written as $A\,\msf{type_0}$, for well-formed
%%   runtime types. Runtime types may be closed under simple type formers.
%% \item We have a type $\Ty_0 : \U_1$ in lieu of the previous $\Lift \U_0$.
%% \item For each $A\,\msf{type}_0$, we have $\Lift A : \Ty_0$.
%% \item We have quoting and splicing for types and terms. For types, we send $A\,
%%   \msf{type}_0$ to $\qut{A} : \Ty_0$. For terms, we send $t : A$ to $\qut{t} :
%%   \Lift A$.
%% \end{itemize}
%% Despite the restriction to simple types at runtime, we can still write arbitrary
%% higher-rank polymorphic functions in 2LTT, such as a function with type $((A : \Ty_0)
%% \to \Lift\,\spl A \to \Lift \spl A) \to \Lift\,\Bool_0$. This function can be
%% only applied to statically known arguments, so the polymorphism can be staged
%% away. The main restriction that programmers have to keep in mind is that
%% polymorphic functions cannot be stored inside runtime data types.

%% %% This is similar to the well-known monomorphization models in the C++ and
%% %% Rust programming languages. The evident difference is that in 2LTT we can use a
%% %% full type theory at compile time. Additionally, we will see in Section
%% %% \ref{sec:intensional-analysis} that this setup is compatible with intensional
%% %% analysis, the ability to analyze the structure of object types and terms.

%% \subsubsection{Memory Representation Polymorphism}

%% This refines monomorphization, so that types are not directly identified with
%% memory representations, but instead representations are internalized in 2LTT as
%% a meta-level type, and runtime types are indexed over representations.
%% \begin{itemize}
%% \item We have $\Rep : \U_1$ as the type of memory representations. We have
%%   considerable freedom in the specification of $\Rep$. A simple setup may
%%   distinguish references from unboxed products, i.e.\ we have $\msf{Ref} : \Rep$
%%   and $\msf{Prod} : \Rep \to \Rep \to \Rep$, and additionally we may assume
%%   any desired primitive machine representation as a value of $\Rep$.
%% \item We have Russell-style $\U_{0,j} : \Rep \to \U_{0, j+1}\,r$, where $r$ is
%%   some chosen runtime representation for types; usually we would mark types are
%%   erased. We leave the meta-level $\U_{1,j}$ hierarchy unchanged.
%% \item We may introduce unboxed $\Sigma$-types and primitive machine types in the
%%   runtime language. For $r : \Rep$, $r' : \Rep$, $A : \U_{0}\,r$ and $B : A \to
%%   \U_{0}\,r'$, we may have $(x : A) \times B\,x :
%%   \U_{0}\,(\msf{Prod\,r\,r'})$. Thus, we have type dependency, but we do not
%%   have dependency in memory representations.
%% \end{itemize}
%% Since $\Rep$ is meta-level, there is no way to abstract over it at runtime, and
%% during staging all $\Rep$ indices are computed to concrete canonical
%% representations. This is a way to reconcile dependent types with some amount of
%% control over memory layouts. The unboxed flavor of $\Sigma$ ends up with a
%% statically known flat memory representation, computed from the representations
%% of the fields.

%% %% In principle, it should be possible to have \emph{dependent memory layouts};
%% %% this would be a more advanced variation on type-passing polymorphism
%% %% \cite{intensional-poly}, where layouts of objects may depend on runtime data. For example,
%% %% length-prefixed flat arrays would not be a primitive notion, but simply defined
%% %% as $(n : \Nat) \times \msf{Vec}\,n\,A$. However, it seems to be highly
%% %% challenging to implement runtime systems and precise garbage collection in this
%% %% setting. See \texttt{sixten} \cite{sixten} for an experimental implementation of
%% %% dependent memory layouts.

%% \section{Formal Setup}\label{sec:formal-2ltt}

%% In this section we describe our approach to formalizing staging and also the
%% \emph{extraction} of algorithms from the formalization. There is a tension
%% here:
%% \begin{itemize}
%%   \item On one hand, we would like to do formal work at a higher level of
%%     abstraction, in order to avoid a deluge of technical details (for which
%%     dependent type theories are infamous).
%%   \item On the other hand, at some point we need to talk about lower-level
%%     operational aspects and extracted algorithms.
%% \end{itemize}
%% In the following, we describe our approach to interfacing between the different levels
%% of abstraction.

%% \subsection{The Algebraic Setting}

%% When defining staging and proving its soundness and stability, we work
%% internally in a constructive type theory, using types and terms that are
%% quotiented by conversion. In this setting, every construction respects syntactic
%% conversion, which is very convenient technically. Here, we use standard
%% definitions and techniques from the categorical and algebraic metatheory of type
%% theories, and reuse prior results which rely on the same style of
%% formalization. Concretely, we use \emph{categories with families} \cite{cwfs}
%% which is a commonly used specification of an \emph{explicit substitution
%% calculus} of types and terms.

%% In the rest of the paper we will mostly work in this setting, and we will
%% explicitly note when we switch to a different metatheory. We summarize
%% the features of this metatheory in the following.

%% \paragraph{Metatheory}
%% The metatheory is an intensional type theory with a cumulative universe
%% hierarchy $\Set_i$, where $i$ is ordinal number such that $i \leq \omega +
%% 1$. The transfinite $i$ levels are a convenience feature for modeling typing
%% contexts, where sometimes we will need universes large enough to fit types at
%% arbitrary finite levels. We also assume uniqueness of identity proofs and
%% function extensionality. We will omit transports along identity proofs in the
%% paper, for the sake of readability. The missing transports can be in principle
%% recovered due to Hofmann's conservativity result \cite{hofmann95extensional};
%% see also \cite{DBLP:conf/cpp/WinterhalterST19}. We assume $\Pi$, $\Sigma$,
%% $\top$ and $\mbb{N}$ types in all universes, and reuse Notation
%% \ref{notation:basics}. We also assume certain quotient inductive-inductive
%% types, to represent the syntax of 2LTT and the object theory. We will detail
%% these shortly in Sections \ref{sec:2lttmod} and \ref{sec:objmod}.

%% \subsection{Algorithm Extraction}

%% In the algebraic setting we can \emph{only} talk about properties which
%% respect syntactic conversion. However, the main motivation of staged compilation is to
%% improve performance and code abstraction, but these notions do \emph{not}
%% respect conversion. For instance, rewriting programs along $\beta$-conversion
%% can improve performance. Hence, at some point we may want to ``escape'' from
%% quotients.

%% Fortunately, our specified meta type theory can be interpreted in
%% \emph{setoids}, as most recently explained in
%% \cite{DBLP:journals/pacmpl/PujetT22}. A setoid is a set together with an
%% equivalence relation on it. We interpret closed types as setoids, and dependent
%% types as setoid fibrations. For inductive base types such $\Bool$ or $\Nat$, the
%% equivalence relation is interpreted as identity. Functions are interpreted as
%% equivalence-preserving functions, and two functions are equivalent if they are
%% pointwise equivalent. A quotient type $A/R$ is interpreted by extending the
%% semantic notion of equivalence for $A$ with the given $R$ relation. We will
%% sketch the interpretation of the 2LTT syntax (given as a quotient
%% inductive-inductive type) in Section \ref{sec:2lttsetoid}.

%% In particular, from $f : A/R \to A/R$ we can extract a function which acts on
%% the underlying set of $A$'s interpretation. Note that an extracted function can
%% be only viewed as an algorithm if the setoid interpretation is itself defined in
%% a constructive theory, but that is the case in the cited work
%% \cite{DBLP:journals/pacmpl/PujetT22}.

%% \begin{definition} Assume a closed function $f : A \to B$ defined
%% in the mentioned type theory. The \textbf{extraction} of $f$ is
%% the underlying function of $f$'s interpretation in the setoid model.
%% \end{definition}

%% In other words, we can view quotients as a convenient shorthand for working with
%% explicit equivalence relations. We can work with quotients when we want to enforce
%% preservation of equivalence relations, but at any point we have the option to
%% switch to an external view and look at underlying sets and functions.

%% %% In other words, we can view the usage of constructive quotients as a convenient
%% %% shorthand for working with explicit equivalence relations. When we reason about
%% %% the syntax of a type theory, this provides a bridge between the
%% %% categorical/algebraic style and the traditional style that uses separate
%% %% equivalence relations.
%% %% It makes sense to work with quotients internally to a type theory whenever we
%% %% only want to talk about constructions and properties which respect
%% %% conversion. However, at any point we have the option to switch to an external
%% %% perspective and look at underlying sets and the underlying maps.

%% \subsection{Models and Syntax of 2LTT}\label{sec:2lttmod}

%% We specify the notion of model for 2LTT in this section. The syntax of 2LTT will
%% be defined as the initial model, as a quotient inductive-inductive type,
%% following \cite{ttintt}. This yields an explicit substitution calculus,
%% quotiented by definitional equality. A major advantage of this representation is
%% that it \emph{only} includes well-typed syntax, so we never have to talk about
%% ``raw'' types and terms.

%% We provide an appendix to this paper, where we present a listing of all rules
%% and equations of 2LTT in two styles: first, in a more traditional style with
%% derivation rules, and also in a very compact manner using a second-order
%% algebraic signature.

%% First, we define the structural scaffolding of 2LTT without type formers, mostly
%% following \cite{twolevel}. The specification is agnostic about the sizes of sets
%% involved, i.e.\ we can model underlying sorts using any metatheoretic $\Set_j$.

%% \begin{definition} A model of \textbf{basic 2LTT} consists of the following.
%% \begin{itemize}
%% \item
%%   A category $\mbbc$ with a terminal object. We denote the set of objects as
%%   $\Con_{\mbbc} : \Set$ and use capital Greek letters starting from $\Gamma$ to
%%   refer to objects. The set of morphisms is $\Sub_{\mbbc} : \Con_{\mbbc} \to
%%   \Con_{\mbbc} \to \Set$, and we use $\sigma$, $\delta$ and so on to refer to
%%   morphisms. The terminal object is written as $\emptycon$ with unique morphism
%%   $\epsilon : \Sub_\mbbc\,\Gamma\,\emptycon$. We omit the $\mbbc$ subscript if
%%   it is clear from context.

%%   In the initial model, an element of $\Con$ is a \emph{typing context}, an
%%   element of $\Sub$ is a \emph{parallel substitution}, $\emptycon$ is the
%%   \emph{empty typing context} and $\epsilon$ is the empty list viewed as a
%%   parallel substitution. The identity substitution $\id : \Sub\,\Gamma\,\Gamma$
%%   maps each variable to itself, and $\sigma \circ \delta$ can be computed
%%   by substituting each term in $\sigma$ (which is a list of terms) with $\delta$.
%% \item
%%   For each $i \in \{0,1\}$ and $j \in \mbb{N}$, we have $\Ty_{i,j} : \Con \to
%%   \Set$ and $\Tm_{i,j} : (\Gamma : \Con) \to \Ty_{i,j}\,\Gamma \to \Set$, as
%%   sets of types and terms. Both types and terms can be substituted:
%%   \begin{alignat*}{3}
%%     & \blank\![\blank\!] &&: \Ty_{i,j}\,\Delta \to \Sub\,\Gamma\,\Delta \to \Ty_{i,j}\,\Gamma \\
%%     & \blank\![\blank\!] &&: \Tm_{i,j}\,\Delta\,A \to (\sigma : \Sub\,\Gamma\,\Delta) \to \Tm_{i,j}\,\Gamma\,(A[\sigma])
%%   \end{alignat*}
%%   Additionally, we have $A[\id] = A$ and $A[\sigma \circ \delta] =
%%   A[\sigma][\delta]$, and we have the same equations for term substitution as
%%   well.

%%   We also have a \emph{comprehension structure}: for each $\Gamma : \Con$
%%   and $A : \Ty_{i,j}\,\Gamma$, we have the extended context $\Gamma \ext A :
%%   \Con$ such that there is a natural isomorphism $\Sub\,\Gamma\,(\Delta\ext A)
%%   \simeq (\sigma : \Sub\,\Gamma\,\Delta) \times
%%   \Tm_{i,j}\,\Gamma\,(A[\sigma])$. We will sometimes use
%%   $\blank\!\ext_0\!\blank$ and $\blank\!\ext_1\!\blank$ to disambiguate
%%   object-level and meta-level context extensions.
%% \item
%%   For each $j$ we have a \emph{lifting structure}, consisting of a natural
%%   transformation $\Lift : \Ty_{0,j}\,\Gamma \to \Ty_{1,j}\,\Gamma$, and an
%%   invertible natural transformation $\qut{\blank} : \Tm_{0,j}\,\Gamma\,A \to
%%   \Tm_{1,j}\,\Gamma\,(\Lift A)$, with inverse $\spl\blank$.
%% \end{itemize}
%% \end{definition}

%% %% In short, for each $i$ and $j$ we have a family structure in the sense of
%% %% categories-with-families \cite{Dybjer96internaltype}, such that there is a
%% %% natural transformation from each $(\Ty_{0,j},\,\Tm_{0,j})$ to $(\Ty_{1,j},\,\Tm_{1,j})$
%% %% which is invertible on terms.

%% The following notions are derivable:
%% \begin{itemize}
%% \item
%%   By moving left-to-right along $\Sub\,\Gamma\,(\Delta\ext A) \simeq (\sigma :
%%   \Sub\,\Gamma\,\Delta) \times \Tm_{i,j}\,\Gamma\,(A[\sigma])$,
%%   and starting from the identity morphism $\id : \Sub\,(\Gamma\ext A)\,(\Gamma\ext A)$, we recover
%%   the \emph{weakening substitution} $\p : \Sub\,(\Gamma\ext A)\,\Gamma$ and the \emph{zero variable}
%%   $\q : \Tm_{i,j}\,(\Gamma\ext A)\,(A[\p])$.
%% \item
%%   By weakening $\q$, we recover a notion of variables as De Bruijn
%%   indices. In general, the $n$-th De Bruijn index is defined as $\q[\p^{n}]$,
%%   where $\p^{n}$ denotes $n$-fold composition of $p$.
%% \item
%%   By moving right-to-left along $\Sub\,\Gamma\,(\Delta\ext A) \simeq (\sigma :
%%   \Sub\,\Gamma\,\Delta) \times \Tm_{i,j}\,\Gamma\,(A[\sigma])$, we recover the
%%   operation which extends a morphism with a term. In the initial model, this
%%   extends a parallel substitution with an extra term, thus justifying the view
%%   of substitutions as lists of terms. We denote the extension operation as
%%   $(\sigma,\,t)$ for $\sigma : \Sub\,\Gamma\,\Delta$ and $t : \Tm_{i,j}\,\Gamma\,(A[\sigma])$.
%% \end{itemize}

%% \begin{notation}
%% De Bruijn indices are rather hard to read, so we will sometimes use nameful notation
%% for binders and substitutions. For example, we may write $\Gamma \ext (x : A)
%% \ext (y : B)$ for a context, and subsequently write $B[x \mapsto t]$ for
%% substituting the $x$ variable for some term $t : \Tm_{i,j}\,\Gamma\,A$. Using
%% nameless notation, we would instead have $B : \Ty_{i,j}\,(\Gamma \ext A)$ and
%% $B[\id,\,t] : \Ty_{i,j}\,\Gamma$; here we recover single substitution by extending the identity
%% substitution $\id : \Sub\,\Gamma\,\Gamma$ with $t$.

%% We may also leave weakening implicit: if a type or term is in a context
%% $\Gamma$, we may use it in an extended context $\Gamma \ext A$ without marking
%% the weakening substitution.
%% \end{notation}

%% \begin{definition}
%% A \textbf{model of 2LTT} is a model of basic 2LTT which supports certain type
%% formers. For the sake of brevity, we specify here only a small collection of
%% type formers. However, we will argue later that our results extend to more
%% general notions of inductive types. We specify type formers in the following. We
%% omit substitution rules for type and term formers; all of these can be specified in
%% an evident structural way, e.g.\ as in $(\Lift A)[\sigma] = \Lift(A[\sigma])$.
%% \begin{itemize}
%% \item \emph{Universes}. For each $i$ and $j$, we have a Coquand-style universe
%%   \cite{coquand2018canonicity} in $\Ty_{i,j}$. This consists of $\U_{i,j} : \Ty_{i,j+1}\,\Gamma$,
%%   together with $\El : \Tm_{i,j+1}\,\Gamma\,\U_{i,j} \to \Ty_{i,j}\,\Gamma$ and $\Code$, where $\Code$
%%   and $\El$ are inverses.
%% \item \emph{$\Sigma$-types}. We have $\Sigma\,(x : A)\,B : \Ty_{i,j}\,\Gamma$ for $A : \Ty_{i,j}\,\Gamma$
%%   and $B : \Ty_{i,j}\,(\Gamma \ext (x : A))$, together with a natural isomorphism consisting of pairing and projections:
%%   \[ \Tm_{i,j}\,\Gamma\,(\Sigma\,(x : A)\,B) \simeq (t : \Tm_{i,j}\,\Gamma\,A) \times \Tm_{i,j}\,\Gamma\,(B[x\mapsto t]) \]
%%   We write $(t,\,u)$ for pairing and $\fst$ and $\snd$ for projections.
%% \item \emph{Function types}.
%%   We have $\Pi\,(x : A)\,B : \Ty_{i,j}\,\Gamma$ for $A : \Ty_{i,j}\,\Gamma$ and
%%   $B : \Ty_{i,j}\,(\Gamma \ext (x : A))$, together with abstraction and application
%%   \begin{alignat*}{3}
%%     & \lam &&: \Tm_{i,j}\,(\Gamma \ext (x : A))\,B \to \Tm_{i,j}\,\Gamma\,(\Pi\,(x : A)\,B) \\
%%     & \app &&: \Tm_{i,j}\,\Gamma\,(\Pi\,(x : A)\,B) \to \Tm_{i,j}\,(\Gamma \ext (x : A))\,B
%%   \end{alignat*}
%%   such that $\lam$ and $\app$ are inverses. Since we have explicit
%%   substitutions, this specification of $\app$ is equivalent to the
%%   ``traditional'' one:
%%   \[ \app' : (t : \Tm_{i,j}\,\Gamma\,(\Pi\,(x :
%%   A)\,B))\to (u : \Tm_{i,j}\,\Gamma\,A) \to \Tm_{i,j}\,\Gamma\,(B[x \mapsto
%%     u]).\]
%%   The traditional version is definable as $(\app\,t)[\id,\,u]$. We use
%%   the $\app$-$\lam$ isomorphism because it is formally more convenient.

%% \item \emph{Natural numbers}. We have $\Nat_{i,j} : \Ty_{i,j}\,\Gamma$, $\zero_{i,j} : \Tm_{i,j}\,\Gamma\,\Nat_{i,j}$,
%%   and $\suc_{i,j} : \Tm_{i,j}\,\Gamma\,\Nat_{i,j} \to \Tm_{i,j}\,\Gamma\,\Nat_{i,j}$. The eliminator is the following.
%%   \begin{alignat*}{5}
%%     & \NatElim :\hspace{0.5em}
%%                 &&  (P &&: \Ty_{i,k}\,(\Gamma \ext (n : \Nat_{i,j}))) \\
%%     &           &&  (z &&: \Tm_{i,k}\,\Gamma\,(P[n\mapsto \zero_{i,j}])) \\
%%     &           &&  (s &&: \Tm_{i,k}\,(\Gamma \ext (n : \Nat_{i,j}) \ext (\mit{pn} : P[n\mapsto n]))\,(P[n\mapsto \suc_{i,j}\,n]))\\
%%     &           &&  (t &&: \Tm_{i,j}\,\Gamma\,\Nat_{i,j}) \\
%%     &\hspace{3em}\to        &&&& \Tm_{i,k}\,\Gamma\,(P[n\mapsto t]))
%%   \end{alignat*}
%%   We also have the $\beta$-rules:
%%   \begin{alignat*}{5}
%%     & \NatElim\,P\,z\,s\,\zero_{i,j}     &&= z \\
%%     & \NatElim\,P\,z\,s\,(\suc_{i,j}\,t) &&= s[n \mapsto t,\,\mit{pn} \mapsto \NatElim\,P\,z\,s\,t]
%%   \end{alignat*}
%%   Note that we can eliminate from a level $j$ to any level $k$.
%% \end{itemize}
%% \end{definition}

%% \begin{definition}
%% The \textbf{syntax of 2LTT} is the initial model of 2LTT, defined as a quotient
%% inductive-inductive type \cite{kaposi2019constructing}. The signature for this
%% type includes four sorts ($\Con$, $\Sub$, $\Ty$, $\Tm$), constructors for
%% contexts, explicit substitutions and type/term formers, and equality
%% constructors for all specified equations. The syntax comes with an
%% \emph{induction principle}, from which we can also derive the \emph{recursion principle} as a
%% non-dependent special case, witnessing the initiality of the syntax.
%% \end{definition}

%% \subsubsection{Comparison to Annekov et al.}
%% Comparing our models to the primary reference on 2LTT \cite{twolevel}, the main
%% difference is the handling of ``sizing'' levels. In ibid.\ there is a cumulative
%% lifting from $\Ty_{i,j}$ to $\Ty_{i,j+1}$, which we do not assume. Instead,
%% we allow elimination from $\Nat_{i,j}$ into any $k$ level. This means that we
%% can manually define lifting maps from $\Nat_{i,j}$ to $\Nat_{i,j+1}$ by
%% elimination. This is more similar to e.g.\ Agda, where we do not have cumulativity,
%% but we can define explicit lifting from $\Nat_j : \Set_j$ to $\Nat_k : \Set_k$.

%% In ibid., ``two-level type theory'' specifically refers to the setup
%% where the object level is a homotopy type theory and the meta level is an
%% extensional type theory. In contrast, we allow a wider range of setups under the
%% 2LTT umbrella. Ibid.\ also considers a range of additional
%% strengthenings and extensions of 2LTT \cite[Section~2.4]{twolevel}, most of which
%% are useful in synthetic homotopy theory. We do not assume any of these and
%% stick to the most basic formulation of 2LTT.

%% \subsubsection{Elaborating Informal Syntax}
%% The justification for the usage of the informal syntax in Section
%% \ref{sec:tour-of-2ltt} is \emph{elaboration}. Elaboration translates surface
%% notation to the formal syntax (often termed ``core syntax'' in the
%% implementation of dependently typed programming languages). Elaboration is
%% partial: it may throw error on invalid input. Elaboration also requires
%% \emph{decidability of conversion} for the formal syntax, since type checking
%% requires comparing types for conversion, which in turn requires comparing terms.
%% We do not detail elaboration in this paper, we only make some observations.
%% \begin{itemize}
%% \item Decidability of conversion can be shown through normalization. Since 2LTT
%%       is a minor variation on ordinary MLTT, we expect that existing proofs for MLTT
%%       can be adapted without issue \cite{coquand2018canonicity,decidableconv}.
%% \item We need to elaborate the surface-level Russell-style universe notation to
%%       Coquand-style universes with explicit $\El$ and $\Code$ annotations. It is
%%       likely that this can be achieved with the use of bidirectional elaboration
%%       \cite{DBLP:journals/csur/DunfieldK21}, using two directions (checking and
%%       inference) both for types and terms, and inserting $\El$ and $\Code$ when
%%       switching between types and terms.
%% \item The formal complexity of elaboration varies wildly, depending on how
%%       explicit the surface syntax is. For example, Agda-style implicit arguments
%%       require metavariables and unification, which dramatically increases the
%%       difficulty of formalization.
%% \end{itemize}

%% \subsubsection{The Setoid Interpretation of the Syntax}\label{sec:2lttsetoid} We need to show
%% that the 2LTT syntax, as a particular inductive type, can be modeled in setoids.
%% Unfortunately, there is no prior work which presents the setoid interpretation
%% of quotient types in the generality of quotient induction-induction. We sketch
%% it for our specific use case, but we argue that the interpretation is fairly
%% mechanical and could be adapted to any quotient inductive-inductive type.
%% We focus on contexts and types here.

%% We mutually inductively define $\Con : \Set$, $\Con^{\sim} : \Con \to \Con \to \Prop$, $\Ty_{i,j} : \Con \to \Set$
%%       and $\Ty^{\sim} : \Ty_{i,j}\,\Gamma \to \Ty_{i,j}\,\Delta \to \Con^{\sim}\,\Gamma\,\Delta \to \Prop$, where
%%       $\Con^{\sim}$ and $\Ty^{\sim}$ are proof-irrelevant relations. Note that $\Ty^{\sim}$ is \emph{heterogeneous},
%%       since we can relate types in different (but provably convertible) contexts.

%%       We specify $\emptycon$ and $\blank\ext\blank$ in $\Con$. To $\Con^{\sim}$ we add that $\emptycon$ and $\blank\ext\blank$ are congruences, and also that $\Con^{\sim}$ is reflexive, symmetric and transitive.

%% We add all type formation rules to $\Ty_{i,j}$, and add all quotient equations to
%% $\Ty^{\sim}$ as homogeneously indexed rules. For example, the substitution rule
%% for $\Nat_0$ is specified as $ \Nat_0[] :
%% \Ty^{\sim}\,(\Nat_0[\sigma])\,\Nat_0\,\refl^{\sim} $, where $\sigma :
%% \Sub\,\Gamma\,\Delta$ and $\refl^{\sim} : \Con^{\sim}\,\Gamma\,\Gamma$. We also
%% add all congruences and reflexivity, symmetry and transitivity to $\Ty^\sim$.

%% We add a \emph{coercion} rule to types, as $\msf{coerce} :
%% \Con^{\sim}\,\Gamma\,\Delta \to \Ty_{i,j}\,\Gamma \to \Ty_{i,j}\,\Delta$. This is probably
%% familiar from the traditional specifications of type theories that use
%% conversion relations. However, we additionally need a \emph{coherence rule}
%% which witnesses $\Ty^{\sim}\,A\,(\msf{coerce}\,p\,A)\,p$ for each $A :
%% \Ty_{i,j}\,\Gamma$ and $p : \Con^{\sim}\,\Gamma\,\Delta$. In short, coherence
%% says that coercion preserves conversion.

%% The concise description of this setup is that $(\Con,\,\Con^{\sim})$ is a
%% setoid, and $(\Ty_{i,j},\,\Ty^{\sim})$ is a \emph{setoid fibration} over
%% $(\Con,\,\Con^{\sim})$, for each $i$ and $j$. This notion of setoid fibration is
%% equivalent to the ordinary notion of fibration, when setoids are viewed as
%% categories with proof-irrelevant invertible morphisms. This is appropriate,
%% since types in general are interpreted as setoid fibrations in the setoid model
%% of a type theory.

%% This scheme generalizes to substitutions and terms in a mechanical way. $\Sub$
%% is indexed over two contexts, so it is interpreted as a fibration over
%% the setoid product of $\Con$ and $\Con$. $\Tm$ is indexed over $(\Gamma :
%% \Con)\times \Ty_{i,j}\,\Gamma$, so it is interpreted as a fibration over a setoid
%% $\Sigma$-type. In every conversion relation we take the congruence closure of the specified
%% quotient equations. Finally, we can use the induction principle for the thus
%% specified sets and relations to interpret the internal elimination principle for
%% the 2LTT syntax.

%% %% We need to show that
%% %% the 2LTT syntax, as a particular inductive type, can be modeled in setoids, if

%% %% The examples in Section \ref{sec:tour-of-2ltt} can be faithfully represented in
%% %% the formal syntax. The only notable difference is that the formal syntax uses
%% %% Coquand-style universes instead of Russell-style ones. We use the former because
%% %% it is much easier to model in the semantics in Section \ref{sec:staging-algorithm}. We note
%% %% though that it is straightforward to elaborate from a Russell-style surface
%% %% syntax to the formal version, by inserting $\El$ and $\Code$ as required.

%% \subsection{Models and Syntax of the Object Theory}\label{sec:objmod}

%% We also need to specify the object theory, which serves as the output language
%% of staging. In general, the object theory corresponding to a particular flavor
%% of 2LTT is simply the type theory that supports only the object-level $\Ty_{0,j}$
%% hierarchy with its type formers.

%% \begin{definition}
%% A \textbf{model of the object theory} is a category-with-families, with types and
%% terms indexed over $j \in \mbb{N}$, supporting Coquand-style universes $\U_j$,
%% type formers $\Pi$, $\Sigma$ and $\Nat_j$, with elimination from $\Nat_j$ to
%% any level $k$.
%% \end{definition}

%% \begin{definition}
%% Like before, the \textbf{syntax of the object theory} is the initial model,
%% given as a quotient inductive-inductive type, and it can be also interpreted
%% in setoids.
%% \end{definition}

%% \begin{notation}
%% From now on, by default we use $\Con$, $\Sub$, $\Ty$ and $\Tm$ to refer to sets
%% in the syntax of 2LTT. We use $\mbbo$ to refer to the object syntax, and
%% $\Con\ob$, $\Sub\ob$, $\Ty\ob$ and $\Tm\ob$ to refer to its underlying sets.
%% \end{notation}

%% \begin{definition}
%% We recursively define the \textbf{embedding} of object syntax into 2LTT syntax,
%% which preserves all structure strictly and consists of the following functions:
%% \begin{alignat*}{4}
%%   & \emb{\blank} : \Con\ob \to \Con  && \emb{\blank} : \Sub\ob\,\Gamma\,\Delta \to \Sub\,\emb{\Gamma}\,\emb{\Delta}\\
%%   & \emb{\blank} : \Ty_{\mbbo j}\,\Gamma \to \Ty_{0,j}\,\emb{\Gamma}\hspace{2em} && \emb{\blank} : \Tm_{\mbbo j}\,\Gamma\,A \to \Tm_{0,j}\,\emb{\Gamma}\,\emb{A}
%% \end{alignat*}
%% \end{definition}

%% \section{The Staging Algorithm}\label{sec:staging-algorithm}

%% In this section we specify what we mean by a staging algorithm, then proceed to
%% define one.

%% \begin{definition}\label{def:staging}
%% A \textbf{staging algorithm} consists of two functions:
%%   \begin{alignat*}{3}
%%     & \Stage : \Ty_{0,j}\,\emb{\Gamma} \to \Ty_{\mbbo j}\,\Gamma\hspace{2em} \Stage : \Tm_{0,j}\,\emb{\Gamma}\,A \to \Tm_{\mbbo j}\,\Gamma\,(\Stage\,A)
%%   \end{alignat*}
%% Note that we can stage open types and terms as long as their contexts are purely
%% object-level. By \emph{closed staging} we mean staging only for closed types and
%% terms.
%% \end{definition}
%% \begin{definition}
%%   The following properties are of interest when considering a staging algorithm:
%%   \begin{enumerate}
%%   \item \emph{Soundness:} $\emb{\Stage\,A} = A$ and $\emb{\Stage\,t} = t$.
%%   \item \emph{Stability:} $\Stage\,\emb{A} = A$ and $\Stage\,\emb{t} = t$.
%%   \item \emph{Strictness:} the function extracted from $\Stage$ preserves all
%%     type and term formers strictly (not just up to conversion).
%%   \end{enumerate}
%% \end{definition}

%% Soundness and stability together state that embedding is a bijection
%% on types and terms (up to conversion). This is a statement of
%% \emph{conservativity} of 2LTT over the object theory. In \cite{twolevel} a
%% significantly weaker conservativity theorem is shown, which only expresses that
%% there exists a function from $\Tm_{0,j}\,\emb{\Gamma}\,\emb{A}$ to $\Tm_{\mbbo
%%   j}\,\Gamma\,A$.

%% Strictness tells us that staging does not perform any computation in the object
%% theory; e.g.\ it does not perform $\beta$-reduction. This is practically quite
%% important; for instance, \emph{full normalization of 2LTT} is a sound and stable
%% staging algorithm, but it is not strict, and it is practically not useful, since
%% it does not provide any control over inlining and $\beta$-reduction in code
%% generation.

%% Note that staging necessarily preserves type and term formers up to conversion,
%% since it is defined on quotiented types and terms. More generally, staging
%% respects conversion because every function must respect conversion in the
%% meta type theory.

%% \subsection{The Presheaf Model}

%% We review the presheaf model described in \cite[Section~2.5.3]{twolevel}.  This
%% will directly yield a closed staging algorithm, by recursive evaluation of 2LTT types and
%% terms in the model.

%% We give a higher-level overview first. Presheaves can be viewed as a
%% generalization of sets, as ``variable sets'' which may vary over a base
%% category. A set-based model of 2LTT yields a standard interpreter, where every
%% syntactic object is mapped to the obvious semantic counterpart. The presheaf
%% model over the category of contexts and substitutions in $\mbbo$ generalizes
%% this to an interpreter where semantic values may depend on object-theoretic
%% contexts. This allows us to handle object-theoretic \emph{types and terms}
%% during interpretation, since these necessarily depend on their contexts.

%% In the presheaf model, every semantic construction must be \emph{stable} under
%% object-level substitution, or using alternative terminology, must be
%% \emph{natural} with respect to substitution. Semantic values can be viewed as
%% runtime objects during interpretation, and naturality means that
%% object-theoretic substitutions can be \emph{applied} to such runtime objects,
%% acting on embedded object-theoretic types and terms. Stability underlies
%% the core trade-off in staging:
%% \begin{itemize}
%% \item On one hand, every syntactic rule and construction in 2LTT must
%%       be stable, which restricts the range of metaprograms
%%       that can be written in 2LTT. For example, we cannot make decisions
%%       based on sizes of the object-theoretic scopes, since these sizes may be changed
%%       by substitution.
%% \item On the other hand, if only stable constructions are possible, we never
%%       have to prove stability, or even explicitly mention object-theoretic
%%       contexts, substitutions and variables. The tedious details of deep embeddings
%%       can be replaced with a handful of staging operations and typing rules.
%% \end{itemize}
%% In principle, the base category in a 2LTT can be any category with finite
%% products; it is not required that it is a category of contexts and substitutions.
%% If the base category has more structure, we get more interesting object
%% theories, but at the same time stability becomes more onerous, because we must
%% be stable under \emph{more} morphisms. If the base category is simpler, with
%% fewer morphisms, then the requirement of stability is less restrictive. In
%% Section \ref{sec:intensional-analysis}, we will look at a concrete example for this.

%% We proceed to summarize the key components of the model in a bit more detail. We
%% denote the components of the model by putting hats on 2LTT components, e.g.\ as
%% in $\hCon$.

%% \begin{notation}
%% In this section, we switch to naming elements of $\Cono$ as $a$, $b$ and $c$,
%% and elements of $\Subo$ as $f$, $g$, and $h$, to avoid name clashing with
%% contexts and substitutions in the presheaf model.
%% \end{notation}

%% \subsubsection{The Syntactic Category and the Meta-Level Fragment}

%% \begin{definition} $\bs{\wh{\Con} : \Set_{\omega+1}}$ is defined as the set of presheaves
%% over $\mbbo$. $\Gamma : \wh{\Con}$ has an action on objects $|\Gamma| :
%% \Cono \to \Set_\omega$ and an action on morphisms $\blank[\blank] : |\Gamma|\,b
%% \to \Subo\,a\,b \to |\Gamma|\,a$, such that $\gamma[\id] = \gamma$ and
%% $\gamma[f\circ g] = \gamma[f][g]$.

%% \begin{notation}
%% Above, we reused the substitution notation $\blank[\blank]$ for the action on
%% morphisms.  Also, we use lowercase $\gamma$ and $\delta$ to denote elements of
%% $|\Gamma|\,a$ and $|\Delta|\,a$ respectively.
%% \end{notation}

%% \end{definition}
%% \begin{definition} $\bs{\hSub\,\Gamma\,\Delta : \Set_\omega}$ is the set of natural transformations
%% from $\Gamma$ to $\Delta$. $\sigma : \hSub\,\Gamma\,\Delta$ has action
%% $|\sigma| : \{a : \Cono\} \to |\Gamma|\,a \to |\Delta|\,a$ such that
%% $|\sigma|\,(\gamma[f]) = (|\sigma|\,\gamma)[f]$.
%% \end{definition}

%% \begin{definition}
%% $\bs{\hTy_{1,j}\,\Gamma : \Set_\omega}$ is the set of \emph{displayed
%% presheaves} over $\Gamma$; see e.g.\ \cite[Chapter~1.2]{huber-thesis}. This is
%% equivalent to the set of presheaves over the category of elements of $\Gamma$,
%% but it is usually more convenient in calculations. An $A : \hTy\,\Gamma$ has
%% an action on objects $|A| : \{a : \Cono\} \to |\Gamma|\,a \to \Set_j$ and an
%% action on morphisms $\blank[\blank] : |A|\,\gamma \to (f : \Subo\,a\,b) \to
%% |A|\,(\gamma[f])$, such that $\alpha[\id] = \alpha$ and $\alpha[f \circ g] =
%% \alpha[f][g]$.

%% \begin{notation}
%%   We write $\alpha$ and $\beta$ respectively for elements of $|A|\,\gamma$ and
%%   $|B|\,\gamma$.
%% \end{notation}
%% \end{definition}

%% \begin{definition}
%%   $\bs{\hTm_{1,j}\,\Gamma\,A : \Set_\omega}$ is the set of sections of the
%%   displayed presheaf $A$. This can be viewed as a dependently typed analogue of
%%   a natural transformation.  A $t : \hTm_{1,j}\,\Gamma\,A$ has action $|t| :
%%   \{a\} \to (\gamma : |\Gamma|\,a) \to |A|\,\gamma$ such that $|t|\,(\gamma[f])
%%   = (|t|\,\gamma)[f]$.
%% \end{definition}

%% We also look at he empty context and context extension with meta-level types,
%% as these will appear in subsequent definitions.

%% \begin{definition}\label{def:psh-emptycon}
%% $\bs{{\widehat{\emptycon}} : \hCon}$ is defined as the presheaf
%% which is constantly $\top$, i.e.\ $|\wh{\emptycon}|\,\_ = \top$.
%% \end{definition}

%% \begin{definition}\label{def:psh-ext} For $A : \hTy_{1,j}\,\Gamma$, we define $\bs{\Gamma\,\,\wh{\ext}\,A}$
%% pointwise by $|\Gamma\,\,\wh{\ext}\,A|\,a := (\gamma : |\Gamma|\,a) \times |A|\,\gamma$
%% and $(\gamma,\,\alpha)[f] := (\gamma[f],\,\alpha[f])$.
%% \end{definition}

%% Using the above definitions, we can model the syntactic category of 2LTT, and
%% also the meta-level family structure and all meta-level type formers. For
%% expositions in previous literature, see \cite[Chapter~1.2]{huber-thesis} and
%% \cite[Section~4.1]{Hofmann97syntaxand}.

%% \subsubsection{The Object-Level Fragment}

%% We move on to modeling the object-level syntactic fragment of 2LTT. We make some
%% preliminary definitions. First, note that types in the object theory yield a
%% presheaf, and terms yield a displayed presheaf over them; this immediately
%% follows from the specification of a family structure in a cwf. Hence, we do a
%% bit of a name overloading, and have $\Ty_{\mbbo j} : \hCon$ and $\Tm_{\mbbo j} : \hTy\,\Ty_{\mbbo j}$.

%% \begin{definition}
%% $\bs{\hTy_{0,j}\,\Gamma : \Set_\omega}$ is defined as $\hSub\,\Gamma\,\Ty_{\mbbo j}$,
%% and $\bs{\hTm_{0,j}\,\Gamma\,A : \Set_\omega}$ is defined as $\hTm\,\Gamma\,(\Tm_{\mbbo j}[A])$.
%% \end{definition}

%% For illustration, if $A : \hTy_{0,j}\,\Gamma$, then $A :
%% \hSub\,\Gamma\,\Ty_{\mbbo j}$, so $|A| : \{a : \Cono\} \to |\Gamma|\,a \to
%% \Ty_{\mbbo j}\,a$. In other words, the action of $A$ on objects maps a semantic
%% context to a syntactic object-level type. Likewise, for $t :
%% \hTm_{0,j}\,\Gamma\,A$, we have $|t| : (\gamma : |\Gamma|\,a) \to \Tm_{\mbbo
%%   j}\,a\,(|A|\,\gamma)$, so we get a syntactic object-level term as output.

%% \begin{definition} For $A : \Ty_{0,j}\,\Gamma$, we define $\bs{\Gamma\,\,\wh{\ext}\,A}$
%% as $|\Gamma\,\,\wh{\ext}\,A|\,a := (\gamma : |\Gamma|\,a) \times \Tm_{\mbbo
%%   j}\,a\,(|A|\,\gamma)$ and $(\gamma,\,t)[f] := (\gamma[f],\,t[f])$. Thus,
%% extending a semantic context with an object-level type appends an object-level
%% term.
%% \end{definition}

%% Using the above definitions and following \cite{twolevel}, we can model all type
%% formers in $\hTy_{0,j}$. Intuitively, this is because $\hTy_{0,j}$ and
%% $\hTm_{0,j}$ return types and terms, so we can reuse the type and term formers
%% in the object theory.

%% \subsubsection{Lifting}

%% \begin{definition}\label{def:psh-lifting}
%% $\bs{\wh{\Lift} A : \hTy_{1,j}\,\Gamma}$ is defined as $\Tm_{\mbbo j}[A]$. With
%%   this, we get that $\hTm_{0,j}\,\Gamma\,A$ is equal to
%%   $\hTm_{1,j}\,\Gamma\,(\wh{\Lift}\,A)$. Hence, we can define both quoting and
%%   splicing as identity functions in the model.
%% \end{definition}

%% \subsection{Closed Staging}

%% \begin{definition}
%% The \textbf{evaluation morphism, denoted $\bs{\ev}$} is a morphism of 2LTT
%% models, between the syntax of 2LTT and $\hato$. It is defined by recursion on
%% the syntax (that is, appealing to the initiality of the syntax), and it strictly
%% preserves all structure.
%% \end{definition}

%% \begin{definition} \textbf{Closed staging} is defined as follows.
%% \begin{alignat*}{4}
%%   & \Stage : \Ty_{0,j}\,\emptycon \to \Ty_{\mbbo j}\,\emptycon \hspace{2em} && \Stage : \Tm_{0,j}\,\emptycon\,A \to \Tm_{\mbbo j}\,\emptycon\,(\Stage\,A) \\
%%   & \Stage\,A := |\ev\,A|\,\{\emptycon\}\,\tt && \Stage\,t := |\ev\,t|\,\{\emptycon\}\,\tt
%% \end{alignat*}
%% Note that $\wh{\emptycon}$ is defined as the terminal presheaf, which is
%% constantly $\top$. Also, $|\ev\,A|\{\emptycon\} : |\ev\,\emptycon|\,\emptycon \to \Ty_{\mbbo
%%   j}\,\emptycon$, therefore $|\ev\,A|\{\emptycon\} : \top \to \Ty_{\mbbo
%%   j}\,\emptycon$ and $|\ev\,t|\,\{\emptycon\} : \top \to \Tm_{\mbbo
%%   j}\,\emptycon\,(|\ev\,A|\,\tt)$.
%% \end{definition}

%% What about general (open) staging though? Given $A : \Ty_{0,j}\,\emb{\Gamma}$,
%% we get $|\ev\,A|\,\{\Gamma\} : |\emb{\Gamma}|\,\Gamma \to \Ty_{\mbbo j}$. We
%% need an element of $|\emb{\Gamma}|\,\Gamma$, in order to obtain object-level
%% type. Such ``generic'' semantic environments should be possible to construct, because
%% elements of $|\emb{\Gamma}|\,\Gamma$ are essentially lists of object-level terms
%% in $\Gamma$, by Definitions \ref{def:psh-ext} and \ref{def:psh-emptycon}, so $|\emb{\Gamma}|\,\Gamma$ should be
%% isomorphic to $\Subo\,\Gamma\,\Gamma$.  It turns out that this falls out from
%% the stability proof of $\ev$.

%% \subsection{Open Staging, Stability and Strictness}

%% We define a family of functions $\blank^P$ by induction on object syntax, such
%% that the interpretation of a context yields a generic semantic environment. The
%% induction motives are as follows.
%% \begin{alignat*}{4}
%%   & (\Gamma : \Cono)^P                    &&: |\ev\,\emb{\Gamma}|\,\Gamma && (\sigma : \Subo\,\Gamma\,\Delta)^P    &&: \Delta^P[\sigma] = |\ev\,\emb{\sigma}|\,\Gamma^P \\
%%   & (A      : \Ty_{\mbbo j}\,\Gamma)^P      &&: A = |\ev\,\emb{A}|\,\Gamma^P\hspace{2em} && (t      : \Tm_{\mbbo j}\,\Gamma\,A)^P   &&: t = |\ev\,\emb{t}|\,\Gamma^P
%% \end{alignat*}
%% We look at the interpretation of contexts.
%% \begin{itemize}
%% \item  For $\emptycon^P$, we need an element of $|\ev\,\emb{\emptycon}|\,\emptycon$, hence
%%   an element of $\top$, so we define $\emptycon^P$ as $\tt$.
%% \item
%%   For $(\Gamma \ext A)^P$, we need an element of
%%     \[(\gamma : |\ev{\emb{\Gamma}}|\,(\Gamma\ext A)) \times \Tm_{\mbbo j}\,(\Gamma \ext A)\,(|\ev\emb{A}|\,\gamma).\]
%%   We have $\Gamma^P : |\ev{\emb{\Gamma}}|\,\Gamma$, which we can weaken as
%%   $\Gamma^P[\p] : |\ev{\emb{\Gamma}}|\,(\Gamma \ext A)$, so we set the first
%%   projection of the result as $\Gamma^P[\p]$. For the second projection, the
%%   goal type can be simplified as follows:
%%   \begin{alignat*}{4}
%%     &       &&\Tm_{\mbbo j}\,(\Gamma \ext A)\,(|\ev\emb{A}|\,(\Gamma^P[\p])) && \\
%%     & =\,\, &&\Tm_{\mbbo j}\,(\Gamma \ext A)\,((|\ev\emb{A}|\,\Gamma^P)[\p]) &&\hspace{2em} \text{by naturality of $\ev\emb{A}$}\\
%%     & =\,\, &&\Tm_{\mbbo j}\,(\Gamma \ext A)\,(A[\p])                        &&\hspace{2em} \text{by $A^P$}
%%   \end{alignat*}
%%   We have the zero de Bruijn variable $\q : \Tm_{\mbbo j}\,(\Gamma \ext
%%   A)\,(A[\p])$.  Hence, we define $(\Gamma \ext A)^P$ as $(\Gamma^P[\p],\,\q)$.
%% \end{itemize}
%% Thus, a generic semantic context $\Gamma^P : |\ev\emb{\Gamma}|\,\Gamma$ is just
%% a list of variables, corresponding to the identity substitution $\id :
%% \Subo\,\Gamma\,\Gamma$ which maps each variable to itself.

%% The rest of the $\blank^P$ interpretation is straightforward and we omit it
%% here. In particular, preservation of definitional equalities is automatic, since
%% types, terms and substitutions are all interpreted as proof-irrelevant
%% equations.

%% \begin{definition}\label{def:open-staging} We define \textbf{open staging} as follows.
%% \begin{alignat*}{4}
%%   & \Stage : \Ty_{0,j}\,\emb{\Gamma} \to \Ty_{\mbbo j}\,\Gamma \hspace{2em} && \Stage : \Tm_{0,j}\,\emb{\Gamma}\,A \to \Tm_{\mbbo j}\,\Gamma\,(\Stage\,A) \\
%%   & \Stage\,A := |\ev\,A|\,\Gamma^P && \Stage\,t := |\ev\,t|\,\Gamma^P
%% \end{alignat*}
%% \end{definition}

%% \begin{theorem} Open staging is stable.
%% \end{theorem}
%% \begin{proof} For $A : \Ty_{\mbbo j}$, $\Stage\,\emb{A}$ is by definition $|\ev\,\emb{A}|\,\Gamma^P$,
%%   hence by $A^P$ it is equal to $A$. Likewise, $\Stage\,\emb{t}$ is equal to $t$ by $t^P$.
%% \end{proof}

%% \begin{theorem} Open staging is strict.
%% \end{theorem}
%% \begin{proof}
%%   This is evident from the definition of the presheaf model, where the action of
%%   each type and term former in $\wh{\Ty}_{0,j}$ and $\wh{\Tm}_{0,j}\,\Gamma\,A$ is
%%   defined using the corresponding type or term former in the object syntax.
%% \end{proof}

%% \subsection{Implementation and Efficiency}

%% We discuss now the operational behavior of the extracted staging algorithm
%% and look at potential adjustments and optimizations that make it more efficient
%% or more convenient to implement. Recall the types of the component functions in $\ev$:
%% \begin{alignat*}{3}
%%   & |\blank| \circ \ev : \Ty_{1 j}\,\Gamma \to \{\Delta : \Cono\} \to |\ev\,\Gamma|\,\Delta \to \Set_j\\
%%   & |\blank| \circ \ev : \Ty_{0 j}\,\Gamma \to \{\Delta : \Cono\} \to |\ev\,\Gamma|\,\Delta \to \Ty_{\mbbo j}\,\Delta\\
%%   & |\blank| \circ \ev : \Tm_{1 j}\,\Gamma\,A \to \{\Delta : \Cono\} \to (\gamma : |\ev\,\Gamma|\,\Delta) \to
%%     |\ev\,A|\,\gamma \\
%%   & |\blank| \circ \ev : \Tm_{0 j}\,\Gamma\,A \to \{\Delta : \Cono\} \to (\gamma : |\ev\,\Gamma|\,\Delta) \to
%%     \Tm_{\mbbo\,j}\,\Delta\,(|\ev\,A|\,\gamma)
%% \end{alignat*}
%% We interpret syntactic types or terms in a semantic environment
%% $\gamma : |\ev\,\Gamma|\,\Delta$. These environments are lists containing
%% a mix of  object-level terms and semantic values. The semantic values are represented
%% using metatheoretic inductive types and function types.
%% \begin{itemize}
%%   \item $|\Nat_1|\,\_$ is simply the set of natural numbers $\mbb{N}$.
%%   \item $|\Sigma_1\,A\,B|\,\gamma$ is simply a set of pairs of values.
%%   \item A non-dependent function type $A\,\wh{\to}\,B$ is defined as the presheaf
%%         exponential. The computational part of $|A\,\wh{\to}\,B|\,\{\Delta\}\,\gamma$
%%         is given by a function with type
%%         \[ (\Theta : \Cono) \to (\sigma : \Subo\,\Theta\,\Delta) \to |A|\,(\gamma[\sigma]) \to |B|\,(\gamma[\sigma])        \]
%%         This may be also familiar as the semantic implication from the Kripke
%%         semantics of intuitionistic logics. Whenever we evaluate a function
%%         application, we supply an extra $\id : \Subo\,\Delta\,\Delta$.
%%         This may incur cost via the $\gamma[\id]$ restrictions in a naive
%%         implementation, but this is easy to optimize, by introducing a formal
%%         representation of $\id$, such that $\gamma[\id]$ immediately computes to
%%         $\gamma$. The case of dependent functions is analogous operationally.
%% \end{itemize}
%% In summary, in the meta-level fragment of 2LTT, $\ev$ yields a reasonably efficient
%% computation of closed values, which reuses functions from the ambient
%% metatheory. Alternatively, instead of using ambient functions, we could use our
%% own implementation of function closures during staging.

%% In contrast, we have a bit of an efficiency problem in the handling of
%% object-level binders: whenever we go under such binder we have to weaken the
%% current semantic environment. Concretely, when moving from $\Delta$ to
%% $\Delta \ext A$, we have to shift $\gamma : |\ev\,\Gamma|\,\Delta$ to
%% $\gamma[\p] : |\ev\,\Gamma|\,(\Delta \ext A)$. This cannot be avoided by an easy
%% optimization trick. For object-level entries in the environment, this is cheap,
%% because we just add an extra explicit weakening, but for semantic values
%% weakening may need to perform deep traversals.

%% The same efficiency issue arises in formal presheaf-based
%% normalization-by-evaluation, such as in \cite{kaposinbe} and
%% \cite{coquand2018canonicity}. However, in practical implementations this issue
%% can be fully solved by using De Bruijn \emph{levels} in the semantic domain,
%% thus arranging that weakening has no operational cost on semantic values; see
%% \cite{coquand1996algorithm}, \cite{modulartc} and \cite{untypedtc} for
%% implementations in this style. We can use the same solution in staging. It is a
%% somewhat unfortunate mismatch that indices are far more convenient in
%% formalization, but levels are more efficient in practice.

%% In our prototype implementation, we use the above optimization with De Bruijn
%% levels, and we drop explicit substitutions from the 2LTT core syntax. In the
%% object-level syntax, we use De Bruijn levels for variables and closures in
%% binders. This lets us completely skip the weakening of environments, which is
%% slightly more efficient than the more faithfully extracted algorithm.

%% Additionally, we use an untyped tagged representation of semantic
%% values, since in Haskell (the implementation language) we do not have
%% strong enough types to represent typed semantic values. For example, there
%% are distinct data constructors storing semantic function values, natural number
%% literals and quoted expressions.

%% \paragraph{Caching} In a production-strength implementation we would need some
%% caching mechanism, to avoid excessive duplication of code. For example, if we
%% use the staged $\mit{map}$ function multiple times with the same type and
%% function arguments, we do not want to generate separate code for each
%% usage. De-duplicating object-level code with function types is usually safe,
%% since function bodies become closed top-level code after closure conversion. We
%% leave this to future work.

%% \section{Soundness of Staging}\label{sec:soundness}

%% In this section we prove soundness of staging. We build a proof-relevant logical
%% relation between the evaluation morphism $\ev$ and a \emph{restriction}
%% morphism, which restricts 2LTT syntax to object-level contexts. The relational
%% interpretations of $\Ty_{0,j}$ and $\Tm_{0,j}$ will yield the soundness
%% property.

%% \subsection{Working in $\hat{\mbbo}$}

%% We have seen that 2LTT can be modeled in the presheaf topos
%% $\hato$. Additionally, $\hato$ supports all type formers of extensional type
%% theory and certain other structures which are stable under object-level
%% substitution. As all constructions in this section must be stable, it makes
%% sense to work internally to $\hato$. This style has been previously used in
%% normalization proofs \cite{coquand2018canonicity} and also in the metatheory of
%% cubical type theories
%% \cite{licata2018internal,orton_et_al:LIPIcs:2016:6564,DBLP:conf/csl/CavalloMS20}.

%% When we work internally in a model of a type theory, we do not explicitly refer
%% to contexts, types, and substitutions. For example, when working in Agda, we do
%% not refer to Agda's typing contexts. Instead, we only work with terms, and use
%% functions and universes to abstract over types and semantic contexts. Hence, we
%% have to convert along certain isomorphisms when we switch between the internal
%% and external views. In the following, we summarize features in $\hato$ and also
%% the internal-external conversions.

%% We write $\whset_j$ for ordinal-indexed Russell-style universes. Formally, we
%% have Coquand-style universes, but for the sake of brevity we omit $\El$ and
%% $\Code$ from internal syntax. Universes are cumulative, and closed under $\Pi$,
%% $\Sigma$, extensional identity $\blank\!=\!\blank$ and inductive types. We use
%% the same conventions as in Notation \ref{basic-notation}.

%% \noindent The basic scheme for internalization is as follows:
%% \begin{alignat*}{3}
%%   & \Gamma : \hCon                  && \hspace{1em}\text{is internalized as}\hspace{1em} \Gamma : \whset_\omega\\
%%   & \sigma : \hSub\,\Gamma\,\Delta  && \hspace{1em}\text{is internalized as}\hspace{1em} \sigma : \Gamma \to \Delta\\
%%   & A : \hTy_{1,j}\,\Gamma           && \hspace{1em}\text{is internalized as}\hspace{1em} A      : \Gamma \to \whset_j \\
%%   & t : \hTm_{1,j}\,\Gamma\,A        && \hspace{1em}\text{is internalized as}\hspace{1em} t      : (\gamma : \Gamma) \to A\,\gamma
%% \end{alignat*}
%% \subsubsection{Object-Theoretic Syntax} The syntax of the object theory is clearly
%% fully stable under object-theoretic substitution, so we can internalize all of
%% its type and term formers. We internalize object-level types and terms as
%% $\Ty_{\mbbo j} : \whset_0$ and $\Tm_{\mbbo j} : \Ty_{\mbbo j} \to
%% \whset_0$. $\Ty_{\mbbo j}$ is closed under type formers. For instance, we have
%% \[ \Nat_j : \Ty_{\mbbo j}\hspace{1.5em}\zero_j : \Tm_{\mbbo j}\,\Nat_j\hspace{1.5em}
%%     \suc_j : \Tm_{\mbbo j}\,\Nat_j \to \Tm_{\mbbo j}\,\Nat_j \]
%% together with $\NatElim$, and likewise we have all other type formers.
%% \subsubsection{Internal $\ev$} We also use an internal view of
%% $\ev$ which maps 2LTT syntax to internal values; i.e.\ we compose $\ev$ with
%% internalization. Below, note that the input to $\ev$ is \emph{external}, so we
%% mark $\ev$ as being parameterized by external input.
%% \begin{alignat*}{3}
%%   &\ev\,(\Gamma &&: \Con)                      &&: \whset_\omega \\
%%   &\ev\,(\sigma &&: \Sub\,\Gamma\,\Delta)      &&: \ev\,\Gamma \to \ev\,\Delta \\
%%   &\ev\,(A      &&: \Ty_{1,j}\,\Gamma)          &&: \ev\,\Gamma \to \whset_j\\
%%   &\ev\,(t      &&: \Tm_{1,j}\,\Gamma)          &&: (\gamma : \ev\,\Gamma) \to \ev\,A\,\gamma\\
%%   &\ev\,(A      &&: \Ty_{0,j}\,\Gamma)          &&: \ev\,\Gamma \to \Ty_{\mbbo_j}\\
%%   &\ev\,(t      &&: \Tm_{0,j}\,\Gamma\,A)       &&: (\gamma : \ev\,\Gamma) \to \Tm_{\mbbo j}\,(\ev\,A\,\gamma)
%% \end{alignat*}
%% \subsubsection{Object-Level Fragment of 2LTT}
%% The purely object-level syntactic fragment of 2LTT can be internalized as
%% follows. We define externally the presheaf of object-level types as
%% $|\Ty_{0,j}|\,\Gamma := \Ty_{0,j}\,\emb{\Gamma}$, and the displayed presheaf of
%% object-level terms over $\Ty_{0,j}$ as $|\Tm_{0,j}|\,\{\Gamma\}\,A :=
%% \Tm_{0,j}\,\emb{\Gamma}\,A$. Hence, internally we have $\Ty_{0,j} : \whset_0$ and
%% $\Tm_{0,j} : \Ty_{0,j} \to \whset_0$. $\Ty_{0,j}$ is closed under all type formers,
%% analogously as we have seen for $\Ty_{\mbbo j}$.

%% %% \begin{alignat*}{6}
%% %%   &\ev\,(\Gamma &&: \Con)                      &&: \whset_\omega && \ev\,(t      &&: \Tm_{1,j}\,\Gamma)          &&: (\gamma : \ev\,\Gamma) \to \ev\,A\,\gamma\\
%% %%   &\ev\,(\sigma &&: \Sub\,\Gamma\,\Delta)      &&: \ev\,\Gamma \to \ev\,\Delta \hspace{3.5em}&& \ev\,(A      &&: \Ty_{0,j}\,\Gamma)          &&: \ev\,\Gamma \to \Ty_{\mbbo_j}\\
%% %%   &\ev\,(A      &&: \Ty_{1,j}\,\Gamma)          &&: \ev\,\Gamma \to \whset_j&& \ev\,(t      &&: \Tm_{0,j}\,\Gamma\,A)       &&: (\gamma : \ev\,\Gamma) \to \Tm_{\mbbo j}\,(\ev\,A\,\gamma)
%% %% \end{alignat*}

%% %% \begin{alignat*}{3}
%% %%   &\ev\,(\Gamma &&: \Con)                      &&: \whset_\omega \\
%% %%   &\ev\,(\sigma &&: \Sub\,\Gamma\,\Delta)      &&: \ev\,\Gamma \to \ev\,\Delta \\
%% %%   &\ev\,(A      &&: \Ty_{1,j}\,\Gamma)          &&: \ev\,\Gamma \to \whset_j\\
%% %%   &\ev\,(t      &&: \Tm_{1,j}\,\Gamma)          &&: (\gamma : \ev\,\Gamma) \to \ev\,A\,\gamma\\
%% %%   &\ev\,(A      &&: \Ty_{0,j}\,\Gamma)          &&: \ev\,\Gamma \to \Ty_{\mbbo_j}\\
%% %%   &\ev\,(t      &&: \Tm_{0,j}\,\Gamma\,A)       &&: (\gamma : \ev\,\Gamma) \to \Tm_{\mbbo j}\,(\ev\,A\,\gamma)
%% %% \end{alignat*}

%% \subsubsection{Embedding}
%% Now, the embedding operation $\emb{\blank}$ can be also internalized on types
%% and terms, as $\emb{\blank} : \Ty_{\mbbo j} \to \Ty_{0,j}$, and $\emb{\blank}
%% :\Tm_{\mbbo\,j}\,A \to \Tm_{0,j}\,\emb{A}$. Embedding strictly preserves all structure.

%% \subsection{The Restriction Morphism}

%% We define a family of functions $\re$ from the 2LTT syntax to objects in
%% $\hato$. We will relate this to the evaluation morphism $\ev$ in the relational
%% interpretation. In short, $\re$ restricts 2LTT syntax so that it can only depend
%% on object-level contexts, which are given as $\emb{\Gamma}$.

%% \begin{definition}\label{def:restriction} We specify the types of the \textbf{restriction operations} internally, and the $|\blank|$
%% components of the operations externally. The naturality of $|\blank|$ is
%% straightforward in each case.
%% \begingroup
%% %% \allowdisplaybreaks
%% \begin{alignat*}{5}
%%   & \re\,(\Gamma : \Con) : \whset_0 && \re\,(\sigma : \Sub\,\Gamma\,\Delta) : \re\,\Gamma \to \re\,\Delta \\
%%   & |\re\,\Gamma|\,\Delta := \Sub\,\emb{\Delta}\,\Gamma && |\re\,\sigma|\,\gamma := \sigma \circ \gamma \\
%%   & \\
%%   & \re\,(A : \Ty_{1,j}\,\Gamma) : \re\,\gamma \to \whset_0 && \re\,(t : \Tm_{1,j}\,\Gamma\,A) : (\gamma : \re\,\Gamma) \to \re\,A\,\gamma \\
%%   & |\re\,A|\,\{\Delta\}\,\gamma := \Tm_{1,j}\,\emb{\Delta}\,(A[\gamma])\hspace{2em} && |\re\,t|\,\gamma := t[\gamma]\\
%%   & \\
%%   & \re\,(A : \Ty_{0,j}\,\Gamma) : \re\,\Gamma \to \Ty_{0,j}\hspace{2em} && \re\,(t : \Tm_{0,j}\,\Gamma\,A) : (\gamma : \re\,\Gamma) \to \Tm_{0,j}\,(\re\,A\,\gamma)\\
%%   & |\re\,A|\,\gamma := A[\gamma] && |\re\,t|\,\gamma := t[\gamma]
%% \end{alignat*}
%% \endgroup
%% \end{definition}

%% \subsubsection{Preservation Properties of $\re$}
%% First, we note that $\re$ strictly preserves $\id$, $\blank\circ\blank$ and
%% type/term substitution, and it preserves $\emptycon$ and $\blank\ext_1\blank$ up
%% to isomorphism. We have the following isomorphisms internally to $\hato$:
%% \begin{alignat*}{3}
%%   & \re_{\emptycon} &&: \re\,\emptycon \simeq \top \\
%%   & \re_{\ext_1}   &&: \re\,(\Gamma \ext_1 A) \simeq ((\gamma : \re\,\Gamma) \times \re\,A\,\gamma)
%% \end{alignat*}

%% \begin{notation} When we have an isomorphism $f : A \simeq B$, we may write $f$ for the
%%   function in $A \to B$, and $f^{-1} : B \to A$ for its inverse.
%% \end{notation}

%% \begin{notation}
%%   We can use a pattern matching notation on isomorphisms. For example, if $f : A
%%   \simeq B$, then we may write $(\lambda\,(f\,a).\,t) : B \to C$, and likewise
%%   $(\lambda\,(f^{-1}\,b).\,t) : A \to C$, where the function bodies can refer to
%%   the bound $a : A$ and $b : B$ variables.
%% \end{notation}

%% The preservation properties $\re_{\emptycon}$ and $\re_{\ext_1}$ mean that $\re$
%% is a \emph{pseudomorphism} in the sense of \cite{gluing}, between the syntactic
%% cwf given by $(\Ty_{1,j},\,\Tm_{1,j})$ and the corresponding cwf structure in
%% $\hato$. In ibid.\ there is an analysis of such cwf morphisms, from which we
%% obtain the following additional preservation properties:
%% \begin{itemize}
%% \item Meta-level $\Sigma$-types are preserved up to isomorphism, so we have
%%   \[ \re_{\Sigma} : \re\,(\Sigma\,A\,B)\,\gamma
%%   \simeq ((\alpha : \re\,A) \times \re\,B\,(\rexti\,(\gamma,\,\alpha))). \]
%%   The semantic values of $\re\,(\Sigma\,A\,B)\,\gamma$
%%   are 2LTT terms with type $(\Sigma\,A\,B)[\gamma]$, restricted to object-level contexts.
%%   We can still perform pairing and projection with such restricted terms; hence
%%   the preservation property.
%% \item Meta-level $\Pi$-types and universes are preserved in a lax way. For $\Pi$, we have
%%   \[ \re_{\app} : \re\,(\Pi\,A\,B)\,\gamma \to (\alpha : \re\,A\,\gamma) \to \re\,B\,(\rexti\,(\gamma,\,\alpha))
%%   \]
%%   such that $\re_{\app}\,(\re\,t\,\gamma)\,\alpha =
%%   \re\,(\app\,t)\,(\rexti\,(\gamma,\,\alpha))$. In this case, we can apply a
%%   restricted term with a $\Pi$-type to a restricted term, but we cannot do
%%   lambda-abstraction, because that would require extending the context with a
%%   meta-level binder. For $\U_{1,j}$, we have
%%   \[
%%     \re_{\El} : \re\,\U_{1,j}\,\gamma \to \whset_j
%%   \]
%%   such that $\re_{\El}\,(\re\,t\,\gamma) = \re\,(\El\,t)\,\gamma$. Here, we only
%%   have lax preservation simply because $\whset_j$ is much larger than the the
%%   set of syntactic 2LTT types, so not every semantic $\whset_j$ has a syntactic
%%   representation.

%% \item Meta-level positive (inductive) types are preserved in an oplax way. In the case of
%%   natural numbers, we have
%%   \[ \re_{\mbb{N}} : \mbb{N} \to \re\,\Nat_{1,j}\,\gamma. \]
%%   This is a ``serialization'' map: from a metatheoretic natural number we
%%   compute a numeral as a closed 2LTT term. This works analogously for other
%%   inductive types, including parameterized ones. For an example, from a semantic
%%   list of restricted terms we would get a syntactic term with list type,
%%   containing the same restricted terms.
%% \end{itemize}
%% \subsubsection{Action on Lifting} We have an isomorphism $\re\,(\Lift\,A)\,\gamma \simeq
%% \Tm_{0,j}\,(\re\,A\,\gamma)$. This is given by quoting and splicing: we convert
%% between restricted meta-level terms with type $\Lift\,A$ and restricted
%% object-level terms with type $A$. Hence, we write components of this isomorphism
%% as $\spl\blank$ and $\qut{\blank}$, as internal analogues of the external
%% operations. With this, we also have $\re\,\qut{t}\,\gamma =
%% \qut{\re\,t\,\gamma}$ and $\re\,\spl\,t\,\gamma = \spl(\re\,t\,\gamma)$.

%% \subsubsection{Action on $\blank\ext_0\blank$} We have preservation up to isomorphism:
%% \[
%%   \re_{\ext_0} : \re\,(\Gamma \ext_0 A) \simeq ((\gamma : \re\,\Gamma)\times \Tm_{0,j}\,(\re\,A\,\gamma))
%% \]
%% This is because substitutions targeting $\Gamma \ext_0 A$ are the same as
%% pairs of substitutions and terms, by the specification of $\blank\!\ext_0\!\blank$.

%% \subsubsection{Action on Object-Level Types and Terms} $\re$ preserves all structure
%% in the object-level fragment of 2LTT. This follows from the $\re$ specification:
%% an external type $A : \Ty_{0,j}\,\Gamma$ is directly internalized as an element
%% of $\Ty_{0,j}$, and the same happens for $t : \Tm_{0,j}\,\Gamma\,A$.

%% \subsection{The Logical Relation}

%% Internally to $\hato$, we define by induction on the syntax of 2LTT a
%% proof-relevant logical relation interpretation, written as $\blank^{\approx}$. The induction
%% motives are specified as follows.
%% \begin{alignat*}{5}
%%   &(\Gamma &&: \Con)\rel && : \ev\,\Gamma \to \re\,\Gamma \to \whset_\omega\\
%%   &(\sigma &&: \Sub\,\Gamma\,\Delta)\rel &&: \Gamma\rel\,\gamma\,\gamma' \to \Delta\rel\,(\ev\,\sigma\,\gamma)\,(\re\,\sigma\,\gamma')\\
%%   &(A      &&: \Ty_{1,j}\,\Gamma)\rel &&: \Gamma\rel\,\gamma\,\gamma' \to \ev\,A\,\gamma \to \re\,A\,\gamma' \to \whset_j\\
%%   &(t      &&: \Tm_{1,j}\,\Gamma\,A)\rel &&: (\gamma\rel : \Gamma\rel\,\gamma\,\gamma') \to A\rel\,\gamma\rel\,(\ev\,t\,\gamma)\,(\re\,t\,\gamma')\\
%%   &(A      &&: \Ty_{0,j}\,\Gamma)\rel &&: \Gamma\rel\,\gamma\,\gamma' \to \emb{\ev\,A\,\gamma} = \re\,A\,\gamma' \\
%%   &(t      &&: \Tm_{0,j}\,\Gamma\,A)\rel &&: \Gamma\rel\,\gamma\,\gamma' \to \emb{\ev\,t\,\gamma} = \re\,t\,\gamma'
%% \end{alignat*}
%% For $\Con$, $\Sub$ and meta-level types and terms, this is a fairly standard
%% logical relation interpretation: contexts are mapped to relations, types to
%% dependent relations, and substitutions and terms respect relations. We will only have
%% modest complications in meta-level type formers because we will sometimes need to use
%% the lax/oplax preservations properties of $\re$. For object-level types and
%% terms, we get soundness statements: evaluation via $\ev$ followed by embedding
%% back to 2LTT is the same as restriction to object-level contexts. We describe
%% the $\blank\rel$ interpretation in the following.

%% \subsubsection{Syntactic Category and Terminal Object}
%% Here, we simply have $\id\rel\,\gamma\rel := \gamma\rel$ and $(\sigma \circ
%% \delta)\rel\,\gamma\rel := \sigma\rel\,(\delta\rel\,\gamma\rel)$. The terminal
%% object is interpreted as $\emptycon\rel\,\gamma\,\gamma' := \top$.

%% \subsubsection{Meta-Level Family Structure}
%% We interpret context extension and type/term substitution as follows. Note the
%% usage of the pattern matching notation on the $\rexti$ isomorphism.
%% \begin{alignat*}{4}
%%   & (\Gamma \ext_1 A)\rel\,(\gamma,\,\alpha)\,(\rexti(\gamma',\,\alpha')) && :=
%%     (\gamma\rel : \Gamma\rel\,\gamma\,\gamma') \times A\rel\,\gamma\rel\,\alpha\,\alpha'\\
%%   & (A[\sigma])\rel\,\gamma\rel\,\alpha\,\alpha' && := A\rel\,(\sigma\rel\,\gamma\rel)\,\alpha\,\alpha'\\
%%   & (t[\sigma])\rel\,\gamma\rel\                 && := t\rel\,(\sigma\rel\,\gamma\rel)
%% \end{alignat*}
%% It is enough to specify the $\blank\rel$ action on extended substitutions
%% $(\sigma,\,t) : \Sub\,\Gamma\,(\Delta\ext A)$, $\p : \Sub\,(\Gamma\ext
%% A)\,\Gamma$, $\p : \Sub\,(\Gamma\ext A)\,\Gamma$ and $\q : \Tm_{1,j}\,(\Gamma
%% \ext A)\,(A[\p])$. The category-with-families equations hold evidently.
%% \begin{alignat*}{4}
%%   & (\sigma,\,t)\rel\,\gamma\rel := (\sigma\rel\,\gamma\rel,\,t\rel\,\gamma\rel) \hspace{2em} \p\rel\,(\gamma\rel,\,t\rel) := \gamma\rel \hspace{2em} \q\rel\,(\gamma\rel,\,t\rel) := t\rel
%% \end{alignat*}

%% \subsubsection{Meta-Level Natural Numbers} First, we have to define a relation:
%% \begin{alignat*}{4}
%%   & (\Nat_{1,j})\rel : \Gamma\rel\,\gamma\,\gamma' \to \mbb{N} \to \re\,\Nat_{1,j}\,\gamma' \to \whset_j\\
%%   & (\Nat_{1,j})\rel\,\gamma\rel\,n\,n' := (\re_{\mbb{N}}\,n = n')
%% \end{alignat*}
%% Note that evaluation sends $\Nat_{1,j}$ to the semantic type of natural numbers,
%% i.e.\ $\ev\,\Nat_{1,j} = \mbb{N}$. We refer to the serialization map
%% $\re_{\mbb{N}} : \mbb{N} \to \re\,\Nat_{1,j}\,\gamma'$ above. In short,
%% $(\Nat_{1,j})\rel$ expresses \emph{canonicity}: $n'$ is canonical precisely if
%% it is of the form $\re_{\mbb{N}}\,n$ for some $n$.

%% For $\zero\rel$ and $\suc\rel$, we need to show that serialization preserves
%% $\zero$ and $\suc$ respectively, which is evident.

%% Let us look at elimination. We need to define the following:
%% \begin{alignat*}{4}
%%   &(\msf{NatElim}\,P\,z\,s\,n)\rel\,\gamma\rel :
%%   P\rel\,(\gamma\rel,\,n\rel\,\gamma\rel)\,(\ev\,(\msf{NatElim}\,P\,z\,s\,n)\,\gamma)\,
%%                                            (\re\,(\msf{NatElim}\,P\,z\,s\,n)\,\gamma')
%% \end{alignat*}
%% Unfolding $\ev$, we can further compute this to the following:
%% \begin{alignat*}{4}
%%   &(\msf{NatElim}\,P\,z\,s\,n)\rel\,\gamma\rel : \\
%%   &\hspace{2em}P\rel\,(\gamma\rel,\,n\rel\,\gamma\rel)\,
%%   (\msf{NatElim}\,(\lambda\,n.\,\ev\,P\,(\gamma,\,n))\,
%%                   (\ev\,z\,\gamma)\,
%%                   (\lambda\,n\,\mit{pn}.\,\ev\,s\,((\gamma,\,n),\,\mit{pn}))
%%                   (\ev\,n\,\gamma))\\
%%   &\hspace{8em}(\re\,(\msf{NatElim}\,P\,z\,s\,n)\,\gamma')
%% \end{alignat*}
%% In short, we need to show that $\msf{NatElim}$ preserves relations. Here we
%% switch to the external view temporarily. By Definition \ref{def:restriction}, we know that
%% \[
%%    |\re\,(\msf{NatElim}\,P\,z\,s\,n)|\,\gamma' = (\msf{NatElim}\,P\,z\,s\,n))[\gamma'].
%% \]
%% At the same time, we have $n\rel\,\gamma\rel : \re_{\mbb{N}}\,(\ev\,n\,\gamma) =
%% \re\,n\,\gamma'$, hence we know externally that $|\ev\,n|\,\gamma = n[\gamma']$.
%% In other words, $n[\gamma']$ is canonical and is obtained as the serialization
%% of $\ev\,n\,\gamma$. Therefore, $(\msf{NatElim}\,P\,z\,s\,n)[\gamma']$ is
%% definitionally equal to $|\ev\,n|\,\gamma$-many applications of $s$ to
%% $z$, and we can use $|\ev\,n|\,\gamma$-many applications of $s\rel$ to $z\rel$
%% to witness the goal type. The $\beta$-rules for $\msf{NatElim}$ are also respected
%% by this definition.

%% \subsubsection{Meta-Level $\Sigma$-Types} We define relatedness pointwise. Pairing and projection
%% are interpreted as meta-level pairing and projection.
%% \begin{alignat*}{3}
%%   &(\Sigma\,A\,B)\rel\,\gamma\rel : ((\alpha : \ev\,A\,\gamma) \times \ev\,B\,(\gamma,\,\alpha))
%%              \to \re\,(\Sigma\,A\,B)\,\gamma' \to \whset_j\\
%%   &(\Sigma\,A\,B)\rel\,\gamma\rel\,(\alpha,\,\beta)\,(\re_{\Sigma}^{-1}(\alpha',\,\beta')) :=
%%              (\alpha\rel : A\rel\,\gamma\rel\,\alpha\,\alpha') \times B\rel\,(\gamma\rel,\,\alpha\rel)\,\beta\,\beta'
%% \end{alignat*}

%% \subsubsection{Meta-Level $\Pi$-Types} We again use a pointwise definition. Note that we need to use
%% $\re_{\app}$ to apply $t' : \re\,(\Pi\,A\,B)\,\gamma'$ to $\alpha'$.
%% \begin{alignat*}{3}
%%   &(\Pi\,A\,B)\rel\,\gamma\rel : ((\alpha : \ev\,A\,\gamma) \to \ev\,B\,(\gamma,\,\alpha))
%%              \to \re\,(\Pi\,A\,B)\,\gamma' \to \whset_j\\
%%   &(\Pi\,A\,B)\rel\,\gamma\rel\,t\,t' :=
%%     (\alpha : \ev\,A\,\gamma)(\alpha' : \re\,A\,\gamma')(\alpha\rel : A\rel\,\gamma\rel\,\alpha\,\alpha') \to
%%     B\rel\,(\gamma\rel,\,\alpha\rel)\,(t\,\alpha)\,(\re_{\app}\,t'\,\alpha')
%% \end{alignat*}
%% For abstraction and application, we use a curry-uncurry definition:
%% \begin{alignat*}{4}
%%   &(\lam\,t)\rel\,\gamma\rel                &&:= \lambda\,\alpha\,\alpha'\,\alpha\rel.\,t\rel\,(\gamma\rel,\,\alpha\rel)\\
%%   &(\app\,t)\rel\,(\gamma\rel,\,\alpha\rel) &&:= t\rel\,\gamma\rel\,\alpha\,\alpha'\,\alpha\rel
%% \end{alignat*}

%% \subsubsection{Meta-Level Universes} We interpret $\U_{1,j}$ as a semantic relation space:
%% \begin{alignat*}{3}
%%   &(\U_{1,j})\rel\,\gamma\rel : \whset_j \to \re\,\U_{1,j}\,\gamma' \to \whset_{j+1} \\
%%   &(\U_{1,j})\rel\,\gamma\rel\,t\,t' := t \to \re_{\El}\,t' \to \whset_{j}
%% \end{alignat*}
%% Note that we have
%% \begin{alignat*}{3}
%%   & (\El\,t)\rel   &&: (\gamma\rel : \Gamma\rel\,\gamma\,\gamma') \to \ev\,t\,\gamma \to \re\,(\El\,t)\,\gamma' \to \whset_j\\
%%   & (\Code\,t)\rel &&: (\gamma\rel : \Gamma\rel\,\gamma\,\gamma') \to \ev\,t\,\gamma \to \re\,(\El\,t)\,\gamma' \to \whset_j.
%% \end{alignat*}
%% The types coincide because of the equation $\re\,(\El\,t)\,\gamma' =
%% \re_{\El}\,(\re\,t\,\gamma')$. Therefore we can interpret $\El$ and $\Code$ as
%% identity maps, as $(\El\,t)\rel := t\rel$ and $(\Code\,t)\rel := t\rel$.

%% \subsubsection{Object-Level Family Structure} We interpret extended contexts as follows.
%% \begin{alignat*}{3}
%%   & (\Gamma \ext_0 A)\rel\,(\gamma,\,\alpha)\,(\rextizero\,(\gamma',\,\alpha')) := (\gamma\rel : \Gamma\rel\,\gamma\,\gamma') \times (\emb{\alpha} = \alpha')
%% \end{alignat*}
%% Note that $\alpha : \Tm_{\mbbo j}\,(\ev\,A\,\gamma)$, so $\emb{\alpha} :
%% \Tm_{0,j}\,\emb{\ev\,A\,\gamma}$, but since $A\rel\,\gamma\rel :
%% \emb{\ev\,A\,\gamma} = \re\,A\,\gamma'$, we also have $\emb{\alpha} :
%% \Tm_{0,j}\,(\re\,A\,\gamma')$. Thus, the equation $\emb{\alpha} = \alpha'$ is
%% well-typed. For type substitution, we need
%% \[
%%   (A[\sigma])\rel\,\gamma\rel : \emb{\ev\,(A[\sigma])\,\gamma} = \re\,(A[\sigma])\,\gamma'.
%% \]
%% The goal type computes to $\emb{\ev\,A\,(\ev\,\sigma\,\gamma)} =
%% \re\,A\,(\re\,\sigma\,\gamma')$. This is obtained directly from
%% $A\rel\,(\sigma\rel\,\gamma\rel)$.  Similarly,
%% \begin{alignat*}{6}
%%   & (t[\sigma])\rel\,\gamma\rel &&:= t\rel\,(\sigma\rel\,\gamma\rel) && \p\rel\,(\gamma\rel,\,\alpha\rel) &&:= \gamma\rel \\
%%   & (\sigma,\,t)\rel\,\gamma\rel &&:= (\sigma\rel\,\gamma\rel,\,t\rel\,\gamma\rel)\hspace{2em}&& \q\rel\,(\gamma\rel,\,\alpha\rel) &&:= \alpha\rel
%% \end{alignat*}

%% \subsubsection{Lifting Structure}
%% \begin{alignat*}{3}
%%   &(\Lift\,A)\rel : \Gamma\rel\,\gamma\,\gamma' \to \Tm_{\mbbo j}\,(\ev\,A\,\gamma) \to \re\,(\Lift\,A)\,\gamma' \to \whset_j\\
%%   &(\Lift\,A)\rel\,\gamma\rel\,t\,t' := (\emb{t} = \spl t')
%% \end{alignat*}
%% This is well-typed by $A\rel\,\gamma\rel : \emb{\ev\,A\,\gamma} = \re\,A\,\gamma'$, which implies
%% that $\emb{t} : \Tm_{\mbbo j}\,(\re\,A\,\gamma')$. For $\qut{t}$, we need
%% \[  \qut{t}\rel\,\gamma\rel : \emb{\ev\,\qut{t}\,\gamma} = \spl(\re\,\qut{t}\,\gamma'). \]
%% The goal type can be further computed to $\emb{\ev\,t\,\gamma} = \re\,t\,\gamma'$, which we prove
%% by $t\rel\,\gamma\rel$. For splicing, we need
%% \[
%%   (\spl t)\rel\,\gamma\rel : \emb{\ev\,\spl t\,\gamma} = \re\,\spl t\,\gamma'
%% \]
%% where the goal type computes to $\emb{F\,t\,\gamma} = \spl(\re\,t\,\gamma')$,
%% but this again follows directly from $t\rel\,\gamma\rel$.

%% \subsubsection{Object-Level Type Formers}
%% Lastly, object-level type formers are straightforward. For types, we need
%% $\emb{\ev\,A\,\gamma} = \re\,A\,\gamma'$, and likewise for terms. Note that
%% $\ev$ and $\emb{\blank}$ preserve all structure, and $\re$ preserves all
%% structure on object-level types and terms. Hence, each object-level $A$ and $t$
%% case in $\blank\rel$ trivially follows from induction hypotheses.

%% This concludes the definition of the $\blank\rel$ interpretation.

%% \subsection{Soundness}

%% \begin{definition} First, we introduce shorthands for
%% external operations that can be obtained from $\blank\rel$.
%% \begin{alignat*}{5}
%%   & \text{For}\hspace{1em} \Gamma &&: \Con \hspace{1em}&&\text{we get}\hspace{1em}
%%          |\Gamma\rel| &&: \{\Delta : \Cono\} \to |\ev\,\Gamma|\,\Delta \to \Sub\,\emb{\Delta}\,\Gamma \to \Set_j\\
%%   & \text{For}\hspace{1em} A &&: \Ty_{0,j}\,\Gamma \hspace{1em}&&\text{we get}\hspace{1em} |A\rel| &&: |\Gamma\rel|\,\gamma\,\gamma' \to \emb{\ev\,A\,\gamma} = A[\gamma']\\
%%   & \text{For}\hspace{1em} t &&: \Tm_{0,j}\,\Gamma\,A \hspace{1em}&&\text{we get}\hspace{1em}|t\rel| &&: |\Gamma\rel|\,\gamma\,\gamma' \to \emb{\ev\,t\,\gamma} = t[\gamma']
%% \end{alignat*}
%% Since $\blank\rel$ was defined in $\hato$, we also know that the above are all
%% stable under object-theoretic substitution.
%% \end{definition}

%% \begin{theorem}[Soundness for generic contexts]
%% For each $\Gamma : \Cono$, we have ${\Gamma^P}\rel : |\Gamma\rel|\,\Gamma^P\,\id$.
%% \end{theorem}
%% \begin{proof}
%% We define ${\blank^P}\rel$ by induction on object-theoretic contexts.
%% ${\emptycon^P}\rel : \top$ is defined trivially as $\tt$. For ${(\Gamma \ext
%%   A)^P}\rel$, we need $(\gamma\rel : |\Gamma|\rel\,(\Gamma^P[\p])\,\p) \times
%% (\emb{\q} = \q)$. We get ${\Gamma^P}\rel : |\Gamma\rel|\,\Gamma^P\,\id$. Because
%% of the naturality of $|\Gamma\rel|$, this can be weakened to ${\Gamma^P}\rel[\p]
%% : |\Gamma\rel|\,(\Gamma^P[\p])\,\p$. Also, $\emb{\q} = \q$ holds immediately.
%% \end{proof}

%% \begin{theorem}[Soundness of staging]
%% The open staging algorithm from Definition \ref{def:open-staging} is sound.
%% \end{theorem}
%% \begin{proof}
%% \mbox{}
%% \begin{itemize}
%% \item For $A : \Ty_{0,j}\,\emb{\Gamma}$, we have that $\emb{\Stage\,A} =
%% \emb{|\ev\,A|\,\Gamma^P}$ by the definition of $\Stage$, and moreover we have
%% $|A\rel|\,{\Gamma^P}\rel : \emb{|\ev\,A|\,\Gamma^P} = A[\id]$, hence $\emb{\Stage\,A} = A$.

%% \item For $t : \Tm_{0,j}\,\emb{\Gamma}\,A$, using $|t\rel|\,{\Gamma^P}\rel : \emb{|\ev\,t|\,\Gamma^P} = t[\id]$, we likewise
%%   have $\emb{\Stage\,t} = \emb{|\ev\,t|\,\Gamma^P} = t[\id] = t$.
%% \end{itemize}
%% \end{proof}

%% \begin{corollary}[Conservativity of 2LTT]\label{conservativity}
%% From the soundness and stability of staging, we get that $\emb{\blank}$ is
%% bijective on types and terms, hence $\Ty_{\mbbo j}\,\Gamma \simeq \Ty_{0,j}\,\emb{\Gamma}$ and
%% $\Tm_{\mbbo j}\,\Gamma\,A \simeq \Tm_{0,j}\,\emb{\Gamma}\,\emb{A}$.
%% \end{corollary}

%% \paragraph{Alternative presentations}
%% The above conservativity proof could be rephrased as an instance of more modern
%% techniques which let us implicitly handle stability under 2LTT substitutions as
%% well, not just $\mbbo$ substitutions. This amounts to working in a \emph{modal}
%% internal language, where modalities control the interaction of the different base
%% categories.
%% \begin{itemize}
%% \item Using \emph{synthetic Tait computability} \cite{sterlingthesis}, we work
%%       in the internal language of the glued category $\msf{Id}_{\hato}\downarrow
%%       \re$.
%% \item Using \emph{relative induction} \cite{bocquet2021relative}, we work in the modal type theory
%%       obtained from the dependent right adjoint functor $\re^* : \wh{\msf{2LTT}} \to \hato$,
%%       where $\wh{\msf{2LTT}}$ denotes presheaves over the syntactic category of 2LTT.
%% \end{itemize}
%% We do not use either of these in this paper, for the following reasons. First,
%% the author of the paper is not sufficiently familiar with the above
%% techniques. Second, the task at hand is not too technically difficult, so using
%% advanced techniques is not essential. Contrast e.g.\ normalization for cubical
%% type theories, which is not feasible to handle without the more synthetic
%% presentations \cite{cubicalnbe}.

%% \section{Intensional Analysis}\label{sec:intensional-analysis}

%% We briefly discuss intensional analysis in this section. This means analyzing
%% the internal structure of object-level terms, i.e.\ values of $\Lift\,A$. Disallowing
%% intensional analysis is a major simplification, which is sometimes enforced
%% in implementations, for example in MetaML \cite{metaml}.

%% If we have sufficiently expressive inductive types in the meta-level language,
%% it is possible to simply use inductive deeply embedded syntaxes, which can be
%% analyzed; our example in Section \ref{sec:staged-programming} for embedding a
%% simply typed lambda calculus is like this. However, native metaprogramming
%% features could be more concise and convenient than deep embeddings, similarly to
%% how staged programming is more convenient than code generation with deeply
%% embedded syntaxes.

%% In the following, we look at the semantics of intensional analysis in the
%% standard presheaf models. We only discuss basic semantics here, and leave
%% practical considerations to future work.

%% \begin{definition}
%% The \textbf{Yoneda embedding} is a functor from $\mbbo$ to $\hato$, sending
%% $\Gamma : \Cono$ to $\msf{y}\Gamma : \hCon$, such that $|\msf{y}\Gamma|\,\Delta
%% = \Subo\,\Delta\,\Gamma$. We say that $\Gamma : \hCon$ is \emph{representable}
%% if it is isomorphic to a Yoneda-embedded context.
%% \end{definition}

%% \begin{lemma}[Yoneda lemma]
%% We have $\hSub\,(\msf{y}\Gamma)\,\Delta \simeq |\Delta|\,\Gamma$
%% \cite[Section~III.2]{maclane98categories}. Also, we have
%% $\hTm_{1,j}\,(\msf{y}\Gamma)\,A \simeq |A|\,\id$, where $\id :
%% \Subo\,\Gamma\,\Gamma$.
%% \end{lemma}

%% The Yoneda lemma restricts possible dependencies on representable contexts.  For
%% example, consider the staging behavior of $t : \Tm_{1}\,(\emptycon \ext (x :
%% \Bool_0))\,\Bool_1$. Staging $t$ yields essentially a natural transformation
%% $\hSub\,(\ev(\emptycon \ext \Bool_0))\,\mbb{B}$, where $\mbb{B}$ is the
%% metatheoretic type with two elements, considered as a constant presheaf.  Now,
%% if $\ev(\emptycon\,\ext\, \Bool_0)$ is representable, then
%% $\hSub\,(\ev(\emptycon \ext \Bool_0))\,\mbb{B} \simeq \mbb{B}$, so $t$ can be
%% staged in at most two different ways.

%% Which contexts are representable? In the specific 2LTT in this paper,
%% $\emb{\Gamma}$ is always representable in the presheaf
%% model. $|\ev{\emb{\Gamma}}|\,\Delta$ contains lists of object-theoretic terms,
%% hence $\ev{\emb{\Gamma}} \simeq \yon \Gamma$.  This implies that intensional
%% analysis is \emph{not compatible} with the standard presheaf model, for our
%% 2LTT.

%% Consider a very basic feature of intensional analysis, \emph{decidability of
%% definitional equality} for object-level Boolean expressions.
%% \[
%%   \msf{decEq} : (x\,y : \Lift\,\Bool_0) \to (x =_1 y) +_1 (x \neq_1 y)
%% \]
%% The Yoneda lemma says that $\msf{decEq}$ cannot actually decide
%% definitional equality. We expect that $\msf{decEq}$ lets us define non-constant
%% maps from $\Lift\,\Bool_0$ to $\Bool_1$, but the Yoneda lemma implies that we
%% only have two terms with distinct semantics in $\Tm_1\,(\emptycon \ext_0
%% \Bool_0)\,\Bool_1$.

%% A more direct way to show infeasibility of $\msf{decEq}$ in $\hato$ is to note
%% that definitional \emph{inequality} for object-level terms is not stable under
%% substitution, since inequal variables can be mapped to equal terms.

%% \paragraph{Stability under weakening only}
%% We can move to different 2LTT setups, where stability under substitution is not
%% required. We may have \emph{weakenings} only as morphisms in $\mbbo$.  A
%% weakening from $\Gamma$ to $\Delta$ describes how $\Gamma$ can be obtained by
%% inserting zero or more entries to $\Delta$. A wide range of intensional analysis
%% features are stable under weakenings. For example, $\msf{decEq}$ now holds in
%% $\hato$ whenever object-theoretic definitional equality is in fact decidable.
%% We also dodge the Yoneda lemma, since $\ev\,\emb{\Gamma}$ is not necessarily
%% representable anymore: $|\ev\,\emb{\Gamma}|\,\Delta$ is still a set of lists of
%% terms, but weakenings from $\Delta$ to $\Gamma$ are not lists of terms.

%% As a trade-off, if there is no notion of substitution in the specification of the
%% object theory, it is not possible to specify dependent or polymorphic types
%% there. We need a substitution operation to describe dependent elimination or
%% polymorphic instantiation.

%% However, stability under weakening is still sufficient for many practical use
%% cases, for example the monomorphization setup in Section
%% \ref{sec:monomorphization}. Here, we only have weakening as definitional
%% equality in the object theory, and we do not have $\beta\eta$-rules.

%% \paragraph{Closed modalities}\label{sec:closed-modalities}
%% Another option is to add a \emph{closed} or \emph{crisp} modality for modeling
%% closed object-level terms \cite{licata2018internal}. Since closed terms are not
%% affected by substitution, we are able to analyze them, and we can also recover
%% open terms by explicitly abstracting over contexts
%% \cite{DBLP:conf/lics/Hofmann99,DBLP:journals/corr/abs-2206-02831}. \emph{Function
%% pointers} in the style of the C programming language could be also an
%% interesting use case, since these must become closed after staging (as they
%% cannot capture closure environments).

%% %% for tracking object-level values which will become
%% %% closed terms after staging. Since closed terms are not affected by substitution,
%% %% we are free to analyze them. Such analysis can be viewed as an internalization
%% %% of canonicity for closed object-level terms. For example, a closed $\Bool_0$
%% %% term will be either $\true_0$ or $\false_0$ after staging, so it makes sense to
%% %% be able to eliminate from it to $\Bool_1$. This can be also viewed as a safe
%% %% implementation of a \emph{run} function in the style of MetaML \cite{metaml}.
%% %% In our case \emph{scope extrusion} would be prevented by the closed modality.

%% \section{Related Work}\label{sec:related-work}

%% The work Annekov et al.\ on two-level type theory \cite{twolevel}, building on
%% \cite{capriotti2017models} and \cite{hts}, is the foremost inspiration in
%% the current work. Interestingly, the above works do mention metaprogramming as a
%% source of intuition for 2LTT, but they only briefly touch on this aspect, and in
%% \cite{twolevel} the main focus is on extensions of basic 2LTT which do not have
%% staging semantics anymore. Ibid.\ conjectures the strong conservativity of 2LTT
%% in Proposition 2.18, and suggests categorical gluing as proof strategy. In this
%% work, we do essentially that: our presheaf model and the logical relation model
%% could be merged into a single instance of gluing along the $\re$ morphism; this
%% merges staging and its soundness proof. We split this construction to two parts
%% in order to get a self-contained presentation of the staging algorithm.

%% \cite{DBLP:conf/lics/Hofmann99} models higher-order abstract syntax of simply
%% typed object languages in presheaves over various syntactic categories. The main
%% difference between HOAS and 2LTT is that the former represents object-level binders
%% using meta-level functions while the latter has a primitive notion of object-level
%% binding.

%% The multi-stage calculus in \cite{multi-stage-calculus} supports staging
%% operations and type dependencies. However, there are major differences to the
%% current work. First, ibid.\ does not have large elimination, hence it does not
%% support computing types by staging. Second, it does not support staged
%% compilation in the sense of Definition \ref{def:staging}; there is no judgment
%% which ensures that staging operations can be eliminated and a purely
%% ``object-level'' program can be extracted from a multi-stage program. Rather, the system
%% supports \emph{runtime} code generation. There are several modal multi-stage
%% systems which support runtime code generation but not staged compilation, such
%% as those based on S4 modalities (e.g.\ \cite{DBLP:journals/jacm/DaviesP01}) or
%% contextual modalities (\cite{DBLP:journals/pacmpl/JangGMP22}).

%% %% We make a
%% %% disctinction between staged compilation and staged computation. For the former,
%% %% we expect staging in the sense of Definition \ref{def:staging}; this is
%% %% essential if we want to use staging as a lowering step in a compilation
%% %% pipeline. Staged computation on the other hand may support runtime code
%% %% construction and evaluation, but may not guarantee that these features can be
%% %% entirely computed away.

%% Idris 1 supports compile-time partial evaluation, through \texttt{static}
%% annotations on function arguments \cite{scrap-your-inefficient-engine}. This can
%% be used for dependently typed code generation, but Idris 1 does not guarantee
%% that partial evaluation adequately progresses. For example, we can pass a neutral
%% value as static argument in Idris 1, while in 2LTT static function arguments
%% are always canonical during staging.

%% Our notation for quoting and splicing is borrowed from MetaML \cite{metaml}.  In
%% the following, we compare 2LTT to MetaML, MetaOCaml \cite{kiselyov14metaocaml}
%% and typed Template Haskell \cite{typed-th}.

%% 2LTT explicitly tracks stages of types, in contrast to the mentioned
%% systems. There, we have type lifting (usually called $\msf{Code}$), quoting and
%% splicing, but lifting does not change the universe of types. We can write a
%% function with type $\Bool \to \msf{Code}\,\Bool$, and the two $\Bool$
%% occurrences refer to the exact same type.

%% This causes some operational confusion down the line, and additional
%% disambiguation is needed for stages of binders. For example, in typed Template
%% Haskell, top-level binders behave as runtime binders when used under a quote,
%% but behave as static binders otherwise. Given a top-level definition $b =
%% \True$, if we write a top-level definition $\msf{expr} = [||\,b\,||]$, then $b$
%% is a \emph{variable} in the staging output, but if we write $\msf{expr} =
%% \msf{if}\,b\,\msf{then}\,[||\,\True\,||]\,\msf{else}\,[||\,\False\,||]$, then
%% $b$ is considered to be a static value. In contrast, 2LTT does not distinguish
%% top-level and local binders, and in fact it has no syntactic or scope-based
%% restrictions; everything is enforced by typing.

%% 2LTT supports staging for types and dependent types, while to our knowledge no
%% previous system for staged compilation does the same. It appears to the author
%% that the main missing component in previous systems is \emph{dependent types at
%% the meta level}, rather than staging operations; staging operations in 2LTT and
%% MetaOCaml are already quite similar. For example, any interesting operation on
%% the static-length vectors in Section \ref{sec:staged-programming} requires
%% meta-level dependent types, even if the object language is simply typed.

%% In MetaOCaml, there is a \emph{run} function for evaluating closed object-level
%% terms, however, this can fail during staging since closedness is not statically
%% ensured. In our setting, this could be reproduced in a safe way using a closed
%% modality; this is part of future work.

%% %% as we mentioned in Section \ref{sec:closed-modalities}, we
%% %% believe that the proper way to implement it in 2LTT would be to use a closed
%% %% modality. This use case is closely related to \cite{licata2018internal}. In
%% %% ibid.\ a semantics is sketched for this modality in a presheaf topos, and we
%% %% could directly adapt it to our presheaf model of 2LTT. That said, working
%% %% out the practical details of this modality remains future work.

%% Additionally, 2LTT only supports two stages, while the other noted systems allow
%% countable stages. It remains future work to develop NLTT (N-level type theory),
%% but this extension does appear to be straightforward. In the specification of
%% NLTT, we would simply need to move from $i \in \{0,1\}$ to $i \in \mbb{N}$,
%% although in the semantics there could be more subtleties.

%% Many existing staged systems also support \emph{side effects} at compile time,
%% while our version of 2LTT does not. Here, general considerations about the
%% mixing of dependent types and effects should apply; see
%% e.g.\ \cite{fire-triangle}. However, it should be possible to use standard
%% effect systems such as monads at the meta level.

%% \section{Conclusions and Future Work}\label{sec:conclusions}

%% We developed the foundational metatheory of 2LTT as a system of
%% two-stage compilation. We view the current results as more like a starting point
%% to more thorough investigation into applications and extensions; in this paper
%% we only sketched these.

%% We emphasize that variants of 2LTT can serve as languages where staging
%% features can be practically implemented, and also as \emph{metatheories}, not
%% only for the formalization for staging techniques, but also more generally for
%% reasoning about constructions in object languages. This purpose is prominent in
%% the original 2LTT use case in synthetic homotopy theory. The meta-level language
%% in a 2LTT can be expressive enough to express general mathematics, and the
%% meta-level propositional equality $\blank\!=_1\!\blank$ can be used to prove
%% $\beta\eta$-equality of object-level types and terms.

%% For example, we could use object theories as \emph{shallowly embedded} target
%% languages of optimization techniques and program translations. In particular, we
%% may choose a low-level object theory without higher-order functions, and
%% formalize closure conversion as an interpretation of an embedded meta-level
%% syntax into the object theory.

%% There is much future work in adapting existing staging techniques to 2LTT
%% (which, as we mentioned, can double as formalization of said techniques), or
%% adding staging where it has not been available previously.
%% \begin{itemize}
%% \item \emph{Let-insertion techniques}. These allow more precise control over
%%   positions of insertion, or smarter selection of such positions. In 2LTT,
%%   meta-level continuation monads could be employed to adapt some of the features
%%   in \cite{DBLP:journals/jfp/KameyamaKS11}. The automatic let-floating feature
%%   of MetaOCaml \cite{DBLP:journals/corr/abs-2201-00495} requires strengthening
%%   of terms, which is not stable under substitution, but it is stable under
%%   weakening, so perhaps it could be implemented as an ``intensional analysis''
%%   feature.
%% \item \emph{Generic programming with dependent types}. There is substantial
%%   literature in this area
%%   (e.g.\ \cite{loh11generic,chapman2010gentle,diehl-thesis,ornaments}), but only
%%   in non-staged settings.  2LTT should immediately yield staging for these
%%   techniques, assuming that the meta level has sufficient type formers
%%   (e.g.\ induction-recursion). We could also try to adapt previous work on
%%   generic treatment of partially static data types
%%   \cite{DBLP:journals/pacmpl/YallopGK18}; fully internalizing this in 2LTT would
%%   also require dependent types.
%% \end{itemize}

%% There are also ways that 2LTT itself could be enhanced.
%% \begin{itemize}
%% \item \emph{More stages, stage polymorphism}. Currently, there is substantial
%%   code duplication if the object-level and meta-level languages are similar.
%%   Stage polymorphism could help reduce this. A simple setup could include three
%%   stages, the runtime one, the static one, and another one which supports
%%   polymorphism over the first two stages.
%% \item \emph{Modalities}. We mentioned the closed modality in Section
%%   \ref{sec:closed-modalities}.  More generally, many \emph{multimodal}
%%   \cite{gratzer20multimodal} type theories could plausibly support staging. 2LTT
%%   itself can be viewed as a very simple multimodal theory with $\Lift$ as the
%%   only modality, which is also degenerate (because it entails no structural
%%   restriction on contexts). We could support multiple object theories, with
%%   modalities for lifting them to the meta level or for representing translations
%%   between them. We could also have modalities for switching between stability
%%   under substitution and stability under weakening.
%% \item \emph{Intensional analysis}. We only touched on the most basic semantics
%%   of intensional analysis in Section \ref{sec:intensional-analysis}. It remains
%%   a substantial challenge to work out the practicalities. For good ergonomics,
%%   we would need something like a ``pattern matching'' operation on object-level
%%   terms, or some induction principle which is more flexible than the plain
%%   assumption of decidable object-level equality.
%% \end{itemize}

%% \section{Data Availability Statement}

%% A standalone prototype implementation is available as \cite{staged-demo}. It
%% implements elaboration and staging for a 2LTT. It includes a tutorial and code
%% examples that expand on the contents of Section \ref{sec:tour-of-2ltt}.

\interlinepenalty=10000
\bibliography{references}

\end{document}
\endinput

### 2024-02-03 newpaper/hs-agda/Notes.hs

@@ -1,0 +1,107 @@
+
+{-# language Strict, LambdaCase, BlockArguments, OverloadedStrings #-}
+{-# options_ghc -Wincomplete-patterns #-}
+
+module Notes where
+
+import Data.Maybe
+import Data.String
+
+type Name = String
+
+data Tm
+  = Var Name
+  | App Tm Tm
+  | Lam Name Tm
+  | Let Name Tm Tm
+  deriving Show
+
+{-
+We want to apply the following transformations everywhere
+
+  let x = y in t   -- where y is a variable
+  ~>
+  t[y/x]
+
+  let x = (let y = t in u) in v
+  ~>
+  let y = t in let x = u in v
+
+  (let x = t in u) v
+  ~>
+  let x = t in u v
+
+  (\x -> t) u
+  ~>
+  let x = u in t
+
+This is possible purely with NbE in an efficient and very easy way, where we
+don't have to manually handle variables and substitution. Instead, every binder
+looks like a plain function.
+
+nThe main limitation is that we can't look "under" a binder to check whether its
+body is of a particular form. However, many things can be implemented without
+looking under binders, and we can work around some limitations by including more
+information in values and Val binders (which I don't demonstrate here).
+-}
+
+data Val
+  = VVar Name
+  | VApp Val Val
+  | VLam Name (Val -> Val)
+  | VLet Name Val (Val -> Val)
+
+type Env = [(Name, Val)]
+
+vapp :: Val -> Val -> Val
+vapp t u = case t of
+  VLam x t   -> vlet x u t
+  VLet x t v -> vlet x t \x -> vapp (v x) u
+  t          -> VApp t u
+
+vlet :: Name -> Val -> (Val -> Val) -> Val
+vlet x t u = case t of
+  VVar _        -> u t
+  VLet x' t' u' -> vlet x' t' \x' -> vlet x (u' x') u
+  t             -> VLet x t u
+
+eval :: Env -> Tm -> Val
+eval e = \case
+  Var x     -> fromJust $ lookup x e
+  App t u   -> vapp (eval e t) (eval e u)
+  Lam x t   -> VLam x \v -> eval ((x, v):e) t
+  Let x t u -> vlet x (eval e t) \v -> eval ((x, v):e) u
+
+fresh :: [Name] -> Name -> Name
+fresh ns x | elem x ns = fresh ns (x ++ "'")
[…]

### 2024-02-03 newpaper/hs-agda/ObjectInterpreter.hs

@@ -1,0 +1,83 @@
+{-# language Strict, LambdaCase #-}
+{-# options_ghc -Wincomplete-patterns #-}
+
+module ObjectInterpreter where
+import Data.Maybe
+import GHC.Stack
+
+data F = K ValTy | Id | ProdF F F | SumF F F
+  deriving Show
+
+data ValTy = One | Mu F | Prod ValTy ValTy | Sum ValTy ValTy
+  deriving Show
+
+data Ty = Fun ValTy Ty | Val ValTy
+  deriving Show
+
+type Name = String
+
+data Tm
+  = Var Name
+  | Let Name Ty Tm Tm
+  | Fix Name Tm            -- fix only for functions!
+  | Lam Name Tm
+  | App Tm Tm
+  | Pair Tm Tm
+  | Fst Tm
+  | Snd Tm
+  | Tt
+  | Inl Tm
+  | Inr Tm
+  | Case Tm Name Tm Name Tm
+  | In Tm
+  | Out Tm
+  deriving Show
+
+--------------------------------------------------------------------------------
+
+data Val = VTt | VPair Val Val | VInl Val | VInr Val | VIn Val
+  deriving Show
+
+type ValEnv   = [(Name, Val)]
+type FunSig   = [(Name, Tm)]
+type ArgStack = [Val]
+
+impossible :: HasCallStack => a
+impossible = error "impossible"
+
+lookupVar :: HasCallStack => Name -> [(Name, a)] -> a
+lookupVar x kvs = case lookup x kvs of
+  Nothing -> impossible
+  Just a  -> a
+
+eval :: ValEnv -> FunSig -> Tm -> ArgStack -> Val
+eval vs fs t args = case t of
+  Var x -> case args of
+    [] -> lookupVar x vs
+    _  -> eval vs fs (lookupVar x fs) args
+  Let x a t u -> case a of
+    Val _ -> let v = eval vs fs t [] in eval ((x,v):vs) fs u args
+    _     -> eval vs ((x,t):fs) u args
+  Fix x t -> eval vs ((x, Fix x t):fs) t args
+  Lam x t -> case args of
+    v:args -> eval ((x,v):vs) fs t args
+    _      -> impossible
+  App t u -> let v = eval vs fs u [] in eval vs fs t (v:args)
+  Pair t u -> VPair (eval vs fs t []) (eval vs fs u [])
+  Fst t -> case eval vs fs t [] of
+    VPair t _ -> t
[…]

### 2024-02-05 dept23prez/prez.tex
just push everything

\documentclass[dvipsnames,aspectratio=169]{beamer}
\usetheme{Madrid}

%% kill footline
\setbeamertemplate{footline}[frame number]{}
\setbeamertemplate{navigation symbols}{}
\setbeamertemplate{footline}{}

%% bibliography
\bibliographystyle{alpha}
\setbeamerfont{bibliography item}{size=\footnotesize}
\setbeamerfont{bibliography entry author}{size=\footnotesize}
\setbeamerfont{bibliography entry title}{size=\footnotesize}
\setbeamerfont{bibliography entry location}{size=\footnotesize}
\setbeamerfont{bibliography entry note}{size=\footnotesize}
\setbeamertemplate{bibliography item}{}

%% kill ball enumeration
\setbeamertemplate{enumerate items}[circle]
\setbeamertemplate{section in toc}[circle]

%% kill block shadows
\setbeamertemplate{blocks}[rounded][shadow=false]
\setbeamertemplate{title page}[default][colsep=-4bp,rounded=true]

%% kill ball itemize
\setbeamertemplate{itemize items}[circle]

%% --------------------------------------------------------------------------------

\usepackage[utf8]{inputenc}
\usepackage{hyperref}
\usepackage{amsmath}
\usepackage{cite}
\usepackage{amsthm}
\usepackage{amssymb}
\usepackage{amsfonts}
\usepackage{mathpartir}
\usepackage{scalerel}
\usepackage{stmaryrd}
\usepackage{bm}
\usepackage{graphicx}
\usepackage{fontspec}

\hypersetup{
  colorlinks=true,
  linkcolor=blue,
  urlcolor=blue}

\setmonofont[SizeFeatures={Size=8}]{DejaVu Sans Mono}

%% --------------------------------------------------------------------------------

\renewcommand{\mit}[1]{\mathit{#1}}
\newcommand{\msf}[1]{\mathsf{#1}}
\newcommand{\mbb}[1]{\mathbb{#1}}
\newcommand{\mbf}[1]{\mathbf{#1}}
\newcommand{\bs}[1]{\boldsymbol{#1}}
\newcommand{\wh}[1]{\widehat{#1}}
\newcommand{\ext}{\triangleright}
\newcommand{\Code}{\msf{Code}}
\newcommand{\El}{\msf{El}}
\newcommand{\lam}{\msf{lam}}
\newcommand{\app}{\msf{app}}
\newcommand{\NatElim}{\msf{NatElim}}
\newcommand{\y}{\msf{y}}

\newcommand{\Lift}{{\Uparrow}}
\newcommand{\spl}{{\sim}}
\newcommand{\qut}[1]{{<}#1{>}}

\newcommand{\mbbc}{\mbb{C}}
\newcommand{\mbbo}{\mbb{O}}
\newcommand{\ob}{_\mbbo}

\newcommand{\U}{\msf{U}}
\newcommand{\Con}{\msf{Con}}
\newcommand{\Sub}{\msf{Sub}}
\newcommand{\Ty}{\msf{Ty}}
\newcommand{\Tm}{\msf{Tm}}
\newcommand{\Cono}{\msf{Con}_{\mbbo}}
\newcommand{\Subo}{\msf{Sub}_{\mbbo}}
\newcommand{\Tyo}{\msf{Ty}_{\mbbo}}
\newcommand{\Tmo}{\msf{Tm}_{\mbbo}}
\newcommand{\hCon}{\wh{\msf{Con}}}
\newcommand{\hSub}{\wh{\msf{Sub}}}
\newcommand{\hTy}{\wh{\msf{Ty}}}
\newcommand{\hTm}{\wh{\msf{Tm}}}

\newcommand{\p}{\mathsf{p}}
\newcommand{\q}{\mathsf{q}}

\newcommand{\refl}{\msf{refl}}
\newcommand{\Bool}{\msf{Bool}}
\newcommand{\true}{\msf{true}}
\newcommand{\false}{\msf{false}}
\newcommand{\True}{\msf{True}}
\newcommand{\False}{\msf{False}}
\newcommand{\List}{\msf{List}}
\newcommand{\nil}{\msf{nil}}
\newcommand{\cons}{\msf{cons}}
\newcommand{\Nat}{\msf{Nat}}
\newcommand{\zero}{\msf{zero}}
\newcommand{\suc}{\msf{suc}}
\renewcommand{\tt}{\msf{tt}}
\newcommand{\fst}{\msf{fst}}
\newcommand{\snd}{\msf{snd}}
\newcommand{\mylet}{\msf{let}}
\newcommand{\emptycon}{\scaleobj{.75}\bullet}
\newcommand{\id}{\msf{id}}

\newcommand{\Set}{\mathsf{Set}}
\newcommand{\Prop}{\mathsf{Prop}}
\newcommand{\Rep}{\msf{Rep}}
\newcommand{\blank}{{\mathord{\hspace{1pt}\text{--}\hspace{1pt}}}}
\newcommand{\emb}[1]{\ulcorner#1\urcorner}

\newcommand{\Stage}{\msf{Stage}}
\newcommand{\hato}{\bm\hat{\mbbo}}
\newcommand{\ev}{\mbb{E}}
\newcommand{\re}{\mbb{R}}

\theoremstyle{remark}
\newtheorem{notation}{Notation}

\newcommand{\whset}{\wh{\Set}}
\newcommand{\rexti}{\re_{\ext_1}^{-1}}
\newcommand{\rextizero}{\re_{\ext_0}^{-1}}

\newcommand{\rel}{^{\approx}}
\newcommand{\yon}{\msf{y}}

\newcommand{\Vect}{\msf{Vec}}
\newcommand{\msA}{\msf{A}}
\newcommand{\msB}{\msf{B}}
\newcommand{\mst}{\msf{t}}
\newcommand{\msu}{\msf{u}}
\newcommand{\msv}{\msf{v}}
\newcommand{\msx}{\msf{x}}
\newcommand{\msy}{\msf{y}}
\newcommand{\msz}{\msf{z}}
\newcommand{\msas}{\msf{as}}
\newcommand{\msbs}{\msf{bs}}
\newcommand{\msxs}{\msf{xs}}
\newcommand{\msys}{\msf{ys}}
\newcommand{\msff}{\msf{f}}
\newcommand{\msg}{\msf{g}}
\newcommand{\msh}{\msf{h}}
\newcommand{\msa}{\msf{a}}
\newcommand{\msn}{\msf{n}}

%% --------------------------------------------------------------------------------

\title{Dysfunctional Programming}
\author{András Kovács}
\institute{
  {Eötvös Loránd University, Dept. of Programming Languages and Compilers}
}
\date{9 February 2023, Department Workshop}
\begin{document}

\frame{\titlepage}

%% ------------------------------------------------------------

\begin{frame}{First-class functions}

A huge amount of software abstraction is based on first-class functions:
\begin{itemize}
\item HOFs in FP
\item Virtual methods in OOP
\item Callbacks, closures in many settings
\end{itemize}
\vspace{1em}
\pause

Lots of headache in compilation:
\begin{itemize}
\item Dynamic calls block static analysis \& inlining
\item Implementation of closures/vtables adds complexity
\end{itemize}
\vspace{1em}
\pause

\begin{block}{}
\textbf{What can we do without first-class functions?}
\end{block}

\end{frame}

%% ------------------------------------------------------------

\begin{frame}[fragile]{}

Do we need function values here?

\begin{exampleblock}{}
\begin{verbatim}
    map : (A → B) → List A → List B
    map f nil         = nil
    map f (cons a as) = cons (f a) (map f as)

    g : List Int → List Int
    g = map (+1)
\end{verbatim}
\end{exampleblock}
\pause

Instead:

\begin{exampleblock}{}
\begin{verbatim}
    g : List Int → List Int
    g nil         = nil
    g (cons a as) = cons (a + 1) (g as)
\end{verbatim}
\end{exampleblock}
\pause
Function elimination by inlining.

\end{frame}

%% ------------------------------------------------------------

\begin{frame}{Getting rid of function values}

  \begin{enumerate}
  \item Metaprogramming: use first-class functions only at compile time, generate first-order code.
        Inlining is a special case of this.
  \item Defunctionalization: replace closures by explicit first-order data.
  \end{enumerate}
\vspace{1em}

We focus on \textbf{1}, using \textbf{two-level type theory}.
\vspace{1em}
\pause

There is a dependently typed compile-time language for metaprogramming.
\vspace{1em}
\pause

Object-level \& meta-level code is freely mixed in programs.
\vspace{1em}
\pause

Generated code is guaranteed to be \emph{well-typed}, \emph{simply typed} and \emph{only have first-order functions}.

\end{frame}

%% ------------------------------------------------------------

\begin{frame}[fragile]{}
\begin{block}{Rules, informally}
  \begin{itemize}
  \item Compile-time types and runtime types have different types (``universes'').
  \item Function and non-function (``value'') types at runtime stage are distinguished.
  \item There are only first-order function types at runtime.
  \item We can't store functions in runtime data.
  \end{itemize}
\end{block}

\begin{exampleblock}{}
\begin{verbatim}
    data List (A : Ty Val) : Ty Val where
      nil  : List A
      cons : A → List A → List A

    map : {A B : Ty Val} → (A → B) → List A → List B

    g : List Int -> List Int
    g := map (+1)
\end{verbatim}
\end{exampleblock}
\end{frame}

%% ------------------------------------------------------------

\begin{frame}[fragile]{Is List a Functor/Applicative/Monad?}

OK:
\begin{exampleblock}{}
\begin{verbatim}
    class Functor (F : Ty Val → Ty Val) where
      fmap : (A → B) → F A → F B
\end{verbatim}
\end{exampleblock}
\pause

Issue with \texttt{Applicative}:
\begin{exampleblock}{}
\begin{verbatim}
    class Applicative (F : Ty Val → Ty Val) where
      pure : A → F A
      ap   : F (A → B) → F A → F B
\end{verbatim}
\end{exampleblock}
\pause

Solution:

\begin{exampleblock}{}
\begin{verbatim}
    class Applicative (F : Ty Val → Ty Val) where
      point : F ()
      zip   : F A → F B → F (A, B)
\end{verbatim}
\end{exampleblock}
\pause

Monad is also OK:

\begin{exampleblock}{}
\begin{verbatim}
    bind : M A → (A → M B) → M B
\end{verbatim}
\end{exampleblock}
\end{frame}

\begin{frame}[fragile]{Is State a Functor/Applicative/Monad?}

\begin{exampleblock}{}
\begin{verbatim}
    State : Ty Val → Ty Val → MetaTy
    State S A = S → (A, S)
\end{verbatim}
\end{exampleblock}
\vspace{1em}

Using the previous Functor/Applicative/Monad definitions, no!
\vspace{1em}
\pause

But it is a \emph{relative} Functor/Applicative/Monad:

\begin{exampleblock}{}
\begin{verbatim}
    class RelFunctor (F : Ty Val → MetaTy) where ...
    class RelApplicative (F : Ty Val → MetaTy) where ...
    class RelMonad (M : Ty Val → MetaTy) where ...
    instance RelMonad (State S) where ...
\end{verbatim}
\end{exampleblock}
\end{frame}

\begin{frame}[fragile]{Defunctionalization}

Difference lists in Haskell:
\vspace{1em}

\begin{exampleblock}{}
\begin{verbatim}
    type DList a = [a] -> [a]

    append :: DList a -> DList a -> DList a
    append xs ys = xs . ys

    toDList a as = a : as
    fromDList xs = xs []
\end{verbatim}
\end{exampleblock}
\end{frame}

\begin{frame}[fragile]{Defunctionalization}

No functions needed:
\vspace{1em}

\begin{exampleblock}{}
\begin{verbatim}
    data DList a = Chunk [a] | Append (DList a) (DList a)

    apply :: DList a -> [a] -> [a]
    apply (Chunk xs)     ys = xs ++ ys
    apply (Append xs ys) zs = apply xs (apply ys zs)

    toDList :: [a] -> DList a
    toDList xs = Chunk xs

    fromDList :: DList a -> [a]
    fromDList xs = apply xs []
\end{verbatim}
\end{exampleblock}
\end{frame}

\begin{frame}{Summary \& further topics}

\begin{itemize}
\item Not having first-class functions makes compilation much easier.
  \begin{itemize}
  \item In most cases: metaprogramming can replace first-class functions.
  \item In most remaining cases: defunctionalization can replace first-class functions.
  \item In the remaining cases: we can distinguish closures \emph{in the type system}.
  \end{itemize}
  \vspace{1em}
\item Trade-off between programming convenience and performance.
  \vspace{1em}
\item Trade-off between code size and performance.
  \vspace{1em}
\item Other things to try to do without functions:
  \begin{itemize}
  \item Fusion optimizations.
  \item More things from the Haskell/ML literature.
  \end{itemize}
\end{itemize}

\end{frame}
\end{document}

### 2024-02-05 icfp22paper/revisions.tex
just push everything

\documentclass[12pt]{article}

\usepackage[left=2cm, right=2cm, top=2cm, bottom=2cm]{geometry}

\begin{document}

\pagenumbering{gobble}

\subsubsection*{Revisions to the paper ``Staged Compilation With Two-Level Type Theory''}
\vspace{1em}

\begin{itemize}
\item
  I added an appendix as a supplement, which includes a complete listing of 2LTT
  rules and equations in a more traditional derivation notation. The appendix
  also includes a more compact specification of the same thing using a
  second-order algebraic signature.
\item
  I extended the beginning of Section 3 with a motivation of algebraic models and quotients, and      a description of extracting algorithms via a setoid interpretation.
\item
  I added a bit more explanation to "Models and Syntax of 2LTT" in Section 3.3,
  trying to explain syntactic aspects.
\item
  I added Section 3.3.2 as a short discussion of elaborating surface syntax to formal syntax,
  as a justification for the usage of surface syntax.
\item
  I added Section 3.3.3 for a sketch of interpreting the syntax of 2LTT in setoids.
\item I rephrased the correctness of staging, now I talk about soundness,
  stability and "strictness" in Definition 4.2. Strictness means that staging
  strictly preserves all object-level type and term formers, where "strictly"
  refers to equality in underlying non-quotiented types and terms. This captures
  that staging should not compute object-level redexes. As far as I see, it is
  easy to show strictness of staging so I added it as Theorem 4.16.
\item
  I reworded Section 4.4 to be hopefully clearer and closer to what I actually
  did in the demo implementation.
\end{itemize}

%% Lorem ipsum dolor sit amet, consectetur adipiscing elit, sed do eiusmod tempor incididunt ut labore et dolore magna aliqua. Ut enim ad minim veniam, quis nostrud exercitation ullamco laboris nisi ut aliquip ex ea commodo consequat. Duis aute irure dolor in reprehenderit in voluptate velit esse cillum dolore eu fugiat nulla pariatur. Excepteur sint occaecat cupidatat non proident, sunt in culpa qui officia deserunt mollit anim id est laborum.

%% Lorem ipsum dolor sit amet, consectetur adipiscing elit, sed do eiusmod tempor incididunt ut labore et dolore magna aliqua. Ut enim ad minim veniam, quis nostrud exercitation ullamco laboris nisi ut aliquip ex ea commodo consequat. Duis aute irure dolor in reprehenderit in voluptate velit esse cillum dolore eu fugiat nulla pariatur. Excepteur sint occaecat cupidatat non proident, sunt in culpa qui officia deserunt mollit anim id est laborum.

%% Lorem ipsum dolor sit amet, consectetur adipiscing elit, sed do eiusmod tempor incididunt ut labore et dolore magna aliqua. Ut enim ad minim veniam, quis nostrud exercitation ullamco laboris nisi ut aliquip ex ea commodo consequat. Duis aute irure dolor in reprehenderit in voluptate velit esse cillum dolore eu fugiat nulla pariatur. Excepteur sint occaecat cupidatat non proident, sunt in culpa qui officia deserunt mollit anim id est laborum.

\end{document}

### 2024-02-05 hipe-doc/lean4/maptest.c

@@ -1,0 +1,81 @@
+// Lean compiler output
+// Module: maptest
+// Imports: Init
+#include <lean/lean.h>
+#if defined(__clang__)
+#pragma clang diagnostic ignored "-Wunused-parameter"
+#pragma clang diagnostic ignored "-Wunused-label"
+#elif defined(__GNUC__) && !defined(__CLANG__)
+#pragma GCC diagnostic ignored "-Wunused-parameter"
+#pragma GCC diagnostic ignored "-Wunused-label"
+#pragma GCC diagnostic ignored "-Wunused-but-set-variable"
+#endif
+#ifdef __cplusplus
+extern "C" {
+#endif
+uint64_t lean_uint64_add(uint64_t, uint64_t);
+LEAN_EXPORT lean_object* l_mymap(lean_object*);
+LEAN_EXPORT lean_object* l_mymap(lean_object* x_1) {
+_start:
+{
+if (lean_obj_tag(x_1) == 0)
+{
+lean_object* x_2; 
+x_2 = lean_box(0);
+return x_2;
+}
+else
+{
+uint8_t x_3; 
+x_3 = !lean_is_exclusive(x_1);
+if (x_3 == 0)
+{
+lean_object* x_4; lean_object* x_5; uint64_t x_6; uint64_t x_7; uint64_t x_8; lean_object* x_9; lean_object* x_10; 
+x_4 = lean_ctor_get(x_1, 0);
+x_5 = lean_ctor_get(x_1, 1);
+x_6 = 100;
+x_7 = lean_unbox_uint64(x_4);
+lean_dec(x_4);
+x_8 = lean_uint64_add(x_7, x_6);
+x_9 = l_mymap(x_5);
+x_10 = lean_box_uint64(x_8);
+lean_ctor_set(x_1, 1, x_9);
+lean_ctor_set(x_1, 0, x_10);
+return x_1;
+}
+else
+{
+lean_object* x_11; lean_object* x_12; uint64_t x_13; uint64_t x_14; uint64_t x_15; lean_object* x_16; lean_object* x_17; lean_object* x_18; 
+x_11 = lean_ctor_get(x_1, 0);
+x_12 = lean_ctor_get(x_1, 1);
+lean_inc(x_12);
+lean_inc(x_11);
+lean_dec(x_1);
+x_13 = 100;
+x_14 = lean_unbox_uint64(x_11);
+lean_dec(x_11);
+x_15 = lean_uint64_add(x_14, x_13);
+x_16 = l_mymap(x_12);
[…]

### 2024-02-05 icfp22paper/sub/demo/Main.hs

@@ -1,0 +1,87 @@
+
+module Main where
+
+import Control.Exception
+import System.Environment
+import System.Exit
+
+import Common
+import Cxt
+import Errors
+import Evaluation
+import Metacontext
+import Parser
+import Pretty
+import Elaboration
+import Staging
+
+import qualified Presyntax as P
+
+--------------------------------------------------------------------------------
+
+helpMsg = unlines [
+  "usage: 2ltt COMMAND",
+  "",
+  "All commands read input from stdin. The input must be a single 2LTT",
+  "expression at stage 0.",
+  "",
+  "Commands:",
+  "  elab          : print elaboration output",
+  "  elab-verbose  : print elaboration output, show metavars, inserted type",
+  "                  annotations, implicit lambdas and applications",
+  "  stage         : print staging output",
+  "  stage-verbose : print staging output, show inserted implicits",
+  "  nf            : print normal form",
+  "  type          : print normal type of input"
+  ]
+
+mainWith :: IO [String] -> IO (P.Tm, String) -> IO ()
+mainWith getOpt getRaw = do
+
+  let elab = do
+        (!t, !file) <- getRaw
+        inferS (emptyCxt (initialPos file)) t S0
+          `catch` \e -> displayError file e >> exitSuccess
+
+  reset
+  getOpt >>= \case
+    ["--help"] -> putStrLn helpMsg
+    ["elab"] -> do
+      (!t, !a) <- elab
+      putStrLn "-- elaboration output"
+      putStrLn "------------------------------------------------------------"
+      putStrLn ""
+      -- putStrLn $ show t
+      putStrLn $ showTopTm V0 t
+    ["elab-verbose"] -> do
+      (!t, !a) <- elab
+      putStrLn "-- metavariables "
+      putStrLn "------------------------------------------------------------"
+      putStrLn ""
+      displayMetas
+      putStrLn "-- elaboration output"
+      putStrLn "------------------------------------------------------------"
+      putStrLn ""
+      putStrLn $ showTopTm V1 t
+    ["nf"]   -> do
+      (!t, !a) <- elab
+      putStrLn $ showTopTm V1 $ nf [] t
+      putStrLn "  :"
+      putStrLn $ showTopTm V1 $ quote 0 a
+    ["type"] -> do
+      (!t, !a) <- elab
[…]

### 2024-02-05 icfp22paper/sub/demo/Pretty.hs

@@ -1,0 +1,165 @@
+
+module Pretty (showTm, showTm0, showTopTm, displayMetas) where
+
+import Control.Monad
+import Data.IORef
+import Text.Printf
+
+import qualified Data.IntMap.Strict as IM
+
+import Common
+import Evaluation
+import Metacontext
+import Syntax
+
+--------------------------------------------------------------------------------
+
+fresh :: [Name] -> Name -> Name
+fresh ns "_" = "_"
+fresh ns x   = if elem x ns then go (1::Int) else x
+  where go n | elem (x ++ show n) ns = go (n + 1)
+             | True                  = x ++ show n
+
+
+-- printing precedences
+atomp = 3  :: Int -- U, var
+appp  = 2  :: Int -- application
+pip   = 1  :: Int -- pi
+letp  = 0  :: Int -- let, lambda
+
+-- Wrap in parens if expression precedence is lower than
+-- enclosing expression precedence.
+par :: Int -> Int -> ShowS -> ShowS
+par p p' = showParen (p' < p)
+
+braces :: ShowS -> ShowS
+braces s = ('{':).s.('}':)
+
+assign :: Stage -> ShowS
+assign = \case S0 -> (":="++); _ -> ("="++)
+
+lamBind x Impl = braces (x++)
+lamBind x Expl = (x++)
+
+goIx :: [Name] -> Ix -> ShowS
+goIx ns topX = go ns topX topX where
+  go ("_":ns) 0 wkx = (("@"++show wkx)++)
+  go ("*":ns) x wkx = go ns (x + 1) (wkx + 1)
+  go (n:ns)   0 _   = (n++)
+  go (n:ns)   x wkx = go ns (x - 1) wkx
+  go _        _ _   = (("@"++show topX)++)
+
+stage S0 = ('0':)
+stage S1 = ('1':)
+ws = (' ':)
+
+goTm :: Verbosity -> Int -> [Name] -> Tm -> ShowS
+goTm v = go where
+
+  piBind ns x Expl a = ('(':) . (x++) . (" : "++) . go letp ns a .(')':)
+  piBind ns x Impl a = braces  ((x++) . (" : "++) . go letp ns a)
+
+  verbose o s1 s2 = if o <= v then s1 else s2
+
+  goPr :: Int -> [Name] -> [Name] -> Tm -> Pruning -> ShowS
+  goPr p topNs ns t pr = goPr' p ns pr (0 :: Int) where
[…]

### 2024-02-05 icfp22paper/sub/demo/Staging.hs

@@ -1,0 +1,126 @@
+
+module Staging (stage) where
+
+import Common
+import Syntax
+import Evaluation (lvl2Ix)
+
+data Env = Nil | Def0 Env Val0 | Def1 Env Val1
+
+data Val1
+  = VLam1 (Val1 -> Val1)
+  | VQuote Val0
+  | VSomeU1 -- ^ Meta-level types are irrelevant during staging, we evaluate them to this
+            --   dummy value.
+  | VZero1
+  | VSuc1 Val1
+
+data Val0
+  = VVar0 Lvl
+  | VApp0 Val0 Val0 Icit Verbosity
+  | VPi0 Name Icit Val0 (Val0 -> Val0)
+  | VLam0 Name Icit Val0 (Val0 -> Val0) Verbosity
+  | VLet0 Name Val0 Val0 (Val0 -> Val0) Verbosity
+  | VU0
+  | VNat0
+  | VZero0
+  | VSuc0
+  | VNatElim0
+
+vVar1 :: Env -> Ix -> Val1
+vVar1 (Def1 _ v) 0 = v
+vVar1 (Def1 e v) x = vVar1 e (x - 1)
+vVar1 (Def0 e v) x = vVar1 e (x - 1)
+vVar1 _          _ = impossible
+
+vApp1 :: Val1 -> Val1 -> Val1
+vApp1 (VLam1 t) ~u = t u
+vApp1 _          _ = impossible
+
+envTail :: Env -> Env
+envTail (Def1 e _) = e
+envTail (Def0 e _) = e
+envTail _          = impossible
+
+metaError :: MetaVar -> a
+metaError m = error (
+    "Unsolved metavariable during staging: ?" ++ show m
+  ++ "\nHint: look at elab-verbose output to find the offending metavariable")
+
+vNatElim1 :: Val1 -> Val1 -> Val1 -> Val1
+vNatElim1 s z t = case t of
+  VZero1  -> z
+  VSuc1 t -> s `vApp1` t `vApp1` vNatElim1 s z t
+  _       -> impossible
+
+eval1Bind :: Env -> Tm -> Val1 -> Val1
+eval1Bind env t u = eval1 (Def1 env u) t
+
+eval1 :: Env -> Tm -> Val1
+eval1 env = \case
+  Var x             -> vVar1 env x
+  Lam x i a t o     -> VLam1 (eval1Bind env t)
+  App t u i o       -> vApp1 (eval1 env t) (eval1 env u)
+  Pi{}              -> VSomeU1
[…]

### 2024-02-05 icfp22paper/sub/demo/Syntax.hs

@@ -1,0 +1,75 @@
+
+module Syntax where
+
+import Common
+
+type Ty = Tm
+
+-- | A `Pruning` represents a spine of variables, which contains a subsequence
+--   of all variables in scope. A `Just` represents application to a var, a `Nothing`
+--   skips over a var.
+type Pruning = [Maybe Icit]
+
+-- | A reversed pruning. Used for pruning Pi domains, where we have to iterate
+--   inside-out.
+newtype RevPruning = RevPruning Pruning
+
+revPruning :: Pruning -> RevPruning
+revPruning = RevPruning . reverse
+
+-- | A "context zipper", used for efficiently creating types for fresh metas.
+data Path
+  = Here
+  | Define Path Name ~Ty Stage ~Tm
+  | Bind Path Name ~Ty Stage
+  deriving Show
+
+-- | Convert type in context to a closed iterated Pi type.  Note: we need `Tm`
+--   and `Ty` in path in order to make this operation efficient. With this, we
+--   can simply move things over from `Path` without having to rename or quote
+--   anything.
+closeTy :: Path -> Ty -> Ty
+closeTy mcl b = case mcl of
+  Here               -> b
+  Bind mcl x a s     -> closeTy mcl (Pi x Expl a b)
+  Define mcl x a s t -> closeTy mcl (Let s x a t b V1)
+
+data Tm
+  = Var Ix
+  | Lam Name Icit Tm Tm Verbosity
+  | App Tm Tm Icit Verbosity
+  | AppPruning Tm Pruning  -- ^ Used for applying a pruned meta to a mask of the scope.
+  | InsertedMeta MetaVar Pruning
+  | Pi Name Icit Ty Ty
+  | Let Stage Name Ty Tm Tm Verbosity
+  | Meta MetaVar
+
+  | U Stage
+  | Quote Tm
+  | Splice Tm
+  | Lift Ty
+
+  | Wk Tm -- ^ Explicit weakening, used in subtyping coercions
+
+  | Nat Stage
+  | Zero Stage
+  | Suc Stage
+  | NatElim Stage
+  deriving Show
+
+tQuote :: Tm -> Tm
+tQuote (Splice t) = t
+tQuote t          = Quote t
+
+tSplice :: Tm -> Tm
+tSplice (Quote t) = t
+tSplice t         = Splice t
+
+appE0 :: Tm -> Tm -> Tm
[…]

### 2024-02-06 newpaper/paper.tex
paper wip

@@
 \usepackage{bm}
+\usepackage{mathtools}
 %% \usepackage{amssymb}
@@
 \newcommand{\msf}[1]{\mathsf{#1}}
+\newcommand{\mbf}[1]{\mathbf{#1}}
 \newcommand{\mbb}[1]{\mathbb{#1}}
-\newcommand{\mbf}[1]{\mathbf{#1}}
 \newcommand{\bs}[1]{\boldsymbol{#1}}
 \newcommand{\wh}[1]{\widehat{#1}}
+\newcommand{\mdo}{\mbf{do}\,}
+\newcommand{\ind}{\hspace{1em}}
+\newcommand{\bif}{\mbf{if}\,}
+\newcommand{\bthen}{\mbf{then}\,}
+\newcommand{\belse}{\mbf{else}\,}
+\newcommand{\return}{\msf{return}\,}
+\newcommand{\lam}{\lambda\,}
+\newcommand{\data}{\mbf{data}\,}
+\newcommand{\where}{\mbf{where}}
+\newcommand{\M}{\msf{M}}
+\newcommand{\letrec}{\mbf{letrec}\,}
+\newcommand{\of}{\mbf{of}\,}
+\newcommand{\go}{\mit{go}}
+\newcommand{\add}{\mit{add}}
+\newcommand{\letdef}{\mbf{let\,}}
+
 \newcommand{\ext}{\triangleright}
-\newcommand{\Code}{\msf{Code}}
-\newcommand{\El}{\msf{El}}
-\newcommand{\lam}{\msf{lam}}
-\newcommand{\app}{\msf{app}}
-\newcommand{\NatElim}{\msf{NatElim}}
-\newcommand{\y}{\msf{y}}
+
+\newcommand{\Int}{\msf{Int}}
+\newcommand{\List}{\msf{List}}
+\newcommand{\Nil}{\msf{Nil}}
+\newcommand{\Cons}{\msf{Cons}}
+\newcommand{\Reader}{\msf{Reader}}
+\newcommand{\Monad}{\msf{Monad}}
+\newcommand{\class}{\msf{class}}
+\newcommand{\Functor}{\msf{Functor}}
+\newcommand{\Bool}{\msf{Bool}}
+\newcommand{\Statel}{\msf{State}}
+\newcommand{\fro}{\leftarrow}
+\newcommand{\case}{\mbf{case\,}}
 
@@
 \newcommand{\spl}{{\sim}}
-\newcommand{\qut}[1]{\langle #1\rangle}
-
-\newcommand{\mbbc}{\mbb{C}}
-\newcommand{\mbbo}{\mbb{O}}
-\newcommand{\ob}{_\mbbo}
-
-\newcommand{\U}{\msf{U}}
-\newcommand{\Con}{\msf{Con}}
-\newcommand{\Sub}{\msf{Sub}}
+\newcommand{\ql}{{\langle}}
+\newcommand{\qr}{{\rangle}}
+\newcommand{\bind}{\mathbin{>\!\!>\mkern-6.7mu=}}
+
+\newcommand{\MTy}{\msf{MetaTy}}
+\newcommand{\VTy}{\msf{ValTy}}
 \newcommand{\Ty}{\msf{Ty}}
-\newcommand{\Tm}{\msf{Tm}}
-\newcommand{\Cono}{\msf{Con}_{\mbbo}}
-\newcommand{\Subo}{\msf{Sub}_{\mbbo}}
-\newcommand{\Tyo}{\msf{Ty}_{\mbbo}}
-\newcommand{\Tmo}{\msf{Tm}_{\mbbo}}
-\newcommand{\hCon}{\wh{\msf{Con}}}
-\newcommand{\hSub}{\wh{\msf{Sub}}}
-\newcommand{\hTy}{\wh{\msf{Ty}}}
-\newcommand{\hTm}{\wh{\msf{Tm}}}
-
-\newcommand{\p}{\mathsf{p}}
-\newcommand{\q}{\mathsf{q}}
-
-\newcommand{\refl}{\msf{refl}}
-\newcommand{\Bool}{\msf{Bool}}
-\newcommand{\true}{\msf{true}}
-\newcommand{\false}{\msf{false}}
+\newcommand{\CTy}{\msf{CompTy}}
 \newcommand{\True}{\msf{True}}
 \newcommand{\False}{\msf{False}}
-\newcommand{\List}{\msf{List}}
-\newcommand{\nil}{\msf{nil}}
-\newcommand{\cons}{\msf{cons}}
-\newcommand{\Nat}{\msf{Nat}}
-\newcommand{\zero}{\msf{zero}}
-\newcommand{\suc}{\msf{suc}}
-\renewcommand{\tt}{\msf{tt}}
 \newcommand{\fst}{\msf{fst}}
 \newcommand{\snd}{\msf{snd}}
-\newcommand{\mylet}{\msf{let}}
-\newcommand{\emptycon}{\scaleobj{.75}\bullet}
-\newcommand{\id}{\msf{id}}
-
-\newcommand{\Set}{\mathsf{Set}}
-\newcommand{\Prop}{\mathsf{Prop}}
-\newcommand{\Rep}{\msf{Rep}}
+
 \newcommand{\blank}{{\mathord{\hspace{1pt}\text{--}\hspace{1pt}}}}
-\newcommand{\emb}[1]{\ulcorner#1\urcorner}
-
-\newcommand{\Stage}{\msf{Stage}}
-\newcommand{\hato}{\bm\hat{\mbbo}}
-\newcommand{\ev}{\mbb{E}}
-\newcommand{\re}{\mbb{R}}
-
-\theoremstyle{remark}
-\newtheorem{notation}{Notation}
-
-\newcommand{\whset}{\wh{\Set}}
-\newcommand{\rexti}{\re_{\ext_1}^{-1}}
-\newcommand{\rextizero}{\re_{\ext_0}^{-1}}
-
-\newcommand{\rel}{^{\approx}}
-\newcommand{\yon}{\msf{y}}
+
+\newcommand{\Nat}{\msf{Nat}}
+\newcommand{\Zero}{\msf{Zero}}
+\newcommand{\Suc}{\msf{Suc}}
+
+%% \newcommand{\spl}{{\sim}}
+%% \newcommand{\qut}[1]{\langle #1\rangle}
+
+%% \newcommand{\mbbc}{\mbb{C}}
+%% \newcommand{\mbbo}{\mbb{O}}
+%% \newcommand{\ob}{_\mbbo}
+
+%% \newcommand{\U}{\msf{U}}
+%% \newcommand{\Con}{\msf{Con}}
+%% \newcommand{\Sub}{\msf{Sub}}
+%% \newcommand{\Ty}{\msf{Ty}}
+%% \newcommand{\Tm}{\msf{Tm}}
+%% \newcommand{\Cono}{\msf{Con}_{\mbbo}}
+%% \newcommand{\Subo}{\msf{Sub}_{\mbbo}}
+%% \newcommand{\Tyo}{\msf{Ty}_{\mbbo}}
+%% \newcommand{\Tmo}{\msf{Tm}_{\mbbo}}
+%% \newcommand{\hCon}{\wh{\msf{Con}}}
+%% \newcommand{\hSub}{\wh{\msf{Sub}}}
+%% \newcommand{\hTy}{\wh{\msf{Ty}}}
+%% \newcommand{\hTm}{\wh{\msf{Tm}}}
+
+%% \newcommand{\p}{\mathsf{p}}
+%% \newcommand{\q}{\mathsf{q}}
+
+%% \newcommand{\refl}{\msf{refl}}
+%% \newcommand{\Bool}{\msf{Bool}}
+%% \newcommand{\true}{\msf{true}}
+%% \newcommand{\false}{\msf{false}}
+%% \newcommand{\True}{\msf{True}}
+%% \newcommand{\False}{\msf{False}}
+%% \newcommand{\List}{\msf{List}}
+%% \newcommand{\nil}{\msf{nil}}
+%% \newcommand{\cons}{\msf{cons}}
+
+%% \renewcommand{\tt}{\msf{tt}}
+%% \newcommand{\fst}{\msf{fst}}
+%% \newcommand{\snd}{\msf{snd}}
+%% \newcommand{\mylet}{\msf{let}}
+%% \newcommand{\emptycon}{\scaleobj{.75}\bullet}
+%% \newcommand{\id}{\msf{id}}
+
+%% \newcommand{\Set}{\mathsf{Set}}
+%% \newcommand{\Prop}{\mathsf{Prop}}
+%% \newcommand{\Rep}{\msf{Rep}}
+
+%% \newcommand{\emb}[1]{\ulcorner#1\urcorner}
+
+%% \newcommand{\Stage}{\msf{Stage}}
+%% \newcommand{\hato}{\bm\hat{\mbbo}}
+%% \newcommand{\ev}{\mbb{E}}
+%% \newcommand{\re}{\mbb{R}}
+
+%% \theoremstyle{remark}
+%% \newtheorem{notation}{Notation}
+
+%% \newcommand{\whset}{\wh{\Set}}
+%% \newcommand{\rexti}{\re_{\ext_1}^{-1}}
+%% \newcommand{\rextizero}{\re_{\ext_0}^{-1}}
+
+%% \newcommand{\rel}{^{\approx}}
+%% \newcommand{\yon}{\msf{y}}
 
@@
 
-\title{Closure-Free Functional Programming in Two-Level Type Theory}
+\title{Closure-Free Functional Programming in a Two-Level Type Theory}
 %% \titlenote{}
@@
 runtime closures. The compile-time language is a dependent type theory. The two
-are integrated as a \emph{two-level type theory}. We develop abstraction tools
-in this setting and demonstrate two use cases. First, we develop monads and
-monad transformers. Second, we develop fusion for push and pull streams. Most of
-our results are also adapted to a proof-of-concept library in typed Template
-Haskell.
+are integrated in a two-level type theory. We develop some abstraction
+tools in this setting. First, we develop monads and monad transformers. Second,
+we develop fusion for push and pull streams. Most of our results are also
+adapted to a proof-of-concept library in typed Template Haskell.
 \end{abstract}
@@
 
+Modern functional programming supports many convenient abstractions. These often
+come with significant runtime overheads. Sometimes the overheads are acceptable,
+but in other cases compiler optimization is crucial. Monads in Haskell is an
+example for the latter. Even the $\Reader$ monad, which is one of the simplest
+in terms of implementation, yields large overheads when compiled without
+optimizations. Consider the following snippet.
+\begin{alignat*}{3}
+  &f :: \Int \to \Reader\,\Bool\,\Int \\
+  &f\,x = \mdo\{b \fro \msf{ask}; \bif x\, \bthen \return (x + 10)\, \belse \return (x + 20)\}
+\end{alignat*}
+With optimizations enabled, GHC compiles this roughly to the following:
+\begin{alignat*}{3}
+  &f :: \Int \to \Bool \to \Int \\
+  &f = \lam x\,b.\, \bif b\, \bthen x + 10\, \belse x + 20
+\end{alignat*}
+Without optimizations we roughly get:
+\begin{alignat*}{3}
+  &f = \lam x.\,(\bind)\,\msf{MonadReaderDict}\,\msf{ask}\,(\lam b.\,\bif b\\
+  &\ind \bthen\return\,\msf{MonadReaderDict}\,(x + 10)\\
+  &\ind \belse\hspace{0.25em}\return\,\msf{MonadReaderDict}\,(x + 20))
+\end{alignat*}
+Here, $\msf{MonadReaderDict}$ is a runtime dictionary, containing the methods of
+the $\Monad$ instance for $\Reader$, and $(\bind)\,\msf{MonadReaderDict}$ is a
+field projection. Here, a runtime closure will be created for the $\lam b.\,...$
+function, and $(\bind)$, $\msf{ask}$ and $\return$ will create additional dynamic
+closures.
+
+The difference between optimized and unoptimized code is already large here, and
+it gets even larger when we consider monad transformers or code that is
+polymorphic over monads. In Haskell, such code is pervasive, even in fairly
+basic programs which do not use fancy abstractions. The $\msf{mapM}$ function in
+the Haskell Prelude, which maps over a list in some monad, is a third-order and
+second-rank polymorphic function in disguise, because its monad dictionary
+argument contains the polymorphic second-order $(\bind)$ method.
+\begin{alignat*}{3}
+  & \msf{mapM} :: \Monad\,m \Rightarrow (a \to m\,b) \to [a] \to m\,[b]
+\end{alignat*}
+Compiling $\msf{mapM}$ efficiently relies on inlining the instance dictionary,
+then inlining the methods contained there, and also inlining the functions
+that the higher-order binding method is applied to.
+
+GHC's optimization efforts are respectable, and it has gotten quite good over
+its long history of development. However, there is no strong guarantee that
+certain optimizations will happen. Control over optimizations remains tricky,
+fragile and non-compositional. \texttt{INLINE} and \texttt{REWRITE} pragmas can
+be used to control code generation, but without any strong guarantee, and their
+sophisticated usage requires knowledge of GHC internals. For example, correctly
+specifying the \emph{ordering} of certain rule applications is often needed. We
+have to also care about function arities. Infamously, the function composition
+operator is defined as $(.)\,f\,g = \lam x \to f\,(g\,x)$ in the base libraries,
+instead of as $(.)\,f\,g\,x = f\,(g\,x)$, to get better inlining behavior ---
+as explained in a source comment right next to the definition \cite{TODO}.
+
+Although there are numerous tricks and idioms that are used in high-performance
+Haskell programming, for reliable performance it is necessary that programmers
+periodically \emph{review} GHC's optimized code output. Needless to say, this is
+rather time-consuming and poorly scalable.
+
+\subsection{Staged Compilation}
+
+In this paper we use staged compilation to address issues of robustness. The
+idea is to shift as much as possible work from general-purpose optimization to
+metaprograms.
+
+Metaprograms can be deterministic, transparent, and can be run efficiently,
+using fast interpreters or machine code compilation. In contrast,
+general-purpose optimizers are slower to run, less transparent and less
+reliable. Also, metaprogramming allows \emph{library authors} to exploit domain
+specific optimizations, while it is not realistic for general-purpose optimizers
+to know about all domains.
+
+On the other hand, metaprogramming requires some additional care and input from
+programmers. Historically, there have been problems with ergonomics as well:
+\begin{itemize}
+\item Code generation might fail \emph{too late} in the pipeline, producing
+      incomprehensible errors; this is often caused by not having enough static
+      guarantees about the well-formedness of code output.
+\item Tooling for the object language (debugging, profiling, IDE support) often does not
+      work for metaprogramming. This is more likely if the object and meta layers are wildly
+      different.
+\item Metaprogramming may introduce heavy noise and boilerplate, obscuring the logic of
+      programs, or impose restrictions on how code can be structured. For instance,
+      Template Haskell mandates that metaprograms used in splices in some module are
+      defined in a different module.
+\end{itemize}
+
+The idea of \textbf{two-level type theory} (2LTT) is to use a highly expressive
+dependent type theory for compile-time computation, but generate code in
+possibly different, simpler object language. In this paper, we use 2LTT to sweeten the
+deal of staged compilation, aiming for a combination of strong guarantees, good
+ergonomics, high level of abstraction and easy-to-optimize code output.
+
+We develop a particular two-level type theory for this purpose, which we
+call \textbf{CFTT}, short for ``closure-free type theory''. This consists
+of
+\begin{itemize}
+\item A simply-typed object theory with first-order functions, general recursion and
+      finitary algebraic data types. This language is easy to optimize and compile
+      downstream in the pipeline, but it lacks many convenience features.
+\item A standard Martin-Löf type theory for the compile-time language. This
+      allows us to recover many features by metaprogramming.
+\end{itemize}
+
+In particular, since the object language is first-order, we guarantee that all
+programs in CFTT can be ultimately compiled without any dynamic closures, using
+only calls and jumps to statically known code locations. Why focus on closures?
+They are the foundation to almost all abstraction tools in functional
+programming:
+\begin{itemize}
+\item Higher-order functions in essentially all functional languages are implemented with closures.
+\item Type classes in Haskell use dictionary-passing, which relies on closures for function methods.
+\item Functors and first-class modules in OCaml and other ML-s rely on closures.
+\end{itemize}
+Hence, doing functional programming without closures is a clear demonstration
+that we can get rid of abstraction overheads.
+
+It turns out that surprisingly little practical programming relies essentially
+on closures. Most of the time, programmers use higher-order functions for
+\emph{abstraction}, such as when mapping over lists, where it is expected that
+the mapping function will be inlined. In other cases, the use of closures can be
+eliminated by small-scale defunctionalization. For example, difference lists
+\cite{TODO} are often implemented with closures, but they can be also
+implemented as binary trees, with the same programming interface.
+
+\todo{Essential closures: CPS? Threaded interpreters?}
+
+Whole-program defunctionalization is also possible, and it is notably used by
+the MLton compiler \cite{TODO}. While this can be practical and effective, it
+can be also expensive and require whole-program processing. Also, conceptually
+speaking, it does not eliminate closures but instead makes them more transparent
+to optimizations. In this paper we do not use defunctionalization.
+
+We note though that our setup is compatible with closures as well, and it can
+support two separate type formers for closure-based and non-closure-based
+(``static'') functions. Having both of these would be desirable in a practical
+system. In the current work we focus on the closure-free case because it is much
+less known and developed, and it is quite interesting to see how far we
+can go with it.
+
+\subsection{Contributions}
+
+\todo{CONTRIBUTIONS}
+
+\section{Basics of CFTT}\label{sec:basics-of-cftt}
+
+In the following we give an overview of CFTT features. Here we focus on examples
+and informal explanations. We defer the more formal details to Sections
+\ref{TODO} and \ref{TODO}. We first review the meta-level language, then the
+object-level one, and finally the staging operations which bridge between the
+two.
+
+\subsection{The Meta Level}\label{sec:the-meta-level}
+
+$\bs{\MTy}$ is the universe of types in the compile-time language. We will often
+use the term ``metatype'' to refer to inhabitants of $\MTy$, and use
+``metaprogram'' for inhabitants of metatypes. $\MTy$ supports dependent
+functions, products and indexed inductive types \cite{TODO}.
+
+Formally, $\MTy$ is additionally indexed by universes levels, and we have
+$\MTy_i : \MTy_{i+1}$. However, this adds a bit of noise, and it is not very
+relevant to the current paper, so we shall omit levels. Note that universe
+levels have nothing to do with staging in the sense of staged compilation; they
+are about sizing for the purpose of logical consistency.
+
+We use Agda-style syntax and implicit arguments. A basic example:
+\begin{alignat*}{3}
+  &id : \{A : \MTy\} \to A \to A\\
+  &id = \lam x.\, x
+\end{alignat*}
+Here, the type argument is implicit, and it gets inferred when we use the
+function. For example, $id\,\True$ is elaborated to $id\,\{\Bool\}\,\True$,
+where the braces mark an explicit application for the implicit argument.
+Similarly, we can introduce implicit lambdas explicitly:
+\begin{alignat*}{3}
+  &id = \lam \{A : \MTy\}(x : A).\,x
+\end{alignat*}
+We write $(a : A) \times B$ for $\Sigma$-types in $\MTy$. For the field
+projections, for $t : (a : A) \times B$, we have $\fst\,t : A$ and $\snd\,t :
+B[a \mapsto \fst\,t]$, and pairing is written simply as $(t,\,u)$. Additionally,
+we use syntactic sugar for field projections: we can write a type as $(field_1 :
+A) \times (field_2 : B) \times (field_3 : C)$, and write $t.field_2$ for a named
+field projection. The type itself denotes a right-nested $\Sigma$-type. We can
+also write $(t,\,u,\,v)$ for a right-nested iterated pairing. In Haskell style,
+we write $()$ for the unit type, and also for its inhabitant.
+
+Inductive types can be introduced using a Haskell-like ADT notation, or with a GADT-style one:
+\begin{alignat*}{3}
+  &                                           &&\hspace{4em}\data\,\Bool_\M &&: \MTy\,\where\\
+  & \data\,\Bool_\M : \MTy = \True\,|\,\False &&\hspace{4em}\ind\ind \True &&: \Bool_\M\\
+  &                                           &&\hspace{4em}\ind\ind \False &&: \Bool_\M
+\end{alignat*}
+Note that we added an $_\M$ subscript to the type; when analogous types can be
+defined both on the meta and object levels, we will use this subscript to
+distinguish the meta-level version.
+
+\subsection{The Object Level}\label{sec:the-object-level}
+
+$\bs{\Ty}$ is the universe of types in the object language. It is itself a
+metatype, so so we have $\Ty : \MTy$. It is further split to two sub-universes.
+
+First, $\bs{\VTy} : \MTy$ is the universe of \emph{value types}. $\VTy$ supports
+parameterized algebraic data types, with two restrictions:
+\begin{itemize}
+\item ADTs can be only parameterized over types in $\VTy$.
+\item Constructor fields must be in $\VTy$.
+\end{itemize}
+Since $\VTy$ is a sub-universe of $\Ty$, we have that when $A : \VTy$ then also
+$A : \Ty$. Formally, this is specified as an explicit embedding operation but
+informally it is nicer to have an implicit subtyping.
+
+Second, $\bs{\CTy} : \MTy$ is the universe of \emph{computation types}. This is
+also a sub-universe of $\Ty$ with implicit coercions. For now, we only specify
+that $\CTy$ contains functions whose domain is a value type:
+\[ \blank\to\blank : \VTy \to \Ty \to \CTy \]
+For instance, if $\Bool : \VTy$ is defined as an object-level ADT, then $\Bool
+\to \Bool : \CTy$, hence also $\Bool \to \Bool : \Ty$. However, $(\Bool \to
+\Bool) \to \Bool$ is ill-formed, since the domain is not a value type. An
+example for an object-level program, where we already have natural numbers
+declared as $\data \Nat : \VTy := \Zero\,|\,\Suc\,\Nat$:
+\begin{alignat*}{3}
+  &\rlap{$\add : \Nat \to \Nat \to \Nat$}\\
+  &\add :=\,&& \mathrlap{\letrec \go := \lam n\,m.\,\case n\,\of}\\
+  &         && \ind \Zero   &&\to m;\\
+  &         && \ind \Suc\,n &&\to \Suc\,(\go\,n\,m);\\
+  &         && \go
+\end{alignat*}
+Every recursive definition must be introduced with $\letrec$. Note also the
+single semicolon after the $\mbf{case}$ branches. The general syntax is $\letrec x :=
+t; u$, optionally with a type annotation for $x$ on a separate line.
+
+Also note that object-level definitions use $:=$ instead of $=$ for meta-level
+ones. Generally, let-definitions are the biggest source of stage ambiguity in
+surface notation. Later on, we will use more implicit staging notation, but we
+will always disambiguate the stages of $\mbf{let}$-s in this way. Non-recursive
+$\mbf{let}$ is also allowed, and can be used to shadow binders:
+\begin{alignat*}{3}
+  &f : \Nat \to \Nat\\
+  &f := \lam x.\,\letdef x := x + 10; \letdef x := x + 20; x * 10
+\end{alignat*}
+$\mbf{let}$-definitions can be used to define values of any type, and the type
+of the $\mbf{let}$ body can be also anything. Additionally, the right hand sides
+of $\case$ branches can also have arbitrary types. So the following is
+well-formed:
+\begin{alignat*}{3}
+  &f : \mathrlap{\Bool \to \Nat \to \Nat}\\
+  &f \mathrlap{:= \lam b.\,\case b\,\of}\\
+  &\ind \True  &&\to \lam x.\, x + 10;\\
+  &\ind \False &&\to \lam x.\, x * 10;
+\end{alignat*}
+Let us discuss the object language. First, notice that there is no polymorphism
+or any kind of type dependency. Although we can define lists as $\data \List\,A
+: \VTy := \Nil\,|\,\Cons\,A\,(\List\,A)$, the parameterization is just a
+shorthand; all concrete instantiations of the type are distinct. This
+monomorphism makes it easy to use different memory layouts for different types.
+For example, types which look like products may be unboxed. We could also make a
+distinction between boxed and unboxed sum types. We do not explore this in
+detail, we just note that monomorphic types make it easy to control memory
+layouts, and we believe that this is an important part of performance
+optimization.
+
+Second, there are no higher-order functions, and functions also cannot be stored
+in data structures. Hence, locally defined functions can never escape their
+scope, and all function calls are to functions defined in the current
+scope. This makes it possible to run object programs without using dynamic
+closures. This latter point might not be completely straightforward; what
+about the previous $f$ function which has $\lambda$-expressions under a
+$\mbf{case}$, should that require closures?
+
+We say that it should not. We make this formal formal in Section \ref{TODO};
+here we only give an intuitive explanation. In short, we choose a call-by-name
+semantics for functions, which means that the only way we can compute with a
+function is by applying it to all arguments and extracting the resulting value.
+Hence, the only way to compute with $f$ is to apply it to \emph{two} arguments,
+so $f$ is operationally equivalent to the following definition:
+\[ \lam b\,x.\,\case b\,\of \True \to x + 10; \False \to x * 10 \]
+In Section \ref{TODO} we show that all object programs can be transformed to a
+\emph{saturated} form, where the arity of every function call matches the number
+of topmost $\lambda$-binders in the function definition.
+
+Why not just make the object language less liberal, e.g.\ by disallowing
+$\lambda$ under $\mbf{case}$ or $\mbf{let}$, thereby making call saturation
+easier or more obvious? There is a trade-off between making the object language
+more restricted, and thus easier to compile, and making \emph{metaprogramming}
+more convenient for the object language. We will see that the ability to insert
+$\mbf{let}$-s without restriction is very convenient in code generation, and
+likewise the ability to have arbitrary object expressions in $\mbf{case}$
+bodies. In this paper we choose to go with the most liberal object syntax, at
+the cost of needing more downstream processing. The call-by-name nature of
+computation types requires a bit of a change of thinking from programmers, but
+we believe that it is well worth to have it for the metaprogramming convenience.
+
+On the other side of the spectrum, one might imagine generating object code in
+low-level A-normal form. This is more laborious, but it could be also
+interesting, because it forces us to invent abstractions for manipulating ANF in
+the metalanguage. In a similar vein, Allais recently proposed metaprogramming
+quantum circuits in a two-level type theory \cite{TODO}. We leave such setups
+with low-level object languages to future investigation.
+
+Finally, one might compare our object language to call-by-push-value
+(CBPV). Indeed, we took inspiration from CBPV, and there are similarities, but
+also differences. In both systems there is a value-computation distinction, and
+values are call-by-value, and computations are call-by-name. However, our object
+language allows variable binding at arbitrary types, while CBPV only supports it
+at value types. In CBPV, a let-definition for a function is only possible by
+first packing it up in a closure value, which clearly does not work for us.
+Also, CBPV makes a judgment-level structural distinction between values and
+computations, while we use type universes for that. Generally speaking,
+type-based restrictions are easier to work with in dependent type theories than
+structural restrictions.
+
+\section{TODOS}
+
+\todo{extend as I have more stuff}
 \todo{FIGURE OUT: monadic tailrec}
-\todo{FIGURE OUT: Traversable}
+\todo{FIGURE OUT: Traversals}
 \todo{DEVELOP: mutual letrec based on computational product types}
-
-Motivation.
-
-Lots of abstraction in FP. Heavy reliance on general-purpose compiler
-optimization, without guarantee of code quality.
-
-Closures introduce heap allocation and dynamic control flow. Compilers
-try hard to get rid of closures.
-
-Here: use staging instead of closures. Essential use of closures is
-remarkably rare.
-
-General-purpose opt is slow, fragile, but automatic.
-
-Metaprog is fast, deterministic, robust but requires user control.
-
-We make metaprog easier and more streamlined.
-
-Defunctionalization. Full defunc requires whole-program compilation.
-Our approach does not use defunc.
-
-Apparent closure use-cases can be locally defunctionalized. DList prime example.
-
-Closure use cases:
-\begin{itemize}
-\item threaded interp, closure interp
-\item Higher-order DSL-s (but they could be also staged to FO!)
-\item ``Monadic reflection'' in effects
-\item ListT working on any monad
-\item CPS
-\end{itemize}
-
-Summary of contributions, sections, incl Typed TH implementation.
 
@@
 
-%% The purpose of staged compilation is to write code-generating programs in a
-%% safe, ergonomic and expressive way. It is always possible to do ad-hoc code
-%% generation, by simply manipulating strings or syntax trees in a sufficiently
-%% expressive programming language. However, these approaches tend to suffer from
-%% verbosity, non-reusability and lack of safety. In staged compilation, there are
-%% certain \emph{restrictions} on which metaprograms are expressible. Usually,
-%% staged systems enforce typing discipline, prohibit arbitrary manipulation of
-%% object-level scopes, and often they also prohibit accessing the internal
-%% structure of object expressions. On the other hand, we get \emph{guarantees}
-%% about the well-scoping or well-typing of the code output, and we are also able
-%% to use concise syntax for embedding object-level code.
-
-%% \emph{Two-level type theory}, or 2LTT in short, was described in
-%% \cite{twolevel}, building on ideas from \cite{hts}. The motivation was to allow
-%% convenient metatheoretical reasoning about a certain mathematical language
-%% (homotopy type theory), and to enable concise and modular ways to extend the
-%% language with axioms.
-
-%% It turns out that metamathematical convenience closely corresponds to
-%% metaprogramming convenience: 2LTT can be directly and effectively employed in
-%% staged compilation. Moreover, semantic ideas underlying 2LTT are also directly
-%% applicable to the theory of staging.
-
-%% \subsection{Overview \& Contributions}\label{sec:overview}
-
-%% \begin{itemize}
-%%   \item In Section \ref{sec:tour-of-2ltt} we present an informal syntax of
-%%     two-level type theory, a dependent type theory with support for two-stage
-%%     compilation. We look at basic use cases involving inlining control, partial
-%%     evaluation and fusion optimizations. We also describe feature variations,
-%%     enabling applications in monomorphization and memory layout control.
-%%   \item In Section \ref{sec:formal-2ltt}, we specify the formal notion of models
-%%     and syntax for the variant of 2LTT used in this paper. We mostly work with a
-%%     high-level algebraic presentation, where the syntax is quotiented by
-%%     conversion. However, we also explain how to extract functions
-%%     operating on non-quotiented syntax, by interpreting type-theoretic
-%%     constructions in a setoid model.
-%%   \item
-%%     In Section \ref{sec:staging-algorithm}, we review the standard presheaf
-%%     model of 2LTT \cite[Section~2.5.3]{twolevel}, which lies over the syntactic
-%%     category of the object theory. We show that evaluation in this model yields
-%%     a staging algorithm for closed types and terms. We then extend staging to
-%%     open types and terms which may depend on object-level variables. We show
-%%     stability of staging, which means that staging is surjective up to
-%%     conversion, and we also show that staging strictly preserves all type and
-%%     term formers. Finally, we discuss efficiency and potential practical implementations
-%%     of staging.
-%%   \item
-%%     In Section \ref{sec:soundness} we show soundness of staging, which roughly
-%%     means that the output of staging is convertible to its input. Staging
-%%     together with its stability and soundness can be viewed as a \emph{strong
-%%     conservativity} theorem of 2LTT over the object theory. This means that the
-%%     same constructions can be expressed in the object theory and the
-%%     object-level fragment of 2LTT, up to conversion, and staging witnesses that
-%%     meta-level constructions can be always computed away. This strengthens the
-%%     weak notion of conservativity shown in \cite{capriotti2017models} and
-%%     \cite{twolevel}.
-%%   \item
-%%     In Section \ref{sec:intensional-analysis}, we discuss possible semantic
-%%     interpretations of intensional code analysis, i.e.\ the ability to look into
-%%     the structure of object-level code.
-%%   \item We provide a small standalone implementation of elaboration and staging
-%%     for a two-level type theory \cite{staged-demo}.
-%%   \item To our knowledge, this work is the first formalization and implementation of
-%%     staged compilation in the presence of full dependent types, with
-%%     universes and large elimination. In particular, we allow unrestricted
-%%     staging for types, so that types can be computed by metaprograms at compile
-%%     time.
-%% \end{itemize}
-
-%% \section{A Tour of Two-Level Type Theory}\label{sec:tour-of-2ltt}
-
-%% In this section, we provide a short overview of 2LTT and its potential
-%% applications in staging. We work in the informal syntax of a dependently typed
-%% language which resembles Agda \cite{agdadocs}. We focus on examples and informal
-%% explanation here; the formal details will be presented in Section \ref{sec:formal-2ltt}.
-
-%% \begin{notation}\label{basic-notation}
-%% We use the following notations throughout the paper. $(x : A) \to B$ denotes a
-%% dependent function type, where $x$ may occur in $B$. We use $\lambda\,x.\,t$ for
-%% abstraction. A $\Sigma$-type is written as $(x : A) \times B$, with pairing as
-%% $(t,\,u)$, projections as $\fst$ and $\snd$, and we may use pattern matching
-%% notation on pairs, e.g.\ as in $\lambda\,(x,\,y).\,t$. The unit type is $\top$
-%% with element $\tt$. We will also use Agda-style notation for implicit arguments,
-%% where $t : \{x : A\} \to B$ implies that the first argument to $t$ is inferred
-%% by default, and we can override this by writing a $t \{u\}$ application. We may
-%% also implicitly quantify over arguments (in the style of Idris and Haskell), for
-%% example when declaring $\mit{id} : A \to A$ with the assumption that $A$ is
-%% universally quantified.
-%% \end{notation}
-
-%% \subsection{Rules of 2LTT}
-
-%% \subsubsection{Universes}
-%% We have universes $\U_{i,j}$, where $i \in \{0,1\}$, and $j \in \mbb{N}$.  The
-%% $i$ index denotes stages, where $0$ is the runtime (object-level) stage, and $1$
-%% is the compile time (meta-level) stage. The $j$ index denotes universe sizes in
-%% the usual sense of type theory. We assume Russell-style universes, with
-%% $\U_{i,j} : \U_{i, j+1}$. However, for the sake of brevity we will usually omit
-%% the $j$ indices in this section, as sizing is orthogonal to our use-cases and
-%% examples.
-%% \begin{itemize}
-%% \item $\U_0$ can be viewed as the \emph{universe of object-level or runtime types}.
-%%     Each closed type $A : \U_0$ can be staged to an actual type in the object language
-%%     (the language of the staging output).
-%%   \item $\U_1$ can be viewed as the \emph{universe of meta-level or static types}. If we
-%%     have $A : \U_1$, then $A$ is guaranteed to be only present at compile time,
-%%     and will be staged away. Elements of $A$ are likewise computed away.
-%% \end{itemize}
-
-%% \subsubsection{Type Formers} $\U_0$ and $\U_1$ may be closed under arbitrary type formers,
-%% such as functions, $\Sigma$-types, identity types or inductive types in general.
-%% However, all constructors and eliminators in type formers must stay at the same
-%% stage. For example:
-%% \begin{itemize}
-%%   \item Function domain and codomain types must be at the same stage.
-%%   \item If we have $\Nat_0 : \U_0$ for the runtime type of natural numbers,
-%%         we can only map from it to a type in $\U_0$ by recursion or induction.
-%% \end{itemize}
-%% It is not required that we have the \emph{same} type formers at both stages. We
-%% will discuss variations of the object-level languages in Section \ref{sec:variations}.
-
-%% \subsubsection{Moving Between Stages}
-%% At this point, our system is rather limited, since there is no interaction
-%% between the stages. We enable such interaction through the following operations.
-%% Note that none of these operations can be expressed as functions, since function
-%% types cannot cross between stages.
-%% \begin{itemize}
-%% \item \emph{Lifting:} for $A : \U_0$, we have $\Lift A : \U_1$.  From the
-%%   staging point of view, $\Lift A$ is the type of metaprograms which compute
-%%   runtime expressions of type $A$.
-%% \item \emph{Quoting:} for $A : \U_0$ and $t : A$, we have $\qut{t} :\,\Lift A$.
-%%   A quoted term $\qut{t}$ represents the metaprogram which immediately yields
-%%   $t$.
-%% \item \emph{Splicing:} for $A : \U_0$ and $t :\,\Lift A$, we have $\spl t : A$.
-%%   During staging, the metaprogram in the splice is executed, and the resulting
-%%   expression is inserted into the output.
-
-%%   \begin{notation} Splicing binds stronger than any operation, including function
-%%     application. For instance, $\spl f\,x$ is parsed as $(\spl f)\,x$. We borrow
-%%     this notation from MetaML \cite{metaml}.
-%%   \end{notation}
-
-%% \item Quoting and splicing are definitional inverses, i.e.\ we have $\spl\qut{t} = t$ and
-%%   $\qut{\spl t}=t$ as definitional equalities.
-%% \end{itemize}
-
-%% Informally, if we have a closed program $t : A$ with $A : \U_0$, \emph{staging}
-%% means computing all metaprograms and recursively replacing all splices in $t$ and $A$
-%% with the resulting runtime expressions. The rules of 2LTT ensure that this is possible,
-%% and we always get a splice-free object program after staging.
-
-%% \paragraph{Remark}
-%% Why do we use the index $0$ for the runtime stage? The reason is that it is not
-%% difficult to generalize 2LTT to multi-level type theory, by allowing to lift
-%% types from $\U_i$ to $\U_{i+1}$. In the semantics, this can be modeled by
-%% having a 2LTT whose object theory is once again a 2LTT, and doing this in an
-%% iterated fashion. But there must be necessarily a bottom-most object theory;
-%% hence our stage indexing scheme. For now though, we leave the multi-level
-%% generalization to future work.
-
-%% \begin{notation}\label{notation:basics} We may disambiguate type formers at different stages by using $0$ or $1$
-%% subscripts. For example, $\Nat_1 : \U_1$ is distinguished from $\Nat_0 : \U_0$,
-%% and likewise we may write $\zero_0 : \Nat_0$ and so on. For function and
-%% $\Sigma$ types, the stage is usually easy to infer, so we do not annotate
-%% them. For example, the type $\Nat_0 \to \Nat_0$ must be at the runtime stage,
-%% since the domain and codomain types are at that stage, and we know that the
-%% function type former stays within a single stage. We may also omit stage annotations
-%% from $\lambda$ and pairing.
-%% \end{notation}
-
-%% \subsection{Staged Programming in 2LTT}\label{sec:staged-programming}
-
-%% In 2LTT, we may have several different polymorphic identity functions. First,
-%% consider the usual identity function at each stage:
-%% \begin{alignat*}{3}
-%%   & \mit{id}_0 : (A : \U_0) \to A \to A\hspace{2em} && \mit{id}_1 : (A : \U_1) \to A \to A\\
-%%   & \mit{id}_0 := \lambda\,A\,x.x       && \mit{id}_1 := \lambda\,A\,x.x
-%% \end{alignat*}
-%% An $\mit{id}_0$ application will simply appear in staging output as it is. In
-%% contrast, $\mit{id}_1$ can be used as a compile-time evaluated function, because
-%% the staging operations allow us to freely apply $\mit{id}_1$ to runtime
-%% arguments. For example, $\mit{id}_1\,(\Lift\,\Bool_0)\,\qut{\true_0}$ has type
-%% $\Lift \Bool_0$, which implies that $\spl(\mit{id}_1\,(\Lift\,\Bool_0)\,\qut{\true_0})$
-%% has type $\Bool_0$. We can stage this expression as follows:
-%% \[
-%% \spl(\mit{id}_1\,(\Lift\,\Bool_0)\,\qut{\true_0}) = \spl\qut{\true_0} = \true_0
-%% \]
-%% There is another identity function, which computes at compile time, but which
-%% can be only used on runtime arguments:
-%% \begin{alignat*}{3}
-%%   & \mit{id_\Lift} : (A : \Lift\U_0) \to \Lift\,\spl A \to \Lift\,\spl A\\
-%%   & \mit{id_\Lift} := \lambda\,A\,x.x
-%% \end{alignat*}
-%% Note that since $A : \Lift\U_0$, we have $\spl A : \U_0$, hence $\Lift\,\spl A
-%% : \U_1$.  Also, $\Lift\U_0 : \U_1$, so all function domain and codomain types in
-%% the type of $\mit{id_\Lift}$ are at the same stage. Now, we may write
-%% $\spl(\mit{id_\Lift}\,\qut{\Bool_0}\,\qut{\true_0})$ for a term which is staged
-%% to $\true_0$. In this specific case $\mit{id_\Lift}$ has no practical advantage
-%% over $\mit{id}_1$, but in some cases we really have to quantify over
-%% $\Lift\U_0$. This brings us to the next example.
-
-%% Assume $\List_0 : \U_0 \to \U_0$ with $\nil_0 : (A : \U_0) \to \List_0\,A$,
-%% $\cons_0 : (A : \U_0) \to A \to \List_0\,A$ and $\msf{foldr}_0 : (A\,B : \U_0)
-%% \to (A \to B \to B) \to B \to \List_0\,A \to B$. We define a mapping function which
-%% inlines its function argument:
-%% \begin{alignat*}{3}
-%%   & \mit{map} : (A\,B : \Lift\U_0) \to (\Lift\,\spl A \to \Lift\,\spl B)
-%%       \to \Lift(\List_0\,\spl A) \to \Lift(\List_0\,\spl B)\\
-%%   & \mit{map} := \lambda\,A\,B\,f\,\mit{as}.\,
-%%       \qut{\msf{foldr}_0\,
-%%         \spl A\,(\List_0 \spl B)\,
-%%         (\lambda\,a\,\mit{bs}.\, \cons_0\,\spl B\,\spl(f\,\qut{a})\,\mit{bs})\,
-%%         (\nil_0\,\spl B)\,
-%%         \spl as
-%%         }
-%% \end{alignat*}
-%% This $\mit{map}$ function can be defined with quantification over $\Lift \U_0$ but not
-%% over $\U_1$, because $\List_0$ expects type parameters in $\U_0$, and there is
-%% no generic way to convert from $\U_1$ to $\U_0$. Now, assuming
-%% $\blank\!+_0\!\blank : \Nat_0 \to \Nat_0 \to \Nat_0$ and $\mit{ns} :
-%% \List_0\,\Nat_0$, we have the following staging behavior:
-%% \begin{alignat*}{3}
-%%   & &&\spl(\mit{map}\,\qut{\Nat_0}\,\qut{\Nat_0}\,(\lambda\,n.\,\qut{\spl n +_0 10})\,\qut{\mit{ns}})\\
-%%   & =\hspace{0.3em}&&\spl\qut{\msf{foldr}_0\,
-%%         \spl \qut{\Nat_0}\,(\List_0\,\spl\qut{\Nat_0})\\
-%%   & && \hspace{3.5em}(\lambda\,a\,\mit{bs}.\, \cons_0\,\spl\qut{\Nat_0}\,\spl\qut{\spl\qut{a} +_0 10}\,\mit{bs})\,
-%%         (\nil_0\,\spl \qut{\Nat_0})\,
-%%         \spl\qut{\mit{ns}}}\\
-%%   & =\hspace{0.3em} &&\msf{foldr_0}\,\Nat_0\,(\List_0\,\Nat_0)\,\,(\lambda\,a\,bs.\,\cons_0\,\Nat_0\,(a +_0 10)\,bs)\,(\nil_0\,\Nat_0)\,\mit{ns}
-%% \end{alignat*}
-
-%% By using meta-level functions and lifted types, we already have control over
-%% inlining. However, if we want to do more complicated meta-level computation, it
-%% is convenient to use recursion or induction on meta-level type formers. A
-%% classic example in staged compilation is the power function for natural numbers,
-%% which evaluates the exponent at compile time. We assume the iterator function
-%% $\msf{iter_1} : \{A : \U_1\} \to \Nat_1 \to (A \to A) \to A \to A$, and runtime
-%% multiplication as $\blank\!*_0\!\blank$.
-%% \begin{alignat*}{3}
-%%   &\mit{exp} : \Nat_1 \to \Lift \Nat_0 \to \Lift \Nat_0 \\
-%%   &\mit{exp} := \lambda\,x\,y.\,
-%%   \msf{iter}_1\,x\,(\lambda\,n.\,\qut{\spl y *_0 \spl n})\,\qut{1}
-%% \end{alignat*}
-%% Now, $\spl(\mit{exp}\,3\,\qut{n})$ stages to $n *_0 n *_0 n *_0 1$ by the computation rules
-%% of $\msf{iter_1}$ and the staging operations.
-
-%% We can also stage \emph{types}. Below, we use iteration to compute the type of
-%% vectors with static length, as a nested pair type.
-%% \begin{alignat*}{3}
-%%   &\mit{Vec} : \Nat_1 \to \Lift \U_0 \to \Lift \U_0\\
-%%   &\mit{Vec} := \lambda\,n\,A.\,\msf{iter}_1\,n\,(\lambda\,B.\,\qut{\spl A \times \spl B})\,\qut{\top_0}
-%% \end{alignat*}
-%% With this definition, $\spl(\mit{Vec}\,3\,\qut{\Nat_0})$ stages to $\Nat_0
-%% \times (\Nat_0 \times (\Nat_0 \times \top_0))$. Now, we can use \emph{induction}
-%% on $\Nat_1$ to implement a map function. For readability, we use an Agda-style
-%% pattern matching definition below (instead of the elimination principle).
-%% \begin{alignat*}{6}
-%%   &\hspace{-3em}\rlap{$\mit{map} : (n : \Nat_1) \to (\Lift\,\spl A \to \Lift\,\spl B) \to \Lift(\mit{Vec}\,n\,A) \to \Lift(\mit{Vec}\,n\,B)$}\\
-%%   &\hspace{-3em}\mit{map}\,\,&&\zero_1      && f\,&&\mit{as} := \qut{\tt_0}\\
-%%   &\hspace{-3em}\mit{map}\,\,&&(\suc_1\,n)\,&& f\,&&\mit{as} :=
-%%      \qut{(\spl(f\,\qut{\fst_0\,\spl\mit{as}}),\,\spl(\mit{map}\,n\,f\,\qut{\snd_0\,\spl \mit{as}}))}
-%% \end{alignat*}
-%% This definition inlines the mapping function for each projected element of the
-%% vector. For instance, staging $\spl(\mit{map}\,2\,(\lambda\,n.\,\qut{\spl n +_0
-%%   10})\,\qut{\mit{ns}})$ yields $(\fst_0\,\mit{ns} +_0
-%% 10,\,(\fst_0(\snd_0\,\mit{ns}) +_0 10,\,\tt_0))$. Sometimes, we do not want to
-%% duplicate the code of the mapping function. In such cases, we can use
-%% \emph{let-insertion}, a common technique in staged compilation. If we bind a runtime
-%% expression to a runtime variable, and only use that variable in subsequent
-%% staging, only the variable itself can be duplicated. One solution is to do an
-%% ad-hoc let-insertion:
-%% \begin{alignat*}{6}
-%%   &   && \mylet_0\,f := \lambda\,n.\, n +_0 10\,\,\msf{in}\,\,
-%%          \spl(\mit{map}\,2\,(\lambda\,n.\,\qut{f\,\spl n})\,\qut{\mit{ns}}) \\
-%%   & =\,\,&&  \mylet_0\,f := \lambda\,n.\, n +_0 10\,\,\msf{in}\,\,
-%%           (f\,(\fst_0\,\mit{ns}),\,(f\,(\fst_0(\snd_0\,\mit{ns})),\,\tt_0))
-%% \end{alignat*}
-%% We include examples of more sophisticated let-insertion in the supplementary
-%% code \cite{staged-demo}.
-
-%% More generally, we are free to use dependent types at the meta-level, so we can
-%% reproduce more complicated staging examples. Any well-typed interpreter can be
-%% rephrased as a \emph{partial evaluator}, as long as we have sufficient type
-%% formers. For instance, we may write a partial evaluator for a simply typed
-%% lambda calculus. We sketch the implementation in the following; the full version
-%% can be found in the supplementary code. First, we inductively define types,
-%% contexts and terms:
-%% \begin{alignat*}{6}
-%%   & \Ty : \U_1  \hspace{2em} \Con : \U_1 \hspace{2em} \Tm : \Con \to \Ty \to \U_1
-%% \end{alignat*}
-%% Then we define the interpretation functions:
-%% \begin{alignat*}{6}
-%%   & \msf{EvalTy}  &&: \Ty \to \Lift \U_0 \\
-%%   & \msf{EvalCon} &&: \Con \to \U_1 \\
-%%   & \msf{EvalTm}  &&: \Tm\,\Gamma\,A \to \msf{EvalCon}\,\Gamma \to \Lift\,\spl(\msf{EvalTy}\,A)
-%% \end{alignat*}
-%% Types are necessarily computed to runtime types, e.g.\ an embedded
-%% representation of the natural number type is evaluated to $\qut{\Nat_0}$.
-%% Contexts are computed as follows:
-%% \begin{alignat*}{4}
-%%   &\msf{EvalCon}\,\,\msf{empty}                &&:= \top_1 \\
-%%   &\msf{EvalCon}\,\,(\msf{extend}\,\Gamma\,A) &&:= \msf{EvalCon}\,\Gamma \times (\Lift \spl(\msf{EvalTy}\,A))
-%% \end{alignat*}
-%% This is an example for the usage of \emph{partially static data} \cite{partial-evaluation}:
-%% semantic contexts are \emph{static} lists storing \emph{runtime}
-%% expressions. This allows us to completely eliminate environment lookups in the
-%% staging output: an embedded lambda expression is staged to the corresponding
-%% lambda expression in the object language. This is similar to the partial
-%% evaluator presented in Idris 1 \cite{scrap-your-inefficient-engine}. However, in
-%% contrast to 2LTT, Idris 1 does not provide a formal guarantee that partial
-%% evaluation does not get stuck.
-
-%% \subsection{Properties of Lifting, Binding Time Improvements}
-
-%% We describe more generally the action of lifting on type formers. While lifting
-%% does not have definitional computation rules, it does preserve negative type formers up
-%% to definitional isomorphism \cite[Section~2.3]{twolevel}:
-%% \begin{alignat*}{5}
-%%   \Lift((x : A) \to B\,x) &\simeq ((x : \Lift A) \to \Lift\,(B\,\spl x))\\
-%%   \Lift ((x : A) \times B\,x) &\simeq ((x : \Lift A) \times \Lift (B\,\spl x))\\
-%%   \Lift \top_0 &\simeq \top_1
-%% \end{alignat*}
-%% For function types, the preservation maps are the following:
-%% \begin{alignat*}{5}
-%%   & \hspace{-6em}\rlap{$\mit{pres}_\to : \Lift((x : A) \to B\,x) \to ((x : \Lift A) \to \Lift\,(B\,\spl x))$}\\
-%%   & \hspace{-6em}\mit{pres}_\to\,f     &&:= \lambda\,x.\,\,\qut{\spl f\,\spl x}\\
-%%   & \hspace{-6em}\mit{pres}_\to^{-1}\,f &&:= \qut{\lambda\,x.\,\,\spl(f\,\qut{x})}
-%% \end{alignat*}
-%% With this, we have that $\mit{pres}_\to\,(\mit{pres}_\to^{-1}\,f)$ is
-%% definitionally equal to $f$, and also the other way around. Preservation maps
-%% for $\Sigma$ and $\top$ work analogously.
-
-%% By rewriting a 2LTT program left-to-right along preservation maps, we perform
-%% what is termed \emph{binding time improvement} in the partial evaluation
-%% literature \cite[Chapter~12]{partial-evaluation}. Note that the output of $\mit{pres}_{\to}$ uses a
-%% meta-level $\lambda$, while going the other way introduces a runtime
-%% binder. Meta-level function types and $\Sigma$-types support more computation during
-%% staging, so in many cases it is beneficial to use the improved forms. In some
-%% cases though we may want to use unimproved forms, to limit the size of generated
-%% code. This is similar to what we have seen with let-insertion. For a minimal
-%% example, consider the following unimproved version of $\mit{id}_\Lift$:
-%% \begin{alignat*}{5}
-%%   & \mit{id}_\Lift : (A : \Lift \U_0) \to \Lift(\spl A \to \spl A) \\
-%%   & \mit{id}_\Lift := \lambda\,A.\,\,\qut{\lambda\,x.\,x}
-%% \end{alignat*}
-%% This can be used at the runtime stage as
-%% $\spl(\mit{id}_\Lift\,\qut{\Bool_0})\,\true_0$, which is staged to
-%% $(\lambda\,x.\,x)\,\true_0$.  This introduces a useless $\beta$-redex, so in
-%% this case the improved version is clearly preferable.
-
-%% For inductive types in general we do not get full preservation, only maps in one
-%% direction. For example, we have $\Bool_1 \to \Lift \Bool_0$, defined as
-%% $\lambda\,b.\,\msf{if}\,b\,\msf{then}\,\qut{\true_0}\,\msf{then}\,\qut{\false_0}$.
-%% In the staging literature, this is called ``serialization'' or ``lifting''
-%% \cite{template-haskell,metaml}.  In the other direction, we can only define
-%% constant functions from $\Lift \Bool_0$ to $\Bool_1$.
-
-%% The lack of elimination principles for $\Lift A$ means that we cannot inspect
-%% the internal structure of expressions. We will briefly discuss ways to lift this
-%% restriction in Section \ref{sec:intensional-analysis}.
-
-%% In particular, we have no serialization map from $\Nat_1 \to \Nat_1$ to
-%% $\Lift(\Nat_0 \to \Nat_0)$. However, when $A : \U_1$ is \emph{finite}, and $B :
-%% \U_1$ can be serialized, then $A \to B$ can be serialized, because it is
-%% equivalent to a finite product. For instance, $\Bool_1 \to \Nat_1 \simeq \Nat_1
-%% \times \Nat_1$.  In 2LTT, $A$ is called \emph{cofibrant}
-%% \cite[Section~3.4]{twolevel}: this means that for each $B$, $A \to \Lift\,B$ is
-%% equivalent to $\Lift C$ for some $C$. This is the 2LTT formalization of the
-%% so-called ``trick'' in partial evaluation, which improves binding times by
-%% $\eta$-expanding functions out of finite sums \cite{eta-expansion-trick}.
-
-%% \subsubsection{Fusion}
-%% Fusion optimizations can be viewed as binding time improvement techniques for
-%% general inductive types. The basic idea is that by lambda-encoding an inductive
-%% type, it is brought to a form which can be binding-time improved. For instance,
-%% consider foldr-build fusion for lists \cite{short-cut}, which is employed in GHC
-%% Haskell.  Starting from $\Lift (\List_0\,A)$, we use Böhm-Berarducci encoding
-%% \cite{boehm-berarducci} under the lifting to get
-%% \begin{alignat*}{3}
-%%          &\Lift((L : \U_0) \to (A \to L \to L) \to L \to L)\\
-%%   \simeq\,\,&((L : \Lift\,U_0) \to (\Lift\,A \to \Lift\,\spl L \to \Lift\,\spl L) \to \Lift\,\spl L \to \Lift\,\spl L).
-%% \end{alignat*}
-%% Alternatively, for \emph{stream fusion} \cite{stream-fusion}, we embed $\List\,A$ into the
-%% coinductive colists (i.e.\ the possibly infinite lists), and use a terminal
-%% lambda-encoding. The embedding into the ``larger'' structure enables some staged
-%% optimizations which are otherwise not possible, such as fusion for the
-%% $\msf{zip}$ function. However, the price we pay is that converting
-%% back to lists from colists is not necessarily total.
-
-%% We do not detail the implementation of fusion in 2LTT here; a small example can
-%% be found in the supplementary code. In short, 2LTT is a natural setting for a
-%% wide range of fusion setups. A major advantage of fusion in 2LTT is the formal
-%% guarantee of staging, in contrast to implementations where compile-time
-%% computation relies on ad-hoc user annotations and general-purpose optimization
-%% passes. For instance, fusion in GHC relies on rewrite rules and inlining
-%% annotations which have to be carefully tuned and ordered, and it is possible to
-%% get pessimized code via failed fusion.
-
-%% \subsubsection{Inferring Staging Operations}
-%% We can extract a coercive subtyping system from the staging operations, which
-%% can be used for inference, and in particular for automatically transporting
-%% definitions along preservation isomorphisms. One choice is to have $A \leq \Lift
-%% A$, $\Lift A \leq A$, a contravariant-covariant rule for functions and a
-%% covariant rule for $\Sigma$. During bidirectional elaboration, when we need to
-%% compare an inferred and an expected type, we can insert coercions.  We
-%% implemented this feature in our prototype. It additionally supports Agda-style
-%% implicit arguments and pattern unification, so it can elaborate the following
-%% definition:
-%% \begin{alignat*}{3}
-%%   & \mit{map} : \{A\,B : \Lift\U_0\} \to (\Lift A \to \Lift B)
-%%       \to \Lift(\List_0\,A) \to \Lift(\List_0\,B)\\
-%%   & \mit{map} := \lambda\,f\,\mit{as}.\,
-%%       \msf{foldr}_0
-%%         (\lambda\,a\,\mit{bs}.\, \cons_0\,(f\,a)\,\mit{bs})\,
-%%         \nil_0\,
-%%         as
-%% \end{alignat*}
-%% We may go a bit further, and also add the coercive subtyping rule $\U_0 \leq
-%% \U_1$, witnessed by $\Lift$. Then, the type of $\mit{map}$ can be written as
-%% $\{A\,B : \Lift\U_0\} \to (A \to B) \to \List_0\,A \to \List_0\,B$. However,
-%% here the elaborator has to make a choice, whether to elaborate to improved or
-%% unimproved types. In this case, the fully unimproved type would be
-%% \[ \{A\,B : \Lift\U_0\} \to \Lift((\spl A \to \spl B) \to \List_0\,\spl A \to \List_0\,\spl B). \]
-%% It seems that improved types are a sensible default, and we can insert explicit
-%% lifting when we want to opt for unimproved types. This is also available in our
-%% prototype.
-
-%% \subsection{Variations of Object-Level Languages}
-%% \label{sec:variations}
-
-%% In the following, we consider variations on object-level languages, with a focus
-%% on applications in downstream compilation after staging. Adding restrictions or
-%% more distinctions to the object language can make it easier to optimize and
-%% compile.
-
-%% \subsubsection{Monomorphization}\label{sec:monomorphization}
-
-%% In this case, the object language is simply typed, so every type is known
-%% statically. This makes it easy to assign different memory layouts to different
-%% types and generate code accordingly for each type. Moving to 2LTT, we still
-%% want to abstract over runtime types at compile time, so we use the following
-%% setup.
-%% \begin{itemize}
-%% \item We have a \emph{jugdment}, written as $A\,\msf{type_0}$, for well-formed
-%%   runtime types. Runtime types may be closed under simple type formers.
-%% \item We have a type $\Ty_0 : \U_1$ in lieu of the previous $\Lift \U_0$.
-%% \item For each $A\,\msf{type}_0$, we have $\Lift A : \Ty_0$.
-%% \item We have quoting and splicing for types and terms. For types, we send $A\,
-%%   \msf{type}_0$ to $\qut{A} : \Ty_0$. For terms, we send $t : A$ to $\qut{t} :
-%%   \Lift A$.
-%% \end{itemize}
-%% Despite the restriction to simple types at runtime, we can still write arbitrary
-%% higher-rank polymorphic functions in 2LTT, such as a function with type $((A : \Ty_0)
-%% \to \Lift\,\spl A \to \Lift \spl A) \to \Lift\,\Bool_0$. This function can be
-%% only applied to statically known arguments, so the polymorphism can be staged
-%% away. The main restriction that programmers have to keep in mind is that
-%% polymorphic functions cannot be stored inside runtime data types.
-
-%% %% This is similar to the well-known monomorphization models in the C++ and
-%% %% Rust programming languages. The evident difference is that in 2LTT we can use a
-%% %% full type theory at compile time. Additionally, we will see in Section
-%% %% \ref{sec:intensional-analysis} that this setup is compatible with intensional
-%% %% analysis, the ability to analyze the structure of object types and terms.
-
-%% \subsubsection{Memory Representation Polymorphism}
-
-%% This refines monomorphization, so that types are not directly identified with
-%% memory representations, but instead representations are internalized in 2LTT as
-%% a meta-level type, and runtime types are indexed over representations.
-%% \begin{itemize}
-%% \item We have $\Rep : \U_1$ as the type of memory representations. We have
-%%   considerable freedom in the specification of $\Rep$. A simple setup may
-%%   distinguish references from unboxed products, i.e.\ we have $\msf{Ref} : \Rep$
-%%   and $\msf{Prod} : \Rep \to \Rep \to \Rep$, and additionally we may assume
-%%   any desired primitive machine representation as a value of $\Rep$.
-%% \item We have Russell-style $\U_{0,j} : \Rep \to \U_{0, j+1}\,r$, where $r$ is
-%%   some chosen runtime representation for types; usually we would mark types are
-%%   erased. We leave the meta-level $\U_{1,j}$ hierarchy unchanged.
-%% \item We may introduce unboxed $\Sigma$-types and primitive machine types in the
-%%   runtime language. For $r : \Rep$, $r' : \Rep$, $A : \U_{0}\,r$ and $B : A \to
-%%   \U_{0}\,r'$, we may have $(x : A) \times B\,x :
-%%   \U_{0}\,(\msf{Prod\,r\,r'})$. Thus, we have type dependency, but we do not
-%%   have dependency in memory representations.
-%% \end{itemize}
-%% Since $\Rep$ is meta-level, there is no way to abstract over it at runtime, and
-%% during staging all $\Rep$ indices are computed to concrete canonical
-%% representations. This is a way to reconcile dependent types with some amount of
-%% control over memory layouts. The unboxed flavor of $\Sigma$ ends up with a
-%% statically known flat memory representation, computed from the representations
-%% of the fields.
-
-%% %% In principle, it should be possible to have \emph{dependent memory layouts};
-%% %% this would be a more advanced variation on type-passing polymorphism
-%% %% \cite{intensional-poly}, where layouts of objects may depend on runtime data. For example,
-%% %% length-prefixed flat arrays would not be a primitive notion, but simply defined
-%% %% as $(n : \Nat) \times \msf{Vec}\,n\,A$. However, it seems to be highly
-%% %% challenging to implement runtime systems and precise garbage collection in this
-%% %% setting. See \texttt{sixten} \cite{sixten} for an experimental implementation of
-%% %% dependent memory layouts.
-
-%% \section{Formal Setup}\label{sec:formal-2ltt}
-
-%% In this section we describe our approach to formalizing staging and also the
-%% \emph{extraction} of algorithms from the formalization. There is a tension
-%% here:
-%% \begin{itemize}
-%%   \item On one hand, we would like to do formal work at a higher level of
-%%     abstraction, in order to avoid a deluge of technical details (for which
-%%     dependent type theories are infamous).
-%%   \item On the other hand, at some point we need to talk about lower-level
-%%     operational aspects and extracted algorithms.
-%% \end{itemize}
-%% In the following, we describe our approach to interfacing between the different levels
-%% of abstraction.
-
-%% \subsection{The Algebraic Setting}
-
-%% When defining staging and proving its soundness and stability, we work
-%% internally in a constructive type theory, using types and terms that are
-%% quotiented by conversion. In this setting, every construction respects syntactic
-%% conversion, which is very convenient technically. Here, we use standard
-%% definitions and techniques from the categorical and algebraic metatheory of type
-%% theories, and reuse prior results which rely on the same style of
-%% formalization. Concretely, we use \emph{categories with families} \cite{cwfs}
-%% which is a commonly used specification of an \emph{explicit substitution
-%% calculus} of types and terms.
-
-%% In the rest of the paper we will mostly work in this setting, and we will
-%% explicitly note when we switch to a different metatheory. We summarize
-%% the features of this metatheory in the following.
-
-%% \paragraph{Metatheory}
-%% The metatheory is an intensional type theory with a cumulative universe
-%% hierarchy $\Set_i$, where $i$ is ordinal number such that $i \leq \omega +
-%% 1$. The transfinite $i$ levels are a convenience feature for modeling typing
-%% contexts, where sometimes we will need universes large enough to fit types at
-%% arbitrary finite levels. We also assume uniqueness of identity proofs and
-%% function extensionality. We will omit transports along identity proofs in the
-%% paper, for the sake of readability. The missing transports can be in principle
-%% recovered due to Hofmann's conservativity result \cite{hofmann95extensional};
-%% see also \cite{DBLP:conf/cpp/WinterhalterST19}. We assume $\Pi$, $\Sigma$,
-%% $\top$ and $\mbb{N}$ types in all universes, and reuse Notation
-%% \ref{notation:basics}. We also assume certain quotient inductive-inductive
-%% types, to represent the syntax of 2LTT and the object theory. We will detail
-%% these shortly in Sections \ref{sec:2lttmod} and \ref{sec:objmod}.
-
-%% \subsection{Algorithm Extraction}
-
-%% In the algebraic setting we can \emph{only} talk about properties which
-%% respect syntactic conversion. However, the main motivation of staged compilation is to
-%% improve performance and code abstraction, but these notions do \emph{not}
-%% respect conversion. For instance, rewriting programs along $\beta$-conversion
-%% can improve performance. Hence, at some point we may want to ``escape'' from
-%% quotients.
-
-%% Fortunately, our specified meta type theory can be interpreted in
-%% \emph{setoids}, as most recently explained in
-%% \cite{DBLP:journals/pacmpl/PujetT22}. A setoid is a set together with an
-%% equivalence relation on it. We interpret closed types as setoids, and dependent
-%% types as setoid fibrations. For inductive base types such $\Bool$ or $\Nat$, the
-%% equivalence relation is interpreted as identity. Functions are interpreted as
-%% equivalence-preserving functions, and two functions are equivalent if they are
-%% pointwise equivalent. A quotient type $A/R$ is interpreted by extending the
-%% semantic notion of equivalence for $A$ with the given $R$ relation. We will
-%% sketch the interpretation of the 2LTT syntax (given as a quotient
-%% inductive-inductive type) in Section \ref{sec:2lttsetoid}.
-
-%% In particular, from $f : A/R \to A/R$ we can extract a function which acts on
-%% the underlying set of $A$'s interpretation. Note that an extracted function can
-%% be only viewed as an algorithm if the setoid interpretation is itself defined in
-%% a constructive theory, but that is the case in the cited work
-%% \cite{DBLP:journals/pacmpl/PujetT22}.
-
-%% \begin{definition} Assume a closed function $f : A \to B$ defined
-%% in the mentioned type theory. The \textbf{extraction} of $f$ is
-%% the underlying function of $f$'s interpretation in the setoid model.
-%% \end{definition}
-
-%% In other words, we can view quotients as a convenient shorthand for working with
-%% explicit equivalence relations. We can work with quotients when we want to enforce
-%% preservation of equivalence relations, but at any point we have the option to
-%% switch to an external view and look at underlying sets and functions.
-
-%% %% In other words, we can view the usage of constructive quotients as a convenient
-%% %% shorthand for working with explicit equivalence relations. When we reason about
-%% %% the syntax of a type theory, this provides a bridge between the
-%% %% categorical/algebraic style and the traditional style that uses separate
-%% %% equivalence relations.
-%% %% It makes sense to work with quotients internally to a type theory whenever we
-%% %% only want to talk about constructions and properties which respect
-%% %% conversion. However, at any point we have the option to switch to an external
-%% %% perspective and look at underlying sets and the underlying maps.
-
-%% \subsection{Models and Syntax of 2LTT}\label{sec:2lttmod}
-
-%% We specify the notion of model for 2LTT in this section. The syntax of 2LTT will
-%% be defined as the initial model, as a quotient inductive-inductive type,
-%% following \cite{ttintt}. This yields an explicit substitution calculus,
-%% quotiented by definitional equality. A major advantage of this representation is
-%% that it \emph{only} includes well-typed syntax, so we never have to talk about
-%% ``raw'' types and terms.
-
-%% We provide an appendix to this paper, where we present a listing of all rules
-%% and equations of 2LTT in two styles: first, in a more traditional style with
-%% derivation rules, and also in a very compact manner using a second-order
-%% algebraic signature.
-
-%% First, we define the structural scaffolding of 2LTT without type formers, mostly
-%% following \cite{twolevel}. The specification is agnostic about the sizes of sets
-%% involved, i.e.\ we can model underlying sorts using any metatheoretic $\Set_j$.
-
-%% \begin{definition} A model of \textbf{basic 2LTT} consists of the following.
-%% \begin{itemize}
-%% \item
-%%   A category $\mbbc$ with a terminal object. We denote the set of objects as
-%%   $\Con_{\mbbc} : \Set$ and use capital Greek letters starting from $\Gamma$ to
-%%   refer to objects. The set of morphisms is $\Sub_{\mbbc} : \Con_{\mbbc} \to
-%%   \Con_{\mbbc} \to \Set$, and we use $\sigma$, $\delta$ and so on to refer to
-%%   morphisms. The terminal object is written as $\emptycon$ with unique morphism
-%%   $\epsilon : \Sub_\mbbc\,\Gamma\,\emptycon$. We omit the $\mbbc$ subscript if
-%%   it is clear from context.
-
-%%   In the initial model, an element of $\Con$ is a \emph{typing context}, an
-%%   element of $\Sub$ is a \emph{parallel substitution}, $\emptycon$ is the
-%%   \emph{empty typing context} and $\epsilon$ is the empty list viewed as a
-%%   parallel substitution. The identity substitution $\id : \Sub\,\Gamma\,\Gamma$
-%%   maps each variable to itself, and $\sigma \circ \delta$ can be computed
-%%   by substituting each term in $\sigma$ (which is a list of terms) with $\delta$.
-%% \item
-%%   For each $i \in \{0,1\}$ and $j \in \mbb{N}$, we have $\Ty_{i,j} : \Con \to
-%%   \Set$ and $\Tm_{i,j} : (\Gamma : \Con) \to \Ty_{i,j}\,\Gamma \to \Set$, as
-%%   sets of types and terms. Both types and terms can be substituted:
-%%   \begin{alignat*}{3}
-%%     & \blank\![\blank\!] &&: \Ty_{i,j}\,\Delta \to \Sub\,\Gamma\,\Delta \to \Ty_{i,j}\,\Gamma \\
-%%     & \blank\![\blank\!] &&: \Tm_{i,j}\,\Delta\,A \to (\sigma : \Sub\,\Gamma\,\Delta) \to \Tm_{i,j}\,\Gamma\,(A[\sigma])
-%%   \end{alignat*}
-%%   Additionally, we have $A[\id] = A$ and $A[\sigma \circ \delta] =
-%%   A[\sigma][\delta]$, and we have the same equations for term substitution as
-%%   well.
-
-%%   We also have a \emph{comprehension structure}: for each $\Gamma : \Con$
-%%   and $A : \Ty_{i,j}\,\Gamma$, we have the extended context $\Gamma \ext A :
-%%   \Con$ such that there is a natural isomorphism $\Sub\,\Gamma\,(\Delta\ext A)
-%%   \simeq (\sigma : \Sub\,\Gamma\,\Delta) \times
-%%   \Tm_{i,j}\,\Gamma\,(A[\sigma])$. We will sometimes use
-%%   $\blank\!\ext_0\!\blank$ and $\blank\!\ext_1\!\blank$ to disambiguate
-%%   object-level and meta-level context extensions.
-%% \item
-%%   For each $j$ we have a \emph{lifting structure}, consisting of a natural
-%%   transformation $\Lift : \Ty_{0,j}\,\Gamma \to \Ty_{1,j}\,\Gamma$, and an
-%%   invertible natural transformation $\qut{\blank} : \Tm_{0,j}\,\Gamma\,A \to
-%%   \Tm_{1,j}\,\Gamma\,(\Lift A)$, with inverse $\spl\blank$.
-%% \end{itemize}
-%% \end{definition}
-
-%% %% In short, for each $i$ and $j$ we have a family structure in the sense of
-%% %% categories-with-families \cite{Dybjer96internaltype}, such that there is a
-%% %% natural transformation from each $(\Ty_{0,j},\,\Tm_{0,j})$ to $(\Ty_{1,j},\,\Tm_{1,j})$
-%% %% which is invertible on terms.
-
-%% The following notions are derivable:
-%% \begin{itemize}
-%% \item
-%%   By moving left-to-right along $\Sub\,\Gamma\,(\Delta\ext A) \simeq (\sigma :
-%%   \Sub\,\Gamma\,\Delta) \times \Tm_{i,j}\,\Gamma\,(A[\sigma])$,
-%%   and starting from the identity morphism $\id : \Sub\,(\Gamma\ext A)\,(\Gamma\ext A)$, we recover
-%%   the \emph{weakening substitution} $\p : \Sub\,(\Gamma\ext A)\,\Gamma$ and the \emph{zero variable}
-%%   $\q : \Tm_{i,j}\,(\Gamma\ext A)\,(A[\p])$.
-%% \item
-%%   By weakening $\q$, we recover a notion of variables as De Bruijn
-%%   indices. In general, the $n$-th De Bruijn index is defined as $\q[\p^{n}]$,
-%%   where $\p^{n}$ denotes $n$-fold composition of $p$.
-%% \item
-%%   By moving right-to-left along $\Sub\,\Gamma\,(\Delta\ext A) \simeq (\sigma :
-%%   \Sub\,\Gamma\,\Delta) \times \Tm_{i,j}\,\Gamma\,(A[\sigma])$, we recover the
-%%   operation which extends a morphism with a term. In the initial model, this
-%%   extends a parallel substitution with an extra term, thus justifying the view
-%%   of substitutions as lists of terms. We denote the extension operation as
-%%   $(\sigma,\,t)$ for $\sigma : \Sub\,\Gamma\,\Delta$ and $t : \Tm_{i,j}\,\Gamma\,(A[\sigma])$.
-%% \end{itemize}
-
-%% \begin{notation}
-%% De Bruijn indices are rather hard to read, so we will sometimes use nameful notation
-%% for binders and substitutions. For example, we may write $\Gamma \ext (x : A)
-%% \ext (y : B)$ for a context, and subsequently write $B[x \mapsto t]$ for
-%% substituting the $x$ variable for some term $t : \Tm_{i,j}\,\Gamma\,A$. Using
-%% nameless notation, we would instead have $B : \Ty_{i,j}\,(\Gamma \ext A)$ and
-%% $B[\id,\,t] : \Ty_{i,j}\,\Gamma$; here we recover single substitution by extending the identity
-%% substitution $\id : \Sub\,\Gamma\,\Gamma$ with $t$.
-
-%% We may also leave weakening implicit: if a type or term is in a context
-%% $\Gamma$, we may use it in an extended context $\Gamma \ext A$ without marking
-%% the weakening substitution.
-%% \end{notation}
-
-%% \begin{definition}
-%% A \textbf{model of 2LTT} is a model of basic 2LTT which supports certain type
-%% formers. For the sake of brevity, we specify here only a small collection of
-%% type formers. However, we will argue later that our results extend to more
-%% general notions of inductive types. We specify type formers in the following. We
-%% omit substitution rules for type and term formers; all of these can be specified in
-%% an evident structural way, e.g.\ as in $(\Lift A)[\sigma] = \Lift(A[\sigma])$.
-%% \begin{itemize}
-%% \item \emph{Universes}. For each $i$ and $j$, we have a Coquand-style universe
-%%   \cite{coquand2018canonicity} in $\Ty_{i,j}$. This consists of $\U_{i,j} : \Ty_{i,j+1}\,\Gamma$,
-%%   together with $\El : \Tm_{i,j+1}\,\Gamma\,\U_{i,j} \to \Ty_{i,j}\,\Gamma$ and $\Code$, where $\Code$
-%%   and $\El$ are inverses.
-%% \item \emph{$\Sigma$-types}. We have $\Sigma\,(x : A)\,B : \Ty_{i,j}\,\Gamma$ for $A : \Ty_{i,j}\,\Gamma$
-%%   and $B : \Ty_{i,j}\,(\Gamma \ext (x : A))$, together with a natural isomorphism consisting of pairing and projections:
-%%   \[ \Tm_{i,j}\,\Gamma\,(\Sigma\,(x : A)\,B) \simeq (t : \Tm_{i,j}\,\Gamma\,A) \times \Tm_{i,j}\,\Gamma\,(B[x\mapsto t]) \]
-%%   We write $(t,\,u)$ for pairing and $\fst$ and $\snd$ for projections.
-%% \item \emph{Function types}.
-%%   We have $\Pi\,(x : A)\,B : \Ty_{i,j}\,\Gamma$ for $A : \Ty_{i,j}\,\Gamma$ and
-%%   $B : \Ty_{i,j}\,(\Gamma \ext (x : A))$, together with abstraction and application
-%%   \begin{alignat*}{3}
-%%     & \lam &&: \Tm_{i,j}\,(\Gamma \ext (x : A))\,B \to \Tm_{i,j}\,\Gamma\,(\Pi\,(x : A)\,B) \\
-%%     & \app &&: \Tm_{i,j}\,\Gamma\,(\Pi\,(x : A)\,B) \to \Tm_{i,j}\,(\Gamma \ext (x : A))\,B
-%%   \end{alignat*}
-%%   such that $\lam$ and $\app$ are inverses. Since we have explicit
-%%   substitutions, this specification of $\app$ is equivalent to the
-%%   ``traditional'' one:
-%%   \[ \app' : (t : \Tm_{i,j}\,\Gamma\,(\Pi\,(x :
-%%   A)\,B))\to (u : \Tm_{i,j}\,\Gamma\,A) \to \Tm_{i,j}\,\Gamma\,(B[x \mapsto
-%%     u]).\]
-%%   The traditional version is definable as $(\app\,t)[\id,\,u]$. We use
-%%   the $\app$-$\lam$ isomorphism because it is formally more convenient.
-
-%% \item \emph{Natural numbers}. We have $\Nat_{i,j} : \Ty_{i,j}\,\Gamma$, $\zero_{i,j} : \Tm_{i,j}\,\Gamma\,\Nat_{i,j}$,
-%%   and $\suc_{i,j} : \Tm_{i,j}\,\Gamma\,\Nat_{i,j} \to \Tm_{i,j}\,\Gamma\,\Nat_{i,j}$. The eliminator is the following.
-%%   \begin{alignat*}{5}
-%%     & \NatElim :\hspace{0.5em}
-%%                 &&  (P &&: \Ty_{i,k}\,(\Gamma \ext (n : \Nat_{i,j}))) \\
-%%     &           &&  (z &&: \Tm_{i,k}\,\Gamma\,(P[n\mapsto \zero_{i,j}])) \\
-%%     &           &&  (s &&: \Tm_{i,k}\,(\Gamma \ext (n : \Nat_{i,j}) \ext (\mit{pn} : P[n\mapsto n]))\,(P[n\mapsto \suc_{i,j}\,n]))\\
-%%     &           &&  (t &&: \Tm_{i,j}\,\Gamma\,\Nat_{i,j}) \\
-%%     &\hspace{3em}\to        &&&& \Tm_{i,k}\,\Gamma\,(P[n\mapsto t]))
-%%   \end{alignat*}
-%%   We also have the $\beta$-rules:
-%%   \begin{alignat*}{5}
-%%     & \NatElim\,P\,z\,s\,\zero_{i,j}     &&= z \\
-%%     & \NatElim\,P\,z\,s\,(\suc_{i,j}\,t) &&= s[n \mapsto t,\,\mit{pn} \mapsto \NatElim\,P\,z\,s\,t]
-%%   \end{alignat*}
-%%   Note that we can eliminate from a level $j$ to any level $k$.
-%% \end{itemize}
-%% \end{definition}
-
-%% \begin{definition}
-%% The \textbf{syntax of 2LTT} is the initial model of 2LTT, defined as a quotient
-%% inductive-inductive type \cite{kaposi2019constructing}. The signature for this
-%% type includes four sorts ($\Con$, $\Sub$, $\Ty$, $\Tm$), constructors for
-%% contexts, explicit substitutions and type/term formers, and equality
-%% constructors for all specified equations. The syntax comes with an
-%% \emph{induction principle}, from which we can also derive the \emph{recursion principle} as a
-%% non-dependent special case, witnessing the initiality of the syntax.
-%% \end{definition}
-
-%% \subsubsection{Comparison to Annekov et al.}
-%% Comparing our models to the primary reference on 2LTT \cite{twolevel}, the main
-%% difference is the handling of ``sizing'' levels. In ibid.\ there is a cumulative
-%% lifting from $\Ty_{i,j}$ to $\Ty_{i,j+1}$, which we do not assume. Instead,
-%% we allow elimination from $\Nat_{i,j}$ into any $k$ level. This means that we
-%% can manually define lifting maps from $\Nat_{i,j}$ to $\Nat_{i,j+1}$ by
-%% elimination. This is more similar to e.g.\ Agda, where we do not have cumulativity,
-%% but we can define explicit lifting from $\Nat_j : \Set_j$ to $\Nat_k : \Set_k$.
-
-%% In ibid., ``two-level type theory'' specifically refers to the setup
-%% where the object level is a homotopy type theory and the meta level is an
-%% extensional type theory. In contrast, we allow a wider range of setups under the
-%% 2LTT umbrella. Ibid.\ also considers a range of additional
-%% strengthenings and extensions of 2LTT \cite[Section~2.4]{twolevel}, most of which
-%% are useful in synthetic homotopy theory. We do not assume any of these and
-%% stick to the most basic formulation of 2LTT.
-
-%% \subsubsection{Elaborating Informal Syntax}
-%% The justification for the usage of the informal syntax in Section
-%% \ref{sec:tour-of-2ltt} is \emph{elaboration}. Elaboration translates surface
-%% notation to the formal syntax (often termed ``core syntax'' in the
-%% implementation of dependently typed programming languages). Elaboration is
-%% partial: it may throw error on invalid input. Elaboration also requires
-%% \emph{decidability of conversion} for the formal syntax, since type checking
-%% requires comparing types for conversion, which in turn requires comparing terms.
-%% We do not detail elaboration in this paper, we only make some observations.
-%% \begin{itemize}
-%% \item Decidability of conversion can be shown through normalization. Since 2LTT
-%%       is a minor variation on ordinary MLTT, we expect that existing proofs for MLTT
-%%       can be adapted without issue \cite{coquand2018canonicity,decidableconv}.
-%% \item We need to elaborate the surface-level Russell-style universe notation to
-%%       Coquand-style universes with explicit $\El$ and $\Code$ annotations. It is
-%%       likely that this can be achieved with the use of bidirectional elaboration
-%%       \cite{DBLP:journals/csur/DunfieldK21}, using two directions (checking and
-%%       inference) both for types and terms, and inserting $\El$ and $\Code$ when
-%%       switching between types and terms.
-%% \item The formal complexity of elaboration varies wildly, depending on how
-%%       explicit the surface syntax is. For example, Agda-style implicit arguments
-%%       require metavariables and unification, which dramatically increases the
-%%       difficulty of formalization.
-%% \end{itemize}
-
-%% \subsubsection{The Setoid Interpretation of the Syntax}\label{sec:2lttsetoid} We need to show
-%% that the 2LTT syntax, as a particular inductive type, can be modeled in setoids.
-%% Unfortunately, there is no prior work which presents the setoid interpretation
-%% of quotient types in the generality of quotient induction-induction. We sketch
-%% it for our specific use case, but we argue that the interpretation is fairly
-%% mechanical and could be adapted to any quotient inductive-inductive type.
-%% We focus on contexts and types here.
-
-%% We mutually inductively define $\Con : \Set$, $\Con^{\sim} : \Con \to \Con \to \Prop$, $\Ty_{i,j} : \Con \to \Set$
-%%       and $\Ty^{\sim} : \Ty_{i,j}\,\Gamma \to \Ty_{i,j}\,\Delta \to \Con^{\sim}\,\Gamma\,\Delta \to \Prop$, where
-%%       $\Con^{\sim}$ and $\Ty^{\sim}$ are proof-irrelevant relations. Note that $\Ty^{\sim}$ is \emph{heterogeneous},
-%%       since we can relate types in different (but provably convertible) contexts.
-
-%%       We specify $\emptycon$ and $\blank\ext\blank$ in $\Con$.
-To $\Con^{\sim}$ we add that $\emptycon$ and $\blank\ext\blank$ are congruences, and also that $\Con^{\sim}$ is reflexive, symmetric and transitive.
-
-%% We add all type formation rules to $\Ty_{i,j}$, and add all quotient equations to
-%% $\Ty^{\sim}$ as homogeneously indexed rules. For example, the substitution rule
-%% for $\Nat_0$ is specified as $ \Nat_0[] :
-%% \Ty^{\sim}\,(\Nat_0[\sigma])\,\Nat_0\,\refl^{\sim} $, where $\sigma :
-%% \Sub\,\Gamma\,\Delta$ and $\refl^{\sim} : \Con^{\sim}\,\Gamma\,\Gamma$. We also
-%% add all congruences and reflexivity, symmetry and transitivity to $\Ty^\sim$.
-
-%% We add a \emph{coercion} rule to types, as $\msf{coerce} :
-%% \Con^{\sim}\,\Gamma\,\Delta \to \Ty_{i,j}\,\Gamma \to \Ty_{i,j}\,\Delta$. This is probably
-%% familiar from the traditional specifications of type theories that use
-%% conversion relations. However, we additionally need a \emph{coherence rule}
-%% which witnesses $\Ty^{\sim}\,A\,(\msf{coerce}\,p\,A)\,p$ for each $A :
-%% \Ty_{i,j}\,\Gamma$ and $p : \Con^{\sim}\,\Gamma\,\Delta$. In short, coherence
-%% says that coercion preserves conversion.
-
-%% The concise description of this setup is that $(\Con,\,\Con^{\sim})$ is a
-%% setoid, and $(\Ty_{i,j},\,\Ty^{\sim})$ is a \emph{setoid fibration} over
-%% $(\Con,\,\Con^{\sim})$, for each $i$ and $j$. This notion of setoid fibration is
-%% equivalent to the ordinary notion of fibration, when setoids are viewed as
-%% categories with proof-irrelevant invertible morphisms. This is appropriate,
-%% since types in general are interpreted as setoid fibrations in the setoid model
-%% of a type theory.
-
-%% This scheme generalizes to substitutions and terms in a mechanical way. $\Sub$
-%% is indexed over two contexts, so it is interpreted as a fibration over
-%% the setoid product of $\Con$ and $\Con$. $\Tm$ is indexed over $(\Gamma :
-%% \Con)\times \Ty_{i,j}\,\Gamma$, so it is interpreted as a fibration over a setoid
-%% $\Sigma$-type. In every conversion relation we take the congruence closure of the specified
-%% quotient equations. Finally, we can use the induction principle for the thus
-%% specified sets and relations to interpret the internal elimination principle for
-%% the 2LTT syntax.
-
-%% %% We need to show that
-%% %% the 2LTT syntax, as a particular inductive type, can be modeled in setoids, if
-
-%% %% The examples in Section \ref{sec:tour-of-2ltt} can be faithfully represented in
-%% %% the formal syntax. The only notable difference is that the formal syntax uses
-%% %% Coquand-style universes instead of Russell-style ones. We use the former because
-%% %% it is much easier to model in the semantics in Section \ref{sec:staging-algorithm}. We note
-%% %% though that it is straightforward to elaborate from a Russell-style surface
-%% %% syntax to the formal version, by inserting $\El$ and $\Code$ as required.
-
-%% \subsection{Models and Syntax of the Object Theory}\label{sec:objmod}
-
-%% We also need to specify the object theory, which serves as the output language
-%% of staging. In general, the object theory corresponding to a particular flavor
-%% of 2LTT is simply the type theory that supports only the object-level $\Ty_{0,j}$
-%% hierarchy with its type formers.
-
-%% \begin{definition}
-%% A \textbf{model of the object theory} is a category-with-families, with types and
-%% terms indexed over $j \in \mbb{N}$, supporting Coquand-style universes $\U_j$,
-%% type formers $\Pi$, $\Sigma$ and $\Nat_j$, with elimination from $\Nat_j$ to
-%% any level $k$.
-%% \end{definition}
-
-%% \begin{definition}
-%% Like before, the \textbf{syntax of the object theory} is the initial model,
-%% given as a quotient inductive-inductive type, and it can be also interpreted
-%% in setoids.
-%% \end{definition}
-
-%% \begin{notation}
-%% From now on, by default we use $\Con$, $\Sub$, $\Ty$ and $\Tm$ to refer to sets
-%% in the syntax of 2LTT. We use $\mbbo$ to refer to the object syntax, and
-%% $\Con\ob$, $\Sub\ob$, $\Ty\ob$ and $\Tm\ob$ to refer to its underlying sets.
-%% \end{notation}
-
-%% \begin{definition}
-%% We recursively define the \textbf{embedding} of object syntax into 2LTT syntax,
-%% which preserves all structure strictly and consists of the following functions:
-%% \begin{alignat*}{4}
-%%   & \emb{\blank} : \Con\ob \to \Con  && \emb{\blank} : \Sub\ob\,\Gamma\,\Delta \to \Sub\,\emb{\Gamma}\,\emb{\Delta}\\
-%%   & \emb{\blank} : \Ty_{\mbbo j}\,\Gamma \to \Ty_{0,j}\,\emb{\Gamma}\hspace{2em} && \emb{\blank} : \Tm_{\mbbo j}\,\Gamma\,A \to \Tm_{0,j}\,\emb{\Gamma}\,\emb{A}
-%% \end{alignat*}
-%% \end{definition}
-
-%% \section{The Staging Algorithm}\label{sec:staging-algorithm}
-
-%% In this section we specify what we mean by a staging algorithm, then proceed to
-%% define one.
-
-%% \begin{definition}\label{def:staging}
-%% A \textbf{staging algorithm} consists of two functions:
-%%   \begin{alignat*}{3}
-%%     & \Stage : \Ty_{0,j}\,\emb{\Gamma} \to \Ty_{\mbbo j}\,\Gamma\hspace{2em} \Stage : \Tm_{0,j}\,\emb{\Gamma}\,A \to \Tm_{\mbbo j}\,\Gamma\,(\Stage\,A)
-%%   \end{alignat*}
-%% Note that we can stage open types and terms as long as their contexts are purely
-%% object-level. By \emph{closed staging} we mean staging only for closed types and
-%% terms.
-%% \end{definition}
-%% \begin{definition}
-%%   The following properties are of interest when considering a staging algorithm:
-%%   \begin{enumerate}
-%%   \item \emph{Soundness:} $\emb{\Stage\,A} = A$ and $\emb{\Stage\,t} = t$.
-%%   \item \emph{Stability:} $\Stage\,\emb{A} = A$ and $\Stage\,\emb{t} = t$.
-%%   \item \emph{Strictness:} the function extracted from $\Stage$ preserves all
-%%     type and term formers strictly (not just up to conversion).
-%%   \end{enumerate}
-%% \end{definition}
-
-%% Soundness and stability together state that embedding is a bijection
-%% on types and terms (up to conversion). This is a statement of
-%% \emph{conservativity} of 2LTT over the object theory. In \cite{twolevel} a
-%% significantly weaker conservativity theorem is shown, which only expresses that
-%% there exists a function from $\Tm_{0,j}\,\emb{\Gamma}\,\emb{A}$ to $\Tm_{\mbbo
-%%   j}\,\Gamma\,A$.
-
-%% Strictness tells us that staging does not perform any computation in the object
-%% theory; e.g.\ it does not perform $\beta$-reduction. This is practically quite
-%% important; for instance, \emph{full normalization of 2LTT} is a sound and stable
-%% staging algorithm, but it is not strict, and it is practically not useful, since
-%% it does not provide any control over inlining and $\beta$-reduction in code
-%% generation.
-
-%% Note that staging necessarily preserves type and term formers up to conversion,
-%% since it is defined on quotiented types and terms. More generally, staging
-%% respects conversion because every function must respect conversion in the
-%% meta type theory.
-
-%% \subsection{The Presheaf Model}
-
-%% We review the presheaf model described in \cite[Section~2.5.3]{twolevel}.  This
-%% will directly yield a closed staging algorithm, by recursive evaluation of 2LTT types and
-%% terms in the model.
-
-%% We give a higher-level overview first. Presheaves can be viewed as a
-%% generalization of sets, as ``variable sets'' which may vary over a base
-%% category. A set-based model of 2LTT yields a standard interpreter, where every
-%% syntactic object is mapped to the obvious semantic counterpart. The presheaf
-%% model over the category of contexts and substitutions in $\mbbo$ generalizes
-%% this to an interpreter where semantic values may depend on object-theoretic
-%% contexts. This allows us to handle object-theoretic \emph{types and terms}
-%% during interpretation, since these necessarily depend on their contexts.
-
-%% In the presheaf model, every semantic construction must be \emph{stable} under
-%% object-level substitution, or using alternative terminology, must be
-%% \emph{natural} with respect to substitution. Semantic values can be viewed as
-%% runtime objects during interpretation, and naturality means that
-%% object-theoretic substitutions can be \emph{applied} to such runtime objects,
-%% acting on embedded object-theoretic types and terms. Stability underlies
-%% the core trade-off in staging:
-%% \begin{itemize}
-%% \item On one hand, every syntactic rule and construction in 2LTT must
-%%       be stable, which restricts the range of metaprograms
-%%       that can be written in 2LTT. For example, we cannot make decisions
-%%       based on sizes of the object-theoretic scopes, since these sizes may be changed
-%%       by substitution.
-%% \item On the other hand, if only stable constructions are possible, we never
-%%       have to prove stability, or even explicitly mention object-theoretic
-%%       contexts, substitutions and variables. The tedious details of deep embeddings
-%%       can be replaced with a handful of staging operations and typing rules.
-%% \end{itemize}
-%% In principle, the base category in a 2LTT can be any category with finite
-%% products; it is not required that it is a category of contexts and substitutions.
-%% If the base category has more structure, we get more interesting object
-%% theories, but at the same time stability becomes more onerous, because we must
-%% be stable under \emph{more} morphisms. If the base category is simpler, with
-%% fewer morphisms, then the requirement of stability is less restrictive. In
-%% Section \ref{sec:intensional-analysis}, we will look at a concrete example for this.
-
-%% We proceed to summarize the key components of the model in a bit more detail. We
-%% denote the components of the model by putting hats on 2LTT components, e.g.\ as
-%% in $\hCon$.
-
-%% \begin{notation}
-%% In this section, we switch to naming elements of $\Cono$ as $a$, $b$ and $c$,
-%% and elements of $\Subo$ as $f$, $g$, and $h$, to avoid name clashing with
-%% contexts and substitutions in the presheaf model.
-%% \end{notation}
-
-%% \subsubsection{The Syntactic Category and the Meta-Level Fragment}
-
-%% \begin{definition} $\bs{\wh{\Con} : \Set_{\omega+1}}$ is defined as the set of presheaves
-%% over $\mbbo$. $\Gamma : \wh{\Con}$ has an action on objects $|\Gamma| :
-%% \Cono \to \Set_\omega$ and an action on morphisms $\blank[\blank] : |\Gamma|\,b
-%% \to \Subo\,a\,b \to |\Gamma|\,a$, such that $\gamma[\id] = \gamma$ and
-%% $\gamma[f\circ g] = \gamma[f][g]$.
-
-%% \begin{notation}
-%% Above, we reused the substitution notation $\blank[\blank]$ for the action on
-%% morphisms.  Also, we use lowercase $\gamma$ and $\delta$ to denote elements of
-%% $|\Gamma|\,a$ and $|\Delta|\,a$ respectively.
-%% \end{notation}
-
-%% \end{definition}
-%% \begin{definition} $\bs{\hSub\,\Gamma\,\Delta : \Set_\omega}$ is the set of natural transformations
-%% from $\Gamma$ to $\Delta$. $\sigma : \hSub\,\Gamma\,\Delta$ has action
-%% $|\sigma| : \{a : \Cono\} \to |\Gamma|\,a \to |\Delta|\,a$ such that
-%% $|\sigma|\,(\gamma[f]) = (|\sigma|\,\gamma)[f]$.
-%% \end{definition}
-
-%% \begin{definition}
-%% $\bs{\hTy_{1,j}\,\Gamma : \Set_\omega}$ is the set of \emph{displayed
-%% presheaves} over $\Gamma$; see e.g.\ \cite[Chapter~1.2]{huber-thesis}. This is
-%% equivalent to the set of presheaves over the category of elements of $\Gamma$,
-%% but it is usually more convenient in calculations. An $A : \hTy\,\Gamma$ has
-%% an action on objects $|A| : \{a : \Cono\} \to |\Gamma|\,a \to \Set_j$ and an
-%% action on morphisms $\blank[\blank] : |A|\,\gamma \to (f : \Subo\,a\,b) \to
-%% |A|\,(\gamma[f])$, such that $\alpha[\id] = \alpha$ and $\alpha[f \circ g] =
-%% \alpha[f][g]$.
-
-%% \begin{notation}
-%%   We write $\alpha$ and $\beta$ respectively for elements of $|A|\,\gamma$ and
-%%   $|B|\,\gamma$.
-%% \end{notation}
-%% \end{definition}
-
-%% \begin{definition}
-%%   $\bs{\hTm_{1,j}\,\Gamma\,A : \Set_\omega}$ is the set of sections of the
-%%   displayed presheaf $A$. This can be viewed as a dependently typed analogue of
-%%   a natural transformation.  A $t : \hTm_{1,j}\,\Gamma\,A$ has action $|t| :
-%%   \{a\} \to (\gamma : |\Gamma|\,a) \to |A|\,\gamma$ such that $|t|\,(\gamma[f])
-%%   = (|t|\,\gamma)[f]$.
-%% \end{definition}
-
-%% We also look at he empty context and context extension with meta-level types,
-%% as these will appear in subsequent definitions.
-
-%% \begin{definition}\label{def:psh-emptycon}
-%% $\bs{{\widehat{\emptycon}} : \hCon}$ is defined as the presheaf
-%% which is constantly $\top$, i.e.\ $|\wh{\emptycon}|\,\_ = \top$.
-%% \end{definition}
-
-%% \begin{definition}\label{def:psh-ext} For $A : \hTy_{1,j}\,\Gamma$, we define $\bs{\Gamma\,\,\wh{\ext}\,A}$
-%% pointwise by $|\Gamma\,\,\wh{\ext}\,A|\,a := (\gamma : |\Gamma|\,a) \times |A|\,\gamma$
-%% and $(\gamma,\,\alpha)[f] := (\gamma[f],\,\alpha[f])$.
-%% \end{definition}
-
-%% Using the above definitions, we can model the syntactic category of 2LTT, and
-%% also the meta-level family structure and all meta-level type formers. For
-%% expositions in previous literature, see \cite[Chapter~1.2]{huber-thesis} and
-%% \cite[Section~4.1]{Hofmann97syntaxand}.
-
-%% \subsubsection{The Object-Level Fragment}
-
-%% We move on to modeling the object-level syntactic fragment of 2LTT. We make some
-%% preliminary definitions. First, note that types in the object theory yield a
-%% presheaf, and terms yield a displayed presheaf over them; this immediately
-%% follows from the specification of a family structure in a cwf. Hence, we do a
-%% bit of a name overloading, and have $\Ty_{\mbbo j} : \hCon$ and $\Tm_{\mbbo j} : \hTy\,\Ty_{\mbbo j}$.
-
-%% \begin{definition}
-%% $\bs{\hTy_{0,j}\,\Gamma : \Set_\omega}$ is defined as $\hSub\,\Gamma\,\Ty_{\mbbo j}$,
-%% and $\bs{\hTm_{0,j}\,\Gamma\,A : \Set_\omega}$ is defined as $\hTm\,\Gamma\,(\Tm_{\mbbo j}[A])$.
-%% \end{definition}
-
-%% For illustration, if $A : \hTy_{0,j}\,\Gamma$, then $A :
-%% \hSub\,\Gamma\,\Ty_{\mbbo j}$, so $|A| : \{a : \Cono\} \to |\Gamma|\,a \to
-%% \Ty_{\mbbo j}\,a$. In other words, the action of $A$ on objects maps a semantic
-%% context to a syntactic object-level type. Likewise, for $t :
-%% \hTm_{0,j}\,\Gamma\,A$, we have $|t| : (\gamma : |\Gamma|\,a) \to \Tm_{\mbbo
-%%   j}\,a\,(|A|\,\gamma)$, so we get a syntactic object-level term as output.
-
-%% \begin{definition} For $A : \Ty_{0,j}\,\Gamma$, we define $\bs{\Gamma\,\,\wh{\ext}\,A}$
-%% as $|\Gamma\,\,\wh{\ext}\,A|\,a := (\gamma : |\Gamma|\,a) \times \Tm_{\mbbo
-%%   j}\,a\,(|A|\,\gamma)$ and $(\gamma,\,t)[f] := (\gamma[f],\,t[f])$. Thus,
-%% extending a semantic context with an object-level type appends an object-level
-%% term.
-%% \end{definition}
-
-%% Using the above definitions and following \cite{twolevel}, we can model all type
-%% formers in $\hTy_{0,j}$. Intuitively, this is because $\hTy_{0,j}$ and
-%% $\hTm_{0,j}$ return types and terms, so we can reuse the type and term formers
-%% in the object theory.
-
-%% \subsubsection{Lifting}
-
-%% \begin{definition}\label{def:psh-lifting}
-%% $\bs{\wh{\Lift} A : \hTy_{1,j}\,\Gamma}$ is defined as $\Tm_{\mbbo j}[A]$. With
-%%   this, we get that $\hTm_{0,j}\,\Gamma\,A$ is equal to
-%%   $\hTm_{1,j}\,\Gamma\,(\wh{\Lift}\,A)$. Hence, we can define both quoting and
-%%   splicing as identity functions in the model.
-%% \end{definition}
-
-%% \subsection{Closed Staging}
-
-%% \begin{definition}
-%% The \textbf{evaluation morphism, denoted $\bs{\ev}$} is a morphism of 2LTT
-%% models, between the syntax of 2LTT and $\hato$. It is defined by recursion on
-%% the syntax (that is, appealing to the initiality of the syntax), and it strictly
-%% preserves all structure.
-%% \end{definition}
-
-%% \begin{definition} \textbf{Closed staging} is defined as follows.
-%% \begin{alignat*}{4}
-%%   & \Stage : \Ty_{0,j}\,\emptycon \to \Ty_{\mbbo j}\,\emptycon \hspace{2em} && \Stage : \Tm_{0,j}\,\emptycon\,A \to \Tm_{\mbbo j}\,\emptycon\,(\Stage\,A) \\
-%%   & \Stage\,A := |\ev\,A|\,\{\emptycon\}\,\tt && \Stage\,t := |\ev\,t|\,\{\emptycon\}\,\tt
-%% \end{alignat*}
-%% Note that $\wh{\emptycon}$ is defined as the terminal presheaf, which is
-%% constantly $\top$. Also, $|\ev\,A|\{\emptycon\} : |\ev\,\emptycon|\,\emptycon \to \Ty_{\mbbo
-%%   j}\,\emptycon$, therefore $|\ev\,A|\{\emptycon\} : \top \to \Ty_{\mbbo
-%%   j}\,\emptycon$ and $|\ev\,t|\,\{\emptycon\} : \top \to \Tm_{\mbbo
-%%   j}\,\emptycon\,(|\ev\,A|\,\tt)$.
-%% \end{definition}
-
-%% What about general (open) staging though? Given $A : \Ty_{0,j}\,\emb{\Gamma}$,
-%% we get $|\ev\,A|\,\{\Gamma\} : |\emb{\Gamma}|\,\Gamma \to \Ty_{\mbbo j}$. We
-%% need an element of $|\emb{\Gamma}|\,\Gamma$, in order to obtain object-level
-%% type. Such ``generic'' semantic environments should be possible to construct, because
-%% elements of $|\emb{\Gamma}|\,\Gamma$ are essentially lists of object-level terms
-%% in $\Gamma$, by Definitions \ref{def:psh-ext} and \ref{def:psh-emptycon}, so $|\emb{\Gamma}|\,\Gamma$ should be
-%% isomorphic to $\Subo\,\Gamma\,\Gamma$.  It turns out that this falls out from
-%% the stability proof of $\ev$.
-
-%% \subsection{Open Staging, Stability and Strictness}
-
-%% We define a family of functions $\blank^P$ by induction on object syntax, such
-%% that the interpretation of a context yields a generic semantic environment. The
-%% induction motives are as follows.
-%% \begin{alignat*}{4}
-%%   & (\Gamma : \Cono)^P                    &&: |\ev\,\emb{\Gamma}|\,\Gamma && (\sigma : \Subo\,\Gamma\,\Delta)^P    &&: \Delta^P[\sigma] = |\ev\,\emb{\sigma}|\,\Gamma^P \\
-%%   & (A      : \Ty_{\mbbo j}\,\Gamma)^P      &&: A = |\ev\,\emb{A}|\,\Gamma^P\hspace{2em} && (t      : \Tm_{\mbbo j}\,\Gamma\,A)^P   &&: t = |\ev\,\emb{t}|\,\Gamma^P
-%% \end{alignat*}
-%% We look at the interpretation of contexts.
-%% \begin{itemize}
-%% \item  For $\emptycon^P$, we need an element of $|\ev\,\emb{\emptycon}|\,\emptycon$, hence
-%%   an element of $\top$, so we define $\emptycon^P$ as $\tt$.
-%% \item
-%%   For $(\Gamma \ext A)^P$, we need an element of
-%%     \[(\gamma : |\ev{\emb{\Gamma}}|\,(\Gamma\ext A)) \times \Tm_{\mbbo j}\,(\Gamma \ext A)\,(|\ev\emb{A}|\,\gamma).\]
-%%   We have $\Gamma^P : |\ev{\emb{\Gamma}}|\,\Gamma$, which we can weaken as
-%%   $\Gamma^P[\p] : |\ev{\emb{\Gamma}}|\,(\Gamma \ext A)$, so we set the first
-%%   projection of the result as $\Gamma^P[\p]$. For the second projection, the
-%%   goal type can be simplified as follows:
-%%   \begin{alignat*}{4}
-%%     &       &&\Tm_{\mbbo j}\,(\Gamma \ext A)\,(|\ev\emb{A}|\,(\Gamma^P[\p])) && \\
-%%     & =\,\, &&\Tm_{\mbbo j}\,(\Gamma \ext A)\,((|\ev\emb{A}|\,\Gamma^P)[\p]) &&\hspace{2em} \text{by naturality of $\ev\emb{A}$}\\
-%%     & =\,\, &&\Tm_{\mbbo j}\,(\Gamma \ext A)\,(A[\p])                        &&\hspace{2em} \text{by $A^P$}
-%%   \end{alignat*}
-%%   We have the zero de Bruijn variable $\q : \Tm_{\mbbo j}\,(\Gamma \ext
-%%   A)\,(A[\p])$.  Hence, we define $(\Gamma \ext A)^P$ as $(\Gamma^P[\p],\,\q)$.
-%% \end{itemize}
-%% Thus, a generic semantic context $\Gamma^P : |\ev\emb{\Gamma}|\,\Gamma$ is just
-%% a list of variables, corresponding to the identity substitution $\id :
-%% \Subo\,\Gamma\,\Gamma$ which maps each variable to itself.
-
-%% The rest of the $\blank^P$ interpretation is straightforward and we omit it
-%% here. In particular, preservation of definitional equalities is automatic, since
-%% types, terms and substitutions are all interpreted as proof-irrelevant
-%% equations.
-
-%% \begin{definition}\label{def:open-staging} We define \textbf{open staging} as follows.
-%% \begin{alignat*}{4}
-%%   & \Stage : \Ty_{0,j}\,\emb{\Gamma} \to \Ty_{\mbbo j}\,\Gamma \hspace{2em} && \Stage : \Tm_{0,j}\,\emb{\Gamma}\,A \to \Tm_{\mbbo j}\,\Gamma\,(\Stage\,A) \\
-%%   & \Stage\,A := |\ev\,A|\,\Gamma^P && \Stage\,t := |\ev\,t|\,\Gamma^P
-%% \end{alignat*}
-%% \end{definition}
-
-%% \begin{theorem} Open staging is stable.
-%% \end{theorem}
-%% \begin{proof} For $A : \Ty_{\mbbo j}$, $\Stage\,\emb{A}$ is by definition $|\ev\,\emb{A}|\,\Gamma^P$,
-%%   hence by $A^P$ it is equal to $A$. Likewise, $\Stage\,\emb{t}$ is equal to $t$ by $t^P$.
-%% \end{proof}
-
-%% \begin{theorem} Open staging is strict.
-%% \end{theorem}
-%% \begin{proof}
-%%   This is evident from the definition of the presheaf model, where the action of
-%%   each type and term former in $\wh{\Ty}_{0,j}$ and $\wh{\Tm}_{0,j}\,\Gamma\,A$ is
-%%   defined using the corresponding type or term former in the object syntax.
-%% \end{proof}
-
-%% \subsection{Implementation and Efficiency}
-
-%% We discuss now the operational behavior of the extracted staging algorithm
-%% and look at potential adjustments and optimizations that make it more efficient
-%% or more convenient to implement. Recall the types of the component functions in $\ev$:
-%% \begin{alignat*}{3}
-%%   & |\blank| \circ \ev : \Ty_{1 j}\,\Gamma \to \{\Delta : \Cono\} \to |\ev\,\Gamma|\,\Delta \to \Set_j\\
-%%   & |\blank| \circ \ev : \Ty_{0 j}\,\Gamma \to \{\Delta : \Cono\} \to |\ev\,\Gamma|\,\Delta \to \Ty_{\mbbo j}\,\Delta\\
-%%   & |\blank| \circ \ev : \Tm_{1 j}\,\Gamma\,A \to \{\Delta : \Cono\} \to (\gamma : |\ev\,\Gamma|\,\Delta) \to
-%%     |\ev\,A|\,\gamma \\
-%%   & |\blank| \circ \ev : \Tm_{0 j}\,\Gamma\,A \to \{\Delta : \Cono\} \to (\gamma : |\ev\,\Gamma|\,\Delta) \to
-%%     \Tm_{\mbbo\,j}\,\Delta\,(|\ev\,A|\,\gamma)
-%% \end{alignat*}
-%% We interpret syntactic types or terms in a semantic environment
-%% $\gamma : |\ev\,\Gamma|\,\Delta$. These environments are lists containing
-%% a mix of  object-level terms and semantic values. The semantic values are represented
-%% using metatheoretic inductive types and function types.
-%% \begin{itemize}
-%%   \item $|\Nat_1|\,\_$ is simply the set of natural numbers $\mbb{N}$.
-%%   \item $|\Sigma_1\,A\,B|\,\gamma$ is simply a set of pairs of values.
-%%   \item A non-dependent function type $A\,\wh{\to}\,B$ is defined as the presheaf
-%%         exponential. The computational part of $|A\,\wh{\to}\,B|\,\{\Delta\}\,\gamma$
-%%         is given by a function with type
-%%         \[ (\Theta : \Cono) \to (\sigma : \Subo\,\Theta\,\Delta) \to |A|\,(\gamma[\sigma]) \to |B|\,(\gamma[\sigma])        \]
-%%         This may be also familiar as the semantic implication from the Kripke
-%%         semantics of intuitionistic logics. Whenever we evaluate a function
-%%         application, we supply an extra $\id : \Subo\,\Delta\,\Delta$.
-%%         This may incur cost via the $\gamma[\id]$ restrictions in a naive
-%%         implementation, but this is easy to optimize, by introducing a formal
-%%         representation of $\id$, such that $\gamma[\id]$ immediately computes to
-%%         $\gamma$. The case of dependent functions is analogous operationally.
-%% \end{itemize}
-%% In summary, in the meta-level fragment of 2LTT, $\ev$ yields a reasonably efficient
-%% computation of closed values, which reuses functions from the ambient
-%% metatheory. Alternatively, instead of using ambient functions, we could use our
-%% own implementation of function closures during staging.
-
-%% In contrast, we have a bit of an efficiency problem in the handling of
-%% object-level binders: whenever we go under such binder we have to weaken the
-%% current semantic environment. Concretely, when moving from $\Delta$ to
-%% $\Delta \ext A$, we have to shift $\gamma : |\ev\,\Gamma|\,\Delta$ to
-%% $\gamma[\p] : |\ev\,\Gamma|\,(\Delta \ext A)$. This cannot be avoided by an easy
-%% optimization trick. For object-level entries in the environment, this is cheap,
-%% because we just add an extra explicit weakening, but for semantic values
-%% weakening may need to perform deep traversals.
-
-%% The same efficiency issue arises in formal presheaf-based
-%% normalization-by-evaluation, such as in \cite{kaposinbe} and
-%% \cite{coquand2018canonicity}. However, in practical implementations this issue
-%% can be fully solved by using De Bruijn \emph{levels} in the semantic domain,
-%% thus arranging that weakening has no operational cost on semantic values; see
-%% \cite{coquand1996algorithm}, \cite{modulartc} and \cite{untypedtc} for
-%% implementations in this style. We can use the same solution in staging. It is a
-%% somewhat unfortunate mismatch that indices are far more convenient in
-%% formalization, but levels are more efficient in practice.
-
-%% In our prototype implementation, we use the above optimization with De Bruijn
-%% levels, and we drop explicit substitutions from the 2LTT core syntax. In the
-%% object-level syntax, we use De Bruijn levels for variables and closures in
-%% binders. This lets us completely skip the weakening of environments, which is
-%% slightly more efficient than the more faithfully extracted algorithm.
-
-%% Additionally, we use an untyped tagged representation of semantic
-%% values, since in Haskell (the implementation language) we do not have
-%% strong enough types to represent typed semantic values. For example, there
-%% are distinct data constructors storing semantic function values, natural number
-%% literals and quoted expressions.
-
-%% \paragraph{Caching} In a production-strength implementation we would need some
-%% caching mechanism, to avoid excessive duplication of code. For example, if we
-%% use the staged $\mit{map}$ function multiple times with the same type and
-%% function arguments, we do not want to generate separate code for each
-%% usage. De-duplicating object-level code with function types is usually safe,
-%% since function bodies become closed top-level code after closure conversion. We
-%% leave this to future work.
-
-%% \section{Soundness of Staging}\label{sec:soundness}
-
-%% In this section we prove soundness of staging. We build a proof-relevant logical
-%% relation between the evaluation morphism $\ev$ and a \emph{restriction}
-%% morphism, which restricts 2LTT syntax to object-level contexts. The relational
-%% interpretations of $\Ty_{0,j}$ and $\Tm_{0,j}$ will yield the soundness
-%% property.
-
-%% \subsection{Working in $\hat{\mbbo}$}
-
-%% We have seen that 2LTT can be modeled in the presheaf topos
-%% $\hato$. Additionally, $\hato$ supports all type formers of extensional type
-%% theory and certain other structures which are stable under object-level
-%% substitution. As all constructions in this section must be stable, it makes
-%% sense to work internally to $\hato$. This style has been previously used in
-%% normalization proofs \cite{coquand2018canonicity} and also in the metatheory of
-%% cubical type theories
-%% \cite{licata2018internal,orton_et_al:LIPIcs:2016:6564,DBLP:conf/csl/CavalloMS20}.
-
-%% When we work internally in a model of a type theory, we do not explicitly refer
-%% to contexts, types, and substitutions. For example, when working in Agda, we do
-%% not refer to Agda's typing contexts. Instead, we only work with terms, and use
-%% functions and universes to abstract over types and semantic contexts. Hence, we
-%% have to convert along certain isomorphisms when we switch between the internal
-%% and external views. In the following, we summarize features in $\hato$ and also
-%% the internal-external conversions.
-
-%% We write $\whset_j$ for ordinal-indexed Russell-style universes. Formally, we
-%% have Coquand-style universes, but for the sake of brevity we omit $\El$ and
-%% $\Code$ from internal syntax. Universes are cumulative, and closed under $\Pi$,
-%% $\Sigma$, extensional identity $\blank\!=\!\blank$ and inductive types. We use
-%% the same conventions as in Notation \ref{basic-notation}.
-
-%% \noindent The basic scheme for internalization is as follows:
-%% \begin{alignat*}{3}
-%%   & \Gamma : \hCon                  && \hspace{1em}\text{is internalized as}\hspace{1em} \Gamma : \whset_\omega\\
-%%   & \sigma : \hSub\,\Gamma\,\Delta  && \hspace{1em}\text{is internalized as}\hspace{1em} \sigma : \Gamma \to \Delta\\
-%%   & A : \hTy_{1,j}\,\Gamma           && \hspace{1em}\text{is internalized as}\hspace{1em} A      : \Gamma \to \whset_j \\
-%%   & t : \hTm_{1,j}\,\Gamma\,A        && \hspace{1em}\text{is internalized as}\hspace{1em} t      : (\gamma : \Gamma) \to A\,\gamma
-%% \end{alignat*}
-%% \subsubsection{Object-Theoretic Syntax} The syntax of the object theory is clearly
-%% fully stable under object-theoretic substitution, so we can internalize all of
-%% its type and term formers. We internalize object-level types and terms as
-%% $\Ty_{\mbbo j} : \whset_0$ and $\Tm_{\mbbo j} : \Ty_{\mbbo j} \to
-%% \whset_0$. $\Ty_{\mbbo j}$ is closed under type formers. For instance, we have
-%% \[ \Nat_j : \Ty_{\mbbo j}\hspace{1.5em}\zero_j : \Tm_{\mbbo j}\,\Nat_j\hspace{1.5em}
-%%     \suc_j : \Tm_{\mbbo j}\,\Nat_j \to \Tm_{\mbbo j}\,\Nat_j \]
-%% together with $\NatElim$, and likewise we have all other type formers.
-%% \subsubsection{Internal $\ev$} We also use an internal view of
-%% $\ev$ which maps 2LTT syntax to internal values; i.e.\ we compose $\ev$ with
-%% internalization. Below, note that the input to $\ev$ is \emph{external}, so we
-%% mark $\ev$ as being parameterized by external input.
-%% \begin{alignat*}{3}
-%%   &\ev\,(\Gamma &&: \Con)                      &&: \whset_\omega \\
-%%   &\ev\,(\sigma &&: \Sub\,\Gamma\,\Delta)      &&: \ev\,\Gamma \to \ev\,\Delta \\
-%%   &\ev\,(A      &&: \Ty_{1,j}\,\Gamma)          &&: \ev\,\Gamma \to \whset_j\\
-%%   &\ev\,(t      &&: \Tm_{1,j}\,\Gamma)          &&: (\gamma : \ev\,\Gamma) \to \ev\,A\,\gamma\\
-%%   &\ev\,(A      &&: \Ty_{0,j}\,\Gamma)          &&: \ev\,\Gamma \to \Ty_{\mbbo_j}\\
-%%   &\ev\,(t      &&: \Tm_{0,j}\,\Gamma\,A)       &&: (\gamma : \ev\,\Gamma) \to \Tm_{\mbbo j}\,(\ev\,A\,\gamma)
-%% \end{alignat*}
-%% \subsubsection{Object-Level Fragment of 2LTT}
-%% The purely object-level syntactic fragment of 2LTT can be internalized as
-%% follows. We define externally the presheaf of object-level types as
-%% $|\Ty_{0,j}|\,\Gamma := \Ty_{0,j}\,\emb{\Gamma}$, and the displayed presheaf of
-%% object-level terms over $\Ty_{0,j}$ as $|\Tm_{0,j}|\,\{\Gamma\}\,A :=
-%% \Tm_{0,j}\,\emb{\Gamma}\,A$. Hence, internally we have $\Ty_{0,j} : \whset_0$ and
-%% $\Tm_{0,j} : \Ty_{0,j} \to \whset_0$. $\Ty_{0,j}$ is closed under all type formers,
-%% analogously as we have seen for $\Ty_{\mbbo j}$.
-
-%% %% \begin{alignat*}{6}
-%% %%   &\ev\,(\Gamma &&: \Con)                      &&: \whset_\omega && \ev\,(t      &&: \Tm_{1,j}\,\Gamma)          &&: (\gamma : \ev\,\Gamma) \to \ev\,A\,\gamma\\
-%% %%   &\ev\,(\sigma &&: \Sub\,\Gamma\,\Delta)      &&: \ev\,\Gamma \to \ev\,\Delta \hspace{3.5em}&& \ev\,(A      &&: \Ty_{0,j}\,\Gamma)          &&: \ev\,\Gamma \to \Ty_{\mbbo_j}\\
-%% %%   &\ev\,(A      &&: \Ty_{1,j}\,\Gamma)          &&: \ev\,\Gamma \to \whset_j&& \ev\,(t      &&: \Tm_{0,j}\,\Gamma\,A)       &&: (\gamma : \ev\,\Gamma) \to \Tm_{\mbbo j}\,(\ev\,A\,\gamma)
-%% %% \end{alignat*}
-
-%% %% \begin{alignat*}{3}
-%% %%   &\ev\,(\Gamma &&: \Con)                      &&: \whset_\omega \\
-%% %%   &\ev\,(\sigma &&: \Sub\,\Gamma\,\Delta)      &&: \ev\,\Gamma \to \ev\,\Delta \\
-%% %%   &\ev\,(A      &&: \Ty_{1,j}\,\Gamma)          &&: \ev\,\Gamma \to \whset_j\\
-%% %%   &\ev\,(t      &&: \Tm_{1,j}\,\Gamma)          &&: (\gamma : \ev\,\Gamma) \to \ev\,A\,\gamma\\
-%% %%   &\ev\,(A      &&: \Ty_{0,j}\,\Gamma)          &&: \ev\,\Gamma \to \Ty_{\mbbo_j}\\
-%% %%   &\ev\,(t      &&: \Tm_{0,j}\,\Gamma\,A)       &&: (\gamma : \ev\,\Gamma) \to \Tm_{\mbbo j}\,(\ev\,A\,\gamma)
-%% %% \end{alignat*}
-
-%% \subsubsection{Embedding}
-%% Now, the embedding operation $\emb{\blank}$ can be also internalized on types
-%% and terms, as $\emb{\blank} : \Ty_{\mbbo j} \to \Ty_{0,j}$, and $\emb{\blank}
-%% :\Tm_{\mbbo\,j}\,A \to \Tm_{0,j}\,\emb{A}$. Embedding strictly preserves all structure.
-
-%% \subsection{The Restriction Morphism}
-
-%% We define a family of functions $\re$ from the 2LTT syntax to objects in
-%% $\hato$. We will relate this to the evaluation morphism $\ev$ in the relational
-%% interpretation. In short, $\re$ restricts 2LTT syntax so that it can only depend
-%% on object-level contexts, which are given as $\emb{\Gamma}$.
-
-%% \begin{definition}\label{def:restriction} We specify the types of the \textbf{restriction operations} internally, and the $|\blank|$
-%% components of the operations externally. The naturality of $|\blank|$ is
-%% straightforward in each case.
-%% \begingroup
-%% %% \allowdisplaybreaks
-%% \begin{alignat*}{5}
-%%   & \re\,(\Gamma : \Con) : \whset_0 && \re\,(\sigma : \Sub\,\Gamma\,\Delta) : \re\,\Gamma \to \re\,\Delta \\
-%%   & |\re\,\Gamma|\,\Delta := \Sub\,\emb{\Delta}\,\Gamma && |\re\,\sigma|\,\gamma := \sigma \circ \gamma \\
-%%   & \\
-%%   & \re\,(A : \Ty_{1,j}\,\Gamma) : \re\,\gamma \to \whset_0 && \re\,(t : \Tm_{1,j}\,\Gamma\,A) : (\gamma : \re\,\Gamma) \to \re\,A\,\gamma \\
-%%   & |\re\,A|\,\{\Delta\}\,\gamma := \Tm_{1,j}\,\emb{\Delta}\,(A[\gamma])\hspace{2em} && |\re\,t|\,\gamma := t[\gamma]\\
-%%   & \\
-%%   & \re\,(A : \Ty_{0,j}\,\Gamma) : \re\,\Gamma \to \Ty_{0,j}\hspace{2em} && \re\,(t : \Tm_{0,j}\,\Gamma\,A) : (\gamma : \re\,\Gamma) \to \Tm_{0,j}\,(\re\,A\,\gamma)\\
-%%   & |\re\,A|\,\gamma := A[\gamma] && |\re\,t|\,\gamma := t[\gamma]
-%% \end{alignat*}
-%% \endgroup
-%% \end{definition}
-
-%% \subsubsection{Preservation Properties of $\re$}
-%% First, we note that $\re$ strictly preserves $\id$, $\blank\circ\blank$ and
-%% type/term substitution, and it preserves $\emptycon$ and $\blank\ext_1\blank$ up
-%% to isomorphism. We have the following isomorphisms internally to $\hato$:
-%% \begin{alignat*}{3}
-%%   & \re_{\emptycon} &&: \re\,\emptycon \simeq \top \\
-%%   & \re_{\ext_1}   &&: \re\,(\Gamma \ext_1 A) \simeq ((\gamma : \re\,\Gamma) \times \re\,A\,\gamma)
-%% \end{alignat*}
-
-%% \begin{notation} When we have an isomorphism $f : A \simeq B$, we may write $f$ for the
-%%   function in $A \to B$, and $f^{-1} : B \to A$ for its inverse.
-%% \end{notation}
-
-%% \begin{notation}
-%%   We can use a pattern matching notation on isomorphisms. For example, if $f : A
-%%   \simeq B$, then we may write $(\lambda\,(f\,a).\,t) : B \to C$, and likewise
-%%   $(\lambda\,(f^{-1}\,b).\,t) : A \to C$, where the function bodies can refer to
-%%   the bound $a : A$ and $b : B$ variables.
-%% \end{notation}
-
-%% The preservation properties $\re_{\emptycon}$ and $\re_{\ext_1}$ mean that $\re$
-%% is a \emph{pseudomorphism} in the sense of \cite{gluing}, between the syntactic
-%% cwf given by $(\Ty_{1,j},\,\Tm_{1,j})$ and the corresponding cwf structure in
-%% $\hato$. In ibid.\ there is an analysis of such cwf morphisms, from which we
-%% obtain the following additional preservation properties:
-%% \begin{itemize}
-%% \item Meta-level $\Sigma$-types are preserved up to isomorphism, so we have
-%%   \[ \re_{\Sigma} : \re\,(\Sigma\,A\,B)\,\gamma
-%%   \simeq ((\alpha : \re\,A) \times \re\,B\,(\rexti\,(\gamma,\,\alpha))). \]
-%%   The semantic values of $\re\,(\Sigma\,A\,B)\,\gamma$
-%%   are 2LTT terms with type $(\Sigma\,A\,B)[\gamma]$, restricted to object-level contexts.
-%%   We can still perform pairing and projection with such restricted terms; hence
-%%   the preservation property.
-%% \item Meta-level $\Pi$-types and universes are preserved in a lax way. For $\Pi$, we have
-%%   \[ \re_{\app} : \re\,(\Pi\,A\,B)\,\gamma \to (\alpha : \re\,A\,\gamma) \to \re\,B\,(\rexti\,(\gamma,\,\alpha))
-%%   \]
-%%   such that $\re_{\app}\,(\re\,t\,\gamma)\,\alpha =
-%%   \re\,(\app\,t)\,(\rexti\,(\gamma,\,\alpha))$. In this case, we can apply a
-%%   restricted term with a $\Pi$-type to a restricted term, but we cannot do
-%%   lambda-abstraction, because that would require extending the context with a
-%%   meta-level binder. For $\U_{1,j}$, we have
-%%   \[
-%%     \re_{\El} : \re\,\U_{1,j}\,\gamma \to \whset_j
-%%   \]
-%%   such that $\re_{\El}\,(\re\,t\,\gamma) = \re\,(\El\,t)\,\gamma$. Here, we only
-%%   have lax preservation simply because $\whset_j$ is much larger than the the
-%%   set of syntactic 2LTT types, so not every semantic $\whset_j$ has a syntactic
-%%   representation.
-
-%% \item Meta-level positive (inductive) types are preserved in an oplax way. In the case of
-%%   natural numbers, we have
-%%   \[ \re_{\mbb{N}} : \mbb{N} \to \re\,\Nat_{1,j}\,\gamma. \]
-%%   This is a ``serialization'' map: from a metatheoretic natural number we
-%%   compute a numeral as a closed 2LTT term. This works analogously for other
-%%   inductive types, including parameterized ones. For an example, from a semantic
-%%   list of restricted terms we would get a syntactic term with list type,
-%%   containing the same restricted terms.
-%% \end{itemize}
-%% \subsubsection{Action on Lifting} We have an isomorphism $\re\,(\Lift\,A)\,\gamma \simeq
-%% \Tm_{0,j}\,(\re\,A\,\gamma)$. This is given by quoting and splicing: we convert
-%% between restricted meta-level terms with type $\Lift\,A$ and restricted
-%% object-level terms with type $A$. Hence, we write components of this isomorphism
-%% as $\spl\blank$ and $\qut{\blank}$, as internal analogues of the external
-%% operations. With this, we also have $\re\,\qut{t}\,\gamma =
-%% \qut{\re\,t\,\gamma}$ and $\re\,\spl\,t\,\gamma = \spl(\re\,t\,\gamma)$.
-
-%% \subsubsection{Action on $\blank\ext_0\blank$} We have preservation up to isomorphism:
-%% \[
-%%   \re_{\ext_0} : \re\,(\Gamma \ext_0 A) \simeq ((\gamma : \re\,\Gamma)\times \Tm_{0,j}\,(\re\,A\,\gamma))
-%% \]
-%% This is because substitutions targeting $\Gamma \ext_0 A$ are the same as
-%% pairs of substitutions and terms, by the specification of $\blank\!\ext_0\!\blank$.
-
-%% \subsubsection{Action on Object-Level Types and Terms} $\re$ preserves all structure
-%% in the object-level fragment of 2LTT. This follows from the $\re$ specification:
-%% an external type $A : \Ty_{0,j}\,\Gamma$ is directly internalized as an element
-%% of $\Ty_{0,j}$, and the same happens for $t : \Tm_{0,j}\,\Gamma\,A$.
-
-%% \subsection{The Logical Relation}
-
-%% Internally to $\hato$, we define by induction on the syntax of 2LTT a
-%% proof-relevant logical relation interpretation, written as $\blank^{\approx}$. The induction
-%% motives are specified as follows.
-%% \begin{alignat*}{5}
-%%   &(\Gamma &&: \Con)\rel && : \ev\,\Gamma \to \re\,\Gamma \to \whset_\omega\\
-%%   &(\sigma &&: \Sub\,\Gamma\,\Delta)\rel &&: \Gamma\rel\,\gamma\,\gamma' \to \Delta\rel\,(\ev\,\sigma\,\gamma)\,(\re\,\sigma\,\gamma')\\
-%%   &(A      &&: \Ty_{1,j}\,\Gamma)\rel &&: \Gamma\rel\,\gamma\,\gamma' \to \ev\,A\,\gamma \to \re\,A\,\gamma' \to \whset_j\\
-%%   &(t      &&: \Tm_{1,j}\,\Gamma\,A)\rel &&: (\gamma\rel : \Gamma\rel\,\gamma\,\gamma') \to A\rel\,\gamma\rel\,(\ev\,t\,\gamma)\,(\re\,t\,\gamma')\\
-%%   &(A      &&: \Ty_{0,j}\,\Gamma)\rel &&: \Gamma\rel\,\gamma\,\gamma' \to \emb{\ev\,A\,\gamma} = \re\,A\,\gamma' \\
-%%   &(t      &&: \Tm_{0,j}\,\Gamma\,A)\rel &&: \Gamma\rel\,\gamma\,\gamma' \to \emb{\ev\,t\,\gamma} = \re\,t\,\gamma'
-%% \end{alignat*}
-%% For $\Con$, $\Sub$ and meta-level types and terms, this is a fairly standard
-%% logical relation interpretation: contexts are mapped to relations, types to
-%% dependent relations, and substitutions and terms respect relations. We will only have
-%% modest complications in meta-level type formers because we will sometimes need to use
-%% the lax/oplax preservations properties of $\re$. For object-level types and
-%% terms, we get soundness statements: evaluation via $\ev$ followed by embedding
-%% back to 2LTT is the same as restriction to object-level contexts. We describe
-%% the $\blank\rel$ interpretation in the following.
-
-%% \subsubsection{Syntactic Category and Terminal Object}
-%% Here, we simply have $\id\rel\,\gamma\rel := \gamma\rel$ and $(\sigma \circ
-%% \delta)\rel\,\gamma\rel := \sigma\rel\,(\delta\rel\,\gamma\rel)$. The terminal
-%% object is interpreted as $\emptycon\rel\,\gamma\,\gamma' := \top$.
-
-%% \subsubsection{Meta-Level Family Structure}
-%% We interpret context extension and type/term substitution as follows. Note the
-%% usage of the pattern matching notation on the $\rexti$ isomorphism.
-%% \begin{alignat*}{4}
-%%   & (\Gamma \ext_1 A)\rel\,(\gamma,\,\alpha)\,(\rexti(\gamma',\,\alpha')) && :=
-%%     (\gamma\rel : \Gamma\rel\,\gamma\,\gamma') \times A\rel\,\gamma\rel\,\alpha\,\alpha'\\
-%%   & (A[\sigma])\rel\,\gamma\rel\,\alpha\,\alpha' && := A\rel\,(\sigma\rel\,\gamma\rel)\,\alpha\,\alpha'\\
-%%   & (t[\sigma])\rel\,\gamma\rel\                 && := t\rel\,(\sigma\rel\,\gamma\rel)
-%% \end{alignat*}
-%% It is enough to specify the $\blank\rel$ action on extended substitutions
-%% $(\sigma,\,t) : \Sub\,\Gamma\,(\Delta\ext A)$, $\p : \Sub\,(\Gamma\ext
-%% A)\,\Gamma$, $\p : \Sub\,(\Gamma\ext A)\,\Gamma$ and $\q : \Tm_{1,j}\,(\Gamma
-%% \ext A)\,(A[\p])$. The category-with-families equations hold evidently.
-%% \begin{alignat*}{4}
-%%   & (\sigma,\,t)\rel\,\gamma\rel := (\sigma\rel\,\gamma\rel,\,t\rel\,\gamma\rel) \hspace{2em} \p\rel\,(\gamma\rel,\,t\rel) := \gamma\rel \hspace{2em} \q\rel\,(\gamma\rel,\,t\rel) := t\rel
-%% \end{alignat*}
-
-%% \subsubsection{Meta-Level Natural Numbers} First, we have to define a relation:
-%% \begin{alignat*}{4}
-%%   & (\Nat_{1,j})\rel : \Gamma\rel\,\gamma\,\gamma' \to \mbb{N} \to \re\,\Nat_{1,j}\,\gamma' \to \whset_j\\
-%%   & (\Nat_{1,j})\rel\,\gamma\rel\,n\,n' := (\re_{\mbb{N}}\,n = n')
-%% \end{alignat*}
-%% Note that evaluation sends $\Nat_{1,j}$ to the semantic type of natural numbers,
-%% i.e.\ $\ev\,\Nat_{1,j} = \mbb{N}$. We refer to the serialization map
-%% $\re_{\mbb{N}} : \mbb{N} \to \re\,\Nat_{1,j}\,\gamma'$ above. In short,
-%% $(\Nat_{1,j})\rel$ expresses \emph{canonicity}: $n'$ is canonical precisely if
-%% it is of the form $\re_{\mbb{N}}\,n$ for some $n$.
-
-%% For $\zero\rel$ and $\suc\rel$, we need to show that serialization preserves
-%% $\zero$ and $\suc$ respectively, which is evident.
-
-%% Let us look at elimination. We need to define the following:
-%% \begin{alignat*}{4}
-%%   &(\msf{NatElim}\,P\,z\,s\,n)\rel\,\gamma\rel :
-%%   P\rel\,(\gamma\rel,\,n\rel\,\gamma\rel)\,(\ev\,(\msf{NatElim}\,P\,z\,s\,n)\,\gamma)\,
-%%                                            (\re\,(\msf{NatElim}\,P\,z\,s\,n)\,\gamma')
-%% \end{alignat*}
-%% Unfolding $\ev$, we can further compute this to the following:
-%% \begin{alignat*}{4}
-%%   &(\msf{NatElim}\,P\,z\,s\,n)\rel\,\gamma\rel : \\
-%%   &\hspace{2em}P\rel\,(\gamma\rel,\,n\rel\,\gamma\rel)\,
-%%   (\msf{NatElim}\,(\lambda\,n.\,\ev\,P\,(\gamma,\,n))\,
-%%                   (\ev\,z\,\gamma)\,
-%%                   (\lambda\,n\,\mit{pn}.\,\ev\,s\,((\gamma,\,n),\,\mit{pn}))
-%%                   (\ev\,n\,\gamma))\\
-%%   &\hspace{8em}(\re\,(\msf{NatElim}\,P\,z\,s\,n)\,\gamma')
-%% \end{alignat*}
-%% In short, we need to show that $\msf{NatElim}$ preserves relations. Here we
-%% switch to the external view temporarily. By Definition \ref{def:restriction}, we know that
-%% \[
-%%    |\re\,(\msf{NatElim}\,P\,z\,s\,n)|\,\gamma' = (\msf{NatElim}\,P\,z\,s\,n))[\gamma'].
-%% \]
-%% At the same time, we have $n\rel\,\gamma\rel : \re_{\mbb{N}}\,(\ev\,n\,\gamma) =
-%% \re\,n\,\gamma'$, hence we know externally that $|\ev\,n|\,\gamma = n[\gamma']$.
-%% In other words, $n[\gamma']$ is canonical and is obtained as the serialization
-%% of $\ev\,n\,\gamma$. Therefore, $(\msf{NatElim}\,P\,z\,s\,n)[\gamma']$ is
-%% definitionally equal to $|\ev\,n|\,\gamma$-many applications of $s$ to
-%% $z$, and we can use $|\ev\,n|\,\gamma$-many applications of $s\rel$ to $z\rel$
-%% to witness the goal type. The $\beta$-rules for $\msf{NatElim}$ are also respected
-%% by this definition.
-
-%% \subsubsection{Meta-Level $\Sigma$-Types} We define relatedness pointwise. Pairing and projection
-%% are interpreted as meta-level pairing and projection.
-%% \begin{alignat*}{3}
-%%   &(\Sigma\,A\,B)\rel\,\gamma\rel : ((\alpha : \ev\,A\,\gamma) \times \ev\,B\,(\gamma,\,\alpha))
-%%              \to \re\,(\Sigma\,A\,B)\,\gamma' \to \whset_j\\
-%%   &(\Sigma\,A\,B)\rel\,\gamma\rel\,(\alpha,\,\beta)\,(\re_{\Sigma}^{-1}(\alpha',\,\beta')) :=
-%%              (\alpha\rel : A\rel\,\gamma\rel\,\alpha\,\alpha') \times B\rel\,(\gamma\rel,\,\alpha\rel)\,\beta\,\beta'
-%% \end{alignat*}
-
-%% \subsubsection{Meta-Level $\Pi$-Types} We again use a pointwise definition. Note that we need to use
-%% $\re_{\app}$ to apply $t' : \re\,(\Pi\,A\,B)\,\gamma'$ to $\alpha'$.
-%% \begin{alignat*}{3}
-%%   &(\Pi\,A\,B)\rel\,\gamma\rel : ((\alpha : \ev\,A\,\gamma) \to \ev\,B\,(\gamma,\,\alpha))
-%%              \to \re\,(\Pi\,A\,B)\,\gamma' \to \whset_j\\
-%%   &(\Pi\,A\,B)\rel\,\gamma\rel\,t\,t' :=
-%%     (\alpha : \ev\,A\,\gamma)(\alpha' : \re\,A\,\gamma')(\alpha\rel : A\rel\,\gamma\rel\,\alpha\,\alpha') \to
-%%     B\rel\,(\gamma\rel,\,\alpha\rel)\,(t\,\alpha)\,(\re_{\app}\,t'\,\alpha')
-%% \end{alignat*}
-%% For abstraction and application, we use a curry-uncurry definition:
-%% \begin{alignat*}{4}
-%%   &(\lam\,t)\rel\,\gamma\rel                &&:= \lambda\,\alpha\,\alpha'\,\alpha\rel.\,t\rel\,(\gamma\rel,\,\alpha\rel)\\
-%%   &(\app\,t)\rel\,(\gamma\rel,\,\alpha\rel) &&:= t\rel\,\gamma\rel\,\alpha\,\alpha'\,\alpha\rel
-%% \end{alignat*}
-
-%% \subsubsection{Meta-Level Universes} We interpret $\U_{1,j}$ as a semantic relation space:
-%% \begin{alignat*}{3}
-%%   &(\U_{1,j})\rel\,\gamma\rel : \whset_j \to \re\,\U_{1,j}\,\gamma' \to \whset_{j+1} \\
-%%   &(\U_{1,j})\rel\,\gamma\rel\,t\,t' := t \to \re_{\El}\,t' \to \whset_{j}
-%% \end{alignat*}
-%% Note that we have
-%% \begin{alignat*}{3}
-%%   & (\El\,t)\rel   &&: (\gamma\rel : \Gamma\rel\,\gamma\,\gamma') \to \ev\,t\,\gamma \to \re\,(\El\,t)\,\gamma' \to \whset_j\\
-%%   & (\Code\,t)\rel &&: (\gamma\rel : \Gamma\rel\,\gamma\,\gamma') \to \ev\,t\,\gamma \to \re\,(\El\,t)\,\gamma' \to \whset_j.
-%% \end{alignat*}
-%% The types coincide because of the equation $\re\,(\El\,t)\,\gamma' =
-%% \re_{\El}\,(\re\,t\,\gamma')$. Therefore we can interpret $\El$ and $\Code$ as
-%% identity maps, as $(\El\,t)\rel := t\rel$ and $(\Code\,t)\rel := t\rel$.
-
-%% \subsubsection{Object-Level Family Structure} We interpret extended contexts as follows.
-%% \begin{alignat*}{3}
-%%   & (\Gamma \ext_0 A)\rel\,(\gamma,\,\alpha)\,(\rextizero\,(\gamma',\,\alpha')) := (\gamma\rel : \Gamma\rel\,\gamma\,\gamma') \times (\emb{\alpha} = \alpha')
-%% \end{alignat*}
-%% Note that $\alpha : \Tm_{\mbbo j}\,(\ev\,A\,\gamma)$, so $\emb{\alpha} :
-%% \Tm_{0,j}\,\emb{\ev\,A\,\gamma}$, but since $A\rel\,\gamma\rel :
-%% \emb{\ev\,A\,\gamma} = \re\,A\,\gamma'$, we also have $\emb{\alpha} :
-%% \Tm_{0,j}\,(\re\,A\,\gamma')$. Thus, the equation $\emb{\alpha} = \alpha'$ is
-%% well-typed. For type substitution, we need
-%% \[
-%%   (A[\sigma])\rel\,\gamma\rel : \emb{\ev\,(A[\sigma])\,\gamma} = \re\,(A[\sigma])\,\gamma'.
-%% \]
-%% The goal type computes to $\emb{\ev\,A\,(\ev\,\sigma\,\gamma)} =
-%% \re\,A\,(\re\,\sigma\,\gamma')$. This is obtained directly from
-%% $A\rel\,(\sigma\rel\,\gamma\rel)$.  Similarly,
-%% \begin{alignat*}{6}
-%%   & (t[\sigma])\rel\,\gamma\rel &&:= t\rel\,(\sigma\rel\,\gamma\rel) && \p\rel\,(\gamma\rel,\,\alpha\rel) &&:= \gamma\rel \\
-%%   & (\sigma,\,t)\rel\,\gamma\rel &&:= (\sigma\rel\,\gamma\rel,\,t\rel\,\gamma\rel)\hspace{2em}&& \q\rel\,(\gamma\rel,\,\alpha\rel) &&:= \alpha\rel
-%% \end{alignat*}
-
-%% \subsubsection{Lifting Structure}
-%% \begin{alignat*}{3}
-%%   &(\Lift\,A)\rel : \Gamma\rel\,\gamma\,\gamma' \to \Tm_{\mbbo j}\,(\ev\,A\,\gamma) \to \re\,(\Lift\,A)\,\gamma' \to \whset_j\\
-%%   &(\Lift\,A)\rel\,\gamma\rel\,t\,t' := (\emb{t} = \spl t')
-%% \end{alignat*}
-%% This is well-typed by $A\rel\,\gamma\rel : \emb{\ev\,A\,\gamma} = \re\,A\,\gamma'$, which implies
-%% that $\emb{t} : \Tm_{\mbbo j}\,(\re\,A\,\gamma')$. For $\qut{t}$, we need
-%% \[  \qut{t}\rel\,\gamma\rel : \emb{\ev\,\qut{t}\,\gamma} = \spl(\re\,\qut{t}\,\gamma'). \]
-%% The goal type can be further computed to $\emb{\ev\,t\,\gamma} = \re\,t\,\gamma'$, which we prove
-%% by $t\rel\,\gamma\rel$. For splicing, we need
-%% \[
-%%   (\spl t)\rel\,\gamma\rel : \emb{\ev\,\spl t\,\gamma} = \re\,\spl t\,\gamma'
-%% \]
-%% where the goal type computes to $\emb{F\,t\,\gamma} = \spl(\re\,t\,\gamma')$,
-%% but this again follows directly from $t\rel\,\gamma\rel$.
-
-%% \subsubsection{Object-Level Type Formers}
-%% Lastly, object-level type formers are straightforward. For types, we need
-%% $\emb{\ev\,A\,\gamma} = \re\,A\,\gamma'$, and likewise for terms. Note that
-%% $\ev$ and $\emb{\blank}$ preserve all structure, and $\re$ preserves all
-%% structure on object-level types and terms. Hence, each object-level $A$ and $t$
-%% case in $\blank\rel$ trivially follows from induction hypotheses.
-
-%% This concludes the definition of the $\blank\rel$ interpretation.
-
-%% \subsection{Soundness}
-
-%% \begin{definition} First, we introduce shorthands for
-%% external operations that can be obtained from $\blank\rel$.
-%% \begin{alignat*}{5}
-%%   & \text{For}\hspace{1em} \Gamma &&: \Con \hspace{1em}&&\text{we get}\hspace{1em}
-%%          |\Gamma\rel| &&: \{\Delta : \Cono\} \to |\ev\,\Gamma|\,\Delta \to \Sub\,\emb{\Delta}\,\Gamma \to \Set_j\\
-%%   & \text{For}\hspace{1em} A &&: \Ty_{0,j}\,\Gamma \hspace{1em}&&\text{we get}\hspace{1em} |A\rel| &&: |\Gamma\rel|\,\gamma\,\gamma' \to \emb{\ev\,A\,\gamma} = A[\gamma']\\
-%%   & \text{For}\hspace{1em} t &&: \Tm_{0,j}\,\Gamma\,A \hspace{1em}&&\text{we get}\hspace{1em}|t\rel| &&: |\Gamma\rel|\,\gamma\,\gamma' \to \emb{\ev\,t\,\gamma} = t[\gamma']
-%% \end{alignat*}
-%% Since $\blank\rel$ was defined in $\hato$, we also know that the above are all
-%% stable under object-theoretic substitution.
-%% \end{definition}
-
-%% \begin{theorem}[Soundness for generic contexts]
-%% For each $\Gamma : \Cono$, we have ${\Gamma^P}\rel : |\Gamma\rel|\,\Gamma^P\,\id$.
-%% \end{theorem}
-%% \begin{proof}
-%% We define ${\blank^P}\rel$ by induction on object-theoretic contexts.
-%% ${\emptycon^P}\rel : \top$ is defined trivially as $\tt$. For ${(\Gamma \ext
-%%   A)^P}\rel$, we need $(\gamma\rel : |\Gamma|\rel\,(\Gamma^P[\p])\,\p) \times
-%% (\emb{\q} = \q)$. We get ${\Gamma^P}\rel : |\Gamma\rel|\,\Gamma^P\,\id$. Because
-%% of the naturality of $|\Gamma\rel|$, this can be weakened to ${\Gamma^P}\rel[\p]
-%% : |\Gamma\rel|\,(\Gamma^P[\p])\,\p$. Also, $\emb{\q} = \q$ holds immediately.
-%% \end{proof}
-
-%% \begin{theorem}[Soundness of staging]
-%% The open staging algorithm from Definition \ref{def:open-staging} is sound.
-%% \end{theorem}
-%% \begin{proof}
-%% \mbox{}
-%% \begin{itemize}
-%% \item For $A : \Ty_{0,j}\,\emb{\Gamma}$, we have that $\emb{\Stage\,A} =
-%% \emb{|\ev\,A|\,\Gamma^P}$ by the definition of $\Stage$, and moreover we have
-%% $|A\rel|\,{\Gamma^P}\rel : \emb{|\ev\,A|\,\Gamma^P} = A[\id]$, hence $\emb{\Stage\,A} = A$.
-
-%% \item For $t : \Tm_{0,j}\,\emb{\Gamma}\,A$, using $|t\rel|\,{\Gamma^P}\rel : \emb{|\ev\,t|\,\Gamma^P} = t[\id]$, we likewise
-%%   have $\emb{\Stage\,t} = \emb{|\ev\,t|\,\Gamma^P} = t[\id] = t$.
-%% \end{itemize}
-%% \end{proof}
-
-%% \begin{corollary}[Conservativity of 2LTT]\label{conservativity}
-%% From the soundness and stability of staging, we get that $\emb{\blank}$ is
-%% bijective on types and terms, hence $\Ty_{\mbbo j}\,\Gamma \simeq \Ty_{0,j}\,\emb{\Gamma}$ and
-%% $\Tm_{\mbbo j}\,\Gamma\,A \simeq \Tm_{0,j}\,\emb{\Gamma}\,\emb{A}$.
-%% \end{corollary}
-
-%% \paragraph{Alternative presentations}
-%% The above conservativity proof could be rephrased as an instance of more modern
-%% techniques which let us implicitly handle stability under 2LTT substitutions as
-%% well, not just $\mbbo$ substitutions. This amounts to working in a \emph{modal}
-%% internal language, where modalities control the interaction of the different base
-%% categories.
-%% \begin{itemize}
-%% \item Using \emph{synthetic Tait computability} \cite{sterlingthesis}, we work
-%%       in the internal language of the glued category $\msf{Id}_{\hato}\downarrow
-%%       \re$.
-%% \item Using \emph{relative induction} \cite{bocquet2021relative}, we work in the modal type theory
-%%       obtained from the dependent right adjoint functor $\re^* : \wh{\msf{2LTT}} \to \hato$,
-%%       where $\wh{\msf{2LTT}}$ denotes presheaves over the syntactic category of 2LTT.
-%% \end{itemize}
-%% We do not use either of these in this paper, for the following reasons. First,
-%% the author of the paper is not sufficiently familiar with the above
-%% techniques. Second, the task at hand is not too technically difficult, so using
-%% advanced techniques is not essential. Contrast e.g.\ normalization for cubical
-%% type theories, which is not feasible to handle without the more synthetic
-%% presentations \cite{cubicalnbe}.
-
-%% \section{Intensional Analysis}\label{sec:intensional-analysis}
-
-%% We briefly discuss intensional analysis in this section. This means analyzing
-%% the internal structure of object-level terms, i.e.\ values of $\Lift\,A$. Disallowing
-%% intensional analysis is a major simplification, which is sometimes enforced
-%% in implementations, for example in MetaML \cite{metaml}.
-
-%% If we have sufficiently expressive inductive types in the meta-level language,
-%% it is possible to simply use inductive deeply embedded syntaxes, which can be
-%% analyzed; our example in Section \ref{sec:staged-programming} for embedding a
-%% simply typed lambda calculus is like this. However, native metaprogramming
-%% features could be more concise and convenient than deep embeddings, similarly to
-%% how staged programming is more convenient than code generation with deeply
-%% embedded syntaxes.
-
-%% In the following, we look at the semantics of intensional analysis in the
-%% standard presheaf models. We only discuss basic semantics here, and leave
-%% practical considerations to future work.
-
-%% \begin{definition}
-%% The \textbf{Yoneda embedding} is a functor from $\mbbo$ to $\hato$, sending
-%% $\Gamma : \Cono$ to $\msf{y}\Gamma : \hCon$, such that $|\msf{y}\Gamma|\,\Delta
-%% = \Subo\,\Delta\,\Gamma$. We say that $\Gamma : \hCon$ is \emph{representable}
-%% if it is isomorphic to a Yoneda-embedded context.
-%% \end{definition}
-
-%% \begin{lemma}[Yoneda lemma]
-%% We have $\hSub\,(\msf{y}\Gamma)\,\Delta \simeq |\Delta|\,\Gamma$
-%% \cite[Section~III.2]{maclane98categories}. Also, we have
-%% $\hTm_{1,j}\,(\msf{y}\Gamma)\,A \simeq |A|\,\id$, where $\id :
-%% \Subo\,\Gamma\,\Gamma$.
-%% \end{lemma}
-
-%% The Yoneda lemma restricts possible dependencies on representable contexts.  For
-%% example, consider the staging behavior of $t : \Tm_{1}\,(\emptycon \ext (x :
-%% \Bool_0))\,\Bool_1$. Staging $t$ yields essentially a natural transformation
-%% $\hSub\,(\ev(\emptycon \ext \Bool_0))\,\mbb{B}$, where $\mbb{B}$ is the
-%% metatheoretic type with two elements, considered as a constant presheaf.  Now,
-%% if $\ev(\emptycon\,\ext\, \Bool_0)$ is representable, then
-%% $\hSub\,(\ev(\emptycon \ext \Bool_0))\,\mbb{B} \simeq \mbb{B}$, so $t$ can be
-%% staged in at most two different ways.
-
-%% Which contexts are representable? In the specific 2LTT in this paper,
-%% $\emb{\Gamma}$ is always representable in the presheaf
-%% model. $|\ev{\emb{\Gamma}}|\,\Delta$ contains lists of object-theoretic terms,
-%% hence $\ev{\emb{\Gamma}} \simeq \yon \Gamma$.  This implies that intensional
-%% analysis is \emph{not compatible} with the standard presheaf model, for our
-%% 2LTT.
-
-%% Consider a very basic feature of intensional analysis, \emph{decidability of
-%% definitional equality} for object-level Boolean expressions.
-%% \[
-%%   \msf{decEq} : (x\,y : \Lift\,\Bool_0) \to (x =_1 y) +_1 (x \neq_1 y)
-%% \]
-%% The Yoneda lemma says that $\msf{decEq}$ cannot actually decide
-%% definitional equality. We expect that $\msf{decEq}$ lets us define non-constant
-%% maps from $\Lift\,\Bool_0$ to $\Bool_1$, but the Yoneda lemma implies that we
-%% only have two terms with distinct semantics in $\Tm_1\,(\emptycon \ext_0
-%% \Bool_0)\,\Bool_1$.
-
-%% A more direct way to show infeasibility of $\msf{decEq}$ in $\hato$ is to note
-%% that definitional \emph{inequality} for object-level terms is not stable under
-%% substitution, since inequal variables can be mapped to equal terms.
-
-%% \paragraph{Stability under weakening only}
-%% We can move to different 2LTT setups, where stability under substitution is not
-%% required. We may have \emph{weakenings} only as morphisms in $\mbbo$.  A
-%% weakening from $\Gamma$ to $\Delta$ describes how $\Gamma$ can be obtained by
-%% inserting zero or more entries to $\Delta$. A wide range of intensional analysis
-%% features are stable under weakenings. For example, $\msf{decEq}$ now holds in
-%% $\hato$ whenever object-theoretic definitional equality is in fact decidable.
-%% We also dodge the Yoneda lemma, since $\ev\,\emb{\Gamma}$ is not necessarily
-%% representable anymore: $|\ev\,\emb{\Gamma}|\,\Delta$ is still a set of lists of
-%% terms, but weakenings from $\Delta$ to $\Gamma$ are not lists of terms.
-
-%% As a trade-off, if there is no notion of substitution in the specification of the
-%% object theory, it is not possible to specify dependent or polymorphic types
-%% there. We need a substitution operation to describe dependent elimination or
-%% polymorphic instantiation.
-
-%% However, stability under weakening is still sufficient for many practical use
-%% cases, for example the monomorphization setup in Section
-%% \ref{sec:monomorphization}. Here, we only have weakening as definitional
-%% equality in the object theory, and we do not have $\beta\eta$-rules.
-
-%% \paragraph{Closed modalities}\label{sec:closed-modalities}
-%% Another option is to add a \emph{closed} or \emph{crisp} modality for modeling
-%% closed object-level terms \cite{licata2018internal}. Since closed terms are not
-%% affected by substitution, we are able to analyze them, and we can also recover
-%% open terms by explicitly abstracting over contexts
-%% \cite{DBLP:conf/lics/Hofmann99,DBLP:journals/corr/abs-2206-02831}. \emph{Function
-%% pointers} in the style of the C programming language could be also an
-%% interesting use case, since these must become closed after staging (as they
-%% cannot capture closure environments).
-
-%% %% for tracking object-level values which will become
-%% %% closed terms after staging. Since closed terms are not affected by substitution,
-%% %% we are free to analyze them. Such analysis can be viewed as an internalization
-%% %% of canonicity for closed object-level terms. For example, a closed $\Bool_0$
-%% %% term will be either $\true_0$ or $\false_0$ after staging, so it makes sense to
-%% %% be able to eliminate from it to $\Bool_1$. This can be also viewed as a safe
-%% %% implementation of a \emph{run} function in the style of MetaML \cite{metaml}.
-%% %% In our case \emph{scope extrusion} would be prevented by the closed modality.
-
-%% \section{Related Work}\label{sec:related-work}
-
-%% The work Annekov et al.\ on two-level type theory \cite{twolevel}, building on
-%% \cite{capriotti2017models} and \cite{hts}, is the foremost inspiration in
-%% the current work. Interestingly, the above works do mention metaprogramming as a
-%% source of intuition for 2LTT, but they only briefly touch on this aspect, and in
-%% \cite{twolevel} the main focus is on extensions of basic 2LTT which do not have
-%% staging semantics anymore. Ibid.\ conjectures the strong conservativity of 2LTT
-%% in Proposition 2.18, and suggests categorical gluing as proof strategy. In this
-%% work, we do essentially that: our presheaf model and the logical relation model
-%% could be merged into a single instance of gluing along the $\re$ morphism; this
-%% merges staging and its soundness proof. We split this construction to two parts
-%% in order to get a self-contained presentation of the staging algorithm.
-
-%% \cite{DBLP:conf/lics/Hofmann99} models higher-order abstract syntax of simply
-%% typed object languages in presheaves over various syntactic categories. The main
-%% difference between HOAS and 2LTT is that the former represents object-level binders
-%% using meta-level functions while the latter has a primitive notion of object-level
-%% binding.
-
-%% The multi-stage calculus in \cite{multi-stage-calculus} supports staging
-%% operations and type dependencies. However, there are major differences to the
-%% current work. First, ibid.\ does not have large elimination, hence it does not
-%% support computing types by staging. Second, it does not support staged
-%% compilation in the sense of Definition \ref{def:staging}; there is no judgment
-%% which ensures that staging operations can be eliminated and a purely
-%% ``object-level'' program can be extracted from a multi-stage program. Rather, the system
-%% supports \emph{runtime} code generation. There are several modal multi-stage
-%% systems which support runtime code generation but not staged compilation, such
-%% as those based on S4 modalities (e.g.\ \cite{DBLP:journals/jacm/DaviesP01}) or
-%% contextual modalities (\cite{DBLP:journals/pacmpl/JangGMP22}).
-
-%% %% We make a
-%% %% disctinction between staged compilation and staged computation. For the former,
-%% %% we expect staging in the sense of Definition \ref{def:staging}; this is
-%% %% essential if we want to use staging as a lowering step in a compilation
-%% %% pipeline. Staged computation on the other hand may support runtime code
-%% %% construction and evaluation, but may not guarantee that these features can be
-%% %% entirely computed away.
-
-%% Idris 1 supports compile-time partial evaluation, through \texttt{static}
-%% annotations on function arguments \cite{scrap-your-inefficient-engine}. This can
-%% be used for dependently typed code generation, but Idris 1 does not guarantee
-%% that partial evaluation adequately progresses. For example, we can pass a neutral
-%% value as static argument in Idris 1, while in 2LTT static function arguments
-%% are always canonical during staging.
-
-%% Our notation for quoting and splicing is borrowed from MetaML \cite{metaml}.  In
-%% the following, we compare 2LTT to MetaML, MetaOCaml \cite{kiselyov14metaocaml}
-%% and typed Template Haskell \cite{typed-th}.
-
-%% 2LTT explicitly tracks stages of types, in contrast to the mentioned
-%% systems. There, we have type lifting (usually called $\msf{Code}$), quoting and
-%% splicing, but lifting does not change the universe of types. We can write a
-%% function with type $\Bool \to \msf{Code}\,\Bool$, and the two $\Bool$
-%% occurrences refer to the exact same type.
-
-%% This causes some operational confusion down the line, and additional
-%% disambiguation is needed for stages of binders. For example, in typed Template
-%% Haskell, top-level binders behave as runtime binders when used under a quote,
-%% but behave as static binders otherwise. Given a top-level definition $b =
-%% \True$, if we write a top-level definition $\msf{expr} = [||\,b\,||]$, then $b$
-%% is a \emph{variable} in the staging output, but if we write $\msf{expr} =
-%% \msf{if}\,b\,\msf{then}\,[||\,\True\,||]\,\msf{else}\,[||\,\False\,||]$, then
-%% $b$ is considered to be a static value. In contrast, 2LTT does not distinguish
-%% top-level and local binders, and in fact it has no syntactic or scope-based
-%% restrictions; everything is enforced by typing.
-
-%% 2LTT supports staging for types and dependent types, while to our knowledge no
-%% previous system for staged compilation does the same. It appears to the author
-%% that the main missing component in previous systems is \emph{dependent types at
-%% the meta level}, rather than staging operations; staging operations in 2LTT and
-%% MetaOCaml are already quite similar. For example, any interesting operation on
-%% the static-length vectors in Section \ref{sec:staged-programming} requires
-%% meta-level dependent types, even if the object language is simply typed.
-
-%% In MetaOCaml, there is a \emph{run} function for evaluating closed object-level
-%% terms, however, this can fail during staging since closedness is not statically
-%% ensured. In our setting, this could be reproduced in a safe way using a closed
-%% modality; this is part of future work.
-
-%% %% as we mentioned in Section \ref{sec:closed-modalities}, we
-%% %% believe that the proper way to implement it in 2LTT would be to use a closed
-%% %% modality. This use case is closely related to \cite{licata2018internal}. In
-%% %% ibid.\ a semantics is sketched for this modality in a presheaf topos, and we
-%% %% could directly adapt it to our presheaf model of 2LTT. That said, working
-%% %% out the practical details of this modality remains future work.
-
-%% Additionally, 2LTT only supports two stages, while the other noted systems allow
-%% countable stages. It remains future work to develop NLTT (N-level type theory),
-%% but this extension does appear to be straightforward. In the specification of
-%% NLTT, we would simply need to move from $i \in \{0,1\}$ to $i \in \mbb{N}$,
-%% although in the semantics there could be more subtleties.
-
-%% Many existing staged systems also support \emph{side effects} at compile time,
-%% while our version of 2LTT does not. Here, general considerations about the
-%% mixing of dependent types and effects should apply; see
-%% e.g.\ \cite{fire-triangle}. However, it should be possible to use standard
-%% effect systems such as monads at the meta level.
-
-%% \section{Conclusions and Future Work}\label{sec:conclusions}
-
-%% We developed the foundational metatheory of 2LTT as a system of
-%% two-stage compilation. We view the current results as more like a starting point
-%% to more thorough investigation into applications and extensions; in this paper
-%% we only sketched these.
-
-%% We emphasize that variants of 2LTT can serve as languages where staging
-%% features can be practically implemented, and also as \emph{metatheories}, not
-%% only for the formalization for staging techniques, but also more generally for
-%% reasoning about constructions in object languages. This purpose is prominent in
-%% the original 2LTT use case in synthetic homotopy theory. The meta-level language
-%% in a 2LTT can be expressive enough to express general mathematics, and the
-%% meta-level propositional equality $\blank\!=_1\!\blank$ can be used to prove
-%% $\beta\eta$-equality of object-level types and terms.
-
-%% For example, we could use object theories as \emph{shallowly embedded} target
-%% languages of optimization techniques and program translations. In particular, we
-%% may choose a low-level object theory without higher-order functions, and
-%% formalize closure conversion as an interpretation of an embedded meta-level
-%% syntax into the object theory.
-
-%% There is much future work in adapting existing staging techniques to 2LTT
-%% (which, as we mentioned, can double as formalization of said techniques), or
-%% adding staging where it has not been available previously.
-%% \begin{itemize}
-%% \item \emph{Let-insertion techniques}. These allow more precise control over
-%%   positions of insertion, or smarter selection of such positions. In 2LTT,
-%%   meta-level continuation monads could be employed to adapt some of the features
-%%   in \cite{DBLP:journals/jfp/KameyamaKS11}. The automatic let-floating feature
-%%   of MetaOCaml \cite{DBLP:journals/corr/abs-2201-00495} requires strengthening
-%%   of terms, which is not stable under substitution, but it is stable under
-%%   weakening, so perhaps it could be implemented as an ``intensional analysis''
-%%   feature.
-%% \item \emph{Generic programming with dependent types}. There is substantial
-%%   literature in this area
-%%   (e.g.\ \cite{loh11generic,chapman2010gentle,diehl-thesis,ornaments}), but only
-%%   in non-staged settings.  2LTT should immediately yield staging for these
-%%   techniques, assuming that the meta level has sufficient type formers
-%%   (e.g.\ induction-recursion). We could also try to adapt previous work on
-%%   generic treatment of partially static data types
-%%   \cite{DBLP:journals/pacmpl/YallopGK18}; fully internalizing this in 2LTT would
-%%   also require dependent types.
-%% \end{itemize}
-
-%% There are also ways that 2LTT itself could be enhanced.
-%% \begin{itemize}
-%% \item \emph{More stages, stage polymorphism}. Currently, there is substantial
-%%   code duplication if the object-level and meta-level languages are similar.
-%%   Stage polymorphism could help reduce this. A simple setup could include three
-%%   stages, the runtime one, the static one, and another one which supports
-%%   polymorphism over the first two stages.
-%% \item \emph{Modalities}. We mentioned the closed modality in Section
-%%   \ref{sec:closed-modalities}.  More generally, many \emph{multimodal}
-%%   \cite{gratzer20multimodal} type theories could plausibly support staging. 2LTT
-%%   itself can be viewed as a very simple multimodal theory with $\Lift$ as the
-%%   only modality, which is also degenerate (because it entails no structural
-%%   restriction on contexts). We could support multiple object theories, with
-%%   modalities for lifting them to the meta level or for representing translations
-%%   between them. We could also have modalities for switching between stability
-%%   under substitution and stability under weakening.
-%% \item \emph{Intensional analysis}. We only touched on the most basic semantics
-%%   of intensional analysis in Section \ref{sec:intensional-analysis}. It remains
-%%   a substantial challenge to work out the practicalities. For good ergonomics,
-%%   we would need something like a ``pattern matching'' operation on object-level
-%%   terms, or some induction principle which is more flexible than the plain
-%%   assumption of decidable object-level equality.
-%% \end{itemize}
-
-%% \section{Data Availability Statement}
-
-%% A standalone prototype implementation is available as \cite{staged-demo}. It
-%% implements elaboration and staging for a 2LTT. It includes a tutorial and code
-%% examples that expand on the contents of Section \ref{sec:tour-of-2ltt}.
-
 \interlinepenalty=10000

### 2024-02-06 newpaper/hs-agda/hs/Lib.hs

@@ -1,0 +1,484 @@
+
+{-# language LambdaCase, TemplateHaskell, BlockArguments, RankNTypes,
+   MultiParamTypeClasses, FunctionalDependencies, TypeApplications,
+   ScopedTypeVariables, UndecidableInstances, TypeFamilies,
+   TypeFamilyDependencies, PartialTypeSignatures, AllowAmbiguousTypes,
+   DataKinds, PolyKinds, MagicHash #-}
+
+module Lib where
+
+import Control.Applicative
+import Control.Monad
+import Control.Monad.Except
+import Control.Monad.Identity
+import Control.Monad.Reader
+import Control.Monad.State.Strict
+import Control.Monad.Trans
+import Control.Monad.Trans.Maybe
+import Data.Coerce
+import Data.Kind
+import GHC.Exts
+import Language.Haskell.TH hiding (ListT, ListE, Type)
+
+import Prelude
+import qualified Prelude as P
+
+import Up
+
+--------------------------------------------------------------------------------
+
+class Val a
+instance Val Int
+instance Val Bool
+instance Val Bot
+instance Val ()
+instance Val a => Val (Identity a)
+instance (Val a, Val b) => Val (Pair a b)
+instance (Val a, Val b) => Val (a, b)
+instance Val a => Val (Maybe a)
+instance (Val a, Val b) => Val (Either a b)
+instance Val a => Val [a]
+
+data Pair a b = Pair !a !b
+
+proj1 (Pair a b) = a
+proj2 (Pair a b) = b
+
+class Improve0 a b | a -> b, b -> a where
+  up0 :: Up a -> b
+  down0 :: b -> Up a
+
+instance Improve0 (Pair a b) (Pair (Up a) (Up b)) where
+  up0 x = Pair [|| proj1 $$x ||] [|| proj2 $$x ||]
+  down0  (Pair a b) = [|| Pair $$a $$b ||]
+
+data Bot
+
+instance Improve0 Bool (Gen Bool) where
+  up0 x = Gen \k -> [|| if $$x then $$(k True) else $$(k False) ||]
+  down0 x = unGen x \x -> if x then [||True||] else [||False||]
+
+instance Improve0 (Maybe a) (Gen (Maybe (Up a))) where
[…]

### 2024-02-06 newpaper/hs-agda/hs/LibTest.hs

@@ -1,0 +1,2 @@
+
+module LibTest where

### 2024-02-06 newpaper/hs-agda/hs/Up.hs

@@ -1,0 +1,59 @@
+
+{-# language TemplateHaskell, UndecidableInstances #-}
+
+module Up where
+
+import Prelude (Num(..))
+import qualified Prelude as P
+import Data.Bits
+
+import Language.Haskell.TH
+import Language.Haskell.TH.Syntax
+
+type Up = CodeQ
+
+instance Num a => Num (Up a) where
+  (+) x y       = [|| $$x + $$y ||]
+  (*) x y       = [|| $$x * $$y ||]
+  abs x         = [|| abs $$x ||]
+  signum x      = [|| signum $$x ||]
+  fromInteger x = [|| fromInteger $$(liftTyped x) ||]
+  negate x      = [|| negate $$x ||]
+
+class Eq a where
+  (==) :: Up a -> Up a -> Up P.Bool
+  (/=) :: Up a -> Up a -> Up P.Bool
+  (/=) x y = [|| P.not $$(x == y) ||]
+
+instance P.Eq a  => Eq a where
+  (==) x y = [|| $$x P.== $$y ||]
+
+not :: Up P.Bool -> Up P.Bool
+not x = [|| P.not $$x ||]
+
+tt :: Up ()
+tt = [|| () ||]
+
+down1 :: (Up a -> Up b) -> Up (a -> b)
+down1 f = [|| \a -> $$(f [||a||]) ||]
+
+up1 :: Up (a -> b) -> Up a -> Up b
+up1 f a = [|| $$f $$a ||]
+
+class Ord a where
+  (<)  :: Up a -> Up a -> Up P.Bool
+  (<=) :: Up a -> Up a -> Up P.Bool
+  (>)  :: Up a -> Up a -> Up P.Bool
+  (>=) :: Up a -> Up a -> Up P.Bool
+
+instance P.Ord a => Ord a where
+  (<)  x y = [|| (P.<)  $$x $$y ||]
+  (<=) x y = [|| (P.<=)  $$x $$y ||]
+  (>)  x y = [|| (P.>)  $$x $$y ||]
+  (>=) x y = [|| (P.>=)  $$x $$y ||]
+
+even :: Up P.Int -> Up P.Bool
+even n = [|| $$n .&. 1 P.== 0 ||]
+
+odd :: Up P.Int -> Up P.Bool
+odd n = [|| $$n .&. 1 P.== 1 ||]

### 2024-02-09 newpaper/paper.tex
stuff

@@
 \newcommand{\belse}{\mbf{else}\,}
-\newcommand{\return}{\msf{return}\,}
+\newcommand{\return}{\mbf{return}\,}
 \newcommand{\lam}{\lambda\,}
@@
 \newcommand{\letdef}{\mbf{let\,}}
+\newcommand{\map}{\mit{map}}
+\newcommand{\vas}{\mit{as}}
+\newcommand{\vbs}{\mit{bs}}
+\newcommand{\vxs}{\mit{xs}}
+\newcommand{\vys}{\mit{ys}}
+\newcommand{\vma}{\mit{ma}}
+\newcommand{\vmb}{\mit{mb}}
 
@@
 \newcommand{\Int}{\msf{Int}}
-\newcommand{\List}{\msf{List}}
+\newcommand{\List}{\msf{List}\,}
 \newcommand{\Nil}{\msf{Nil}}
@@
 \newcommand{\Reader}{\msf{Reader}}
+\newcommand{\ReaderT}{\msf{ReaderT}}
 \newcommand{\Monad}{\msf{Monad}}
+\newcommand{\Applicative}{\msf{Monad}}
 \newcommand{\class}{\msf{class}}
@@
 \newcommand{\Lift}{{\Uparrow}}
-\newcommand{\spl}{{\sim}}
-\newcommand{\ql}{{\langle}}
-\newcommand{\qr}{{\rangle}}
+\newcommand{\Up}{{\Uparrow}}
+\newcommand{\spl}{{\bs{\sim}}}
+\newcommand{\ql}{{\bs{\langle}}}
+\newcommand{\qr}{{\bs{\rangle}}}
 \newcommand{\bind}{\mathbin{>\!\!>\mkern-6.7mu=}}
@@
 \newcommand{\Suc}{\msf{Suc}}
+\newcommand{\Maybe}{\msf{Maybe}}
+\newcommand{\MaybeT}{\msf{MaybeT}}
+\newcommand{\Nothing}{\msf{Nothing}}
+\newcommand{\Just}{\msf{Just}}
 
-%% \newcommand{\spl}{{\sim}}
-%% \newcommand{\qut}[1]{\langle #1\rangle}
-
-%% \newcommand{\mbbc}{\mbb{C}}
-%% \newcommand{\mbbo}{\mbb{O}}
-%% \newcommand{\ob}{_\mbbo}
-
-%% \newcommand{\U}{\msf{U}}
-%% \newcommand{\Con}{\msf{Con}}
-%% \newcommand{\Sub}{\msf{Sub}}
-%% \newcommand{\Ty}{\msf{Ty}}
-%% \newcommand{\Tm}{\msf{Tm}}
-%% \newcommand{\Cono}{\msf{Con}_{\mbbo}}
-%% \newcommand{\Subo}{\msf{Sub}_{\mbbo}}
-%% \newcommand{\Tyo}{\msf{Ty}_{\mbbo}}
-%% \newcommand{\Tmo}{\msf{Tm}_{\mbbo}}
-%% \newcommand{\hCon}{\wh{\msf{Con}}}
-%% \newcommand{\hSub}{\wh{\msf{Sub}}}
-%% \newcommand{\hTy}{\wh{\msf{Ty}}}
-%% \newcommand{\hTm}{\wh{\msf{Tm}}}
-
-%% \newcommand{\p}{\mathsf{p}}
-%% \newcommand{\q}{\mathsf{q}}
-
-%% \newcommand{\refl}{\msf{refl}}
-%% \newcommand{\Bool}{\msf{Bool}}
-%% \newcommand{\true}{\msf{true}}
-%% \newcommand{\false}{\msf{false}}
-%% \newcommand{\True}{\msf{True}}
-%% \newcommand{\False}{\msf{False}}
-%% \newcommand{\List}{\msf{List}}
-%% \newcommand{\nil}{\msf{nil}}
-%% \newcommand{\cons}{\msf{cons}}
-
-%% \renewcommand{\tt}{\msf{tt}}
-%% \newcommand{\fst}{\msf{fst}}
-%% \newcommand{\snd}{\msf{snd}}
-%% \newcommand{\mylet}{\msf{let}}
-%% \newcommand{\emptycon}{\scaleobj{.75}\bullet}
-%% \newcommand{\id}{\msf{id}}
-
-%% \newcommand{\Set}{\mathsf{Set}}
-%% \newcommand{\Prop}{\mathsf{Prop}}
-%% \newcommand{\Rep}{\msf{Rep}}
-
-%% \newcommand{\emb}[1]{\ulcorner#1\urcorner}
-
-%% \newcommand{\Stage}{\msf{Stage}}
-%% \newcommand{\hato}{\bm\hat{\mbbo}}
-%% \newcommand{\ev}{\mbb{E}}
-%% \newcommand{\re}{\mbb{R}}
-
-%% \theoremstyle{remark}
-%% \newtheorem{notation}{Notation}
-
-%% \newcommand{\whset}{\wh{\Set}}
-%% \newcommand{\rexti}{\re_{\ext_1}^{-1}}
-%% \newcommand{\rextizero}{\re_{\ext_0}^{-1}}
-
-%% \newcommand{\rel}{^{\approx}}
-%% \newcommand{\yon}{\msf{y}}
+\theoremstyle{remark}
+\newtheorem{notation}{Notation}
+\newcommand{\id}{\mit{id}}
+\newcommand{\mup}{\mit{up}}
+\newcommand{\mdown}{\mit{down}}
+\newcommand{\tyclass}{\mbf{class}}
+\newcommand{\instance}{\mbf{instance}\,}
+\newcommand{\Improve}{\msf{Improve}}
+\newcommand{\Gen}{\msf{Gen}}
+\newcommand{\unGen}{\mit{unGen}}
+\renewcommand{\Vec}{\msf{Vec}}
+\newcommand{\gen}{\mit{gen}}
+\newcommand{\genRec}{\mit{genRec}}
+\newcommand{\fmap}{<\!\!\$\!\!>}
+\newcommand{\ap}{<\!\!*\!\!>}
+\newcommand{\runGen}{\mit{runGen}}
+\newcommand{\qt}[1]{\ql#1\qr}
+\newcommand{\lift}{\mit{lift}}
+\newcommand{\liftGen}{\mit{liftGen}}
+\newcommand{\MonadGen}{\msf{MonadGen}}
+\newcommand{\RA}{\Rightarrow}
+\newcommand{\EitherT}{\msf{EitherT}}
+\newcommand{\StateT}{\msf{StateT}}
+\newcommand{\Identity}{\msf{Identity}}
+\newcommand{\runIdentity}{\mit{runIdentity}}
+\newcommand{\newtype}{\mbf{newtype}\,}
+\newcommand{\runMaybeT}{\mit{runMaybeT}}
+\newcommand{\dlr}{\,\$\,}
+\newcommand{\ImproveF}{\msf{ImproveF}}
+\newcommand{\ExceptT}{\msf{ExceptT}}
 
@@
 general-purpose optimizers are slower to run, less transparent and less
-reliable. Also, metaprogramming allows \emph{library authors} to exploit domain
-specific optimizations, while it is not realistic for general-purpose optimizers
-to know about all domains.
+reliable. Also, metaprogramming allows library authors to exploit
+domain-specific optimizations, while it is not realistic for general-purpose
+optimizers to know about all domains.
 
@@
 relevant to the current paper, so we shall omit levels. Note that universe
-levels have nothing to do with staging in the sense of staged compilation; they
-are about sizing for the purpose of logical consistency.
+levels have nothing to do with staged compilation; they are about sizing for the
+purpose of logical consistency.
 
-We use Agda-style syntax and implicit arguments. A basic example:
+We use Agda-like syntax for functions and implicit arguments. A basic example:
 \begin{alignat*}{3}
-  &id : \{A : \MTy\} \to A \to A\\
-  &id = \lam x.\, x
+  &\id : \{A : \MTy\} \to A \to A\\
+  &\id = \lam x.\, x
 \end{alignat*}
@@
 Note that we added an $_\M$ subscript to the type; when analogous types can be
-defined both on the meta and object levels, we will use this subscript to
-distinguish the meta-level version.
+defined both on the meta and object levels, we will sometimes use this subscript
+to disambiguate the meta-level version. We will also use Haskell-like newtype
+notation, such as in $\newtype \msf{Wrap}\,A = \msf{Wrap}\,\{\mit{unWrap} : A
+\}$.
+
+Importantly, we can only produce meta-level values by induction on meta-level
+data. In general, all construction and elimination rules for type formers in
+$\MTy$ stay within $\MTy$.
 
@@
 $\bs{\Ty}$ is the universe of types in the object language. It is itself a
-metatype, so so we have $\Ty : \MTy$. It is further split to two sub-universes.
+metatype, so so we have $\Ty : \MTy$. Similarly as in the case of $\MTy$, all
+construction and elimination rules of the object language stay within $\Ty$.
+However, we further split $\Ty$ to two sub-universes.
 
 First, $\bs{\VTy} : \MTy$ is the universe of \emph{value types}. $\VTy$ supports
-parameterized algebraic data types, with two restrictions:
-\begin{itemize}
-\item ADTs can be only parameterized over types in $\VTy$.
-\item Constructor fields must be in $\VTy$.
-\end{itemize}
+parameterized algebraic data types, where parameters can have arbitrary types,
+but all constructor field types must be in $\VTy$.
+
 Since $\VTy$ is a sub-universe of $\Ty$, we have that when $A : \VTy$ then also
@@
 also a sub-universe of $\Ty$ with implicit coercions. For now, we only specify
-that $\CTy$ contains functions whose domain is a value type:
+that $\CTy$ contains functions whose domains are value types:
 \[ \blank\to\blank : \VTy \to \Ty \to \CTy \]
@@
 \to \Bool : \CTy$, hence also $\Bool \to \Bool : \Ty$. However, $(\Bool \to
-\Bool) \to \Bool$ is ill-formed, since the domain is not a value type. An
+\Bool) \to \Bool$ is ill-formed, since the domain is not a value type. Let us look at an
 example for an object-level program, where we already have natural numbers
-declared as $\data \Nat : \VTy := \Zero\,|\,\Suc\,\Nat$:
+declared as $\data \Nat := \Zero\,|\,\Suc\,\Nat$:
 \begin{alignat*}{3}
-  &\rlap{$\add : \Nat \to \Nat \to \Nat$}\\
-  &\add :=\,&& \mathrlap{\letrec \go := \lam n\,m.\,\case n\,\of}\\
-  &         && \ind \Zero   &&\to m;\\
-  &         && \ind \Suc\,n &&\to \Suc\,(\go\,n\,m);\\
-  &         && \go
+  &\hspace{-1em}\rlap{$\add : \Nat \to \Nat \to \Nat$}\\
+  &\hspace{-1em}\add :=\,&& \mathrlap{\letrec \go\,n\,m := \case n\,\of}\\
+  &\hspace{-1em}         && \ind \Zero   &&\to m;\\
+  &\hspace{-1em}         && \ind \Suc\,n &&\to \Suc\,(\go\,n\,m);\\
+  &\hspace{-1em}         && \go
 \end{alignat*}
-Every recursive definition must be introduced with $\letrec$. Note also the
-single semicolon after the $\mbf{case}$ branches. The general syntax is $\letrec x :=
-t; u$, optionally with a type annotation for $x$ on a separate line.
+Every recursive definition must be introduced with $\letrec$. The general syntax
+is $\letrec x : A := t; u$, where the $A$ type annotation can be
+omitted. $\mbf{letrec}$ can be only used to define computations, not values
+(hence, only functions can be recursive so far).
 
-Also note that object-level definitions use $:=$ instead of $=$ for meta-level
-ones. Generally, let-definitions are the biggest source of stage ambiguity in
-surface notation. Later on, we will use more implicit staging notation, but we
+Object-level definitions use $:=$ as notation, instead of the $=$ that is used
+for meta-level ones. Later on, we may use some implicit CFTT notations, but we
 will always disambiguate the stages of $\mbf{let}$-s in this way. Non-recursive
@@
   &f : \Nat \to \Nat\\
-  &f := \lam x.\,\letdef x := x + 10; \letdef x := x + 20; x * 10
+  &f\,x := \letdef x := x + 10; \letdef x := x + 20; x * 10
 \end{alignat*}
-$\mbf{let}$-definitions can be used to define values of any type, and the type
-of the $\mbf{let}$ body can be also anything. Additionally, the right hand sides
-of $\case$ branches can also have arbitrary types. So the following is
-well-formed:
+Values are call-by-value at runtime; they are computed eagerly in function
+applications and $\mbf{let}$-s. $\mbf{let}$-definitions can be used to define
+inhabitants of any type, and the type of the $\mbf{let}$ body can be also
+arbitrary. Additionally, the right hand sides of $\case$ branches can also have
+arbitrary types. So the following is well-formed:
 \begin{alignat*}{3}
-  &f : \mathrlap{\Bool \to \Nat \to \Nat}\\
-  &f \mathrlap{:= \lam b.\,\case b\,\of}\\
-  &\ind \True  &&\to \lam x.\, x + 10;\\
-  &\ind \False &&\to \lam x.\, x * 10;
+  &\hspace{-1em}f : \mathrlap{\Bool \to \Nat \to \Nat}\\
+  &\hspace{-1em}f\,b\,\mathrlap{:= \case b\,\of}\\
+  &\hspace{-1em}\ind \True  &&\to \lam x.\, x + 10;\\
+  &\hspace{-1em}\ind \False &&\to \lam x.\, x * 10;
 \end{alignat*}
+
+What about \textbf{definitional equality} at the object level? This is a
+distinct notion from the runtime semantics; object programs are embedded in
+CFTT, which is dependently typed, and we need to decide definitional equalities
+during type checking. The setup is simple: we have no $\beta$ or $\eta$ rules
+for object programs at all, nor any rule for $\mbf{let}$-unfolding. The reasons
+are the following. First, since the object language has general recursion, most
+$\beta$-rules are only valid in the operational semantics up to some
+restrictions, and $\beta$-conversion is not decidable to begin with. Second, in
+metaprogramming we care about the size and efficiency of generated code, and
+these properties are not stable under $\beta$ and $\eta$.  Hence, the sensible
+choice is to do metaprogramming up to strict syntactic equality of object
+programs.
+
 Let us discuss the object language. First, notice that there is no polymorphism
@@
 here we only give an intuitive explanation. In short, we choose a call-by-name
-semantics for functions, which means that the only way we can compute with a
-function is by applying it to all arguments and extracting the resulting value.
-Hence, the only way to compute with $f$ is to apply it to \emph{two} arguments,
-so $f$ is operationally equivalent to the following definition:
-\[ \lam b\,x.\,\case b\,\of \True \to x + 10; \False \to x * 10 \]
-In Section \ref{TODO} we show that all object programs can be transformed to a
-\emph{saturated} form, where the arity of every function call matches the number
-of topmost $\lambda$-binders in the function definition.
+runtime semantics for functions, which means that the only way we can compute
+with a function is by applying it to all arguments and extracting the resulting
+value.  Hence, the only way to compute with $f$ is to apply it to \emph{two}
+arguments, so $f$ is operationally equivalent to $\lam b\,x.\,\case b\,\of \True
+\to x + 10; \False \to x * 10$.  In Section \ref{TODO} we show that all object
+programs can be transformed to a \emph{saturated} form, where the arity of every
+function call matches the number of topmost $\lambda$-binders in the definition
+of the called function. Then, the local functions can be compiled either to
+top-level functions by lambda-lifting, or if they are only tail-called, left in
+place as join points \cite{TODO}.
 
@@
 more restricted, and thus easier to compile, and making \emph{metaprogramming}
-more convenient for the object language. We will see that the ability to insert
-$\mbf{let}$-s without restriction is very convenient in code generation, and
-likewise the ability to have arbitrary object expressions in $\mbf{case}$
-bodies. In this paper we choose to go with the most liberal object syntax, at
-the cost of needing more downstream processing. The call-by-name nature of
-computation types requires a bit of a change of thinking from programmers, but
-we believe that it is well worth to have it for the metaprogramming convenience.
+more convenient. We will see that the ability to insert $\mbf{let}$-s without
+restriction is very convenient in code generation, and likewise the ability to
+have arbitrary object expressions in $\mbf{case}$ bodies. In this paper we go
+with the most liberal object syntax, at the cost of needing more downstream
+processing. The call-by-name nature of computation types requires a bit of a
+change of thinking from programmers, but we believe that it is well worth to
+have for the metaprogramming convenience.
 
@@
 at value types. In CBPV, a let-definition for a function is only possible by
-first packing it up in a closure value, which clearly does not work for us.
-Also, CBPV makes a judgment-level structural distinction between values and
-computations, while we use type universes for that. Generally speaking,
-type-based restrictions are easier to work with in dependent type theories than
-structural restrictions.
+first packing it up as a closure value (or: ``thunk''), which clearly does not
+work for us.  Also, CBPV makes a judgment-level structural distinction between
+values and computations, while we use type universes for that. Generally
+speaking, type-based restrictions are easier to work with in dependent type
+theories than structural restrictions.
+
+\subsection{Staging}\label{sec:staging}
+
+With what we have seen so far, there is no interaction between the meta and
+object levels. We make such interaction possible with \emph{staging operations}.
+
+\begin{itemize}
+\item For $A : \Ty$, we have $\Lift A : \MTy$, pronounced as ``lift $A$''. This is
+      the type of metaprograms that produce $A$-typed object programs.
+\item For $A : \Ty$ and $t : A$, we have $\ql t \qr : \Lift A$, pronounced ``quote $t$''. This
+      is the metaprogram which immediately returns $t$.
+\item For $t : \Lift A$, we have $\spl t : A$, pronounced ``splice $t$''. This
+  inserts the result of a metaprogram into an object term. \emph{Notation:}
+  splicing binds stronger than function application, so $f\,\spl x$ is parsed as
+  $f\,(\spl x)$.
+\item We have $\ql \spl t \qr \equiv t$ and $\spl \ql t \qr \equiv t$ as definitional equalities.
+\end{itemize}
+
+A CFTT program is a mixture of object-level and meta-level top-level definitions
+and declarations. \textbf{Staging} means running all metaprograms in splices and
+inserting their output into object code, keeping all object-level top entries
+and discarding all meta-level ones. Thus, staging takes a CFTT program as input,
+and produces output which is purely in the object-level fragment, with no
+metatypes and metaprograms remaining. The output is guaranteed to be
+well-typed. This staging is formalized in detail in \cite{TODO}. In Section
+\ref{TODO} we describe the modifications to ibid.\ that are used in this paper.
+
+Let us look at some basic staging examples. Recall the meta-level identity
+function; it can be used at the object-level too, by applying it to quoted
+terms:
+  \[ \letdef n : \Nat := \spl(\id\,\ql 10 + 10 \qr); ... \]
+Here, $\id$ is used at type $\Lift \Nat$. During staging, the expression in the
+splice is evaluated, so we get $\spl\ql 10 + 10 \qr$, which is definitionally
+the same as $10 + 10$, which is our staging output here. Boolean
+short-circuiting is another basic use-case:
+\begin{alignat*}{3}
+  &\mit{and} : \Up\Bool \to \Up\Bool \to \Up\Bool\\
+  &\mit{and}\,x\,y = \ql\case \spl x\,\of \True \to \spl y; \False \to \False\qr
+\end{alignat*}
+Since the $y$ expression is inlined under a $\case$ branch at every use site, it
+is only computed at runtime when $x$ evaluates to $\True$. In many situations,
+staging can be used instead of laziness to implement short-circuiting, and with
+generally better runtime performance, avoiding the overhead of thunking. Consider
+the $\map$ function now:
+\begin{alignat*}{3}
+  & \hspace{-5em}\mathrlap{\map : \{A\,B : \VTy\} \to (\Up A \to \Up B) \to \Up (\List A) \to \Up(\List B)}\\
+  & \hspace{-5em}\mathrlap{\map\,f\,\vas = \ql\letrec \go\,\vas := \case \vas\,\of}\\
+  & \hspace{-5em}\ind\ind\ind\ind\ind \ind\Nil           &&\to \Nil;\\
+  & \hspace{-5em}\ind\ind\ind\ind\ind \ind\Cons\,a\,\vas &&\to \Cons\,\spl(f\,\ql a \qr)\,(\go\,\vas);\\
+  & \hspace{-5em}\ind\ind\ind\ind\ind\go\,\spl\vas \qr
+\end{alignat*}
+For example, this can be used as $\letdef f\,\vas : \List \Nat \to \List \Nat :=
+\spl(\map\,(\lam x.\,\ql \spl x + 10 \qr)\,\qt{\vas}) $.  This is staged to a recursive
+definition where the mapping function is inlined into the $\Cons$ case as
+$\Cons\,a\,\vas \to \Cons\,(a + 10)\,(\go\,\vas)$.  Note that $\map$ has to
+abstract over value types, since lists can only contain values, not functions. Also, the
+mapping function has type $\Up A \to \Up B$, instead of $\Up (A \to B)$. The
+former type is often preferable to the latter in staging; the former is
+a metafunction with useful computational content, while the latter is merely a
+black box that computes object code. If we have $f : \Up(A \to B)$, and $f$ is
+staged to $\ql \lam x.\,t\qr$, then $\spl f u$ is staged to an
+undesirable ``administrative'' $\beta$-redex $(\lam x.\,t)\,u$.
+
+\section{Staged Programming in CFTT}\label{sec:abstractions-in-cftt}
+
+\subsection{Binding-Time Improvements}\label{sec:binding-time-improvements}
+
+We saw that $\Up A \to \Up B$ is usually preferable to $\Up(A \to B)$. The two
+types are actually equivalent up to the runtime semantics of object programs,
+and we can convert back and forth in CFTT:
+\begin{alignat*}{3}
+  &\mup : \Up (A \to B) \to \Up A \to \Up B && \ind\ind \mdown : (\Up A \to \Up B) \to \Up (A \to B) \\
+  &\mup\,f\,a = \ql \spl f\, \spl a\qr   && \ind\ind \mdown\,f = \ql \lam a.\,\spl(f\,\ql a \qr) \qr
+\end{alignat*}
+We can not show internally, using propositional equality, that these functions are
+inverses, since we do not have $\beta\eta$-rules for object functions; but we
+will not need that in the rest of the paper.
+
+In the staged compilation and partial evaluation literature, the term
+\textbf{binding time improvement} is used to refer to such conversions, where
+the ``improved'' version supports more compile-time computation. A general
+strategy for generating efficient ``fused'' programs, is to try to work as much
+as possible with improved representations, and only convert back to object code
+at points where runtime control dependencies are unavoidable.
+
+We use a Haskell-style type class for binding time improvements:
+\begin{alignat*}{3}
+  &\hspace{-8em}\mathrlap{\tyclass\,\Improve\,(A : \Ty)\,(B : \MTy)\,\where} \\
+  &\hspace{-8em}\ind \mup   &&: \Up A \to B\\
+  &\hspace{-8em}\ind \mdown &&: B \to \Up A
+  &\\
+  &\hspace{-8em}\mathrlap{\instance \Improve\,(A \to B)\,(\Up A \to \Up B)\,\where\,\,...}
+\end{alignat*}
+We will use type classes in an informal way, without precisely specifying how
+they work. However, in the Haskell adaptation of this paper we do make crucial
+use of type classes. There are some necessary differences between the CFTT and
+Haskell versions though, which we will summarize in Section \ref{TODO}. Let us
+look at improvement for product types now:
+\begin{alignat*}{3}
+  &\hspace{-4em}\mathrlap{\instance \Improve\,(A,\,B)\,(\Up A,\,\Up B)\,\where}\\
+  &\hspace{-4em}\ind \mup\,x   && = (\ql \fst\,\spl x \qr,\, \ql \snd\,\spl x \qr)\\
+  &\hspace{-4em}\ind \mdown\,(a,\,b) && = \ql(\spl a,\, \spl b)\qr
+\end{alignat*}
+Here we overload Haskell-style product type notation for types and pairing, at
+both levels (products are definable as a value type at the object level). There
+is a problem with this conversion though: $\mup$ uses $x : \Up(A,\,B)$ twice,
+which can increase code size and duplicate runtime computations. For example,
+$\mdown\,(\mup\,\ql f\,x \qr)$ is staged to $\ql (\fst\,(f\,x),\,\snd\,(f\,x))
+\qr$. It would be safer to first $\mbf{let}$-bind an expression with type
+$\Up(A,\,B)$, and then only use projections of the newly bound variable. This is
+called \textbf{let-insertion} in staged compilation. But it is impossible to use
+let-insertion in $\mup$ because the return type is a product in $\MTy$, and we
+cannot introduce object binders in meta-level code.
+
+\subsection{The Code Generation Monad}\label{sec:the-code-generation-monad}
+
+Our solution is to use a monad which freely extends $\MTy$ with the ability to
+generate object-level code and introduce object binders.
+\begin{alignat*}{3}
+  & \newtype \Gen\,A = \Gen\,\{\unGen : \{R : \Ty\} \to (A \to \Up R) \to \Up R\}
+\end{alignat*}
+This is a monad in $\MTy$ in a very standard sense, and we reuse Haskell type classes
+and $\mdo$-notation.
+\begin{alignat*}{3}
+  &\instance \Monad\,\Gen\,\where\\
+  &\ind \return a = \Gen\,(\lam k.\,k\,a)\\
+  &\ind \mit{ga} \bind f = \Gen\,(\lam k.\,\unGen\,\mit{ga}\,(\lam a.\,(f\,a\,k)))
+\end{alignat*}
+From the $\Monad$ instance, the $\Functor$ and $\Applicative$ instances can be also derived.
+We will reuse $(\fmap)$ and $(\ap)$ for applicative notation
+as well.
+An inhabitant of $\Gen\,A$ can be viewed as an action whose effect is to produce
+some object code and returns an $A$ value that can depend on produced object
+binders. $\Gen$ can be only ``run'' when the result type is an object type:
+\begin{alignat*}{3}
+  &\runGen : \Gen\,(\Up A) \to \Up A\\
+  &\runGen\,\vma = \unGen\,\vma\,\id
+\end{alignat*}
+We can let-bind object expressions in $\Gen$:
+\begin{alignat*}{3}
+  & \gen : \Up A \to \Gen\,(\Up A) \\
+  & \gen\,a = \Gen\,(\lam k.\,\ql \letdef x : A := \spl a; \spl(k\,\ql x \qr) \qr)
+\end{alignat*}
+And also recursive function definitions:
+\begin{alignat*}{3}
+  & \genRec : \{A : \CTy\} \to (\Up A \to \Up A) \to \Gen\,(\Up A) \\
+  & \genRec\,f = \Gen\,(\lam k.\,\qt{\letrec x : A := \spl(f\,\qt{x}); \spl(k\,\qt{x})})
+\end{alignat*}
+Now, using do-notation, we may write $\mdo \{x \fro \gen\,\qt{10 + 10}; y \fro
+\gen\,\qt{20 + 20}\};\return \qt{x + y}\}$, for a $\Gen\,(\Up \Nat)$
+action. Running this with $\runGen$ yields $\qt{\letdef x := 10 + 10; \letdef y
+  := 20 + 20; x + y}$. We can also define a ``safer'' binding-time improvement for
+products, using let-insertion:
+\begin{alignat*}{3}
+  &\hspace{-4em}\mathrlap{\instance \Improve\,(A,\,B)\,(\Gen\,(\Up A,\,\Up B))\,\where}\\
+  &\hspace{-4em}\ind \mup\,x   && = \mdo\{x \fro \gen\,x; \return (\qt{\fst\,\spl x},\,\qt{\snd\,\spl x})\}\\
+  &\hspace{-4em}\ind \mdown\,x && = \runGen\,\$\,\mdo \{(a,\,b) \fro x; \return \qt{(\spl a,\, \spl b)}\}
+\end{alignat*}
+
+Working in $\Gen$ is convenient, since we can freely generate object code and
+also have access to the full metalanguage. Also, the whole point of staging is
+that \emph{eventually} all metaprograms will be used for the purpose of code
+generation, so ultimately we always want to $\runGen$ our actions. So why not
+just always work in $\Gen$? We may not want to do that, because we would like to
+reason about the size and content of the code we are generating, and the
+implicit code generation in $\Gen$ is not very transparent. This is a bit
+similar to the $\msf{IO}$ monad in Haskell, where eventually everything needs
+to run in $\msf{IO}$, but we may prefer to not write most of our program in
+$\msf{IO}$.
+
+\subsection{Monads}\label{sec:monads}
+
+Let us start with the $\Maybe$ monad. We have $\data \Maybe\,A : \VTy :=
+\Nothing\,|\,\Just\,A$, and $\Maybe$ itself is available as a $\VTy \to \VTy$
+metafunction. However, we cannot directly fashion a monad out of $\Maybe$, since
+we do not have enough type formers in $\VTy$. We could try to use the following
+type for binding:
+\[ (\bind) : \Up(\Maybe\,A) \to (\Up\,A \to \Up\,(\Maybe\,B)) \to \Up(\Maybe\,B) \]
+This works, but the definition necessarily uses runtime case splits on $\Maybe$
+values, many of which could be optimized away during staging. Also, not having
+a ``real'' monad is inconvenient for the purpose of code reuse.
+
+Instead, our strategy is to only use ordinary monads in $\MTy$, and convert
+between object types and meta-monads when necessary, as a form of binding-time
+improvement. Assume that $\Maybe_M$ is a meta-level monad, and $\MaybeT_\M$ is
+the monad transformer defined as follows:
+\[ \newtype \MaybeT_\M\,M\,A = \MaybeT_\M\,\{\runMaybeT_\M : M\,(\Maybe_\M\,A)\} \]
+Then the object-level $\Maybe$ is binding-time improved as follows:
+\begin{alignat*}{3}
+  &\instance \Improve\,(\Maybe\,A)\,(\MaybeT_\M\,\Gen\,(\Up\,A))\,\where\\
+  %% &\ind \mup : \Up\,(\Maybe\,A) \to \MaybeT_\M\,\Gen\,(\Up\,A) \\
+  &\ind \mup\,x = \MaybeT_\M \dlr \Gen \dlr \lam k.\\
+  &\ind\ind \qt{\case \spl x\,\of \Nothing \to \spl(k\,\Nothing_\M);\Just\,a \to \spl(k\,(\Just_\M\qt{a}))}\\
+  &\\
+  %% &\ind \mdown : \MaybeT_\M\,\Gen\,(\Up\,A) \to \Up\,(\Maybe\,A) \\
+  &\ind \mdown\,(\MaybeT_\M\,(\Gen\,\vma)) = \\
+  &\ind\ind \vma\,(\lam x.\,\case x\,\of \Nothing_\M \to \qt{\Nothing}; \Just_\M\,a \to \qt{\Just\,\spl a})
+\end{alignat*}
+With this, we get the $\Monad$ instance for free from $\MaybeT_\M$ and $\Gen$. A small example:
+\[ \letdef n : \Maybe\,\Nat := \spl(\mdown\,\$\,\mdo \{x \fro \return\,\qt{10}; y \fro \return\,\qt{20}; \return\,\qt{x + y})\});\,... \]
+Since $\MaybeT_\M$ is meta-level, its binding always fully computes at staging time. Thus, the above code is staged to
+\[ \letdef n : \Maybe\,\Nat := \Just\,(10 + 20);\;... \]
+Assume also a $\lift : \Monad\,M \RA M\,A \to \MaybeT_\M\,M\,A$ operation, as in
+Haskell. We shall assume this for all monad transformers henceforth. We can do
+let-insertion in $\MaybeT_\M\,\Gen$ by simply lifting:
+\begin{alignat*}{3}
+  &\gen' : \Up A \to \MaybeT_\M\,\Gen\,(\Up A) \\
+  &\gen'\,a = \lift\,(\gen\,a)
+\end{alignat*}
+However, it is more convenient to proceed in the style of Haskell's monad
+transformer library \cite{TODO}, and have a class for monads that can do code
+generation:
+\begin{alignat*}{3}
+&\tyclass\,\Monad\,M \RA \MonadGen\,M\,\where\\
+&\ind \liftGen : \Gen\,A \to \M\,A\\
+&\instance \MonadGen\,\Gen\,\where\,\liftGen = \id \\
+&\instance \MonadGen\,M \RA \MonadGen\,(\MaybeT_\M\,M)\,\where\,\liftGen = \lift \circ \liftGen
+\end{alignat*}
+We also redefine $\gen$ and $\genRec$ to work in any $\MonadGen$, so from now on we have:
+\begin{alignat*}{3}
+ & \gen   &&: \MonadGen\,M \RA \Up A \to M\,(\Up A)\\
+ &\genRec &&: \MonadGen\,M \RA (\Up A \to \Up A) \to M\,(\Up A)
+\end{alignat*}
+
+\subsubsection{Case splitting in monads}
+Trying to write more substantial code, we quickly bump into an issue. We
+want to case-split on object-level data inside a monadic action, like in the following:
+\begin{alignat*}{3}
+  & \hspace{-4em}\mathrlap{f : \Nat \to \Maybe\,\Nat}\\
+  & \hspace{-4em}\mathrlap{f\,x := \spl(\mdown\,\$\, \case x == 10\,\of}\\
+  & \hspace{-4em}\ind \True  &&\to \return\,\qt{x + 5};\\
+  & \hspace{-4em}\ind \False &&\to \Nothing_\M)
+\end{alignat*}
+This is ill-typed as written, since we cannot compute meta-level actions from an
+object-level case split. Fortunately, with a little bit more work, case
+splitting on object values is actually possible in any $\MonadGen$, on any value
+type.
+
+We demonstrate this for lists first. An object-level $\mbf{case}$ on lists introduces
+two points where code generation can continue. We define a metatype which gives
+us a ``view'' on these points:
+\[ \data \msf{SplitList}\,A = \Nil'\,|\,\Cons'\,(\Up A)\,(\Up (\List\,A)) \]
+We can generate code for a case split, returning a view on it:
+\begin{alignat*}{3}
+  &\mit{split} : \Up (\List\,A) \to \Gen\,(\msf{SplitList}\,A)\\
+  &\mit{split}\,\vas = \Gen \dlr \lam k.\,\qt{\case \spl \vas\,\of \Nil \to \spl(k\,\Nil'); \Cons\,a\,\vas \to \spl(k\,(\Cons'\,\qt{a}\,\qt{\vas}))}
+\end{alignat*}
+Now, in any $\MonadGen$, we may write
+\begin{alignat*}{3}
+  &\mdo && \mit{sp} \fro \liftGen \dlr \mit{split}\,\vas\\
+  &     && \case \mit{sp}\,\of \\
+  &     && \ind \Nil' \to ... \\
+  &     && \ind \Cons'\,a\,\vas \to ...
+\end{alignat*}
+This can be generalized to any splitting on object values. In the supplemental
+Haskell implementation, we overload $\mit{split}$ with a type class. However, in
+a native implementation of CFTT it might make sense to simply extend
+$\mdo$-notation with an elaboration of $\mbf{case}$ on $\Up A$-typed data to an
+application of the appropriate $\mit{split}$ function. We adopt this in
+the rest of this paper. So we will simply write
+\[ \mdo \{\case \vas\,\of \Nil \to ...; \Cons\,a\,\vas \to ...\}. \]
+
+\subsection{Building a Monad Transformer Library}\label{monad-transformers}
+
+At this point, it makes sense to aim for a monad transformer library where
+binding-time improvement is defined compositionally, by recursion on the
+transformer stack. The base case is the following:
+\begin{alignat*}{3}
+  & \newtype \Identity\,A := \Identity\,\{\runIdentity : A\} \\
+  & \instance \Improve\,(\Identity\,A)\,(\Gen\,(\Up A))\,\where\\
+  & \ind \mup\,x = \Gen \dlr \lam k.\,k\,\qt{\runIdentity\,\spl x}\\
+  & \ind \mdown\,x = \unGen\,x \dlr \lam a.\,\qt{\Identity\,\spl a}
+\end{alignat*}
+We also switch from the object-level $\Maybe$ to a ``transformer'' form as well,
+and recover it as $\MaybeT\,\Identity$.
+\[ \newtype \MaybeT\,M\,A := \MaybeT\,\{\runMaybeT : M\,(\Maybe\,A)\} \]
+Generally, we want to improve an operator $F : \VTy \to \VTy$ to some $M : \MTy
+\to \MTy$, where $M$ is a $\MonadGen$. We define a \emph{quantified class
+constraint} \cite{TODO} for this, that we will often use as an assumption:
+\[ \ImproveF\,F\,M = \MonadGen\,M \times ((A : \VTy) \to \Improve\,(F\,A)\,(M\,(\Lift\,A))) \]
+With this, improvement can be generally defined for $\MaybeT$:
+\begin{alignat*}{3}
+  &\instance \ImproveF\,F\,M \RA \Improve\,(\MaybeT\,F\,A)\,(\MaybeT_\M\,M\,(\Up A))\,\where\\
+  &\ind \mup\,x = \MaybeT_\M \dlr \mdo\\
+  &\ind\ind \vma \fro \mup\,\qt{\runMaybeT\,\spl x}\\
+  &\ind\ind \case \vma\,\of \Nothing \to \return \Nothing_\M\\
+  &\hspace{6.5em}         \Just\,a \hspace{1.15em}\to \return (\Just_\M\,a)\\
+  &\\
+  &\ind \mdown\,(\MaybeT_\M\,x) = \qt{\MaybeT\,\spl(\mdown \dlr x \bind \lambda\case\\
+  &\ind\ind\Nothing_\M \to \return \qt{\Nothing}\\
+  &\ind\ind\Just_\M\,a\hspace{1.15em} \to \return \qt{\Just\,\spl a})}
+\end{alignat*}
+In the $\mbf{case}$ in $\mup$, we already use our syntax sugar for matching on a $\Maybe$
+inside an $\M$ action. This is legal, since we know from the $\ImproveF\,F\,M$ assumption
+that $M$ is a $\MonadGen$.
+
+In the meta level, we can reuse essentially all definitions from Haskell's
+monad transformer library $\msf{mtl}$. The additional work that we have to do in
+the CFTT setting:
+\begin{itemize}
+  \item Adding $\MonadGen$ instances, but these are trivially defined as
+        $\liftGen = \lift \circ \liftGen$ everywhere.
+  \item Defining the object-level counterparts of monad transformers, and
+        also the $\Improve$ instances.
+\end{itemize}
+From $\msf{mtl}$, only the continuation monad transformer fails to support
+binding-time-improvement in CFTT, because of the obvious need for dynamic
+closures. In the following we present only $\ReaderT$ and $\StateT$.
+\begin{alignat*}{3}
+  &\instance \ImproveF\,F\,M \RA \Improve\,(\MaybeT\,F\,A)\,(\MaybeT_\M\,M\,(\Up A))\,\where\\
+  &\ind \mup\,x = \MaybeT_\M \dlr \mdo\\
+  &\ind\ind \vma \fro \mup\,\qt{\runMaybeT\,\spl x}\\
+  &\ind\ind \case \vma\,\of \Nothing \to \return \Nothing_\M\\
+  &\hspace{6.5em}         \Just\,a \hspace{1.15em}\to \return (\Just_\M\,a)\\
+  &\\
+  &\ind \mdown\,(\MaybeT_\M\,x) = \qt{\MaybeT\,\spl(\mdown \dlr x \bind \lambda\case\\
+  &\ind\ind\Nothing_\M \to \return \qt{\Nothing}\\
+  &\ind\ind\Just_\M\,a\hspace{1.15em} \to \return \qt{\Just\,\spl a})}
+\end{alignat*}
+
+%% instance (Improve m n) => Improve (StateT s m) (StateT (Up s) n) where
+
+%%   up x = StateT \s ->
+%%     do as <- up [|| runStateT $$x $$s ||]
+%%        caseGen as pure
+
+%%   down x = [|| StateT \s -> $$(down
+%%     do (a, s) <- runStateT x [||s||]
+%%        pure [||($$a, $$s)||]
+%%        )||]
+
+%% instance Improve m n => Improve (MaybeT m) (MaybeT n) where
+%%   up x = MaybeT do
+%%     ma <- up [||runMaybeT $$x||]
+%%     caseGen ma pure
+
+%%   down (MaybeT x) =
+%%     [|| MaybeT $$(down (x >>= \case
+%%           Nothing -> pure [||Nothing||]
+%%           Just a -> pure [||Just $$a||])) ||]
+
+%% instance Improve m n => Improve (ReaderT r m) (ReaderT (Up r) n) where
+%%   up   x = ReaderT \r -> up [||runReaderT $$x $$r||]
+%%   down x = [|| ReaderT \r -> $$(down (runReaderT x [||r||])) ||]
 
@@
 
+\todo{Introduce newtypes accordingly}
+\todo{allow arbitrary parameters (including ValTy -> ValTy functions) for object ADTs}
 \todo{extend as I have more stuff}
@@
 
-\section{The Object Language}\label{sec:the-object-language}
+%% \section{The Object Language}\label{sec:the-object-language}
 
-Describe object language. Simple types + first-order fun + fixpoints + ADT-s.
+%% Describe object language. Simple types + first-order fun + fixpoints + ADT-s.
 
-Show Agda formalization basically, types and syntax.
+%% Show Agda formalization basically, types and syntax.
 
-Show Agda definitional interpreter. Discuss closure-free evaluation
+%% Show Agda definitional interpreter. Discuss closure-free evaluation
 
-Examples, properties. Maybe big-step reduction? Maybe program equivalence?  Look
-at Harper PFPL for program equivalence reference.
+%% Examples, properties. Maybe big-step reduction? Maybe program equivalence?  Look
+%% at Harper PFPL for program equivalence reference.
 
-Sketch downstream compilation.
-\begin{itemize}
-\item Eta-expand function definitions to arity.
-\item Turn functions in immediate beta-redexes to let-bindings.
-\item Lambda-lift non-tail-called functions to top.
-\item Keep only-tail-called local functions as join points.
-\end{itemize}
+%% Sketch downstream compilation.
+%% \begin{itemize}
+%% \item Eta-expand function definitions to arity.
+%% \item Turn functions in immediate beta-redexes to let-bindings.
+%% \item Lambda-lift non-tail-called functions to top.
+%% \item Keep only-tail-called local functions as join points.
+%% \end{itemize}
 
-Downstream compilation should erase the unit type and implement products as
-``flat'' unboxed structures.
+%% Downstream compilation should erase the unit type and implement products as
+%% ``flat'' unboxed structures.
 
-Discuss object language design choices.
-\begin{itemize}
-\item Can be more obviously first-order: e.g.\ no currying, only Val->Val
-  functions, or ANF functions. This requires more bureaucracy in
-  metaprogramming.
-\item Can be more flexible but less obviously first-order: allow function let
-  bodies, allow function case bodies. Requires more translation. More convenient
-  to program in. In future work.
-\end{itemize}
+%% Discuss object language design choices.
+%% \begin{itemize}
+%% \item Can be more obviously first-order: e.g.\ no currying, only Val->Val
+%%   functions, or ANF functions. This requires more bureaucracy in
+%%   metaprogramming.
+%% \item Can be more flexible but less obviously first-order: allow function let
+%%   bodies, allow function case bodies. Requires more translation. More convenient
+%%   to program in. In future work.
+%% \end{itemize}
 
-\section{The Staged Language}\label{sec:the-staged-language}
+%% \section{The Staged Language}\label{sec:the-staged-language}
 
-Don't give horizontal-bar rules.
+%% Don't give horizontal-bar rules.
 
-Meta-features:
-\begin{itemize}
-\item Pi, Sigma, Unit, Bottom, Universes (indices elided), ADT-s. Ind families probably not needed.
-\item Agda notation.
-\item Type classes informally in Haskell notation?
-\end{itemize}
+%% Meta-features:
+%% \begin{itemize}
+%% \item Pi, Sigma, Unit, Bottom, Universes (indices elided), ADT-s. Ind families probably not needed.
+%% \item Agda notation.
+%% \item Type classes informally in Haskell notation?
+%% \end{itemize}
 
-Explain staging ops, staging algorithm. Explain object embedding. Refer to 2LTT
-paper for correctness.
+%% Explain staging ops, staging algorithm. Explain object embedding. Refer to 2LTT
+%% paper for correctness.
 
-Mention stage annotation inference, but we'll not use it often! To make things
-explicit.
+%% Mention stage annotation inference, but we'll not use it often! To make things
+%% explicit.
 
-We should include a generous amount of staging examples, and some traced staging.
+%% We should include a generous amount of staging examples, and some traced staging.
 
-Basic staging examples: identity, map, power function. Boolean short-circuiting.
+%% Basic staging examples: identity, map, power function. Boolean short-circuiting.
 
-Basic staged classes, Eq, Ord, Show, Monoid, etc. Vectors with computed length.
+%% Basic staged classes, Eq, Ord, Show, Monoid, etc. Vectors with computed length.
 
-\section{Programming With Monads}\label{sec:programming-with-monads}
+%% \section{Programming With Monads}\label{sec:programming-with-monads}
 
-\subsection{Basic Binding Time Improvements}\label{sec:basic-binding-time-improvements}
+%% \subsection{Basic Binding Time Improvements}\label{sec:basic-binding-time-improvements}
 
-Notion.
-For functions.
-For products.
-Note that for products there is a computation duplication problem.
-Mention let-insertion.
+%% Notion.
+%% For functions.
+%% For products.
+%% Note that for products there is a computation duplication problem.
+%% Mention let-insertion.
 
-\subsection{The Code Generation Monad}\label{sec:the-code-generation-monad}
+%% \subsection{The Code Generation Monad}\label{sec:the-code-generation-monad}
 
-Is fundamental to this work.
+%% Is fundamental to this work.
 
-Definition. Explanation. Define Functor, Applicative, Monad instances.
-Define runGen, runGen1, runGenN etc. Define let-insertion.
+%% Definition. Explanation. Define Functor, Applicative, Monad instances.
+%% Define runGen, runGen1, runGenN etc. Define let-insertion.
 
-Basic examples.
+%% Basic examples.
 
-Explanation in terms of Church-coding.
+%% Explanation in terms of Church-coding.
 
-Gen Vector mapping or folding example from 2LTT repo. Without Gen, it's a
-big pain to get ideal linear-size map code in N. With Gen it's easy-peasy.
+%% Gen Vector mapping or folding example from 2LTT repo. Without Gen, it's a
+%% big pain to get ideal linear-size map code in N. With Gen it's easy-peasy.
 
-\subsection{Monad Transformers}\label{sec:monad-transformers}
+%% \subsection{Monad Transformers}\label{sec:monad-transformers}
 
-Identity, ReaderT, StateT, MaybeT, ExceptT
+%% Identity, ReaderT, StateT, MaybeT, ExceptT
 
-Strict versions of put, modify, local, runner functions.
+%% Strict versions of put, modify, local, runner functions.
 
-MTL-style overloading. MonadReader, MonadState, MonadError, MonadGen, liftGen
+%% MTL-style overloading. MonadReader, MonadState, MonadError, MonadGen, liftGen
 
-Binding time improvement for monads. Up/down overloading.
+%% Binding time improvement for monads. Up/down overloading.
 
-\subsection{Case Splitting in MonadGen}\label{sec:case-splitting-in-monadgen}
+%% \subsection{Case Splitting in MonadGen}\label{sec:case-splitting-in-monadgen}
 
-It works for any object ADT.
+%% It works for any object ADT.
 
-Introduce sums-of-values (SOV).
+%% Introduce sums-of-values (SOV).
 
-Introduce syntax sugar for object case splits in MonadGen blocks.
+%% Introduce syntax sugar for object case splits in MonadGen blocks.
 
-Basic examples.
+%% Basic examples.
 
-\subsection{Joining Control Flow}\label{sec:rejoining-control-flow}
+%% \subsection{Joining Control Flow}\label{sec:rejoining-control-flow}
 
-The problem of join points. Motivation in terms of MaybeT Gen.
+%% The problem of join points. Motivation in terms of MaybeT Gen.
 
-We can let-bind computations by tripping through up/down of monads,
-but that introduces unbox/rebox cost for Maybe \& Either.
+%% We can let-bind computations by tripping through up/down of monads,
+%% but that introduces unbox/rebox cost for Maybe \& Either.
 
-Works OK for Reader/State though.
+%% Works OK for Reader/State though.
 
-\begin{verbatim}
-f : Bool -> MaybeT0 Gen ^Int
-f = \b. ~(down do
-  x <- case <b> of
-    True  -> gen <10>
-    False -> gen <20>
-  y <- gen <~x + 10>
-  z <- gen <~x * 10>
-  pure <~x + ~y + ~z>)
-\end{verbatim}
+%% \begin{verbatim}
+%% f : Bool -> MaybeT0 Gen ^Int
+%% f = \b. ~(down do
+%%   x <- case <b> of
+%%     True  -> gen <10>
+%%     False -> gen <20>
+%%   y <- gen <~x + 10>
+%%   z <- gen <~x * 10>
+%%   pure <~x + ~y + ~z>)
+%% \end{verbatim}
 
-Join points using SOV-s. Description of SOV machinery. MonadJoin.
-instance MonadJoin Gen. MonadJoin MaybeT.
+%% Join points using SOV-s. Description of SOV machinery. MonadJoin.
+%% instance MonadJoin Gen. MonadJoin MaybeT.
 
-\subsection{Monadic Tail Recursion}\label{sec:monadic-tail-recursion}
+%% \subsection{Monadic Tail Recursion}\label{sec:monadic-tail-recursion}
 
-\texttt{replicateM\_}, \texttt{find} functions working in ``any'' monad.
-Non-tail recursion is not possible to fuse, return addresses are dynamic.
-Tail recursion can return to statically known points.
+%% \texttt{replicateM\_}, \texttt{find} functions working in ``any'' monad.
+%% Non-tail recursion is not possible to fuse, return addresses are dynamic.
+%% Tail recursion can return to statically known points.
 
-\subsection{Discussion}
+%% \subsection{Discussion}
 
-Limitations. We can't put non-value actions into structures, can't accumulate them.
+%% Limitations. We can't put non-value actions into structures, can't accumulate them.
 
-Guarantees/properties. Closure-freedom. No administrative redexes. No boxing in join points.
-Only boxing in monadic code comes from ``down''. No code blowup.
+%% Guarantees/properties. Closure-freedom. No administrative redexes. No boxing in join points.
+%% Only boxing in monadic code comes from ``down''. No code blowup.
 
-\section{Fusion}\label{sec:push-pull-fusion}
+%% \section{Fusion}\label{sec:push-pull-fusion}
 
-Motivation.
+%% Motivation.
 
-\subsection{Push streams}\label{sec:push-streams}
+%% \subsection{Push streams}\label{sec:push-streams}
 
-Definition.
+%% Definition.
 
-\subsection{Pull streams}\label{sec:pull-streams}
+%% \subsection{Pull streams}\label{sec:pull-streams}
 
-Pull streams are definitely not monadic. \texttt{Nat -> Stream a} contains
-``infinite'' amount of code internally.
+%% Pull streams are definitely not monadic. \texttt{Nat -> Stream a} contains
+%% ``infinite'' amount of code internally.
 
-Pull streams are also not selective functors.
+%% Pull streams are also not selective functors.
 
-We do have monadic binding for lifted types, plus ``case'' matching
-on arbitrary object values! This is practically almost as good as
-having unlimited monadic bind! We need to assume a ``non-dependency axiom'',
-that any $f : \Lift A \to \msf{ValTy}$ must be constant.
+%% We do have monadic binding for lifted types, plus ``case'' matching
+%% on arbitrary object values! This is practically almost as good as
+%% having unlimited monadic bind! We need to assume a ``non-dependency axiom'',
+%% that any $f : \Lift A \to \msf{ValTy}$ must be constant.
 
-\todo{Validate non-dependency in the psh model!}
-\todo{Add meta-identity type to the psh model}
+%% \todo{Validate non-dependency in the psh model!}
+%% \todo{Add meta-identity type to the psh model}
 

### 2024-02-13 newpaper/paper.tex
stuff

@@
 \newcommand{\Int}{\msf{Int}}
-\newcommand{\List}{\msf{List}\,}
+\newcommand{\List}{\msf{List}}
 \newcommand{\Nil}{\msf{Nil}}
@@
 \newcommand{\MonadGen}{\msf{MonadGen}}
+\newcommand{\MonadState}{\msf{MonadState}}
+\newcommand{\MonadReader}{\msf{MonadReader}}
 \newcommand{\RA}{\Rightarrow}
@@
 \newcommand{\Identity}{\msf{Identity}}
+
+\newcommand{\Stop}{\msf{Stop}}
+\newcommand{\Skip}{\msf{Skip}}
+\newcommand{\Yield}{\msf{Yield}}
+
 \newcommand{\runIdentity}{\mit{runIdentity}}
+\newcommand{\runReaderT}{\mit{runReaderT}}
 \newcommand{\newtype}{\mbf{newtype}\,}
 \newcommand{\runMaybeT}{\mit{runMaybeT}}
+\newcommand{\runStateT}{\mit{runStateT}}
+\newcommand{\runState}{\mit{runState}}
 \newcommand{\dlr}{\,\$\,}
@@
 \newcommand{\ExceptT}{\msf{ExceptT}}
+\newcommand{\State}{\msf{State}}
+\newcommand{\SumVS}{\msf{SumVS}}
+\newcommand{\ProdCS}{\msf{ProdCS}}
+\newcommand{\Here}{\msf{Here}}
+\newcommand{\There}{\msf{There}}
+\newcommand{\IsSumVS}{\msf{IsSumVS}}
+\newcommand{\MonadJoin}{\msf{MonadJoin}}
+\newcommand{\Stream}{\msf{Stream}}
+\newcommand{\join}{\mit{join}}
+\newcommand{\modify}{\mit{modify}}
+\newcommand{\get}{\mit{get}}
+\newcommand{\mput}{\mit{put}}
+\newcommand{\mAs}{\mit{As}}
+\newcommand{\Rep}{\mit{Rep}}
+\newcommand{\encode}{\mit{encode}}
+\newcommand{\decode}{\mit{decode}}
+\newcommand{\mindex}{\mit{index}}
+\newcommand{\mtabulate}{\mit{tabulate}}
+\newcommand{\States}{\mit{States}}
+\newcommand{\seed}{\mit{seed}}
+\newcommand{\step}{\mit{step}}
+\newcommand{\Step}{\msf{Step}}
+\newcommand{\Pull}{\msf{Pull}}
+\newcommand{\MkPull}{\msf{MkPull}}
 
@@
 
+\todo{REORG intro, trim, contrib, motivate monadtrans \& stream fusion}
+
 GHC's optimization efforts are respectable, and it has gotten quite good over
@@
 
+\todo{If we won't use Sigma notation, get rid of it}
+
 We use Agda-like syntax for functions and implicit arguments. A basic example:
@@
   &                                           &&\hspace{4em}\data\,\Bool_\M &&: \MTy\,\where\\
-  & \data\,\Bool_\M : \MTy = \True\,|\,\False &&\hspace{4em}\ind\ind \True &&: \Bool_\M\\
-  &                                           &&\hspace{4em}\ind\ind \False &&: \Bool_\M
+  & \data\,\Bool_\M : \MTy = \True_\M\,|\,\False_\M &&\hspace{4em}\ind\ind \True_\M &&: \Bool_\M\\
+  &                                           &&\hspace{4em}\ind\ind \False_\M &&: \Bool_\M
 \end{alignat*}
@@
 \end{alignat*}
+We also allow $\mbf{newtype}$ definitions, both in $\VTy$ and $\CTy$. These are
+assumed to be erased at runtime. In the Haskell adaptation they are important
+for aiding type class resolution, and we think that the explicit wrapping
+makes many definitions more comprehensible in CFTT as well.
+
 Values are call-by-value at runtime; they are computed eagerly in function
@@
 \begin{alignat*}{3}
-  &\hspace{-1em}f : \mathrlap{\Bool \to \Nat \to \Nat}\\
-  &\hspace{-1em}f\,b\,\mathrlap{:= \case b\,\of}\\
-  &\hspace{-1em}\ind \True  &&\to \lam x.\, x + 10;\\
-  &\hspace{-1em}\ind \False &&\to \lam x.\, x * 10;
+  &f : \mathrlap{\Bool \to \Nat \to \Nat}\\
+  &f\,b := \case b\,\of \True \to (\lam x.\, x + 10); \False &&\to (\lam x.\, x * 10)
 \end{alignat*}
@@
 \begin{alignat*}{3}
-  & \hspace{-5em}\mathrlap{\map : \{A\,B : \VTy\} \to (\Up A \to \Up B) \to \Up (\List A) \to \Up(\List B)}\\
+  & \hspace{-5em}\mathrlap{\map : \{A\,B : \VTy\} \to (\Up A \to \Up B) \to \Up (\List\,A) \to \Up(\List\,B)}\\
   & \hspace{-5em}\mathrlap{\map\,f\,\vas = \ql\letrec \go\,\vas := \case \vas\,\of}\\
@@
 \end{alignat*}
-For example, this can be used as $\letdef f\,\vas : \List \Nat \to \List \Nat :=
+For example, this can be used as $\letdef f\,\vas : \List\,\Nat \to \List\,\Nat :=
 \spl(\map\,(\lam x.\,\ql \spl x + 10 \qr)\,\qt{\vas}) $.  This is staged to a recursive
@@
 
-\section{Staged Programming in CFTT}\label{sec:abstractions-in-cftt}
+\section{Monads \& Monad Transformers}\label{sec:monad-transformers}
+
+In this section we build a library for monads and monad transformers. We believe
+that this is a good demonstration of CFTT-s abilities, since monads are
+ubiquitous in practical Haskell programming, and they also introduce a great
+amount of abstraction and that should be optimized away.
 
@@
 
-We saw that $\Up A \to \Up B$ is usually preferable to $\Up(A \to B)$. The two
-types are actually equivalent up to the runtime semantics of object programs,
-and we can convert back and forth in CFTT:
+We start with some preparatory work before getting to monads. We saw that $\Up
+A \to \Up B$ is usually preferable to $\Up(A \to B)$. The two types are actually
+equivalent up to the runtime semantics of object programs, and we can convert
+back and forth in CFTT:
 \begin{alignat*}{3}
@@
 inverses, since we do not have $\beta\eta$-rules for object functions; but we
-will not need that in the rest of the paper.
+will not need this proof in the rest of the paper.
 
@@
 
-Our solution is to use a monad which freely extends $\MTy$ with the ability to
-generate object-level code and introduce object binders.
+Our solution is to use a monad which extends $\MTy$ with the ability to
+freely generate object-level code and introduce object binders.
 \begin{alignat*}{3}
@@
 just always work in $\Gen$? We may not want to do that, because we would like to
-reason about the size and content of the code we are generating, and the
+reason about the size and makeup of the code we are generating, and the
 implicit code generation in $\Gen$ is not very transparent. This is a bit
@@
 
-Let us start with the $\Maybe$ monad. We have $\data \Maybe\,A : \VTy :=
+Let us start with the $\Maybe$ monad. We have $\data \Maybe\,A :=
 \Nothing\,|\,\Just\,A$, and $\Maybe$ itself is available as a $\VTy \to \VTy$
@@
 
-Instead, our strategy is to only use ordinary monads in $\MTy$, and convert
+Instead, our strategy is to only use proper monads in $\MTy$, and convert
 between object types and meta-monads when necessary, as a form of binding-time
 improvement. Assume that $\Maybe_M$ is a meta-level monad, and $\MaybeT_\M$ is
-the monad transformer defined as follows:
+the monad transformer with the standard definition:
 \[ \newtype \MaybeT_\M\,M\,A = \MaybeT_\M\,\{\runMaybeT_\M : M\,(\Maybe_\M\,A)\} \]
@@
   &\ind\ind \qt{\case \spl x\,\of \Nothing \to \spl(k\,\Nothing_\M);\Just\,a \to \spl(k\,(\Just_\M\qt{a}))}\\
-  &\\
   %% &\ind \mdown : \MaybeT_\M\,\Gen\,(\Up\,A) \to \Up\,(\Maybe\,A) \\
@@
 \begin{alignat*}{3}
-  &\mdo && \mit{sp} \fro \liftGen \dlr \mit{split}\,\vas\\
-  &     && \case \mit{sp}\,\of \\
-  &     && \ind \Nil' \to ... \\
-  &     && \ind \Cons'\,a\,\vas \to ...
-\end{alignat*}
+  &\mdo \{\mit{sp} \fro \liftGen\;(\!\mit{split}\,\vas);(\case \mit{sp}\,\of \Nil' \to ...;\Cons'\,a\,\vas \to ...)\}
+\end{alignat*}
+%% \begin{alignat*}{3}
+%%   &\mdo && \mit{sp} \fro \liftGen \dlr \mit{split}\,\vas\\
+%%   &     && \case \mit{sp}\,\of \\
+%%   &     && \ind \Nil' \to ... \\
+%%   &     && \ind \Cons'\,a\,\vas \to ...
+%% \end{alignat*}
 This can be generalized to any splitting on object values. In the supplemental
@@
 $\mdo$-notation with an elaboration of $\mbf{case}$ on $\Up A$-typed data to an
-application of the appropriate $\mit{split}$ function. We adopt this in
-the rest of this paper. So we will simply write
-\[ \mdo \{\case \vas\,\of \Nil \to ...; \Cons\,a\,\vas \to ...\}. \]
-
-\subsection{Building a Monad Transformer Library}\label{monad-transformers}
+application of the appropriate $\mit{split}$ function, so we can simply write a
+$\case\,\vas\,\of\,...$ in a $\mdo$-block. We adopt this in the rest of this
+paper.
+
+\subsection{Monad Transformers}\label{monad-transformers}
 
@@
   &\hspace{6.5em}         \Just\,a \hspace{1.15em}\to \return (\Just_\M\,a)\\
-  &\\
-  &\ind \mdown\,(\MaybeT_\M\,x) = \qt{\MaybeT\,\spl(\mdown \dlr x \bind \lambda\case\\
+  &\ind \mdown\,(\MaybeT_\M\,x) = \qt{\MaybeT\,\spl(\mdown \dlr x \bind \lam x.\,\case x\,\of\\
   &\ind\ind\Nothing_\M \to \return \qt{\Nothing}\\
@@
 binding-time-improvement in CFTT, because of the obvious need for dynamic
-closures. In the following we present only $\ReaderT$ and $\StateT$.
-\begin{alignat*}{3}
-  &\instance \ImproveF\,F\,M \RA \Improve\,(\MaybeT\,F\,A)\,(\MaybeT_\M\,M\,(\Up A))\,\where\\
-  &\ind \mup\,x = \MaybeT_\M \dlr \mdo\\
-  &\ind\ind \vma \fro \mup\,\qt{\runMaybeT\,\spl x}\\
-  &\ind\ind \case \vma\,\of \Nothing \to \return \Nothing_\M\\
-  &\hspace{6.5em}         \Just\,a \hspace{1.15em}\to \return (\Just_\M\,a)\\
+closures. In the following we present only $\StateT$ and $\ReaderT$. Starting
+with $\StateT$, we assume $\StateT_\M$ as the standard meta-level definition.
+The object-level $\StateT$ has type $(S : \VTy)(F : \VTy \to \VTy)(A : \VTy) \to
+\VTy$; the state parameter $S$ has to be a value type, since it is an input to
+an object-level function.
+\begin{alignat*}{3}
+  &\instance \ImproveF\,F\,M \RA \Improve\,(\StateT\,S\,F\,A)\,(\StateT_\M\,(\Up S)\,M\,(\Up A))\,\where\\
+  &\ind \mup\,x = \StateT_\M \dlr \lam s.\,\mdo\\
+  &\ind\ind \mit{as} \fro \mup\,\qt{\runStateT\,\spl x\,\spl s}\\
+  &\ind\ind \case \mit{as}\,\,\of (a,\,s) \to \return (a,\,s)\\
+  &\ind \mdown\,x = \qt{ \StateT\,(\lam s.\, \spl(\mdown \dlr \mdo\\\
+  &\ind \ind (a,\,s) \fro \runStateT_\M\,x\,\qt{s}\\
+  &\ind \ind \return \qt{(\spl a,\, \spl s)}))}
+\end{alignat*}
+Like before in $\MaybeT$, we rely on object-level case splitting in the
+definition of $\mup$. For $\Reader$, the environment parameter also has to be a
+value type, and we define improvement as follows.
+\begin{alignat*}{3}
+  &\instance \ImproveF\,F\,M \RA \Improve\,(\ReaderT\,R\,F\,A)\,(\ReaderT_\M\,(\Up R)\,M\,(\Up A))\,\where\\
+  &\ind \mup\,x   = \ReaderT_\M \dlr \lam r.\, \mup\,\qt{\runReaderT\,\spl x\,\spl r}\\
+  &\ind \mdown\,x = \qt{\ReaderT\,(\lam r.\,\spl(\mdown\,(\runReaderT_M\,x\,\qt{r})))}
+\end{alignat*}
+
+\subsubsection{$\State$ and $\Reader$ operations} Now, if we try to use
+the $\modify$ function that we already have for $\State_\M$, a curious thing
+happens. The meaning of $\modify\,(\lam x.\,\qt{\spl x + \spl x})$ is to replace
+the current state $x$, as an object expression, with the expression $\qt{\spl x
+  + \spl x}$, and this happens at staging time. This behaves as an ``inline''
+modification which replaces every subsequent mention of the state with a different
+expression. For instance, ignoring newtype wrappers for now,
+\[ \mdown \dlr \mdo \{\modify\,(\lam x.\,\qt{\spl x + \spl x}); \modify\,(\lam x.\,\qt{\spl x + \spl x});\return\,\qt{()}\} \]
+is staged to
+\[ \qt{\lam x.\,((),\,(x + x) + (x + x))} \]
+which duplicates the evaluation of $x + x$. The solution is to force the
+evaluation of the new state in the object language, by let-insertion. A similar
+phenomenon happens with the $\mit{local}$ function in $\Reader$. So we define
+``stricter'' versions of these operations. We also return $\Up ()$ from
+actions instead of $()$ --- the former is more convenient, because it can
+be immediately lowered with $\mdown$.
+\begingroup
+\allowdisplaybreaks
+\begin{alignat*}{3}
+  & \mput' : (\MonadState\,(\Up S)\,M,\,\MonadGen\,M) \RA \Up S \to M\,(\Up ()) \\
+  & \mput'\,s = \mdo \{s \fro \gen\,s; \mput\,s; \return \qt{()}\}\\
   &\\
-  &\ind \mdown\,(\MaybeT_\M\,x) = \qt{\MaybeT\,\spl(\mdown \dlr x \bind \lambda\case\\
-  &\ind\ind\Nothing_\M \to \return \qt{\Nothing}\\
-  &\ind\ind\Just_\M\,a\hspace{1.15em} \to \return \qt{\Just\,\spl a})}
-\end{alignat*}
-
-%% instance (Improve m n) => Improve (StateT s m) (StateT (Up s) n) where
-
-%%   up x = StateT \s ->
-%%     do as <- up [|| runStateT $$x $$s ||]
-%%        caseGen as pure
-
-%%   down x = [|| StateT \s -> $$(down
-%%     do (a, s) <- runStateT x [||s||]
-%%        pure [||($$a, $$s)||]
-%%        )||]
-
-%% instance Improve m n => Improve (MaybeT m) (MaybeT n) where
-%%   up x = MaybeT do
-%%     ma <- up [||runMaybeT $$x||]
-%%     caseGen ma pure
-
-%%   down (MaybeT x) =
-%%     [|| MaybeT $$(down (x >>= \case
-%%           Nothing -> pure [||Nothing||]
-%%           Just a -> pure [||Just $$a||])) ||]
-
-%% instance Improve m n => Improve (ReaderT r m) (ReaderT (Up r) n) where
-%%   up   x = ReaderT \r -> up [||runReaderT $$x $$r||]
-%%   down x = [|| ReaderT \r -> $$(down (runReaderT x [||r||])) ||]
+  & \modify' : (\MonadState\,(\Up S)\,M,\,\MonadGen\,M) \RA (\Up S \to \Up S) \to M\,(\Up ()) \\
+  & \modify'\,f = \mdo \{s \fro \get; \mput'\,(f\,s)\}\\
+  &\\
+  & \mit{local'} : (\MonadReader\,(\Up R)\,M,\,\MonadGen\,M) \RA (\Up R \to \Up R) \to M\,A \to M\,A\\
+  & \mit{local'}\,f\,\vma = \mdo \{r \fro \mit{ask}; r \fro \gen\,(f\,r);\mit{local}\,(\lam \_.\,r)\,\vma\}
+\end{alignat*}
+\endgroup
+Now,
+\[ \mdown \dlr \mdo \{\modify'\,(\lam x.\,\qt{\spl x + \spl x}); \modify'\,(\lam x.\,\qt{\spl x + \spl x})\} \]
+is staged to
+\[ \qt{\lam x.\,\letdef x := x + x; \letdef x := x + x; ((),\,x)} \]
+
+\subsection{Joining Control Flow in Monads}\label{sec:joining-control-flow-in-monads}
+
+There is a deficiency in our library so far. Assuming $b : \Bool$, consider:
+\begin{alignat*}{3}
+  & \mdown \dlr \mdo \\
+  & \ind \case b\,\of\\
+  & \ind \ind \True  \hspace{0.2em}\to \mput' \qt{10}\\
+  & \ind \ind \False \to \mput' \qt{20}\\
+  & \ind \modify' (\lam x. \qt{\spl x + 10})
+\end{alignat*}
+The $\modify' (\lam x. \qt{\spl x + 10})$ action gets inlined in both case
+branches during staging! This follows from the definition of monadic binding in
+$\Gen$ and the $\mit{split}$ function in the desugaring of $\mbf{case}$. Code
+generation is continued in two branches with the same action. If we have
+multiple $\mbf{case}$ splits sequenced after each other, that yields exponential
+code size. Right now, we can fix this by let-binding the problematic action:
+\begin{alignat*}{3}
+  & \mdown \dlr \mdo \\
+  & \ind x \fro \gen \dlr \mdown \dlr \case b\,\of\\
+  & \ind \ind \True  \hspace{0.2em}\to \mput' \qt{10}\\
+  & \ind \ind \False \to \mput' \qt{20}\\
+  & \ind \mup\,x\\
+  & \ind \modify' (\lam x. \qt{\spl x + 10})
+\end{alignat*}
+This removes code duplication by round-tripping through an object-level
+$\mbf{let}$. However, while this solution is fairly good in a state monad, where
+we can reasonably expect that product types can be unboxed by the object
+compiler, it introduces unnecessary runtime constructors for $\Maybe$ and
+other monads containing sum types. For $\Maybe$, $\mdown$ introduces a runtime
+$\Just$ or $\Nothing$, and $\mup$ introduces a runtime case split. A better
+solution would be to introduce two let-bound \emph{join points} before the
+offending $\mbf{case}$, one for returning a $\Just$ and one for returning
+$\Nothing$, but fusing away the actual runtime constructors. So, we would like
+to produce object code for a $\mbf{case}$ which looks like the following:
+\begin{alignat*}{3}
+  & \letdef\,\mit{joinJust}\,n := ...\\
+  & \letdef\,\mit{joinNothing}\,() := ...\\
+  & \case x == 10\,\of\\
+  & \ind \True \hspace{0.2em}\to \mit{joinJust}\,(x + 10) \\
+  & \ind \False \to \mit{joinNothing}\,()
+\end{alignat*}
+Such fused returns are possible whenever we have a $\Gen\,A$ action at the
+bottom of the transformer stack, such that $A$ is isomorphic to a finite
+meta-level sum of value types. Recall that $\Gen\,A$ is defined as $\{R : \VTy\}
+\to (A \to \Up R) \to \Up R$. Here, the $A \to \Up R$ continuation represents
+all possible code points where we can return to, and if $A$ is a finite sum, we
+can rearrange $A \to \Up R$ to a finite product of functions. Hence, we need a modest
+amount of generic programming with sums of values and products of computations.
+\begin{alignat*}{3}
+  & \data \SumVS : \List\,\VTy \to \MTy\,\where\,...\\
+  & \data \ProdCS : \List\,\CTy \to \MTy\,\where\,...
+\end{alignat*}
+%% \begin{alignat*}{3}
+%%   & \mathrlap{\data \SumVS : \List\,\VTy \to \MTy\,\where}\\
+%%   & \ind \Here &&: \Lift A \to \SumVS\,(\Cons\,A\,\mAs) \\
+%%   & \ind \There &&: \SumVS\,\mAs \to \SumVS\,(\Cons\,A\,\mAs)\\
+%%   & \mathrlap{\data \ProdCS : \List\,\CTy \to \MTy\,\where}\\
+%%   & \ind \Nil  &&: \ProdCS\,\Nil \\
+%%   & \ind \Cons &&: \Lift A \to \ProdCS\,\mAs \to \SumVS\,(\Cons\,A\,\mAs)
+%% \end{alignat*}
+%% Here, we overload $\List$ to refer to meta-level lists, dropping the $_\M$
+%% subscripts. We also overload the list constructors for $\ProdCS$.
+We use a class for types that have a $\SumVS$ representation:
+\begin{alignat*}{3}
+  & \hspace{-3em}\mathrlap{\tyclass\,\IsSumVS\,(A : \MTy)\,\where} \\
+  & \hspace{-3em}\ind \Rep    &&: \List\,\VTy \\
+  & \hspace{-3em}\ind \encode &&: A \to \SumVS\,\Rep \\
+  & \hspace{-3em}\ind \decode &&: \SumVS\,\Rep \to A
+\end{alignat*}
+We also define the mentioned isomorphism for $\Gen$ continuations.
+\begin{alignat*}{3}
+  &\mtabulate &&: (\SumVS\,\mAs \to \Lift B) \to \ProdCS\,(\map\,(\lam\,A.\,A\to B)\,\mAs)\\
+  &\mindex    &&: \ProdCS\,(\mit{map}\,(\lam\,A.\,A\to B)\,\mAs) \to (\SumVS\,\mAs \to \Lift B)
+\end{alignat*}
+We omit the definitions of $\mtabulate$, $\mindex$ and the $\IsSumVS$ instances,
+since they amount to standard generic programming; see e.g. \cite{TODO}. The
+last basic definition we need is to let-bind all computations in a generic
+product:
+\begin{alignat*}{3}
+  & \mathrlap{\mit{genProdCS} : \ProdCS\,\mAs \to \Gen\,(\ProdCS\,\mAs)}\\
+  & \mit{genProdCS}\,\Nil                 &&= \return\,\Nil\\
+  & \mit{genProdCS}\,(\Cons\,c\,\mit{cs}) &&= \Cons \fmap \gen\,c \ap \mit{genProdCS}\,\mit{cs}
+\end{alignat*}
+We introduce a class for monads that support control flow joining.
+\[ \tyclass\,\MonadJoin\,M\,\where\,\join : \IsSumVS\,A \RA M\,A \to M\,A \]
+The most interesting instance is for $\Gen$.
+\begin{alignat*}{3}
+  &\instance \MonadJoin\,\Gen\,\where\\
+  &\ind \join\,\vma = \Gen \dlr \lam k.\,\runGen \dlr \mdo\\
+  &\ind \ind \mit{joinpoints} \fro \mit{genProdCS}\,(\mtabulate\,(k \circ \decode))\\
+  &\ind \ind a \fro \vma\\
+  &\ind \ind \return \dlr \mindex\,\mit{joinpoints}\,(\encode\,a)
+\end{alignat*}
+Here we first convert $k : A \to \Lift R$ to a product of join points and
+let-bind each of them. Then we adjust the return code points of $\vma$ by
+looking up the corresponding continuation from the list of join points, using
+$\mindex$. We also need $\MonadJoin$ instances for the other transformers in our
+library, but these are all essentially just lifting $\join$. In $\StateT_\M$ we
+must have $\IsSumVS$ for $S$ because $S$ appears in the transformer stack's
+return value.
+\begin{alignat*}{3}
+  & \instance\,(\MonadJoin\,M) \RA \MonadJoin\,(\MaybeT_\M\,M)\,\where\\
+  & \ind \join\,(\MaybeT_\M\,\vma) = \MaybeT_\M\,(\join\,\vma)\\
+  & \instance\,(\MonadJoin\,M) \RA \MonadJoin\,(\ReaderT_\M\,R\,M)\,\where\\
+  & \ind \join\,(\ReaderT_\M\,\vma) = \ReaderT_\M\,(\join \circ \vma)\\
+  & \instance\,(\MonadJoin\,M,\,\IsSumVS\,S) \RA \MonadJoin\,(\StateT_\M\,S\,M)\,\where\\
+  & \ind \join\,(\StateT_\M\,\vma) = \StateT_\M\,(\join \circ \vma)
+\end{alignat*}
+Now, whenever the return value of a $\mbf{case}$-splitting action is a sum of
+values (this includes just returning an object value, which is by far the most
+common situation), we can use $\join \dlr \case x\,\of ...$ to block inlining
+into the branches, without creating any runtime sum types.
+
+\todo{bigger example}
+
+\subsection{Discussion}
+
+So far, we have a monad transformer library with the following features:
+\begin{itemize}
+\item Almost all definitions from the well-known Haskell ecosystem of monads and monad transformers
+      can be directly reused, in the meta level.
+\item We can quite liberally pattern match on object-level values in monadic code, insert object-level $\mbf{let}$-s
+      with $\gen$ and control code size by $\join$-s.
+\item In monadic code, object-level data constructors are only ever created by $\mdown$, and switching on object-level data is only
+      created by $\mit{split}$ and $\mup$. Everything else is fully fused, and all function calls can be compiled to
+      statically known saturated calls.
+\end{itemize}
+
+As to potential weaknesses, first, the system as described in this section has
+quite a bit of syntactic noise and requires extra attention from programmers.
+We believe that the code noise can be mitigated very effectively in a native
+CFTT implementation. Kovács \cite{TODO} demonstrated in a prototype elaborator
+that almost all quotes and splices are unambiguously inferable in 2LTT programs,
+if we make the concession that stages of let-definitions are always specified
+(as we do here). Moreover, $\mup$ and $\mdown$ should be also effectively
+inferable, using bidirectional elaboration, but it would make sense to limit
+inferred binding-time improvement to monads only. With stage inference and
+binding-time improvement inference, code in CFTT would look only modestly more
+complicated than in Haskell, and we would get far more robust and controllable
+code generation.
+
+Second, in CFTT we cannot store most monadic actions in runtime data structures,
+those that are implemented using functions, nor can we have functions in
+$\State$ state, or in $\Reader$ environments. However:
+\begin{itemize}
+  \item Essential usage of this is rare in practice.
+  \item It is possible to extend CFTT with a closure type former, converting computations to values,
+        in which case there is no such limitation anymore. Here, closure-freedom is still available;
+        we can choose to use or abstain from the explicit closure type.
+\end{itemize}
+
+\subsection{Haskell Implementation}
+
+We implemented all features in this section in typed Template Haskell
+\cite{TODO}, using only the strongly typed interface. The organization of the
+code is essentially the same as here, but some changes are necessary:
+\begin{itemize}
+  \item TH cannot properly handle scoping of type variables across quotations,
+        generating ill-typed code in some cases. It is fairly easy to work around
+        this using term-level proxy values.
+  \item We do not bother tracking value and computation types. This could be done,
+        using dummy type classes, but we would not get any extra guarantees from it
+        in downstream compilation.
+  \item When we take the product of generic sums in CFTT, we simply take the
+        object-level product of component types, pairwise. We do this because we
+        assume that the downstream compiler can represent products without
+        overheads, in an unboxed manner, and also erase unit types. In Haskell we do
+        not rely on this, and instead use a slightly more complicated
+        sum-of-products-of-values representation, and also fuse away product types
+        during staging.
+  \item We make heavy use of \emph{singleton types} \cite{TODO} in generic sums-of-products, to emulate
+        dependent types. This adds significant noise compared to a native dependently typed
+        version.
+  \item We have two separate type classes for binding-time improvement, one for
+        base types and one for monads specifically. This is because the quantified
+        constraints in \ref{TODO} yield poor type inference for $\mup$ and $\mdown$,
+        by preventing \emph{functional dependencies} \cite{TODO} between the class
+        parameters. We expect that in a native CFTT implementation, the extra staging
+        information in type universes would help inference in this case.
+\end{itemize}
+
+\section{Stream Fusion}
+
+Stream fusion refers to a collection of techniques for generating efficient code
+from declarative definitions involving streams of values, where intermediate
+data structures are eliminated. For example, mapping twice over a list produces
+an intermediate list in a naive implementation but does not do so in a fused
+implementation.
+
+Fusion techniques can be broadly grouped into \emph{push} fusion, which is based
+on Church-encodings of inductive lists, and $\emph{pull}$ fusion, which is based
+on Church-encodings of coinductive lists, which manifests as \emph{state
+machines}. The two styles have different trade-offs, and in practical
+programming it is a good idea to support both, but in this section we focus on
+pull fusion.
+
+The reason is partly lack of space, but also that pull fusion has been
+historically the more difficult to reliably compile, and we can demonstrate
+significant improvements in CFTT, in robustness and expressive power. Our fusion
+always goes through, and we also support a $\mit{concatMap}$ operation, which
+has been elusive in guaranteed fusion. We believe that our implementation also
+provides valuable conceptual clarity.
+
+\subsection{Streams}
+
+A pull stream is a meta-level specification of a state machine:
+\begin{alignat*}{3}
+  & \data \Step\,S\,A = \Stop\,|\,\Yield\,A\,S\,|\,\Skip\,S\\
+  & \data \Pull\,(A : \MTy) : \MTy\,\where\\
+  & \ind \MkPull : (S : \MTy) \to \IsSumVS\,S \RA S \to (S \to \Gen\,(\Step\,S\,A)) \to \Pull\,A
+\end{alignat*}
+In $\MkPull$, $S$ is the type of the internal state which is required to be a
+finite sum of values by the $\IsSumVS\,S$ constraint. The next $S$-typed field
+is the initial state, while $S \to \Gen\,(\Step\,S\,A)$ represents all
+transitions. Possible transitions are stopping ($\Stop$), transitioning to a
+new state while outputting a value ($\Yield$) and making a silent transition to
+a new state ($\Skip$).
 

### 2024-02-13 newpaper/hs-agda/hs/SOP.hs

@@ -1,0 +1,255 @@
+
+{-# language
+  DataKinds, PolyKinds, GADTs, TypeFamilies, TypeApplications,
+  ScopedTypeVariables, TypeOperators, AllowAmbiguousTypes,
+  TypeFamilyDependencies, UndecidableInstances, BlockArguments,
+  DefaultSignatures, TemplateHaskell
+  #-}
+
+{-# options_ghc -Wincomplete-patterns #-}
+
+module SOP where
+
+import Data.Kind
+import Up
+
+--------------------------------------------------------------------------------
+
+data family Sing (a :: k)
+
+data instance Sing (a :: Type) where
+  SType :: Sing a
+
+data instance Sing (a :: [k]) where
+  SNil  :: Sing '[]
+  SCons :: forall x xs. Sing x -> Sing xs -> Sing (x ': xs)
+
+class                           SingI a           where sing :: Sing a
+instance                        SingI (a :: Type) where sing = SType
+instance                        SingI '[]         where sing = SNil
+instance (SingI a, SingI as) => SingI (a ': as)   where sing = SCons sing sing
+
+type family (++) (xs :: [k]) (ys :: [k]) :: [k] where
+  (++) '[]       ys = ys
+  (++) (x ': xs) ys = x ': (++) xs ys
+
+infixr 5 %++
+(%++) :: Sing xs -> Sing ys -> Sing (xs ++ ys)
+(%++) SNil         ys = ys
+(%++) (SCons x xs) ys = SCons x (xs %++ ys)
+
+data Id :: forall k. k -> k -> Type where
+  Refl :: forall k (x :: k). Id x x
+
+--------------------------------------------------------------------------------
+
+data P :: [Type] -> Type where
+  Nil  :: P '[]
+  Cons :: forall a as. Up a -> P as -> P (a ': as)
+
+data S :: [[Type]] -> Type where
+  Here  :: forall as ass. P as -> S (as ': ass)
+  There :: forall as ass. S ass -> S (as ': ass)
+
+-- ProdP
+--------------------------------------------------------------------------------
+
+type ProdP :: [Type] -> [Type] -> [Type]
+type ProdP as bs = as ++ bs
+
+sProdP :: Sing as -> Sing bs -> Sing (ProdP as bs)
+sProdP = (%++)
+
[…]

### 2024-02-13 newpaper/hs-agda/hs/ixedsop/SOP.hs

@@ -1,0 +1,284 @@
+
+{-# language
+  DataKinds, PolyKinds, GADTs, TypeFamilies, TypeApplications,
+  ScopedTypeVariables, TypeOperators, AllowAmbiguousTypes,
+  TypeFamilyDependencies, UndecidableInstances, BlockArguments,
+  DefaultSignatures, TemplateHaskell
+  #-}
+
+{-# options_ghc -Wincomplete-patterns #-}
+
+module SOP where
+
+import Data.Kind
+import Up
+
+--------------------------------------------------------------------------------
+
+data family Sing (a :: k)
+
+data instance Sing (a :: Type) where
+  SType :: Sing a
+
+data instance Sing (a :: [k]) where
+  SNil  :: Sing '[]
+  SCons :: forall x xs. Sing x -> Sing xs -> Sing (x ': xs)
+
+class                           SingI a           where sing :: Sing a
+instance                        SingI (a :: Type) where sing = SType
+instance                        SingI '[]         where sing = SNil
+instance (SingI a, SingI as) => SingI (a ': as)   where sing = SCons sing sing
+
+data Nat = Zero | Suc Nat
+
+data instance Sing (n :: Nat) where
+  SZero :: Sing 'Zero
+  SSuc  :: Sing n -> Sing ('Suc n)
+
+instance SingI Zero               where sing = SZero
+instance SingI n => SingI (Suc n) where sing = SSuc sing
+
+data Vec (a :: Type) :: Nat -> Type where
+  VNil  :: Vec a Zero
+  VCons :: a -> Vec a n -> Vec a (Suc n)
+
+data instance Sing (v :: Vec a n) where
+  SVNil :: Sing VNil
+  SVCons :: Sing (x :: a) -> Sing (v :: Vec a n) -> Sing (VCons x v)
+
+instance SingI VNil where
+  sing = SVNil
+instance (SingI (x :: a), SingI (v :: Vec a n)) => SingI (VCons x v) where
+  sing = SVCons sing sing
+
+data Fin (n :: Nat) where
+  FZero :: Fin (Suc n)
+  FSuc  :: Fin n -> Fin (Suc n)
+
+type Ix :: forall (a :: Type)(n :: Nat)(as :: Vec a n)(i :: Fin n) -> a
+type family Ix a n as i where
+  Ix a (Suc n) (VCons x xs) FZero    = x
[…]

### 2024-02-20 newpaper/paper.tex
wip

--- 2024-02-09 newpaper/paper.tex
@@
 \usepackage{mathtools}
-%% \usepackage{amssymb}
-
-\newcommand{\mit}[1]{\mathit{#1}}
-\newcommand{\msf}[1]{\mathsf{#1}}
-\newcommand{\mbf}[1]{\mathbf{#1}}
+
+\newcommand{\mit}[1]{{\mathsf{#1}}}
+\newcommand{\msf}[1]{{\mathsf{#1}}}
+\newcommand{\mbf}[1]{{\mathbf{#1}}}
 \newcommand{\mbb}[1]{\mathbb{#1}}
@@
 \newcommand{\return}{\mbf{return}\,}
+\newcommand{\pure}{\mbf{pure}\,}
 \newcommand{\lam}{\lambda\,}
@@
 \newcommand{\map}{\mit{map}}
-\newcommand{\vas}{\mit{as}}
-\newcommand{\vbs}{\mit{bs}}
-\newcommand{\vxs}{\mit{xs}}
-\newcommand{\vys}{\mit{ys}}
-\newcommand{\vma}{\mit{ma}}
-\newcommand{\vmb}{\mit{mb}}
+
+\newcommand{\vas}{\mathsf{as}}
+\newcommand{\vbs}{\mathsf{bs}}
+\newcommand{\vcs}{\mathsf{cs}}
+\newcommand{\vxs}{\mathsf{xs}}
+\newcommand{\vys}{\mathsf{ys}}
+\newcommand{\vma}{\mathsf{ma}}
+\newcommand{\vm}{\mathsf{m}}
+\newcommand{\vn}{\mathsf{n}}
+\newcommand{\vk}{\mathsf{k}}
+\newcommand{\vA}{\mathsf{A}}
+\newcommand{\vB}{\mathsf{B}}
+\newcommand{\vC}{\mathsf{C}}
+\newcommand{\vS}{\mathsf{S}}
+\newcommand{\vF}{\mathsf{F}}
+\newcommand{\vR}{\mathsf{R}}
+\newcommand{\vM}{\mathsf{M}}
+\newcommand{\vmb}{\mathsf{mb}}
+\newcommand{\mAs}{\mathsf{As}}
+\newcommand{\va}{\mathsf{a}}
+\newcommand{\vb}{\mathsf{b}}
+\newcommand{\vc}{\mathsf{c}}
+\newcommand{\vd}{\mathsf{d}}
+\newcommand{\vx}{\mathsf{x}}
+\newcommand{\vy}{\mathsf{y}}
+\newcommand{\vz}{\mathsf{z}}
+\newcommand{\vf}{\mathsf{f}}
+\newcommand{\vfs}{\mathsf{fs}}
+\newcommand{\vg}{\mathsf{g}}
+\newcommand{\vh}{\mathsf{h}}
+\newcommand{\vt}{\mathsf{t}}
+\newcommand{\vs}{\mathsf{s}}
+\newcommand{\vr}{\mathsf{r}}
+\newcommand{\vu}{\mathsf{u}}
+
+\newcommand{\SOP}{\msf{SOP}}
+\newcommand{\El}{\msf{El}}
+\newcommand{\USOP}{\msf{U}_{\msf{SOP}}}
+\newcommand{\Uprod}{\msf{U_P}}
+\newcommand{\Elprod}{\msf{El_{P}}}
+\newcommand{\IsSOP}{\msf{IsSOP}}
+\newcommand{\forEach}{\msf{forEach}}
+\newcommand{\single}{\msf{single}}
+\newcommand{\msplit}{\msf{split}}
+\newcommand{\mapGen}{\msf{mapGen}}
 
@@
 \newcommand{\Monad}{\msf{Monad}}
-\newcommand{\Applicative}{\msf{Monad}}
+\newcommand{\Applicative}{\msf{Applicative}}
 \newcommand{\class}{\msf{class}}
@@
 \newcommand{\case}{\mbf{case\,}}
+\newcommand{\foldr}{\msf{foldr}}
+\newcommand{\foldl}{\msf{foldl}}
+\newcommand{\rep}{\msf{rep}}
+\newcommand{\concatMap}{\msf{concatMap}}
 
@@
 \newtheorem{notation}{Notation}
+\newtheorem*{axiom}{Axiom}
+
 \newcommand{\id}{\mit{id}}
@@
 \newcommand{\fmap}{<\!\!\$\!\!>}
-\newcommand{\ap}{<\!\!*\!\!>}
+\newcommand{\ap}{{<\!\!*\!\!>}}
 \newcommand{\runGen}{\mit{runGen}}
@@
 \newcommand{\MonadGen}{\msf{MonadGen}}
+\newcommand{\MonadState}{\msf{MonadState}}
+\newcommand{\MonadReader}{\msf{MonadReader}}
 \newcommand{\RA}{\Rightarrow}
 \newcommand{\EitherT}{\msf{EitherT}}
+\newcommand{\Either}{\msf{Either}}
+\newcommand{\Left}{\msf{Left}}
+\newcommand{\Right}{\msf{Right}}
 \newcommand{\StateT}{\msf{StateT}}
 \newcommand{\Identity}{\msf{Identity}}
+
+\newcommand{\Stop}{\msf{Stop}}
+\newcommand{\Skip}{\msf{Skip}}
+\newcommand{\Yield}{\msf{Yield}}
+
 \newcommand{\runIdentity}{\mit{runIdentity}}
+\newcommand{\runReaderT}{\mit{runReaderT}}
 \newcommand{\newtype}{\mbf{newtype}\,}
 \newcommand{\runMaybeT}{\mit{runMaybeT}}
+\newcommand{\runStateT}{\mit{runStateT}}
+\newcommand{\runState}{\mit{runState}}
 \newcommand{\dlr}{\,\$\,}
@@
 \newcommand{\ExceptT}{\msf{ExceptT}}
+\newcommand{\State}{\msf{State}}
+\newcommand{\SumVS}{\msf{SumVS}}
+\newcommand{\ProdCS}{\msf{ProdCS}}
+\newcommand{\Here}{\msf{Here}}
+\newcommand{\There}{\msf{There}}
+\newcommand{\IsSumVS}{\msf{IsSumVS}}
+\newcommand{\MonadJoin}{\msf{MonadJoin}}
+\newcommand{\Stream}{\msf{Stream}}
+\newcommand{\join}{\mit{join}}
+\newcommand{\modify}{\mit{modify}}
+\newcommand{\get}{\mit{get}}
+\newcommand{\mput}{\mit{put}}
+\newcommand{\Rep}{\mit{Rep}}
+\newcommand{\encode}{\mit{encode}}
+\newcommand{\decode}{\mit{decode}}
+\newcommand{\mindex}{\mit{index}}
+\newcommand{\mtabulate}{\mit{tabulate}}
+\newcommand{\States}{\mit{States}}
+\newcommand{\seed}{\mit{seed}}
+\newcommand{\step}{\mit{step}}
+\newcommand{\Step}{\msf{Step}}
+\newcommand{\Pull}{\msf{Pull}}
+\newcommand{\MkPull}{\msf{MkPull}}
 
@@
 \begin{abstract}
-There are many abstraction tools in modern functional programming which heavily
-rely on general-purpose compiler optimization to achieve adequate
-performance. For example, monadic binding is a higher-order function which
-yields runtime closures in the absence of sufficient compile-time inlining and
-beta-reductions, thereby significantly degrading performance. In current systems
-such as the Glasgow Haskell Compiler, there is no strong guarantee that
-general-purpose optimization can eliminate abstraction overheads, and users only
-have indirect and fragile control over code generation through inlining
-directives and compiler options. In this paper we propose using a two-stage
-language to simultaneously get strong code generation guarantees and strong
-abstraction features. The object language is a simply typed first-order language
-where all function calls are statically known, and which can be compiled without
-runtime closures. The compile-time language is a dependent type theory. The two
-are integrated in a two-level type theory. We develop some abstraction
-tools in this setting. First, we develop monads and monad transformers. Second,
-we develop fusion for push and pull streams. Most of our results are also
-adapted to a proof-of-concept library in typed Template Haskell.
+Many abstraction tools in functional programming rely heavily on general-purpose
+compiler optimization to achieve adequate performance. For example, monadic
+binding is a higher-order function which yields runtime closures in the absence
+of sufficient compile-time inlining and beta-reductions, thereby significantly
+degrading performance. In current systems such as the Glasgow Haskell Compiler,
+there is no strong guarantee that general-purpose optimization can eliminate
+abstraction overheads, and users only have indirect and fragile control over
+code generation through inlining directives and compiler options. We propose a
+two-stage language to simultaneously get strong guarantees about code generation
+and strong abstraction features. The object language is a simply-typed
+first-order language which can be compiled without runtime closures. The
+compile-time language is a dependent type theory. The two are integrated in a
+two-level type theory.
+
+We demonstrate two applications of the system. First, we develop monads and
+monad transformers. Here, abstraction overheads are eliminated by staging and we
+can reuse almost all definitions from the existing Haskell ecosystem. Second,
+we develop pull-based stream fusion. Here we make essential use of dependent
+types to give a concise definition of a $\mathsf{concatMap}$ operation with
+guaranteed fusion. We provide an Agda implementation and a typed Template
+Haskell implementation of these developments.
+
 \end{abstract}
@@
 in terms of implementation, yields large overheads when compiled without
-optimizations. Consider the following snippet.
-\begin{alignat*}{3}
-  &f :: \Int \to \Reader\,\Bool\,\Int \\
-  &f\,x = \mdo\{b \fro \msf{ask}; \bif x\, \bthen \return (x + 10)\, \belse \return (x + 20)\}
-\end{alignat*}
-With optimizations enabled, GHC compiles this roughly to the following:
-\begin{alignat*}{3}
-  &f :: \Int \to \Bool \to \Int \\
-  &f = \lam x\,b.\, \bif b\, \bthen x + 10\, \belse x + 20
+optimizations. Consider the following:
+\begin{alignat*}{3}
+  &\vf :: \Int \to \Reader\,\Bool\,\Int \\
+  &\vf\,\vx = \mdo\{\vb \fro \msf{ask}; \bif \vx\, \bthen \return (\vx + 10)\, \belse \return (\vx + 20)\}
+\end{alignat*}
+With optimizations enabled, GHC compiles this roughly to the code below:
+\begin{alignat*}{3}
+  &\vf :: \Int \to \Bool \to \Int \\
+  &\vf = \lam \vx\,\vb.\, \bif \vb\, \bthen \vx + 10\, \belse \vx + 20
 \end{alignat*}
@@
 \begin{alignat*}{3}
-  &f = \lam x.\,(\bind)\,\msf{MonadReaderDict}\,\msf{ask}\,(\lam b.\,\bif b\\
-  &\ind \bthen\return\,\msf{MonadReaderDict}\,(x + 10)\\
-  &\ind \belse\hspace{0.25em}\return\,\msf{MonadReaderDict}\,(x + 20))
+  &\vf = \lam \vx.\,(\bind)\,\msf{MonadReaderDict}\,\msf{ask}\,(\lam \vb.\,\bif \vb\\
+  &\ind \bthen\return\,\msf{MonadReaderDict}\,(\vx + 10)\\
+  &\ind \belse\hspace{0.25em}\return\,\msf{MonadReaderDict}\,(\vx + 20))
 \end{alignat*}
@@
 the $\Monad$ instance for $\Reader$, and $(\bind)\,\msf{MonadReaderDict}$ is a
-field projection. Here, a runtime closure will be created for the $\lam b.\,...$
+field projection. Here, a runtime closure will be created for the $\lam \vb.\,...$
 function, and $(\bind)$, $\msf{ask}$ and $\return$ will create additional dynamic
@@
 polymorphic over monads. In Haskell, such code is pervasive, even in fairly
-basic programs which do not use fancy abstractions. The $\msf{mapM}$ function in
-the Haskell Prelude, which maps over a list in some monad, is a third-order and
-second-rank polymorphic function in disguise, because its monad dictionary
-argument contains the polymorphic second-order $(\bind)$ method.
-\begin{alignat*}{3}
-  & \msf{mapM} :: \Monad\,m \Rightarrow (a \to m\,b) \to [a] \to m\,[b]
-\end{alignat*}
-Compiling $\msf{mapM}$ efficiently relies on inlining the instance dictionary,
-then inlining the methods contained there, and also inlining the functions
-that the higher-order binding method is applied to.
+basic programs which do not use fancy abstractions. Consider the $\msf{mapM}$
+function from the Haskell Prelude:
+\begin{alignat*}{3}
+  & \msf{mapM} :: \Monad\,\vm \Rightarrow (\va \to \vm\,\vb) \to [\va] \to \vm\,[\vb]
+\end{alignat*}
+This is a third-order and second-rank polymorphic function in disguise, because
+its monad dictionary argument contains the polymorphic second-order $(\bind)$
+method. Compiling $\msf{mapM}$ efficiently relies on inlining the instance
+dictionary, then inlining the methods contained there, and also inlining the
+functions that the higher-order binding method is applied to.
 
@@
 be used to control code generation, but without any strong guarantee, and their
-sophisticated usage requires knowledge of GHC internals. For example, correctly
+advanced usage requires knowledge of GHC internals. For example, correctly
 specifying the \emph{ordering} of certain rule applications is often needed. We
-have to also care about function arities. Infamously, the function composition
-operator is defined as $(.)\,f\,g = \lam x \to f\,(g\,x)$ in the base libraries,
-instead of as $(.)\,f\,g\,x = f\,(g\,x)$, to get better inlining behavior ---
-as explained in a source comment right next to the definition \cite{TODO}.
-
-Although there are numerous tricks and idioms that are used in high-performance
-Haskell programming, for reliable performance it is necessary that programmers
-periodically \emph{review} GHC's optimized code output. Needless to say, this is
-rather time-consuming and poorly scalable.
-
-\subsection{Staged Compilation}
+also have to care about formal function arities. Infamously, the function
+composition operator is defined as $(.)\,\vf\,\vg = \lam \vx \to \vf\,(\vg\,\vx)$ in the
+base libraries, instead of as $(.)\,\vf\,\vg\,\vx = \vf\,(\vg\,\vx)$, to get better inlining
+behavior \cite{TODO}. It remains a common practice in high-performance Haskell
+programming to manually review GHC's optimized code output.
+
+\subsection{Closure-Free Staged Compilation}
 
@@
 general-purpose optimizers are slower to run, less transparent and less
-reliable. Also, metaprogramming allows library authors to exploit
+robust. Also, metaprogramming allows library authors to exploit
 domain-specific optimizations, while it is not realistic for general-purpose
@@
 On the other hand, metaprogramming requires some additional care and input from
-programmers. Historically, there have been problems with ergonomics as well:
-\begin{itemize}
-\item Code generation might fail \emph{too late} in the pipeline, producing
-      incomprehensible errors; this is often caused by not having enough static
-      guarantees about the well-formedness of code output.
-\item Tooling for the object language (debugging, profiling, IDE support) often does not
-      work for metaprogramming. This is more likely if the object and meta layers are wildly
-      different.
-\item Metaprogramming may introduce heavy noise and boilerplate, obscuring the logic of
-      programs, or impose restrictions on how code can be structured. For instance,
-      Template Haskell mandates that metaprograms used in splices in some module are
-      defined in a different module.
-\end{itemize}
+programmers. Historically, there have been problems with ergonomics as well. In
+weakly-typed staged systems, code generation might fail \emph{too late} in the
+pipeline, producing incomprehensible errors. Or, tooling that works for an
+object language (like debugging, profiling, IDEs) may not work for metaprogramming, or
+metaprogramming may introduce heavy noise and boilerplate, obscuring the logic
+of programs and imposing restrictions on code structure.
 
@@
       finitary algebraic data types. This language is easy to optimize and compile
-      downstream in the pipeline, but it lacks many convenience features.
+      in the downstream pipeline, but it lacks many convenience features.
 \item A standard Martin-Löf type theory for the compile-time language. This
@@
 
-In particular, since the object language is first-order, we guarantee that all
-programs in CFTT can be ultimately compiled without any dynamic closures, using
-only calls and jumps to statically known code locations. Why focus on closures?
-They are the foundation to almost all abstraction tools in functional
-programming:
+Since the object language is first-order, we guarantee that all programs in CFTT
+can be ultimately compiled without any dynamic closures, using only calls and
+jumps to statically known code. Why focus on closures?  They are the
+foundation to almost all abstraction tools in functional programming:
 \begin{itemize}
@@
 
-It turns out that surprisingly little practical programming relies essentially
-on closures. Most of the time, programmers use higher-order functions for
+Perhaps surprisingly, little practical programming relies essentially on
+closures. Most of the time, programmers use higher-order functions for
 \emph{abstraction}, such as when mapping over lists, where it is expected that
-the mapping function will be inlined. In other cases, the use of closures can be
-eliminated by small-scale defunctionalization. For example, difference lists
-\cite{TODO} are often implemented with closures, but they can be also
-implemented as binary trees, with the same programming interface.
-
-\todo{Essential closures: CPS? Threaded interpreters?}
-
-Whole-program defunctionalization is also possible, and it is notably used by
-the MLton compiler \cite{TODO}. While this can be practical and effective, it
-can be also expensive and require whole-program processing. Also, conceptually
-speaking, it does not eliminate closures but instead makes them more transparent
-to optimizations. In this paper we do not use defunctionalization.
-
-We note though that our setup is compatible with closures as well, and it can
-support two separate type formers for closure-based and non-closure-based
-(``static'') functions. Having both of these would be desirable in a practical
-system. In the current work we focus on the closure-free case because it is much
-less known and developed, and it is quite interesting to see how far we
-can go with it.
+the mapping function will be inlined. We note though that our setup is
+compatible with closures as well, and it can support two separate type formers
+for closure-based and non-closure-based (``static'') functions. Having both of
+these would be desirable in a practical system. In the current work we focus on
+the closure-free case because it is much less known and developed, and it is
+interesting to see how far we can go with it.
+
+%% For example, difference lists
+%% \cite{TODO} are often implemented with closures, but they can be also
+%% implemented as binary trees, with the same programming interface.
+
+%% \todo{Essential closures: CPS? Threaded interpreters?}
+
+%% Whole-program defunctionalization is also possible, and it is notably used by
+%% the MLton compiler \cite{TODO}. While this can be practical and effective, it
+%% can be also expensive and require whole-program processing. Also, conceptually
+%% speaking, it does not eliminate closures but instead makes them more transparent
+%% to optimizations. In this paper we do not use defunctionalization.
 
@@
 
-\todo{CONTRIBUTIONS}
-
-\section{Basics of CFTT}\label{sec:basics-of-cftt}
-
-In the following we give an overview of CFTT features. Here we focus on examples
-and informal explanations. We defer the more formal details to Sections
-\ref{TODO} and \ref{TODO}. We first review the meta-level language, then the
-object-level one, and finally the staging operations which bridge between the
-two.
+\begin{itemize}
+\item In Section \ref{TODO} we present the two-level type theory CFTT, where the
+  object language is first-order simply-typed and the metalanguage is
+  dependently typed. The object language admits an operational semantics without
+  runtime closures, and can be compiled in a way such that all function calls
+  are statically known. We provide a supplementary Agda formalization of the
+  operational semantics.
+\item In Section \ref{TODO} we build a monad transformer library. We believe
+  that this is a good demonstration, because monads and monad transformers are
+  the most widely used effect system in Haskell, and at the same time their
+  compilation to efficient code can be surprisingly difficult. The main idea is
+  to have monads and monad transformers \emph{only} in the metalanguage, and try
+  to express as much as possible at the meta level, and funnel the object-meta
+  interactions through specific \emph{binding-time improvements}.  We show how
+  almost all definitions from the Haskell ecosystem of monad transformers can be
+  reused in the meta level of CFTT. We also develop let-insertion, join points
+  and case switching on object-level data, in the generality of monad transformers.
+\item In Section \ref{TODO} we build a pull-based stream fusion library. Here,
+  the motivation is to demonstrate essential usage of dependent types to
+  concisely solve a difficult problem, namely guaranteed fusion for unrestricted
+  combinations of $\concatMap$ and $\mit{zip}$. We use a state machine
+  representation that is based on \emph{sums-of-products} of object-level
+  values. We show that CFTT is compatible with a \emph{generativity} axiom,
+  which internalizes the fact that metaprograms cannot inspect the structure of
+  object-level terms. We use this to show that the universe of sums-of-products
+  is closed under $\Sigma$-types. This in turn enables a very
+  concise definition of $\concatMap$.
+\item
+  We adapt the contents of the paper to a Template Haskell, with some
+  modifications, simplifications and fewer guarantees about generated code.  In
+  particular, Haskell does not have enough dependent types for the simple
+  $\concatMap$ definition, but we can still work around this limitation. We also
+  provide a more faithful Agda embedding of CFTT and our libraries. Here, the
+  object theory is embedded as a collection of postulated operations, and we can
+  use Agda's normalization command to print out generated object code.
+\end{itemize}
+
+\section{Overview of CFTT}\label{sec:overview-of-cftt}
+
+In the following we give an overview of CFTT features. We focus on informal
+explanations and examples here. We first review the meta-level language, then
+the object-level one, and finally the staging operations which bridge between
+the two.
 
@@
 relevant to the current paper, so we shall omit levels. Note that universe
-levels have nothing to do with staged compilation; they are about sizing for the
-purpose of logical consistency.
-
-We use Agda-like syntax for functions and implicit arguments. A basic example:
-\begin{alignat*}{3}
-  &\id : \{A : \MTy\} \to A \to A\\
-  &\id = \lam x.\, x
+levels are orthogonal to staging; they are instead used to increase the strength
+of the system while maintaining logical consistency.
+
+Throughout this paper we use a mix of Agda and Haskell syntax for
+CFTT. Dependent functions and implicit arguments follow Agda. A basic example:
+\begin{alignat*}{3}
+  &\id : \{\vA : \MTy\} \to \vA \to \vA\\
+  &\id = \lam \vx.\, \vx
 \end{alignat*}
 Here, the type argument is implicit, and it gets inferred when we use the
-function. For example, $id\,\True$ is elaborated to $id\,\{\Bool\}\,\True$,
+function. For example, $\id\,\True$ is elaborated to $\id\,\{\Bool\}\,\True$,
 where the braces mark an explicit application for the implicit argument.
-Similarly, we can introduce implicit lambdas explicitly:
-\begin{alignat*}{3}
-  &id = \lam \{A : \MTy\}(x : A).\,x
-\end{alignat*}
-We write $(a : A) \times B$ for $\Sigma$-types in $\MTy$. For the field
-projections, for $t : (a : A) \times B$, we have $\fst\,t : A$ and $\snd\,t :
-B[a \mapsto \fst\,t]$, and pairing is written simply as $(t,\,u)$. Additionally,
-we use syntactic sugar for field projections: we can write a type as $(field_1 :
-A) \times (field_2 : B) \times (field_3 : C)$, and write $t.field_2$ for a named
-field projection. The type itself denotes a right-nested $\Sigma$-type. We can
-also write $(t,\,u,\,v)$ for a right-nested iterated pairing. In Haskell style,
-we write $()$ for the unit type, and also for its inhabitant.
-
-Inductive types can be introduced using a Haskell-like ADT notation, or with a GADT-style one:
+Inductive types can be introduced using a Haskell-like ADT notation, or with a
+GADT-style one:
 \begin{alignat*}{3}
   &                                           &&\hspace{4em}\data\,\Bool_\M &&: \MTy\,\where\\
-  & \data\,\Bool_\M : \MTy = \True\,|\,\False &&\hspace{4em}\ind\ind \True &&: \Bool_\M\\
-  &                                           &&\hspace{4em}\ind\ind \False &&: \Bool_\M
+  & \data\,\Bool_\M : \MTy = \True_\M\,|\,\False_\M &&\hspace{4em}\ind\ind \True_\M &&: \Bool_\M\\
+  &                                           &&\hspace{4em}\ind\ind \False_\M &&: \Bool_\M
 \end{alignat*}
@@
 defined both on the meta and object levels, we will sometimes use this subscript
-to disambiguate the meta-level version. We will also use Haskell-like newtype
-notation, such as in $\newtype \msf{Wrap}\,A = \msf{Wrap}\,\{\mit{unWrap} : A
-\}$.
-
-Importantly, we can only produce meta-level values by induction on meta-level
-data. In general, all construction and elimination rules for type formers in
-$\MTy$ stay within $\MTy$.
+to disambiguate the meta-level version.
+
+We use Haskell-like newtype notation, such as in $\newtype \msf{Wrap}\,\vA =
+\msf{Wrap}\,\{\mit{unWrap} : \vA \}$, and also a similar notation for (dependent)
+record types, for instance as in
+\[\data \msf{Record} = \msf{Record}\,\{\mit{field1} : \vA,\,\mit{field2} : \vB\}.\]
+
+All construction and elimination rules for type formers in $\MTy$ stay within
+$\MTy$. For example, induction on meta-level values can only produce meta-level
+values.
 
@@
 
-Since $\VTy$ is a sub-universe of $\Ty$, we have that when $A : \VTy$ then also
-$A : \Ty$. Formally, this is specified as an explicit embedding operation but
-informally it is nicer to have an implicit subtyping.
+Since $\VTy$ is a sub-universe of $\Ty$, we have that when $\vA : \VTy$ then also
+$\vA : \Ty$. Formally, this is specified as an explicit embedding operation, but
+we will use implicit subtyping for convenience.
 
@@
   &\hspace{-1em}\rlap{$\add : \Nat \to \Nat \to \Nat$}\\
-  &\hspace{-1em}\add :=\,&& \mathrlap{\letrec \go\,n\,m := \case n\,\of}\\
-  &\hspace{-1em}         && \ind \Zero   &&\to m;\\
-  &\hspace{-1em}         && \ind \Suc\,n &&\to \Suc\,(\go\,n\,m);\\
+  &\hspace{-1em}\add :=\,&& \mathrlap{\letrec \go\,\vn\,\vm := \case \vn\,\of}\\
+  &\hspace{-1em}         && \ind \Zero     &&\to \vm;\\
+  &\hspace{-1em}         && \ind \Suc\,\vn &&\to \Suc\,(\go\,\vn\,\vm);\\
   &\hspace{-1em}         && \go
@@
 Every recursive definition must be introduced with $\letrec$. The general syntax
-is $\letrec x : A := t; u$, where the $A$ type annotation can be
+is $\letrec \vx : \vA := \vt; \vu$, where the $\vA$ type annotation can be
 omitted. $\mbf{letrec}$ can be only used to define computations, not values
@@
 Object-level definitions use $:=$ as notation, instead of the $=$ that is used
-for meta-level ones. Later on, we may use some implicit CFTT notations, but we
-will always disambiguate the stages of $\mbf{let}$-s in this way. Non-recursive
-$\mbf{let}$ is also allowed, and can be used to shadow binders:
-\begin{alignat*}{3}
-  &f : \Nat \to \Nat\\
-  &f\,x := \letdef x := x + 10; \letdef x := x + 20; x * 10
-\end{alignat*}
+for meta-level ones. We also have non-recursive $\mbf{let}$, which can be used
+to define computations and values as well, and can be used to shadow binders:
+\begin{alignat*}{3}
+  &\vf : \Nat \to \Nat\\
+  &\vf\,\vx := \letdef \vx := \vx + 10; \letdef \vx := \vx + 20; \vx * 10
+\end{alignat*}
+We also allow $\mbf{newtype}$ definitions, both in $\VTy$ and $\CTy$. These are
+assumed to be erased at runtime. In the Haskell implementation they are
+important for guiding type class resolution, and we think that the explicit
+wrapping makes many definitions more comprehensible in CFTT as well.
+
 Values are call-by-value at runtime; they are computed eagerly in function
@@
 \begin{alignat*}{3}
-  &\hspace{-1em}f : \mathrlap{\Bool \to \Nat \to \Nat}\\
-  &\hspace{-1em}f\,b\,\mathrlap{:= \case b\,\of}\\
-  &\hspace{-1em}\ind \True  &&\to \lam x.\, x + 10;\\
-  &\hspace{-1em}\ind \False &&\to \lam x.\, x * 10;
-\end{alignat*}
-
-What about \textbf{definitional equality} at the object level? This is a
-distinct notion from the runtime semantics; object programs are embedded in
-CFTT, which is dependently typed, and we need to decide definitional equalities
-during type checking. The setup is simple: we have no $\beta$ or $\eta$ rules
-for object programs at all, nor any rule for $\mbf{let}$-unfolding. The reasons
-are the following. First, since the object language has general recursion, most
-$\beta$-rules are only valid in the operational semantics up to some
-restrictions, and $\beta$-conversion is not decidable to begin with. Second, in
-metaprogramming we care about the size and efficiency of generated code, and
-these properties are not stable under $\beta$ and $\eta$.  Hence, the sensible
-choice is to do metaprogramming up to strict syntactic equality of object
-programs.
-
-Let us discuss the object language. First, notice that there is no polymorphism
-or any kind of type dependency. Although we can define lists as $\data \List\,A
-: \VTy := \Nil\,|\,\Cons\,A\,(\List\,A)$, the parameterization is just a
-shorthand; all concrete instantiations of the type are distinct. This
-monomorphism makes it easy to use different memory layouts for different types.
-For example, types which look like products may be unboxed. We could also make a
-distinction between boxed and unboxed sum types. We do not explore this in
-detail, we just note that monomorphic types make it easy to control memory
-layouts, and we believe that this is an important part of performance
-optimization.
-
-Second, there are no higher-order functions, and functions also cannot be stored
-in data structures. Hence, locally defined functions can never escape their
-scope, and all function calls are to functions defined in the current
-scope. This makes it possible to run object programs without using dynamic
-closures. This latter point might not be completely straightforward; what
-about the previous $f$ function which has $\lambda$-expressions under a
-$\mbf{case}$, should that require closures?
-
-We say that it should not. We make this formal formal in Section \ref{TODO};
-here we only give an intuitive explanation. In short, we choose a call-by-name
-runtime semantics for functions, which means that the only way we can compute
-with a function is by applying it to all arguments and extracting the resulting
-value.  Hence, the only way to compute with $f$ is to apply it to \emph{two}
-arguments, so $f$ is operationally equivalent to $\lam b\,x.\,\case b\,\of \True
-\to x + 10; \False \to x * 10$.  In Section \ref{TODO} we show that all object
-programs can be transformed to a \emph{saturated} form, where the arity of every
-function call matches the number of topmost $\lambda$-binders in the definition
-of the called function. Then, the local functions can be compiled either to
-top-level functions by lambda-lifting, or if they are only tail-called, left in
-place as join points \cite{TODO}.
-
-Why not just make the object language less liberal, e.g.\ by disallowing
-$\lambda$ under $\mbf{case}$ or $\mbf{let}$, thereby making call saturation
-easier or more obvious? There is a trade-off between making the object language
-more restricted, and thus easier to compile, and making \emph{metaprogramming}
-more convenient. We will see that the ability to insert $\mbf{let}$-s without
-restriction is very convenient in code generation, and likewise the ability to
-have arbitrary object expressions in $\mbf{case}$ bodies. In this paper we go
-with the most liberal object syntax, at the cost of needing more downstream
-processing. The call-by-name nature of computation types requires a bit of a
-change of thinking from programmers, but we believe that it is well worth to
-have for the metaprogramming convenience.
-
-On the other side of the spectrum, one might imagine generating object code in
-low-level A-normal form. This is more laborious, but it could be also
-interesting, because it forces us to invent abstractions for manipulating ANF in
-the metalanguage. In a similar vein, Allais recently proposed metaprogramming
-quantum circuits in a two-level type theory \cite{TODO}. We leave such setups
-with low-level object languages to future investigation.
-
-Finally, one might compare our object language to call-by-push-value
-(CBPV). Indeed, we took inspiration from CBPV, and there are similarities, but
-also differences. In both systems there is a value-computation distinction, and
-values are call-by-value, and computations are call-by-name. However, our object
-language allows variable binding at arbitrary types, while CBPV only supports it
-at value types. In CBPV, a let-definition for a function is only possible by
-first packing it up as a closure value (or: ``thunk''), which clearly does not
-work for us.  Also, CBPV makes a judgment-level structural distinction between
-values and computations, while we use type universes for that. Generally
-speaking, type-based restrictions are easier to work with in dependent type
-theories than structural restrictions.
+  &\vf : \mathrlap{\Bool \to \Nat \to \Nat}\\
+  &\vf\,\vb := \case \vb\,\of \True \to (\lam \vx.\, \vx + 10); \False &&\to (\lam \vx.\, \vx * 10)
+\end{alignat*}
+In contrast, computations are call-by-name, and the only way we can compute with
+functions is to apply them to value arguments. The call-by-name strategy is
+fairly benign here and does not lead to significant duplication of computation,
+because functions cannot escape their scope; they cannot be passed as arguments
+or stored in data constructors. This makes it possible to run object programs
+without using dynamic closures. This point is not completely straightforward;
+consider the previous $\vf$ function which has $\lambda$-expressions under a
+$\mbf{case}$.
+
+However, the call-by-name semantics lets us transform $\vf$ to $\lam
+\vb\,\vx.\,\case \vb\,\of \True \to \vx + 10; \False \to \vx * 10$, and more
+generally we can transform programs so that every function call is
+\emph{saturated}. This means that every function call is of the form
+$\vf\,\vt_1\,\vt_2\,...\,\vt_n$, where $\vf$ is a function variable and the
+definition of $\vf$ immediately $\lambda$-binds $n$ arguments. We do not detail
+this here. We provide formal syntax and operational semantics of the object
+language in the Agda supplement. We formalized the specific translation steps
+that are involved in call saturation, but only specified the full translation
+informally.
+
+\subsubsection{Object-level definitional equality} This is a distinct notion
+from runtime semantics. Object programs are embedded in CFTT, which is a
+dependently typed language, so we may need to decide definitional equality of
+object programs during type checking. The setup is simple: we have no $\beta$ or
+$\eta$ rules for object programs at all, nor any rule for $\mbf{let}$-unfolding.
+The main reason is the following: we care about the size and efficiency of
+generated code, and these properties are not stable under $\beta\eta$-conversion
+and $\mbf{let}$-unfolding. Moreover, since the object language has general
+recursion, we do not have a sensible and decidable notion of program equivalence
+anyway.
+
+\subsubsection{Comparison to call-by-push-value}
+We took inspiration from call-by-push-value (CBPV), and there are similarities,
+but there are also significant differences. Both systems have a
+value-computation distinction, with call-by-name computations and call-by-value
+values. However, our object theory supports variable binding at arbitrary types
+while CBPV only supports value variables. In CBPV, a let-definition for a
+function is only possible by first packing it up as a closure value (or
+``thunk''), which clearly does not suit us. In future work we could make
+a closer look at the connections to CBPV.
+
+%% \subsubsection{Spectrum of possible object languages}
+%% There is a trade-off between making the object language more restricted, and
+%% thus easier to compile, and making metaprogramming more convenient. We will see
+%% that the ability to insert $\mbf{let}$-s without restriction is very convenient,
+%% and likewise the ability to have $\lambda$-abstraction under $\mbf{case}$
+%% bodies, although these features necessitate more downstream processing, to bring
+%% programs to saturated form. In this paper we go for metaprogramming convenience,
+
+%% more restricted, and thus easier to compile, and making \emph{metaprogramming}
+%% more convenient. We will see that the ability to insert $\mbf{let}$-s without
+%% restriction is very convenient in code generation, and likewise the ability to
+%% have arbitrary object expressions in $\mbf{case}$ bodies. In this paper we go
+%% with the most liberal object syntax, at the cost of needing more downstream
+%% processing. The call-by-name nature of computation types requires a bit of a
+%% change of thinking from programmers, but we believe that it is well worth to
+%% have for the metaprogramming convenience.
+
+%% On the other side of the spectrum, one might imagine generating object code in
+%% low-level A-normal form. This is more laborious, but it could be also
+%% interesting, because it forces us to invent abstractions for manipulating ANF in
+%% the metalanguage. In a similar vein, Allais recently proposed metaprogramming
+%% quantum circuits in a two-level type theory \cite{TODO}. We leave such setups
+%% with low-level object languages to future investigation.
+
+%% Finally, one might compare our object language to call-by-push-value
+%% (CBPV). Indeed, we took inspiration from CBPV, and there are similarities, but
+%% also differences. In both systems there is a value-computation distinction, and
+%% values are call-by-value, and computations are call-by-name. However, our object
+%% language allows variable binding at arbitrary types, while CBPV only supports it
+%% at value types. In CBPV, a let-definition for a function is only possible by
+%% first packing it up as a closure value (or: ``thunk''), which clearly does not
+%% work for us.  Also, CBPV makes a judgment-level structural distinction between
+%% values and computations, while we use type universes for that. Generally
+%% speaking, type-based restrictions are easier to work with in dependent type
+%% theories than structural restrictions.
+
+%% The reasons are the following. First, since the object
+%% language has general recursion, most $\beta$-rules are only valid in the
+%% operational semantics up to some restrictions, and $\beta$-conversion is not
+%% decidable to begin with. Second, in metaprogramming we care about the size and
+%% efficiency of generated code, and these properties are not stable under $\beta$
+%% and $\eta$. Hence, the sensible choice is to do metaprogramming up to strict
+%% syntactic equality of object programs.
+
+%% Let us discuss the object language. First, notice that there is no polymorphism
+%% or any kind of type dependency. Although we can define lists as $\data \List\,A
+%% : \VTy := \Nil\,|\,\Cons\,A\,(\List\,A)$, the parameterization is just a
+%% shorthand; all concrete instantiations of the type are distinct. This
+%% monomorphism makes it easy to use different memory layouts for different types.
+%% For example, types which look like products may be unboxed. We could also make a
+%% distinction between boxed and unboxed sum types. We do not explore this in
+%% detail, we just note that monomorphic types make it easy to control memory
+%% layouts, and we believe that this is an important part of performance
+%% optimization.
+
+%% Second, there are no higher-order functions, and functions also cannot be stored
+%% in data structures. Hence, locally defined functions can never escape their
+%% scope, and all function calls are to functions defined in the current
+%% scope. This makes it possible to run object programs without using dynamic
+%% closures. This latter point might not be completely straightforward; what
+%% about the previous $f$ function which has $\lambda$-expressions under a
+%% $\mbf{case}$, should that require closures?
+
+%% We say that it should not. We make this formal formal in Section \ref{TODO};
+%% here we only give an intuitive explanation. In short, we choose a call-by-name
+%% runtime semantics for functions, which means that the only way we can compute
+%% with a function is by applying it to all arguments and extracting the resulting
+%% value.  Hence, the only way to compute with $f$ is to apply it to \emph{two}
+%% arguments, so $f$ is operationally equivalent to $\lam b\,x.\,\case b\,\of \True
+%% \to x + 10; \False \to x * 10$.  In Section \ref{TODO} we show that all object
+%% programs can be transformed to a \emph{saturated} form, where the arity of every
+%% function call matches the number of topmost $\lambda$-binders in the definition
+%% of the called function. Then, the local functions can be compiled either to
+%% top-level functions by lambda-lifting, or if they are only tail-called, left in
+%% place as join points \cite{TODO}.
+
+%% Why not just make the object language less liberal, e.g.\ by disallowing
+%% $\lambda$ under $\mbf{case}$ or $\mbf{let}$, thereby making call saturation
+%% easier or more obvious? There is a trade-off between making the object language
+%% more restricted, and thus easier to compile, and making \emph{metaprogramming}
+%% more convenient. We will see that the ability to insert $\mbf{let}$-s without
+%% restriction is very convenient in code generation, and likewise the ability to
+%% have arbitrary object expressions in $\mbf{case}$ bodies. In this paper we go
+%% with the most liberal object syntax, at the cost of needing more downstream
+%% processing. The call-by-name nature of computation types requires a bit of a
+%% change of thinking from programmers, but we believe that it is well worth to
+%% have for the metaprogramming convenience.
+
+%% On the other side of the spectrum, one might imagine generating object code in
+%% low-level A-normal form. This is more laborious, but it could be also
+%% interesting, because it forces us to invent abstractions for manipulating ANF in
+%% the metalanguage. In a similar vein, Allais recently proposed metaprogramming
+%% quantum circuits in a two-level type theory \cite{TODO}. We leave such setups
+%% with low-level object languages to future investigation.
+
+%% Finally, one might compare our object language to call-by-push-value
+%% (CBPV). Indeed, we took inspiration from CBPV, and there are similarities, but
+%% also differences. In both systems there is a value-computation distinction, and
+%% values are call-by-value, and computations are call-by-name. However, our object
+%% language allows variable binding at arbitrary types, while CBPV only supports it
+%% at value types. In CBPV, a let-definition for a function is only possible by
+%% first packing it up as a closure value (or: ``thunk''), which clearly does not
+%% work for us.  Also, CBPV makes a judgment-level structural distinction between
+%% values and computations, while we use type universes for that. Generally
+%% speaking, type-based restrictions are easier to work with in dependent type
+%% theories than structural restrictions.
 
@@
 \begin{itemize}
-\item For $A : \Ty$, we have $\Lift A : \MTy$, pronounced as ``lift $A$''. This is
-      the type of metaprograms that produce $A$-typed object programs.
-\item For $A : \Ty$ and $t : A$, we have $\ql t \qr : \Lift A$, pronounced ``quote $t$''. This
-      is the metaprogram which immediately returns $t$.
-\item For $t : \Lift A$, we have $\spl t : A$, pronounced ``splice $t$''. This
+\item For $\vA : \Ty$, we have $\Lift \vA : \MTy$, pronounced as ``lift $\vA$''. This is
+      the type of metaprograms that produce $\vA$-typed object programs.
+\item For $\vA : \Ty$ and $\vt : \vA$, we have $\ql \vt \qr : \Lift \vA$, pronounced ``quote $\vt$''. This
+      is the metaprogram which immediately returns $\vt$.
+\item For $\vt : \Lift \vA$, we have $\spl \vt : A$, pronounced ``splice $\vt$''. This
   inserts the result of a metaprogram into an object term. \emph{Notation:}
-  splicing binds stronger than function application, so $f\,\spl x$ is parsed as
-  $f\,(\spl x)$.
-\item We have $\ql \spl t \qr \equiv t$ and $\spl \ql t \qr \equiv t$ as definitional equalities.
+  splicing binds stronger than function application, so $\vf\,\spl \vx$ is parsed as
+  $\vf\,(\spl \vx)$.
+\item We have $\ql \spl \vt \qr \equiv \vt$ and $\spl \ql \vt \qr \equiv \vt$ as definitional equalities.
 \end{itemize}
@@
 terms:
-  \[ \letdef n : \Nat := \spl(\id\,\ql 10 + 10 \qr); ... \]
+  \[ \letdef \vn : \Nat := \spl(\id\,\ql 10 + 10 \qr); ... \]
 Here, $\id$ is used at type $\Lift \Nat$. During staging, the expression in the
@@
   &\mit{and} : \Up\Bool \to \Up\Bool \to \Up\Bool\\
-  &\mit{and}\,x\,y = \ql\case \spl x\,\of \True \to \spl y; \False \to \False\qr
+  &\mit{and}\,\vx\,\vy = \ql\case \spl \vx\,\of \True \to \spl \vy; \False \to \False\qr
 \end{alignat*}
@@
 \begin{alignat*}{3}
-  & \hspace{-5em}\mathrlap{\map : \{A\,B : \VTy\} \to (\Up A \to \Up B) \to \Up (\List A) \to \Up(\List B)}\\
-  & \hspace{-5em}\mathrlap{\map\,f\,\vas = \ql\letrec \go\,\vas := \case \vas\,\of}\\
+  & \hspace{-5em}\mathrlap{\map : \{\vA\,\vB : \VTy\} \to (\Up \vA \to \Up \vB) \to \Up (\List\,\vA) \to \Up(\List\,\vB)}\\
+  & \hspace{-5em}\mathrlap{\map\,\vf\,\vas = \ql\letrec \go\,\vas := \case \vas\,\of}\\
   & \hspace{-5em}\ind\ind\ind\ind\ind \ind\Nil             &&\to \Nil;\\
-  & \hspace{-5em}\ind\ind\ind\ind\ind \ind\Cons\,a\,\vas &&\to \Cons\,\spl(f\,\ql a \qr)\,(\go\,\vas);\\
+  & \hspace{-5em}\ind\ind\ind\ind\ind \ind\Cons\,\va\,\vas &&\to \Cons\,\spl(\vf\,\ql \va \qr)\,(\go\,\vas);\\
   & \hspace{-5em}\ind\ind\ind\ind\ind\go\,\spl\vas \qr
 \end{alignat*}
-For example, this can be used as $\letdef f\,\vas : \List \Nat \to \List \Nat :=
-\spl(\map\,(\lam x.\,\ql \spl x + 10 \qr)\,\qt{\vas}) $.  This is staged to a recursive
-definition where the mapping function is inlined into the $\Cons$ case as
-$\Cons\,a\,\vas \to \Cons\,(a + 10)\,(\go\,\vas)$.  Note that $\map$ has to
-abstract over value types, since lists can only contain values, not functions. Also, the
-mapping function has type $\Up A \to \Up B$, instead of $\Up (A \to B)$. The
-former type is often preferable to the latter in staging; the former is
-a metafunction with useful computational content, while the latter is merely a
-black box that computes object code. If we have $f : \Up(A \to B)$, and $f$ is
-staged to $\ql \lam x.\,t\qr$, then $\spl f u$ is staged to an
-undesirable ``administrative'' $\beta$-redex $(\lam x.\,t)\,u$.
-
-\section{Staged Programming in CFTT}\label{sec:abstractions-in-cftt}
+For example, this can be used as $\letdef f\,\vas : \List\,\Nat \to \List\,\Nat
+:= \spl(\map\,(\lam \vx.\,\ql \spl \vx + 10 \qr)\,\qt{\vas}) $.  This is staged to a
+recursive definition where the mapping function is inlined into the $\Cons$ case
+as $\Cons\,\va\,\vas \to \Cons\,(\va + 10)\,(\go\,\vas)$.  Note that $\map$ has to
+abstract over value types, since lists can only contain values, not
+functions. Also, the mapping function has type $\Up \vA \to \Up \vB$, instead of
+$\Up (\vA \to \vB)$. The former type is often preferable to the latter in staging;
+the former is a metafunction with useful computational content, while the latter
+is merely a black box that computes object code. If we have $\vf : \Up(\vA \to \vB)$,
+and $\vf$ is staged to $\ql \lam \vx.\,\vt\qr$, then $\spl \vf\,\vu$ is staged to an
+undesirable ``administrative'' $\beta$-redex $(\lam \vx.\,\vt)\,\vu$.
+
+\section{Monads \& Monad Transformers}\label{sec:monad-transformers}
+
+In this section we build a library for monads and monad transformers. We believe
+that this is a good demonstration of CFTT-s abilities, since monads are
+ubiquitous in Haskell programming, and they also introduce a great amount of
+abstraction that should be optimized away.
 
@@
 
-We saw that $\Up A \to \Up B$ is usually preferable to $\Up(A \to B)$. The two
-types are actually equivalent up to the runtime semantics of object programs,
+We start with some preparatory work before getting to monads. We saw that $\Up
+\vA \to \Up \vB$ is usually preferable to $\Up(\vA \to \vB)$. The two types are
+actually equivalent during staging, up to the runtime equivalence of object programs,
 and we can convert back and forth in CFTT:
 \begin{alignat*}{3}
-  &\mup : \Up (A \to B) \to \Up A \to \Up B && \ind\ind \mdown : (\Up A \to \Up B) \to \Up (A \to B) \\
-  &\mup\,f\,a = \ql \spl f\, \spl a\qr   && \ind\ind \mdown\,f = \ql \lam a.\,\spl(f\,\ql a \qr) \qr
-\end{alignat*}
-We can not show internally, using propositional equality, that these functions are
+  &\mup : \Up (\vA \to \vB) \to \Up \vA \to \Up \vB && \ind\ind \mdown : (\Up \vA \to \Up \vB) \to \Up (\vA \to \vB) \\
+  &\mup\,\vf\,\va = \ql \spl \vf\, \spl \va\qr   && \ind\ind \mdown\,\vf = \ql \lam \va.\,\spl(\vf\,\ql \va \qr) \qr
+\end{alignat*}
+We cannot show internally, using propositional equality, that these functions are
 inverses, since we do not have $\beta\eta$-rules for object functions; but we
-will not need that in the rest of the paper.
+will not need this proof in the rest of the paper.
 
 In the staged compilation and partial evaluation literature, the term
-\textbf{binding time improvement} is used to refer to such conversions, where
+\emph{binding time improvement} is used to refer to such conversions, where
 the ``improved'' version supports more compile-time computation. A general
@@
 as possible with improved representations, and only convert back to object code
-at points where runtime control dependencies are unavoidable.
-
-We use a Haskell-style type class for binding time improvements:
-\begin{alignat*}{3}
-  &\hspace{-8em}\mathrlap{\tyclass\,\Improve\,(A : \Ty)\,(B : \MTy)\,\where} \\
-  &\hspace{-8em}\ind \mup   &&: \Up A \to B\\
-  &\hspace{-8em}\ind \mdown &&: B \to \Up A
-  &\\
-  &\hspace{-8em}\mathrlap{\instance \Improve\,(A \to B)\,(\Up A \to \Up B)\,\where\,\,...}
-\end{alignat*}
+at points where runtime dependencies are unavoidable. Let us look at
+binding-time-improvement for product types now:
+\begin{alignat*}{3}
+  &\mup : \Up (\vA,\,\vB) \to (\Up \vA,\,\Up \vB) && \ind\ind \mdown : (\Up \vA,\,\Up \vB) \to \Up(\vA,\,\vB) \\
+  &\mup\,\vx = (\qt{\fst\,\spl \vx},\, \qt{\snd\,\spl \vx})   && \ind\ind \mdown\,(\vx,\,\vy) = \qt{(\spl \vx,\,\spl \vy)}
+\end{alignat*}
+
+%% We use a Haskell-style type class for binding time improvements:
+%% \begin{alignat*}{3}
+%%   &\hspace{-8em}\mathrlap{\tyclass\,\Improve\,(A : \Ty)\,(B : \MTy)\,\where} \\
+%%   &\hspace{-8em}\ind \mup   &&: \Up A \to B\\
+%%   &\hspace{-8em}\ind \mdown &&: B \to \Up A
+%%   &\\
+%%   &\hspace{-8em}\mathrlap{\instance \Improve\,(A \to B)\,(\Up A \to \Up B)\,\where\,\,...}
+%% \end{alignat*}
+%% We will use type classes in an informal way, without precisely specifying how
+%% they work. However, in the Haskell adaptation of this paper we do make crucial
+%% use of type classes. There are some necessary differences between the CFTT and
+%% Haskell versions though, which we will summarize in Section \ref{TODO}. Let us
+%% look at improvement for product types now:
+%% \begin{alignat*}{3}
+%%   &\hspace{-4em}\mathrlap{\instance \Improve\,(A,\,B)\,(\Up A,\,\Up B)\,\where}\\
+%%   &\hspace{-4em}\ind \mup\,x   && = (\ql \fst\,\spl x \qr,\, \ql \snd\,\spl x \qr)\\
+%%   &\hspace{-4em}\ind \mdown\,(a,\,b) && = \ql(\spl a,\, \spl b)\qr
+%% \end{alignat*}
+
+Here we overload Haskell-style product type notation, both for types and the
+pair constructor, at both levels (products are definable as a value type at the
+object level). There is a problem with this conversion though: $\mup$ uses $\vx
+: \Up(\vA,\,\vB)$ twice, which can increase code size and duplicate runtime
+computations. For example, $\mdown\,({\mup\,\ql \vf\,\vx \qr})$ is staged to
+$\ql (\fst\,(\vf\,\vx),\,\snd\,(\vf\,\vx)) \qr$. It would be safer to first
+let-bind an expression with type $\Up(\vA,\,\vB)$, and then only use projections
+of the newly bound variable. This is called \emph{let-insertion} in staged
+compilation. But it is impossible to use let-insertion in $\mup$ because the
+return type is in $\MTy$, and we cannot introduce object binders in meta-level
+code.  Fortunately, there is a principled solution.
+
+\subsection{The Code Generation Monad}\label{sec:the-code-generation-monad}
+
+Our solution is to use a monad which extends $\MTy$ with the ability to
+freely generate object-level code and introduce object binders.
+\begin{alignat*}{3}
+  & \newtype \Gen\,\vA = \Gen\,\{\unGen : \{\vR : \Ty\} \to (\vA \to \Up \vR) \to \Up \vR\}
+\end{alignat*}
+This is a monad in $\MTy$ in a standard sense:
+\begin{alignat*}{3}
+  &\mathrlap{\instance \Monad\,\Gen\,\where}\\
+  &\ind \return \va        &&= \Gen \dlr \lam \vk.\,\vk\,\va\\
+  &\ind \mit{ga} \bind \vf &&= \Gen \dlr \lam \vk.\,\unGen\,\mit{ga}\,(\lam \va.\,(\vf\,\va\,\vk))
+\end{alignat*}
+From now on, we reuse Haskell-style type classes and $\mdo$-notation in CFTT.
 We will use type classes in an informal way, without precisely specifying how
-they work. However, in the Haskell adaptation of this paper we do make crucial
-use of type classes. There are some necessary differences between the CFTT and
-Haskell versions though, which we will summarize in Section \ref{TODO}. Let us
-look at improvement for product types now:
-\begin{alignat*}{3}
-  &\hspace{-4em}\mathrlap{\instance \Improve\,(A,\,B)\,(\Up A,\,\Up B)\,\where}\\
-  &\hspace{-4em}\ind \mup\,x   && = (\ql \fst\,\spl x \qr,\, \ql \snd\,\spl x \qr)\\
-  &\hspace{-4em}\ind \mdown\,(a,\,b) && = \ql(\spl a,\, \spl b)\qr
-\end{alignat*}
-Here we overload Haskell-style product type notation for types and pairing, at
-both levels (products are definable as a value type at the object level). There
-is a problem with this conversion though: $\mup$ uses $x : \Up(A,\,B)$ twice,
-which can increase code size and duplicate runtime computations. For example,
-$\mdown\,(\mup\,\ql f\,x \qr)$ is staged to $\ql (\fst\,(f\,x),\,\snd\,(f\,x))
-\qr$. It would be safer to first $\mbf{let}$-bind an expression with type
-$\Up(A,\,B)$, and then only use projections of the newly bound variable. This is
-called \textbf{let-insertion} in staged compilation. But it is impossible to use
-let-insertion in $\mup$ because the return type is a product in $\MTy$, and we
-cannot introduce object binders in meta-level code.
-
-\subsection{The Code Generation Monad}\label{sec:the-code-generation-monad}
-
-Our solution is to use a monad which freely extends $\MTy$ with the ability to
-generate object-level code and introduce object binders.
-\begin{alignat*}{3}
-  & \newtype \Gen\,A = \Gen\,\{\unGen : \{R : \Ty\} \to (A \to \Up R) \to \Up R\}
-\end{alignat*}
-This is a monad in $\MTy$ in a very standard sense, and we reuse Haskell type classes
-and $\mdo$-notation.
-\begin{alignat*}{3}
-  &\instance \Monad\,\Gen\,\where\\
-  &\ind \return a = \Gen\,(\lam k.\,k\,a)\\
-  &\ind \mit{ga} \bind f = \Gen\,(\lam k.\,\unGen\,\mit{ga}\,(\lam a.\,(f\,a\,k)))
-\end{alignat*}
-From the $\Monad$ instance, the $\Functor$ and $\Applicative$ instances can be also derived.
-We will reuse $(\fmap)$ and $(\ap)$ for applicative notation
-as well.
-An inhabitant of $\Gen\,A$ can be viewed as an action whose effect is to produce
-some object code and returns an $A$ value that can depend on produced object
-binders. $\Gen$ can be only ``run'' when the result type is an object type:
-\begin{alignat*}{3}
-  &\runGen : \Gen\,(\Up A) \to \Up A\\
+they work. However, type classes are used in essentially the same way in the
+Haskell implementation of the paper, and also in the Agda implementation, with
+modest technical differences.
+
+From the $\Monad$ instance, the $\Functor$ and $\Applicative$ instances can be
+also derived. We reuse $({\fmap})$ and $({\ap})$ for applicative notation as well.
+An inhabitant of $\Gen\,\vA$ can be viewed as an action whose effect is to produce
+some object code. $\Gen$ can be only ``run'' when the result type is an object
+type:
+\begin{alignat*}{3}
+  &\runGen : \Gen\,(\Up \vA) \to \Up \vA\\
   &\runGen\,\vma = \unGen\,\vma\,\id
@@
 \begin{alignat*}{3}
-  & \gen : \Up A \to \Gen\,(\Up A) \\
-  & \gen\,a = \Gen\,(\lam k.\,\ql \letdef x : A := \spl a; \spl(k\,\ql x \qr) \qr)
-\end{alignat*}
-And also recursive function definitions:
-\begin{alignat*}{3}
-  & \genRec : \{A : \CTy\} \to (\Up A \to \Up A) \to \Gen\,(\Up A) \\
-  & \genRec\,f = \Gen\,(\lam k.\,\qt{\letrec x : A := \spl(f\,\qt{x}); \spl(k\,\qt{x})})
-\end{alignat*}
-Now, using do-notation, we may write $\mdo \{x \fro \gen\,\qt{10 + 10}; y \fro
-\gen\,\qt{20 + 20}\};\return \qt{x + y}\}$, for a $\Gen\,(\Up \Nat)$
-action. Running this with $\runGen$ yields $\qt{\letdef x := 10 + 10; \letdef y
-  := 20 + 20; x + y}$. We can also define a ``safer'' binding-time improvement for
+  & \gen : \Up \vA \to \Gen\,(\Up \vA) \\
+  & \gen\,\va = \Gen \dlr \lam \vk.\,\ql \letdef \vx : \vA := \spl \va; \spl(\vk\,\ql \vx \qr) \qr
+\end{alignat*}
+And also recursive definitions of computations:
+\begin{alignat*}{3}
+  & \genRec : \{\vA : \CTy\} \to (\Up \vA \to \Up \vA) \to \Gen\,(\Up \vA) \\
+  & \genRec\,\vf = \Gen\ \dlr \lam \vk.\,\qt{\letrec \vx : \vA := \spl(\vf\,\qt{\vx}); \spl(\vk\,\qt{\vx})}
+\end{alignat*}
+Now, using do-notation, we may write $\mdo \{\vx \fro \gen\,\qt{10 + 10}; \vy \fro
+\gen\,\qt{20 + 20}\};\return \qt{\vx + \vy}\}$, for a $\Gen\,(\Up \Nat)$
+action. Running this with $\runGen$ yields $\qt{\letdef \vx := 10 + 10; \letdef \vy
+  := 20 + 20; \vx + \vy}$. We can also define a ``safer'' binding-time improvement for
 products, using let-insertion:
-\begin{alignat*}{3}
-  &\hspace{-4em}\mathrlap{\instance \Improve\,(A,\,B)\,(\Gen\,(\Up A,\,\Up B))\,\where}\\
-  &\hspace{-4em}\ind \mup\,x   && = \mdo\{x \fro \gen\,x; \return (\qt{\fst\,\spl x},\,\qt{\snd\,\spl x})\}\\
-  &\hspace{-4em}\ind \mdown\,x && = \runGen\,\$\,\mdo \{(a,\,b) \fro x; \return \qt{(\spl a,\, \spl b)}\}
+\begin{alignat*}{6}
+  &\hspace{-2em}\mathrlap{\mup : \Up (\vA,\,\vB) \to \Gen\,(\Up \vA,\,\Up \vB)            \hspace{5.45em} \mdown : \Gen\,(\Up \vA,\, \Up \vB) \to \Up(\vA,\,\vB)} \\
+  &\hspace{-2em}\mup\,\vx = \mdo &&\vx \fro \gen\,\vx                                    && \ind\ind \mdown\,x = \mdo &&(\va,\,\vb) \fro x\\
+  &\hspace{-2em}               &&\return (\qt{\fst\,\spl \vx},\,\qt{\snd\,\spl \vx})\} &&                           &&\return \qt{(\spl \va,\, \spl \vb)}
 \end{alignat*}
@@
 generation, so ultimately we always want to $\runGen$ our actions. So why not
-just always work in $\Gen$? We may not want to do that, because we would like to
-reason about the size and content of the code we are generating, and the
-implicit code generation in $\Gen$ is not very transparent. This is a bit
-similar to the $\msf{IO}$ monad in Haskell, where eventually everything needs
-to run in $\msf{IO}$, but we may prefer to not write most of our program in
-$\msf{IO}$.
+just always work in $\Gen$? The implicitness of $\Gen$ may make it harder to
+reason about the size and content of generated code. This is a bit similar to
+the $\msf{IO}$ monad in Haskell, where eventually everything needs to run in
+$\msf{IO}$, but we may prefer to not write most of our program in $\msf{IO}$.
 
@@
 
-Let us start with the $\Maybe$ monad. We have $\data \Maybe\,A : \VTy :=
-\Nothing\,|\,\Just\,A$, and $\Maybe$ itself is available as a $\VTy \to \VTy$
+Let us start with the $\Maybe$ monad. We have $\data \Maybe\,\vA :=
+\Nothing\,|\,\Just\,\vA$, and $\Maybe$ itself is available as a $\VTy \to \VTy$
 metafunction. However, we cannot directly fashion a monad out of $\Maybe$, since
@@
 type for binding:
-\[ (\bind) : \Up(\Maybe\,A) \to (\Up\,A \to \Up\,(\Maybe\,B)) \to \Up(\Maybe\,B) \]
+\[ (\bind) : \Up(\Maybe\,\vA) \to (\Up\,\vA \to \Up\,(\Maybe\,\vB)) \to \Up(\Maybe\,\vB) \]
 This works, but the definition necessarily uses runtime case splits on $\Maybe$
@@
 
-Instead, our strategy is to only use ordinary monads in $\MTy$, and convert
+Instead, our strategy is to only use proper monads in $\MTy$, and convert
 between object types and meta-monads when necessary, as a form of binding-time
-improvement. Assume that $\Maybe_M$ is a meta-level monad, and $\MaybeT_\M$ is
-the monad transformer defined as follows:
-\[ \newtype \MaybeT_\M\,M\,A = \MaybeT_\M\,\{\runMaybeT_\M : M\,(\Maybe_\M\,A)\} \]
-Then the object-level $\Maybe$ is binding-time improved as follows:
-\begin{alignat*}{3}
-  &\instance \Improve\,(\Maybe\,A)\,(\MaybeT_\M\,\Gen\,(\Up\,A))\,\where\\
-  %% &\ind \mup : \Up\,(\Maybe\,A) \to \MaybeT_\M\,\Gen\,(\Up\,A) \\
-  &\ind \mup\,x = \MaybeT_\M \dlr \Gen \dlr \lam k.\\
-  &\ind\ind \qt{\case \spl x\,\of \Nothing \to \spl(k\,\Nothing_\M);\Just\,a \to \spl(k\,(\Just_\M\qt{a}))}\\
-  &\\
+improvement. We define a class for this conversion:
+\begin{alignat*}{3}
+  &\hspace{-9em}\mathrlap{\tyclass\,\MonadGen\,\vM \RA \Improve\,(\vF : \VTy \to \Ty)\,(\vM : \MTy \to \MTy)\,\where} \\
+  &\hspace{-9em}\ind \mup   &&: \{\vA : \VTy\} \to \Up(\vF\,\vA) \to \vM\,(\Up \vA)\\
+  &\hspace{-9em}\ind \mdown &&: \{\vA : \VTy\} \to \vM\,(\Up \vA) \to \Up(\vF\,\vA)
+\end{alignat*}
+Assume that $\Maybe_\M$ is the standard meta-level monad, and $\MaybeT_\M$ is
+the standard monad transformer:
+\[ \newtype \MaybeT_\M\,\vM\,\vA = \MaybeT_\M\,\{\runMaybeT_\M : \vM\,(\Maybe_\M\,\vA)\} \]
+Now, the binding-time improvement of $\Maybe$ is as follows:
+\begin{alignat*}{3}
+  &\instance \Improve\,\Maybe\,(\MaybeT_\M\,\Gen)\,\where\\
+  %% &\ind \mup : \Up\,(\Maybe\,\vA) \to \MaybeT_\M\,\Gen\,(\Up\,\vA) \\
+  &\ind \mup\,\vx = \MaybeT_\M \dlr \Gen \dlr \lam \vk.\\
+  &\ind\ind \qt{\case \spl \vx\,\of \Nothing \to \spl(\vk\,\Nothing_\M);\Just\,\va \to \spl(\vk\,(\Just_\M\qt{\va}))}\\
   %% &\ind \mdown : \MaybeT_\M\,\Gen\,(\Up\,A) \to \Up\,(\Maybe\,A) \\
   &\ind \mdown\,(\MaybeT_\M\,(\Gen\,\vma)) = \\
-  &\ind\ind \vma\,(\lam x.\,\case x\,\of \Nothing_\M \to \qt{\Nothing}; \Just_\M\,a \to \qt{\Just\,\spl a})
+  &\ind\ind \vma\,(\lam \vx.\,\case \vx\,\of \Nothing_\M \to \qt{\Nothing}; \Just_\M\,\va \to \qt{\Just\,\spl \va})
 \end{alignat*}
 With this, we get the $\Monad$ instance for free from $\MaybeT_\M$ and $\Gen$. A small example:
-\[ \letdef n : \Maybe\,\Nat := \spl(\mdown\,\$\,\mdo \{x \fro \return\,\qt{10}; y \fro \return\,\qt{20}; \return\,\qt{x + y})\});\,... \]
-Since $\MaybeT_\M$ is meta-level, its binding always fully computes at staging time. Thus, the above code is staged to
-\[ \letdef n : \Maybe\,\Nat := \Just\,(10 + 20);\;... \]
-Assume also a $\lift : \Monad\,M \RA M\,A \to \MaybeT_\M\,M\,A$ operation, as in
-Haskell. We shall assume this for all monad transformers henceforth. We can do
-let-insertion in $\MaybeT_\M\,\Gen$ by simply lifting:
-\begin{alignat*}{3}
-  &\gen' : \Up A \to \MaybeT_\M\,\Gen\,(\Up A) \\
-  &\gen'\,a = \lift\,(\gen\,a)
+\[ \letdef \vn : \Maybe\,\Nat := \spl(\mdown\,\$\,\mdo \{\vx \fro \return\,\qt{10}; \vy \fro \return\,\qt{20}; \return\,\qt{\vx + \vy})\});\,... \]
+Since $\MaybeT_\M$ is meta-level, its monadic binding fully computes at staging time. Thus, the above code is staged to
+\[ \letdef \vn : \Maybe\,\Nat := \Just\,(10 + 20);\;... \]
+Assume also a $\lift : \Monad\,\vM \RA \vM\,\vA \to \MaybeT_\M\,\vM\,\vA$ operation which
+comes from $\MaybeT_\M$ being a monad transformer. We can do let-insertion in
+$\MaybeT_\M\,\Gen$ by simply lifting:
+\begin{alignat*}{3}
+  &\gen' : \Up \vA \to \MaybeT_\M\,\Gen\,(\Up \vA) \\
+  &\gen'\,\va = \lift\,(\gen\,\va)
 \end{alignat*}
@@
 \begin{alignat*}{3}
-&\tyclass\,\Monad\,M \RA \MonadGen\,M\,\where\\
-&\ind \liftGen : \Gen\,A \to \M\,A\\
+&\tyclass\,\Monad\,\vM \RA \MonadGen\,\vM\,\where\\
+&\ind \liftGen : \Gen\,\vA \to \M\,\vA\\
 &\instance \MonadGen\,\Gen\,\where\,\liftGen = \id \\
-&\instance \MonadGen\,M \RA \MonadGen\,(\MaybeT_\M\,M)\,\where\,\liftGen = \lift \circ \liftGen
-\end{alignat*}
-We also redefine $\gen$ and $\genRec$ to work in any $\MonadGen$, so from now on we have:
-\begin{alignat*}{3}
- & \gen   &&: \MonadGen\,M \RA \Up A \to M\,(\Up A)\\
- &\genRec &&: \MonadGen\,M \RA (\Up A \to \Up A) \to M\,(\Up A)
+&\instance \MonadGen\,\vM \RA \MonadGen\,(\MaybeT_\M\,\vM)\,\where\,\liftGen = \lift \circ \liftGen
+\end{alignat*}
+The $\MonadGen$ instance can be defined uniformly for every monad transformer as
+$\liftGen = \lift \circ \liftGen$. We also redefine $\gen$ and $\genRec$ to work
+in any $\MonadGen$, so from now on we have:
+\begin{alignat*}{3}
+ &\gen   &&: \MonadGen\,\vM \RA \Up \vA \to \vM\,(\Up \vA)\\
+ &\genRec &&: \MonadGen\,\vM \RA (\Up \vA \to \Up \vA) \to \vM\,(\Up \vA)
 \end{alignat*}
@@
 \subsubsection{Case splitting in monads}
-Trying to write more substantial code, we quickly bump into an issue. We
-want to case-split on object-level data inside a monadic action, like in the following:
-\begin{alignat*}{3}
-  & \hspace{-4em}\mathrlap{f : \Nat \to \Maybe\,\Nat}\\
-  & \hspace{-4em}\mathrlap{f\,x := \spl(\mdown\,\$\, \case x == 10\,\of}\\
-  & \hspace{-4em}\ind \True  &&\to \return\,\qt{x + 5};\\
+We often want to case-split on object-level data inside a monadic action, like
+in the following:
+\begin{alignat*}{3}
+  & \hspace{-4em}\mathrlap{\vf : \Nat \to \Maybe\,\Nat}\\
+  & \hspace{-4em}\mathrlap{\vf\,\vx := \spl(\mdown\,\$\, \case \vx == 10\,\of}\\
+  & \hspace{-4em}\ind \True  &&\to \return\,\qt{\vx + 5};\\
   & \hspace{-4em}\ind \False &&\to \Nothing_\M)
@@
 us a ``view'' on these points:
-\[ \data \msf{SplitList}\,A = \Nil'\,|\,\Cons'\,(\Up A)\,(\Up (\List\,A)) \]
+\[ \data \msf{SplitList}\,\vA = \Nil'\,|\,\Cons'\,(\Up \vA)\,(\Up (\List\,\vA)) \]
 We can generate code for a case split, returning a view on it:
 \begin{alignat*}{3}
-  &\mit{split} : \Up (\List\,A) \to \Gen\,(\msf{SplitList}\,A)\\
-  &\mit{split}\,\vas = \Gen \dlr \lam k.\,\qt{\case \spl \vas\,\of \Nil \to \spl(k\,\Nil'); \Cons\,a\,\vas \to \spl(k\,(\Cons'\,\qt{a}\,\qt{\vas}))}
+  &\mit{split} : \Up (\List\,\vA) \to \Gen\,(\msf{SplitList}\,\vA)\\
+  &\mit{split}\,\vas = \Gen \dlr \lam \vk.\,\qt{\case \spl \vas\,\of \Nil \to \spl(\vk\,\Nil'); \Cons\,\va\,\vas \to \spl(\vk\,(\Cons'\,\qt{a}\,\qt{\vas}))}
 \end{alignat*}
@@
 \begin{alignat*}{3}
-  &\mdo && \mit{sp} \fro \liftGen \dlr \mit{split}\,\vas\\
-  &     && \case \mit{sp}\,\of \\
-  &     && \ind \Nil' \to ... \\
-  &     && \ind \Cons'\,a\,\vas \to ...
-\end{alignat*}
-This can be generalized to any splitting on object values. In the supplemental
-Haskell implementation, we overload $\mit{split}$ with a type class. However, in
-a native implementation of CFTT it might make sense to simply extend
-$\mdo$-notation with an elaboration of $\mbf{case}$ on $\Up A$-typed data to an
-application of the appropriate $\mit{split}$ function. We adopt this in
-the rest of this paper. So we will simply write
-\[ \mdo \{\case \vas\,\of \Nil \to ...; \Cons\,a\,\vas \to ...\}. \]
-
-\subsection{Building a Monad Transformer Library}\label{monad-transformers}
+  &\mdo \{\mit{sp} \fro \liftGen\;(\!\mit{split}\,\vas);(\case \mit{sp}\,\of \Nil' \to ...;\Cons'\,\va\,\vas \to ...)\}
+\end{alignat*}
+%% \begin{alignat*}{3}
+%%   &\mdo && \mit{sp} \fro \liftGen \dlr \mit{split}\,\vas\\
+%%   &     && \case \mit{sp}\,\of \\
+%%   &     && \ind \Nil' \to ... \\
+%%   &     && \ind \Cons'\,a\,\vas \to ...
+%% \end{alignat*}
+This can be generalized to splitting on any object value. In the supplemental
+Haskell and Agda implementations, we overload $\mit{split}$ with a type
+class. However, in a native implementation of CFTT it might make sense to extend
+$\mdo$-notation, so that we elaborate $\mbf{case}$ on object values to an
+application of the appropriate $\mit{split}$ function. We adopt this in the rest
+of the paper, so we will be able to write $\case \vx\,\of ...$ in a $\mdo$-block,
+whenever we work in a $\MonadGen$ and $\vx$ has a value type.
+
+\subsection{Monad Transformers}\label{monad-transformers}
 
@@
 \begin{alignat*}{3}
-  & \newtype \Identity\,A := \Identity\,\{\runIdentity : A\} \\
-  & \instance \Improve\,(\Identity\,A)\,(\Gen\,(\Up A))\,\where\\
-  & \ind \mup\,x = \Gen \dlr \lam k.\,k\,\qt{\runIdentity\,\spl x}\\
-  & \ind \mdown\,x = \unGen\,x \dlr \lam a.\,\qt{\Identity\,\spl a}
-\end{alignat*}
-We also switch from the object-level $\Maybe$ to a ``transformer'' form as well,
-and recover it as $\MaybeT\,\Identity$.
-\[ \newtype \MaybeT\,M\,A := \MaybeT\,\{\runMaybeT : M\,(\Maybe\,A)\} \]
-Generally, we want to improve an operator $F : \VTy \to \VTy$ to some $M : \MTy
-\to \MTy$, where $M$ is a $\MonadGen$. We define a \emph{quantified class
-constraint} \cite{TODO} for this, that we will often use as an assumption:
-\[ \ImproveF\,F\,M = \MonadGen\,M \times ((A : \VTy) \to \Improve\,(F\,A)\,(M\,(\Lift\,A))) \]
+  & \newtype \Identity\,\vA := \Identity\,\{\runIdentity : \vA\} \\
+  & \instance \Improve\,\Identity\,\Gen\,\where\\
+  & \ind \mup\,\vx = \Gen \dlr \lam \vk.\,\vk\,\qt{\runIdentity\,\spl \vx}\\
+  & \ind \mdown\,\vx = \unGen\,\vx \dlr \lam \va.\,\qt{\Identity\,\spl \va}
+\end{alignat*}
+We also recover the object-level $\Maybe$ as $\MaybeT\,\Identity$, from the following $\MaybeT$:
+\[ \newtype \MaybeT\,(\vM : \VTy \to \Ty)\,(\vA : \VTy) := \MaybeT\,\{\runMaybeT : \vM\,(\Maybe\,\vA)\} \]
 With this, improvement can be generally defined for $\MaybeT$:
 \begin{alignat*}{3}
-  &\instance \ImproveF\,F\,M \RA \Improve\,(\MaybeT\,F\,A)\,(\MaybeT_\M\,M\,(\Up A))\,\where\\
-  &\ind \mup\,x = \MaybeT_\M \dlr \mdo\\
-  &\ind\ind \vma \fro \mup\,\qt{\runMaybeT\,\spl x}\\
+  &\instance \Improve\,\vF\,\vM \RA \Improve\,(\MaybeT\,\vF)\,(\MaybeT_\M\,\vM)\,\where\\
+  &\ind \mup\,\vx = \MaybeT_\M \dlr \mdo\\
+  &\ind\ind \vma \fro \mup\,\qt{\runMaybeT\,\spl \vx}\\
   &\ind\ind \case \vma\,\of \Nothing \to \return \Nothing_\M\\
-  &\hspace{6.5em}         \Just\,a \hspace{1.15em}\to \return (\Just_\M\,a)\\
+  &\hspace{6.5em}         \Just\,\va \hspace{1.15em}\to \return (\Just_\M\,\va)\\
+  &\ind \mdown\,(\MaybeT_\M\,\vx) = \qt{\MaybeT\,\spl(\mdown \dlr \vx \bind \lam\case\\
+  &\ind\ind\Nothing_\M \to \return \qt{\Nothing}\\
+  &\ind\ind\Just_\M\,\va\hspace{1.15em} \to \return \qt{\Just\,\spl \va})}
+\end{alignat*}
+In the $\mbf{case}$ in $\mup$, we already use our syntax sugar for matching on a
+$\Maybe$ value inside an $\vM$ action. This is legal, since we know from the
+$\Improve\,\vF\,\vM$ assumption that $\vM$ is a $\MonadGen$. In $\mdown$ we also use
+$\lam\case ...$ to shorten $\lam \vx.\,\case \vx\,\of ...$.
+
+In the meta level, we can reuse essentially all definitions from Haskell's monad
+transformer library $\msf{mtl}$. From $\msf{mtl}$, only the continuation monad
+transformer fails to support binding-time-improvement in CFTT, because of the
+obvious need for dynamic closures. In the following we present only $\StateT$
+and $\ReaderT$. Starting with $\StateT$, we assume $\StateT_\M$ as the standard
+meta-level definition. The object-level $\StateT$ has type $(\vS : \VTy)(\vF : \VTy
+\to \Ty)(\vA : \VTy) \to \VTy$; the state parameter $\vS$ has to be a value type,
+since it is an input to an object-level function.
+\begin{alignat*}{3}
+  &\instance \Improve\,\vF\,\vM \RA \Improve\,(\StateT\,\vS\,\vF)\,(\StateT_\M\,(\Up \vS)\,\vM)\,\where\\
+  &\ind \mup\,\vx = \StateT_\M \dlr \lam \vs.\,\mdo\\
+  &\ind\ind \mit{as} \fro \mup\,\qt{\runStateT\,\spl \vx\,\spl \vs}\\
+  &\ind\ind \case \mit{as}\,\,\of (\va,\,\vs) \to \return (\va,\,\vs)\\
+  &\ind \mdown\,\vx = \qt{ \StateT\,(\lam \vs.\, \spl(\mdown \dlr \mdo\\\
+  &\ind \ind (\va,\,\vs) \fro \runStateT_\M\,\vx\,\qt{\vs}\\
+  &\ind \ind \return \qt{(\spl \va,\, \spl \vs)}))}
+\end{alignat*}
+Like before in $\MaybeT$, we rely on object-level case splitting in the
+definition of $\mup$. For $\Reader$, the environment parameter also has to be a
+value type, and we define improvement as follows.
+\begin{alignat*}{3}
+  &\instance \Improve\,\vF\,\vM \RA \Improve\,(\ReaderT\,\vR\,\vF)\,(\ReaderT_\M\,(\Up \vR)\,\vM)\,\where\\
+  &\ind \mup\,\vx   = \ReaderT_\M \dlr \lam \vr.\, \mup\,\qt{\runReaderT\,\spl \vx\,\spl \vr}\\
+  &\ind \mdown\,\vx = \qt{\ReaderT\,(\lam \vr.\,\spl(\mdown\,(\runReaderT_\M\,\vx\,\qt{\vr})))}
+\end{alignat*}
+
+\subsubsection{$\State$ and $\Reader$ operations} If we try to use
+the $\modify$ function that we already have for $\State_\M$, a curious thing
+happens. The meaning of $\modify\,(\lam \vx.\,\qt{\spl \vx + \spl \vx})$ is to replace
+the current state $\vx$, as an object expression, with the expression $\qt{\spl \vx
+  + \spl \vx}$, and this happens at staging time. This behaves as an ``inline''
+modification which replaces every subsequent mention of the state with a
+different expression. For instance, ignoring newtype wrappers for now,
+\[ \mdown \dlr \mdo \{\modify\,(\lam \vx.\,\qt{\spl \vx + \spl \vx}); \modify\,(\lam \vx.\,\qt{\spl \vx + \spl \vx});\return\,\qt{()}\} \]
+is staged to
+\[ \qt{\lam \vx.\,((),\,(\vx + \vx) + (\vx + \vx))} \]
+which duplicates the evaluation of $\vx + \vx$. The solution is to force the
+evaluation of the new state in the object language, by let-insertion. A similar
+phenomenon happens with the $\mit{local}$ function in $\Reader$. So we define
+``stricter'' versions of these operations. We also return $\Up ()$ from
+actions instead of $()$ --- the former is more convenient, because the $\mdown$
+operation can be immediately used on it.
+\begingroup
+\allowdisplaybreaks
+\begin{alignat*}{3}
+  & \mput' : (\MonadState\,(\Up \vS)\,\vM,\,\MonadGen\,\vM) \RA \Up \vS \to \vM\,(\Up ()) \\
+  & \mput'\,\vs = \mdo \{\vs \fro \gen\,\vs; \mput\,\vs; \return \qt{()}\}\\
   &\\
-  &\ind \mdown\,(\MaybeT_\M\,x) = \qt{\MaybeT\,\spl(\mdown \dlr x \bind \lambda\case\\
-  &\ind\ind\Nothing_\M \to \return \qt{\Nothing}\\
-  &\ind\ind\Just_\M\,a\hspace{1.15em} \to \return \qt{\Just\,\spl a})}
-\end{alignat*}
-In the $\mbf{case}$ in $\mup$, we already use our syntax sugar for matching on a $\Maybe$
-inside an $\M$ action. This is legal, since we know from the $\ImproveF\,F\,M$ assumption
-that $M$ is a $\MonadGen$.
-
-In the meta level, we can reuse essentially all definitions from Haskell's
-monad transformer library $\msf{mtl}$. The additional work that we have to do in
-the CFTT setting:
+  & \modify' : (\MonadState\,(\Up \vS)\,\vM,\,\MonadGen\,\vM) \RA (\Up \vS \to \Up \vS) \to \vM\,(\Up ()) \\
+  & \modify'\,\vf = \mdo \{\vs \fro \get; \mput'\,(\vf\,\vs)\}\\
+  &\\
+  & \mit{local'} : (\MonadReader\,(\Up \vR)\,\vM,\,\MonadGen\,\vM) \RA (\Up \vR \to \Up \vR) \to \vM\,\vA \to \vM\,\vA\\
+  & \mit{local'}\,\vf\,\vma = \mdo \{\vr \fro \mit{ask}; \vr \fro \gen\,(\vf\,\vr);\mit{local}\,(\lam \_.\,\vr)\,\vma\}
+\end{alignat*}
+\endgroup
+Now,
+\[ \mdown \dlr \mdo \{\modify'\,(\lam \vx.\,\qt{\spl \vx + \spl \vx}); \modify'\,(\lam \vx.\,\qt{\spl \vx + \spl \vx})\} \]
+is staged to
+\[ \qt{\lam \vx.\,\letdef x := \vx + \vx; \letdef \vx := \vx + \vx; ((),\,\vx)} \]
+
+\subsection{Joining Control Flow in Monads}\label{sec:joining-control-flow-in-monads}
+
+There is a deficiency in our library so far. Assuming $\vb : \Bool$ as an object value, consider:
+\begin{alignat*}{3}
+  & \mdown \dlr \mdo \\
+  & \ind \case \vb\,\of\\
+  & \ind \ind \True  \hspace{0.2em}\to \mput' \qt{10}\\
+  & \ind \ind \False \to \mput' \qt{20}\\
+  & \ind \modify' (\lam \vx. \qt{\spl \vx + 10})
+\end{alignat*}
+The $\modify' (\lam \vx. \qt{\spl \vx + 10})$ action gets inlined to both $\mbf{case}$
+branches during staging. This follows from the definition of monadic binding in
+$\Gen$ and the $\mit{split}$ function in the desugaring of $\mbf{case}$. Code
+generation is continued in both branches with the same action. If we have
+multiple $\mbf{case}$ splits sequenced after each other, that yields exponential
+code size. Right now, we can fix this by let-binding the problematic action:
+\begin{alignat*}{3}
+  & \mdown \dlr \mdo \\
+  & \ind \vx \fro \gen \dlr \mdown \dlr \case \vb\,\of\\
+  & \ind \ind \True  \hspace{0.2em}\to \mput' \qt{10}\\
+  & \ind \ind \False \to \mput' \qt{20}\\
+  & \ind \mup\,\vx\\
+  & \ind \modify' (\lam \vx. \qt{\spl \vx + 10})
+\end{alignat*}
+This removes code duplication by round-tripping through an object-level
+$\mbf{let}$. This solution is fairly good in a state monad, since $\mdown$ only
+introduces a runtime pair constructor, and it is feasible to compile
+object-level pairs as unboxed data, without overheads. However, for $\Maybe$,
+$\mdown$ introduces a runtime $\Just$ or $\Nothing$, and $\mup$ introduces a
+runtime case split. A better solution would be to introduce two let-bound
+\emph{join points} before the offending $\mbf{case}$, one for returning a
+$\Just$ and one for returning $\Nothing$, but fusing away the actual runtime
+constructors. So, for example, we would like to produce object code which looks
+like the following:
+\begin{alignat*}{3}
+  & \letdef\,\mit{joinJust}\,\vn := ...\\
+  & \letdef\,\mit{joinNothing}\,() := ...\\
+  & \case \vx == 10\,\of\\
+  & \ind \True \hspace{0.2em}\to \mit{joinJust}\,(\vx + 10) \\
+  & \ind \False \to \mit{joinNothing}\,()
+\end{alignat*}
+Such fused returns are possible whenever we have a $\Gen\,\vA$ action at the
+bottom of the transformer stack, such that $\vA$ is isomorphic to a meta-level
+finite sum of value types. Recall that $\Gen\,\vA$ is defined as $\{\vR : \VTy\} \to
+(\vA \to \Up \vR) \to \Up \vR$. Here, if $\vA$ is a finite sum, we can rearrange $\vA \to
+\Up \vR$ to a finite product of functions.
+
+We could proceed with finite sums, but we will need finite
+\emph{sums-of-products} (SOP in short) later in Section in \ref{TODO}, so we
+develop SOP-s. We view SOP-s as a Tarski-style universe consisting of a type of
+descriptions, and a way of interpreting descriptions into $\MTy$ (``$\El$'' for
+``elements'' of the type).
+\begin{alignat*}{3}
+  &\USOP : \MTy                    &&\ind \El_\SOP : \USOP \to \MTy \\
+  &\USOP = \List\,(\List\,\VTy)    &&\ind \El_\SOP\,\vA = ...
+\end{alignat*}
+A type description is a list of lists of value types. We decode this to a sum of
+products of value types. We omit the definition here. $\SOP$ is closed under
+value types, finite product types and finite sum types. For instance, we have
+$\Either_\SOP : \SOP \to \SOP \to \SOP$ together with $\Left_\SOP : \El_\SOP\,\vA
+\to \El_\SOP\,(\Either_\SOP\,\vA\,\vB)$, $\Right_\SOP : \El_\SOP\,\vB \to
+\El_\SOP\,(\Either_\SOP\,\vA\,\vB)$ and a case splitting operation. It is more
+convenient to work with type formers in $\MTy$, and only convert to $\SOP$
+representations when needed, so we define a class for the representable types:
+\begin{alignat*}{3}
+  & \hspace{-5em}\mathrlap{\tyclass\,\IsSOP\,(\vA : \MTy)\,\where} \\
+  & \hspace{-5em}\ind \Rep      &&: \USOP \\
+  & \hspace{-5em}\ind \mit{rep} &&: \vA \simeq \El_\SOP\,\Rep
+\end{alignat*}
+Above, $\vA \simeq \Rep$ denotes a record containing a pair of back-and-forth
+functions, together with proofs (as propositional equalities) that they are
+inverses. We will overload $\mit{rep}$ as the forward conversion function with
+type $\vA \to \Rep$, and write $\mit{rep}^{-1}$ for its inverse. Now we define
+an isomorphic representation of $\El_\SOP\,\vA \to \Lift \vR$ as a product of
+object-level functions:
+\begin{alignat*}{3}
+  &\mathrlap{\mit{Fun}_{\SOP\Lift} : \USOP \to \Ty \to \MTy}\\
+  &\mit{Fun}_{\SOP\Lift}\,\Nil\,&& \vR = ()\\
+  &\mit{Fun}_{\SOP\Lift}\,(\Cons\,\vA\,\vB)\,&& \vR = (\Lift(\mit{foldr}\,(\to)\,\vR\,\vA),\,\mit{Fun}_{\SOP\Lift}\,\vB\,\vR)
+\end{alignat*}
+\begin{alignat*}{3}
+  &\mtabulate &&: (\El_\SOP\,\vA \to \Lift \vR) \to \mit{Fun}_{\SOP\Lift}\,\vA\,\vR\\
+  &\mindex    &&: \mit{Fun}_{\SOP\Lift}\,\vA\,\vR \to (\El_\SOP\,\vA \to \Lift \vR)
+\end{alignat*}
+We omit here the definitions of $\mtabulate$ and $\mindex$. We will also need to
+let-bind all functions in a $\mit{Fun}_{\SOP\Lift}$:
+\begin{alignat*}{3}
+  & \mathrlap{\mit{genFun}_{\SOP\Lift} : \{\vA : \USOP\} \to \mit{Fun}_{\SOP\Lift}\,\vA\,\vR \to \mit{Fun}_{\SOP\Lift}\,\vA\,\vR}\\
+  & \mit{genFun}_{\SOP\Lift}\,\{\Nil\}        && ()             &&= \return\,()\\
+  & \mit{genFun}_{\SOP\Lift}\,\{\Cons\,\_\,\vA\}&& (\vf,\,\mit{\vfs}) &&= ({,})\,{\fmap}\,\gen\,\vf\,{\ap}\,\mit{genFun}_{\SOP\Lift}\,\{\vA\}\,\vfs
+\end{alignat*}
+We introduce a class for monads that support control flow joining.
+\[ \tyclass\,\Monad\,\vM \RA \MonadJoin\,\vM\,\where\,\join : \IsSOP\,\vA \RA \vM\,\vA \to \vM\,\vA \]
+The most interesting instance is the ``base case'' for $\Gen$:
+\begin{alignat*}{3}
+  &\instance \MonadJoin\,\Gen\,\where\\
+  &\ind \join\,\vma = \Gen \dlr \lam \vk.\,\runGen \dlr \mdo\\
+  &\ind \ind \mit{joinPoints} \fro \mit{genFun}_{\SOP\Lift}\,(\mtabulate\,(\vk \circ \mit{rep}^{-1}))\\
+  &\ind \ind \va \fro \vma\\
+  &\ind \ind \return \dlr \mindex\,\mit{joinPoints}\,(\mit{rep}\,\va)
+\end{alignat*}
+Here we first convert $\vk : \vA \to \Lift \vR$ to a product of join points and
+let-bind each one of them. Then we generate code that returns to the appropriate
+join point for each return value. Other $\MonadJoin$ instances are straightforward. In
+$\StateT_\M$, we need the extra $\IsSOP\,\vS$ constraint because $\vS$ is returned
+as a result value.
+\begin{alignat*}{3}
+  & \instance\,(\MonadJoin\,\vM) \RA \MonadJoin\,(\MaybeT_\M\,\vM)\,\where\\
+  & \ind \join\,(\MaybeT_\M\,\vma) = \MaybeT_\M\,(\join\,\vma)\\
+  & \instance\,(\MonadJoin\,\vM) \RA \MonadJoin\,(\ReaderT_\M\,\vR\,\vM)\,\where\\
+  & \ind \join\,(\ReaderT_\M\,\vma) = \ReaderT_\M\,(\join \circ \vma)\\
+  & \instance\,(\MonadJoin\,\vM,\,\IsSOP\,\vS) \RA \MonadJoin\,(\StateT_\M\,\vS\,\vM)\,\where\\
+  & \ind \join\,(\StateT_\M\,\vma) = \StateT_\M\,(\join \circ \vma)
+\end{alignat*}
+
+Now, whenever the return value of a $\mbf{case}$-splitting action is a sum of
+values (this includes just returning an object value, which is by far the most
+common situation), we can use $\join \dlr \case \vx\,\of ...$ to eliminate code
+duplication, without creating runtime sum types.
+
+\todo{bigger example, non-tail-recursive list function with some effects, input, output, mentioned tail rec}
+
+\subsection{Discussion}
+
+So far, we have a monad transformer library with the following features:
 \begin{itemize}
-  \item Adding $\MonadGen$ instances, but these are trivially defined as
-        $\liftGen = \lift \circ \liftGen$ everywhere.
-  \item Defining the object-level counterparts of monad transformers, and
-        also the $\Improve$ instances.
+\item Almost all definitions from the well-known Haskell ecosystem of monads and monad transformers
+      can be directly reused, in the meta level.
+\item We can pattern match on object-level values in monadic code, insert object-level $\mbf{let}$-s
+      with $\gen$ and avoid code duplication with $\join$.
+\item In monadic code, object-level data constructors are only ever created by
+      $\mdown$, and switching on object-level data is only created by $\mit{split}$
+      and $\mup$. Monadic operations are fully fused, and all function calls can be
+      compiled to statically known saturated calls.
 \end{itemize}
-From $\msf{mtl}$, only the continuation monad transformer fails to support
-binding-time-improvement in CFTT, because of the obvious need for dynamic
-closures. In the following we present only $\ReaderT$ and $\StateT$.
-\begin{alignat*}{3}
-  &\instance \ImproveF\,F\,M \RA \Improve\,(\MaybeT\,F\,A)\,(\MaybeT_\M\,M\,(\Up A))\,\where\\
-  &\ind \mup\,x = \MaybeT_\M \dlr \mdo\\
-  &\ind\ind \vma \fro \mup\,\qt{\runMaybeT\,\spl x}\\
-  &\ind\ind \case \vma\,\of \Nothing \to \return \Nothing_\M\\
-  &\hspace{6.5em}         \Just\,a \hspace{1.15em}\to \return (\Just_\M\,a)\\
-  &\\
-  &\ind \mdown\,(\MaybeT_\M\,x) = \qt{\MaybeT\,\spl(\mdown \dlr x \bind \lambda\case\\
-  &\ind\ind\Nothing_\M \to \return \qt{\Nothing}\\
-  &\ind\ind\Just_\M\,a\hspace{1.15em} \to \return \qt{\Just\,\spl a})}
-\end{alignat*}
-
-%% instance (Improve m n) => Improve (StateT s m) (StateT (Up s) n) where
-
-%%   up x = StateT \s ->
-%%     do as <- up [|| runStateT $$x $$s ||]
-%%        caseGen as pure
-
-%%   down x = [|| StateT \s -> $$(down
-%%     do (a, s) <- runStateT x [||s||]
-%%        pure [||($$a, $$s)||]
-%%        )||]
-
-%% instance Improve m n => Improve (MaybeT m) (MaybeT n) where
-%%   up x = MaybeT do
-%%     ma <- up [||runMaybeT $$x||]
-%%     caseGen ma pure
-
-%%   down (MaybeT x) =
-%%     [|| MaybeT $$(down (x >>= \case
-%%           Nothing -> pure [||Nothing||]
-%%           Just a -> pure [||Just $$a||])) ||]
-
-%% instance Improve m n => Improve (ReaderT r m) (ReaderT (Up r) n) where
-%%   up   x = ReaderT \r -> up [||runReaderT $$x $$r||]
-%%   down x = [|| ReaderT \r -> $$(down (runReaderT x [||r||])) ||]
+
+As to potential weaknesses, first, the system as described in this section has
+some syntactic noise and requires extra attention from programmers.  We believe
+that the code noise can be mitigated very effectively in a native CFTT
+implementation. Kovács \cite{TODO} demonstrated in a prototype elaborator that
+almost all quotes and splices are unambiguously inferable in 2LTT programs, if
+we require that stages of let-definitions are always specified (as we do
+here). Moreover, $\mup$ and $\mdown$ should be also effectively inferable, using
+bidirectional elaboration. With stage inference and binding-time improvement
+inference, monadic code in CFTT would look only modestly more complicated than
+in Haskell.
+
+Second, in CFTT we cannot store computations (e.g.\ functions or $\State$
+actions) in runtime data structures, nor can we have computations in $\State$
+state or in $\Reader$ environments. However, it would be possible to extend CFTT
+with a closure type former that converts computations to values, in which case
+there is no such limitation anymore. Here, closure-freedom would be still
+available; we would be able to pick where to use or avoid the closure type
+former.
+
+\subsection{Agda \& Haskell Implementations}
+
+We implemented everything in this section in both Agda and typed Template
+Haskell. We summarize features and differences:
+\begin{itemize}
+\item The Haskell implementation can be used to generate code that can be
+      further compiled by GHC; here the object language is taken to be Haskell
+      itself. Since Haskell does not distinguish value and computation types, we do
+      not track them in the library, and we do not get guaranteed closure-freedom
+      from GHC. We do get high quality code though, typically.
+\item In Agda, we postulate all types and terms of the object theory in a
+  faithful way (i.e.\ equivalently to the CFTT syntax presented here), and take
+  Agda itself to be the metalanguage. Here, we can test ``staging'' by running Agda
+  programs which compute object expressions. However, we can only inspect
+  staging output and cannot compile or run object programs.
+\item For sums-of-products in Haskell, we make heavy use of \emph{singleton
+  types} \cite{TODO} to emulate dependent types. This adds significant
+  noise. Also, in $\IsSOP$ instances we can only define the conversion functions
+  and cannot prove that they are inverses, because Haskell does not have enough support
+  for dependent types.
+\end{itemize}
+
+%% We implemented all features in this section in typed Template Haskell
+%% \cite{TODO}, using only the strongly typed interface. The organization of the
+%% code is essentially the same as here, but some changes are necessary:
+%% \begin{itemize}
+%%   \item TH cannot properly handle scoping of type variables across quotations,
+%%         generating ill-typed code in some cases. It is fairly easy to work around
+%%         this using term-level proxy values.
+%%   \item We do not bother tracking value and computation types. This could be done,
+%%         using dummy type classes, but we would not get any extra guarantees from it
+%%         in downstream compilation.
+%%   \item When we take the product of generic sums in CFTT, we simply take the
+%%         object-level product of component types, pairwise. We do this because we
+%%         assume that the downstream compiler can represent products without
+%%         overheads, in an unboxed manner, and also erase unit types. In Haskell we do
+%%         not rely on this, and instead use a slightly more complicated
+%%         sum-of-products-of-values representation, and also fuse away product types
+%%         during staging.
+%%   \item We make heavy use of \emph{singleton types} \cite{TODO} in generic sums-of-products, to emulate
+%%         dependent types. This adds significant noise compared to a native dependently typed
+%%         version.
+%%   \item We have two separate type classes for binding-time improvement, one for
+%%         base types and one for monads specifically. This is because the quantified
+%%         constraints in \ref{TODO} yield poor type inference for $\mup$ and $\mdown$,
+%%         by preventing \emph{functional dependencies} \cite{TODO} between the class
+%%         parameters. We expect that in a native CFTT implementation, the extra staging
+%%         information in type universes would help inference in this case.
+%% \end{itemize}
+
+\section{Stream Fusion}
+
+Stream fusion refers to a collection of techniques for generating efficient code
+from declarative definitions involving streams of values, where intermediate
+data structures are eliminated. Stream fusion can be broadly grouped into \emph{push}
+fusion, which is based on Church-encodings of inductive lists, and $\emph{pull}$
+fusion, which is based on Church-encodings of coinductive lists \cite{TODO}. The
+two styles have different trade-offs, and in practical programming it is a good
+idea to support both, but in this section we focus on pull streams.
+
+The reason is that pull streams have been historically more difficult to
+efficiently compile, and we can demonstrate significant improvements in CFTT.
+We also use dependent types in a more essential way than in the previous
+section.
+
+%% Fusion techniques can be broadly grouped into \emph{push} fusion, which is based
+%% on Church-encodings of inductive lists, and $\emph{pull}$ fusion, which is based
+%% on Church-encodings of coinductive lists, which manifests as \emph{state
+%% machines} \cite{TODO}. The two styles have different trade-offs, and in
+%% practical programming it is a good idea to support both, but in this section we
+%% focus on pull fusion.
+
+%% The reason is partly lack of space, but also that pull fusion has been
+%% historically the more difficult to reliably compile, and we can demonstrate
+%% significant improvements in CFTT, in robustness and expressive power. Our fusion
+%% always succeeds, and we also support a $\mit{concatMap}$ operation, which
+%% has been elusive in guaranteed fusion. We believe that our implementation also
+%% provides valuable conceptual clarity.
+
+\subsection{Streams}
+
+A pull stream is a meta-level specification of a state machine:
+\begin{alignat*}{3}
+  & \data \Step\,\vS\,\vA = \Stop\,|\,\Skip\,\vS\,|\,\Yield\,\vA\,\vS\\
+  & \data \Pull\,(\vA : \MTy) : \MTy\,\where\\
+  & \ind \Pull : (\vS : \MTy) \to \IsSOP\,\vS \RA \vS \to (\vS \to \Gen\,(\Step\,\vS\,\vA)) \to \Pull\,\vA
+\end{alignat*}
+In the $\Pull$ constructor, $\vS$ is the type of the internal state which is
+required to be a sum-of-products of value types by the $\IsSumVS\,\vS$
+constraint. The next field with type $S$ is the initial state, while transitions
+are represented by the $\vS \to \Gen\,(\Step\,\vS\,\vA)$ field. Possible transitions
+are stopping ($\Stop$), transitioning to a new state while outputting a value
+($\Yield$) and making a silent transition to a new state ($\Skip$).
+
+Let us see some operations on streams now. First, $\Pull$ is a ``zippy''
+applicative functor, where $\pure$ infinitely repeats a value and $(\!\ap\!)$
+applies a stream of functions to a stream of values, pointwise.
+\begin{alignat*}{3}
+  &\mit{pure} : \vA \to \Pull\,\vA\\
+  &\mit{pure} = \Pull\,()\,()\,(\lam \_.\,\return \dlr \Stop)
+\end{alignat*}
+\begin{alignat*}{3}
+  &(\!\ap\!) : \Pull\,(\vA \to \vB) \to \Pull\,\vA \to \Pull\,\vB \\
+  &(\!\ap\!)\,(\Pull\,\vS\,\seed\,\step)\,(\Pull\,\vS'\,\seed'\,\step') =\\
+  & \ind \Pull\,(\vS,\,\vS')\,(\seed,\,\seed') \dlr \lam (\vs,\,\vs').\,\step\,\vs \bind \lam \case \\
+  & \ind \ind \Stop \hspace{1.20em}        \to \return \Stop \\
+  & \ind \ind \Skip\,\vs \hspace{0.75em}     \to \return \dlr \Skip\,(\vs,\,\vs') \\
+  & \ind \ind \Yield\,\vf\,\vs \to \step'\,\vs' \bind \lam \case \\
+  & \ind \ind \ind \Stop \hspace{1.7em} \to \return \Stop \\
+  & \ind \ind \ind \Skip\,\vs'\hspace{0.9em}\to \return \dlr \Skip\,(\vs,\,\vs') \\
+  & \ind \ind \ind \Yield\,\va\,\vs' \to \return \dlr \Yield\,(\vf\,\va)\,(\vs,\,\vs')
+\end{alignat*}
+In $\mit{pure}$, the state is the unit type, while in $(\!\ap\!)$ we take the
+product of the stream states and do synchronous transitions. In both
+definitions, the $\IsSOP\,S$ constraint is implicitly dispatched by instance
+resolution; this works in the Agda and Haskell versions too. We can derive the
+usual $\mit{zip}$ and $\mit{zipWith}$ operations from $(\!\ap\!)$. $\Pull$
+is also a monoid with stream appending as the binary operation.
+\begin{alignat*}{3}
+  & \mit{empty} : \Pull\,\vA\\
+  & \mit{empty} = \Pull\,()\,()\,(\lam \_.\,\return \Stop)
+\end{alignat*}
+\begin{alignat*}{5}
+  & \hspace{-4em}\mathrlap{(\!<>\!) : \Pull\,\vA \to \Pull\,\vA \to \Pull\,\vA} \\
+  & \hspace{-4em}\mathrlap{(\!<>\!)\,(\Pull\,\vS\,\seed\,\step)\,(\Pull\,\vS'\,\seed'\,\step') = \Pull\,(\Either\,\vS\,\vS')\,(\Left\,\seed)\,\lam \case} \\
+  & \hspace{-4em}\ind \Left\,\vs &&\to \step\,s \bind \lam \case &&\Stop             &&\to \return \dlr \Skip\,(\Right\,\seed')\\
+  & \hspace{-4em}                    &&                          &&\Skip\,\vs        &&\to \return \dlr \Skip\,(\Left\,\vs) \\
+  & \hspace{-4em}                    &&                        &&\Yield\,\va\,\vs    &&\to \return \dlr \Yield\,\va\,(\Left\,\vs)\\
+  & \hspace{-4em}\ind \Right\,\vs' &&\to \step\,s' \bind \lam \case &&\Stop          &&\to \return \Stop\\
+  & \hspace{-4em}                    &&                        &&\Skip\,\vs'         &&\to \return \dlr \Skip\,(\Right\,\vs') \\
+  & \hspace{-4em}                    &&                        &&\Yield\,\va\,\vs'   &&\to \return \dlr \Yield\,\va\,(\Right\,\vs')
+\end{alignat*}
+These definitions are standard for streams; note though that compared to the
+unstaged definitions in previous literature, the only additional noise is just
+the $\Gen$ monad in the transitions. Likewise, we can give standard definitions
+for the usual stream functions such as $\msf{filter}$, $\msf{take}$ or
+$\msf{drop}$.
+
+\subsection{Running Streams with Mutual Recursion}
+
+How do we generate object code from streams? The $\vS$ state is given as a
+finite sums of products, but the sums and the products are on the meta
+level, so we cannot directly use $\vS$ in object code. Similarly as in the the
+treatment of join points, we tabulate the $\vS \to \Gen\,\Step\,\vS\,\vA$
+transition function to a product of functions. However, these functions need to
+be \emph{mutually recursive}, since it is possible to transition from any state
+to any other state, and each such transition is represented as a function call.
+The problem of generating well-typed mutual blocks was addressed by Yallop and
+Kiselyov in \cite{TODO}. Unlike ibid., which used control effects and mutable
+references in MetaOCaml, we present a solution that does not use side effects
+in the metalanguage.
+
+The solution is to extend $\CTy$ with finite products of computations,
+i.e.\ assume $() : \CTy$ and $({,}) : \CTy \to \CTy \to \CTy$, together with
+pairing and projections. These types, like functions, are call-by-name in the
+runtime semantics, and they also cannot escape the scope of their definition.
+Hence, we can also ``saturate'' programs that involve computational product types:
+every computation definition at type $(\vA,\,\vB)$ can be translated to a pair,
+and every projection of a let-defined variable can be statically matched up with
+a pairing in the variable definition. Thus, a recursive let-definition at type
+$(\vA \to \vB,\,\vA \to \vB)$ can be always compiled to a pair of mutual
+functions.
+
+We redefine the previous $\mit{Fun}_{\SOP\Lift}$ to return a computation type
+instead:
+\begin{alignat*}{3}
+  &\mathrlap{\mit{Fun}_{\SOP\Lift} : \USOP \to \Ty \to \CTy}\\
+  &\mit{Fun}_{\SOP\Lift}\,\Nil\,&& \vR = ()\\
+  &\mit{Fun}_{\SOP\Lift}\,(\Cons\,\vA\,\vB)\,&& \vR = (\mit{foldr}\,(\to)\,\vR\,\vA,\,\mit{Fun}_{\SOP\Lift}\,\vB\,\vR)
+\end{alignat*}
+\begin{alignat*}{3}
+  &\mtabulate &&: (\El_\SOP\,\vA \to \Lift \vR) \to \Lift(\mit{Fun}_{\SOP\Lift}\,\vA\,\vR)\\
+  &\mindex    &&: \Lift(\mit{Fun}_{\SOP\Lift}\,\vA\,\vR) \to (\El_\SOP\,\vA \to \Lift \vR)
+\end{alignat*}
+For join points, this is just as efficient as the previous
+$\mit{Fun}_{\SOP\Lift}$, since a definition of a product of functions gets
+compiled to a sequence of function definitions. Now, we can generate object code
+from streams. There are several choices for this, but in CFTT the $\msf{foldr}$
+function is as good as we can get.
+\begin{alignat*}{3}
+  &\hspace{-13em}\mathrlap{\foldr : \{\vA : \MTy\}\{\vB : \Ty\} \to (\vA \to \Lift \vB \to \Lift \vB) \to \Lift \vB \to \Pull\,\vA \to \Lift\,\vB}\\
+  &\hspace{-13em}\mathrlap{\foldr\,\{\vA\}\,\{\vB\}\,(\Pull\,\vS\,\seed\,\step)\,\vf\,\vb = \ql}\\
+  &\hspace{-13em}\mathrlap{\ind \letrec \vfs : \mit{Fun}_{\SOP\Lift}\,(\Rep\,\{\vS\})\,\vB := \spl\big(\mtabulate \dlr \lam \vs.\,\unGen\,(\step\,(\rep^{-1}\,\vs)) \dlr \lam \case} \\
+  &\hspace{-13em}\ind \ind \Stop            &&\to \vb \\
+  &\hspace{-13em}\ind \ind \Skip\,\vs       &&\to \mindex\,\qt{\vfs}\,(\rep\,\vs) \\
+  &\hspace{-13em}\ind \ind \Yield\,\va\,\vs &&\to \vf\,\va\,(\mindex\,\qt{\vfs}\,(\rep\,\vs))\big);\\
+  &\hspace{-13em}\mathrlap{\ind \spl(\mindex\,\qt{\vfs}\,(\rep\,\seed))\qr}
+\end{alignat*}
+This $\foldr$ is quite flexible, becuse we can eliminate into any object type,
+including function types. For instance, we can define $\foldl$ from $\foldr$:
+\begin{alignat*}{3}
+  &\foldl : \{\vA : \MTy\}\{\vB : \VTy\} \to (\Lift \vB \to \vA \to \Lift \vB) \to \Lift \vB \to \Pull\,\vA \to \Lift\,\vB\\
+  &\foldl\,\vf\,\vb\,\vas = \spl(\foldr\,(\lam \va\,\vg.\,\qt{\lam \vb.\,\spl \vg\,\spl(\vf\,\qt{\vb}\,\spl \va)})\,\qt{\lam \vb.\,\vb}\,\vas)\,\spl \vb
+\end{alignat*}
+Note that since we abstract $\vB$ in a runtime function, it must be a value
+type. Here, each $\Stop$ and $\Yield$ in the transition function gets
+interpreted as a $\lambda$-expression in the output. However, those $\lambda$-s
+will be lifted out in the scope, yielding a proper mutually tail-recursive
+definition with an accummulator for $\vB$. Contrast this to the GHC base
+library, where $\foldl$ for lists is also defined from $\foldr$, to enable
+push-based fusion, but where a substantial \emph{arity analysis} is required in
+the compiler to eliminate the intermediate closures \cite{TODO}.
+
+\subsection{$\mbf{concatMap}$ for Streams}
+
+We saw that $\Pull$ is $\Applicative$, but what about having a list-like
+$\Monad$ instance as well, with singleton streams for $\return$ and $\concatMap$
+for binding? This is not possible. If we have
+\[ \concatMap : (\vA \to \Pull\,\vB) \to \Pull\,\vA \to \Pull\,\vB, \]
+that means that $\vA \to \Pull\,\vB$ can contain an infinite number of different
+machine state types, which can never be represented in a finite amount of object
+code.\footnote{Although in two-level type theories that are not used for
+staging, the representability of countably infinite products of object types is
+sometimes assumed as an axiom; this axiom is called ``cofibrancy of
+$\mathbb{N}$'' \cite{TODO}.} Here by ``infinite'' we mean the notion that is
+internally available in the meta type-theory. For instance, we can define a
+$\Nat_\M \to \Pull\,\vB$ function which for each $\vn : \Nat_\M$ produces a
+concatenation of $\vn$ streams. Hence, we shall have the following function
+instead:
+\[ \concatMap : \IsSOP\,\vA \RA (\vA \to \Pull\,\vB) \to \Pull\,\vA \to \Pull\,\vB, \]
+The idea is the following: if $\USOP$ is closed under dependent sum types, we
+can directly define this $\concatMap$, by taking the appropriate dependent sum
+of the $\vA \to \USOP$ family of machine states, which we extract from the $\vA
+\to \Pull\,\vB$ function. Let us write $\Sigma\,\vA\,\vB : \MTy$ for dependent
+sums, for $\vA : \MTy$ and $\vB : \vA \to \MTy$, and reuse $({,})$ for pairing
+and $\fst$ and $\snd$ for projections. We use the following field projection functions:
+$\mit{projS} : \Pull\,\vA \to \MTy$, $\mit{projSeed} : (\vas : \Pull\,\vA) \to
+\mit{projS}\,\vas$, and $\mit{projStep} : (\vas : \Pull\,\vA) \to
+\mit{projS}\,\vas \to \Gen\,(\Step\,(\mit{projS}\,\vas)\,\vA)$.
+For now, we assume that the following instance exists:
+\[ \instance (\IsSOP\,\vA,\,\{\va : \vA\} \to \IsSOP\,(\vB\,\va)) \RA \IsSOP\,(\Sigma\,\vA\,\vB) \]
+Above, $\{\va : \vA\} \to \IsSOP\,(\vB\,\va)$ is a universally quantified
+instance constraint; this is available in Agda and, less generally, in Haskell
+\cite{TODO}. The definition of $\concatMap$ is as follows.
+\begin{alignat*}{3}
+  & \concatMap : \IsSOP\,\vA \RA (\vA \to \Pull\,\vB) \to \Pull\,\vA \to \Pull\,\vB \\
+  & \concatMap\,\{\vA\}\,\{\vB\}\,\vf\,(\Pull\,\vS\,\seed\,\step) = \\
+  & \ind \Pull\,(\vS,\,\Maybe\,(\Sigma\,\vA\,(\mit{projS} \circ \vf)))\,(\seed,\,\Nothing) \dlr \lam \case \\
+  & \ind \ind (\vs,\,\Nothing) \to \step\,s \bind \lam \case \\
+  & \ind \ind \ind \Stop      \hspace{1.4em}   \to \return \Stop\\
+  & \ind \ind \ind \Skip\,\vs \hspace{0.9em} \to \return \dlr \Skip\,(\vs,\, \Nothing)\\
+  & \ind \ind \ind \Yield\,\va\,\vs          \to \return \dlr \Skip\,(\vs,\, \Just\,(\va,\,\mit{projSeed}\,(\vf\,\va)))\\
+  & \ind \ind (\vs,\,\Just\,(\va,\,\vs')) \to \mit{projStep}\,(\vf\,\va)\,\vs' \bind \lam \case \\
+  & \ind \ind \ind \Stop       \hspace{1.75em}\to \return \dlr \Skip\,(\vs,\,\Nothing) \\
+  & \ind \ind \ind \Skip\,\vs' \hspace{0.94em}\to \return \dlr \Skip\,(\vs,\,\Just\,(\va,\,\vs'))\\
+  & \ind \ind \ind \Yield\,\vb\,\vs' \to \return \dlr \Yield\,\vb\,(\vs,\, \Just\,(\va,\,\vs'))\\
+\end{alignat*}
+Here, $\Nothing$ marks the states where we are in the ``outer'' loop, trying to
+run the $\Pull\,\vA$ stream until we get the next value. $\Just$ marks the
+states of the ``inner'' loop, where we have a concrete $\va : \vA$ value and
+we run the $\vf\,\va$ stream until it ends. In the inner loop, the machine state
+type depends on the $\va : \vA$ value, hence the need for $\Sigma$.
+
+How do we get $\IsSOP$ for $\Sigma$? The key observations are:
+\begin{itemize}
+ \item Metaprograms cannot inspect the structure of object terms.
+ \item Object types do not depend on object terms.
+\end{itemize}
+Hence, we expect that during staging, every $\vf : \Lift\,\vA \to \VTy$ function
+has to be constant. This is actually true in the staging semantics of CFTT. In
+the semantics, all metafunctions are stable under object-level parallel
+substitutions. Also, object types are untouched by substitution. Hence, a
+straightforward unfolding of the definitions in the staging semantics validates
+the constancy of $\vf$.
+
+Generally speaking, every function whose domain is a product of object types and
+whose codomain is a constant presheaf in the semantics, is a constant function
+in the the semantics. We may call these constancy statements \emph{generativity
+axioms}, since they reflect the inability of metaprograms to inspect terms, and
+``generativity'' refers to this property in the staging literature \cite{TODO}.
+
+Let us write $\Uprod = \List\,\VTy$ and $\Elprod : \Uprod \to \MTy$ for a
+universe of finite products of value types. Concretely in CFTT and the Agda
+implementation, we assume the following:
+\\\\
+\noindent \textbf{Axiom (generativity).}\textit{ Every $\vf : \Elprod\,\vA \to \USOP$ is constant. }
+\\\\
+We remark that there is no risk of staging getting \emph{stuck} on this axiom,
+because propositional equality proofs get \emph{erased} in the staging semantics,
+as we noted in Section \ref{TODO}.
+
+We also remark that although generativity is inconsistent with inspecting the
+structure of object terms, it is consistent with inspecting the structure of
+object types.
+
+From generativity, we derive $\Sigma_\SOP : (\vA : \USOP) \to (\vB : \El_\SOP\,\vA \to \USOP)
+\to \USOP$ as follows. First, for each $\vA : \Uprod$, we define $\msf{loop_{\vA}} :
+\Elprod\,\vA$ as a product of non-terminating object programs. This is only needed
+to get arbitrary inhabitants with which we can call $\Elprod\,A \to \USOP$
+functions. Conveniently, every object type is inhabited by infinite loops.
+
+Then, $\Sigma_\SOP\,\vA\,\vB$ is defined as the concatenation of $\vA_i \times
+\vB\,(\msf{inject_i}\,\msf{loop_{\vA_i}})$ for each $\vA_i \in \vA$, where
+$({\times})$ is the product type former in $\USOP$ and $\msf{inject_i} :
+\Elprod\,\vA_i \to \El_\SOP\,\vA$. This is similar to the definition of
+non-dependent products in $\USOP$, except that we have to get rid of the type
+dependency by instantiating $\vB$ with looping programs.
+
+Then, we can show using the generativity axiom that $\Sigma_\SOP$ supports
+projections, pairing and the $\beta\eta$-rules. These are all needed when we
+define the $\IsSOP$ instance, when we have to prove that encoding via $\rep$
+is an isomorphism. Concretely, assuming $\IsSOP\,A$ and $\{\va : \vA\} \to
+\IsSOP\,(\vB\,\va)\}$, we define $\Rep$ for $\Sigma\,\vA\,\vB$ as follows:
+\[ \Rep\,\{\Sigma\,\vA\,\vB\} = \Sigma\,(\Rep\,\{\vA\})\,(\lam \vx.\,\Rep\,\{\vB\,(\rep^{-1}\,\vx)\}) \]
+Then, the definition of encoding for $\Sigma$ is only well-typed up to the fact
+that $\rep^{-1} \circ \rep = \id$:
+\[  \rep\{\Sigma\,\vA\,\vB\}\,(\va,\,\vb) = (\rep\,\{\vA\}\,\va,\,\rep\,\{\vB\,(\rep^{-1}\,(\rep\,\vx))\}\,\vb) \]
+This is, in fact, our reason for including the isomorphism equations as well in
+$\IsSOP$. This, in turn, necessitates using SOP instead of finite sums. We can
+define a product type former for finite sums of value types, by taking the
+pairwise products of components. However, we can only take the
+\emph{object-level} products here, and since object-level products have no
+$\beta\eta$-rules, we cannot prove $\beta\eta$ for the derived product type in
+finite sums, and likewise for the derived $\Sigma$-type. When we use SOP
+instead, we do not have to use object-level products and this issue does not
+appear.
+
+We do not detail the full definition of $\Sigma_\SOP$ here. The reader may refer
+to the SOP module in the Agda implementation, which is altogether 260 lines with
+all $\IsSOP$ instances.
+
+\subsection{Let-Insertion and Case Splitting}
+
+Since $\Pull$ is not a monad, it cannot be $\MonadGen$ in our framework,
+but we still support some forms of let-insertion. First, we note that $\Pull$ can be
+mapped over by functions returning in $\Gen$:
+\begin{alignat*}{3}
+  &\hspace{-3em}\mathrlap{\mit{mapGen} : \Pull\,\vA \to (\vA \to \Gen\,\vB) \to \Pull\,\vB }\\
+  &\hspace{-3em}\mathrlap{\mit{mapGen}\,(\Pull\,\vS\,\seed\,\step)\,\vf = \Pull\,\vS\,\seed \dlr \lam \case}\\
+  &\hspace{-3em} \ind \Stop            &&\to \return\,\Stop \\
+  &\hspace{-3em} \ind \Skip\,\vs       &&\to \return \dlr \Skip\,\vs\\
+  &\hspace{-3em} \ind \Yield\,\va\,\vs &&\to \mdo \{\vb \fro \vf\,\va; \return \dlr \Yield\,\vb\,\vs\}
+\end{alignat*}
+Now, for some $\vas : \Pull\,(\Lift \vA)$, we may use
+\[ \mit{mapGen}\,\vas\,\gen : \Pull\,(\Lift \vA) \]
+to let-bind all yielded expressions in the stream. This can be useful when some
+stream operations duplicate yielded values. For example, the function with type
+$\Pull\,\vA \to \Pull\,(\vA,\,\vA)$ is often used to avoid duplicating whole streams,
+and when $\vA$ is an object type we can use let-insertion in the definition to also
+avoid duplicating object expressions.
+
+Let us now consider a more ``monadic'' style of stream programming. For this we
+need singleton streams:
+\begin{alignat*}{3}
+  & \mit{single} : \vA \to \Pull\,\vA \\
+  & \mit{single}\,\va = \Pull\,\Bool_\M\,\True \dlr \lam \vb.\,\return \dlr\mbf{if}\,\vb\,\mbf{then}\,\Yield\,\va\,\False\,\mbf{else}\,\Stop
+\end{alignat*}
+We also write $\mit{forEach}$ for $\concatMap$ with flipped arguments. Now, a
+more monadic flavor of let-insertion is given as
+\begin{alignat*}{3}
+  &\gen_{\Pull} : \Lift \vA \to \Pull\,(\Lift \vA) \\
+  &\gen_{\Pull}\,\va = \mit{mapGen}\,(\mit{single}\,\va)\,\gen.
+\end{alignat*}
+We can also implement case splitting on object values. Recall from Section
+\ref{TODO} that for each value type we have a splitting function that returns a
+meta-level view on the constructors. Importantly, all such views are $\SOP$-s, and
+therefore can be bound by $\forEach$. A small example:
+\begin{alignat*}{3}
+  & \hspace{-5em}\mathrlap{\forEach\,(\mit{countFrom}\,\qt{0}) \dlr \lam \vn. }\\
+  & \hspace{-5em}\mathrlap{\forEach\,(\gen_{\Pull}\,\qt{\spl \vn * 2}) \dlr \lam \vm. }\\
+  & \hspace{-5em}\mathrlap{\forEach\,(\single\,(\msplit\,\qt{\spl \vm < 20})) \dlr \lam \case }\\
+  & \hspace{-5em}\ind \True_\M  &&\to \single\,\qt{\spl \vm + 10}   \\
+  & \hspace{-5em}\ind \False_\M &&\to \single\,\qt{\spl \vm + 20}
+\end{alignat*}
+Here $(\gen_{\Pull}\,\qt{\spl \vn * 2})$ behaves like a let-insertion, and
+$(\single\,(\msplit\,\qt{\spl \vm < 20}))$ implements the case splitting.
+
+\subsection{Discussion}
+
+Our stream library has a fairly strong support for programming in a monadic
+style, even though $\Pull$ is not literally a monad. We can bind object values
+with $\concatMap$, and we can also do let-insertion and case splitting for
+them. We also get guaranteed well-typing, closure-freedom, and arbitrary mixing
+of zipping and $\concatMap$.
+
+We highlight the usage of the generativity axiom as well. Previously in staged
+compilation, intensional analysis (i.e. the ability to analyze object code) has
+been viewed as a desirable feature that increases the expressive power of the
+system. To our knowledge, our work is the first one that exploits the
+\emph{lack} of intensional analysis in metaprogramming. This is a bit similar to
+parametricity in type theories, where the inability to analyze types has a
+payoff in the form of ``free theorems'' \cite{TODO}.
+
+Regarding the practical application of our stream library, we think that it
+would make sense to support both push and pull fusion in a realistic
+implementation, and allow users to benefit from the strong points of both. Push
+streams, which we do not present in this paper, have proper $\Monad$ and
+$\MonadGen$ instances and are often more convenient to use in CFTT. They are
+also better for deep traversals of structures where they can utilize unbounded
+stack allocations, while pull streams need heap allocations for unbounded space
+in the machine state.
+
+\subsection{Agda \& Haskell Implementations}
+
+The Agda implementation follows this section very closely, so we do not detail it.
+
+In the Haskell, there are some limitations. First, in $\concatMap$, the
+projection function $\mit{projS}$ cannot be defined, nbecause Haskell is not
+dependently typed, and the other field projections are also out of reach. We
+only have a weaker ``positive'' recursion principle for existential types. It
+might be the case that a strongly typed $\concatMap$ is possible with only weak
+existentials, but we attempted this and found that it introduces too much
+technical complication.
+
+So instead of giving a single generic definition for $\concatMap$ for $\IsSOP$
+types, we define $\concatMap$ just for object types,\footnote{Recall that we do
+not distinguish value and computation types in Haskell.} and then define a case
+splitting function separately for each type. In each of these functions, we only
+need to deal with a concrete finite number of different machine state types,
+which is feasible with weak existentials. For example, we define
+\[ \mit{split_{\Pull}} : \Lift (\List\,\vA) \to (\Maybe_\M\,(\Lift \vA,\,\Lift (\List\,\vA)) \to \Pull\,\vB) \to \Pull\,\vB. \]
+Also, the generativity axiom is \emph{false} in Template Haskell, since it is
+possible to look inside quoted expressions. Instead, we use type coercions that
+can fail at staging time, when users of the library violate generativity.
 

### 2024-02-20 newpaper/paper.tex
wip

@@
 Now, whenever the return value of a $\mbf{case}$-splitting action is a sum of
-values (this includes just returning an object value, which is by far the most
-common situation), we can use $\join \dlr \case \vx\,\of ...$ to eliminate code
-duplication, without creating runtime sum types.
+products of values (this includes just returning an object value, which is by
+far the most common situation), we can use $\join \dlr \case \vx\,\of ...$ to
+eliminate code duplication, without creating runtime sum types.
 
@@
 Let us see some operations on streams now. First, $\Pull$ is a ``zippy''
-applicative functor, where $\pure$ infinitely repeats a value and $(\!\ap\!)$
+applicative functor, where $\mit{pure}$ infinitely repeats a value and $({\ap})$
 applies a stream of functions to a stream of values, pointwise.
@@
 product of the stream states and do synchronous transitions. In both
-definitions, the $\IsSOP\,S$ constraint is implicitly dispatched by instance
+definitions, the $\IsSOP\,\vS$ constraint is implicitly dispatched by instance
 resolution; this works in the Agda and Haskell versions too. We can derive the
@@
 the $\Gen$ monad in the transitions. Likewise, we can give standard definitions
-for the usual stream functions such as $\msf{filter}$, $\msf{take}$ or
+for usual stream functions such as $\msf{filter}$, $\msf{take}$ or
 $\msf{drop}$.
@@
 to any other state, and each such transition is represented as a function call.
-The problem of generating well-typed mutual blocks was addressed by Yallop and
+This problem of generating well-typed mutual blocks was addressed by Yallop and
 Kiselyov in \cite{TODO}. Unlike ibid., which used control effects and mutable
@@
 $\Monad$ instance as well, with singleton streams for $\return$ and $\concatMap$
-for binding? This is not possible. If we have
+for binding? This is not possible. Aiming at
 \[ \concatMap : (\vA \to \Pull\,\vB) \to \Pull\,\vA \to \Pull\,\vB, \]
-that means that $\vA \to \Pull\,\vB$ can contain an infinite number of different
-machine state types, which can never be represented in a finite amount of object
+the $\vA \to \Pull\,\vB$ function can contain an infinite number of different
+machine state types, which cannot be represented in a finite amount of object
 code.\footnote{Although in two-level type theories that are not used for
@@
 $\mathbb{N}$'' \cite{TODO}.} Here by ``infinite'' we mean the notion that is
-internally available in the meta type-theory. For instance, we can define a
+internally available in the meta type theory. For instance, we can define a
 $\Nat_\M \to \Pull\,\vB$ function which for each $\vn : \Nat_\M$ produces a
@@
 \to \Pull\,\vB$ function. Let us write $\Sigma\,\vA\,\vB : \MTy$ for dependent
-sums, for $\vA : \MTy$ and $\vB : \vA \to \MTy$, and reuse $({,})$ for pairing
-and $\fst$ and $\snd$ for projections. We use the following field projection functions:
-$\mit{projS} : \Pull\,\vA \to \MTy$, $\mit{projSeed} : (\vas : \Pull\,\vA) \to
-\mit{projS}\,\vas$, and $\mit{projStep} : (\vas : \Pull\,\vA) \to
-\mit{projS}\,\vas \to \Gen\,(\Step\,(\mit{projS}\,\vas)\,\vA)$.
-For now, we assume that the following instance exists:
+sums, for $\vA : \MTy$ and $\vB : \vA \to \MTy$, and reuse $({,})$ for pairing.
+We also use the following field projection functions: $\mit{projS} : \Pull\,\vA
+\to \MTy$, $\mit{projSeed} : (\vas : \Pull\,\vA) \to \mit{projS}\,\vas$, and
+$\mit{projStep} : (\vas : \Pull\,\vA) \to \mit{projS}\,\vas \to
+\Gen\,(\Step\,(\mit{projS}\,\vas)\,\vA)$.  For now, we assume that the following
+instance exists:
 \[ \instance (\IsSOP\,\vA,\,\{\va : \vA\} \to \IsSOP\,(\vB\,\va)) \RA \IsSOP\,(\Sigma\,\vA\,\vB) \]
@@
 states of the ``inner'' loop, where we have a concrete $\va : \vA$ value and
-we run the $\vf\,\va$ stream until it ends. In the inner loop, the machine state
+we run the $(\vf\,\va)$ stream until it stops. In the inner loop, the machine state
 type depends on the $\va : \vA$ value, hence the need for $\Sigma$.
@@
 \begin{alignat*}{3}
-  &\hspace{-3em}\mathrlap{\mit{mapGen} : \Pull\,\vA \to (\vA \to \Gen\,\vB) \to \Pull\,\vB }\\
-  &\hspace{-3em}\mathrlap{\mit{mapGen}\,(\Pull\,\vS\,\seed\,\step)\,\vf = \Pull\,\vS\,\seed \dlr \lam \case}\\
+  &\hspace{-3em}\mathrlap{\mit{mapGen} : (\vA \to \Gen\,\vB) \to \Pull\,\vA \to \Pull\,\vB }\\
+  &\hspace{-3em}\mathrlap{\mit{mapGen}\,\vf\,(\Pull\,\vS\,\seed\,\step) = \Pull\,\vS\,\seed \dlr \lam \case}\\
   &\hspace{-3em} \ind \Stop            &&\to \return\,\Stop \\
@@
 \end{alignat*}
-Now, for some $\vas : \Pull\,(\Lift \vA)$, we may use
-\[ \mit{mapGen}\,\vas\,\gen : \Pull\,(\Lift \vA) \]
-to let-bind all yielded expressions in the stream. This can be useful when some
-stream operations duplicate yielded values. For example, the function with type
-$\Pull\,\vA \to \Pull\,(\vA,\,\vA)$ is often used to avoid duplicating whole streams,
-and when $\vA$ is an object type we can use let-insertion in the definition to also
-avoid duplicating object expressions.
+Now, we may use $\mit{mapGen}\,\gen : \Pull\,(\Lift \vA) \to \Pull\,(\Lift \vA)$
+to let-bind all yielded expressions in a stream. This can be useful in stream
+operations that would otherwise duplicate yielded expressions. For example, the
+function with type $\Pull\,(\Lift \vA) \to \Pull\,(\Lift \vA,\,\Lift \vA)$
+should be defined with let-insertion.
 
@@
   &\gen_{\Pull} : \Lift \vA \to \Pull\,(\Lift \vA) \\
-  &\gen_{\Pull}\,\va = \mit{mapGen}\,(\mit{single}\,\va)\,\gen.
+  &\gen_{\Pull}\,\va = \mit{mapGen}\,\gen\,(\mit{single}\,\va).
 \end{alignat*}
@@
 
-In the Haskell, there are some limitations. First, in $\concatMap$, the
-projection function $\mit{projS}$ cannot be defined, nbecause Haskell is not
+In Haskell there are some limitations. First, in $\concatMap$, the
+projection function $\mit{projS}$ cannot be defined, because Haskell is not
 dependently typed, and the other field projections are also out of reach. We
@@
 
-\section{TODOS}
-
-\todo{Introduce newtypes accordingly}
-\todo{allow arbitrary parameters (including ValTy -> ValTy functions) for object ADTs}
-\todo{extend as I have more stuff}
-\todo{FIGURE OUT: monadic tailrec}
-\todo{FIGURE OUT: Traversals}
-\todo{DEVELOP: mutual letrec based on computational product types}
-
-%% \section{The Object Language}\label{sec:the-object-language}
-
-%% Describe object language. Simple types + first-order fun + fixpoints + ADT-s.
-
-%% Show Agda formalization basically, types and syntax.
-
-%% Show Agda definitional interpreter. Discuss closure-free evaluation
-
-%% Examples, properties. Maybe big-step reduction? Maybe program equivalence?  Look
-%% at Harper PFPL for program equivalence reference.
-
-%% Sketch downstream compilation.
-%% \begin{itemize}
-%% \item Eta-expand function definitions to arity.
-%% \item Turn functions in immediate beta-redexes to let-bindings.
-%% \item Lambda-lift non-tail-called functions to top.
-%% \item Keep only-tail-called local functions as join points.
-%% \end{itemize}
-
-%% Downstream compilation should erase the unit type and implement products as
-%% ``flat'' unboxed structures.
-
-%% Discuss object language design choices.
-%% \begin{itemize}
-%% \item Can be more obviously first-order: e.g.\ no currying, only Val->Val
-%%   functions, or ANF functions. This requires more bureaucracy in
-%%   metaprogramming.
-%% \item Can be more flexible but less obviously first-order: allow function let
-%%   bodies, allow function case bodies. Requires more translation. More convenient
-%%   to program in. In future work.
-%% \end{itemize}
-
-%% \section{The Staged Language}\label{sec:the-staged-language}
-
-%% Don't give horizontal-bar rules.
-
-%% Meta-features:
-%% \begin{itemize}
-%% \item Pi, Sigma, Unit, Bottom, Universes (indices elided), ADT-s. Ind families probably not needed.
-%% \item Agda notation.
-%% \item Type classes informally in Haskell notation?
-%% \end{itemize}
-
-%% Explain staging ops, staging algorithm. Explain object embedding. Refer to 2LTT
-%% paper for correctness.
-
-%% Mention stage annotation inference, but we'll not use it often! To make things
-%% explicit.
-
-%% We should include a generous amount of staging examples, and some traced staging.
-
-%% Basic staging examples: identity, map, power function. Boolean short-circuiting.
-
-%% Basic staged classes, Eq, Ord, Show, Monoid, etc. Vectors with computed length.
-
-%% \section{Programming With Monads}\label{sec:programming-with-monads}
-
-%% \subsection{Basic Binding Time Improvements}\label{sec:basic-binding-time-improvements}
-
-%% Notion.
-%% For functions.
-%% For products.
-%% Note that for products there is a computation duplication problem.
-%% Mention let-insertion.
-
-%% \subsection{The Code Generation Monad}\label{sec:the-code-generation-monad}
-
-%% Is fundamental to this work.
-
-%% Definition. Explanation. Define Functor, Applicative, Monad instances.
-%% Define runGen, runGen1, runGenN etc. Define let-insertion.
-
-%% Basic examples.
-
-%% Explanation in terms of Church-coding.
-
-%% Gen Vector mapping or folding example from 2LTT repo. Without Gen, it's a
-%% big pain to get ideal linear-size map code in N. With Gen it's easy-peasy.
-
-%% \subsection{Monad Transformers}\label{sec:monad-transformers}
-
-%% Identity, ReaderT, StateT, MaybeT, ExceptT
-
-%% Strict versions of put, modify, local, runner functions.
-
-%% MTL-style overloading. MonadReader, MonadState, MonadError, MonadGen, liftGen
-
-%% Binding time improvement for monads. Up/down overloading.
-
-%% \subsection{Case Splitting in MonadGen}\label{sec:case-splitting-in-monadgen}
-
-%% It works for any object ADT.
-
-%% Introduce sums-of-values (SOV).
-
-%% Introduce syntax sugar for object case splits in MonadGen blocks.
-
-%% Basic examples.
-
-%% \subsection{Joining Control Flow}\label{sec:rejoining-control-flow}
-
-%% The problem of join points. Motivation in terms of MaybeT Gen.
-
-%% We can let-bind computations by tripping through up/down of monads,
-%% but that introduces unbox/rebox cost for Maybe \& Either.
-
-%% Works OK for Reader/State though.
-
-%% \begin{verbatim}
-%% f : Bool -> MaybeT0 Gen ^Int
-%% f = \b. ~(down do
-%%   x <- case <b> of
-%%     True  -> gen <10>
-%%     False -> gen <20>
-%%   y <- gen <~x + 10>
-%%   z <- gen <~x * 10>
-%%   pure <~x + ~y + ~z>)
-%% \end{verbatim}
-
-%% Join points using SOV-s. Description of SOV machinery. MonadJoin.
-%% instance MonadJoin Gen. MonadJoin MaybeT.
-
-%% \subsection{Monadic Tail Recursion}\label{sec:monadic-tail-recursion}
-
-%% \texttt{replicateM\_}, \texttt{find} functions working in ``any'' monad.
-%% Non-tail recursion is not possible to fuse, return addresses are dynamic.
-%% Tail recursion can return to statically known points.
-
-%% \subsection{Discussion}
-
-%% Limitations. We can't put non-value actions into structures, can't accumulate them.
-
-%% Guarantees/properties. Closure-freedom. No administrative redexes. No boxing in join points.
-%% Only boxing in monadic code comes from ``down''. No code blowup.
-
-%% \section{Fusion}\label{sec:push-pull-fusion}
-
-%% Motivation.
-
-%% \subsection{Push streams}\label{sec:push-streams}
-
-%% Definition.
-
-%% \subsection{Pull streams}\label{sec:pull-streams}
-
-%% Pull streams are definitely not monadic. \texttt{Nat -> Stream a} contains
-%% ``infinite'' amount of code internally.
-
-%% Pull streams are also not selective functors.
-
-%% We do have monadic binding for lifted types, plus ``case'' matching
-%% on arbitrary object values! This is practically almost as good as
-%% having unlimited monadic bind! We need to assume a ``non-dependency axiom'',
-%% that any $f : \Lift A \to \msf{ValTy}$ must be constant.
-
-%% \todo{Validate non-dependency in the psh model!}
-%% \todo{Add meta-identity type to the psh model}
-
-\interlinepenalty=10000
+%% \interlinepenalty=10000
 \bibliography{references}

### 2024-02-23 newpaper/paper.tex
lots of stream fusion fiddling

@@
 \newcommand{\vys}{\mathsf{ys}}
+\newcommand{\vsp}{\mathsf{sp}}
 \newcommand{\vma}{\mathsf{ma}}
@@
 \newcommand{\mapGen}{\msf{mapGen}}
+\newcommand{\genPull}{\msf{gen_{Pull}}}
+\newcommand{\casePull}{\msf{case_{Pull}}}
 
@@
 \end{alignat*}
-%% \begin{alignat*}{3}
-%%   &\mdo && \mit{sp} \fro \liftGen \dlr \mit{split}\,\vas\\
-%%   &     && \case \mit{sp}\,\of \\
-%%   &     && \ind \Nil' \to ... \\
-%%   &     && \ind \Cons'\,a\,\vas \to ...
-%% \end{alignat*}
-This can be generalized to splitting on any object value. In the supplemental
-Haskell and Agda implementations, we overload $\mit{split}$ with a type
-class. However, in a native implementation of CFTT it might make sense to extend
-$\mdo$-notation, so that we elaborate $\mbf{case}$ on object values to an
-application of the appropriate $\mit{split}$ function. We adopt this in the rest
-of the paper, so we will be able to write $\case \vx\,\of ...$ in a $\mdo$-block,
-whenever we work in a $\MonadGen$ and $\vx$ has a value type.
+This can be generalized to splitting on any object value. In the Agda and
+Haskell implementations, we overload $\mit{split}$ with a class similar to the
+following:
+\begin{alignat*}{3}
+  & \hspace{-2em}\mathrlap{\tyclass\,\mit{Split}\,(\vA : \VTy)\,\where}\\
+  & \hspace{-2em}\ind \mit{SplitTo} &&: \MTy \\
+  & \hspace{-2em}\ind \mit{split}   &&: \Lift \vA \to \Gen\,\mit{SplitTo}
+\end{alignat*}
+In a native implementation of CFTT it may make sense to extend $\mdo$-notation,
+so that we elaborate $\mbf{case}$ on object values to an application of the
+appropriate $\mit{split}$ function. We adopt this in the rest of the paper, so
+we will be able to write $\case \vx\,\of ...$ in a $\mdo$-block, whenever we
+work in a $\MonadGen$ and $\vx$ has a value type.
 
@@
 
-%% We implemented all features in this section in typed Template Haskell
-%% \cite{TODO}, using only the strongly typed interface. The organization of the
-%% code is essentially the same as here, but some changes are necessary:
-%% \begin{itemize}
-%%   \item TH cannot properly handle scoping of type variables across quotations,
-%%         generating ill-typed code in some cases. It is fairly easy to work around
-%%         this using term-level proxy values.
-%%   \item We do not bother tracking value and computation types. This could be done,
-%%         using dummy type classes, but we would not get any extra guarantees from it
-%%         in downstream compilation.
-%%   \item When we take the product of generic sums in CFTT, we simply take the
-%%         object-level product of component types, pairwise. We do this because we
-%%         assume that the downstream compiler can represent products without
-%%         overheads, in an unboxed manner, and also erase unit types. In Haskell we do
-%%         not rely on this, and instead use a slightly more complicated
-%%         sum-of-products-of-values representation, and also fuse away product types
-%%         during staging.
-%%   \item We make heavy use of \emph{singleton types} \cite{TODO} in generic sums-of-products, to emulate
-%%         dependent types. This adds significant noise compared to a native dependently typed
-%%         version.
-%%   \item We have two separate type classes for binding-time improvement, one for
-%%         base types and one for monads specifically. This is because the quantified
-%%         constraints in \ref{TODO} yield poor type inference for $\mup$ and $\mdown$,
-%%         by preventing \emph{functional dependencies} \cite{TODO} between the class
-%%         parameters. We expect that in a native CFTT implementation, the extra staging
-%%         information in type universes would help inference in this case.
-%% \end{itemize}
-
 \section{Stream Fusion}
@@
 
-%% Fusion techniques can be broadly grouped into \emph{push} fusion, which is based
-%% on Church-encodings of inductive lists, and $\emph{pull}$ fusion, which is based
-%% on Church-encodings of coinductive lists, which manifests as \emph{state
-%% machines} \cite{TODO}. The two styles have different trade-offs, and in
-%% practical programming it is a good idea to support both, but in this section we
-%% focus on pull fusion.
-
-%% The reason is partly lack of space, but also that pull fusion has been
-%% historically the more difficult to reliably compile, and we can demonstrate
-%% significant improvements in CFTT, in robustness and expressive power. Our fusion
-%% always succeeds, and we also support a $\mit{concatMap}$ operation, which
-%% has been elusive in guaranteed fusion. We believe that our implementation also
-%% provides valuable conceptual clarity.
-
 \subsection{Streams}
@@
   & \data \Pull\,(\vA : \MTy) : \MTy\,\where\\
-  & \ind \Pull : (\vS : \MTy) \to \IsSOP\,\vS \RA \vS \to (\vS \to \Gen\,(\Step\,\vS\,\vA)) \to \Pull\,\vA
+  & \ind \Pull : (\vS : \MTy) \to \IsSOP\,\vS \RA \Gen\,\vS \to (\vS \to \Gen\,(\Step\,\vS\,\vA)) \to \Pull\,\vA
 \end{alignat*}
@@
 required to be a sum-of-products of value types by the $\IsSumVS\,\vS$
-constraint. The next field with type $S$ is the initial state, while transitions
+constraint. The next field with type $\Gen\,S$ is the initial state, while transitions
 are represented by the $\vS \to \Gen\,(\Step\,\vS\,\vA)$ field. Possible transitions
@@
   &\mit{pure} : \vA \to \Pull\,\vA\\
-  &\mit{pure} = \Pull\,()\,()\,(\lam \_.\,\return \dlr \Stop)
+  &\mit{pure}\,\va = \Pull\,()\,(\return\,())\,(\lam \_.\,\return \dlr \Yield\,\va\,())
 \end{alignat*}
@@
   &(\!\ap\!)\,(\Pull\,\vS\,\seed\,\step)\,(\Pull\,\vS'\,\seed'\,\step') =\\
-  & \ind \Pull\,(\vS,\,\vS')\,(\seed,\,\seed') \dlr \lam (\vs,\,\vs').\,\step\,\vs \bind \lam \case \\
+  & \ind \Pull\,(\vS,\,\vS')\,(({,}) \fmap \seed\,\ap\,\seed') \dlr \lam (\vs,\,\vs').\,\step\,\vs \bind \lam \case \\
   & \ind \ind \Stop \hspace{1.20em}        \to \return \Stop \\
@@
   & \mit{empty} : \Pull\,\vA\\
-  & \mit{empty} = \Pull\,()\,()\,(\lam \_.\,\return \Stop)
+  & \mit{empty} = \Pull\,()\,(\return\,())\,(\lam \_.\,\return \Stop)
 \end{alignat*}
 \begin{alignat*}{5}
-  & \hspace{-4em}\mathrlap{(\!<>\!) : \Pull\,\vA \to \Pull\,\vA \to \Pull\,\vA} \\
-  & \hspace{-4em}\mathrlap{(\!<>\!)\,(\Pull\,\vS\,\seed\,\step)\,(\Pull\,\vS'\,\seed'\,\step') = \Pull\,(\Either\,\vS\,\vS')\,(\Left\,\seed)\,\lam \case} \\
-  & \hspace{-4em}\ind \Left\,\vs &&\to \step\,s \bind \lam \case &&\Stop             &&\to \return \dlr \Skip\,(\Right\,\seed')\\
-  & \hspace{-4em}                    &&                          &&\Skip\,\vs        &&\to \return \dlr \Skip\,(\Left\,\vs) \\
-  & \hspace{-4em}                    &&                        &&\Yield\,\va\,\vs    &&\to \return \dlr \Yield\,\va\,(\Left\,\vs)\\
-  & \hspace{-4em}\ind \Right\,\vs' &&\to \step\,s' \bind \lam \case &&\Stop          &&\to \return \Stop\\
-  & \hspace{-4em}                    &&                        &&\Skip\,\vs'         &&\to \return \dlr \Skip\,(\Right\,\vs') \\
-  & \hspace{-4em}                    &&                        &&\Yield\,\va\,\vs'   &&\to \return \dlr \Yield\,\va\,(\Right\,\vs')
+  & \hspace{-5em}\mathrlap{(\!<>\!) : \Pull\,\vA \to \Pull\,\vA \to \Pull\,\vA} \\
+  & \hspace{-5em}\mathrlap{(\!<>\!)\,(\Pull\,\vS\,\seed\,\step)\,(\Pull\,\vS'\,\seed'\,\step') = \Pull\,(\Either\,\vS\,\vS')\,(\Left \fmap \seed)\,\lam \case} \\
+  & \hspace{-5em}\ind \Left\,\vs &&\to \step\,s \bind \lam \case &&\Stop             &&\to (\Skip \circ \Right) \fmap \seed'\\
+  & \hspace{-5em}                    &&                          &&\Skip\,\vs        &&\to \return \dlr \Skip\,(\Left\,\vs) \\
+  & \hspace{-5em}                    &&                        &&\Yield\,\va\,\vs    &&\to \return \dlr \Yield\,\va\,(\Left\,\vs)\\
+  & \hspace{-5em}\ind \Right\,\vs' &&\to \step\,s' \bind \lam \case &&\Stop          &&\to \return \Stop\\
+  & \hspace{-5em}                    &&                        &&\Skip\,\vs'         &&\to \return \dlr \Skip\,(\Right\,\vs') \\
+  & \hspace{-5em}                    &&                        &&\Yield\,\va\,\vs'   &&\to \return \dlr \Yield\,\va\,(\Right\,\vs')
 \end{alignat*}
@@
 unstaged definitions in previous literature, the only additional noise is just
-the $\Gen$ monad in the transitions. Likewise, we can give standard definitions
-for usual stream functions such as $\msf{filter}$, $\msf{take}$ or
-$\msf{drop}$.
+the $\Gen$ monad in the initial states and the transitions. Likewise, we can
+give standard definitions for usual stream functions such as $\msf{filter}$,
+$\msf{take}$ or $\msf{drop}$.
 
@@
 \end{alignat*}
-For join points, this is just as efficient as the previous
-$\mit{Fun}_{\SOP\Lift}$, since a definition of a product of functions gets
-compiled to a sequence of function definitions. Now, we can generate object code
-from streams. There are several choices for this, but in CFTT the $\msf{foldr}$
-function is as good as we can get.
+Even for join points, this is just as efficient as the previous
+$\mit{Fun}_{\SOP\Lift}$ version, since a definition of a product of functions
+gets compiled to a sequence of function definitions. Now, we can generate object
+code from streams. There are several choices for this, but in CFTT the
+$\msf{foldr}$ function is as good as we can get.
 \begin{alignat*}{3}
@@
   &\hspace{-13em}\ind \ind \Yield\,\va\,\vs &&\to \vf\,\va\,(\mindex\,\qt{\vfs}\,(\rep\,\vs))\big);\\
-  &\hspace{-13em}\mathrlap{\ind \spl(\mindex\,\qt{\vfs}\,(\rep\,\seed))\qr}
+  &\hspace{-13em}\mathrlap{\ind \spl\big(\runGen \dlr \mdo \{s \fro \seed; \return \dlr \mindex\,\qt{\vfs}\,(\rep\,\vs)\}\big)\qr}
 \end{alignat*}
@@
 \begin{alignat*}{3}
-  & \concatMap : \IsSOP\,\vA \RA (\vA \to \Pull\,\vB) \to \Pull\,\vA \to \Pull\,\vB \\
-  & \concatMap\,\{\vA\}\,\{\vB\}\,\vf\,(\Pull\,\vS\,\seed\,\step) = \\
-  & \ind \Pull\,(\vS,\,\Maybe\,(\Sigma\,\vA\,(\mit{projS} \circ \vf)))\,(\seed,\,\Nothing) \dlr \lam \case \\
-  & \ind \ind (\vs,\,\Nothing) \to \step\,s \bind \lam \case \\
-  & \ind \ind \ind \Stop      \hspace{1.4em}   \to \return \Stop\\
-  & \ind \ind \ind \Skip\,\vs \hspace{0.9em} \to \return \dlr \Skip\,(\vs,\, \Nothing)\\
-  & \ind \ind \ind \Yield\,\va\,\vs          \to \return \dlr \Skip\,(\vs,\, \Just\,(\va,\,\mit{projSeed}\,(\vf\,\va)))\\
-  & \ind \ind (\vs,\,\Just\,(\va,\,\vs')) \to \mit{projStep}\,(\vf\,\va)\,\vs' \bind \lam \case \\
-  & \ind \ind \ind \Stop       \hspace{1.75em}\to \return \dlr \Skip\,(\vs,\,\Nothing) \\
-  & \ind \ind \ind \Skip\,\vs' \hspace{0.94em}\to \return \dlr \Skip\,(\vs,\,\Just\,(\va,\,\vs'))\\
-  & \ind \ind \ind \Yield\,\vb\,\vs' \to \return \dlr \Yield\,\vb\,(\vs,\, \Just\,(\va,\,\vs'))\\
-\end{alignat*}
-Here, $\Nothing$ marks the states where we are in the ``outer'' loop, trying to
-run the $\Pull\,\vA$ stream until we get the next value. $\Just$ marks the
-states of the ``inner'' loop, where we have a concrete $\va : \vA$ value and
-we run the $(\vf\,\va)$ stream until it stops. In the inner loop, the machine state
-type depends on the $\va : \vA$ value, hence the need for $\Sigma$.
+  & \mathrlap{\concatMap : \IsSOP\,\vA \RA (\vA \to \Pull\,\vB) \to \Pull\,\vA \to \Pull\,\vB }\\
+  & \mathrlap{\concatMap\,\{\vA\}\,\{\vB\}\,\vf\,(\Pull\,\vS\,\seed\,\step) = }\\
+  & \mathrlap{\ind \Pull\,(\vS,\,\Maybe\,(\Sigma\,\vA\,(\mit{projS} \circ \vf)))\,(({,}) \fmap \seed\,\ap\,\return\,\Nothing) \dlr \lam \case }\\
+  & \mathrlap{\ind \ind (\vs,\,\Nothing) \to \step\,\vs \bind \lam \case }\\
+  & \ind \ind \ind \ind \Stop                 &&\to \return \Stop\\
+  & \ind \ind \ind \ind \Skip\,\vs            &&\to \return \dlr \Skip\,(\vs,\, \Nothing)\\
+  & \ind \ind \ind \ind \Yield\,\va\,\vs      &&\to \mdo \{\vs' \fro \mit{projSeed}\,(\vf\,\va); \return \dlr \Skip\,(\vs,\,\Just\,(\va,\,\vs'))\}\\
+  & \mathrlap{\ind \ind (\vs,\,\Just\,(\va,\,\vs')) \to \mit{projStep}\,(\vf\,\va)\,\vs' \bind \lam \case }\\
+  & \ind \ind \ind \ind \Stop                &&\to \return \dlr \Skip\,(\vs,\,\Nothing) \\
+  & \ind \ind \ind \ind \Skip\,\vs'          &&\to \return \dlr \Skip\,(\vs,\,\Just\,(\va,\,\vs'))\\
+  & \ind \ind \ind \ind \Yield\,\vb\,\vs'    &&\to \return \dlr \Yield\,\vb\,(\vs,\, \Just\,(\va,\,\vs'))
+\end{alignat*}
+Here, $\Nothing$ marks the states where we are in the ``outer'' loop, running
+the $\Pull\,\vA$ stream until we get its next value. $\Just$ marks the states of
+the ``inner'' loop, where we have a concrete $\va : \vA$ value and we run the
+$(\vf\,\va)$ stream until it stops. In the inner loop, the machine state type
+depends on the $\va : \vA$ value, hence the need for $\Sigma$.
 
@@
 
-\subsection{Let-Insertion and Case Splitting}
-
-Since $\Pull$ is not a monad, it cannot be $\MonadGen$ in our framework,
-but we still support some forms of let-insertion. First, we note that $\Pull$ can be
-mapped over by functions returning in $\Gen$:
-\begin{alignat*}{3}
-  &\hspace{-3em}\mathrlap{\mit{mapGen} : (\vA \to \Gen\,\vB) \to \Pull\,\vA \to \Pull\,\vB }\\
-  &\hspace{-3em}\mathrlap{\mit{mapGen}\,\vf\,(\Pull\,\vS\,\seed\,\step) = \Pull\,\vS\,\seed \dlr \lam \case}\\
-  &\hspace{-3em} \ind \Stop            &&\to \return\,\Stop \\
-  &\hspace{-3em} \ind \Skip\,\vs       &&\to \return \dlr \Skip\,\vs\\
-  &\hspace{-3em} \ind \Yield\,\va\,\vs &&\to \mdo \{\vb \fro \vf\,\va; \return \dlr \Yield\,\vb\,\vs\}
-\end{alignat*}
-Now, we may use $\mit{mapGen}\,\gen : \Pull\,(\Lift \vA) \to \Pull\,(\Lift \vA)$
-to let-bind all yielded expressions in a stream. This can be useful in stream
-operations that would otherwise duplicate yielded expressions. For example, the
-function with type $\Pull\,(\Lift \vA) \to \Pull\,(\Lift \vA,\,\Lift \vA)$
-should be defined with let-insertion.
-
-Let us now consider a more ``monadic'' style of stream programming. For this we
-need singleton streams:
+\subsection{Let-Insertion \& Case Splitting in Monadic Style}
+
+While $\Pull$ is not a monad, and hence also not a $\MonadGen$, we can still use
+a monadic style of stream programming with good ergonomics. First, we need
+singleton streams for ``returning'':
 \begin{alignat*}{3}
@@
 \end{alignat*}
-We also write $\mit{forEach}$ for $\concatMap$ with flipped arguments. Now, a
-more monadic flavor of let-insertion is given as
-\begin{alignat*}{3}
-  &\gen_{\Pull} : \Lift \vA \to \Pull\,(\Lift \vA) \\
-  &\gen_{\Pull}\,\va = \mit{mapGen}\,\gen\,(\mit{single}\,\va).
-\end{alignat*}
-We can also implement case splitting on object values. Recall from Section
-\ref{TODO} that for each value type we have a splitting function that returns a
-meta-level view on the constructors. Importantly, all such views are $\SOP$-s, and
-therefore can be bound by $\forEach$. A small example:
-\begin{alignat*}{3}
-  & \hspace{-5em}\mathrlap{\forEach\,(\mit{countFrom}\,\qt{0}) \dlr \lam \vn. }\\
-  & \hspace{-5em}\mathrlap{\forEach\,(\gen_{\Pull}\,\qt{\spl \vn * 2}) \dlr \lam \vm. }\\
-  & \hspace{-5em}\mathrlap{\forEach\,(\single\,(\msplit\,\qt{\spl \vm < 20})) \dlr \lam \case }\\
-  & \hspace{-5em}\ind \True_\M  &&\to \single\,\qt{\spl \vm + 10}   \\
-  & \hspace{-5em}\ind \False_\M &&\to \single\,\qt{\spl \vm + 20}
-\end{alignat*}
-Here $(\gen_{\Pull}\,\qt{\spl \vn * 2})$ behaves like a let-insertion, and
-$(\single\,(\msplit\,\qt{\spl \vm < 20}))$ implements the case splitting.
+Now, this operation should be avoided when possible, since it has two states and
+can contribute to a blow-up of state size in combination with state-multiplying
+operations such as $\mit{zip}$ or $\concatMap$. Both let-insertion and case
+splitting could be defined generically from $\single$, $\concatMap$ and the
+following operation (definition omitted):
+\[ \mit{mapGen} : (\vA \to \Gen\,\vB) \to \Pull\,\vA \to \Pull\,\vB \]
+However, $\single$ would introduce superfluous states that way. Instead, we give
+more efficient specialized definitions. Let-insertion is as follows:
+\begin{alignat*}{3}
+  & \hspace{-8em}\mathrlap{\genPull : \{\vA : \VTy\}\{\vB : \MTy\} \to \Lift \vA \to (\Lift \vA \to \Pull\,\vB) \to \Pull\,\vB}\\
+  & \hspace{-8em}\mathrlap{\genPull\,\va\,\vf = \Pull\,(\Sigma\,(\Lift\,\vA)\,(\mit{projS} \circ \vf))\,(({\va,}) \fmap \mit{projSeed}\,(\vf\,\va)) \dlr \lam \case}\\
+  & \hspace{-8em}\ind \Stop            &&\to \return\,\Stop \\
+  & \hspace{-8em}\ind \Skip\,\vs       &&\to \return \dlr \Skip\,(\va,\,\vs)\\
+  & \hspace{-8em}\ind \Yield\,\vb\,\vs &&\to \return \dlr \Yield\,\vb\,(\va,\,\vs)
+\end{alignat*}
+For case splitting, we have:
+\begin{alignat*}{3}
+  &\hspace{-12em}\mathrlap{\casePull : (\mit{Split}\,\vA,\,\IsSOP\,(\mit{SplitTo}\,\{\vA\})) \RA \Lift \vA \to (\mit{SplitTo}\,\{\vA\} \to \Pull\,\vB) \to \Pull\,\vB }\\
+  &\hspace{-12em}\mathrlap{\casePull\,\va\,\vf = \Pull\,(\Sigma\,(\mit{SplitTo}\,\{\vA\})\,(\mit{projS} \circ \vf))}\\
+  &\hspace{-12em}\mathrlap{    \hspace{7.1em}         (\mdo \{\va' \fro \mit{split}\,\va; \vs \fro \mit{projSeed}\,(\vf\,\va');\return (\va',\,\vs)\}) \dlr \lam \case}\\
+  &\hspace{-4em}\mathrlap{  \ind   (\va',\,\vs) \to \mit{projStep}\,(\vf\,\va')\,\vs \bind \lam \case} \\
+  &\hspace{-4em} \ind \ind \ind \Stop            &&\to \return \Stop \\
+  &\hspace{-4em} \ind \ind \ind \Skip\,\vs       &&\to \return \dlr \Skip\,(\va',\,\vs) \\
+  &\hspace{-4em} \ind \ind \ind \Yield\,\vb\,\vs &&\to \return \dlr \Yield\,\vb\,(\va',\,\vs)
+\end{alignat*}
+
+%% casePull' : ∀ {A B}⦃ _ : Split A ⦄ ⦃ _ : IsSOP (SplitTo {A}) ⦄ → ↑V A → (SplitTo {A} → Pull B) → Pull B
+%% casePull' {A} {B} {{_}}{{sopA}} a f =
+%%   pull (Σ (SplitTo {A}) (St ∘ f)) {{SOPΣ{{sopA}}{{λ {x} → StSOP (f x)}}}}
+%%     (do sp ← splitGen a; s ← seed (f sp); pure (_,_ {A = SplitTo {A}}{B = _} sp s)) λ where
+%%       (sp , s) → step (f sp) s >>= λ where
+%%          stop        → pure stop
+%%          (skip s)    → pure $ skip (sp , s)
+%%          (yield b s) → pure $ yield b (sp , s)
+
+%% To get good ergonomics in a ``monadic'' style of stream definitions, we need to
+%% reproduce let-insertion and case splitting on object values. First, we need
+%% singleton streams:
+
+%% genLetPull' : ∀ {A B} → ↑V A → (↑V A → Pull B) → Pull B
+%% genLetPull' {A} {B} a f =
+%%   pull (Σ (↑V A) (St ∘ f)) {{SOPΣ {{↑SOP}} {{λ {x} → StSOP (f x)}}}} ((a ,_) <$> (seed (f a))) λ where
+%%     (a , s) → step (f a) s <&> λ where
+%%       stop        → stop
+%%       (skip s)    → skip (a , s)
+%%       (yield b s) → yield b (a , s)
+
+%% Since $\Pull$ is not a monad, it cannot be $\MonadGen$ in our framework,
+%% but we still support some forms of let-insertion. First, we note that $\Pull$ can be
+%% mapped over by functions returning in $\Gen$:
+%% \begin{alignat*}{3}
+%%   &\hspace{-3em}\mathrlap{\mit{mapGen} : (\vA \to \Gen\,\vB) \to \Pull\,\vA \to \Pull\,\vB }\\
+%%   &\hspace{-3em}\mathrlap{\mit{mapGen}\,\vf\,(\Pull\,\vS\,\seed\,\step) = \Pull\,\vS\,\seed \dlr \lam \case}\\
+%%   &\hspace{-3em} \ind \Stop            &&\to \return\,\Stop \\
+%%   &\hspace{-3em} \ind \Skip\,\vs       &&\to \return \dlr \Skip\,\vs\\
+%%   &\hspace{-3em} \ind \Yield\,\va\,\vs &&\to \mdo \{\vb \fro \vf\,\va; \return \dlr \Yield\,\vb\,\vs\}
+%% \end{alignat*}
+%% Now, we may use $\mit{mapGen}\,\gen : \Pull\,(\Lift \vA) \to \Pull\,(\Lift \vA)$
+%% to let-bind all yielded expressions in a stream. This can be useful in stream
+%% operations that would otherwise duplicate yielded expressions. For example, the
+%% function with type $\Pull\,(\Lift \vA) \to \Pull\,(\Lift \vA,\,\Lift \vA)$
+%% should be defined with let-insertion.
+
+%% Let us now consider a more ``monadic'' style of stream programming. For this we
+%% need singleton streams:
+%% \begin{alignat*}{3}
+%%   & \mit{single} : \vA \to \Pull\,\vA \\
+%%   & \mit{single}\,\va = \Pull\,\Bool_\M\,\True \dlr \lam \vb.\,\return \dlr\mbf{if}\,\vb\,\mbf{then}\,\Yield\,\va\,\False\,\mbf{else}\,\Stop
+%% \end{alignat*}
+%% We also write $\mit{forEach}$ for $\concatMap$ with flipped arguments. Now, a
+%% more monadic flavor of let-insertion is given as
+%% \begin{alignat*}{3}
+%%   &\gen_{\Pull} : \Lift \vA \to \Pull\,(\Lift \vA) \\
+%%   &\gen_{\Pull}\,\va = \mit{mapGen}\,\gen\,(\mit{single}\,\va).
+%% \end{alignat*}
+%% We can also implement case splitting on object values. Recall from Section
+%% \ref{TODO} that for each value type we have a splitting function that returns a
+%% meta-level view on the constructors. Importantly, all such views are $\SOP$-s, and
+%% therefore can be bound by $\forEach$. A small example:
+%% \begin{alignat*}{3}
+%%   & \hspace{-5em}\mathrlap{\forEach\,(\mit{countFrom}\,\qt{0}) \dlr \lam \vn. }\\
+%%   & \hspace{-5em}\mathrlap{\forEach\,(\gen_{\Pull}\,\qt{\spl \vn * 2}) \dlr \lam \vm. }\\
+%%   & \hspace{-5em}\mathrlap{\forEach\,(\single\,(\msplit\,\qt{\spl \vm < 20})) \dlr \lam \case }\\
+%%   & \hspace{-5em}\ind \True_\M  &&\to \single\,\qt{\spl \vm + 10}   \\
+%%   & \hspace{-5em}\ind \False_\M &&\to \single\,\qt{\spl \vm + 20}
+%% \end{alignat*}
+%% Here $(\gen_{\Pull}\,\qt{\spl \vn * 2})$ behaves like a let-insertion, and
+%% $(\single\,(\msplit\,\qt{\spl \vm < 20}))$ implements the case splitting.
 

### 2024-02-23 newpaper/paper.tex
wip

@@
 \newcommand{\genPull}{\msf{gen_{Pull}}}
+\newcommand{\Set}{\msf{Set}}
 \newcommand{\casePull}{\msf{case_{Pull}}}
+\newcommand{\appull}{\ap_{\Pull}}
 
@@
 
+\subsection{Semantics of Staging}
+
+We give a short summary of the semantics of staging here. We reuse most of the
+semantics from \cite{TODO}, also borrowing notation and definitions, and only
+mention additions and differences. The reader may refer to ibid.\ for technical
+details.
+
+Staging is given by evaluation of CFTT into a presheaf model, where the base
+category is given as follows: objects are typing contexts of the object language
+and morphisms are parallel substitutions. The equality of morphisms is given by
+strict syntactic equality of object terms. $\Ty$ is interpreted as the presheaf
+which is constantly the set of object-theoretic types. Soundness of staging is
+proven by a separate logical relations model, internally to the mentioned
+presheaf model. Soundness means that the output of staging, viewed as a CFTT
+term, is definitionally equal to the input.
+
+The main addition in CFTT is the assumption of all inductive families in the
+meta-level. However, based on the results in \cite{TODO}, it is enough to
+extend the meta-level syntax in \cite{TODO} with W-types and identity types to
+derive all inductive families, so we consider these in the following.
+
+\subsubsection{W-types} W-types have a standard object-wise inductive definition in
+the presheaf model \cite{TODO}. Additionally, we need to give a logical relation
+interpretation to W-types, as required for the soundness proof for staging. In
+Section 5.2.\ of \cite{TODO} a serialization map is used to define the relation
+for natural numbers, however, there is no analogous map for W-types because we
+cannot build finite terms from infinitely branching trees.
+
+Instead, we give an inductive definition for the logical relation at
+W-types. First, assuming $A : \wh{\Set}$ and $B : A \to \wh{\Set}$, we have the
+usual constructor for the semantic W-type in the internal language of the
+presheaf model, as $\msf{sup} : (a : A) \to (B\,a \to \msf{W}\,A\,B) \to
+\msf{W}\,A\,B$.  Second, assuming $A : \Ty\,\Gamma$ and $B : \Ty\,(\Gamma \ext
+A)$ in CFTT, we internalize the $\msf{sup}$ constructor which constructs a term
+of a syntactic W-type from restricted terms:
+\[ \mathbb{R}_{\msf{sup}} : (\alpha : \mathbb{R}\,A\,\gamma) \to \mathbb{R}\,(B \RA (\msf{W_{CFTT}}\,A\,B)[\msf{wk}])\,(\gamma,\,\alpha) \to \mathbb{R}\,(\msf{W_{CFTT}}\,A\,B)\,\gamma \]
+Here, $\msf{W_{CFTT}}$ denotes the W-type former in CFTT, $B \RA
+(\msf{W_{CFTT}}\,A\,B)[\msf{wk}]$ is a non-dependent function type in CFTT, and
+$\msf{wk}$ is a weakening substitution.
+
+Then, assuming $\gamma^{\approx} : \Gamma^{\approx}\,\gamma\,\gamma'$, we
+inductively define a relation $\msf{W}^{\approx}$ between
+$\msf{W}\,(\mbb{E}\,A\,\gamma)\,(\lam \alpha.\,\mbb{E}\,B\,(\gamma,\,\alpha))$
+and $\mbb{R}\,(\msf{W_{CFTT}}\,A\,B)\,\gamma'$, which relates
+$\msf{sup}\,\alpha\,f$ to $\mathbb{R}_{\msf{sup}}\,\alpha'\,f'$ if $\alpha$ is
+related to $\alpha'$ at type $A$ and $f$ is related to $f'$ at type $B[\alpha']
+\RA \msf{W_{CFTT}}\,A\,B$. The relation at the non-dependent function type $\blank{\RA}\blank$
+is given as pointwise preservation of relations. With this, we can also give
+relational interpretation to the syntactic $\msf{sup}$ rule in CFTT,
+and we can use induction on $\msf{W}^{\approx}$ to interpret the syntactic
+induction rule for W. Thus, we get soundness of staging with W-types.
+
+\subsubsection{Identity type} Our treatment of the identity type is
+influenced by the following: we would like to assume identities as axioms in
+CFTT, which are true in the presheaf model (i.e.\ they are true at
+staging-time), without blocking staging as a computation. In Section \ref{TODO}
+we use such an axiom to good effect. However, axioms do block computation in
+CFTT up to definitional equality, which implies that the soundness of staging
+fails. Our solution:
+\begin{itemize}
+\item Have a standard intensional identity type in CFTT, perhaps with uniqueness of identity proofs assumed.
+\item Before staging, translate CFTT to CFTT plus equality reflection. Operationally,
+      this means erasing all identity proofs and transports.
+\item Perform staging after erasure of identity proofs.
+\end{itemize}
+In the presheaf model, we interpret the extensional identity in a standard way,
+as strict equality of sections of presheaves. With this, soundness of staging is
+trivial to show; we can pick the trivial relation for the interpretation of
+identity. Hence, we weaken the statement of soundness: every CFTT program is
+convertible to the staging output, up to erasure of identity proofs.
+
 \section{Monads \& Monad Transformers}\label{sec:monad-transformers}
@@
 required to be a sum-of-products of value types by the $\IsSumVS\,\vS$
-constraint. The next field with type $\Gen\,S$ is the initial state, while transitions
+constraint. Borrowing terminology from Coutts \cite{TODO}, we call each product-of-value-types in the
+state a \emph{state shape}.
+
+The next field with type $\Gen\,\vS$ is the initial state, while transitions
 are represented by the $\vS \to \Gen\,(\Step\,\vS\,\vA)$ field. Possible transitions
@@
 
-Let us see some operations on streams now. First, $\Pull$ is a ``zippy''
-applicative functor, where $\mit{pure}$ infinitely repeats a value and $({\ap})$
-applies a stream of functions to a stream of values, pointwise.
-\begin{alignat*}{3}
-  &\mit{pure} : \vA \to \Pull\,\vA\\
-  &\mit{pure}\,\va = \Pull\,()\,(\return\,())\,(\lam \_.\,\return \dlr \Yield\,\va\,())
-\end{alignat*}
-\begin{alignat*}{3}
-  &(\!\ap\!) : \Pull\,(\vA \to \vB) \to \Pull\,\vA \to \Pull\,\vB \\
-  &(\!\ap\!)\,(\Pull\,\vS\,\seed\,\step)\,(\Pull\,\vS'\,\seed'\,\step') =\\
-  & \ind \Pull\,(\vS,\,\vS')\,(({,}) \fmap \seed\,\ap\,\seed') \dlr \lam (\vs,\,\vs').\,\step\,\vs \bind \lam \case \\
-  & \ind \ind \Stop \hspace{1.20em}        \to \return \Stop \\
-  & \ind \ind \Skip\,\vs \hspace{0.75em}     \to \return \dlr \Skip\,(\vs,\,\vs') \\
-  & \ind \ind \Yield\,\vf\,\vs \to \step'\,\vs' \bind \lam \case \\
-  & \ind \ind \ind \Stop \hspace{1.7em} \to \return \Stop \\
-  & \ind \ind \ind \Skip\,\vs'\hspace{0.9em}\to \return \dlr \Skip\,(\vs,\,\vs') \\
-  & \ind \ind \ind \Yield\,\va\,\vs' \to \return \dlr \Yield\,(\vf\,\va)\,(\vs,\,\vs')
-\end{alignat*}
-In $\mit{pure}$, the state is the unit type, while in $(\!\ap\!)$ we take the
-product of the stream states and do synchronous transitions. In both
-definitions, the $\IsSOP\,\vS$ constraint is implicitly dispatched by instance
-resolution; this works in the Agda and Haskell versions too. We can derive the
-usual $\mit{zip}$ and $\mit{zipWith}$ operations from $(\!\ap\!)$. $\Pull$
-is also a monoid with stream appending as the binary operation.
+Let us see some operations on streams now. $\Pull$ is evidently a $\Functor$. It
+is not quite a ``zippy'' applicative functor, because its application operator
+requires an extra $\IsSOP$ constraint:
+\begin{alignat*}{3}
+  &\mit{repeat} : \vA \to \Pull\,\vA\\
+  &\mit{repeat}\,\va = \Pull\,()\,(\return\,())\,(\lam \_.\,\return \dlr \Yield\,\va\,())
+\end{alignat*}
+\begin{alignat*}{3}
+  &\hspace{-7em}\mathrlap{({\ap_{\Pull}}) : \IsSOP\,\vA \RA \Pull\,(\vA \to \vB) \to \Pull\,\vA \to \Pull\,\vB }\\
+  &\hspace{-7em}\mathrlap{({\ap_{\Pull}})\,(\Pull\,\vS\,\seed\,\step)\,(\Pull\,\vS'\,\seed'\,\step') =}\\
+  &\hspace{-7em}\mathrlap{ \ind \Pull\,(\vS,\,\vS',\,\Maybe\,\vA)\,(({,}) \fmap \seed \ap ((,\Nothing) \fmap \seed') \dlr \lam \case }\\
+  &\hspace{-7em}\mathrlap{ \ind \ind (\vs,\,\vs',\,\Just\,\va) \to \step\,\vs \bind \lam \case}\\
+  &\hspace{-7em} \ind \ind \ind \Stop        &&\to \return \Stop \\
+  &\hspace{-7em} \ind \ind \ind \Skip\,\vs     &&\to \return \dlr \Skip\,(\vs,\,\vs',\,\Just\,\va) \\
+  &\hspace{-7em} \ind \ind \ind \Yield\,\vf\,\vs &&\to \return \dlr \Yield\,(\vf\,\va)\,(\vs,\,\vs',\,\Nothing) \\
+  &\hspace{-7em} \mathrlap{\ind \ind (\vs,\,\vs',\,\Nothing) \to \step'\,\vs' \bind \lam \case }\\
+  &\hspace{-7em} \ind \ind \ind \Stop         &&\to \return \Stop \\
+  &\hspace{-7em} \ind \ind \ind \Skip\,\vs'     &&\to \return \dlr \Skip\,(\vs,\,\vs',\,\Nothing) \\
+  &\hspace{-7em} \ind \ind \ind \Yield\,\va\,\vs' &&\to \return \dlr \Skip\,(\vs,\,\vs',\,\Just\,\va)
+\end{alignat*}
+
+In $\mit{repeat}$, the state is the unit type, while in $({\appull})$, we take
+the product of states, and also add another $\Maybe\,\vA$ that is used to buffer
+a single $\vA$ value while we are stepping the $\Pull\,(\vA \to \vB)$ stream.
+The $\IsSOP$ constraints for the new machine states are implicitly dispatched
+by instance resolution; this works in the Agda and Haskell versions too. We can
+derive $\mit{zip}$ and $\mit{zipWith}$ from $({\appull})$ and $({\fmap})$ for
+streams, with the restriction that the zipped streams must all produce $\IsSOP$
+values.
+
+$\Pull$ is also a monoid with stream appending as the binary operation.
 \begin{alignat*}{3}
@@
 \begin{alignat*}{5}
-  & \hspace{-5em}\mathrlap{(\!<>\!) : \Pull\,\vA \to \Pull\,\vA \to \Pull\,\vA} \\
-  & \hspace{-5em}\mathrlap{(\!<>\!)\,(\Pull\,\vS\,\seed\,\step)\,(\Pull\,\vS'\,\seed'\,\step') = \Pull\,(\Either\,\vS\,\vS')\,(\Left \fmap \seed)\,\lam \case} \\
+  & \hspace{-5em}\mathrlap{({<>}) : \Pull\,\vA \to \Pull\,\vA \to \Pull\,\vA} \\
+  & \hspace{-5em}\mathrlap{({<>})\,(\Pull\,\vS\,\seed\,\step)\,(\Pull\,\vS'\,\seed'\,\step') = \Pull\,(\Either\,\vS\,\vS')\,(\Left \fmap \seed)\,\lam \case} \\
   & \hspace{-5em}\ind \Left\,\vs &&\to \step\,s \bind \lam \case &&\Stop             &&\to (\Skip \circ \Right) \fmap \seed'\\
@@
 \end{alignat*}
-These definitions are standard for streams; note though that compared to the
-unstaged definitions in previous literature, the only additional noise is just
-the $\Gen$ monad in the initial states and the transitions. Likewise, we can
-give standard definitions for usual stream functions such as $\msf{filter}$,
-$\msf{take}$ or $\msf{drop}$.
+These definitions are standard for streams; compared to the unstaged definitions
+in previous literature, the only additional noise is just the $\Gen$ monad in
+the initial states and the transitions. Likewise, we can give standard
+definitions for usual stream functions such as $\msf{filter}$, $\msf{take}$ or
+$\msf{drop}$.
 
@@
 $\mit{Fun}_{\SOP\Lift}$ version, since a definition of a product of functions
-gets compiled to a sequence of function definitions. Now, we can generate object
-code from streams. There are several choices for this, but in CFTT the
-$\msf{foldr}$ function is as good as we can get.
+will get compiled to a sequence of function definitions. In addition, we can use
+it to generate object code from streams. There are several choices for this, but
+in CFTT the $\msf{foldr}$ function is as good as we can get.
 \begin{alignat*}{3}
@@
   &\hspace{-13em}\mathrlap{\ind \letrec \vfs : \mit{Fun}_{\SOP\Lift}\,(\Rep\,\{\vS\})\,\vB := \spl\big(\mtabulate \dlr \lam \vs.\,\unGen\,(\step\,(\rep^{-1}\,\vs)) \dlr \lam \case} \\
-  &\hspace{-13em}\ind \ind \Stop            &&\to \vb \\
-  &\hspace{-13em}\ind \ind \Skip\,\vs       &&\to \mindex\,\qt{\vfs}\,(\rep\,\vs) \\
-  &\hspace{-13em}\ind \ind \Yield\,\va\,\vs &&\to \vf\,\va\,(\mindex\,\qt{\vfs}\,(\rep\,\vs))\big);\\
-  &\hspace{-13em}\mathrlap{\ind \spl\big(\runGen \dlr \mdo \{s \fro \seed; \return \dlr \mindex\,\qt{\vfs}\,(\rep\,\vs)\}\big)\qr}
-\end{alignat*}
-This $\foldr$ is quite flexible, becuse we can eliminate into any object type,
-including function types. For instance, we can define $\foldl$ from $\foldr$:
+  &\hspace{-13em}\ind \ind \ind \Stop            &&\to \vb \\
+  &\hspace{-13em}\ind \ind \ind \Skip\,\vs       &&\to \mindex\,\qt{\vfs}\,(\rep\,\vs) \\
+  &\hspace{-13em}\ind \ind \ind \Yield\,\va\,\vs &&\to \vf\,\va\,(\mindex\,\qt{\vfs}\,(\rep\,\vs))\big);\\
+  &\hspace{-13em}\mathrlap{\ind \spl\big(\unGen\,\seed \dlr \lam s.\,\mindex\,\qt{\vfs}\,(\rep\,\vs)\big)\qr}
+\end{alignat*}
+This $\foldr$ is quite flexible becuse we can compute any object value from it,
+including functions. For instance, we can define $\foldl$ from $\foldr$:
 \begin{alignat*}{3}
@@
 
-We saw that $\Pull$ is $\Applicative$, but what about having a list-like
-$\Monad$ instance as well, with singleton streams for $\return$ and $\concatMap$
-for binding? This is not possible. Aiming at
+Can we have a list-like $\Monad$ instance for streams, with singleton streams
+for $\return$ and $\concatMap$ for binding? This is not possible. Aiming at
 \[ \concatMap : (\vA \to \Pull\,\vB) \to \Pull\,\vA \to \Pull\,\vB, \]
@@
 machine state types, which cannot be represented in a finite amount of object
-code.\footnote{Although in two-level type theories that are not used for
-staging, the representability of countably infinite products of object types is
-sometimes assumed as an axiom; this axiom is called ``cofibrancy of
-$\mathbb{N}$'' \cite{TODO}.} Here by ``infinite'' we mean the notion that is
-internally available in the meta type theory. For instance, we can define a
-$\Nat_\M \to \Pull\,\vB$ function which for each $\vn : \Nat_\M$ produces a
-concatenation of $\vn$ streams. Hence, we shall have the following function
-instead:
+code. Here by ``infinite'' we mean the notion that is internally available in
+the meta type theory. For instance, we can define a $\Nat_\M \to \Pull\,\vB$
+function which for each $\vn : \Nat_\M$ produces a concatenation of $\vn$
+streams. Hence, we shall have the following function instead:
 \[ \concatMap : \IsSOP\,\vA \RA (\vA \to \Pull\,\vB) \to \Pull\,\vA \to \Pull\,\vB, \]
@@
 axioms}, since they reflect the inability of metaprograms to inspect terms, and
-``generativity'' refers to this property in the staging literature \cite{TODO}.
+``generativity'' often refers to this property in the staging literature
+\cite{TODO}.
 
@@
 \vB\,(\msf{inject_i}\,\msf{loop_{\vA_i}})$ for each $\vA_i \in \vA$, where
-$({\times})$ is the product type former in $\USOP$ and $\msf{inject_i} :
-\Elprod\,\vA_i \to \El_\SOP\,\vA$. This is similar to the definition of
+$({\times})$ is the product type former in $\USOP$ and $\msf{inject_i}$
+has type $\Elprod\,\vA_i \to \El_\SOP\,\vA$. This is similar to the definition of
 non-dependent products in $\USOP$, except that we have to get rid of the type
@@
 Now, this operation should be avoided when possible, since it has two states and
-can contribute to a blow-up of state size in combination with state-multiplying
-operations such as $\mit{zip}$ or $\concatMap$. Both let-insertion and case
-splitting could be defined generically from $\single$, $\concatMap$ and the
-following operation (definition omitted):
-\[ \mit{mapGen} : (\vA \to \Gen\,\vB) \to \Pull\,\vA \to \Pull\,\vB \]
-However, $\single$ would introduce superfluous states that way. Instead, we give
-more efficient specialized definitions. Let-insertion is as follows:
-\begin{alignat*}{3}
-  & \hspace{-8em}\mathrlap{\genPull : \{\vA : \VTy\}\{\vB : \MTy\} \to \Lift \vA \to (\Lift \vA \to \Pull\,\vB) \to \Pull\,\vB}\\
-  & \hspace{-8em}\mathrlap{\genPull\,\va\,\vf = \Pull\,(\Sigma\,(\Lift\,\vA)\,(\mit{projS} \circ \vf))\,(({\va,}) \fmap \mit{projSeed}\,(\vf\,\va)) \dlr \lam \case}\\
-  & \hspace{-8em}\ind \Stop            &&\to \return\,\Stop \\
-  & \hspace{-8em}\ind \Skip\,\vs       &&\to \return \dlr \Skip\,(\va,\,\vs)\\
-  & \hspace{-8em}\ind \Yield\,\vb\,\vs &&\to \return \dlr \Yield\,\vb\,(\va,\,\vs)
-\end{alignat*}
-For case splitting, we have:
-\begin{alignat*}{3}
-  &\hspace{-12em}\mathrlap{\casePull : (\mit{Split}\,\vA,\,\IsSOP\,(\mit{SplitTo}\,\{\vA\})) \RA \Lift \vA \to (\mit{SplitTo}\,\{\vA\} \to \Pull\,\vB) \to \Pull\,\vB }\\
-  &\hspace{-12em}\mathrlap{\casePull\,\va\,\vf = \Pull\,(\Sigma\,(\mit{SplitTo}\,\{\vA\})\,(\mit{projS} \circ \vf))}\\
-  &\hspace{-12em}\mathrlap{    \hspace{7.1em}         (\mdo \{\va' \fro \mit{split}\,\va; \vs \fro \mit{projSeed}\,(\vf\,\va');\return (\va',\,\vs)\}) \dlr \lam \case}\\
-  &\hspace{-4em}\mathrlap{  \ind   (\va',\,\vs) \to \mit{projStep}\,(\vf\,\va')\,\vs \bind \lam \case} \\
-  &\hspace{-4em} \ind \ind \ind \Stop            &&\to \return \Stop \\
-  &\hspace{-4em} \ind \ind \ind \Skip\,\vs       &&\to \return \dlr \Skip\,(\va',\,\vs) \\
-  &\hspace{-4em} \ind \ind \ind \Yield\,\vb\,\vs &&\to \return \dlr \Yield\,\vb\,(\va',\,\vs)
-\end{alignat*}
-
-%% casePull' : ∀ {A B}⦃ _ : Split A ⦄ ⦃ _ : IsSOP (SplitTo {A}) ⦄ → ↑V A → (SplitTo {A} → Pull B) → Pull B
-%% casePull' {A} {B} {{_}}{{sopA}} a f =
-%%   pull (Σ (SplitTo {A}) (St ∘ f)) {{SOPΣ{{sopA}}{{λ {x} → StSOP (f x)}}}}
-%%     (do sp ← splitGen a; s ← seed (f sp); pure (_,_ {A = SplitTo {A}}{B = _} sp s)) λ where
-%%       (sp , s) → step (f sp) s >>= λ where
-%%          stop        → pure stop
-%%          (skip s)    → pure $ skip (sp , s)
-%%          (yield b s) → pure $ yield b (sp , s)
-
-%% To get good ergonomics in a ``monadic'' style of stream definitions, we need to
-%% reproduce let-insertion and case splitting on object values. First, we need
-%% singleton streams:
-
-%% genLetPull' : ∀ {A B} → ↑V A → (↑V A → Pull B) → Pull B
-%% genLetPull' {A} {B} a f =
-%%   pull (Σ (↑V A) (St ∘ f)) {{SOPΣ {{↑SOP}} {{λ {x} → StSOP (f x)}}}} ((a ,_) <$> (seed (f a))) λ where
-%%     (a , s) → step (f a) s <&> λ where
-%%       stop        → stop
-%%       (skip s)    → skip (a , s)
-%%       (yield b s) → yield b (a , s)
-
-%% Since $\Pull$ is not a monad, it cannot be $\MonadGen$ in our framework,
-%% but we still support some forms of let-insertion. First, we note that $\Pull$ can be
-%% mapped over by functions returning in $\Gen$:
-%% \begin{alignat*}{3}
-%%   &\hspace{-3em}\mathrlap{\mit{mapGen} : (\vA \to \Gen\,\vB) \to \Pull\,\vA \to \Pull\,\vB }\\
-%%   &\hspace{-3em}\mathrlap{\mit{mapGen}\,\vf\,(\Pull\,\vS\,\seed\,\step) = \Pull\,\vS\,\seed \dlr \lam \case}\\
-%%   &\hspace{-3em} \ind \Stop            &&\to \return\,\Stop \\
-%%   &\hspace{-3em} \ind \Skip\,\vs       &&\to \return \dlr \Skip\,\vs\\
-%%   &\hspace{-3em} \ind \Yield\,\va\,\vs &&\to \mdo \{\vb \fro \vf\,\va; \return \dlr \Yield\,\vb\,\vs\}
-%% \end{alignat*}
-%% Now, we may use $\mit{mapGen}\,\gen : \Pull\,(\Lift \vA) \to \Pull\,(\Lift \vA)$
-%% to let-bind all yielded expressions in a stream. This can be useful in stream
-%% operations that would otherwise duplicate yielded expressions. For example, the
-%% function with type $\Pull\,(\Lift \vA) \to \Pull\,(\Lift \vA,\,\Lift \vA)$
-%% should be defined with let-insertion.
-
-%% Let us now consider a more ``monadic'' style of stream programming. For this we
-%% need singleton streams:
-%% \begin{alignat*}{3}
-%%   & \mit{single} : \vA \to \Pull\,\vA \\
-%%   & \mit{single}\,\va = \Pull\,\Bool_\M\,\True \dlr \lam \vb.\,\return \dlr\mbf{if}\,\vb\,\mbf{then}\,\Yield\,\va\,\False\,\mbf{else}\,\Stop
-%% \end{alignat*}
-%% We also write $\mit{forEach}$ for $\concatMap$ with flipped arguments. Now, a
-%% more monadic flavor of let-insertion is given as
-%% \begin{alignat*}{3}
-%%   &\gen_{\Pull} : \Lift \vA \to \Pull\,(\Lift \vA) \\
-%%   &\gen_{\Pull}\,\va = \mit{mapGen}\,\gen\,(\mit{single}\,\va).
-%% \end{alignat*}
-%% We can also implement case splitting on object values. Recall from Section
-%% \ref{TODO} that for each value type we have a splitting function that returns a
-%% meta-level view on the constructors. Importantly, all such views are $\SOP$-s, and
-%% therefore can be bound by $\forEach$. A small example:
-%% \begin{alignat*}{3}
-%%   & \hspace{-5em}\mathrlap{\forEach\,(\mit{countFrom}\,\qt{0}) \dlr \lam \vn. }\\
-%%   & \hspace{-5em}\mathrlap{\forEach\,(\gen_{\Pull}\,\qt{\spl \vn * 2}) \dlr \lam \vm. }\\
-%%   & \hspace{-5em}\mathrlap{\forEach\,(\single\,(\msplit\,\qt{\spl \vm < 20})) \dlr \lam \case }\\
-%%   & \hspace{-5em}\ind \True_\M  &&\to \single\,\qt{\spl \vm + 10}   \\
-%%   & \hspace{-5em}\ind \False_\M &&\to \single\,\qt{\spl \vm + 20}
-%% \end{alignat*}
-%% Here $(\gen_{\Pull}\,\qt{\spl \vn * 2})$ behaves like a let-insertion, and
-%% $(\single\,(\msplit\,\qt{\spl \vm < 20}))$ implements the case splitting.
+can contribute to a blow-up of state size. For example, $\concatMap$ over
+$\single$ introduces at least a doubling of the number of state shapes. Although
+let-insertion and case splitting could be derived as $\concatMap$ over
+$\single$, to avoid the size explosion we give more specialized definitions
+instead. First, we define a helper function for binding for a single $\IsSOP$ value.
+\begin{alignat*}{3}
+  & \hspace{-11em}\mathrlap{\mit{bind_{single}} : \IsSOP\,\vA' \RA \Lift \vA \to (\Lift \vA \to \Gen\,\vA') \to (\vA' \to \Pull\,\vB) \to \Pull\,\vB}\\
+  & \hspace{-11em}\mathrlap{\mit{bind_{single}}\,\va\,\vf\,\vg =}\\
+  & \hspace{-11em}\mathrlap{\ind \Pull\,(\Sigma\,\vA'\,(\mit{projS} \circ \vg))\,(\mdo \{\va' \fro \vf\,\va; \vs \fro \mit{projSeed}\,(\vg\,\va'); \return (\va',\,\vs)) \dlr \lam \case}\\
+  & \hspace{-11em}\mathrlap{\ind \ind (\va',\,\vs) \to \mit{projStep}\,(\vg\,\va')\,\vs \bind \lam \case }\\
+  & \hspace{-11em}\ind \ind \ind \Stop            &&\to \return \Stop\\
+  & \hspace{-11em}\ind \ind \ind \Skip\,\vs       &&\to \return \dlr \Skip\,(\va',\,\vs) \\
+  & \hspace{-11em}\ind \ind \ind \Yield\,\vb\,\vs &&\to \return \dlr \Yield\,\vb\,(\va',\,\vs)
+\end{alignat*}
+We recover let-insertion and case splitting as follows:
+\begin{alignat*}{3}
+  &\mit{gen_\Pull} : \Lift \vA \to (\Lift \vA \to \Pull\,\vB) \to \Pull\,\vB\\
+  &\mit{gen_\Pull}\,\va = \mit{bind_{single}}\,\va\,\gen
+\end{alignat*}
+\begin{alignat*}{3}
+  &\mit{case_\Pull} : (\mit{Split}\,\vA,\,\IsSOP\,(\mit{SplitTo}\,\{\vA\})) \RA \Lift \vA \to (\mit{SplitTo}\,\{\vA\} \to \Pull\,\vB) \to \Pull\,\vB\\
+  &\mit{case_\Pull}\,\va = \mit{bind_{single}}\,\va\,\mit{split}
+\end{alignat*}
+Let us look at small example. We write $\mit{forEach}$ for $\concatMap$ with
+flipped arguments.
+\begin{alignat*}{3}
+  &\hspace{-4em}\mathrlap{\mit{forEach}\,(\mit{take}\,\qt{100}\,(\mit{countFrom}\,\qt{0})) \dlr \lam \vx.}\\
+  &\hspace{-4em}\mathrlap{\gen_\Pull\, \qt{\spl \vx * 2} \dlr \lam \vy.}\\
+  &\hspace{-4em}\mathrlap{\mit{case_\Pull}\,\qt{\spl \vx < 50} \dlr \lam \case}\\
+  &\hspace{-4em} \ind \True  &&\to \mit{take}\,\vy\,(\mit{countFrom}\,x)\\
+  &\hspace{-4em} \ind \False &&\to \single\,\vy
+\end{alignat*}
+Here, in every $\forEach$ iteration, $\gen_\Pull\,\qt{\spl \vx * 2}$ evaluates
+the given expression and saves the result to the machine state. Then,
+$\mit{case_\Pull}$ branches on an object-level Boolean expression. If we use
+$\foldr$ to generate object code from this definition, we get \emph{four}
+mutually defined functions. We get two state shapes from $\single\,\vy$ and one
+from $\mit{take}\,\vy\,(\mit{countFrom}\,x)$; we sum these to get three for the
+$\mit{case_\Pull}$, then we get an extra shape from $\mit{forEach}$ which
+introduces an additional $\Maybe$.
 
@@
 
-The Agda implementation follows this section very closely, so we do not detail it.
+In Agda, to avoid computation getting stuck on the generativity axiom, we use
+the \texttt{primTrustMe} built-in \cite{TODO} to automatically erase the axiom
+when the sides of the equation are definitionally equal. Otherwise the
+implementation closely follows this section.
 
@@
 types, we define $\concatMap$ just for object types,\footnote{Recall that we do
-not distinguish value and computation types in Haskell.} and then define a case
-splitting function separately for each type. In each of these functions, we only
-need to deal with a concrete finite number of different machine state types,
-which is feasible with weak existentials. For example, we define
-\[ \mit{split_{\Pull}} : \Lift (\List\,\vA) \to (\Maybe_\M\,(\Lift \vA,\,\Lift (\List\,\vA)) \to \Pull\,\vB) \to \Pull\,\vB. \]
+not distinguish value and computation types in Haskell.} and define
+$\mit{case_\Pull}$ separately for each object type, as an overloaded class
+method. In each of these definitions, we only need to deal with a concrete
+finite number of machine state types, which is feasible to handle with weak
+existentials.
+
 Also, the generativity axiom is \emph{false} in Template Haskell, since it is
 possible to look inside quoted expressions. Instead, we use type coercions that
-can fail at staging time, when users of the library violate generativity.
+can fail at staging time. If library users do not violate generativity, these
+coercions disappear and the staging output will not contain unsafe coercions.
+
+\section{Related Work}\label{sec:related-work}
 

### 2024-02-26 newpaper/paper.tex
first draft of paper version

@@
 \newcommand{\vu}{\mathsf{u}}
+\newcommand{\vl}{\mathsf{l}}
+\newcommand{\vns}{\mathsf{ns}}
 
@@
 \newcommand{\List}{\msf{List}}
+\newcommand{\Tree}{\msf{Tree}}
+\newcommand{\Node}{\msf{Node}}
+\newcommand{\Leaf}{\msf{Leaf}}
 \newcommand{\Nil}{\msf{Nil}}
@@
 \newcommand{\id}{\mit{id}}
-\newcommand{\mup}{\mit{up}}
-\newcommand{\mdown}{\mit{down}}
+\newcommand{\mup}{\mbf{up}}
+\newcommand{\mdown}{\mbf{down}}
 \newcommand{\tyclass}{\mbf{class}}
@@
 fragile and non-compositional. \texttt{INLINE} and \texttt{REWRITE} pragmas can
-be used to control code generation, but without any strong guarantee, and their
-advanced usage requires knowledge of GHC internals. For example, correctly
-specifying the \emph{ordering} of certain rule applications is often needed. We
-also have to care about formal function arities. Infamously, the function
-composition operator is defined as $(.)\,\vf\,\vg = \lam \vx \to \vf\,(\vg\,\vx)$ in the
-base libraries, instead of as $(.)\,\vf\,\vg\,\vx = \vf\,(\vg\,\vx)$, to get better inlining
-behavior \cite{TODO}. It remains a common practice in high-performance Haskell
-programming to manually review GHC's optimized code output.
+be used to control code generation \cite{ghcdocs}, but without strong
+guarantees, and their advanced usage requires knowledge of GHC internals. For
+example, correctly specifying the \emph{ordering} of certain rule applications
+is often needed. We also have to care about formal function arities. Infamously,
+the function composition operator is defined as $(.)\,\vf\,\vg = \lam \vx \to
+\vf\,(\vg\,\vx)$ in the base libraries, instead of as $(.)\,\vf\,\vg\,\vx =
+\vf\,(\vg\,\vx)$, to get better inlining behavior \cite{ghcbase}. It remains a
+common practice in high-performance Haskell programming to manually review GHC's
+optimized code output.
 
@@
 
-The idea of \textbf{two-level type theory} (2LTT) is to use a highly expressive
-dependent type theory for compile-time computation, but generate code in
-possibly different, simpler object language. In this paper, we use 2LTT to sweeten the
+\textbf{Two-level type theory} (2LTT) \cite{twolevel,staged2ltt} makes it
+possible to use expressive dependent type theories for metaprogramming
+for a wide range of object languages. In this paper, we use 2LTT to sweeten the
 deal of staged compilation, aiming for a combination of strong guarantees, good
@@
 call \textbf{CFTT}, short for ``closure-free type theory''. This consists
-of
+of:
 \begin{itemize}
@@
       in the downstream pipeline, but it lacks many convenience features.
-\item A standard Martin-Löf type theory for the compile-time language. This
+\item A dependent type theory for the compile-time language. This
       allows us to recover many features by metaprogramming.
 \end{itemize}
-
 Since the object language is first-order, we guarantee that all programs in CFTT
 can be ultimately compiled without any dynamic closures, using only calls and
-jumps to statically known code. Why focus on closures?  They are the
+jumps to statically known code. Why emphasize closures?  They are the
 foundation to almost all abstraction tools in functional programming:
@@
 \item Higher-order functions in essentially all functional languages are implemented with closures.
-\item Type classes in Haskell use dictionary-passing, which relies on closures for function methods.
+\item Type classes in Haskell use dictionary-passing, which relies on closures for function methods \cite{DBLP:conf/popl/WadlerB89}.
 \item Functors and first-class modules in OCaml and other ML-s rely on closures.
@@
 %% \todo{Essential closures: CPS? Threaded interpreters?}
-
 %% Whole-program defunctionalization is also possible, and it is notably used by
@@
 \begin{itemize}
-\item In Section \ref{TODO} we present the two-level type theory CFTT, where the
-  object language is first-order simply-typed and the metalanguage is
-  dependently typed. The object language admits an operational semantics without
+\item In Section \ref{sec:overview-of-cftt} we present the two-level type theory
+  CFTT, where the object level is first-order simply-typed and the meta level is
+  dependently typed. The object language supports an operational semantics without
   runtime closures, and can be compiled in a way such that all function calls
@@
   operational semantics.
-\item In Section \ref{TODO} we build a monad transformer library. We believe
+\item In Section \ref{sec:monad-transformers} we build a monad transformer library. We believe
   that this is a good demonstration, because monads and monad transformers are
@@
   and case switching on object-level data, in the generality of monad transformers.
-\item In Section \ref{TODO} we build a pull-based stream fusion library. Here,
-  the motivation is to demonstrate essential usage of dependent types to
-  concisely solve a difficult problem, namely guaranteed fusion for unrestricted
-  combinations of $\concatMap$ and $\mit{zip}$. We use a state machine
-  representation that is based on \emph{sums-of-products} of object-level
-  values. We show that CFTT is compatible with a \emph{generativity} axiom,
-  which internalizes the fact that metaprograms cannot inspect the structure of
-  object-level terms. We use this to show that the universe of sums-of-products
-  is closed under $\Sigma$-types. This in turn enables a very
-  concise definition of $\concatMap$.
+\item In Section \ref{sec:stream-fusion} we build a pull-based stream fusion library. Here,
+  we demonstrate essential usage of dependent types, for providing guaranteed
+  fusion for arbitrary combinations of $\concatMap$ and $\mit{zip}$. We use a
+  state machine representation that is based on \emph{sums-of-products} of
+  object-level values. We show that CFTT is compatible with a
+  \emph{generativity} axiom, which internalizes the fact that metaprograms
+  cannot inspect the structure of object-level terms. We use this to show that
+  the universe of sums-of-products is closed under $\Sigma$-types. This in turn
+  enables a very concise definition of $\concatMap$.
 \item
-  We adapt the contents of the paper to a Template Haskell, with some
-  modifications, simplifications and fewer guarantees about generated code.  In
-  particular, Haskell does not have enough dependent types for the simple
-  $\concatMap$ definition, but we can still work around this limitation. We also
-  provide a more faithful Agda embedding of CFTT and our libraries. Here, the
-  object theory is embedded as a collection of postulated operations, and we can
-  use Agda's normalization command to print out generated object code.
+  We adapt the contents of the paper to typed Template Haskell
+  \cite{DBLP:journals/pacmpl/XiePLWYW22}, with some modifications,
+  simplifications and fewer guarantees about generated code.  In particular,
+  Haskell does not have enough dependent types for the simple $\concatMap$
+  definition, but we can still work around this limitation. We also provide a
+  precise Agda embedding of CFTT and our libraries as described in this
+  paper. Here, the object theory is embedded as a collection of postulated
+  operations, and we can use Agda's normalization command to print out generated
+  object code.
 \end{itemize}
@@
 ``metaprogram'' for inhabitants of metatypes. $\MTy$ supports dependent
-functions, products and indexed inductive types \cite{TODO}.
+functions, products and indexed inductive types \cite{inductivefamilies}.
 
 Formally, $\MTy$ is additionally indexed by universes levels, and we have
-$\MTy_i : \MTy_{i+1}$. However, this adds a bit of noise, and it is not very
-relevant to the current paper, so we shall omit levels. Note that universe
-levels are orthogonal to staging; they are instead used to increase the strength
-of the system while maintaining logical consistency.
+$\MTy_i : \MTy_{i+1}$. This is orthogonal to the levels of staging. However,
+universe levels add noise and they are not too relevant to the current paper, so
+we will omit them.
 
@@
 \[\data \msf{Record} = \msf{Record}\,\{\mit{field1} : \vA,\,\mit{field2} : \vB\}.\]
-
 All construction and elimination rules for type formers in $\MTy$ stay within
@@
 $\bs{\Ty}$ is the universe of types in the object language. It is itself a
-metatype, so so we have $\Ty : \MTy$. Similarly as in the case of $\MTy$, all
-construction and elimination rules of the object language stay within $\Ty$.
-However, we further split $\Ty$ to two sub-universes.
+metatype, so so we have $\Ty : \MTy$. All construction and elimination rules of
+type formers in $\Ty$ stay within $\Ty$. We further split $\Ty$ to two
+sub-universes.
 
@@
 for meta-level ones. We also have non-recursive $\mbf{let}$, which can be used
-to define computations and values as well, and can be used to shadow binders:
+to define computations and values alike, and can be used to shadow binders:
 \begin{alignat*}{3}
@@
 \vb\,\vx.\,\case \vb\,\of \True \to \vx + 10; \False \to \vx * 10$, and more
-generally we can transform programs so that every function call is
+generally we can transform programs so that every function call becomes
 \emph{saturated}. This means that every function call is of the form
@@
 \subsubsection{Comparison to call-by-push-value}
-We took inspiration from call-by-push-value (CBPV), and there are similarities,
-but there are also significant differences. Both systems have a
-value-computation distinction, with call-by-name computations and call-by-value
-values. However, our object theory supports variable binding at arbitrary types
-while CBPV only supports value variables. In CBPV, a let-definition for a
-function is only possible by first packing it up as a closure value (or
-``thunk''), which clearly does not suit us. In future work we could make
-a closer look at the connections to CBPV.
+We took inspiration from call-by-push-value (CBPV) \cite{DBLP:conf/tlca/Levy99},
+and there are similarities, but there are also significant differences. Both
+systems have a value-computation distinction, with call-by-name computations and
+call-by-value values. However, our object theory supports variable binding at
+arbitrary types while CBPV only supports value variables. In CBPV, a
+let-definition for a function is only possible by first packing it up as a
+closure value (or ``thunk''), which clearly does not suit our applications.
+Investigating the relation between CBPV and our object language could be future work.
 
@@
 %% thus easier to compile, and making metaprogramming more convenient. We will see
-%% that the ability to insert $\mbf{let}$-s without restriction is very convenient,
-%% and likewise the ability to have $\lambda$-abstraction under $\mbf{case}$
-%% bodies, although these features necessitate more downstream processing, to bring
-%% programs to saturated form. In this paper we go for metaprogramming convenience,
+%% that the ability to insert $\mbf{let}$-s without restriction is convenient, and
+%% likewise the ability to have $\lambda$-abstraction under $\mbf{case}$ bodies,
+%% although these features necessitate more downstream processing, to bring
+%% programs to saturated form. On the other side of the spectrum, one might imagine
+%% generating object code in low-level A-normal form. This is more laborious, but
+%% it could be also interesting, because it forces us to invent abstractions for
+%% manipulating ANF in the metalanguage. In a similar vein, Allais recently
+%% proposed metaprogramming quantum circuits in a two-level type theory
+%% \cite{TODO}.
 
@@
       is the metaprogram which immediately returns $\vt$.
-\item For $\vt : \Lift \vA$, we have $\spl \vt : A$, pronounced ``splice $\vt$''. This
-  inserts the result of a metaprogram into an object term. \emph{Notation:}
-  splicing binds stronger than function application, so $\vf\,\spl \vx$ is parsed as
-  $\vf\,(\spl \vx)$.
+\item For $\vt : \Lift \vA$, we have $\spl \vt : A$, pronounced ``splice
+  $\vt$''. This inserts the result of a metaprogram into an object
+  term. \emph{Notation:} splicing binds stronger than function application, so
+  $\vf\,\spl \vx$ is parsed as $\vf\,(\spl \vx)$. We borrow this notation from
+  MetaML \cite{metaml}.
 \item We have $\ql \spl \vt \qr \equiv \vt$ and $\spl \ql \vt \qr \equiv \vt$ as definitional equalities.
@@
 and produces output which is purely in the object-level fragment, with no
-metatypes and metaprograms remaining. The output is guaranteed to be
-well-typed. This staging is formalized in detail in \cite{TODO}. In Section
-\ref{TODO} we describe the modifications to ibid.\ that are used in this paper.
+metatypes and metaprograms remaining. The output is guaranteed to be well-typed.
 
@@
 
-\subsection{Semantics of Staging}
+\subsection{Semantics of Staging}\label{sec:semantics-of-staging}
 
 We give a short summary of the semantics of staging here. We reuse most of the
-semantics from \cite{TODO}, also borrowing notation and definitions, and only
-mention additions and differences. The reader may refer to ibid.\ for technical
-details.
+semantics from \cite{staged2ltt} and only mention additions and differences. The
+reader may refer to ibid.\ for technical background.
 
@@
 which is constantly the set of object-theoretic types. Soundness of staging is
-proven by a separate logical relations model, internally to the mentioned
-presheaf model. Soundness means that the output of staging, viewed as a CFTT
-term, is definitionally equal to the input.
+proven by a logical relation model, internally to the mentioned presheaf
+model. Soundness means that the output of staging, viewed as a CFTT term, is
+definitionally equal to the input.
 
 The main addition in CFTT is the assumption of all inductive families in the
-meta-level. However, based on the results in \cite{TODO}, it is enough to
-extend the meta-level syntax in \cite{TODO} with W-types and identity types to
-derive all inductive families, so we consider these in the following.
+meta-level. However, based on the results in \cite{whynotw}, it is enough to
+extend the meta-level syntax in \cite{staged2ltt} with W-types and identity types to
+derive all inductive families. We consider the semantics of these in the following.
 
 \subsubsection{W-types} W-types have a standard object-wise inductive definition in
-the presheaf model \cite{TODO}. Additionally, we need to give a logical relation
+the presheaf model \cite{DBLP:journals/apal/MoerdijkP00}. Additionally, we need to give a logical relation
 interpretation to W-types, as required for the soundness proof for staging. In
-Section 5.2.\ of \cite{TODO} a serialization map is used to define the relation
+Section 5.2.\ of \cite{staged2ltt} a serialization map is used to define the relation
 for natural numbers, however, there is no analogous map for W-types because we
@@
 \subsubsection{Identity type} Our treatment of the identity type is
-influenced by the following: we would like to assume identities as axioms in
+influenced by the following: we would like to assume equations as axioms in
 CFTT, which are true in the presheaf model (i.e.\ they are true at
-staging-time), without blocking staging as a computation. In Section \ref{TODO}
+staging-time), without blocking staging as a computation. In Section \ref{sec:concatmap-for-streams}
 we use such an axiom to good effect. However, axioms do block computation in
@@
 \begin{itemize}
-\item Have a standard intensional identity type in CFTT, perhaps with uniqueness of identity proofs assumed.
-\item Before staging, translate CFTT to CFTT plus equality reflection. Operationally,
-      this means erasing all identity proofs and transports.
-\item Perform staging after erasure of identity proofs.
+\item Have a standard intensional identity type in CFTT.
+\item Before staging, translate the intensional identity to extensional
+      identity. This amounts to erasing all identity proofs and transports.
+\item Perform staging after erasure of identity proofs, i.e.\ model extensional identity in the semantics.
 \end{itemize}
 In the presheaf model, we interpret the extensional identity in a standard way,
-as strict equality of sections of presheaves. With this, soundness of staging is
-trivial to show; we can pick the trivial relation for the interpretation of
-identity. Hence, we weaken the statement of soundness: every CFTT program is
-convertible to the staging output, up to erasure of identity proofs.
+as strict equality of sections of presheaves, and soundness for extensional
+identity becomes trivial to show. Hence, the soundness property for our overall
+staging pipeline becomes the following: the staging output is definitionally
+equal to the identity-erased version of the staging input.
 
@@
 that this is a good demonstration of CFTT-s abilities, since monads are
-ubiquitous in Haskell programming, and they also introduce a great amount of
+ubiquitous in Haskell programming and they also introduce a great amount of
 abstraction that should be optimized away.
@@
 In the staged compilation and partial evaluation literature, the term
-\emph{binding time improvement} is used to refer to such conversions, where
-the ``improved'' version supports more compile-time computation. A general
-strategy for generating efficient ``fused'' programs, is to try to work as much
-as possible with improved representations, and only convert back to object code
-at points where runtime dependencies are unavoidable. Let us look at
-binding-time-improvement for product types now:
+\emph{binding time improvement} is used to refer to such conversions, where the
+``improved'' version supports more compile-time computation
+\cite{partial-evaluation}. A general strategy for generating efficient ``fused''
+programs, is to try to work as much as possible with improved representations,
+and only convert back to object code at points where runtime dependencies are
+unavoidable. Let us look at binding-time-improvement for product types now:
 \begin{alignat*}{3}
@@
 \end{alignat*}
-
-%% We use a Haskell-style type class for binding time improvements:
-%% \begin{alignat*}{3}
-%%   &\hspace{-8em}\mathrlap{\tyclass\,\Improve\,(A : \Ty)\,(B : \MTy)\,\where} \\
-%%   &\hspace{-8em}\ind \mup   &&: \Up A \to B\\
-%%   &\hspace{-8em}\ind \mdown &&: B \to \Up A
-%%   &\\
-%%   &\hspace{-8em}\mathrlap{\instance \Improve\,(A \to B)\,(\Up A \to \Up B)\,\where\,\,...}
-%% \end{alignat*}
-%% We will use type classes in an informal way, without precisely specifying how
-%% they work. However, in the Haskell adaptation of this paper we do make crucial
-%% use of type classes. There are some necessary differences between the CFTT and
-%% Haskell versions though, which we will summarize in Section \ref{TODO}. Let us
-%% look at improvement for product types now:
-%% \begin{alignat*}{3}
-%%   &\hspace{-4em}\mathrlap{\instance \Improve\,(A,\,B)\,(\Up A,\,\Up B)\,\where}\\
-%%   &\hspace{-4em}\ind \mup\,x   && = (\ql \fst\,\spl x \qr,\, \ql \snd\,\spl x \qr)\\
-%%   &\hspace{-4em}\ind \mdown\,(a,\,b) && = \ql(\spl a,\, \spl b)\qr
-%% \end{alignat*}
 
@@
 \end{alignat*}
-From now on, we reuse Haskell-style type classes and $\mdo$-notation in CFTT.
-We will use type classes in an informal way, without precisely specifying how
-they work. However, type classes are used in essentially the same way in the
-Haskell implementation of the paper, and also in the Agda implementation, with
-modest technical differences.
+From now on, we reuse Haskell-style type classes and $\mbf{do}$-notation in
+CFTT. We will use type classes in an informal way, without precisely specifying
+how they work. However, type classes are used in essentially the same way in the
+Agda and Haskell implementations of the paper, with modest technical
+differences.
 
@@
   & \genRec : \{\vA : \CTy\} \to (\Up \vA \to \Up \vA) \to \Gen\,(\Up \vA) \\
-  & \genRec\,\vf = \Gen\ \dlr \lam \vk.\,\qt{\letrec \vx : \vA := \spl(\vf\,\qt{\vx}); \spl(\vk\,\qt{\vx})}
+  & \genRec\,\vf = \Gen \dlr \lam \vk.\,\qt{\letrec \vx : \vA := \spl(\vf\,\qt{\vx}); \spl(\vk\,\qt{\vx})}
 \end{alignat*}
@@
   &\hspace{-2em}\mup\,\vx = \mdo &&\vx \fro \gen\,\vx                                    && \ind\ind \mdown\,x = \mdo &&(\va,\,\vb) \fro x\\
-  &\hspace{-2em}               &&\return (\qt{\fst\,\spl \vx},\,\qt{\snd\,\spl \vx})\} &&                           &&\return \qt{(\spl \va,\, \spl \vb)}
+  &\hspace{-2em}               &&\return (\qt{\fst\,\spl \vx},\,\qt{\snd\,\spl \vx})     &&                           &&\return \qt{(\spl \va,\, \spl \vb)}
 \end{alignat*}
@@
 Working in $\Gen$ is convenient, since we can freely generate object code and
-also have access to the full metalanguage. Also, the whole point of staging is
-that \emph{eventually} all metaprograms will be used for the purpose of code
-generation, so ultimately we always want to $\runGen$ our actions. So why not
-just always work in $\Gen$? The implicitness of $\Gen$ may make it harder to
+also have access to the full metalanguage. Also, the point of staging is that
+\emph{eventually} all metaprograms will be used for the purpose of code
+generation, so we eventually $\runGen$ all of our actions. So why not
+just always work in $\Gen$? The implicit nature of $\Gen$ may make it harder to
 reason about the size and content of generated code. This is a bit similar to
 the $\msf{IO}$ monad in Haskell, where eventually everything needs to run in
-$\msf{IO}$, but we may prefer to not write most of our program in $\msf{IO}$.
+$\msf{IO}$, but we may not want to write all of our code in $\msf{IO}$.
 
@@
 type for binding:
-\[ (\bind) : \Up(\Maybe\,\vA) \to (\Up\,\vA \to \Up\,(\Maybe\,\vB)) \to \Up(\Maybe\,\vB) \]
+\[ (\bind) : \Up(\Maybe\,\vA) \to (\Up\vA \to \Up(\Maybe\,\vB)) \to \Up(\Maybe\,\vB) \]
 This works, but the definition necessarily uses runtime case splits on $\Maybe$
@@
 Assume that $\Maybe_\M$ is the standard meta-level monad, and $\MaybeT_\M$ is
-the standard monad transformer:
+the standard monad transformer, defined as follows:
 \[ \newtype \MaybeT_\M\,\vM\,\vA = \MaybeT_\M\,\{\runMaybeT_\M : \vM\,(\Maybe_\M\,\vA)\} \]
@@
   %% &\ind \mup : \Up\,(\Maybe\,\vA) \to \MaybeT_\M\,\Gen\,(\Up\,\vA) \\
-  &\ind \mup\,\vx = \MaybeT_\M \dlr \Gen \dlr \lam \vk.\\
-  &\ind\ind \qt{\case \spl \vx\,\of \Nothing \to \spl(\vk\,\Nothing_\M);\Just\,\va \to \spl(\vk\,(\Just_\M\qt{\va}))}\\
+  &\ind \mup\,\vma = \MaybeT_\M \dlr \Gen \dlr \lam \vk.\\
+  &\ind\ind \qt{\case \spl \vma\,\of \Nothing \to \spl(\vk\,\Nothing_\M);\Just\,\va \to \spl(\vk\,(\Just_\M\qt{\va}))}\\
   %% &\ind \mdown : \MaybeT_\M\,\Gen\,(\Up\,A) \to \Up\,(\Maybe\,A) \\
@@
 However, it is more convenient to proceed in the style of Haskell's monad
-transformer library \cite{TODO}, and have a class for monads that can do code
-generation:
+transformer library \texttt{mtl} \cite{mtl}, and have a class for monads that
+support code generation:
 \begin{alignat*}{3}
@@
 \subsubsection{Case splitting in monads}
-We often want to case-split on object-level data inside a monadic action, like
-in the following:
-\begin{alignat*}{3}
-  & \hspace{-4em}\mathrlap{\vf : \Nat \to \Maybe\,\Nat}\\
-  & \hspace{-4em}\mathrlap{\vf\,\vx := \spl(\mdown\,\$\, \case \vx == 10\,\of}\\
-  & \hspace{-4em}\ind \True  &&\to \return\,\qt{\vx + 5};\\
-  & \hspace{-4em}\ind \False &&\to \Nothing_\M)
-\end{alignat*}
-This is ill-typed as written, since we cannot compute meta-level actions from an
-object-level case split. Fortunately, with a little bit more work, case
-splitting on object values is actually possible in any $\MonadGen$, on any value
-type.
+We often want to case-split on object-level data inside a monadic action, and
+perform different actions in different branches. At first, this may seem
+problematic, because we cannot directly compute metaprograms from object-level
+case splits. Fortunately, with a little bit more work, case splitting on object
+values is actually possible in any $\MonadGen$, on any value type.
 
@@
 \end{alignat*}
-Now, in any $\MonadGen$, we may write
+Now, in any $\MonadGen$, assuming $\vas : \Up(\List\,\vA)$, we may write
 \begin{alignat*}{3}
@@
 \end{alignat*}
-In a native implementation of CFTT it may make sense to extend $\mdo$-notation,
-so that we elaborate $\mbf{case}$ on object values to an application of the
+In a native implementation of CFTT it may make sense to extend do-notation, so
+that we elaborate $\mbf{case}$ on object values to an application of the
 appropriate $\mit{split}$ function. We adopt this in the rest of the paper, so
-we will be able to write $\case \vx\,\of ...$ in a $\mdo$-block, whenever we
-work in a $\MonadGen$ and $\vx$ has a value type.
+when working in a $\MonadGen$, we can write $\case \vas\,\of \Nil \to ...;
+\Cons\,\va\,\vas \to ...$, binding $\va : \Up \vA$ and $\vas : \Up(\List\,\vA)$
+in the $\Cons$ case.
 
@@
 \begin{alignat*}{3}
-  & \newtype \Identity\,\vA := \Identity\,\{\runIdentity : \vA\} \\
-  & \instance \Improve\,\Identity\,\Gen\,\where\\
-  & \ind \mup\,\vx = \Gen \dlr \lam \vk.\,\vk\,\qt{\runIdentity\,\spl \vx}\\
-  & \ind \mdown\,\vx = \unGen\,\vx \dlr \lam \va.\,\qt{\Identity\,\spl \va}
+  & \hspace{-2em}\mathrlap{\newtype \Identity\,\vA := \Identity\,\{\runIdentity : \vA\} }\\
+  & \hspace{-2em}\mathrlap{\instance \Improve\,\Identity\,\Gen\,\where}\\
+  & \hspace{-2em}\ind \mup\,\vx   &&= \return \qt{\runIdentity\,\spl \vx}\\
+  & \hspace{-2em}\ind \mdown\,\vx &&= \unGen\,\vx \dlr \lam \va.\,\qt{\Identity\,\spl \va}
 \end{alignat*}
@@
 \end{alignat*}
-In the $\mbf{case}$ in $\mup$, we already use our syntax sugar for matching on a
+In the $\mbf{case}$ in $\mup$, we use our syntax sugar for matching on a
 $\Maybe$ value inside an $\vM$ action. This is legal, since we know from the
@@
 
-In the meta level, we can reuse essentially all definitions from Haskell's monad
-transformer library $\msf{mtl}$. From $\msf{mtl}$, only the continuation monad
-transformer fails to support binding-time-improvement in CFTT, because of the
-obvious need for dynamic closures. In the following we present only $\StateT$
-and $\ReaderT$. Starting with $\StateT$, we assume $\StateT_\M$ as the standard
-meta-level definition. The object-level $\StateT$ has type $(\vS : \VTy)(\vF : \VTy
-\to \Ty)(\vA : \VTy) \to \VTy$; the state parameter $\vS$ has to be a value type,
+In the meta level, we can reuse essentially all definitions from \texttt{mtl}.
+From there, only the continuation monad transformer fails to support
+binding-time-improvement in CFTT, because of the obvious need for dynamic
+closures. In the following we only present $\StateT$ and $\ReaderT$.
+
+We start with $\StateT$. We assume $\StateT_\M$ as the standard meta-level
+definition. The object-level $\StateT$ has type $(\vS : \VTy)(\vF : \VTy \to
+\Ty)(\vA : \VTy) \to \VTy$; the state parameter $\vS$ has to be a value type,
 since it is an input to an object-level function.
@@
 \begin{alignat*}{3}
-  &\instance \Improve\,\vF\,\vM \RA \Improve\,(\ReaderT\,\vR\,\vF)\,(\ReaderT_\M\,(\Up \vR)\,\vM)\,\where\\
-  &\ind \mup\,\vx   = \ReaderT_\M \dlr \lam \vr.\, \mup\,\qt{\runReaderT\,\spl \vx\,\spl \vr}\\
-  &\ind \mdown\,\vx = \qt{\ReaderT\,(\lam \vr.\,\spl(\mdown\,(\runReaderT_\M\,\vx\,\qt{\vr})))}
-\end{alignat*}
-
-\subsubsection{$\State$ and $\Reader$ operations} If we try to use
-the $\modify$ function that we already have for $\State_\M$, a curious thing
-happens. The meaning of $\modify\,(\lam \vx.\,\qt{\spl \vx + \spl \vx})$ is to replace
-the current state $\vx$, as an object expression, with the expression $\qt{\spl \vx
-  + \spl \vx}$, and this happens at staging time. This behaves as an ``inline''
-modification which replaces every subsequent mention of the state with a
-different expression. For instance, ignoring newtype wrappers for now,
+  &\hspace{-5em}\mathrlap{\instance \Improve\,\vF\,\vM \RA \Improve\,(\ReaderT\,\vR\,\vF)\,(\ReaderT_\M\,(\Up \vR)\,\vM)\,\where}\\
+  &\hspace{-5em}\ind \mup\,\vx   &&= \ReaderT_\M \dlr \lam \vr.\, \mup\,\qt{\runReaderT\,\spl \vx\,\spl \vr}\\
+  &\hspace{-5em}\ind \mdown\,\vx &&= \qt{\ReaderT\,(\lam \vr.\,\spl(\mdown\,(\runReaderT_\M\,\vx\,\qt{\vr})))}
+\end{alignat*}
+
+\subsubsection{$\State$ and $\Reader$ operations} If we use
+the $\modify$ function that we already have in $\State_\M$, a curious thing
+happens. The meaning of $\modify\,(\lam \vx.\,\qt{\spl \vx + \spl \vx})$ is to
+replace the current state $\vx$, as an object expression, with the expression
+$\qt{\spl \vx + \spl \vx}$, and this happens at staging time. This behaves as an
+``inline'' modification which replaces every subsequent mention of the state
+with a different expression. For instance, ignoring newtype wrappers for now,
 \[ \mdown \dlr \mdo \{\modify\,(\lam \vx.\,\qt{\spl \vx + \spl \vx}); \modify\,(\lam \vx.\,\qt{\spl \vx + \spl \vx});\return\,\qt{()}\} \]
@@
 There is a deficiency in our library so far. Assuming $\vb : \Bool$ as an object value, consider:
-\begin{alignat*}{3}
-  & \mdown \dlr \mdo \\
-  & \ind \case \vb\,\of\\
-  & \ind \ind \True  \hspace{0.2em}\to \mput' \qt{10}\\
-  & \ind \ind \False \to \mput' \qt{20}\\
-  & \ind \modify' (\lam \vx. \qt{\spl \vx + 10})
-\end{alignat*}
-The $\modify' (\lam \vx. \qt{\spl \vx + 10})$ action gets inlined to both $\mbf{case}$
+\begin{alignat*}{5}
+  & \hspace{-1em}\mdo && (\case \vb\,\of\,\True\to \mput' \qt{10}; \False &&\to \mput' \qt{20})\\
+  & \hspace{-1em}     && \mathrlap{\modify' (\lam \vx.\,\qt{\spl \vx + 10})}
+\end{alignat*}
+The $\modify' (\lam \vx. \qt{\spl \vx + 10})$ action gets inlined in both case
 branches during staging. This follows from the definition of monadic binding in
@@
 generation is continued in both branches with the same action. If we have
-multiple $\mbf{case}$ splits sequenced after each other, that yields exponential
-code size. Right now, we can fix this by let-binding the problematic action:
-\begin{alignat*}{3}
-  & \mdown \dlr \mdo \\
-  & \ind \vx \fro \gen \dlr \mdown \dlr \case \vb\,\of\\
-  & \ind \ind \True  \hspace{0.2em}\to \mput' \qt{10}\\
-  & \ind \ind \False \to \mput' \qt{20}\\
-  & \ind \mup\,\vx\\
-  & \ind \modify' (\lam \vx. \qt{\spl \vx + 10})
+multiple Boolean $\mbf{case}$ splits sequenced after each other, that yields
+exponential code size. Right now, we can fix this by let-binding the problematic
+action:
+%% \begin{alignat*}{5}
+%%   & \mdo && (\gen \dlr \mdown\,(\case \vb\,\of \True \to \mput' \qt{10};\False &&\to \mput' \qt{20})) \bind \mup \\
+%%   &      && \mathrlap{\modify' (\lam \vx.\,\qt{\spl \vx + 10})}
+%% \end{alignat*}
+\begin{alignat*}{5}
+  & \mdo && \msf{action} \fro \gen \dlr \mdown\,(\case \vb\,\of \True \to \mput' \qt{10};\False &&\to \mput' \qt{20}) \\
+  &      && \mup\,\msf{action}\\
+  &      && \mathrlap{\modify' (\lam \vx.\,\qt{\spl \vx + 10})}
 \end{alignat*}
@@
 We could proceed with finite sums, but we will need finite
-\emph{sums-of-products} (SOP in short) later in Section in \ref{TODO}, so we
-develop SOP-s. We view SOP-s as a Tarski-style universe consisting of a type of
-descriptions, and a way of interpreting descriptions into $\MTy$ (``$\El$'' for
+\emph{sums-of-products} ($\SOP$ in short) later in Section in \ref{sec:stream-fusion}, so we
+develop SOP-s. We view $\SOP$-s as a Tarski-style universe consisting of a type of
+descriptions and a way of interpreting descriptions into $\MTy$ (``$\El$'' for
 ``elements'' of the type).
@@
 A type description is a list of lists of value types. We decode this to a sum of
-products of value types. We omit the definition here. $\SOP$ is closed under
+products of value types. We omit the definition here. $\USOP$ is closed under
 value types, finite product types and finite sum types. For instance, we have
-$\Either_\SOP : \SOP \to \SOP \to \SOP$ together with $\Left_\SOP : \El_\SOP\,\vA
+$\Either_\SOP : \USOP \to \USOP \to \USOP$ together with $\Left_\SOP : \El_\SOP\,\vA
 \to \El_\SOP\,(\Either_\SOP\,\vA\,\vB)$, $\Right_\SOP : \El_\SOP\,\vB \to
@@
 inverses. We will overload $\mit{rep}$ as the forward conversion function with
-type $\vA \to \Rep$, and write $\mit{rep}^{-1}$ for its inverse. Now we define
-an isomorphic representation of $\El_\SOP\,\vA \to \Lift \vR$ as a product of
+type $\vA \to \El_\SOP\,\Rep$, and write $\mit{rep}^{-1}$ for its inverse. Now we define
+an isomorphic presentation of $\El_\SOP\,\vA \to \Lift \vR$ as a product of
 object-level functions:
@@
 We introduce a class for monads that support control flow joining.
-\[ \tyclass\,\Monad\,\vM \RA \MonadJoin\,\vM\,\where\,\join : \IsSOP\,\vA \RA \vM\,\vA \to \vM\,\vA \]
+\begin{alignat*}{3}
+  & \tyclass\,\Monad\,\vM \RA \MonadJoin\,\vM\,\where\\
+  & \ind \join : \IsSOP\,\vA \RA \vM\,\vA \to \vM\,\vA
+\end{alignat*}
 The most interesting instance is the ``base case'' for $\Gen$:
@@
 
-\todo{bigger example, non-tail-recursive list function with some effects, input, output, mentioned tail rec}
-
+\subsection{Code Example}
+
+We look at a slightly larger code example. We define annotated binary trees as
+\[ \data \Tree\,\vA := \Leaf\,|\,\Node\,\vA\,(\Tree\,\vA)\,(\Tree\,\vA). \]
+We write $\msf{fail : M\,A}$ for returning $\Nothing$ in any monad transformer
+stack that contains $\MaybeT$. We define a function which traverses a tree and
+replaces values with values taken from a list. If the tree contains $0$, we
+throw an error. If we run out of list elements, we leave values unchanged.
+\begin{alignat*}{4}
+  &  \letrec &&\vf : \Tree\,\Nat \to \StateT\,(\List\,\Nat)\,(\MaybeT\,\Identity)\,(\Tree\,\Nat) \\
+  &          &&\vf\,\vt := \spl \big(\mdown \dlr \case \vt\,\of\\
+  &          &&\ind \Leaf \to \return \qt{\Leaf} \\
+  &          &&\ind \Node\,\vn\,\vl\,\vr \to \mdo\\
+  &          &&\ind \ind \case \qt{\spl \vn == 0}\,\of\,\True \hspace{0.2em}\to \msf{fail}\\
+  &          &&\ind \ind \ind \hspace{6.7em}             \False \to \return () \\
+  &          &&\ind \ind \vns \fro \get \\
+  &          &&\ind \ind \vn\hspace{0.35em}\fro\join \dlr \case \vns\,\of\,\Nil \hspace{2.65em}\to \return \vn \\
+  &          &&\ind \ind \ind \hspace{8.2em}\Cons\,\vn\,\vns \to \mdo \{\msf{put}\,\vns; \return \vn\} \\
+  &          &&\ind \ind \vl\hspace{0.1em} \fro \mup\,\qt{\vf\,\spl \vl} \\
+  &          &&\ind \ind \vr \fro \mup\,\qt{\vf\,\spl \vr} \\
+  &          &&\ind \ind \return \qt{\Node\,\spl \vn\,\spl \vl\,\spl \vr}\big)
+\end{alignat*}
+Here, all of the $\case$ matches are done on object-level values, so they are
+all desugared to applications of $\msf{split}$.
+\begin{itemize}
+\item Note the $\join$: without it, the recursive $\Node$ traversal code would be inlined in both case branches.
+\item We do not use $\join$ when checking $\qt{\spl n == 0}$. Here, it
+      happens to be superfluous, since the $\msf{fail}$ ``destroys'' all subsequent
+      code generation in the branch, by short-circuiting at the meta-level.
+\item Also, we can use $\msf{put}$ instead of the ``stricter'' $\msf{put'}$ because
+      the state modification immediately gets sequenced by the enclosing join point.
+\item To make object-level recursive calls, we just need to wrap them in $\mup$.
+\end{itemize}
+Here, we intentionally tuned the definition for a nice-looking staging output.
+However, we could also just default to ``safe'' choices everywhere: we could use
+a joint point in every $\mbf{case}$ with two or more branches, and use the
+strict state modifications everywhere. This would only add a small amount of
+administrative noise that can be easily cleaned up by the downstream compiler.
+
+Let us look at the staging output. We omit newtype wrappers and use nested
+pattern matching on pairs, but otherwise this is exactly the code that we get in
+our Agda implementation.
+\begin{alignat*}{4}
+  &  \letrec &&\vf : \Tree\,\Nat \to \StateT\,(\List\,\Nat)\,(\MaybeT\,\Identity)\,(\Tree\,\Nat) \\
+  &          &&\vf := \lam \vt\,\vns.\, \case\,\vt\,\of\\
+  &          &&\ind \Leaf \to \Just\,(\Leaf,\,\vns) \\
+  &          &&\ind \Node\,\vn\,\vl\,\vr \to \case (\vn == 0)\,\of\\
+  &          &&\ind \ind \True \hspace{0.25em}\to \Nothing \\
+  &          &&\ind \ind \False \to \\
+  &          &&\ind \ind \ind \letdef \msf{joinNothing} := \lam \_.\,\Nothing;\\
+  &          &&\ind \ind \ind \letdef \msf{joinJust} := \lam \vn\,\vns.\,\case\,\vf\,\vl\,\vns\,\of\\
+  &          &&\ind \ind \ind \ind \ind \ind \Nothing\hspace{0.9em} \to \Nothing\\
+  &          &&\ind \ind \ind \ind \ind \ind \Just\,(\vr,\,\vns) \to \case\,\vf\,\vr\,\vns\,\of\\
+  &          &&\ind \ind \ind \ind \ind \ind \ind \Nothing\hspace{.95em} \to \Nothing\\
+  &          &&\ind \ind \ind \ind \ind \ind \ind \Just\,(\vr,\,\vns) \to \Just\,(\Node\,\vn\,\vl\,\vr,\,\vns);\\
+  &          &&\ind \ind \ind \case \vns\,\of\\
+  &          &&\ind \ind \ind \ind \Nil \hspace{2.65em}\to \msf{joinJust}\,\vn\,\vns\\
+  &          &&\ind \ind \ind \ind \Cons\,\vn\,\vns \to \msf{joinJust}\,\vn\,\vns
+\end{alignat*}
+The only flaw in this code is the $\msf{joinNothing}$, which is never called,
+becase we joined an action that never throws an error. But this is also
+very easy to clean up after staging.
 \subsection{Discussion}
@@
 \item In monadic code, object-level data constructors are only ever created by
-      $\mdown$, and switching on object-level data is only created by $\mit{split}$
+      $\mdown$, and matching on object-level data is only created by $\mit{split}$
       and $\mup$. Monadic operations are fully fused, and all function calls can be
@@
 some syntactic noise and requires extra attention from programmers.  We believe
-that the code noise can be mitigated very effectively in a native CFTT
-implementation. Kovács \cite{TODO} demonstrated in a prototype elaborator that
-almost all quotes and splices are unambiguously inferable in 2LTT programs, if
-we require that stages of let-definitions are always specified (as we do
-here). Moreover, $\mup$ and $\mdown$ should be also effectively inferable, using
-bidirectional elaboration. With stage inference and binding-time improvement
-inference, monadic code in CFTT would look only modestly more complicated than
-in Haskell.
+that the noise can be mitigated very effectively in a native CFTT
+implementation. It was demonstrated in a 2LTT implementation in
+\cite{staged-demo} that almost all quotes and splices are unambiguously
+inferable, if we require that stages of let-definitions are always specified (as
+we do here). Moreover, $\mup$ and $\mdown$ should be also effectively inferable,
+using bidirectional elaboration. With such inference, monadic code in CFTT would
+look only modestly more complicated than in Haskell.
 
@@
       not track them in the library, and we do not get guaranteed closure-freedom
-      from GHC. We do get high quality code though, typically.
+      from GHC.
 \item In Agda, we postulate all types and terms of the object theory in a
@@
 \item For sums-of-products in Haskell, we make heavy use of \emph{singleton
-  types} \cite{TODO} to emulate dependent types. This adds significant
+  types} \cite{DBLP:conf/haskell/EisenbergW12} to emulate dependent types. This adds significant
   noise. Also, in $\IsSOP$ instances we can only define the conversion functions
@@
 
-\section{Stream Fusion}
+\section{Stream Fusion}\label{sec:stream-fusion}
 
@@
 from declarative definitions involving streams of values, where intermediate
-data structures are eliminated. Stream fusion can be broadly grouped into \emph{push}
-fusion, which is based on Church-encodings of inductive lists, and $\emph{pull}$
-fusion, which is based on Church-encodings of coinductive lists \cite{TODO}. The
-two styles have different trade-offs, and in practical programming it is a good
-idea to support both, but in this section we focus on pull streams.
+data structures are eliminated. Stream fusion can be broadly grouped into
+\emph{push} fusion, which is based on Church-encodings of inductive lists, and
+$\emph{pull}$ fusion, which is based on Church-encodings of coinductive lists;
+see e.g.\ \cite{DBLP:conf/ifl/HinzeHJ10}. The two styles have different
+trade-offs, and in practical programming it is a good idea to support both, but
+in this section we focus on pull streams.
 
@@
 required to be a sum-of-products of value types by the $\IsSumVS\,\vS$
-constraint. Borrowing terminology from Coutts \cite{TODO}, we call each product-of-value-types in the
-state a \emph{state shape}.
+constraint. Borrowing terminology from Coutts \cite{DBLP:phd/ethos/Coutts11}, we
+call each product in the state a \emph{state shape}.
 
@@
 
-In $\mit{repeat}$, the state is the unit type, while in $({\appull})$, we take
+In $\mit{repeat}$, the state is the unit type, while in $({\appull})$ we take
 the product of states, and also add another $\Maybe\,\vA$ that is used to buffer
@@
 
-How do we generate object code from streams? The $\vS$ state is given as a
-finite sums of products, but the sums and the products are on the meta
-level, so we cannot directly use $\vS$ in object code. Similarly as in the the
-treatment of join points, we tabulate the $\vS \to \Gen\,\Step\,\vS\,\vA$
-transition function to a product of functions. However, these functions need to
-be \emph{mutually recursive}, since it is possible to transition from any state
-to any other state, and each such transition is represented as a function call.
+How do we generate object code from streams? The state $\vS$ is given as a
+finite sums of products, but the sums and the products are on the meta level, so
+we cannot directly use $\vS$ in object code. Similarly as in the the treatment
+of join points, we tabulate the $\vS \to \Gen\,\Step\,\vS\,\vA$ transition
+function to a product of functions. However, these functions need to be
+\emph{mutually recursive}, since it is possible to transition from any state to
+any other state, and each such transition is represented as a function call.
 This problem of generating well-typed mutual blocks was addressed by Yallop and
-Kiselyov in \cite{TODO}. Unlike ibid., which used control effects and mutable
-references in MetaOCaml, we present a solution that does not use side effects
-in the metalanguage.
+Kiselyov in \cite{DBLP:conf/pepm/YallopK19}. Unlike ibid., which used control
+effects and mutable references in MetaOCaml, we present a solution that does not
+use side effects in the metalanguage.
 
@@
 library, where $\foldl$ for lists is also defined from $\foldr$, to enable
-push-based fusion, but where a substantial \emph{arity analysis} is required in
-the compiler to eliminate the intermediate closures \cite{TODO}.
-
-\subsection{$\mbf{concatMap}$ for Streams}
+push-based fusion, but where a substantial \emph{arity analysis} is used in
+the compiler to (incompletely) eliminate the intermediate closures
+\cite{DBLP:conf/sfp/Breitner14}.
+
+\subsection{$\mbf{concatMap}$ for Streams}\label{sec:concatmap-for-streams}
 
@@
 streams. Hence, we shall have the following function instead:
-\[ \concatMap : \IsSOP\,\vA \RA (\vA \to \Pull\,\vB) \to \Pull\,\vA \to \Pull\,\vB, \]
-The idea is the following: if $\USOP$ is closed under dependent sum types, we
-can directly define this $\concatMap$, by taking the appropriate dependent sum
-of the $\vA \to \USOP$ family of machine states, which we extract from the $\vA
-\to \Pull\,\vB$ function. Let us write $\Sigma\,\vA\,\vB : \MTy$ for dependent
-sums, for $\vA : \MTy$ and $\vB : \vA \to \MTy$, and reuse $({,})$ for pairing.
-We also use the following field projection functions: $\mit{projS} : \Pull\,\vA
-\to \MTy$, $\mit{projSeed} : (\vas : \Pull\,\vA) \to \mit{projS}\,\vas$, and
+\[ \concatMap : \IsSOP\,\vA \RA (\vA \to \Pull\,\vB) \to \Pull\,\vA \to \Pull\,\vB \]
+The idea is the following: if $\USOP$ is closed $\Sigma$-types, we can directly
+define this $\concatMap$, by taking the appropriate dependent sum of the $\vA
+\to \USOP$ family of machine states, which we extract from the $\vA \to
+\Pull\,\vB$ function. Let us write $\Sigma\,\vA\,\vB : \MTy$ for dependent sums,
+for $\vA : \MTy$ and $\vB : \vA \to \MTy$, and reuse $({,})$ for pairing.  We
+also use the following field projection functions: $\mit{projS} : \Pull\,\vA \to
+\MTy$, $\mit{projSeed} : (\vas : \Pull\,\vA) \to \mit{projS}\,\vas$, and
 $\mit{projStep} : (\vas : \Pull\,\vA) \to \mit{projS}\,\vas \to
@@
 Above, $\{\va : \vA\} \to \IsSOP\,(\vB\,\va)$ is a universally quantified
-instance constraint; this is available in Agda and, less generally, in Haskell
-\cite{TODO}. The definition of $\concatMap$ is as follows.
+instance constraint; this can be also represented in Agda. The definition of
+$\concatMap$ is as follows.
 \begin{alignat*}{3}
@@
 $(\vf\,\va)$ stream until it stops. In the inner loop, the machine state type
-depends on the $\va : \vA$ value, hence the need for $\Sigma$.
-
-How do we get $\IsSOP$ for $\Sigma$? The key observations are:
+depends on the $\va : \vA$ value, hence the need for $\Sigma$. How do we get
+$\IsSOP$ for $\Sigma$? The key observations are:
 \begin{itemize}
@@
 Hence, we expect that during staging, every $\vf : \Lift\,\vA \to \VTy$ function
-has to be constant. This is actually true in the staging semantics of CFTT. In
-the semantics, all metafunctions are stable under object-level parallel
+has to be constant. This is actually true in the staging semantics of
+CFTT. There, all metafunctions are stable under object-level parallel
 substitutions. Also, object types are untouched by substitution. Hence, a
-straightforward unfolding of the definitions in the staging semantics validates
-the constancy of $\vf$.
-
-Generally speaking, every function whose domain is a product of object types and
-whose codomain is a constant presheaf in the semantics, is a constant function
-in the the semantics. We may call these constancy statements \emph{generativity
-axioms}, since they reflect the inability of metaprograms to inspect terms, and
-``generativity'' often refers to this property in the staging literature
-\cite{TODO}.
+straightforward unfolding of semantics definitions validates the constancy of
+$\vf$.
+
+Generally speaking, in the semantics, every function whose domain is a product
+of object types and whose codomain is a constant presheaf, is a constant
+function. We may call these constancy statements \emph{generativity axioms},
+since they reflect the inability of metaprograms to inspect terms, and
+``generativity'' often refers to this property in the staged compilation
+literature.
 
@@
 \\\\
-We remark that there is no risk of staging getting \emph{stuck} on this axiom,
-because propositional equality proofs get \emph{erased} in the staging semantics,
-as we noted in Section \ref{TODO}.
-
-We also remark that although generativity is inconsistent with inspecting the
-structure of object terms, it is consistent with inspecting the structure of
-object types.
+We remark that there is no risk of staging getting stuck on this axiom, because
+propositional equality proofs get erased, as we described in Section
+\ref{sec:semantics-of-staging}.
+
+%% We also remark that although generativity is inconsistent with inspecting the
+%% structure of object terms, it is consistent with inspecting the structure of
+%% object \emph{types}.
 
@@
 to get arbitrary inhabitants with which we can call $\Elprod\,A \to \USOP$
-functions. Conveniently, every object type is inhabited by infinite loops.
+functions.
 
@@
 non-dependent products in $\USOP$, except that we have to get rid of the type
-dependency by instantiating $\vB$ with looping programs.
+dependency by instantiating $\vB$ with arbitrary programs.
 
@@
 define the $\IsSOP$ instance, when we have to prove that encoding via $\rep$
-is an isomorphism. Concretely, assuming $\IsSOP\,A$ and $\{\va : \vA\} \to
-\IsSOP\,(\vB\,\va)\}$, we define $\Rep$ for $\Sigma\,\vA\,\vB$ as follows:
+is an isomorphism. Concretely, assuming $\IsSOP\,\vA$ and $\{\va : \vA\} \to
+\IsSOP\,(\vB\,\va)$, we define $\Rep$ and $\msf{rep}$ for $\Sigma\,\vA\,\vB$ as follows:
 \[ \Rep\,\{\Sigma\,\vA\,\vB\} = \Sigma\,(\Rep\,\{\vA\})\,(\lam \vx.\,\Rep\,\{\vB\,(\rep^{-1}\,\vx)\}) \]
-Then, the definition of encoding for $\Sigma$ is only well-typed up to the fact
-that $\rep^{-1} \circ \rep = \id$:
-\[  \rep\{\Sigma\,\vA\,\vB\}\,(\va,\,\vb) = (\rep\,\{\vA\}\,\va,\,\rep\,\{\vB\,(\rep^{-1}\,(\rep\,\vx))\}\,\vb) \]
-This is, in fact, our reason for including the isomorphism equations as well in
+\[  \rep\,\{\Sigma\,\vA\,\vB\}\,(\va,\,\vb) = (\rep\,\{\vA\}\,\va,\,\rep\,\{\vB\,(\rep^{-1}\,(\rep\,\vx))\}\,\vb) \]
+Note that $\rep$ is only well-typed up to the fact that $\rep^{-1} \circ \rep =
+\id$; the Agda definition contains an additional transport that we omit here.
+This is, in fact, our reason for including the isomorphism equations in
 $\IsSOP$. This, in turn, necessitates using SOP instead of finite sums. We can
@@
 
-We do not detail the full definition of $\Sigma_\SOP$ here. The reader may refer
-to the SOP module in the Agda implementation, which is altogether 260 lines with
+We skip the full definition of $\Sigma_\SOP$ here. The reader may refer
+to the $\SOP$ module in the Agda implementation, which is altogether 260 lines with
 all $\IsSOP$ instances.
@@
 \end{alignat*}
-Now, this operation should be avoided when possible, since it has two states and
-can contribute to a blow-up of state size. For example, $\concatMap$ over
-$\single$ introduces at least a doubling of the number of state shapes. Although
+Now, we should use this operations with care, since it has two states and can
+contribute to a code size blow-up. For example, $\concatMap$ over $\single$
+introduces at least a doubling of the number of state shapes. Although
 let-insertion and case splitting could be derived as $\concatMap$ over
 $\single$, to avoid the size explosion we give more specialized definitions
-instead. First, we define a helper function for binding for a single $\IsSOP$ value.
+instead. First, we define a helper function for binding a single $\IsSOP$ value.
 \begin{alignat*}{3}
@@
 from $\mit{take}\,\vy\,(\mit{countFrom}\,x)$; we sum these to get three for the
-$\mit{case_\Pull}$, then we get an extra shape from $\mit{forEach}$ which
+$\mit{case_\Pull}$, then finally we get an extra shape from $\mit{forEach}$ which
 introduces an additional $\Maybe$.
@@
 
-Our stream library has a fairly strong support for programming in a monadic
-style, even though $\Pull$ is not literally a monad. We can bind object values
-with $\concatMap$, and we can also do let-insertion and case splitting for
-them. We also get guaranteed well-typing, closure-freedom, and arbitrary mixing
-of zipping and $\concatMap$.
+Our stream library has a strong support for programming in a monadic style, even
+though $\Pull$ is not literally a monad. We can bind object values with
+$\concatMap$, and we can also do let-insertion and case splitting for them. We
+also get guaranteed well-typing, closure-freedom, and arbitrary mixing of
+zipping and $\concatMap$.
 
@@
 compilation, intensional analysis (i.e. the ability to analyze object code) has
-been viewed as a desirable feature that increases the expressive power of the
+been viewed as a desirable feature that increases the expressive power of a
 system. To our knowledge, our work is the first one that exploits the
@@
 parametricity in type theories, where the inability to analyze types has a
-payoff in the form of ``free theorems'' \cite{TODO}.
+payoff in the form of ``free theorems'' \cite{wadler89theoremsforfree}.
 
@@
 In Agda, to avoid computation getting stuck on the generativity axiom, we use
-the \texttt{primTrustMe} built-in \cite{TODO} to automatically erase the axiom
+the \texttt{primTrustMe} built-in \cite{agdadocs} to automatically erase the axiom
 when the sides of the equation are definitionally equal. Otherwise the
@@
 
-%% \interlinepenalty=10000
+\subsubsection*{Two-level type theory.} Two-level lambda calculi were first developed in
+the context of abstract interpretation and partial evaluation
+\cite{DBLP:phd/ethos/Nielson84a,DBLP:books/daglib/0071307}.  This line of
+research is characterized by simple types, having the same language features at
+different stages, and an emphasis on \emph{binding time analysis},
+i.e.\ automatically inferring stage annotations as part of a pipeline for
+partial evaluation and program optimization.
+
+Later and independently, the same notion of level appeared in homotopy type
+theory, first in Voevodsky's Homotopy Type System \cite{hts}, and subsequently developed as
+2LTT in \cite{twolevel}. Here, dependent types are essential, and we have different
+theories at the two stages. The application of 2LTT to staged compilation was
+developed in \cite{staged2ltt}. Binding-time analysis has not been developed in this
+setting; 2LTT-s have been meant to be used as surface languages to directly work
+in.
+
+We build on both traditions. CFTT itself is a 2LTT and makes heavy use of
+dependent type theory. However, we are mostly concerned with code generation and
+optimization and borrow concepts from the partial evaluation literature.
+
+\subsubsection*{Tracking function arities and closures in types} Downen et al.'s
+$\mathcal{IL}$ intermediate language has similar motivations as our object
+language \cite{DBLP:journals/pacmpl/DownenAJE20}. In both systems, function
+types are distinguished from closure types, enjoy universal $\eta$-conversion
+and have an explicit calling arity. In fact, our object language can be almost
+viewed as a small simply-typed fragment of $\mathcal{IL}$, with only
+$\mbf{letrec}$ missing from $\mathcal{IL}$.
+
+\subsubsection*{Sums-of-products} $\SOP$ was proposed for generic Haskell programming
+by De Vries and Löh \cite{DBLP:conf/icfp/VriesL14}. Our own $\SOP$ implementation in Haskell is
+mostly the same as in ibid. More specifically, the $\SOP$-of-object-code
+representation appeared in typed Template Haskell in \cite{DBLP:conf/haskell/PickeringLW20}.
+
+\subsubsection*{CPS and binding-time improvement}
+Our $\Gen$ monad can be viewed as a variation on the continuation-passing-style
+that has been often used in staged programming. Flanagan et al.'s ANF
+translation algorithm uses let-insertion in CPS that works the same way as our
+$\gen$ \cite{DBLP:conf/pldi/FlanaganSDF93}. Kiselyov et al.\ likewise uses
+essentially the same CPS, in MetaOCaml
+\cite{DBLP:conf/popl/KiselyovBPS17}. Jones et al.\ discuss CPS and binding-time
+improvement in partial evaluation \cite{partial-evaluation}. $\eta$-expansion
+for object-level finite sums is known as ``the trick'' \cite{eta-expansion-trick}.
+
+\subsubsection*{Join points}
+The use join points in GHC's core language is partly motivated by avoiding code
+duplication in case-of-case transformations
+\cite{DBLP:conf/pldi/MaurerDAJ17}. In our monad library, case-of-case is
+implicitly and eagerly computed during staging, and we similarly use join points
+to avoid code duplication.
+
+\subsubsection*{Stream fusion}
+The staged stream fusion library $\mathit{strymonas}$ by Kiselyov et al.\ is the
+primary prior art \cite{DBLP:conf/popl/KiselyovBPS17}. Here, streams are represented as a sum type of
+push and pull representations, allowing both zipping and
+$\concatMap$-ing. However, fusion is not guaranteed for all combinations;
+zipping two $\concatMap$-s reifies one of the streams in a runtime closure. In
+newer versions of the library, fusion is reported to be complete and guaranteed
+\cite{DBLP:conf/pepm/0001K24}, however, an exposition of this has not yet been published. Another
+notable difference is that $\mathit{strymonas}$ relies on mutable references in
+the object language, while we do not.
+
+\emph{Machine fusion} \cite{DBLP:conf/ppdp/RobinsonL17} supports splitting
+streams to multiple streams while avoiding duplicated computation; this is not
+possible in $\mathit{strymonas}$ or our library. Machine fusion also supports
+$\concatMap$ and zipping. However, its state machine representation is
+significantly more complicated than ours, and its fusion algorithm is not
+guaranteed to succeed on arbitrary stream programs.
+
+Coutts developed pull stream fusion in depth in
+\cite{DBLP:phd/ethos/Coutts11}. We borrowed the basic design and the basic
+stream combinator definitions from there. This account is also close to existing
+stream implementations in Haskell. Ibid.\ characterizes fusibility in a
+non-staged setting, in terms of ``good consumers'' and ``good producers''.  This
+scheme does not cover $\concatMap$.
+
+\section{Conclusions and Future Work}
+
+We believe that a programming language in the style of this paper would be
+highly useful, especially in high-performance functional programming and
+domain-specific programming. In particular, the pipeline for using and compiling
+monads could look like the following.
+\begin{itemize}
+\item Users write definitions in a style similar to Haskell, without quotes,
+      splices, $\mup$-s and $\mdown$-s, only marking stages in type annotations and
+      in let-definitions, with $:=$ and $=$.  Storing monadic actions at runtime
+      requires an explicit boxing operation that yields a closure type.
+\item Type- and stage-directed elaboration adds the missing operations, and also desugars
+      case splitting using $\join$ and the underlying splitting function.
+\item In the downstream compiler, fast \& conservative optimization passes are
+      sufficient, since abstraction overheads are eliminated by staging. Eliminating
+      dead code and unused arguments would be important, since we have not yet addressed
+      this in metaprogramming.
+\item A systematic way to de-duplicate code would be also needed, probably already
+      during staging.
+\end{itemize}
+
+Going a bit further, we believe that a 2LTT-based
+language could be ergonomic enough for programming in general, getting us
+valuable control over code generation for a modest amount of extra complexity.
+
+We also think that rebuilding known abstractions in staged programming is
+valuable because it provides a \emph{semantic explanation} of how abstractions
+can be compiled, and provides insights that could be reused even in non-staged
+settings. For instance, in this paper we demonstrated that compiling monadic code to
+efficient code can be done by using \emph{the same monads} in the metalanguage,
+extended with code generation as an effect.
+
+Following this line of thought, it could be interesting to also adapt to 2LTT
+the style of \emph{binding-time analysis} that is well-known in partial
+evaluation. For example, we might do program optimization in the following way.
+First, start with monadic code in a non-staged language. Second, try to infer
+stages, inventing quotes, splices, $\mup$-s and $\mdown$-s, thereby translating
+definitions to a 2LTT. Third, we perform staging and proceed from there. This would
+be more fragile than unambiguous stage inference, but we would still benefit from
+shifting a lot of machinery into staging (which is deterministic and efficient).
+
+In this paper we only briefly looked at two applications. Both monad
+transformers and streams could be further developed, and many other programming
+abstraction could be revisited in the staged setting.
+\begin{itemize}
+\item
+     We did not discuss the issue of \emph{tail calls} in monadic code. For
+     example, a tail call in $\Maybe$ should not case split on the result, while
+     in our current library, making a call with $\mup$ always does so. We did
+     develop an abstraction for tail calls in the Agda and Haskell
+     implementations, but omitted it here partly for lack of space, and partly
+     because we have not yet explored much of the design space.
+
+\item In a practical stream library, it would be useful to further try to
+     minimize the size of the machine state. In the Agda and Haskell versions, we
+     additionally track whether streams can $\Skip$, to provide more efficient
+     zipping for non-skipping ``synchronous'' streams. However, this could be
+     refined in many ways, and we could also try to represent the full transition
+     graph in an observable way, thereby enabling merging or deleting states.  We
+     could also support more optimization for restricted domains such as integer
+     ranges.
+
+\item It seems promising to find synergies between push streams, pull streams
+     and monad transformers. It seems that pull streams could be generalized from
+     $\Gen$ to different monads, in the representation of transitions and initial
+     states. This would allow putting a $\Pull$ \emph{on the top} of a monad
+     transformer stack. On the other hand, push streams form a proper monad, but
+     they cannot be used to ``transform'' other monads, so they can be placed
+     at the \emph{bottom} of a transformer stack.
+\item We could try to lean more heavily on datatype-generic programming and
+     generalize abstractions to arbitrary object types. For example, push and pull
+     streams can be generalized as inductive and coinductive Church-codings of
+     object types.
+\end{itemize}
+
+%% \begin{itemize}
+%%   \item Start with monadic code in a non-staged language.
+%%   \item Try to infer stages, inventing quotes, splices, $\mup$-s and $\mdown$-s,
+%%         thereby translating definitions to 2LTT.
+%%   \item Stage the obtained definitions.
+%% \end{itemize}
+%% In this approach, much complexity is factored out to staging
+
+\interlinepenalty=10000
 \bibliography{references}

### 2024-02-26 newpaper/paper.tex
edits

@@
 This is a third-order and second-rank polymorphic function in disguise, because
-its monad dictionary argument contains the polymorphic second-order $(\bind)$
-method. Compiling $\msf{mapM}$ efficiently relies on inlining the instance
+its monad dictionary argument contains the polymorphic second-order method
+$(\bind)$.  Compiling $\msf{mapM}$ efficiently relies on inlining the instance
 dictionary, then inlining the methods contained there, and also inlining the
@@
 
-GHC's optimization efforts are respectable, and it has gotten quite good over
+GHC's optimization efforts are respectable, and it has gotten quite adept over
 its long history of development. However, there is no strong guarantee that
@@
 \vf\,(\vg\,\vx)$ in the base libraries, instead of as $(.)\,\vf\,\vg\,\vx =
-\vf\,(\vg\,\vx)$, to get better inlining behavior \cite{ghcbase}. It remains a
-common practice in high-performance Haskell programming to manually review GHC's
-optimized code output.
+\vf\,(\vg\,\vx)$, to get better inlining behavior \cite{ghcbase}. Tellingly, it
+is common practice in high-performance Haskell programming to visually review
+GHC's optimized code output.
 
@@
 \item Type classes in Haskell use dictionary-passing, which relies on closures for function methods \cite{DBLP:conf/popl/WadlerB89}.
-\item Functors and first-class modules in OCaml and other ML-s rely on closures.
+\item Functors and first-class modules in OCaml \cite{leroy2023ocaml} and other ML-s rely on closures.
 \end{itemize}
@@
 compatible with closures as well, and it can support two separate type formers
-for closure-based and non-closure-based (``static'') functions. Having both of
+for closure-based and non-closure-based ``static'' functions. Having both of
 these would be desirable in a practical system. In the current work we focus on
@@
   CFTT, where the object level is first-order simply-typed and the meta level is
-  dependently typed. The object language supports an operational semantics without
-  runtime closures, and can be compiled in a way such that all function calls
-  are statically known. We provide a supplementary Agda formalization of the
-  operational semantics.
+  dependently typed. The object language supports an operational semantics
+  without runtime closures, and can be compiled with only statically known
+  function calls. We provide a supplementary Agda formalization of the
+  operational semantics of the object language.
 \item In Section \ref{sec:monad-transformers} we build a monad transformer library. We believe
@@
   reused in the meta level of CFTT. We also develop let-insertion, join points
-  and case switching on object-level data, in the generality of monad transformers.
+  and case switching on object-level data.
 \item In Section \ref{sec:stream-fusion} we build a pull-based stream fusion library. Here,
-  we demonstrate essential usage of dependent types, for providing guaranteed
+  we demonstrate essential usage of dependent types, in providing guaranteed
   fusion for arbitrary combinations of $\concatMap$ and $\mit{zip}$. We use a
@@
 
-In the following we give an overview of CFTT features. We focus on informal
-explanations and examples here. We first review the meta-level language, then
-the object-level one, and finally the staging operations which bridge between
-the two.
+In the following we give an overview of CFTT features. We first review the
+meta-level language, then the object-level one, and finally the staging
+operations which bridge between the two.
 
@@
 ``metaprogram'' for inhabitants of metatypes. $\MTy$ supports dependent
-functions, products and indexed inductive types \cite{inductivefamilies}.
-
-Formally, $\MTy$ is additionally indexed by universes levels, and we have
-$\MTy_i : \MTy_{i+1}$. This is orthogonal to the levels of staging. However,
-universe levels add noise and they are not too relevant to the current paper, so
-we will omit them.
+functions, $\Sigma$-types and indexed inductive types \cite{inductivefamilies}.
+
+Formally, $\MTy$ is additionally indexed by universes levels (orthogonally to
+staging), and we have $\MTy_i : \MTy_{i+1}$. However, universe levels add noise
+and they are not too relevant to the current paper, so we will omit them.
 
@@
 parameterized algebraic data types, where parameters can have arbitrary types,
-but all constructor field types must be in $\VTy$.
-
-Since $\VTy$ is a sub-universe of $\Ty$, we have that when $\vA : \VTy$ then also
-$\vA : \Ty$. Formally, this is specified as an explicit embedding operation, but
-we will use implicit subtyping for convenience.
+but all constructor field types must be in $\VTy$. Since $\VTy$ is a
+sub-universe of $\Ty$, we have that when $\vA : \VTy$ then also $\vA :
+\Ty$. Formally, this is specified as an explicit embedding operation, but we
+will use implicit subtyping for convenience.
 
@@
 \end{alignat*}
-Every recursive definition must be introduced with $\letrec$. The general syntax
-is $\letrec \vx : \vA := \vt; \vu$, where the $\vA$ type annotation can be
+Recursive definitions are introduced with $\letrec$. The general syntax is
+$\letrec \vx : \vA := \vt; \vu$, where the $\vA$ type annotation can be
 omitted. $\mbf{letrec}$ can be only used to define computations, not values
@@
 inhabitants of any type, and the type of the $\mbf{let}$ body can be also
-arbitrary. Additionally, the right hand sides of $\case$ branches can also have
-arbitrary types. So the following is well-formed:
+arbitrary. Additionally, the right hand sides of $\mbf{case}$ branches can also
+have arbitrary types. So the following is well-formed:
 \begin{alignat*}{3}
@@
 from runtime semantics. Object programs are embedded in CFTT, which is a
-dependently typed language, so we may need to decide definitional equality of
+dependently typed language, so sometimes we need to decide definitional equality of
 object programs during type checking. The setup is simple: we have no $\beta$ or
@@
 We took inspiration from call-by-push-value (CBPV) \cite{DBLP:conf/tlca/Levy99},
-and there are similarities, but there are also significant differences. Both
-systems have a value-computation distinction, with call-by-name computations and
-call-by-value values. However, our object theory supports variable binding at
-arbitrary types while CBPV only supports value variables. In CBPV, a
-let-definition for a function is only possible by first packing it up as a
-closure value (or ``thunk''), which clearly does not suit our applications.
-Investigating the relation between CBPV and our object language could be future work.
+and there are similarities to our object language, but there are also
+significant differences. Both systems have a value-computation distinction, with
+call-by-name computations and call-by-value values. However, our object theory
+supports variable binding at arbitrary types while CBPV only supports value
+variables. In CBPV, a let-definition for a function is only possible by first
+packing it up as a closure value (or ``thunk''), which clearly does not suit our
+applications.  Investigating the relation between CBPV and our object language
+could be future work.
 
@@
 is only computed at runtime when $x$ evaluates to $\True$. In many situations,
-staging can be used instead of laziness to implement short-circuiting, and with
-generally better runtime performance, avoiding the overhead of thunking. Consider
-the $\map$ function now:
+staging can be used instead of laziness to implement short-circuiting, and often
+with better runtime performance, avoiding the overhead of thunking. Consider the
+$\map$ function now:
 \begin{alignat*}{3}
@@
 semantics from \cite{staged2ltt} and only mention additions and differences. The
-reader may refer to ibid.\ for technical background.
+reader may refer to ibid.\ for technical details.
 
@@
 Here we overload Haskell-style product type notation, both for types and the
-pair constructor, at both levels (products are definable as a value type at the
-object level). There is a problem with this conversion though: $\mup$ uses $\vx
-: \Up(\vA,\,\vB)$ twice, which can increase code size and duplicate runtime
-computations. For example, $\mdown\,({\mup\,\ql \vf\,\vx \qr})$ is staged to
-$\ql (\fst\,(\vf\,\vx),\,\snd\,(\vf\,\vx)) \qr$. It would be safer to first
-let-bind an expression with type $\Up(\vA,\,\vB)$, and then only use projections
-of the newly bound variable. This is called \emph{let-insertion} in staged
-compilation. But it is impossible to use let-insertion in $\mup$ because the
-return type is in $\MTy$, and we cannot introduce object binders in meta-level
-code.  Fortunately, there is a principled solution.
+pair constructor, at both levels. There is a problem with this conversion
+though: $\mup$ uses $\vx : \Up(\vA,\,\vB)$ twice, which can increase code size
+and duplicate runtime computations. For example, $\mdown\,({\mup\,\ql \vf\,\vx
+  \qr})$ is staged to $\ql (\fst\,(\vf\,\vx),\,\snd\,(\vf\,\vx)) \qr$. It would
+be safer to first let-bind an expression with type $\Up(\vA,\,\vB)$, and then
+only use projections of the newly bound variable. This is called
+\emph{let-insertion} in staged compilation. But it is impossible to use
+let-insertion in $\mup$ because the return type is in $\MTy$, and we cannot
+introduce object binders in meta-level code.  Fortunately, there is a principled
+solution.
 
@@
 
-Our solution is to use a monad which extends $\MTy$ with the ability to
+The solution is to use a monad which extends $\MTy$ with the ability to
 freely generate object-level code and introduce object binders.
 \begin{alignat*}{3}
-  & \newtype \Gen\,\vA = \Gen\,\{\unGen : \{\vR : \Ty\} \to (\vA \to \Up \vR) \to \Up \vR\}
+  & \newtype \Gen\,(\vA : \MTy) = \Gen\,\{\unGen : \{\vR : \Ty\} \to (\vA \to \Up \vR) \to \Up \vR\}
 \end{alignat*}
@@
   &\ind \return \va        &&= \Gen \dlr \lam \vk.\,\vk\,\va\\
-  &\ind \mit{ga} \bind \vf &&= \Gen \dlr \lam \vk.\,\unGen\,\mit{ga}\,(\lam \va.\,(\vf\,\va\,\vk))
+  &\ind \mit{ga} \bind \vf &&= \Gen \dlr \lam \vk.\,\unGen\,\mit{ga}\,(\lam \va.\,\unGen\,(\vf\,\va)\,\vk))
 \end{alignat*}
@@
 \begin{alignat*}{6}
-  &\hspace{-2em}\mathrlap{\mup : \Up (\vA,\,\vB) \to \Gen\,(\Up \vA,\,\Up \vB)            \hspace{5.45em} \mdown : \Gen\,(\Up \vA,\, \Up \vB) \to \Up(\vA,\,\vB)} \\
-  &\hspace{-2em}\mup\,\vx = \mdo &&\vx \fro \gen\,\vx                                    && \ind\ind \mdown\,x = \mdo &&(\va,\,\vb) \fro x\\
+  &\hspace{-2em}\mathrlap{\mup : \Up (\vA,\,\vB) \to \Gen\,(\Up \vA,\,\Up \vB)            \hspace{5em} \mdown : \Gen\,(\Up \vA,\, \Up \vB) \to \Up(\vA,\,\vB)} \\
+  &\hspace{-2em}\mup\,\vx = \mdo &&\vx \fro \gen\,\vx                                    && \ind\ind \mdown\,\vx = \mdo &&(\va,\,\vb) \fro \vx\\
   &\hspace{-2em}               &&\return (\qt{\fst\,\spl \vx},\,\qt{\snd\,\spl \vx})     &&                           &&\return \qt{(\spl \va,\, \spl \vb)}
@@
 problematic, because we cannot directly compute metaprograms from object-level
-case splits. Fortunately, with a little bit more work, case splitting on object
-values is actually possible in any $\MonadGen$, on any value type.
+case splits. Fortunately, with a little bit more work, this is actually possible
+in any $\MonadGen$, for any value type.
 
@@
 
-There is a deficiency in our library so far. Assuming $\vb : \Bool$ as an object value, consider:
+There is a deficiency in our library so far. Assuming $\vb : \Up\Bool$, consider:
 \begin{alignat*}{5}
@@
 This removes code duplication by round-tripping through an object-level
-$\mbf{let}$. This solution is fairly good in a state monad, since $\mdown$ only
+$\mbf{let}$. This solution is fairly good in a state monad, where $\mdown$ only
 introduces a runtime pair constructor, and it is feasible to compile
@@
 Let us write $\Uprod = \List\,\VTy$ and $\Elprod : \Uprod \to \MTy$ for a
-universe of finite products of value types. Concretely in CFTT and the Agda
+universe of finite products of value types. Concretely, in CFTT and the Agda
 implementation, we assume the following:
@@
 \end{alignat*}
-Now, we should use this operations with care, since it has two states and can
+Now, we should use this operation with care, since it has two states and can
 contribute to a code size blow-up. For example, $\concatMap$ over $\single$
@@
 \subsubsection*{Tracking function arities and closures in types} Downen et al.'s
-$\mathcal{IL}$ intermediate language has similar motivations as our object
+intermediate language $\mathcal{IL}$ has similar motivations as our object
 language \cite{DBLP:journals/pacmpl/DownenAJE20}. In both systems, function
@@
 \subsubsection*{Sums-of-products} $\SOP$ was proposed for generic Haskell programming
-by De Vries and Löh \cite{DBLP:conf/icfp/VriesL14}. Our own $\SOP$ implementation in Haskell is
-mostly the same as in ibid. More specifically, the $\SOP$-of-object-code
-representation appeared in typed Template Haskell in \cite{DBLP:conf/haskell/PickeringLW20}.
+by De Vries and Löh \cite{DBLP:conf/icfp/VriesL14}. Our own $\SOP$
+implementation in Haskell is mostly the same as in ibid. The
+$\SOP$-of-object-code representation appeared as well in typed Template Haskell
+in \cite{DBLP:conf/haskell/PickeringLW20}.
 
@@
 \item In the downstream compiler, fast \& conservative optimization passes are
-      sufficient, since abstraction overheads are eliminated by staging. Eliminating
+      sufficient, since abstraction overheads are already eliminated by staging. Eliminating
       dead code and unused arguments would be important, since we have not yet addressed
-      this in metaprogramming.
+      this through staging.
 \item A systematic way to de-duplicate code would be also needed, probably already
@@
 
-Going a bit further, we believe that a 2LTT-based
-language could be ergonomic enough for programming in general, getting us
-valuable control over code generation for a modest amount of extra complexity.
-
-We also think that rebuilding known abstractions in staged programming is
-valuable because it provides a \emph{semantic explanation} of how abstractions
-can be compiled, and provides insights that could be reused even in non-staged
-settings. For instance, in this paper we demonstrated that compiling monadic code to
-efficient code can be done by using \emph{the same monads} in the metalanguage,
-extended with code generation as an effect.
-
-Following this line of thought, it could be interesting to also adapt to 2LTT
+Going a bit further, we believe that a 2LTT-based language could be a good design point
+for programming in general, buying us plenty of control over code generation for
+a modest amount of extra complexity.
+
+Also, rebuilding known abstractions in staged programming is valuable because it
+provides a \emph{semantic explanation} of how abstractions can be compiled, and
+provides insights that could be reused even in non-staged settings. For
+instance, in this paper we demonstrated that compiling monadic code can be done
+by using \emph{the same monads} in the metalanguage, extended with code
+generation as an effect.
+
+Continuing this line of thought, it could be interesting to also adapt to 2LTT
 the style of \emph{binding-time analysis} that is well-known in partial
@@
 stages, inventing quotes, splices, $\mup$-s and $\mdown$-s, thereby translating
-definitions to a 2LTT. Third, we perform staging and proceed from there. This would
+definitions to a 2LTT. Third, perform staging and proceed from there. This would
 be more fragile than unambiguous stage inference, but we would still benefit from
@@
 
+We also find it interesting how little impact closure-freedom had on the
+developments in this paper. This provides some evidence that in staged
+functional programming, closures can be an opt-in feature instead of the
+pervasive default.
+
 In this paper we only briefly looked at two applications. Both monad
 transformers and streams could be further developed, and many other programming
-abstraction could be revisited in the staged setting.
+abstractions could be revisited in the staged setting.
 \begin{itemize}
@@
      refined in many ways, and we could also try to represent the full transition
-     graph in an observable way, thereby enabling merging or deleting states.  We
-     could also support more optimization for restricted domains such as integer
-     ranges.
-
-\item It seems promising to find synergies between push streams, pull streams
+     graph in an observable way, thereby enabling merging or deleting states.
+
+\item It seems promising to look for synergies between push streams, pull streams
      and monad transformers. It seems that pull streams could be generalized from
@@
      generalize abstractions to arbitrary object types. For example, push and pull
-     streams can be generalized as inductive and coinductive Church-codings of
-     object types.
+     streams could be generalized to inductive and coinductive Church-codings of
+     arbitrary value types.
 \end{itemize}
-
-%% \begin{itemize}
-%%   \item Start with monadic code in a non-staged language.
-%%   \item Try to infer stages, inventing quotes, splices, $\mup$-s and $\mdown$-s,
-%%         thereby translating definitions to 2LTT.
-%%   \item Stage the obtained definitions.
-%% \end{itemize}
-%% In this approach, much complexity is factored out to staging
 

### 2024-02-29 newpaper/paper.tex
stuff

@@
 field projection. Here, a runtime closure will be created for the $\lam \vb.\,...$
-function, and $(\bind)$, $\msf{ask}$ and $\return$ will create additional dynamic
+function, and $(\bind)$, $\msf{ask}$ and $\mbf{return}$ will create additional dynamic
 closures.
@@
 \vf\,(\vg\,\vx)$ in the base libraries, instead of as $(.)\,\vf\,\vg\,\vx =
-\vf\,(\vg\,\vx)$, to get better inlining behavior \cite{ghcbase}. Tellingly, it
-is common practice in high-performance Haskell programming to visually review
-GHC's optimized code output.
+\vf\,(\vg\,\vx)$, to get better inlining behavior \cite{ghcbase}. It is common
+practice in high-performance Haskell programming to visually review GHC's
+optimized code output.
 
@@
 A CFTT program is a mixture of object-level and meta-level top-level definitions
-and declarations. \textbf{Staging} means running all metaprograms in splices and
+and declarations. \emph{Staging} means running all metaprograms in splices and
 inserting their output into object code, keeping all object-level top entries
@@
 \end{alignat*}
-For example, this can be used as $\letdef f\,\vas : \List\,\Nat \to \List\,\Nat
+For example, this can be used as $\letdef \vf\,\vas : \List\,\Nat \to \List\,\Nat
 := \spl(\map\,(\lam \vx.\,\ql \spl \vx + 10 \qr)\,\qt{\vas}) $.  This is staged to a
@@
 how they work. However, type classes are used in essentially the same way in the
-Agda and Haskell implementations of the paper, with modest technical
-differences.
+Agda and Haskell implementations, with modest technical differences.
 
@@
 \begin{alignat*}{3}
-  & \gen : \Up \vA \to \Gen\,(\Up \vA) \\
+  & \gen : \{\vA : \Ty\} \to \Up \vA \to \Gen\,(\Up \vA) \\
   & \gen\,\va = \Gen \dlr \lam \vk.\,\ql \letdef \vx : \vA := \spl \va; \spl(\vk\,\ql \vx \qr) \qr
@@
 
-We demonstrate this for lists first. An object-level $\mbf{case}$ on lists introduces
+We demonstrate this for lists. An object-level $\mbf{case}$ on lists introduces
 two points where code generation can continue. We define a metatype which gives
@@
 \end{alignat*}
-We also recover the object-level $\Maybe$ as $\MaybeT\,\Identity$, from the following $\MaybeT$:
+We recover the object-level $\Maybe$ as $\MaybeT\,\Identity$, from the following $\MaybeT$:
 \[ \newtype \MaybeT\,(\vM : \VTy \to \Ty)\,(\vA : \VTy) := \MaybeT\,\{\runMaybeT : \vM\,(\Maybe\,\vA)\} \]
@@
   &          &&\ind \ind \ind \ind \ind \ind \Nothing\hspace{0.9em} \to \Nothing\\
-  &          &&\ind \ind \ind \ind \ind \ind \Just\,(\vr,\,\vns) \to \case\,\vf\,\vr\,\vns\,\of\\
+  &          &&\ind \ind \ind \ind \ind \ind \Just\,(\vl,\,\vns) \to \case\,\vf\,\vr\,\vns\,\of\\
   &          &&\ind \ind \ind \ind \ind \ind \ind \Nothing\hspace{.95em} \to \Nothing\\
@@
 substitutions. Also, object types are untouched by substitution. Hence, a
-straightforward unfolding of semantics definitions validates the constancy of
+straightforward unfolding of semantic definitions validates the constancy of
 $\vf$.
@@
 \\\\
-\noindent \textbf{Axiom (generativity).}\textit{ Every $\vf : \Elprod\,\vA \to \USOP$ is constant. }
+\noindent \textbf{Axiom (generativity).}\textit{ For every $\vf : \Elprod\,\vA \to \USOP$ and $\va,\,\va' : \Elprod\,\vA$,
+we have that $\vf\,\va = \vf\,\va'$.}
 \\\\

### 2024-02-29 newpaper/old/submission/haskell-cftt/CFTT/Gen.hs

@@ -1,0 +1,68 @@
+
+module CFTT.Gen where
+
+import Control.Monad.Except
+import Control.Monad.Reader
+import Control.Monad.State.Strict hiding (modify')
+import Control.Monad.Trans.Maybe
+import CFTT.Up
+
+newtype Gen a = Gen {unGen :: forall r. (a -> Up r) -> Up r}
+
+runGen :: Gen (Up a) -> Up a
+runGen (Gen f) = f id
+
+instance Functor Gen where
+  fmap f ma = Gen \k -> unGen ma \a -> k (f a)
+
+instance Applicative Gen where
+  pure a = Gen \k -> k a
+  (<*>) gf ga = Gen \k -> unGen gf \f -> unGen ga \a -> k (f a)
+
+instance Monad Gen where
+  return = pure
+  (>>=) ga f = Gen \k -> unGen ga \a -> unGen (f a) k
+
+instance MonadFail Gen where
+  fail = error
+
+class Monad m => MonadGen m where
+  liftGen :: Gen a -> m a
+
+instance MonadGen Gen where
+  liftGen = id
+
+instance MonadGen m => MonadGen (StateT s m) where
+  liftGen = lift . liftGen
+
+instance MonadGen m => MonadGen (ReaderT r m) where
+  liftGen = lift . liftGen
+
+instance MonadGen m => MonadGen (ExceptT e m) where
+  liftGen = lift . liftGen
+
+instance MonadGen m => MonadGen (MaybeT m) where
+  liftGen = lift . liftGen
+
+gen :: MonadGen m => Up a -> m (Up a)
+gen a = liftGen $ Gen \k -> [|| let x = $$a in seq x $$(k [||x||]) ||]
+
+genRec :: MonadGen m => (Up a -> Up a) -> m (Up a)
+genRec a = liftGen $ Gen \k -> [|| let x = $$(a [||x||]) in $$(k [||x||]) ||]
+
+put' :: (MonadState (Up s) m, MonadGen m) => Up s -> m (Up ())
+put' s = do
+  s <- gen s
+  put s
+  pure [||()||]
+
+modify' :: (MonadState (Up s) m, MonadGen m) => (Up s -> Up s) -> m (Up ())
+modify' f = do
+  s <- get
+  put' (f s)
+
+local' :: (MonadReader (Up r) m, MonadGen m) => (Up r -> Up r) -> m a -> m a
+local' f ma = do
+  r <- ask
+  r <- gen (f r)
+  local (const r) ma

### 2024-05-27 newpaper/paper.tex
revision wip

@@
 \newcommand{\vma}{\mathsf{ma}}
+\newcommand{\vga}{\mathsf{ga}}
 \newcommand{\vm}{\mathsf{m}}
@@
 In the $\Pull$ constructor, $\vS$ is the type of the internal state which is
-required to be a sum-of-products of value types by the $\IsSumVS\,\vS$
+required to be a sum-of-products of value types by the $\IsSOP\,\vS$
 constraint. Borrowing terminology from Coutts \cite{DBLP:phd/ethos/Coutts11}, we
@@
 This problem of generating well-typed mutual blocks was addressed by Yallop and
-Kiselyov in \cite{DBLP:conf/pepm/YallopK19}. Unlike ibid., which used control
-effects and mutable references in MetaOCaml, we present a solution that does not
-use side effects in the metalanguage.
+Kiselyov in \cite{DBLP:conf/pepm/YallopK19}. In contrast to this work, which
+used control effects and mutable references in MetaOCaml, we present a solution
+that does not use side effects in the metalanguage.
 
@@
   &\hspace{-13em}\mathrlap{\foldr : \{\vA : \MTy\}\{\vB : \Ty\} \to (\vA \to \Lift \vB \to \Lift \vB) \to \Lift \vB \to \Pull\,\vA \to \Lift\,\vB}\\
-  &\hspace{-13em}\mathrlap{\foldr\,\{\vA\}\,\{\vB\}\,(\Pull\,\vS\,\seed\,\step)\,\vf\,\vb = \ql}\\
+  &\hspace{-13em}\mathrlap{\foldr\,\{\vA\}\,\{\vB\}\,\vf\,\vb\,(\Pull\,\vS\,\seed\,\step) = \ql}\\
   &\hspace{-13em}\mathrlap{\ind \letrec \vfs : \mit{Fun}_{\SOP\Lift}\,(\Rep\,\{\vS\})\,\vB := \spl\big(\mtabulate \dlr \lam \vs.\,\unGen\,(\step\,(\rep^{-1}\,\vs)) \dlr \lam \case} \\
@@
 \end{alignat*}
-This $\foldr$ is quite flexible becuse we can compute any object value from it,
-including functions. For instance, we can define $\foldl$ from $\foldr$:
+This $\foldr$ is quite flexible because it can produce object terms with computation types.
+For instance, we can define $\foldl$ from $\foldr$:
 \begin{alignat*}{3}
@@
 \[ \concatMap : \IsSOP\,\vA \RA (\vA \to \Pull\,\vB) \to \Pull\,\vA \to \Pull\,\vB \]
-The idea is the following: if $\USOP$ is closed $\Sigma$-types, we can directly
+The idea is the following: if $\USOP$ is closed under $\Sigma$-types, we can directly
 define this $\concatMap$, by taking the appropriate dependent sum of the $\vA
@@
 $\single$, to avoid the size explosion we give more specialized definitions
-instead. First, we define a helper function for binding a single $\IsSOP$ value.
-\begin{alignat*}{3}
-  & \hspace{-11em}\mathrlap{\mit{bind_{single}} : \IsSOP\,\vA' \RA \Lift \vA \to (\Lift \vA \to \Gen\,\vA') \to (\vA' \to \Pull\,\vB) \to \Pull\,\vB}\\
-  & \hspace{-11em}\mathrlap{\mit{bind_{single}}\,\va\,\vf\,\vg =}\\
-  & \hspace{-11em}\mathrlap{\ind \Pull\,(\Sigma\,\vA'\,(\mit{projS} \circ \vg))\,(\mdo \{\va' \fro \vf\,\va; \vs \fro \mit{projSeed}\,(\vg\,\va'); \return (\va',\,\vs)) \dlr \lam \case}\\
-  & \hspace{-11em}\mathrlap{\ind \ind (\va',\,\vs) \to \mit{projStep}\,(\vg\,\va')\,\vs \bind \lam \case }\\
+instead. First, we define an embedding of $\Gen$ operations in streams.
+\begin{alignat*}{3}
+  & \hspace{-11em}\mathrlap{\mit{bind_{\Gen}} : \IsSOP\,\vA \RA \Gen\,\vA \to (\vA \to \Pull\,\vB) \to \Pull\,\vB}\\
+  & \hspace{-11em}\mathrlap{\mit{bind_{\Gen}}\,\vga\,\vf\,=}\\
+  & \hspace{-11em}\mathrlap{\ind \Pull\,(\Sigma\,\vA\,(\mit{projS} \circ \vf))\,(\mdo \{\va \fro \vga; \vs \fro \mit{projSeed}\,(\vf\,\va); \return (\va,\,\vs)) \dlr \lam \case}\\
+  & \hspace{-11em}\mathrlap{\ind \ind (\va,\,\vs) \to \mit{projStep}\,(\vf\,\va)\,\vs \bind \lam \case }\\
   & \hspace{-11em}\ind \ind \ind \Stop            &&\to \return \Stop\\
-  & \hspace{-11em}\ind \ind \ind \Skip\,\vs       &&\to \return \dlr \Skip\,(\va',\,\vs) \\
-  & \hspace{-11em}\ind \ind \ind \Yield\,\vb\,\vs &&\to \return \dlr \Yield\,\vb\,(\va',\,\vs)
+  & \hspace{-11em}\ind \ind \ind \Skip\,\vs       &&\to \return \dlr \Skip\,(\va,\,\vs) \\
+  & \hspace{-11em}\ind \ind \ind \Yield\,\vb\,\vs &&\to \return \dlr \Yield\,\vb\,(\va,\,\vs)
 \end{alignat*}
@@
   &\mit{gen_\Pull} : \Lift \vA \to (\Lift \vA \to \Pull\,\vB) \to \Pull\,\vB\\
-  &\mit{gen_\Pull}\,\va = \mit{bind_{single}}\,\va\,\gen
+  &\mit{gen_\Pull}\,\va = \mit{bind_{\Gen}}\,(\gen\,\va)
 \end{alignat*}
@@
   &\mit{case_\Pull} : (\mit{Split}\,\vA,\,\IsSOP\,(\mit{SplitTo}\,\{\vA\})) \RA \Lift \vA \to (\mit{SplitTo}\,\{\vA\} \to \Pull\,\vB) \to \Pull\,\vB\\
-  &\mit{case_\Pull}\,\va = \mit{bind_{single}}\,\va\,\mit{split}
+  &\mit{case_\Pull}\,\va = \mit{bind_{\Gen}}\,(\mit{split}\,\va)
 \end{alignat*}

### 2024-06-02 newpaper/paper.tex
wip

@@
 \usepackage{mathtools}
+\usepackage{stmaryrd}
+\usepackage[title]{appendix}
 
@@
 \newcommand{\mbb}[1]{\mathbb{#1}}
+\newcommand{\U}{\mathsf{U}}
+\newcommand{\code}{\mathsf{code}}
 \newcommand{\bs}[1]{\boldsymbol{#1}}
@@
 \newcommand{\map}{\mit{map}}
+\newcommand{\emptycon}{\scaleobj{.75}\bullet}
+\newcommand{\Tyo}{\msf{Ty}_{\mbbo}}
+\newcommand{\Tmo}{\msf{Tm}_{\mbbo}}
+\newcommand{\whset}{\wh{\Set}}
+\newcommand{\ev}{\mbb{E}}
+\newcommand{\re}{\mbb{R}}
+
+\newcommand{\mbbc}{\mbb{C}}
+\newcommand{\mbbo}{\mbb{O}}
 
@@
 \newcommand{\vns}{\mathsf{ns}}
+\newcommand{\vW}{\mathsf{W}}
+\newcommand{\vsup}{\mathsf{sup}}
+\newcommand{\vid}{\mathsf{id}}
+\newcommand{\whW}{\wh{\vW}}
 
@@
 \newcommand{\appull}{\ap_{\Pull}}
+
+\newcommand{\Con}{\msf{Con}}
+\newcommand{\Sub}{\msf{Sub}}
+\newcommand{\Tm}{\msf{Tm}}
+\newcommand{\Mor}{\msf{Tm}}
 
@@
 \newcommand{\MTy}{\msf{MetaTy}}
+\newcommand{\MTm}{\msf{MetaTm}}
 \newcommand{\VTy}{\msf{ValTy}}
@@
 \begin{alignat*}{3}
-  &\vf = \lam \vx.\,(\bind)\,\msf{MonadReaderDict}\,\msf{ask}\,(\lam \vb.\,\bif \vb\\
+  &\vf = \lam \vx.\,(\bind)\,\msf{MonadReaderDict}\,(\msf{ask}\,\msf{MonadReaderDict})\,(\lam \vb.\,\bif \vb\\
   &\ind \bthen\return\,\msf{MonadReaderDict}\,(\vx + 10)\\
@@
 Here, $\msf{MonadReaderDict}$ is a runtime dictionary, containing the methods of
-the $\Monad$ instance for $\Reader$, and $(\bind)\,\msf{MonadReaderDict}$ is a
-field projection. Here, a runtime closure will be created for the $\lam \vb.\,...$
-function, and $(\bind)$, $\msf{ask}$ and $\mbf{return}$ will create additional dynamic
-closures.
+the $\Monad$ instance for $\Reader$, and $(\bind)\,\msf{MonadReaderDict}$ is the
+projection of the binding function. Here, a runtime closure will be created for
+the $\lam \vb.\,...$ function, and $(\bind)$, $\msf{ask}$ and $\mbf{return}$
+will create additional dynamic closures.
 
@@
 dictionary, then inlining the methods contained there, and also inlining the
-functions that the higher-order binding method is applied to.
+functions that the higher-order binding is applied to.
 
@@
 
-\subsection{Semantics of Staging}\label{sec:semantics-of-staging}
-
-We give a short summary of the semantics of staging here. We reuse most of the
-semantics from \cite{staged2ltt} and only mention additions and differences. The
-reader may refer to ibid.\ for technical details.
-
-Staging is given by evaluation of CFTT into a presheaf model, where the base
-category is given as follows: objects are typing contexts of the object language
-and morphisms are parallel substitutions. The equality of morphisms is given by
-strict syntactic equality of object terms. $\Ty$ is interpreted as the presheaf
-which is constantly the set of object-theoretic types. Soundness of staging is
-proven by a logical relation model, internally to the mentioned presheaf
-model. Soundness means that the output of staging, viewed as a CFTT term, is
-definitionally equal to the input.
-
-The main addition in CFTT is the assumption of all inductive families in the
-meta-level. However, based on the results in \cite{whynotw}, it is enough to
-extend the meta-level syntax in \cite{staged2ltt} with W-types and identity types to
-derive all inductive families. We consider the semantics of these in the following.
-
-\subsubsection{W-types} W-types have a standard object-wise inductive definition in
-the presheaf model \cite{DBLP:journals/apal/MoerdijkP00}. Additionally, we need to give a logical relation
-interpretation to W-types, as required for the soundness proof for staging. In
-Section 5.2.\ of \cite{staged2ltt} a serialization map is used to define the relation
-for natural numbers, however, there is no analogous map for W-types because we
-cannot build finite terms from infinitely branching trees.
-
-Instead, we give an inductive definition for the logical relation at
-W-types. First, assuming $A : \wh{\Set}$ and $B : A \to \wh{\Set}$, we have the
-usual constructor for the semantic W-type in the internal language of the
-presheaf model, as $\msf{sup} : (a : A) \to (B\,a \to \msf{W}\,A\,B) \to
-\msf{W}\,A\,B$.  Second, assuming $A : \Ty\,\Gamma$ and $B : \Ty\,(\Gamma \ext
-A)$ in CFTT, we internalize the $\msf{sup}$ constructor which constructs a term
-of a syntactic W-type from restricted terms:
-\[ \mathbb{R}_{\msf{sup}} : (\alpha : \mathbb{R}\,A\,\gamma) \to \mathbb{R}\,(B \RA (\msf{W_{CFTT}}\,A\,B)[\msf{wk}])\,(\gamma,\,\alpha) \to \mathbb{R}\,(\msf{W_{CFTT}}\,A\,B)\,\gamma \]
-Here, $\msf{W_{CFTT}}$ denotes the W-type former in CFTT, $B \RA
-(\msf{W_{CFTT}}\,A\,B)[\msf{wk}]$ is a non-dependent function type in CFTT, and
-$\msf{wk}$ is a weakening substitution.
-
-Then, assuming $\gamma^{\approx} : \Gamma^{\approx}\,\gamma\,\gamma'$, we
-inductively define a relation $\msf{W}^{\approx}$ between
-$\msf{W}\,(\mbb{E}\,A\,\gamma)\,(\lam \alpha.\,\mbb{E}\,B\,(\gamma,\,\alpha))$
-and $\mbb{R}\,(\msf{W_{CFTT}}\,A\,B)\,\gamma'$, which relates
-$\msf{sup}\,\alpha\,f$ to $\mathbb{R}_{\msf{sup}}\,\alpha'\,f'$ if $\alpha$ is
-related to $\alpha'$ at type $A$ and $f$ is related to $f'$ at type $B[\alpha']
-\RA \msf{W_{CFTT}}\,A\,B$. The relation at the non-dependent function type $\blank{\RA}\blank$
-is given as pointwise preservation of relations. With this, we can also give
-relational interpretation to the syntactic $\msf{sup}$ rule in CFTT,
-and we can use induction on $\msf{W}^{\approx}$ to interpret the syntactic
-induction rule for W. Thus, we get soundness of staging with W-types.
-
-\subsubsection{Identity type} Our treatment of the identity type is
-influenced by the following: we would like to assume equations as axioms in
-CFTT, which are true in the presheaf model (i.e.\ they are true at
-staging-time), without blocking staging as a computation. In Section \ref{sec:concatmap-for-streams}
-we use such an axiom to good effect. However, axioms do block computation in
-CFTT up to definitional equality, which implies that the soundness of staging
-fails. Our solution:
+\subsection{Staged Semantics of CFTT}\label{sec:semantics-of-staging}
+
+We give an overview of the unstaging algorithm together with its correctness.
+We reuse the concepts and most of the results from \cite{staged2ltt}. Our main
+task is to extend ibid.\, with identity types and inductive types in the
+metalanguage, since we use them in this paper. For inductive types, it
+is the most sensible to develop W-types, since all inductive families can be
+faithfully derived from them in our setting \cite{whynotw}.
+
+This however requires a substantial amount of technical background from
+\cite{staged2ltt}, so we relegate it to Appendix \ref{appendix:A}. Here we only
+review the basic notions.
 \begin{itemize}
-\item Have a standard intensional identity type in CFTT.
-\item Before staging, translate the intensional identity to extensional
-      identity. This amounts to erasing all identity proofs and transports.
-\item Perform staging after erasure of identity proofs, i.e.\ model extensional identity in the semantics.
+\item The input of unstaging is a CFTT term with an object-level type, which
+      only depends on object-level free variables. The output is a term in the
+      object theory. In the other direction, there is an evident embedding of
+      object-theoretic terms as CFTT terms.
+\item \emph{Soundness} means that unstaging followed by embedding is the
+      identity map, up to conversion. In other words, unstaging respects conversion
+      of CFTT terms.
+\item \emph{Stability} means that embedding followed by unstaging is the
+      identity map. In other words, unstaging has no action on terms which are
+      already purely object-level (i.e.\ contain no splices).
 \end{itemize}
-In the presheaf model, we interpret the extensional identity in a standard way,
-as strict equality of sections of presheaves, and soundness for extensional
-identity becomes trivial to show. Hence, the soundness property for our overall
-staging pipeline becomes the following: the staging output is definitionally
-equal to the identity-erased version of the staging input.
+In \cite{staged2ltt}, there is an extra property, called \emph{strictness},
+which means that unstaging does not perform $\beta\eta$-conversion in the
+object theory, but this holds trivially in our case since we do not have such
+conversion rules.
+
+However, for CFTT we slightly depart from the above notion of soundness, for the
+following reason. We would like to assume certain propositional identities as
+\emph{axioms} in the metalanguage, without blocking unstaging as a
+computation. We use such an axiom to good effect in Section
+\ref{sec:concatmap-for-streams}.
+
+Now, axioms clearly block computation, so they are incompatible with the
+mentioned notion of soundness. But we are only interested in equations which
+hold definitionally during unstaging. Hence, we can do the following: first, we
+erase all identity proofs and their transports from CFTT programs, then we
+proceed with unstaging. The erasure of identity proofs can be defined as a
+syntactic translation \cite{next700} from CFTT into CFTT with an extensional
+identity type. Thus, what we get is that unstaging is sound only after the
+erasure of identity proofs.
+
+%% \subsubsection{Identity type} Our treatment of the identity type is
+%% influenced by the following: we would like to assume equations as axioms in
+%% CFTT, which are true in the presheaf model (i.e.\ they are true at
+%% staging-time), without blocking staging as a computation. In Section \ref{sec:concatmap-for-streams}
+%% we use such an axiom to good effect. However, axioms do block computation in
+%% CFTT up to definitional equality, which implies that the soundness of staging
+%% fails. Our solution:
+%% \begin{itemize}
+%% \item Have a standard intensional identity type in CFTT.
+%% \item Before staging, translate the intensional identity to extensional
+%%       identity. This amounts to erasing all identity proofs and transports.
+%% \item Perform staging after erasure of identity proofs, i.e.\ model extensional identity in the semantics.
+%% \end{itemize}
+%% In the presheaf model, we interpret the extensional identity in a standard way,
+%% as strict equality of sections of presheaves, and soundness for extensional
+%% identity becomes trivial to show. Hence, the soundness property for our overall
+%% staging pipeline becomes the following: the staging output is definitionally
+%% equal to the identity-erased version of the staging input.
+
+%% In the
+%% current work we use identity types and some inductive types and the
+%% metalanguage, which are not covered in \cite{staged2ltt}. Hence, our main task
+
+%% We give an overview of the semantics of CFTT. This yields a algorithm for
+%% evaluating
+%% extracting object programs from CFTT
+
+%% We give a short summary of the semantics of staging here. We reuse most of the
+%% semantics from \cite{staged2ltt} and only mention additions and differences. The
+%% reader may refer to ibid.\ for technical details.
+
+%% Staging is given by evaluation of CFTT into a presheaf model, where the base
+%% category is given as follows: objects are typing contexts of the object language
+%% and morphisms are parallel substitutions. The equality of morphisms is given by
+%% strict syntactic equality of object terms. $\Ty$ is interpreted as the presheaf
+%% which is constantly the set of object-theoretic types. Soundness of staging is
+%% proven by a logical relation model, internally to the mentioned presheaf
+%% model. Soundness means that the output of staging, viewed as a CFTT term, is
+%% definitionally equal to the input.
+
+%% The main addition in CFTT is the assumption of all inductive families in the
+%% meta-level. However, based on the results in \cite{whynotw}, it is enough to
+%% extend the meta-level syntax in \cite{staged2ltt} with W-types and identity types to
+%% derive all inductive families. We consider the semantics of these in the following.
+
+%% \subsubsection{W-types} W-types have a standard object-wise inductive definition in
+%% the presheaf model \cite{DBLP:journals/apal/MoerdijkP00}. Additionally, we need to give a logical relation
+%% interpretation to W-types, as required for the soundness proof for staging. In
+%% Section 5.2.\ of \cite{staged2ltt} a serialization map is used to define the relation
+%% for natural numbers, however, there is no analogous map for W-types because we
+%% cannot build finite terms from infinitely branching trees.
+
+%% Instead, we give an inductive definition for the logical relation at
+%% W-types. First, assuming $A : \wh{\Set}$ and $B : A \to \wh{\Set}$, we have the
+%% usual constructor for the semantic W-type in the internal language of the
+%% presheaf model, as $\msf{sup} : (a : A) \to (B\,a \to \msf{W}\,A\,B) \to
+%% \msf{W}\,A\,B$.  Second, assuming $A : \Ty\,\Gamma$ and $B : \Ty\,(\Gamma \ext
+%% A)$ in CFTT, we internalize the $\msf{sup}$ constructor which constructs a term
+%% of a syntactic W-type from restricted terms:
+%% \[ \mathbb{R}_{\msf{sup}} : (\alpha : \mathbb{R}\,A\,\gamma) \to \mathbb{R}\,(B \RA (\msf{W_{CFTT}}\,A\,B)[\msf{wk}])\,(\gamma,\,\alpha) \to \mathbb{R}\,(\msf{W_{CFTT}}\,A\,B)\,\gamma \]
+%% Here, $\msf{W_{CFTT}}$ denotes the W-type former in CFTT, $B \RA
+%% (\msf{W_{CFTT}}\,A\,B)[\msf{wk}]$ is a non-dependent function type in CFTT, and
+%% $\msf{wk}$ is a weakening substitution.
+
+%% Then, assuming $\gamma^{\approx} : \Gamma^{\approx}\,\gamma\,\gamma'$, we
+%% inductively define a relation $\msf{W}^{\approx}$ between
+%% $\msf{W}\,(\mbb{E}\,A\,\gamma)\,(\lam \alpha.\,\mbb{E}\,B\,(\gamma,\,\alpha))$
+%% and $\mbb{R}\,(\msf{W_{CFTT}}\,A\,B)\,\gamma'$, which relates
+%% $\msf{sup}\,\alpha\,f$ to $\mathbb{R}_{\msf{sup}}\,\alpha'\,f'$ if $\alpha$ is
+%% related to $\alpha'$ at type $A$ and $f$ is related to $f'$ at type $B[\alpha']
+%% \RA \msf{W_{CFTT}}\,A\,B$. The relation at the non-dependent function type $\blank{\RA}\blank$
+%% is given as pointwise preservation of relations. With this, we can also give
+%% relational interpretation to the syntactic $\msf{sup}$ rule in CFTT,
+%% and we can use induction on $\msf{W}^{\approx}$ to interpret the syntactic
+%% induction rule for W. Thus, we get soundness of staging with W-types.
+
+%% \subsubsection{Identity type} Our treatment of the identity type is
+%% influenced by the following: we would like to assume equations as axioms in
+%% CFTT, which are true in the presheaf model (i.e.\ they are true at
+%% staging-time), without blocking staging as a computation. In Section \ref{sec:concatmap-for-streams}
+%% we use such an axiom to good effect. However, axioms do block computation in
+%% CFTT up to definitional equality, which implies that the soundness of staging
+%% fails. Our solution:
+%% \begin{itemize}
+%% \item Have a standard intensional identity type in CFTT.
+%% \item Before staging, translate the intensional identity to extensional
+%%       identity. This amounts to erasing all identity proofs and transports.
+%% \item Perform staging after erasure of identity proofs, i.e.\ model extensional identity in the semantics.
+%% \end{itemize}
+%% In the presheaf model, we interpret the extensional identity in a standard way,
+%% as strict equality of sections of presheaves, and soundness for extensional
+%% identity becomes trivial to show. Hence, the soundness property for our overall
+%% staging pipeline becomes the following: the staging output is definitionally
+%% equal to the identity-erased version of the staging input.
 
@@
 \end{itemize}
+\newpage
+
+\appendices
+\section{Semantics of CFTT}\label{appendix:A}
+
+In this appendix we describe the semantics of CFTT, from which the unstaging
+algorithm is obtained, and also describe the soundness and completeness of the
+algorithm. We borrow the setup from \cite{staged2ltt}; we use terminology and
+definitions from there by default, and focus on differences and extensions.
+
+\subsection{Models and Syntax of CFTT}
+
+We give an algebraic \cite{TODO} specification for the models of CFTT and define
+the syntax of CFTT to be the initial model.
+
+\begin{definition}[] A model of CFTT consists of the following.
+
+\begin{itemize}
+\item A category with a terminal object; in the syntax this is the category
+  of contexts and parallel substitutions. We use $\Con : \Set$ for contexts,
+  $\Mor : \Con \to \Con \to \Set$ for morphisms and $\emptycon : \Con$ for the
+  terminal object (which is the empty context in the syntax).
+\item A family structure in the sense of category-with-families \cite{TODO},
+  specifying metatypes and terms. This structure is additionally indexed by
+  universe levels, so we have $\MTy_i : \Con \to \Set$ and $\msf{MetaTm}_i :
+  (\Gamma : \Con) \to \MTy_i\,\Gamma \to \Set$ for $i \in \mbb{N}$. We support
+  basic type formers, identity types and W-types. Additionally, we have
+  Coquand-style universes, i.e.\ $\U_i : \MTy_{i+1}\,\Gamma$ such that there is
+  a definitional isomorphism $\msf{MetaTm}_{i+1}\,\Gamma\,\U_i \simeq \MTy_i\,\Gamma$.
+  We use $\El$ and $\code$ to denote the components of this isomorphism.
+\item
+  For object-level types, we have $\Ty : \MTy_0\,\Gamma$, $\VTy :
+  \MTy_0\,\Gamma$ and $\CTy : \MTy_0\,\Gamma$, together with all object-level
+  type formers and inclusion operations $\msf{V} : \MTm\,\Gamma\,\VTy \to \MTm\,\Gamma\,\Ty$
+  and $\msf{C} : \MTm\,\Gamma\,\CTy \to \MTm\,\Gamma\,\Ty$.
+\item For object-level terms, we have $\Tm : \MTm\,\Gamma\,\Ty \to \Set$, closed under
+  all object-level term formers. Here, $\MTm\,\blank\,\Ty$ and $\Tm$ also form a
+  family structure, so that we can substitute object terms and extend the context
+  with object term variables.
+\item Staging operations: we have $\Lift : \MTm\,\Gamma\,\Ty \to \MTy\,\Gamma$ together
+  with definitional isomorphisms $\Tm\,\Gamma\,A \simeq \MTm\,\Gamma\,(\Lift A)$.
+  We write $\ql\!\blank\!\qr$ and $\spl\!\blank$ for the components of the isomorphism.
+\end{itemize}
+\end{definition}
+This presentation is a bit more verbose than the ``surface'' syntax used in the
+paper. First, the meta level uses explicit coding and decoding operations for
+universes, while the surface syntax assumes Russell-style universes. Second, the
+object level has $\VTy$ and $\CTy$ as Tarski-style universes with explicit
+inclusion into $\Ty$, while the surface syntax used implicit coercions.
+
+Also, this is not the only possible presentation. Here, we follow the style of
+the informal syntax by having a separate \emph{sort} for object-level terms,
+together with explicit quoting and splicing, but we specify object-level types
+as certain meta-level terms. This a hybrid solution of two styles:
+\begin{itemize}
+\item One would have separate sorts for $\Ty$, $\CTy$ and $\VTy$, and would have
+  separate staging operations for all of these sorts. This follows the ``classic''
+  2LTT presentation \cite{TODO}, where the two levels are specified independently
+  and level-mixing is only possible through the staging operations.
+\item The other one would only have metatypes and terms, no lifting, quoting or
+  splicing, and would use metafunctions to represent object-level binders. This
+  is the HOAS (``higher-order abstract syntax'') or ``logical framework''
+  flavor. Our Agda implementation actually follows this style.
+\end{itemize}
+In the context of CFTT, the choice between HOAS and 2LTT styles is a matter of
+taste, and the two are inter-derivable.
+
+\subsection{Presheaf Model}
+
+Unstaging is defined as evaluation of the syntax of CFTT in a particular
+presheaf model.
+\begin{itemize}
+  \item The \emph{object theory} is the simple type theory supporting only value
+    types and computation types as described in Section
+    \ref{sec:the-object-level} and also in the Agda
+    formalization. Object-theoretic terms are not quotiented by
+    $\beta\eta$-conversion. Substitution is a recursively defined operation on
+    terms.
+  \item $\mbbo$ is the category whose objects are contexts of the object theory,
+    and morphisms are parallel substitutions, as lists of terms. With this,
+    object-theoretic types form a constant presheaf which we denote
+    as $\Tyo : \mbbo^{\msf{op}} \to \Set$, and terms form a dependent presheaf over $\Tyo$ which we write as
+    $\Tmo$.
+  \item
+  In the presheaf model of CFTT, every typing context (and also every closed
+  type) is modeled as a presheaf over $\mbbo$. The interpretation of meta-level
+  type formers is standard, following Hofmann \cite{Hofmann97syntaxand} and
+  Huber \cite{huber-thesis}.
+\item $\Ty : \MTy_0\,\Gamma$ is the displayed presheaf which is constantly
+  $\Tyo$. $\VTy$ and $\CTy$ are modeled analogously.
+\item $\Lift A : \MTy_0\,\Gamma$ is modeled as follows. We have that $A :
+  \MTy\,\Gamma\,\Ty$, but since $\Ty$ is defined as constantly $\Tyo$, it
+  follows that $A$ is a natural transformation from $\Gamma$ to $\Tyo$. Hence,
+  we define $\Lift A : \MTy_0\,\Gamma$ to be $\Tmo[A]$, where $[\blank]$ is
+  substitution in presheaves, i.e.\ the reindexing of a displayed presheaf by a
+  natural transformation.
+\item
+  $\Tm\,\Gamma\,A$ is simply defined to be $\MTy\,\Gamma\,(\Lift A)$ in the model, which
+  means that quotation and splicing are both identity maps.
+\end{itemize}
+Now, we can take a closed term $t : \Tm\,\emptycon\,A$ in CFTT and evaluate it
+in the presheaf model to obtain an object-theoretic term whose type is the
+result of evaluating $A$.
+
+\subsection{Soundness for W-types}
+So far we have stayed close to \cite{staged2ltt}, with only cosmetic
+changes. Supporting W-types, as a way to get inductive families \cite{TODO}, is
+the biggest extension here. We extend Section 5 of \cite{staged2ltt} here.  In
+the syntax of CFTT, we have
+\begin{alignat*}{3}
+  &\vW   &&: (A : \MTy_i\,\Gamma) \to \MTy_j\,(\Gamma,\,A) \to \MTy_{\msf{max}(i,j)}\,\Gamma\\
+  &\vsup &&: (a : \MTm\,\Gamma\,A) \to \MTm\,(\Gamma,B[\vid,a])\,(\vW\,A\,B) \to \MTm\,\Gamma\,(\vW\,A\,B),
+\end{alignat*}
+where $\vid$ is the identity substitution from $\Gamma$ to $\Gamma$. We also have the eliminator for $\vW$
+whose specification we omit here.
+
+Internally to $\whset$, i.e.\ internally to presheaves over $\mbbo$, we also
+have the standard W-type, defined pointwise as the W-type in sets:
+\begin{alignat*}{3}
+  & \wh{\vW}   &&: (A : \whset) \to (A \to \whset) \to \whset\\
+  & \wh{\vsup} &&: (a : A) \to (B\,a \to \wh{\vW}\,A\,B) \to \wh{\vW}\,A\,B
+\end{alignat*}
+In the presheaf model we can evidently interpret $\vW$ using $\whW$.
+
+%% The \emph{restriction} of this
+
+%% \subsubsection{Stability} Stability means that if we take any object-theoretic term,
+%% embed it as a CFTT term, then unstage it
 

### 2024-06-07 newpaper/paper.tex
wip

@@
 \newcommand{\mbb}[1]{\mathbb{#1}}
+\newcommand{\rel}{^{\approx}}
 \newcommand{\U}{\mathsf{U}}
+\newcommand{\p}{\mathsf{p}}
+\newcommand{\q}{\mathsf{q}}
 \newcommand{\code}{\mathsf{code}}
@@
 \newcommand{\re}{\mbb{R}}
+\newcommand{\welim}{\vW{-}\msf{elim}}
 
@@
 \newcommand{\Tm}{\msf{Tm}}
-\newcommand{\Mor}{\msf{Tm}}
 
@@
 With what we have seen so far, there is no interaction between the meta and
-object levels. We make such interaction possible with \emph{staging operations}.
+object levels. We make such interaction possible with \emph{staging primitives}.
 
@@
   term. \emph{Notation:} splicing binds stronger than function application, so
-  $\vf\,\spl \vx$ is parsed as $\vf\,(\spl \vx)$. We borrow this notation from
+  $\vf\,\spl \vx$ is parsed as $\vf\,(\spl \vx)$. We borrow this from
   MetaML \cite{metaml}.
@@
 A CFTT program is a mixture of object-level and meta-level top-level definitions
-and declarations. \emph{Staging} means running all metaprograms in splices and
-inserting their output into object code, keeping all object-level top entries
-and discarding all meta-level ones. Thus, staging takes a CFTT program as input,
-and produces output which is purely in the object-level fragment, with no
-metatypes and metaprograms remaining. The output is guaranteed to be well-typed.
-
-Let us look at some basic staging examples. Recall the meta-level identity
-function; it can be used at the object-level too, by applying it to quoted
-terms:
+and declarations. We use \emph{unstaging} to refer to the process of evaluating
+metaprograms in a CFTT program in order to extract purely object-level code.
+The output is guaranteed to be well-typed.
+
+Let us look at some basic examples. Recall the meta-level identity function; it
+can be used at the object-level too, by applying it to quoted terms:
   \[ \letdef \vn : \Nat := \spl(\id\,\ql 10 + 10 \qr); ... \]
-Here, $\id$ is used at type $\Lift \Nat$. During staging, the expression in the
-splice is evaluated, so we get $\spl\ql 10 + 10 \qr$, which is definitionally
-the same as $10 + 10$, which is our staging output here. Boolean
+Here, $\id$ is used at type $\Lift \Nat$. During unstaging, the expression in
+the splice is evaluated, so we get $\spl\ql 10 + 10 \qr$, which is
+definitionally the same as $10 + 10$, which is our code output here. Boolean
 short-circuiting is another basic use-case:
@@
 \end{alignat*}
-For example, this can be used as $\letdef \vf\,\vas : \List\,\Nat \to \List\,\Nat
-:= \spl(\map\,(\lam \vx.\,\ql \spl \vx + 10 \qr)\,\qt{\vas}) $.  This is staged to a
-recursive definition where the mapping function is inlined into the $\Cons$ case
-as $\Cons\,\va\,\vas \to \Cons\,(\va + 10)\,(\go\,\vas)$.  Note that $\map$ has to
-abstract over value types, since lists can only contain values, not
-functions. Also, the mapping function has type $\Up \vA \to \Up \vB$, instead of
-$\Up (\vA \to \vB)$. The former type is often preferable to the latter in staging;
-the former is a metafunction with useful computational content, while the latter
-is merely a black box that computes object code. If we have $\vf : \Up(\vA \to \vB)$,
-and $\vf$ is staged to $\ql \lam \vx.\,\vt\qr$, then $\spl \vf\,\vu$ is staged to an
-undesirable ``administrative'' $\beta$-redex $(\lam \vx.\,\vt)\,\vu$.
+For example, this can be used as $\letdef \vf\,\vas : \List\,\Nat \to
+\List\,\Nat := \spl(\map\,(\lam \vx.\,\ql \spl \vx + 10 \qr)\,\qt{\vas}) $.
+This is unstaged to a recursive definition where the mapping function is inlined
+into the $\Cons$ case as $\Cons\,\va\,\vas \to \Cons\,(\va + 10)\,(\go\,\vas)$.
+Note that $\map$ has to abstract over value types, since lists can only contain
+values, not functions. Also, the mapping function has type $\Up \vA \to \Up
+\vB$, instead of $\Up (\vA \to \vB)$. The former type is often preferable to the
+latter; the former is a metafunction with useful computational content, while
+the latter is merely a black box that computes object code. If we have $\vf :
+\Up(\vA \to \vB)$, and $\vf$ computes to $\ql \lam \vx.\,\vt\qr$, then $\spl
+\vf\,\vu$ is unstaged to an undesirable ``administrative'' $\beta$-redex $(\lam
+\vx.\,\vt)\,\vu$.
 
@@
 We reuse the concepts and most of the results from \cite{staged2ltt}. Our main
-task is to extend ibid.\, with identity types and inductive types in the
-metalanguage, since we use them in this paper. For inductive types, it
-is the most sensible to develop W-types, since all inductive families can be
-faithfully derived from them in our setting \cite{whynotw}.
+task is to extend ibid.\,with identity types and inductive types in the
+metalanguage, since we use them in this paper. For inductive types, it is the
+most sensible to add W-types, since all inductive families can be faithfully
+derived from them in our setting \cite{whynotw}.
 
@@
 \cite{staged2ltt}, so we relegate it to Appendix \ref{appendix:A}. Here we only
-review the basic notions.
+review the basics.
 \begin{itemize}
@@
 \end{itemize}
-In \cite{staged2ltt}, there is an extra property, called \emph{strictness},
+In \cite{staged2ltt}, there is one more property, called \emph{strictness},
 which means that unstaging does not perform $\beta\eta$-conversion in the
@@
 
-Now, axioms clearly block computation, so they are incompatible with the
-mentioned notion of soundness. But we are only interested in equations which
-hold definitionally during unstaging. Hence, we can do the following: first, we
-erase all identity proofs and their transports from CFTT programs, then we
-proceed with unstaging. The erasure of identity proofs can be defined as a
-syntactic translation \cite{next700} from CFTT into CFTT with an extensional
-identity type. Thus, what we get is that unstaging is sound only after the
-erasure of identity proofs.
-
-%% \subsubsection{Identity type} Our treatment of the identity type is
-%% influenced by the following: we would like to assume equations as axioms in
-%% CFTT, which are true in the presheaf model (i.e.\ they are true at
-%% staging-time), without blocking staging as a computation. In Section \ref{sec:concatmap-for-streams}
-%% we use such an axiom to good effect. However, axioms do block computation in
-%% CFTT up to definitional equality, which implies that the soundness of staging
-%% fails. Our solution:
-%% \begin{itemize}
-%% \item Have a standard intensional identity type in CFTT.
-%% \item Before staging, translate the intensional identity to extensional
-%%       identity. This amounts to erasing all identity proofs and transports.
-%% \item Perform staging after erasure of identity proofs, i.e.\ model extensional identity in the semantics.
-%% \end{itemize}
-%% In the presheaf model, we interpret the extensional identity in a standard way,
-%% as strict equality of sections of presheaves, and soundness for extensional
-%% identity becomes trivial to show. Hence, the soundness property for our overall
-%% staging pipeline becomes the following: the staging output is definitionally
-%% equal to the identity-erased version of the staging input.
-
-%% In the
-%% current work we use identity types and some inductive types and the
-%% metalanguage, which are not covered in \cite{staged2ltt}. Hence, our main task
-
-%% We give an overview of the semantics of CFTT. This yields a algorithm for
-%% evaluating
-%% extracting object programs from CFTT
-
-%% We give a short summary of the semantics of staging here. We reuse most of the
-%% semantics from \cite{staged2ltt} and only mention additions and differences. The
-%% reader may refer to ibid.\ for technical details.
-
-%% Staging is given by evaluation of CFTT into a presheaf model, where the base
-%% category is given as follows: objects are typing contexts of the object language
-%% and morphisms are parallel substitutions. The equality of morphisms is given by
-%% strict syntactic equality of object terms. $\Ty$ is interpreted as the presheaf
-%% which is constantly the set of object-theoretic types. Soundness of staging is
-%% proven by a logical relation model, internally to the mentioned presheaf
-%% model. Soundness means that the output of staging, viewed as a CFTT term, is
-%% definitionally equal to the input.
-
-%% The main addition in CFTT is the assumption of all inductive families in the
-%% meta-level. However, based on the results in \cite{whynotw}, it is enough to
-%% extend the meta-level syntax in \cite{staged2ltt} with W-types and identity types to
-%% derive all inductive families. We consider the semantics of these in the following.
-
-%% \subsubsection{W-types} W-types have a standard object-wise inductive definition in
-%% the presheaf model \cite{DBLP:journals/apal/MoerdijkP00}. Additionally, we need to give a logical relation
-%% interpretation to W-types, as required for the soundness proof for staging. In
-%% Section 5.2.\ of \cite{staged2ltt} a serialization map is used to define the relation
-%% for natural numbers, however, there is no analogous map for W-types because we
-%% cannot build finite terms from infinitely branching trees.
-
-%% Instead, we give an inductive definition for the logical relation at
-%% W-types. First, assuming $A : \wh{\Set}$ and $B : A \to \wh{\Set}$, we have the
-%% usual constructor for the semantic W-type in the internal language of the
-%% presheaf model, as $\msf{sup} : (a : A) \to (B\,a \to \msf{W}\,A\,B) \to
-%% \msf{W}\,A\,B$.  Second, assuming $A : \Ty\,\Gamma$ and $B : \Ty\,(\Gamma \ext
-%% A)$ in CFTT, we internalize the $\msf{sup}$ constructor which constructs a term
-%% of a syntactic W-type from restricted terms:
-%% \[ \mathbb{R}_{\msf{sup}} : (\alpha : \mathbb{R}\,A\,\gamma) \to \mathbb{R}\,(B \RA (\msf{W_{CFTT}}\,A\,B)[\msf{wk}])\,(\gamma,\,\alpha) \to \mathbb{R}\,(\msf{W_{CFTT}}\,A\,B)\,\gamma \]
-%% Here, $\msf{W_{CFTT}}$ denotes the W-type former in CFTT, $B \RA
-%% (\msf{W_{CFTT}}\,A\,B)[\msf{wk}]$ is a non-dependent function type in CFTT, and
-%% $\msf{wk}$ is a weakening substitution.
-
-%% Then, assuming $\gamma^{\approx} : \Gamma^{\approx}\,\gamma\,\gamma'$, we
-%% inductively define a relation $\msf{W}^{\approx}$ between
-%% $\msf{W}\,(\mbb{E}\,A\,\gamma)\,(\lam \alpha.\,\mbb{E}\,B\,(\gamma,\,\alpha))$
-%% and $\mbb{R}\,(\msf{W_{CFTT}}\,A\,B)\,\gamma'$, which relates
-%% $\msf{sup}\,\alpha\,f$ to $\mathbb{R}_{\msf{sup}}\,\alpha'\,f'$ if $\alpha$ is
-%% related to $\alpha'$ at type $A$ and $f$ is related to $f'$ at type $B[\alpha']
-%% \RA \msf{W_{CFTT}}\,A\,B$. The relation at the non-dependent function type $\blank{\RA}\blank$
-%% is given as pointwise preservation of relations. With this, we can also give
-%% relational interpretation to the syntactic $\msf{sup}$ rule in CFTT,
-%% and we can use induction on $\msf{W}^{\approx}$ to interpret the syntactic
-%% induction rule for W. Thus, we get soundness of staging with W-types.
-
-%% \subsubsection{Identity type} Our treatment of the identity type is
-%% influenced by the following: we would like to assume equations as axioms in
-%% CFTT, which are true in the presheaf model (i.e.\ they are true at
-%% staging-time), without blocking staging as a computation. In Section \ref{sec:concatmap-for-streams}
-%% we use such an axiom to good effect. However, axioms do block computation in
-%% CFTT up to definitional equality, which implies that the soundness of staging
-%% fails. Our solution:
-%% \begin{itemize}
-%% \item Have a standard intensional identity type in CFTT.
-%% \item Before staging, translate the intensional identity to extensional
-%%       identity. This amounts to erasing all identity proofs and transports.
-%% \item Perform staging after erasure of identity proofs, i.e.\ model extensional identity in the semantics.
-%% \end{itemize}
-%% In the presheaf model, we interpret the extensional identity in a standard way,
-%% as strict equality of sections of presheaves, and soundness for extensional
-%% identity becomes trivial to show. Hence, the soundness property for our overall
-%% staging pipeline becomes the following: the staging output is definitionally
-%% equal to the identity-erased version of the staging input.
+Axioms clearly block computation, so they are incompatible with the mentioned
+notion of soundness. But we are only interested in equations which hold
+definitionally during unstaging. Hence, we can do the following: first, we erase
+all identity proofs and their transports from CFTT programs, then we proceed
+with unstaging as in \cite{staged2ltt}. This erasure can be formalzied as a
+syntactic translation \cite{next700} from CFTT to CFTT extended with the
+equality reflection principle \cite{TODO}. Thus, soundness of staging for CFTT
+holds up to erasure of identity proofs.
 
@@
 \vA \to \Up \vB$ is usually preferable to $\Up(\vA \to \vB)$. The two types are
-actually equivalent during staging, up to the runtime equivalence of object programs,
+equivalent during unstaging, up to the runtime equivalence of object programs,
 and we can convert back and forth in CFTT:
@@
 \[ \letdef \vn : \Maybe\,\Nat := \spl(\mdown\,\$\,\mdo \{\vx \fro \return\,\qt{10}; \vy \fro \return\,\qt{20}; \return\,\qt{\vx + \vy})\});\,... \]
-Since $\MaybeT_\M$ is meta-level, its monadic binding fully computes at staging time. Thus, the above code is staged to
+Since $\MaybeT_\M$ is meta-level, its monadic binding fully computes at unstaging time. Thus, the above code is computed to
 \[ \letdef \vn : \Maybe\,\Nat := \Just\,(10 + 20);\;... \]
@@
 replace the current state $\vx$, as an object expression, with the expression
-$\qt{\spl \vx + \spl \vx}$, and this happens at staging time. This behaves as an
+$\qt{\spl \vx + \spl \vx}$, and this happens at unstaging time. This behaves as an
 ``inline'' modification which replaces every subsequent mention of the state
@@
 \[ \mdown \dlr \mdo \{\modify\,(\lam \vx.\,\qt{\spl \vx + \spl \vx}); \modify\,(\lam \vx.\,\qt{\spl \vx + \spl \vx});\return\,\qt{()}\} \]
-is staged to
+is unstaged to
 \[ \qt{\lam \vx.\,((),\,(\vx + \vx) + (\vx + \vx))} \]
@@
 \[ \mdown \dlr \mdo \{\modify'\,(\lam \vx.\,\qt{\spl \vx + \spl \vx}); \modify'\,(\lam \vx.\,\qt{\spl \vx + \spl \vx})\} \]
-is staged to
+is unstaged to
 \[ \qt{\lam \vx.\,\letdef x := \vx + \vx; \letdef \vx := \vx + \vx; ((),\,\vx)} \]
@@
 The $\modify' (\lam \vx. \qt{\spl \vx + 10})$ action gets inlined in both case
-branches during staging. This follows from the definition of monadic binding in
+branches during unstaging. This follows from the definition of monadic binding in
 $\Gen$ and the $\mit{split}$ function in the desugaring of $\mbf{case}$. Code
@@
 
-Let us look at the staging output. We omit newtype wrappers and use nested
+Let us look at the unstaging output. We omit newtype wrappers and use nested
 pattern matching on pairs, but otherwise this is exactly the code that we get in
@@
 The only flaw in this code is the $\msf{joinNothing}$, which is never called,
-becase we joined an action that never throws an error. But this is also
-very easy to clean up after staging.
+becase we joined an action that never throws an error. But this is also easy to
+clean up afterwards.
 \subsection{Discussion}
@@
 using bidirectional elaboration. With such inference, monadic code in CFTT would
-look only modestly more complicated than in Haskell.
+look only slightly more complicated than in Haskell.
 
@@
 the product of states, and also add another $\Maybe\,\vA$ that is used to buffer
-a single $\vA$ value while we are stepping the $\Pull\,(\vA \to \vB)$ stream.
+a single value while we are stepping the $\Pull\,(\vA \to \vB)$ stream.
 The $\IsSOP$ constraints for the new machine states are implicitly dispatched
@@
 \end{alignat*}
-These definitions are standard for streams; compared to the unstaged definitions
-in previous literature, the only additional noise is just the $\Gen$ monad in
-the initial states and the transitions. Likewise, we can give standard
+These definitions are standard for streams; compared to the non-staged
+definitions in previous literature, the only additional noise is just the $\Gen$
+monad in the initial states and the transitions. Likewise, we can give standard
 definitions for usual stream functions such as $\msf{filter}$, $\msf{take}$ or
@@
 possible to look inside quoted expressions. Instead, we use type coercions that
-can fail at staging time. If library users do not violate generativity, these
-coercions disappear and the staging output will not contain unsafe coercions.
+can fail at unstaging time. If library users do not violate generativity, these
+coercions disappear and the code output will not contain unsafe coercions.
 
@@
 \item We could try to lean more heavily on datatype-generic programming and
-     generalize abstractions to arbitrary object types. For example, push and pull
+     generalize abstractions for more object types. For example, push and pull
      streams could be generalized to inductive and coinductive Church-codings of
@@
 \appendices
-\section{Semantics of CFTT}\label{appendix:A}
+\section{Staged Semantics of CFTT}\label{appendix:A}
 
@@
 
+\begin{notation} In the following, we work in a mostly unspecified extensional
+type theory, writing $\Set$ for universes (omitting universe levels), $(x : A) \to B$
+for $\Pi$-types, $(x : A) \times B$ for $\Sigma$-types and $\blank\!=\!\blank$ for
+extensional identity.
+
+\end{notation}
+
 \subsection{Models and Syntax of CFTT}
@@
 We give an algebraic \cite{TODO} specification for the models of CFTT and define
-the syntax of CFTT to be the initial model.
+the syntax of CFTT to be the initial model. In the following, all type and term
+formers are implicitly assumed to be stable under substitution.
 
@@
   of contexts and parallel substitutions. We use $\Con : \Set$ for contexts,
-  $\Mor : \Con \to \Con \to \Set$ for morphisms and $\emptycon : \Con$ for the
+  $\Sub : \Con \to \Con \to \Set$ for morphisms and $\emptycon : \Con$ for the
   terminal object (which is the empty context in the syntax).
@@
   (\Gamma : \Con) \to \MTy_i\,\Gamma \to \Set$ for $i \in \mbb{N}$. We support
-  basic type formers, identity types and W-types. Additionally, we have
-  Coquand-style universes, i.e.\ $\U_i : \MTy_{i+1}\,\Gamma$ such that there is
-  a definitional isomorphism $\msf{MetaTm}_{i+1}\,\Gamma\,\U_i \simeq \MTy_i\,\Gamma$.
-  We use $\El$ and $\code$ to denote the components of this isomorphism.
+  basic type formers, identity types and W-types. We write $\Gamma \ext A$ for
+  an extended context. Additionally, we have Coquand-style universes,
+  i.e.\ $\U_i : \MTy_{i+1}\,\Gamma$ such that there is an isomorphism
+  $\msf{MetaTm}_{i+1}\,\Gamma\,\U_i \simeq \MTy_i\,\Gamma$.  We use $\El$ and
+  $\code$ to denote the components of this isomorphism.
 \item
@@
   and $\msf{C} : \MTm\,\Gamma\,\CTy \to \MTm\,\Gamma\,\Ty$.
-\item For object-level terms, we have $\Tm : \MTm\,\Gamma\,\Ty \to \Set$, closed under
-  all object-level term formers. Here, $\MTm\,\blank\,\Ty$ and $\Tm$ also form a
-  family structure, so that we can substitute object terms and extend the context
-  with object term variables.
+\item For object-level terms, we have $\Tm : \MTm\,\Gamma\,\Ty \to \Set$,
+  supporting context extension, i.e.\ we have $\Gamma,\ext\,A$ for
+  $A : \MTm\,\Gamma\,\Ty$ such that $\Sub\,\Gamma\,(\Delta\,\ext\,A)$ is isomorphic
+  to $(\sigma : \Sub\,\Gamma\,\Delta) \times \Tm\,\Gamma\,(A[\sigma])$. All
+  object-level term formers are supported in $\Tm$.
 \item Staging operations: we have $\Lift : \MTm\,\Gamma\,\Ty \to \MTy\,\Gamma$ together
-  with definitional isomorphisms $\Tm\,\Gamma\,A \simeq \MTm\,\Gamma\,(\Lift A)$.
+  with isomorphisms $\Tm\,\Gamma\,A \simeq \MTm\,\Gamma\,(\Lift A)$.
   We write $\ql\!\blank\!\qr$ and $\spl\!\blank$ for the components of the isomorphism.
@@
 \end{definition}
-This presentation is a bit more verbose than the ``surface'' syntax used in the
+This presentation is a bit more verbose than the surface syntax used in the
 paper. First, the meta level uses explicit coding and decoding operations for
@@
 the informal syntax by having a separate \emph{sort} for object-level terms,
-together with explicit quoting and splicing, but we specify object-level types
-as certain meta-level terms. This a hybrid solution of two styles:
+together with explicit quoting and splicing. However, we do not have a sort for
+object types and instead specify them as terms of metatypes. This is a mixture
+of two styles:
 \begin{itemize}
-\item One would have separate sorts for $\Ty$, $\CTy$ and $\VTy$, and would have
-  separate staging operations for all of these sorts. This follows the ``classic''
-  2LTT presentation \cite{TODO}, where the two levels are specified independently
-  and level-mixing is only possible through the staging operations.
-\item The other one would only have metatypes and terms, no lifting, quoting or
-  splicing, and would use metafunctions to represent object-level binders. This
-  is the HOAS (``higher-order abstract syntax'') or ``logical framework''
-  flavor. Our Agda implementation actually follows this style.
+\item The ``classic'' 2LTT style would have separate sorts and staging
+  operations for all of $\Ty$, $\CTy$ and $\VTy$. Here the two levels are
+  specified independently and level-mixing is only possible through staging
+  operations.
+\item The ``higher-order abstract syntax'' (HOAS) or ``logical framework'' style
+  would only have metatypes and metaterms, no quoting or splicing, and would
+  reuse metafunctions to represent object-level binders.Our Agda implementation
+  actually follows this style.
 \end{itemize}
 In the context of CFTT, the choice between HOAS and 2LTT styles is a matter of
-taste, and the two are inter-derivable.
+taste, and the two are inter-derivable \cite{TODO}.
 
@@
 \item
-  $\Tm\,\Gamma\,A$ is simply defined to be $\MTy\,\Gamma\,(\Lift A)$ in the model, which
-  means that quotation and splicing are both identity maps.
+  $\Tm\,\Gamma\,A$ is simply defined to be $\MTy\,\Gamma\,(\Lift A)$ in the model. Hence,
+  quotation and splicing are both interpreted as identity maps.
 \end{itemize}
@@
 
-\subsection{Soundness for W-types}
+\subsection{W-types}
 So far we have stayed close to \cite{staged2ltt}, with only cosmetic
@@
 the biggest extension here. We extend Section 5 of \cite{staged2ltt} here. In
-the syntax of CFTT, we have
-\begin{alignat*}{3}
-  &\vW   &&: (A : \MTy_i\,\Gamma) \to \MTy_j\,(\Gamma,\,A) \to \MTy_{\msf{max}(i,j)}\,\Gamma\\
-  &\vsup &&: (a : \MTm\,\Gamma\,A) \to \MTm\,(\Gamma,B[\vid,a])\,(\vW\,A\,B) \to \MTm\,\Gamma\,(\vW\,A\,B),
-\end{alignat*}
-where $\vid$ is the identity substitution from $\Gamma$ to $\Gamma$. We also have the eliminator for $\vW$
-whose specification we omit here.
-
-Internally to $\whset$, i.e.\ internally to presheaves over $\mbbo$, we also
-have the standard W-type, defined pointwise as the W-type in sets:
+the syntax of CFTT, $\vW$-types are as follows. The formal specification of
+syntactic $\vW{-}\msf{elim}$ is rather gnarly, because of the usage of explicit
+substitutions, but we write it out here anways.
+
+\begin{notation} We use $\p^N$ for $N$-fold
+weakening and we also use numerals as De Bruijn indices, to abbreviate
+$\q[\p^N]$. We write $\msf{app}$ and $\msf{lam}$ for $\Pi$ application and
+abstraction. We write $\blank{\Rightarrow}\blank$ for non-dependent function
+types. Finally, for the sake of brevity we omit the component maps of the isomorphism
+$\re_\ext : \re\,(\Gamma \ext A) \simeq ((\gamma : \re\,\Gamma) \times \re\,A\,\gamma)$.
+\end{notation}
+\begin{alignat*}{3}
+  &\vW   &&: (A : \MTy_i\,\Gamma) \to \MTy_j\,(\Gamma\,\ext\,A) \to \MTy_{\msf{max}(i,j)}\,\Gamma\\
+  &\vsup &&: (a : \MTm\,\Gamma\,A) \to \MTm\,\Gamma\,(B[\vid,\,a] \Rightarrow \vW\,A\,B) \to \MTm\,\Gamma\,(\vW\,A\,B)\\
+  &\vW{-}\msf{elim}  &&: (P : \MTy\,(\Gamma\,\ext\,\vW\,A\,B))\\
+  &                  && \hspace{0.6em}\to \MTm\,(\Gamma\,\ext\,A\,\ext\,B \Rightarrow (\vW\,A\,B)[p]\,\ext\,\Pi\,(B[p])\,(P[\p^3,\,\msf{app}\,1\,0]))\,(P[\p^3,\,\vsup\,2\,1))\\
+  &                  && \hspace{0.6em}\to (w : \MTm\,\Gamma\,(\vW\,A\,B)) \to \MTm\,\Gamma\,(P[\id,\,w])\\
+  & \vW{\beta}       &&:
+    \vW{-}\msf{elim}\,P\,s\,(\vsup\,a\,f) = s[\id,\,a,\,f,\,\msf{lam}\,(\vW{-}\msf{elim}\,(P[\p^2,\,\q])\,(s[\p])\,(\msf{app}\,f\,0)]
+\end{alignat*}
+We shall omit the $i$ and $j$ universe levels in the following. In $\whset$,
+i.e.\ internally to presheaves over $\mbbo$, we also have the standard W-type.
 \begin{alignat*}{3}
   & \wh{\vW}   &&: (A : \whset) \to (A \to \whset) \to \whset\\
-  & \wh{\vsup} &&: (a : A) \to (B\,a \to \wh{\vW}\,A\,B) \to \wh{\vW}\,A\,B
-\end{alignat*}
-In the presheaf model we can evidently interpret $\vW$ using $\whW$.
-
-%% The \emph{restriction} of this
-
-%% \subsubsection{Stability} Stability means that if we take any object-theoretic term,
-%% embed it as a CFTT term, then unstage it
+  & \wh{\vsup} &&: (a : A) \to (B\,a \to \wh{\vW}\,A\,B) \to \wh{\vW}\,A\,B\\
+  & \wh{\vW{-}\msf{elim}} &&: (P : \wh{\vW}\,A\,B \to \whset)\\
+  & && \hspace{0.6em}\to ((a : A)(f : B\,a \to \wh{\vW}\,A\,B) \to ((b : B\,a) \to P\,(f\,b)) \to P\,(\wh{\vsup}\,a\,f))\\
+  & && \hspace{0.6em}\to (w : \wh{\vW}\,A\,B) \to P\,w\\
+  & \wh{\vW}{\beta} &&: \wh{\vW{-}\msf{elim}}\,P\,s\,(\wh{\vsup}\,a\,f) = s\,a\,f\,(\lambda\,b.\,\wh{\vW{-}\msf{elim}}\,P\,s\,(f\,b))
+\end{alignat*}
+In the presheaf model we interpret $\vW$ using $\whW$.
+\begin{alignat*}{3}
+  & \ev\,(\vW\,A\,B)\,\gamma   &&:= \wh{\vW}\,(\ev\,A\,\gamma)\,(\lambda\,\alpha.\,\ev\,B\,(\gamma,\,\alpha))\\
+  & \ev\,(\vsup\,a\,f)\,\gamma &&:= \wh{\vsup}\,(\ev\,a\,\gamma)\,(\lambda\,\beta.\,\ev\,f\,(\gamma, \beta))\\
+  & \ev\,(\vW{-}\msf{elim}\,P\,s\,w)\,\gamma &&:= \wh{\vW{-}\msf{elim}}\,(\lambda\,w.\,\ev\,P\,(\gamma,\,w))\,(\lambda\,a\,f\,g.\,\ev\,s\,(\gamma,\,a,\,f,\,g))\,(\ev\,w\,\gamma)
+\end{alignat*}
+
+%% W-elim̂ (λ w. E P (γ, w)) (λ a f g. E s (γ, a, f, g)) (E w γ)
+Now, assuming $A : \MTy\,\Gamma$ and $B : \MTy\,(\Gamma\,\ext\,A)$, we consider
+$\re\,(\vW\,A\,B) : (\gamma : \re\,\Gamma) \to \whset$. Externally, the elements
+of $\re\,(\vW\,A\,B)\,\gamma$ are CFTT terms with a $\vW$-type in purely
+object-level contexts. We call such terms \emph{restricted}. Now, we internalize
+the restriction of the $\vsup$ constructor:
+\[ \re_{\vsup} : (\alpha : \re\,A\,\gamma) \to \re\,(B \Rightarrow (W\,A\,B)[\p])\,(\gamma,\,\alpha) \to \re\,(W\,A\,B)\,\gamma \]
+Externally, $\re_{\vsup}$ applies $\vsup$ to restricted terms. Note that in
+$\re\,(B \Rightarrow (W\,A\,B)[\p])\,(\gamma,\,\alpha)$, the type $B \Rightarrow (W\,A\,B)[\p]$ is in
+the context $(\Gamma\,\ext\,A)$, which makes it possible to pass $(\gamma,\,\alpha)$ as
+argument; this ``instantiates'' $B$ to $\alpha$. With this, we can also
+show that $\re$ preserves $\vsup$:
+\[ \re\,(\vsup\,t\,u)\,\gamma = \re_\vsup\,(\re\,t\,\gamma)\,(\re\,u\,\gamma) \]
+Note that the type of $\re\,u\,\gamma$ is $\re\,(B[\id,\,t] \Rightarrow
+\vW\,A\,B)\,\gamma$, which is equal to $\re\,((B \Rightarrow
+(\vW\,A\,B)[\p])[\id,\,t])\,\gamma$, which is in turn equal to $\re\,(B
+\Rightarrow (\vW\,A\,B)[\p])\,(\gamma,\,\re\,t\,\gamma)$ by $\re$'s preservation
+of substitutions, which makes the above equation well-typed.
+
+Next, we give a logical relation interpretation for $\vW$. We define an
+inductive relation in $\whset$ as follows.
+\begin{alignat*}{3}
+  & \vW\rel &&: \Gamma\rel\,\gamma\,\gamma' \to \ev\,(\vW\,A\,B)\,\gamma \to \re\,(\vW\,A\,B)\,\gamma' \to \whset\\
+  & \vsup\rel &&: \hspace{0.8em}\{\alpha : \ev\,A\,\gamma\}\{\alpha' : \re\,A\,\gamma'\}(\alpha\rel : A\rel \gamma\rel\,\alpha\,\alpha')\\
+  &           &&\to \{f \hspace{0.5em}: \ev\,B\,(\gamma,\,\alpha) \to \ev\,(\vW\,A\,B)\,\gamma\}\\
+  &           &&\to \{f' \hspace{0.2em}: \re\,(B \Rightarrow (\vW\,A\,B)[\p])\,(\gamma',\,\alpha')\}\\
+  &           &&\to (f\rel : (B \Rightarrow (\vW\,A\,B)[\p])\rel\,(\gamma\rel,\,\alpha\rel)\,f\,f')\\
+  &           &&\to \vW\rel\,\gamma\rel\,(\wh{\vsup}\,\alpha\,f)\,(\re_\vsup\,\alpha'\,f')
+\end{alignat*}
+Here, $(B \Rightarrow (\vW\,A\,B)[\p])\rel$ is the relational interpretation of the function type, which unfolds to
+pointwise preservation of relations. In short, $\vW\rel$ expresses that $\wh{\vsup}\,\alpha\,f$ is related $\re_\vsup\,\alpha'\,f'$
+when the arguments (inductively) related. We can immediately interpret $\vW$ and $\vsup$:
+\begin{alignat*}{3}
+  & (\vW\,A\,B)\rel\,\gamma\rel\,t\,t' &&:= \vW\rel\,\gamma\rel\,t\,t'\\
+  & (\vsup\,t\,u)\rel\,\gamma\rel      &&:= \vsup\rel\,(t\rel\,\gamma\rel)\,(u\rel\,\gamma\rel)
+\end{alignat*}
+It only remains to interpret $\vW$-elimination and its $\beta$-rule. We need to define
+\[ (\vW{-}\msf{elim}\,P\,s\,w)\rel\,\gamma\rel : P\rel\,(\gamma\rel,\,w\rel\,\gamma\rel)\,(\ev\,(\welim\,P\,s\,w)\,\gamma)\,(\re\,(\welim\,P\,s\,w)\,\gamma') \]
+We do this by induction on $w\rel\,\gamma\rel :
+\vW\rel\,\gamma\rel\,(\ev\,w\,\gamma)\,(\re\,w\,\gamma')$. At this point it
+becomes rather tedious to compute and unfold the involved types, so we only give
+a compact definition here together with some informal explanation:
+\begin{alignat*}{3}
+  &(\vW{-}\msf{elim}\,P\,s\,w)\rel\,\gamma\rel :=\\
+  &\hspace{4em}\vW\rel{-}\msf{elim}\,(\lambda\,w\rel.\,\,P\rel\,(\gamma\rel,\,w\rel)\,(\ev\,(\welim\,P\,s\,w)\,\gamma)\,(\re\,(\welim\,P\,s\,w)\,\gamma'))\\
+  &\hspace{8.1em}(\lambda\,a\rel\,f\rel\,\mathit{hyp}.\,\,s\rel\,(\gamma\rel,\,a\rel,\,f\rel,\,\mathit{hyp}))\\
+  &\hspace{8.1em}(w\rel\,\gamma\rel)
+\end{alignat*}
+$\vW\rel{-}\msf{elim}$ is the eliminator for $\vW\rel$. Its first argument is
+the induction motive, which matches the goal type. The second argument is the
+method for $\vsup\rel$; its $a\rel$ and $f\rel$ inputs are the relatedness
+witnesses stored in a $\vsup\rel$, while $\mathit{hyp}$ is induction hypothesis,
+witnessing the inductive motive for every output of $f\rel$. In the body here,
+we need to witness the motive in the case where $w$ is a $\vsup$, which causes
+both $\ev\,(\welim\,P\,s\,w)\,\gamma)$ and $\re\,(\welim\,P\,s\,w)\,\gamma')$ to
+compute further to applications of $\ev\,s\,\gamma$ and $\re\,s\,\gamma'$
+respectively. Hence, we use $s\rel$ to relate the two sides. The
+$\vW{\beta}$-rule also evidently holds with this definition.
+
+\subsection{Identity type}
 

### 2024-06-07 newpaper/paper.tex
after appendix finish

@@
 \newcommand{\U}{\mathsf{U}}
+\newcommand{\Id}{\msf{Id}}
 \newcommand{\p}{\mathsf{p}}
@@
 \newcommand{\Tmo}{\msf{Tm}_{\mbbo}}
+\newcommand{\Cono}{\msf{Con}_{\mbbo}}
+\newcommand{\Subo}{\msf{Sub}_{\mbbo}}
 \newcommand{\whset}{\wh{\Set}}
@@
 \appendices
-\section{Staged Semantics of CFTT}\label{appendix:A}
+\section{Staging in CFTT}\label{appendix:A}
 
@@
 compute further to applications of $\ev\,s\,\gamma$ and $\re\,s\,\gamma'$
-respectively. Hence, we use $s\rel$ to relate the two sides. The
-$\vW{\beta}$-rule also evidently holds with this definition.
+respectively. Hence, we use $s\rel$ to relate the two sides. This definition
+also respects $\vW\beta$, by the $\beta$ rule for $\vW\rel$.
 
 \subsection{Identity type}
+
+In this section we assume an \emph{extensional} identity type in CFTT.  As we
+mentioned in Section \ref{TODO}, we can show soundness of staging with
+extensional identity and certain \emph{postulated} identities, but we cannot do
+the same with intensional identity. Hence, we formulate unstaging as consisting
+of two steps: first, a translation of intensional identity to extensional (which
+amounts to proof erasure), followed by evaluation in the presheaf model.
+
+Extensional identity in CFTT type is interpreted as the standard extensional
+identity in presheaves, i.e.\ we have $\ev\,(\msf{Id}\,A\,t\,u)\,\gamma :=
+\ev\,t\,\gamma = \ev\,u\,\gamma$. This supports $\msf{refl}$, transport,
+uniqueness of identity and equality reflection; we omit the fairly trivial
+definitions here. We choose the relational interpretation to be trivial as well,
+i.e.\ $(\msf{Id}\,t\,u)\rel\,\gamma\rel\,p\,p' := \top$.
+
+This is because both $\ev\,(\Id\,t\,u)\,\gamma$ and $\re\,(\Id\,t\,u)\,\gamma'$
+are mere propositions --- the latter because of equality reflection in the
+syntax.  In particular, if we have a witness of $\re\,(\Id\,t\,u)\,\gamma'$, we
+also have $\re\,t\,\gamma' = \re\,u\,\gamma'$ by equality reflection. Thus, the
+rest of the relational interpretation of $\Id$ is also trivial. Another way to
+think about this is that soundness expresses a form of canonicity, but if we
+have equality reflection then syntactic identities are always canonical.
+
+\subsection{Generativity}
+
+The generativity axiom from Section \ref{TODO} has the following type, in internal
+CFTT syntax:
+\[ (A : \U_\msf{P})(f : \Elprod\,A \to \USOP)(x\,y : \Elprod\,A) \to f\,x = f\,y  \]
+In the presheaf model, we have the following:
+\begin{itemize}
+\item $\wh{\Uprod}$ is the set (constant presheaf) of lists of object-theoretic types.
+\item $\wh{\USOP}$ is the set of lists of lists of object-theoretic types.
+\item $\wh{\Elprod\,A}$ is the presheaf such that $\wh{\Elprod\,A}\,\Gamma$ is
+  the set of products of object-theoretic terms in context $\Gamma$, with types
+  of components taken from $A$.
+\end{itemize}
+Unfolding the type of generativity in the presheaf model, it suffices to show
+that assuming
+\begin{itemize}
+\item $A : \wh{\Uprod}$, $\Gamma : \Cono$, $x,\,y : \wh{\Elprod\,A}\,\Gamma$
+\item and $f : (\Delta : \Cono) \to \Subo\,\Delta\,\Gamma \to \wh{\Elprod\,A}\,\Delta \to \wh{\USOP}$ such that $f$ is natural with respect to object-theoretic substitution,
+\end{itemize}
+we have that $f\,\Gamma\,\id\,x = f\,\Gamma\,\id\,y$.
+
+Now, $\wh{\Elprod\,A}$ is a representable presheaf since it is a finite product
+of object-theoretic terms. Also, since $f$ is a natural transformation from the
+representable presheaf $\Subo\blank\Gamma \times \wh{\Elprod\,A}$ to the
+constant presheaf $\wh{\USOP}$, the Yoneda lemma implies that $f$ is constant and
+hence the generativity axiom is validated.
+
+Since the logical relation interpretation of identity is trivial, it is also trivial
+for the generativity axiom.
+
+\subsection{Summary}
+
+By the presheaf and logical relations interpretation of CFTT as shown in this appendix,
+we obtain \emph{unstaging} and \emph{soundness} of unstaging, following \cite{staged2ltt}.
+
+\emph{Completeness} of unstaging is shown by straightforward induction on the
+syntax of the object theory, the same way as in \cite{staged2ltt}, and we omit
+it here.
 

### 2024-06-10 newpaper/paper.tex
wip

@@
 \begin{alignat*}{3}
-  &\vf = \lam \vx.\,(\bind)\,\msf{MonadReaderDict}\,(\msf{ask}\,\msf{MonadReaderDict})\,(\lam \vb.\,\bif \vb\\
-  &\ind \bthen\return\,\msf{MonadReaderDict}\,(\vx + 10)\\
-  &\ind \belse\hspace{0.25em}\return\,\msf{MonadReaderDict}\,(\vx + 20))
-\end{alignat*}
-Here, $\msf{MonadReaderDict}$ is a runtime dictionary, containing the methods of
-the $\Monad$ instance for $\Reader$, and $(\bind)\,\msf{MonadReaderDict}$ is the
-projection of the binding function. Here, a runtime closure will be created for
-the $\lam \vb.\,...$ function, and $(\bind)$, $\msf{ask}$ and $\mbf{return}$
-will create additional dynamic closures.
+  &\vf = \lam \vx.\,(\bind)\,\msf{monadReader}\,(\msf{ask}\,\msf{monadReaderReader})\,(\lam \vb.\,\bif \vb\\
+  &\ind \bthen\return\,\msf{monadReader}\,(\vx + 10)\\
+  &\ind \belse\hspace{0.25em}\return\,\msf{monadReader}\,(\vx + 20))
+\end{alignat*}
+Here, $\msf{monadReader}$ and $\msf{monadReaderReader}$ are runtime
+dictionaries, respectively for the $\msf{Monad}$ and $\msf{MonadReader}$
+instances, and, for example, $(\bind)\,\msf{monadReader}$ is a field projection
+from the dictionary. We get a runtime closure from the $\lam \vb.\,...$
+function, and $(\bind)$, $\msf{ask}$ and $\mbf{return}$ also produce additional
+closures.
 
@@
 \end{alignat*}
-This is a third-order and second-rank polymorphic function in disguise, because
+This is a third-order rank-2 polymorphic function in disguise, because
 its monad dictionary argument contains the polymorphic second-order method
@@
 In this paper we use staged compilation to address issues of robustness. The
-idea is to shift as much as possible work from general-purpose optimization to
+idea is to shift as much work as possible from general-purpose optimization to
 metaprograms.
@@
   operational semantics of the object language.
-\item In Section \ref{sec:monad-transformers} we build a monad transformer library. We believe
-  that this is a good demonstration, because monads and monad transformers are
-  the most widely used effect system in Haskell, and at the same time their
-  compilation to efficient code can be surprisingly difficult. The main idea is
-  to have monads and monad transformers \emph{only} in the metalanguage, and try
-  to express as much as possible at the meta level, and funnel the object-meta
-  interactions through specific \emph{binding-time improvements}.  We show how
-  almost all definitions from the Haskell ecosystem of monad transformers can be
-  reused in the meta level of CFTT. We also develop let-insertion, join points
-  and case switching on object-level data.
+\item
+  In Section \ref{sec:monad-transformers} we build a staged library for monad
+  transformers \cite{DBLP:conf/popl/LiangHJ95}. We believe that this is a good
+  demonstration, because monads and monad transformers are the most widely used
+  effect system in Haskell, and at the same time their compilation to efficient
+  code can be surprisingly difficult. The continuation monad is well-known in
+  staged compilation \cite{DBLP:conf/lfp/Bondorf92}, and staged state monads
+  have also been used
+  \cite{DBLP:conf/pepm/SwadiTKP06,DBLP:conf/emsoft/KiselyovST04,DBLP:journals/scp/CaretteK11}.
+  These works used specific monads as tools for domain-specific code
+  generation. In contrast, we propose ubiquitous staging for general-purpose
+  monadic programming, where users can write code that looks similar to monadic
+  code in Haskell but with deterministic and robust compilation to efficient
+  code.
 \item In Section \ref{sec:stream-fusion} we build a pull-based stream fusion library. Here,
@@
 $\bs{\Ty}$ is the universe of types in the object language. It is itself a
-metatype, so so we have $\Ty : \MTy$. All construction and elimination rules of
+metatype, so we have $\Ty : \MTy$. All construction and elimination rules of
 type formers in $\Ty$ stay within $\Ty$. We further split $\Ty$ to two
@@
 With what we have seen so far, there is no interaction between the meta and
-object levels. We make such interaction possible with \emph{staging primitives}.
+object levels. We make such interaction possible with the following primitives.
 
@@
 
-A CFTT program is a mixture of object-level and meta-level top-level definitions
-and declarations. We use \emph{unstaging} to refer to the process of evaluating
-metaprograms in a CFTT program in order to extract purely object-level code.
-The output is guaranteed to be well-typed.
+We use \textbf{unstaging} to refer to the process of extracting object code from
+CFTT programs, by evaluating all metaprograms in splices. This term has been
+sporadically used in the literature, e.g.\ by Odersky and Rompf
+\cite{DBLP:journals/cacm/RompfO12} and also by Choi et
+al.\ \cite{DBLP:conf/popl/ChoiAYT11} with a somewhat different meaning in a
+multi-stage context. The main precursor to this paper used ``staging'' instead
+of our ``unstaging'' \cite{staged2ltt}; we use the latter because the former
+conflicts with other common usages of ``staging''.
 
@@
 
+\subsubsection{Comparison to defunctionalization}
+Defunctionalization is a program translation which represents higher-order
+functions using only first-order functions and inductive data
+\cite{DBLP:journals/lisp/Reynolds98a,DBLP:conf/ppdp/DanvyN01}. The idea is to
+represent each closure as a constructor of an inductive type, and implement
+closure application as a top-level first-order function which switches on all
+possible closure constructors. Hence, like our unstaging algorithm,
+defunctionalization produces first-order code from higher-order code. We summarize
+the differences.
+\begin{itemize}
+\item Defunctionalization mainly supports \emph{compiler optimizations}, by making
+  closures transparent to analysis. It is not an optimization by itself; it does
+  not change the amount of dynamic control flow or dynamic allocations in a
+  program. A dynamic closure call becomes a dynamic case-switch and a
+  heap-allocated ``functional'' closure becomes a heap-allocated inductive
+  constructor.
+\item Staging mainly supports optimization by \emph{programmers} by providing
+  control over code generation. Unstaging does not translate higher-order
+  functions to anything in particular. Instead, it is the execution of
+  higher-order metaprograms that produce first-order programs.
+\end{itemize}
+
 \subsection{Staged Semantics of CFTT}\label{sec:semantics-of-staging}
@@
 We give an overview of the unstaging algorithm together with its correctness.
-We reuse the concepts and most of the results from \cite{staged2ltt}. Our main
-task is to extend ibid.\,with identity types and inductive types in the
-metalanguage, since we use them in this paper. For inductive types, it is the
-most sensible to add W-types, since all inductive families can be faithfully
-derived from them in our setting \cite{whynotw}.
+Our main task is to extend \cite{staged2ltt} with identity types and inductive
+types in the metalanguage, since we use them in this paper. For inductive types,
+it is the most sensible to add W-types, since all inductive families can be
+faithfully derived from them in our setting \cite{whynotw}.
 
@@
 \cite{staged2ltt}, so we relegate it to Appendix \ref{appendix:A}. Here we only
-review the basics.
+give an overview.
 \begin{itemize}
@@
 syntactic translation \cite{next700} from CFTT to CFTT extended with the
-equality reflection principle \cite{TODO}. Thus, soundness of staging for CFTT
-holds up to erasure of identity proofs.
+equality reflection principle, which says that propositional equality implies
+definitional equality (see e.g.\ \cite{hofmann95conservativity}). Thus,
+soundness of staging for CFTT holds up to erasure of identity proofs.
 
@@
 In this section we build a library for monads and monad transformers. We believe
-that this is a good demonstration of CFTT-s abilities, since monads are
+that this is a good demonstration of CFTT's abilities, since monads are
 ubiquitous in Haskell programming and they also introduce a great amount of
@@
 
-We start with some preparatory work before getting to monads. We saw that $\Up
-\vA \to \Up \vB$ is usually preferable to $\Up(\vA \to \vB)$. The two types are
-equivalent during unstaging, up to the runtime equivalence of object programs,
-and we can convert back and forth in CFTT:
+We do a preliminary overview before getting to monads. In the staged compilation
+and partial evaluation literature, the term \emph{binding time improvement} is
+used to refer to such conversions, where the ``improved'' version supports more
+compile-time computation \cite[Chapter 12]{partial-evaluation}.
+
+Translation from $\Up \vA \to \Up \vB$ to $\Up(\vA \to \vB)$ is a basic
+binding-time improvement, as an $\eta$-expansion \cite{eta-expansion-trick}. The
+two types are equivalent during unstaging, up to the runtime equivalence of
+object programs, and we can convert back and forth in CFTT, the same way as in
+MetaML \cite[Section 9]{metaml}:
 \begin{alignat*}{3}
@@
 
-In the staged compilation and partial evaluation literature, the term
-\emph{binding time improvement} is used to refer to such conversions, where the
-``improved'' version supports more compile-time computation
-\cite{partial-evaluation}. A general strategy for generating efficient ``fused''
+ A general strategy for generating efficient ``fused''
 programs, is to try to work as much as possible with improved representations,
@@
 let-insertion in $\mup$ because the return type is in $\MTy$, and we cannot
-introduce object binders in meta-level code.  Fortunately, there is a principled
-solution.
+introduce object binders in meta-level code.
 
@@
 
-The solution is to use a monad which extends $\MTy$ with the ability to
-freely generate object-level code and introduce object binders.
+A principled solution to the previous issue is to write code generators in
+continuation-passing style, as first proposed by Bondorf
+\cite{DBLP:conf/lfp/Bondorf92}. This can be structured as a continuation monad,
+and later work used explicit monadic notation for it
+\cite{DBLP:conf/emsoft/KiselyovST04,DBLP:journals/scp/CaretteK11,DBLP:conf/pepm/SwadiTKP06}. Our
+definition is as follows:
 \begin{alignat*}{3}
@@
 \end{alignat*}
+In \cite{DBLP:conf/emsoft/KiselyovST04},\cite{DBLP:journals/scp/CaretteK11} and
+\cite{DBLP:conf/pepm/SwadiTKP06}, the answer type is a parameter to the
+continuation monad, while we have it as polymorphic. We have found that
+polymorphic answer types are more convenient, because we do not have to
+anticipate the type of the code output when are defining a code generator.
+Hence, we can ``run'' actions as follows:
+\begin{alignat*}{3}
+  &\runGen : \Gen\,(\Up \vA) \to \Up \vA\\
+  &\runGen\,\vma = \unGen\,\vma\,\id
+\end{alignat*}
 From now on, we reuse Haskell-style type classes and $\mbf{do}$-notation in
@@
 how they work. However, type classes are used in essentially the same way in the
-Agda and Haskell implementations, with modest technical differences.
-
-From the $\Monad$ instance, the $\Functor$ and $\Applicative$ instances can be
-also derived. We reuse $({\fmap})$ and $({\ap})$ for applicative notation as well.
-An inhabitant of $\Gen\,\vA$ can be viewed as an action whose effect is to produce
-some object code. $\Gen$ can be only ``run'' when the result type is an object
-type:
-\begin{alignat*}{3}
-  &\runGen : \Gen\,(\Up \vA) \to \Up \vA\\
-  &\runGen\,\vma = \unGen\,\vma\,\id
-\end{alignat*}
+Agda and Haskell implementations, with modest technical differences.  From the
+$\Monad$ instance, the $\Functor$ and $\Applicative$ instances can be also
+derived. We reuse $({\fmap})$ and $({\ap})$ for applicative notation as well.
+
 We can let-bind object expressions in $\Gen$:
@@
 \end{alignat*}
-Now, using do-notation, we may write $\mdo \{\vx \fro \gen\,\qt{10 + 10}; \vy \fro
+Now, using $\mbf{do}$-notation, we may write $\mdo \{\vx \fro \gen\,\qt{10 + 10}; \vy \fro
 \gen\,\qt{20 + 20}\};\return \qt{\vx + \vy}\}$, for a $\Gen\,(\Up \Nat)$
@@
 \emph{eventually} all metaprograms will be used for the purpose of code
-generation, so we eventually $\runGen$ all of our actions. So why not
-just always work in $\Gen$? The implicit nature of $\Gen$ may make it harder to
+generation, so we eventually $\runGen$ all of our actions. So why not just
+always work in $\Gen$? The implicit nature of $\Gen$ may make it harder to
 reason about the size and content of generated code. This is a bit similar to
-the $\msf{IO}$ monad in Haskell, where eventually everything needs to run in
-$\msf{IO}$, but we may not want to write all of our code in $\msf{IO}$.
+the $\msf{IO}$ monad in Haskell \cite{jones2001tackling}, where eventually
+everything needs to run in $\msf{IO}$, but we may not want to write all of our
+code in $\msf{IO}$.
 
@@
 metafunction. However, we cannot directly fashion a monad out of $\Maybe$, since
-we do not have enough type formers in $\VTy$. We could try to use the following
-type for binding:
+we do not have enough type formers in $\VTy$. In particular, we cannot store
+functions inside $\Maybe$ which rules out the applicative $\ap$ operation.  We
+could try to use the following type for binding:
 \[ (\bind) : \Up(\Maybe\,\vA) \to (\Up\vA \to \Up(\Maybe\,\vB)) \to \Up(\Maybe\,\vB) \]
@@
 problematic, because we cannot directly compute metaprograms from object-level
-case splits. Fortunately, with a little bit more work, this is actually possible
-in any $\MonadGen$, for any value type.
+case splits. Fortunately, with a little bit more work, this is possible in any
+$\MonadGen$, for any value type, using a variation of the so-called ``trick''
+\cite{eta-expansion-trick}.
 
@@
 \begin{alignat*}{3}
-  & \hspace{-2em}\mathrlap{\newtype \Identity\,\vA := \Identity\,\{\runIdentity : \vA\} }\\
-  & \hspace{-2em}\mathrlap{\instance \Improve\,\Identity\,\Gen\,\where}\\
-  & \hspace{-2em}\ind \mup\,\vx   &&= \return \qt{\runIdentity\,\spl \vx}\\
-  & \hspace{-2em}\ind \mdown\,\vx &&= \unGen\,\vx \dlr \lam \va.\,\qt{\Identity\,\spl \va}
+  & \hspace{-4em}\mathrlap{\newtype \Identity\,\vA := \Identity\,\{\runIdentity : \vA\} }\\
+  & \hspace{-4em}\mathrlap{\instance \Improve\,\Identity\,\Gen\,\where}\\
+  & \hspace{-4em}\ind \mup\,\vx   &&= \return \qt{\runIdentity\,\spl \vx}\\
+  & \hspace{-4em}\ind \mdown\,\vx &&= \unGen\,\vx \dlr \lam \va.\,\qt{\Identity\,\spl \va}
 \end{alignat*}
@@
 \[ \qt{\lam \vx.\,((),\,(\vx + \vx) + (\vx + \vx))} \]
-which duplicates the evaluation of $\vx + \vx$. The solution is to force the
-evaluation of the new state in the object language, by let-insertion. A similar
-phenomenon happens with the $\mit{local}$ function in $\Reader$. So we define
-``stricter'' versions of these operations. We also return $\Up ()$ from
-actions instead of $()$ --- the former is more convenient, because the $\mdown$
-operation can be immediately used on it.
+which duplicates the evaluation of $\vx + \vx$. The duplication can be avoided
+by let-binding the result in the object language. A similar phenomenon happens
+with the $\mit{local}$ function in $\Reader$. So we define ``stricter'' versions
+of these operations. We also return $\Up ()$ from actions instead of $()$ ---
+the former is more convenient, because the $\mdown$ operation can be immediately
+used on it.
 \begingroup
@@
 The $\modify' (\lam \vx. \qt{\spl \vx + 10})$ action gets inlined in both case
-branches during unstaging. This follows from the definition of monadic binding in
-$\Gen$ and the $\mit{split}$ function in the desugaring of $\mbf{case}$. Code
-generation is continued in both branches with the same action. If we have
-multiple Boolean $\mbf{case}$ splits sequenced after each other, that yields
-exponential code size. Right now, we can fix this by let-binding the problematic
-action:
+branches during unstaging, yielding the following:
+\begin{alignat*}{3}
+   & boooo
+\end{alignat*}
+The duplication follows from the definition of monadic binding in $\Gen$ and the
+$\mit{split}$ function in the desugaring of $\mbf{case}$. Code generation is
+continued in both branches with the same action. If we have multiple Boolean
+$\mbf{case}$ splits sequenced after each other, that yields exponential code
+size. Right now, we can fix this by let-binding the problematic action:
+
 %% \begin{alignat*}{5}
@@
 
-\emph{Completeness} of unstaging is shown by straightforward induction on the
-syntax of the object theory, the same way as in \cite{staged2ltt}, and we omit
-it here.
+\emph{Strictness} of unstaging holds trivially, since there are no $\beta$ and
+$\eta$ rules in the object theory. \emph{Completeness} of unstaging is shown by
+straightforward induction on the syntax of the object theory, the same way as in
+\cite{staged2ltt}, and we omit it here.
 

### 2024-06-10 newpaper/paper.tex
wip

@@
 
-Let us start with the $\Maybe$ monad. We have $\data \Maybe\,\vA :=
+First, following the style of Haskell's monad transformer library \texttt{mtl}
+\cite{mtl}, we define a class for monads that support code generation.
+\begin{alignat*}{3}
+&\tyclass\,\Monad\,\vM \RA \MonadGen\,\vM\,\where\\
+&\ind \liftGen : \Gen\,\vA \to \M\,\vA
+\end{alignat*}
+We also redefine $\gen$ and $\genRec$ to work in any $\MonadGen$, so from now on
+we have:
+\begin{alignat*}{3}
+ &\gen   &&: \MonadGen\,\vM \RA \Up \vA \to \vM\,(\Up \vA)\\
+ &\genRec &&: \MonadGen\,\vM \RA (\Up \vA \to \Up \vA) \to \vM\,(\Up \vA)
+\end{alignat*}
+In the rest of this paper, we shall only mention $\ReaderT$, $\MaybeT$ and
+$\StateT$ as monad transformers. For all of these, we can define the $\MonadGen$
+instance simply by $\liftGen = \lift \circ \liftGen$.
+
+Let us look at the $\Maybe$ monad now. We have $\data \Maybe\,\vA :=
 \Nothing\,|\,\Just\,\vA$, and $\Maybe$ itself is available as a $\VTy \to \VTy$
 metafunction. However, we cannot directly fashion a monad out of $\Maybe$, since
-we do not have enough type formers in $\VTy$. In particular, we cannot store
-functions inside $\Maybe$ which rules out the applicative $\ap$ operation.  We
-could try to use the following type for binding:
+we do not have polymorphism in $\VTy$, nor can we store functions inside
+$\Maybe$. We could try to use the following type for binding:
 \[ (\bind) : \Up(\Maybe\,\vA) \to (\Up\vA \to \Up(\Maybe\,\vB)) \to \Up(\Maybe\,\vB) \]
-This works, but the definition necessarily uses runtime case splits on $\Maybe$
+This would work, but the definition would necessarily use runtime case splits on $\Maybe$
 values, many of which could be optimized away during staging. Also, not having
@@
 \[ \letdef \vn : \Maybe\,\Nat := \Just\,(10 + 20);\;... \]
-Assume also a $\lift : \Monad\,\vM \RA \vM\,\vA \to \MaybeT_\M\,\vM\,\vA$ operation which
-comes from $\MaybeT_\M$ being a monad transformer. We can do let-insertion in
-$\MaybeT_\M\,\Gen$ by simply lifting:
-\begin{alignat*}{3}
-  &\gen' : \Up \vA \to \MaybeT_\M\,\Gen\,(\Up \vA) \\
-  &\gen'\,\va = \lift\,(\gen\,\va)
-\end{alignat*}
-However, it is more convenient to proceed in the style of Haskell's monad
-transformer library \texttt{mtl} \cite{mtl}, and have a class for monads that
-support code generation:
-\begin{alignat*}{3}
-&\tyclass\,\Monad\,\vM \RA \MonadGen\,\vM\,\where\\
-&\ind \liftGen : \Gen\,\vA \to \M\,\vA\\
-&\instance \MonadGen\,\Gen\,\where\,\liftGen = \id \\
-&\instance \MonadGen\,\vM \RA \MonadGen\,(\MaybeT_\M\,\vM)\,\where\,\liftGen = \lift \circ \liftGen
-\end{alignat*}
-The $\MonadGen$ instance can be defined uniformly for every monad transformer as
-$\liftGen = \lift \circ \liftGen$. We also redefine $\gen$ and $\genRec$ to work
-in any $\MonadGen$, so from now on we have:
-\begin{alignat*}{3}
- &\gen   &&: \MonadGen\,\vM \RA \Up \vA \to \vM\,(\Up \vA)\\
- &\genRec &&: \MonadGen\,\vM \RA (\Up \vA \to \Up \vA) \to \vM\,(\Up \vA)
-\end{alignat*}
+We can also use $\gen$ for let-insertion in $\MaybeT\,\Gen$, since it has a $\MonadGen$
+instance.
+%% Assume also a $\lift : \Monad\,\vM \RA \vM\,\vA \to \MaybeT_\M\,\vM\,\vA$ operation which
+%% comes from $\MaybeT_\M$ being a monad transformer. We can do let-insertion in
+%% $\MaybeT_\M\,\Gen$ by simply lifting:
+%% \begin{alignat*}{3}
+%%   &\gen' : \Up \vA \to \MaybeT_\M\,\Gen\,(\Up \vA) \\
+%%   &\gen'\,\va = \lift\,(\gen\,\va)
+%% \end{alignat*}
 
@@
 
-There is a deficiency in our library so far. Assuming $\vb : \Up\Bool$, consider:
-\begin{alignat*}{5}
-  & \hspace{-1em}\mdo && (\case \vb\,\of\,\True\to \mput' \qt{10}; \False &&\to \mput' \qt{20})\\
-  & \hspace{-1em}     && \mathrlap{\modify' (\lam \vx.\,\qt{\spl \vx + 10})}
-\end{alignat*}
-The $\modify' (\lam \vx. \qt{\spl \vx + 10})$ action gets inlined in both case
-branches during unstaging, yielding the following:
-\begin{alignat*}{3}
-   & boooo
-\end{alignat*}
-The duplication follows from the definition of monadic binding in $\Gen$ and the
+There is a deficiency in our library so far. Consider:
+\begin{alignat*}{3}
+  & \vf : \Bool \to \StateT\,\Nat\,(\MaybeT\,\Identity)\,()\\
+  & \vf\,\vb := \spl(\mdown \dlr \mdo \\
+  & \ind \case \qt{\vb}\,\of\,\True\to \mput' \qt{10}; \False \to \mput' \qt{20}\\
+  & \ind\modify'\,(\lambda\,x.\,\qt{\spl x + \spl x}))
+\end{alignat*}
+This is unstaged to the following (omitting newtype wrappers):
+\begin{alignat*}{3}
+  & \vf\,\vb\,\vs := \case\,\vb\,\of \\
+  & \ind \True\hspace{0.2em} \to \letdef \vs := 10; \letdef \vs := \vs + \vs; \Just\,((),\vs)\\
+  & \ind \False \to \letdef \vs := 20; \letdef \vs := \vs + \vs; \Just\,((),\vs)
+\end{alignat*}
+Notice that the final state modification gets inlined into both branches. The
+duplication follows from the definition of monadic binding in $\Gen$ and the
 $\mit{split}$ function in the desugaring of $\mbf{case}$. Code generation is
@@
 $\mbf{case}$ splits sequenced after each other, that yields exponential code
-size. Right now, we can fix this by let-binding the problematic action:
-
-%% \begin{alignat*}{5}
-%%   & \mdo && (\gen \dlr \mdown\,(\case \vb\,\of \True \to \mput' \qt{10};\False &&\to \mput' \qt{20})) \bind \mup \\
-%%   &      && \mathrlap{\modify' (\lam \vx.\,\qt{\spl \vx + 10})}
-%% \end{alignat*}
-\begin{alignat*}{5}
-  & \mdo && \msf{action} \fro \gen \dlr \mdown\,(\case \vb\,\of \True \to \mput' \qt{10};\False &&\to \mput' \qt{20}) \\
-  &      && \mup\,\msf{action}\\
-  &      && \mathrlap{\modify' (\lam \vx.\,\qt{\spl \vx + 10})}
-\end{alignat*}
-This removes code duplication by round-tripping through an object-level
-$\mbf{let}$. This solution is fairly good in a state monad, where $\mdown$ only
-introduces a runtime pair constructor, and it is feasible to compile
-object-level pairs as unboxed data, without overheads. However, for $\Maybe$,
-$\mdown$ introduces a runtime $\Just$ or $\Nothing$, and $\mup$ introduces a
-runtime case split. A better solution would be to introduce two let-bound
-\emph{join points} before the offending $\mbf{case}$, one for returning a
-$\Just$ and one for returning $\Nothing$, but fusing away the actual runtime
-constructors. So, for example, we would like to produce object code which looks
-like the following:
-\begin{alignat*}{3}
-  & \letdef\,\mit{joinJust}\,\vn := ...\\
-  & \letdef\,\mit{joinNothing}\,() := ...\\
-  & \case \vx == 10\,\of\\
-  & \ind \True \hspace{0.2em}\to \mit{joinJust}\,(\vx + 10) \\
-  & \ind \False \to \mit{joinNothing}\,()
-\end{alignat*}
-Such fused returns are possible whenever we have a $\Gen\,\vA$ action at the
-bottom of the transformer stack, such that $\vA$ is isomorphic to a meta-level
-finite sum of value types. Recall that $\Gen\,\vA$ is defined as $\{\vR : \VTy\} \to
-(\vA \to \Up \vR) \to \Up \vR$. Here, if $\vA$ is a finite sum, we can rearrange $\vA \to
-\Up \vR$ to a finite product of functions.
+size. A possible fix is to let-bind the branching action:
+\begin{alignat*}{3}
+  & \vf\,\vb := \spl(\mdown \dlr \mdo \\
+  & \ind \msf{act} \fro \gen \dlr \mdown \dlr \case \qt{\vb}\,\of\,\True\to \mput' \qt{10}; \False \to \mput' \qt{20} \\
+  & \ind \mup\,\msf{act} \\
+  & \ind\modify\,(\lambda\,x.\,\qt{\spl x + \spl x}))
+\end{alignat*}
+However, this yields suboptimal code:
+\begin{alignat*}{3}
+  & \vf\,\vb\,\vs :=\\
+  & \ind \letdef\,\msf{act}\,\vs := \case\,\vb\,\of \True\hspace{0.2em}\to \letdef \vs := 10; \Just((), \vs)\\
+  & \ind\hspace{8.5em}                               \False \to \letdef \vs := 20; \Just((), \vs);\\
+  & \ind \case \msf{act}\,\vs\,\of\\
+  & \ind\ind \Just\,(\_,\,\vs) \to \letdef \vs := \vs + \vs; \Just\,((),\,\vs)\\
+  & \ind\ind \Nothing\hspace{0.6em} \to \Nothing
+\end{alignat*}
+This solution is fairly good in a state monad, where $\mdown$ only introduces a
+runtime pair constructor, and it is feasible to compile object-level pairs as
+unboxed data, without overheads. However, for $\Maybe$, $\mdown$ introduces a
+runtime $\Just$ or $\Nothing$, and $\mup$ introduces a runtime case split. A
+better solution would be to introduce two let-bound join points before
+the offending $\mbf{case}$, one for returning a $\Just$ and one for returning
+$\Nothing$, but fusing away the actual runtime constructors:
+\begin{alignat*}{3}
+  & \vf\,\vb\,\vs :=\\
+  & \ind \letdef\,\msf{joinNothing}\,\vs := \Nothing;\\
+  & \ind \letdef\,\msf{joinJust}\,\hspace{1.8em}\vs := \letdef \vs := \vs + \vs; \Just\,((),\,\vs);\\
+  & \ind \case \msf{b}\,\of\\
+  & \ind \ind \True\hspace{0.2em}  \to \letdef \vs := 10; \msf{joinJust}\,\vs\\
+  & \ind \ind \False \to \letdef \vs := 20; \msf{joinJust}\,\vs
+\end{alignat*}
+Here, $\msf{joinNothing}$ happens to be dead code, but it is easy to clean that up in
+downstream compilation. Such ``fused'' returns are possible whenever we have a
+$\Gen\,\vA$ action at the bottom of the transformer stack, such that $\vA$ is
+isomorphic to a meta-level finite sum of value types. Recall that $\Gen\,\vA$ is
+defined as $\{\vR : \VTy\} \to (\vA \to \Up \vR) \to \Up \vR$. Here, if $\vA$ is
+a finite sum, we can rearrange $\vA \to \Up \vR$ to a finite product of
+functions.
 
 We could proceed with finite sums, but we will need finite
-\emph{sums-of-products} ($\SOP$ in short) later in Section in \ref{sec:stream-fusion}, so we
-develop SOP-s. We view $\SOP$-s as a Tarski-style universe consisting of a type of
-descriptions and a way of interpreting descriptions into $\MTy$ (``$\El$'' for
-``elements'' of the type).
+\emph{sums-of-products} ($\SOP$ in short) later in Section in
+\ref{sec:stream-fusion}, so we develop SOP-s. We view $\SOP$-s as a Tarski-style
+universe consisting of a type of descriptions and a way of interpreting
+descriptions into $\MTy$ (``$\El$'' for ``elements'' of the type).
 \begin{alignat*}{3}
@@
 \end{alignat*}
-
-Now, whenever the return value of a $\mbf{case}$-splitting action is a sum of
-products of values (this includes just returning an object value, which is by
-far the most common situation), we can use $\join \dlr \case \vx\,\of ...$ to
-eliminate code duplication, without creating runtime sum types.
+Now, we can write the following in our running example:
+\begin{alignat*}{3}
+  & \vf : \Bool \to \StateT\,\Nat\,(\MaybeT\,\Identity)\,()\\
+  & \vf\,\vb := \spl(\mdown \dlr \mdo \\
+  & \ind \join \dlr \case \qt{\vb}\,\of\,\True\to \mput' \qt{10}; \False \to \mput' \qt{20}\\
+  & \ind\modify'\,(\lambda\,x.\,\qt{\spl x + \spl x}))
+\end{alignat*}
+It might make sense to also have a simple ``desugaring'' rule that inserts a
+$\join$ whenever we have a object-level case split with more than one branch. At
+worst, this generates some noise and dead code that is easy to remove by
+conservative code optimization.
 
@@
 
-We look at a slightly larger code example. We define annotated binary trees as
+We look at a slightly larger code example in Figure \ref{fig:codeinp}. We define
+annotated binary trees as
 \[ \data \Tree\,\vA := \Leaf\,|\,\Node\,\vA\,(\Tree\,\vA)\,(\Tree\,\vA). \]
@@
 throw an error. If we run out of list elements, we leave values unchanged.
+\begin{figure}
 \begin{alignat*}{4}
@@
 \end{alignat*}
+\caption{Monadic source code}
+\label{fig:codeinp}
+\end{figure}
 Here, all of the $\case$ matches are done on object-level values, so they are
@@
 a joint point in every $\mbf{case}$ with two or more branches, and use the
-strict state modifications everywhere. This would only add a small amount of
-administrative noise that can be easily cleaned up by the downstream compiler.
-
-Let us look at the unstaging output. We omit newtype wrappers and use nested
-pattern matching on pairs, but otherwise this is exactly the code that we get in
-our Agda implementation.
+strict state modifications everywhere, and leave it up to the downstream compiler
+to clean up the code.
+
+We show the unstaging output on Figure \ref{fig:codeout}. We omit newtype
+wrappers and use nested pattern matching on pairs, but otherwise this is exactly
+the code that we get in our Agda implementation. Again, the only flaw in this
+code is the dead binding for $\msf{joinNothing}$.
+\begin{figure}\label{fig:codeout}
 \begin{alignat*}{4}
@@
 \end{alignat*}
-The only flaw in this code is the $\msf{joinNothing}$, which is never called,
-becase we joined an action that never throws an error. But this is also easy to
-clean up afterwards.
+\caption{Unstaged monadic code}
+\label{fig:codeout}
+\end{figure}
 \subsection{Discussion}
@@
 syntactic $\vW{-}\msf{elim}$ is rather gnarly, because of the usage of explicit
-substitutions, but we write it out here anways.
+substitutions, but we write it out here anyways.
 

### 2024-06-10 newpaper/paper.tex
wip

@@
 that we elaborate $\mbf{case}$ on object values to an application of the
-appropriate $\mit{split}$ function. We adopt this in the rest of the paper, so
+appropriate splitting function. We adopt this in the rest of the paper, so
 when working in a $\MonadGen$, we can write $\case \vas\,\of \Nil \to ...;
@@
 
-In the meta level, we can reuse essentially all definitions from \texttt{mtl}.
-From there, only the continuation monad transformer fails to support
+In the meta level, we can borrow essentially all definitions from
+\texttt{mtl}. Only the continuation monad transformer fails to support
 binding-time-improvement in CFTT, because of the obvious need for dynamic
@@
 A type description is a list of lists of value types. We decode this to a sum of
-products of value types. We omit the definition here. $\USOP$ is closed under
+products of value types. For example, $\El_\SOP\,[[\Bool,\,\Bool],\,[]]$ (using
+Haskell-like list notation) is isomorphic to $\Either\,(\Lift\Bool,\,\Lift\Bool)\,()$.
+$\USOP$ is closed under
 value types, finite product types and finite sum types. For instance, we have
@@
 \end{alignat*}
-Now, we can write the following in our running example:
-\begin{alignat*}{3}
-  & \vf : \Bool \to \StateT\,\Nat\,(\MaybeT\,\Identity)\,()\\
+Now, we can modify our running example as follows.
+\begin{alignat*}{3}
   & \vf\,\vb := \spl(\mdown \dlr \mdo \\
@@
 
-Since the logical relation interpretation of identity is trivial, it is also trivial
-for the generativity axiom.
+Finally, since the logical relation interpretation of the identity type is trivial,
+the logical relation interpretation for the generativity axiom is also trivial.
 
@@
 \emph{Strictness} of unstaging holds trivially, since there are no $\beta$ and
-$\eta$ rules in the object theory. \emph{Completeness} of unstaging is shown by
+$\eta$ rules in the object theory. \emph{Stability} of unstaging is shown by
 straightforward induction on the syntax of the object theory, the same way as in

### 2024-06-11 newpaper/paper.tex
stuff

@@
 
-\textbf{Two-level type theory} (2LTT) \cite{twolevel,staged2ltt} makes it
-possible to use expressive dependent type theories for metaprogramming
-for a wide range of object languages. In this paper, we use 2LTT to sweeten the
-deal of staged compilation, aiming for a combination of strong guarantees, good
-ergonomics, high level of abstraction and easy-to-optimize code output.
-
-We develop a particular two-level type theory for this purpose, which we
-call \textbf{CFTT}, short for ``closure-free type theory''. This consists
-of:
+We use \textbf{two-level type theory} (2LTT) \cite{twolevel,staged2ltt} to
+sweeten the deal of staged compilation, aiming for a combination of strong
+guarantees, good ergonomics, high level of abstraction and easy-to-optimize code
+output. We develop a particular two-level type theory for this purpose, which we
+call \textbf{CFTT}, short for ``closure-free type theory''. This consists of:
 \begin{itemize}
@@
 
-Formally, $\MTy$ is additionally indexed by universes levels (orthogonally to
+Formally, $\MTy$ is additionally indexed by universe levels (orthogonally to
 staging), and we have $\MTy_i : \MTy_{i+1}$. However, universe levels add noise
@@
 We also allow $\mbf{newtype}$ definitions, both in $\VTy$ and $\CTy$. These are
-assumed to be erased at runtime. In the Haskell implementation they are
-important for guiding type class resolution, and we think that the explicit
+assumed to be erased at runtime. In the Haskell and Agda implementations they
+are important for guiding instance resolution, and we think that the explicit
 wrapping makes many definitions more comprehensible in CFTT as well.
@@
   control over code generation. Unstaging does not translate higher-order
-  functions to anything in particular. Instead, it is the execution of
-  higher-order metaprograms that produce first-order programs.
+  functions to anything in particular. Instead, unstaging is the execution of
+  higher-order metaprograms which produce first-order programs.
 \end{itemize}
@@
 
-We give an overview of the unstaging algorithm together with its correctness.
-Our main task is to extend \cite{staged2ltt} with identity types and inductive
-types in the metalanguage, since we use them in this paper. For inductive types,
-it is the most sensible to add W-types, since all inductive families can be
-faithfully derived from them in our setting \cite{whynotw}.
+To obtain an unstaging algorithm, together with its correctness, the main task
+is to extend \cite{staged2ltt} with identity types and inductive families in the
+metalanguage, since we use those types in this paper. For inductive types, it is the
+most sensible to add W-types, since all inductive families can be faithfully
+derived from them in our setting \cite{whynotw}.
 

### 2024-06-12 newpaper/paper.tex
wip

@@
 instances, and, for example, $(\bind)\,\msf{monadReader}$ is a field projection
-from the dictionary. We get a runtime closure from the $\lam \vb.\,...$
-function, and $(\bind)$, $\msf{ask}$ and $\mbf{return}$ also produce additional
-closures.
+from the dictionary. This results from the dictionary-passing elaboration of
+type classes \cite{DBLP:conf/popl/WadlerB89}. We get a runtime closure from the
+$\lam \vb.\,...$ function, and $(\bind)$, $\msf{ask}$ and $\mbf{return}$ also
+produce additional closures.
 
@@
 
+Why not just make the object language less liberal, e.g.\ by disallowing
+$\lambda$ under $\mbf{case}$ or $\mbf{let}$, thereby making call saturation
+easier or more obvious? There is a trade-off between making the object language
+more restricted, and thus easier to compile, and making \emph{metaprogramming}
+more convenient. We will see that the ability to insert $\mbf{let}$-s without
+restriction is very convenient in code generation, and likewise the ability to
+have arbitrary object expressions in $\mbf{case}$ bodies. In this paper we go
+with the most liberal object syntax, at the cost of needing more downstream
+processing.
+
+On the other side of the spectrum, one might imagine generating object code in
+low-level A-normal form. This is more laborious, but it could be also
+interesting, because it forces us to invent abstractions for manipulating ANF in
+the metalanguage.
+
 \subsubsection{Object-level definitional equality} This is a distinct notion
@@
 %% the metalanguage. In a similar vein, Allais recently proposed metaprogramming
-%% quantum circuits in a two-level type theory \cite{TODO}. We leave such setups
+%% quantum circuits in a 2LTT \cite{TODO}. We leave such setups
 %% with low-level object languages to future investigation.
@@
 is unstaged to
-\[ \qt{\lam \vx.\,\letdef x := \vx + \vx; \letdef \vx := \vx + \vx; ((),\,\vx)} \]
+\[ \qt{\lam \vx.\,\letdef x := \vx + \vx; \letdef \vx := \vx + \vx; ((),\,\vx)}. \]
 
@@
 \emph{sums-of-products} ($\SOP$ in short) later in Section in
-\ref{sec:stream-fusion}, so we develop SOP-s. We view $\SOP$-s as a Tarski-style
+\ref{sec:stream-fusion}, so we develop SOP-s. SOPs have been used in generic and
+staged programming; see e.g.\ \cite{sop} and
+\cite{DBLP:conf/haskell/PickeringLW20}. We view $\SOP$-s as a Tarski-style
 universe consisting of a type of descriptions and a way of interpreting
@@
 \end{alignat*}
-Now, we can modify our running example as follows.
+Now, the following definition yields the previously seen code output with
+$\msf{joinNothing}$ and $\msf{joinJust}$.
 \begin{alignat*}{3}
@@
 This is, in fact, our reason for including the isomorphism equations in
-$\IsSOP$. This, in turn, necessitates using SOP instead of finite sums. We can
-define a product type former for finite sums of value types, by taking the
-pairwise products of components. However, we can only take the
-\emph{object-level} products here, and since object-level products have no
-$\beta\eta$-rules, we cannot prove $\beta\eta$ for the derived product type in
-finite sums, and likewise for the derived $\Sigma$-type. When we use SOP
-instead, we do not have to use object-level products and this issue does not
-appear.
-
-We skip the full definition of $\Sigma_\SOP$ here. The reader may refer
-to the $\SOP$ module in the Agda implementation, which is altogether 260 lines with
-all $\IsSOP$ instances.
+$\IsSOP$.
+
+This, in turn, necessitates using SOP instead of finite sums, for the following
+reason. We can define a product type former for finite sums of value types, by
+taking the pairwise products of components, but in this case the products would
+be object-level products, which do not support $\beta\eta$ rules. This implies
+that we cannot prove $\beta\eta$-rules for the derived product type in finite
+sums. In contrast, when we define products for SOP-s, we take \emph{meta-level}
+pairwise products of components, which does support $\beta$ and $\eta$.
+
+We skip the full definition of $\Sigma_\SOP$ here. The reader may refer to the
+$\SOP$ module in the Agda implementation, which is altogether 260 lines with all
+$\IsSOP$ instances.
 
@@
 
-\subsubsection*{Two-level type theory.} Two-level lambda calculi were first developed in
-the context of abstract interpretation and partial evaluation
+\subsubsection*{Two-level calculi.}
+Two-level lambda calculi were first developed in the context of abstract
+interpretation and partial evaluation
 \cite{DBLP:phd/ethos/Nielson84a,DBLP:books/daglib/0071307}.  This line of
@@
 Later and independently, the same notion of level appeared in homotopy type
-theory, first in Voevodsky's Homotopy Type System \cite{hts}, and subsequently developed as
-2LTT in \cite{twolevel}. Here, dependent types are essential, and we have different
-theories at the two stages. The application of 2LTT to staged compilation was
-developed in \cite{staged2ltt}. Binding-time analysis has not been developed in this
-setting; 2LTT-s have been meant to be used as surface languages to directly work
-in.
-
-We build on both traditions. CFTT itself is a 2LTT and makes heavy use of
-dependent type theory. However, we are mostly concerned with code generation and
-optimization and borrow concepts from the partial evaluation literature.
+theory, first in Voevodsky's Homotopy Type System \cite{hts}, and subsequently
+developed as 2LTT in \cite{twolevel}. Here, dependent types are essential, and
+we have different theories at the two stages. The application of 2LTT to staged
+compilation was developed in \cite{staged2ltt}. Binding-time analysis has not
+been developed in this setting; 2LTT-s have been meant to be used as surface
+languages to directly work in.
+
+\subsubsection*{Higher-order abstract syntax and logical frameworks.}
+Hofmann's work on the semantics of higher-order abstract syntax (HOAS) is an
+important precursor to our work \cite{Hofmann:1999:SAH:788021.788940}. It
+anticipates many of the later developments in logical frameworks and two-level
+type theories. In particular, we build on Hofmann's presheaf model for our 2LTT
+semantics, and his sketch of adequacy for HOAS corresponds to our definition of
+unstaging and its soundness proof. In general, there is a close correspondence
+between logical frameworks that represent object languages in HOAS style
+(e.g.\ \cite{Harper93lf}) and two-level type theories, and in some cases the two
+kinds of presentations are inter-derivable \cite[Section
+  3.3]{DBLP:journals/corr/abs-2302-08837}.
+
+Cocon is a logical framework that has a dependently typed meta level
+(``computation'' level) and an object language that is configured by a signature
+\cite{DBLP:journals/corr/abs-1901-03378}. Cocon, like 2LTTs, can be used as a
+dependently typed metaprogramming language, but there are several
+differences. First, Cocon supports and emphasizes intensional analysis
+(i.e.\ the ability to do structural induction on object code), which is made
+possible through a contextual modality \cite{nanevski2008contextual}. In
+contrast, intensional analysis is missing from 2LTTs. However, in Cocon, we can
+only metaprogram object-level terms by specifying their object-level contexts
+(i.e.\ possible free variables), while in 2LTTs the object-level contexts are
+implicit and are only computed during code generation. The latter style is
+more convenient for staged compilation applications, where intensional analysis
+is rarely used and explicit context tracking would add a significant amount of
+bureaucracy.
+
+\subsubsection*{CPS and monads in staged programming} Writing code generators
+in continuation-passing style was first proposed by Bondorf
+\cite{DBLP:conf/lfp/Bondorf92}. Flanagan et al.'s ANF translation algorithm uses
+CPS for let-insertion as well \cite{DBLP:conf/pldi/FlanaganSDF93}, similarly to
+our $\gen$ function. These sources do no use explicit monadic notation
+though. Jones et al.\ discuss CPS and binding-time improvement in partial
+evaluation in \cite{partial-evaluation}. In
+\cite{DBLP:conf/emsoft/KiselyovST04}, \cite{DBLP:conf/pepm/SwadiTKP06} and
+\cite{DBLP:journals/scp/CaretteK11}, the composition of state and continuation
+monads was used, similarly to our $\StateT\,S\,\Gen$, but without polymorphic
+answer types. The definition of let-insertion here is again similar to ours.
 
@@
 $\SOP$-of-object-code representation appeared as well in typed Template Haskell
-in \cite{DBLP:conf/haskell/PickeringLW20}.
-
-\subsubsection*{CPS and binding-time improvement}
-Our $\Gen$ monad can be viewed as a variation on the continuation-passing-style
-that has been often used in staged programming. Flanagan et al.'s ANF
-translation algorithm uses let-insertion in CPS that works the same way as our
-$\gen$ \cite{DBLP:conf/pldi/FlanaganSDF93}. Kiselyov et al.\ likewise uses
-essentially the same CPS, in MetaOCaml
-\cite{DBLP:conf/popl/KiselyovBPS17}. Jones et al.\ discuss CPS and binding-time
-improvement in partial evaluation \cite{partial-evaluation}. $\eta$-expansion
-for object-level finite sums is known as ``the trick'' \cite{eta-expansion-trick}.
+in \cite{DBLP:conf/haskell/PickeringLW20}. $\eta$-expansion for object-level
+finite sums is known as ``the trick'' \cite{eta-expansion-trick}.
 
@@
 The staged stream fusion library $\mathit{strymonas}$ by Kiselyov et al.\ is the
-primary prior art \cite{DBLP:conf/popl/KiselyovBPS17}. Here, streams are represented as a sum type of
-push and pull representations, allowing both zipping and
-$\concatMap$-ing. However, fusion is not guaranteed for all combinations;
-zipping two $\concatMap$-s reifies one of the streams in a runtime closure. In
-newer versions of the library, fusion is reported to be complete and guaranteed
-\cite{DBLP:conf/pepm/0001K24}, however, an exposition of this has not yet been published. Another
-notable difference is that $\mathit{strymonas}$ relies on mutable references in
-the object language, while we do not.
+primary prior art \cite{DBLP:conf/popl/KiselyovBPS17}. Here, streams are
+represented as a sum type of push and pull representations, allowing both
+zipping and $\concatMap$-ing. However, fusion is not guaranteed for all
+combinations; zipping two $\concatMap$-s reifies one of the streams in a runtime
+closure. In newer versions of the library, fusion is complete
+\cite{DBLP:conf/pepm/0001K24}, however, an exposition of this has not yet been
+published. Additionally, $\mathit{strymonas}$ heavily relies on mutable
+references in the object language, and also uses some switching on object-level
+data to implement control flow. This style of code output can be still reliably
+optimized by downstream compilers (C or OCaml), but our solution is more
+conservative in that it does not use any mutation or runtime switching.
+This ``purity'' of our streams also makes them easier to generalize, e.g.\
+by using arbitrary $\MonadGen$ monads instead of $\Gen$ in stream internals.
 
@@
 stream combinator definitions from there. This account is also close to existing
-stream implementations in Haskell. Ibid.\ characterizes fusibility in a
+stream implementations in Haskell. Coutts characterized fusibility in a
 non-staged setting, in terms of ``good consumers'' and ``good producers''.  This
@@
 
-We give an algebraic \cite{TODO} specification for the models of CFTT and define
-the syntax of CFTT to be the initial model. In the following, all type and term
-formers are implicitly assumed to be stable under substitution.
+We give an algebraic \cite{cartmellthesis,DBLP:journals/corr/abs-2302-08837}
+specification for the models of CFTT and define the syntax of CFTT to be the
+initial model. In the following, all type and term formers are implicitly
+assumed to be stable under substitution.
 
@@
   terminal object (which is the empty context in the syntax).
-\item A family structure in the sense of category-with-families \cite{TODO},
+\item A family structure in the sense of category-with-families \cite{Dybjer96internaltype},
   specifying metatypes and terms. This structure is additionally indexed by
@@
 In the context of CFTT, the choice between HOAS and 2LTT styles is a matter of
-taste, and the two are inter-derivable \cite{TODO}.
+taste, and the two are inter-derivable \cite[Section
+  3.3]{DBLP:journals/corr/abs-2302-08837}.
 
@@
 So far we have stayed close to \cite{staged2ltt}, with only cosmetic
-changes. Supporting W-types, as a way to get inductive families \cite{TODO}, is
-the biggest extension here. We extend Section 5 of \cite{staged2ltt} here. In
-the syntax of CFTT, $\vW$-types are as follows. The formal specification of
-syntactic $\vW{-}\msf{elim}$ is rather gnarly, because of the usage of explicit
-substitutions, but we write it out here anyways.
+changes. Supporting W-types, as a way to get inductive families
+\cite{inductivefamilies}, is the biggest extension here. We extend Section 5 of
+\cite{staged2ltt} here. In the syntax of CFTT, $\vW$-types are as follows. The
+formal specification of syntactic $\vW{-}\msf{elim}$ is rather gnarly, because
+of the usage of explicit substitutions, but we write it out here anyways.
 
@@
 In this section we assume an \emph{extensional} identity type in CFTT. As we
-mentioned in Section \ref{TODO}, we can show soundness of staging with
+mentioned in Section \ref{sec:semantics-of-staging}, we can show soundness of staging with
 extensional identity and certain \emph{postulated} identities, but we cannot do
@@
 
-The generativity axiom from Section \ref{TODO} has the following type, in internal
+The generativity axiom from Section \ref{sec:concatmap-for-streams} has the following type, in internal
 CFTT syntax:

### 2024-06-12 newpaper/paper.tex
wip

@@
 
-\documentclass[acmsmall,screen,review,anonymous]{acmart}
-%% \documentclass[nonacm,acmsmall,review]{acmart}
+%% \documentclass[acmsmall,screen,review,anonymous]{acmart}
+%% \documentclass[nonacm,acmsmall]{acmart}
+\documentclass[acmsmall,screen]{acmart}
 %% \raggedbottom
@@
 %% used to denote shared contribution to the research.
-\author{András Kovács}
-\email{andrask@chalmers.se}
-\orcid{0000-0002-6375-9781}
+
+\author{First Last}
 \affiliation{%
-  \institution{University of Gothenburg}
-  \country{Sweden}
-  \city{Gothenburg}
+  \institution{Institution}
+  \country{Country}
+  \city{City}
 }
+
+%% \author{András Kovács}
+%% \email{andrask@chalmers.se}
+%% \orcid{0000-0002-6375-9781}
+%% \affiliation{%
+%%   \institution{University of Gothenburg}
+%%   \country{Sweden}
+%%   \city{Gothenburg}
+%% }
 
@@
 with the most liberal object syntax, at the cost of needing more downstream
-processing.
-
-On the other side of the spectrum, one might imagine generating object code in
-low-level A-normal form. This is more laborious, but it could be also
-interesting, because it forces us to invent abstractions for manipulating ANF in
-the metalanguage.
+processing. On the other side of the spectrum, one might imagine generating
+object code in low-level A-normal form. This is more laborious, but it could be
+also interesting, because it would force us to invent abstractions for
+manipulating ANF in the metalanguage. We leave such lower-level object languages
+for future investigation.
 

### 2024-06-12 newpaper/paper.tex
paper revision

@@
 easier or more obvious? There is a trade-off between making the object language
-more restricted, and thus easier to compile, and making \emph{metaprogramming}
-more convenient. We will see that the ability to insert $\mbf{let}$-s without
+more restricted, and thus easier to compile, and making metaprogramming more
+convenient. We will see that the ability to insert $\mbf{let}$-s without
 restriction is very convenient in code generation, and likewise the ability to
@@
 with the most liberal object syntax, at the cost of needing more downstream
-processing. On the other side of the spectrum, one might imagine generating
-object code in low-level A-normal form. This is more laborious, but it could be
-also interesting, because it would force us to invent abstractions for
-manipulating ANF in the metalanguage. We leave such lower-level object languages
-for future investigation.
+processing.
+
+%% On the other side of the spectrum, one might imagine generating
+%% object code in low-level A-normal form. This is more laborious, but it could be
+%% also interesting, because it would force us to invent abstractions for
+%% manipulating ANF in the metalanguage. We leave such lower-level object languages
+%% for future investigation.
 
@@
 
-In the meta level, we can borrow essentially all definitions from
+On the meta level, we can borrow essentially all definitions from
 \texttt{mtl}. Only the continuation monad transformer fails to support
@@
 \end{alignat*}
-Notice that the final state modification gets inlined into both branches. The
+Notice that the final state modification gets inlined into both branches. This
 duplication follows from the definition of monadic binding in $\Gen$ and the
@@
 \end{alignat*}
-This solution is fairly good in a state monad, where $\mdown$ only introduces a
-runtime pair constructor, and it is feasible to compile object-level pairs as
-unboxed data, without overheads. However, for $\Maybe$, $\mdown$ introduces a
-runtime $\Just$ or $\Nothing$, and $\mup$ introduces a runtime case split. A
-better solution would be to introduce two let-bound join points before
-the offending $\mbf{case}$, one for returning a $\Just$ and one for returning
-$\Nothing$, but fusing away the actual runtime constructors:
+This solution is fairly good when we only have $\State$ or $\Reader$ effects, where
+$\mdown$ only introduces a runtime pair constructor, and it is feasible to
+compile object-level pairs as unboxed data, without overheads. However, for
+$\Maybe$, $\mdown$ introduces a runtime $\Just$ or $\Nothing$, and $\mup$
+introduces a runtime case split. A better solution would be to introduce two
+let-bound join points before the offending $\mbf{case}$, one for returning a
+$\Just$ and one for returning $\Nothing$, but fusing away the actual runtime
+constructors:
 \begin{alignat*}{3}
@@
 inverses. We will overload $\mit{rep}$ as the forward conversion function with
-type $\vA \to \El_\SOP\,\Rep$, and write $\mit{rep}^{-1}$ for its inverse. Now we define
-an isomorphic presentation of $\El_\SOP\,\vA \to \Lift \vR$ as a product of
-object-level functions:
+type $\vA \to \El_\SOP\,\Rep$, and write $\mit{rep}^{-1}$ for its inverse.
+
+In 2LTT literature, a metatype $\vA$ is called \emph{cofibrant} if $\vA \to \vB$ is
+isomorphic to an object type whenever $\vB$ is isomorphic to an object type
+\cite{twolevel}. We conjecture that our $\IsSOP$ types are exactly the cofibrant
+types in this sense. However, ``cofibrant'' is not very descriptive in our
+setting, so we shall keep writing $\IsSOP$.
+
+Accordingly, we define an isomorphic presentation of $\El_\SOP\,\vA \to
+\Lift \vR$ as a product of object-level functions:
 \begin{alignat*}{3}
@@
 required to be a sum-of-products of value types by the $\IsSOP\,\vS$
-constraint. Borrowing terminology from Coutts \cite{DBLP:phd/ethos/Coutts11}, we
+constraint. The next field with type $\Gen\,\vS$ is the initial state, while
+transitions are represented by the $\vS \to \Gen\,(\Step\,\vS\,\vA)$
+field. Possible transitions are stopping ($\Stop$), transitioning to a new state
+while outputting a value ($\Yield$) and making a silent transition to a new
+state ($\Skip$).
+
+Our definition roughly follows the non-staged stream fusion setup of
+Coutts \cite{DBLP:phd/ethos/Coutts11}; the difference is the addition of
+$\IsSOP$ and $\Gen$ in our version. Also, borrowing terminology from Coutts, we
 call each product in the state a \emph{state shape}.
-
-The next field with type $\Gen\,\vS$ is the initial state, while transitions
-are represented by the $\vS \to \Gen\,(\Step\,\vS\,\vA)$ field. Possible transitions
-are stopping ($\Stop$), transitioning to a new state while outputting a value
-($\Yield$) and making a silent transition to a new state ($\Skip$).
 
@@
 kinds of presentations are inter-derivable \cite[Section
-  3.3]{DBLP:journals/corr/abs-2302-08837}.
+  3.3]{DBLP:journals/corr/abs-2302-08837}. In fact, our Agda implementation
+of CFTT also uses a HOAS embedding of the object theory.
 
@@
 possible through a contextual modality \cite{nanevski2008contextual}. In
-contrast, intensional analysis is missing from 2LTTs. However, in Cocon, we can
-only metaprogram object-level terms by specifying their object-level contexts
-(i.e.\ possible free variables), while in 2LTTs the object-level contexts are
-implicit and are only computed during code generation. The latter style is
-more convenient for staged compilation applications, where intensional analysis
-is rarely used and explicit context tracking would add a significant amount of
+contrast, intensional analysis is missing from 2LTTs so far. However, in Cocon,
+we can only manipulate object terms at the meta level by specifying their
+concrete object-level contexts (i.e.\ possible free variables), while in 2LTTs
+the object-level contexts are implicit and are only computed during code
+generation. The latter style is more convenient for staged compilation
+applications, where explicit context tracking would add a significant amount of
 bureaucracy.
+
+Nonetheless, it seems to be possible and useful to have a
+system which simultaneously supports intensional analysis through contextual
+types and the 2LTT-style implicit contexts.
 
@@
 stream combinator definitions from there. This account is also close to existing
-stream implementations in Haskell. Coutts characterized fusibility in a
-non-staged setting, in terms of ``good consumers'' and ``good producers''.  This
-scheme does not cover $\concatMap$.
+stream implementations in Haskell. We find it interesting that our staged
+definitions are very close to the non-staged versions there; essentially all of
+the additional complexity is compartmentalized in our $\SOP$ module.  Coutts
+characterized fusibility in a non-staged setting, in terms of ``good consumers''
+and ``good producers'', but this scheme does not cover $\concatMap$.
 
@@
 instance, in this paper we demonstrated that compiling monadic code can be done
-by using \emph{the same monads} in the metalanguage, extended with code
+by using the same monads in the metalanguage, extended with code
 generation as an effect.
@@
 \appendices
-\section{Staging in CFTT}\label{appendix:A}
+\section{Staged Semantics for CFTT}\label{appendix:A}
 
@@
   i.e.\ $\U_i : \MTy_{i+1}\,\Gamma$ such that there is an isomorphism
-  $\msf{MetaTm}_{i+1}\,\Gamma\,\U_i \simeq \MTy_i\,\Gamma$.  We use $\El$ and
+  $\msf{MetaTm}\,\Gamma\,\U_i \simeq \MTy_i\,\Gamma$.  We use $\El$ and
   $\code$ to denote the components of this isomorphism.
@@
 \item For object-level terms, we have $\Tm : \MTm\,\Gamma\,\Ty \to \Set$,
-  supporting context extension, i.e.\ we have $\Gamma,\ext\,A$ for
-  $A : \MTm\,\Gamma\,\Ty$ such that $\Sub\,\Gamma\,(\Delta\,\ext\,A)$ is isomorphic
+  supporting context extension, i.e.\ we have $\Gamma\ext A$ for
+  $A : \MTm\,\Gamma\,\Ty$ such that $\Sub\,\Gamma\,(\Delta\ext A)$ is isomorphic
   to $(\sigma : \Sub\,\Gamma\,\Delta) \times \Tm\,\Gamma\,(A[\sigma])$. All
@@
 This presentation is a bit more verbose than the surface syntax used in the
-paper. First, the meta level uses explicit coding and decoding operations for
-universes, while the surface syntax assumes Russell-style universes. Second, the
-object level has $\VTy$ and $\CTy$ as Tarski-style universes with explicit
-inclusion into $\Ty$, while the surface syntax used implicit coercions.
+paper. First, the meta level uses explicit coding and decoding for universes,
+while the surface syntax assumes Russell-style universes. Second, the object
+level has $\VTy$ and $\CTy$ as Tarski-style universes with explicit inclusions
+$\msf{V}$ and $\msf{C}$ into $\Ty$, while the surface syntax used implicit
+coercions.
 
@@
   would only have metatypes and metaterms, no quoting or splicing, and would
-  reuse metafunctions to represent object-level binders.Our Agda implementation
+  reuse metafunctions to represent object-level binders. Our Agda implementation
   actually follows this style.
@@
 changes. Supporting W-types, as a way to get inductive families
-\cite{inductivefamilies}, is the biggest extension here. We extend Section 5 of
+\cite{whynotw}, is the biggest extension here. We extend Section 5 of
 \cite{staged2ltt} here. In the syntax of CFTT, $\vW$-types are as follows. The
@@
 
-Next, we give a logical relation interpretation for $\vW$. We define an
-inductive relation in $\whset$ as follows.
+Next, we give a logical relation interpretation for $\vW$. We inductively define
+a relation in $\whset$ as follows.
 \begin{alignat*}{3}
@@
 Here, $(B \Rightarrow (\vW\,A\,B)[\p])\rel$ is the relational interpretation of the function type, which unfolds to
-pointwise preservation of relations. In short, $\vW\rel$ expresses that $\wh{\vsup}\,\alpha\,f$ is related $\re_\vsup\,\alpha'\,f'$
-when the arguments (inductively) related. We can immediately interpret $\vW$ and $\vsup$:
+pointwise preservation of relations. In short, $\vW\rel$ expresses that $\wh{\vsup}\,\alpha\,f$ is related to $\re_\vsup\,\alpha'\,f'$
+when the arguments are (inductively) related. We can immediately interpret $\vW$ and $\vsup$:
 \begin{alignat*}{3}
@@
 becomes rather tedious to compute and unfold the involved types, so we only give
-a compact definition here together with some informal explanation:
+a compact definition together with some informal explanation:
 \begin{alignat*}{3}
@@
 rest of the relational interpretation of $\Id$ is also trivial. Another way to
-think about this is that soundness expresses a form of canonicity, but if we
-have equality reflection then syntactic identities are always canonical.
+think about this: soundness expresses a form of canonicity, but if we have
+equality reflection then terms with identity types are always canonical.
 

### 2024-06-18 icfp24paper/paper.tex
icfp24 final version folder

--- 2024-06-12 newpaper/paper.tex
@@
 
-\author{First Last}
+\author{András Kovács}
+\orcid{0000-0002-6375-9781}
+\email{andrask@chalmers.se}
 \affiliation{%
-  \institution{Institution}
-  \country{Country}
-  \city{City}
+  \institution{University of Gothenburg}
+  \country{Sweden}
+  \city{Gothenburg}
 }
@@
 CFTT programs, by evaluating all metaprograms in splices. This term has been
-sporadically used in the literature, e.g.\ by Odersky and Rompf
-\cite{DBLP:journals/cacm/RompfO12} and also by Choi et
-al.\ \cite{DBLP:conf/popl/ChoiAYT11} with a somewhat different meaning in a
-multi-stage context. The main precursor to this paper used ``staging'' instead
-of our ``unstaging'' \cite{staged2ltt}; we use the latter because the former
-conflicts with other common usages of ``staging''.
+sporadically used in the literature, see
+e.g.\ \cite{DBLP:journals/cacm/RompfO12}, or \cite{DBLP:conf/popl/ChoiAYT11},
+with a somewhat different meaning in a multi-stage context. The main precursor
+to this paper used ``staging'' instead of our ``unstaging'' \cite{staged2ltt};
+we use the latter because the former conflicts with other common usages of
+``staging''.
 
@@
 all identity proofs and their transports from CFTT programs, then we proceed
-with unstaging as in \cite{staged2ltt}. This erasure can be formalzied as a
+with unstaging as in \cite{staged2ltt}. This erasure can be formalized as a
 syntactic translation \cite{next700} from CFTT to CFTT extended with the
@@
 \cite{DBLP:conf/lfp/Bondorf92}. This can be structured as a continuation monad,
-and later work used explicit monadic notation for it
+and later works used explicit monadic notation for it
 \cite{DBLP:conf/emsoft/KiselyovST04,DBLP:journals/scp/CaretteK11,DBLP:conf/pepm/SwadiTKP06}. Our
@@
 polymorphic answer types are more convenient, because we do not have to
-anticipate the type of the code output when are defining a code generator.
-Hence, we can ``run'' actions as follows:
+anticipate the type of the code output when are defining a code generator. We
+can ``run'' actions as follows:
 \begin{alignat*}{3}
@@
   &\hspace{-2em}\mathrlap{\mup : \Up (\vA,\,\vB) \to \Gen\,(\Up \vA,\,\Up \vB)             \hspace{5em} \mdown : \Gen\,(\Up \vA,\, \Up \vB) \to \Up(\vA,\,\vB)} \\
-  &\hspace{-2em}\mup\,\vx = \mdo &&\vx \fro \gen\,\vx                                    && \ind\ind \mdown\,\vx = \mdo &&(\va,\,\vb) \fro \vx\\
+  &\hspace{-2em}\mup\,\vx = \mdo &&\vx \fro \gen\,\vx                                    && \ind\ind \mdown\,\vx = \runGen \dlr \mdo &&(\va,\,\vb) \fro \vx\\
   &\hspace{-2em}               &&\return (\qt{\fst\,\spl \vx},\,\qt{\snd\,\spl \vx})     &&                           &&\return \qt{(\spl \va,\, \spl \vb)}
@@
 \end{alignat*}
-In the rest of this paper, we shall only mention $\ReaderT$, $\MaybeT$ and
+In the rest of this paper, we only present $\ReaderT$, $\MaybeT$ and
 $\StateT$ as monad transformers. For all of these, we can define the $\MonadGen$
@@
   &\mit{split} : \Up (\List\,\vA) \to \Gen\,(\msf{SplitList}\,\vA)\\
-  &\mit{split}\,\vas = \Gen \dlr \lam \vk.\,\qt{\case \spl \vas\,\of \Nil \to \spl(\vk\,\Nil'); \Cons\,\va\,\vas \to \spl(\vk\,(\Cons'\,\qt{a}\,\qt{\vas}))}
+  &\mit{split}\,\vas = \Gen \dlr \lam \vk.\,\qt{\case \spl \vas\,\of \Nil \to \spl(\vk\,\Nil'); \Cons\,\va\,\vas \to \spl(\vk\,(\Cons'\,\qt{\va}\,\qt{\vas}))}
 \end{alignat*}
@@
 This can be generalized to splitting on any object value. In the Agda and
-Haskell implementations, we overload $\mit{split}$ with a class similar to the
-following:
+Haskell implementations, we overload it with a class similar to the following:
 \begin{alignat*}{3}
@@
   & \vf\,\vb\,\vs :=\\
-  & \ind \letdef\,\msf{act}\,\vs := \case\,\vb\,\of \True\hspace{0.2em}\to \letdef \vs := 10; \Just((), \vs)\\
-  & \ind\hspace{8.5em}                               \False \to \letdef \vs := 20; \Just((), \vs);\\
+  & \ind \letdef\,\msf{act}\,\vs := \case\,\vb\,\of \True\hspace{0.2em}\to \letdef \vs := 10; \Just\,((), \vs)\\
+  & \ind\hspace{8.5em}                               \False \to \letdef \vs := 20; \Just\,((), \vs);\\
   & \ind \case \msf{act}\,\vs\,\of\\
@@
 \end{alignat*}
-Here, $\msf{joinNothing}$ happens to be dead code, but it is easy to clean that up in
+Here, $\msf{joinNothing}$ happens to be dead code, but it is easy to clean up in
 downstream compilation. Such ``fused'' returns are possible whenever we have a
@@
 any other state, and each such transition is represented as a function call.
-This problem of generating well-typed mutual blocks was addressed by Yallop and
-Kiselyov in \cite{DBLP:conf/pepm/YallopK19}. In contrast to this work, which
-used control effects and mutable references in MetaOCaml, we present a solution
-that does not use side effects in the metalanguage.
+This problem of generating well-typed mutual blocks was addressed in
+\cite{DBLP:conf/pepm/YallopK19}. In contrast to this work, which used control
+effects and mutable references in MetaOCaml, we present a solution that does not
+use side effects in the metalanguage.
 
@@
 \end{itemize}
-Hence, we expect that during staging, every $\vf : \Lift\,\vA \to \VTy$ function
-has to be constant. This is actually true in the staging semantics of
-CFTT. There, all metafunctions are stable under object-level parallel
-substitutions. Also, object types are untouched by substitution. Hence, a
-straightforward unfolding of semantic definitions validates the constancy of
-$\vf$.
-
-Generally speaking, in the semantics, every function whose domain is a product
+Hence, we expect that during unstaging, every $\vf : \Lift\,\vA \to \VTy$
+function has to be constant. More generally in the semantics in Appendix \ref{appendix:A}, every function whose domain is a product
 of object types and whose codomain is a constant presheaf, is a constant
@@
 \\\\
-We remark that there is no risk of staging getting stuck on this axiom, because
-propositional equality proofs get erased, as we described in Section
-\ref{sec:semantics-of-staging}.
-
-%% We also remark that although generativity is inconsistent with inspecting the
-%% structure of object terms, it is consistent with inspecting the structure of
-%% object \emph{types}.
+See Section \ref{appendix:generativity} in Appendix $\ref{appendix:A}$ for a
+validation of this axiom. We remark that there is no risk of staging getting
+stuck, because propositional equality proofs get erased, as we described in
+Section \ref{sec:semantics-of-staging}. We also remark that although
+generativity is inconsistent with inspecting the structure of object terms, it
+is consistent with inspecting the structure of object types.
 
@@
 system which simultaneously supports intensional analysis through contextual
-types and the 2LTT-style implicit contexts.
+types and the 2LTT-style implicit contexts, and it would be interesting
+to consider in future work.
 
@@
       sufficient, since abstraction overheads are already eliminated by staging. Eliminating
-      dead code and unused arguments would be important, since we have not yet addressed
+      dead code and unused arguments would be important, as we have not yet addressed
       this through staging.
@@
      zipping for non-skipping ``synchronous'' streams. However, this could be
-     refined in many ways, and we could also try to represent the full transition
+     refined in many ways, and we could also try to represent the transition
      graph in an observable way, thereby enabling merging or deleting states.
@@
 \end{itemize}
+
+\section*{Acknowledgements}
+I thank the anonymous reviewers and Sebastian Graf, for their comments and suggestions
+on draft versions of this paper.
+
 \newpage
@@
   reuse metafunctions to represent object-level binders. Our Agda implementation
-  actually follows this style.
+  follows this style.
 \end{itemize}
@@
 So far we have stayed close to \cite{staged2ltt}, with only cosmetic
-changes. Supporting W-types, as a way to get inductive families
-\cite{whynotw}, is the biggest extension here. We extend Section 5 of
-\cite{staged2ltt} here. In the syntax of CFTT, $\vW$-types are as follows. The
-formal specification of syntactic $\vW{-}\msf{elim}$ is rather gnarly, because
-of the usage of explicit substitutions, but we write it out here anyways.
+changes. Supporting W-types, as a way to get inductive families \cite{whynotw},
+is the biggest extension here. We extend Section 5 of \cite{staged2ltt}. In the
+syntax of CFTT, $\vW$-types are as follows. The formal specification of
+syntactic $\vW{-}\msf{elim}$ is rather gnarly, because of the usage of explicit
+substitutions, but we write it out here anyways.
 
@@
 \end{alignat*}
-Here, $(B \Rightarrow (\vW\,A\,B)[\p])\rel$ is the relational interpretation of the function type, which unfolds to
-pointwise preservation of relations. In short, $\vW\rel$ expresses that $\wh{\vsup}\,\alpha\,f$ is related to $\re_\vsup\,\alpha'\,f'$
-when the arguments are (inductively) related. We can immediately interpret $\vW$ and $\vsup$:
-\begin{alignat*}{3}
-  & (\vW\,A\,B)\rel\,\gamma\rel\,t\,t' &&:= \vW\rel\,\gamma\rel\,t\,t'\\
-  & (\vsup\,t\,u)\rel\,\gamma\rel      &&:= \vsup\rel\,(t\rel\,\gamma\rel)\,(u\rel\,\gamma\rel)
-\end{alignat*}
-It only remains to interpret $\vW$-elimination and its $\beta$-rule. We need to define
-\[ (\vW{-}\msf{elim}\,P\,s\,w)\rel\,\gamma\rel : P\rel\,(\gamma\rel,\,w\rel\,\gamma\rel)\,(\ev\,(\welim\,P\,s\,w)\,\gamma)\,(\re\,(\welim\,P\,s\,w)\,\gamma') \]
+Here, $(B \Rightarrow (\vW\,A\,B)[\p])\rel$ is the relational interpretation of
+the function type, which unfolds to pointwise preservation of relations. We also
+define $(\vW\,A\,B)\rel\,\gamma\rel\,t\,t'$ to be $\vW\rel\,\gamma\rel\,t\,t'$.
+In short, $\vW\rel$ expresses that $\wh{\vsup}\,\alpha\,f$ is related to
+$\re_\vsup\,\alpha'\,f'$ when the arguments are (inductively) related. Similarly, we
+interpret $(\vsup\,t\,u)\rel\,\gamma\rel$ as
+$\vsup\rel\,(t\rel\,\gamma\rel)\,(u\rel\,\gamma\rel)$. It only remains to
+interpret $\vW$-elimination and its $\beta$-rule. We need to define
+\[ (\vW{-}\msf{elim}\,P\,s\,w)\rel\,\gamma\rel : P\rel\,(\gamma\rel,\,w\rel\,\gamma\rel)\,(\ev\,(\welim\,P\,s\,w)\,\gamma)\,(\re\,(\welim\,P\,s\,w)\,\gamma'). \]
 We do this by induction on $w\rel\,\gamma\rel :
@@
 
-\subsection{Generativity}
+\subsection{Generativity} \label{appendix:generativity}
 

### 2024-06-18 old/icfp24_old/response.txt
move old icfp24 files

Review A
--------------------------------------------------------------------------------

Questions:

- What are the main formal/"soundness" results from the previous paper(s) on
  two-level type theory that section 2.4 extends?"

The main result is soundness, which means that in each program that has
an object-level type and only depends on object-level free variables, all quotes
and splices can be computed away, and the result is definitionally equal to what
we started with. This is theorem 5.2. in Kovács 22. There's also "completeness"
(roughly meaning that every object program can be represented in the staged
language) and "strictness" (meaning that the staging algorithm performs no
beta-eta conversion on object terms).

However, we glossed over completeness for CFTT (because it's easy and
mechanically follows Kovács22) and strictness holds trivially for CFTT since
there are no beta-eta rules in the object theory. Although now we think that we
should include these points in the paper revisions.

- How does the code size/complexity of your approach compare to say Haskell
  standard library definitions of monad / monad transformer operations?

For each supported transformer in the library, we need an extra Join and an
Improve instance. This is around 5 lines for each transformer in the our code
supplement. We also need let-inserting versions of operations like "put" and
"modify".

- How does performance compare on a like-for-like basis (say, with all normal
  GHC optimizations enabled and without)?
  ...
  What degree of change at the level of library/interface clients is/would be
  necessary for "normal" Haskell programs to make use of your approach?

The Template Haskell implementation is useful for TH power users but we do not
think that a) it can fully replace GHC optimizations b) ordinary users of monads
in Haskell would want to use the staged version.

For an example for a), if we write a staged "State (↑Int)" program, although the
overhead of the monadic binding will be eliminated, we still need to use GHC's
-O1 optimization to also unbox the involved "Int"-s. There are limitations on
memory layout control, data specialization and code deduplication in Haskell
which are annoying even when we use staging, and it's more convenient to just
leave some things to GHC's optimization.

For b), users would need to write explicit quotes and splices and reorganize
their code to comply with GHC stage restrictions (which means that TH code
cannot be in the same module as the splice that uses it!).

This is not to say that the TH implementation is useless; we're using it in some
of our TH code to good effect, it's just that our vision in Section 6 about
staged monads is more realistic in a standalone CFTT implementation or in a
major extension of an existing language.

Addressing other questions
------------------------------------------------------------

- discussion of ... defunctionalization ... lambda-lifting

We did cut out a discussion of defunctionalization from the paper, for lack of
space. In short, defunc by itself does not eliminate closures nor dynamic
control; a dynamic call to a heap-allocated closure becomes a dynamic switch on
a tag in a heap-allocated defunctionalized closure. In contrast, CFTT programs
compile to truly closure-free code where the only source of dynamic control flow is
case-switching on user-introduced algebraic data types. That said, if we add
closure types to CFTT, defunc can be still used in the usual way for those
closures.

As to lambda-lifting, it's actually required in CFTT compilation, for the
let-defined local functions that are not join points, i.e. called somewhere
in non-tail position.

- we get soundness of W-types

Yes, it is the soundness mentioned in our previous answer here. Perhaps more
precisely, it's "soundness of staging in the presence of W-types".

- cross-stage persistence

(This also addresses Reviewer B's related question)

CFTT in the current form does not support any CSP, like previous 2LTT-s. By this
we mean that it's not possible to write a definition that can be later reused at
both levels, without explicit conversions. For such a thing to be possible, we
would need syntax for a language fragment which can be interpreted into both
levels, perhaps as a form of "stage-polymorphism". Technically, the meta type
theory is expressive enough to embed languages that can be then interpreted into
both levels, but that would likely be less convenient to use than "native" stage
polymorphism.

We think though that the issue of code duplication across levels is mitigated in
CFTT. First, lots of code can't be duplicated because they're not even
expressible in the object language. The monads in the paper are like this: we
only have a meta-monad library. In other cases we don't really want to use the
same definitions across stages. E.g. we would use operations on push/pull
streams instead of operations on object-level lists, for most list
functions. For meta-level code we'd use ordinary non-fused lists instead,
because we expect to get sufficient performance at compile time without fancy
optimizations. Our impression of prior works in staging is that compile-time
performance tends to be good enough, and at the meta level it's better to focus on
abstraction, clarity and modularity instead of performance.

This still leaves plenty of code to be potentially duplicated,
e.g. implementations of standard immutable data structures, so there's certainly
reason to look at CSP or stage polymorphism in the future.

Thank you for the other comments as well, we'll take them into account in
revisions.

Review B
--------------------------------------------------------------------------------

Questions:

- In multiple systems with staging, only one of <~t> = t and ~<t> = t is true,
  and not always definitionally. Is this unconditionally true?

It is. It holds in the semantics in Kovács 22. Also, from an implementation
perspective, there is nothing fancy in the equational theories of 2LTT-s and
CFTT, compared to plain MLTT. Supporting ↑_ in a 2LTT is not more difficult
than supporting the following type in Agda:

  -- with eta rule
  record ↑ (A : Set₀) : Set₁ where
    constructor <_>
    field
      ~_ : A

- What about 'cross-stage persistence'?

See my answer to Reviewer A regarding CSP.

- Finally tagless for syntactic noise

The most important thing for noise-reduction is inference of quotes and
splices. It works very effectively with bidirectional elaboration, as
demonstrated in the Kovács22 supplement, mainly because universes of types carry
stage information (unlike in TH and MetaOCaml). Generally we only need to
disambiguate let-definitions, as we do in the current paper with "=" and ":=",
to be able to infer quotes and splices unambiguously. Hence, we consider
splice/quote noise to be basically solved but there may be other kinds of
staging noise in practice, like the up/down conversions.

Addressing other questions/comments
------------------------------------------------------------

- on "staging" terminology

Skimming through the literature, we're surprised and slightly horrified that
indeed, our "staging" is idiosyncratic terminology. We are open to changing
"staging" to "unstaging". Are there other choices though that we could use?

- Extra citations

We are very grateful for these pointers. The "functors and monads" and the
"combinatorial circuits" papers seem to be very relevant and we'll be sure to
mention these in the contributions section already.

As to Section 3.2, we did not intend to suggest that we invented anything
related to let-insertion and the Trick, but we do agree that we should make this
more obvious by including citations there (and not just in the related works
section). We'll also include the SOP & strymonas citations.

- "universe reflection" & "finitely presented"

In the HoTT-flavored 2LTT literature, the former is called "fibrant" and the
latter is called "cofibrant". We found these terms un-enlightening in the staged
programming context, so we skipped them, but we could mention them in a
footnote. About IsSOP, we conjecture that in CFTT it precisely captures the
notion of "finite presentation" or "cofibrancy", i.e. every cofibrant type is
isomorphic to a SOP.

- Section 2.4

This section is intended to be read as an appendix of Kovács22; it's a modest
extension but still requires a lot of technical background from there. This is a
bit awkward but we're not sure what to do. The addition of W-types and identity
is not a significant contribution by itself but we also don't want to just
handwave it. Maybe it could be turned into an appendix of this paper?

- exotic terms in HOAS

Exotic terms in CFTT are ruled out by the soundness & completeness of staging.
This is closely related to "adequacy" of HOAS in logical frameworks.

- The mixture of meta-language and object language can be very hard to read at times.

We agree that coloring could help and we could add it in a revision of the paper.

Also thanks for the other comments, we'll take them into account.

Review C
--------------------------------------------------------------------------------

Questions:

- In section 3.5, what becomes of the original example?

  The code output looks as if "modify' (λ x. <~x + 10>)"
  was inlined in the case branches:
  λ s. case b of
    True  → let s := 10; let s := s + 10; ((), s)
    False → let s := 20; let s := s + 10; ((), s)

  The fix would indeed have "join $ case b of ..."
  In a revision of the paper we could include these examples too.

- Extraction from Agda code

It wouldn't be hard! We could write a non-total interpreter for the object
language in Agda, and extract code for that, or just print Haskell code as
strings. We didn't implement these because we already had most of the examples
in executable form in Template Haskell.

- This example looks very much toyish and unmotivated

We don't think that it's toyish. In GHC it took significant engineering effort
and lots of tuning to reliably get similar code optimization for similar code
examples. There are lots of redexes and unfoldings in the non-optimized monadic
code, and GHC has to compute them using a general algorithm that also has to avoid
code size blow-ups.

- Rupicola: thanks, we will check this work out.

Addressing other questions/comments
------------------------------------------------------------

- "A weakness of the paper is that the intended model underlying the soundness
  of the approach is sketched very briefly."

We think that Section 2.4 is not very technically difficult when taken together with
Kovács22, for which it is basically an addendum; see also our answer to "Section
2.4" for Review B. We could spell out more details in an extended revision of
the paper or perhaps put them in an appendix.

- "MonadGen M": yes, we introduced things in a wrong order
- "reuse essentially all definitions from mtl": We'll write something clearer,
   here we meant mtl in a more language-agnostic way.

We'll act on the other comments as well.

### 2024-06-18 old/icfp24_old/old/submission/haskell-cftt/CFTT/Join.hs

@@ -1,0 +1,53 @@
+
+module CFTT.Join where
+
+import Data.Kind
+import Control.Monad.Except
+import Control.Monad.Reader
+import Control.Monad.State.Strict
+import Control.Monad.Trans.Maybe
+
+import CFTT.Up
+import CFTT.Gen
+import CFTT.SOP
+
+-- Functions from Elₛ to Up
+--------------------------------------------------------------------------------
+
+type family FunSU (a ∷ Uₛ) (b ∷ Type) ∷ Type where
+  FunSU '[]       b = ()
+  FunSU (a ': as) b = (Up (Funₚₜ a b), FunSU as b)
+
+tabulate ∷ ∀ a b. Sing a → (Elₛ a → Up b) → FunSU a b
+tabulate SNil           f = ()
+tabulate (SCons as ass) f = (lamₚₜ as (f . Here), tabulate ass (f . There))
+
+index ∷ FunSU a b → Elₛ a → Up b
+index fs (Here x)  = appₚₜ (fst fs) x
+index fs (There x) = index (snd fs) x
+
+genFunSU ∷ ∀ a b. Sing a → FunSU a b → Gen (FunSU a b)
+genFunSU SNil         fs = pure ()
+genFunSU (SCons a as) fs = (,) <$> gen (fst fs) <*> genFunSU @_ @b as (snd fs)
+
+class MonadJoin m where
+  join :: IsSOP a => m a -> m a
+
+instance MonadJoin Gen where
+  join ma = Gen \(k :: a -> Up r) -> runGen do
+    let sa = singRep @a
+    conts <- genFunSU @_ @r sa (tabulate sa (k . decode))
+    a <- ma
+    pure $ index conts (encode a)
+
+instance (IsSOP e, MonadJoin m) => MonadJoin (ExceptT e m) where
+  join (ExceptT ma) = ExceptT $ join ma
+
+instance (MonadJoin m, IsSOP s) => MonadJoin (StateT s m) where
+  join (StateT ma) = StateT (join . ma)
+
+instance MonadJoin m => MonadJoin (MaybeT m) where
+  join (MaybeT ma) = MaybeT $ join ma
+
+instance (MonadJoin m) => MonadJoin (ReaderT r m) where
+  join (ReaderT ma) = ReaderT (join . ma)

### 2024-06-18 icfp24paper/supplement/haskell-cftt/CFTT/Examples.hs
stuff

@@ -1,0 +1,221 @@
+
+module CFTT.Examples where
+
+import Control.Monad.Identity
+import Control.Monad.Except
+import Control.Monad.State.Strict hiding (modify')
+import qualified Control.Monad.State.Strict as State
+import Control.Monad.Trans.Maybe
+import Language.Haskell.TH
+
+import CFTT.Gen
+import CFTT.Split
+import CFTT.Join
+import CFTT.Up
+import CFTT.Pull
+import CFTT.Improve
+import CFTT.Tree
+import qualified CFTT.Up as Up
+import qualified CFTT.Pull as P
+
+--------------------------------------------------------------------------------
+
+printUp :: Up a -> IO ()
+printUp a = do
+  x <- unType <$> runQ (examineCode a)
+  print $ ppr x
+
+
+--------------------------------------------------------------------------------
+
+-- Code size exponential in the number of caseM-s here, because
+-- everything gets inlined in case branches.
+exM1 :: Int -> StateT Int (MaybeT Identity) ()
+exM1 x = $$(down do
+  caseM ([||x||] Up.== 10) \case
+    True -> modify'  (+10)
+    False -> modify' (+20)
+  caseM ([||x||] Up.== 10) \case
+    True -> modify'  (+10)
+    False -> modify' (+20)
+  caseM ([||x||] Up.== 10) \case
+    True -> modify'  (+10)
+    False -> modify' (+20)
+  caseM ([||x||] Up.== 10) \case
+    True  -> modify' (+10)
+    False -> modify' (+20)
+  )
+
+
+-- Code size is linear.
+exM2 :: Int -> StateT Int (MaybeT Identity) ()
+exM2 x = $$(down do
+  join $ caseM ([||x||] Up.== 10) \case
+    True  -> modify  (+10)
+    False -> modify (+20)
+  join $ caseM ([||x||] Up.== 10) \case
+    True -> modify  (+10)
+    False -> modify (+20)
+  join $ caseM ([||x||] Up.== 10) \case
+    True -> modify  (+10)
+    False -> modify (+20)
+  caseM ([||x||] Up.== 10) \case
+    True  -> modify' (+10)
+    False -> modify' (+20)
+  )
+
+type M = StateT Int (ExceptT () Identity)
+
+-- The "fail" branches jump immediately to the "catch" code
+exM3 :: Int -> M ()
+exM3 x = $$(down do
+  catchError (join $ do
[…]

### 2024-06-18 icfp24paper/supplement/haskell-cftt/CFTT/Tree.hs

@@ -1,0 +1,43 @@
+{-# language ScopedTypeVariables #-}
+
+module CFTT.Tree where
+
+import Data.Typeable
+import CFTT.Up
+import CFTT.Gen
+import CFTT.Split
+import CFTT.Pull
+import CFTT.SOP
+
+data Tree a = Leaf | Node a (Tree a) (Tree a)
+
+instance Split (Tree a) (Maybe (Up a, Up (Tree a), Up (Tree a))) where
+  splitGen x = Gen \k -> [|| case $$x of
+    Leaf -> $$(k Nothing)
+    Node a l r -> $$(k (Just ([||a||], [||l||], [||r||]))) ||]
+
+instance Typeable a => CasePull (Tree a) (Maybe (Up a, Up (Tree a), Up (Tree a))) where
+  casePull x f = case (f Nothing, unravel3 (f . Just)) of
+    (Pull @s skip seed step, Pull' @s' skip' seed' step') ->
+      Pull @( (Either s (Up a, (Up (Tree a), Up (Tree a)), s'))) (skip || skip')
+           (caseM x \case Nothing        -> Left <$> seed
+                          Just (a, l, r) -> do s' <- seed' $ Cons a $ Cons l $ Cons r Nil
+                                               pure $ Right (a, (l, r), s'))
+         \case
+         (Left s) -> step s >>= \case
+          Stop      -> pure Stop
+          Skip s    -> pure $ Skip ( (Left s))
+          Yield c s -> pure $ Yield c ( (Left s))
+         (Right (a, (l, r), s)) -> step' (Cons a $ Cons l $ Cons r Nil) s >>= \case
+          Stop      -> pure Stop
+          Skip s    -> pure $ Skip (Right (a, (l, r), s))
+          Yield c s -> pure $ Yield c ((Right (a, (l, r), s)))
+
+
+-- myMap :: forall (a :: Up Type)(b :: Up Type). (Up $$a -> Up $$b) -> Up [$$a] -> Up [$$b]
+-- myMap f as = [||
+--    let go :: [$$a] -> [$$b]
+--        go []     = []
+--        go (a:as) = $$(f [||a||]) : go as
+--    in go $$as
+--   ||]

### 2024-06-18 icfp24paper/paper.tex
fix

@@
   &\vf :: \Int \to \Reader\,\Bool\,\Int \\
-  &\vf\,\vx = \mdo\{\vb \fro \msf{ask}; \bif \vx\, \bthen \return (\vx + 10)\, \belse \return (\vx + 20)\}
+  &\vf\,\vx = \mdo\{\vb \fro \msf{ask}; \bif \vb\, \bthen \return (\vx + 10)\, \belse \return (\vx + 20)\}
 \end{alignat*}
@@
 In this appendix we describe the semantics of CFTT, from which the unstaging
-algorithm is obtained, and also describe the soundness and completeness of the
+algorithm is obtained, and also describe the soundness and stability of the
 algorithm. We borrow the setup from \cite{staged2ltt}; we use terminology and

### 2024-06-18 icfp24paper/paper.tex
fix

@@
 we need to witness the motive in the case where $w$ is a $\vsup$, which causes
-both $\ev\,(\welim\,P\,s\,w)\,\gamma)$ and $\re\,(\welim\,P\,s\,w)\,\gamma')$ to
+both $\ev\,(\welim\,P\,s\,w)\,\gamma$ and $\re\,(\welim\,P\,s\,w)\,\gamma'$ to
 compute further to applications of $\ev\,s\,\gamma$ and $\re\,s\,\gamma'$

### 2024-06-18 icfp24paper/paper.tex
fix

@@
 witnessing the inductive motive for every output of $f\rel$. In the body here,
-we need to witness the motive in the case where $w$ is a $\vsup$, which causes
+we need to witness the motive in the case where $w\rel\,\gamma\rel$ is a $\vsup\rel$, which causes
 both $\ev\,(\welim\,P\,s\,w)\,\gamma$ and $\re\,(\welim\,P\,s\,w)\,\gamma'$ to

### 2024-06-19 icfp24paper/paper.tex
fix

@@
 It might make sense to also have a simple ``desugaring'' rule that inserts a
-$\join$ whenever we have a object-level case split with more than one branch. At
+$\join$ whenever we have an object-level case split with more than one branch. At
 worst, this generates some noise and dead code that is easy to remove by

### 2024-07-02 icfp24paper/paper.tex
fix
acknowledgements

--- 2024-06-18 icfp24paper/paper.tex
@@
 \title{Closure-Free Functional Programming in a Two-Level Type Theory}
-%% \titlenote{}
 
@@
   &\vf :: \Int \to \Reader\,\Bool\,\Int \\
-  &\vf\,\vx = \mdo\{\vb \fro \msf{ask}; \bif \vb\, \bthen \return (\vx + 10)\, \belse \return (\vx + 20)\}
+  &\vf\,\vx = \mdo\{\vb \fro \msf{ask}; \bif \vx\, \bthen \return (\vx + 10)\, \belse \return (\vx + 20)\}
 \end{alignat*}
@@
 \section*{Acknowledgements}
-I thank the anonymous reviewers and Sebastian Graf, for their comments and suggestions
-on draft versions of this paper.
+I thank the anonymous reviewers, Sebastian Graf and Jacques Carette, for their
+comments and suggestions on draft versions of this paper. This work was supported
+by grant 2019.0116 of the Knut and Allice Wallenberg Foundation.
 
@@
 In this appendix we describe the semantics of CFTT, from which the unstaging
-algorithm is obtained, and also describe the soundness and stability of the
+algorithm is obtained, and also describe the soundness and completeness of the
 algorithm. We borrow the setup from \cite{staged2ltt}; we use terminology and
@@
 we need to witness the motive in the case where $w$ is a $\vsup$, which causes
-both $\ev\,(\welim\,P\,s\,w)\,\gamma$ and $\re\,(\welim\,P\,s\,w)\,\gamma'$ to
+both $\ev\,(\welim\,P\,s\,w)\,\gamma)$ and $\re\,(\welim\,P\,s\,w)\,\gamma')$ to
 compute further to applications of $\ev\,s\,\gamma$ and $\re\,s\,\gamma'$

### 2024-07-02 old/icfp24_old/hs-agda/haskell-cftt/CFTT/Pull.hs

@@ -1,322 +1,0 @@
-{-# language MagicHash #-}
-
-module CFTT.Pull where
-
-import Data.Typeable
-import Data.Kind
-
-import CFTT.Gen
-import CFTT.SOP
-import CFTT.Split
-import CFTT.Up
-
-import GHC.Exts
-
---------------------------------------------------------------------------------
-
-impossible :: a
-impossible = error "impossible"
-
-data Step s a = Stop | Yield a s | Skip s deriving Functor
-
-data Pull a where
-  Pull :: forall s a. (IsSOP s, Typeable s) => Bool -> Gen s -> (s -> Gen (Step s a)) -> Pull a
-
---------------------------------------------------------------------------------
-
-infixl 1 <&>
-(<&>) :: Functor f => f a -> (a -> b) -> f b
-(<&>) = flip fmap
-
-repeat :: a -> Pull a
-repeat a = Pull False (pure ()) \_ -> pure $ Yield a ()
-
-instance Functor Pull where
-  fmap f (Pull skips seed step) = Pull skips seed (fmap (fmap (fmap f)) step)
-
-infixl 4 <:>
-(<:>) :: ∀ a b. (IsSOP a, Typeable a) => Pull (a → b) → Pull a → Pull b
-(<:>) (Pull @s False seed step) (Pull @s' False seed' step') =
-  Pull False ((,) <$> seed <*> seed') \(s, s') -> step s >>= \case
-    Stop      -> pure Stop
-    Skip _    -> impossible
-    Yield f s -> step' s' >>= \case
-      Stop       -> pure Stop
-      Skip _     -> impossible
-      Yield a s' -> pure $ Yield (f a) (s, s')
-(<:>) (Pull @s True seed step) (Pull @s' False seed' step') =
-  Pull True ((,) <$> seed <*> seed') \(s, s') ->
-    step s >>= \case
-      Stop        → pure Stop
-      (Skip s)    → pure $ Skip (s , s')
-      (Yield f s) → step' s' >>= \case
-        Stop         → pure Stop
-        (Skip s')    → impossible
-        (Yield a s') → pure $ Yield (f a) (s , s')
-(<:>) (Pull @s False seed step) (Pull @s' True seed' step') =
-  Pull True ((,) <$> seed <*> seed') \case
-    (s, s') → step' s' >>= \case
-      Stop         → pure Stop
-      (Skip s')    → impossible
[…]

### 2024-07-02 old/icfp24_old/hs-agda/haskell-cftt/CFTT/Tree.hs

@@ -1,33 +1,0 @@
-
-module CFTT.Tree where
-
-import Data.Typeable
-import CFTT.Up
-import CFTT.Gen
-import CFTT.Split
-import CFTT.Pull
-import CFTT.SOP
-
-data Tree a = Leaf | Node a (Tree a) (Tree a)
-
-instance Split (Tree a) (Maybe (Up a, Up (Tree a), Up (Tree a))) where
-  splitGen x = Gen \k -> [|| case $$x of
-    Leaf -> $$(k Nothing)
-    Node a l r -> $$(k (Just ([||a||], [||l||], [||r||]))) ||]
-
-instance Typeable a => CasePull (Tree a) (Maybe (Up a, Up (Tree a), Up (Tree a))) where
-  casePull x f = case (f Nothing, unravel3 (f . Just)) of
-    (Pull @s skip seed step, Pull' @s' skip' seed' step') ->
-      Pull @( (Either s (Up a, (Up (Tree a), Up (Tree a)), s'))) (skip || skip')
-           (caseM x \case Nothing        -> Left <$> seed
-                          Just (a, l, r) -> do s' <- seed' $ Cons a $ Cons l $ Cons r Nil
-                                               pure $ Right (a, (l, r), s'))
-         \case
-         (Left s) -> step s >>= \case
-          Stop      -> pure Stop
-          Skip s    -> pure $ Skip ( (Left s))
-          Yield c s -> pure $ Yield c ( (Left s))
-         (Right (a, (l, r), s)) -> step' (Cons a $ Cons l $ Cons r Nil) s >>= \case
-          Stop      -> pure Stop
-          Skip s    -> pure $ Skip (Right (a, (l, r), s))
-          Yield c s -> pure $ Yield c ((Right (a, (l, r), s)))

### 2024-07-02 old/icfp24_old/old/hs/Lib.hs

@@ -1,422 +1,0 @@
-
-{-# language LambdaCase, TemplateHaskell, BlockArguments, RankNTypes,
-   MultiParamTypeClasses, FunctionalDependencies, TypeApplications,
-   ScopedTypeVariables, UndecidableInstances, TypeFamilies,
-   TypeFamilyDependencies, PartialTypeSignatures, AllowAmbiguousTypes,
-   DataKinds, PolyKinds, MagicHash, ViewPatterns #-}
-
-{-# options_ghc -Wincomplete-patterns #-}
-
-module Lib where
-
-import Control.Applicative
-import Control.Monad hiding (join)
-import Control.Monad.Except
-import Control.Monad.Identity
-import Control.Monad.Reader
-import Control.Monad.State.Strict hiding (modify')
-import Control.Monad.Trans
-import Control.Monad.Trans.Maybe
-import Data.Coerce
-import Data.Kind
-import GHC.Exts
-import Language.Haskell.TH hiding (ListT, ListE, Type, BindS)
-import Data.Typeable
-
-import Prelude
-import qualified Prelude as P
-
-import Up
-import SOP
-
---------------------------------------------------------------------------------
-
-newtype Gen a = Gen {unGen :: forall r. (a -> Up r) -> Up r}
-
-runGen :: Gen (Up a) -> Up a
-runGen (Gen f) = f id
-
-instance Functor Gen where
-  fmap f ma = Gen \k -> unGen ma \a -> k (f a)
-
-instance Applicative Gen where
-  pure a = Gen \k -> k a
-  (<*>) gf ga = Gen \k -> unGen gf \f -> unGen ga \a -> k (f a)
-
-instance Monad Gen where
-  return = pure
-  (>>=) ga f = Gen \k -> unGen ga \a -> unGen (f a) k
-
-instance MonadFail Gen where
-  fail = error
-
-class Monad m => MonadGen m where
-  liftGen :: Gen a -> m a
-
-instance MonadGen Gen where
-  liftGen = id
-
-instance MonadGen m => MonadGen (StateT s m) where
-  liftGen = lift . liftGen
-
-instance MonadGen m => MonadGen (ReaderT r m) where
-  liftGen = lift . liftGen
-
-instance MonadGen m => MonadGen (ExceptT e m) where
-  liftGen = lift . liftGen
-
-instance MonadGen m => MonadGen (MaybeT m) where
[…]

### 2024-07-02 old/icfp24_old/old/hs/ixedsop/LibTest.hs

@@ -1,23 +1,0 @@
-
-{-# language LambdaCase, TemplateHaskell, BlockArguments, RankNTypes,
-   MultiParamTypeClasses, FunctionalDependencies, TypeApplications,
-   ScopedTypeVariables, UndecidableInstances, TypeFamilies,
-   TypeFamilyDependencies, PartialTypeSignatures, AllowAmbiguousTypes,
-   DataKinds, PolyKinds, MagicHash #-}
-
-module LibTest where
-
-import Prelude
-import qualified Prelude as P
-
-import Up
-import Lib
-import Control.Monad.State.Strict
-
-foo :: Int -> ((), Int)
-foo = runState $$(down $ do
-  -- modifyG (\x -> x + 10)
-  -- modifyG (\x -> x + 10)
-  modify (\x -> x + x)
-  modify (\x -> x + x)
-  pure Up.tt)

### 2024-07-02 old/icfp24_old/old/submission/haskell-cftt/CFTT/Examples.hs

@@ -1,212 +1,0 @@
-
-module CFTT.Examples where
-
-import Control.Monad.Identity
-import Control.Monad.Except
-import Control.Monad.State.Strict hiding (modify')
-import qualified Control.Monad.State.Strict as State
-import Control.Monad.Trans.Maybe
-import Language.Haskell.TH
-
-import CFTT.Gen
-import CFTT.Split
-import CFTT.Join
-import CFTT.Up
-import CFTT.Pull
-import CFTT.MonadTailCall
-import CFTT.Improve
-import CFTT.Tree
-import qualified CFTT.Up as Up
-import qualified CFTT.Pull as P
-
---------------------------------------------------------------------------------
-
-printUp :: Up a -> IO ()
-printUp a = do
-  x <- unType <$> runQ (examineCode a)
-  print $ ppr x
-
-
---------------------------------------------------------------------------------
-
--- Code size exponential in the number of caseM-s here, because
--- everything gets inlined in case branches.
-exM1 :: Int -> StateT Int (MaybeT Identity) ()
-exM1 x = $$(down do
-  caseM ([||x||] Up.== 10) \case
-    True -> modify'  (+10)
-    False -> modify' (+20)
-  caseM ([||x||] Up.== 10) \case
-    True -> modify'  (+10)
-    False -> modify' (+20)
-  caseM ([||x||] Up.== 10) \case
-    True -> modify'  (+10)
-    False -> modify' (+20)
-  caseM ([||x||] Up.== 10) \case
-    True  -> modify' (+10)
-    False -> modify' (+20)
-  )
-
-
--- Code size is linear.
-exM2 :: Int -> StateT Int (MaybeT Identity) ()
-exM2 x = $$(down do
-  join $ caseM ([||x||] Up.== 10) \case
-    True  -> modify  (+10)
-    False -> modify (+20)
-  join $ caseM ([||x||] Up.== 10) \case
-    True -> modify  (+10)
-    False -> modify (+20)
-  join $ caseM ([||x||] Up.== 10) \case
-    True -> modify  (+10)
-    False -> modify (+20)
-  caseM ([||x||] Up.== 10) \case
-    True  -> modify' (+10)
-    False -> modify' (+20)
-  )
-
-type M = StateT Int (ExceptT () Identity)
-
--- The "fail" branches jump immediately to the "catch" code
-exM3 :: Int -> M ()
[…]

### 2024-07-02 old/icfp24_old/old/submission/haskell-cftt/CFTT/Gen.hs

@@ -1,68 +1,0 @@
-
-module CFTT.Gen where
-
-import Control.Monad.Except
-import Control.Monad.Reader
-import Control.Monad.State.Strict hiding (modify')
-import Control.Monad.Trans.Maybe
-import CFTT.Up
-
-newtype Gen a = Gen {unGen :: forall r. (a -> Up r) -> Up r}
-
-runGen :: Gen (Up a) -> Up a
-runGen (Gen f) = f id
-
-instance Functor Gen where
-  fmap f ma = Gen \k -> unGen ma \a -> k (f a)
-
-instance Applicative Gen where
-  pure a = Gen \k -> k a
-  (<*>) gf ga = Gen \k -> unGen gf \f -> unGen ga \a -> k (f a)
-
-instance Monad Gen where
-  return = pure
-  (>>=) ga f = Gen \k -> unGen ga \a -> unGen (f a) k
-
-instance MonadFail Gen where
-  fail = error
-
-class Monad m => MonadGen m where
-  liftGen :: Gen a -> m a
-
-instance MonadGen Gen where
-  liftGen = id
-
-instance MonadGen m => MonadGen (StateT s m) where
-  liftGen = lift . liftGen
-
-instance MonadGen m => MonadGen (ReaderT r m) where
-  liftGen = lift . liftGen
-
-instance MonadGen m => MonadGen (ExceptT e m) where
-  liftGen = lift . liftGen
-
-instance MonadGen m => MonadGen (MaybeT m) where
-  liftGen = lift . liftGen
-
-gen :: MonadGen m => Up a -> m (Up a)
-gen a = liftGen $ Gen \k -> [|| let x = $$a in seq x $$(k [||x||]) ||]
-
-genRec :: MonadGen m => (Up a -> Up a) -> m (Up a)
-genRec a = liftGen $ Gen \k -> [|| let x = $$(a [||x||]) in $$(k [||x||]) ||]
-
-put' :: (MonadState (Up s) m, MonadGen m) => Up s -> m (Up ())
-put' s = do
-  s <- gen s
-  put s
-  pure [||()||]
-
-modify' :: (MonadState (Up s) m, MonadGen m) => (Up s -> Up s) -> m (Up ())
-modify' f = do
-  s <- get
-  put' (f s)
-
-local' :: (MonadReader (Up r) m, MonadGen m) => (Up r -> Up r) -> m a -> m a
-local' f ma = do
-  r <- ask
-  r <- gen (f r)
-  local (const r) ma

### 2024-07-02 old/icfp24_old/old/submission/haskell-cftt/CFTT/Split.hs

@@ -1,42 +1,0 @@
-
-module CFTT.Split where
-
-import CFTT.Up
-import CFTT.Gen
-
-class Split a b | a -> b, b -> a where
-  splitGen :: Up a -> Gen b
-
-split :: MonadGen m => Split a b => Up a -> m b
-split a = liftGen (splitGen a)
-
-caseM :: (MonadGen m) => Split a b => Up a -> (b -> m c) -> m c
-caseM a f = split a >>= f
-
-instance Split Bool Bool where
-  splitGen x = Gen \k -> [|| case $$x of
-    True -> $$(k True)
-    _    -> $$(k False) ||]
-
-instance Split [a] (Maybe (Up a, Up [a])) where
-  splitGen x = Gen \k -> [|| case $$x of
-    []   -> $$(k Nothing)
-    a:as -> $$(k (Just ([||a||], [||as||]))) ||]
-
-instance Split (a, b) (Up a, Up b) where
-  splitGen x = Gen \k -> [|| case $$x of
-    (a, b) -> $$(k ([||a||], [||b||])) ||]
-
-instance Split (Pair a b) (Pair (Up a) (Up b)) where
-  splitGen x = Gen \k -> [|| case $$x of
-    Pair a b -> $$(k (Pair [||a||] [||b||])) ||]
-
-instance Split (Either a b) (Either (Up a) (Up b)) where
-  splitGen x = Gen \k -> [|| case $$x of
-    Left a  -> $$(k (Left [||a||]))
-    Right b -> $$(k (Right [||b||])) ||]
-
-instance Split (Maybe a) (Maybe (Up a)) where
-  splitGen x = Gen \k -> [|| case $$x of
-    Nothing -> $$(k Nothing)
-    Just a  -> $$(k (Just [||a||])) ||]

### 2024-07-02 old/icfp24_old/supplement/haskell-cftt/CFTT/Examples.hs

@@ -1,221 +1,0 @@
-
-module CFTT.Examples where
-
-import Control.Monad.Identity
-import Control.Monad.Except
-import Control.Monad.State.Strict hiding (modify')
-import qualified Control.Monad.State.Strict as State
-import Control.Monad.Trans.Maybe
-import Language.Haskell.TH
-
-import CFTT.Gen
-import CFTT.Split
-import CFTT.Join
-import CFTT.Up
-import CFTT.Pull
-import CFTT.Improve
-import CFTT.Tree
-import qualified CFTT.Up as Up
-import qualified CFTT.Pull as P
-
---------------------------------------------------------------------------------
-
-printUp :: Up a -> IO ()
-printUp a = do
-  x <- unType <$> runQ (examineCode a)
-  print $ ppr x
-
-
---------------------------------------------------------------------------------
-
--- Code size exponential in the number of caseM-s here, because
--- everything gets inlined in case branches.
-exM1 :: Int -> StateT Int (MaybeT Identity) ()
-exM1 x = $$(down do
-  caseM ([||x||] Up.== 10) \case
-    True -> modify'  (+10)
-    False -> modify' (+20)
-  caseM ([||x||] Up.== 10) \case
-    True -> modify'  (+10)
-    False -> modify' (+20)
-  caseM ([||x||] Up.== 10) \case
-    True -> modify'  (+10)
-    False -> modify' (+20)
-  caseM ([||x||] Up.== 10) \case
-    True  -> modify' (+10)
-    False -> modify' (+20)
-  )
-
-
--- Code size is linear.
-exM2 :: Int -> StateT Int (MaybeT Identity) ()
-exM2 x = $$(down do
-  join $ caseM ([||x||] Up.== 10) \case
-    True  -> modify  (+10)
-    False -> modify (+20)
-  join $ caseM ([||x||] Up.== 10) \case
-    True -> modify  (+10)
-    False -> modify (+20)
-  join $ caseM ([||x||] Up.== 10) \case
-    True -> modify  (+10)
-    False -> modify (+20)
-  caseM ([||x||] Up.== 10) \case
-    True  -> modify' (+10)
-    False -> modify' (+20)
-  )
-
-type M = StateT Int (ExceptT () Identity)
-
--- The "fail" branches jump immediately to the "catch" code
-exM3 :: Int -> M ()
-exM3 x = $$(down do
-  catchError (join $ do
[…]

### 2024-07-02 icfp24paper/paper.tex

--- 2024-06-19 icfp24paper/paper.tex
@@
 \end{itemize}
-Here, we intentionally tuned the definition for a nice-looking staging output.
+Here, we intentionally tuned the definition for a nice-looking unstaged output.
 However, we could also just default to ``safe'' choices everywhere: we could use

### 2024-07-04 icfp24paper/paper.tex
update artifact, add data availability, add acknowledgement

@@
 \end{alignat*}
-Since the $y$ expression is inlined under a $\case$ branch at every use site, it
+Since the $y$ expression is inlined under a $\mbf{case}$ branch at every use site, it
 is only computed at runtime when $x$ evaluates to $\True$. In many situations,
@@
 by grant 2019.0116 of the Knut and Allice Wallenberg Foundation.
+
+\section*{Data Availability Statement}
+The Agda and Haskell artifacts are available in \cite{cftt-artifacts}.
 
@@
   type) is modeled as a presheaf over $\mbbo$. The interpretation of meta-level
-  type formers is standard, following Hofmann \cite{Hofmann97syntaxand} and
-  Huber \cite{huber-thesis}.
+  type formers is standard, following \cite{Hofmann97syntaxand} and \cite{huber-thesis}.
 \item $\Ty : \MTy_0\,\Gamma$ is the displayed presheaf which is constantly

### 2024-07-04 icfp24paper/paper.tex
cosmetics
camera ready paper

@@
 %% \documentclass[nonacm,acmsmall]{acmart}
-\documentclass[acmsmall,screen]{acmart}
+\documentclass[acmsmall]{acmart}
 %% \raggedbottom
@@
 
-%% \setcopyright{rightsretained}
-%% \acmPrice{}
-%% \acmDOI{10.1145/3547641}
-%% \acmYear{2022}
-%% \copyrightyear{2022}
-%% \acmSubmissionID{icfp22main-p52-p}
-%% \acmJournal{PACMPL}
-%% \acmVolume{6}
-%% \acmNumber{ICFP}
-%% \acmArticle{110}
-%% \acmMonth{8}
-
-%% These commands are for a JOURNAL article.
-
-%% \acmJournal{JACM}
-%% \acmVolume{37}
-%% \acmNumber{4}
-%% \acmArticle{111}
-%% \acmMonth{8}
-
-%%
-%% Submission ID.
-%% Use this when submitting an article to a sponsored event. You'll
-%% receive a unique submission ID from the organizers
-%% of the event, and this ID should be used as the parameter to this command.
-%%\acmSubmissionID{123-A56-BU3}
+\setcopyright{acmlicensed}
+\acmDOI{10.1145/3674648}
+\acmYear{2024}
+\copyrightyear{2024}
+\acmSubmissionID{icfp24main-p75-p}
+\acmJournal{PACMPL}
+\acmVolume{8}
+\acmNumber{ICFP}
+\acmArticle{259}
+\acmMonth{8}
+\received{2024-02-28}
+\received[accepted]{2024-06-18}
 
@@
 \title{Closure-Free Functional Programming in a Two-Level Type Theory}
-
-%% The "author" command and its associated commands are used to define
-%% the authors and their affiliations.
-%% Of note is the shared affiliation of the first two authors, and the
-%% "authornote" and "authornotemark" commands
-%% used to denote shared contribution to the research.
+\titlenote{This work was supported
+by grant 2019.0116 of the Knut and Alice Wallenberg Foundation.}
 
@@
 \orcid{0000-0002-6375-9781}
-\email{andrask@chalmers.se}
 \affiliation{%
   \institution{University of Gothenburg}
+  \city{Gothenburg}
   \country{Sweden}
-  \city{Gothenburg}
 }
-
-%% \author{András Kovács}
-%% \email{andrask@chalmers.se}
-%% \orcid{0000-0002-6375-9781}
-%% \affiliation{%
-%%   \institution{University of Gothenburg}
-%%   \country{Sweden}
-%%   \city{Gothenburg}
-%% }
+\email{andrask@chalmers.se}
 
@@
 I thank the anonymous reviewers, Sebastian Graf and Jacques Carette, for their
-comments and suggestions on draft versions of this paper. This work was supported
-by grant 2019.0116 of the Knut and Allice Wallenberg Foundation.
+comments and suggestions on draft versions of this paper.
 
@@
 
-\newpage
+%% --------------------------------------------------------------------------------
+%% --------------------------------------------------------------------------------
 
@@
 changes. Supporting W-types, as a way to get inductive families \cite{whynotw},
-is the biggest extension here. We extend Section 5 of \cite{staged2ltt}. In the
-syntax of CFTT, $\vW$-types are as follows. The formal specification of
-syntactic $\vW{-}\msf{elim}$ is rather gnarly, because of the usage of explicit
-substitutions, but we write it out here anyways.
+is the biggest extension here. We extend Section 5 of \cite{staged2ltt}. The
+formal specification of syntactic $\vW{-}\msf{elim}$ is rather gnarly, because
+of the usage of explicit substitutions, but we write it out here anyways.
 
@@
 \end{notation}
+
+In the syntax of CFTT, $\vW$-types are as follows.
 \begin{alignat*}{3}
@@
 Note that the type of $\re\,u\,\gamma$ is $\re\,(B[\id,\,t] \Rightarrow
-\vW\,A\,B)\,\gamma$, which is equal to $\re\,((B \Rightarrow
-(\vW\,A\,B)[\p])[\id,\,t])\,\gamma$, which is in turn equal to $\re\,(B
-\Rightarrow (\vW\,A\,B)[\p])\,(\gamma,\,\re\,t\,\gamma)$ by $\re$'s preservation
-of substitutions, which makes the above equation well-typed.
+\vW\,A\,B)\,\gamma$, which is equal to
+\[\re\,((B \Rightarrow(\vW\,A\,B)[\p])[\id,\,t])\,\gamma\]
+which is in turn equal to $\re\,(B \Rightarrow
+(\vW\,A\,B)[\p])\,(\gamma,\,\re\,t\,\gamma)$ by $\re$'s preservation of
+substitutions, which makes the above equation well-typed.
 

### 2024-07-04 icfp24paper/submission/haskell-cftt/CFTT/Improve.hs

@@ -1,59 +1,0 @@
-
-module CFTT.Improve where
-
-import Control.Monad.Identity
-import Control.Monad.Except
-import Control.Monad.Reader
-import Control.Monad.State.Strict
-import Control.Monad.Trans.Maybe
-
-import CFTT.Up
-import CFTT.Gen
-import CFTT.Split
-
--- paper 3.4
---------------------------------------------------------------------------------
-
-class MonadGen n => Improve m n | m -> n, n -> m where
-  up   :: Up (m a) -> n (Up a)
-  down :: n (Up a) -> Up (m a)
-
-instance Improve Identity Gen where
-  up x = Gen \k -> k [||runIdentity $$x||]
-  down x = unGen x \a -> [||Identity $$a||]
-
-instance (Improve m n) => Improve (StateT s m) (StateT (Up s) n) where
-
-  up x = StateT \s ->
-    do as <- up [|| runStateT $$x $$s ||]
-       split as
-
-  down x = [|| StateT \s -> $$(down
-    do (a, s) <- runStateT x [||s||]
-       pure [||($$a, $$s)||]
-       )||]
-
-instance (Improve m n) => Improve (ExceptT e m) (ExceptT (Up e) n) where
-  up x = ExceptT do
-    ea <- up [||runExceptT $$x||]
-    split ea
-
-  down (ExceptT x) =
-    [|| ExceptT $$(down (x >>= \case
-          Left e  -> pure [||Left $$e||]
-          Right a -> pure [||Right $$a||]
-                        )) ||]
-
-instance Improve m n => Improve (MaybeT m) (MaybeT n) where
-  up x = MaybeT do
-    ma <- up [||runMaybeT $$x||]
-    split ma
-
-  down (MaybeT x) =
-    [|| MaybeT $$(down (x >>= \case
-          Nothing -> pure [||Nothing||]
-          Just a -> pure [||Just $$a||])) ||]
-
-instance Improve m n => Improve (ReaderT r m) (ReaderT (Up r) n) where
-  up   x = ReaderT \r -> up [||runReaderT $$x $$r||]
-  down x = [|| ReaderT \r -> $$(down (runReaderT x [||r||])) ||]

### 2024-07-04 icfp24paper/submission/haskell-cftt/CFTT/Join.hs

@@ -1,54 +1,0 @@
-
-module CFTT.Join where
-
-import Data.Kind
-import Control.Monad.Except
-import Control.Monad.Reader
-import Control.Monad.State.Strict
-import Control.Monad.Trans.Maybe
-
-import CFTT.Up
-import CFTT.Gen
-import CFTT.SOP
-
--- paper 3.5
---------------------------------------------------------------------------------
-
--- Functions from Elₛ to Up
-type family FunSU (a ∷ Uₛ) (b ∷ Type) ∷ Type where
-  FunSU '[]       b = ()
-  FunSU (a ': as) b = (Up (Funₚₜ a b), FunSU as b)
-
-tabulate ∷ ∀ a b. Sing a → (Elₛ a → Up b) → FunSU a b
-tabulate SNil           f = ()
-tabulate (SCons as ass) f = (lamₚₜ as (f . Here), tabulate ass (f . There))
-
-index ∷ FunSU a b → Elₛ a → Up b
-index fs (Here x)  = appₚₜ (fst fs) x
-index fs (There x) = index (snd fs) x
-
-genFunSU ∷ ∀ a b. Sing a → FunSU a b → Gen (FunSU a b)
-genFunSU SNil         fs = pure ()
-genFunSU (SCons a as) fs = (,) <$> gen (fst fs) <*> genFunSU @_ @b as (snd fs)
-
-class MonadJoin m where
-  join :: IsSOP a => m a -> m a
-
-instance MonadJoin Gen where
-  join ma = Gen \(k :: a -> Up r) -> runGen do
-    let sa = singRep @a
-    conts <- genFunSU @_ @r sa (tabulate sa (k . decode))
-    a <- ma
-    pure $ index conts (encode a)
-
-instance (IsSOP e, MonadJoin m) => MonadJoin (ExceptT e m) where
-  join (ExceptT ma) = ExceptT $ join ma
-
-instance (MonadJoin m, IsSOP s) => MonadJoin (StateT s m) where
-  join (StateT ma) = StateT (join . ma)
-
-instance MonadJoin m => MonadJoin (MaybeT m) where
-  join (MaybeT ma) = MaybeT $ join ma
-
-instance (MonadJoin m) => MonadJoin (ReaderT r m) where
-  join (ReaderT ma) = ReaderT (join . ma)

### 2024-07-04 icfp24paper/submission/haskell-cftt/CFTT/Pull.hs

@@ -1,334 +1,0 @@
-{-# language MagicHash #-}
-
-module CFTT.Pull where
-
-import Data.Typeable
-import Data.Kind
-
-import CFTT.Gen
-import CFTT.SOP
-import CFTT.Split
-import CFTT.Up
-
-import GHC.Exts
-
--- paper section 4
---------------------------------------------------------------------------------
-
-impossible :: a
-impossible = error "impossible"
-
-data Step s a = Stop | Yield a s | Skip s deriving Functor
-
-data Pull a where
-  Pull :: forall s a. (IsSOP s, Typeable s) => Bool -> Gen s -> (s -> Gen (Step s a)) -> Pull a
-
-infixl 1 <&>
-(<&>) :: Functor f => f a -> (a -> b) -> f b
-(<&>) = flip fmap
-
-repeat :: a -> Pull a
-repeat a = Pull False (pure ()) \_ -> pure $ Yield a ()
-
-instance Functor Pull where
-  fmap f (Pull skips seed step) = Pull skips seed (fmap (fmap (fmap f)) step)
-
-infixl 4 <:>
-(<:>) :: ∀ a b. (IsSOP a, Typeable a) => Pull (a → b) → Pull a → Pull b
-(<:>) (Pull @s False seed step) (Pull @s' False seed' step') =
-  Pull False ((,) <$> seed <*> seed') \(s, s') -> step s >>= \case
-    Stop      -> pure Stop
-    Skip _    -> impossible
-    Yield f s -> step' s' >>= \case
-      Stop       -> pure Stop
-      Skip _     -> impossible
-      Yield a s' -> pure $ Yield (f a) (s, s')
-(<:>) (Pull @s True seed step) (Pull @s' False seed' step') =
-  Pull True ((,) <$> seed <*> seed') \(s, s') ->
-    step s >>= \case
-      Stop        → pure Stop
-      (Skip s)    → pure $ Skip (s , s')
-      (Yield f s) → step' s' >>= \case
-        Stop         → pure Stop
-        (Skip s')    → impossible
-        (Yield a s') → pure $ Yield (f a) (s , s')
-(<:>) (Pull @s False seed step) (Pull @s' True seed' step') =
-  Pull True ((,) <$> seed <*> seed') \case
-    (s, s') → step' s' >>= \case
-      Stop         → pure Stop
-      (Skip s')    → impossible
-      (Yield a s') → step s >>= \case
-        Stop        → pure Stop
[…]

### 2024-07-04 icfp24paper/submission/haskell-cftt/CFTT/Split.hs

@@ -1,45 +1,0 @@
-
-module CFTT.Split where
-
-import CFTT.Up
-import CFTT.Gen
-
--- paper 3.3.1
---------------------------------------------------------------------------------
-
-class Split a b | a -> b, b -> a where
-  splitGen :: Up a -> Gen b
-
-split :: MonadGen m => Split a b => Up a -> m b
-split a = liftGen (splitGen a)
-
-caseM :: (MonadGen m) => Split a b => Up a -> (b -> m c) -> m c
-caseM a f = split a >>= f
-
-instance Split Bool Bool where
-  splitGen x = Gen \k -> [|| case $$x of
-    True -> $$(k True)
-    _    -> $$(k False) ||]
-
-instance Split [a] (Maybe (Up a, Up [a])) where
-  splitGen x = Gen \k -> [|| case $$x of
-    []   -> $$(k Nothing)
-    a:as -> $$(k (Just ([||a||], [||as||]))) ||]
-
-instance Split (a, b) (Up a, Up b) where
-  splitGen x = Gen \k -> [|| case $$x of
-    (a, b) -> $$(k ([||a||], [||b||])) ||]
-
-instance Split (Pair a b) (Pair (Up a) (Up b)) where
-  splitGen x = Gen \k -> [|| case $$x of
-    Pair a b -> $$(k (Pair [||a||] [||b||])) ||]
-
-instance Split (Either a b) (Either (Up a) (Up b)) where
-  splitGen x = Gen \k -> [|| case $$x of
-    Left a  -> $$(k (Left [||a||]))
-    Right b -> $$(k (Right [||b||])) ||]
-
-instance Split (Maybe a) (Maybe (Up a)) where
-  splitGen x = Gen \k -> [|| case $$x of
-    Nothing -> $$(k Nothing)
-    Just a  -> $$(k (Just [||a||])) ||]

### 2024-07-04 icfp24paper/submission/haskell-cftt/CFTT/Up.hs

@@ -1,63 +1,0 @@
-
-{-# options_ghc -Wno-orphans #-}
-
-module CFTT.Up where
-
-import Prelude (Num(..))
-import qualified Prelude as P
-import Data.Bits
-
-import Language.Haskell.TH
-import Language.Haskell.TH.Syntax
-
-type Up = CodeQ
-
-data Pair a b = Pair !a !b
-proj1 (Pair a b) = a
-proj2 (Pair a b) = b
-
-instance Num a => Num (Up a) where
-  (+) x y       = [|| $$x + $$y ||]
-  (*) x y       = [|| $$x * $$y ||]
-  abs x         = [|| abs $$x ||]
-  signum x      = [|| signum $$x ||]
-  fromInteger x = [|| fromInteger $$(liftTyped x) ||]
-  negate x      = [|| negate $$x ||]
-
-class Eq a where
-  (==) :: Up a -> Up a -> Up P.Bool
-  (/=) :: Up a -> Up a -> Up P.Bool
-  (/=) x y = [|| P.not $$(x == y) ||]
-
-instance P.Eq a  => Eq a where
-  (==) x y = [|| $$x P.== $$y ||]
-
-not :: Up P.Bool -> Up P.Bool
-not x = [|| P.not $$x ||]
-
-tt :: Up ()
-tt = [|| () ||]
-
-down1 :: (Up a -> Up b) -> Up (a -> b)
-down1 f = [|| \a -> $$(f [||a||]) ||]
-
-up1 :: Up (a -> b) -> Up a -> Up b
-up1 f a = [|| $$f $$a ||]
-
-class Ord a where
-  (<)  :: Up a -> Up a -> Up P.Bool
-  (<=) :: Up a -> Up a -> Up P.Bool
-  (>)  :: Up a -> Up a -> Up P.Bool
-  (>=) :: Up a -> Up a -> Up P.Bool
-
-instance P.Ord a => Ord a where
-  (<)  x y = [|| (P.<)  $$x $$y ||]
-  (<=) x y = [|| (P.<=)  $$x $$y ||]
-  (>)  x y = [|| (P.>)  $$x $$y ||]
-  (>=) x y = [|| (P.>=)  $$x $$y ||]
-
-even :: Up P.Int -> Up P.Bool
-even n = [|| $$n .&. 1 P.== 0 ||]
-
-odd :: Up P.Int -> Up P.Bool
-odd n = [|| $$n .&. 1 P.== 1 ||]

### 2024-07-18 icfp24paper/paper.tex
camera ready
stuff

@@
 \title{Closure-Free Functional Programming in a Two-Level Type Theory}
-\titlenote{This work was supported
-by grant 2019.0116 of the Knut and Alice Wallenberg Foundation.}
 
@@
 
-\subsubsection{Comparison to defunctionalization}
+\subsubsection*{Comparison to defunctionalization}
 Defunctionalization is a program translation which represents higher-order
@@
 
-\subsubsection{Case splitting in monads}
+\subsubsection*{Case splitting in monads}
 We often want to case-split on object-level data inside a monadic action, and
@@
 
-\subsubsection{$\State$ and $\Reader$ operations} If we use
+\subsubsection*{$\State$ and $\Reader$ operations} If we use
 the $\modify$ function that we already have in $\State_\M$, a curious thing
@@
 is unstaged to
-\[ \qt{\lam \vx.\,\letdef x := \vx + \vx; \letdef \vx := \vx + \vx; ((),\,\vx)}. \]
+\[ \qt{\lam \vx.\,\letdef \vx := \vx + \vx; \letdef \vx := \vx + \vx; ((),\,\vx)}. \]
 
@@
   & \ind \case \qt{\vb}\,\of\,\True\to \mput' \qt{10}; \False \to \mput' \qt{20}\\
-  & \ind\modify'\,(\lambda\,x.\,\qt{\spl x + \spl x}))
+  & \ind\modify'\,(\lambda\,\vx.\,\qt{\spl \vx + \spl \vx}))
 \end{alignat*}
@@
   & \ind \mup\,\msf{act} \\
-  & \ind\modify\,(\lambda\,x.\,\qt{\spl x + \spl x}))
+  & \ind\modify\,(\lambda\,\vx.\,\qt{\spl \vx + \spl \vx}))
 \end{alignat*}
@@
   & \ind \join \dlr \case \qt{\vb}\,\of\,\True\to \mput' \qt{10}; \False \to \mput' \qt{20}\\
-  & \ind\modify'\,(\lambda\,x.\,\qt{\spl x + \spl x}))
+  & \ind\modify'\,(\lambda\,\vx.\,\qt{\spl \vx + \spl \vx}))
 \end{alignat*}
@@
 
-\subsection{Running Streams with Mutual Recursion}
+\subsection{Running Streams With Mutual Recursion}
 
@@
   & \hspace{-11em}\mathrlap{\mit{bind_{\Gen}}\,\vga\,\vf\,=}\\
-  & \hspace{-11em}\mathrlap{\ind \Pull\,(\Sigma\,\vA\,(\mit{projS} \circ \vf))\,(\mdo \{\va \fro \vga; \vs \fro \mit{projSeed}\,(\vf\,\va); \return (\va,\,\vs)) \dlr \lam \case}\\
+  & \hspace{-11em}\mathrlap{\ind \Pull\,(\Sigma\,\vA\,(\mit{projS} \circ \vf))\,(\mdo \{\va \fro \vga; \vs \fro \mit{projSeed}\,(\vf\,\va); \return (\va,\,\vs)\}) \dlr \lam \case}\\
   & \hspace{-11em}\mathrlap{\ind \ind (\va,\,\vs) \to \mit{projStep}\,(\vf\,\va)\,\vs \bind \lam \case }\\
@@
 I thank the anonymous reviewers, Sebastian Graf and Jacques Carette, for their
-comments and suggestions on draft versions of this paper.
+comments and suggestions on draft versions of this paper. This work was supported
+by grant 2019.0116 of the Knut and Alice Wallenberg Foundation.
 
@@
 
-\subsection{W-types}
+\subsection{W-Types}
 So far we have stayed close to \cite{staged2ltt}, with only cosmetic
@@
 
-\subsection{Identity type}
+\subsection{Identity Type}
 

### 2024-08-27 opt/CSE1.hs
stuff

@@ -1,0 +1,197 @@
+{-# language Strict, LambdaCase, BlockArguments, BangPatterns, DerivingVia #-}
+{-# options_ghc -Wincomplete-patterns -Wunused-imports #-}
+
+import Data.Map.Strict (Map)
+import qualified Data.Map.Strict as M
+import Control.Monad.State.Strict
+import qualified Data.IntSet as IS
+
+-- global CSE + closure conversion for CBV untyped ANF LC in O(n*log n). The log factor comes
+-- from local scope size (env lookups, free variables sets) so it's not very prominent in
+-- non-pathological input.
+
+-- Raw terms
+--------------------------------------------------------------------------------
+
+data RTm
+  = RLetLam String String RTm RTm
+  | RLetApp String String String RTm
+  | RRet String
+  deriving Show
+
+elabVar :: Lvl -> [(String, Lvl)] -> String -> Int
+elabVar l xs x = case lookup x xs of
+  Nothing -> error $ "name not in scope: " ++ x
+  Just x  -> l - x - 1
+
+elab :: Lvl -> [(String, Lvl)] -> RTm -> Tm
+elab l xs = \case
+  RLetLam f x t u ->
+    LetLam (elab (l + 1) ((x, l):xs) t) (elab (l + 1) ((f, l):xs) u)
+  RLetApp x y z t ->
+    LetApp (elabVar l xs y) (elabVar l xs z) (elab (l + 1) ((x, l):xs) t)
+  RRet x ->
+    Ret (elabVar l xs x)
+
+elab0 = elab 0 []
+
+--------------------------------------------------------------------------------
+
+type Ix     = Int
+type Lvl    = Int
+type LvlSet = IS.IntSet
+type OTmId  = Int
+
+data Tm
+  = LetLam Tm Tm     -- let f = \x. t; u
+  | LetApp Ix Ix Tm  -- let x = f y; t
+  | Ret Ix           -- return x
+  deriving (Show)
+
+data Val
+  = VApp Lvl Lvl
+  | VLam LvlSet OTmId
+  deriving (Eq, Show, Ord)
+
+data OTm
+  = OLet Val OTm
+  | ORet Lvl
+  deriving (Show, Eq, Ord)
+
+
+-- global store, memoizes closure bodies
+--------------------------------------------------------------------------------
+
+type Store = Map OTm OTmId
+
[…]

### 2024-09-01 icfp24prez/abstract.tex
icfp24 prez wip

--- 2024-07-04 icfp24paper/paper.tex
@@
-%% build: latexmk -pdf -pvc paper.tex
+\documentclass{easychair}
+\raggedbottom
 
-%% \documentclass[acmsmall,screen,review,anonymous]{acmart}
-%% \documentclass[nonacm,acmsmall]{acmart}
-\documentclass[acmsmall]{acmart}
-%% \raggedbottom
-
-%% \BibTeX command to typeset BibTeX logo in the docs
-\AtBeginDocument{%
-  \providecommand\BibTeX{{%
-    \normalfont B\kern-0.5em{\scshape i\kern-0.25em b}\kern-0.8em\TeX}}}
-
-\setcopyright{acmlicensed}
-\acmDOI{10.1145/3674648}
-\acmYear{2024}
-\copyrightyear{2024}
-\acmSubmissionID{icfp24main-p75-p}
-\acmJournal{PACMPL}
-\acmVolume{8}
-\acmNumber{ICFP}
-\acmArticle{259}
-\acmMonth{8}
-\received{2024-02-28}
-\received[accepted]{2024-06-18}
-
-\bibliographystyle{ACM-Reference-Format}
-%%
-%% The majority of ACM publications use numbered citations and
-%% references.  The command \citestyle{authoryear} switches to the
-%% "author year" style.
-%%
-%% If you are preparing content for an event
-%% sponsored by ACM SIGGRAPH, you must use the "author year" style of
-%% citations and references.
-%% Uncommenting
-%% the next command will enable that style.
-\citestyle{acmauthoryear}
-
-%% --------------------------------------------------------------------------------
-
+\usepackage{doc}
 \usepackage{xcolor}
@@
 \usepackage{bm}
-\usepackage{mathtools}
-\usepackage{stmaryrd}
-\usepackage[title]{appendix}
+\usepackage{amssymb}
 
-\newcommand{\mit}[1]{{\mathsf{#1}}}
+\renewcommand{\mit}[1]{{\mathit{#1}}}
 \newcommand{\msf}[1]{{\mathsf{#1}}}
@@
 \newcommand{\mbb}[1]{\mathbb{#1}}
-\newcommand{\rel}{^{\approx}}
-\newcommand{\U}{\mathsf{U}}
-\newcommand{\Id}{\msf{Id}}
-\newcommand{\p}{\mathsf{p}}
-\newcommand{\q}{\mathsf{q}}
-\newcommand{\code}{\mathsf{code}}
 \newcommand{\bs}[1]{\boldsymbol{#1}}
@@
 \newcommand{\map}{\mit{map}}
-\newcommand{\emptycon}{\scaleobj{.75}\bullet}
-\newcommand{\Tyo}{\msf{Ty}_{\mbbo}}
-\newcommand{\Tmo}{\msf{Tm}_{\mbbo}}
-\newcommand{\Cono}{\msf{Con}_{\mbbo}}
-\newcommand{\Subo}{\msf{Sub}_{\mbbo}}
-\newcommand{\whset}{\wh{\Set}}
-\newcommand{\ev}{\mbb{E}}
-\newcommand{\re}{\mbb{R}}
-\newcommand{\welim}{\vW{-}\msf{elim}}
-
-\newcommand{\mbbc}{\mbb{C}}
-\newcommand{\mbbo}{\mbb{O}}
 
@@
 \newcommand{\vma}{\mathsf{ma}}
-\newcommand{\vga}{\mathsf{ga}}
 \newcommand{\vm}{\mathsf{m}}
@@
 \newcommand{\vns}{\mathsf{ns}}
-\newcommand{\vW}{\mathsf{W}}
-\newcommand{\vsup}{\mathsf{sup}}
-\newcommand{\vid}{\mathsf{id}}
-\newcommand{\whW}{\wh{\vW}}
 
@@
 \newcommand{\appull}{\ap_{\Pull}}
-
-\newcommand{\Con}{\msf{Con}}
-\newcommand{\Sub}{\msf{Sub}}
-\newcommand{\Tm}{\msf{Tm}}
 
@@
 \newcommand{\MTy}{\msf{MetaTy}}
-\newcommand{\MTm}{\msf{MetaTm}}
 \newcommand{\VTy}{\msf{ValTy}}
@@
 
-%% --------------------------------------------------------------------------------
+\title{Polarized Lambda-Calculus at Runtime, Dependent Types at Compile Time}
+\author{Andr\'as Kov\'acs}
+\authorrunning{Kov\'acs}
+\titlerunning{Polarized Lambda-Calculus at Runtime, Dependent Types at Compile Time}
 
-%%
-%% end of the preamble, start of the body of the document source.
-%% \hypersetup{draft}
+\institute{
+  University of Gothenburg
+  Gothenburg, Sweden \\
+  \email{andrask@chalmers.se}
+}
+
+\pagenumbering{gobble}
 \begin{document}
-
-\title{Closure-Free Functional Programming in a Two-Level Type Theory}
-\titlenote{This work was supported
-by grant 2019.0116 of the Knut and Alice Wallenberg Foundation.}
-
-\author{András Kovács}
-\orcid{0000-0002-6375-9781}
-\affiliation{%
-  \institution{University of Gothenburg}
-  \city{Gothenburg}
-  \country{Sweden}
-}
-\email{andrask@chalmers.se}
-
-\begin{abstract}
-Many abstraction tools in functional programming rely heavily on general-purpose
-compiler optimization to achieve adequate performance. For example, monadic
-binding is a higher-order function which yields runtime closures in the absence
-of sufficient compile-time inlining and beta-reductions, thereby significantly
-degrading performance. In current systems such as the Glasgow Haskell Compiler,
-there is no strong guarantee that general-purpose optimization can eliminate
-abstraction overheads, and users only have indirect and fragile control over
-code generation through inlining directives and compiler options. We propose a
-two-stage language to simultaneously get strong guarantees about code generation
-and strong abstraction features. The object language is a simply-typed
-first-order language which can be compiled without runtime closures. The
-compile-time language is a dependent type theory. The two are integrated in a
-two-level type theory.
-
-We demonstrate two applications of the system. First, we develop monads and
-monad transformers. Here, abstraction overheads are eliminated by staging and we
-can reuse almost all definitions from the existing Haskell ecosystem. Second,
-we develop pull-based stream fusion. Here we make essential use of dependent
-types to give a concise definition of a $\mathsf{concatMap}$ operation with
-guaranteed fusion. We provide an Agda implementation and a typed Template
-Haskell implementation of these developments.
-
-\end{abstract}
-
-\begin{CCSXML}
-<ccs2012>
-   <concept>
-       <concept_id>10003752.10003790.10011740</concept_id>
-       <concept_desc>Theory of computation~Type theory</concept_desc>
-       <concept_significance>500</concept_significance>
-       </concept>
-   <concept>
-       <concept_id>10011007.10011006.10011041.10011047</concept_id>
-       <concept_desc>Software and its engineering~Source code generation</concept_desc>
-       <concept_significance>500</concept_significance>
-       </concept>
- </ccs2012>
-\end{CCSXML}
-
-\ccsdesc[500]{Theory of computation~Type theory}
-\ccsdesc[500]{Software and its engineering~Source code generation}
-
-\keywords{two-level type theory, staged compilation}
-
 \maketitle
 
-\section{Introduction}\label{sec:introduction}
+We describe a particular two-level type theory \cite{staged2ltt}:
+\begin{itemize}
+\item The object level is a polarized simply-typed calculus, with
+      computation types (functions, computational products) and value types
+      (inductive types, closures).
+\item The meta level is a standard dependent type theory.
+\end{itemize}
+This supports two-stage compilation, where we can use dependent types in the
+surface language, but only object-level constructions remain after staging. It
+appears to be an excellent setting for performance-focused staged programming.
 
-Modern functional programming supports many convenient abstractions. These often
-come with significant runtime overheads. Sometimes the overheads are acceptable,
-but in other cases compiler optimization is crucial. Monads in Haskell is an
-example for the latter. Even the $\Reader$ monad, which is one of the simplest
-in terms of implementation, yields large overheads when compiled without
-optimizations. Consider the following:
+The polarization lets us control closures and function arities in a fairly
+lightweight way. In particular, if we do not use the explicit closure type
+former, then all function calls can be compiled to calls and jumps to statically
+known code locations. Hence, we migth ask: how much can we reproduce
+from the abstraction tools of functional programming, without using any runtime
+closures? Perhaps surprisingly, closures are rarely used in an essential way.
+\begin{itemize}
+\item A $\mit{map}$ function for lists is meant to be inlined, and after
+      inlining no runtime closures remain.
+\item Monadic binding in Haskell is a higher-order function. Without compiler optimizations,
+      we get a big overhead from a deluge of runtime closures; but with optimizations
+      we expect that no closures remain.
+\end{itemize}
+In the current work, we start to build up libraries in the mentioned two-level
+theory, aiming to minimize the usage of runtime closures and eliminate
+abstraction overheads. We want to shift work from general-purpose optimizing
+compilers to deterministic \& extensible metaprogramming. For example, instead
+of expecting GHC to sufficiently inline our monadic code, we implement efficient
+monadic code generation ourselves.
+
+\subsubsection*{Staged Monad Transformers}
+
+Despite recent competition from a variety of algebraic effect systems, monads
+and monad transformers remain the most widely used strongly-typed effect system,
+so it make sense for us to develop their staged flavor.
+
+In our approach, we \emph{don't use monads at all in the object
+language}. Instead, we use meta-level monads, and only convert down to the
+object level when runtime control dependencies force us to do so. Generally, the
+staged version of a monad is obtained by extending it with \emph{code generation
+as an effect}. The code generation monad is the following:
 \begin{alignat*}{3}
-  &\vf :: \Int \to \Reader\,\Bool\,\Int \\
-  &\vf\,\vx = \mdo\{\vb \fro \msf{ask}; \bif \vb\, \bthen \return (\vx + 10)\, \belse \return (\vx + 20)\}
+  &\newtype \Gen : \MTy \to \MTy \\
+  &\newtype \Gen\,A = \Gen\,(\{R : \msf{ObjTy}\} \to (A \to \Up R) \to \Up R)
 \end{alignat*}
-With optimizations enabled, GHC compiles this roughly to the code below:
-\begin{alignat*}{3}
-  &\vf :: \Int \to \Bool \to \Int \\
-  &\vf = \lam \vx\,\vb.\, \bif \vb\, \bthen \vx + 10\, \belse \vx + 20
-\end{alignat*}
-Without optimizations we roughly get:
-\begin{alignat*}{3}
-  &\vf = \lam \vx.\,(\bind)\,\msf{monadReader}\,(\msf{ask}\,\msf{monadReaderReader})\,(\lam \vb.\,\bif \vb\\
-  &\ind \bthen\return\,\msf{monadReader}\,(\vx + 10)\\
-  &\ind \belse\hspace{0.25em}\return\,\msf{monadReader}\,(\vx + 20))
-\end{alignat*}
-Here, $\msf{monadReader}$ and $\msf{monadReaderReader}$ are runtime
-dictionaries, respectively for the $\msf{Monad}$ and $\msf{MonadReader}$
-instances, and, for example, $(\bind)\,\msf{monadReader}$ is a field projection
-from the dictionary. This results from the dictionary-passing elaboration of
-type classes \cite{DBLP:conf/popl/WadlerB89}. We get a runtime closure from the
-$\lam \vb.\,...$ function, and $(\bind)$, $\msf{ask}$ and $\mbf{return}$ also
-produce additional closures.
+Here, $\msf{ObjTy}$ is the universe of object-level types, $\MTy$ is a
+meta-level universe, and $\Up R$ is (intuitively) the type of metaprograms which
+generate object expressions of type $R$.
 
-The difference between optimized and unoptimized code is already large here, and
-it gets even larger when we consider monad transformers or code that is
-polymorphic over monads. In Haskell, such code is pervasive, even in fairly
-basic programs which do not use fancy abstractions. Consider the $\msf{mapM}$
-function from the Haskell Prelude:
-\begin{alignat*}{3}
-  & \msf{mapM} :: \Monad\,\vm \Rightarrow (\va \to \vm\,\vb) \to [\va] \to \vm\,[\vb]
-\end{alignat*}
-This is a third-order rank-2 polymorphic function in disguise, because
-its monad dictionary argument contains the polymorphic second-order method
-$(\bind)$.  Compiling $\msf{mapM}$ efficiently relies on inlining the instance
-dictionary, then inlining the methods contained there, and also inlining the
-functions that the higher-order binding is applied to.
+This is a continuation monad where \emph{eventually} we have to return an
+object-level value, so we can only ``run'' actions of type $\Gen\,(\Up A)$ to
+extract a result with type $\Up A$. However, while working in the monad, we are
+free to use both object-level and meta-level constructions, and introduce
+object-level let-binders. In general, for some monad transformer stack $M$, we
+obtain its staged version by putting a $\Gen$ at the bottom of the stack. For
+example, $\State\,(\Up\msf{Int})$ becomes $\StateT\,(\Up\msf{Int})\,\Gen$. In
+this monad in particular, we can modify an object expression in the state, and
+also generate object code via $\Gen$.
 
-GHC's optimization efforts are respectable, and it has gotten quite adept over
-its long history of development. However, there is no strong guarantee that
-certain optimizations will happen. Control over optimizations remains tricky,
-fragile and non-compositional. \texttt{INLINE} and \texttt{REWRITE} pragmas can
-be used to control code generation \cite{ghcdocs}, but without strong
-guarantees, and their advanced usage requires knowledge of GHC internals. For
-example, correctly specifying the \emph{ordering} of certain rule applications
-is often needed. We also have to care about formal function arities. Infamously,
-the function composition operator is defined as $(.)\,\vf\,\vg = \lam \vx \to
-\vf\,(\vg\,\vx)$ in the base libraries, instead of as $(.)\,\vf\,\vg\,\vx =
-\vf\,(\vg\,\vx)$, to get better inlining behavior \cite{ghcbase}. It is common
-practice in high-performance Haskell programming to visually review GHC's
-optimized code output.
+Using our transformer library, we program in meta-monads, and insert conversions
+to and from object code at the points of dynamic dependencies
+(e.g.\ object-level function calls). Such conversions are defined in a
+compositional manner, by recursion on transformer stacks. Overall, this style of
+programming requires modest extra noise compared to Haskell, but it guarantees
+high-quality code output by staging.
 
-\subsection{Closure-Free Staged Compilation}
+\subsubsection*{Staged Stream Fusion}
 
-In this paper we use staged compilation to address issues of robustness. The
-idea is to shift as much work as possible from general-purpose optimization to
-metaprograms.
+\emph{Stream fusion} \cite{DBLP:phd/ethos/Coutts11} is another application that we developed in this
+setting. The idea here is to give a convenient list-like interface for
+programming with meta-level state machines. Such machines can be turned into
+efficient object-level code as blocks of mutually recursive functions, without
+storing intermediate results in runtime lists. We demonstrate very concise
+solutions to two well-known problems.
 
-Metaprograms can be deterministic, transparent, and can be run efficiently,
-using fast interpreters or machine code compilation. In contrast,
-general-purpose optimizers are slower to run, less transparent and less
-robust. Also, metaprogramming allows library authors to exploit
-domain-specific optimizations, while it is not realistic for general-purpose
-optimizers to know about all domains.
+\textbf{First}, we need to generate mutually recursive blocks. One part of the
+challenge is to have guaranteed well-scoping and well-typing, although the
+number (and types) of definitions in a mutual block is only known at staging
+time; see e.g.\ \cite{DBLP:conf/pepm/YallopK19}. We also want to avoid any runtime
+overheads. We represent a mutual block as a single recursive definition with a
+computational product type. The polarization guarantees that this can be
+compiled without downstream overheads to an actual mutual block.
 
-On the other hand, metaprogramming requires some additional care and input from
-programmers. Historically, there have been problems with ergonomics as well. In
-weakly-typed staged systems, code generation might fail \emph{too late} in the
-pipeline, producing incomprehensible errors. Or, tooling that works for an
-object language (like debugging, profiling, IDEs) may not work for metaprogramming, or
-metaprogramming may introduce heavy noise and boilerplate, obscuring the logic
-of programs and imposing restrictions on code structure.
+\textbf{Second}, fusion for arbitrary combinations of zipping and
+$\mit{concatMap}$ has been a long-standing challenge. Currently, the
+$\mit{strymonas}$ library \cite{DBLP:conf/pepm/0001K24} supports such fusion
+but its solution is much more complex than ours, and it also relies on mutable
+references in the object language. In contrast, we compile to pure object
+code, which also enables us to parameterize streams over arbitrary monads
+and embed monadic effects in stream definitions.
 
-We use \textbf{two-level type theory} (2LTT) \cite{twolevel,staged2ltt} to
-sweeten the deal of staged compilation, aiming for a combination of strong
-guarantees, good ergonomics, high level of abstraction and easy-to-optimize code
-output. We develop a particular two-level type theory for this purpose, which we
-call \textbf{CFTT}, short for ``closure-free type theory''. This consists of:
-\begin{itemize}
-\item A simply-typed object theory with first-order functions, general recursion and
-      finitary algebraic data types. This language is easy to optimize and compile
-      in the downstream pipeline, but it lacks many convenience features.
-\item A dependent type theory for the compile-time language. This
-      allows us to recover many features by metaprogramming.
-\end{itemize}
-Since the object language is first-order, we guarantee that all programs in CFTT
-can be ultimately compiled without any dynamic closures, using only calls and
-jumps to statically known code. Why emphasize closures?  They are the
-foundation to almost all abstraction tools in functional programming:
-\begin{itemize}
-\item Higher-order functions in essentially all functional languages are implemented with closures.
-\item Type classes in Haskell use dictionary-passing, which relies on closures for function methods \cite{DBLP:conf/popl/WadlerB89}.
-\item Functors and first-class modules in OCaml \cite{leroy2023ocaml} and other ML-s rely on closures.
-\end{itemize}
-Hence, doing functional programming without closures is a clear demonstration
-that we can get rid of abstraction overheads.
+Our streams are tuples containing a type for an internal state, an initial
+state and a representation of state transitions. Moreover, the internal state is
+required to be a finite sum-of-products of object-level value types. Now, if
+such sums-of-products are closed under $\Sigma$-types, $\concatMap$ is easily
+definable by using a $\Sigma$-type of the family of internal states that we get
+from an $A \to \msf{Stream}\,B$ function. We construct such $\Sigma$-types from
+an internal \emph{generativity axiom}, which expresses that certain metaprograms
+can't depend on object-level terms. This axiom does hold in the staging
+semantics and we can erase it during staging. From this, we construct
+\emph{transient} $\Sigma$-types which end up as non-dependent product types
+after staging.
 
-Perhaps surprisingly, little practical programming relies essentially on
-closures. Most of the time, programmers use higher-order functions for
-\emph{abstraction}, such as when mapping over lists, where it is expected that
-the mapping function will be inlined. We note though that our setup is
-compatible with closures as well, and it can support two separate type formers
-for closure-based and non-closure-based ``static'' functions. Having both of
-these would be desirable in a practical system. In the current work we focus on
-the closure-free case because it is much less known and developed, and it is
-interesting to see how far we can go with it.
+The ability to analyze object code has been often viewed as a desirable feature
+in metaprogramming. In contrast, we demonstrate a use-case for the explicit
+\emph{lack} of intensional analysis, through our generativity axiom. This might
+be compared to \emph{internal parametricity} statements in type theories.
 
-%% For example, difference lists
-%% \cite{TODO} are often implemented with closures, but they can be also
-%% implemented as binary trees, with the same programming interface.
-
-%% \todo{Essential closures: CPS? Threaded interpreters?}
-%% Whole-program defunctionalization is also possible, and it is notably used by
-%% the MLton compiler \cite{TODO}. While this can be practical and effective, it
-%% can be also expensive and require whole-program processing. Also, conceptually
-%% speaking, it does not eliminate closures but instead makes them more transparent
-%% to optimizations. In this paper we do not use defunctionalization.
-
-\subsection{Contributions}
-
-\begin{itemize}
-\item In Section \ref{sec:overview-of-cftt} we present the two-level type theory
-  CFTT, where the object level is first-order simply-typed and the meta level is
-  dependently typed. The object language supports an operational semantics
-  without runtime closures, and can be compiled with only statically known
-  function calls. We provide a supplementary Agda formalization of the
-  operational semantics of the object language.
-\item
-  In Section \ref{sec:monad-transformers} we build a staged library for monad
-  transformers \cite{DBLP:conf/popl/LiangHJ95}. We believe that this is a good
-  demonstration, because monads and monad transformers are the most widely used
-  effect system in Haskell, and at the same time their compilation to efficient
-  code can be surprisingly difficult. The continuation monad is well-known in
-  staged compilation \cite{DBLP:conf/lfp/Bondorf92}, and staged state monads
-  have also been used
-  \cite{DBLP:conf/pepm/SwadiTKP06,DBLP:conf/emsoft/KiselyovST04,DBLP:journals/scp/CaretteK11}.
-  These works used specific monads as tools for domain-specific code
-  generation. In contrast, we propose ubiquitous staging for general-purpose
-  monadic programming, where users can write code that looks similar to monadic
-  code in Haskell but with deterministic and robust compilation to efficient
-  code.
-\item In Section \ref{sec:stream-fusion} we build a pull-based stream fusion library. Here,
-  we demonstrate essential usage of dependent types, in providing guaranteed
-  fusion for arbitrary combinations of $\concatMap$ and $\mit{zip}$. We use a
-  state machine representation that is based on \emph{sums-of-products} of
-  object-level values. We show that CFTT is compatible with a
-  \emph{generativity} axiom, which internalizes the fact that metaprograms
-  cannot inspect the structure of object-level terms. We use this to show that
-  the universe of sums-of-products is closed under $\Sigma$-types. This in turn
-  enables a very concise definition of $\concatMap$.
-\item
-  We adapt the contents of the paper to typed Template Haskell
-  \cite{DBLP:journals/pacmpl/XiePLWYW22}, with some modifications,
-  simplifications and fewer guarantees about generated code.  In particular,
-  Haskell does not have enough dependent types for the simple $\concatMap$
-  definition, but we can still work around this limitation. We also provide a
-  precise Agda embedding of CFTT and our libraries as described in this
-  paper. Here, the object theory is embedded as a collection of postulated
-  operations, and we can use Agda's normalization command to print out generated
-  object code.
-\end{itemize}
-
-\section{Overview of CFTT}\label{sec:overview-of-cftt}
-
-In the following we give an overview of CFTT features. We first review the
-meta-level language, then the object-level one, and finally the staging
-operations which bridge between the two.
-
-\subsection{The Meta Level}\label{sec:the-meta-level}
-
-$\bs{\MTy}$ is the universe of types in the compile-time language. We will often
-use the term ``metatype'' to refer to inhabitants of $\MTy$, and use
-``metaprogram'' for inhabitants of metatypes. $\MTy$ supports dependent
-functions, $\Sigma$-types and indexed inductive types \cite{inductivefamilies}.
-
-Formally, $\MTy$ is additionally indexed by universe levels (orthogonally to
-staging), and we have $\MTy_i : \MTy_{i+1}$. However, universe levels add noise
-and they are not too relevant to the current paper, so we will omit them.
-
-Throughout this paper we use a mix of Agda and Haskell syntax for
-CFTT. Dependent functions and implicit arguments follow Agda. A basic example:
-\begin{alignat*}{3}
-  &\id : \{\vA : \MTy\} \to \vA \to \vA\\
-  &\id = \lam \vx.\, \vx
-\end{alignat*}
-Here, the type argument is implicit, and it gets inferred when we use the
-function. For example, $\id\,\True$ is elaborated to $\id\,\{\Bool\}\,\True$,
-where the braces mark an explicit application for the implicit argument.
-Inductive types can be introduced using a Haskell-like ADT notation, or with a
-GADT-style one:
-\begin{alignat*}{3}
-  &                                           &&\hspace{4em}\data\,\Bool_\M &&: \MTy\,\where\\
-  & \data\,\Bool_\M : \MTy = \True_\M\,|\,\False_\M &&\hspace{4em}\ind\ind \True_\M &&: \Bool_\M\\
-  &                                           &&\hspace{4em}\ind\ind \False_\M &&: \Bool_\M
-\end{alignat*}
-Note that we added an $_\M$ subscript to the type; when analogous types can be
-defined both on the meta and object levels, we will sometimes use this subscript
-to disambiguate the meta-level version.
-
-We use Haskell-like newtype notation, such as in $\newtype \msf{Wrap}\,\vA =
-\msf{Wrap}\,\{\mit{unWrap} : \vA \}$, and also a similar notation for (dependent)
-record types, for instance as in
-\[\data \msf{Record} = \msf{Record}\,\{\mit{field1} : \vA,\,\mit{field2} : \vB\}.\]
-All construction and elimination rules for type formers in $\MTy$ stay within
-$\MTy$. For example, induction on meta-level values can only produce meta-level
-values.
-
-\subsection{The Object Level}\label{sec:the-object-level}
-
-$\bs{\Ty}$ is the universe of types in the object language. It is itself a
-metatype, so we have $\Ty : \MTy$. All construction and elimination rules of
-type formers in $\Ty$ stay within $\Ty$. We further split $\Ty$ to two
-sub-universes.
-
-First, $\bs{\VTy} : \MTy$ is the universe of \emph{value types}. $\VTy$ supports
-parameterized algebraic data types, where parameters can have arbitrary types,
-but all constructor field types must be in $\VTy$. Since $\VTy$ is a
-sub-universe of $\Ty$, we have that when $\vA : \VTy$ then also $\vA :
-\Ty$. Formally, this is specified as an explicit embedding operation, but we
-will use implicit subtyping for convenience.
-
-Second, $\bs{\CTy} : \MTy$ is the universe of \emph{computation types}. This is
-also a sub-universe of $\Ty$ with implicit coercions. For now, we only specify
-that $\CTy$ contains functions whose domains are value types:
-\[ \blank\to\blank : \VTy \to \Ty \to \CTy \]
-For instance, if $\Bool : \VTy$ is defined as an object-level ADT, then $\Bool
-\to \Bool : \CTy$, hence also $\Bool \to \Bool : \Ty$. However, $(\Bool \to
-\Bool) \to \Bool$ is ill-formed, since the domain is not a value type. Let us look at an
-example for an object-level program, where we already have natural numbers
-declared as $\data \Nat := \Zero\,|\,\Suc\,\Nat$:
-\begin{alignat*}{3}
-  &\hspace{-1em}\rlap{$\add : \Nat \to \Nat \to \Nat$}\\
-  &\hspace{-1em}\add :=\,&& \mathrlap{\letrec \go\,\vn\,\vm := \case \vn\,\of}\\
-  &\hspace{-1em}         && \ind \Zero     &&\to \vm;\\
-  &\hspace{-1em}         && \ind \Suc\,\vn &&\to \Suc\,(\go\,\vn\,\vm);\\
-  &\hspace{-1em}         && \go
-\end{alignat*}
-Recursive definitions are introduced with $\letrec$. The general syntax is
-$\letrec \vx : \vA := \vt; \vu$, where the $\vA$ type annotation can be
-omitted. $\mbf{letrec}$ can be only used to define computations, not values
-(hence, only functions can be recursive so far).
-
-Object-level definitions use $:=$ as notation, instead of the $=$ that is used
-for meta-level ones. We also have non-recursive $\mbf{let}$, which can be used
-to define computations and values alike, and can be used to shadow binders:
-\begin{alignat*}{3}
-  &\vf : \Nat \to \Nat\\
-  &\vf\,\vx := \letdef \vx := \vx + 10; \letdef \vx := \vx + 20; \vx * 10
-\end{alignat*}
-We also allow $\mbf{newtype}$ definitions, both in $\VTy$ and $\CTy$. These are
-assumed to be erased at runtime. In the Haskell and Agda implementations they
-are important for guiding instance resolution, and we think that the explicit
-wrapping makes many definitions more comprehensible in CFTT as well.
-
-Values are call-by-value at runtime; they are computed eagerly in function
-applications and $\mbf{let}$-s. $\mbf{let}$-definitions can be used to define
-inhabitants of any type, and the type of the $\mbf{let}$ body can be also
-arbitrary. Additionally, the right hand sides of $\mbf{case}$ branches can also
-have arbitrary types. So the following is well-formed:
-\begin{alignat*}{3}
-  &\vf : \mathrlap{\Bool \to \Nat \to \Nat}\\
-  &\vf\,\vb := \case \vb\,\of \True \to (\lam \vx.\, \vx + 10); \False &&\to (\lam \vx.\, \vx * 10)
-\end{alignat*}
-In contrast, computations are call-by-name, and the only way we can compute with
-functions is to apply them to value arguments. The call-by-name strategy is
-fairly benign here and does not lead to significant duplication of computation,
-because functions cannot escape their scope; they cannot be passed as arguments
-or stored in data constructors. This makes it possible to run object programs
-without using dynamic closures. This point is not completely straightforward;
-consider the previous $\vf$ function which has $\lambda$-expressions under a
-$\mbf{case}$.
-
-However, the call-by-name semantics lets us transform $\vf$ to $\lam
-\vb\,\vx.\,\case \vb\,\of \True \to \vx + 10; \False \to \vx * 10$, and more
-generally we can transform programs so that every function call becomes
-\emph{saturated}. This means that every function call is of the form
-$\vf\,\vt_1\,\vt_2\,...\,\vt_n$, where $\vf$ is a function variable and the
-definition of $\vf$ immediately $\lambda$-binds $n$ arguments. We do not detail
-this here. We provide formal syntax and operational semantics of the object
-language in the Agda supplement. We formalized the specific translation steps
-that are involved in call saturation, but only specified the full translation
-informally.
-
-Why not just make the object language less liberal, e.g.\ by disallowing
-$\lambda$ under $\mbf{case}$ or $\mbf{let}$, thereby making call saturation
-easier or more obvious? There is a trade-off between making the object language
-more restricted, and thus easier to compile, and making metaprogramming more
-convenient. We will see that the ability to insert $\mbf{let}$-s without
-restriction is very convenient in code generation, and likewise the ability to
-have arbitrary object expressions in $\mbf{case}$ bodies. In this paper we go
-with the most liberal object syntax, at the cost of needing more downstream
-processing.
-
-%% On the other side of the spectrum, one might imagine generating
-%% object code in low-level A-normal form. This is more laborious, but it could be
-%% also interesting, because it would force us to invent abstractions for
-%% manipulating ANF in the metalanguage. We leave such lower-level object languages
-%% for future investigation.
-
-\subsubsection{Object-level definitional equality} This is a distinct notion
-from runtime semantics. Object programs are embedded in CFTT, which is a
-dependently typed language, so sometimes we need to decide definitional equality of
-object programs during type checking. The setup is simple: we have no $\beta$ or
-$\eta$ rules for object programs at all, nor any rule for $\mbf{let}$-unfolding.
-The main reason is the following: we care about the size and efficiency of
-generated code, and these properties are not stable under $\beta\eta$-conversion
-and $\mbf{let}$-unfolding. Moreover, since the object language has general
-recursion, we do not have a sensible and decidable notion of program equivalence
-anyway.
-
-\subsubsection{Comparison to call-by-push-value}
-We took inspiration from call-by-push-value (CBPV) \cite{DBLP:conf/tlca/Levy99},
-and there are similarities to our object language, but there are also
-significant differences. Both systems have a value-computation distinction, with
-call-by-name computations and call-by-value values. However, our object theory
-supports variable binding at arbitrary types while CBPV only supports value
-variables. In CBPV, a let-definition for a function is only possible by first
-packing it up as a closure value (or ``thunk''), which clearly does not suit our
-applications.  Investigating the relation between CBPV and our object language
-could be future work.
-
-%% \subsubsection{Spectrum of possible object languages}
-%% There is a trade-off between making the object language more restricted, and
-%% thus easier to compile, and making metaprogramming more convenient. We will see
-%% that the ability to insert $\mbf{let}$-s without restriction is convenient, and
-%% likewise the ability to have $\lambda$-abstraction under $\mbf{case}$ bodies,
-%% although these features necessitate more downstream processing, to bring
-%% programs to saturated form. On the other side of the spectrum, one might imagine
-%% generating object code in low-level A-normal form. This is more laborious, but
-%% it could be also interesting, because it forces us to invent abstractions for
-%% manipulating ANF in the metalanguage. In a similar vein, Allais recently
-%% proposed metaprogramming quantum circuits in a two-level type theory
-%% \cite{TODO}.
-
-%% more restricted, and thus easier to compile, and making \emph{metaprogramming}
-%% more convenient. We will see that the ability to insert $\mbf{let}$-s without
-%% restriction is very convenient in code generation, and likewise the ability to
-%% have arbitrary object expressions in $\mbf{case}$ bodies. In this paper we go
-%% with the most liberal object syntax, at the cost of needing more downstream
-%% processing. The call-by-name nature of computation types requires a bit of a
-%% change of thinking from programmers, but we believe that it is well worth to
-%% have for the metaprogramming convenience.
-
-%% On the other side of the spectrum, one might imagine generating object code in
-%% low-level A-normal form. This is more laborious, but it could be also
-%% interesting, because it forces us to invent abstractions for manipulating ANF in
-%% the metalanguage. In a similar vein, Allais recently proposed metaprogramming
-%% quantum circuits in a two-level type theory \cite{TODO}. We leave such setups
-%% with low-level object languages to future investigation.
-
-%% Finally, one might compare our object language to call-by-push-value
-%% (CBPV). Indeed, we took inspiration from CBPV, and there are similarities, but
-%% also differences. In both systems there is a value-computation distinction, and
-%% values are call-by-value, and computations are call-by-name. However, our object
-%% language allows variable binding at arbitrary types, while CBPV only supports it
-%% at value types. In CBPV, a let-definition for a function is only possible by
-%% first packing it up as a closure value (or: ``thunk''), which clearly does not
-%% work for us.  Also, CBPV makes a judgment-level structural distinction between
-%% values and computations, while we use type universes for that. Generally
-%% speaking, type-based restrictions are easier to work with in dependent type
-%% theories than structural restrictions.
-
-%% The reasons are the following. First, since the object
-%% language has general recursion, most $\beta$-rules are only valid in the
-%% operational semantics up to some restrictions, and $\beta$-conversion is not
-%% decidable to begin with. Second, in metaprogramming we care about the size and
-%% efficiency of generated code, and these properties are not stable under $\beta$
-%% and $\eta$. Hence, the sensible choice is to do metaprogramming up to strict
-%% syntactic equality of object programs.
-
-%% Let us discuss the object language. First, notice that there is no polymorphism
-%% or any kind of type dependency. Although we can define lists as $\data \List\,A
-%% : \VTy := \Nil\,|\,\Cons\,A\,(\List\,A)$, the parameterization is just a
-%% shorthand; all concrete instantiations of the type are distinct. This
-%% monomorphism makes it easy to use different memory layouts for different types.
-%% For example, types which look like products may be unboxed. We could also make a
-%% distinction between boxed and unboxed sum types. We do not explore this in
-%% detail, we just note that monomorphic types make it easy to control memory
-%% layouts, and we believe that this is an important part of performance
-%% optimization.
-
-%% Second, there are no higher-order functions, and functions also cannot be stored
-%% in data structures. Hence, locally defined functions can never escape their
-%% scope, and all function calls are to functions defined in the current
-%% scope. This makes it possible to run object programs without using dynamic
-%% closures. This latter point might not be completely straightforward; what
-%% about the previous $f$ function which has $\lambda$-expressions under a
-%% $\mbf{case}$, should that require closures?
-
-%% We say that it should not. We make this formal formal in Section \ref{TODO};
-%% here we only give an intuitive explanation. In short, we choose a call-by-name
-%% runtime semantics for functions, which means that the only way we can compute
-%% with a function is by applying it to all arguments and extracting the resulting
-%% value.  Hence, the only way to compute with $f$ is to apply it to \emph{two}
-%% arguments, so $f$ is operationally equivalent to $\lam b\,x.\,\case b\,\of \True
-%% \to x + 10; \False \to x * 10$.  In Section \ref{TODO} we show that all object
-%% programs can be transformed to a \emph{saturated} form, where the arity of every
-%% function call matches the number of topmost $\lambda$-binders in the definition
-%% of the called function. Then, the local functions can be compiled either to
-%% top-level functions by lambda-lifting, or if they are only tail-called, left in
-%% place as join points \cite{TODO}.
-
-%% Why not just make the object language less liberal, e.g.\ by disallowing
-%% $\lambda$ under $\mbf{case}$ or $\mbf{let}$, thereby making call saturation
-%% easier or more obvious? There is a trade-off between making the object language
-%% more restricted, and thus easier to compile, and making \emph{metaprogramming}
-%% more convenient. We will see that the ability to insert $\mbf{let}$-s without
-%% restriction is very convenient in code generation, and likewise the ability to
-%% have arbitrary object expressions in $\mbf{case}$ bodies. In this paper we go
-%% with the most liberal object syntax, at the cost of needing more downstream
-%% processing. The call-by-name nature of computation types requires a bit of a
-%% change of thinking from programmers, but we believe that it is well worth to
-%% have for the metaprogramming convenience.
-
-%% On the other side of the spectrum, one might imagine generating object code in
-%% low-level A-normal form. This is more laborious, but it could be also
-%% interesting, because it forces us to invent abstractions for manipulating ANF in
-%% the metalanguage. In a similar vein, Allais recently proposed metaprogramming
-%% quantum circuits in a 2LTT \cite{TODO}. We leave such setups
-%% with low-level object languages to future investigation.
-
-%% Finally, one might compare our object language to call-by-push-value
-%% (CBPV). Indeed, we took inspiration from CBPV, and there are similarities, but
-%% also differences. In both systems there is a value-computation distinction, and
-%% values are call-by-value, and computations are call-by-name. However, our object
-%% language allows variable binding at arbitrary types, while CBPV only supports it
-%% at value types. In CBPV, a let-definition for a function is only possible by
-%% first packing it up as a closure value (or: ``thunk''), which clearly does not
-%% work for us.  Also, CBPV makes a judgment-level structural distinction between
-%% values and computations, while we use type universes for that. Generally
-%% speaking, type-based restrictions are easier to work with in dependent type
-%% theories than structural restrictions.
-
-\subsection{Staging}\label{sec:staging}
-
-With what we have seen so far, there is no interaction between the meta and
-object levels. We make such interaction possible with the following primitives.
-
-\begin{itemize}
-\item For $\vA : \Ty$, we have $\Lift \vA : \MTy$, pronounced as ``lift $\vA$''. This is
-      the type of metaprograms that produce $\vA$-typed object programs.
-\item For $\vA : \Ty$ and $\vt : \vA$, we have $\ql \vt \qr : \Lift \vA$, pronounced ``quote $\vt$''. This
-      is the metaprogram which immediately returns $\vt$.
-\item For $\vt : \Lift \vA$, we have $\spl \vt : A$, pronounced ``splice
-  $\vt$''. This inserts the result of a metaprogram into an object
-  term. \emph{Notation:} splicing binds stronger than function application, so
-  $\vf\,\spl \vx$ is parsed as $\vf\,(\spl \vx)$. We borrow this from
-  MetaML \cite{metaml}.
-\item We have $\ql \spl \vt \qr \equiv \vt$ and $\spl \ql \vt \qr \equiv \vt$ as definitional equalities.
-\end{itemize}
-
-We use \textbf{unstaging} to refer to the process of extracting object code from
-CFTT programs, by evaluating all metaprograms in splices. This term has been
-sporadically used in the literature, see
-e.g.\ \cite{DBLP:journals/cacm/RompfO12}, or \cite{DBLP:conf/popl/ChoiAYT11},
-with a somewhat different meaning in a multi-stage context. The main precursor
-to this paper used ``staging'' instead of our ``unstaging'' \cite{staged2ltt};
-we use the latter because the former conflicts with other common usages of
-``staging''.
-
-Let us look at some basic examples. Recall the meta-level identity function; it
-can be used at the object-level too, by applying it to quoted terms:
-  \[ \letdef \vn : \Nat := \spl(\id\,\ql 10 + 10 \qr); ... \]
-Here, $\id$ is used at type $\Lift \Nat$. During unstaging, the expression in
-the splice is evaluated, so we get $\spl\ql 10 + 10 \qr$, which is
-definitionally the same as $10 + 10$, which is our code output here. Boolean
-short-circuiting is another basic use-case:
-\begin{alignat*}{3}
-  &\mit{and} : \Up\Bool \to \Up\Bool \to \Up\Bool\\
-  &\mit{and}\,\vx\,\vy = \ql\case \spl \vx\,\of \True \to \spl \vy; \False \to \False\qr
-\end{alignat*}
-Since the $y$ expression is inlined under a $\mbf{case}$ branch at every use site, it
-is only computed at runtime when $x$ evaluates to $\True$. In many situations,
-staging can be used instead of laziness to implement short-circuiting, and often
-with better runtime performance, avoiding the overhead of thunking. Consider the
-$\map$ function now:
-\begin{alignat*}{3}
-  & \hspace{-5em}\mathrlap{\map : \{\vA\,\vB : \VTy\} \to (\Up \vA \to \Up \vB) \to \Up (\List\,\vA) \to \Up(\List\,\vB)}\\
-  & \hspace{-5em}\mathrlap{\map\,\vf\,\vas = \ql\letrec \go\,\vas := \case \vas\,\of}\\
-  & \hspace{-5em}\ind\ind\ind\ind\ind \ind\Nil             &&\to \Nil;\\
-  & \hspace{-5em}\ind\ind\ind\ind\ind \ind\Cons\,\va\,\vas &&\to \Cons\,\spl(\vf\,\ql \va \qr)\,(\go\,\vas);\\
-  & \hspace{-5em}\ind\ind\ind\ind\ind\go\,\spl\vas \qr
-\end{alignat*}
-For example, this can be used as $\letdef \vf\,\vas : \List\,\Nat \to
-\List\,\Nat := \spl(\map\,(\lam \vx.\,\ql \spl \vx + 10 \qr)\,\qt{\vas}) $.
-This is unstaged to a recursive definition where the mapping function is inlined
-into the $\Cons$ case as $\Cons\,\va\,\vas \to \Cons\,(\va + 10)\,(\go\,\vas)$.
-Note that $\map$ has to abstract over value types, since lists can only contain
-values, not functions. Also, the mapping function has type $\Up \vA \to \Up
-\vB$, instead of $\Up (\vA \to \vB)$. The former type is often preferable to the
-latter; the former is a metafunction with useful computational content, while
-the latter is merely a black box that computes object code. If we have $\vf :
-\Up(\vA \to \vB)$, and $\vf$ computes to $\ql \lam \vx.\,\vt\qr$, then $\spl
-\vf\,\vu$ is unstaged to an undesirable ``administrative'' $\beta$-redex $(\lam
-\vx.\,\vt)\,\vu$.
-
-\subsubsection{Comparison to defunctionalization}
-Defunctionalization is a program translation which represents higher-order
-functions using only first-order functions and inductive data
-\cite{DBLP:journals/lisp/Reynolds98a,DBLP:conf/ppdp/DanvyN01}. The idea is to
-represent each closure as a constructor of an inductive type, and implement
-closure application as a top-level first-order function which switches on all
-possible closure constructors. Hence, like our unstaging algorithm,
-defunctionalization produces first-order code from higher-order code. We summarize
-the differences.
-\begin{itemize}
-\item Defunctionalization mainly supports \emph{compiler optimizations}, by making
-  closures transparent to analysis. It is not an optimization by itself; it does
-  not change the amount of dynamic control flow or dynamic allocations in a
-  program. A dynamic closure call becomes a dynamic case-switch and a
-  heap-allocated ``functional'' closure becomes a heap-allocated inductive
-  constructor.
-\item Staging mainly supports optimization by \emph{programmers} by providing
-  control over code generation. Unstaging does not translate higher-order
-  functions to anything in particular. Instead, unstaging is the execution of
-  higher-order metaprograms which produce first-order programs.
-\end{itemize}
-
-\subsection{Staged Semantics of CFTT}\label{sec:semantics-of-staging}
-
-To obtain an unstaging algorithm, together with its correctness, the main task
-is to extend \cite{staged2ltt} with identity types and inductive families in the
-metalanguage, since we use those types in this paper. For inductive types, it is the
-most sensible to add W-types, since all inductive families can be faithfully
-derived from them in our setting \cite{whynotw}.
-
-This however requires a substantial amount of technical background from
-\cite{staged2ltt}, so we relegate it to Appendix \ref{appendix:A}. Here we only
-give an overview.
-\begin{itemize}
-\item The input of unstaging is a CFTT term with an object-level type, which
-      only depends on object-level free variables. The output is a term in the
-      object theory. In the other direction, there is an evident embedding of
-      object-theoretic terms as CFTT terms.
-\item \emph{Soundness} means that unstaging followed by embedding is the
-      identity map, up to conversion. In other words, unstaging respects conversion
-      of CFTT terms.
-\item \emph{Stability} means that embedding followed by unstaging is the
-      identity map. In other words, unstaging has no action on terms which are
-      already purely object-level (i.e.\ contain no splices).
-\end{itemize}
-In \cite{staged2ltt}, there is one more property, called \emph{strictness},
-which means that unstaging does not perform $\beta\eta$-conversion in the
-object theory, but this holds trivially in our case since we do not have such
-conversion rules.
-
-However, for CFTT we slightly depart from the above notion of soundness, for the
-following reason. We would like to assume certain propositional identities as
-\emph{axioms} in the metalanguage, without blocking unstaging as a
-computation. We use such an axiom to good effect in Section
-\ref{sec:concatmap-for-streams}.
-
-Axioms clearly block computation, so they are incompatible with the mentioned
-notion of soundness. But we are only interested in equations which hold
-definitionally during unstaging. Hence, we can do the following: first, we erase
-all identity proofs and their transports from CFTT programs, then we proceed
-with unstaging as in \cite{staged2ltt}. This erasure can be formalized as a
-syntactic translation \cite{next700} from CFTT to CFTT extended with the
-equality reflection principle, which says that propositional equality implies
-definitional equality (see e.g.\ \cite{hofmann95conservativity}). Thus,
-soundness of staging for CFTT holds up to erasure of identity proofs.
-
-\section{Monads \& Monad Transformers}\label{sec:monad-transformers}
-
-In this section we build a library for monads and monad transformers. We believe
-that this is a good demonstration of CFTT's abilities, since monads are
-ubiquitous in Haskell programming and they also introduce a great amount of
-abstraction that should be optimized away.
-
-\subsection{Binding-Time Improvements}\label{sec:binding-time-improvements}
-
-We do a preliminary overview before getting to monads. In the staged compilation
-and partial evaluation literature, the term \emph{binding time improvement} is
-used to refer to such conversions, where the ``improved'' version supports more
-compile-time computation \cite[Chapter 12]{partial-evaluation}.
-
-Translation from $\Up \vA \to \Up \vB$ to $\Up(\vA \to \vB)$ is a basic
-binding-time improvement, as an $\eta$-expansion \cite{eta-expansion-trick}. The
-two types are equivalent during unstaging, up to the runtime equivalence of
-object programs, and we can convert back and forth in CFTT, the same way as in
-MetaML \cite[Section 9]{metaml}:
-\begin{alignat*}{3}
-  &\mup : \Up (\vA \to \vB) \to \Up \vA \to \Up \vB && \ind\ind \mdown : (\Up \vA \to \Up \vB) \to \Up (\vA \to \vB) \\
-  &\mup\,\vf\,\va = \ql \spl \vf\, \spl \va\qr   && \ind\ind \mdown\,\vf = \ql \lam \va.\,\spl(\vf\,\ql \va \qr) \qr
-\end{alignat*}
-We cannot show internally, using propositional equality, that these functions are
-inverses, since we do not have $\beta\eta$-rules for object functions; but we
-will not need this proof in the rest of the paper.
-
- A general strategy for generating efficient ``fused''
-programs, is to try to work as much as possible with improved representations,
-and only convert back to object code at points where runtime dependencies are
-unavoidable. Let us look at binding-time-improvement for product types now:
-\begin{alignat*}{3}
-  &\mup : \Up (\vA,\,\vB) \to (\Up \vA,\,\Up \vB) && \ind\ind \mdown : (\Up \vA,\,\Up \vB) \to \Up(\vA,\,\vB) \\
-  &\mup\,\vx = (\qt{\fst\,\spl \vx},\, \qt{\snd\,\spl \vx})   && \ind\ind \mdown\,(\vx,\,\vy) = \qt{(\spl \vx,\,\spl \vy)}
-\end{alignat*}
-
-Here we overload Haskell-style product type notation, both for types and the
-pair constructor, at both levels. There is a problem with this conversion
-though: $\mup$ uses $\vx : \Up(\vA,\,\vB)$ twice, which can increase code size
-and duplicate runtime computations. For example, $\mdown\,({\mup\,\ql \vf\,\vx
-  \qr})$ is staged to $\ql (\fst\,(\vf\,\vx),\,\snd\,(\vf\,\vx)) \qr$. It would
-be safer to first let-bind an expression with type $\Up(\vA,\,\vB)$, and then
-only use projections of the newly bound variable. This is called
-\emph{let-insertion} in staged compilation. But it is impossible to use
-let-insertion in $\mup$ because the return type is in $\MTy$, and we cannot
-introduce object binders in meta-level code.
-
-\subsection{The Code Generation Monad}\label{sec:the-code-generation-monad}
-
-A principled solution to the previous issue is to write code generators in
-continuation-passing style, as first proposed by Bondorf
-\cite{DBLP:conf/lfp/Bondorf92}. This can be structured as a continuation monad,
-and later works used explicit monadic notation for it
-\cite{DBLP:conf/emsoft/KiselyovST04,DBLP:journals/scp/CaretteK11,DBLP:conf/pepm/SwadiTKP06}. Our
-definition is as follows:
-\begin{alignat*}{3}
-  & \newtype \Gen\,(\vA : \MTy) = \Gen\,\{\unGen : \{\vR : \Ty\} \to (\vA \to \Up \vR) \to \Up \vR\}
-\end{alignat*}
-This is a monad in $\MTy$ in a standard sense:
-\begin{alignat*}{3}
-  &\mathrlap{\instance \Monad\,\Gen\,\where}\\
-  &\ind \return \va        &&= \Gen \dlr \lam \vk.\,\vk\,\va\\
-  &\ind \mit{ga} \bind \vf &&= \Gen \dlr \lam \vk.\,\unGen\,\mit{ga}\,(\lam \va.\,\unGen\,(\vf\,\va)\,\vk))
-\end{alignat*}
-In \cite{DBLP:conf/emsoft/KiselyovST04},\cite{DBLP:journals/scp/CaretteK11} and
-\cite{DBLP:conf/pepm/SwadiTKP06}, the answer type is a parameter to the
-continuation monad, while we have it as polymorphic. We have found that
-polymorphic answer types are more convenient, because we do not have to
-anticipate the type of the code output when are defining a code generator. We
-can ``run'' actions as follows:
-\begin{alignat*}{3}
-  &\runGen : \Gen\,(\Up \vA) \to \Up \vA\\
-  &\runGen\,\vma = \unGen\,\vma\,\id
-\end{alignat*}
-From now on, we reuse Haskell-style type classes and $\mbf{do}$-notation in
-CFTT. We will use type classes in an informal way, without precisely specifying
-how they work. However, type classes are used in essentially the same way in the
-Agda and Haskell implementations, with modest technical differences.  From the
-$\Monad$ instance, the $\Functor$ and $\Applicative$ instances can be also
-derived. We reuse $({\fmap})$ and $({\ap})$ for applicative notation as well.
-
-We can let-bind object expressions in $\Gen$:
-\begin{alignat*}{3}
-  & \gen : \{\vA : \Ty\} \to \Up \vA \to \Gen\,(\Up \vA) \\
-  & \gen\,\va = \Gen \dlr \lam \vk.\,\ql \letdef \vx : \vA := \spl \va; \spl(\vk\,\ql \vx \qr) \qr
-\end{alignat*}
-And also recursive definitions of computations:
-\begin{alignat*}{3}
-  & \genRec : \{\vA : \CTy\} \to (\Up \vA \to \Up \vA) \to \Gen\,(\Up \vA) \\
-  & \genRec\,\vf = \Gen \dlr \lam \vk.\,\qt{\letrec \vx : \vA := \spl(\vf\,\qt{\vx}); \spl(\vk\,\qt{\vx})}
-\end{alignat*}
-Now, using $\mbf{do}$-notation, we may write $\mdo \{\vx \fro \gen\,\qt{10 + 10}; \vy \fro
-\gen\,\qt{20 + 20}\};\return \qt{\vx + \vy}\}$, for a $\Gen\,(\Up \Nat)$
-action. Running this with $\runGen$ yields $\qt{\letdef \vx := 10 + 10; \letdef \vy
-  := 20 + 20; \vx + \vy}$. We can also define a ``safer'' binding-time improvement for
-products, using let-insertion:
-\begin{alignat*}{6}
-  &\hspace{-2em}\mathrlap{\mup : \Up (\vA,\,\vB) \to \Gen\,(\Up \vA,\,\Up \vB)             \hspace{5em} \mdown : \Gen\,(\Up \vA,\, \Up \vB) \to \Up(\vA,\,\vB)} \\
-  &\hspace{-2em}\mup\,\vx = \mdo &&\vx \fro \gen\,\vx                                    && \ind\ind \mdown\,\vx = \runGen \dlr \mdo &&(\va,\,\vb) \fro \vx\\
-  &\hspace{-2em}               &&\return (\qt{\fst\,\spl \vx},\,\qt{\snd\,\spl \vx})     &&                           &&\return \qt{(\spl \va,\, \spl \vb)}
-\end{alignat*}
-
-Working in $\Gen$ is convenient, since we can freely generate object code and
-also have access to the full metalanguage. Also, the point of staging is that
-\emph{eventually} all metaprograms will be used for the purpose of code
-generation, so we eventually $\runGen$ all of our actions. So why not just
-always work in $\Gen$? The implicit nature of $\Gen$ may make it harder to
-reason about the size and content of generated code. This is a bit similar to
-the $\msf{IO}$ monad in Haskell \cite{jones2001tackling}, where eventually
-everything needs to run in $\msf{IO}$, but we may not want to write all of our
-code in $\msf{IO}$.
-
-\subsection{Monads}\label{sec:monads}
-
-First, following the style of Haskell's monad transformer library \texttt{mtl}
-\cite{mtl}, we define a class for monads that support code generation.
-\begin{alignat*}{3}
-&\tyclass\,\Monad\,\vM \RA \MonadGen\,\vM\,\where\\
-&\ind \liftGen : \Gen\,\vA \to \M\,\vA
-\end{alignat*}
-We also redefine $\gen$ and $\genRec$ to work in any $\MonadGen$, so from now on
-we have:
-\begin{alignat*}{3}
- &\gen   &&: \MonadGen\,\vM \RA \Up \vA \to \vM\,(\Up \vA)\\
- &\genRec &&: \MonadGen\,\vM \RA (\Up \vA \to \Up \vA) \to \vM\,(\Up \vA)
-\end{alignat*}
-In the rest of this paper, we only present $\ReaderT$, $\MaybeT$ and
-$\StateT$ as monad transformers. For all of these, we can define the $\MonadGen$
-instance simply by $\liftGen = \lift \circ \liftGen$.
-
-Let us look at the $\Maybe$ monad now. We have $\data \Maybe\,\vA :=
-\Nothing\,|\,\Just\,\vA$, and $\Maybe$ itself is available as a $\VTy \to \VTy$
-metafunction. However, we cannot directly fashion a monad out of $\Maybe$, since
-we do not have polymorphism in $\VTy$, nor can we store functions inside
-$\Maybe$. We could try to use the following type for binding:
-\[ (\bind) : \Up(\Maybe\,\vA) \to (\Up\vA \to \Up(\Maybe\,\vB)) \to \Up(\Maybe\,\vB) \]
-This would work, but the definition would necessarily use runtime case splits on $\Maybe$
-values, many of which could be optimized away during staging. Also, not having
-a ``real'' monad is inconvenient for the purpose of code reuse.
-
-Instead, our strategy is to only use proper monads in $\MTy$, and convert
-between object types and meta-monads when necessary, as a form of binding-time
-improvement. We define a class for this conversion:
-\begin{alignat*}{3}
-  &\hspace{-9em}\mathrlap{\tyclass\,\MonadGen\,\vM \RA \Improve\,(\vF : \VTy \to \Ty)\,(\vM : \MTy \to \MTy)\,\where} \\
-  &\hspace{-9em}\ind \mup   &&: \{\vA : \VTy\} \to \Up(\vF\,\vA) \to \vM\,(\Up \vA)\\
-  &\hspace{-9em}\ind \mdown &&: \{\vA : \VTy\} \to \vM\,(\Up \vA) \to \Up(\vF\,\vA)
-\end{alignat*}
-Assume that $\Maybe_\M$ is the standard meta-level monad, and $\MaybeT_\M$ is
-the standard monad transformer, defined as follows:
-\[ \newtype \MaybeT_\M\,\vM\,\vA = \MaybeT_\M\,\{\runMaybeT_\M : \vM\,(\Maybe_\M\,\vA)\} \]
-Now, the binding-time improvement of $\Maybe$ is as follows:
-\begin{alignat*}{3}
-  &\instance \Improve\,\Maybe\,(\MaybeT_\M\,\Gen)\,\where\\
-  %% &\ind \mup : \Up\,(\Maybe\,\vA) \to \MaybeT_\M\,\Gen\,(\Up\,\vA) \\
-  &\ind \mup\,\vma = \MaybeT_\M \dlr \Gen \dlr \lam \vk.\\
-  &\ind\ind \qt{\case \spl \vma\,\of \Nothing \to \spl(\vk\,\Nothing_\M);\Just\,\va \to \spl(\vk\,(\Just_\M\qt{\va}))}\\
-  %% &\ind \mdown : \MaybeT_\M\,\Gen\,(\Up\,A) \to \Up\,(\Maybe\,A) \\
-  &\ind \mdown\,(\MaybeT_\M\,(\Gen\,\vma)) = \\
-  &\ind\ind \vma\,(\lam \vx.\,\case \vx\,\of \Nothing_\M \to \qt{\Nothing}; \Just_\M\,\va \to \qt{\Just\,\spl \va})
-\end{alignat*}
-With this, we get the $\Monad$ instance for free from $\MaybeT_\M$ and $\Gen$. A small example:
-\[ \letdef \vn : \Maybe\,\Nat := \spl(\mdown\,\$\,\mdo \{\vx \fro \return\,\qt{10}; \vy \fro \return\,\qt{20}; \return\,\qt{\vx + \vy})\});\,... \]
-Since $\MaybeT_\M$ is meta-level, its monadic binding fully computes at unstaging time. Thus, the above code is computed to
-\[ \letdef \vn : \Maybe\,\Nat := \Just\,(10 + 20);\;... \]
-We can also use $\gen$ for let-insertion in $\MaybeT\,\Gen$, since it has a $\MonadGen$
-instance.
-%% Assume also a $\lift : \Monad\,\vM \RA \vM\,\vA \to \MaybeT_\M\,\vM\,\vA$ operation which
-%% comes from $\MaybeT_\M$ being a monad transformer. We can do let-insertion in
-%% $\MaybeT_\M\,\Gen$ by simply lifting:
-%% \begin{alignat*}{3}
-%%   &\gen' : \Up \vA \to \MaybeT_\M\,\Gen\,(\Up \vA) \\
-%%   &\gen'\,\va = \lift\,(\gen\,\va)
-%% \end{alignat*}
-
-\subsubsection{Case splitting in monads}
-We often want to case-split on object-level data inside a monadic action, and
-perform different actions in different branches. At first, this may seem
-problematic, because we cannot directly compute metaprograms from object-level
-case splits. Fortunately, with a little bit more work, this is possible in any
-$\MonadGen$, for any value type, using a variation of the so-called ``trick''
-\cite{eta-expansion-trick}.
-
-We demonstrate this for lists. An object-level $\mbf{case}$ on lists introduces
-two points where code generation can continue. We define a metatype which gives
-us a ``view'' on these points:
-\[ \data \msf{SplitList}\,\vA = \Nil'\,|\,\Cons'\,(\Up \vA)\,(\Up (\List\,\vA)) \]
-We can generate code for a case split, returning a view on it:
-\begin{alignat*}{3}
-  &\mit{split} : \Up (\List\,\vA) \to \Gen\,(\msf{SplitList}\,\vA)\\
-  &\mit{split}\,\vas = \Gen \dlr \lam \vk.\,\qt{\case \spl \vas\,\of \Nil \to \spl(\vk\,\Nil'); \Cons\,\va\,\vas \to \spl(\vk\,(\Cons'\,\qt{\va}\,\qt{\vas}))}
-\end{alignat*}
-Now, in any $\MonadGen$, assuming $\vas : \Up(\List\,\vA)$, we may write
-\begin{alignat*}{3}
-  &\mdo \{\mit{sp} \fro \liftGen\;(\!\mit{split}\,\vas);(\case \mit{sp}\,\of \Nil' \to ...;\Cons'\,\va\,\vas \to ...)\}
-\end{alignat*}
-This can be generalized to splitting on any object value. In the Agda and
-Haskell implementations, we overload it with a class similar to the following:
-\begin{alignat*}{3}
-  & \hspace{-2em}\mathrlap{\tyclass\,\mit{Split}\,(\vA : \VTy)\,\where}\\
-  & \hspace{-2em}\ind \mit{SplitTo} &&: \MTy \\
-  & \hspace{-2em}\ind \mit{split}   &&: \Lift \vA \to \Gen\,\mit{SplitTo}
-\end{alignat*}
-In a native implementation of CFTT it may make sense to extend do-notation, so
-that we elaborate $\mbf{case}$ on object values to an application of the
-appropriate splitting function. We adopt this in the rest of the paper, so
-when working in a $\MonadGen$, we can write $\case \vas\,\of \Nil \to ...;
-\Cons\,\va\,\vas \to ...$, binding $\va : \Up \vA$ and $\vas : \Up(\List\,\vA)$
-in the $\Cons$ case.
-
-\subsection{Monad Transformers}\label{monad-transformers}
-
-At this point, it makes sense to aim for a monad transformer library where
-binding-time improvement is defined compositionally, by recursion on the
-transformer stack. The base case is the following:
-\begin{alignat*}{3}
-  & \hspace{-4em}\mathrlap{\newtype \Identity\,\vA := \Identity\,\{\runIdentity : \vA\} }\\
-  & \hspace{-4em}\mathrlap{\instance \Improve\,\Identity\,\Gen\,\where}\\
-  & \hspace{-4em}\ind \mup\,\vx   &&= \return \qt{\runIdentity\,\spl \vx}\\
-  & \hspace{-4em}\ind \mdown\,\vx &&= \unGen\,\vx \dlr \lam \va.\,\qt{\Identity\,\spl \va}
-\end{alignat*}
-We recover the object-level $\Maybe$ as $\MaybeT\,\Identity$, from the following $\MaybeT$:
-\[ \newtype \MaybeT\,(\vM : \VTy \to \Ty)\,(\vA : \VTy) := \MaybeT\,\{\runMaybeT : \vM\,(\Maybe\,\vA)\} \]
-With this, improvement can be generally defined for $\MaybeT$:
-\begin{alignat*}{3}
-  &\instance \Improve\,\vF\,\vM \RA \Improve\,(\MaybeT\,\vF)\,(\MaybeT_\M\,\vM)\,\where\\
-  &\ind \mup\,\vx = \MaybeT_\M \dlr \mdo\\
-  &\ind\ind \vma \fro \mup\,\qt{\runMaybeT\,\spl \vx}\\
-  &\ind\ind \case \vma\,\of \Nothing \to \return \Nothing_\M\\
-  &\hspace{6.5em}         \Just\,\va \hspace{1.15em}\to \return (\Just_\M\,\va)\\
-  &\ind \mdown\,(\MaybeT_\M\,\vx) = \qt{\MaybeT\,\spl(\mdown \dlr \vx \bind \lam\case\\
-  &\ind\ind\Nothing_\M \to \return \qt{\Nothing}\\
-  &\ind\ind\Just_\M\,\va\hspace{1.15em} \to \return \qt{\Just\,\spl \va})}
-\end{alignat*}
-In the $\mbf{case}$ in $\mup$, we use our syntax sugar for matching on a
-$\Maybe$ value inside an $\vM$ action. This is legal, since we know from the
-$\Improve\,\vF\,\vM$ assumption that $\vM$ is a $\MonadGen$. In $\mdown$ we also use
-$\lam\case ...$ to shorten $\lam \vx.\,\case \vx\,\of ...$.
-
-On the meta level, we can borrow essentially all definitions from
-\texttt{mtl}. Only the continuation monad transformer fails to support
-binding-time-improvement in CFTT, because of the obvious need for dynamic
-closures. In the following we only present $\StateT$ and $\ReaderT$.
-
-We start with $\StateT$. We assume $\StateT_\M$ as the standard meta-level
-definition. The object-level $\StateT$ has type $(\vS : \VTy)(\vF : \VTy \to
-\Ty)(\vA : \VTy) \to \VTy$; the state parameter $\vS$ has to be a value type,
-since it is an input to an object-level function.
-\begin{alignat*}{3}
-  &\instance \Improve\,\vF\,\vM \RA \Improve\,(\StateT\,\vS\,\vF)\,(\StateT_\M\,(\Up \vS)\,\vM)\,\where\\
-  &\ind \mup\,\vx = \StateT_\M \dlr \lam \vs.\,\mdo\\
-  &\ind\ind \mit{as} \fro \mup\,\qt{\runStateT\,\spl \vx\,\spl \vs}\\
-  &\ind\ind \case \mit{as}\,\,\of (\va,\,\vs) \to \return (\va,\,\vs)\\
-  &\ind \mdown\,\vx = \qt{ \StateT\,(\lam \vs.\, \spl(\mdown \dlr \mdo\\\
-  &\ind \ind (\va,\,\vs) \fro \runStateT_\M\,\vx\,\qt{\vs}\\
-  &\ind \ind \return \qt{(\spl \va,\, \spl \vs)}))}
-\end{alignat*}
-Like before in $\MaybeT$, we rely on object-level case splitting in the
-definition of $\mup$. For $\Reader$, the environment parameter also has to be a
-value type, and we define improvement as follows.
-\begin{alignat*}{3}
-  &\hspace{-5em}\mathrlap{\instance \Improve\,\vF\,\vM \RA \Improve\,(\ReaderT\,\vR\,\vF)\,(\ReaderT_\M\,(\Up \vR)\,\vM)\,\where}\\
-  &\hspace{-5em}\ind \mup\,\vx   &&= \ReaderT_\M \dlr \lam \vr.\, \mup\,\qt{\runReaderT\,\spl \vx\,\spl \vr}\\
-  &\hspace{-5em}\ind \mdown\,\vx &&= \qt{\ReaderT\,(\lam \vr.\,\spl(\mdown\,(\runReaderT_\M\,\vx\,\qt{\vr})))}
-\end{alignat*}
-
-\subsubsection{$\State$ and $\Reader$ operations} If we use
-the $\modify$ function that we already have in $\State_\M$, a curious thing
-happens. The meaning of $\modify\,(\lam \vx.\,\qt{\spl \vx + \spl \vx})$ is to
-replace the current state $\vx$, as an object expression, with the expression
-$\qt{\spl \vx + \spl \vx}$, and this happens at unstaging time. This behaves as an
-``inline'' modification which replaces every subsequent mention of the state
-with a different expression. For instance, ignoring newtype wrappers for now,
-\[ \mdown \dlr \mdo \{\modify\,(\lam \vx.\,\qt{\spl \vx + \spl \vx}); \modify\,(\lam \vx.\,\qt{\spl \vx + \spl \vx});\return\,\qt{()}\} \]
-is unstaged to
-\[ \qt{\lam \vx.\,((),\,(\vx + \vx) + (\vx + \vx))} \]
-which duplicates the evaluation of $\vx + \vx$. The duplication can be avoided
-by let-binding the result in the object language. A similar phenomenon happens
-with the $\mit{local}$ function in $\Reader$. So we define ``stricter'' versions
-of these operations. We also return $\Up ()$ from actions instead of $()$ ---
-the former is more convenient, because the $\mdown$ operation can be immediately
-used on it.
-\begingroup
-\allowdisplaybreaks
-\begin{alignat*}{3}
-  & \mput' : (\MonadState\,(\Up \vS)\,\vM,\,\MonadGen\,\vM) \RA \Up \vS \to \vM\,(\Up ()) \\
-  & \mput'\,\vs = \mdo \{\vs \fro \gen\,\vs; \mput\,\vs; \return \qt{()}\}\\
-  &\\
-  & \modify' : (\MonadState\,(\Up \vS)\,\vM,\,\MonadGen\,\vM) \RA (\Up \vS \to \Up \vS) \to \vM\,(\Up ()) \\
-  & \modify'\,\vf = \mdo \{\vs \fro \get; \mput'\,(\vf\,\vs)\}\\
-  &\\
-  & \mit{local'} : (\MonadReader\,(\Up \vR)\,\vM,\,\MonadGen\,\vM) \RA (\Up \vR \to \Up \vR) \to \vM\,\vA \to \vM\,\vA\\
-  & \mit{local'}\,\vf\,\vma = \mdo \{\vr \fro \mit{ask}; \vr \fro \gen\,(\vf\,\vr);\mit{local}\,(\lam \_.\,\vr)\,\vma\}
-\end{alignat*}
-\endgroup
-Now,
-\[ \mdown \dlr \mdo \{\modify'\,(\lam \vx.\,\qt{\spl \vx + \spl \vx}); \modify'\,(\lam \vx.\,\qt{\spl \vx + \spl \vx})\} \]
-is unstaged to
-\[ \qt{\lam \vx.\,\letdef x := \vx + \vx; \letdef \vx := \vx + \vx; ((),\,\vx)}. \]
-
-\subsection{Joining Control Flow in Monads}\label{sec:joining-control-flow-in-monads}
-
-There is a deficiency in our library so far. Consider:
-\begin{alignat*}{3}
-  & \vf : \Bool \to \StateT\,\Nat\,(\MaybeT\,\Identity)\,()\\
-  & \vf\,\vb := \spl(\mdown \dlr \mdo \\
-  & \ind \case \qt{\vb}\,\of\,\True\to \mput' \qt{10}; \False \to \mput' \qt{20}\\
-  & \ind\modify'\,(\lambda\,x.\,\qt{\spl x + \spl x}))
-\end{alignat*}
-This is unstaged to the following (omitting newtype wrappers):
-\begin{alignat*}{3}
-  & \vf\,\vb\,\vs := \case\,\vb\,\of \\
-  & \ind \True\hspace{0.2em} \to \letdef \vs := 10; \letdef \vs := \vs + \vs; \Just\,((),\vs)\\
-  & \ind \False \to \letdef \vs := 20; \letdef \vs := \vs + \vs; \Just\,((),\vs)
-\end{alignat*}
-Notice that the final state modification gets inlined into both branches. This
-duplication follows from the definition of monadic binding in $\Gen$ and the
-$\mit{split}$ function in the desugaring of $\mbf{case}$. Code generation is
-continued in both branches with the same action. If we have multiple Boolean
-$\mbf{case}$ splits sequenced after each other, that yields exponential code
-size. A possible fix is to let-bind the branching action:
-\begin{alignat*}{3}
-  & \vf\,\vb := \spl(\mdown \dlr \mdo \\
-  & \ind \msf{act} \fro \gen \dlr \mdown \dlr \case \qt{\vb}\,\of\,\True\to \mput' \qt{10}; \False \to \mput' \qt{20} \\
-  & \ind \mup\,\msf{act} \\
-  & \ind\modify\,(\lambda\,x.\,\qt{\spl x + \spl x}))
-\end{alignat*}
-However, this yields suboptimal code:
-\begin{alignat*}{3}
-  & \vf\,\vb\,\vs :=\\
-  & \ind \letdef\,\msf{act}\,\vs := \case\,\vb\,\of \True\hspace{0.2em}\to \letdef \vs := 10; \Just\,((), \vs)\\
-  & \ind\hspace{8.5em}                               \False \to \letdef \vs := 20; \Just\,((), \vs);\\
-  & \ind \case \msf{act}\,\vs\,\of\\
-  & \ind\ind \Just\,(\_,\,\vs) \to \letdef \vs := \vs + \vs; \Just\,((),\,\vs)\\
-  & \ind\ind \Nothing\hspace{0.6em} \to \Nothing
-\end{alignat*}
-This solution is fairly good when we only have $\State$ or $\Reader$ effects, where
-$\mdown$ only introduces a runtime pair constructor, and it is feasible to
-compile object-level pairs as unboxed data, without overheads. However, for
-$\Maybe$, $\mdown$ introduces a runtime $\Just$ or $\Nothing$, and $\mup$
-introduces a runtime case split. A better solution would be to introduce two
-let-bound join points before the offending $\mbf{case}$, one for returning a
-$\Just$ and one for returning $\Nothing$, but fusing away the actual runtime
-constructors:
-\begin{alignat*}{3}
-  & \vf\,\vb\,\vs :=\\
-  & \ind \letdef\,\msf{joinNothing}\,\vs := \Nothing;\\
-  & \ind \letdef\,\msf{joinJust}\,\hspace{1.8em}\vs := \letdef \vs := \vs + \vs; \Just\,((),\,\vs);\\
-  & \ind \case \msf{b}\,\of\\
-  & \ind \ind \True\hspace{0.2em}  \to \letdef \vs := 10; \msf{joinJust}\,\vs\\
-  & \ind \ind \False \to \letdef \vs := 20; \msf{joinJust}\,\vs
-\end{alignat*}
-Here, $\msf{joinNothing}$ happens to be dead code, but it is easy to clean up in
-downstream compilation. Such ``fused'' returns are possible whenever we have a
-$\Gen\,\vA$ action at the bottom of the transformer stack, such that $\vA$ is
-isomorphic to a meta-level finite sum of value types. Recall that $\Gen\,\vA$ is
-defined as $\{\vR : \VTy\} \to (\vA \to \Up \vR) \to \Up \vR$. Here, if $\vA$ is
-a finite sum, we can rearrange $\vA \to \Up \vR$ to a finite product of
-functions.
-
-We could proceed with finite sums, but we will need finite
-\emph{sums-of-products} ($\SOP$ in short) later in Section in
-\ref{sec:stream-fusion}, so we develop SOP-s. SOPs have been used in generic and
-staged programming; see e.g.\ \cite{sop} and
-\cite{DBLP:conf/haskell/PickeringLW20}. We view $\SOP$-s as a Tarski-style
-universe consisting of a type of descriptions and a way of interpreting
-descriptions into $\MTy$ (``$\El$'' for ``elements'' of the type).
-\begin{alignat*}{3}
-  &\USOP : \MTy                    &&\ind \El_\SOP : \USOP \to \MTy \\
-  &\USOP = \List\,(\List\,\VTy)    &&\ind \El_\SOP\,\vA = ...
-\end{alignat*}
-A type description is a list of lists of value types. We decode this to a sum of
-products of value types. For example, $\El_\SOP\,[[\Bool,\,\Bool],\,[]]$ (using
-Haskell-like list notation) is isomorphic to $\Either\,(\Lift\Bool,\,\Lift\Bool)\,()$.
-$\USOP$ is closed under
-value types, finite product types and finite sum types. For instance, we have
-$\Either_\SOP : \USOP \to \USOP \to \USOP$ together with $\Left_\SOP : \El_\SOP\,\vA
-\to \El_\SOP\,(\Either_\SOP\,\vA\,\vB)$, $\Right_\SOP : \El_\SOP\,\vB \to
-\El_\SOP\,(\Either_\SOP\,\vA\,\vB)$ and a case splitting operation. It is more
-convenient to work with type formers in $\MTy$, and only convert to $\SOP$
-representations when needed, so we define a class for the representable types:
-\begin{alignat*}{3}
-  & \hspace{-5em}\mathrlap{\tyclass\,\IsSOP\,(\vA : \MTy)\,\where} \\
-  & \hspace{-5em}\ind \Rep      &&: \USOP \\
-  & \hspace{-5em}\ind \mit{rep} &&: \vA \simeq \El_\SOP\,\Rep
-\end{alignat*}
-Above, $\vA \simeq \Rep$ denotes a record containing a pair of back-and-forth
-functions, together with proofs (as propositional equalities) that they are
-inverses. We will overload $\mit{rep}$ as the forward conversion function with
-type $\vA \to \El_\SOP\,\Rep$, and write $\mit{rep}^{-1}$ for its inverse.
-
-In 2LTT literature, a metatype $\vA$ is called \emph{cofibrant} if $\vA \to \vB$ is
-isomorphic to an object type whenever $\vB$ is isomorphic to an object type
-\cite{twolevel}. We conjecture that our $\IsSOP$ types are exactly the cofibrant
-types in this sense. However, ``cofibrant'' is not very descriptive in our
-setting, so we shall keep writing $\IsSOP$.
-
-Accordingly, we define an isomorphic presentation of $\El_\SOP\,\vA \to
-\Lift \vR$ as a product of object-level functions:
-\begin{alignat*}{3}
-  &\mathrlap{\mit{Fun}_{\SOP\Lift} : \USOP \to \Ty \to \MTy}\\
-  &\mit{Fun}_{\SOP\Lift}\,\Nil\,&& \vR = ()\\
-  &\mit{Fun}_{\SOP\Lift}\,(\Cons\,\vA\,\vB)\,&& \vR = (\Lift(\mit{foldr}\,(\to)\,\vR\,\vA),\,\mit{Fun}_{\SOP\Lift}\,\vB\,\vR)
-\end{alignat*}
-\begin{alignat*}{3}
-  &\mtabulate &&: (\El_\SOP\,\vA \to \Lift \vR) \to \mit{Fun}_{\SOP\Lift}\,\vA\,\vR\\
-  &\mindex    &&: \mit{Fun}_{\SOP\Lift}\,\vA\,\vR \to (\El_\SOP\,\vA \to \Lift \vR)
-\end{alignat*}
-We omit here the definitions of $\mtabulate$ and $\mindex$. We will also need to
-let-bind all functions in a $\mit{Fun}_{\SOP\Lift}$:
-\begin{alignat*}{3}
-  & \mathrlap{\mit{genFun}_{\SOP\Lift} : \{\vA : \USOP\} \to \mit{Fun}_{\SOP\Lift}\,\vA\,\vR \to \mit{Fun}_{\SOP\Lift}\,\vA\,\vR}\\
-  & \mit{genFun}_{\SOP\Lift}\,\{\Nil\}        && ()             &&= \return\,()\\
-  & \mit{genFun}_{\SOP\Lift}\,\{\Cons\,\_\,\vA\}&& (\vf,\,\mit{\vfs}) &&= ({,})\,{\fmap}\,\gen\,\vf\,{\ap}\,\mit{genFun}_{\SOP\Lift}\,\{\vA\}\,\vfs
-\end{alignat*}
-We introduce a class for monads that support control flow joining.
-\begin{alignat*}{3}
-  & \tyclass\,\Monad\,\vM \RA \MonadJoin\,\vM\,\where\\
-  & \ind \join : \IsSOP\,\vA \RA \vM\,\vA \to \vM\,\vA
-\end{alignat*}
-The most interesting instance is the ``base case'' for $\Gen$:
-\begin{alignat*}{3}
-  &\instance \MonadJoin\,\Gen\,\where\\
-  &\ind \join\,\vma = \Gen \dlr \lam \vk.\,\runGen \dlr \mdo\\
-  &\ind \ind \mit{joinPoints} \fro \mit{genFun}_{\SOP\Lift}\,(\mtabulate\,(\vk \circ \mit{rep}^{-1}))\\
-  &\ind \ind \va \fro \vma\\
-  &\ind \ind \return \dlr \mindex\,\mit{joinPoints}\,(\mit{rep}\,\va)
-\end{alignat*}
-Here we first convert $\vk : \vA \to \Lift \vR$ to a product of join points and
-let-bind each one of them. Then we generate code that returns to the appropriate
-join point for each return value. Other $\MonadJoin$ instances are straightforward. In
-$\StateT_\M$, we need the extra $\IsSOP\,\vS$ constraint because $\vS$ is returned
-as a result value.
-\begin{alignat*}{3}
-  & \instance\,(\MonadJoin\,\vM) \RA \MonadJoin\,(\MaybeT_\M\,\vM)\,\where\\
-  & \ind \join\,(\MaybeT_\M\,\vma) = \MaybeT_\M\,(\join\,\vma)\\
-  & \instance\,(\MonadJoin\,\vM) \RA \MonadJoin\,(\ReaderT_\M\,\vR\,\vM)\,\where\\
-  & \ind \join\,(\ReaderT_\M\,\vma) = \ReaderT_\M\,(\join \circ \vma)\\
-  & \instance\,(\MonadJoin\,\vM,\,\IsSOP\,\vS) \RA \MonadJoin\,(\StateT_\M\,\vS\,\vM)\,\where\\
-  & \ind \join\,(\StateT_\M\,\vma) = \StateT_\M\,(\join \circ \vma)
-\end{alignat*}
-Now, the following definition yields the previously seen code output with
-$\msf{joinNothing}$ and $\msf{joinJust}$.
-\begin{alignat*}{3}
-  & \vf\,\vb := \spl(\mdown \dlr \mdo \\
-  & \ind \join \dlr \case \qt{\vb}\,\of\,\True\to \mput' \qt{10}; \False \to \mput' \qt{20}\\
-  & \ind\modify'\,(\lambda\,x.\,\qt{\spl x + \spl x}))
-\end{alignat*}
-It might make sense to also have a simple ``desugaring'' rule that inserts a
-$\join$ whenever we have an object-level case split with more than one branch. At
-worst, this generates some noise and dead code that is easy to remove by
-conservative code optimization.
-
-\subsection{Code Example}
-
-We look at a slightly larger code example in Figure \ref{fig:codeinp}. We define
-annotated binary trees as
-\[ \data \Tree\,\vA := \Leaf\,|\,\Node\,\vA\,(\Tree\,\vA)\,(\Tree\,\vA). \]
-We write $\msf{fail : M\,A}$ for returning $\Nothing$ in any monad transformer
-stack that contains $\MaybeT$. We define a function which traverses a tree and
-replaces values with values taken from a list. If the tree contains $0$, we
-throw an error. If we run out of list elements, we leave values unchanged.
-\begin{figure}
-\begin{alignat*}{4}
-  &  \letrec &&\vf : \Tree\,\Nat \to \StateT\,(\List\,\Nat)\,(\MaybeT\,\Identity)\,(\Tree\,\Nat) \\
-  &          &&\vf\,\vt := \spl \big(\mdown \dlr \case \vt\,\of\\
-  &          &&\ind \Leaf \to \return \qt{\Leaf} \\
-  &          &&\ind \Node\,\vn\,\vl\,\vr \to \mdo\\
-  &          &&\ind \ind \case \qt{\spl \vn == 0}\,\of\,\True \hspace{0.2em}\to \msf{fail}\\
-  &          &&\ind \ind \ind \hspace{6.7em}             \False \to \return () \\
-  &          &&\ind \ind \vns \fro \get \\
-  &          &&\ind \ind \vn\hspace{0.35em}\fro\join \dlr \case \vns\,\of\,\Nil \hspace{2.65em}\to \return \vn \\
-  &          &&\ind \ind \ind \hspace{8.2em}\Cons\,\vn\,\vns \to \mdo \{\msf{put}\,\vns; \return \vn\} \\
-  &          &&\ind \ind \vl\hspace{0.1em} \fro \mup\,\qt{\vf\,\spl \vl} \\
-  &          &&\ind \ind \vr \fro \mup\,\qt{\vf\,\spl \vr} \\
-  &          &&\ind \ind \return \qt{\Node\,\spl \vn\,\spl \vl\,\spl \vr}\big)
-\end{alignat*}
-\caption{Monadic source code}
-\label{fig:codeinp}
-\end{figure}
-Here, all of the $\case$ matches are done on object-level values, so they are
-all desugared to applications of $\msf{split}$.
-\begin{itemize}
-\item Note the $\join$: without it, the recursive $\Node$ traversal code would be inlined in both case branches.
-\item We do not use $\join$ when checking $\qt{\spl n == 0}$. Here, it
-      happens to be superfluous, since the $\msf{fail}$ ``destroys'' all subsequent
-      code generation in the branch, by short-circuiting at the meta-level.
-\item Also, we can use $\msf{put}$ instead of the ``stricter'' $\msf{put'}$ because
-      the state modification immediately gets sequenced by the enclosing join point.
-\item To make object-level recursive calls, we just need to wrap them in $\mup$.
-\end{itemize}
-Here, we intentionally tuned the definition for a nice-looking unstaged output.
-However, we could also just default to ``safe'' choices everywhere: we could use
-a joint point in every $\mbf{case}$ with two or more branches, and use the
-strict state modifications everywhere, and leave it up to the downstream compiler
-to clean up the code.
-
-We show the unstaging output on Figure \ref{fig:codeout}. We omit newtype
-wrappers and use nested pattern matching on pairs, but otherwise this is exactly
-the code that we get in our Agda implementation. Again, the only flaw in this
-code is the dead binding for $\msf{joinNothing}$.
-\begin{figure}\label{fig:codeout}
-\begin{alignat*}{4}
-  &  \letrec &&\vf : \Tree\,\Nat \to \StateT\,(\List\,\Nat)\,(\MaybeT\,\Identity)\,(\Tree\,\Nat) \\
-  &          &&\vf := \lam \vt\,\vns.\, \case\,\vt\,\of\\
-  &          &&\ind \Leaf \to \Just\,(\Leaf,\,\vns) \\
-  &          &&\ind \Node\,\vn\,\vl\,\vr \to \case (\vn == 0)\,\of\\
-  &          &&\ind \ind \True \hspace{0.25em}\to \Nothing \\
-  &          &&\ind \ind \False \to \\
-  &          &&\ind \ind \ind \letdef \msf{joinNothing} := \lam \_.\,\Nothing;\\
-  &          &&\ind \ind \ind \letdef \msf{joinJust} := \lam \vn\,\vns.\,\case\,\vf\,\vl\,\vns\,\of\\
-  &          &&\ind \ind \ind \ind \ind \ind \Nothing\hspace{0.9em} \to \Nothing\\
-  &          &&\ind \ind \ind \ind \ind \ind \Just\,(\vl,\,\vns) \to \case\,\vf\,\vr\,\vns\,\of\\
-  &          &&\ind \ind \ind \ind \ind \ind \ind \Nothing\hspace{.95em} \to \Nothing\\
-  &          &&\ind \ind \ind \ind \ind \ind \ind \Just\,(\vr,\,\vns) \to \Just\,(\Node\,\vn\,\vl\,\vr,\,\vns);\\
-  &          &&\ind \ind \ind \case \vns\,\of\\
-  &          &&\ind \ind \ind \ind \Nil \hspace{2.65em}\to \msf{joinJust}\,\vn\,\vns\\
-  &          &&\ind \ind \ind \ind \Cons\,\vn\,\vns \to \msf{joinJust}\,\vn\,\vns
-\end{alignat*}
-\caption{Unstaged monadic code}
-\label{fig:codeout}
-\end{figure}
-\subsection{Discussion}
-
-So far, we have a monad transformer library with the following features:
-\begin{itemize}
-\item Almost all definitions from the well-known Haskell ecosystem of monads and monad transformers
-      can be directly reused, in the meta level.
-\item We can pattern match on object-level values in monadic code, insert object-level $\mbf{let}$-s
-      with $\gen$ and avoid code duplication with $\join$.
-\item In monadic code, object-level data constructors are only ever created by
-      $\mdown$, and matching on object-level data is only created by $\mit{split}$
-      and $\mup$. Monadic operations are fully fused, and all function calls can be
-      compiled to statically known saturated calls.
-\end{itemize}
-
-As to potential weaknesses, first, the system as described in this section has
-some syntactic noise and requires extra attention from programmers.  We believe
-that the noise can be mitigated very effectively in a native CFTT
-implementation. It was demonstrated in a 2LTT implementation in
-\cite{staged-demo} that almost all quotes and splices are unambiguously
-inferable, if we require that stages of let-definitions are always specified (as
-we do here). Moreover, $\mup$ and $\mdown$ should be also effectively inferable,
-using bidirectional elaboration. With such inference, monadic code in CFTT would
-look only slightly more complicated than in Haskell.
-
-Second, in CFTT we cannot store computations (e.g.\ functions or $\State$
-actions) in runtime data structures, nor can we have computations in $\State$
-state or in $\Reader$ environments. However, it would be possible to extend CFTT
-with a closure type former that converts computations to values, in which case
-there is no such limitation anymore. Here, closure-freedom would be still
-available; we would be able to pick where to use or avoid the closure type
-former.
-
-\subsection{Agda \& Haskell Implementations}
-
-We implemented everything in this section in both Agda and typed Template
-Haskell. We summarize features and differences:
-\begin{itemize}
-\item The Haskell implementation can be used to generate code that can be
-      further compiled by GHC; here the object language is taken to be Haskell
-      itself. Since Haskell does not distinguish value and computation types, we do
-      not track them in the library, and we do not get guaranteed closure-freedom
-      from GHC.
-\item In Agda, we postulate all types and terms of the object theory in a
-  faithful way (i.e.\ equivalently to the CFTT syntax presented here), and take
-  Agda itself to be the metalanguage. Here, we can test ``staging'' by running Agda
-  programs which compute object expressions. However, we can only inspect
-  staging output and cannot compile or run object programs.
-\item For sums-of-products in Haskell, we make heavy use of \emph{singleton
-  types} \cite{DBLP:conf/haskell/EisenbergW12} to emulate dependent types. This adds significant
-  noise. Also, in $\IsSOP$ instances we can only define the conversion functions
-  and cannot prove that they are inverses, because Haskell does not have enough support
-  for dependent types.
-\end{itemize}
-
-\section{Stream Fusion}\label{sec:stream-fusion}
-
-Stream fusion refers to a collection of techniques for generating efficient code
-from declarative definitions involving streams of values, where intermediate
-data structures are eliminated. Stream fusion can be broadly grouped into
-\emph{push} fusion, which is based on Church-encodings of inductive lists, and
-$\emph{pull}$ fusion, which is based on Church-encodings of coinductive lists;
-see e.g.\ \cite{DBLP:conf/ifl/HinzeHJ10}. The two styles have different
-trade-offs, and in practical programming it is a good idea to support both, but
-in this section we focus on pull streams.
-
-The reason is that pull streams have been historically more difficult to
-efficiently compile, and we can demonstrate significant improvements in CFTT.
-We also use dependent types in a more essential way than in the previous
-section.
-
-\subsection{Streams}
-
-A pull stream is a meta-level specification of a state machine:
-\begin{alignat*}{3}
-  & \data \Step\,\vS\,\vA = \Stop\,|\,\Skip\,\vS\,|\,\Yield\,\vA\,\vS\\
-  & \data \Pull\,(\vA : \MTy) : \MTy\,\where\\
-  & \ind \Pull : (\vS : \MTy) \to \IsSOP\,\vS \RA \Gen\,\vS \to (\vS \to \Gen\,(\Step\,\vS\,\vA)) \to \Pull\,\vA
-\end{alignat*}
-In the $\Pull$ constructor, $\vS$ is the type of the internal state which is
-required to be a sum-of-products of value types by the $\IsSOP\,\vS$
-constraint. The next field with type $\Gen\,\vS$ is the initial state, while
-transitions are represented by the $\vS \to \Gen\,(\Step\,\vS\,\vA)$
-field. Possible transitions are stopping ($\Stop$), transitioning to a new state
-while outputting a value ($\Yield$) and making a silent transition to a new
-state ($\Skip$).
-
-Our definition roughly follows the non-staged stream fusion setup of
-Coutts \cite{DBLP:phd/ethos/Coutts11}; the difference is the addition of
-$\IsSOP$ and $\Gen$ in our version. Also, borrowing terminology from Coutts, we
-call each product in the state a \emph{state shape}.
-
-Let us see some operations on streams now. $\Pull$ is evidently a $\Functor$. It
-is not quite a ``zippy'' applicative functor, because its application operator
-requires an extra $\IsSOP$ constraint:
-\begin{alignat*}{3}
-  &\mit{repeat} : \vA \to \Pull\,\vA\\
-  &\mit{repeat}\,\va = \Pull\,()\,(\return\,())\,(\lam \_.\,\return \dlr \Yield\,\va\,())
-\end{alignat*}
-\begin{alignat*}{3}
-  &\hspace{-7em}\mathrlap{({\ap_{\Pull}}) : \IsSOP\,\vA \RA \Pull\,(\vA \to \vB) \to \Pull\,\vA \to \Pull\,\vB }\\
-  &\hspace{-7em}\mathrlap{({\ap_{\Pull}})\,(\Pull\,\vS\,\seed\,\step)\,(\Pull\,\vS'\,\seed'\,\step') =}\\
-  &\hspace{-7em}\mathrlap{ \ind \Pull\,(\vS,\,\vS',\,\Maybe\,\vA)\,(({,}) \fmap \seed \ap ((,\Nothing) \fmap \seed') \dlr \lam \case }\\
-  &\hspace{-7em}\mathrlap{ \ind \ind (\vs,\,\vs',\,\Just\,\va) \to \step\,\vs \bind \lam \case}\\
-  &\hspace{-7em} \ind \ind \ind \Stop        &&\to \return \Stop \\
-  &\hspace{-7em} \ind \ind \ind \Skip\,\vs     &&\to \return \dlr \Skip\,(\vs,\,\vs',\,\Just\,\va) \\
-  &\hspace{-7em} \ind \ind \ind \Yield\,\vf\,\vs &&\to \return \dlr \Yield\,(\vf\,\va)\,(\vs,\,\vs',\,\Nothing) \\
-  &\hspace{-7em} \mathrlap{\ind \ind (\vs,\,\vs',\,\Nothing) \to \step'\,\vs' \bind \lam \case }\\
-  &\hspace{-7em} \ind \ind \ind \Stop         &&\to \return \Stop \\
-  &\hspace{-7em} \ind \ind \ind \Skip\,\vs'     &&\to \return \dlr \Skip\,(\vs,\,\vs',\,\Nothing) \\
-  &\hspace{-7em} \ind \ind \ind \Yield\,\va\,\vs' &&\to \return \dlr \Skip\,(\vs,\,\vs',\,\Just\,\va)
-\end{alignat*}
-
-In $\mit{repeat}$, the state is the unit type, while in $({\appull})$ we take
-the product of states, and also add another $\Maybe\,\vA$ that is used to buffer
-a single value while we are stepping the $\Pull\,(\vA \to \vB)$ stream.
-The $\IsSOP$ constraints for the new machine states are implicitly dispatched
-by instance resolution; this works in the Agda and Haskell versions too. We can
-derive $\mit{zip}$ and $\mit{zipWith}$ from $({\appull})$ and $({\fmap})$ for
-streams, with the restriction that the zipped streams must all produce $\IsSOP$
-values.
-
-$\Pull$ is also a monoid with stream appending as the binary operation.
-\begin{alignat*}{3}
-  & \mit{empty} : \Pull\,\vA\\
-  & \mit{empty} = \Pull\,()\,(\return\,())\,(\lam \_.\,\return \Stop)
-\end{alignat*}
-\begin{alignat*}{5}
-  & \hspace{-5em}\mathrlap{({<>}) : \Pull\,\vA \to \Pull\,\vA \to \Pull\,\vA} \\
-  & \hspace{-5em}\mathrlap{({<>})\,(\Pull\,\vS\,\seed\,\step)\,(\Pull\,\vS'\,\seed'\,\step') = \Pull\,(\Either\,\vS\,\vS')\,(\Left \fmap \seed)\,\lam \case} \\
-  & \hspace{-5em}\ind \Left\,\vs &&\to \step\,s \bind \lam \case &&\Stop             &&\to (\Skip \circ \Right) \fmap \seed'\\
-  & \hspace{-5em}                    &&                          &&\Skip\,\vs        &&\to \return \dlr \Skip\,(\Left\,\vs) \\
-  & \hspace{-5em}                    &&                        &&\Yield\,\va\,\vs    &&\to \return \dlr \Yield\,\va\,(\Left\,\vs)\\
-  & \hspace{-5em}\ind \Right\,\vs' &&\to \step\,s' \bind \lam \case &&\Stop          &&\to \return \Stop\\
-  & \hspace{-5em}                    &&                        &&\Skip\,\vs'         &&\to \return \dlr \Skip\,(\Right\,\vs') \\
-  & \hspace{-5em}                    &&                        &&\Yield\,\va\,\vs'   &&\to \return \dlr \Yield\,\va\,(\Right\,\vs')
-\end{alignat*}
-These definitions are standard for streams; compared to the non-staged
-definitions in previous literature, the only additional noise is just the $\Gen$
-monad in the initial states and the transitions. Likewise, we can give standard
-definitions for usual stream functions such as $\msf{filter}$, $\msf{take}$ or
-$\msf{drop}$.
-
-\subsection{Running Streams with Mutual Recursion}
-
-How do we generate object code from streams? The state $\vS$ is given as a
-finite sums of products, but the sums and the products are on the meta level, so
-we cannot directly use $\vS$ in object code. Similarly as in the the treatment
-of join points, we tabulate the $\vS \to \Gen\,\Step\,\vS\,\vA$ transition
-function to a product of functions. However, these functions need to be
-\emph{mutually recursive}, since it is possible to transition from any state to
-any other state, and each such transition is represented as a function call.
-This problem of generating well-typed mutual blocks was addressed in
-\cite{DBLP:conf/pepm/YallopK19}. In contrast to this work, which used control
-effects and mutable references in MetaOCaml, we present a solution that does not
-use side effects in the metalanguage.
-
-The solution is to extend $\CTy$ with finite products of computations,
-i.e.\ assume $() : \CTy$ and $({,}) : \CTy \to \CTy \to \CTy$, together with
-pairing and projections. These types, like functions, are call-by-name in the
-runtime semantics, and they also cannot escape the scope of their definition.
-Hence, we can also ``saturate'' programs that involve computational product types:
-every computation definition at type $(\vA,\,\vB)$ can be translated to a pair,
-and every projection of a let-defined variable can be statically matched up with
-a pairing in the variable definition. Thus, a recursive let-definition at type
-$(\vA \to \vB,\,\vA \to \vB)$ can be always compiled to a pair of mutual
-functions.
-
-We redefine the previous $\mit{Fun}_{\SOP\Lift}$ to return a computation type
-instead:
-\begin{alignat*}{3}
-  &\mathrlap{\mit{Fun}_{\SOP\Lift} : \USOP \to \Ty \to \CTy}\\
-  &\mit{Fun}_{\SOP\Lift}\,\Nil\,&& \vR = ()\\
-  &\mit{Fun}_{\SOP\Lift}\,(\Cons\,\vA\,\vB)\,&& \vR = (\mit{foldr}\,(\to)\,\vR\,\vA,\,\mit{Fun}_{\SOP\Lift}\,\vB\,\vR)
-\end{alignat*}
-\begin{alignat*}{3}
-  &\mtabulate &&: (\El_\SOP\,\vA \to \Lift \vR) \to \Lift(\mit{Fun}_{\SOP\Lift}\,\vA\,\vR)\\
-  &\mindex    &&: \Lift(\mit{Fun}_{\SOP\Lift}\,\vA\,\vR) \to (\El_\SOP\,\vA \to \Lift \vR)
-\end{alignat*}
-Even for join points, this is just as efficient as the previous
-$\mit{Fun}_{\SOP\Lift}$ version, since a definition of a product of functions
-will get compiled to a sequence of function definitions. In addition, we can use
-it to generate object code from streams. There are several choices for this, but
-in CFTT the $\msf{foldr}$ function is as good as we can get.
-\begin{alignat*}{3}
-  &\hspace{-13em}\mathrlap{\foldr : \{\vA : \MTy\}\{\vB : \Ty\} \to (\vA \to \Lift \vB \to \Lift \vB) \to \Lift \vB \to \Pull\,\vA \to \Lift\,\vB}\\
-  &\hspace{-13em}\mathrlap{\foldr\,\{\vA\}\,\{\vB\}\,\vf\,\vb\,(\Pull\,\vS\,\seed\,\step) = \ql}\\
-  &\hspace{-13em}\mathrlap{\ind \letrec \vfs : \mit{Fun}_{\SOP\Lift}\,(\Rep\,\{\vS\})\,\vB := \spl\big(\mtabulate \dlr \lam \vs.\,\unGen\,(\step\,(\rep^{-1}\,\vs)) \dlr \lam \case} \\
-  &\hspace{-13em}\ind \ind \ind \Stop            &&\to \vb \\
-  &\hspace{-13em}\ind \ind \ind \Skip\,\vs       &&\to \mindex\,\qt{\vfs}\,(\rep\,\vs) \\
-  &\hspace{-13em}\ind \ind \ind \Yield\,\va\,\vs &&\to \vf\,\va\,(\mindex\,\qt{\vfs}\,(\rep\,\vs))\big);\\
-  &\hspace{-13em}\mathrlap{\ind \spl\big(\unGen\,\seed \dlr \lam s.\,\mindex\,\qt{\vfs}\,(\rep\,\vs)\big)\qr}
-\end{alignat*}
-This $\foldr$ is quite flexible because it can produce object terms with computation types.
-For instance, we can define $\foldl$ from $\foldr$:
-\begin{alignat*}{3}
-  &\foldl : \{\vA : \MTy\}\{\vB : \VTy\} \to (\Lift \vB \to \vA \to \Lift \vB) \to \Lift \vB \to \Pull\,\vA \to \Lift\,\vB\\
-  &\foldl\,\vf\,\vb\,\vas = \spl(\foldr\,(\lam \va\,\vg.\,\qt{\lam \vb.\,\spl \vg\,\spl(\vf\,\qt{\vb}\,\spl \va)})\,\qt{\lam \vb.\,\vb}\,\vas)\,\spl \vb
-\end{alignat*}
-Note that since we abstract $\vB$ in a runtime function, it must be a value
-type. Here, each $\Stop$ and $\Yield$ in the transition function gets
-interpreted as a $\lambda$-expression in the output. However, those $\lambda$-s
-will be lifted out in the scope, yielding a proper mutually tail-recursive
-definition with an accummulator for $\vB$. Contrast this to the GHC base
-library, where $\foldl$ for lists is also defined from $\foldr$, to enable
-push-based fusion, but where a substantial \emph{arity analysis} is used in
-the compiler to (incompletely) eliminate the intermediate closures
-\cite{DBLP:conf/sfp/Breitner14}.
-
-\subsection{$\mbf{concatMap}$ for Streams}\label{sec:concatmap-for-streams}
-
-Can we have a list-like $\Monad$ instance for streams, with singleton streams
-for $\return$ and $\concatMap$ for binding? This is not possible. Aiming at
-\[ \concatMap : (\vA \to \Pull\,\vB) \to \Pull\,\vA \to \Pull\,\vB, \]
-the $\vA \to \Pull\,\vB$ function can contain an infinite number of different
-machine state types, which cannot be represented in a finite amount of object
-code. Here by ``infinite'' we mean the notion that is internally available in
-the meta type theory. For instance, we can define a $\Nat_\M \to \Pull\,\vB$
-function which for each $\vn : \Nat_\M$ produces a concatenation of $\vn$
-streams. Hence, we shall have the following function instead:
-\[ \concatMap : \IsSOP\,\vA \RA (\vA \to \Pull\,\vB) \to \Pull\,\vA \to \Pull\,\vB \]
-The idea is the following: if $\USOP$ is closed under $\Sigma$-types, we can directly
-define this $\concatMap$, by taking the appropriate dependent sum of the $\vA
-\to \USOP$ family of machine states, which we extract from the $\vA \to
-\Pull\,\vB$ function. Let us write $\Sigma\,\vA\,\vB : \MTy$ for dependent sums,
-for $\vA : \MTy$ and $\vB : \vA \to \MTy$, and reuse $({,})$ for pairing.  We
-also use the following field projection functions: $\mit{projS} : \Pull\,\vA \to
-\MTy$, $\mit{projSeed} : (\vas : \Pull\,\vA) \to \mit{projS}\,\vas$, and
-$\mit{projStep} : (\vas : \Pull\,\vA) \to \mit{projS}\,\vas \to
-\Gen\,(\Step\,(\mit{projS}\,\vas)\,\vA)$.  For now, we assume that the following
-instance exists:
-\[ \instance (\IsSOP\,\vA,\,\{\va : \vA\} \to \IsSOP\,(\vB\,\va)) \RA \IsSOP\,(\Sigma\,\vA\,\vB) \]
-Above, $\{\va : \vA\} \to \IsSOP\,(\vB\,\va)$ is a universally quantified
-instance constraint; this can be also represented in Agda. The definition of
-$\concatMap$ is as follows.
-\begin{alignat*}{3}
-  & \mathrlap{\concatMap : \IsSOP\,\vA \RA (\vA \to \Pull\,\vB) \to \Pull\,\vA \to \Pull\,\vB }\\
-  & \mathrlap{\concatMap\,\{\vA\}\,\{\vB\}\,\vf\,(\Pull\,\vS\,\seed\,\step) = }\\
-  & \mathrlap{\ind \Pull\,(\vS,\,\Maybe\,(\Sigma\,\vA\,(\mit{projS} \circ \vf)))\,(({,}) \fmap \seed\,\ap\,\return\,\Nothing) \dlr \lam \case }\\
-  & \mathrlap{\ind \ind (\vs,\,\Nothing) \to \step\,\vs \bind \lam \case }\\
-  & \ind \ind \ind \ind \Stop                 &&\to \return \Stop\\
-  & \ind \ind \ind \ind \Skip\,\vs            &&\to \return \dlr \Skip\,(\vs,\, \Nothing)\\
-  & \ind \ind \ind \ind \Yield\,\va\,\vs      &&\to \mdo \{\vs' \fro \mit{projSeed}\,(\vf\,\va); \return \dlr \Skip\,(\vs,\,\Just\,(\va,\,\vs'))\}\\
-  & \mathrlap{\ind \ind (\vs,\,\Just\,(\va,\,\vs')) \to \mit{projStep}\,(\vf\,\va)\,\vs' \bind \lam \case }\\
-  & \ind \ind \ind \ind \Stop                &&\to \return \dlr \Skip\,(\vs,\,\Nothing) \\
-  & \ind \ind \ind \ind \Skip\,\vs'          &&\to \return \dlr \Skip\,(\vs,\,\Just\,(\va,\,\vs'))\\
-  & \ind \ind \ind \ind \Yield\,\vb\,\vs'    &&\to \return \dlr \Yield\,\vb\,(\vs,\, \Just\,(\va,\,\vs'))
-\end{alignat*}
-Here, $\Nothing$ marks the states where we are in the ``outer'' loop, running
-the $\Pull\,\vA$ stream until we get its next value. $\Just$ marks the states of
-the ``inner'' loop, where we have a concrete $\va : \vA$ value and we run the
-$(\vf\,\va)$ stream until it stops. In the inner loop, the machine state type
-depends on the $\va : \vA$ value, hence the need for $\Sigma$. How do we get
-$\IsSOP$ for $\Sigma$? The key observations are:
-\begin{itemize}
- \item Metaprograms cannot inspect the structure of object terms.
- \item Object types do not depend on object terms.
-\end{itemize}
-Hence, we expect that during unstaging, every $\vf : \Lift\,\vA \to \VTy$
-function has to be constant. More generally in the semantics in Appendix \ref{appendix:A}, every function whose domain is a product
-of object types and whose codomain is a constant presheaf, is a constant
-function. We may call these constancy statements \emph{generativity axioms},
-since they reflect the inability of metaprograms to inspect terms, and
-``generativity'' often refers to this property in the staged compilation
-literature.
-
-Let us write $\Uprod = \List\,\VTy$ and $\Elprod : \Uprod \to \MTy$ for a
-universe of finite products of value types. Concretely, in CFTT and the Agda
-implementation, we assume the following:
-\\\\
-\noindent \textbf{Axiom (generativity).}\textit{ For every $\vf : \Elprod\,\vA \to \USOP$ and $\va,\,\va' : \Elprod\,\vA$,
-we have that $\vf\,\va = \vf\,\va'$.}
-\\\\
-See Section \ref{appendix:generativity} in Appendix $\ref{appendix:A}$ for a
-validation of this axiom. We remark that there is no risk of staging getting
-stuck, because propositional equality proofs get erased, as we described in
-Section \ref{sec:semantics-of-staging}. We also remark that although
-generativity is inconsistent with inspecting the structure of object terms, it
-is consistent with inspecting the structure of object types.
-
-From generativity, we derive $\Sigma_\SOP : (\vA : \USOP) \to (\vB : \El_\SOP\,\vA \to \USOP)
-\to \USOP$ as follows. First, for each $\vA : \Uprod$, we define $\msf{loop_{\vA}} :
-\Elprod\,\vA$ as a product of non-terminating object programs. This is only needed
-to get arbitrary inhabitants with which we can call $\Elprod\,A \to \USOP$
-functions.
-
-Then, $\Sigma_\SOP\,\vA\,\vB$ is defined as the concatenation of $\vA_i \times
-\vB\,(\msf{inject_i}\,\msf{loop_{\vA_i}})$ for each $\vA_i \in \vA$, where
-$({\times})$ is the product type former in $\USOP$ and $\msf{inject_i}$
-has type $\Elprod\,\vA_i \to \El_\SOP\,\vA$. This is similar to the definition of
-non-dependent products in $\USOP$, except that we have to get rid of the type
-dependency by instantiating $\vB$ with arbitrary programs.
-
-Then, we can show using the generativity axiom that $\Sigma_\SOP$ supports
-projections, pairing and the $\beta\eta$-rules. These are all needed when we
-define the $\IsSOP$ instance, when we have to prove that encoding via $\rep$
-is an isomorphism. Concretely, assuming $\IsSOP\,\vA$ and $\{\va : \vA\} \to
-\IsSOP\,(\vB\,\va)$, we define $\Rep$ and $\msf{rep}$ for $\Sigma\,\vA\,\vB$ as follows:
-\[ \Rep\,\{\Sigma\,\vA\,\vB\} = \Sigma\,(\Rep\,\{\vA\})\,(\lam \vx.\,\Rep\,\{\vB\,(\rep^{-1}\,\vx)\}) \]
-\[  \rep\,\{\Sigma\,\vA\,\vB\}\,(\va,\,\vb) = (\rep\,\{\vA\}\,\va,\,\rep\,\{\vB\,(\rep^{-1}\,(\rep\,\vx))\}\,\vb) \]
-Note that $\rep$ is only well-typed up to the fact that $\rep^{-1} \circ \rep =
-\id$; the Agda definition contains an additional transport that we omit here.
-This is, in fact, our reason for including the isomorphism equations in
-$\IsSOP$.
-
-This, in turn, necessitates using SOP instead of finite sums, for the following
-reason. We can define a product type former for finite sums of value types, by
-taking the pairwise products of components, but in this case the products would
-be object-level products, which do not support $\beta\eta$ rules. This implies
-that we cannot prove $\beta\eta$-rules for the derived product type in finite
-sums. In contrast, when we define products for SOP-s, we take \emph{meta-level}
-pairwise products of components, which does support $\beta$ and $\eta$.
-
-We skip the full definition of $\Sigma_\SOP$ here. The reader may refer to the
-$\SOP$ module in the Agda implementation, which is altogether 260 lines with all
-$\IsSOP$ instances.
-
-\subsection{Let-Insertion \& Case Splitting in Monadic Style}
-
-While $\Pull$ is not a monad, and hence also not a $\MonadGen$, we can still use
-a monadic style of stream programming with good ergonomics. First, we need
-singleton streams for ``returning'':
-\begin{alignat*}{3}
-  & \mit{single} : \vA \to \Pull\,\vA \\
-  & \mit{single}\,\va = \Pull\,\Bool_\M\,\True \dlr \lam \vb.\,\return \dlr\mbf{if}\,\vb\,\mbf{then}\,\Yield\,\va\,\False\,\mbf{else}\,\Stop
-\end{alignat*}
-Now, we should use this operation with care, since it has two states and can
-contribute to a code size blow-up. For example, $\concatMap$ over $\single$
-introduces at least a doubling of the number of state shapes. Although
-let-insertion and case splitting could be derived as $\concatMap$ over
-$\single$, to avoid the size explosion we give more specialized definitions
-instead. First, we define an embedding of $\Gen$ operations in streams.
-\begin{alignat*}{3}
-  & \hspace{-11em}\mathrlap{\mit{bind_{\Gen}} : \IsSOP\,\vA \RA \Gen\,\vA \to (\vA \to \Pull\,\vB) \to \Pull\,\vB}\\
-  & \hspace{-11em}\mathrlap{\mit{bind_{\Gen}}\,\vga\,\vf\,=}\\
-  & \hspace{-11em}\mathrlap{\ind \Pull\,(\Sigma\,\vA\,(\mit{projS} \circ \vf))\,(\mdo \{\va \fro \vga; \vs \fro \mit{projSeed}\,(\vf\,\va); \return (\va,\,\vs)) \dlr \lam \case}\\
-  & \hspace{-11em}\mathrlap{\ind \ind (\va,\,\vs) \to \mit{projStep}\,(\vf\,\va)\,\vs \bind \lam \case }\\
-  & \hspace{-11em}\ind \ind \ind \Stop            &&\to \return \Stop\\
-  & \hspace{-11em}\ind \ind \ind \Skip\,\vs       &&\to \return \dlr \Skip\,(\va,\,\vs) \\
-  & \hspace{-11em}\ind \ind \ind \Yield\,\vb\,\vs &&\to \return \dlr \Yield\,\vb\,(\va,\,\vs)
-\end{alignat*}
-We recover let-insertion and case splitting as follows:
-\begin{alignat*}{3}
-  &\mit{gen_\Pull} : \Lift \vA \to (\Lift \vA \to \Pull\,\vB) \to \Pull\,\vB\\
-  &\mit{gen_\Pull}\,\va = \mit{bind_{\Gen}}\,(\gen\,\va)
-\end{alignat*}
-\begin{alignat*}{3}
-  &\mit{case_\Pull} : (\mit{Split}\,\vA,\,\IsSOP\,(\mit{SplitTo}\,\{\vA\})) \RA \Lift \vA \to (\mit{SplitTo}\,\{\vA\} \to \Pull\,\vB) \to \Pull\,\vB\\
-  &\mit{case_\Pull}\,\va = \mit{bind_{\Gen}}\,(\mit{split}\,\va)
-\end{alignat*}
-Let us look at small example. We write $\mit{forEach}$ for $\concatMap$ with
-flipped arguments.
-\begin{alignat*}{3}
-  &\hspace{-4em}\mathrlap{\mit{forEach}\,(\mit{take}\,\qt{100}\,(\mit{countFrom}\,\qt{0})) \dlr \lam \vx.}\\
-  &\hspace{-4em}\mathrlap{\gen_\Pull\, \qt{\spl \vx * 2} \dlr \lam \vy.}\\
-  &\hspace{-4em}\mathrlap{\mit{case_\Pull}\,\qt{\spl \vx < 50} \dlr \lam \case}\\
-  &\hspace{-4em} \ind \True  &&\to \mit{take}\,\vy\,(\mit{countFrom}\,x)\\
-  &\hspace{-4em} \ind \False &&\to \single\,\vy
-\end{alignat*}
-Here, in every $\forEach$ iteration, $\gen_\Pull\,\qt{\spl \vx * 2}$ evaluates
-the given expression and saves the result to the machine state. Then,
-$\mit{case_\Pull}$ branches on an object-level Boolean expression. If we use
-$\foldr$ to generate object code from this definition, we get \emph{four}
-mutually defined functions. We get two state shapes from $\single\,\vy$ and one
-from $\mit{take}\,\vy\,(\mit{countFrom}\,x)$; we sum these to get three for the
-$\mit{case_\Pull}$, then finally we get an extra shape from $\mit{forEach}$ which
-introduces an additional $\Maybe$.
-
-\subsection{Discussion}
-
-Our stream library has a strong support for programming in a monadic style, even
-though $\Pull$ is not literally a monad. We can bind object values with
-$\concatMap$, and we can also do let-insertion and case splitting for them. We
-also get guaranteed well-typing, closure-freedom, and arbitrary mixing of
-zipping and $\concatMap$.
-
-We highlight the usage of the generativity axiom as well. Previously in staged
-compilation, intensional analysis (i.e. the ability to analyze object code) has
-been viewed as a desirable feature that increases the expressive power of a
-system. To our knowledge, our work is the first one that exploits the
-\emph{lack} of intensional analysis in metaprogramming. This is a bit similar to
-parametricity in type theories, where the inability to analyze types has a
-payoff in the form of ``free theorems'' \cite{wadler89theoremsforfree}.
-
-Regarding the practical application of our stream library, we think that it
-would make sense to support both push and pull fusion in a realistic
-implementation, and allow users to benefit from the strong points of both. Push
-streams, which we do not present in this paper, have proper $\Monad$ and
-$\MonadGen$ instances and are often more convenient to use in CFTT. They are
-also better for deep traversals of structures where they can utilize unbounded
-stack allocations, while pull streams need heap allocations for unbounded space
-in the machine state.
-
-\subsection{Agda \& Haskell Implementations}
-
-In Agda, to avoid computation getting stuck on the generativity axiom, we use
-the \texttt{primTrustMe} built-in \cite{agdadocs} to automatically erase the axiom
-when the sides of the equation are definitionally equal. Otherwise the
-implementation closely follows this section.
-
-In Haskell there are some limitations. First, in $\concatMap$, the
-projection function $\mit{projS}$ cannot be defined, because Haskell is not
-dependently typed, and the other field projections are also out of reach. We
-only have a weaker ``positive'' recursion principle for existential types. It
-might be the case that a strongly typed $\concatMap$ is possible with only weak
-existentials, but we attempted this and found that it introduces too much
-technical complication.
-
-So instead of giving a single generic definition for $\concatMap$ for $\IsSOP$
-types, we define $\concatMap$ just for object types,\footnote{Recall that we do
-not distinguish value and computation types in Haskell.} and define
-$\mit{case_\Pull}$ separately for each object type, as an overloaded class
-method. In each of these definitions, we only need to deal with a concrete
-finite number of machine state types, which is feasible to handle with weak
-existentials.
-
-Also, the generativity axiom is \emph{false} in Template Haskell, since it is
-possible to look inside quoted expressions. Instead, we use type coercions that
-can fail at unstaging time. If library users do not violate generativity, these
-coercions disappear and the code output will not contain unsafe coercions.
-
-\section{Related Work}\label{sec:related-work}
-
-\subsubsection*{Two-level calculi.}
-Two-level lambda calculi were first developed in the context of abstract
-interpretation and partial evaluation
-\cite{DBLP:phd/ethos/Nielson84a,DBLP:books/daglib/0071307}.  This line of
-research is characterized by simple types, having the same language features at
-different stages, and an emphasis on \emph{binding time analysis},
-i.e.\ automatically inferring stage annotations as part of a pipeline for
-partial evaluation and program optimization.
-
-Later and independently, the same notion of level appeared in homotopy type
-theory, first in Voevodsky's Homotopy Type System \cite{hts}, and subsequently
-developed as 2LTT in \cite{twolevel}. Here, dependent types are essential, and
-we have different theories at the two stages. The application of 2LTT to staged
-compilation was developed in \cite{staged2ltt}. Binding-time analysis has not
-been developed in this setting; 2LTT-s have been meant to be used as surface
-languages to directly work in.
-
-\subsubsection*{Higher-order abstract syntax and logical frameworks.}
-Hofmann's work on the semantics of higher-order abstract syntax (HOAS) is an
-important precursor to our work \cite{Hofmann:1999:SAH:788021.788940}. It
-anticipates many of the later developments in logical frameworks and two-level
-type theories. In particular, we build on Hofmann's presheaf model for our 2LTT
-semantics, and his sketch of adequacy for HOAS corresponds to our definition of
-unstaging and its soundness proof. In general, there is a close correspondence
-between logical frameworks that represent object languages in HOAS style
-(e.g.\ \cite{Harper93lf}) and two-level type theories, and in some cases the two
-kinds of presentations are inter-derivable \cite[Section
-  3.3]{DBLP:journals/corr/abs-2302-08837}. In fact, our Agda implementation
-of CFTT also uses a HOAS embedding of the object theory.
-
-Cocon is a logical framework that has a dependently typed meta level
-(``computation'' level) and an object language that is configured by a signature
-\cite{DBLP:journals/corr/abs-1901-03378}. Cocon, like 2LTTs, can be used as a
-dependently typed metaprogramming language, but there are several
-differences. First, Cocon supports and emphasizes intensional analysis
-(i.e.\ the ability to do structural induction on object code), which is made
-possible through a contextual modality \cite{nanevski2008contextual}. In
-contrast, intensional analysis is missing from 2LTTs so far. However, in Cocon,
-we can only manipulate object terms at the meta level by specifying their
-concrete object-level contexts (i.e.\ possible free variables), while in 2LTTs
-the object-level contexts are implicit and are only computed during code
-generation. The latter style is more convenient for staged compilation
-applications, where explicit context tracking would add a significant amount of
-bureaucracy.
-
-Nonetheless, it seems to be possible and useful to have a
-system which simultaneously supports intensional analysis through contextual
-types and the 2LTT-style implicit contexts, and it would be interesting
-to consider in future work.
-
-\subsubsection*{CPS and monads in staged programming} Writing code generators
-in continuation-passing style was first proposed by Bondorf
-\cite{DBLP:conf/lfp/Bondorf92}. Flanagan et al.'s ANF translation algorithm uses
-CPS for let-insertion as well \cite{DBLP:conf/pldi/FlanaganSDF93}, similarly to
-our $\gen$ function. These sources do no use explicit monadic notation
-though. Jones et al.\ discuss CPS and binding-time improvement in partial
-evaluation in \cite{partial-evaluation}. In
-\cite{DBLP:conf/emsoft/KiselyovST04}, \cite{DBLP:conf/pepm/SwadiTKP06} and
-\cite{DBLP:journals/scp/CaretteK11}, the composition of state and continuation
-monads was used, similarly to our $\StateT\,S\,\Gen$, but without polymorphic
-answer types. The definition of let-insertion here is again similar to ours.
-
-\subsubsection*{Tracking function arities and closures in types} Downen et al.'s
-intermediate language $\mathcal{IL}$ has similar motivations as our object
-language \cite{DBLP:journals/pacmpl/DownenAJE20}. In both systems, function
-types are distinguished from closure types, enjoy universal $\eta$-conversion
-and have an explicit calling arity. In fact, our object language can be almost
-viewed as a small simply-typed fragment of $\mathcal{IL}$, with only
-$\mbf{letrec}$ missing from $\mathcal{IL}$.
-
-\subsubsection*{Sums-of-products} $\SOP$ was proposed for generic Haskell programming
-by De Vries and Löh \cite{DBLP:conf/icfp/VriesL14}. Our own $\SOP$
-implementation in Haskell is mostly the same as in ibid. The
-$\SOP$-of-object-code representation appeared as well in typed Template Haskell
-in \cite{DBLP:conf/haskell/PickeringLW20}. $\eta$-expansion for object-level
-finite sums is known as ``the trick'' \cite{eta-expansion-trick}.
-
-\subsubsection*{Join points}
-The use join points in GHC's core language is partly motivated by avoiding code
-duplication in case-of-case transformations
-\cite{DBLP:conf/pldi/MaurerDAJ17}. In our monad library, case-of-case is
-implicitly and eagerly computed during staging, and we similarly use join points
-to avoid code duplication.
-
-\subsubsection*{Stream fusion}
-The staged stream fusion library $\mathit{strymonas}$ by Kiselyov et al.\ is the
-primary prior art \cite{DBLP:conf/popl/KiselyovBPS17}. Here, streams are
-represented as a sum type of push and pull representations, allowing both
-zipping and $\concatMap$-ing. However, fusion is not guaranteed for all
-combinations; zipping two $\concatMap$-s reifies one of the streams in a runtime
-closure. In newer versions of the library, fusion is complete
-\cite{DBLP:conf/pepm/0001K24}, however, an exposition of this has not yet been
-published. Additionally, $\mathit{strymonas}$ heavily relies on mutable
-references in the object language, and also uses some switching on object-level
-data to implement control flow. This style of code output can be still reliably
-optimized by downstream compilers (C or OCaml), but our solution is more
-conservative in that it does not use any mutation or runtime switching.
-This ``purity'' of our streams also makes them easier to generalize, e.g.\
-by using arbitrary $\MonadGen$ monads instead of $\Gen$ in stream internals.
-
-\emph{Machine fusion} \cite{DBLP:conf/ppdp/RobinsonL17} supports splitting
-streams to multiple streams while avoiding duplicated computation; this is not
-possible in $\mathit{strymonas}$ or our library. Machine fusion also supports
-$\concatMap$ and zipping. However, its state machine representation is
-significantly more complicated than ours, and its fusion algorithm is not
-guaranteed to succeed on arbitrary stream programs.
-
-Coutts developed pull stream fusion in depth in
-\cite{DBLP:phd/ethos/Coutts11}. We borrowed the basic design and the basic
-stream combinator definitions from there. This account is also close to existing
-stream implementations in Haskell. We find it interesting that our staged
-definitions are very close to the non-staged versions there; essentially all of
-the additional complexity is compartmentalized in our $\SOP$ module.  Coutts
-characterized fusibility in a non-staged setting, in terms of ``good consumers''
-and ``good producers'', but this scheme does not cover $\concatMap$.
-
-\section{Conclusions and Future Work}
-
-We believe that a programming language in the style of this paper would be
-highly useful, especially in high-performance functional programming and
-domain-specific programming. In particular, the pipeline for using and compiling
-monads could look like the following.
-\begin{itemize}
-\item Users write definitions in a style similar to Haskell, without quotes,
-      splices, $\mup$-s and $\mdown$-s, only marking stages in type annotations and
-      in let-definitions, with $:=$ and $=$.  Storing monadic actions at runtime
-      requires an explicit boxing operation that yields a closure type.
-\item Type- and stage-directed elaboration adds the missing operations, and also desugars
-      case splitting using $\join$ and the underlying splitting function.
-\item In the downstream compiler, fast \& conservative optimization passes are
-      sufficient, since abstraction overheads are already eliminated by staging. Eliminating
-      dead code and unused arguments would be important, as we have not yet addressed
-      this through staging.
-\item A systematic way to de-duplicate code would be also needed, probably already
-      during staging.
-\end{itemize}
-
-Going a bit further, we believe that a 2LTT-based language could be a good design point
-for programming in general, buying us plenty of control over code generation for
-a modest amount of extra complexity.
-
-Also, rebuilding known abstractions in staged programming is valuable because it
-provides a \emph{semantic explanation} of how abstractions can be compiled, and
-provides insights that could be reused even in non-staged settings. For
-instance, in this paper we demonstrated that compiling monadic code can be done
-by using the same monads in the metalanguage, extended with code
-generation as an effect.
-
-Continuing this line of thought, it could be interesting to also adapt to 2LTT
-the style of \emph{binding-time analysis} that is well-known in partial
-evaluation. For example, we might do program optimization in the following way.
-First, start with monadic code in a non-staged language. Second, try to infer
-stages, inventing quotes, splices, $\mup$-s and $\mdown$-s, thereby translating
-definitions to a 2LTT. Third, perform staging and proceed from there. This would
-be more fragile than unambiguous stage inference, but we would still benefit from
-shifting a lot of machinery into staging (which is deterministic and efficient).
-
-We also find it interesting how little impact closure-freedom had on the
-developments in this paper. This provides some evidence that in staged
-functional programming, closures can be an opt-in feature instead of the
-pervasive default.
-
-In this paper we only briefly looked at two applications. Both monad
-transformers and streams could be further developed, and many other programming
-abstractions could be revisited in the staged setting.
-\begin{itemize}
-\item
-     We did not discuss the issue of \emph{tail calls} in monadic code. For
-     example, a tail call in $\Maybe$ should not case split on the result, while
-     in our current library, making a call with $\mup$ always does so. We did
-     develop an abstraction for tail calls in the Agda and Haskell
-     implementations, but omitted it here partly for lack of space, and partly
-     because we have not yet explored much of the design space.
-
-\item In a practical stream library, it would be useful to further try to
-     minimize the size of the machine state. In the Agda and Haskell versions, we
-     additionally track whether streams can $\Skip$, to provide more efficient
-     zipping for non-skipping ``synchronous'' streams. However, this could be
-     refined in many ways, and we could also try to represent the transition
-     graph in an observable way, thereby enabling merging or deleting states.
-
-\item It seems promising to look for synergies between push streams, pull streams
-     and monad transformers. It seems that pull streams could be generalized from
-     $\Gen$ to different monads, in the representation of transitions and initial
-     states. This would allow putting a $\Pull$ \emph{on the top} of a monad
-     transformer stack. On the other hand, push streams form a proper monad, but
-     they cannot be used to ``transform'' other monads, so they can be placed
-     at the \emph{bottom} of a transformer stack.
-\item We could try to lean more heavily on datatype-generic programming and
-     generalize abstractions for more object types. For example, push and pull
-     streams could be generalized to inductive and coinductive Church-codings of
-     arbitrary value types.
-\end{itemize}
-
-\section*{Acknowledgements}
-I thank the anonymous reviewers, Sebastian Graf and Jacques Carette, for their
-comments and suggestions on draft versions of this paper.
-
-\section*{Data Availability Statement}
-The Agda and Haskell artifacts are available in \cite{cftt-artifacts}.
-
-%% --------------------------------------------------------------------------------
-%% --------------------------------------------------------------------------------
-
-\appendices
-\section{Staged Semantics for CFTT}\label{appendix:A}
-
-In this appendix we describe the semantics of CFTT, from which the unstaging
-algorithm is obtained, and also describe the soundness and stability of the
-algorithm. We borrow the setup from \cite{staged2ltt}; we use terminology and
-definitions from there by default, and focus on differences and extensions.
-
-\begin{notation} In the following, we work in a mostly unspecified extensional
-type theory, writing $\Set$ for universes (omitting universe levels), $(x : A) \to B$
-for $\Pi$-types, $(x : A) \times B$ for $\Sigma$-types and $\blank\!=\!\blank$ for
-extensional identity.
-
-\end{notation}
-
-\subsection{Models and Syntax of CFTT}
-
-We give an algebraic \cite{cartmellthesis,DBLP:journals/corr/abs-2302-08837}
-specification for the models of CFTT and define the syntax of CFTT to be the
-initial model. In the following, all type and term formers are implicitly
-assumed to be stable under substitution.
-
-\begin{definition}[] A model of CFTT consists of the following.
-
-\begin{itemize}
-\item A category with a terminal object; in the syntax this is the category
-  of contexts and parallel substitutions. We use $\Con : \Set$ for contexts,
-  $\Sub : \Con \to \Con \to \Set$ for morphisms and $\emptycon : \Con$ for the
-  terminal object (which is the empty context in the syntax).
-\item A family structure in the sense of category-with-families \cite{Dybjer96internaltype},
-  specifying metatypes and terms. This structure is additionally indexed by
-  universe levels, so we have $\MTy_i : \Con \to \Set$ and $\msf{MetaTm}_i :
-  (\Gamma : \Con) \to \MTy_i\,\Gamma \to \Set$ for $i \in \mbb{N}$. We support
-  basic type formers, identity types and W-types. We write $\Gamma \ext A$ for
-  an extended context. Additionally, we have Coquand-style universes,
-  i.e.\ $\U_i : \MTy_{i+1}\,\Gamma$ such that there is an isomorphism
-  $\msf{MetaTm}\,\Gamma\,\U_i \simeq \MTy_i\,\Gamma$.  We use $\El$ and
-  $\code$ to denote the components of this isomorphism.
-\item
-  For object-level types, we have $\Ty : \MTy_0\,\Gamma$, $\VTy :
-  \MTy_0\,\Gamma$ and $\CTy : \MTy_0\,\Gamma$, together with all object-level
-  type formers and inclusion operations $\msf{V} : \MTm\,\Gamma\,\VTy \to \MTm\,\Gamma\,\Ty$
-  and $\msf{C} : \MTm\,\Gamma\,\CTy \to \MTm\,\Gamma\,\Ty$.
-\item For object-level terms, we have $\Tm : \MTm\,\Gamma\,\Ty \to \Set$,
-  supporting context extension, i.e.\ we have $\Gamma\ext A$ for
-  $A : \MTm\,\Gamma\,\Ty$ such that $\Sub\,\Gamma\,(\Delta\ext A)$ is isomorphic
-  to $(\sigma : \Sub\,\Gamma\,\Delta) \times \Tm\,\Gamma\,(A[\sigma])$. All
-  object-level term formers are supported in $\Tm$.
-\item Staging operations: we have $\Lift : \MTm\,\Gamma\,\Ty \to \MTy\,\Gamma$ together
-  with isomorphisms $\Tm\,\Gamma\,A \simeq \MTm\,\Gamma\,(\Lift A)$.
-  We write $\ql\!\blank\!\qr$ and $\spl\!\blank$ for the components of the isomorphism.
-\end{itemize}
-\end{definition}
-This presentation is a bit more verbose than the surface syntax used in the
-paper. First, the meta level uses explicit coding and decoding for universes,
-while the surface syntax assumes Russell-style universes. Second, the object
-level has $\VTy$ and $\CTy$ as Tarski-style universes with explicit inclusions
-$\msf{V}$ and $\msf{C}$ into $\Ty$, while the surface syntax used implicit
-coercions.
-
-Also, this is not the only possible presentation. Here, we follow the style of
-the informal syntax by having a separate \emph{sort} for object-level terms,
-together with explicit quoting and splicing. However, we do not have a sort for
-object types and instead specify them as terms of metatypes. This is a mixture
-of two styles:
-\begin{itemize}
-\item The ``classic'' 2LTT style would have separate sorts and staging
-  operations for all of $\Ty$, $\CTy$ and $\VTy$. Here the two levels are
-  specified independently and level-mixing is only possible through staging
-  operations.
-\item The ``higher-order abstract syntax'' (HOAS) or ``logical framework'' style
-  would only have metatypes and metaterms, no quoting or splicing, and would
-  reuse metafunctions to represent object-level binders. Our Agda implementation
-  follows this style.
-\end{itemize}
-In the context of CFTT, the choice between HOAS and 2LTT styles is a matter of
-taste, and the two are inter-derivable \cite[Section
-  3.3]{DBLP:journals/corr/abs-2302-08837}.
-
-\subsection{Presheaf Model}
-
-Unstaging is defined as evaluation of the syntax of CFTT in a particular
-presheaf model.
-\begin{itemize}
-  \item The \emph{object theory} is the simple type theory supporting only value
-    types and computation types as described in Section
-    \ref{sec:the-object-level} and also in the Agda
-    formalization. Object-theoretic terms are not quotiented by
-    $\beta\eta$-conversion. Substitution is a recursively defined operation on
-    terms.
-  \item $\mbbo$ is the category whose objects are contexts of the object theory,
-    and morphisms are parallel substitutions, as lists of terms. With this,
-    object-theoretic types form a constant presheaf which we denote
-    as $\Tyo : \mbbo^{\msf{op}} \to \Set$, and terms form a dependent presheaf over $\Tyo$ which we write as
-    $\Tmo$.
-  \item
-  In the presheaf model of CFTT, every typing context (and also every closed
-  type) is modeled as a presheaf over $\mbbo$. The interpretation of meta-level
-  type formers is standard, following \cite{Hofmann97syntaxand} and \cite{huber-thesis}.
-\item $\Ty : \MTy_0\,\Gamma$ is the displayed presheaf which is constantly
-  $\Tyo$. $\VTy$ and $\CTy$ are modeled analogously.
-\item $\Lift A : \MTy_0\,\Gamma$ is modeled as follows. We have that $A :
-  \MTy\,\Gamma\,\Ty$, but since $\Ty$ is defined as constantly $\Tyo$, it
-  follows that $A$ is a natural transformation from $\Gamma$ to $\Tyo$. Hence,
-  we define $\Lift A : \MTy_0\,\Gamma$ to be $\Tmo[A]$, where $[\blank]$ is
-  substitution in presheaves, i.e.\ the reindexing of a displayed presheaf by a
-  natural transformation.
-\item
-  $\Tm\,\Gamma\,A$ is simply defined to be $\MTy\,\Gamma\,(\Lift A)$ in the model. Hence,
-  quotation and splicing are both interpreted as identity maps.
-\end{itemize}
-Now, we can take a closed term $t : \Tm\,\emptycon\,A$ in CFTT and evaluate it
-in the presheaf model to obtain an object-theoretic term whose type is the
-result of evaluating $A$.
-
-\subsection{W-types}
-So far we have stayed close to \cite{staged2ltt}, with only cosmetic
-changes. Supporting W-types, as a way to get inductive families \cite{whynotw},
-is the biggest extension here. We extend Section 5 of \cite{staged2ltt}. The
-formal specification of syntactic $\vW{-}\msf{elim}$ is rather gnarly, because
-of the usage of explicit substitutions, but we write it out here anyways.
-
-\begin{notation} We use $\p^N$ for $N$-fold
-weakening and we also use numerals as De Bruijn indices, to abbreviate
-$\q[\p^N]$. We write $\msf{app}$ and $\msf{lam}$ for $\Pi$ application and
-abstraction. We write $\blank{\Rightarrow}\blank$ for non-dependent function
-types. Finally, for the sake of brevity we omit the component maps of the isomorphism
-$\re_\ext : \re\,(\Gamma \ext A) \simeq ((\gamma : \re\,\Gamma) \times \re\,A\,\gamma)$.
-\end{notation}
-
-In the syntax of CFTT, $\vW$-types are as follows.
-\begin{alignat*}{3}
-  &\vW   &&: (A : \MTy_i\,\Gamma) \to \MTy_j\,(\Gamma\,\ext\,A) \to \MTy_{\msf{max}(i,j)}\,\Gamma\\
-  &\vsup &&: (a : \MTm\,\Gamma\,A) \to \MTm\,\Gamma\,(B[\vid,\,a] \Rightarrow \vW\,A\,B) \to \MTm\,\Gamma\,(\vW\,A\,B)\\
-  &\vW{-}\msf{elim}  &&: (P : \MTy\,(\Gamma\,\ext\,\vW\,A\,B))\\
-  &                  && \hspace{0.6em}\to \MTm\,(\Gamma\,\ext\,A\,\ext\,B \Rightarrow (\vW\,A\,B)[p]\,\ext\,\Pi\,(B[p])\,(P[\p^3,\,\msf{app}\,1\,0]))\,(P[\p^3,\,\vsup\,2\,1))\\
-  &                  && \hspace{0.6em}\to (w : \MTm\,\Gamma\,(\vW\,A\,B)) \to \MTm\,\Gamma\,(P[\id,\,w])\\
-  & \vW{\beta}       &&:
-    \vW{-}\msf{elim}\,P\,s\,(\vsup\,a\,f) = s[\id,\,a,\,f,\,\msf{lam}\,(\vW{-}\msf{elim}\,(P[\p^2,\,\q])\,(s[\p])\,(\msf{app}\,f\,0)]
-\end{alignat*}
-We shall omit the $i$ and $j$ universe levels in the following. In $\whset$,
-i.e.\ internally to presheaves over $\mbbo$, we also have the standard W-type.
-\begin{alignat*}{3}
-  & \wh{\vW}   &&: (A : \whset) \to (A \to \whset) \to \whset\\
-  & \wh{\vsup} &&: (a : A) \to (B\,a \to \wh{\vW}\,A\,B) \to \wh{\vW}\,A\,B\\
-  & \wh{\vW{-}\msf{elim}} &&: (P : \wh{\vW}\,A\,B \to \whset)\\
-  & && \hspace{0.6em}\to ((a : A)(f : B\,a \to \wh{\vW}\,A\,B) \to ((b : B\,a) \to P\,(f\,b)) \to P\,(\wh{\vsup}\,a\,f))\\
-  & && \hspace{0.6em}\to (w : \wh{\vW}\,A\,B) \to P\,w\\
-  & \wh{\vW}{\beta} &&: \wh{\vW{-}\msf{elim}}\,P\,s\,(\wh{\vsup}\,a\,f) = s\,a\,f\,(\lambda\,b.\,\wh{\vW{-}\msf{elim}}\,P\,s\,(f\,b))
-\end{alignat*}
-In the presheaf model we interpret $\vW$ using $\whW$.
-\begin{alignat*}{3}
-  & \ev\,(\vW\,A\,B)\,\gamma   &&:= \wh{\vW}\,(\ev\,A\,\gamma)\,(\lambda\,\alpha.\,\ev\,B\,(\gamma,\,\alpha))\\
-  & \ev\,(\vsup\,a\,f)\,\gamma &&:= \wh{\vsup}\,(\ev\,a\,\gamma)\,(\lambda\,\beta.\,\ev\,f\,(\gamma, \beta))\\
-  & \ev\,(\vW{-}\msf{elim}\,P\,s\,w)\,\gamma &&:= \wh{\vW{-}\msf{elim}}\,(\lambda\,w.\,\ev\,P\,(\gamma,\,w))\,(\lambda\,a\,f\,g.\,\ev\,s\,(\gamma,\,a,\,f,\,g))\,(\ev\,w\,\gamma)
-\end{alignat*}
-
-%% W-elim̂ (λ w. E P (γ, w)) (λ a f g. E s (γ, a, f, g)) (E w γ)
-Now, assuming $A : \MTy\,\Gamma$ and $B : \MTy\,(\Gamma\,\ext\,A)$, we consider
-$\re\,(\vW\,A\,B) : (\gamma : \re\,\Gamma) \to \whset$. Externally, the elements
-of $\re\,(\vW\,A\,B)\,\gamma$ are CFTT terms with a $\vW$-type in purely
-object-level contexts. We call such terms \emph{restricted}. Now, we internalize
-the restriction of the $\vsup$ constructor:
-\[ \re_{\vsup} : (\alpha : \re\,A\,\gamma) \to \re\,(B \Rightarrow (W\,A\,B)[\p])\,(\gamma,\,\alpha) \to \re\,(W\,A\,B)\,\gamma \]
-Externally, $\re_{\vsup}$ applies $\vsup$ to restricted terms. Note that in
-$\re\,(B \Rightarrow (W\,A\,B)[\p])\,(\gamma,\,\alpha)$, the type $B \Rightarrow (W\,A\,B)[\p]$ is in
-the context $(\Gamma\,\ext\,A)$, which makes it possible to pass $(\gamma,\,\alpha)$ as
-argument; this ``instantiates'' $B$ to $\alpha$. With this, we can also
-show that $\re$ preserves $\vsup$:
-\[ \re\,(\vsup\,t\,u)\,\gamma = \re_\vsup\,(\re\,t\,\gamma)\,(\re\,u\,\gamma) \]
-Note that the type of $\re\,u\,\gamma$ is $\re\,(B[\id,\,t] \Rightarrow
-\vW\,A\,B)\,\gamma$, which is equal to
-\[\re\,((B \Rightarrow(\vW\,A\,B)[\p])[\id,\,t])\,\gamma\]
-which is in turn equal to $\re\,(B \Rightarrow
-(\vW\,A\,B)[\p])\,(\gamma,\,\re\,t\,\gamma)$ by $\re$'s preservation of
-substitutions, which makes the above equation well-typed.
-
-Next, we give a logical relation interpretation for $\vW$. We inductively define
-a relation in $\whset$ as follows.
-\begin{alignat*}{3}
-  & \vW\rel &&: \Gamma\rel\,\gamma\,\gamma' \to \ev\,(\vW\,A\,B)\,\gamma \to \re\,(\vW\,A\,B)\,\gamma' \to \whset\\
-  & \vsup\rel &&: \hspace{0.8em}\{\alpha : \ev\,A\,\gamma\}\{\alpha' : \re\,A\,\gamma'\}(\alpha\rel : A\rel \gamma\rel\,\alpha\,\alpha')\\
-  &           &&\to \{f \hspace{0.5em}: \ev\,B\,(\gamma,\,\alpha) \to \ev\,(\vW\,A\,B)\,\gamma\}\\
-  &           &&\to \{f' \hspace{0.2em}: \re\,(B \Rightarrow (\vW\,A\,B)[\p])\,(\gamma',\,\alpha')\}\\
-  &           &&\to (f\rel : (B \Rightarrow (\vW\,A\,B)[\p])\rel\,(\gamma\rel,\,\alpha\rel)\,f\,f')\\
-  &           &&\to \vW\rel\,\gamma\rel\,(\wh{\vsup}\,\alpha\,f)\,(\re_\vsup\,\alpha'\,f')
-\end{alignat*}
-Here, $(B \Rightarrow (\vW\,A\,B)[\p])\rel$ is the relational interpretation of
-the function type, which unfolds to pointwise preservation of relations. We also
-define $(\vW\,A\,B)\rel\,\gamma\rel\,t\,t'$ to be $\vW\rel\,\gamma\rel\,t\,t'$.
-In short, $\vW\rel$ expresses that $\wh{\vsup}\,\alpha\,f$ is related to
-$\re_\vsup\,\alpha'\,f'$ when the arguments are (inductively) related. Similarly, we
-interpret $(\vsup\,t\,u)\rel\,\gamma\rel$ as
-$\vsup\rel\,(t\rel\,\gamma\rel)\,(u\rel\,\gamma\rel)$. It only remains to
-interpret $\vW$-elimination and its $\beta$-rule. We need to define
-\[ (\vW{-}\msf{elim}\,P\,s\,w)\rel\,\gamma\rel : P\rel\,(\gamma\rel,\,w\rel\,\gamma\rel)\,(\ev\,(\welim\,P\,s\,w)\,\gamma)\,(\re\,(\welim\,P\,s\,w)\,\gamma'). \]
-We do this by induction on $w\rel\,\gamma\rel :
-\vW\rel\,\gamma\rel\,(\ev\,w\,\gamma)\,(\re\,w\,\gamma')$. At this point it
-becomes rather tedious to compute and unfold the involved types, so we only give
-a compact definition together with some informal explanation:
-\begin{alignat*}{3}
-  &(\vW{-}\msf{elim}\,P\,s\,w)\rel\,\gamma\rel :=\\
-  &\hspace{4em}\vW\rel{-}\msf{elim}\,(\lambda\,w\rel.\,\,P\rel\,(\gamma\rel,\,w\rel)\,(\ev\,(\welim\,P\,s\,w)\,\gamma)\,(\re\,(\welim\,P\,s\,w)\,\gamma'))\\
-  &\hspace{8.1em}(\lambda\,a\rel\,f\rel\,\mathit{hyp}.\,\,s\rel\,(\gamma\rel,\,a\rel,\,f\rel,\,\mathit{hyp}))\\
-  &\hspace{8.1em}(w\rel\,\gamma\rel)
-\end{alignat*}
-$\vW\rel{-}\msf{elim}$ is the eliminator for $\vW\rel$. Its first argument is
-the induction motive, which matches the goal type. The second argument is the
-method for $\vsup\rel$; its $a\rel$ and $f\rel$ inputs are the relatedness
-witnesses stored in a $\vsup\rel$, while $\mathit{hyp}$ is induction hypothesis,
-witnessing the inductive motive for every output of $f\rel$. In the body here,
-we need to witness the motive in the case where $w\rel\,\gamma\rel$ is a $\vsup\rel$, which causes
-both $\ev\,(\welim\,P\,s\,w)\,\gamma$ and $\re\,(\welim\,P\,s\,w)\,\gamma'$ to
-compute further to applications of $\ev\,s\,\gamma$ and $\re\,s\,\gamma'$
-respectively. Hence, we use $s\rel$ to relate the two sides. This definition
-also respects $\vW\beta$, by the $\beta$ rule for $\vW\rel$.
-
-\subsection{Identity type}
-
-In this section we assume an \emph{extensional} identity type in CFTT. As we
-mentioned in Section \ref{sec:semantics-of-staging}, we can show soundness of staging with
-extensional identity and certain \emph{postulated} identities, but we cannot do
-the same with intensional identity. Hence, we formulate unstaging as consisting
-of two steps: first, a translation of intensional identity to extensional (which
-amounts to proof erasure), followed by evaluation in the presheaf model.
-
-Extensional identity in CFTT type is interpreted as the standard extensional
-identity in presheaves, i.e.\ we have $\ev\,(\msf{Id}\,A\,t\,u)\,\gamma :=
-\ev\,t\,\gamma = \ev\,u\,\gamma$. This supports $\msf{refl}$, transport,
-uniqueness of identity and equality reflection; we omit the fairly trivial
-definitions here. We choose the relational interpretation to be trivial as well,
-i.e.\ $(\msf{Id}\,t\,u)\rel\,\gamma\rel\,p\,p' := \top$.
-
-This is because both $\ev\,(\Id\,t\,u)\,\gamma$ and $\re\,(\Id\,t\,u)\,\gamma'$
-are mere propositions --- the latter because of equality reflection in the
-syntax.  In particular, if we have a witness of $\re\,(\Id\,t\,u)\,\gamma'$, we
-also have $\re\,t\,\gamma' = \re\,u\,\gamma'$ by equality reflection. Thus, the
-rest of the relational interpretation of $\Id$ is also trivial. Another way to
-think about this: soundness expresses a form of canonicity, but if we have
-equality reflection then terms with identity types are always canonical.
-
-\subsection{Generativity} \label{appendix:generativity}
-
-The generativity axiom from Section \ref{sec:concatmap-for-streams} has the following type, in internal
-CFTT syntax:
-\[ (A : \U_\msf{P})(f : \Elprod\,A \to \USOP)(x\,y : \Elprod\,A) \to f\,x = f\,y  \]
-In the presheaf model, we have the following:
-\begin{itemize}
-\item $\wh{\Uprod}$ is the set (constant presheaf) of lists of object-theoretic types.
-\item $\wh{\USOP}$ is the set of lists of lists of object-theoretic types.
-\item $\wh{\Elprod\,A}$ is the presheaf such that $\wh{\Elprod\,A}\,\Gamma$ is
-  the set of products of object-theoretic terms in context $\Gamma$, with types
-  of components taken from $A$.
-\end{itemize}
-Unfolding the type of generativity in the presheaf model, it suffices to show
-that assuming
-\begin{itemize}
-\item $A : \wh{\Uprod}$, $\Gamma : \Cono$, $x,\,y : \wh{\Elprod\,A}\,\Gamma$
-\item and $f : (\Delta : \Cono) \to \Subo\,\Delta\,\Gamma \to \wh{\Elprod\,A}\,\Delta \to \wh{\USOP}$ such that $f$ is natural with respect to object-theoretic substitution,
-\end{itemize}
-we have that $f\,\Gamma\,\id\,x = f\,\Gamma\,\id\,y$.
-
-Now, $\wh{\Elprod\,A}$ is a representable presheaf since it is a finite product
-of object-theoretic terms. Also, since $f$ is a natural transformation from the
-representable presheaf $\Subo\blank\Gamma \times \wh{\Elprod\,A}$ to the
-constant presheaf $\wh{\USOP}$, the Yoneda lemma implies that $f$ is constant and
-hence the generativity axiom is validated.
-
-Finally, since the logical relation interpretation of the identity type is trivial,
-the logical relation interpretation for the generativity axiom is also trivial.
-
-\subsection{Summary}
-
-By the presheaf and logical relations interpretation of CFTT as shown in this appendix,
-we obtain \emph{unstaging} and \emph{soundness} of unstaging, following \cite{staged2ltt}.
-
-\emph{Strictness} of unstaging holds trivially, since there are no $\beta$ and
-$\eta$ rules in the object theory. \emph{Stability} of unstaging is shown by
-straightforward induction on the syntax of the object theory, the same way as in
-\cite{staged2ltt}, and we omit it here.
-
-\interlinepenalty=10000
+\bibliographystyle{plain}
 \bibliography{references}
@@
 \end{document}
-\endinput

### 2024-09-02 icfp24prez/pres/Notes.hs
icfp24prez wip

@@ -8,3 +8,6 @@ f = do
   b <- ask
   if b then return 10
        else return 20
+
+
+data Mallac

### 2024-09-02 icfp24prez/pres/Notes.hs
icfp24prez wip

@@ -9,5 +9,8 @@ f = do
   if b then return 10
        else return 20
 
-
 data Mallac
+
+data Foo = Foo {mallac :: Int}
+
+newtype

### 2024-09-08 opt/SimpleANF2.hs
stuff

@@ -1,0 +1,291 @@
+{-# language Strict, LambdaCase, BlockArguments, ImplicitParams, ConstraintKinds #-}
+{-# options_ghc -Wincomplete-patterns #-}
+
+module SimpleANF2 where
+
+import Control.Monad.State.Strict
+import Data.Foldable
+
+{-
+If the continuation k has no Case,
+push the Case and eval the scrutinee
+
+   eval (case t u v) k =
+     eval t (Case * u v ; k)
+
+If the continuation has a Case, split the it off and reify it.
+
+   eval (case t u v) (k1 ; Case * u' v' ; k2) =
+     let k x = ACase x (eval u' k2) (eval v' k2)
+     eval t (Case * u v ; k1 ; k *)
+
+We have at most one Case in the continuation, somewhere in the middle.
+
+Continuations can be of the following form:
+
+   - apps
+   - apps, Case, apps
+   - apps, Case, apps ; k *
+
+
+
+
+-}
+
+--------------------------------------------------------------------------------
+
+type Ix = Int
+type Lvl = Int
+type Arity = Int
+
+data Tm
+  = Var Ix
+  | LetV Tm Tm
+  | LetC Arity Tm Tm
+  | In1 Tm
+  | In2 Tm
+  | Case Tm Tm Tm
+  | Lam Tm
+  | App Tm Tm
+  deriving Show
+
+--------------------------------------------------------------------------------
+
+data Comp = CLam Comp | CBody ANF
+  deriving Show
+
+data Let = LComp Comp | LApp Lvl [Lvl] | LIn1 Lvl | LIn2 Lvl
+  deriving Show
+
+data ANF = ARet Lvl | ACase Lvl ANF ANF | ALet Let ANF
+  deriving Show
+
+--------------------------------------------------------------------------------
+
+data Cont
+  = Apps      [Lvl]
+  | AppsCase  [Lvl] ANF ANF [Lvl]
+  | AppsCaseK [Lvl] ANF ANF [Lvl] Lvl
+  deriving Show
+
+type M = State (Lvl, [Let])
+
+addLet :: Let -> M Lvl
+addLet lt = do
+  (l, lets) <- get
+  put (l+1, lt:lets)
+  pure l
+
+locally :: M a -> M a
+locally act = do
+  (fresh, lets) <- get
+  put (fresh, [])
+  a <- act
+  put (fresh, lets)
+  pure a
+
[…]

### 2024-09-08 runtime_codegen/proto2/Common.hs

@@ -1,0 +1,74 @@
+{-# options_ghc -Wno-unused-imports #-}
+
+module Common (
+    module Common
+  , SourcePos(..)
+  , Pos
+  , unPos
+  , module Data.Coerce
+  , initialPos) where
+
+import Data.Coerce
+import Data.Kind
+import Data.List
+import Debug.Trace
+import GHC.Stack
+import Text.Megaparsec
+import Text.Printf
+
+--------------------------------------------------------------------------------
+-- We use a more custom debugging function here. We can comment out one
+-- definition to toggle debug printing.
+
+-- debug :: (Applicative f) => [String] -> f ()
+-- debug strs = traceM (intercalate " | " strs ++ " END")
+
+-- type Dbg :: Constraint
+-- type Dbg = HasCallStack
+
+debug :: (Applicative f) => [String] -> f ()
+debug strs = pure ()
+
+type Dbg :: Constraint
+type Dbg = ()
+
+--------------------------------------------------------------------------------
+
+
+
+impossible :: Dbg => a
+impossible = error "impossible"
+
+type Name = String
+
+data Icit = Impl | Expl deriving (Eq)
+
+icit :: Icit -> a -> a -> a
+icit Impl a _ = a
+icit Expl _ b = b
+
+instance Show Icit where
+  show Impl = "implicit"
+  show Expl = "explicit"
+
+-- | De Bruijn index.
+newtype Ix  = Ix {unIx :: Int} deriving (Eq, Show, Num) via Int
+
+-- | De Bruijn level.
+newtype Lvl = Lvl {unLvl :: Int} deriving (Eq, Ord, Show, Num) via Int
+
+-- | Metavariable.
+newtype MetaVar = MetaVar {unMetaVar :: Int} deriving (Eq, Show, Num) via Int
+
+-- | Identifier of a delayed checking problem.
+newtype CheckVar = CheckVar {unCheckVar :: Int} deriving (Eq, Show, Num, Ord) via Int
+
+
+-- Snoc lists
+--------------------------------------------------------------------------------
+
+infixl 4 :>
+
+pattern (:>) :: [a] -> a -> [a]
+pattern xs :> x <- x:xs where (:>) xs ~x = x:xs
+{-# complete (:>), [] #-}

### 2024-09-08 runtime_codegen/proto2/Presyntax.hs

@@ -1,0 +1,28 @@
+
+module Presyntax where
+
+import Common
+
+data Tm
+  = Var Name                                  -- x
+  | Lam Name (Either Name Icit) (Maybe Tm) Tm -- \x. t | \{x}. t | \{x = y}. t |
+                                              -- \(x : A). t | \{x : A}. t
+  | App Tm Tm (Either Name Icit)              -- t u  | t {u} | t {x = u}
+  | U                                         -- U
+  | Pi Name Icit Tm Tm                        -- (x : A) -> B | {x : A} -> B
+  | Let Name Tm Tm Tm                         -- let x : A = t; u
+  | SrcPos SourcePos Tm                       -- source position for error reporting
+  | Hole                                      -- _
+  deriving Show
+
+-- | Get rid of source positions, for better debug printing.
+stripPos :: Tm -> Tm
+stripPos = \case
+  Var x        -> Var x
+  Lam x i ma t -> Lam x i (stripPos <$> ma) (stripPos t)
+  App t u i    -> App (stripPos t) (stripPos u) i
+  U            -> U
+  Pi x i a b   -> Pi x i (stripPos a) (stripPos b)
+  Let x a t u  -> Let x (stripPos a) (stripPos t) (stripPos u)
+  SrcPos _ t   -> stripPos t
+  Hole         -> Hole

### 2024-12-13 icfp22paper/paper.tex
just push everything
wip
stuff

--- 2022-09-20 paper/paper.tex
@@
 Assume $\List_0 : \U_0 \to \U_0$ with $\nil_0 : (A : \U_0) \to \List_0\,A$,
-$\cons_0 : (A : \U_0) \to A \to \List_0\,A$ and $\msf{foldr}_0 : (A\,B : \U_0)
+$\cons_0 : (A : \U_0) \to A \to \List_0\,A \to \List_0\,A$ and $\msf{foldr}_0 : (A\,B : \U_0)
 \to (A \to B \to B) \to B \to \List_0\,A \to B$. We define a mapping function which
@@
 modest complications in meta-level type formers because we will sometimes need to use
-the lax/oplax preservations properties of $\re$. For object-level types and
+the lax/oplax preservation properties of $\re$. For object-level types and
 terms, we get soundness statements: evaluation via $\ev$ followed by embedding

### 2025-02-10 yonedaLF/draft/paper.tex
stuff

\documentclass[acmsmall,review,anonymous]{acmart}
%%
%% \BibTeX command to typeset BibTeX logo in the docs
\AtBeginDocument{%
  \providecommand\BibTeX{{%
    Bib\TeX}}}

%% Rights management information.  This information is sent to you
%% when you complete the rights form.  These commands have SAMPLE
%% values in them; it is your responsibility as an author to replace
%% the commands and values with those provided to you when you
%% complete the rights form.
\setcopyright{acmlicensed}
\copyrightyear{2018}
\acmYear{2018}
\acmDOI{XXXXXXX.XXXXXXX}

%%
%% These commands are for a JOURNAL article.
\acmJournal{JACM}
\acmVolume{37}
\acmNumber{4}
\acmArticle{111}
\acmMonth{8}

%%
%% Submission ID.
%% Use this when submitting an article to a sponsored event. You'll
%% receive a unique submission ID from the organizers
%% of the event, and this ID should be used as the parameter to this command.
%%\acmSubmissionID{123-A56-BU3}

%%
%% For managing citations, it is recommended to use bibliography
%% files in BibTeX format.
%%
%% You can then either use BibTeX with the ACM-Reference-Format style,
%% or BibLaTeX with the acmnumeric or acmauthoryear sytles, that include
%% support for advanced citation of software artefact from the
%% biblatex-software package, also separately available on CTAN.
%%
%% Look at the sample-*-biblatex.tex files for templates showcasing
%% the biblatex styles.
%%

%%
%% The majority of ACM publications use numbered citations and
%% references.  The command \citestyle{authoryear} switches to the
%% "author year" style.
%%
%% If you are preparing content for an event
%% sponsored by ACM SIGGRAPH, you must use the "author year" style of
%% citations and references.
%% Uncommenting
%% the next command will enable that style.
%%\citestyle{acmauthoryear}

%%
%% end of the preamble, start of the body of the document source.
\begin{document}

\title{A Multitheory Logical Framework}

\author{András Kovács}
\orcid{0000-0002-6375-9781}
\affiliation{%
  \institution{University of Gothenburg \& Chalmers University of Technology}
  \city{Gothenburg}
  \country{Sweden}
}
\email{andrask@chalmers.se}

%%
%% The abstract is a short summary of the work to be presented in the
%% article.
\begin{abstract}
  A clear and well-documented \LaTeX\ document is presented as an
  article formatted for publication by ACM in a conference proceedings
  or journal publication. Based on the ``acmart'' document class, this
  article presents and explains many of the common variations, as well
  as many of the formatting elements an author may use in the
  preparation of the documentation of their work.
\end{abstract}

%%
%% The code below is generated by the tool at http://dl.acm.org/ccs.cfm.
%% Please copy and paste the code instead of the example below.
%%
\begin{CCSXML}
<ccs2012>
 <concept>
  <concept_id>00000000.0000000.0000000</concept_id>
  <concept_desc>Do Not Use This Code, Generate the Correct Terms for Your Paper</concept_desc>
  <concept_significance>500</concept_significance>
 </concept>
 <concept>
  <concept_id>00000000.00000000.00000000</concept_id>
  <concept_desc>Do Not Use This Code, Generate the Correct Terms for Your Paper</concept_desc>
  <concept_significance>300</concept_significance>
 </concept>
 <concept>
  <concept_id>00000000.00000000.00000000</concept_id>
  <concept_desc>Do Not Use This Code, Generate the Correct Terms for Your Paper</concept_desc>
  <concept_significance>100</concept_significance>
 </concept>
 <concept>
  <concept_id>00000000.00000000.00000000</concept_id>
  <concept_desc>Do Not Use This Code, Generate the Correct Terms for Your Paper</concept_desc>
  <concept_significance>100</concept_significance>
 </concept>
</ccs2012>
\end{CCSXML}

\ccsdesc[500]{Do Not Use This Code~Generate the Correct Terms for Your Paper}
\ccsdesc[300]{Do Not Use This Code~Generate the Correct Terms for Your Paper}
\ccsdesc{Do Not Use This Code~Generate the Correct Terms for Your Paper}
\ccsdesc[100]{Do Not Use This Code~Generate the Correct Terms for Your Paper}

%%
%% Keywords. The author(s) should pick words that accurately describe
%% the work being presented. Separate the keywords with commas.
\keywords{Do, Not, Us, This, Code, Put, the, Correct, Terms, for,
  Your, Paper}

\received{20 February 2007}
\received[revised]{12 March 2009}
\received[accepted]{5 June 2009}

%%
%% This command processes the author and affiliation and title
%% information and builds the first part of the formatted document.
\maketitle

\section{Introduction}
ACM's consolidated article template, introduced in 2017, provides a
consistent \LaTeX\ style for use across ACM publications, and
incorporates accessibility and metadata-extraction functionality
necessary for future Digital Library endeavors. Numerous ACM and
SIG-specific \LaTeX\ templates have been examined, and their unique
features incorporated into this single new template.

If you are new to publishing with ACM, this document is a valuable
guide to the process of preparing your work for publication. If you
have published with ACM before, this document provides insight and
instruction into more recent changes to the article template.

The ``\verb|acmart|'' document class can be used to prepare articles
for any ACM publication --- conference or journal, and for any stage
of publication, from review to final ``camera-ready'' copy, to the
author's own version, with {\itshape very} few changes to the source.

\section{Template Overview}
As noted in the introduction, the ``\verb|acmart|'' document class can
be used to prepare many different kinds of documentation --- a
double-anonymous initial submission of a full-length technical paper, a
two-page SIGGRAPH Emerging Technologies abstract, a ``camera-ready''
journal article, a SIGCHI Extended Abstract, and more --- all by
selecting the appropriate {\itshape template style} and {\itshape
  template parameters}.

This document will explain the major features of the document
class. For further information, the {\itshape \LaTeX\ User's Guide} is
available from
\url{https://www.acm.org/publications/proceedings-template}.

\subsection{Template Styles}

The primary parameter given to the ``\verb|acmart|'' document class is
the {\itshape template style} which corresponds to the kind of publication
or SIG publishing the work. This parameter is enclosed in square
brackets and is a part of the {\verb|documentclass|} command:
\begin{verbatim}
  \documentclass[STYLE]{acmart}
\end{verbatim}

Journals use one of three template styles. All but three ACM journals
use the {\verb|acmsmall|} template style:
\begin{itemize}
\item {\texttt{acmsmall}}: The default journal template style.
\item {\texttt{acmlarge}}: Used by JOCCH and TAP.
\item {\texttt{acmtog}}: Used by TOG.
\end{itemize}

The majority of conference proceedings documentation will use the {\verb|acmconf|} template style.
\begin{itemize}
\item {\texttt{sigconf}}: The default proceedings template style.
\item{\texttt{sigchi}}: Used for SIGCHI conference articles.
\item{\texttt{sigplan}}: Used for SIGPLAN conference articles.
\end{itemize}

\subsection{Template Parameters}

In addition to specifying the {\itshape template style} to be used in
formatting your work, there are a number of {\itshape template parameters}
which modify some part of the applied template style. A complete list
of these parameters can be found in the {\itshape \LaTeX\ User's Guide.}

Frequently-used parameters, or combinations of parameters, include:
\begin{itemize}
\item {\texttt{anonymous,review}}: Suitable for a ``double-anonymous''
  conference submission. Anonymizes the work and includes line
  numbers. Use with the \texttt{\string\acmSubmissionID} command to print the
  submission's unique ID on each page of the work.
\item{\texttt{authorversion}}: Produces a version of the work suitable
  for posting by the author.
\item{\texttt{screen}}: Produces colored hyperlinks.
\end{itemize}

This document uses the following string as the first command in the
source file:
\begin{verbatim}
\documentclass[acmsmall]{acmart}
\end{verbatim}

\section{Modifications}

Modifying the template --- including but not limited to: adjusting
margins, typeface sizes, line spacing, paragraph and list definitions,
and the use of the \verb|\vspace| command to manually adjust the
vertical spacing between elements of your work --- is not allowed.

{\bfseries Your document will be returned to you for revision if
  modifications are discovered.}

\section{Typefaces}

The ``\verb|acmart|'' document class requires the use of the
``Libertine'' typeface family. Your \TeX\ installation should include
this set of packages. Please do not substitute other typefaces. The
``\verb|lmodern|'' and ``\verb|ltimes|'' packages should not be used,
as they will override the built-in typeface families.

\section{Title Information}

The title of your work should use capital letters appropriately -
\url{https://capitalizemytitle.com/} has useful rules for
capitalization. Use the {\verb|title|} command to define the title of
your work. If your work has a subtitle, define it with the
{\verb|subtitle|} command.  Do not insert line breaks in your title.

If your title is lengthy, you must define a short version to be used
in the page headers, to prevent overlapping text. The \verb|title|
command has a ``short title'' parameter:
\begin{verbatim}
  \title[short title]{full title}
\end{verbatim}

\section{Authors and Affiliations}

Each author must be defined separately for accurate metadata
identification.  As an exception, multiple authors may share one
affiliation. Authors' names should not be abbreviated; use full first
names wherever possible. Include authors' e-mail addresses whenever
possible.

Grouping authors' names or e-mail addresses, or providing an ``e-mail
alias,'' as shown below, is not acceptable:
\begin{verbatim}
  \author{Brooke Aster, David Mehldau}
  \email{dave,judy,steve@university.edu}
  \email{firstname.lastname@phillips.org}
\end{verbatim}

The \verb|authornote| and \verb|authornotemark| commands allow a note
to apply to multiple authors --- for example, if the first two authors
of an article contributed equally to the work.

If your author list is lengthy, you must define a shortened version of
the list of authors to be used in the page headers, to prevent
overlapping text. The following command should be placed just after
the last \verb|\author{}| definition:
\begin{verbatim}
  \renewcommand{\shortauthors}{McCartney, et al.}
\end{verbatim}
Omitting this command will force the use of a concatenated list of all
of the authors' names, which may result in overlapping text in the
page headers.

The article template's documentation, available at
\url{https://www.acm.org/publications/proceedings-template}, has a
complete explanation of these commands and tips for their effective
use.

Note that authors' addresses are mandatory for journal articles.

\section{Rights Information}

Authors of any work published by ACM will need to complete a rights
form. Depending on the kind of work, and the rights management choice
made by the author, this may be copyright transfer, permission,
license, or an OA (open access) agreement.

Regardless of the rights management choice, the author will receive a
copy of the completed rights form once it has been submitted. This
form contains \LaTeX\ commands that must be copied into the source
document. When the document source is compiled, these commands and
their parameters add formatted text to several areas of the final
document:
\begin{itemize}
\item the ``ACM Reference Format'' text on the first page.
\item the ``rights management'' text on the first page.
\item the conference information in the page header(s).
\end{itemize}

Rights information is unique to the work; if you are preparing several
works for an event, make sure to use the correct set of commands with
each of the works.

The ACM Reference Format text is required for all articles over one
page in length, and is optional for one-page articles (abstracts).

\section{CCS Concepts and User-Defined Keywords}

Two elements of the ``acmart'' document class provide powerful
taxonomic tools for you to help readers find your work in an online
search.

The ACM Computing Classification System ---
\url{https://www.acm.org/publications/class-2012} --- is a set of
classifiers and concepts that describe the computing
discipline. Authors can select entries from this classification
system, via \url{https://dl.acm.org/ccs/ccs.cfm}, and generate the
commands to be included in the \LaTeX\ source.

User-defined keywords are a comma-separated list of words and phrases
of the authors' choosing, providing a more flexible way of describing
the research being presented.

CCS concepts and user-defined keywords are required for for all
articles over two pages in length, and are optional for one- and
two-page articles (or abstracts).

\section{Sectioning Commands}

Your work should use standard \LaTeX\ sectioning commands:
\verb|\section|, \verb|\subsection|, \verb|\subsubsection|,
\verb|\paragraph|, and \verb|\subparagraph|. The sectioning levels up to
\verb|\subsusection| should be numbered; do not remove the numbering
from the commands.

Simulating a sectioning command by setting the first word or words of
a paragraph in boldface or italicized text is {\bfseries not allowed.}

Below are examples of sectioning commands.

\subsection{Subsection}
\label{sec:subsection}

This is a subsection.

\subsubsection{Subsubsection}
\label{sec:subsubsection}

This is a subsubsection.

\paragraph{Paragraph}

This is a paragraph.

\subparagraph{Subparagraph}

This is a subparagraph.

\section{Tables}

The ``\verb|acmart|'' document class includes the ``\verb|booktabs|''
package --- \url{https://ctan.org/pkg/booktabs} --- for preparing
high-quality tables.

Table captions are placed {\itshape above} the table.

Because tables cannot be split across pages, the best placement for
them is typically the top of the page nearest their initial cite.  To
ensure this proper ``floating'' placement of tables, use the
environment \textbf{table} to enclose the table's contents and the
table caption.  The contents of the table itself must go in the
\textbf{tabular} environment, to be aligned properly in rows and
columns, with the desired horizontal and vertical rules.  Again,
detailed instructions on \textbf{tabular} material are found in the
\textit{\LaTeX\ User's Guide}.

Immediately following this sentence is the point at which
Table~\ref{tab:freq} is included in the input file; compare the
placement of the table here with the table in the printed output of
this document.

\begin{table}
  \caption{Frequency of Special Characters}
  \label{tab:freq}
  \begin{tabular}{ccl}
    \toprule
    Non-English or Math&Frequency&Comments\\
    \midrule
    \O & 1 in 1,000& For Swedish names\\
    $\pi$ & 1 in 5& Common in math\\
    \$ & 4 in 5 & Used in business\\
    $\Psi^2_1$ & 1 in 40,000& Unexplained usage\\
  \bottomrule
\end{tabular}
\end{table}

To set a wider table, which takes up the whole width of the page's
live area, use the environment \textbf{table*} to enclose the table's
contents and the table caption.  As with a single-column table, this
wide table will ``float'' to a location deemed more
desirable. Immediately following this sentence is the point at which
Table~\ref{tab:commands} is included in the input file; again, it is
instructive to compare the placement of the table here with the table
in the printed output of this document.

\begin{table*}
  \caption{Some Typical Commands}
  \label{tab:commands}
  \begin{tabular}{ccl}
    \toprule
    Command &A Number & Comments\\
    \midrule
    \texttt{{\char'134}author} & 100& Author \\
    \texttt{{\char'134}table}& 300 & For tables\\
    \texttt{{\char'134}table*}& 400& For wider tables\\
    \bottomrule
  \end{tabular}
\end{table*}

Always use midrule to separate table header rows from data rows, and
use it only for this purpose. This enables assistive technologies to
recognise table headers and support their users in navigating tables
more easily.

\section{Math Equations}
You may want to display math equations in three distinct styles:
inline, numbered or non-numbered display.  Each of the three are
discussed in the next sections.

\subsection{Inline (In-text) Equations}
A formula that appears in the running text is called an inline or
in-text formula.  It is produced by the \textbf{math} environment,
which can be invoked with the usual
\texttt{{\char'134}begin\,\ldots{\char'134}end} construction or with
the short form \texttt{\$\,\ldots\$}. You can use any of the symbols
and structures, from $\alpha$ to $\omega$, available in
\LaTeX~\cite{Lamport:LaTeX}; this section will simply show a few
examples of in-text equations in context. Notice how this equation:
\begin{math}
  \lim_{n\rightarrow \infty}x=0
\end{math},
set here in in-line math style, looks slightly different when
set in display style.  (See next section).

\subsection{Display Equations}
A numbered display equation---one set off by vertical space from the
text and centered horizontally---is produced by the \textbf{equation}
environment. An unnumbered display equation is produced by the
\textbf{displaymath} environment.

Again, in either environment, you can use any of the symbols and
structures available in \LaTeX\@; this section will just give a couple
of examples of display equations in context.  First, consider the
equation, shown as an inline equation above:
\begin{equation}
  \lim_{n\rightarrow \infty}x=0
\end{equation}
Notice how it is formatted somewhat differently in
the \textbf{displaymath}
environment.  Now, we'll enter an unnumbered equation:
\begin{displaymath}
  \sum_{i=0}^{\infty} x + 1
\end{displaymath}
and follow it with another numbered equation:
\begin{equation}
  \sum_{i=0}^{\infty}x_i=\int_{0}^{\pi+2} f
\end{equation}
just to demonstrate \LaTeX's able handling of numbering.

\section{Figures}

%% The ``\verb|figure|'' environment should be used for figures. One or
%% more images can be placed within a figure. If your figure contains
%% third-party material, you must clearly identify it as such, as shown
%% in the example below.
%% \begin{figure}[h]
%%   \centering
%%   \includegraphics[width=\linewidth]{sample-franklin}
%%   \caption{1907 Franklin Model D roadster. Photograph by Harris \&
%%     Ewing, Inc. [Public domain], via Wikimedia
%%     Commons. (\url{https://goo.gl/VLCRBB}).}
%%   \Description{A woman and a girl in white dresses sit in an open car.}
%% \end{figure}

Your figures should contain a caption which describes the figure to
the reader.

Figure captions are placed {\itshape below} the figure.

Every figure should also have a figure description unless it is purely
decorative. These descriptions convey what’s in the image to someone
who cannot see it. They are also used by search engine crawlers for
indexing images, and when images cannot be loaded.

A figure description must be unformatted plain text less than 2000
characters long (including spaces).  {\bfseries Figure descriptions
  should not repeat the figure caption – their purpose is to capture
  important information that is not already provided in the caption or
  the main text of the paper.} For figures that convey important and
complex new information, a short text description may not be
adequate. More complex alternative descriptions can be placed in an
appendix and referenced in a short figure description. For example,
provide a data table capturing the information in a bar chart, or a
structured list representing a graph.  For additional information
regarding how best to write figure descriptions and why doing this is
so important, please see
\url{https://www.acm.org/publications/taps/describing-figures/}.

\subsection{The ``Teaser Figure''}

A ``teaser figure'' is an image, or set of images in one figure, that
are placed after all author and affiliation information, and before
the body of the article, spanning the page. If you wish to have such a
figure in your article, place the command immediately before the
\verb|\maketitle| command:
\begin{verbatim}
  \begin{teaserfigure}
    \includegraphics[width=\textwidth]{sampleteaser}
    \caption{figure caption}
    \Description{figure description}
  \end{teaserfigure}
\end{verbatim}

\section{Citations and Bibliographies}

The use of \BibTeX\ for the preparation and formatting of one's
references is strongly recommended. Authors' names should be complete
--- use full first names (``Donald E. Knuth'') not initials
(``D. E. Knuth'') --- and the salient identifying features of a
reference should be included: title, year, volume, number, pages,
article DOI, etc.

The bibliography is included in your source document with these two
commands, placed just before the \verb|\end{document}| command:
\begin{verbatim}
  \bibliographystyle{ACM-Reference-Format}
  \bibliography{bibfile}
\end{verbatim}
where ``\verb|bibfile|'' is the name, without the ``\verb|.bib|''
suffix, of the \BibTeX\ file.

Citations and references are numbered by default. A small number of
ACM publications have citations and references formatted in the
``author year'' style; for these exceptions, please include this
command in the {\bfseries preamble} (before the command
``\verb|\begin{document}|'') of your \LaTeX\ source:
\begin{verbatim}
  \citestyle{acmauthoryear}
\end{verbatim}

  Some examples.  A paginated journal article \cite{Abril07}, an
  enumerated journal article \cite{Cohen07}, a reference to an entire
  issue \cite{JCohen96}, a monograph (whole book) \cite{Kosiur01}, a
  monograph/whole book in a series (see 2a in spec. document)
  \cite{Harel79}, a divisible-book such as an anthology or compilation
  \cite{Editor00} followed by the same example, however we only output
  the series if the volume number is given \cite{Editor00a} (so
  Editor00a's series should NOT be present since it has no vol. no.),
  a chapter in a divisible book \cite{Spector90}, a chapter in a
  divisible book in a series \cite{Douglass98}, a multi-volume work as
  book \cite{Knuth97}, a couple of articles in a proceedings (of a
  conference, symposium, workshop for example) (paginated proceedings
  article) \cite{Andler79, Hagerup1993}, a proceedings article with
  all possible elements \cite{Smith10}, an example of an enumerated
  proceedings article \cite{VanGundy07}, an informally published work
  \cite{Harel78}, a couple of preprints \cite{Bornmann2019,
    AnzarootPBM14}, a doctoral dissertation \cite{Clarkson85}, a
  master's thesis: \cite{anisi03}, an online document / world wide web
  resource \cite{Thornburg01, Ablamowicz07, Poker06}, a video game
  (Case 1) \cite{Obama08} and (Case 2) \cite{Novak03} and \cite{Lee05}
  and (Case 3) a patent \cite{JoeScientist001}, work accepted for
  publication \cite{rous08}, 'YYYYb'-test for prolific author
  \cite{SaeediMEJ10} and \cite{SaeediJETC10}. Other cites might
  contain 'duplicate' DOI and URLs (some SIAM articles)
  \cite{Kirschmer:2010:AEI:1958016.1958018}. Boris / Barbara Beeton:
  multi-volume works as books \cite{MR781536} and \cite{MR781537}. A
  couple of citations with DOIs:
  \cite{2004:ITE:1009386.1010128,Kirschmer:2010:AEI:1958016.1958018}. Online
  citations: \cite{TUGInstmem, Thornburg01, CTANacmart}.
  Artifacts: \cite{R} and \cite{UMassCitations}.

\section{Acknowledgments}

Identification of funding sources and other support, and thanks to
individuals and groups that assisted in the research and the
preparation of the work should be included in an acknowledgment
section, which is placed just before the reference section in your
document.

This section has a special environment:
\begin{verbatim}
  \begin{acks}
  ...
  \end{acks}
\end{verbatim}
so that the information contained therein can be more easily collected
during the article metadata extraction phase, and to ensure
consistency in the spelling of the section heading.

Authors should not prepare this section as a numbered or unnumbered {\verb|\section|}; please use the ``{\verb|acks|}'' environment.

\section{Appendices}

If your work needs an appendix, add it before the
``\verb|\end{document}|'' command at the conclusion of your source
document.

Start the appendix with the ``\verb|appendix|'' command:
\begin{verbatim}
  \appendix
\end{verbatim}
and note that in the appendix, sections are lettered, not
numbered. This document has two appendices, demonstrating the section
and subsection identification method.

\section{Multi-language papers}

Papers may be written in languages other than English or include
titles, subtitles, keywords and abstracts in different languages (as a
rule, a paper in a language other than English should include an
English title and an English abstract).  Use \verb|language=...| for
every language used in the paper.  The last language indicated is the
main language of the paper.  For example, a French paper with
additional titles and abstracts in English and German may start with
the following command
\begin{verbatim}
\documentclass[sigconf, language=english, language=german,
               language=french]{acmart}
\end{verbatim}

The title, subtitle, keywords and abstract will be typeset in the main
language of the paper.  The commands \verb|\translatedXXX|, \verb|XXX|
begin title, subtitle and keywords, can be used to set these elements
in the other languages.  The environment \verb|translatedabstract| is
used to set the translation of the abstract.  These commands and
environment have a mandatory first argument: the language of the
second argument.  See \verb|sample-sigconf-i13n.tex| file for examples
of their usage.

\section{SIGCHI Extended Abstracts}

The ``\verb|sigchi-a|'' template style (available only in \LaTeX\ and
not in Word) produces a landscape-orientation formatted article, with
a wide left margin. Three environments are available for use with the
``\verb|sigchi-a|'' template style, and produce formatted output in
the margin:
\begin{description}
\item[\texttt{sidebar}:]  Place formatted text in the margin.
\item[\texttt{marginfigure}:] Place a figure in the margin.
\item[\texttt{margintable}:] Place a table in the margin.
\end{description}

%%
%% The acknowledgments section is defined using the "acks" environment
%% (and NOT an unnumbered section). This ensures the proper
%% identification of the section in the article metadata, and the
%% consistent spelling of the heading.
\begin{acks}
To Robert, for the bagels and explaining CMYK and color spaces.
\end{acks}

%%
%% The next two lines define the bibliography style to be used, and
%% the bibliography file.
\bibliographystyle{ACM-Reference-Format}
\bibliography{sample-base}

%%
%% If your work has an appendix, this is the place to put it.
\appendix

\section{Research Methods}

\subsection{Part One}

Lorem ipsum dolor sit amet, consectetur adipiscing elit. Morbi
malesuada, quam in pulvinar varius, metus nunc fermentum urna, id
sollicitudin purus odio sit amet enim. Aliquam ullamcorper eu ipsum
vel mollis. Curabitur quis dictum nisl. Phasellus vel semper risus, et
lacinia dolor. Integer ultricies commodo sem nec semper.

\subsection{Part Two}

Etiam commodo feugiat nisl pulvinar pellentesque. Etiam auctor sodales
ligula, non varius nibh pulvinar semper. Suspendisse nec lectus non
ipsum convallis congue hendrerit vitae sapien. Donec at laoreet
eros. Vivamus non purus placerat, scelerisque diam eu, cursus
ante. Etiam aliquam tortor auctor efficitur mattis.

\section{Online Resources}

Nam id fermentum dui. Suspendisse sagittis tortor a nulla mollis, in
pulvinar ex pretium. Sed interdum orci quis metus euismod, et sagittis
enim maximus. Vestibulum gravida massa ut felis suscipit
congue. Quisque mattis elit a risus ultrices commodo venenatis eget
dui. Etiam sagittis eleifend elementum.

Nam interdum magna at lectus dignissim, ac dignissim lorem
rhoncus. Maecenas eu arcu ac neque placerat aliquam. Nunc pulvinar
massa et mattis lacinia.

\end{document}
\endinput
%%
%% End of file `sample-acmsmall.tex'.

### 2025-03-11 yonedaLF/types2025/abstract.tex
stuff

\documentclass{easychair}
\raggedbottom

\usepackage{doc}
\usepackage{xcolor}
\usepackage{mathpartir}
\usepackage{todonotes}
\presetkeys{todonotes}{inline}{}
\usepackage{scalerel}
\usepackage{bm}
\usepackage{amssymb}

\renewcommand{\mit}[1]{{\mathit{#1}}}
\newcommand{\msf}[1]{{\mathsf{#1}}}
\newcommand{\mbf}[1]{{\mathbf{#1}}}
\newcommand{\mbb}[1]{\mathbb{#1}}
\newcommand{\bs}[1]{\boldsymbol{#1}}
\newcommand{\ext}{\triangleright}
\newcommand{\blank}{{\mathord{\hspace{1pt}\text{--}\hspace{1pt}}}}

\newcommand{\MetaTy}{\msf{MetaTy}}
\newcommand{\Base}{\msf{Base}}
\newcommand{\PSh}{\msf{PSh}}
\newcommand{\El}{\msf{El}}
\newcommand{\Cat}{\msf{Cat}}
\newcommand{\In}{\msf{In}}
\newcommand{\base}{\msf{base}}
\newcommand{\Tm}{\msf{Tm}}
\newcommand{\U}{\msf{U}}
\newcommand{\Con}{\msf{Con}}
\newcommand{\Sub}{\msf{Sub}}
\newcommand{\Y}{\msf{Y}}
\renewcommand{\S}{\msf{S}}
\newcommand{\Fib}{\msf{Fib}}
\newcommand{\Fin}{\msf{Fin}}
\newcommand{\Tree}{\msf{Tree}}
\newcommand{\Disc}{\msf{Disc}}

\title{A Generalized Logical Framework}
\author{Andr\'as Kov\'acs$^{1}$ \and Christian Sattler$^{2}$}
\titlerunning{A Generalized Logical Framework}

\institute{
  Chalmers University of Technology \& University of Gothenburg, Sweden \\
  $^{1}$\,\email{andrask@chalmers.se}\,\,\,$^{2}$\,\email{sattler@chalmers.se}
}

\pagenumbering{gobble}
\begin{document}
\maketitle

Logical frameworks (LFs \cite{Harper93lf}) and the closely related two-level type
theories (2LTTs \cite{twolevel}) let us work in a mixed syntax of a metatheory and a
chosen object theory. Here, we have a second-order view on the object theory,
where contexts, variables and substitutions are implicit, and binders are
represented as meta-level functions. There are some well-known limitations to
LFs. First, we have to pick a model of the object theory externally. Second,
since we only have a second-order view on that model, many constructions cannot
be expressed; for example, the induction principle for the syntax of an object
theory requires a notion of first-order model, where contexts and substitutions
are explicit. Various ways have been described to make logical frameworks more
expressive by extending them with modalities (e.g.\ \cite{sterlingthesis,DBLP:conf/lics/Hofmann99,DBLP:journals/corr/abs-1901-03378,orton_et_al:LIPIcs:2016:6564}). In the current work we
describe an LF with the following features:
\begin{itemize}
\item We can work with multiple models of multiple object theories at the same
  time. By ``theory'' we mean a second-order generalized algebraic theory
  (SOGAT \cite{uemura,DBLP:conf/fscd/KaposiX24}); this includes all type theories and programming languages that
  only use structural binders.
\item We have both an ``external'' first-order view and an ``internal'' second-order view on each model,
      and we can freely switch between perspectives.
\item Models of object theories are themselves internally defined in the LF.
\item The LF is fully structural as a type theory; no substructural modalities
  are used.
\end{itemize}

\noindent\textbf{The Logical Framework.} The basic structure is as follows.
\begin{itemize}
\item We have a universe $\MetaTy$\footnote{More precisely, a $\mbb{N}$-indexed universe
hierarchy, but we shall omit ``sizing'' levels in this abstract.} closed under the type formers
  of extensional type theory.
\item We have $\Base : \MetaTy$, $1 : \Base$, $\PSh : \Base \to \MetaTy$ and $\El : \{i :
  \Base\} \to \PSh\,i \to \MetaTy$ such that each $\PSh\,i$ and $\El$ constitutes
  a Tarski-style universe closed under ETT type formers.
\item Let us define $\Cat\,i : \PSh\,i$ as the type of categories internally to
  $\PSh\,i$. Then, we have $\In : \{i : \Base\} \to \Cat\,i \to \MetaTy$ and $\msf{base} : \In\,C \to \Base$.
\end{itemize}
We give some semantic intuition in the following. Each $\PSh\,i$ is a universe
of presheaves over some base category. In the empty context, only $\PSh\,1$ is
available, which is the universe of sets. Internally to $\PSh\,1$, we can define
some $C : \El\,(\Cat\,1)$. Now, if we have $i : \In\,C$, we can form $\PSh\,(\base\,i)$
as the universe of presheaves over $C$.
\begin{enumerate}
\item We can define $\msf{PShExt\,C} : \PSh\,1$ as the \emph{external} type of presheaves over $C$.
\item Our semantics supports the isomorphism $\El\,(\msf{PShExt}\,C) \simeq ((i
  : \In\,C) \to \PSh\,(\base\,i))$.  In other words, external and internal
  notions of presheaves coincide. More generally, we have this isomorphism for
  any $C : \Cat\,j$, i.e.\ starting from a category that's internal to any
  previously defined presheaf universe.
\end{enumerate}
\textbf{Yoneda embeddings}. Our semantics actually supports a more general
notion of internalization than the above one, which we don't describe here. We
have not yet finalized which operations to enshrine in the LF's syntax, but the
special case of \emph{Yoneda embeddings} seems to be especially useful. This
works in the generality of SOGATs but we shall focus on the example of pure
lambda calculus. A second-order model of pure LC in some universe $\U$ is simply
$\Tm : \U$ together with an isomorphism $\Tm \simeq (\Tm \to \Tm)$. A
first-order model is a unityped category with families \cite{cwfs}, where we
write $\msf{Con} : \U$ for the type of contexts, $\Tm : \Con \to \U$ for the
type of terms, $\Gamma+ : \Con$ for the extension of $\Gamma : \Con$ with a
binding, and we have a natural isomorphism $\Tm\,\Gamma \simeq \Tm\,(\Gamma+)$.
\begin{itemize}
\item For each $M$ a first-order model in $\PSh\,i$ and $j : \In\,M$\footnote{We
implicitly take the underlying category of $M$ here.}, we have $\msf{S}_j$ as a
  second-order model in $\PSh\,j$. In other words, internally to presheaves over
  a model of lambda calculus, we have a second-order model of lambda
  calculus. In fact, this is the standard semantics of traditional LFs/2LTTs,
  and we get all such LFs/2LTTs as syntactic fragments of our generalized LF, by
  working under an assumption of $j : \In\,M$.
\item Yoneda embedding has action on contexts, substitutions and terms:
  \begin{alignat*}{3}
    & \Y : \El\,\Con_M \to (\{j : \In\,M\} \to \PSh\,j)\\
    & \Y : \El\,(\Sub_M\,\Gamma\,\Delta) \simeq (\{j : \In\,M\} \to \El\,(\Y\,\Gamma) \to \El\,(\Y\,\Delta))\\
    & \Y : \El\,(\Tm_M\,\Gamma) \simeq (\{j : \In\,M\} \to \El\,(\Y\,\Gamma) \to \El\,\Tm_{\S_j})
  \end{alignat*}
\end{itemize}
Additionally, $\Y$ preserves empty contexts and extended contexts up to
isomorphism and preserves all other structure strictly. $\Y$ allows ad-hoc
switching between first-order and second-order syntax. For example, the identity
substitution $\msf{id} : \El\,(\Sub_M\,\Gamma\,\Gamma)$ can be alternatively defined as
$\Y^{-1}(\lambda\,\gamma.\,\gamma)$, where we use $\Y^{-1}$ to externalize
$(\lambda\,\gamma.\,\gamma) : (\{j : \In\,M\} \to \El\,(\Y\,\Gamma) \to
\El\,(\Y\,\Gamma))$. More generally, by using a modest amount of syntactic sugar
and elaboration, we can develop $\Y$ and $\Y^{-1}$ into a ``second-order
notation'' for any SOGAT, which constitutes a rigorous and nicely readable
alternative to De Bruijn indices and explicit substitution operations.
\\

\noindent \textbf{Sketch of the semantics}. The model of LF is constructed in
two steps. First, we give a model for the theory that has $\PSh$, $\Base$ and
$\In$ as sorts but does not support $\MetaTy$, and then take presheaves over
that model to obtain a model of a 2LTT where $\MetaTy$ represents the outer
layer. In the inner model, we start with an inductive definition of certain
``trees in categories'':
\begin{alignat*}{3}
  & \mbf{data}\,\Tree\,(B : \Cat) : \msf{Set}\,\mbf{where}\\
  & \hspace{1em} \msf{node} : (\Gamma : \PSh\,B)(n : \mbb{N})(C : \Fin\,n \to \Fib\,(B \ext \Disc\,\Gamma))\\
  & \hspace{3.8em} ((i : \Fin\,n) \to \Tree\,(B \ext \Disc\,\Gamma \ext C\,i)) \to \Tree\,B
\end{alignat*}
Here, $\PSh$ means presheaves in sets, $\Fib$ is cartesian fibrations, $\Disc$
creates a discrete fibration from a presheaf and $\blank\!\ext\!\blank$ takes
the total category of a fibration. Now, the objects of the semantic base
category are elements of $\Tree\,1$, and morphisms between trees are level-wise
natural transformations between the $\Gamma$ components together with $\Fin\,n
\to \Fin\,m$ renamings of subtree indices. A semantic $\Base$ points to a
subtree of a context, an $\In$ is a $\Fin\,n$ index pointing to a child of a
given node, and a $\PSh$ is a dependent presheaf a $\Gamma$ inside a given node.
This semantics can be viewed as a way of inductively carving out a subcategory
of $\mbf{Cat}$ where ETT type formers in presheaf categories are strictly stable
under substitution.

\bibliographystyle{plain}
\bibliography{references}

\end{document}

### 2025-03-11 yonedaLF/types2025/abstract.tex
types25 abstract typo

@@
   only use structural binders.
-\item We have both an ``external'' first-order view and an ``internal'' second-order view on each model,
-      and we can freely switch between perspectives.
-\item Models of object theories are themselves internally defined in the LF.
+\item We have both an ``external'' first-order view and an ``internal''
+  second-order view on each model, and we can freely switch between
+  perspectives. All models of object theories are defined internally in the LF.
 \item The LF is fully structural as a type theory; no substructural modalities
@@
 \item Let us define $\Cat\,i : \PSh\,i$ as the type of categories internally to
-  $\PSh\,i$. Then, we have $\In : \{i : \Base\} \to \Cat\,i \to \MetaTy$ and $\msf{base} : \In\,C \to \Base$.
+  $\PSh\,i$. Then, we have $\In : \{i : \Base\} \to \El\,(\Cat\,i) \to \MetaTy$ and $\msf{base} : \In\,C \to \Base$.
 \end{itemize}
@@
   notions of presheaves coincide. More generally, we have this isomorphism for
-  any $C : \Cat\,j$, i.e.\ starting from a category that's internal to any
+  any $C : \El\,(\Cat\,j)$, i.e.\ starting from a category that's internal to any
   previously defined presheaf universe.
@@
     & \Y : \El\,\Con_M \to (\{j : \In\,M\} \to \PSh\,j)\\
-    & \Y : \El\,(\Sub_M\,\Gamma\,\Delta) \simeq (\{j : \In\,M\} \to \El\,(\Y\,\Gamma) \to \El\,(\Y\,\Delta))\\
+    & \Y : \El\,(\Sub_M\,\Gamma\,\Delta) \simeq (\{j : \In\,M\} \to \El\,(\Y\,\Gamma\,\{j\}) \to \El\,(\Y\,\Delta\,\{j\}))\\
     & \Y : \El\,(\Tm_M\,\Gamma) \simeq (\{j : \In\,M\} \to \El\,(\Y\,\Gamma) \to \El\,\Tm_{\S_j})
@@
 creates a discrete fibration from a presheaf and $\blank\!\ext\!\blank$ takes
-the total category of a fibration. Now, the objects of the semantic base
-category are elements of $\Tree\,1$, and morphisms between trees are level-wise
-natural transformations between the $\Gamma$ components together with $\Fin\,n
-\to \Fin\,m$ renamings of subtree indices. A semantic $\Base$ points to a
-subtree of a context, an $\In$ is a $\Fin\,n$ index pointing to a child of a
-given node, and a $\PSh$ is a dependent presheaf a $\Gamma$ inside a given node.
-This semantics can be viewed as a way of inductively carving out a subcategory
-of $\mbf{Cat}$ where ETT type formers in presheaf categories are strictly stable
-under substitution.
+the total category of a fibration. Now, the objects of
+the semantic base category are elements of $\Tree\,1$, and morphisms between
+trees are level-wise natural transformations between the $\Gamma$ components
+together with $\Fin\,n \to \Fin\,m$ renamings of subtree indices. The
+non-discrete $\Fib$ components are preserved by morphisms. A semantic $\Base$
+points to a subtree of a context, an $\In$ is a $\Fin\,n$ index pointing to a
+child of a given node, and a $\PSh$ is a dependent presheaf over a $\Gamma$ inside a
+given node. Extending a context with an $\In$ binding adds a new empty subtree
+to a given node. Extending with an $\El$ binding extends the $\Gamma$ presheaf
+in a node with a dependent presheaf.
 
